### 2019-03-15 notebooks/general_info.Rmd
initial commit, setting up hierarchy and putting in some info. the data for 7.2K package runs is in data/full_sigs/..., and this is the data from the late stage of the analysis. analysis code itself can be found in code/data_processing.R

---
title: "Overview"
output: html_notebook
---

Ok.
First, load the full signatures, and make sure we source the data processing file.
There's an error here, not sure what's causing it.
This is a work-in-progress.

```{r, echo=FALSE}
source("../code/data_processing.R")

los <- list.files("../data/full_sigs", full.names=TRUE)
names <- list.files("../data/full_sigs")
names(los) <- substr(names, 1, nchar(names)-4)
los <- lapply(los, readRDS)
los <- remove_empties(los)
# annoying strange error that I should fix pops up
```

These are the updated counts.
```{r}
total_arguments <- apply_count_to_lopkg_c(los, args_count_all)
total_mono_type <- apply_count_to_lopkg_c(los, args_count_all_mono_type)
total_mono_attr <- apply_count_to_lopkg_c(los, args_count_all_mono_attr_pattern)
total_mono_clas <- apply_count_to_lopkg_c(los, args_count_all_mono_class)

total_functions <- apply_count_to_lopkg_c(los, length)
total_mono_type_functions <- apply_count_to_lopkg_c(los, fun_count_all_mono_type)
total_mono_attr_functions <- apply_count_to_lopkg_c(los, fun_count_all_mono_attr_pattern)
total_mono_clas_functions <- apply_count_to_lopkg_c(los, fun_count_all_mono_class)
```

This is how you get the top sigs.
```{r}
poly_type_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_type_args))
poly_type_args <- poly_type_args[order(-poly_type_args$count),] # sort by count
poly_type_args$perc <- round(poly_type_args$count/sum(poly_type_args$count) * 100, 2)
```

Then, pick out the ones you want.
```{r}
print_me <- head(poly_type_args, 20)
print_me$type <- sapply(print_me$type, function(x) paste0(x[[1]], collapse=", "))
print_me
# I need to streamline this, too many weird lists appearing...
```

Similarly for attributes, and classes.
```{r}
poly_attr_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_attr_args))
poly_attr_args <- poly_attr_args[order(-poly_attr_args$count),] # sort by count
poly_attr_args$perc <- round(poly_attr_args$count/sum(poly_attr_args$count) * 100, 2)
```

```{r}
poly_class_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_class_args))
poly_class_args <- poly_class_args[order(-poly_class_args$count),] # sort by count
poly_class_args$perc <- round(poly_class_args$count/sum(poly_class_args$count) * 100, 2)
```

Printing is as before.
Note that ATM I also have attribute type information in here.
I'll get it to display just the names as well.
I figured this might be interesting, and I don't expect many functions to be polymorphic in this way.
```{r}
# will fix printing, its annoying atm
print_me_a <- head(poly_attr_args, 20)
print_me_a$attr <- sapply(print_me_a$attr, function(x) paste0(x[[1]], collapse=", "))
print_me_a
```

```{r}
print_me_c <- head(poly_class_args, 20)
print_me_c$class <- sapply(print_me_c$class, function(x) paste0(x[[1]], collapse=", "))
print_me_c
```

### 2019-03-15 paper/body.tex
initial commit, setting up hierarchy and putting in some info. the data for 7.2K package runs is in data/full_sigs/..., and this is the data from the late stage of the analysis. analysis code itself can be found in code/data_processing.R

\section{High-Level Discussion}

I set up author notes:
\KS{for Konrad}, \JV{for Jan}, and \AT{for myself}.
Just do backslash and your initials (see source).

So, what do we want to say in this paper?
We would like to produce a high-level paper saying what the status of the language is in terms of types, and possibly give direction to future work on the language.

%
%
%
%
%
%
\section{Introduction}

\begin{itemize}
    \item R is a cool and unusual language that isn't studied a lot.
    \item People really like it, even if many people in our community scoff at it. \textit{``What? It's lazy \textbf{and} side-effecting?''}
    \item We should study it because it can give us some insights on how people without a formal training in programming program.
    \item Deepening our understanding of the language will enable us to further improve it, while continuing to support common paradigms.
    \item In this work, we aim to shed some light on how R programmers program by analyzing how they use \textit{type information} (either types themselves, or attributes which are a form of RTTI).
\end{itemize}

%
%
%
%
%
%
\section{Background}

%
%
%
%
\subsection{The R Programming Language}

\AT{Talk about R.}
R is a lazy, side-effecting, dynamically-typed, ... .

Typical usage of the language is ... . \AT{Do we have something to cite for this?}
\AT{Should we motivate why we only look at R packages?
Some people might wonder why we don't look at non-library code.}

Most R libraries (called \textit{packages}) are stored in the Comprehensive R Archive Network (CRAN) and Bioconductor package repositories.
\AT{Talk about important packages? tidyverse?}

\AT{Kind of ... want to combine the above two paragraphs somehow? Together, they lead in nicely to the genthat talk.}

%
%
%
%
\subsection{The \texttt{genthat} Package}

Unfortunately, getting our hands on ``front-line'' R code is tricky:
R is not like many other languages, where large systems are built up and stored in a repository such as Github.
Instead, R programmers are liable to leverage a number of packages and work them into their data analysis pipelines.

Interestingly, we are somewhat able to replicate this using only package code.
CRAN, for instance, has a policy that packages made available on the CRAN platform should be accompanied by a series of examples, tests, and/or vignettes showing off how the code is intended to be used.
Nominally unrelated, the {\tt genthat} project aims to trace this swath of example code to generate new examples which package designers may not have thought of.
Crucially, {\tt genthat} outputs complete trace information for each of the functions that it runs so that they can be run again in a fresh R instance (e.g., by capturing the random seed, environment variables, etc.).
We can leverage this trace information to glean type and attribute patterns of functions, and since these traces correspond to examples of the packages being used (rather than just arbitrarily running package code), we have a reasonable approximation of how R programmers would use the packages.

\AT{We use/modify the {\tt genthat} tracer, probably worth going in to more detail here.}

%
%
%
%
%
%
\section{Motivation and High-level Overview}

Our community is centred on building, improving, and reasoning about programming languages.
A key component in all this is to truly \textit{understand} the languages we're working on, so that we may make the best decisions for the benefit of all users of a programming language.
In many cases, we as researchers can appeal to our own intuition to at least build a sense for how a language is used, as language users typically have some training from an experienced programmer or software engineer (e.g., computer science studies, online tutorials written by language experts, etc.).
\AT{We claim that this is emphatically not the case in R.}

R is a languages written by and for \textit{statisticians}.
It traces its roots to the S language, developed by ... .
R programmers don't write big servers or software systems; instead, they write and modify small data processing pipelines as a means to facilitate interaction with digital data.
This leads to ``front-line'' R code (i.e., code written for users, not programmers) not being made available on typical platforms like Github.
\AT{Maybe front-line code isn't a good word, is there a good way to describe non-library code?}
Put differently, R is essentially a really big domain-specific language.

The design of R is also rife with questionable decisions.
For example, the language is lazy \textit{and} side-effecting, which one could describe as an antagonistic pair of features.
In addition, data structures must be entirely copied when being modified, which appears to be far from ideal in a language meant to process (thus, modify) big data.

In spite of these flaws, the language is incredibly popular. 
\AT{We'll want to cite something here.}

One way or another, R is here to stay, so we may as well figure out how to deal with it.
One aspect of this is understanding R programmers and how they interact with the language, and a step towards this goal is to understand \AT{how types arise in R programs.}
We will shed some light on this by analyzing the usage of R functions, and build a picture of how polymorphic R code is.

\begin{itemize}
    \item Understanding R can help us improve the language, as well as better equip us to build languages which have an appeal outside of the realm of computer science.
    \item We're gonna tackle this by looking at how R programmers use types and type information in their programs.
    \item We will be mindful of the runtime representations of values to give more context to the information we glean from programmer-visible type information.
\end{itemize}

%
%
%
%
%
%
\section{The Method}

In this section, we will detail our methodology for collecting data.

First, we leverage {\tt genthat}'s tracer to trace function executions.
The idea here is that we don't really have access to non-library code written in R, as general use patterns are (possibly?) to write small scripts which analyze some bit of data and possibly visualize results.
The goal is likely not to build big working systems, instead to explore data with by writing and rewriting small scripts, ad infinitum.
In looking at package tests, vignettes, and examples, we are painting a picture of how the package designers intended their packages to be used, which we believe is a close approximation of what R users would do.

{\tt genthat} generates a trace file which can be loaded into R and contains all of the necessary information to rerun the tested function.
This includes the random seed and all relevant environment data.
We utilize this trace data to collect type information:
we run each argument in the specified environment, and collect type information through R's {\tt typeof} function, and attribute information through R's {\tt attributes} function.

Then, we aggregate the information in all traces for a particular function. 
If there was only one function trace, we discard the result as the function is trivially monomorphic.
We can see which arguments had which types over how many traces, and from here we can build a \textit{signature} for each function argument, and consider them together to create a function signature.
A \textbf{polymorphic argument} is one which has been inhabited by values of at least two different types, and a \textbf{polymorphic function} is a function with at least one polymorphic argument or a polymorphic return.

We repeat the above for each package on the Comprehensive R Archive Network (CRAN), the premier source for R libraries (called packages), and the Bioconductor package repository.
Now, depending on the data point we're after, our methods from here differ slightly.

\begin{itemize}
    \item for \textbf{counting signatures}: for each signature, we count how many functions or arguments have the requisite signature;
    \item for \textbf{finding the most common signatures}: we enumerate as before, and sort the signatures based on how often they appear;
\end{itemize}

%
%
%
%
\subsection{Attributes}
\label{sec:method:attributes}

Types (in the sense of the result of a call to {\tt typeof}) are easy to analyze as values can only have one type.
Attributes, however, are a different story, as values can have an arbitrary number of attributes.
Essentially, we're not interesting by function arguments which have had multiple attributes, we're interested in function arguments which have had multiple attribute \textit{patterns}. \AT{is this defined earlier?}
Arguments are said to be polymorphic in attribute if they have been inhabited with values which have had different attribute patterns.
\AT{We have not investigated whether functions which are polymorphic in attribute use attributes extensively inside the function code.}

\AT{Paragraph justifying our construction of an attribute pattern.
Either just name, or (name, type).}

\AT{Paragraph about so-called ``naturally-occurring'' attribute patterns.}

%
%
%
%
\subsection{{\tt genthat} Tracer Quirks}

First and foremost, {\tt genthat} captures arguments only on function exit, so as to not force promises during function execution.
\AT{Uh oh:} One clear limitation here is that if an argument \textit{gains} attributes over the functione execution, then they will appear in our signatures.
As R is side-effecting, this is an entirely valid practice (i.e., passing objects to functions to modify them), though \AT{it's unclear if people actually do this}.

\AT{Maybe talk about the quirks of genthat in here?
For example, how it deals with default arguments, and stuff like that.}

%
%
%
%
\subsection{Goals}

Ultimately, our goal is to collect data and report on patterns which emerge organically.
That said, the comprehensive picture of type usage we are developing can also be used to inform the design of a set of \textit{type annotations} for the language, and this goal informs our analysis.
As we collect our data, we keep in mind the idea that the data we produce should clearly suggest what sorts of type annotations would reflect language usage patterns.

% %
% %
% %
% %
% \subsection{Front-End Data Collection}

% The following describes our data collection methodology:

% \begin{enumerate}
%     \item use the {\tt genthat} tracer to run tests, vignettes, and examples for a package, and optionally repeat for package's reverse dependencies;
%     \item gather type and attribute information of all arguments and return values from all calls to all functions by looking at generated trace data;
%     \item per function, consolidate information to generate a type signature.
% \end{enumerate}

% In order to get a global picture of R usage, we do the above over \textit{all} CRAN and Bioconductor packages, and aggregate type and attribute signature results over all of them.

%
%
%
%
\subsection{Back-End Data Collection}

\AT{Do we want to go down this route with this paper?
This would be Konrad's bit, if we want to try to stick our data together.}

%
%
%
%
%
%
\section{Results}

\AT{Tables, graphs, code examples should go in here.}
In this section, we ... .

%
%
%
%
\subsection{Usage Patterns}

In this section, we will discuss function usage patterns which arose in our analysis.
We will start by looking at the morphicity of functions.

\subsubsection{Function Argument and Return Morphicity}

First and foremost, we would like to know how often R programmers create polymorphic functions.
Recall that we define a polymorphic function to be a function with at least one polymorphic argument, or a polymorphic return.
We will turn our attention now to the data in \todo{Figure}, and go through each entry in the table.

Here, we see that the vast majority of function arguments are indeed monomorphic.
Monomorphic arguments are easy to annotate, as the {\tt typeof} the arguments is exactly the most precise annotation we could give, at least in terms of \textit{type}.
That said, types alone don't always paint the whole picture:
recall that \textit{attributes} are an R language feature which allows programmers to stick metadata onto values.
So in which circumstances \textit{do} types paint the whole picture?

The second data point in \textbf{TODO FIGURE} indicates that a \isit{majority} of function arguments are monomorphic in type \textit{and} have no attributes.
These represent arguments which are truly trivial to annotate, as the {\tt typeof} an argument perfectly describes the usage of that argument.
That said, some attributes arise naturally in R:
For instance, names in a named list (e.g. {\tt x} and {\tt y} in {\tt list(x=0, y=0)}) appear as an attribute on the value.
In Section~\ref{sec:method:attributes}, we outlined these naturally-occurring attribute patterns, and the third data point in \textbf{TODO FIGURE} shows that a nontrivial amount of functions are monomorphic in type with said natural attribute patterns.

Another facet of type information in R is in the {\tt class} attribute.
Recall that values have a \textit{class} in addition to a type:
For example, a {\tt data.frame} has type {\tt list} and class {\tt data.frame}.
\todo{Plug classes earlier}
R has a number of built-in classes, such as \todo{X, Y, and Z}, but users are free to redefine the class of any value at runtime, and easily define new classes.
The next data point in \todo{Figure} shows that user-defined classes don't appear altogether often, though they do indeed feature.
\AT{In a following section, we will discuss how the usage of these classes manifests itself.}

\AT{Separate section for functions? Right now, this is at argument granularity.}

\subsubsection{Type Signatures}

In the last section, we presented a high-level overview of the morphicity ... .

\subsubsection{Attribute Signatures}

Recall that attributes are a way for programmers to store metadata on values in R.
What are the common attribute patterns?
And how often is the attribute pattern polymorphic?

\subsubsection{Takeaways}

\begin{itemize}
    \item the vast majority of arguments are monomorphic in type;
    \item of those, over 60\% have no attribute information;
    \item of the 40\% with attribute information, roughly 1/2 have fairly simple attributes corresponding to base R constructs (named lists and vectors, matrices, and data frames);
    \item now, of those arguments which are polymorphic, a sizable chunk (well over half) have defensible signatures (e.g., double and character for named list indexing, double and integer for obvious reasons, etc.).
\end{itemize}

In short, it looks like R (package) programmers are reasonable.
I'd conjecture that a lot of the polymorphism (e.g., double and list) is coming from how easy it is to use either type in a given situation (e.g., converting from vector of doubles to list or vice versa is simple).

%
%
%
%
%
%
\section{Synthesis}

In this section, we will discuss the conclusions that we draw from our data.

%
%
%
%
\subsection{Suggested Annotations}

\AT{How can we capture these patterns with annotations?}

Some possible annotations:

\begin{itemize}
    \item \textit{real}: for \textit{double} and \textit{integer} values
    \item \textit{function}: for \textit{closure}, \textit{special}, and \textit{builtin} values
    \item \textit{vector}: to indicate that something should be vectorized
    \item \textit{scalar}: to indicate that something should \textbf{not} be vectorized
    \item \textit{index}: for \textit{real} and \textit{character} values
\end{itemize}

%
%
\subsubsection{Struct-Like Attribute Declarations}

\AT{What's a convenient way to annotate attributes?
Should investigate how often attributes are consistent.}

%
%
\subsubsection{Coverage}

\AT{What is the coverage of these new annotations?}

%
%
%
%
%
%
\section{Conclusions and Future Work}

### 2019-03-15 paper/main.tex
initial commit, setting up hierarchy and putting in some info. the data for 7.2K package runs is in data/full_sigs/..., and this is the data from the late stage of the analysis. analysis code itself can be found in code/data_processing.R

\documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

\usepackage{booktabs} % For formal tables

\usepackage{listings}
\lstset{language=R}

% Style Stuff
\usepackage{my_style}

% Copyright
%\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
\acmDOI{10.475/123_4}

% ISBN
\acmISBN{123-4567-24-567/08/06}

%Conference
\acmConference[WOODSTOCK'97]{ACM Woodstock conference}{July 1997}{El
  Paso, Texas USA}
\acmYear{1997}
\copyrightyear{2016}

\acmPrice{15.00}

\acmSubmissionID{123-A12-B3}

\begin{document}
\title{How R R's Types Used}
% \titlenote{Produces the permission block, and
%   copyright information}
\subtitle{A Study of Type Usage in R}
% \subtitlenote{The full version of the author's guide is available as
%   \texttt{acmart.pdf} document}

\author{Alexi Turcotte}
\authornote{Work of this author was partially supported by NSERC.}
% \orcid{1234-5678-9012}
\affiliation{%
  \institution{Northeastern University}
%   \streetaddress{P.O. Box 1212}
%   \city{Dublin}
%   \state{Ohio}
%   \postcode{43017-6221}
}
\email{alexi@ccs.neu.edu}

% if the default list of authors is too long for headers:
% \renewcommand{\shortauthors}{B. Trovato et al.}

\begin{abstract}

\end{abstract}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below.
%
% \begin{CCSXML}
% <ccs2012>
%  <concept>
%   <concept_id>10010520.10010553.10010562</concept_id>
%   <concept_desc>Computer systems organization~Embedded systems</concept_desc>
%   <concept_significance>500</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10010520.10010575.10010755</concept_id>
%   <concept_desc>Computer systems organization~Redundancy</concept_desc>
%   <concept_significance>300</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10010520.10010553.10010554</concept_id>
%   <concept_desc>Computer systems organization~Robotics</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10003033.10003083.10003095</concept_id>
%   <concept_desc>Networks~Network reliability</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
% </ccs2012>
% \end{CCSXML}

% \ccsdesc[500]{Computer systems organization~Embedded systems}
% \ccsdesc[300]{Computer systems organization~Redundancy}
% \ccsdesc{Computer systems organization~Robotics}
% \ccsdesc[100]{Networks~Network reliability}

\keywords{Type Systems, R}

\maketitle

\input{body.tex}

\bibliographystyle{ACM-Reference-Format}
\bibliography{biblio}

\end{document}

### 2019-03-15 notebooks/general_info.Rmd
added function to display data frames in notebook, and functions to get fully monomorphic things (i.e., monomorphic in class, attribue, and type). the counts are available in the notebook, just not being printed yet (you might need to open it with RStudio to see)

@@
 
+```{r}
+source("../code/data_processing.R") # get the functions we need
+```
+
 ```{r, echo=FALSE}
-source("../code/data_processing.R")
-
 los <- list.files("../data/full_sigs", full.names=TRUE)
@@
 Then, pick out the ones you want.
+The `format_df_for_print` function is just for letting the notebook display the data frame properly.
+The type, attr, and class rows are lists of lists.
 ```{r}
-print_me <- head(poly_type_args, 20)
-print_me$type <- sapply(print_me$type, function(x) paste0(x[[1]], collapse=", "))
-print_me
-# I need to streamline this, too many weird lists appearing...
+format_df_for_print(poly_type_args[1:20,])
 ```
@@
 ```{r}
-# will fix printing, its annoying atm
-print_me_a <- head(poly_attr_args, 20)
-print_me_a$attr <- sapply(print_me_a$attr, function(x) paste0(x[[1]], collapse=", "))
-print_me_a
+format_df_for_print(poly_attr_args[1:20,])
 ```
@@
 ```{r}
-print_me_c <- head(poly_class_args, 20)
-print_me_c$class <- sapply(print_me_c$class, function(x) paste0(x[[1]], collapse=", "))
-print_me_c
+format_df_for_print(poly_class_args[1:20,])
 ```

### 2019-03-15 notebooks/general_info.Rmd
added scalar counting. 40% of arguments are just scalar

@@
 ```{r}
+# argument counts
 total_arguments <- apply_count_to_lopkg_c(los, args_count_all)
@@
 total_mono_clas <- apply_count_to_lopkg_c(los, args_count_all_mono_class)
+total_mono_all <- apply_count_to_lopkg_c(los, args_count_all_mono_all)
+```
 
+```{r}
+# function counts
 total_functions <- apply_count_to_lopkg_c(los, length)
@@
 total_mono_clas_functions <- apply_count_to_lopkg_c(los, fun_count_all_mono_class)
+total_mono_all_functions <- apply_count_to_lopkg_c(los, fun_count_all_mono_all)
 ```
@@
 ```
+
+How many things are just scalar?
+```{r}
+lo_scalars <- apply_row_extractor_to_lopkg_c(los, get_rows_from_df_only_scalar)
+total_scalar_arguments <- apply_count_to_lopkg_c(lo_scalars, args_count_all)
+```

### 2019-03-15 notebooks/general_info.Rmd
added functionality to translate between type systems. numbers up for a type system corresponding to Rs as it stands, next up will be rolling NULLs into all other types

@@
 ```
+
+Let's see what happens if we change type systems!
+```{r}
+los_basic <- lapply(los, function(lof) {lapply(lof, function(df) {change_type_systems_df(df, type_map_T0_to_r)})})
+```
+
+Let's get some counts.
+This is for a type system which is in direct correspondence with R's typeof.
+```{r}
+# argument counts
+total_mono_type_basic <- apply_count_to_lopkg_c(los_basic, args_count_all_mono_type)
+total_mono_all_basic <- apply_count_to_lopkg_c(los_basic, args_count_all_mono_all)
+```
+
+What if we roll NULL into every type?

### 2019-03-15 notebooks/general_info.Rmd
got numbers all the way up to folding ints and doubles together. barely any poly left. time to pull up colocation of character/real poly and lists

@@
 What if we roll NULL into every type?
+```{r}
+los_basic_no_null <- lapply(los_basic, function(lof) {lapply(lof, function(df) {fold_NULL_into_other_types_df(df)})})
+```
+
+What are the counts now?
+```{r}
+total_mono_type_basic_no_null <- apply_count_to_lopkg_c(los_basic_no_null, args_count_all_mono_type)
+total_mono_all_basic_no_null <- apply_count_to_lopkg_c(los_basic_no_null, args_count_all_mono_all)
+```
+
+Great.
+Now, what are the top instances of type-based polymorphism?
+```{r}
+poly_type_args_basic_no_null <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los_basic_no_null, extract_poly_type_args))
+poly_type_args_basic_no_null <- poly_type_args_basic_no_null[order(-poly_type_args_basic_no_null$count),] # sort by count
+poly_type_args_basic_no_null$perc <- round(poly_type_args_basic_no_null$count/sum(poly_type_args_basic_no_null$count) * 100, 2)
+```
+
+What's the top?
+```{r}
+format_df_for_print(poly_type_args_basic_no_null[1:20,])
+```
+
+So, clearly, "real number polymorphism" is the top choice.
+What if we fold ints and doubles together into reals?
+
+```{r}
+los_basic_no_null_real <- lapply(los_basic_no_null, function(lof) {lapply(lof, function(df) {fold_together_int_double_df(df)})})
+```

### 2019-03-15 paper/main.tex
minor

@@
 
-%
-% The code below should be generated by the tool at
-% http://dl.acm.org/ccs.cfm
-% Please copy and paste the code instead of the example below.
-%
-% \begin{CCSXML}
-% <ccs2012>
-%  <concept>
-%   <concept_id>10010520.10010553.10010562</concept_id>
-%   <concept_desc>Computer systems organization~Embedded systems</concept_desc>
-%   <concept_significance>500</concept_significance>
-%  </concept>
-%  <concept>
-%   <concept_id>10010520.10010575.10010755</concept_id>
-%   <concept_desc>Computer systems organization~Redundancy</concept_desc>
-%   <concept_significance>300</concept_significance>
-%  </concept>
-%  <concept>
-%   <concept_id>10010520.10010553.10010554</concept_id>
-%   <concept_desc>Computer systems organization~Robotics</concept_desc>
-%   <concept_significance>100</concept_significance>
-%  </concept>
-%  <concept>
-%   <concept_id>10003033.10003083.10003095</concept_id>
-%   <concept_desc>Networks~Network reliability</concept_desc>
-%   <concept_significance>100</concept_significance>
-%  </concept>
-% </ccs2012>
-% \end{CCSXML}
-
-% \ccsdesc[500]{Computer systems organization~Embedded systems}
-% \ccsdesc[300]{Computer systems organization~Redundancy}
-% \ccsdesc{Computer systems organization~Robotics}
-% \ccsdesc[100]{Networks~Network reliability}
-
 \keywords{Type Systems, R}
@@
 
-\input{body.tex}
+\section{High-Level Discussion}
+
+I set up author notes:
+\KS{for Konrad}, \JV{for Jan}, and \AT{for myself}.
+Just do backslash and your initials (see source).
+
+So, what do we want to say in this paper?
+We would like to produce a high-level paper saying what the status of the language is in terms of types, and possibly give direction to future work on the language.
+
+%
+%
+%
+%
+%
+%
+\section{Introduction}
+
+\begin{itemize}
+    \item R is a cool and unusual language that isn't studied a lot.
+    \item People really like it, even if many people in our community scoff at it. \textit{``What? It's lazy \textbf{and} side-effecting?''}
+    \item We should study it because it can give us some insights on how people without a formal training in programming program.
+    \item Deepening our understanding of the language will enable us to further improve it, while continuing to support common paradigms.
+    \item In this work, we aim to shed some light on how R programmers program by analyzing how they use \textit{type information} (either types themselves, or attributes which are a form of RTTI).
+\end{itemize}
+
+%
+%
+%
+%
+%
+%
+\section{Background}
+
+%
+%
+%
+%
+\subsection{The R Programming Language}
+
+\AT{Talk about R.}
+R is a lazy, side-effecting, dynamically-typed, ... .
+
+Typical usage of the language is ... . \AT{Do we have something to cite for this?}
+\AT{Should we motivate why we only look at R packages?
+Some people might wonder why we don't look at non-library code.}
+
+Most R libraries (called \textit{packages}) are stored in the Comprehensive R Archive Network (CRAN) and Bioconductor package repositories.
+\AT{Talk about important packages? tidyverse?}
+
+\AT{Kind of ... want to combine the above two paragraphs somehow? Together, they lead in nicely to the genthat talk.}
+
+%
+%
+%
+%
+\subsection{The \texttt{genthat} Package}
+
+Unfortunately, getting our hands on ``front-line'' R code is tricky:
+R is not like many other languages, where large systems are built up and stored in a repository such as Github.
+Instead, R programmers are liable to leverage a number of packages and work them into their data analysis pipelines.
+
+Interestingly, we are somewhat able to replicate this using only package code.
+CRAN, for instance, has a policy that packages made available on the CRAN platform should be accompanied by a series of examples, tests, and/or vignettes showing off how the code is intended to be used.
+Nominally unrelated, the {\tt genthat} project aims to trace this swath of example code to generate new examples which package designers may not have thought of.
+Crucially, {\tt genthat} outputs complete trace information for each of the functions that it runs so that they can be run again in a fresh R instance (e.g., by capturing the random seed, environment variables, etc.).
+We can leverage this trace information to glean type and attribute patterns of functions, and since these traces correspond to examples of the packages being used (rather than just arbitrarily running package code), we have a reasonable approximation of how R programmers would use the packages.
+
+\AT{We use/modify the {\tt genthat} tracer, probably worth going in to more detail here.}
+
+%
+%
+%
+%
+%
+%
+\section{Motivation and High-level Overview}
+
+Our community is centred on building, improving, and reasoning about programming languages.
+A key component in all this is to truly \textit{understand} the languages we're working on, so that we may make the best decisions for the benefit of all users of a programming language.
+In many cases, we as researchers can appeal to our own intuition to at least build a sense for how a language is used, as language users typically have some training from an experienced programmer or software engineer (e.g., computer science studies, online tutorials written by language experts, etc.).
+\AT{We claim that this is emphatically not the case in R.}
+
+R is a languages written by and for \textit{statisticians}.
+It traces its roots to the S language, developed by ... .
+R programmers don't write big servers or software systems; instead, they write and modify small data processing pipelines as a means to facilitate interaction with digital data.
+This leads to ``front-line'' R code (i.e., code written for users, not programmers) not being made available on typical platforms like Github.
+\AT{Maybe front-line code isn't a good word, is there a good way to describe non-library code?}
+Put differently, R is essentially a really big domain-specific language.
+
+The design of R is also rife with questionable decisions.
+For example, the language is lazy \textit{and} side-effecting, which one could describe as an antagonistic pair of features.
+In addition, data structures must be entirely copied when being modified, which appears to be far from ideal in a language meant to process (thus, modify) big data.
+
+In spite of these flaws, the language is incredibly popular. 
+\AT{We'll want to cite something here.}
+
+One way or another, R is here to stay, so we may as well figure out how to deal with it.
+One aspect of this is understanding R programmers and how they interact with the language, and a step towards this goal is to understand \AT{how types arise in R programs.}
+We will shed some light on this by analyzing the usage of R functions, and build a picture of how polymorphic R code is.
+
+\begin{itemize}
+    \item Understanding R can help us improve the language, as well as better equip us to build languages which have an appeal outside of the realm of computer science.
+    \item We're gonna tackle this by looking at how R programmers use types and type information in their programs.
+    \item We will be mindful of the runtime representations of values to give more context to the information we glean from programmer-visible type information.
+\end{itemize}
+
+%
+%
+%
+%
+%
+%
+\section{The Method}
+
+In this section, we will detail our methodology for collecting data.
+
+First, we leverage {\tt genthat}'s tracer to trace function executions.
+The idea here is that we don't really have access to non-library code written in R, as general use patterns are (possibly?) to write small scripts which analyze some bit of data and possibly visualize results.
+The goal is likely not to build big working systems, instead to explore data with by writing and rewriting small scripts, ad infinitum.
+In looking at package tests, vignettes, and examples, we are painting a picture of how the package designers intended their packages to be used, which we believe is a close approximation of what R users would do.
+
+{\tt genthat} generates a trace file which can be loaded into R and contains all of the necessary information to rerun the tested function.
+This includes the random seed and all relevant environment data.
+We utilize this trace data to collect type information:
+we run each argument in the specified environment, and collect type information through R's {\tt typeof} function, and attribute information through R's {\tt attributes} function.
+
+Then, we aggregate the information in all traces for a particular function. 
+If there was only one function trace, we discard the result as the function is trivially monomorphic.
+We can see which arguments had which types over how many traces, and from here we can build a \textit{signature} for each function argument, and consider them together to create a function signature.
+A \textbf{polymorphic argument} is one which has been inhabited by values of at least two different types, and a \textbf{polymorphic function} is a function with at least one polymorphic argument or a polymorphic return.
+
+We repeat the above for each package on the Comprehensive R Archive Network (CRAN), the premier source for R libraries (called packages), and the Bioconductor package repository.
+Now, depending on the data point we're after, our methods from here differ slightly.
+
+\begin{itemize}
+    \item for \textbf{counting signatures}: for each signature, we count how many functions or arguments have the requisite signature;
+    \item for \textbf{finding the most common signatures}: we enumerate as before, and sort the signatures based on how often they appear;
+\end{itemize}
+
+%
+%
+%
+%
+\subsection{Attributes}
+\label{sec:method:attributes}
+
+Types (in the sense of the result of a call to {\tt typeof}) are easy to analyze as values can only have one type.
+Attributes, however, are a different story, as values can have an arbitrary number of attributes.
+Essentially, we're not interesting by function arguments which have had multiple attributes, we're interested in function arguments which have had multiple attribute \textit{patterns}.
+\AT{is this defined earlier?}
+Arguments are said to be polymorphic in attribute if they have been inhabited with values which have had different attribute patterns.
+\AT{We have not investigated whether functions which are polymorphic in attribute use attributes extensively inside the function code.}
+
+\AT{Paragraph justifying our construction of an attribute pattern.
+Either just name, or (name, type).}
+
+\AT{Paragraph about so-called ``naturally-occurring'' attribute patterns.}
+
+%
+%
+%
+%
+\subsection{{\tt genthat} Tracer Quirks}
+
+First and foremost, {\tt genthat} captures arguments only on function exit, so as to not force promises during function execution.
+\AT{Uh oh:} One clear limitation here is that if an argument \textit{gains} attributes over the functione execution, then they will appear in our signatures.
+As R is side-effecting, this is an entirely valid practice (i.e., passing objects to functions to modify them), though \AT{it's unclear if people actually do this}.
+
+\AT{Maybe talk about the quirks of genthat in here?
+For example, how it deals with default arguments, and stuff like that.}
+
+%
+%
+%
+%
+\subsection{Goals}
+
+Ultimately, our goal is to collect data and report on patterns which emerge organically.
+That said, the comprehensive picture of type usage we are developing can also be used to inform the design of a set of \textit{type annotations} for the language, and this goal informs our analysis.
+As we collect our data, we keep in mind the idea that the data we produce should clearly suggest what sorts of type annotations would reflect language usage patterns.
+
+% %
+% %
+% %
+% %
+% \subsection{Front-End Data Collection}
+
+% The following describes our data collection methodology:
+
+% \begin{enumerate}
+%     \item use the {\tt genthat} tracer to run tests, vignettes, and examples for a package, and optionally repeat for package's reverse dependencies;
+%     \item gather type and attribute information of all arguments and return values from all calls to all functions by looking at generated trace data;
+%     \item per function, consolidate information to generate a type signature.
+% \end{enumerate}
+
+% In order to get a global picture of R usage, we do the above over \textit{all} CRAN and Bioconductor packages, and aggregate type and attribute signature results over all of them.
+
+%
+%
+%
+%
+\subsection{Back-End Data Collection}
+
+\AT{Do we want to go down this route with this paper?
+This would be Konrad's bit, if we want to try to stick our data together.}
+
+%
+%
+%
+%
+%
+%
+\section{Results}
+
+\AT{Tables, graphs, code examples should go in here.}
+In this section, we ... .
+
+%
+%
+%
+%
+\subsection{Usage Patterns}
+
+In this section, we will discuss function usage patterns which arose in our analysis.
+We will start by looking at the morphicity of functions.
+
+\subsubsection{Function Argument and Return Morphicity}
+
+First and foremost, we would like to know how often R programmers create polymorphic functions.
+Recall that we define a polymorphic function to be a function with at least one polymorphic argument, or a polymorphic return.
+We will turn our attention now to the data in \todo{Figure}, and go through each entry in the table.
+
+Here, we see that the vast majority of function arguments are indeed monomorphic.
+Monomorphic arguments are easy to annotate, as the {\tt typeof} the arguments is exactly the most precise annotation we could give, at least in terms of \textit{type}.
+That said, types alone don't always paint the whole picture:
+recall that \textit{attributes} are an R language feature which allows programmers to stick metadata onto values.
+So in which circumstances \textit{do} types paint the whole picture?
+
+The second data point in \textbf{TODO FIGURE} indicates that a \isit{majority} of function arguments are monomorphic in type \textit{and} have no attributes.
+These represent arguments which are truly trivial to annotate, as the {\tt typeof} an argument perfectly describes the usage of that argument.
+That said, some attributes arise naturally in R:
+For instance, names in a named list (e.g. {\tt x} and {\tt y} in {\tt list(x=0, y=0)}) appear as an attribute on the value.
+In Section~\ref{sec:method:attributes}, we outlined these naturally-occurring attribute patterns, and the third data point in \textbf{TODO FIGURE} shows that a nontrivial amount of functions are monomorphic in type with said natural attribute patterns.
+
+Another facet of type information in R is in the {\tt class} attribute.
+Recall that values have a \textit{class} in addition to a type:
+For example, a {\tt data.frame} has type {\tt list} and class {\tt data.frame}.
+\todo{Plug classes earlier}
+R has a number of built-in classes, such as \todo{X, Y, and Z}, but users are free to redefine the class of any value at runtime, and easily define new classes.
+The next data point in \todo{Figure} shows that user-defined classes don't appear altogether often, though they do indeed feature.
+\AT{In a following section, we will discuss how the usage of these classes manifests itself.}
+
+\AT{Separate section for functions? Right now, this is at argument granularity.}
+
+\subsubsection{Type Signatures}
+
+In the last section, we presented a high-level overview of the morphicity ... .
+
+\subsubsection{Attribute Signatures}
+
+Recall that attributes are a way for programmers to store metadata on values in R.
+What are the common attribute patterns?
+And how often is the attribute pattern polymorphic?
+
+\subsubsection{Takeaways}
+
+\begin{itemize}
+    \item the vast majority of arguments are monomorphic in type;
+    \item of those, over 60\% have no attribute information;
+    \item of the 40\% with attribute information, roughly 1/2 have fairly simple attributes corresponding to base R constructs (named lists and vectors, matrices, and data frames);
+    \item now, of those arguments which are polymorphic, a sizable chunk (well over half) have defensible signatures (e.g., double and character for named list indexing, double and integer for obvious reasons, etc.).
+\end{itemize}
+
+In short, it looks like R (package) programmers are reasonable.
+I'd conjecture that a lot of the polymorphism (e.g., double and list) is coming from how easy it is to use either type in a given situation (e.g., converting from vector of doubles to list or vice versa is simple).
+
+%
+%
+%
+%
+%
+%
+\section{Synthesis}
+
+In this section, we will discuss the conclusions that we draw from our data.
+
+%
+%
+%
+%
+\subsection{Suggested Annotations}
+
+\AT{How can we capture these patterns with annotations?}
+
+Some possible annotations:
+
+\begin{itemize}
+    \item \textit{real}: for \textit{double} and \textit{integer} values
+    \item \textit{function}: for \textit{closure}, \textit{special}, and \textit{builtin} values
+    \item \textit{vector}: to indicate that something should be vectorized
+    \item \textit{scalar}: to indicate that something should \textbf{not} be vectorized
+    \item \textit{index}: for \textit{real} and \textit{character} values
+\end{itemize}
+
+%
+%
+\subsubsection{Struct-Like Attribute Declarations}
+
+\AT{What's a convenient way to annotate attributes?
+Should investigate how often attributes are consistent.}
+
+%
+%
+\subsubsection{Coverage}
+
+\AT{What is the coverage of these new annotations?}
+
+%
+%
+%
+%
+%
+%
+\section{Conclusions and Future Work}
 

### 2019-03-15 paper/main.tex
started discussion of type systems, and how we can explore the space of potential type systems. i really like how parameterizing the analysis with a type system sounds, we should definitely harp on that somewhere in the paper.

@@
 We would like to produce a high-level paper saying what the status of the language is in terms of types, and possibly give direction to future work on the language.
+
+\subsection{What is a type?}
+
+\AT{this should be moved to a relevant place later}
+
+R is a unityped language, and as such is trivially monomorphic, and this notion of monomorphism is far from useful.
+We are interested in expanding the notion of ``type'' in R, and thankfully the language has some tools on offer.
+\AT{this is discussed later on}
+
+\subsection{What are possible type systems?}
+
+Great question.
+Some things:
+
+\begin{itemize}
+
+\item {\it The Obvious Type System (OTS)}: this type system corresponds directly with the type information that one can fetch using the {\tt typeof}, {\tt attributes}, and {\tt class} functions.
+In this system, there is no distinction between vectors and scalars, and lists of differently-typed elements (e.g., a list of integers vs.
+a list of strings), and there is a distinction between integers and doubles, and the type of NULL and all other types.
+This type system isn't particularly informative to a programmer, though it does reflect the treatment of values in the R runtime, for the {\tt typeof} a value is the type that the value has according to the R internals.
+
+\item {\it The Fine-Grained Type System (FGTS)}: this type systems includes a number of relevant additions.
+Chiefly, it distinguishes between vectors and scalars, as it is conceivable that boxing scalars as single-element vectors might be overkill.
+In addition, list types are parameterized over the types of their elements (e.g., list<integer> vs. list<character>).
+Finally, NAs ``in the wild'' are given the type {\tt raw\_NA}, as {\tt typeof(NA) = logical} in R.
+
+\item {\it The Null-and-NA-Free Type System (NNTS)}: most languages consider NULLs to inhabit all types, and R does so for NAs \AT{elaborate}.
+This gives us ... .
+
+\end{itemize}
+
+Of course, one could conceive of a more fine-grained type system than FGTS, but it's unclear that we would gain much in doing so.
+Indeed, there is a space to explore in slightly less fine-grained type systems, which make modifications including:
+
+\begin{itemize}
+
+\item {\bf collapsing vectors and scalars}: this yields a type system slightly closer to OTS.
+
+\item {\bf rolling NULL into other types}: in most languages, NULL is an inhabitant of all types, and this adjustment seems interesting.
+Note that this goes beyond even OTS, as in R {\tt typeof(NULL) = NULL}.
+
+\item {\bf rolling NA into other types}: \AT{i should try this}
+similarly, NA is an inhabitant of all types, though NA standing alone has type {\tt logical}.
+FGTS distinguishes such NAs with the {\tt raw\_NA} type, and if instead we consider NA to inhabit all types we may \AT{win}.
+
+\item {\bf combining integers and doubles}: as in most languages, doubles and ints can stand-in for each other in nearly every scenario.
+Even if not, it is trivial to convert from one to the other, particularly converting ints to doubles.
+Aside, in R a double can be used to index a list, and R will floor it when performing the lookup.
+
+\end{itemize}
 

### 2019-03-15 paper/main.tex
minor

--- 2019-03-15 paper/main.tex
@@
 \documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-
 \usepackage{booktabs} % For formal tables
-
 \usepackage{listings}
 \lstset{language=R}
-
-% Style Stuff
 \usepackage{my_style}
 
-% Copyright
-%\setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-\setcopyright{rightsretained}
-%\setcopyright{usgov}
-%\setcopyright{usgovmixed}
-%\setcopyright{cagov}
-%\setcopyright{cagovmixed}
+\setcopyright{none}
+%\setcopyright{acmcopyright}%\setcopyright{acmlicensed}
+%\acmDOI{10.475/123_4}
+%\acmISBN{123-4567-24-567/08/06}
+%\acmConference[OOPSLAs]{Woodstock conference}{July 1997}{El Paso, Texas USA}
+%\acmYear{1997}%\copyrightyear{2016}%\acmPrice{15.00}
+\begin{document}
 
-% DOI
-\acmDOI{10.475/123_4}
-
-% ISBN
-\acmISBN{123-4567-24-567/08/06}
-
-%Conference
-\acmConference[WOODSTOCK'97]{ACM Woodstock conference}{July 1997}{El
-  Paso, Texas USA}
-\acmYear{1997}
-\copyrightyear{2016}
-
-\acmPrice{15.00}
-
-\acmSubmissionID{123-A12-B3}
-
-\begin{document}
-\title{How R R's Types Used}
-% \titlenote{Produces the permission block, and
-%   copyright information}
-\subtitle{A Study of Type Usage in R}
-% \subtitlenote{The full version of the author's guide is available as
-%   \texttt{acmart.pdf} document}
-
-\author{Alexi Turcotte}
-\authornote{Work of this author was partially supported by NSERC.}
-% \orcid{1234-5678-9012}
-\affiliation{%
-  \institution{Northeastern University}
-%   \streetaddress{P.O. Box 1212}
-%   \city{Dublin}
-%   \state{Ohio}
-%   \postcode{43017-6221}
-}
-\email{alexi@ccs.neu.edu}
-
-% if the default list of authors is too long for headers:
-% \renewcommand{\shortauthors}{B. Trovato et al.}
+\title{A large-scale study of polymorphism}
 
 \begin{abstract}
+The R programming language is widely used in a variety of scientfic domains
+for tasks related to data science. The language is geared towards exploratory
+data
 
+The R programming language is widely used in a variety of scientific domains
+for tasks related to data analysis. Over the last two decades many libraries
+have been developed to facilitate all steps of the data analysis process.
+These libraries are stored in the CRAN reposot
 \end{abstract}
@@
 
-I set up author notes:
-\KS{for Konrad}, \JV{for Jan}, and \AT{for myself}.
-Just do backslash and your initials (see source).
+So, what do we want to say in this paper?  We would like to produce a
+high-level paper saying what the status of the language is in terms of
+types, and possibly give direction to future work on the language.
 
-So, what do we want to say in this paper?
-We would like to produce a high-level paper saying what the status of the language is in terms of types, and possibly give direction to future work on the language.
-
-%
-%
-%
-%
-%
 %
@@
 
-%
-%
-%
-%
-%
-%
 \section{Conclusions and Future Work}

### 2019-03-15 paper/main.tex
abstract

@@
 \documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-\usepackage{booktabs} % For formal tables
-\usepackage{listings}
+\usepackage{booktabs,listings}
 \lstset{language=R}
@@
 %\acmDOI{10.475/123_4}
-%\acmISBN{123-4567-24-567/08/06}
 %\acmConference[OOPSLAs]{Woodstock conference}{July 1997}{El Paso, Texas USA}
@@
 
-\title{A large-scale study of polymorphism}
+\title{A large-scale study of polymorphism in R}
 
@@
 The R programming language is widely used in a variety of scientfic domains
-for tasks related to data science. The language is geared towards exploratory
-data
-
-The R programming language is widely used in a variety of scientific domains
-for tasks related to data analysis. Over the last two decades many libraries
-have been developed to facilitate all steps of the data analysis process.
-These libraries are stored in the CRAN reposot
+for tasks related to data science. The language was designed to favor an
+interactive style of programming with minimal syntactic and conceptual
+overhead. This design is well suited to support interactive data analysis,
+but is not well suited to generating performant code or catching programming
+errors.  In particular, R has no type annotations and all operations are
+dynamically checked at runtime. The starting point for our work is the
+question: \emph{what could a static type system for R look like?}  To answer
+that question we study the polymorphism that is present in over X millions
+of lines of R code, written over a period of Y years by Z programmers.  We
+perform a dynamic analysis, leveraging tests and use-cases, to determine the
+level of polymprhism that is present in the code. We do this for several
+potential notions of types.
 \end{abstract}
-
-\keywords{Type Systems, R}
-
 \maketitle
 
-\section{High-Level Discussion}
-
-So, what do we want to say in this paper?  We would like to produce a
-high-level paper saying what the status of the language is in terms of
-types, and possibly give direction to future work on the language.
-
-%
 \section{Introduction}
 
-\begin{itemize}
-    \item R is a cool and unusual language that isn't studied a lot.
-    \item People really like it, even if many people in our community scoff at it. \textit{``What? It's lazy \textbf{and} side-effecting?''}
-    \item We should study it because it can give us some insights on how people without a formal training in programming program.
-    \item Deepening our understanding of the language will enable us to further improve it, while continuing to support common paradigms.
-    \item In this work, we aim to shed some light on how R programmers program by analyzing how they use \textit{type information} (either types themselves, or attributes which are a form of RTTI).
-\end{itemize}
-
-%
-%
-%
 %
@@
 
-\AT{Talk about R.}
 R is a lazy, side-effecting, dynamically-typed, ... .
@@
 
-%
-%
 %
@@
 
-Our community is centred on building, improving, and reasoning about programming languages.
-A key component in all this is to truly \textit{understand} the languages we're working on, so that we may make the best decisions for the benefit of all users of a programming language.
-In many cases, we as researchers can appeal to our own intuition to at least build a sense for how a language is used, as language users typically have some training from an experienced programmer or software engineer (e.g., computer science studies, online tutorials written by language experts, etc.).
-\AT{We claim that this is emphatically not the case in R.}
+Our community is centred on building, improving, and reasoning about
+programming languages.  A key component in all this is to truly
+\textit{understand} the languages we're working on, so that we may make the
+best decisions for the benefit of all users of a programming language.  In
+many cases, we as researchers can appeal to our own intuition to at least
+build a sense for how a language is used, as language users typically have
+some training from an experienced programmer or software engineer (e.g.,
+computer science studies, online tutorials written by language experts,
+etc.).  \AT{We claim that this is emphatically not the case in R.}
 
-R is a languages written by and for \textit{statisticians}.
-It traces its roots to the S language, developed by ... .
-R programmers don't write big servers or software systems; instead, they write and modify small data processing pipelines as a means to facilitate interaction with digital data.
-This leads to ``front-line'' R code (i.e., code written for users, not programmers) not being made available on typical platforms like Github.
-\AT{Maybe front-line code isn't a good word, is there a good way to describe non-library code?}
-Put differently, R is essentially a really big domain-specific language.
+R is a languages written by and for \textit{statisticians}.  It traces its
+roots to the S language, developed by ... .  R programmers don't write big
+servers or software systems; instead, they write and modify small data
+processing pipelines as a means to facilitate interaction with digital data.
+This leads to ``front-line'' R code (i.e., code written for users, not
+programmers) not being made available on typical platforms like Github.
+\AT{Maybe front-line code isn't a good word, is there a good way to describe
+  non-library code?}  Put differently, R is essentially a really big
+domain-specific language.
 
-The design of R is also rife with questionable decisions.
-For example, the language is lazy \textit{and} side-effecting, which one could describe as an antagonistic pair of features.
-In addition, data structures must be entirely copied when being modified, which appears to be far from ideal in a language meant to process (thus, modify) big data.
+The design of R is also rife with questionable decisions.  For example, the
+language is lazy \textit{and} side-effecting, which one could describe as an
+antagonistic pair of features.  In addition, data structures must be
+entirely copied when being modified, which appears to be far from ideal in a
+language meant to process (thus, modify) big data.
 
@@
 
-One way or another, R is here to stay, so we may as well figure out how to deal with it.
-One aspect of this is understanding R programmers and how they interact with the language, and a step towards this goal is to understand \AT{how types arise in R programs.}
-We will shed some light on this by analyzing the usage of R functions, and build a picture of how polymorphic R code is.
+One way or another, R is here to stay, so we may as well figure out how to
+deal with it.  One aspect of this is understanding R programmers and how
+they interact with the language, and a step towards this goal is to
+understand \AT{how types arise in R programs.}  We will shed some light on
+this by analyzing the usage of R functions, and build a picture of how
+polymorphic R code is.
 
 \begin{itemize}
-    \item Understanding R can help us improve the language, as well as better equip us to build languages which have an appeal outside of the realm of computer science.
-    \item We're gonna tackle this by looking at how R programmers use types and type information in their programs.
-    \item We will be mindful of the runtime representations of values to give more context to the information we glean from programmer-visible type information.
+    \item Understanding R can help us improve the language, as well as
+      better equip us to build languages which have an appeal outside of the
+      realm of computer science.
+    \item We're gonna tackle this by looking at how R programmers use types
+      and type information in their programs.
+    \item We will be mindful of the runtime representations of values to
+      give more context to the information we glean from programmer-visible
+      type information.
 \end{itemize}
@@
 
-First, we leverage {\tt genthat}'s tracer to trace function executions.
-The idea here is that we don't really have access to non-library code written in R, as general use patterns are (possibly?) to write small scripts which analyze some bit of data and possibly visualize results.
-The goal is likely not to build big working systems, instead to explore data with by writing and rewriting small scripts, ad infinitum.
-In looking at package tests, vignettes, and examples, we are painting a picture of how the package designers intended their packages to be used, which we believe is a close approximation of what R users would do.
+First, we leverage {\tt genthat}'s tracer to trace function executions.  The
+idea here is that we don't really have access to non-library code written in
+R, as general use patterns are (possibly?) to write small scripts which
+analyze some bit of data and possibly visualize results.  The goal is likely
+not to build big working systems, instead to explore data with by writing
+and rewriting small scripts, ad infinitum.  In looking at package tests,
+vignettes, and examples, we are painting a picture of how the package
+designers intended their packages to be used, which we believe is a close
+approximation of what R users would do.
 
-{\tt genthat} generates a trace file which can be loaded into R and contains all of the necessary information to rerun the tested function.
-This includes the random seed and all relevant environment data.
-We utilize this trace data to collect type information:
-we run each argument in the specified environment, and collect type information through R's {\tt typeof} function, and attribute information through R's {\tt attributes} function.
+{\tt genthat} generates a trace file which can be loaded into R and contains
+all of the necessary information to rerun the tested function.  This
+includes the random seed and all relevant environment data.  We utilize this
+trace data to collect type information: we run each argument in the
+specified environment, and collect type information through R's {\tt typeof}
+function, and attribute information through R's {\tt attributes} function.
 
 Then, we aggregate the information in all traces for a particular function.
-If there was only one function trace, we discard the result as the function is trivially monomorphic.
-We can see which arguments had which types over how many traces, and from here we can build a \textit{signature} for each function argument, and consider them together to create a function signature.
-A \textbf{polymorphic argument} is one which has been inhabited by values of at least two different types, and a \textbf{polymorphic function} is a function with at least one polymorphic argument or a polymorphic return.
+If there was only one function trace, we discard the result as the function
+is trivially monomorphic.  We can see which arguments had which types over
+how many traces, and from here we can build a \textit{signature} for each
+function argument, and consider them together to create a function
+signature.  A \textbf{polymorphic argument} is one which has been inhabited
+by values of at least two different types, and a \textbf{polymorphic
+  function} is a function with at least one polymorphic argument or a
+polymorphic return.
 
-We repeat the above for each package on the Comprehensive R Archive Network (CRAN), the premier source for R libraries (called packages), and the Bioconductor package repository.
-Now, depending on the data point we're after, our methods from here differ slightly.
+We repeat the above for each package on the Comprehensive R Archive Network
+(CRAN), the premier source for R libraries (called packages), and the
+Bioconductor package repository.  Now, depending on the data point we're
+after, our methods from here differ slightly.
 
 \begin{itemize}
-    \item for \textbf{counting signatures}: for each signature, we count how many functions or arguments have the requisite signature;
+    \item for \textbf{counting signatures}: for each signature, we count how
+      many functions or arguments have the requisite signature;
     \item for \textbf{finding the most common signatures}: we enumerate as before, and sort the signatures based on how often they appear;
@@
 
-The second data point in \textbf{TODO FIGURE} indicates that a \isit{majority} of function arguments are monomorphic in type \textit{and} have no attributes.
-These represent arguments which are truly trivial to annotate, as the {\tt typeof} an argument perfectly describes the usage of that argument.
-That said, some attributes arise naturally in R:
-For instance, names in a named list (e.g. {\tt x} and {\tt y} in {\tt list(x=0, y=0)}) appear as an attribute on the value.
-In Section~\ref{sec:method:attributes}, we outlined these naturally-occurring attribute patterns, and the third data point in \textbf{TODO FIGURE} shows that a nontrivial amount of functions are monomorphic in type with said natural attribute patterns.
+The second data point in \textbf{TODO FIGURE} indicates that a
+\isit{majority} of function arguments are monomorphic in type \textit{and}
+have no attributes.  These represent arguments which are truly trivial to
+annotate, as the {\tt typeof} an argument perfectly describes the usage of
+that argument.  That said, some attributes arise naturally in R: For
+instance, names in a named list (e.g. {\tt x} and {\tt y} in {\tt list(x=0,
+  y=0)}) appear as an attribute on the value.  In
+Section~\ref{sec:method:attributes}, we outlined these naturally-occurring
+attribute patterns, and the third data point in \textbf{TODO FIGURE} shows
+that a nontrivial amount of functions are monomorphic in type with said
+natural attribute patterns.
 
 Another facet of type information in R is in the {\tt class} attribute.
-Recall that values have a \textit{class} in addition to a type:
-For example, a {\tt data.frame} has type {\tt list} and class {\tt data.frame}.
-\todo{Plug classes earlier}
-R has a number of built-in classes, such as \todo{X, Y, and Z}, but users are free to redefine the class of any value at runtime, and easily define new classes.
-The next data point in \todo{Figure} shows that user-defined classes don't appear altogether often, though they do indeed feature.
-\AT{In a following section, we will discuss how the usage of these classes manifests itself.}
+Recall that values have a \textit{class} in addition to a type: For example,
+a {\tt data.frame} has type {\tt list} and class {\tt data.frame}.
+\todo{Plug classes earlier} R has a number of built-in classes, such as
+\todo{X, Y, and Z}, but users are free to redefine the class of any value at
+runtime, and easily define new classes.  The next data point in
+\todo{Figure} shows that user-defined classes don't appear altogether often,
+though they do indeed feature.  \AT{In a following section, we will discuss
+  how the usage of these classes manifests itself.}
 
@@
     \item of those, over 60\% have no attribute information;
-    \item of the 40\% with attribute information, roughly 1/2 have fairly simple attributes corresponding to base R constructs (named lists and vectors, matrices, and data frames);
+    \item of the 40\% with attribute information, roughly 1/2 have fairly
+      simple attributes corresponding to base R constructs (named lists and
+      vectors, matrices, and data frames);
     \item now, of those arguments which are polymorphic, a sizable chunk (well over half) have defensible signatures (e.g., double and character for named list indexing, double and integer for obvious reasons, etc.).
@@
 
-In short, it looks like R (package) programmers are reasonable.
-I'd conjecture that a lot of the polymorphism (e.g., double and list) is coming from how easy it is to use either type in a given situation (e.g., converting from vector of doubles to list or vice versa is simple).
+In short, it looks like R (package) programmers are reasonable.  I'd
+conjecture that a lot of the polymorphism (e.g., double and list) is coming
+from how easy it is to use either type in a given situation (e.g.,
+converting from vector of doubles to list or vice versa is simple).
 
@@
 
-\AT{What's a convenient way to annotate attributes?
-Should investigate how often attributes are consistent.}
+\AT{What's a convenient way to annotate attributes?  Should investigate how
+  often attributes are consistent.}
 
-%
 %

### 2019-03-15 paper/main.tex
minor

@@
 that question we study the polymorphism that is present in over X millions
-of lines of R code, written over a period of Y years by Z programmers.  We
-perform a dynamic analysis, leveraging tests and use-cases, to determine the
-level of polymprhism that is present in the code. We do this for several
-potential notions of types.
+of lines of R code, XX packages, written over a period of Y years by Z
+programmers.  We perform a dynamic analysis, leveraging tests and use-cases,
+to determine the level of polymoprhism that is present in the code. We do
+this for several potential notions of types. Our result suggest that
+polymorphism is important in some key parts of the system but that
+relatively simple type annotations could be used to capture most of the
+interesting cases.
 \end{abstract}

### 2019-03-15 paper/main.tex
intro wip

@@
 \section{Introduction}
+
+A number of programming languages have gained popularity for data science
+tasks, they include MATLAB, Python, Julia and R.
+
+These languages are all dynamically typed.
+
+This is because they emphasize ease of programming, interactive data
+exploration, and light-weight syntax.
+
+To faciliate programming these language offer all kinds of short cuts and
+facilities that increase the likelyhood that code will ``just work''.
+
+For example abbreviation of name arguments and automatic conversions.
+
+But invariably some applications written in those languages survive past the prototype stage.
+
+The results obtained are published and even sometimes drive policy decisions.
+
+Given that code almost never stops, and that any computation is likely to deliver a bunch of numbers, how can we trust those numbers?
+
+Years of work in programming languages suggest that one cost efficient way to catch some errors is to add types to a language.
+
+In the last decade much attention has been given to the incremental addition of types. Successes like Hack, TypeScript and Racket suggest that this may be feasible for wide spectrum languages.
 

### 2019-03-15 paper/main.tex
accidentally deleted the bit about type systems, i put it back in in the results section for now

--- 2019-03-15 paper/main.tex
@@
 
-% %
-% %
-% %
-% %
-% \subsection{Front-End Data Collection}
-
-% The following describes our data collection methodology:
-
-% \begin{enumerate}
-%     \item use the {\tt genthat} tracer to run tests, vignettes, and examples for a package, and optionally repeat for package's reverse dependencies;
-%     \item gather type and attribute information of all arguments and return values from all calls to all functions by looking at generated trace data;
-%     \item per function, consolidate information to generate a type signature.
-% \end{enumerate}
-
-% In order to get a global picture of R usage, we do the above over \textit{all} CRAN and Bioconductor packages, and aggregate type and attribute signature results over all of them.
-
-%
-%
-%
-%
-\subsection{Back-End Data Collection}
-
-\AT{Do we want to go down this route with this paper?
-This would be Konrad's bit, if we want to try to stick our data together.}
-
 %
@@
 In this section, we ... .
+
+%
+%
+%
+%
+\subsection{Types in R}
+
+\AT{may not be the correct place for this. i think i accidentally deleted it.}
+
+R is a unityped language, and as such is trivially monomorphic, and this notion of monomorphism is far from useful.
+We are interested in expanding the notion of ``type'' in R, and thankfully the language has some tools on offer.
+\AT{this is discussed later on}
+
+%
+%
+\subsubsection{What are possible type systems?}
+
+Great question.
+Some things:
+
+\begin{itemize}
+
+\item {\it The Obvious Type System (OTS)}: this type system corresponds directly with the type information that one can fetch using the {\tt typeof}, {\tt attributes}, and {\tt class} functions.
+In this system, there is no distinction between vectors and scalars, and lists of differently-typed elements (e.g., a list of integers vs.
+a list of strings), and there is a distinction between integers and doubles, and the type of NULL and all other types.
+This type system isn't particularly informative to a programmer, though it does reflect the treatment of values in the R runtime, for the {\tt typeof} a value is the type that the value has according to the R internals.
+
+\item {\it The Fine-Grained Type System (FGTS)}: this type systems includes a number of relevant additions.
+Chiefly, it distinguishes between vectors and scalars, as it is conceivable that boxing scalars as single-element vectors might be overkill.
+In addition, list types are parameterized over the types of their elements (e.g., list<integer> vs. list<character>).
+Finally, NAs ``in the wild'' are given the type {\tt raw\_NA}, as {\tt typeof(NA) = logical} in R.
+
+\item {\it The Null-and-NA-Free Type System (NNTS)}: most languages consider NULLs to inhabit all types, and R does so for NAs \AT{elaborate}.
+This gives us ... .
+
+\end{itemize}
+
+Of course, one could conceive of a more fine-grained type system than FGTS, but it's unclear that we would gain much in doing so.
+Indeed, there is a space to explore in slightly less fine-grained type systems, which make modifications including:
+
+\begin{itemize}
+
+\item {\bf collapsing vectors and scalars}: this yields a type system slightly closer to OTS.
+
+\item {\bf rolling NULL into other types}: in most languages, NULL is an inhabitant of all types, and this adjustment seems interesting.
+Note that this goes beyond even OTS, as in R {\tt typeof(NULL) = NULL}.
+
+\item {\bf rolling NA into other types}: \AT{i should try this}
+similarly, NA is an inhabitant of all types, though NA standing alone has type {\tt logical}.
+FGTS distinguishes such NAs with the {\tt raw\_NA} type, and if instead we consider NA to inhabit all types we may \AT{win}.
+
+\item {\bf combining integers and doubles}: as in most languages, doubles and ints can stand-in for each other in nearly every scenario.
+Even if not, it is trivial to convert from one to the other, particularly converting ints to doubles.
+Aside, in R a double can be used to index a list, and R will floor it when performing the lookup.
+
+\end{itemize}
 

### 2019-03-15 notebooks/general_info.Rmd
framework for per-package tallies is available. will get top poly packages soon

@@
 ```
+
+What are the counts now?
+```{r}
+total_mono_type_basic_no_null_real <- apply_count_to_lopkg_c(los_basic_no_null_real, args_count_all_mono_type)
+total_mono_all_basic_no_null_real <- apply_count_to_lopkg_c(los_basic_no_null_real, args_count_all_mono_all)
+```
+
+Great.
+Now, what are the top instances of type-based polymorphism?
+```{r}
+poly_type_args_basic_no_null_real <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los_basic_no_null_real, extract_poly_type_args))
+poly_type_args_basic_no_null_real <- poly_type_args_basic_no_null_real[order(-poly_type_args_basic_no_null_real$count),] # sort by count
+poly_type_args_basic_no_null_real$perc <- round(poly_type_args_basic_no_null_real$count/sum(poly_type_args_basic_no_null_real$count) * 100, 2)
+```
+
+What's the top?
+```{r}
+format_df_for_print(poly_type_args_basic_no_null_real[1:20,])
+```
+
+Ok.
+What if we want some per package information?
+Well, to do it on one package (let's try rlang):
+
+```{r}
+rlang_tally <- tally_up_lofun_c(los$rlang)
+rlang_tally$tot_args
+rlang_tally$tot_mono_args_type
+format_df_for_print(rlang_tally$top_poly_type)
+```
+
+As you'll see, the numbers here are low because we didn't actually trace rlang itself, just that some other packages used rlang, giving us limited results.

### 2019-03-16 notebooks/general_info.Rmd
breakdown by package functionality added, will have numbers tomorrow. also, did primitive classes

@@
 As you'll see, the numbers here are low because we didn't actually trace rlang itself, just that some other packages used rlang, giving us limited results.
+
+Now, how about if we take all of R's base classes (i.e., those which occur naturally in R) and lump them all into a "primitive" class.
+What sort of class-based polymorphism do we see, now?
+
+```{r}
+los_primitive_class <- lapply(los, function(lof) {lapply(lof, class_to_primitive_df)})
+```
+
+```{r}
+total_mono_clas_prim <- apply_count_to_lopkg_c(los_primitive_class, args_count_all_mono_class)
+total_mono_clas_functions_prim <- apply_count_to_lopkg_c(los_primitive_class, fun_count_all_mono_class)
+```
+
+Let's get the top poly class now.
+```{r}
+poly_class_args_prim <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los_primitive_class, extract_poly_class_args))
+poly_class_args_prim <- poly_class_args_prim[order(-poly_class_args_prim$count),] # sort by count
+poly_class_args_prim$perc <- round(poly_class_args_prim$count/sum(poly_class_args_prim$count) * 100, 2)
+```
+
+```{r}
+format_df_for_print(poly_class_args_prim[1:20,])
+```
+
+The low percentage numbers suggest that no one user-defined pattern contributes too much to the class-based polymorphism.

### 2019-03-16 paper/main.tex
minor
mior
talked a bit about corpus, added some code stuff

@@
 programmers.  We perform a dynamic analysis, leveraging tests and use-cases,
-to determine the level of polymoprhism that is present in the code. We do
+to determine the level of polymorphism that is present in the code. We do
 this for several potential notions of types. Our result suggest that
@@
 %
+\subsection{The Corpus}
+
+\AT{I'm hesitant to go into too much detail here, as I'm not sure if we're quite done running our analysis.
+I can imagine a scenario where we will try to rerun with a longer timeout just to get a little bit more data (because why not).}
+
+We ran our analysis over a subset of CRAN, the Comprehensive R Archive Network.
+We analyzed \AT{over 7500} packages, over half of all packages available on CRAN.
+Our analysis had an explicit timeout of \AT{30 minutes} per package, as a number of R packages feature functions which have a nigh-eternal running time (e.g., machine learning packages and complex simulations) \AT{cite those packages/examples}.
+
+\AT{Should we talk about some of the packages?
+What else do we have to say about the corpus?
+We can restate the numbers from the abstract.}
+
+%
+%
+%
+%
 %

### 2019-03-16 paper/main.tex
working on biblio, and added small paper things

@@
 
-But invariably some applications written in those languages survive past the prototype stage.
+But invariably some applications written in those languages survive past the
+prototype stage.
 
@@
 
-Given that code almost never stops, and that any computation is likely to deliver a bunch of numbers, how can we trust those numbers?
+Given that code almost never stops, and that any computation is likely to
+deliver a bunch of numbers, how can we trust those numbers?
 
-Years of work in programming languages suggest that one cost efficient way to catch some errors is to add types to a language.
+Years of work in programming languages suggest that one cost efficient way
+to catch some errors is to add types to a language.
 
-In the last decade much attention has been given to the incremental addition of types. Successes like Hack, TypeScript and Racket suggest that this may be feasible for wide spectrum languages.
-
-%
-%
-%
-\section{Background}
-
-%
-%
-%
-%
-\subsection{The R Programming Language}
-
-R is a lazy, side-effecting, dynamically-typed, ... .
-
-Typical usage of the language is ... . \AT{Do we have something to cite for this?}
-\AT{Should we motivate why we only look at R packages?
-Some people might wonder why we don't look at non-library code.}
-
-Most R libraries (called \textit{packages}) are stored in the Comprehensive R Archive Network (CRAN) and Bioconductor package repositories.
-\AT{Talk about important packages? tidyverse?}
-
-\AT{Kind of ... want to combine the above two paragraphs somehow? Together, they lead in nicely to the genthat talk.}
-
-%
-%
-\subsection{The \texttt{genthat} Package}
-
-Unfortunately, getting our hands on ``front-line'' R code is tricky:
-R is not like many other languages, where large systems are built up and stored in a repository such as Github.
-Instead, R programmers are liable to leverage a number of packages and work them into their data analysis pipelines.
-
-Interestingly, we are somewhat able to replicate this using only package code.
-CRAN, for instance, has a policy that packages made available on the CRAN platform should be accompanied by a series of examples, tests, and/or vignettes showing off how the code is intended to be used.
-Nominally unrelated, the {\tt genthat} project aims to trace this swath of example code to generate new examples which package designers may not have thought of.
-Crucially, {\tt genthat} outputs complete trace information for each of the functions that it runs so that they can be run again in a fresh R instance (e.g., by capturing the random seed, environment variables, etc.).
-We can leverage this trace information to glean type and attribute patterns of functions, and since these traces correspond to examples of the packages being used (rather than just arbitrarily running package code), we have a reasonable approximation of how R programmers would use the packages.
-
-\AT{We use/modify the {\tt genthat} tracer, probably worth going in to more detail here.}
-
-%
-%
-%
-%
-%
-%
-\section{Motivation and High-level Overview}
+In the last decade much attention has been given to the incremental addition
+of types. Successes like Hack, TypeScript and Racket suggest that this may
+be feasible for wide spectrum languages.
 
@@
 R is a languages written by and for \textit{statisticians}.  It traces its
-roots to the S language, developed by ... .  R programmers don't write big
+roots to the S language~\cite{becker1988new}, developed by Richard Becker, John Chambers, and Allan Wilks.  R programmers don't write big
 servers or software systems; instead, they write and modify small data
@@
 %
+\section{The R Programming Language}
+
+R is a lazy, side-effecting, dynamically-typed, ... .
+
+Typical usage of the language is ... . \AT{Do we have something to cite for
+  this?}  \AT{Should we motivate why we only look at R packages?  Some
+  people might wonder why we don't look at non-library code.}
+
+Most R libraries (called \textit{packages}) are stored in the Comprehensive
+R Archive Network (CRAN) and Bioconductor package repositories.  \AT{Talk
+  about important packages? tidyverse?}
+
+\AT{Kind of ... want to combine the above two paragraphs somehow? Together,
+  they lead in nicely to the genthat talk.}
+
+\subsection{The \texttt{genthat} Package}
+
+Unfortunately, getting our hands on ``front-line'' R code is tricky: R is
+not like many other languages, where large systems are built up and stored
+in a repository such as Github.  Instead, R programmers are liable to
+leverage a number of packages and work them into their data analysis
+pipelines.
+
+Interestingly, we are somewhat able to replicate this using only package
+code.  CRAN, for instance, has a policy that packages made available on the
+CRAN platform should be accompanied by a series of examples, tests, and/or
+vignettes showing off how the code is intended to be used.  Nominally
+unrelated, the {\tt genthat} project aims to trace this swath of example
+code to generate new examples which package designers may not have thought
+of.  Crucially, {\tt genthat} outputs complete trace information for each of
+the functions that it runs so that they can be run again in a fresh R
+instance (e.g., by capturing the random seed, environment variables, etc.).
+We can leverage this trace information to glean type and attribute patterns
+of functions, and since these traces correspond to examples of the packages
+being used (rather than just arbitrarily running package code), we have a
+reasonable approximation of how R programmers would use the packages.
+
+\AT{We use/modify the {\tt genthat} tracer, probably worth going in to more
+  detail here.}
 %
-%
-%
-%
-%
+
 \section{The Method}
@@
       many functions or arguments have the requisite signature;
-    \item for \textbf{finding the most common signatures}: we enumerate as before, and sort the signatures based on how often they appear;
+    \item for \textbf{finding the most common signatures}: we enumerate as
+      before, and sort the signatures based on how often they appear;
 \end{itemize}
@@
 
-Types (in the sense of the result of a call to {\tt typeof}) are easy to analyze as values can only have one type.
-Attributes, however, are a different story, as values can have an arbitrary number of attributes.
-Essentially, we're not interesting by function arguments which have had multiple attributes, we're interested in function arguments which have had multiple attribute \textit{patterns}.
-\AT{is this defined earlier?}
-Arguments are said to be polymorphic in attribute if they have been inhabited with values which have had different attribute patterns.
-\AT{We have not investigated whether functions which are polymorphic in attribute use attributes extensively inside the function code.}
+Types (in the sense of the result of a call to {\tt typeof}) are easy to
+analyze as values can only have one type.  Attributes, however, are a
+different story, as values can have an arbitrary number of attributes.
+Essentially, we're not interesting by function arguments which have had
+multiple attributes, we're interested in function arguments which have had
+multiple attribute \textit{patterns}. \AT{is this defined earlier?}
+Arguments are said to be polymorphic in attribute if they have been
+inhabited with values which have had different attribute patterns.  \AT{We
+  have not investigated whether functions which are polymorphic in attribute
+  use attributes extensively inside the function code.}
 
@@
 
-First and foremost, {\tt genthat} captures arguments only on function exit, so as to not force promises during function execution.
-\AT{Uh oh:} One clear limitation here is that if an argument \textit{gains} attributes over the functione execution, then they will appear in our signatures.
-As R is side-effecting, this is an entirely valid practice (i.e., passing objects to functions to modify them), though \AT{it's unclear if people actually do this}.
+First and foremost, {\tt genthat} captures arguments only on function exit,
+so as to not force promises during function execution.  \AT{Uh oh:} One
+clear limitation here is that if an argument \textit{gains} attributes over
+the functione execution, then they will appear in our signatures.  As R is
+side-effecting, this is an entirely valid practice (i.e., passing objects to
+functions to modify them), though \AT{it's unclear if people actually do
+  this}.
 
@@
 
-Ultimately, our goal is to collect data and report on patterns which emerge organically.
-That said, the comprehensive picture of type usage we are developing can also be used to inform the design of a set of \textit{type annotations} for the language, and this goal informs our analysis.
-As we collect our data, we keep in mind the idea that the data we produce should clearly suggest what sorts of type annotations would reflect language usage patterns.
+Ultimately, our goal is to collect data and report on patterns which emerge
+organically.  That said, the comprehensive picture of type usage we are
+developing can also be used to inform the design of a set of \textit{type
+  annotations} for the language, and this goal informs our analysis.  As we
+collect our data, we keep in mind the idea that the data we produce should
+clearly suggest what sorts of type annotations would reflect language usage
+patterns.
 
@@
 
+%
+%
+%
+%
+\section{Related Work}
+
+\AT{Not sure what the best place for this is, but we should discuss these papers somewhere.}
+
+The idea of measuring polymorphism via some program analysis isn't new.
+For instance, some work~\cite{aakerblom2015measuring} ... .
+
+More generally, there is an existing literature on analyzing usage patterns of language features.
+For instance, the dynamic features of Smalltalk were analyzed in some work~\cite{callau2011howdevelopers}, ...
+
+Other analysis work~\cite{milojkovic2017duck} ... .
+
 \section{Conclusions and Future Work}

### 2019-03-16 paper/main.tex
intro

@@
 
+Our community builds, improves, and reasons about programming languages.  To
+make design decisions that benefit most users of a language, we need to
+understand the language we are working with as well as the real-world needs
+it answers. Often, we, as researchers, can appeal to our intuition as many
+languages are intended to be general purpose and appeal to users with some 
+training.
+
 A number of programming languages have gained popularity for data science
@@
 be feasible for wide spectrum languages.
-
-Our community is centred on building, improving, and reasoning about
-programming languages.  A key component in all this is to truly
-\textit{understand} the languages we're working on, so that we may make the
-best decisions for the benefit of all users of a programming language.  In
-many cases, we as researchers can appeal to our own intuition to at least
-build a sense for how a language is used, as language users typically have
-some training from an experienced programmer or software engineer (e.g.,
-computer science studies, online tutorials written by language experts,
-etc.).  \AT{We claim that this is emphatically not the case in R.}
 

### 2019-03-16 notebooks/general_info.Rmd
added per-package tally, and a size metric to start off with. looks like the top packages we looked at (with a rudimentary notion of polymorphism size) are plotmo, photobiology, phylosim, imager, IsoplotR, and ggplot2. lots of plotters
coinciding up and running, will generalize the function next. gonna look into other occurences of char/real that are unexplained by coinciding list signatures

@@
 The low percentage numbers suggest that no one user-defined pattern contributes too much to the class-based polymorphism.
+
+Alright, let's do a package tally for all packages.
+First, let's make all the classes primitive in our previous collapsed list.
+
+```{r}
+los_basic_no_null_real_primitive_class <- lapply(los_basic_no_null_real, function(lof) {lapply(lof, class_to_primitive_df)})
+```
+
+Note: this takes a long time to run.
+```{r}
+all_tallies_all_reduced <- lapply(los_basic_no_null_real_primitive_class, tally_up_lofun_c)
+```
+
+```{r}
+all_tallies_sizes <- lapply(all_tallies_all_reduced, function(x) x$tot_size_of_poly)
+```
+
+Ok.
+It's certainly not perfect, but one metric I used for figuring out the size of polymorphism is just by summing up the type-, attribute-, and class-based polymorphism of each argument in a function.
+This might give us some leads.
+The top 20 polymorphic packages by this metric is:
+
+```{r}
+head(sort(unlist(all_tallies_sizes), decreasing=TRUE), 20)
+```
+
+Ok.
+Let's see what we can get out of coinciding type signatures.
+With out reduced type system, we might want to know a little bit more about the "real", "character" type signature, which is by far one of the most common signatures.
+We hypothesize that these should appear alongside lists (and possibly vectors).
+For lists only, we can do this easily:
+```{r}
+# for list coinciding
+coincide_list_index <- get_fun_sigs_matching_pred_from_lopkg(los_basic_no_null_real_primitive_class, does_df_have_coinciding_types_index_and_list_reduced_TS)
+
+# just real, character
+char_real_sigs <- get_fun_sigs_matching_pred_from_lopkg(los_basic_no_null_real_primitive_class, does_df_have_X_types, types_to_find=c("real", "character"), strict=TRUE)
+```
+
+Let's look for vectors, also.
+```{r}
+coincide_list_vector_index <- get_fun_sigs_matching_pred_from_lopkg(los, does_df_have_coinciding_types_index_and_list_or_vector)
+```

### 2019-03-17 paper/main.tex
working on the results section

@@
 \begin{abstract}
-The R programming language is widely used in a variety of scientfic domains
+The R programming language is widely used in a variety of scientific domains
 for tasks related to data science. The language was designed to favor an
@@
 that question we study the polymorphism that is present in over X millions
-of lines of R code, XX packages, written over a period of Y years by Z
+of lines of R code spread among some 7500 packages, written over a period of over 20 years by over 10000
 programmers.  We perform a dynamic analysis, leveraging tests and use-cases,
@@
 \end{abstract}
+
 \maketitle
@@
 languages are intended to be general purpose and appeal to users with some 
-training.
+training. Unfortunately, these intuitions don't always apply to domain-specific
+languages.
 
@@
 
-To faciliate programming these language offer all kinds of short cuts and
-facilities that increase the likelyhood that code will ``just work''.
+To facilitate programming these language offer all kinds of short cuts and
+facilities that increase the likelihood that code will ``just work''.
 
@@
 The results obtained are published and even sometimes drive policy decisions.
+\AT{There's a cool stat in the blog I linked which says that a lot of the Stack
+Overflow traffic R gets is from government officials.}
 
@@
 R is a languages written by and for \textit{statisticians}.  It traces its
-roots to the S language~\cite{becker1988new}, developed by Richard Becker, John Chambers, and Allan Wilks.  R programmers don't write big
+roots to the S language~\cite{becker1988new}, developed by Richard Becker, 
+John Chambers, and Allan Wilks.  R programmers don't write big
 servers or software systems; instead, they write and modify small data
@@
 In spite of these flaws, the language is incredibly popular.
-\AT{We'll want to cite something here.}
+\AT{We'll want to cite something here, maybe the blog I linked?}
 
@@
       type information.
+\end{itemize}
+
+The main contributions of this work are:
+
+\begin{itemize}
+
+	\item \AT{maybe} a tool to analyze the polymorphicity of R code;
+
+	\item a study of polymorphism in the R language \AT{using said tool};
+
+	\item a number of proposed static type systems for R which cover most of the 
+	type usage patterns which arise in R code.
+
 \end{itemize}
@@
 
+There are a number of interesting packages in our corpus.
+\AT{References for all of these.}
+For example, we have analyzed some packages for plotting data: among others, the {\tt ggplot2} package is an immensely popular package for generating highly-customizable plots, and the {\tt plotmo} package is an interesting package aimed at displaying models.
+We also analyzed {\tt dplyr}, an integral package to the {\tt tidyverse} set of R packages \AT{talk about tidyverse earlier, or here}.
+{\tt dplyr} is a popular package that makes full use of R's features to create what is tantamount to a small DSL for data manipulation within R.
+Another interesting package is {\tt rquery}, a package for generating SQL-like queries.
+
 \AT{Should we talk about some of the packages?
@@
 
-\AT{Tables, graphs, code examples should go in here.}
-In this section, we ... .
+\AT{PSA:} There are a few ways we could do this section.
+One way would be to go through our data in a sort of linear fashion, akin to how the main notebook is set up.
+We can explore each of our type systems one at a time, and present them (thus breaking the section up by type system).
+Another option is to just break down by-result, so have a section for the counts, for the top polymorphism...
+I like the first option better.
+
+\AT{Draft}
+In this section, we will present the results of our analysis. 
+We approach the data with the purpose of informing the design of a static type system for R.
+We propose to begin with a very basic type system, discussed next.
+
+\subsection{Base R Type System}
+
+The {\it Base R Type System} (BRTS) ...
 

### 2019-03-17 notebooks/general_info.Rmd
small additions to the code. date lookup was not impressive, but i used classes---dates show up in class info in a few ways---and not too much overlap was there. mind you, some arguments still claim that they are dates, even without class info to corroborate. ill dig deeper

@@
 ```
+
+73% coincide.
+Nice.
+Dates might be another possibility?
+Let's check:
+
+```{r}
+sigs_with_real_char <- get_fun_sigs_matching_pred_from_lopkg(los_basic_no_null_real_primitive_class, does_df_have_X_types, types_to_find=c("real", "character"), strict=TRUE)
+```
+
+```{r}
+sigs_with_dates <- get_fun_sigs_matching_pred_from_lopkg(los_basic_no_null_real_primitive_class, does_df_have_date_real_char)
+```
+
+Hmm... Not as much as we would have liked, just 20% of the leftover (or 5% of the total).
+Why don't we just see what's leftover:
+```{r}
+just_index <- get_fun_sigs_matching_pred_from_lopkg(los, does_df_have_coinciding_types_index_and_NOT_list_or_vector)
+```
+
+This isn't working the way I was expecting.

### 2019-03-17 paper/main.tex

@@
 
-\title{A large-scale study of polymorphism in R}
+\title{A Large-scale Study of Polymorphism in R}
 
@@
 that question we study the polymorphism that is present in over X millions
-of lines of R code spread among some 7500 packages, written over a period of over 20 years by over 10000
-programmers.  We perform a dynamic analysis, leveraging tests and use-cases,
-to determine the level of polymorphism that is present in the code. We do
-this for several potential notions of types. Our result suggest that
-polymorphism is important in some key parts of the system but that
-relatively simple type annotations could be used to capture most of the
-interesting cases.
+of lines of R code spread among some 7,500 packages, written over a period of
+over 20 years by over 10,000 programmers.  We perform a dynamic analysis,
+leveraging tests and use-cases, to determine the level of polymorphism that
+is present in the code. We do this for several potential notions of
+types. Our result suggest that polymorphism is important in some key parts
+of the system but that relatively simple type annotations could be used to
+capture most of the interesting cases.
 \end{abstract}
@@
 Our community builds, improves, and reasons about programming languages.  To
-make design decisions that benefit most users of a language, we need to
-understand the language we are working with as well as the real-world needs
-it answers. Often, we, as researchers, can appeal to our intuition as many
+make design decisions that benefit most users, we need to understand the
+language we are working with as well as the real-world needs it
+answers. Often, we, as researchers, can appeal to our intuition as many
 languages are intended to be general purpose and appeal to users with some
-training. Unfortunately, these intuitions don't always apply to domain-specific
-languages.
+computer science training. Unfortunately, these intuitions don't always
+apply to domain-specific languages, languages designed for and by a specific
+group of users to solve very specific needs. This is the case of the data
+science language R.
 
-A number of programming languages have gained popularity for data science
-tasks, they include MATLAB, Python, Julia and R.
+R and it's ancestor S are languages designed, implemented and maintained by
+statisticians. Originally they were designed as glue languages, languages
+that would allow to read data into vectors and call statistical routines
+written in Fortran. Over three decades, the languages became widely used
+across many fields of science and in industry for data analysis and data
+visualization; with time additional features were added.  Modern R, as a
+linguistic object of study, is fascinating. It is a vectorized, dynamically
+typed, lazy functional language with limited side-effects, extensive
+reflective facilities and retrofitted object-oriented programming support.
 
-These languages are all dynamically typed.
+Many of the design decisions that gave us R were intended to foster an
+interactive, exploratory, programming style. This includes, to name a few,
+the lack of type annotations on variables and functions, the ability to use
+syntactic shortcut, and the automatic conversion between data types.  While
+these choices have led to a language that is surprisingly easy to use by
+beginners --many data science programs do not teach the language itself but
+simply introduce some of its key libraries-- they have also created a
+language where almost all computations yield a numeric result and where
+errors can go undetected.
 
-This is because they emphasize ease of programming, interactive data
-exploration, and light-weight syntax.
-
-To facilitate programming these language offer all kinds of short cuts and
-facilities that increase the likelihood that code will ``just work''.
-
-For example abbreviation of name arguments and automatic conversions.
-
-But invariably some applications written in those languages survive past the
-prototype stage.
-
-The results obtained are published and even sometimes drive policy decisions.
-\AT{There's a cool stat in the blog I linked which says that a lot of the Stack
-Overflow traffic R gets is from government officials.}
-
-Given that code almost never stops, and that any computation is likely to
-deliver a bunch of numbers, how can we trust those numbers?
-
-Years of work in programming languages suggest that one cost efficient way
-to catch some errors is to add types to a language.
-
-In the last decade much attention has been given to the incremental addition
-of types. Successes like Hack, TypeScript and Racket suggest that this may
-be feasible for wide spectrum languages.
-
-R is a languages written by and for \textit{statisticians}.  It traces its
-roots to the S language~\cite{becker1988new}, developed by Richard Becker, 
-John Chambers, and Allan Wilks.  R programmers don't write big
-servers or software systems; instead, they write and modify small data
-processing pipelines as a means to facilitate interaction with digital data.
-This leads to ``front-line'' R code (i.e., code written for users, not
-programmers) not being made available on typical platforms like Github.
-\AT{Maybe front-line code isn't a good word, is there a good way to describe
-  non-library code?}  Put differently, R is essentially a really big
-domain-specific language.
-
-The design of R is also rife with questionable decisions.  For example, the
-language is lazy \textit{and} side-effecting, which one could describe as an
-antagonistic pair of features.  In addition, data structures must be
-entirely copied when being modified, which appears to be far from ideal in a
-language meant to process (thus, modify) big data.
-
-In spite of these flaws, the language is incredibly popular.
-\AT{We'll want to cite something here, maybe the blog I linked?}
+One way to increase assurance in the results obtained when using R would be
+to add type annotations to functions and variable declarations. These
+annotations could then be used, either statically or (more likely)
+dynamically, to catch mismatches between expected and provided data values.
+The nature of R is such that it is unlikely to be ever fully statically
+checked, furthermore end users may not be willing to write types when
+carrying out exploratory programming tasks. So, we are looking for an
+optional type system that would allow us to capture as much of behavior of
+library functions as possible while remaining easy to understand for
+end-users and library developers alike.
 

### 2019-03-17 paper/main.tex
talked about genthat, and working on method

@@
 \section{The R Programming Language}
+\label{sec:rlang}
 
@@
 Interestingly, we are somewhat able to replicate this using only package
-code.  CRAN, for instance, has a policy that packages made available on the
+code.  CRAN, has a policy that packages made available on the
 CRAN platform should be accompanied by a series of examples, tests, and/or
 vignettes showing off how the code is intended to be used.  Nominally
-unrelated, the {\tt genthat} project aims to trace this swath of example
+unrelated, the {\tt genthat} project~\cite{kvrikava2018tests} traces this swath of example
 code to generate new examples which package designers may not have thought
-of.  Crucially, {\tt genthat} outputs complete trace information for each of
-the functions that it runs so that they can be run again in a fresh R
-instance (e.g., by capturing the random seed, environment variables, etc.).
-We can leverage this trace information to glean type and attribute patterns
-of functions, and since these traces correspond to examples of the packages
-being used (rather than just arbitrarily running package code), we have a
-reasonable approximation of how R programmers would use the packages.
+of.  As it happens, we can leverage {\tt genthat} to extract dynamic type
+information from running package test code, which presumably indicates
+how package code is intended to be used.
 
-\AT{We use/modify the {\tt genthat} tracer, probably worth going in to more
-  detail here.}
-%
+\AT{Draft}
+
+We modify the {\tt genthat} tracer to capture the type information of arguments.
+{\tt genthat} generates traces for each unique function invocation with the express purpose of re-invoking the function at a later time.
+Instead of generating traces for these invocations, we can instead capture type information of the arguments and return values of functions.
+This is done by evaluating the arguments and return values of functions at the end of the function's scope in the relevant environment.
+The sort of types and granularity of the type information that are generated can be specified in our tracer:
+Essentially, our tracer is parameterized over a user-specified type system.
+For instance, if desired our tracer can dig into lists and data frames to get the types of list elements and fields.
+Our tracer produces trace results for each unique type signature encountered, which may then be used for data analysis.
+
+An unfortunate consequence of R's most common usage pattern is the lack of "deployed" R code.
+R is most often used for exploratory data analysis, and very little of this analysis code is made publicly available on e.g. GitHub.
+The {\tt genthat} angle is thus desirable:
+With the scarcity of non-library code, examples of package usage is about as representative as one can hope for of the sort of code people will write in their analysis pipelines.
+
+We can do a little bit better than merely package usage examples, though.
+Available through CRAN is a list of {\it reverse dependencies} of packages:
+for some package {\tt p}, a list of all packages which require {\tt p} in some way.
+Our tracer is able to leverage and trace this code as well.
+When running a package, we also trace how all inherited functions are used, thus painting an even clearer picture of function usage.
+This additional data is considered alongside similar data collected for all other packages, as well as the data obtained by tracing the package's examples.
+
+% Some prose we might want to pull from.
+%First, we leverage {\tt genthat}'s tracer to trace function executions.  The
+%idea here is that we don't really have access to non-library code written in
+%R, as general use patterns are (possibly?) to write small scripts which
+%analyze some bit of data and possibly visualize results.  The goal is likely
+%not to build big working systems, instead to explore data with by writing
+%and rewriting small scripts, ad infinitum.  In looking at package tests,
+%vignettes, and examples, we are painting a picture of how the package
+%designers intended their packages to be used, which we believe is a close
+%approximation of what R users would do.
 
@@
 
-First, we leverage {\tt genthat}'s tracer to trace function executions.  The
-idea here is that we don't really have access to non-library code written in
-R, as general use patterns are (possibly?) to write small scripts which
-analyze some bit of data and possibly visualize results.  The goal is likely
-not to build big working systems, instead to explore data with by writing
-and rewriting small scripts, ad infinitum.  In looking at package tests,
-vignettes, and examples, we are painting a picture of how the package
-designers intended their packages to be used, which we believe is a close
-approximation of what R users would do.
+%
+%
+%
+%
+\subsection{Data Collection Pipeline}
 
-{\tt genthat} generates a trace file which can be loaded into R and contains
-all of the necessary information to rerun the tested function.  This
-includes the random seed and all relevant environment data.  We utilize this
-trace data to collect type information: we run each argument in the
-specified environment, and collect type information through R's {\tt typeof}
-function, and attribute information through R's {\tt attributes} function.
+As we mentioned in the previous section, we modify the {\tt genthat} tracer to collect and output the type information for all function invocations.
+
+\AT{somewhere}
+Our analysis can be parameterized by a user-specified ``type system'', simply a map between existing dynamically observable type information and user-defined types.
+If no type system is specified, the analysis gains as much information as it can.
+For instance, if the analysis encounters {\tt typeof(x) == "double"}, it will look at the length of {\tt x} to determine if {\tt x} is a scalar or a vector, generating the annotation {\tt scalar/double} or {\tt vector/double} accordingly.
+Note that this is imperfect, as a vector of unit length would (perhaps incorrectly) show up as a {\tt scalar/X}, but any other invocation where the vector does not have unit length will generate the correct {\tt vector/X} type, and the distinction between scalars and vectors can be later collapsed if desired.
+Note also that the type system described here is the \AT{SOMETHING} Type System, described in detail in Section~\AT{TODO}.
+
+The process for tracing one package {\tt p} is as follows:
+
+\begin{enumerate}
+
+	\item first, all functions in {\tt p} are instrumented with on-exit hooks.
+	These hooks call a number of R's available reflection functions on each of the arguments, as well as the return type:
+	The called functions are {\tt typeof}, {\tt class}, and {\tt attributes}.
+	See Section~\ref{sec:rlang} for details on these functions.
+
+	\item then, depending on the specified type
+
+\end{enumerate}
 

### 2019-03-18 paper/main.tex
intro

@@
 
+This papers is a data-driven study of what a type system for the R language
+could look like. Longer term, our intention is to propose changes to
+language, but for any changes to be accepted by the user community, they
+must clearly benefit the language without endangering backwards
+compatibility. Our goal is thus to find a compromise between simplicity and
+usefulness; the proposed type system should cover most common programming
+idioms while remaining easy to use. In order to do this, we need to
+understand the degree of polymorphism present in R code, that is to say, how
+programmers leverage the dynamic nature of R to write code that can accept
+arguments of different types.  This understanding will drive our design.
+
 One way or another, R is here to stay, so we may as well figure out how to
@@
 \begin{itemize}
-
 \item \AT{maybe} a tool to analyze the polymorphicity of R code;
-
 \item a study of polymorphism in the R language \AT{using said tool};
-
 \item a number of proposed static type systems for R which cover most of the
   type usage patterns which arise in R code.
-
 \end{itemize}

### 2019-03-18 paper/main.tex
intro + r lang + bib

@@
 \documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-\usepackage{booktabs,listings}
+\usepackage{booktabs,listings,xspace}
 \lstset{language=R}
@@
 
+\newcommand{\PACKAGES}{7,500\xspace}
+\newcommand{\PROGRAMMERS}{10,000\xspace}
+
 \begin{abstract}
@@
 that question we study the polymorphism that is present in over X millions
-of lines of R code spread among some 7,500 packages, written over a period of
-over 20 years by over 10,000 programmers.  We perform a dynamic analysis,
-leveraging tests and use-cases, to determine the level of polymorphism that
-is present in the code. We do this for several potential notions of
-types. Our result suggest that polymorphism is important in some key parts
-of the system but that relatively simple type annotations could be used to
-capture most of the interesting cases.
+of lines of R code spread among some \PACKAGES packages, written over a
+period of over 20 years by over 10,000 programmers.  We perform a dynamic
+analysis, leveraging tests and use-cases, to determine the level of
+polymorphism that is present in the code. We do this for several potential
+notions of types. Our result suggest that polymorphism is important in some
+key parts of the system but that relatively simple type annotations could be
+used to capture most of the interesting cases.
 \end{abstract}
@@
 
-One way or another, R is here to stay, so we may as well figure out how to
-deal with it.  One aspect of this is understanding R programmers and how
-they interact with the language, and a step towards this goal is to
-understand \AT{how types arise in R programs.}  We will shed some light on
-this by analyzing the usage of R functions, and build a picture of how
-polymorphic R code is.
+We propose to capture the degree of polymorphism present in R by the means
+of a dynamic analysis of widely used libraries. For each function call we
+can record the types of its arguments and of its return value. This allows
+us to observe how many different combination of types are accepted by any
+given function. Unlike many other languages, R has a carefully curated
+software repository called CRAN. To be deposited in CRAN, a package must
+come with sample dataset, tests and executable use-cases. As part of normal
+operations these tests are run regularly and failing packages are removed.
+This allowed us to have access to \PACKAGES libraries and about an order of
+magnitude more runnable scripts that exercise those libraries.
 
-\begin{itemize}
-    \item Understanding R can help us improve the language, as well as
-      better equip us to build languages which have an appeal outside of the
-      realm of computer science.
-    \item We're gonna tackle this by looking at how R programmers use types
-      and type information in their programs.
-    \item We will be mindful of the runtime representations of values to
-      give more context to the information we glean from programmer-visible
-      type information.
-\end{itemize}
+\section{The R Programming Language}\label{sec:rlang}
 
-The main contributions of this work are:
+Over the last decade, the R Project has become a key tool for implementing
+sophisticated data analysis algorithms in fields ranging from Computational
+Biology~\cite{R05} to Political Science~\cite{R:Keele:2008}. At the heart of
+the R project is a \emph{dynamic, lazy, functional, object-oriented}
+programming language with a rather unusual combination of
+features~\cite{ecoop12} designed to ease learning by non-programmer and
+enable rapid development of new statistical methods.  The language, commonly
+referred to as R was designed in 1993 by Ross Ihaka and Robert
+Gentleman~\cite{R96} as a successor to S~\cite{S88}.  First released in
+1995, under a GNU license, R rapidly became the lingua franca for
+statistical data analysis. Today, there are over 13,000 R packages available
+from repositories such as CRAN and Bioconductor.  With 55 R user groups
+world-wide, Smith~\cite{eco11} estimates that there are over 2 million
+end-users.
 
-\begin{itemize}
-\item \AT{maybe} a tool to analyze the polymorphicity of R code;
-\item a study of polymorphism in the R language \AT{using said tool};
-\item a number of proposed static type systems for R which cover most of the
-  type usage patterns which arise in R code.
-\end{itemize}
-
-%
-\section{The R Programming Language}
-\label{sec:rlang}
-
-R is a lazy, side-effecting, dynamically-typed, ... .
-
-Typical usage of the language is ... . \AT{Do we have something to cite for
-  this?}  \AT{Should we motivate why we only look at R packages?  Some
-  people might wonder why we don't look at non-library code.}
-
-Most R libraries (called \textit{packages}) are stored in the Comprehensive
-R Archive Network (CRAN) and Bioconductor package repositories.  \AT{Talk
-  about important packages? tidyverse?}
-
-\AT{Kind of ... want to combine the above two paragraphs somehow? Together,
-  they lead in nicely to the genthat talk.}
-
-\subsection{The \texttt{genthat} Package}
+\section{The \texttt{genthat} Package}
 
@@
 Interestingly, we are somewhat able to replicate this using only package
-code.  CRAN, has a policy that packages made available on the
-CRAN platform should be accompanied by a series of examples, tests, and/or
-vignettes showing off how the code is intended to be used.  Nominally
-unrelated, the {\tt genthat} project~\cite{kvrikava2018tests} traces this swath of example
-code to generate new examples which package designers may not have thought
-of.  As it happens, we can leverage {\tt genthat} to extract dynamic type
-information from running package test code, which presumably indicates
-how package code is intended to be used.
+code.  CRAN, has a policy that packages made available on the CRAN platform
+should be accompanied by a series of examples, tests, and/or vignettes
+showing off how the code is intended to be used.  Nominally unrelated, the
+{\tt genthat} project~\cite{issta18} traces this swath of example code to
+generate new examples which package designers may not have thought of.  As
+it happens, we can leverage {\tt genthat} to extract dynamic type
+information from running package test code, which presumably indicates how
+package code is intended to be used.
 
@@
 
-We modify the {\tt genthat} tracer to capture the type information of arguments.
-{\tt genthat} generates traces for each unique function invocation with the express purpose of re-invoking the function at a later time.
-Instead of generating traces for these invocations, we can instead capture type information of the arguments and return values of functions.
-This is done by evaluating the arguments and return values of functions at the end of the function's scope in the relevant environment.
-The sort of types and granularity of the type information that are generated can be specified in our tracer:
-Essentially, our tracer is parameterized over a user-specified type system.
-For instance, if desired our tracer can dig into lists and data frames to get the types of list elements and fields.
-Our tracer produces trace results for each unique type signature encountered, which may then be used for data analysis.
+We modify the {\tt genthat} tracer to capture the type information of
+arguments.  {\tt genthat} generates traces for each unique function
+invocation with the express purpose of re-invoking the function at a later
+time.  Instead of generating traces for these invocations, we can instead
+capture type information of the arguments and return values of functions.
+This is done by evaluating the arguments and return values of functions at
+the end of the function's scope in the relevant environment.  The sort of
+types and granularity of the type information that are generated can be
+specified in our tracer: Essentially, our tracer is parameterized over a
+user-specified type system.  For instance, if desired our tracer can dig
+into lists and data frames to get the types of list elements and fields.
+Our tracer produces trace results for each unique type signature
+encountered, which may then be used for data analysis.
 
-An unfortunate consequence of R's most common usage pattern is the lack of "deployed" R code.
-R is most often used for exploratory data analysis, and very little of this analysis code is made publicly available on e.g. GitHub.
-The {\tt genthat} angle is thus desirable:
-With the scarcity of non-library code, examples of package usage is about as representative as one can hope for of the sort of code people will write in their analysis pipelines.
+An unfortunate consequence of R's most common usage pattern is the lack of
+"deployed" R code.  R is most often used for exploratory data analysis, and
+very little of this analysis code is made publicly available on e.g. GitHub.
+The {\tt genthat} angle is thus desirable: With the scarcity of non-library
+code, examples of package usage is about as representative as one can hope
+for of the sort of code people will write in their analysis pipelines.
 
@@
 Available through CRAN is a list of {\it reverse dependencies} of packages:
-for some package {\tt p}, a list of all packages which require {\tt p} in some way.
-Our tracer is able to leverage and trace this code as well.
-When running a package, we also trace how all inherited functions are used, thus painting an even clearer picture of function usage.
-This additional data is considered alongside similar data collected for all other packages, as well as the data obtained by tracing the package's examples.
+for some package {\tt p}, a list of all packages which require {\tt p} in
+some way.  Our tracer is able to leverage and trace this code as well.  When
+running a package, we also trace how all inherited functions are used, thus
+painting an even clearer picture of function usage.  This additional data is
+considered alongside similar data collected for all other packages, as well
+as the data obtained by tracing the package's examples.
 
@@
 %
-%
-%
-%
 \subsection{Data Collection Pipeline}
 
-As we mentioned in the previous section, we modify the {\tt genthat} tracer to collect and output the type information for all function invocations.
+As we mentioned in the previous section, we modify the {\tt genthat} tracer
+to collect and output the type information for all function invocations.
 
-\AT{somewhere}
-Our analysis can be parameterized by a user-specified ``type system'', simply a map between existing dynamically observable type information and user-defined types.
-If no type system is specified, the analysis gains as much information as it can.
-For instance, if the analysis encounters {\tt typeof(x) == "double"}, it will look at the length of {\tt x} to determine if {\tt x} is a scalar or a vector, generating the annotation {\tt scalar/double} or {\tt vector/double} accordingly.
-Note that this is imperfect, as a vector of unit length would (perhaps incorrectly) show up as a {\tt scalar/X}, but any other invocation where the vector does not have unit length will generate the correct {\tt vector/X} type, and the distinction between scalars and vectors can be later collapsed if desired.
-Note also that the type system described here is the \AT{SOMETHING} Type System, described in detail in Section~\AT{TODO}.
+\AT{somewhere} Our analysis can be parameterized by a user-specified ``type
+system'', simply a map between existing dynamically observable type
+information and user-defined types.  If no type system is specified, the
+analysis gains as much information as it can.  For instance, if the analysis
+encounters {\tt typeof(x) == "double"}, it will look at the length of {\tt
+  x} to determine if {\tt x} is a scalar or a vector, generating the
+annotation {\tt scalar/double} or {\tt vector/double} accordingly.  Note
+that this is imperfect, as a vector of unit length would (perhaps
+incorrectly) show up as a {\tt scalar/X}, but any other invocation where the
+vector does not have unit length will generate the correct {\tt vector/X}
+type, and the distinction between scalars and vectors can be later collapsed
+if desired.  Note also that the type system described here is the
+\AT{SOMETHING} Type System, described in detail in Section~\AT{TODO}.
 
@@
 \begin{enumerate}
-
 \item first, all functions in {\tt p} are instrumented with on-exit hooks.
-	These hooks call a number of R's available reflection functions on each of the arguments, as well as the return type:
-	The called functions are {\tt typeof}, {\tt class}, and {\tt attributes}.
-	See Section~\ref{sec:rlang} for details on these functions.
+  These hooks call a number of R's available reflection functions on each of
+  the arguments, as well as the return type: The called functions are {\tt
+    typeof}, {\tt class}, and {\tt attributes}.  See Section~\ref{sec:rlang}
+  for details on these functions.
 
 \item then, depending on the specified type 
-
 \end{enumerate}
@@
 %
-%
-%
-\subsection{Attributes}
-\label{sec:method:attributes}
+\subsection{Attributes}\label{sec:method:attributes}
 
@@
 
-More generally, there is an existing literature on analyzing usage patterns of language features.
-For instance, the dynamic features of Smalltalk were analyzed in some work~\cite{callau2011howdevelopers}, ...
+More generally, there is an existing literature on analyzing usage patterns
+of language features.  For instance, the dynamic features of Smalltalk were
+analyzed in some work~\cite{callau2011howdevelopers}, ...
 
@@
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{biblio}
+\bibliography{bib/biblio,bib/jv,bib/r,bib/new}
 

### 2019-03-18 paper/main.tex
intro

@@
 \usepackage{my_style}
+
+\definecolor{LightGray}{rgb}{.95,.95,.95}
+\definecolor{Gray}{rgb}{.3,.3,.3}
+\definecolor{DarkGray}{rgb}{.5,.5,.5}
+
+\lstset{ %
+  columns=flexible,
+  captionpos=b,
+  frame=single,
+  framerule=0pt,
+  framexleftmargin=-1mm,
+  framexrightmargin=-1mm,
+  tabsize=2,
+  belowskip=0pt,
+  basicstyle=\small\ttfamily,
+  backgroundcolor=\color{LightGray},
+  emphstyle=\sffamily,
+  keywordstyle=\bfseries,
+  commentstyle=\color{Gray}\em,
+  stringstyle=\color{Gray}
+}
+
+\lstdefinestyle{R}{ %
+  language=R,
+  deletekeywords={env, equal, c, runif, trace, args},
+  breaklines=true
+}
+
+\lstdefinestyle{Rin}{ %
+  style=R,
+  numberstyle=none,
+  basicstyle=\normalsize\ttfamily,
+  breaklines=false
+}
+\newcommand{\code}[1]{\lstinline|#1|\xspace}
 
@@
 magnitude more runnable scripts that exercise those libraries.
+
+The contributions of this paper are thus as follows:
+\begin{itemize}
+\item A large-scale analysis of the polymorphism present in function
+  signatures of \PACKAGES widely used and actively maintained R packages.
+\item A tracing and analysis pipeline that extends a previously published
+  test generation tool named \code{genthat}.
+\item Manual analysis of 100 functions to validate the dynamic analysis
+  results.
+\end{itemize}
+
+One threat to validity of our work is that we rely on dynamic analysis, so
+our conclusions are only as good as the coverage of the possibly function
+behaviors. Previous work~\cite{issta18}, reported that running all the
+scripts that come with CRAN packages gives, on average, 68\% test coverage.
+We attempted to mitigate the threat coming from the fact that only part of
+the code is being exercised by manual analysis. It would be reasonable to
+ask for confirmation of the data by static analysis of the code, but sound
+static analysis of R is difficult because of the extensive use of reflective
+features such as \code{eval} and of the ability to redefine the meaning of
+operators such as \code{+} and \code{if}.  Another threat to validity is
+that we only have access to code that has been deposited in the CRAN
+repository. While this may bias our findings towards code written to be
+reusable and, possibly, better engineered than typical user code. This is
+also the code that would most benefit from type annotations.
 
@@
 
-We ran our analysis over a subset of CRAN, the Comprehensive R Archive Network.
-We analyzed \AT{over 7500} packages, over half of all packages available on CRAN.
-Our analysis had an explicit timeout of \AT{30 minutes} per package, as a number of R packages feature functions which have a nigh-eternal running time (e.g., machine learning packages and complex simulations) \AT{cite those packages/examples}.
+We ran our analysis over a subset of CRAN, the Comprehensive R Archive
+Network.  We analyzed \AT{over 7500} packages, over half of all packages
+available on CRAN.  Our analysis had an explicit timeout of \AT{30 minutes}
+per package, as a number of R packages feature functions which have a
+nigh-eternal running time (e.g., machine learning packages and complex
+simulations) \AT{cite those packages/examples}.
 
-There are a number of interesting packages in our corpus.
-\AT{References for all of these.}
-For example, we have analyzed some packages for plotting data: among others, the {\tt ggplot2} package is an immensely popular package for generating highly-customizable plots, and the {\tt plotmo} package is an interesting package aimed at displaying models.
-We also analyzed {\tt dplyr}, an integral package to the {\tt tidyverse} set of R packages \AT{talk about tidyverse earlier, or here}.
-{\tt dplyr} is a popular package that makes full use of R's features to create what is tantamount to a small DSL for data manipulation within R.
-Another interesting package is {\tt rquery}, a package for generating SQL-like queries.
+There are a number of interesting packages in our corpus.  \AT{References
+  for all of these.}  For example, we have analyzed some packages for
+plotting data: among others, the {\tt ggplot2} package is an immensely
+popular package for generating highly-customizable plots, and the {\tt
+  plotmo} package is an interesting package aimed at displaying models.  We
+also analyzed {\tt dplyr}, an integral package to the {\tt tidyverse} set of
+R packages \AT{talk about tidyverse earlier, or here}.  {\tt dplyr} is a
+popular package that makes full use of R's features to create what is
+tantamount to a small DSL for data manipulation within R.  Another
+interesting package is {\tt rquery}, a package for generating SQL-like
+queries.
 

### 2019-03-18 paper/main.tex
sec 2

@@
 Biology~\cite{R05} to Political Science~\cite{R:Keele:2008}. At the heart of
-the R project is a \emph{dynamic, lazy, functional, object-oriented}
-programming language with a rather unusual combination of
+the R project is a \emph{vectorized, dynamic, lazy, functional,
+  object-oriented} programming language with a rather unusual combination of
 features~\cite{ecoop12} designed to ease learning by non-programmer and
@@
 end-users.
+
+As an introduction to R, consider the code snippet in Fig.~\ref{sample} from
+a top-level interaction where the user defines a function \code{normSum}
+that accepts vectors of integers, logicals, doubles and complex values and
+normalizes the vector with respect to its sum and rounds the results. The
+function definition does not require type annotations, and all operations
+transparently work on vectors of any length and different types.
+
+\begin{figure}{\small
+\begin{lstlisting}[style=R]
+> normSum <- function( m )  round( m / sum(m), 2)
+> normSum(c(1L,3L,6L))
+[1] 0.1 0.3 0.6
+> normSum(c(1.1,3.3,6.6))
+[1] 0.1 0.3 0.6
+> normSum(c(1.6,3.3,6.1))
+[1] 0.15 0.30 0.55
+> normSum(complex(r=rnorm(3),i=rnorm(3))
+[1] 0.49+0.21i 0.30-0.18i 0.22-0.03i
+\end{lstlisting}}
+\caption{Sample R code}\label{sample}
+\end{figure}
 

### 2019-03-18 paper/main.tex
sec2

@@
 \end{figure}
+
+\subsection{Types of Data}
+
+Before attempting to define a type system for R, we should understand the
+different kinds of values that programs operate on.  As we will see
+different notions of type may emerge depending on how granular we want to
+be.
+
+The first distinction that can be made is based on the value of the
+\code{typeof} function.  This function, given a value, will return one of
+the following types: \code{logical} (vector of booleans), \code{integer}
+(vector of integers), \code{double} (vector of floating points),
+\code{complex} (vector of complex numbers), \code{character} (vector of
+character strings), \code{raw} (vector of bytes) and \code{list}
+(polymorphic vector), \x`code{NULL} (vector of NULL),
+\code{closure} (function), \code{environment} (naming environment),
+\code{S4} (S4 object) and others that are used within the implementation of
+R (\code{special}, \code{builtin}, \code{symbol}, \code{pairlist},
+\code{promise}, \code{language}, \code{char}, \code{...}, \code{any},
+\code{expression}, \code{externalptr}, \code{bytecode} and \code{weakref}).
+
+%% TODO : make the types all bold.
+
+attributes:
+
+  - dim
+
+class:
 

### 2019-03-18 paper/main.tex
minor

@@
 character strings), \code{raw} (vector of bytes) and \code{list}
-(polymorphic vector), \x`code{NULL} (vector of NULL),
+(polymorphic vector), \code{NULL} (vector of NULL),
 \code{closure} (function), \code{environment} (naming environment),
@@
 
-\bibliographystyle{ACM-Reference-Format}
+\bibliographystyle{boilerplate/ACM-Reference-Format}
 \bibliography{bib/biblio,bib/jv,bib/r,bib/new}

### 2019-03-18 paper/main.tex
few type examples

@@
 [1] 0.15 0.30 0.55
-> normSum(complex(r=rnorm(3),i=rnorm(3))
+> normSum(complex(r=rnorm(3),i=rnorm(3)))
 [1] 0.49+0.21i 0.30-0.18i 0.22-0.03i
@@
 %% TODO : make the types all bold.
+\AT{Some examples of use of \code{typeof}:}
+To build a sense of how these types arise in R code, consider the following:
+\AT{We could think of another way to do this also.}
 
-attributes:
+\begin{itemize}
+
+	\item \code{typeof(5) == typeof(c(5)) == typeof(c(5, 5)) == "double"}.
+	As we mentioned, R is vectorized, so even apparent scalars are vectors.
+	\AT{Just for fun,} note that \code{5[1] == 5} in R.
+
+	\item to actually get an integer, you need \code{typeof(5L) == "integer"}.
+	Doubles are by far the most common numeric type in R \AT{I can get data for this}, and you can even use them to index lists and vectors: \code{list(1, 2)\[\[1.9\]\] == 1}.
+	Note the double brackets for list indexing.
+
+	\item NAs are common in data-centric languages (e.g. MATLAB, \AT{others?}), and R is no exception.
+	As for their type, we see that \code{typeof(NA) == "logical"}, but that isn't to say that NA only has type logical.
+	Consider \code{typeof(c(1, NA, 3)\[2\])}: NA actually inhabits every type in R, \AT{details}.
+
+	\item \AT{other examples?}
+
+\end{itemize}
+
+Another form of type information relates to metadata on values, and R allows programmers to add said metadata in the form of {\it attributes}.
+The functionality to do so in R is through the \code{attributes(x)} and \code{attr(x, "name")}  functions.
+\code{attributes}
 

### 2019-03-19 notebooks/corpus_stats_and_plots.Rmd
got counts data, starting to plot the data. also wrangled some more data, about 500 packages, which ill pull in soon

---
title: "Corpus Stats"
output: html_notebook
---

First order of business: the lines of code of the corpus.
Load.

```{r}
r_counts <- readRDS("../data/R_loc_by_package.RDS")
c_counts <- readRDS("../data/c_loc_by_package.RDS")

# just named lists. some elements (in either) are integer(0) i believe, as some packages
# just have data.
```

...I guess let's try to plot it??

```{r}
library(ggplot2)

r_c_df <- data.frame(package_name = names(r_counts), loc_r = r_counts)
# looks weird sorted
# r_c_df <- r_c_df[order(r_c_df$loc_r),]

plot_r_counts <- ggplot( data=r_c_df, aes(x=1:length(r_counts), y=loc_r, group=1))
plot_r_counts <- plot_r_counts + geom_line() + geom_point()
plot_r_counts <- plot_r_counts + labs(x = "Corpus of Analyzed Packages", y = "Lines of R Code in Package")

plot_r_counts <- plot_r_counts + theme(axis.text.x = element_blank())

# plot it
plot_r_counts

```

### 2019-03-19 notebooks/general_info.Rmd
got counts data, starting to plot the data. also wrangled some more data, about 500 packages, which ill pull in soon

@@
 los_basic_no_null <- lapply(los_basic, function(lof) {lapply(lof, function(df) {fold_NULL_into_other_types_df(df)})})
+
+# also curious if we roll raw_NAs in
+los_basic_no_null_no_NA <- lapply(los_basic, function(lof) {lapply(lof, function(df) {fold_NULL_and_NA_into_other_types_df(df)})})
 ```

### 2019-03-19 notebooks/corpus_stats_and_plots.Rmd
fixed small error in c locs and worked more on notebook

@@
 # looks weird sorted
-# r_c_df <- r_c_df[order(r_c_df$loc_r),]
+r_c_df <- r_c_df[order(r_c_df$loc_r),]
+# looks weirder log10-scaled
+r_c_df$loc_r <- log10(r_c_df$loc_r)
 
@@
 plot_r_counts <- plot_r_counts + geom_line() + geom_point()
-plot_r_counts <- plot_r_counts + labs(x = "Corpus of Analyzed Packages", y = "Lines of R Code in Package")
+plot_r_counts <- plot_r_counts + labs(
+  title = "Analysis of R Corpus (R Code Portion)",
+  x = "Corpus of Analyzed Packages", 
+  y = "Lines of R Code in Package (Log10-scaled)"
+)
 
@@
 ```
+
+The outlier is the spatstat package.
+How about for C?
+
+```{r}
+library(ggplot2)
+
+c_c_df <- data.frame(package_name = names(c_counts), loc_c = c_counts)
+# removing packages w/ no C code (-1)
+c_c_df$loc_c <- lapply(c_c_df$loc_c, function(loc) if (loc == -1) 0 else loc)
+# looks weird sorted
+c_c_df <- c_c_df[order(c_c_df$loc_c),]
+# looks weirder log10-scaled
+c_c_df$loc_c <- log10(c_c_df$loc_c)
+
+plot_c_counts <- ggplot( data=c_c_df, aes(x=1:length(c_counts), y=loc_c, group=1))
+plot_c_counts <- plot_c_counts + geom_line() + geom_point()
+plot_c_counts <- plot_c_counts + labs(
+  title = "Analysis of R Corpus (C Code Portion)",
+  x = "Corpus of Analyzed Packages", 
+  y = "Lines of C Code in Package (Log10-scaled)"
+)
+
+plot_c_counts <- plot_c_counts + theme(axis.text.x = element_blank())
+
+# plot it
+plot_c_counts
+
+```

### 2019-03-19 notebooks/corpus_stats_and_plots.Rmd
working on the plot notebook, have the data just have to fight with ggplot enough to get it to display the thing that i want

--- 2019-03-19 notebooks/corpus_stats_and_plots.Rmd
@@
 ```{r}
+library(reshape2)
+library(ggplot2)
+
 r_counts <- readRDS("../data/R_loc_by_package.RDS")
@@
 ```{r}
-library(ggplot2)
-
 r_c_df <- data.frame(package_name = names(r_counts), loc_r = r_counts)
 # looks weird sorted
-# r_c_df <- r_c_df[order(r_c_df$loc_r),]
+r_c_df_sub <- r_c_df[order(r_c_df$loc_r),]
+# just get the entries with > 0 lines of code
+r_c_df_sub <- r_c_df_sub[r_c_df_sub$loc_r > 0, ]
+# looks weirder log10-scaled
+r_c_df_sub$loc_r <- log10(r_c_df_sub$loc_r)
 
-plot_r_counts <- ggplot( data=r_c_df, aes(x=1:length(r_counts), y=loc_r, group=1))
+plot_r_counts <- ggplot( data=r_c_df_sub, aes(x=1:length(r_c_df_sub$loc_r), y=loc_r, group=1))
 plot_r_counts <- plot_r_counts + geom_line() + geom_point()
-plot_r_counts <- plot_r_counts + labs(x = "Corpus of Analyzed Packages", y = "Lines of R Code in Package")
+plot_r_counts <- plot_r_counts + labs(
+  title = "Analysis of R Corpus (R Code Portion)",
+  x = "Corpus of Analyzed Packages", 
+  y = "Lines of R Code in Package (Log10-scaled)"
+)
 
@@
 ```
+
+The outlier is the spatstat package.
+How about for C?
+
+```{r}
+library(ggplot2)
+
+c_c_df <- data.frame(package_name = names(c_counts), loc_c = c_counts)
+# looks weird sorted
+c_c_df_sub <- c_c_df[order(c_c_df$loc_c),]
+# get rid of 0s
+c_c_df_sub <- c_c_df_sub[c_c_df_sub$loc_c > 0, ]
+# looks weirder log10-scaled
+c_c_df_sub$loc_c <- log10(c_c_df_sub$loc_c)
+
+plot_c_counts <- ggplot( data=c_c_df_sub, aes(x=1:length(c_c_df_sub$loc_c), y=loc_c, group=1))
+plot_c_counts <- plot_c_counts + geom_line() + geom_point()
+plot_c_counts <- plot_c_counts + labs(
+  title = "Analysis of R Corpus (C Code Portion)",
+  x = "Corpus of Analyzed Packages", 
+  y = "Lines of C Code in Package (Log10-scaled)"
+)
+
+plot_c_counts <- plot_c_counts + theme(axis.text.x = element_blank())
+
+# plot it
+plot_c_counts
+
+```
+
+Similar to the R stuff.
+Now, we should combine these plots.
+```{r}
+combo_df <- r_c_df
+combo_df$loc_c <- c_c_df$loc_c
+combo_df$loc_r <- sapply(combo_df$loc_r, function(x) if (x == 0) x else log10(x))
+combo_df$loc_c <- sapply(combo_df$loc_c, function(x) if (x == 0) x else log10(x))
+
+# sort
+combo_df <- combo_df[order(combo_df$loc_r), ]
+
+# TODO: not staying sorted, probably id.vars
+zzz <- melt(combo_df, id.vars="package_name", variable.name = "language")
+
+plot_both <- ggplot( data=zzz, aes(package_name, value, group=1))
+plot_both <- plot_both + geom_line(aes(colour = language)) + geom_point(aes(colour = language))
+plot_both <- plot_both + labs(
+  title = "Analysis of R Corpus (R and C)",
+  x = "Corpus of Analyzed Packages", 
+  y = "Lines of Code in Package (Log10-scaled)"
+)
+
+plot_both <- plot_both + theme(axis.text.x = element_blank())
+
+# plot it
+plot_both
+```

### 2019-03-19 notebooks/corpus_stats_and_plots.Rmd
got plots working, need to sort out the many many points. but the sorting is there

@@
 Now, we should combine these plots.
+Messy graphs, beware!
 ```{r}
@@
 combo_df <- combo_df[order(combo_df$loc_r), ]
+# this makes sure the package_names are in the right order, ggplot2 plots according to factors
+# and levels
+combo_df$package_name <- factor(combo_df$package_name, levels=combo_df$package_name)
+
+combo_df_c <- combo_df[order(combo_df$loc_c), ]
+combo_df_c$package_name <- factor(combo_df_c$package_name, levels=combo_df_c$package_name)
+
+combo_df_c <- combo_df_c[combo_df_c$loc_c > 0, ]
 
@@
 zzz <- melt(combo_df, id.vars="package_name", variable.name = "language")
+zzz_c <- melt(combo_df_c, id.vars="package_name", variable.name = "language")
 
@@
 
+plot_both_c <- ggplot( data=zzz_c, aes(package_name, value, group=1))
+plot_both_c <- plot_both_c + geom_line(aes(colour = language)) + geom_point(aes(colour = language))
+plot_both_c <- plot_both_c + labs(
+  title = "Analysis of R Corpus (R and C)",
+  x = "Corpus of Analyzed Packages", 
+  y = "Lines of Code in Package (Log10-scaled)"
+)
+
 plot_both <- plot_both + theme(axis.text.x = element_blank())
+
+plot_both_c <- plot_both_c + theme(axis.text.x = element_blank())
 
@@
 plot_both
+
+plot_both_c
 ```

### 2019-03-19 paper/main.tex
got counts data, starting to plot the data. also wrangled some more data, about 500 packages, which ill pull in soon
working on corpus

@@
 
-\newcommand{\PACKAGES}{7,500\xspace}
+\newcommand{\PACKAGES}{7,763\xspace}
 \newcommand{\PROGRAMMERS}{10,000\xspace}
+\newcommand{\PERCENTCRAN}{56\%\xspace}
+\newcommand{\CRANTOTAL}{13,841\xspace}
+\newcommand{\RLOC}{8,996,572\xspace}
+\newcommand{\YEARS}{20\xspace}
 
@@
 question: \emph{what could a static type system for R look like?}  To answer
-that question we study the polymorphism that is present in over X millions
-of lines of R code spread among some \PACKAGES packages, written over a
-period of over 20 years by over 10,000 programmers.  We perform a dynamic
+that question we study the polymorphism that is present in \RLOC lines of R 
+code spread among some \PACKAGES packages, written over a
+period of over \YEARS years by \PROGRAMMERS programmers.  We perform a dynamic
 analysis, leveraging tests and use-cases, to determine the level of
@@
 
+\AT{Note: some stuff I wrote that might belong somewhere in here, if not, delete.}
+As mentioned, the syntax is highly unrestrictive to make this data analysis easy.
+All functions in R are vectorized, and consequently operators and functions are multi-purpose.
+For instance, the syntax for adding numbers, vectors, and matrices uses the same "+" operator.
+\AT{End note.}
+
 Over the last decade, the R Project has become a key tool for implementing
@@
 
+%
+% Corpus Section
+%
+% > talk about CRAN
+% > what is a package 
+% > look at genthat paper for inspiration
+% > communicate the size of the adventure
+% > convince people that this is real
+% > lines of C and R code per project? big graph could be nice
+% > give people a sense of our data set
+
+\section{The Corpus}
+
+\AT{Draft}
+
+In this section, we will discuss the R language ecosystem with a focus on our dataset.
+
+First and foremost, recall that R is essentially a DSL for statisticians, designed to help them in their day-to-day tasks of data analytics.
+Among other things, R has an out-of-the-box data type called the data frame for storing data, and this data type is essential to a large part of R's functionality.
+A data frame is a two-dimensional array where each row represents an observation, with columns indicating the values of some observation.
+For example, our own analysis is written in R, and one of our data frames has a row for each package analyzed, with columns for values ranging from number of polymorphic functions, number of monomorphic arguments, unique attribute patterns, etc.
+\AT{Put a figure?}
+Much of R's base functionality, and indeed much of the functionality provided by packages, operate on data frames.
+
+In general, the data analysis workflow tightly integrates the analyst and the environment.
+\AT{I think the tidyverse "manifesto" has some stuff we could inspire ourselves from. But, generally:}
+For tasks related to data visualization, an analyst will need to: (1) import a data set, (2) sanitize or otherwise transform the data, (3) plot the data, (4) repeat as necessary.
+For tasks related to testing hypotheses or otherwise validating conclusions: (1) import a data set, (2) sanitize or otherwise transform the data, (3) call some functions to perform necessary computations for the test, (4) repeat as necessary.
+The pattern of import, transform, do something, repeat is referred to as a {\it pipeline} in the R community.
+Indeed, RStudio, one of the most popular IDEs for developing R, is designed to support this iterative behaviour.
+In it, users create {\it notebooks} (akin to Jupyter notebooks for those familiar) wherein they may write text alongside their computation pipeline, and these notebooks expose the code written by the user, altogether facilitating collaboration with other users.
+
+Now, there is a wealth of literature built up over centuries for statistics and the foundation of what we refer to as data science, and a data analyst might be interested in leveraging that in their pipeline:
+Thanks to R's {\it package} ecosystem, this is easy.
+In R, using a library (called a package in R) is straightforward and can be done at any point in an R script.
+After installing the package, a simple call to \code{library(pkg\_name)} exposes all of \code{pkg\_name}'s exported functions for use in the rest of the script.
+All package code is available for download from CRAN, the Comprehensive R Archive Network, and CRAN sports over 13,000 packages.
+CRAN is an actively maintained and curated environment, and there are a number of reasonable requirements for a package to be accepted for publication.
+Package authors are encouraged to write tests for their code, as well as examples showing off basic usage and functionality, and vignettes which show how one might use the package in a realistic scenario.
+Bioconductor is one other popular package repository for R, focusing on supporting analysis of genomic data.
+
+Our aim is to analyze a large portion of existing R code to build a sense of how programmers use types in the language, to perhaps inform the design of a static type system for the language.
+Unfortunately, there isn't much non-library code available for our analysis to peruse:
+Besides CRAN and Bioconductor, most R scripts are not built to be exported, and are part of an iterative and interactive process of data analysis.
+Unlike other programming languages, R code has a very small footprint on GitHub, and most of the R code available there is referenced through CRAN \AT{maybe we can get numbers here?
+or just take out the sentence}.
+
+ \AT{Still working on this, meeting speaker now.}
+
+%We ran our analysis over a subset of CRAN, the Comprehensive R Archive
+%Network.  We analyzed \AT{over 7500} packages, over half of all packages
+%available on CRAN.  Our analysis had an explicit timeout of \AT{30 minutes}
+%per package, as a number of R packages feature functions which have a
+%nigh-eternal running time (e.g., machine learning packages and complex
+%simulations) \AT{cite those packages/examples}.
+%
+%There are a number of interesting packages in our corpus.  \AT{References
+%for all of these.}  For example, we have analyzed some packages for
+%plotting data: among others, the {\tt ggplot2} package is an immensely
+%popular package for generating highly-customizable plots, and the {\tt
+%plotmo} package is an interesting package aimed at displaying models.  We
+%also analyzed {\tt dplyr}, an integral package to the {\tt tidyverse} set of
+%R packages \AT{talk about tidyverse earlier, or here}.  {\tt dplyr} is a
+%popular package that makes full use of R's features to create what is
+%tantamount to a small DSL for data manipulation within R.  Another
+%interesting package is {\tt rquery}, a package for generating SQL-like
+%queries.
+
+%
+%
+%
+%
 \section{The \texttt{genthat} Package}
@@
 patterns.
-
-%
-%
-%
-%
-\subsection{The Corpus}
-
-\AT{I'm hesitant to go into too much detail here, as I'm not sure if we're quite done running our analysis.
-I can imagine a scenario where we will try to rerun with a longer timeout just to get a little bit more data (because why not).}
-
-We ran our analysis over a subset of CRAN, the Comprehensive R Archive
-Network.  We analyzed \AT{over 7500} packages, over half of all packages
-available on CRAN.  Our analysis had an explicit timeout of \AT{30 minutes}
-per package, as a number of R packages feature functions which have a
-nigh-eternal running time (e.g., machine learning packages and complex
-simulations) \AT{cite those packages/examples}.
-
-There are a number of interesting packages in our corpus.  \AT{References
-  for all of these.}  For example, we have analyzed some packages for
-plotting data: among others, the {\tt ggplot2} package is an immensely
-popular package for generating highly-customizable plots, and the {\tt
-  plotmo} package is an interesting package aimed at displaying models.  We
-also analyzed {\tt dplyr}, an integral package to the {\tt tidyverse} set of
-R packages \AT{talk about tidyverse earlier, or here}.  {\tt dplyr} is a
-popular package that makes full use of R's features to create what is
-tantamount to a small DSL for data manipulation within R.  Another
-interesting package is {\tt rquery}, a package for generating SQL-like
-queries.
-
-\AT{Should we talk about some of the packages?
-What else do we have to say about the corpus?
-We can restate the numbers from the abstract.}
 

### 2019-03-19 paper/main.tex
corpus and some tidbits added to the r section

@@
 For instance, the syntax for adding numbers, vectors, and matrices uses the same "+" operator.
+
+\AT{Also:}
+The R language has a very easy-to-use FFI to C, and even to C++ through another R package, {\tt Rcpp}.
+
 \AT{End note.}
@@
 \end{figure}
+
+\AT{I wrote a paragraph in the next section which seems to belong here, maybe work this in?}
+Besides these primitive data types, R has an out-of-the-box data type called the data frame for storing data, and this data type is essential to a large part of R's functionality.
+A data frame is a two-dimensional array where each row represents an observation, with columns indicating the values of some observation.
+For example, our own analysis is written in R, and one of our data frames has a row for each package analyzed, with columns for values ranging from number of polymorphic functions, number of monomorphic arguments, unique attribute patterns, etc.
+\AT{Put a figure?}
+Much of R's base functionality, and indeed much of the functionality provided by packages, operate on data frames.
 
@@
 
-First and foremost, recall that R is essentially a DSL for statisticians, designed to help them in their day-to-day tasks of data analytics.
-Among other things, R has an out-of-the-box data type called the data frame for storing data, and this data type is essential to a large part of R's functionality.
-A data frame is a two-dimensional array where each row represents an observation, with columns indicating the values of some observation.
-For example, our own analysis is written in R, and one of our data frames has a row for each package analyzed, with columns for values ranging from number of polymorphic functions, number of monomorphic arguments, unique attribute patterns, etc.
-\AT{Put a figure?}
-Much of R's base functionality, and indeed much of the functionality provided by packages, operate on data frames.
-
-In general, the data analysis workflow tightly integrates the analyst and the environment.
-\AT{I think the tidyverse "manifesto" has some stuff we could inspire ourselves from. But, generally:}
-For tasks related to data visualization, an analyst will need to: (1) import a data set, (2) sanitize or otherwise transform the data, (3) plot the data, (4) repeat as necessary.
-For tasks related to testing hypotheses or otherwise validating conclusions: (1) import a data set, (2) sanitize or otherwise transform the data, (3) call some functions to perform necessary computations for the test, (4) repeat as necessary.
+The R environment is designed to accommodate the workflow of the data analyst, a workflow which tightly integrates the analyst and the data.
+As an example example, for a task related to data visualization, an analyst will need to: (1) import a data set, (2) sanitize or otherwise transform the data, (3) plot the data, (4) repeat as necessary.
+Similarly, for a task related to testing hypotheses or otherwise validating conclusions: (1) import a data set, (2) sanitize or otherwise transform the data, (3) call some functions to perform necessary computations for the test, (4) repeat as necessary.
 The pattern of import, transform, do something, repeat is referred to as a {\it pipeline} in the R community.
@@
 
-Now, there is a wealth of literature built up over centuries for statistics and the foundation of what we refer to as data science, and a data analyst might be interested in leveraging that in their pipeline:
-Thanks to R's {\it package} ecosystem, this is easy.
-In R, using a library (called a package in R) is straightforward and can be done at any point in an R script.
-After installing the package, a simple call to \code{library(pkg\_name)} exposes all of \code{pkg\_name}'s exported functions for use in the rest of the script.
-All package code is available for download from CRAN, the Comprehensive R Archive Network, and CRAN sports over 13,000 packages.
-CRAN is an actively maintained and curated environment, and there are a number of reasonable requirements for a package to be accepted for publication.
-Package authors are encouraged to write tests for their code, as well as examples showing off basic usage and functionality, and vignettes which show how one might use the package in a realistic scenario.
-Bioconductor is one other popular package repository for R, focusing on supporting analysis of genomic data.
+By and large, the most extensive repository of R code is CRAN, the Comprehensive R Archive Network.
+After installing a package, a simple call to \code{library(pkg\_name)} exposes all of \code{pkg\_name}'s exported functions for use in the rest of the script.
+CRAN sports over 13,00 packages, and is an actively maintained and curated environment, in that there are a number of requirements for a package to be accepted for publication, particularly with regards to proper package documentation.
+One such requirement is the presence of at least one vignette, a documented walk-through of some of the package's functions.
+\AT{Do we want an example of a vignette? We could put one.}
+Besides vignettes, package authors are encouraged to write tests for their code as well as examples showing off basic usage and functionality.
 
-Our aim is to analyze a large portion of existing R code to build a sense of how programmers use types in the language, to perhaps inform the design of a static type system for the language.
-Unfortunately, there isn't much non-library code available for our analysis to peruse:
-Besides CRAN and Bioconductor, most R scripts are not built to be exported, and are part of an iterative and interactive process of data analysis.
-Unlike other programming languages, R code has a very small footprint on GitHub, and most of the R code available there is referenced through CRAN \AT{maybe we can get numbers here?
+One goal of our analysis is to build a sense of how programmers use types in R, and unfortunately there isn't much non-library code available for our analysis to peruse.
+Most R scripts are not built to be exported, and are part of an iterative and interactive process of data analysis, and unlike other programming languages, R has a very small footprint on GitHub, and most of the R code available there is referenced through CRAN \AT{maybe we can get numbers here?
 or just take out the sentence}.
+There are a number of potential explanations for this, including the proprietary nature of certain data sets (e.g., governments working with sensitive census data, or medical researchers working with private patient data).
 
- \AT{Still working on this, meeting speaker now.}
+The lack of non-package code is somewhat mitigated by the good culture of packages published on CRAN, in that they typically provide tests and examples of intended package use.
+We modify the {\tt genthat} tracer, discussed in more detail in the next section, which essentially traces these tests, examples, and vignettes to generate new package tests.
+\AT{Not sure how to word the second half of this paragraph.}
+In focusing on these exemplary snippets, we should capture the intent of package designers, which gives up hope that our proposed type annotations would be adopted.
+
+%
+%
+%
+%
+\subsection{Our Test Suite}
+
+We ran our analysis over a subset of CRAN, analyzing \PACKAGES packages, account for some \PERCENTCRAN of all packages available on CRAN.
+\AT{Talk about timeout?}  Due to the nature of some packages (e.g., packages implementing a computationally-intensive model with many tests), we ran our analysis with a 1 hour timeout.
+Our tracer also traces reverse-dependencies: if some package utilizes some other package, both are instrumented and traced.
+This serves to further mitigate the lack of non-library R code.
+
+To get a better sense of the corpus we're dealing with, we will highlight a few packages that we have analyzed:
+
+\begin{itemize}
+
+	\item \textbf{\texttt{ggplot2}} is an important package for building graphics and plotting data in R.
+	{\tt ggplot2} is part of the tidyverse, which bills itself as an ``opinionated collection of R packages designed for data science''.
+	Essentially, packages in the tidyverse use a common grammar and are all designed to address the main issues faced in data analysis.
+	{\tt ggplot2}'s role in the tidyverse is to make creating elegant plots easy: in it, graphics are created declaratively in a manner based on The Grammar of Graphics \AT{cite}.
+
+	\item \textbf{\texttt{dplyr}} is another integral package to the tidyverse.
+	In it is defined a convenient piece of R syntax: the pipe operator {\tt \%>\%}.
+	If using {\tt dplyr}, one can write \code{f(x) \%>\% g()}, which is equivalent to \code{g(f(x))}.
+	{\tt dplyr} also defines a number of highly useful functions for manipulating data frames, and considered alongside the pipe operator form the underpinning of the tidyverse data pipeline philosophy.
+
+	\item \textbf{\texttt{zoo}} is a popular R package for dealing with time series, themselves a common occurrence in statistics.
+
+	\item \textbf{\texttt{spatstat}} is the largest package in our corpus in terms of lines of R code.
+	This package is designed for spatial statistics, focusing on the analysis of spatial point patterns.
+	It is, in some sense, the definitive spatial statistics package in R.
+
+	\item \textbf{\texttt{multinet}} is the largest package in our corpus in terms of lines of C code.
+	{\tt multinet} is a package designed to store and analyze multilayer social networks.
+
+\end{itemize}
 

### 2019-03-20 paper/main.tex
corpus, method

@@
 
-\AT{Draft}
-
 In this section, we will discuss the R language ecosystem with a focus on our dataset.
@@
 \AT{Not sure how to word the second half of this paragraph.}
-In focusing on these exemplary snippets, we should capture the intent of package designers, which gives up hope that our proposed type annotations would be adopted.
+In focusing on these exemplary snippets, we should capture the intent of package designers, which gives us hope that our proposed type annotations would be adopted.
 
@@
 
-We ran our analysis over a subset of CRAN, analyzing \PACKAGES packages, account for some \PERCENTCRAN of all packages available on CRAN.
+We ran our analysis over a subset of CRAN, analyzing \PACKAGES packages, accounting for some \PERCENTCRAN of all packages available on CRAN.
+These packages have a total of \RLOC lines of code, and were written by \PROGRAMMERS programmers over a period of \YEARS years.
 \AT{Talk about timeout?}  Due to the nature of some packages (e.g., packages implementing a computationally-intensive model with many tests), we ran our analysis with a 1 hour timeout.
@@
 	\item \textbf{\texttt{zoo}} is a popular R package for dealing with time series, themselves a common occurrence in statistics.
+	One of the main functions in this package is the eponymous {\tt zoo} function, which takes in a numeric vector or list along with some parameters, and produces a ``zoo'' object for use in other package functions.
 
@@
 
-%We ran our analysis over a subset of CRAN, the Comprehensive R Archive
-%Network.  We analyzed \AT{over 7500} packages, over half of all packages
-%available on CRAN.  Our analysis had an explicit timeout of \AT{30 minutes}
-%per package, as a number of R packages feature functions which have a
-%nigh-eternal running time (e.g., machine learning packages and complex
-%simulations) \AT{cite those packages/examples}.
-%
-%There are a number of interesting packages in our corpus.  \AT{References
-%for all of these.}  For example, we have analyzed some packages for
-%plotting data: among others, the {\tt ggplot2} package is an immensely
-%popular package for generating highly-customizable plots, and the {\tt
-%plotmo} package is an interesting package aimed at displaying models.  We
-%also analyzed {\tt dplyr}, an integral package to the {\tt tidyverse} set of
-%R packages \AT{talk about tidyverse earlier, or here}.  {\tt dplyr} is a
-%popular package that makes full use of R's features to create what is
-%tantamount to a small DSL for data manipulation within R.  Another
-%interesting package is {\tt rquery}, a package for generating SQL-like
-%queries.
+Most of the packages we looked at have between 100 and 1000 lines of R code.
 
@@
 \section{The \texttt{genthat} Package}
+\label{sec:genthat}
 
-Unfortunately, getting our hands on ``front-line'' R code is tricky: R is
-not like many other languages, where large systems are built up and stored
-in a repository such as Github.  Instead, R programmers are liable to
-leverage a number of packages and work them into their data analysis
-pipelines.
+As we mentioned, getting our hands on ``front-line'' non-library R code is tricky: 
+R is not like many other languages, where large systems are built up and stored in a repository such as GitHub.  
+Instead, R programmers are liable to leverage a number of packages and work them into their data analysis pipelines.
 
-Interestingly, we are somewhat able to replicate this using only package
-code.  CRAN, has a policy that packages made available on the CRAN platform
-should be accompanied by a series of examples, tests, and/or vignettes
-showing off how the code is intended to be used.  Nominally unrelated, the
-{\tt genthat} project~\cite{issta18} traces this swath of example code to
-generate new examples which package designers may not have thought of.  As
-it happens, we can leverage {\tt genthat} to extract dynamic type
-information from running package test code, which presumably indicates how
-package code is intended to be used.
+Thankfully, we are somewhat able to make up for this lack of code.
+CRAN has a policy that published packages are to be accompanied by at least one vignette, and typically by tests and examples showing off how the code is intended to be used.  
+Nominally unrelated, the {\tt genthat} project~\cite{issta18} traces this example code to generate new unit tests which package designers may not have thought of.
+As it happens, we can leverage {\tt genthat} to extract dynamic type information from running package test code, which presumably indicates how package code is intended to be used.
 
@@
 
-We modify the {\tt genthat} tracer to capture the type information of
-arguments.  {\tt genthat} generates traces for each unique function
-invocation with the express purpose of re-invoking the function at a later
-time.  Instead of generating traces for these invocations, we can instead
-capture type information of the arguments and return values of functions.
-This is done by evaluating the arguments and return values of functions at
-the end of the function's scope in the relevant environment.  The sort of
-types and granularity of the type information that are generated can be
-specified in our tracer: Essentially, our tracer is parameterized over a
-user-specified type system.  For instance, if desired our tracer can dig
-into lists and data frames to get the types of list elements and fields.
-Our tracer produces trace results for each unique type signature
-encountered, which may then be used for data analysis.
+Our tracer is built on of the {\tt genthat}~\cite{issta18} tracer, which we modified to capture the type information of arguments.  
+{\tt genthat} generates traces for each unique function invocation with the express purpose of re-invoking the function at a later time, and instead of generating traces for these invocations we capture type information of the arguments and return values of functions.
+This is done by evaluating the arguments and return values of functions at the end of the function's scope in the relevant environment.  
+The sort of types and granularity of the type information that are generated can be specified in our tracer: 
+Essentially, our tracer is parameterized over a user-specified type system.  
+For instance, if desired our tracer can dig into lists and data frames to get the types of list elements and fields.
+Our tracer produces trace results for each unique type signature encountered, which may then be used for data analysis.
 
-An unfortunate consequence of R's most common usage pattern is the lack of
-"deployed" R code.  R is most often used for exploratory data analysis, and
-very little of this analysis code is made publicly available on e.g. GitHub.
-The {\tt genthat} angle is thus desirable: With the scarcity of non-library
-code, examples of package usage is about as representative as one can hope
-for of the sort of code people will write in their analysis pipelines.
+An unfortunate consequence of R's most common usage pattern is the lack of "deployed" R code. 
+R is most often used for exploratory data analysis, and very little of this analysis code is made publicly available on e.g. GitHub.
+The {\tt genthat} angle is thus desirable: 
+With the scarcity of non-library code, examples of package usage is about as representative as one can hope for of the sort of code people will write in their analysis pipelines.
 
@@
 Available through CRAN is a list of {\it reverse dependencies} of packages: 
-for some package {\tt p}, a list of all packages which require {\tt p} in
-some way.  Our tracer is able to leverage and trace this code as well.  When
-running a package, we also trace how all inherited functions are used, thus
-painting an even clearer picture of function usage.  This additional data is
-considered alongside similar data collected for all other packages, as well
-as the data obtained by tracing the package's examples.
+for some package {\tt p}, a list of all packages which require {\tt p} in some way.  
+Our tracer is able to leverage and trace this code as well. 
+When running a package, we also trace how all inherited functions are used, thus painting an even clearer picture of function usage.  
+This additional data is considered alongside similar data collected for all other packages, as well as the data obtained by tracing the package's examples.
 
@@
 %
+%
+%
+%
+\subsection{Granularity of Type Information}
+
+As we mentioned in Section~\ref{sec:rlang}, R has several notions of ``type'', ranging from the results of \code{typeof} to metadata on values obtainable via the \code{attributes} function.
+Individually, these paint an incomplete picture of the type of a value:
+For example, recall that \code{typeof(5) == typeof(c(5, 5)) == "double"}.
+
+To be as specific as possible, our analysis collects additional information to create a more specific type than the result of any of R's reflection functions.
+For instance, if the analysis encounters {\tt typeof(x) == \~ "double"}, it will look at the length of {\tt x} to determine if {\tt x} is a scalar or a vector, generating the annotation {\tt scalar/double} or {\tt vector/double} accordingly.
+One notable imperfection in this example is that unit-length vectors will appear as scalars, but if some function was called with unit-length vectors {\it as well as} larger ones, both scalar and vector types will appear, and this distinction can be later collapsed.
+Another example is that our analysis builds parametric list types by capturing the types of list elements.
+We also distinguish NAs ``in the wild'' with a {\tt raw\_NA} type.
+
+%
+%
+%
+%
 \subsection{Data Collection Pipeline}
 
-As we mentioned in the previous section, we modify the {\tt genthat} tracer
-to collect and output the type information for all function invocations.
+As we mentioned in Section~\ref{sec:genthat}, we modify the {\tt genthat} tracer to collect and output the type information for all function invocations.
 
-\AT{somewhere} Our analysis can be parameterized by a user-specified ``type
-system'', simply a map between existing dynamically observable type
-information and user-defined types.  If no type system is specified, the
-analysis gains as much information as it can.  For instance, if the analysis
-encounters {\tt typeof(x) == "double"}, it will look at the length of {\tt
-  x} to determine if {\tt x} is a scalar or a vector, generating the
-annotation {\tt scalar/double} or {\tt vector/double} accordingly.  Note
-that this is imperfect, as a vector of unit length would (perhaps
-incorrectly) show up as a {\tt scalar/X}, but any other invocation where the
-vector does not have unit length will generate the correct {\tt vector/X}
-type, and the distinction between scalars and vectors can be later collapsed
-if desired.  Note also that the type system described here is the
-\AT{SOMETHING} Type System, described in detail in Section~\AT{TODO}.
-
-The process for tracing one package {\tt p} is as follows:
+The high-level process for tracing one package {\tt p} is as follows:
 
@@
 	\item first, all functions in {\tt p} are instrumented with on-exit hooks.
-  These hooks call a number of R's available reflection functions on each of
-  the arguments, as well as the return type: The called functions are {\tt
-    typeof}, {\tt class}, and {\tt attributes}.  See Section~\ref{sec:rlang}
-  for details on these functions.
+  	These hooks call a number of R's available reflection functions on each of the arguments, as well as the return type: 
+  	The called functions are {\tt typeof}, {\tt class}, and {\tt attributes}.  
+  	See Section~\ref{sec:rlang} for details on these functions.
 
-\item then, depending on the specified type 
+  	\item next, runnable code is extracted and run from package tests, examples, and vignettes, and \AT{what about rev deps}.
+
+	\item each function invocation triggers the on-exit hook, and it is this on-exit code which collects the information.
+	Depending on the results of {\tt typeof}, {\tt class}, and {\tt attributes}, more analysis may be performed.
+	For example, if an argument is a list, type information is collected on list elements until a more accurate type (e.g., {\tt list<double>}) can be constructed.
+	Indeed, in a later section we do propose some possible type systems for R, and our analysis can be parameterized with these type systems to generate type information according to the types specified, and this will be discussed further in Section \AT{yap}.
+
+	\item all of the recorded type information makes up a function {\it trace}.
+	This trace is passed to a C++ function (for performance reasons) which hashes it and checks it against already seen trace hashes.
+	If the trace is determined to be new, the hash is saved and the trace is written to a file.
+
+	\item finally, after all test code has finished executing, we are left with a directory full of all of the unique type signatures that were observed while running the package code.
+
 \end{enumerate}

### 2019-03-20 paper/main.tex
corpus

@@
 
-\AT{Note: some stuff I wrote that might belong somewhere in here, if not, delete.}
-As mentioned, the syntax is highly unrestrictive to make this data analysis easy.
-All functions in R are vectorized, and consequently operators and functions are multi-purpose.
-For instance, the syntax for adding numbers, vectors, and matrices uses the same "+" operator.
+\AT{Note: some stuff I wrote that might belong somewhere in here, if not,
+  delete.}  As mentioned, the syntax is highly unrestrictive to make this
+data analysis easy.  All functions in R are vectorized, and consequently
+operators and functions are multi-purpose.  For instance, the syntax for
+adding numbers, vectors, and matrices uses the same "+" operator.
 
@@
 %% TODO : make the types all bold.
-\AT{Some examples of use of \code{typeof}:}
-To build a sense of how these types arise in R code, consider the following:
-\AT{We could think of another way to do this also.}
+\AT{Some examples of use of \code{typeof}:} To build a sense of how these
+types arise in R code, consider the following: \AT{We could think of another
+  way to do this also.}
 
 \begin{itemize}
-
-	\item \code{typeof(5) == typeof(c(5)) == typeof(c(5, 5)) == "double"}.
-	As we mentioned, R is vectorized, so even apparent scalars are vectors.
+\item \code{typeof(5) == typeof(c(5)) == typeof(c(5, 5)) == "double"}.  As
+  we mentioned, R is vectorized, so even apparent scalars are vectors.
   \AT{Just for fun,} note that \code{5[1] == 5} in R.
-
 \item to actually get an integer, you need \code{typeof(5L) == "integer"}.
-	Doubles are by far the most common numeric type in R \AT{I can get data for this}, and you can even use them to index lists and vectors: \code{list(1, 2)\[\[1.9\]\] == 1}.
-	Note the double brackets for list indexing.
-
-	\item NAs are common in data-centric languages (e.g. MATLAB, \AT{others?}), and R is no exception.
-	As for their type, we see that \code{typeof(NA) == "logical"}, but that isn't to say that NA only has type logical.
-	Consider \code{typeof(c(1, NA, 3)\[2\])}: NA actually inhabits every type in R, \AT{details}.
-
+  Doubles are by far the most common numeric type in R \AT{I can get data
+    for this}, and you can even use them to index lists and vectors:
+  \code{list(1, 2)\[\[1.9\]\] == 1}.  Note the double brackets for list
+  indexing.
+\item NAs are common in data-centric languages (e.g. MATLAB, \AT{others?}),
+  and R is no exception.  As for their type, we see that \code{typeof(NA) ==
+    "logical"}, but that isn't to say that NA only has type logical.
+  Consider \code{typeof(c(1, NA, 3)\[2\])}: NA actually inhabits every type
+  in R, \AT{details}.
 \item \AT{other examples?}
-
 \end{itemize}
 
-Another form of type information relates to metadata on values, and R allows programmers to add said metadata in the form of {\it attributes}.
-The functionality to do so in R is through the \code{attributes(x)} and \code{attr(x, "name")}  functions.
-\code{attributes}
+Another form of type information relates to metadata on values, and R allows
+programmers to add said metadata in the form of {\it attributes}.  The
+functionality to do so in R is through the \code{attributes(x)} and
+\code{attr(x, "name")} functions.  \code{attributes}
 
@@
 
-%
-% Corpus Section
-%
-% > talk about CRAN
-% > what is a package 
-% > look at genthat paper for inspiration
-% > communicate the size of the adventure
-% > convince people that this is real
-% > lines of C and R code per project? big graph could be nice
-% > give people a sense of our data set
-
 \section{The Corpus}
 
-In this section, we will discuss the R language ecosystem with a focus on our dataset.
+In this section, we present our dataset. R is designed to accommodate the
+workflow of data analysts.  Typical workflows go from data import, to data
+cleaning, the steps of modeling, transformation and visualization. Often
+this code is stored in notebooks which combine code and its output.
+Unfortunately notebooks are typically not available on GitHub, for this our
+analysis focuses on R packages, these bundle reusable units of R code with
+documentation, sample data, use-cases and tests.
 
-The R environment is designed to accommodate the workflow of the data analyst, a workflow which tightly integrates the analyst and the data.
-As an example example, for a task related to data visualization, an analyst will need to: (1) import a data set, (2) sanitize or otherwise transform the data, (3) plot the data, (4) repeat as necessary.
-Similarly, for a task related to testing hypotheses or otherwise validating conclusions: (1) import a data set, (2) sanitize or otherwise transform the data, (3) call some functions to perform necessary computations for the test, (4) repeat as necessary.
-The pattern of import, transform, do something, repeat is referred to as a {\it pipeline} in the R community.
-Indeed, RStudio, one of the most popular IDEs for developing R, is designed to support this iterative behaviour.
-In it, users create {\it notebooks} (akin to Jupyter notebooks for those familiar) wherein they may write text alongside their computation pipeline, and these notebooks expose the code written by the user, altogether facilitating collaboration with other users.
+We focus on packages hosted on the \emph{Comprehensive R Archive Network} or
+CRAN.  With over 13,000 packages, CRAN is the largest repository of software
+written in R. Unlike sites like GitHub, CRAN is a curated collection of
+packages. A package is only accepted to CRAN if it abides by a number of
+well-formdness rules.  Most relevant for our purposes, packages must have
+data, examples, vignettes and tests. All code associated with the package
+must succesfully run. This means that each of the package in CRAN comes with
+up to a dozen executable scripts that exercise some of its functionality.
 
-By and large, the most extensive repository of R code is CRAN, the Comprehensive R Archive Network.
-After installing a package, a simple call to \code{library(pkg\_name)} exposes all of \code{pkg\_name}'s exported functions for use in the rest of the script.
-CRAN sports over 13,00 packages, and is an actively maintained and curated environment, in that there are a number of requirements for a package to be accepted for publication, particularly with regards to proper package documentation.
-One such requirement is the presence of at least one vignette, a documented walk-through of some of the package's functions.
-\AT{Do we want an example of a vignette? We could put one.}
-Besides vignettes, package authors are encouraged to write tests for their code as well as examples showing off basic usage and functionality.
+Our corpus is a subset of CRAN, consisting of \PACKAGES packages, accounting
+for some \PERCENTCRAN of all packages.  These packages have a total of \RLOC
+lines of R code and ??? lines of C code. For each package, we extracted all
+executable code snippets from documentation, vignettes and tests and ran
+them independently recording all call to R functions.  The rationale for
+this particular subset of CRAN was pragmatic, our infrastructure adds some
+overhead to the execution of scripts, the packages we selected we all those
+that could be run in one hour or less.
 
-One goal of our analysis is to build a sense of how programmers use types in R, and unfortunately there isn't much non-library code available for our analysis to peruse.
-Most R scripts are not built to be exported, and are part of an iterative and interactive process of data analysis, and unlike other programming languages, R has a very small footprint on GitHub, and most of the R code available there is referenced through CRAN \AT{maybe we can get numbers here?
-or just take out the sentence}.
-There are a number of potential explanations for this, including the proprietary nature of certain data sets (e.g., governments working with sensitive census data, or medical researchers working with private patient data).
+Figure~\ref{most} shows the ten most downloaded CRAN packages
+%% TODO make sure we are analyzing them, and cut the list to ten
+and for each package it shows the number of lines of R, the number of lines
+of C/C++, and the number of executable scripts that could be run.
 
-The lack of non-package code is somewhat mitigated by the good culture of packages published on CRAN, in that they typically provide tests and examples of intended package use.
-We modify the {\tt genthat} tracer, discussed in more detail in the next section, which essentially traces these tests, examples, and vignettes to generate new package tests.
-\AT{Not sure how to word the second half of this paragraph.}
-In focusing on these exemplary snippets, we should capture the intent of package designers, which gives us hope that our proposed type annotations would be adopted.
+Figure~\ref{allcloc} shows the size of the code in each package in C and R.
 
-%
-%
-%
-%
-\subsection{Our Test Suite}
+Figure~\ref{recorded} gives the number of function signatures that we
+recorded while executing the scripts of each package.
 
-We ran our analysis over a subset of CRAN, analyzing \PACKAGES packages, accounting for some \PERCENTCRAN of all packages available on CRAN.
-These packages have a total of \RLOC lines of code, and were written by \PROGRAMMERS programmers over a period of \YEARS years.
-\AT{Talk about timeout?}  Due to the nature of some packages (e.g., packages implementing a computationally-intensive model with many tests), we ran our analysis with a 1 hour timeout.
-Our tracer also traces reverse-dependencies: if some package utilizes some other package, both are instrumented and traced.
-This serves to further mitigate the lack of non-library R code.
-
-To get a better sense of the corpus we're dealing with, we will highlight a few packages that we have analyzed:
-
-\begin{itemize}
-
-	\item \textbf{\texttt{ggplot2}} is an important package for building graphics and plotting data in R.
-	{\tt ggplot2} is part of the tidyverse, which bills itself as an ``opinionated collection of R packages designed for data science''.
-	Essentially, packages in the tidyverse use a common grammar and are all designed to address the main issues faced in data analysis.
-	{\tt ggplot2}'s role in the tidyverse is to make creating elegant plots easy: in it, graphics are created declaratively in a manner based on The Grammar of Graphics \AT{cite}.
-
-	\item \textbf{\texttt{dplyr}} is another integral package to the tidyverse.
-	In it is defined a convenient piece of R syntax: the pipe operator {\tt \%>\%}.
-	If using {\tt dplyr}, one can write \code{f(x) \%>\% g()}, which is equivalent to \code{g(f(x))}.
-	{\tt dplyr} also defines a number of highly useful functions for manipulating data frames, and considered alongside the pipe operator form the underpinning of the tidyverse data pipeline philosophy.
-
-	\item \textbf{\texttt{zoo}} is a popular R package for dealing with time series, themselves a common occurrence in statistics.
-	One of the main functions in this package is the eponymous {\tt zoo} function, which takes in a numeric vector or list along with some parameters, and produces a ``zoo'' object for use in other package functions.
-
-	\item \textbf{\texttt{spatstat}} is the largest package in our corpus in terms of lines of R code.
-	This package is designed for spatial statistics, focusing on the analysis of spatial point patterns.
-	It is, in some sense, the definitive spatial statistics package in R.
-
-	\item \textbf{\texttt{multinet}} is the largest package in our corpus in terms of lines of C code.
-	{\tt multinet} is a package designed to store and analyze multilayer social networks.
-
-\end{itemize}
-
-Most of the packages we looked at have between 100 and 1000 lines of R code.
-
+% https://www.r-pkg.org/downloaded
+\begin{figure}[!th]
+  {\footnotesize\begin{tabular}{@{}r||l|r|r|r@{}}
+\hline
+\bf Package & \bf Description & \bf R LOC &\bf C LOC &\bf Scripts\\\hline
+\tt Rcpp  & Seamless C++ integration & 10K & 3K & 12 \\
+\tt rlang & Functions for 'Tidyverse'&&& \\
+\tt glue  & Interpreted string literals&&& \\
+\tt tibble & Simple data frames&& &\\
+\tt stringi &  String processing &&& \\
+\tt ggplot2 & Data visualisations&& &\\
+\tt dplyr  &  Data manipulation&&& \\
+\tt pillar & Formatting for columns&&& \\
+\tt R6 & Classes w. ref. semantics&&& \\
+\tt stringr & String operations&&& \\
+\tt data.table & Extension of `data.frame`&&& \\
+\tt yaml & Convert R data to YAML&&& \\
+\tt cli &  Command line interfaces&&& \\
+\tt purrr & Functional programming&&& \\\hline
+\end{tabular}}
+\caption{Most downloaded packages.}\label{most}
+\end{figure}
 %

### 2019-03-20 paper/main.tex
method, pipeline, before class

@@
 \newcommand{\RLOC}{8,996,572\xspace}
+\newcommand{\CLOC}{5,100,719\xspace}
 \newcommand{\YEARS}{20\xspace}
@@
 \section{The Corpus}
+\label{sec:corpus}
 
@@
 packages. A package is only accepted to CRAN if it abides by a number of
-well-formdness rules.  Most relevant for our purposes, packages must have
+well-formedness rules.  Most relevant for our purposes, packages must have
 data, examples, vignettes and tests. All code associated with the package
-must succesfully run. This means that each of the package in CRAN comes with
+must successfully run. This means that each of the package in CRAN comes with
 up to a dozen executable scripts that exercise some of its functionality.
@@
 for some \PERCENTCRAN of all packages.  These packages have a total of \RLOC
-lines of R code and ??? lines of C code. For each package, we extracted all
+lines of R code and \CLOC lines of C code. For each package, we extracted all
 executable code snippets from documentation, vignettes and tests and ran
@@
 Figure~\ref{most} shows the ten most downloaded CRAN packages
-%% TODO make sure we are analyzing them, and cut the list to ten
 and for each package it shows the number of lines of R, the number of lines
@@
 \tt stringr & String operations&&& \\
-\tt data.table & Extension of `data.frame`&&& \\
-\tt yaml & Convert R data to YAML&&& \\
-\tt cli &  Command line interfaces&&& \\
-\tt purrr & Functional programming&&& \\\hline
+%\tt data.table & Extension of `data.frame`&&& \\
+%\tt yaml & Convert R data to YAML&&& \\
+%\tt cli &  Command line interfaces&&& \\
+%\tt purrr & Functional programming&&& \\\hline
 \end{tabular}}
-\caption{Most downloaded packages.}\label{most}
+\caption{10 Most Downloaded Packages.}\label{most}
 \end{figure}
+
 %
@@
 To be as specific as possible, our analysis collects additional information to create a more specific type than the result of any of R's reflection functions.
-For instance, if the analysis encounters {\tt typeof(x) == \~ "double"}, it will look at the length of {\tt x} to determine if {\tt x} is a scalar or a vector, generating the annotation {\tt scalar/double} or {\tt vector/double} accordingly.
+For instance, if the analysis encounters {\tt typeof(x) == "double"}, it will look at the length of {\tt x} to determine if {\tt x} is a scalar or a vector, generating the annotation {\tt scalar/double} or {\tt vector/double} accordingly.
 One notable imperfection in this example is that unit-length vectors will appear as scalars, but if some function was called with unit-length vectors {\it as well as} larger ones, both scalar and vector types will appear, and this distinction can be later collapsed.
@@
 
-The high-level process for tracing one package {\tt p} is as follows:
+The high-level process for tracing one package is as follows:
 
 \begin{enumerate}
-	\item first, all functions in {\tt p} are instrumented with on-exit hooks.
+	\item first, all functions in the package are instrumented with on-exit hooks.
   	These hooks call a number of R's available reflection functions on each of the arguments, as well as the return type: 
   	The called functions are {\tt typeof}, {\tt class}, and {\tt attributes}.  
-  	See Section~\ref{sec:rlang} for details on these functions.
+  	Recording is done on function exit to avoid forcing promises before they are needed (as these promises may cause side effects).
 
@@
 
-Then, we aggregate the information in all traces for a particular function.
-If there was only one function trace, we discard the result as the function
-is trivially monomorphic.  We can see which arguments had which types over
-how many traces, and from here we can build a \textit{signature} for each
-function argument, and consider them together to create a function
-signature.  A \textbf{polymorphic argument} is one which has been inhabited
-by values of at least two different types, and a \textbf{polymorphic
-  function} is a function with at least one polymorphic argument or a
-polymorphic return.
+For each package function, we then consolidate all observed signatures: this gives us a {\it function signature}.
+\AT{Example?}
+In the function signature, each function argument is accompanied by a list of types, attributes, and classes that have inhabited it.
 
-We repeat the above for each package on the Comprehensive R Archive Network
-(CRAN), the premier source for R libraries (called packages), and the
-Bioconductor package repository.  Now, depending on the data point we're
-after, our methods from here differ slightly.
+We will need a notion of polymorphism when undertaking this analysis.
+As R is unityped, and the only notion of polymorphism in the language is its use of dynamic dispatch on the class of an argument.
+For our purposes, we will define 3 simple types of polymorphism, closely connected with the reflection functions we use.
+We start by defining what it means for an {\it argument} to be polymorphic.
+\AT{Could be valuable to have an example here.}
 
 \begin{itemize}
-    \item for \textbf{counting signatures}: for each signature, we count how
-      many functions or arguments have the requisite signature;
-    \item for \textbf{finding the most common signatures}: we enumerate as
-      before, and sort the signatures based on how often they appear;
+
+	\item an argument is {\it polymorphic in type} if it has been called with at least two distinct ``types'' according to the \code{typeof} function and our extra analysis.
+
+	\item similarly, an argument is {\it polymorphic in class} if it has been called with at least two distinct classes according to the \code{class} function.
+
+	\item attributes are a bit more tricky.
+	We can't simply count the number of attributes that have inhabited an argument, since values can contain arbitrarily many values (and having several attributes isn't polymorphism---think records).
+	Instead, we define the {\it attribute pattern} of a value to be the set of all attribute name and type pairs in the attribute list of a value (obtained via \code{attributes}).
+	And so an argument is {\it polymorphic in attribute} if it has been called with at least two distinct attribute patterns.
+
 \end{itemize}
 
-%
-%
-\subsection{Attributes}\label{sec:method:attributes}
+There is an important distinction to be made between the result of \code{typeof} and that of \code{class} and \code{attributes}:
+In R, the \code{typeof} a value cannot change, and this is not the case with class and attributes.
+As mentioned in \AT{an earlier section}, the class of values can be easily redefined by assigning to the value's class attribute.
+Similarly, attributes can be dynamically added, removed, or changed.
 
-Types (in the sense of the result of a call to {\tt typeof}) are easy to
-analyze as values can only have one type.  Attributes, however, are a
-different story, as values can have an arbitrary number of attributes.
-Essentially, we're not interesting by function arguments which have had
-multiple attributes, we're interested in function arguments which have had
-multiple attribute \textit{patterns}. \AT{is this defined earlier?}
-Arguments are said to be polymorphic in attribute if they have been
-inhabited with values which have had different attribute patterns.  \AT{We
-  have not investigated whether functions which are polymorphic in attribute
-  use attributes extensively inside the function code.}
+This begs the question: what sort of classes and attribute patterns occur naturally, before possible user tampering?
+In terms of classes, the discussion can be found in Section~\ref{sec:rlang}.
+As for attributes:
 
-\AT{Paragraph justifying our construction of an attribute pattern.
-Either just name, or (name, type).}
+\begin{itemize}
 
-\AT{Paragraph about so-called ``naturally-occurring'' attribute patterns.}
+	\item data frames have three attributes: {\tt names} for column names, {\tt row.names} for row names, and {\tt class} (which is instantiated to {\tt "data.frame"} upon creation).
+
+	\item matrices have at least a {\tt dim} attribute with matrix dimensions, and optionally a {\tt dimnames} attribute for dimension names.
+
+	\item lists and vectors may have a {\tt names} attribute, which assigns names to locations.
+
+	\item to ease dealing with large sorted sets of non-numeric values, R offers the \code{factor} and \code{levels} functions.
+	\code{factor(x, levels=someOrder)} adds a {\tt "levels"} attribute to \code{x}, which specifies an ordering for \code{x} according to the \code{sortedOrder} parameter.
+	\code{factor} also changes the class of the parameter being factored to \code{"factor"}.
+
+\end{itemize}
 

### 2019-03-20 paper/main.tex
minor

@@
 
-\AT{Note: some stuff I wrote that might belong somewhere in here, if not,
-  delete.}  As mentioned, the syntax is highly unrestrictive to make this
-data analysis easy.  All functions in R are vectorized, and consequently
-operators and functions are multi-purpose.  For instance, the syntax for
-adding numbers, vectors, and matrices uses the same "+" operator.
-
-\AT{Also:}
-The R language has a very easy-to-use FFI to C, and even to C++ through another R package, {\tt Rcpp}.
-
-\AT{End note.}
-
 Over the last decade, the R Project has become a key tool for implementing
@@
 
-\AT{I wrote a paragraph in the next section which seems to belong here, maybe work this in?}
-Besides these primitive data types, R has an out-of-the-box data type called the data frame for storing data, and this data type is essential to a large part of R's functionality.
-A data frame is a two-dimensional array where each row represents an observation, with columns indicating the values of some observation.
-For example, our own analysis is written in R, and one of our data frames has a row for each package analyzed, with columns for values ranging from number of polymorphic functions, number of monomorphic arguments, unique attribute patterns, etc.
-\AT{Put a figure?}
-Much of R's base functionality, and indeed much of the functionality provided by packages, operate on data frames.
+R has a number of features that are not crucial to the present
+discussion. We will mention some of them here for completeness.  In R, data
+structures are reference counted and have copy-on-write semantics, thus the
+assignment \code{x[12]<-3} results in an update to a copy of \code{x} unless
+the reference count on that object is 1.  This semantics gives R a
+functional flavor while allowing updating in place within loops (the first
+update copies, subsequent updates are performed on the copy). Arguments to
+functions are evaluated only when needed, they are bundled in so-called
+promises which package the original expression (as an AST), its environment
+as well as the result of evaluating the expression. Promises can be
+leveraged for meta-programming as it is possible to retrieve the text of a
+promise and evaluate that in a different environment.
 
@@
 be.
+
+\renewcommand{\k}[1]{{\tt #1}\xspace}
+
+R has one builtin notion of type that can be queried by the \k{typeof}
+function. Over the years, programmers have found the need for a richer type
+structure and have added attributes. The best way to think of attributes is
+as an optional map from name to values that can be attached to any object.
+Attributes are used to encode various type structures. They can be queried
+with functions such as \k{attributes} and \k{class}.
+
+\begin{figure}[!ht]
+\begin{tabular}{l|l}\hline
+\multicolumn{2}{l}{\bf Vectorized data types:}  \\\hline
+\k{logical}  & vector of boolean values\\
+\k{integer}   & vector of 32 bit integer values\\
+\k{double} & vector of 64 bit floating point values\\
+\k{complex} & vector of complex values\\
+\k{character} & vector of strings values\\
+\k{raw} & vector of bytes\\
+\k{list} & vector of values of any type\\\hline
+\multicolumn{2}{l}{\bf Scalar data types:}\\\hline
+\k{NULL}  &  singleton null value\\
+\k{closure} & a function with its environment\\
+\k{environment} & a mapping from symbol to value \\\hline
+\multicolumn{2}{l}{\bf Implementation data types:}\\\hline
+&\k{special},
+\k{builtin},
+\k{symbol},
+\k{pairlist},
+\k{promise},
+\k{language},
+\k{char},
+\k{...}\\
+&\k{any},
+\k{expression},
+\k{externalprt},
+\k{bytecode},
+\k{weakref}\\\hline
+\end{tabular}\caption{Builtin Types}\label{types}\end{figure}
 
@@
 
-\section{The Corpus}
-\label{sec:corpus}
+Besides these primitive data types, R has an out-of-the-box data type called
+the data frame for storing data, and this data type is essential to a large
+part of R's functionality.  A data frame is a two-dimensional array where
+each row represents an observation, with columns indicating the values of
+some observation.  For example, our own analysis is written in R, and one of
+our data frames has a row for each package analyzed, with columns for values
+ranging from number of polymorphic functions, number of monomorphic
+arguments, unique attribute patterns, etc.  \AT{Put a figure?}  Much of R's
+base functionality, and indeed much of the functionality provided by
+packages, operate on data frames.
 
-In this section, we present our dataset. R is designed to accommodate the
-workflow of data analysts.  Typical workflows go from data import, to data
-cleaning, the steps of modeling, transformation and visualization. Often
-this code is stored in notebooks which combine code and its output.
-Unfortunately notebooks are typically not available on GitHub, for this our
-analysis focuses on R packages, these bundle reusable units of R code with
-documentation, sample data, use-cases and tests.
+\section{The Corpus}\label{sec:corpus}
+
+In this section, we present our dataset. R is designed to accommodate data
+analysts.  Typical workflows go from data import, to data cleaning, followed
+by steps of modeling, transformation and visualization. Often the code that
+implement these analysis pipelines together with their results resides in
+notebooks. Unfortunately notebooks are typically not shared in public
+repositories such as GitHub, and if they were, the data required to run them
+would surely not be shared. For this reason our analysis focuses on
+packages; these bundle reusable units of R code with documentation, sample
+data, use-cases and tests.
 
@@
 CRAN.  With over 13,000 packages, CRAN is the largest repository of software
-written in R. Unlike sites like GitHub, CRAN is a curated collection of
-packages. A package is only accepted to CRAN if it abides by a number of
-well-formedness rules.  Most relevant for our purposes, packages must have
-data, examples, vignettes and tests. All code associated with the package
-must successfully run. This means that each of the package in CRAN comes with
-up to a dozen executable scripts that exercise some of its functionality.
+written in R. It is experiencing sustained growth with an average of size
+new packages a day~\cite{LIgges2017}.  Unlike sites like GitHub, CRAN is a
+curated collection of packages. A package is only accepted to CRAN if it
+abides by a number of well-formedness rules.  Most relevant for our
+purposes, packages must have data, examples, vignettes and tests. All of
+which must successfully run. From our perspective this means that each of
+the package in CRAN comes with up to a dozen executable scripts that
+exercise some of its functionality.
 
@@
 for some \PERCENTCRAN of all packages.  These packages have a total of \RLOC
-lines of R code and \CLOC lines of C code. For each package, we extracted all
-executable code snippets from documentation, vignettes and tests and ran
-them independently recording all call to R functions.  The rationale for
-this particular subset of CRAN was pragmatic, our infrastructure adds some
-overhead to the execution of scripts, the packages we selected we all those
-that could be run in one hour or less.
+lines of R code and \CLOC lines of C code. For each package, we extracted
+all executable code snippets from documentation, vignettes and tests and ran
+them independently recording all call to R functions.  It is noteworthy that
+in order to run the scripts in one package, it is often necessary to load a
+number of other packages.  In~\cite{issta18}, the authors estimated code
+coverage to be around 60\% when including reverse dependencies.  The
+rationale for our choice of packages was pragmatic, our infrastructure adds
+overhead to the execution of scripts, we eliminated any package that would
+take more than one hour to run. We expect to extend our coverage of CRAN for
+the next version of this paper.
 
-Figure~\ref{most} shows the ten most downloaded CRAN packages
-and for each package it shows the number of lines of R, the number of lines
-of C/C++, and the number of executable scripts that could be run.
-
-Figure~\ref{allcloc} shows the size of the code in each package in C and R.
-
-Figure~\ref{recorded} gives the number of function signatures that we
-recorded while executing the scripts of each package.
+Figure~\ref{most} shows the ten most downloaded CRAN packages.  For each
+one, we list how many lines of R and C/C++ the packages contains.
+Typically, C/C++ code is used to implement performance critical portions of
+the code. We show the number of scripts that could be extracted from the
+package. Each script corresponds to either one use-case or a set of unit
+tests.  Figure~\ref{allcloc} shows the size of the code across all the
+packages we have used.  Lastly, Figure~\ref{recorded} gives the number of
+function signatures that we recorded while executing the scripts of each
+package.
 
 % https://www.r-pkg.org/downloaded
-\begin{figure}[!th]
-  {\footnotesize\begin{tabular}{@{}r||l|r|r|r@{}}
-\hline
+\begin{figure}[!th]{\footnotesize\begin{tabular}{@{}r||l|r|r|r@{}}\hline
 \bf Package & \bf Description & \bf R LOC &\bf C LOC &\bf Scripts\\\hline
@@
 \tt stringr & String operations&&& \\
-%\tt data.table & Extension of `data.frame`&&& \\
-%\tt yaml & Convert R data to YAML&&& \\
-%\tt cli &  Command line interfaces&&& \\
-%\tt purrr & Functional programming&&& \\\hline
-\end{tabular}}
-\caption{10 Most Downloaded Packages.}\label{most}
+\end{tabular}}\caption{10 Most Downloaded Packages.}\label{most}
 \end{figure}

### 2019-03-20 paper/main.tex
results

@@
 
-%
-%
-%
-%
-\subsection{{\tt genthat} Tracer Quirks}
-
-First and foremost, {\tt genthat} captures arguments only on function exit,
-so as to not force promises during function execution.  \AT{Uh oh:} One
-clear limitation here is that if an argument \textit{gains} attributes over
-the functione execution, then they will appear in our signatures.  As R is
-side-effecting, this is an entirely valid practice (i.e., passing objects to
-functions to modify them), though \AT{it's unclear if people actually do
-  this}.
-
-\AT{Maybe talk about the quirks of genthat in here?
-For example, how it deals with default arguments, and stuff like that.}
-
-%
-%
-%
-%
-\subsection{Goals}
-
-Ultimately, our goal is to collect data and report on patterns which emerge
-organically.  That said, the comprehensive picture of type usage we are
-developing can also be used to inform the design of a set of \textit{type
-  annotations} for the language, and this goal informs our analysis.  As we
-collect our data, we keep in mind the idea that the data we produce should
-clearly suggest what sorts of type annotations would reflect language usage
-patterns.
+These naturally-occurring attribute patterns and classes can be separately accounted for to highlight user-defined patterns and behaviours.
+This will be made clear in Section~\ref{sec:results}.
 
@@
 \section{Results}
+\label{sec:results}
 
-\AT{PSA:} There are a few ways we could do this section.
-One way would be to go through our data in a sort of linear fashion, akin to how the main notebook is set up.
-We can explore each of our type systems one at a time, and present them (thus breaking the section up by type system).
-Another option is to just break down by-result, so have a section for the counts, for the top polymorphism...
-I like the first option better.
+\AT{Meta-Comment:} 
+My plan for this section is to linearly go through the data, and show logical transitions between various type systems.
+I'll start with the finest-grained one.
 
 \AT{Draft}
-In this section, we will present the results of our analysis. 
+Having built an understanding of our data collection and analysis pipeline, we will turn our attention to the results of our analysis.
 We approach the data with the purpose of informing the design of a static type system for R.
-We propose to begin with a very basic type system, discussed next.
+To that end, we propose to begin with a very basic type system, the {\it Fine-Grained Type System} (FGTS).
 
-\subsection{Base R Type System}
+The logical ``baseline'' for any design of a static type system for R is promotion of the language's dynamic types (in the sense of \code{typeof}) to static types, but there are some clear limitations to this approach.
+For example, \code{typeof} makes no distinction between vectors and scalars, as indeed there is no such distinction in R.
+But that doesn't mean that there {\it shouldn't} be a distinction, as an all-new runtime environment design for R might benefit from such a distinction.
 
-The {\it Base R Type System} (BRTS) ...
+To address these limitations, we augment the logical naive type system with more information which was alluded to in the discussion of our data collection methodology.
+We will consider unit-length vectors to be scalars, collect list member type information to build an accurate list type, and promote NAs to their own type.
+In addition, we collect attribute and class information to distinguish a type for data frames and matrices, as the \code{typeof} these is simply {\tt list} in R.
+This type system is referred to as the Fine-Grained Type System, or FGTS.
+An example of a signature in this type system can be found in Figure~\AT{figggg}.
+
+The first question we will address is: {\it how polymorphic is R?}
+To answer that, we will first look at the number of monomorphic arguments across all of the packages we analyzed.
+The data can be found in Figure~\AT{fiGgGgGgGGg}.
+
+\AT{We should figure out the pipeline to get the notebook output into here.}
+
+\AT{I'm fairly sure the numbers aren't going to be insanely different, so I'll just manually write them for now.}
+
+We see that roughly 10\% of all argument positions in R are polymorphic.
+The numbers for attribute and class polymorphism are included for completeness, but the important number is that of type polymorphism (as our method of constructing types sometimes factors in class and attribute).
+Specifically for classes, often the class and \code{typeof} a value will line up (refer to Section~\ref{sec:rlang}).
+
+As for functions, roughly 25\% are polymorphic, or put differently 25\% have at least one polymorphic argument or a polymorphic return.
+The discrepancy between this and the polymorphicity of arguments is explained by the tendency of some functions, particularly those dealing with models, to have {\it many} arguments.
+
+These numbers alone are insufficient, so we ask: {\it of the polymorphic arguments, what are the most common patterns?}
+The answer to that question can be found in Figure~\AT{eyyy}.
+
+By and large, the most common pattern is {\tt scalar/X} and {\tt vector/X}.
+One way to account for this would be to collapse scalar and vector types into a vector type:
+If a function argument is observed to be both a unit-length and non-unit-length vector, then it's reasonable to assume that the argument is intended to be a vector.
+
+One other interesting question is: {\it how many arguments are strictly scalar?}
+As it happens, over 40\% of monomorphic arguments with primitive vectorized types are always scalar (so, 40\% of {\it monomorphic} occurrences of {\tt scalar/X} and {\tt vector/X} are scalar).
+This is important as it shows that there is value in a scalar annotation, which could be used to communicate to an R runtime that vectorization of a value is not necessary, which may in turn ead to a performance improvement.
+
+We will collapse the distinction between vectors and scalars in a moment, but imagine we did so and refer back to Figure~\AT{eyyy}.
+Another major source of polymorphism is the {\tt double} and {\tt integer} distinction.
+This distinction is mainly just relevant to the implementation, as integers and doubles are stored and dealt with differently in the runtime.
+For a programmer, this distinction is meaningless, as both types can be used interchangeably in every situation.
+We propose to also collapse this distinction, instead considering doubles and integers as part of a real number type, {\tt real}.
+
+\AT{TODO: do vector/scalar collapse and real collapse and get numbers.}
 
@@
 
-\AT{may not be the correct place for this. i think i accidentally deleted it.}
-
-R is a unityped language, and as such is trivially monomorphic, and this notion of monomorphism is far from useful.
-We are interested in expanding the notion of ``type'' in R, and thankfully the language has some tools on offer.
-\AT{this is discussed later on}
+\AT{I'll pull from this section when appropriate, as the discussion of the type systems becomes relevant.}
 

### 2019-03-21 paper/main.tex
minor
types

@@
 \documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-\usepackage{booktabs,listings,xspace}
+\usepackage{booktabs,listings,xspace,wrapfig}
 \lstset{language=R}
@@
 
+\newpage  %%Leave here
+
 \section{The R Programming Language}\label{sec:rlang}
@@
 
-\begin{figure}{\small
+\begin{figure}[!hb]{\small
 \begin{lstlisting}[style=R]
@@
 
-\begin{figure}[!ht]
-\begin{tabular}{l|l}\hline
+\begin{wrapfigure}{r}{6.1cm}
+\footnotesize\begin{tabular}{l|l@{}}\hline
 \multicolumn{2}{l}{\bf Vectorized data types:}  \\\hline
@@
 \k{integer}   & vector of 32 bit integer values\\
-\k{double} & vector of 64 bit floating point values\\
+\k{double} & vector of 64 bit floating points\\
 \k{complex} & vector of complex values\\
@@
 \k{NULL}  &  singleton null value\\
+\k{S4}    &  instance of a S4 class \\
 \k{closure} & a function with its environment\\
@@
 \multicolumn{2}{l}{\bf Implementation data types:}\\\hline
-&\k{special},
+\multicolumn{2}{l}{\k{special},
 \k{builtin},
@@
 \k{pairlist},
-\k{promise},
+\k{promise}}\\
+\multicolumn{2}{l}{
 \k{language},
 \k{char},
-\k{...}\\
-&\k{any},
+\k{...}, 
+\k{any},
 \k{expression},
+}\\
+\multicolumn{2}{l}{
 \k{externalprt},
 \k{bytecode},
-\k{weakref}\\\hline
-\end{tabular}\caption{Builtin Types}\label{types}\end{figure}
+\k{weakref}}\\\hline
+\end{tabular}\caption{Builtin Types}\label{types}\end{wrapfigure}
 
-The first distinction that can be made is based on the value of the
-\code{typeof} function.  This function, given a value, will return one of
-the following types: \code{logical} (vector of booleans), \code{integer}
-(vector of integers), \code{double} (vector of floating points),
-\code{complex} (vector of complex numbers), \code{character} (vector of
-character strings), \code{raw} (vector of bytes) and \code{list}
-(polymorphic vector), \code{NULL} (vector of NULL),
-\code{closure} (function), \code{environment} (naming environment),
-\code{S4} (S4 object) and others that are used within the implementation of
-R (\code{special}, \code{builtin}, \code{symbol}, \code{pairlist},
-\code{promise}, \code{language}, \code{char}, \code{...}, \code{any},
-\code{expression}, \code{externalptr}, \code{bytecode} and \code{weakref}).
+Figure~\ref{types} lists all of the builtin types that are provided by the
+language. They are the possible return values of \k{typeof}. There is no
+intrinsic notion of subtyping in R. But, in many context a \k{logical} will
+convert to \k{integer}, and an \k{integer} will convert to \k{double}.  Some
+off conversion can occur in corner cases, such as \k{1<"2"} holds and
+\k{c(1,2)[1.6]} returns the first element of the vector, as the double is
+converted to an integer. R does not distinguish between scalars and vectors
+(they are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
+  typeof(c(5,5))} \code{ == "double"}. Finally all vectorized data types have a
+distinguished missing value denoted by \code{NA}. The default type of
+\code{NA} is \k{logical}. We can see that \code{typeof(NA)=="logical"}, but
+NA inhabits every type: \code{typeof(c(1,NA)[2])=="double"}.
 
-%% TODO : make the types all bold.
-\AT{Some examples of use of \code{typeof}:} To build a sense of how these
-types arise in R code, consider the following: \AT{We could think of another
-  way to do this also.}
+With one exception all vectorized data types are monomorphic, the exception
+is the \k{list} type which can hold values of any other type including
+\k{list}. For all monomorphic data types, attempting to store a value of a
+different type will cause a conversion. Either the value is converted to the
+type of vector, or the vector is converted to the type of the value.
 
-\begin{itemize}
-\item \code{typeof(5) == typeof(c(5)) == typeof(c(5, 5)) == "double"}.  As
-  we mentioned, R is vectorized, so even apparent scalars are vectors.
-  \AT{Just for fun,} note that \code{5[1] == 5} in R.
-\item to actually get an integer, you need \code{typeof(5L) == "integer"}.
-  Doubles are by far the most common numeric type in R \AT{I can get data
-    for this}, and you can even use them to index lists and vectors:
-  \code{list(1, 2)\[\[1.9\]\] == 1}.  Note the double brackets for list
-  indexing.
-\item NAs are common in data-centric languages (e.g. MATLAB, \AT{others?}),
-  and R is no exception.  As for their type, we see that \code{typeof(NA) ==
-    "logical"}, but that isn't to say that NA only has type logical.
-  Consider \code{typeof(c(1, NA, 3)\[2\])}: NA actually inhabits every type
-  in R, \AT{details}.
-\item \AT{other examples?}
-\end{itemize}
+Scalar data types include the distinguished \k{NULL} value, which is also of
+type \k{NULL}, instance of classes written using the S4 object system,
+closures and environments.  The implementation of R has a number of other
+types that are mostly not used by user code, they are listed in
+Figure~\ref{types} for reference.
 
@@
 \code{attr(x, "name")} functions.  \code{attributes}
-
-  - dim
-
-class:
 
@@
 number of other packages.  In~\cite{issta18}, the authors estimated code
-coverage to be around 60\% when including reverse dependencies.  The
+coverage to be around 68\% when including reverse dependencies.  The
 rationale for our choice of packages was pragmatic, our infrastructure adds

### 2019-03-21 paper/main.tex
types

@@
 
-Another form of type information relates to metadata on values, and R allows
-programmers to add said metadata in the form of {\it attributes}.  The
-functionality to do so in R is through the \code{attributes(x)} and
-\code{attr(x, "name")} functions.  \code{attributes}
+The addition of attributes lets programmers extend the set of types by
+tagging data with user-defined attributes. For example, one could define a
+vector of four values, \code{x<-c(1,2,3,4)} and then attach the attribute
+\k{dim} with a pair of numbers as value: \code{attr(x,"dim")<-c(2,2)}.  From
+that point, arithmetic functions will treat \k{x} as a 2x2 matrix. Another
+attribute that can be set is the \k{class}.  This attribute can be bound to
+a list of class names. For instance, \code{class(x)<-"human"}, set the class
+of \k{x} to be \k{human}.  Attributes are thus used for object-oriented
+programming. The S3 object system support single dispatch on the class of
+the first argument of a function, whereas the S4 object system allows
+multiple dispatch (on all arguments). Some of the most widely used data
+type, such as data frames, leverage attributes. A data frame, for instance,
+is a list of vectors with a class and a column name attribute.
 
-Besides these primitive data types, R has an out-of-the-box data type called
-the data frame for storing data, and this data type is essential to a large
-part of R's functionality.  A data frame is a two-dimensional array where
-each row represents an observation, with columns indicating the values of
-some observation.  For example, our own analysis is written in R, and one of
-our data frames has a row for each package analyzed, with columns for values
-ranging from number of polymorphic functions, number of monomorphic
-arguments, unique attribute patterns, etc.  \AT{Put a figure?}  Much of R's
-base functionality, and indeed much of the functionality provided by
-packages, operate on data frames.
+\paragraph{Summary.} The most common values in R computations are vectorized
+types. R programs do not have a way to constrain values to be scalar.
+\k{NULL} is sometimes used to represent the case when no value is
+available. \k{NA} is used within vector to represent missing observations.
+Attributes can decorate values and are used as building blocks for
+object-oriented programming. A potential type system for R could focus only
+on the builtin types, if one wanted to strive for simplicity, or it could
+try to capture attributes at the risk of increased complexity.
 
+\newpage
 \section{The Corpus}\label{sec:corpus}

### 2019-03-21 paper/main.tex
m

@@
 
-%
-%
-%
-%
-\section{The \texttt{genthat} Package}
-\label{sec:genthat}
-
-As we mentioned, getting our hands on ``front-line'' non-library R code is tricky: 
-R is not like many other languages, where large systems are built up and stored in a repository such as GitHub.  
-Instead, R programmers are liable to leverage a number of packages and work them into their data analysis pipelines.
-
-Thankfully, we are somewhat able to make up for this lack of code.
-CRAN has a policy that published packages are to be accompanied by at least one vignette, and typically by tests and examples showing off how the code is intended to be used.  
-Nominally unrelated, the {\tt genthat} project~\cite{issta18} traces this example code to generate new unit tests which package designers may not have thought of.
-As it happens, we can leverage {\tt genthat} to extract dynamic type information from running package test code, which presumably indicates how package code is intended to be used.
-
-\AT{Draft}
-
-Our tracer is built on of the {\tt genthat}~\cite{issta18} tracer, which we modified to capture the type information of arguments.  
-{\tt genthat} generates traces for each unique function invocation with the express purpose of re-invoking the function at a later time, and instead of generating traces for these invocations we capture type information of the arguments and return values of functions.
-This is done by evaluating the arguments and return values of functions at the end of the function's scope in the relevant environment.  
-The sort of types and granularity of the type information that are generated can be specified in our tracer: 
-Essentially, our tracer is parameterized over a user-specified type system.  
-For instance, if desired our tracer can dig into lists and data frames to get the types of list elements and fields.
-Our tracer produces trace results for each unique type signature encountered, which may then be used for data analysis.
-
-An unfortunate consequence of R's most common usage pattern is the lack of "deployed" R code. 
-R is most often used for exploratory data analysis, and very little of this analysis code is made publicly available on e.g. GitHub.
-The {\tt genthat} angle is thus desirable: 
-With the scarcity of non-library code, examples of package usage is about as representative as one can hope for of the sort of code people will write in their analysis pipelines.
-
-We can do a little bit better than merely package usage examples, though.
-Available through CRAN is a list of {\it reverse dependencies} of packages: 
-for some package {\tt p}, a list of all packages which require {\tt p} in some way.  
-Our tracer is able to leverage and trace this code as well. 
-When running a package, we also trace how all inherited functions are used, thus painting an even clearer picture of function usage.  
-This additional data is considered alongside similar data collected for all other packages, as well as the data obtained by tracing the package's examples.
-
-% Some prose we might want to pull from.
-%First, we leverage {\tt genthat}'s tracer to trace function executions.  The
-%idea here is that we don't really have access to non-library code written in
-%R, as general use patterns are (possibly?) to write small scripts which
-%analyze some bit of data and possibly visualize results.  The goal is likely
-%not to build big working systems, instead to explore data with by writing
-%and rewriting small scripts, ad infinitum.  In looking at package tests,
-%vignettes, and examples, we are painting a picture of how the package
-%designers intended their packages to be used, which we believe is a close
-%approximation of what R users would do.
-
-\section{The Method}
+\newpage
+\section{Methods}
 
@@
 
+Our tracer is built on of the {\tt genthat}~\cite{issta18} tracer, which we
+modified to capture the type information of arguments.  {\tt genthat}
+generates traces for each unique function invocation with the express
+purpose of re-invoking the function at a later time, and instead of
+generating traces for these invocations we capture type information of the
+arguments and return values of functions.  This is done by evaluating the
+arguments and return values of functions at the end of the function's scope
+in the relevant environment.  The sort of types and granularity of the type
+information that are generated can be specified in our tracer: Essentially,
+our tracer is parameterized over a user-specified type system.  For
+instance, if desired our tracer can dig into lists and data frames to get
+the types of list elements and fields.  Our tracer produces trace results
+for each unique type signature encountered, which may then be used for data
+analysis.
+
+%
+The high-level process for tracing one package is as follows:
+
+\begin{enumerate}
+\item first, all functions in the package are instrumented with on-exit
+  hooks.  These hooks call a number of R's available reflection functions on
+  each of the arguments, as well as the return type: The called functions
+  are {\tt typeof}, {\tt class}, and {\tt attributes}.  Recording is done on
+  function exit to avoid forcing promises before they are needed (as these
+  promises may cause side effects).
+
+\item next, runnable code is extracted and run from package tests, examples,
+  and vignettes.
+
+\item each function invocation triggers the on-exit hook, and it is this
+  on-exit code which collects the information.  Depending on the results of
+  {\tt typeof}, {\tt class}, and {\tt attributes}, more analysis may be
+  performed.  For example, if an argument is a list, type information is
+  collected on list elements until a more accurate type (e.g., {\tt
+    list<double>}) can be constructed.  Indeed, in a later section we do
+  propose some possible type systems for R, and our analysis can be
+  parameterized with these type systems to generate type information
+  according to the types specified, and this will be discussed further in
+  Section \AT{yap}.
+
+\item all of the recorded type information makes up a function {\it trace}.
+  This trace is passed to a C++ function (for performance reasons) which
+  hashes it and checks it against already seen trace hashes.  If the trace
+  is determined to be new, the hash is saved and the trace is written to a
+  file.
+
+\item finally, after all test code has finished executing, we are left with
+  a directory full of all of the unique type signatures that were observed
+  while running the package code.
+\end{enumerate}
+
+For each package function, we then consolidate all observed signatures: this
+gives us a {\it function signature}.  \AT{Example?}  In the function
+signature, each function argument is accompanied by a list of types,
+attributes, and classes that have inhabited it.
+
+We will need a notion of polymorphism when undertaking this analysis.  As R
+is unityped, and the only notion of polymorphism in the language is its use
+of dynamic dispatch on the class of an argument.  For our purposes, we will
+define 3 simple types of polymorphism, closely connected with the reflection
+functions we use.  We start by defining what it means for an {\it argument}
+to be polymorphic.  \AT{Could be valuable to have an example here.}
+
+\begin{itemize}
+\item an argument is {\it polymorphic in type} if it has been called with at
+  least two distinct ``types'' according to the \code{typeof} function and
+  our extra analysis.
+\item similarly, an argument is {\it polymorphic in class} if it has been
+  called with at least two distinct classes according to the \code{class}
+  function.
+
+\item attributes are a bit more tricky.  We can't simply count the number of
+  attributes that have inhabited an argument, since values can contain
+  arbitrarily many values (and having several attributes isn't
+  polymorphism---think records).  Instead, we define the {\it attribute
+    pattern} of a value to be the set of all attribute name and type pairs
+  in the attribute list of a value (obtained via \code{attributes}).  And so
+  an argument is {\it polymorphic in attribute} if it has been called with
+  at least two distinct attribute patterns.
+
+\end{itemize}
+
+There is an important distinction to be made between the result of
+\code{typeof} and that of \code{class} and \code{attributes}: In R, the
+\code{typeof} a value cannot change, and this is not the case with class and
+attributes.  As mentioned in \AT{an earlier section}, the class of values
+can be easily redefined by assigning to the value's class attribute.
+Similarly, attributes can be dynamically added, removed, or changed.
+
+This begs the question: what sort of classes and attribute patterns occur
+naturally, before possible user tampering?  In terms of classes, the
+discussion can be found in Section~\ref{sec:rlang}.  As for attributes:
+
+\begin{itemize}
+
+\item data frames have three attributes: {\tt names} for column names, {\tt
+  row.names} for row names, and {\tt class} (which is instantiated to {\tt
+  "data.frame"} upon creation).
+
+\item matrices have at least a {\tt dim} attribute with matrix dimensions,
+  and optionally a {\tt dimnames} attribute for dimension names.
+
+\item lists and vectors may have a {\tt names} attribute, which assigns
+  names to locations.
+
+\item to ease dealing with large sorted sets of non-numeric values, R offers
+  the \code{factor} and \code{levels} functions.  \code{factor(x,
+    levels=someOrder)} adds a {\tt "levels"} attribute to \code{x}, which
+  specifies an ordering for \code{x} according to the \code{sortedOrder}
+  parameter.  \code{factor} also changes the class of the parameter being
+  factored to \code{"factor"}.
+
+\end{itemize}
+
+These naturally-occurring attribute patterns and classes can be separately
+accounted for to highlight user-defined patterns and behaviours.  This will
+be made clear in Section~\ref{sec:results}.
+
 %
 %
-%
+
 %
@@
 
-As we mentioned in Section~\ref{sec:rlang}, R has several notions of ``type'', ranging from the results of \code{typeof} to metadata on values obtainable via the \code{attributes} function.
-Individually, these paint an incomplete picture of the type of a value:
-For example, recall that \code{typeof(5) == typeof(c(5, 5)) == "double"}.
+As we mentioned in Section~\ref{sec:rlang}, R has several notions of
+``type'', ranging from the results of \code{typeof} to metadata on values
+obtainable via the \code{attributes} function.  Individually, these paint an
+incomplete picture of the type of a value: For example, recall that
+\code{typeof(5) == typeof(c(5, 5)) == "double"}.
 
-To be as specific as possible, our analysis collects additional information to create a more specific type than the result of any of R's reflection functions.
-For instance, if the analysis encounters {\tt typeof(x) == "double"}, it will look at the length of {\tt x} to determine if {\tt x} is a scalar or a vector, generating the annotation {\tt scalar/double} or {\tt vector/double} accordingly.
-One notable imperfection in this example is that unit-length vectors will appear as scalars, but if some function was called with unit-length vectors {\it as well as} larger ones, both scalar and vector types will appear, and this distinction can be later collapsed.
-Another example is that our analysis builds parametric list types by capturing the types of list elements.
-We also distinguish NAs ``in the wild'' with a {\tt raw\_NA} type.
+To be as specific as possible, our analysis collects additional information
+to create a more specific type than the result of any of R's reflection
+functions.  For instance, if the analysis encounters {\tt typeof(x) ==
+  "double"}, it will look at the length of {\tt x} to determine if {\tt x}
+is a scalar or a vector, generating the annotation {\tt scalar/double} or
+{\tt vector/double} accordingly.  One notable imperfection in this example
+is that unit-length vectors will appear as scalars, but if some function was
+called with unit-length vectors {\it as well as} larger ones, both scalar
+and vector types will appear, and this distinction can be later collapsed.
+Another example is that our analysis builds parametric list types by
+capturing the types of list elements.  We also distinguish NAs ``in the
+wild'' with a {\tt raw\_NA} type.
 
 %
-%
-%
-%
-\subsection{Data Collection Pipeline}
+\section{Results}\label{sec:results}
 
-As we mentioned in Section~\ref{sec:genthat}, we modify the {\tt genthat} tracer to collect and output the type information for all function invocations.
+\AT{Meta-Comment:} My plan for this section is to linearly go through the
+data, and show logical transitions between various type systems.  I'll start
+with the finest-grained one.
 
-The high-level process for tracing one package is as follows:
+Having built an understanding of our data collection and analysis
+pipeline, we will turn our attention to the results of our analysis.  We
+approach the data with the purpose of informing the design of a static type
+system for R.  To that end, we propose to begin with a very basic type
+system, the {\it Fine-Grained Type System} (FGTS).
 
-\begin{enumerate}
-	\item first, all functions in the package are instrumented with on-exit hooks.
-  	These hooks call a number of R's available reflection functions on each of the arguments, as well as the return type: 
-  	The called functions are {\tt typeof}, {\tt class}, and {\tt attributes}.  
-  	Recording is done on function exit to avoid forcing promises before they are needed (as these promises may cause side effects).
+The logical ``baseline'' for any design of a static type system for R is
+promotion of the language's dynamic types (in the sense of \code{typeof}) to
+static types, but there are some clear limitations to this approach.  For
+example, \code{typeof} makes no distinction between vectors and scalars, as
+indeed there is no such distinction in R.  But that doesn't mean that there
+{\it shouldn't} be a distinction, as an all-new runtime environment design
+for R might benefit from such a distinction.
 
-  	\item next, runnable code is extracted and run from package tests, examples, and vignettes, and \AT{what about rev deps}.
+To address these limitations, we augment the logical naive type system with
+more information which was alluded to in the discussion of our data
+collection methodology.  We will consider unit-length vectors to be scalars,
+collect list member type information to build an accurate list type, and
+promote NAs to their own type.  In addition, we collect attribute and class
+information to distinguish a type for data frames and matrices, as the
+\code{typeof} these is simply {\tt list} in R.  This type system is referred
+to as the Fine-Grained Type System, or FGTS.  An example of a signature in
+this type system can be found in Figure~\AT{figggg}.
 
-	\item each function invocation triggers the on-exit hook, and it is this on-exit code which collects the information.
-	Depending on the results of {\tt typeof}, {\tt class}, and {\tt attributes}, more analysis may be performed.
-	For example, if an argument is a list, type information is collected on list elements until a more accurate type (e.g., {\tt list<double>}) can be constructed.
-	Indeed, in a later section we do propose some possible type systems for R, and our analysis can be parameterized with these type systems to generate type information according to the types specified, and this will be discussed further in Section \AT{yap}.
-
-	\item all of the recorded type information makes up a function {\it trace}.
-	This trace is passed to a C++ function (for performance reasons) which hashes it and checks it against already seen trace hashes.
-	If the trace is determined to be new, the hash is saved and the trace is written to a file.
-
-	\item finally, after all test code has finished executing, we are left with a directory full of all of the unique type signatures that were observed while running the package code.
-
-\end{enumerate}
-
-For each package function, we then consolidate all observed signatures: this gives us a {\it function signature}.
-\AT{Example?}
-In the function signature, each function argument is accompanied by a list of types, attributes, and classes that have inhabited it.
-
-We will need a notion of polymorphism when undertaking this analysis.
-As R is unityped, and the only notion of polymorphism in the language is its use of dynamic dispatch on the class of an argument.
-For our purposes, we will define 3 simple types of polymorphism, closely connected with the reflection functions we use.
-We start by defining what it means for an {\it argument} to be polymorphic.
-\AT{Could be valuable to have an example here.}
-
-\begin{itemize}
-
-	\item an argument is {\it polymorphic in type} if it has been called with at least two distinct ``types'' according to the \code{typeof} function and our extra analysis.
-
-	\item similarly, an argument is {\it polymorphic in class} if it has been called with at least two distinct classes according to the \code{class} function.
-
-	\item attributes are a bit more tricky.
-	We can't simply count the number of attributes that have inhabited an argument, since values can contain arbitrarily many values (and having several attributes isn't polymorphism---think records).
-	Instead, we define the {\it attribute pattern} of a value to be the set of all attribute name and type pairs in the attribute list of a value (obtained via \code{attributes}).
-	And so an argument is {\it polymorphic in attribute} if it has been called with at least two distinct attribute patterns.
-
-\end{itemize}
-
-There is an important distinction to be made between the result of \code{typeof} and that of \code{class} and \code{attributes}:
-In R, the \code{typeof} a value cannot change, and this is not the case with class and attributes.
-As mentioned in \AT{an earlier section}, the class of values can be easily redefined by assigning to the value's class attribute.
-Similarly, attributes can be dynamically added, removed, or changed.
-
-This begs the question: what sort of classes and attribute patterns occur naturally, before possible user tampering?
-In terms of classes, the discussion can be found in Section~\ref{sec:rlang}.
-As for attributes:
-
-\begin{itemize}
-
-	\item data frames have three attributes: {\tt names} for column names, {\tt row.names} for row names, and {\tt class} (which is instantiated to {\tt "data.frame"} upon creation).
-
-	\item matrices have at least a {\tt dim} attribute with matrix dimensions, and optionally a {\tt dimnames} attribute for dimension names.
-
-	\item lists and vectors may have a {\tt names} attribute, which assigns names to locations.
-
-	\item to ease dealing with large sorted sets of non-numeric values, R offers the \code{factor} and \code{levels} functions.
-	\code{factor(x, levels=someOrder)} adds a {\tt "levels"} attribute to \code{x}, which specifies an ordering for \code{x} according to the \code{sortedOrder} parameter.
-	\code{factor} also changes the class of the parameter being factored to \code{"factor"}.
-
-\end{itemize}
-
-These naturally-occurring attribute patterns and classes can be separately accounted for to highlight user-defined patterns and behaviours.
-This will be made clear in Section~\ref{sec:results}.
-
-%
-%
-%
-%
-%
-%
-\section{Results}
-\label{sec:results}
-
-\AT{Meta-Comment:} 
-My plan for this section is to linearly go through the data, and show logical transitions between various type systems.
-I'll start with the finest-grained one.
-
-\AT{Draft}
-Having built an understanding of our data collection and analysis pipeline, we will turn our attention to the results of our analysis.
-We approach the data with the purpose of informing the design of a static type system for R.
-To that end, we propose to begin with a very basic type system, the {\it Fine-Grained Type System} (FGTS).
-
-The logical ``baseline'' for any design of a static type system for R is promotion of the language's dynamic types (in the sense of \code{typeof}) to static types, but there are some clear limitations to this approach.
-For example, \code{typeof} makes no distinction between vectors and scalars, as indeed there is no such distinction in R.
-But that doesn't mean that there {\it shouldn't} be a distinction, as an all-new runtime environment design for R might benefit from such a distinction.
-
-To address these limitations, we augment the logical naive type system with more information which was alluded to in the discussion of our data collection methodology.
-We will consider unit-length vectors to be scalars, collect list member type information to build an accurate list type, and promote NAs to their own type.
-In addition, we collect attribute and class information to distinguish a type for data frames and matrices, as the \code{typeof} these is simply {\tt list} in R.
-This type system is referred to as the Fine-Grained Type System, or FGTS.
-An example of a signature in this type system can be found in Figure~\AT{figggg}.
-
-The first question we will address is: {\it how polymorphic is R?}
-To answer that, we will first look at the number of monomorphic arguments across all of the packages we analyzed.
-The data can be found in Figure~\AT{fiGgGgGgGGg}.
+The first question we will address is: {\it how polymorphic is R?}  To
+answer that, we will first look at the number of monomorphic arguments
+across all of the packages we analyzed.  The data can be found in
+Figure~\AT{fiGgGgGgGGg}.
 
@@
 
-\AT{I'm fairly sure the numbers aren't going to be insanely different, so I'll just manually write them for now.}
+\AT{I'm fairly sure the numbers aren't going to be insanely different, so
+  I'll just manually write them for now.}
 
 We see that roughly 10\% of all argument positions in R are polymorphic.
-The numbers for attribute and class polymorphism are included for completeness, but the important number is that of type polymorphism (as our method of constructing types sometimes factors in class and attribute).
-Specifically for classes, often the class and \code{typeof} a value will line up (refer to Section~\ref{sec:rlang}).
+The numbers for attribute and class polymorphism are included for
+completeness, but the important number is that of type polymorphism (as our
+method of constructing types sometimes factors in class and attribute).
+Specifically for classes, often the class and \code{typeof} a value will
+line up (refer to Section~\ref{sec:rlang}).
 

### 2019-03-21 paper/main.tex
methods

@@
 
+\newcommand{\genthat}{{\sc genthat}\xspace}
+
 In this section, we will detail our methodology for collecting data.
 
-Our tracer is built on of the {\tt genthat}~\cite{issta18} tracer, which we
+The basic idea we are exploring is to observe arguments and return values of
+function calls, and from these generalize possible type signatures for the
+called functions. To this end, we build on an open source tool called
+\genthat whose purpose was to generate unit tests for R
+libraries~\cite{issta18}.  To generate test cases, one must record and
+recreate accurately arguments and return values and avoid generating
+redundant tests (i.e. tests that do not increase code coverage). For our
+purposes, we change the existing tool in two main ways, we record argument
+\emph{shapes} rather than their values, given the limited number of shapes
+this allowed us to avoid having to run code coverage minimization phase of
+the tool. Both changes were beneficial for scalability -- allowing us to
+trace more calls than was reported in~\cite{issta18}.
+
+\subsection{Example}
+
+To illustrate our approach consider the following script which adds a double
+to an integer, and the creates a matrix and a vector to i.
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+> 1 + 2L
+[1] 3
+> x <- c(1,2,3,4)
+> attr(x,"dim") <- c(2,2)
+> x + c(1,2)
+     [,1] [,2]
+[1,]    2    4
+[2,]    4    6
+\end{lstlisting}}\caption{Example script}\label{example}\end{figure}
+
+The functions being called here are \k{`+`}, \k{`<-`}, \k{c} and
+\k{`attr<-`}.  Some functions syntactic support, addition is infix, and the
+assignment to an attribute is written mixfix. The shapes we would expect to
+record are as follows:
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+`+`: D[1], I[1] -> D[1]
+`c`: D[1], D[1], D[1], D[1] -> D[4]
+`<-`: S, D[3] -> D[3]
+`c`: D[1], D[1] -> D[2]
+`attr<-1`:  S, C, D[2] -> 
+`c`: D[1], D[1] -> D
+`+`: D[4]{dim=(2,2)}, D[2] -> D[4]{dim=(2,2)}
+\end{lstlisting}}\caption{Recorded shapes}\label{shapes}\end{figure}
+
+In the above we abreviate types, \k{double}, \k{integer}, \k{symbol} and
+\k{character}. For vectorized types we record their length and for all types
+we record attributes with some of their values. Looking at the signatures
+observed for addition, it is clear that the function is polymorphic as it
+starts with the addition of two scalar numbers of different types, and then
+adds a matrix to a vector returning a matrix.
+
+\subsection{Implementation}
+
+tracer, which we
 modified to capture the type information of arguments.  {\tt genthat}

### 2019-03-21 paper/main.tex
minor

@@
 
+In R, function can be called with named parameters, R support variable
+argument lists, and arguments can have default values. Putting all of these
+together consider the following declaration:
+
+\begin{lstlisting}[style=R]
+f <- function(x, ..., y=3) x + y
+\end{lstlisting}
+
+\noindent
+Function \k{f} can be called with a single argument \code{f(3)}, with named
+argument \code{f(y=4,x=2)} and with a variable number of arguments,
+\code{f(1,2,3,4,y=5)}, all of these calls will return \code{6}.
+
 R has a number of features that are not crucial to the present
@@
 \newpage
-\section{The Corpus}\label{sec:corpus}
+\section{Corpus}\label{sec:corpus}
 
@@
 
-In this section, we will detail our methodology for collecting data.
-
-The basic idea we are exploring is to observe arguments and return values of
+In this section, we will detail our methodology for collecting data.  The
+basic idea we are exploring is to observe arguments and return values of
 function calls, and from these generalize possible type signatures for the
@@
 To illustrate our approach consider the following script which adds a double
-to an integer, and the creates a matrix and a vector to i.
+to an integer, and then creates a matrix and a vector, adds them together to
+yield a new matrix.
 
@@
 
-tracer, which we
-modified to capture the type information of arguments.  {\tt genthat}
-generates traces for each unique function invocation with the express
-purpose of re-invoking the function at a later time, and instead of
-generating traces for these invocations we capture type information of the
-arguments and return values of functions.  This is done by evaluating the
-arguments and return values of functions at the end of the function's scope
-in the relevant environment.  The sort of types and granularity of the type
-information that are generated can be specified in our tracer: Essentially,
-our tracer is parameterized over a user-specified type system.  For
-instance, if desired our tracer can dig into lists and data frames to get
-the types of list elements and fields.  Our tracer produces trace results
-for each unique type signature encountered, which may then be used for data
-analysis.
+>>>>
 
-%
 The high-level process for tracing one package is as follows:

### 2019-03-21 paper/main.tex
implementation

@@
 \newcommand{\code}[1]{\lstinline|#1|\xspace}
+
+\newcommand{\genthat}{{\sc Genthat}\xspace}
 
@@
 \item A tracing and analysis pipeline that extends a previously published
-  test generation tool named \code{genthat}.
+  test generation tool named \genthat.
 \item Manual analysis of 100 functions to validate the dynamic analysis
@@
 
-\newcommand{\genthat}{{\sc genthat}\xspace}
-
 In this section, we will detail our methodology for collecting data.  The
@@
 `+`: D[1], I[1] -> D[1]
-`c`: D[1], D[1], D[1], D[1] -> D[4]
+c: D[1], D[1], D[1], D[1] -> D[4]
 `<-`: S, D[3] -> D[3]
-`c`: D[1], D[1] -> D[2]
+c: D[1], D[1] -> D[2]
 `attr<-1`:  S, C, D[2] -> 
-`c`: D[1], D[1] -> D
-`+`: D[4]{dim=(2,2)}, D[2] -> D[4]{dim=(2,2)}
+c: D[1], D[1] -> D[2]
+`+`: D[4]{dim=D[2]}, D[2] -> D[4]{dim=D[2]}
 \end{lstlisting}}\caption{Recorded shapes}\label{shapes}\end{figure}
@@
 
->>>>
-
-The high-level process for tracing one package is as follows:
+A high-level description of the workflow of our tool for one package
+retrieved from CRAN is as follows:
 
 \begin{enumerate}
-\item first, all functions in the package are instrumented with on-exit
-  hooks.  These hooks call a number of R's available reflection functions on
-  each of the arguments, as well as the return type: The called functions
-  are {\tt typeof}, {\tt class}, and {\tt attributes}.  Recording is done on
-  function exit to avoid forcing promises before they are needed (as these
-  promises may cause side effects).
-
-\item next, runnable code is extracted and run from package tests, examples,
-  and vignettes.
-
-\item each function invocation triggers the on-exit hook, and it is this
-  on-exit code which collects the information.  Depending on the results of
-  {\tt typeof}, {\tt class}, and {\tt attributes}, more analysis may be
-  performed.  For example, if an argument is a list, type information is
-  collected on list elements until a more accurate type (e.g., {\tt
-    list<double>}) can be constructed.  Indeed, in a later section we do
-  propose some possible type systems for R, and our analysis can be
-  parameterized with these type systems to generate type information
-  according to the types specified, and this will be discussed further in
-  Section \AT{yap}.
-
-\item all of the recorded type information makes up a function {\it trace}.
-  This trace is passed to a C++ function (for performance reasons) which
-  hashes it and checks it against already seen trace hashes.  If the trace
-  is determined to be new, the hash is saved and the trace is written to a
-  file.
-
-\item finally, after all test code has finished executing, we are left with
-  a directory full of all of the unique type signatures that were observed
-  while running the package code.
+\item {\bf Exec generation:} All runnable code in the package is extracted
+  from its tests, examples and vignettes. The code snippets are combined
+  into a single file.
+\item {\bf Installation:} All packages that are required for execution of
+  the current package are downloaded and installed.
+\item {\bf Instrumentation:} As code is loaded into the R, every function
+  definition is instrumented with an \code{on-exit} hook which is invoked
+  when the function returns either normally or through an exception.
+\item {\bf Recording:} When a hook is called, arguments and return value are
+  inspected. We record \k{typeof}, \k{class} and \k{attributes} recursively.
+  For \code{list} values, an extra bit of analysis is performed to record
+  the element type.
+\item {\bf Writing:} Unique signatures are recorded to file with information
+  about which package triggered the recording.
 \end{enumerate}
 
-For each package function, we then consolidate all observed signatures: this
-gives us a {\it function signature}.  \AT{Example?}  In the function
-signature, each function argument is accompanied by a list of types,
-attributes, and classes that have inhabited it.
+Our recording mechanism does not remember the order in which arguments were
+passed, nor does it record which arguments were not passed (and for which
+default values were used). If an argument was not passed, and no default
+value was specified, we record it as \emph{missing}. In R, trying to use
+such a missing value results in an error. For practical reasons, we do not
+record the contents of environments. These can be used as hash tables and
+may be big and are quite likely different from one another.
+
+In R all arguments are passed as promises, unused arguments will be
+unevaluated, for these we can not report a type. THUS WE DO WHAT? SAY MORE.
+
+For each function of each package, we consolidate all recorded observations
+into a function signature. The details of how consolidation is parameterized
+by the definition of a potential type system for R and will be discussed in
+the next section.
+
+\section{Stuff}
 

### 2019-03-22 notebooks/corpus_stats_and_plots.Rmd
put in plots, i will tinker with them over the next bit

--- 2019-03-19 notebooks/corpus_stats_and_plots.Rmd
@@
 combo_df$loc_r <- sapply(combo_df$loc_r, function(x) if (x == 0) x else log10(x))
-combo_df$loc_c <- sapply(combo_df$loc_c, function(x) if (x == 0) x else log10(x))
+combo_df$loc_c <- - sapply(combo_df$loc_c, function(x) if (x == 0) x else log10(x))
 
@@
 combo_df <- combo_df[order(combo_df$loc_r), ]
+# this makes sure the package_names are in the right order, ggplot2 plots according to factors
+# and levels
+combo_df$package_name <- factor(combo_df$package_name, levels=combo_df$package_name)
 
@@
 zzz <- melt(combo_df, id.vars="package_name", variable.name = "language")
+# zzz_c <- melt(combo_df_c, id.vars="package_name", variable.name = "language")
 
 plot_both <- ggplot( data=zzz, aes(package_name, value, group=1))
-plot_both <- plot_both + geom_line(aes(colour = language)) + geom_point(aes(colour = language))
+plot_both <- plot_both + geom_line(aes(colour = language), size=0.001) # +  geom_point(aes(colour = language))
 plot_both <- plot_both + labs(
   title = "Analysis of R Corpus (R and C)",
-  x = "Corpus of Analyzed Packages", 
   y = "Lines of Code in Package (Log10-scaled)"
@@
 
-plot_both <- plot_both + theme(axis.text.x = element_blank())
+plot_both <- plot_both + theme(axis.text.x = element_blank(), axis.title.x=element_blank(), axis.ticks.x=element_blank()) + scale_y_continuous(name="Lines of Code in Package") # , breaks = c(-4, -3, -2, -1, 0, 1, 2, 3, 4), labels = function(x) abs(x^10)
 
@@
 plot_both
+
+ggsave("../paper/plots/linesofrandccode.pdf", plot_both)
+
 ```
+
+Changing gears, lets get the number of recorded events by package.
+Final numbers will become available as the trace completes.
+```{r}
+
+records_by_fun <- unlist(readRDS("../data/total_list_of_records_by_pkg.RDS"))
+records_by_fun_df <- data.frame(pname=names(records_by_fun), records=unname(records_by_fun))
+# records_by_fun_df <- records_by_fun_df[order(records_by_fun_df$records), ]
+# records_by_fun_df$pname <- factor(records_by_fun_df$pname, levels=records_by_fun_df$pname)
+
+plot_records <- ggplot(data=records_by_fun_df, aes(x=pname, y=records, group=1)) +
+                geom_line() + 
+                labs( title = "Numbers of Recorded Signatures",
+                      y = "Recorded Traces in a Package (Thousands)") +
+                theme(axis.text.x = element_blank(), axis.title.x=element_blank(),
+                      axis.ticks.x=element_blank())
+
+# plot
+plot_records
+
+# generate file
+ggsave("../paper/plots/recordsbypkg.pdf", plot_records)
+```

### 2019-03-22 paper/main.tex
put in plots, i will tinker with them over the next bit
counts in top 10 table, some stuff left that i will get once data avail

@@
 \definecolor{DarkGray}{rgb}{.5,.5,.5}
+
+\graphicspath{ {./plots/} }
 
@@
 analysts.  Typical workflows go from data import, to data cleaning, followed
-by steps of modeling, transformation and visualization. Often the code that
-implement these analysis pipelines together with their results resides in
+by steps of modeling, transformation and visualization. Often, the code that
+implements these analysis pipelines together with their results resides in
 notebooks. Unfortunately notebooks are typically not shared in public
@@
 new packages a day~\cite{LIgges2017}.  Unlike sites like GitHub, CRAN is a
-curated collection of packages. A package is only accepted to CRAN if it
+curated collection of packages: A package is only accepted to CRAN if it
 abides by a number of well-formedness rules.  Most relevant for our
-purposes, packages must have data, examples, vignettes and tests. All of
-which must successfully run. From our perspective this means that each of
-the package in CRAN comes with up to a dozen executable scripts that
-exercise some of its functionality.
+purposes, packages must have data, examples, vignettes and tests, all of
+which must successfully run. From our perspective this means that each package 
+in CRAN comes with several executable scripts that exercise some of its functionality.
+Notable exceptions to this rule are packages only containing data, which have no runnable
+code but are referenced by other packages \AT{I should collect stats on this}.
 
@@
 all executable code snippets from documentation, vignettes and tests and ran
-them independently recording all call to R functions.  It is noteworthy that
+them independently recording all calls to R functions.  It is noteworthy that
 in order to run the scripts in one package, it is often necessary to load a
 number of other packages.  In~\cite{issta18}, the authors estimated code
-coverage to be around 68\% when including reverse dependencies.  The
-rationale for our choice of packages was pragmatic, our infrastructure adds
-overhead to the execution of scripts, we eliminated any package that would
-take more than one hour to run. We expect to extend our coverage of CRAN for
-the next version of this paper.
+coverage to be around 68\% when including reverse dependencies.  As our infrastructure
+adds overhead to script execution, coupled with the fact that some scripts take 
+inordinate amounts of time to run, we limit our analysis of any given package to one hour.
+We capture and record function signatures on-the-fly, though, so we do collect (partial) data even 
+for these long-running packages. \AT{We expect to extend our coverage of CRAN for the next version of this paper.}
 
@@
 
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[scale=0.6]{linesofrandccode.pdf}
+\caption{Lines of Code (R and C) in the Analyzed Corpus}
+\label{allcloc}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[scale=0.6]{recordsbypkg}
+\caption{Numbers of Recorded Function Invocations in the Analyzed Corpus}
+\label{allcloc}
+\end{center}
+\end{figure}
+
 % https://www.r-pkg.org/downloaded
@@
 \bf Package & \bf Description & \bf R LOC &\bf C LOC &\bf Scripts\\\hline
-\tt Rcpp  & Seamless C++ integration & 10K & 3K & 12 \\
-\tt rlang & Functions for 'Tidyverse'&&& \\
-\tt glue  & Interpreted string literals&&& \\
-\tt tibble & Simple data frames&& &\\
-\tt stringi &  String processing &&& \\
-\tt ggplot2 & Data visualisations&& &\\
-\tt dplyr  &  Data manipulation&&& \\
-\tt pillar & Formatting for columns&&& \\
-\tt R6 & Classes w. ref. semantics&&& \\
-\tt stringr & String operations&&& \\
+\tt Rcpp  & Seamless C++ integration & 2.1K & 4K & 12 \\
+\tt rlang & Functions for 'Tidyverse' & 4.4K & 3.8K & ? \\
+\tt glue  & Interpreted string literals & 0.2K & 0.2K & 8 \\
+\tt tibble & Simple data frames & 1.3K & 0.3K & ? \\
+\tt stringi &  String processing & 1.5K & 515K & ? \\
+\tt ggplot2 & Data visualisations & 10.8K & 0 & 130 \\
+\tt dplyr  &  Data manipulation & 3.8K & 4.7K & 78 \\
+\tt pillar & Formatting for columns & 1.3K & 0 & ? \\
+\tt R6 & Classes w. ref. semantics & 0.6K & 0 & ? \\
+\tt stringr & String operations & 0.5K & 0 & ? \\
 \end{tabular}}\caption{10 Most Downloaded Packages.}\label{most}

### 2019-03-22 notebooks/corpus_stats_and_plots.Rmd
results, some tables and stuff

@@
 plot_both <- ggplot( data=zzz, aes(package_name, value, group=1))
-plot_both <- plot_both + geom_line(aes(colour = language), size=0.001) # +  geom_point(aes(colour = language))
+plot_both <- plot_both + geom_line(aes(colour = language), size=0.05) # +  geom_point(aes(colour = language))
 plot_both <- plot_both + labs(
@@
   y = "Lines of Code in Package (Log10-scaled)"
-)
+) + guides(colour=F)
 
-plot_both <- plot_both + theme(axis.text.x = element_blank(), axis.title.x=element_blank(), axis.ticks.x=element_blank()) + scale_y_continuous(name="Lines of Code in Package") # , breaks = c(-4, -3, -2, -1, 0, 1, 2, 3, 4), labels = function(x) abs(x^10)
+# cut_top <- data.frame( x = c(-Inf, Inf), y = 2, cutoff = factor(2) )
+# cut_bot <- data.frame( x = c(-Inf, Inf), y = -2, cutoff = factor(-2) )
+
+plot_both <- plot_both + theme(
+  axis.text.x  = element_blank(), 
+  axis.title.x = element_blank(),
+  axis.ticks.x = element_blank(),
+  axis.ticks.y = element_line(), 
+  panel.grid.major.y = element_line(colour = "red")) + # ,
+  # panel.ontop = T) + 
+  scale_y_continuous(name="Lines of Code in Package", breaks = c(-4, -3, -2, -1, 0, 1, 2, 3, 4), 
+                     labels = c(10000, 1000, 100, 10, 0, 10, 100, 1000, 10000))
+  # geom_line(aes( x, y), cut_top) +
+  # geom_line(aes( x, y), cut_bot)
 
@@
 
-ggsave("../paper/plots/linesofrandccode.pdf", plot_both)
+ggsave("../paper/plots/linesofrandccode.jpg", plot_both)
 
@@
 records_by_fun <- unlist(readRDS("../data/total_list_of_records_by_pkg.RDS"))
-records_by_fun_df <- data.frame(pname=names(records_by_fun), records=unname(records_by_fun))
+records_by_fun_df <- data.frame(pname=names(records_by_fun), records=unname(records_by_fun)/1000)
 # records_by_fun_df <- records_by_fun_df[order(records_by_fun_df$records), ]
@@
 # generate file
-ggsave("../paper/plots/recordsbypkg.pdf", plot_records)
+ggsave("../paper/plots/recordsbypkg.jpg", plot_records)
 ```

### 2019-03-22 paper/main.tex
results, some tables and stuff

@@
 \begin{center}
-\includegraphics[scale=0.6]{linesofrandccode.pdf}
-\caption{Lines of Code (R and C) in the Analyzed Corpus}
+\includegraphics[scale=0.15]{linesofrandccode.jpg}
+\caption{Lines of code in the analyzed corpus. \# of lines of R code are above 0, \# of lines of C code are below 0.}
 \label{allcloc}
@@
 \begin{center}
-\includegraphics[scale=0.6]{recordsbypkg}
+\includegraphics[scale=0.15]{recordsbypkg}
 \caption{Numbers of Recorded Function Invocations in the Analyzed Corpus}
@@
 called functions. To this end, we build on an open source tool called
-\genthat whose purpose was to generate unit tests for R
+\genthat whose purpose is to generate unit tests for R
 libraries~\cite{issta18}.  To generate test cases, one must record and
-recreate accurately arguments and return values and avoid generating
+accurately recreate arguments and return values and avoid generating
 redundant tests (i.e. tests that do not increase code coverage). For our
-purposes, we change the existing tool in two main ways, we record argument
-\emph{shapes} rather than their values, given the limited number of shapes
-this allowed us to avoid having to run code coverage minimization phase of
-the tool. Both changes were beneficial for scalability -- allowing us to
+purposes, we change the existing tool in two main ways: We record argument
+\emph{shapes} rather than their values, and given the limited number of shapes
+this allowed us to avoid having to run the code coverage minimization phase of
+the tool. Both changes were beneficial for scalability---allowing us to
 trace more calls than was reported in~\cite{issta18}.
@@
 
-To illustrate our approach consider the following script which adds a double
-to an integer, and then creates a matrix and a vector, adds them together to
-yield a new matrix.
+To illustrate our approach, consider the following script which adds a double
+to an integer, and then creates a matrix and a vector, and adds them together to
+get a new matrix.
 
@@
 The functions being called here are \k{`+`}, \k{`<-`}, \k{c} and
-\k{`attr<-`}.  Some functions syntactic support, addition is infix, and the
+\k{`attr<-`}.  Some functions syntactic support \AT{?}, addition is infix, and the
 assignment to an attribute is written mixfix. The shapes we would expect to
@@
 \begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-`+`: D[1], I[1] -> D[1]
+`+`: D[1], I[1] -> D[1] # (args: double size 1, integer size 1, return: double size 1)
 c: D[1], D[1], D[1], D[1] -> D[4]
@@
 
-We will need a notion of polymorphism when undertaking this analysis.  As R
+We will need a notion of polymorphism when undertaking this analysis. R
 is unityped, and the only notion of polymorphism in the language is its use
@@
   attributes that have inhabited an argument, since values can contain
-  arbitrarily many values (and having several attributes isn't
+  arbitrarily many attributes (and having several attributes isn't
   polymorphism---think records).  Instead, we define the {\it attribute
@@
   parameter.  \code{factor} also changes the class of the parameter being
-  factored to \code{"factor"}.
+  factored to \texttt{"factor"}.
 
@@
 
-The logical ``baseline'' for any design of a static type system for R is
+The logical ``baseline'' for the design of a static type system for R is
 promotion of the language's dynamic types (in the sense of \code{typeof}) to
@@
 indeed there is no such distinction in R.  But that doesn't mean that there
-{\it shouldn't} be a distinction, as an all-new runtime environment design
-for R might benefit from such a distinction.
+{\it couldn't} be a distinction, as an all-new runtime environment design
+for R might benefit from e.g. not needing to vectorize scalar values.
 
@@
 to as the Fine-Grained Type System, or FGTS.  An example of a signature in
-this type system can be found in Figure~\AT{figggg}.
+this type system can be found in Figure~\ref{fig:exFGTS}.
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+> cat_and_list <- function(x, y) as.list(c(x, y)) # c(2, c(3, 1)) => c(2, 3, 1)
+> cat_and_list(2, c(3, 1)) # => list(2, 3, 1)
+\end{lstlisting}}
+\begin{tabular}{@{}r|l|l|l@{}}\hline
+\bf Argument & \bf Type & \bf Class &\bf Attributes \\
+x & \tt scalar/double & \tt numeric & \tt \{\} \\
+y & \tt vector/double & \tt numeric & \tt \{\} \\
+retv & \tt list<double> & \tt list & \tt \{\}  
+\end{tabular}
+\caption{Example call and FGTS signature}\label{fig:exFGTS}\end{figure}
 
@@
 across all of the packages we analyzed.  The data can be found in
-Figure~\AT{fiGgGgGgGGg}.
+Tables~\ref{tab:argcountsFGTS}~and~\ref{tab:funcountsFGTS}.
+\AT{TODO: have these tables be generated automatically}
+\AT{TODO: side by side in minipage?}
+
+% latex table generated in R 3.5.1 by xtable 1.8-3 package
+% Fri Mar 22 18:03:30 2019
+\begin{table}[ht]
+\label{tab:argcountsFGTS}
+\centering
+\begin{tabular}{rlrr}
+  \hline
+ & Argument Polymorphism & Count & Percentage \\ 
+  \hline
+1 & Full Monomorphic & 448K & 85.50 \\ 
+  2 & Monomorphic in Type & 471K & 90.28 \\ 
+  3 & Monomorphic in Class & 480K & 92.16 \\ 
+  4 & Monomorphic in Attribute Pattern & 480K & 91.99 \\ 
+  5 & Total Seen & 521K & -- \\ 
+   \hline
+\end{tabular}
+\caption{Account of {\it argument} polymorphism in the FGTS.}
+\end{table}
+
+% latex table generated in R 3.5.1 by xtable 1.8-3 package
+% Fri Mar 22 18:10:46 2019
+\begin{table}[ht]
+\label{tab:funcountsFGTS}
+\centering
+\begin{tabular}{rlrr}
+  \hline
+ & Function Polymorphism & Count & Percentage \\ 
+  \hline
+1 & Full Monomorphic & 82K & 66.27 \\ 
+  2 & Monomorphic in Type & 94K & 75.74 \\ 
+  3 & Monomorphic in Class & 97K & 78.45 \\ 
+  4 & Monomorphic in Attribute Pattern & 96K & 77.47 \\ 
+  5 & Total Seen & 124K & -- \\ 
+   \hline
+\end{tabular}
+\caption{Account of {\it function} polymorphism in the FGTS.}
+\end{table}
 
@@
 
-We see that roughly 10\% of all argument positions in R are polymorphic.
+In the tables, ``Full Monomorphic'' denotes monomorphism in type, class, and attribute.
+We see that roughly 10\% of all argument positions in R are polymorphic in type.
 The numbers for attribute and class polymorphism are included for
@@
 
-As for functions, roughly 25\% are polymorphic, or put differently 25\% have at least one polymorphic argument or a polymorphic return.
+As for functions, roughly 25\% are polymorphic in type, or put differently 25\% have at least one polymorphic argument or a polymorphic return.
 The discrepancy between this and the polymorphicity of arguments is explained by the tendency of some functions, particularly those dealing with models, to have {\it many} arguments.

### 2019-03-23 notebooks/paper_results_section.Rmd
plots and new data, with vector/scalar dealt with

--- 2019-03-16 notebooks/general_info.Rmd
@@
 ---
-title: "Overview"
+title: "Paper Results Section"
 output: html_notebook
@@
 
-Ok.
-First, load the full signatures, and make sure we source the data processing file.
-There's an error here, not sure what's causing it.
-This is a work-in-progress.
+This Notebook walks through the results section of the paper.
 
+First, administrivia.
 ```{r}
 source("../code/data_processing.R") # get the functions we need
+
+library(xtable)
 ```
 
+Load the data.
 ```{r, echo=FALSE}
@@
 
-These are the updated counts.
+Get the counts.
 ```{r}
@@
 total_mono_all <- apply_count_to_lopkg_c(los, args_count_all_mono_all)
+```
+
+```{r}
+tot_arg_df <- data.frame(
+  Type=c("Full Monomorphic", "Monomorphic in Type", "Monomorphic in Class", "Monomorphic in Attribute Pattern", "Total Seen"),
+  Count=c(total_mono_all, total_mono_type, total_mono_clas, total_mono_attr, total_arguments),
+  Percentage=round(c(total_mono_all, total_mono_type, total_mono_clas, total_mono_attr, total_arguments) / total_arguments * 100, 2)
+)
+
+xtable(tot_arg_df)
 ```
@@
 
+```{r}
+tot_fun_df <- data.frame(
+  Type=c("Full Monomorphic", "Monomorphic in Type", "Monomorphic in Class", "Monomorphic in Attribute Pattern", "Total Seen"),
+  Count=c(total_mono_all_functions, total_mono_type_functions, total_mono_clas_functions, total_mono_attr_functions, total_functions),
+  Percentage=round(c(total_mono_all_functions, total_mono_type_functions, total_mono_clas_functions, total_mono_attr_functions, total_functions) / total_functions * 100, 2)
+)
+
+xtable(tot_fun_df)
+```
+
+We Have one thing we want to do first.
+Namely, let's get rid of vector/X, scalar/X polymorphism.
+```{r}
+los_vs <- lapply(los, function(lofuns) {lapply(lofuns, combine_scalar_vector_where_appropriate)})
+```
+
+Top signatures, with printing.
 This is how you get the top sigs.
@@
 
-Then, pick out the ones you want.
-The `format_df_for_print` function is just for letting the notebook display the data frame properly.
-The type, attr, and class rows are lists of lists.
+Print top 10.
 ```{r}
-format_df_for_print(poly_type_args[1:20,])
+format_df_for_print(poly_type_args[1:10,]) -> put_me
+xtable(put_me)
 ```
 
-Similarly for attributes, and classes.
+Now, having noted that:
+(1) vector/X, scalar/X is common, and
+(2) ints and doubles are basically interchangable
+while noting that together these make up a huge portion of the polymorphism, we transform.
 ```{r}
-poly_attr_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_attr_args))
-poly_attr_args <- poly_attr_args[order(-poly_attr_args$count),] # sort by count
-poly_attr_args$perc <- round(poly_attr_args$count/sum(poly_attr_args$count) * 100, 2)
+
 ```
-
-```{r}
-poly_class_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_class_args))
-poly_class_args <- poly_class_args[order(-poly_class_args$count),] # sort by count
-poly_class_args$perc <- round(poly_class_args$count/sum(poly_class_args$count) * 100, 2)
-```
-
-Printing is as before.
-Note that ATM I also have attribute type information in here.
-I'll get it to display just the names as well.
-I figured this might be interesting, and I don't expect many functions to be polymorphic in this way.
-```{r}
-format_df_for_print(poly_attr_args[1:20,])
-```
-
-```{r}
-format_df_for_print(poly_class_args[1:20,])
-```
-
-How many things are just scalar?
-```{r}
-lo_scalars <- apply_row_extractor_to_lopkg_c(los, get_rows_from_df_only_scalar)
-total_scalar_arguments <- apply_count_to_lopkg_c(lo_scalars, args_count_all)
-```
-
-Let's see what happens if we change type systems!
-```{r}
-los_basic <- lapply(los, function(lof) {lapply(lof, function(df) {change_type_systems_df(df, type_map_T0_to_r)})})
-```
-
-Let's get some counts.
-This is for a type system which is in direct correspondence with R's typeof.
-```{r}
-# argument counts
-total_mono_type_basic <- apply_count_to_lopkg_c(los_basic, args_count_all_mono_type)
-total_mono_all_basic <- apply_count_to_lopkg_c(los_basic, args_count_all_mono_all)
-```
-
-What if we roll NULL into every type?
-```{r}
-los_basic_no_null <- lapply(los_basic, function(lof) {lapply(lof, function(df) {fold_NULL_into_other_types_df(df)})})
-```
-
-What are the counts now?
-```{r}
-total_mono_type_basic_no_null <- apply_count_to_lopkg_c(los_basic_no_null, args_count_all_mono_type)
-total_mono_all_basic_no_null <- apply_count_to_lopkg_c(los_basic_no_null, args_count_all_mono_all)
-```
-
-Great.
-Now, what are the top instances of type-based polymorphism?
-```{r}
-poly_type_args_basic_no_null <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los_basic_no_null, extract_poly_type_args))
-poly_type_args_basic_no_null <- poly_type_args_basic_no_null[order(-poly_type_args_basic_no_null$count),] # sort by count
-poly_type_args_basic_no_null$perc <- round(poly_type_args_basic_no_null$count/sum(poly_type_args_basic_no_null$count) * 100, 2)
-```
-
-What's the top?
-```{r}
-format_df_for_print(poly_type_args_basic_no_null[1:20,])
-```
-
-So, clearly, "real number polymorphism" is the top choice.
-What if we fold ints and doubles together into reals?
-
-```{r}
-los_basic_no_null_real <- lapply(los_basic_no_null, function(lof) {lapply(lof, function(df) {fold_together_int_double_df(df)})})
-```
-
-What are the counts now?
-```{r}
-total_mono_type_basic_no_null_real <- apply_count_to_lopkg_c(los_basic_no_null_real, args_count_all_mono_type)
-total_mono_all_basic_no_null_real <- apply_count_to_lopkg_c(los_basic_no_null_real, args_count_all_mono_all)
-```
-
-Great.
-Now, what are the top instances of type-based polymorphism?
-```{r}
-poly_type_args_basic_no_null_real <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los_basic_no_null_real, extract_poly_type_args))
-poly_type_args_basic_no_null_real <- poly_type_args_basic_no_null_real[order(-poly_type_args_basic_no_null_real$count),] # sort by count
-poly_type_args_basic_no_null_real$perc <- round(poly_type_args_basic_no_null_real$count/sum(poly_type_args_basic_no_null_real$count) * 100, 2)
-```
-
-What's the top?
-```{r}
-format_df_for_print(poly_type_args_basic_no_null_real[1:20,])
-```
-
-Ok.
-What if we want some per package information?
-Well, to do it on one package (let's try rlang):
-
-```{r}
-rlang_tally <- tally_up_lofun_c(los$rlang)
-rlang_tally$tot_args
-rlang_tally$tot_mono_args_type
-format_df_for_print(rlang_tally$top_poly_type)
-```
-
-As you'll see, the numbers here are low because we didn't actually trace rlang itself, just that some other packages used rlang, giving us limited results.
-
-Now, how about if we take all of R's base classes (i.e., those which occur naturally in R) and lump them all into a "primitive" class.
-What sort of class-based polymorphism do we see, now?
-
-```{r}
-los_primitive_class <- lapply(los, function(lof) {lapply(lof, class_to_primitive_df)})
-```
-
-```{r}
-total_mono_clas_prim <- apply_count_to_lopkg_c(los_primitive_class, args_count_all_mono_class)
-total_mono_clas_functions_prim <- apply_count_to_lopkg_c(los_primitive_class, fun_count_all_mono_class)
-```
-
-Let's get the top poly class now.
-```{r}
-poly_class_args_prim <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los_primitive_class, extract_poly_class_args))
-poly_class_args_prim <- poly_class_args_prim[order(-poly_class_args_prim$count),] # sort by count
-poly_class_args_prim$perc <- round(poly_class_args_prim$count/sum(poly_class_args_prim$count) * 100, 2)
-```
-
-```{r}
-format_df_for_print(poly_class_args_prim[1:20,])
-```
-
-The low percentage numbers suggest that no one user-defined pattern contributes too much to the class-based polymorphism.
-
-Alright, let's do a package tally for all packages.
-First, let's make all the classes primitive in our previous collapsed list.
-
-```{r}
-los_basic_no_null_real_primitive_class <- lapply(los_basic_no_null_real, function(lof) {lapply(lof, class_to_primitive_df)})
-```
-
-Note: this takes a long time to run.
-```{r}
-all_tallies_all_reduced <- lapply(los_basic_no_null_real_primitive_class, tally_up_lofun_c)
-```
-
-```{r}
-all_tallies_sizes <- lapply(all_tallies_all_reduced, function(x) x$tot_size_of_poly)
-```
-
-Ok.
-It's certainly not perfect, but one metric I used for figuring out the size of polymorphism is just by summing up the type-, attribute-, and class-based polymorphism of each argument in a function.
-This might give us some leads.
-The top 20 polymorphic packages by this metric is:
-
-```{r}
-head(sort(unlist(all_tallies_sizes), decreasing=TRUE), 20)
-```
-
-Ok.
-Let's see what we can get out of coinciding type signatures.
-With out reduced type system, we might want to know a little bit more about the "real", "character" type signature, which is by far one of the most common signatures.
-We hypothesize that these should appear alongside lists (and possibly vectors).
-For lists only, we can do this easily:
-```{r}
-# for list coinciding
-coincide_list_index <- get_fun_sigs_matching_pred_from_lopkg(los_basic_no_null_real_primitive_class, does_df_have_coinciding_types_index_and_list_reduced_TS)
-
-# just real, character
-char_real_sigs <- get_fun_sigs_matching_pred_from_lopkg(los_basic_no_null_real_primitive_class, does_df_have_X_types, types_to_find=c("real", "character"), strict=TRUE)
-```
-
-Let's look for vectors, also.
-```{r}
-coincide_list_vector_index <- get_fun_sigs_matching_pred_from_lopkg(los, does_df_have_coinciding_types_index_and_list_or_vector)
-```

### 2019-03-23 paper/main.tex
plots and new data, with vector/scalar dealt with

@@
 
+\AT{TODO: adjust prose to talk about how vector/scalar is dealt with before this point.}
+
 Having built an understanding of our data collection and analysis
@@
 \AT{TODO: have these tables be generated automatically}
-\AT{TODO: side by side in minipage?}
+\AT{TODO: put all counts in one table?}
 
@@
   \hline
-1 & Full Monomorphic & 448K & 85.50 \\ 
-  2 & Monomorphic in Type & 471K & 90.28 \\ 
+1 & Full Monomorphic & 457K & 87.75 \\ 
+  2 & Monomorphic in Type & 485K & 93.08 \\ 
   3 & Monomorphic in Class & 480K & 92.16 \\ 
   4 & Monomorphic in Attribute Pattern & 480K & 91.99 \\ 
-  5 & Total Seen & 521K & -- \\ 
+  5 & Total Seen & 521K & --- \\ 
      \hline
@@
   \hline
-1 & Full Monomorphic & 82K & 66.27 \\ 
-  2 & Monomorphic in Type & 94K & 75.74 \\ 
+1 & Full Monomorphic & 86K & 69.44 \\ 
+  2 & Monomorphic in Type & 100K & 81.01 \\ 
   3 & Monomorphic in Class & 97K & 78.45 \\ 
   4 & Monomorphic in Attribute Pattern & 96K & 77.47 \\ 
-  5 & Total Seen & 124K & -- \\ 
+  5 & Total Seen & 124K & --- \\ 
    \hline
@@
 In the tables, ``Full Monomorphic'' denotes monomorphism in type, class, and attribute.
-We see that roughly 10\% of all argument positions in R are polymorphic in type.
+We see that roughly 7\% of all argument positions in R are polymorphic in type.
 The numbers for attribute and class polymorphism are included for
@@
 
-As for functions, roughly 25\% are polymorphic in type, or put differently 25\% have at least one polymorphic argument or a polymorphic return.
+As for functions, roughly 20\% are polymorphic in type, or put differently 20\% have at least one polymorphic argument or a polymorphic return.
 The discrepancy between this and the polymorphicity of arguments is explained by the tendency of some functions, particularly those dealing with models, to have {\it many} arguments.
@@
 These numbers alone are insufficient, so we ask: {\it of the polymorphic arguments, what are the most common patterns?}
-The answer to that question can be found in Figure~\AT{eyyy}.
+The answer to that question can be found in Table~\ref{tab:toppolyFDTS}.
 
-By and large, the most common pattern is {\tt scalar/X} and {\tt vector/X}.
-One way to account for this would be to collapse scalar and vector types into a vector type:
+% latex table generated in R 3.5.1 by xtable 1.8-3 package
+% Fri Mar 22 21:27:37 2019
+\begin{table}[ht]
+\label{tab:toppolyFDTS}
+\centering
+\begin{tabular}{lrr}
+  \hline
+ Polymorphic Type Signature & \# Occurrences & \% of Total Polymorphism \\ 
+  \hline
+  vector/double, vector/integer & 4838 & 13.42 \\ 
+  scalar/double, scalar/integer & 3666 & 10.17 \\ 
+  NULL, vector/double & 1463 & 4.06 \\ 
+  data.frame, vector/double & 1324 & 3.67 \\ 
+  NULL, scalar/character & 1237 & 3.43 \\ 
+   NULL, scalar/double & 1174 & 3.26 \\ 
+   list$<$any$>$, list$<$double$>$ & 1092 & 3.03 \\ 
+   list$<$any$>$, list$<$list$>$ & 767 & 2.13 \\ 
+   scalar/character, scalar/double & 674 & 1.87 \\ 
+   NULL, vector/character & 640 & 1.78 \\ 
+     \hline
+\end{tabular}
+\caption{Top polymorphic argument signatures in the FDTS.}
+\end{table}
+
+By and large, the most common pattern is {\tt scalar/X} and {\tt vector/X}, but we deal with this polymorphism separately:
 If a function argument is observed to be both a unit-length and non-unit-length vector, then it's reasonable to assume that the argument is intended to be a vector.

### 2019-03-23 paper/main.tex
results: adjusted discussion to forget about vec/sca

@@
 
-\AT{TODO: adjust prose to talk about how vector/scalar is dealt with before this point.}
-
 Having built an understanding of our data collection and analysis
@@
 approach the data with the purpose of informing the design of a static type
-system for R.  To that end, we propose to begin with a very basic type
-system, the {\it Fine-Grained Type System} (FGTS).
+system for R.
 
@@
 \code{typeof} these is simply {\tt list} in R.  This type system is referred
-to as the Fine-Grained Type System, or FGTS.  An example of a signature in
+to as the {\it Fine-Grained Type System}, or FGTS.  An example of a signature in
 this type system can be found in Figure~\ref{fig:exFGTS}.
@@
 \begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-> cat_and_list <- function(x, y) as.list(c(x, y)) # c(2, c(3, 1)) => c(2, 3, 1)
-> cat_and_list(2, c(3, 1)) # => list(2, 3, 1)
+> add_make_list <- function(x, y) as.list(x + y)
+> add_make_list(2, c(3, 1)) # => list(5, 3)
 \end{lstlisting}}
@@
 
-\AT{I'm fairly sure the numbers aren't going to be insanely different, so
-  I'll just manually write them for now.}
-
 In the tables, ``Full Monomorphic'' denotes monomorphism in type, class, and attribute.
@@
 As for functions, roughly 20\% are polymorphic in type, or put differently 20\% have at least one polymorphic argument or a polymorphic return.
-The discrepancy between this and the polymorphicity of arguments is explained by the tendency of some functions, particularly those dealing with models, to have {\it many} arguments.
+The discrepancy between this and the polymorphicity of arguments is explained by the tendency of some functions to have {\it many} arguments.
 
@@
 
-By and large, the most common pattern is {\tt scalar/X} and {\tt vector/X}, but we deal with this polymorphism separately:
-If a function argument is observed to be both a unit-length and non-unit-length vector, then it's reasonable to assume that the argument is intended to be a vector.
+\AT{This should probably be discussed earlier.}
+Not pictured here, the most common polymorphic pattern in the FGTS is the {\tt scalar/X}, {\tt vector/X} pattern, but this is dealt with separately:
+If a function argument is observed to be both a unit-length and non-unit-length vector, then it's reasonable to assume that the argument is intended to be a vector, so we collapse the distinction in such cases.
 
-One other interesting question is: {\it how many arguments are strictly scalar?}
-As it happens, over 40\% of monomorphic arguments with primitive vectorized types are always scalar (so, 40\% of {\it monomorphic} occurrences of {\tt scalar/X} and {\tt vector/X} are scalar).
-This is important as it shows that there is value in a scalar annotation, which could be used to communicate to an R runtime that vectorization of a value is not necessary, which may in turn ead to a performance improvement.
+In Table~\ref{tab:toppolyFDTS}, we see the prominence of {\tt double} and {\tt integer} polymorphism.
+In many dynamic languages, primitive double and integer types can easily stand-in for each other as coercion between them is easy.
+In R, the distinction is mainly just relevant to the implementation, as integers and doubles are stored and dealt with differently in the runtime but programmers only see the distinction when printing.
+We address this by creating a new type, {\it real}, encompassing both.
 
-We will collapse the distinction between vectors and scalars in a moment, but imagine we did so and refer back to Figure~\AT{eyyy}.
-Another major source of polymorphism is the {\tt double} and {\tt integer} distinction.
-This distinction is mainly just relevant to the implementation, as integers and doubles are stored and dealt with differently in the runtime.
-For a programmer, this distinction is meaningless, as both types can be used interchangeably in every situation.
-We propose to also collapse this distinction, instead considering doubles and integers as part of a real number type, {\tt real}.
+As we make the distinction between vectors and scalars, one other interesting question that arises is: {\it how many arguments are strictly scalar?}
+As it happens, over 40\% of monomorphic arguments with primitive vectorized types are always scalar (so, 40\% of {\it monomorphic} occurrences of {\tt scalar/X} or {\tt vector/X} are scalar).
+This is important as it shows that there is value in a scalar annotation, which could be used to communicate to an R runtime that vectorization of a value is not necessary, which may in turn lead to a performance improvement.
 
-\AT{TODO: do vector/scalar collapse and real collapse and get numbers.}
+\AT{Next: get real.}
 

### 2019-03-23 paper/main.tex
results: up to next part of results, also working on generating some better data factoring in missing types

@@
 \caption{Numbers of Recorded Function Invocations in the Analyzed Corpus}
-\label{allcloc}
+\label{recorded}
 \end{center}
@@
 We address this by creating a new type, {\it real}, encompassing both.
+The expected coercion is evident throughout R.
+Consider the snippet in Figure~\ref{fig:realex} for details.
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+> 5L + 1L # => 6L, an integer
+> 5L + 1.2 # => 6.2, a double (5L coerced to 5.0)
+> c(10, 20, 30)[1.2] # => 10, 1.2 coerced to 1L
+\end{lstlisting}}\caption{Example of {\it real} type usage.}\label{fig:realex}\end{figure}
 
@@
 
-\AT{Next: get real.}
+Another clear pattern in Table~\ref{tab:toppolyFDTS} is that of NULL, X.
+In many programming languages, NULLs inhabit every type, and this is not the case in our FGTS.
+The FGTS reflects many of R's implementation details, and this distinction is no exception:
+NULLs are represented as \AT{fill in} in the R runtime.
+
+In terms of programmer behaviour, we observe a pattern of optional arguments throughout several R packages, wherein programmers specify a default argument value which is sometimes NULL.
+Consider the snippet in Figure~\ref{fig:optnull} for an example.
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+# w here is weights
+frac.ranks <- function(x, w=NULL) {
+  if (is.null(w)) w <- rep(1, length(x)) # if no weights passed, take all weights = 1
+  ...
+  \end{lstlisting}}\caption{Example of optional argument.}\label{fig:optnull}\end{figure}
+The distinction between NULL and other types is not particularly useful to the programmer, so we propose to also collapse this distinction.
+
+Not pictured in Table~\ref{tab:toppolyFDTS} are NAs, which are ascribed the type {\tt raw\_NA} in FGTS; NAs in fact inhabit every type, though the default NA type is {\tt logical}.
+Just as we fold NULL into all types, we will do the same with NAs.
+
+The next type system design we will consider is ... .
+\AT{Generate tables. The stuff we do is: real, NULL fold, NA fold.}
 

### 2019-03-24 paper/main.tex
a few examples, more to come. waiting on latest prl3 run results

@@
 
+\AT{Collection of examples of polymorphism...}
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+# w here is weights
+frac.ranks <- function(x, w=NULL) {
+  if (is.null(w)) w <- rep(1, length(x)) # if no weights passed, take all weights = 1
+  ...
+\end{lstlisting}}\caption{Example of optional argument (from {\tt acid} package).}\label{fig:optnull}\end{figure}
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+is.point.inside <- function (point, polyg) {
+    p <- as.numeric(point) # as.numeric(list(1, 2)) => c(1, 2)
+    ...
+\end{lstlisting}}\caption{Example of list/vector argument (from {\tt bivrp} package).}\label{fig:listvec}\end{figure}
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+function (object, fn = mean, par.name, comp.label, chain.no,  ...) 
+{
+    ...
+    if (is.character(fn)) {
+        if (fn == "MODE" | fn == "MAP") {
+            do_MAP <- TRUE
+        }
+        else {
+            do_MAP <- FALSE
+            fn <- match.fun(fn) # looks up the function name
+        }
+    }
+    ...
+\end{lstlisting}}\caption{Example of char/closure argument (from {\tt BAMBI} package).}\label{fig:charclos}\end{figure}
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+nret.translator <- function(dat, spss="out", ...){
+  ...
+  if(identical(spss, "out")) {
+    if(!is.vector(dat)) {
+        stop(simpleError("If 'corr=TRUE', 'dat' must be a vector!"))
+      }
+    ...
+  } else {
+    ...
+    # here, dat is data.frame
+    items.idx <- items.idx[order(names(dat[, items.idx]))]
+  }
+
+\end{lstlisting}}\caption{Example of data.frame/vector/double argument (from {\tt klausR} package).}\label{fig:charclos}\end{figure}
+
 \AT{Meta-Comment:} My plan for this section is to linearly go through the
@@
 Consider the snippet in Figure~\ref{fig:optnull} for an example.
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-# w here is weights
-frac.ranks <- function(x, w=NULL) {
-  if (is.null(w)) w <- rep(1, length(x)) # if no weights passed, take all weights = 1
-  ...
-  \end{lstlisting}}\caption{Example of optional argument.}\label{fig:optnull}\end{figure}
+\AT{TODO put that figure here}
 The distinction between NULL and other types is not particularly useful to the programmer, so we propose to also collapse this distinction.

### 2019-03-24 paper/main.tex
more examples, with some descriptions. more to come

@@
   commentstyle=\color{Gray}\em,
-  stringstyle=\color{Gray}
+  stringstyle=\color{Gray},
+  numbers=left
 }
@@
 %
-\section{Results}\label{sec:results}
-
-\AT{Collection of examples of polymorphism...}
+%
+%
+%
+%
+%
+\section{Examples of Polymorphism}
+\label{sec:polyex}
+
+Figure~\ref{fig:optnull} shows an example of a function with an optional argument.
+In the function, {\tt x} is a sorted vector of income values, and {\tt w} is a vector of weights, optionally NULL.
+The function computes fractional ranks required for the computation of some coefficient implemented by the package.
+We see that if {\tt w} is NULL, then default unit weights are generated.
 
 \begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-# w here is weights
 frac.ranks <- function(x, w=NULL) {
@@
 \end{lstlisting}}\caption{Example of optional argument (from {\tt acid} package).}\label{fig:optnull}\end{figure}
+
+Figure~\ref{fig:listvec} shows an example of a function with list and vector polymorphism.
+The function takes a list or vector {\tt point} and a data frame {\tt polyg} representing a polygon.
+In it, we see that {\tt point} can be either a list or a vector, and the function code casts the argument to a double vector.
 
@@
 
+\AT{I'll explain this better, maybe?}
+In Figure~\ref{fig:dfdbl}, we see a function that, among other things, takes in some data ({\tt dat}) and a character string ({\tt spss}) specifying either {\tt "in"} or {\tt "out"}.
+Now, depending on the value of {\tt spss}, {\tt dat} will either be a data frame or a vector of doubles.
+\AT{This ties in to the (arg, retv) coupling, which we will have data on soon.}
+
 \begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
@@
     if(!is.vector(dat)) {
-        stop(simpleError("If 'corr=TRUE', 'dat' must be a vector!"))
+        stop(simpleError("...'dat' must be a vector!"))
       }
@@
   }
-
-\end{lstlisting}}\caption{Example of data.frame/vector/double argument (from {\tt klausR} package).}\label{fig:charclos}\end{figure}
+\end{lstlisting}}\caption{Example of a data.frame/(vector/double) argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}
+
+In Figure~\ref{fig:chardbl}, we see a function which takes in a list ({\tt network}), a vector of indices of that list ({\tt fixIndices}), and a vector of values ({\tt values}).
+The locations specified by {\tt fixIndicies} in {\tt network} are updated with {\tt values}.
+Here, {\tt fixIndicies} has been observed to be either a character or double vector:
+In R, list indices are typically doubles, but can be characters (if the list or vector being indexed has a {\tt names} attribute).
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+fixGenes <- function (network, fixIndices, values)  {
+  ...
+  network$fixed[fixIndices] <- as.integer(values)
+  ...
+}
+\end{lstlisting}}\caption{Example of character/double argument (from {\tt BoolNet} package).}\label{fig:chardbl}\end{figure}
+
+%
+%
+%
+%
+%
+%
+\section{Results}\label{sec:results}
 

### 2019-03-25 paper/main.tex
filled in some prelim numbers in corpus

@@
 
+\AT{TODO Change "recorded" to "observed" in recordsbypkg}
+
 \begin{figure}[htbp]
@@
 % https://www.r-pkg.org/downloaded
-\begin{figure}[!th]{\footnotesize\begin{tabular}{@{}r||l|r|r|r@{}}\hline
-\bf Package & \bf Description & \bf R LOC &\bf C LOC &\bf Scripts\\\hline
-\tt Rcpp  & Seamless C++ integration & 2.1K & 4K & 12 \\
-\tt rlang & Functions for 'Tidyverse' & 4.4K & 3.8K & ? \\
-\tt glue  & Interpreted string literals & 0.2K & 0.2K & 8 \\
-\tt tibble & Simple data frames & 1.3K & 0.3K & ? \\
-\tt stringi &  String processing & 1.5K & 515K & ? \\
-\tt ggplot2 & Data visualisations & 10.8K & 0 & 130 \\
-\tt dplyr  &  Data manipulation & 3.8K & 4.7K & 78 \\
-\tt pillar & Formatting for columns & 1.3K & 0 & ? \\
-\tt R6 & Classes w. ref. semantics & 0.6K & 0 & ? \\
-\tt stringr & String operations & 0.5K & 0 & ? \\
+\begin{figure}[!th]{\footnotesize\begin{tabular}{@{}r||l|r|r|r|r|r@{}}\hline
+\bf Package & \bf Description & \bf R LOC &\bf C LOC &\bf Scripts & \bf Calls Observed & \bf Calls Recorded \\
+\hline
+\tt Rcpp  & Seamless C++ integration & 2.1K & 4K & 25 & 51K & 282 \\
+\tt rlang & Functions for 'Tidyverse' & 4.4K & 3.8K & 122 & 2,240K & 5,171 \\
+\tt glue  & Interpreted string literals & 0.2K & 0.2K & 8 & 2K & 120 \\
+\tt tibble & Simple data frames & 1.3K & 0.3K & 16 & 839K & 4,148 \\
+\tt stringi &  String processing & 1.5K & 515K & 64 & 692K & 666 \\
+\tt ggplot2 & Data visualisations & 10.8K & 0 & 130 & 110K & 2,636\\
+\tt dplyr  &  Data manipulation & 3.8K & 4.7K & 78 & 157K & 1,967 \\
+\tt pillar & Formatting for columns & 1.3K & 0 & 13 & 549K & 1,273 \\
+\tt R6 & Classes w. ref. semantics & 0.6K & 0 & 2 & 1K & 226 \\
+\tt stringr & String operations & 0.5K & 0 & 32 & 1,551K & 371 \\
 \end{tabular}}\caption{10 Most Downloaded Packages.}\label{most}
@@
 
+\AT{Putting examples of polymorphism in this section, we can talk about them and where they should go.}
+
+First, Figure~\ref{fig:realex} shows an example of how integers and doubles are very compatible.
+\AT{TODO add matrix (dimensions) to the example.}
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+> 5L + 1L # => 6L, an integer
+> 5L + 1.2 # => 6.2, a double (5L coerced to 5.0)
+> c(10, 20, 30)[1.2] # => 10, 1.2 coerced to 1L
+\end{lstlisting}}\caption{Example of {\it real} type usage.}\label{fig:realex}\end{figure}
+
 Figure~\ref{fig:optnull} shows an example of a function with an optional argument.
@@
 
+Figure~\ref{fig:charclos} shows an example of a function with character and function polymorphism.
+\AT{Bluh.}
+At a high level, this function calculates point estimates for an {\tt angmcmc} object (specific to some packages).
+The caller specifies {\tt fn}, either a function or a function name which will be evaluated on the object samples to estimate parameters.
+The call to lookup the function name is on line 10.
+
 \begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-function (object, fn = mean, par.name, comp.label, chain.no,  ...) 
-{
+pointest <- function (object, fn = mean, par.name, comp.label, chain.no,  ...) {
     ...
@@
 \begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-nret.translator <- function(dat, spss="out", ...){
+nret.translator <- function(dat, spss="out", ...) {
   ...
@@
 The expected coercion is evident throughout R.
-Consider the snippet in Figure~\ref{fig:realex} for details.
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-> 5L + 1L # => 6L, an integer
-> 5L + 1.2 # => 6.2, a double (5L coerced to 5.0)
-> c(10, 20, 30)[1.2] # => 10, 1.2 coerced to 1L
-\end{lstlisting}}\caption{Example of {\it real} type usage.}\label{fig:realex}\end{figure}
+Refer to the snippet in Figure~\ref{fig:realex} for details.
 

### 2019-03-26 notebooks/corpus.Rmd
updated counts and new signatures, dunno if i should push the data (it big)

---
title: "Corpus Analysis"
output: html_notebook
---

Always load data processing.
```{r}
source("../code/data_processing.R")
```

```{r}
get_counts_for_pkg_in_counts_dir("rlang", path_to_counts = "../data/genthat_counts_final")
```

Get counts for observations.
```{r}

pkgs_to_count <- c("Rcpp", "rlang", "glue", "tibble", "stringi", "ggplot2", "dplyr", "pillar", "R6", "stringr")
names(pkgs_to_count) <- pkgs_to_count

top_obs <- lapply(pkgs_to_count, function(n) get_counts_for_pkg_in_counts_dir(n, path_to_counts = "../data/genthat_counts_final"))

top_obs

```

```{r}

```

### 2019-03-26 paper/main.tex
updated counts and new signatures, dunno if i should push the data (it big)

@@
 \hline
-\tt Rcpp  & Seamless C++ integration & 2.1K & 4K & 25 & 51K & 282 \\
-\tt rlang & Functions for 'Tidyverse' & 4.4K & 3.8K & 122 & 2,240K & 5,171 \\
-\tt glue  & Interpreted string literals & 0.2K & 0.2K & 8 & 2K & 120 \\
-\tt tibble & Simple data frames & 1.3K & 0.3K & 16 & 839K & 4,148 \\
-\tt stringi &  String processing & 1.5K & 515K & 64 & 692K & 666 \\
-\tt ggplot2 & Data visualisations & 10.8K & 0 & 130 & 110K & 2,636\\
-\tt dplyr  &  Data manipulation & 3.8K & 4.7K & 78 & 157K & 1,967 \\
-\tt pillar & Formatting for columns & 1.3K & 0 & 13 & 549K & 1,273 \\
+\tt Rcpp  & Seamless C++ integration & 2.1K & 4K & 25 & 55K & 282 \\
+\tt rlang & Functions for 'Tidyverse' & 4.4K & 3.8K & 122 & 3,924K & 5,171 \\
+\tt glue  & Interpreted string literals & 0.2K & 0.2K & 8 & 4K & 120 \\
+\tt tibble & Simple data frames & 1.3K & 0.3K & 16 & 1,332K & 4,148 \\
+\tt stringi &  String processing & 1.5K & 515K & 64 & 923K & 666 \\
+\tt ggplot2 & Data visualisations & 10.8K & 0 & 130 & 153K & 2,636\\
+\tt dplyr  &  Data manipulation & 3.8K & 4.7K & 78 & 233K & 1,967 \\
+\tt pillar & Formatting for columns & 1.3K & 0 & 13 & 803K & 1,273 \\
 \tt R6 & Classes w. ref. semantics & 0.6K & 0 & 2 & 1K & 226 \\
-\tt stringr & String operations & 0.5K & 0 & 32 & 1,551K & 371 \\
+\tt stringr & String operations & 0.5K & 0 & 32 & 1,764K & 371 \\
 \end{tabular}}\caption{10 Most Downloaded Packages.}\label{most}
@@
   \hline
-1 & Full Monomorphic & 457K & 87.75 \\ 
-  2 & Monomorphic in Type & 485K & 93.08 \\ 
-  3 & Monomorphic in Class & 480K & 92.16 \\ 
-  4 & Monomorphic in Attribute Pattern & 480K & 91.99 \\ 
-  5 & Total Seen & 521K & --- \\ 
+  1 & Full Monomorphic & 784K & 85.19 \\ 
+  2 & Monomorphic in Type & 839K & 91.27 \\ 
+  3 & Monomorphic in Class & 831K & 90.31 \\ 
+  4 & Monomorphic in Attribute Pattern & 830K & 90.25 \\ 
+  5 & Total Seen & 920K & --- \\ 
      \hline
@@
   \hline
-1 & Full Monomorphic & 86K & 69.44 \\ 
-  2 & Monomorphic in Type & 100K & 81.01 \\ 
-  3 & Monomorphic in Class & 97K & 78.45 \\ 
-  4 & Monomorphic in Attribute Pattern & 96K & 77.47 \\ 
-  5 & Total Seen & 124K & --- \\ 
+  1 & Full Monomorphic & 143K & 65.64 \\ 
+  2 & Monomorphic in Type & 169K & 77.65 \\ 
+  3 & Monomorphic in Class & 163K & 75.13 \\ 
+  4 & Monomorphic in Attribute Pattern & 161K & 74.31 \\ 
+  5 & Total Seen & 217K & --- \\ 
    \hline
@@
 In the tables, ``Full Monomorphic'' denotes monomorphism in type, class, and attribute.
-We see that roughly 7\% of all argument positions in R are polymorphic in type.
+We see that roughly 9\% of all argument positions in R are polymorphic in type.
 The numbers for attribute and class polymorphism are included for
@@
 
-As for functions, roughly 20\% are polymorphic in type, or put differently 20\% have at least one polymorphic argument or a polymorphic return.
+As for functions, roughly 22\% are polymorphic in type, or put differently 22\% have at least one polymorphic argument or a polymorphic return.
 The discrepancy between this and the polymorphicity of arguments is explained by the tendency of some functions to have {\it many} arguments.
@@
   \hline
-  vector/double, vector/integer & 4838 & 13.42 \\ 
-  scalar/double, scalar/integer & 3666 & 10.17 \\ 
-  NULL, vector/double & 1463 & 4.06 \\ 
-  data.frame, vector/double & 1324 & 3.67 \\ 
-  NULL, scalar/character & 1237 & 3.43 \\ 
-   NULL, scalar/double & 1174 & 3.26 \\ 
-   list$<$any$>$, list$<$double$>$ & 1092 & 3.03 \\ 
-   list$<$any$>$, list$<$list$>$ & 767 & 2.13 \\ 
-   scalar/character, scalar/double & 674 & 1.87 \\ 
-   NULL, vector/character & 640 & 1.78 \\ 
+vector/double, vector/integer & 10,316 & 12.85 \\ 
+scalar/double, scalar/integer & 7,194 & 8.96 \\ 
+NULL, vector/double & 3,087 & 3.85 \\ 
+data.frame, vector/double & 2,604 & 3.24 \\ 
+NULL, scalar/character & 2,333 & 2.91 \\ 
+list$<$any$>$, list$<$double$>$ & 2,224 & 2.77 \\ 
+NULL, scalar/double & 2,066 & 2.57 \\ 
+raw\_NA, vector/double & 1,465 & 1.83 \\ 
+list$<$any$>$, list$<$list$>$ & 1,426 & 1.78 \\ 
+NULL, vector/character & 1,385 & 1.73 \\ 
      \hline

### 2019-03-26 notebooks/corpus_stats_and_plots.Rmd
notebooks, new plots, data up to date. still debating if i should push it ... it big

@@
 
-r_counts <- readRDS("../data/R_loc_by_package.RDS")
-c_counts <- readRDS("../data/c_loc_by_package.RDS")
+count_df <- readRDS("../data/final_loc_counts.RDS")
+
+r_counts <- count_df[, 1]; names(r_counts) <- rownames(count_df)
+c_counts <- count_df[, 2]; names(c_counts) <- rownames(count_df)
+
+# for observations
+records_by_fun <- readRDS("../data/final_obs_counts_df.RDS")
+pkgs_to_plot <- intersect(rownames(records_by_fun), names(r_counts))
+
+r_counts <- r_counts[pkgs_to_plot]
+c_counts <- c_counts[pkgs_to_plot]
+records_by_fun[ "pkg" ] <- rownames(records_by_fun) # needed for plotting
+
+records_by_fun <- records_by_fun[records_by_fun[,"pkg"] %in% pkgs_to_plot, ]
+
+top_10 <- c("Rcpp", "rlang", "glue", "tibble", "stringi", "ggplot2", "dplyr", "pillar", "R6", "stringr")
 
@@
 
+loc_of_top_10 <- lapply(1:length(combo_df$package_name), function(x) if (combo_df$package_name[[x]] %in% top_10) x else FALSE)
+
 # TODO: not staying sorted, probably id.vars
@@
   scale_y_continuous(name="Lines of Code in Package", breaks = c(-4, -3, -2, -1, 0, 1, 2, 3, 4), 
-                     labels = c(10000, 1000, 100, 10, 0, 10, 100, 1000, 10000))
+                     labels = c(10000, 1000, 100, 10, 0, 10, 100, 1000, 10000)) # +
+  # scale_x_discrete(name="Package", breaks = unlist(loc_of_top_10[!(loc_of_top_10 == FALSE)]), labels = top_10)
   # geom_line(aes( x, y), cut_top) +
@@
 
-Changing gears, lets get the number of recorded events by package.
-Final numbers will become available as the trace completes.
+Plot:
 ```{r}
 
-records_by_fun <- unlist(readRDS("../data/total_list_of_records_by_pkg.RDS"))
-records_by_fun_df <- data.frame(pname=names(records_by_fun), records=unname(records_by_fun)/1000)
-# records_by_fun_df <- records_by_fun_df[order(records_by_fun_df$records), ]
-# records_by_fun_df$pname <- factor(records_by_fun_df$pname, levels=records_by_fun_df$pname)
+records_by_fun$pkg <- factor(records_by_fun$pkg, levels=combo_df$package_name)
+# log scale?
+records_by_fun$invo_obs <- sapply(records_by_fun$invo_obs, function(x) if (x == 0) x else log10(x))
 
-plot_records <- ggplot(data=records_by_fun_df, aes(x=pname, y=records, group=1)) +
-                geom_line() + 
-                labs( title = "Numbers of Recorded Signatures",
-                      y = "Recorded Traces in a Package (Thousands)") +
+plot_records <- ggplot(data=records_by_fun, aes(x=pkg, y=invo_obs, group=1)) +
+                geom_line(size=0.25) + 
+                labs( title = "Numbers of Observed Signatures",
+                      y = "Observed Invocations in Pkg (Thousands, log10-Scaled)") +
                 theme(axis.text.x = element_blank(), axis.title.x=element_blank(),
-                      axis.ticks.x=element_blank())
+                      axis.ticks.x=element_blank()) +
+                scale_y_continuous(breaks = c(0, 1, 2, 3, 4, 5, 6), 
+                     labels = c(0, 10, 100, 1000, 10000, 100000, 1000000))
 

### 2019-03-26 notebooks/paper_results_section.Rmd
notebooks, new plots, data up to date. still debating if i should push it ... it big

@@
 # annoying strange error that I should fix pops up
+
+# There's at least one transformation that we're always gonna want to apply, namly the vector/scalar
+# collapse where both are present in a signature.
+los <- lapply(los, function(lofuns) {lapply(lofuns, combine_scalar_vector_where_appropriate)})
 ```
@@
   Type=c("Full Monomorphic", "Monomorphic in Type", "Monomorphic in Class", "Monomorphic in Attribute Pattern", "Total Seen"),
-  Count=c(total_mono_all, total_mono_type, total_mono_clas, total_mono_attr, total_arguments),
+  Count=round(c(total_mono_all, total_mono_type, total_mono_clas, total_mono_attr, total_arguments) / 1000, 0),
   Percentage=round(c(total_mono_all, total_mono_type, total_mono_clas, total_mono_attr, total_arguments) / total_arguments * 100, 2)
@@
   Type=c("Full Monomorphic", "Monomorphic in Type", "Monomorphic in Class", "Monomorphic in Attribute Pattern", "Total Seen"),
-  Count=c(total_mono_all_functions, total_mono_type_functions, total_mono_clas_functions, total_mono_attr_functions, total_functions),
+  Count=round(c(total_mono_all_functions, total_mono_type_functions, total_mono_clas_functions, total_mono_attr_functions, total_functions) / 1000, 0),
   Percentage=round(c(total_mono_all_functions, total_mono_type_functions, total_mono_clas_functions, total_mono_attr_functions, total_functions) / total_functions * 100, 2)
@@
 xtable(tot_fun_df)
-```
-
-We Have one thing we want to do first.
-Namely, let's get rid of vector/X, scalar/X polymorphism.
-```{r}
-los_vs <- lapply(los, function(lofuns) {lapply(lofuns, combine_scalar_vector_where_appropriate)})
 ```
@@
 format_df_for_print(poly_type_args[1:10,]) -> put_me
+put_me
 xtable(put_me)
@@
 Now, having noted that:
-(1) vector/X, scalar/X is common, and
-(2) ints and doubles are basically interchangable
-while noting that together these make up a huge portion of the polymorphism, we transform.
+(1) ints and doubles are basically interchangable, and
+(2) null and NA should be a part of every type,
+while also noting that together these make up a huge portion of the polymorphism, we transform.
 ```{r}
+los <- lapply(los, function(s) lapply(s, fold_NULL_and_NA_into_other_types_df))
+null_na_fold_arg_counts <- build_arg_count_df_from_lopkgs(los)
+null_na_fold_fun_counts <- build_fun_count_df_from_lopkgs(los)
 
+los <- lapply(los, function(lof) {lapply(lof, function(df) {change_type_systems_df(df, type_map_sv_to_real)})})
+null_na_fold_real_arg_counts <- build_arg_count_df_from_lopkgs(los)
+null_na_fold_real_fun_counts <- build_fun_count_df_from_lopkgs(los)
 ```

### 2019-03-26 paper/main.tex
data and playing with plots

@@
 \centering
-\begin{tabular}{rlrr}
-  \hline
- & Argument Polymorphism & Count & Percentage \\ 
-  \hline
-  1 & Full Monomorphic & 784K & 85.19 \\ 
-  2 & Monomorphic in Type & 839K & 91.27 \\ 
-  3 & Monomorphic in Class & 831K & 90.31 \\ 
-  4 & Monomorphic in Attribute Pattern & 830K & 90.25 \\ 
-  5 & Total Seen & 920K & --- \\ 
+\begin{tabular}{lrr}
+  \hline
+ Argument Polymorphism & Count & Percentage \\ 
+  \hline
+  Full Monomorphic & 784K & 85.19 \\ 
+  Monomorphic in Type & 839K & 91.27 \\ 
+  Monomorphic in Class & 831K & 90.31 \\ 
+  Monomorphic in Attribute Pattern & 830K & 90.25 \\ 
+  Total Seen & 920K & --- \\ 
      \hline
@@
 \centering
-\begin{tabular}{rlrr}
-  \hline
- & Function Polymorphism & Count & Percentage \\ 
-  \hline
-  1 & Full Monomorphic & 143K & 65.64 \\ 
-  2 & Monomorphic in Type & 169K & 77.65 \\ 
-  3 & Monomorphic in Class & 163K & 75.13 \\ 
-  4 & Monomorphic in Attribute Pattern & 161K & 74.31 \\ 
-  5 & Total Seen & 217K & --- \\ 
+\begin{tabular}{lrr}
+  \hline
+ Function Polymorphism & Count & Percentage \\ 
+  \hline
+  Full Monomorphic & 143K & 65.64 \\ 
+  Monomorphic in Type & 169K & 77.65 \\ 
+  Monomorphic in Class & 163K & 75.13 \\ 
+  Monomorphic in Attribute Pattern & 161K & 74.31 \\ 
+  Total Seen & 217K & --- \\ 
    \hline
@@
 The next type system design we will consider is ... .
-\AT{Generate tables. The stuff we do is: real, NULL fold, NA fold.}
+
+% latex table generated in R 3.5.1 by xtable 1.8-3 package
+% Fri Mar 22 18:03:30 2019
+\begin{table}[ht]
+\label{tab:argcountsWHATTS}
+\centering
+\begin{tabular}{lrr}
+  \hline
+ Argument Polymorphism & Count & Percentage \\ 
+  \hline
+  Full Monomorphic & 785K & 85.38 \\ 
+  Monomorphic in Type & 874K & 95.08 \\ 
+%  Monomorphic in Class & 831K & 90.31 \\ 
+%  Monomorphic in Attribute Pattern & 830K & 90.25 \\ 
+  Total Seen & 920K & --- \\ 
+     \hline
+\end{tabular}
+\caption{Account of {\it argument} polymorphism in the NEW TYPE SYSTEM.}
+\end{table}
+
+% latex table generated in R 3.5.1 by xtable 1.8-3 package
+% Fri Mar 22 18:10:46 2019
+\begin{table}[ht]
+\label{tab:funcountsWHATTS}
+\centering
+\begin{tabular}{lrr}
+  \hline
+ Function Polymorphism & Count & Percentage \\ 
+  \hline
+  Full Monomorphic & 143K & 65.95 \\ 
+  Monomorphic in Type & 186K & 85.85 \\ 
+%  Monomorphic in Class & 163K & 75.13 \\ 
+%  Monomorphic in Attribute Pattern & 161K & 74.31 \\ 
+  Total Seen & 217K & --- \\ 
+   \hline
+\end{tabular}
+\caption{Account of {\it function} polymorphism in the NEW TYPE SYSTEM.}
+\end{table}
+
+We had some big wins here.
+
+How do the top signatures shake out now?
 

### 2019-03-26 paper/main.tex
minor

--- 2019-03-26 paper/main.tex
@@
 
-In this section, we present our dataset. R is designed to accommodate data
-analysts.  Typical workflows go from data import, to data cleaning, followed
-by steps of modeling, transformation and visualization. Often, the code that
-implements these analysis pipelines together with their results resides in
-notebooks. Unfortunately notebooks are typically not shared in public
-repositories such as GitHub, and if they were, the data required to run them
-would surely not be shared. For this reason our analysis focuses on
-packages; these bundle reusable units of R code with documentation, sample
-data, use-cases and tests.
+In this section, we present our dataset. R aims to accommodate data
+analysts, their workflows start with data import, followed by cleaning, and
+then by steps of modeling, transformation and visualization. Often, the code
+of these analysis pipelines resides, together with the data and results, in
+notebooks. Few notebooks are publicly shared, and when they are, the data
+isn't. For this reason our
+analysis focuses on packages; these bundle reusable units of R code with
+documentation, sample data and use-cases.
 
@@
 purposes, packages must have data, examples, vignettes and tests, all of
-which must successfully run. From our perspective this means that each package 
-in CRAN comes with several executable scripts that exercise some of its functionality.
-Notable exceptions to this rule are packages only containing data, which have no runnable
-code but are referenced by other packages \AT{I should collect stats on this}.
+which must successfully run. From our perspective this means that each
+package in CRAN comes with several executable scripts that exercise some of
+its functionality.  Notable exceptions to this rule are packages only
+containing data, which have no runnable code but are referenced by other
+packages \AT{I should collect stats on this}.
 

### 2019-03-26 notebooks/corpus_stats_and_plots.Rmd

@@
 library(ggplot2)
+library(ggrepel)
 
 count_df <- readRDS("../data/final_loc_counts.RDS")
+
+total_recorded_sigs <- readRDS("../data/final_recorded_sigs_by_package.RDS")
 
@@
 c_counts <- c_counts[pkgs_to_plot]
+total_recorded_sigs <- total_recorded_sigs[pkgs_to_plot]
+
 records_by_fun[ "pkg" ] <- rownames(records_by_fun) # needed for plotting
-
 records_by_fun <- records_by_fun[records_by_fun[,"pkg"] %in% pkgs_to_plot, ]
+records_by_fun$invo_rec <- unlist(total_recorded_sigs)
 
@@
   scale_y_continuous(name="Lines of Code in Package", breaks = c(-4, -3, -2, -1, 0, 1, 2, 3, 4), 
-                     labels = c(10000, 1000, 100, 10, 0, 10, 100, 1000, 10000)) # +
-  # scale_x_discrete(name="Package", breaks = unlist(loc_of_top_10[!(loc_of_top_10 == FALSE)]), labels = top_10)
-  # geom_line(aes( x, y), cut_top) +
-  # geom_line(aes( x, y), cut_bot)
+                     labels = c(10000, 1000, 100, 10, 0, 10, 100, 1000, 10000)) +
+  geom_text_repel(data=subset(zzz, package_name %in% top_10 & language == "loc_r"),
+                  aes(y = 0, label=package_name), position=position_dodge(0.9), direction="y")
+  # trying to get some text to show up
 
@@
 # log scale?
-records_by_fun$invo_obs <- sapply(records_by_fun$invo_obs, function(x) if (x == 0) x else log10(x))
+# records_by_fun$invo_obs <- sapply(records_by_fun$invo_obs, function(x) if (x == 0) x else log10(x))
 
-plot_records <- ggplot(data=records_by_fun, aes(x=pkg, y=invo_obs, group=1)) +
-                geom_line(size=0.25) + 
+plot_records <- ggplot(data=records_by_fun, aes(x=invo_obs, y=invo_rec)) +
+                geom_point(size=0.25) + 
                 labs( title = "Numbers of Observed Signatures",
-                      y = "Observed Invocations in Pkg (Thousands, log10-Scaled)") +
+                      y = "Recorded Invocations in Pkg (Thousands)",
+                      x = "Observed Invocations in Pkg (Thousands)") +
                 theme(axis.text.x = element_blank(), axis.title.x=element_blank(),
-                      axis.ticks.x=element_blank()) +
-                scale_y_continuous(breaks = c(0, 1, 2, 3, 4, 5, 6), 
-                     labels = c(0, 10, 100, 1000, 10000, 100000, 1000000))
+                      axis.ticks.x=element_blank()) #+
+                # scale_y_continuous(breaks = c(0, 1, 2, 3, 4, 5, 6),
+                #      labels = c(0, 10, 100, 1000, 10000, 100000, 1000000))
 

### 2019-03-26 notebooks/corpus_stats_and_plots.Rmd
notebooks

@@
 # records_by_fun$invo_obs <- sapply(records_by_fun$invo_obs, function(x) if (x == 0) x else log10(x))
+# records_by_fun$invo_rec <- sapply(records_by_fun$invo_rec, function(x) if (x == 0) x else log10(x))
 
-plot_records <- ggplot(data=records_by_fun, aes(x=invo_obs, y=invo_rec)) +
+plot_records <- ggplot(data=records_by_fun, aes(x=invo_rec, y=invo_obs)) +
                 geom_point(size=0.25) + 
                 labs( title = "Numbers of Observed Signatures",
-                      y = "Recorded Invocations in Pkg (Thousands)",
-                      x = "Observed Invocations in Pkg (Thousands)") +
-                theme(axis.text.x = element_blank(), axis.title.x=element_blank(),
-                      axis.ticks.x=element_blank()) #+
+                      y = "Obs Invocations in Pkg",
+                      x = "Rec Invocations in Pkg")
+                # theme(axis.text.x = element_blank(), axis.title.x=element_blank(),
+                #       axis.ticks.x=element_blank()) #+
                 # scale_y_continuous(breaks = c(0, 1, 2, 3, 4, 5, 6),

### 2019-03-26 notebooks/corpus_stats_and_plots.Rmd

@@
 library(ggrepel)
+library(tidyverse)
 
-count_df <- readRDS("../data/final_loc_counts.RDS")
-
-total_recorded_sigs <- readRDS("../data/final_recorded_sigs_by_package.RDS")
+count_df <- read_rds("../data/final_loc_counts.RDS")
+total_recorded_sigs <- read_rds("../data/final_recorded_sigs_by_package.RDS")
 

### 2019-03-26 notebooks/corpus_stats_and_plots.Rmd

@@
 count_df <- read_rds("../data/final_loc_counts.RDS")
-total_recorded_sigs <- read_rds("../data/final_recorded_sigs_by_package.RDS")
-
-r_counts <- count_df[, 1]; names(r_counts) <- rownames(count_df)
-c_counts <- count_df[, 2]; names(c_counts) <- rownames(count_df)
+as.tibble(count_df) %>%
+  mutate(names = rownames(count_df)) -> cloc
 
 # for observations
-records_by_fun <- readRDS("../data/final_obs_counts_df.RDS")
-pkgs_to_plot <- intersect(rownames(records_by_fun), names(r_counts))
+t_r_s <- read_rds("../data/final_recorded_sigs_by_package.RDS")
+tibble(names=names(t_r_s), rec = unlist(t_r_s)) -> total_recorded_sigs
 
-r_counts <- r_counts[pkgs_to_plot]
-c_counts <- c_counts[pkgs_to_plot]
-total_recorded_sigs <- total_recorded_sigs[pkgs_to_plot]
+r_b_f <- read_rds("../data/final_obs_counts_df.RDS")
+as.tibble(r_b_f) %>%
+  mutate(names = row.names(records_by_fun)) %>%
+  rename(obs = invo_obs) %>%
+  select(names, obs) -> records_by_fun
 
-records_by_fun[ "pkg" ] <- rownames(records_by_fun) # needed for plotting
-records_by_fun <- records_by_fun[records_by_fun[,"pkg"] %in% pkgs_to_plot, ]
-records_by_fun$invo_rec <- unlist(total_recorded_sigs)
+total_recorded_sigs %>%
+  inner_join(records_by_fun, by="names") %>%
+  filter(obs != 0, rec !=0) %>% 
+  inner_join(cloc, by="names") %>%
+  filter(r_counts != 0) -> df
 
 top_10 <- c("Rcpp", "rlang", "glue", "tibble", "stringi", "ggplot2", "dplyr", "pillar", "R6", "stringr")
-
 # just named lists. some elements (in either) are integer(0) i believe, as some packages
@@
 ```{r}
-r_c_df <- data.frame(package_name = names(r_counts), loc_r = r_counts)
-# looks weird sorted
-r_c_df_sub <- r_c_df[order(r_c_df$loc_r),]
-# just get the entries with > 0 lines of code
-r_c_df_sub <- r_c_df_sub[r_c_df_sub$loc_r > 0, ]
-# looks weirder log10-scaled
-r_c_df_sub$loc_r <- log10(r_c_df_sub$loc_r)
+df %>%
+  arrange(r_counts) %>%
+  mutate(names_o = factor(names, names) , r_c_log= log10(r_counts), c_c_log = ifelse((c_counts==0),0, log10(c_counts)))  -> df_rc
 
-plot_r_counts <- ggplot( data=r_c_df_sub, aes(x=1:length(r_c_df_sub$loc_r), y=loc_r, group=1))
-plot_r_counts <- plot_r_counts + geom_line() + geom_point()
-plot_r_counts <- plot_r_counts + labs(
-  title = "Analysis of R Corpus (R Code Portion)",
+df_rc %>% 
+  ggplot(aes(x=names_o, y=r_c_log, group=1)) +
+  geom_line() + geom_point() + labs(title = "Analysis of R Corpus (R Code Portion)",
     x = "Corpus of Analyzed Packages", 
-  y = "Lines of R Code in Package (Log10-scaled)"
-)
-
-plot_r_counts <- plot_r_counts + theme(axis.text.x = element_blank())
-
-# plot it
-plot_r_counts
-
+    y = "Lines of R Code in Package (Log10-scaled)") + theme(axis.text.x = element_blank())
 ```
 
-The outlier is the spatstat package.
-How about for C?
+```{r}
 
-```{r}
-library(ggplot2)
-
-c_c_df <- data.frame(package_name = names(c_counts), loc_c = c_counts)
-# looks weird sorted
-c_c_df_sub <- c_c_df[order(c_c_df$loc_c),]
-# get rid of 0s
-c_c_df_sub <- c_c_df_sub[c_c_df_sub$loc_c > 0, ]
-# looks weirder log10-scaled
-c_c_df_sub$loc_c <- log10(c_c_df_sub$loc_c)
-
-plot_c_counts <- ggplot( data=c_c_df_sub, aes(x=1:length(c_c_df_sub$loc_c), y=loc_c, group=1))
-plot_c_counts <- plot_c_counts + geom_line() + geom_point()
-plot_c_counts <- plot_c_counts + labs(
-  title = "Analysis of R Corpus (C Code Portion)",
+df_rc %>% 
+  ggplot(aes(x=names_o, y=c_c_log, group=1)) +
+  geom_line() + geom_point() + labs(title = "Analysis of R Corpus (C Code Portion)",
     x = "Corpus of Analyzed Packages", 
-  y = "Lines of C Code in Package (Log10-scaled)"
-)
-
-plot_c_counts <- plot_c_counts + theme(axis.text.x = element_blank())
-
-# plot it
-plot_c_counts
-
+    y = "Lines of C Code in Package (Log10-scaled)") + theme(axis.text.x = element_blank())
 ```
@@
 ```{r}
-combo_df <- r_c_df
-combo_df$loc_c <- c_c_df$loc_c
-combo_df$loc_r <- sapply(combo_df$loc_r, function(x) if (x == 0) x else log10(x))
-combo_df$loc_c <- - sapply(combo_df$loc_c, function(x) if (x == 0) x else log10(x))
-
-# sort
-combo_df <- combo_df[order(combo_df$loc_r), ]
-# this makes sure the package_names are in the right order, ggplot2 plots according to factors
-# and levels
-combo_df$package_name <- factor(combo_df$package_name, levels=combo_df$package_name)
-
 loc_of_top_10 <- lapply(1:length(combo_df$package_name), function(x) if (combo_df$package_name[[x]] %in% top_10) x else FALSE)
 
-# TODO: not staying sorted, probably id.vars
-zzz <- melt(combo_df, id.vars="package_name", variable.name = "language")
-# zzz_c <- melt(combo_df_c, id.vars="package_name", variable.name = "language")
+df_rc %>% 
+  select(names_o, r_c_log, c_c_log) %>% 
+  mutate(c_c_log=-c_c_log) %>%
+  melt(id.vars="names_o",variable.name="language") -> zzz
 
-plot_both <- ggplot( data=zzz, aes(package_name, value, group=1))
-plot_both <- plot_both + geom_line(aes(colour = language), size=0.05) # +  geom_point(aes(colour = language))
-plot_both <- plot_both + labs(
-  title = "Analysis of R Corpus (R and C)",
-  y = "Lines of Code in Package (Log10-scaled)"
-) + guides(colour=F)
+zzz %>%
+ggplot(aes(x=names_o, value, group=1)) +
+ geom_line(aes(colour = language), size=0.05) +
+ labs(title = "Analysis of R Corpus (R and C)", y = "Lines of Code in Package (Log10-scaled)") + guides(colour=F) + 
+ theme( axis.text.x  = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(),
+        axis.ticks.y = element_line(),  panel.grid.major.y = element_line(colour = "red")) + # ,
+  scale_y_continuous(name="Lines of Code in Package", breaks = c(-4, -3, -2, -1, 0, 1, 2, 3, 4), 
+                     labels = c(10000, 1000, 100, 10, 0, 10, 100, 1000, 10000))
 
-# cut_top <- data.frame( x = c(-Inf, Inf), y = 2, cutoff = factor(2) )
-# cut_bot <- data.frame( x = c(-Inf, Inf), y = -2, cutoff = factor(-2) )
+#+  geom_text_repel(data=subset(zzz, names %in% top_10 & language == "loc_r"), 
+#                  aes(y = 0, label=names), position=position_dodge(0.9), direction="y")
 
-plot_both <- plot_both + theme(
-  axis.text.x  = element_blank(), 
-  axis.title.x = element_blank(),
-  axis.ticks.x = element_blank(),
-  axis.ticks.y = element_line(), 
-  panel.grid.major.y = element_line(colour = "red")) + # ,
-  # panel.ontop = T) + 
-  scale_y_continuous(name="Lines of Code in Package", breaks = c(-4, -3, -2, -1, 0, 1, 2, 3, 4), 
-                     labels = c(10000, 1000, 100, 10, 0, 10, 100, 1000, 10000)) +
-  geom_text_repel(data=subset(zzz, package_name %in% top_10 & language == "loc_r"),
-                  aes(y = 0, label=package_name), position=position_dodge(0.9), direction="y")
-  # trying to get some text to show up
-
-# plot it
-plot_both
-
-ggsave("../paper/plots/linesofrandccode.jpg", plot_both)
-
+ggsave("../paper/plots/linesofrandccode.pdf")
 ```
@@
 ```{r}
-
-records_by_fun$pkg <- factor(records_by_fun$pkg, levels=combo_df$package_name)
-# log scale?
-# records_by_fun$invo_obs <- sapply(records_by_fun$invo_obs, function(x) if (x == 0) x else log10(x))
-# records_by_fun$invo_rec <- sapply(records_by_fun$invo_rec, function(x) if (x == 0) x else log10(x))
-
-plot_records <- ggplot(data=records_by_fun, aes(x=invo_rec, y=invo_obs)) +
-                geom_point(size=0.25) + 
+df_rc %>%
+  mutate(rec=log(rec),obs=log(obs))  %>%
+  ggplot(aes(x=rec, y=obs)) +
+                geom_point(size=0.1) + 
                 labs( title = "Numbers of Observed Signatures",
@@
 
-# plot
-plot_records
-
 # generate file
-ggsave("../paper/plots/recordsbypkg.jpg", plot_records)
+ggsave("../paper/plots/recordsbypkg.pdf")
 ```

### 2019-03-26 notebooks/corpus_stats_and_plots.Rmd
code

@@
 ggplot(aes(x=names_o, value, group=1)) +
- geom_line(aes(colour = language), size=0.05) +
+ geom_line(aes(colour = language), size=0.008) +
  labs(title = "Analysis of R Corpus (R and C)", y = "Lines of Code in Package (Log10-scaled)") + guides(colour=F) + 

### 2019-03-26 notebooks/corpus_stats_and_plots.Rmd
added filter to plot, updated some counts

--- 2019-03-26 notebooks/corpus_stats_and_plots.Rmd
@@
 as.tibble(r_b_f) %>%
-  mutate(names = row.names(records_by_fun)) %>%
+  mutate(names = row.names(r_b_f)) %>%
   rename(obs = invo_obs) %>%
@@
   inner_join(cloc, by="names") %>%
+  filter(rec <= obs) %>%                      # Added this
   filter(r_counts != 0) -> df

### 2019-03-26 notebooks/paper_results_section.Rmd
data and playing with plots
notebooks
added filter to plot, updated some counts

@@
 # argument counts
-total_arguments <- apply_count_to_lopkg_c(los, args_count_all)
-total_mono_type <- apply_count_to_lopkg_c(los, args_count_all_mono_type)
-total_mono_attr <- apply_count_to_lopkg_c(los, args_count_all_mono_attr_pattern)
-total_mono_clas <- apply_count_to_lopkg_c(los, args_count_all_mono_class)
-total_mono_all <- apply_count_to_lopkg_c(los, args_count_all_mono_all)
-```
+c_basic_df <- build_arg_count_df_from_lopkgs(los)
 
-```{r}
-tot_arg_df <- data.frame(
-  Type=c("Full Monomorphic", "Monomorphic in Type", "Monomorphic in Class", "Monomorphic in Attribute Pattern", "Total Seen"),
-  Count=round(c(total_mono_all, total_mono_type, total_mono_clas, total_mono_attr, total_arguments) / 1000, 0),
-  Percentage=round(c(total_mono_all, total_mono_type, total_mono_clas, total_mono_attr, total_arguments) / total_arguments * 100, 2)
-)
-
-xtable(tot_arg_df)
+xtable(c_basic_df)
 ```
@@
 # function counts
-total_functions <- apply_count_to_lopkg_c(los, length)
-total_mono_type_functions <- apply_count_to_lopkg_c(los, fun_count_all_mono_type)
-total_mono_attr_functions <- apply_count_to_lopkg_c(los, fun_count_all_mono_attr_pattern)
-total_mono_clas_functions <- apply_count_to_lopkg_c(los, fun_count_all_mono_class)
-total_mono_all_functions <- apply_count_to_lopkg_c(los, fun_count_all_mono_all)
-```
+c_f_basic_df <- build_fun_count_df_from_lopkgs(los)
 
-```{r}
-tot_fun_df <- data.frame(
-  Type=c("Full Monomorphic", "Monomorphic in Type", "Monomorphic in Class", "Monomorphic in Attribute Pattern", "Total Seen"),
-  Count=round(c(total_mono_all_functions, total_mono_type_functions, total_mono_clas_functions, total_mono_attr_functions, total_functions) / 1000, 0),
-  Percentage=round(c(total_mono_all_functions, total_mono_type_functions, total_mono_clas_functions, total_mono_attr_functions, total_functions) / total_functions * 100, 2)
-)
-
-xtable(tot_fun_df)
+xtable(c_f_basic_df)
 ```
@@
 ```
+
+Generate...
+```{r}
+xtable(null_na_fold_real_arg_counts)
+xtable(null_na_fold_real_fun_counts)
+```
+
+Top sig time.
+```{r}
+poly_type_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_type_args))
+poly_type_args <- poly_type_args[order(-poly_type_args$count),] # sort by count
+poly_type_args$perc <- round(poly_type_args$count/sum(poly_type_args$count) * 100, 2)
+```
+
+Print top 10.
+```{r}
+format_df_for_print(poly_type_args[1:10,]) -> put_me
+put_me
+xtable(put_me)
+```

### 2019-03-27 notebooks/paper_results_section.Rmd
updated counts w matrix

@@
 los <- lapply(los, function(lofuns) {lapply(lofuns, combine_scalar_vector_where_appropriate)})
+# Figure out which ones are matrices. Remove if the analysis picks this up (and we rerun).
+los_m <- lapply(los, function(lofun) lapply(lofun, distinguish_matrices_df))
 ```
@@
 # argument counts
-c_basic_df <- build_arg_count_df_from_lopkgs(los)
+c_basic_df <- build_arg_count_df_from_lopkgs(los_m)
 
@@
 ```{r}
-xtable(null_na_fold_real_arg_counts)
-xtable(null_na_fold_real_fun_counts)
+xtable(null_na_fold_real_arg_counts) %>% print(include.rownames=F)
+xtable(null_na_fold_real_fun_counts) %>% print(include.rownames=F)
 ```
@@
 put_me
-xtable(put_me)
+xtable(put_me) %>% print(include.rownames=F)
 ```
+
+I think something funky is happening.
+We should have matrix types, maybe I forgot.
+Dealing with that now:
+```{r}
+# NOTE matrix stuff moved up, oopsie
+```
+
+Top sig time.
+```{r}
+# poly_type_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_type_args))
+# poly_type_args <- poly_type_args[order(-poly_type_args$count),] # sort by count
+# poly_type_args$perc <- round(poly_type_args$count/sum(poly_type_args$count) * 100, 2)
+```
+
+Print top 10.
+```{r}
+# format_df_for_print(poly_type_args[1:10,]) -> put_me
+# put_me
+# xtable(put_me) %>% print(include.rownames=F)
+```
+
+```{r}
+# sigs_df_vec <- get_fun_sigs_matching_pred_from_lopkg(los, pred=does_df_have_X_types, types_to_find=c("matrix", "vector/real"), strict=T)
+```

### 2019-03-27 paper/main.tex
minor

@@
 notebooks. Few notebooks are publicly shared, and when they are, the data
-isn't. For this reason our
-analysis focuses on packages; these bundle reusable units of R code with
-documentation, sample data and use-cases.
+isn't. For this reason our analysis focuses on packages which bundle
+reusable units of R code with documentation, sample data and use-cases.
 
@@
 \begin{center}
-\includegraphics[scale=0.15]{linesofrandccode.jpg}
+\includegraphics[scale=0.6]{linesofrandccode FILE MISSING IN GIT}
 \caption{Lines of code in the analyzed corpus. \# of lines of R code are above 0, \# of lines of C code are below 0.}
@@
 \begin{center}
-\includegraphics[scale=0.15]{recordsbypkg}
+\includegraphics[scale=0.1]{recordsbypkg}
 \caption{Numbers of Recorded Function Invocations in the Analyzed Corpus}
@@
 
+In Figure~\ref{fig:matvec}, we see that the function can take in a vector, but immediately transforms it (transpose, with {\tt t}) into a matrix.
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+tee <- function (x, theta, D1, D2, phi) 
+{
+    if (is.vector(x)) {
+        x <- t(x)
+    }
+    ...
+\end{lstlisting}}\caption{Example of matrix/(vector/double) argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}
+
 %
@@
   \hline
-  Full Monomorphic & 784K & 85.19 \\ 
-  Monomorphic in Type & 839K & 91.27 \\ 
+  Full Monomorphic & 784K & 85.28 \\ 
+  Monomorphic in Type & 836K & 90.91 \\ 
   Monomorphic in Class & 831K & 90.31 \\ 
@@
   \hline
-  Full Monomorphic & 143K & 65.64 \\ 
-  Monomorphic in Type & 169K & 77.65 \\ 
+  Full Monomorphic & 143K & 65.80 \\ 
+  Monomorphic in Type & 167K & 76.84 \\ 
   Monomorphic in Class & 163K & 75.13 \\ 
@@
   \hline
-vector/double, vector/integer & 10,316 & 12.85 \\ 
-scalar/double, scalar/integer & 7,194 & 8.96 \\ 
-NULL, vector/double & 3,087 & 3.85 \\ 
-data.frame, vector/double & 2,604 & 3.24 \\ 
-NULL, scalar/character & 2,333 & 2.91 \\ 
-list$<$any$>$, list$<$double$>$ & 2,224 & 2.77 \\ 
-NULL, scalar/double & 2,066 & 2.57 \\ 
-raw\_NA, vector/double & 1,465 & 1.83 \\ 
-list$<$any$>$, list$<$list$>$ & 1,426 & 1.78 \\ 
-NULL, vector/character & 1,385 & 1.73 \\ 
+  vector/double, vector/integer & 7677 & 9.18 \\ 
+  scalar/double, scalar/integer & 7194 & 8.61 \\ 
+  matrix, vector/double & 5354 & 6.40 \\ 
+  NULL, scalar/character & 2333 & 2.79 \\ 
+  list$<$any$>$, list$<$double$>$ & 2224 & 2.66 \\ 
+  data.frame, matrix & 2115 & 2.53 \\ 
+  NULL, scalar/double & 2066 & 2.47 \\ 
+  NULL, vector/double & 2042 & 2.44 \\ 
+  list$<$any$>$, list$<$list$>$ & 1426 & 1.71 \\ 
+  NULL, vector/character & 1385 & 1.66 \\ 
      \hline
@@
 The next type system design we will consider is ... .
+The number of monomorphic arguments and functions can be found in Tables~\ref{tab:argcountsWHATTS} and~\ref{tab:funcountsWHATTS} respectively.
 
@@
   Full Monomorphic & 785K & 85.38 \\ 
-  Monomorphic in Type & 874K & 95.08 \\ 
+  Monomorphic in Type & 871K & 94.65 \\ 
 %  Monomorphic in Class & 831K & 90.31 \\ 
@@
   Full Monomorphic & 143K & 65.95 \\ 
-  Monomorphic in Type & 186K & 85.85 \\ 
+  Monomorphic in Type & 184K & 84.778 \\
 %  Monomorphic in Class & 163K & 75.13 \\ 
@@
 
-We had some big wins here.
-
-How do the top signatures shake out now?
+We had some big wins here, with only ~5\% of argument polymorphism unaccounted for.
+To get an idea of what more we can do, let's see what the most frequent remaining polymorphic signatures are.
+See Table~\ref{tab:toppolyWHATTS}.
+
+% latex table generated in R 3.5.1 by xtable 1.8-3 package
+% Tue Mar 26 17:46:45 2019
+\begin{table}[ht]
+\label{tab:toppolyWHATTS}
+\centering
+\begin{tabular}{lrr}
+  \hline
+Polymorphic Type Signature & \# Occurrences & \% of Total Polymorphism \\ 
+  \hline
+  matrix, vector/real & 6307 & 12.82 \\ 
+  list$<$any$>$, list$<$real$>$ & 2792 & 5.68 \\ 
+  vector/character, vector/real & 2719 & 5.53 \\ 
+  scalar/character, scalar/real & 2536 & 5.16 \\ 
+  data.frame, matrix & 2163 & 4.40 \\ 
+  list$<$any$>$, list$<$list$>$ & 1492 & 3.03 \\ 
+  list$<$any$>$, list$<$character$>$ & 1217 & 2.47 \\ 
+  data.frame, vector/real & 1025 & 2.08 \\ 
+  scalar/logical, scalar/real & 917 & 1.86 \\ 
+  scalar/character, vector/real & 844 & 1.72 \\ 
+   \hline
+\end{tabular}
+\caption{Top 10 argument signatures in the NEW TYPE SYSTEM.}
+\end{table}
+
+What's up with matrix, vector/real?
 

### 2019-03-27 paper/main.tex
minor
typos, behaviour->behavior, updated corpus counts

@@
 
-\newcommand{\PACKAGES}{7,763\xspace}
-\newcommand{\PROGRAMMERS}{10,000\xspace}
-\newcommand{\PERCENTCRAN}{56\%\xspace}
+\newcommand{\PACKAGES}{11,463\xspace}
+\newcommand{\PROGRAMMERS}{?\xspace}
+\newcommand{\PERCENTCRAN}{83\%\xspace}
 \newcommand{\CRANTOTAL}{13,841\xspace}
-\newcommand{\RLOC}{8,996,572\xspace}
-\newcommand{\CLOC}{5,100,719\xspace}
-\newcommand{\YEARS}{20\xspace}
+\newcommand{\RLOC}{15,050,267\xspace}
+\newcommand{\CLOC}{9,373,542\xspace}
+\newcommand{\YEARS}{?\xspace}
 
@@
 
-R and it's ancestor S are languages designed, implemented and maintained by
+R and its ancestor S are languages designed, implemented and maintained by
 statisticians. Originally they were designed as glue languages, languages
@@
 \begin{center}
-\includegraphics[scale=0.6]{linesofrandccode FILE MISSING IN GIT}
+\includegraphics[scale=0.6]{linesofrandccode}
 \caption{Lines of code in the analyzed corpus. \# of lines of R code are above 0, \# of lines of C code are below 0.}
@@
 \begin{center}
-\includegraphics[scale=0.1]{recordsbypkg}
+\includegraphics[scale=0.6]{recordsbypkg}
 \caption{Numbers of Recorded Function Invocations in the Analyzed Corpus}
@@
 These naturally-occurring attribute patterns and classes can be separately
-accounted for to highlight user-defined patterns and behaviours.  This will
+accounted for to highlight user-defined patterns and behaviors.  This will
 be made clear in Section~\ref{sec:results}.
@@
 In Figure~\ref{fig:matvec}, we see that the function can take in a vector, but immediately transforms it (transpose, with {\tt t}) into a matrix.
+As an idea, matrix/vector polymorphism seems sensible, as mathematically vectors are matrices.
+R echoes this by making conversion between the two ``types'' easy:
+\code{as.vector(m)} flattens a matrix \code{m} into a vector (e.g., \code{as.vector(matrix(2, 2, 2)) == c(2, 2, 2, 2)}), and \code{as.matrix(v)} builds a {\tt length(v) x 1} matrix (e.g., \code{as.matrix(c(1, 2)) == matrix(data=c(1, 2))}).
+\AT{TODO fix line break in code at end of last sentence}
 
 \begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-tee <- function (x, theta, D1, D2, phi) 
-{
+tee <- function (x, theta, D1, D2, phi)  {
     if (is.vector(x)) {
@@
 
-In terms of programmer behaviour, we observe a pattern of optional arguments throughout several R packages, wherein programmers specify a default argument value which is sometimes NULL.
+In terms of programmer behavior, we observe a pattern of optional arguments throughout several R packages, wherein programmers specify a default argument value which is sometimes NULL.
 Consider the snippet in Figure~\ref{fig:optnull} for an example.

### 2019-03-28 notebooks/paper_results_section.Rmd
small changes, additions, new functions

@@
 # Figure out which ones are matrices. Remove if the analysis picks this up (and we rerun).
-los_m <- lapply(los, function(lofun) lapply(lofun, distinguish_matrices_df))
+los <- lapply(los, function(lofun) lapply(lofun, distinguish_matrices_df))
 ```
@@
 # argument counts
-c_basic_df <- build_arg_count_df_from_lopkgs(los_m)
+c_basic_df <- build_arg_count_df_from_lopkgs(los)
 
-xtable(c_basic_df)
+xtable(c_basic_df) %>% print(include.rownames=F)
 ```
@@
 
-xtable(c_f_basic_df)
+xtable(c_f_basic_df) %>% print(include.rownames=F)
 ```
@@
 put_me
-xtable(put_me)
+xtable(put_me) %>% print(include.rownames=F)
 ```

### 2019-03-28 paper/main.tex
small changes, additions, new functions

@@
 \newcommand{\YEARS}{?\xspace}
+\newcommand{\INDEXCOINCIDENCE}{3,426\xspace}
+\newcommand{\TOTALINDEXY}{6,461\xspace}
+\newcommand{\INDEXYPERC}{53\%\xspace}		% TODO Not an accurate count.
 
@@
 
-What's up with matrix, vector/real?
+A few of the top polymorphic signatures stand out.
+First, the matrix/(vector/real) signature is interesting: \AT{as we showed in our polymorphism examples section}, converting between vectors and matrices is very simple, and indeed numeric vectors are mathematically {\tt 1 x n} matrices.
+In fact, in R matrices can be used identically to vectors in many cases, and we refer the reader to Figure~\ref{fig:matasvec} for examples.
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+v <- c(1, 2, 3, 4)
+ma <- matrix(data=v, nrow=4, ncol=1)
+v[2] == m[2] == m[2, 1]
+
+mb <- matrix(data=v, nrow=2, ncol=2)
+v[3] == m[3] == m[1, 2] 
+\end{lstlisting}}\caption{Using a matrix like a vector.}\label{fig:matasvec}\end{figure}
+
+\AT{We should figure out how we want to deal with this polymorphism.}
+
+Another significant source of apparently strange polymorphism is the {\tt character}, {\tt real} polymorphism.
+Recall Figure~\ref{fig:chardbl}, which laid out an example of this polymorphic signature: lists and vectors can be accessed by strings as well as by numerics.
+This is a somewhat unusual example of polymorphism, and our argument depends on there being perhaps a list or vector passed alongside the {\tt character}- or {\tt real}-typed argument.
+So, just how often is this the case?
+Well, as it happens, character/real polymorphism coincides with some list-flavored type in \INDEXCOINCIDENCE instances (\INDEXYPERC of occurrences).
+
+Another option is a date.
+In R, dates are often represented as either strings or numbers, and many packages make use of dates.
+
+\AT{We might want to discuss some of the most polymorphic functions in the language.}
+\subsection{Most Polymorphic Functions}
+
+\AT{Draft}
+
+The most polymorphic function in the corpus of code we analyzed is {\tt magrittr}'s {\tt \%>\%} (pipe) function.
+{\tt magrittr} is part of the Tidyverse, and the pipe function is frequently used to chain function calls.
+The fact that the pipe is the most polymorphic function is not shocking, as its goal is simply to take the right hand argument and apply it to the left hand argument, and this in a sense irrespective of type.
 

### 2019-03-28 paper/main.tex
small code updates to deal with failed traces, and light reformatting of paper, writing a bit about function level polymorphism

@@
 
+\AT{Do we want to talk about how sometimes we can't eval an argument, even if a function executes successfully?
+Like pipe from magrittr.}
+
 We will need a notion of polymorphism when undertaking this analysis. R
@@
 
-The first question we will address is: {\it how polymorphic is R?}  To
-answer that, we will first look at the number of monomorphic arguments
-across all of the packages we analyzed.  The data can be found in
-Tables~\ref{tab:argcountsFGTS}~and~\ref{tab:funcountsFGTS}.
+Broadly, the first question we will address is: {\it how polymorphic is R?} 
+We will answer this in two parts, and first we will look at the number of monomorphic {\it arguments} across all of the packages we analyzed.
+
+%
+%
+%
+%
+\subsection{Argument Polymorphism}
+
+In this section, we will explore the polymorphicity of function argument positions and return values.
+Exploring argument polymorphism will help to motivate a design for whole function signatures.
+
+To begin, let's look at the {\it amount} of polymorphism by way of {\it number of polymorphic arguments}.
+The data can be found in Tables~\ref{tab:argcountsFGTS}~and~\ref{tab:funcountsFGTS}.
 \AT{TODO: have these tables be generated automatically}
-\AT{TODO: put all counts in one table?}
 
@@
 In R, dates are often represented as either strings or numbers, and many packages make use of dates.
-
+Dates (sometimes) manifest themselves in the {\tt class} of a value, for examples refer to Figure~\ref{fig:date}.
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+class(date()) # => character
+class(Sys.Date()) # => Date
+class(as.POSIXlt(Sys.Date())) # => POSIXlt, POSIXt
+as.integer(as.Date(1, origin = "1900-01-01"))
+\end{lstlisting}}\caption{Dates in R.}\label{fig:date}\end{figure}
+
+This figure reveals an unfortunate truth that date ``data type'' in R is not standardized.
+That said, translating between the data types is straigthforward.
+
+%
+%
+%
+%
 \AT{We might want to discuss some of the most polymorphic functions in the language.}
@@
 
-%
-%
-%
-%
-\subsection{Types in R}
-
-\AT{I'll pull from this section when appropriate, as the discussion of the type systems becomes relevant.}
-
-%
-%
-\subsubsection{What are possible type systems?}
-
-Great question.
-Some things:
-
-\begin{itemize}
-
-\item {\it The Obvious Type System (OTS)}: this type system corresponds directly with the type information that one can fetch using the {\tt typeof}, {\tt attributes}, and {\tt class} functions.
-In this system, there is no distinction between vectors and scalars, and lists of differently-typed elements (e.g., a list of integers vs.
-a list of strings), and there is a distinction between integers and doubles, and the type of NULL and all other types.
-This type system isn't particularly informative to a programmer, though it does reflect the treatment of values in the R runtime, for the {\tt typeof} a value is the type that the value has according to the R internals.
-
-\item {\it The Fine-Grained Type System (FGTS)}: this type systems includes a number of relevant additions.
-Chiefly, it distinguishes between vectors and scalars, as it is conceivable that boxing scalars as single-element vectors might be overkill.
-In addition, list types are parameterized over the types of their elements (e.g., list<integer> vs. list<character>).
-Finally, NAs ``in the wild'' are given the type {\tt raw\_NA}, as {\tt typeof(NA) = logical} in R.
-
-\item {\it The Null-and-NA-Free Type System (NNTS)}: most languages consider NULLs to inhabit all types, and R does so for NAs \AT{elaborate}.
-This gives us ... .
-
-\end{itemize}
-
-Of course, one could conceive of a more fine-grained type system than FGTS, but it's unclear that we would gain much in doing so.
-Indeed, there is a space to explore in slightly less fine-grained type systems, which make modifications including:
-
-\begin{itemize}
-
-\item {\bf collapsing vectors and scalars}: this yields a type system slightly closer to OTS.
-
-\item {\bf rolling NULL into other types}: in most languages, NULL is an inhabitant of all types, and this adjustment seems interesting.
-Note that this goes beyond even OTS, as in R {\tt typeof(NULL) = NULL}.
-
-\item {\bf rolling NA into other types}: \AT{i should try this}
-similarly, NA is an inhabitant of all types, though NA standing alone has type {\tt logical}.
-FGTS distinguishes such NAs with the {\tt raw\_NA} type, and if instead we consider NA to inhabit all types we may \AT{win}.
-
-\item {\bf combining integers and doubles}: as in most languages, doubles and ints can stand-in for each other in nearly every scenario.
-Even if not, it is trivial to convert from one to the other, particularly converting ints to doubles.
-Aside, in R a double can be used to index a list, and R will floor it when performing the lookup.
-
-\end{itemize}
-
-%
-%
-%
-%
-\subsection{Usage Patterns}
-
-In this section, we will discuss function usage patterns which arose in our analysis.
-We will start by looking at the morphicity of functions.
-
-\subsubsection{Function Argument and Return Morphicity}
-
-First and foremost, we would like to know how often R programmers create polymorphic functions.
-Recall that we define a polymorphic function to be a function with at least one polymorphic argument, or a polymorphic return.
-We will turn our attention now to the data in \todo{Figure}, and go through each entry in the table.
-
-Here, we see that the vast majority of function arguments are indeed monomorphic.
-Monomorphic arguments are easy to annotate, as the {\tt typeof} the arguments is exactly the most precise annotation we could give, at least in terms of \textit{type}.
-That said, types alone don't always paint the whole picture:
-recall that \textit{attributes} are an R language feature which allows programmers to stick metadata onto values.
-So in which circumstances \textit{do} types paint the whole picture?
-
-The second data point in \textbf{TODO FIGURE} indicates that a
-\isit{majority} of function arguments are monomorphic in type \textit{and}
-have no attributes.  These represent arguments which are truly trivial to
-annotate, as the {\tt typeof} an argument perfectly describes the usage of
-that argument.  That said, some attributes arise naturally in R: For
-instance, names in a named list (e.g. {\tt x} and {\tt y} in {\tt list(x=0,
-  y=0)}) appear as an attribute on the value.  In
-Section~\ref{sec:method:attributes}, we outlined these naturally-occurring
-attribute patterns, and the third data point in \textbf{TODO FIGURE} shows
-that a nontrivial amount of functions are monomorphic in type with said
-natural attribute patterns.
-
-Another facet of type information in R is in the {\tt class} attribute.
-Recall that values have a \textit{class} in addition to a type: For example,
-a {\tt data.frame} has type {\tt list} and class {\tt data.frame}.
-\todo{Plug classes earlier} R has a number of built-in classes, such as
-\todo{X, Y, and Z}, but users are free to redefine the class of any value at
-runtime, and easily define new classes.  The next data point in
-\todo{Figure} shows that user-defined classes don't appear altogether often,
-though they do indeed feature.  \AT{In a following section, we will discuss
-  how the usage of these classes manifests itself.}
-
-\AT{Separate section for functions? Right now, this is at argument granularity.}
-
-\subsubsection{Type Signatures}
-
-In the last section, we presented a high-level overview of the morphicity ... .
-
-\subsubsection{Attribute Signatures}
-
-Recall that attributes are a way for programmers to store metadata on values in R.
-What are the common attribute patterns?
-And how often is the attribute pattern polymorphic?
-
-\subsubsection{Takeaways}
-
-\begin{itemize}
-    \item the vast majority of arguments are monomorphic in type;
-    \item of those, over 60\% have no attribute information;
-    \item of the 40\% with attribute information, roughly 1/2 have fairly
-      simple attributes corresponding to base R constructs (named lists and
-      vectors, matrices, and data frames);
-    \item now, of those arguments which are polymorphic, a sizable chunk (well over half) have defensible signatures (e.g., double and character for named list indexing, double and integer for obvious reasons, etc.).
-\end{itemize}
-
-In short, it looks like R (package) programmers are reasonable.  I'd
-conjecture that a lot of the polymorphism (e.g., double and list) is coming
-from how easy it is to use either type in a given situation (e.g.,
-converting from vector of doubles to list or vice versa is simple).
-
-%
-%
-%
-%
-%
-%
-\section{Synthesis}
-
-In this section, we will discuss the conclusions that we draw from our data.
-
-%
-%
-\subsection{Suggested Annotations}
-
-\AT{How can we capture these patterns with annotations?}
-
-Some possible annotations:
-
-\begin{itemize}
-    \item \textit{real}: for \textit{double} and \textit{integer} values
-    \item \textit{function}: for \textit{closure}, \textit{special}, and \textit{builtin} values
-    \item \textit{vector}: to indicate that something should be vectorized
-    \item \textit{scalar}: to indicate that something should \textbf{not} be vectorized
-    \item \textit{index}: for \textit{real} and \textit{character} values
-\end{itemize}
-
-%
-%
-\subsubsection{Struct-Like Attribute Declarations}
-
-\AT{What's a convenient way to annotate attributes?  Should investigate how
-  often attributes are consistent.}
-
-%
-\subsubsection{Coverage}
+...
+
+%
+%
+%
+%
+\subsection{Function Polymorphism}
+
+Now, we will turn our attention to whole function polymorphism.
+
+This is different because we can now look at argument polymorphism across several arguments, like we did when thinking about character/real polymorphism.
+
+We're interested to see what kinds of ``meta-annotations'' we can make at the function level.
+For instance, should we link argument types with return types?
+Consider Figure~\ref{fig:funanno}.
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+iden <- function(x) x;
+
+# observed calls:
+f(2); # x: double -> double
+f(list(1, 2, 3)) # x: list<double> -> list<double>
+
+# possible annotations:
+# [1] x: {double, list<double>} -> {double, list<double>}
+# [2] x: {double -> double, x: list<double> -> list<double>}
+\end{lstlisting}}\caption{Possible function annotations.}\label{fig:funanno}\end{figure}
+The first annotation option is
+
+%
+%
+%
+%
+\subsection{Coverage}
 
@@
 
+%%
+%%
+%%
+%%
+%\subsection{Types in R}
+%
+%\AT{I'll pull from this section when appropriate, as the discussion of the type systems becomes relevant.}
+%
+%%
+%%
+%\subsubsection{What are possible type systems?}
+%
+%Great question.
+%Some things:
+%
+%\begin{itemize}
+%
+%\item {\it The Obvious Type System (OTS)}: this type system corresponds directly with the type information that one can fetch using the {\tt typeof}, {\tt attributes}, and {\tt class} functions.
+%In this system, there is no distinction between vectors and scalars, and lists of differently-typed elements (e.g., a list of integers vs.
+a list of strings), and there is a distinction between integers and doubles, and the type of NULL and all other types.
+%This type system isn't particularly informative to a programmer, though it does reflect the treatment of values in the R runtime, for the {\tt typeof} a value is the type that the value has according to the R internals.
+%
+%\item {\it The Fine-Grained Type System (FGTS)}: this type systems includes a number of relevant additions.
+%Chiefly, it distinguishes between vectors and scalars, as it is conceivable that boxing scalars as single-element vectors might be overkill.
+%In addition, list types are parameterized over the types of their elements (e.g., list<integer> vs. list<character>).
+%Finally, NAs ``in the wild'' are given the type {\tt raw\_NA}, as {\tt typeof(NA) = logical} in R.
+%
+%\item {\it The Null-and-NA-Free Type System (NNTS)}: most languages consider NULLs to inhabit all types, and R does so for NAs \AT{elaborate}.
+%This gives us ... .
+%
+%\end{itemize}
+%
+%Of course, one could conceive of a more fine-grained type system than FGTS, but it's unclear that we would gain much in doing so.
+%Indeed, there is a space to explore in slightly less fine-grained type systems, which make modifications including:
+%
+%\begin{itemize}
+%
+%\item {\bf collapsing vectors and scalars}: this yields a type system slightly closer to OTS.
+%
+%\item {\bf rolling NULL into other types}: in most languages, NULL is an inhabitant of all types, and this adjustment seems interesting.
+%Note that this goes beyond even OTS, as in R {\tt typeof(NULL) = NULL}.
+%
+%\item {\bf rolling NA into other types}: \AT{i should try this}
+%similarly, NA is an inhabitant of all types, though NA standing alone has type {\tt logical}.
+%FGTS distinguishes such NAs with the {\tt raw\_NA} type, and if instead we consider NA to inhabit all types we may \AT{win}.
+%
+%\item {\bf combining integers and doubles}: as in most languages, doubles and ints can stand-in for each other in nearly every scenario.
+%Even if not, it is trivial to convert from one to the other, particularly converting ints to doubles.
+%Aside, in R a double can be used to index a list, and R will floor it when performing the lookup.
+%
+%\end{itemize}
+%\subsection{Usage Patterns}
+%
+%In this section, we will discuss function usage patterns which arose in our analysis.
+%We will start by looking at the morphicity of functions.
+%
+%\subsubsection{Function Argument and Return Morphicity}
+%
+%First and foremost, we would like to know how often R programmers create polymorphic functions.
+%Recall that we define a polymorphic function to be a function with at least one polymorphic argument, or a polymorphic return.
+%We will turn our attention now to the data in \todo{Figure}, and go through each entry in the table.
+%
+%Here, we see that the vast majority of function arguments are indeed monomorphic.
+%Monomorphic arguments are easy to annotate, as the {\tt typeof} the arguments is exactly the most precise annotation we could give, at least in terms of \textit{type}.
+%That said, types alone don't always paint the whole picture:
+%recall that \textit{attributes} are an R language feature which allows programmers to stick metadata onto values.
+%So in which circumstances \textit{do} types paint the whole picture?
+%
+%The second data point in \textbf{TODO FIGURE} indicates that a
+%\isit{majority} of function arguments are monomorphic in type \textit{and}
+%have no attributes.  These represent arguments which are truly trivial to
+%annotate, as the {\tt typeof} an argument perfectly describes the usage of
+%that argument.  That said, some attributes arise naturally in R: For
+%instance, names in a named list (e.g. {\tt x} and {\tt y} in {\tt list(x=0,
+%  y=0)}) appear as an attribute on the value.  In
+%Section~\ref{sec:method:attributes}, we outlined these naturally-occurring
+%attribute patterns, and the third data point in \textbf{TODO FIGURE} shows
+%that a nontrivial amount of functions are monomorphic in type with said
+%natural attribute patterns.
+%
+%Another facet of type information in R is in the {\tt class} attribute.
+%Recall that values have a \textit{class} in addition to a type: For example,
+%a {\tt data.frame} has type {\tt list} and class {\tt data.frame}.
+%\todo{Plug classes earlier} R has a number of built-in classes, such as
+%\todo{X, Y, and Z}, but users are free to redefine the class of any value at
+%runtime, and easily define new classes.  The next data point in
+%\todo{Figure} shows that user-defined classes don't appear altogether often,
+%though they do indeed feature.  \AT{In a following section, we will discuss
+%  how the usage of these classes manifests itself.}
+%
+%\AT{Separate section for functions? Right now, this is at argument granularity.}
+%
+%\subsubsection{Type Signatures}
+%
+%In the last section, we presented a high-level overview of the morphicity ... .
+%
+%\subsubsection{Attribute Signatures}
+%
+%Recall that attributes are a way for programmers to store metadata on values in R.
+%What are the common attribute patterns?
+%And how often is the attribute pattern polymorphic?
+%
+%\subsubsection{Takeaways}
+%
+%\begin{itemize}
+%    \item the vast majority of arguments are monomorphic in type;
+%    \item of those, over 60\% have no attribute information;
+%    \item of the 40\% with attribute information, roughly 1/2 have fairly
+%      simple attributes corresponding to base R constructs (named lists and
+%      vectors, matrices, and data frames);
+%    \item now, of those arguments which are polymorphic, a sizable chunk (well over half) have defensible signatures (e.g., double and character for named list indexing, double and integer for obvious reasons, etc.).
+%\end{itemize}
+%
+%In short, it looks like R (package) programmers are reasonable.  I'd
+%conjecture that a lot of the polymorphism (e.g., double and list) is coming
+%from how easy it is to use either type in a given situation (e.g.,
+%converting from vector of doubles to list or vice versa is simple).
+
 \end{document}

### 2019-03-28 paper/main.tex
discussing possible function annotations

@@
 \end{lstlisting}}\caption{Possible function annotations.}\label{fig:funanno}\end{figure}
-The first annotation option is
+\AT{Say more about this?}
+
+Harkening back to our discussion of character/real polymorphism, we mentioned that we might want to consider such a type signature to be a list or vector index only if a list or vector was present as another argument.
+Now, this may be moving well beyond the purview of R programmers, but perhaps there is value in such a scenario, particularly if we include even {\it more} information in a type system.
+For example, say we have a function which takes in a 2D point list.
+There are a plethora of useful annotations, here, and of varying degrees of granularity.
+We might want simply a "list" annotation, or we might want a "list<double>" annotation to be more specific.
+In addition, though, we might want to know if the list has a names attribute, yielding "list<double>@names".
+Then, perhaps it would be useful to know what the names were, so "list<double>@names\{x, y\}".
+
+Now, depending on the granularity of individual argument annotations, we might want to link certain arguments together.
+Consider the code snippet in Figure~\ref{fig:indexof}
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+access <- function(l, i) l[[i]]
+
+L <- list(a=1, b=2, c=3)
+access(L, 2) # => 2
+access(L, "c") # => 3
+
+# possible annotations:
+# [1] l: list, i: {character, real} -> real
+# [2] l: list<double>, i: {character, real} -> real
+# [3] l: list<X>, i: {character, real} -> X
+# [4] l: list<double>@names, i: index -> double
+# [5] l: list<double>@names{a, b, c}, i: index@indexOf(l) -> double
+\end{lstlisting}}\caption{Linking annotations.}\label{fig:indexof}\end{figure}
+The annotations could be used to generate code to check for compliance, like a contract.
 

### 2019-03-28 notebooks/paper_results_section.Rmd
updated counts in paper after controlling for failed traces, will push data separately later

@@
 
-I think something funky is happening.
-We should have matrix types, maybe I forgot.
-Dealing with that now:
+Get the biggest signature.
 ```{r}
-# NOTE matrix stuff moved up, oopsie
+big <- get_biggest_function(los, metric=size_of_signature_type)
+
+# it's %>%
 ```
-
-Top sig time.
-```{r}
-# poly_type_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_type_args))
-# poly_type_args <- poly_type_args[order(-poly_type_args$count),] # sort by count
-# poly_type_args$perc <- round(poly_type_args$count/sum(poly_type_args$count) * 100, 2)
-```
-
-Print top 10.
-```{r}
-# format_df_for_print(poly_type_args[1:10,]) -> put_me
-# put_me
-# xtable(put_me) %>% print(include.rownames=F)
-```
-
-```{r}
-# sigs_df_vec <- get_fun_sigs_matching_pred_from_lopkg(los, pred=does_df_have_X_types, types_to_find=c("matrix", "vector/real"), strict=T)
-```

### 2019-03-28 paper/main.tex
updated counts in paper after controlling for failed traces, will push data separately later

@@
   \hline
-  Full Monomorphic & 784K & 85.28 \\ 
-  Monomorphic in Type & 836K & 90.91 \\ 
-  Monomorphic in Class & 831K & 90.31 \\ 
-  Monomorphic in Attribute Pattern & 830K & 90.25 \\ 
-  Total Seen & 920K & --- \\ 
+  Full Monomorphic & 780K & 85.36 \\ 
+  Monomorphic in Type & 832K & 91.03 \\ 
+  Monomorphic in Class & 827K & 90.41 \\ 
+  Monomorphic in Attribute Pattern & 825K & 90.27 \\ 
+  Total Seen & 914K & --- \\ 
      \hline
@@
   \hline
-  Full Monomorphic & 143K & 65.80 \\ 
-  Monomorphic in Type & 167K & 76.84 \\ 
-  Monomorphic in Class & 163K & 75.13 \\ 
-  Monomorphic in Attribute Pattern & 161K & 74.31 \\ 
-  Total Seen & 217K & --- \\ 
+  Full Monomorphic & 143K & 65.87 \\ 
+  Monomorphic in Type & 167K & 76.95 \\ 
+  Monomorphic in Class & 163K & 75.21 \\ 
+  Monomorphic in Attribute Pattern & 161K & 74.37 \\ 
+  Total Seen & 216K & --- \\ 
    \hline
@@
   \hline
-  vector/double, vector/integer & 7677 & 9.18 \\ 
-  scalar/double, scalar/integer & 7194 & 8.61 \\ 
-  matrix, vector/double & 5354 & 6.40 \\ 
-  NULL, scalar/character & 2333 & 2.79 \\ 
-  list$<$any$>$, list$<$double$>$ & 2224 & 2.66 \\ 
-  data.frame, matrix & 2115 & 2.53 \\ 
-  NULL, scalar/double & 2066 & 2.47 \\ 
-  NULL, vector/double & 2042 & 2.44 \\ 
-  list$<$any$>$, list$<$list$>$ & 1426 & 1.71 \\ 
-  NULL, vector/character & 1385 & 1.66 \\ 
+  vector/double, vector/integer & 8228 & 10.03 \\ 
+  scalar/double, scalar/integer & 7190 & 8.76 \\ 
+  matrix, vector/double & 5362 & 6.54 \\ 
+  NULL, scalar/character & 2308 & 2.81 \\ 
+  list$<$any$>$, list$<$double$>$ & 2220 & 2.71 \\ 
+  data.frame, matrix & 2115 & 2.58 \\ 
+  NULL, scalar/double & 2064 & 2.52 \\ 
+  NULL, vector/double & 2041 & 2.49 \\ 
+  list$<$any$>$, list$<$list$>$ & 1420 & 1.73 \\ 
+  NULL, vector/character & 1374 & 1.67 \\ 
      \hline
@@
   \hline
-  Full Monomorphic & 785K & 85.38 \\ 
-  Monomorphic in Type & 871K & 94.65 \\ 
+  Full Monomorphic & 781K & 85.46 \\
+  Monomorphic in Type & 867K & 94.78 \\
 %  Monomorphic in Class & 831K & 90.31 \\ 
 %  Monomorphic in Attribute Pattern & 830K & 90.25 \\ 
-  Total Seen & 920K & --- \\ 
+  Total Seen & 914K & --- \\ 
      \hline
@@
   \hline
-  Full Monomorphic & 143K & 65.95 \\ 
-  Monomorphic in Type & 184K & 84.778 \\
+  Full Monomorphic & 143K & 66.02 \\ 
+  Monomorphic in Type & 184K & 84.91 \\
 %  Monomorphic in Class & 163K & 75.13 \\ 
 %  Monomorphic in Attribute Pattern & 161K & 74.31 \\ 
-  Total Seen & 217K & --- \\ 
+  Total Seen & 216K & --- \\ 
    \hline
@@
   \hline
-  matrix, vector/real & 6307 & 12.82 \\ 
-  list$<$any$>$, list$<$real$>$ & 2792 & 5.68 \\ 
-  vector/character, vector/real & 2719 & 5.53 \\ 
-  scalar/character, scalar/real & 2536 & 5.16 \\ 
-  data.frame, matrix & 2163 & 4.40 \\ 
-  list$<$any$>$, list$<$list$>$ & 1492 & 3.03 \\ 
-  list$<$any$>$, list$<$character$>$ & 1217 & 2.47 \\ 
-  data.frame, vector/real & 1025 & 2.08 \\ 
-  scalar/logical, scalar/real & 917 & 1.86 \\ 
-  scalar/character, vector/real & 844 & 1.72 \\ 
+  matrix, vector/real & 6320 & 13.25 \\ 
+  list$<$any$>$, list$<$real$>$ & 2782 & 5.83 \\ 
+  vector/character, vector/real & 2706 & 5.67 \\ 
+  scalar/character, scalar/real & 2523 & 5.29 \\ 
+  data.frame, matrix & 2163 & 4.53 \\ 
+  list$<$any$>$, list$<$list$>$ & 1484 & 3.11 \\ 
+  list$<$any$>$, list$<$character$>$ & 1211 & 2.54 \\ 
+  data.frame, vector/real & 1027 & 2.15 \\ 
+  scalar/logical, scalar/real & 911 & 1.91 \\ 
+  scalar/character, vector/real & 839 & 1.76 \\ 
    \hline

### 2019-03-28 notebooks/paper_results_section.Rmd
new data, and some top poly by size info

@@
 
-Get the biggest signature.
+Get the biggest signatures.
 ```{r}
-big <- get_biggest_function(los, metric=size_of_signature_type)
+lo_sizes <- get_X_biggest_functions(los, metric = size_of_signature_type)
 
-# it's %>%
+top_10_big <- head(lo_sizes, 10)
+
+# make df
+top_10_big_df <- data.frame(
+  pkg = sapply(names(top_10_big), function(n) strsplit(n, split="::")[[1]][1]),
+  fun = sapply(names(top_10_big), function(n) strsplit(n, split="::")[[1]][2]),
+  size = top_10_big
+)
+
+rownames(top_10_big_df) <- NULL
+
+top_10_big_df %>% xtable
+
 ```

### 2019-03-28 paper/main.tex
new data, and some top poly by size info

@@
 Now, depending on the value of {\tt spss}, {\tt dat} will either be a data frame or a vector of doubles.
-\AT{This ties in to the (arg, retv) coupling, which we will have data on soon.}
 
@@
 
-\AT{Meta-Comment:} My plan for this section is to linearly go through the
+\AT{Note:} My plan for this section is to linearly go through the
 data, and show logical transitions between various type systems.  I'll start
 with the finest-grained one.
+\AT{End Note.}
 
@@
 \begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-> add_make_list <- function(x, y) as.list(x + y)
-> add_make_list(2, c(3, 1)) # => list(5, 3)
+add_make_list <- function(x, y) {
+  l <- as.list(x + y)
+  attr(l, "example") <- "this is an attribute"
+  l
+}
+add_make_list(2, c(3, 1)) # => list(5, 3)
 \end{lstlisting}}
@@
 y & \tt vector/double & \tt numeric & \tt \{\} \\
-retv & \tt list<double> & \tt list & \tt \{\}  
+retv & \tt list<double> & \tt list & \tt \{example: scalar/character\}  
 \end{tabular}
@@
 \AT{We might want to discuss some of the most polymorphic functions in the language.}
-\subsection{Most Polymorphic Functions}
+\subsection{R's Most Polymorphic Functions}
+
+In this section, our goal is to shed some light on the most polymorphic functions in R.
+Table~\ref{tab:bigpolyfuns} shows the 10 functions with the biggest recorded signatures.
+
+% latex table generated in R 3.5.1 by xtable 1.8-3 package
+% Thu Mar 28 16:01:37 2019
+\begin{table}[ht]
+\label{tab:bigpolyfuns}
+\centering
+\begin{tabular}{llrl}
+  \hline
+Package & Function & Size & Description \\ 
+  \hline
+magrittr & \%$>$\% & 23.00 & Tidyverse pipe function \\ 
+  magrittr & freduce & 17.67 & Apply a list of functions sequentially \\ 
+  rlang & captureArgInfo & 16.50 & Capture arg info \\ 
+  rlang & enquo & 16.00 & Create quosure \\ 
+  rlang & get\_expr & 15.00 & Extract expression \\ 
+  data.table & replace\_dot\_alias & 13.50 & Deals with '.' in passed expr \\ 
+  tibble & strip\_dim & 13.00 & Removes 'dim' attribute \\ 
+  data.table & copy & 11.00 & Create copy of passed value \\ 
+  htmlwidgets & shouldEval & 11.00 & Eval passed value depending on tag \\ 
+  magrittr & \%$<$$>$\% & 11.00 & Like \%$>$\% but assigns \\ 
+   \hline
+\end{tabular}
+\caption{10 most polymorphic functions.}
+\end{table}
+
+The size metric we use is roughly the size of each individual argument signature, divided by the number of function arguments.
+The division is performed to avoid unfairly favouring functions with many arguments.
 
@@
 Now, depending on the granularity of individual argument annotations, we might want to link certain arguments together.
-Consider the code snippet in Figure~\ref{fig:indexof}
+Consider the code snippet in Figure~\ref{fig:indexof}.
 \begin{figure}[!hb]{\small\begin{lstlisting}[style=R]

### 2019-03-28 paper/main.tex
more data, more discussion

@@
 \newcommand{\YEARS}{?\xspace}
-\newcommand{\INDEXCOINCIDENCE}{3,426\xspace}
-\newcommand{\TOTALINDEXY}{6,461\xspace}
+\newcommand{\INDEXCOINCIDENCE}{3,499\xspace}
+\newcommand{\TOTALINDEXY}{6,561\xspace}
 \newcommand{\INDEXYPERC}{53\%\xspace}		% TODO Not an accurate count.
@@
   \hline
-  Full Monomorphic & 780K & 85.36 \\ 
-  Monomorphic in Type & 832K & 91.03 \\ 
-  Monomorphic in Class & 827K & 90.41 \\ 
+  Full Monomorphic & 781K & 85.37 \\ 
+  Monomorphic in Type & 834K & 91.23 \\ 
+  Monomorphic in Class & 827K & 90.42 \\ 
   Monomorphic in Attribute Pattern & 825K & 90.27 \\ 
@@
   \hline
-  Full Monomorphic & 143K & 65.87 \\ 
-  Monomorphic in Type & 167K & 76.95 \\ 
-  Monomorphic in Class & 163K & 75.21 \\ 
+  Full Monomorphic & 143K & 65.88 \\ 
+  Monomorphic in Type & 167K & 77.25 \\ 
+  Monomorphic in Class & 163K & 75.24 \\ 
   Monomorphic in Attribute Pattern & 161K & 74.37 \\ 
@@
   \hline
-  vector/double, vector/integer & 8228 & 10.03 \\ 
-  scalar/double, scalar/integer & 7190 & 8.76 \\ 
-  matrix, vector/double & 5362 & 6.54 \\ 
-  NULL, scalar/character & 2308 & 2.81 \\ 
-  list$<$any$>$, list$<$double$>$ & 2220 & 2.71 \\ 
-  data.frame, matrix & 2115 & 2.58 \\ 
-  NULL, scalar/double & 2064 & 2.52 \\ 
-  NULL, vector/double & 2041 & 2.49 \\ 
-  list$<$any$>$, list$<$list$>$ & 1420 & 1.73 \\ 
-  NULL, vector/character & 1374 & 1.67 \\ 
+  vector/double, vector/integer & 7742 & 9.66 \\ 
+  scalar/double, scalar/integer & 7201 & 8.99 \\ 
+  matrix, vector/double & 5384 & 6.72 \\ 
+  NULL, scalar/character & 2325 & 2.90 \\ 
+  list$<$any$>$, list$<$double$>$ & 2231 & 2.78 \\ 
+  data.frame, matrix & 2116 & 2.64 \\ 
+  NULL, scalar/double & 2082 & 2.60 \\ 
+  NULL, vector/double & 2069 & 2.58 \\ 
+  list$<$any$>$, list$<$list$>$ & 1432 & 1.79 \\ 
+  NULL, vector/character & 1380 & 1.72 \\
      \hline
@@
   \hline
-  Full Monomorphic & 781K & 85.46 \\
-  Monomorphic in Type & 867K & 94.78 \\
+  Full Monomorphic & 781K & 85.47 \\
+  Monomorphic in Type & 867K & 95.01 \\
 %  Monomorphic in Class & 831K & 90.31 \\ 
@@
   \hline
-  Full Monomorphic & 143K & 66.02 \\ 
-  Monomorphic in Type & 184K & 84.91 \\
+  Full Monomorphic & 143K & 66.04 \\ 
+  Monomorphic in Type & 184K & 85.30 \\
 %  Monomorphic in Class & 163K & 75.13 \\ 
@@
   \hline
-  matrix, vector/real & 6320 & 13.25 \\ 
-  list$<$any$>$, list$<$real$>$ & 2782 & 5.83 \\ 
-  vector/character, vector/real & 2706 & 5.67 \\ 
-  scalar/character, scalar/real & 2523 & 5.29 \\ 
-  data.frame, matrix & 2163 & 4.53 \\ 
-  list$<$any$>$, list$<$list$>$ & 1484 & 3.11 \\ 
-  list$<$any$>$, list$<$character$>$ & 1211 & 2.54 \\ 
-  data.frame, vector/real & 1027 & 2.15 \\ 
-  scalar/logical, scalar/real & 911 & 1.91 \\ 
-  scalar/character, vector/real & 839 & 1.76 \\ 
+  matrix, vector/real & 6360 & 13.94 \\ 
+  list$<$any$>$, list$<$real$>$ & 2799 & 6.13 \\ 
+  vector/character, vector/real & 2774 & 6.08 \\ 
+  scalar/character, scalar/real & 2552 & 5.59 \\ 
+  data.frame, matrix & 2167 & 4.75 \\ 
+  list$<$any$>$, list$<$list$>$ & 1496 & 3.28 \\ 
+  list$<$any$>$, list$<$character$>$ & 1214 & 2.66 \\ 
+  data.frame, vector/real & 1046 & 2.29 \\ 
+  scalar/logical, scalar/real & 920 & 2.02 \\ 
+  scalar/character, vector/real & 868 & 1.90 \\
    \hline
@@
 This figure reveals an unfortunate truth that date ``data type'' in R is not standardized.
-That said, translating between the data types is straigthforward.
+That said, translating between the data types is straightforward.
 
@@
 The division is performed to avoid unfairly favouring functions with many arguments.
+
+Which packages are the most polymorphic?
+Figure~\ref{tab:packagepolysize} has the answer.
+
+% latex table generated in R 3.5.1 by xtable 1.8-3 package
+% Thu Mar 28 18:15:26 2019
+\begin{table}[ht]
+\label{tab:packagepolysize}
+\centering
+\begin{tabular}{lrl}
+  \hline
+Package & Total Size & Description \\ 
+  \hline
+  magrittr & 4.16 & Tidyverse pipe suite \\ 
+  types & 3.25 & Simple type annotations for R \\ 
+  r6extended & 3.00 & Utilities extending barebones R6 package \\ 
+  utils & 2.93 & Utilities for R developers \\ 
+  zeallot & 2.85 & Metaprogramming package for assignment \\ 
+  listviewer & 2.72 & R list visualizer \\ 
+  proto & 2.60 & Prototype object-based programming \\ 
+  apercu & 2.49 & Display a short view of a complex R value (list, vector, etc.) \\ 
+  pipeR & 2.46 & Another pipe suite \\ 
+  rethinker & 2.44 & RethinkDB client for R \\ 
+   \hline
+\end{tabular}
+\caption{10 most polymorphic packages.}
+\end{table}
 

### 2019-03-29 paper/main.tex
going through, consolidating, dealing with comments. at 5.1

@@
 \newcommand{\INDEXYPERC}{53\%\xspace}		% TODO Not an accurate count.
+\newcommand{\DATAPKGS}{206\xspace}
+\newcommand{\DATAPKGSPERC}{1.5\%\xspace}
 
@@
 containing data, which have no runnable code but are referenced by other
-packages \AT{I should collect stats on this}.
+packages.
+In our experience, only \DATAPKGS packages had no executable test or example code,
+accounting for \DATAPKGSPERC of CRAN.
 
@@
 We capture and record function signatures on-the-fly, though, so we do collect (partial) data even 
-for these long-running packages. \AT{We expect to extend our coverage of CRAN for the next version of this paper.}
+for these long-running packages.
 
@@
 
-\AT{TODO Change "recorded" to "observed" in recordsbypkg}
+\AT{TODO Touch up graphs.}
 
@@
 In R all arguments are passed as promises, unused arguments will be
-unevaluated, for these we can not report a type. THUS WE DO WHAT? SAY MORE.
+unevaluated, for these we can not report a type. 
+\AT{Unused arguments are still picked up by genthat, so we evaluate them.
+I suppose this might lead to issues in some rare circumstances.}
+
+\AT{Do we care about this discussion? This is probably a bit too specific to the implementation.}
+Packages which implement metaprogramming mechanisms are also a bit challenging.
+For example, consider the {\tt magrittr} package, which implements the Tidyverse's pipe functions.
+Refer to Figure~\ref{fig:magpipeex} for a translation of a call to {\tt magrittr}'s pipe ({\tt \%>\%}) function.
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+mul_by <- function(x, y) { x * y }
+
+4 %>% mul_by(3) === `%>%`(lhs = 4, rhs = mul_by(3)) === mul_by(4, 3)
+\end{lstlisting}}\caption{Example and rough translation of \%$>$\% call.}\label{fig:realex}\end{figure}
+Our tracer evaluates function arguments, and here \code{rhs = mul_by(3)}, but that's erroneous code (you can't call {\tt mul\_by} without supplying two arguments).
+In this situation, our tracer will still register the entire call, but flag the {\tt rhs} argument to indicate to later stages of the analysis that there was an issue evaluating that argument.
 
@@
 
-\AT{Do we want to talk about how sometimes we can't eval an argument, even if a function executes successfully?
-Like pipe from magrittr.}
-
-We will need a notion of polymorphism when undertaking this analysis. R
-is unityped, and the only notion of polymorphism in the language is its use
-of dynamic dispatch on the class of an argument.  For our purposes, we will
-define 3 simple types of polymorphism, closely connected with the reflection
-functions we use.  We start by defining what it means for an {\it argument}
-to be polymorphic.  \AT{Could be valuable to have an example here.}
+We will need a notion of polymorphism, and thus a notion of type, when undertaking this analysis. 
+R is unityped, and the only notion of polymorphism in the language is its use of dynamic dispatch on the class of an argument.  
+For our purposes, we will define a notion in type roughly in line with R's \code{typeof} reflection function, in some cases augmented by information obtained by \code{class} and \code{attributes} reflection.
+We start by defining what it means for an {\it argument} to be polymorphic.
 
@@
   \item an argument is {\it polymorphic in type} if it has been called with at
-  least two distinct ``types'' according to the \code{typeof} function and
-  our extra analysis.
-\item similarly, an argument is {\it polymorphic in class} if it has been
-  called with at least two distinct classes according to the \code{class}
-  function.
+  least two distinct types according to our definition of type.
+
+   \item an argument is {\it polymorphic in class} if it has been called with at least two distinct classes according to the \code{class} function.
 
@@
 
-There is an important distinction to be made between the result of
-\code{typeof} and that of \code{class} and \code{attributes}: In R, the
-\code{typeof} a value cannot change, and this is not the case with class and
-attributes.  As mentioned in \AT{an earlier section}, the class of values
-can be easily redefined by assigning to the value's class attribute.
-Similarly, attributes can be dynamically added, removed, or changed.
-
-This begs the question: what sort of classes and attribute patterns occur
-naturally, before possible user tampering?  In terms of classes, the
-discussion can be found in Section~\ref{sec:rlang}.  As for attributes:
+The main ingredient of our types is the {\tt typeof} a value, and there is an important distinction to be made between the result of \code{typeof} and that of \code{class} and \code{attributes}.
+In R, the \code{typeof} a value cannot change, and this is not the case with class and attributes.  
+As mentioned in \AT{an earlier section}, the class of values can be easily redefined by assigning to the value's class attribute, and similarly attributes can be dynamically added, removed, or changed.
+
+While we do consider some class and attribute information when constructing our types, it is also valuable to consider them entirely separate from types.
+Users are free to define their own classes, and arbitrarily add and remove attributes, and interesting patterns might arise in considering them.
+That said, not attribute patterns and classes are necessarily interesting, and indeed some of R's built-in data structures use them to some extend.
+For one, data frames have a {\tt data.frame} class (explicitly on the class attribute).
+Other standard patterns are:
 
@@
   and optionally a {\tt dimnames} attribute for dimension names.
+Curiously, the {\tt dim} attribute is all that is needed to define a matrix (i.e., a list with a {\tt dim} attribute {\it is} a matrix).
 

### 2019-03-29 paper/main.tex
cleaned up stuff section, added bit about what data we collect, dunno if needed

@@
 %
-
+%
 %
@@
 
-As we mentioned in Section~\ref{sec:rlang}, R has several notions of
-``type'', ranging from the results of \code{typeof} to metadata on values
-obtainable via the \code{attributes} function.  Individually, these paint an
-incomplete picture of the type of a value: For example, recall that
-\code{typeof(5) == typeof(c(5, 5)) == "double"}.
+When constructing a type for a value, we begin with the {\tt typeof} the value.
+This alone paints an incomplete picture of the value, though, especially when it comes to vectors (recall that \code{typeof(5) == typeof(c(5, 5)) == "double"}).
+In some cases, the class or attributes of a value might enlighten us, but in many cases we can be a little more precise.
 
@@
 to create a more specific type than the result of any of R's reflection
-functions.  For instance, if the analysis encounters {\tt typeof(x) ==
-  "double"}, it will look at the length of {\tt x} to determine if {\tt x}
-is a scalar or a vector, generating the annotation {\tt scalar/double} or
-{\tt vector/double} accordingly.  One notable imperfection in this example
-is that unit-length vectors will appear as scalars, but if some function was
-called with unit-length vectors {\it as well as} larger ones, both scalar
-and vector types will appear, and this distinction can be later collapsed.
-Another example is that our analysis builds parametric list types by
-capturing the types of list elements.  We also distinguish NAs ``in the
-wild'' with a {\tt raw\_NA} type.
+functions.  The details of the information we collect follows.
+
+\begin{itemize}
+
+	\item if the analysis encounters \code{ typeof(x) == "double"}, it will look at the length of {\tt x} to determine if {\tt x} is a scalar or a vector, generating the annotation {\tt scalar/double} or {\tt vector/double} accordingly.
+	This is also true for integer, complex, raw, logical, and character {\tt typeof}s (i.e., the primitive vectorized data types).
+	A notable imperfection here is is that unit length vectors will appear as scalars, so later in our analysis we ensure that argument signatures containing both scalar/X and vector/X are collapsed down to contain only vector/X.
+
+	\item if the analysis encounters \code{typeof(x) == "list"}, we collect type information on all list elements in order to ascribe a general type to the contents of the list.
+	To avoid undue slowdowns, we only collect the \code{typeof} the contents, and ascribe type {\tt any} if two or more {\tt typeof}s are observed.
+
+	\item if the analysis encounters \code{NA}, we ascribe a unique {\tt raw\_NA} type.
+	In R, \code{NA} inhabits all types, but \code{typeof(NA) == "logical"}.
+	The analysis deals with these {\tt raw\_NA}s at a later stage.
+
+	\item if the analysis encounters a matrix (i.e. the value has matrix class and has a dims and optionally dimnames attribute), we ascribe the type {\tt matrix}.
+
+	\item similarly, if it encounters a data frame (i.e., class {\tt data.frame} with appropriate attributes), we ascribe the type {\tt data.frame}.
+
+\end{itemize}
 

### 2019-03-29 notebooks/corpus_stats_and_plots.Rmd
small changes, additions, new functions
updated loc counts

@@
 count_df <- read_rds("../data/final_loc_counts.RDS")
+# rows might need to be cast to integers
+class(count_df$r_counts) <- "integer"
+class(count_df$c_counts) <- "integer"
+
 as.tibble(count_df) %>%
@@
 # just have data.
+```
+
+First: let's get some of the counts for the beginning of the paper.
+```{r}
+tot_r <- sum(df$r_counts)
+tot_c <- sum(df$c_counts)
+nrow(df)
 ```

### 2019-03-29 paper/main.tex
fighting git
updated loc counts
merging
more corpus prose

@@
 \usepackage{my_style}
-
 \definecolor{LightGray}{rgb}{.95,.95,.95}
@@
 package. Each script corresponds to either one use-case or a set of unit
-tests.  Figure~\ref{allcloc} shows the size of the code across all the
-packages we have used.  Lastly, Figure~\ref{recorded} gives the number of
-function signatures that we recorded while executing the scripts of each
-package.
+tests.
+
+Figure~\ref{allcloc} shows the size of the code across all the packages we have used. 
+The lines of R code are above the X-axis, and lines of C code below.
+All counts are $log_{10}$-scaled, and entries are in ascending order by number of lines of R.
+Roughly speaking, the majority of R packages have between 100 and 1000 lines of code.
+In terms of C, only ~25\% of packages have any C code.
+C and C++ are often used to implement parts of a package where performance is critical.
+
+Lastly, Figure~\ref{recorded} plots the relationship between recorded and observed signatures.
+Here, \# observed $>$ \# recorded since it's impossible to record more traces than those which pass through the tracer.
+Numbers are again $log_{10}$-scaled.
 
 \AT{TODO Touch up graphs.}
+\AT{TODO Other interesting plots? Recorded vs \# scripts?}
+
+Having built a sense for the corpus of data that we're dealing with, we will turn our attention to our methods of data collection.
 
@@
 \hline
-\tt Rcpp  & Seamless C++ integration & 2.1K & 4K & 25 & 55K & 282 \\
-\tt rlang & Functions for 'Tidyverse' & 4.4K & 3.8K & 122 & 3,924K & 5,171 \\
-\tt glue  & Interpreted string literals & 0.2K & 0.2K & 8 & 4K & 120 \\
-\tt tibble & Simple data frames & 1.3K & 0.3K & 16 & 1,332K & 4,148 \\
-\tt stringi &  String processing & 1.5K & 515K & 64 & 923K & 666 \\
-\tt ggplot2 & Data visualisations & 10.8K & 0 & 130 & 153K & 2,636\\
-\tt dplyr  &  Data manipulation & 3.8K & 4.7K & 78 & 233K & 1,967 \\
-\tt pillar & Formatting for columns & 1.3K & 0 & 13 & 803K & 1,273 \\
-\tt R6 & Classes w. ref. semantics & 0.6K & 0 & 2 & 1K & 226 \\
-\tt stringr & String operations & 0.5K & 0 & 32 & 1,764K & 371 \\
+\tt Rcpp  & Seamless C++ integration & 2.2K & 4.2K & 25 & 55K & 340 \\
+\tt rlang & Functions for 'Tidyverse' & 7.0K & 6.1K & 122 & 3,924K & 8,422 \\
+\tt glue  & Interpreted string literals & 0.3K & 0.3K & 8 & 4K & 145 \\
+\tt tibble & Simple data frames & 2.0K & 0.3K & 16 & 1,332K & 6,367 \\
+\tt stringi &  String processing & 1.5K & 515K & 64 & 923K & 873 \\
+\tt ggplot2 & Data visualisations & 14K & 0 & 130 & 153K & 4,608 \\
+\tt dplyr  &  Data manipulation & 4.5K & 4.7K & 78 & 233K & 3,099 \\
+\tt pillar & Formatting for columns & 1.4K & 0 & 13 & 803K & 1,514 \\
+\tt R6 & Classes w. ref. semantics & 0.7K & 0 & 2 & 1K & 330 \\
+\tt stringr & String operations & 0.5K & 0 & 32 & 1,764K & 534 \\
 \end{tabular}}\caption{10 Most Downloaded Packages.}\label{most}

### 2019-03-29 notebooks/corpus_stats_and_plots.Rmd
merged airplane changes. rip wow air

@@
   inner_join(cloc, by="names") %>%
-  filter(rec <= obs) %>%                      # Added this
+  filter(rec <= obs) %>%                      # Added this: FIXME, understand why
   filter(r_counts != 0) -> df
@@
 ```{r}
-loc_of_top_10 <- lapply(1:length(combo_df$package_name), function(x) if (combo_df$package_name[[x]] %in% top_10) x else FALSE)
+#loc_of_top_10 <- lapply(1:length(combo_df$package_name), function(x) if (combo_df$package_name[[x]] %in% top_10) x else FALSE)
 
@@
 ggplot(aes(x=names_o, value, group=1)) +
- geom_line(aes(colour = language), size=0.008) +
- labs(title = "Analysis of R Corpus (R and C)", y = "Lines of Code in Package (Log10-scaled)") + guides(colour=F) + 
+ geom_line(aes(colour = language), size=0.008) + guides(colour=F) + 
  theme( axis.text.x  = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(),
         axis.ticks.y = element_line(),  panel.grid.major.y = element_line(colour = "red")) + # ,
-  scale_y_continuous(name="Lines of Code in Package", breaks = c(-4, -3, -2, -1, 0, 1, 2, 3, 4), 
+  scale_y_continuous(name="", breaks = c(-4, -3, -2, -1, 0, 1, 2, 3, 4), 
                      labels = c(10000, 1000, 100, 10, 0, 10, 100, 1000, 10000))
@@
 
-ggsave("../paper/plots/linesofrandccode.pdf")
+ggsave("../paper/plots/linesofrandccode.pdf",width=10,height=4)
 ```
@@
 df_rc %>%
-  mutate(rec=log(rec),obs=log(obs))  %>%
-  ggplot(aes(x=rec, y=obs)) +
-                geom_point(size=0.1) + 
-                labs( title = "Numbers of Observed Signatures",
-                      y = "Obs Invocations in Pkg",
-                      x = "Rec Invocations in Pkg")
-                # theme(axis.text.x = element_blank(), axis.title.x=element_blank(),
-                #       axis.ticks.x=element_blank()) #+
-                # scale_y_continuous(breaks = c(0, 1, 2, 3, 4, 5, 6),
-                #      labels = c(0, 10, 100, 1000, 10000, 100000, 1000000))
+  mutate(rec=log10(rec),obs=log10(obs))  %>%
+  ggplot(aes(y=rec, x=obs)) + geom_point(size=0.1) + 
+    labs( x = "Observed", y = "Recorded") +
+    scale_x_continuous(breaks = c( 1, 2, 3, 4, 5, 6, 7),
+                       labels = c( 10, 100, 1000, 10000, 100000, 1000000, 10000000))+
+    scale_y_continuous(breaks = c( 1, 2, 3, 4, 5, 6, 7),
+                       labels = c( 10, 100, 1000, 10000, 100000, 1000000, 10000000))
 
 # generate file
-ggsave("../paper/plots/recordsbypkg.pdf")
+ggsave("../paper/plots/recordsbypkg.pdf",width=9,height=5)
 ```

### 2019-03-29 paper/main.tex
merged airplane changes. rip wow air

@@
 
-In this section, we present our dataset. R aims to accommodate data
-analysts, their workflows start with data import, followed by cleaning, and
-then by steps of modeling, transformation and visualization. Often, the code
-of these analysis pipelines resides, together with the data and results, in
-notebooks. Few notebooks are publicly shared, and when they are, the data
-isn't. For this reason our analysis focuses on packages which bundle
-reusable units of R code with documentation, sample data and use-cases.
+In this section, we present our dataset. The R language aims to accommodate
+data analysts; their workflows start with data import, followed by cleaning,
+and then by steps of modeling, transformation and visualization. Often, the
+code of these analysis pipelines resides, together with the data and
+results, in notebooks. Few notebooks are publicly shared, and when they are,
+the data isn't. For this reason our analysis focuses on packages which
+bundle reusable units of R code with documentation, sample data and
+use-cases.
 
@@
 new packages a day~\cite{LIgges2017}.  Unlike sites like GitHub, CRAN is a
-curated collection of packages: A package is only accepted to CRAN if it
-abides by a number of well-formedness rules.  Most relevant for our
-purposes, packages must have data, examples, vignettes and tests, all of
-which must successfully run. From our perspective this means that each
-package in CRAN comes with several executable scripts that exercise some of
-its functionality.  Notable exceptions to this rule are packages only
-containing data, which have no runnable code but are referenced by other
-packages.
-In our experience, only \DATAPKGS packages had no executable test or example code,
-accounting for \DATAPKGSPERC of CRAN.
-
-Our corpus is a subset of CRAN, consisting of \PACKAGES packages, accounting
-for some \PERCENTCRAN of all packages.  These packages have a total of \RLOC
-lines of R code and \CLOC lines of C code. For each package, we extracted
-all executable code snippets from documentation, vignettes and tests and ran
-them independently recording all calls to R functions.  It is noteworthy that
-in order to run the scripts in one package, it is often necessary to load a
-number of other packages.  In~\cite{issta18}, the authors estimated code
-coverage to be around 68\% when including reverse dependencies.  As our infrastructure
-adds overhead to script execution, coupled with the fact that some scripts take 
-inordinate amounts of time to run, we limit our analysis of any given package to one hour.
-We capture and record function signatures on-the-fly, though, so we do collect (partial) data even 
-for these long-running packages.
-
-Figure~\ref{most} shows the ten most downloaded CRAN packages.  For each
-one, we list how many lines of R and C/C++ the packages contains.
-Typically, C/C++ code is used to implement performance critical portions of
-the code. We show the number of scripts that could be extracted from the
-package. Each script corresponds to either one use-case or a set of unit
-tests.
-
-Figure~\ref{allcloc} shows the size of the code across all the packages we have used. 
-The lines of R code are above the X-axis, and lines of C code below.
-All counts are $log_{10}$-scaled, and entries are in ascending order by number of lines of R.
-Roughly speaking, the majority of R packages have between 100 and 1000 lines of code.
-In terms of C, only ~25\% of packages have any C code.
-C and C++ are often used to implement parts of a package where performance is critical.
-
-Lastly, Figure~\ref{recorded} plots the relationship between recorded and observed signatures.
-Here, \# observed $>$ \# recorded since it's impossible to record more traces than those which pass through the tracer.
-Numbers are again $log_{10}$-scaled.
-
-\AT{TODO Touch up graphs.}
-\AT{TODO Other interesting plots? Recorded vs \# scripts?}
-
-Having built a sense for the corpus of data that we're dealing with, we will turn our attention to our methods of data collection.
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[scale=0.6]{linesofrandccode}
-\caption{Lines of code in the analyzed corpus. \# of lines of R code are above 0, \# of lines of C code are below 0.}
-\label{allcloc}
-\end{center}
+\emph{curated} collection: A package is only accepted to CRAN if it abides
+by a number of well-formedness rules.  Most relevant for our purposes,
+packages must have data, examples, vignettes and tests, all of which must
+successfully run. From our perspective this means that each package in CRAN
+comes with several executable scripts that exercise some of its
+functionality.  Notable exceptions to this rule are packages only containing
+data, which have no runnable code but are referenced by other packages.
+Only \DATAPKGS packages had no executable code, accounting for \DATAPKGSPERC
+of CRAN.
+
+The corpus used in this paper is a subset of CRAN. We retained packages that
+could be run by our infrastructure in less than one hour. This corpus
+consists of \PACKAGES packages, accounting for some \PERCENTCRAN of all
+packages.  These packages have a total of \RLOC lines of R code and \CLOC
+lines of C code. Figure~\ref{allcloc} shows a per-package breakdown of the
+size of each package sorted by increasing numbers of lines of R. The figure
+suggests that there is little correlation between the C/C++ parts and how
+many lines of R a package contains. The median size of a package is 541
+lines of R code and the largest package has 86K LOC. Typically, C/C++ code
+is used to implement performance critical portions of the code. The majority
+of package, 8,375 to be precise, have no C or C++ code.  The remaining 3,078
+packages, have a median 572 lines of C and the largest package has 385,839
+lines of C.
+
+\begin{figure}[!b]\begin{center}
+\includegraphics[width=.9\textwidth]{linesofrandccode}
+\caption{Lines of code (log scal); for each package, R is above 0, and C/C++
+  below}\label{allcloc}\end{center}
 \end{figure}
 
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[scale=0.6]{recordsbypkg}
-\caption{Numbers of Recorded Function Invocations in the Analyzed Corpus}
-\label{recorded}
-\end{center}
-\end{figure}
+For each package, we extracted all executable code snippets from
+documentation, vignettes and tests and ran them independently recording all
+calls to R functions.  It is noteworthy that in order to run the scripts in
+one package, it is often necessary to load a number of other packages.
+In~\cite{issta18}, the authors estimated code coverage to be around 68\%
+when including reverse dependencies.  As our infrastructure adds overhead to
+script execution, coupled with the fact that some scripts take inordinate
+amounts of time to run, this is why we limited our analysis of any given
+package to one hour.
 
@@
 
+Figure~\ref{most} shows the ten most downloaded CRAN packages.  For each
+one, we list how many lines of R and C/C++ the packages contains.  We show
+the number of scripts that could be extracted from the package. Each script
+corresponds to either one use-case or a set of unit tests. We print the
+number of function calls that were observed by our infastructure, and the
+number of unique signatures that were recorded.
+
+Our infrastructure only retains unique argument/return combinations. Thus,
+while we observe large number of functions being called with different
+values, the types of these function call are often similar. Over the entire
+corpus, we can see the relation between observed and recorded calls in
+Figure~\ref{recorded}.  The median number of observed calls is 82 and
+maximum is 19 million.  The median number of recorded signatures is 16 and
+the maxium is 8,422. These numbers are skewed by a number of scripts doing
+very few calls before plunging into C code.
+
+\begin{figure}[htbp]\begin{center}
+\includegraphics[width=.9\textwidth]{recordsbypkg}
+\caption{Numbers of Recorded Function Invocations in the Analyzed Corpus}
+\label{recorded}\end{center}
+\end{figure}
+
 \newpage
@@
 
-In this section, we will detail our methodology for collecting data.  The
-basic idea we are exploring is to observe arguments and return values of
-function calls, and from these generalize possible type signatures for the
-called functions. To this end, we build on an open source tool called
-\genthat whose purpose is to generate unit tests for R
-libraries~\cite{issta18}.  To generate test cases, one must record and
-accurately recreate arguments and return values and avoid generating
-redundant tests (i.e. tests that do not increase code coverage). For our
-purposes, we change the existing tool in two main ways: We record argument
-\emph{shapes} rather than their values, and given the limited number of shapes
-this allowed us to avoid having to run the code coverage minimization phase of
-the tool. Both changes were beneficial for scalability---allowing us to
-trace more calls than was reported in~\cite{issta18}.
+In this section, we will detail our methodology for collecting data.  
+Basically, our aim is to observe arguments and return values of function calls, and from these generalize possible type signatures for the called functions. 
+To this end, we build on an open source tool called \genthat whose purpose is to generate unit tests for R libraries~\cite{issta18}.  
+\genthat achieves this by synthesizing unit tests from recorded function argument and return values, comparing the test against existing ones to avoid generating tests which do not increase code coverage.
+To suit our purposes, we change the existing tool in two main ways: 
+We record argument \emph{shapes} rather than values, and given the limited number of shapes this allowed us to avoid having to run the code coverage minimization phase of the tool. 
+Both changes were beneficial for scalability, allowing us to trace far more calls than was reported in the \genthat work~\cite{issta18}.
 
@@
 
-To illustrate our approach, consider the following script which adds a double
-to an integer, and then creates a matrix and a vector, and adds them together to
-get a new matrix.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-> 1 + 2L
+To illustrate our approach, consider the following script which adds a double to an integer, and then creates a matrix from a vector, and adds them together to get a new matrix.
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+> 1 + 2L                        # add double and integer
 [1] 3
-> x <- c(1,2,3,4)
-> attr(x,"dim") <- c(2,2)
-> x + c(1,2)
+> x <- c(1,2,3,4)               # save double vector
+> attr(x,"dim") <- c(2,2)       # assign to "dim" attribute
+> x + c(1,2)                    # add double 3 to each matrix element matrix
      [,1] [,2]
@@
 [2,]    4    6
-\end{lstlisting}}\caption{Example script}\label{example}\end{figure}
+\end{lstlisting}}\caption{Example script.}\label{example}\end{figure}
 

### 2019-03-29 notebooks/paper_L123_results_section.Rmd
small updates, new notebook for new layout and numbers

---
title: "R Notebook"
output: html_notebook
---

This is the notebook for the NEW results section.

Step 0: Administrivia.
```{r}
source("../code/data_processing.R") # get the functions we need

library(xtable)
```

Step 1: Load Data
```{r}
los <- list.files("../data/full_sigs", full.names=TRUE)
names <- list.files("../data/full_sigs")
names(los) <- substr(names, 1, nchar(names)-4)
los <- lapply(los, readRDS)
los <- remove_empties(los)
```

Step 2: Get data ready for L1.
So: Translate the type system to R.
```{r}
los <- change_type_systems_lopkg(los, type_map_T0_to_r, unparam=T)
```

Step 2.1: Counts.
```{r}
L1_c_basic_df <- build_arg_count_df_from_lopkgs(los)

L1_c_f_basic_df <- build_fun_count_df_from_lopkgs(los)
```

```{r}
xtable(L1_c_basic_df) %>% print(include.rownames=F)
xtable(L1_c_f_basic_df) %>% print(include.rownames=F)
```

Step 2.2: Top.
```{r}
L1_poly_type_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_type_args))
L1_poly_type_args <- L1_poly_type_args[order(-L1_poly_type_args$count),] # sort by count
L1_poly_type_args$perc <- round(L1_poly_type_args$count/sum(L1_poly_type_args$count) * 100, 2)
```

Step 2.3: Print top 10.
```{r}
format_df_for_print(L1_poly_type_args[1:10,]) %>%
xtable %>% print(include.rownames=F)
```

Step 3: Translate to L2.
```{r}
# We need to reload b/c we lost information in L1.
los <- list.files("../data/full_sigs", full.names=TRUE)
names <- list.files("../data/full_sigs")
names(los) <- substr(names, 1, nchar(names)-4)
los <- lapply(los, readRDS)
los <- remove_empties(los)
# annoying strange error that I should fix pops up

# There's at least one transformation that we're always gonna want to apply, namly the vector/scalar
# collapse where both are present in a signature.
los <- lapply(los, function(lofuns) {lapply(lofuns, combine_scalar_vector_where_appropriate)})
# Figure out which ones are matrices. Remove if the analysis picks this up (and we rerun).
los <- lapply(los, function(lofun) lapply(lofun, distinguish_matrices_df))
# remove "errors"
los <- lapply(los, function(lofun) lapply(lofun, remove_err_df))
```

Step 3.1: Counts.
```{r}
# argument counts
c_basic_df <- build_arg_count_df_from_lopkgs(los)

# function counts
c_f_basic_df <- build_fun_count_df_from_lopkgs(los)

xtable(c_f_basic_df) %>% print(include.rownames=F)
xtable(c_basic_df) %>% print(include.rownames=F)

```

Step 3.2: Top.
```{r}
poly_type_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_type_args))
poly_type_args <- poly_type_args[order(-poly_type_args$count),] # sort by count
poly_type_args$perc <- round(poly_type_args$count/sum(poly_type_args$count) * 100, 2)
```

Step 3.3: Print top 10.
```{r}
format_df_for_print(poly_type_args[1:10,]) -> put_me
put_me
xtable(put_me) %>% print(include.rownames=F)
```

### 2019-03-29 paper/main.tex
small updates, new notebook for new layout and numbers

@@
 To suit our purposes, we change the existing tool in two main ways: 
-We record argument \emph{shapes} rather than values, and given the limited number of shapes this allowed us to avoid having to run the code coverage minimization phase of the tool. 
+We record argument \emph{shapes} rather than values, and given the limited number of shapes this allowed us to avoid having to run \genthat's code coverage minimization phase. 
 Both changes were beneficial for scalability, allowing us to trace far more calls than was reported in the \genthat work~\cite{issta18}.
@@
 
-To illustrate our approach, consider the following script which adds a double to an integer, and then creates a matrix from a vector, and adds them together to get a new matrix.
+To illustrate our approach, consider the following script which adds a double to an integer, and then creates a matrix from a vector, finally adding them together to get a new matrix.
 
@@
 
-In R all arguments are passed as promises, unused arguments will be
-unevaluated, for these we can not report a type. 
-\AT{Unused arguments are still picked up by genthat, so we evaluate them.
-I suppose this might lead to issues in some rare circumstances.}
-
-\AT{Do we care about this discussion? This is probably a bit too specific to the implementation.}
-Packages which implement metaprogramming mechanisms are also a bit challenging.
+In R all arguments are passed as promises, and unused arguments will be unevaluated.
+Our tracer does still collect information on these arguments, as whether or not they were used doesn't change the fact that they were in fact passed. 
+In some cases, arguments aren't even intended to be evaluated.
 For example, consider the {\tt magrittr} package, which implements the Tidyverse's pipe functions.
@@
 4 %>% mul_by(3) === `%>%`(lhs = 4, rhs = mul_by(3)) === mul_by(4, 3)
-\end{lstlisting}}\caption{Example and rough translation of \%$>$\% call.}\label{fig:realex}\end{figure}
+\end{lstlisting}}\caption{Example and rough translation of \%$>$\% call.}\label{fig:magpipeex}\end{figure}
 Our tracer evaluates function arguments, and here \code{rhs = mul_by(3)}, but that's erroneous code (you can't call {\tt mul\_by} without supplying two arguments).
@@
 
-\AT{Note:} My plan for this section is to linearly go through the
-data, and show logical transitions between various type systems.  I'll start
-with the finest-grained one.
-\AT{End Note.}
+\AT{This is the (partial) data for the new layout.
+We need to decide how to tackle L3.}
+
+Figure~\ref{fig:L1argcounts} has arg counts for L1.
+
+Figure~\ref{fig:L1funcounts} has fun counts for L1.
+
+% latex table generated in R 3.5.1 by xtable 1.8-3 package
+% Fri Mar 29 18:10:23 2019
+\begin{table}[ht]
+\label{fig:L1argcounts}
+\centering
+\begin{tabular}{lrr}
+  \hline
+Type & Count & Percentage \\ 
+  \hline
+Full Monomorphic & 784K & 85.76 \\ 
+  Monomorphic in Type & 845K & 92.46 \\ 
+  Monomorphic in Class & 827K & 90.41 \\ 
+  Monomorphic in Attribute Pattern & 825K & 90.27 \\ 
+  Total Seen & 914K & --- \\ 
+   \hline
+\end{tabular}
+\caption{typeof type system (L1) (arguments)}
+\end{table}
+% latex table generated in R 3.5.1 by xtable 1.8-3 package
+% Fri Mar 29 18:10:23 2019
+\begin{table}[ht]
+\label{fig:L1funcounts}
+\centering
+\begin{tabular}{lrr}
+  \hline
+Type & Count & Percentage \\ 
+  \hline
+Full Monomorphic & 144K & 66.54 \\ 
+  Monomorphic in Type & 173K & 80.16 \\ 
+  Monomorphic in Class & 163K & 75.21 \\ 
+  Monomorphic in Attribute Pattern & 161K & 74.37 \\ 
+  Total Seen & 216K & --- \\ 
+   \hline
+\end{tabular}
+\caption{typeof type system (L1) (functions)}
+\end{table}
+
+Table~\ref{tab:L1toppoly} has the top polymorphic signatures for L1.
+
+% latex table generated in R 3.5.1 by xtable 1.8-3 package
+% Fri Mar 29 18:13:17 2019
+\begin{table}[ht]
+\label{tab:L1toppoly}
+\centering
+\begin{tabular}{lrr}
+  \hline
+type & count & perc \\ 
+  \hline
+double, integer & 19799 & 28.73 \\ 
+  double, list & 5324 & 7.72 \\ 
+  double, NULL & 5152 & 7.48 \\ 
+  character, NULL & 3682 & 5.34 \\ 
+  double, logical & 3388 & 4.92 \\ 
+  character, double & 2842 & 4.12 \\ 
+  list, NULL & 2799 & 4.06 \\ 
+  character, list & 1836 & 2.66 \\ 
+  character, integer & 1663 & 2.41 \\ 
+  character, double, integer & 1362 & 1.98 \\ 
+   \hline
+\end{tabular}
+\caption{Top polymorphic signatures for the L1 type system.}
+\end{table}
+
+\AT{!!! The layout of this section will change drastically. !!!}
+\AT{Old stuff beyond.}
 

### 2019-03-30 paper/main.tex
updating prose in results, getting through L0

@@
 
-\AT{This is the (partial) data for the new layout.
-We need to decide how to tackle L3.}
-
-Figure~\ref{fig:L1argcounts} has arg counts for L1.
-
-Figure~\ref{fig:L1funcounts} has fun counts for L1.
+Having built an understanding of our data collection and analysis pipeline, we will turn our attention to the results of our analysis.
+
+The logical ``baseline'' for the design of a static type system for R is promotion of the language's dynamic types (in the sense of \code{typeof}) to static types; we call this type system L0.
+Such a type system exactly reflects the types of values in the R runtime environment, as the \code{typeof} an object is in direct correspondence with the runtime type tag of the value.
+
+In this section, we will begin by breaking down our corpus of data with L0 in mind.
+From this perspective, we will see how current R programmers see types in the language, and we will leverage our newfound insight to design a type system which should be relevant to them.
+
+Without further ado, let's look at the data, first focusing on a broad overview of the polymorphism in L0.
+Figure~\ref{fig:L1argcounts} shows the polymorphicity of argument positions in R function.
+From there, we build up a picture of the polymorphicity of functions as a whole, and the data for this function polymorphism can be found in Figure~\ref{fig:L1funcounts}.
 
@@
 \end{tabular}
-\caption{typeof type system (L1) (arguments)}
+\caption{typeof type system (L0) (arguments)}
 \end{table}
@@
 \end{tabular}
-\caption{typeof type system (L1) (functions)}
+\caption{typeof type system (L0) (functions)}
 \end{table}
 
-Table~\ref{tab:L1toppoly} has the top polymorphic signatures for L1.
+From Figure~\ref{fig:L1argcounts}, we glean that argument positions aren't very polymorphic in L0.
+An argument which is "Monomorphic in Type" is monomorphic w.r.t. \code{typeof}, in that it was only ever observed to have one \code{typeof}.
+Similarly, an argument is "Monomorphic in Class" if it is only ever called with one {\tt class}.
+As we said earlier, attributes are a little more complicated, and an argument which is "Monomorphic in Attribute Pattern" was only ever observed to have one attribute pattern (and that pattern may itself contain many attributes).
+
+As for the polymorphism of functions as a whole, we glean from Figure~\ref{fig:L1funcounts} that a sizeable number of functions are polymorphic.
+Functions are said to be polymorphic if at least one of their arguments or return are polymorphic.
+\AT{Say more?}
+
+At present, we will only concern ourselves with \code{typeof} polymorphism, and we will later worry about classes and attributes.
+To get an idea of what kinds of polymorphism occur, we will look at the most commonly occurring polymorphic argument signatures.
+Table~\ref{tab:L1toppoly} has the top polymorphic signatures for L0.
 
@@
   \hline
-double, integer & 19799 & 28.73 \\ 
-  double, list & 5324 & 7.72 \\ 
-  double, NULL & 5152 & 7.48 \\ 
-  character, NULL & 3682 & 5.34 \\ 
-  double, logical & 3388 & 4.92 \\ 
-  character, double & 2842 & 4.12 \\ 
-  list, NULL & 2799 & 4.06 \\ 
-  character, list & 1836 & 2.66 \\ 
-  character, integer & 1663 & 2.41 \\ 
-  character, double, integer & 1362 & 1.98 \\ 
+  double, integer & 19,799 & 28.73 \\ 
+  double, list & 5,324 & 7.72 \\ 
+  double, NULL & 5,152 & 7.48 \\ 
+  character, NULL & 3,682 & 5.34 \\ 
+  double, logical & 3,388 & 4.92 \\ 
+  character, double & 2,842 & 4.12 \\ 
+  list, NULL & 2,799 & 4.06 \\ 
+  character, list & 1,836 & 2.66 \\ 
+  character, integer & 1,663 & 2.41 \\ 
+  character, double, integer & 1,362 & 1.98 \\ 
    \hline
 \end{tabular}
-\caption{Top polymorphic signatures for the L1 type system.}
+\caption{Top polymorphic signatures for the L0 type system.}
 \end{table}
+
+Some entries in the Table~\ref{tab:L1toppoly} are clear (e.g., double, integer polymorphism is unsurprising), others are strange (e.g., character, double), but in fact they all have some explanation.
+
+\begin{itemize}
+
+	\item \{double, integer\}: in R, doubles and integers are effectively interchangeable, as casting from one to the other is straightforward and automatically performed when appropriate.
+	For example, \code{c(1, 2)[1.9] == 1}.
+	We can easily consolidate these into one type, say a real number type.
+
+	\item \{double, list\} and \{character, list\}: at first glance, these type signatures are suspect.
+	Why would so many programmers want to pass a double or a list to some function?
+	The key insight here is that doubles and characters are {\it vectorized} in R, meaning that a double is really a vector of doubles.
+	Thus, these seemingly strange signatures are revealed to be entirely reasonable: these signatures are likely to describe arguments which are either vectors or lists.
+
+	\item \{X, NULL\}: in most languages, NULL is an inhabitant of all types, and we see that this is not the case in R.
+	That said, there is a pattern of many functions having default or optional arguments, and in many of these cases NULL is the default value.
+	This is a distinction that might be worth collapsing, and further this indicates that some annotation for optional arguments would be useful.
+
+	\item \{double, logical\}: the first values of type logical that come to mind are TRUE and FALSE, which does little to explain this strange signature.
+	Indeed, the key here is an implementation detail of R's: recall that NAs inhabit every type.
+	As it happens, the default type of NA is logical (e.g., \code{typeof(NA) == "logical"}), and this is likely what is being seen here.
+	We will control for this by distinguishing NAs with their own type, {\tt raw\_NA}, but we will get to that in time.
+
+	\item \{character, double\}, \{character, integer\} and \{character, double, integer\}: yet more strange signatures.
+	These signatures are a little more subtle to explain.
+	For one, indexing lists and vectors can be done with either strings (i.e., to access a named index) or with numbers.
+	Another option is dates: R has many ways to present dates, and strings and numbers feature among them.
+	Finally, many \AT{get numbers} functions in R return a string if an error occurs during execution.
+
+\end{itemize}
 
@@
 
-Having built an understanding of our data collection and analysis
-pipeline, we will turn our attention to the results of our analysis.  We
-approach the data with the purpose of informing the design of a static type
-system for R.
-
-The logical ``baseline'' for the design of a static type system for R is
-promotion of the language's dynamic types (in the sense of \code{typeof}) to
-static types, but there are some clear limitations to this approach.  For
+\AT{Limitations of L0.}
+... but there are some clear limitations to this approach.  For
 example, \code{typeof} makes no distinction between vectors and scalars, as

### 2019-03-30 paper/main.tex
getting through L0

@@
 
-\begin{itemize}
+\begin{enumerate}
 
@@
 	For example, \code{c(1, 2)[1.9] == 1}.
-	We can easily consolidate these into one type, say a real number type.
 
@@
 	That said, there is a pattern of many functions having default or optional arguments, and in many of these cases NULL is the default value.
-	This is a distinction that might be worth collapsing, and further this indicates that some annotation for optional arguments would be useful.
 
@@
 
-\end{itemize}
+\end{enumerate}
+
+How will we deal with this?
+
+\begin{enumerate}
+
+	\item consolidating doubles and integers into a single real-number type would account for all of this polymorphism.
+
+	\item vectors!
+
+	\item if we roll NULLs into every type, we can cleanly deal with these X, NULL signatures.
+	As we have established that an annotation for an optional argument is useful, this will allow us to reveal other interesting patterns.
+
+	\item similarly with NULLs, if we encounter a ...
+
+	\item indices?
+
+\end{enumerate}
 

### 2019-03-30 paper/main.tex
working L1 into sec:results

@@
 
+%
+%
+%
+%
+\subsection{L0}
+
 The logical ``baseline'' for the design of a static type system for R is promotion of the language's dynamic types (in the sense of \code{typeof}) to static types; we call this type system L0.
@@
 
-	\item vectors!
+	\item to get an idea of how often vectors and lists co-occur in signatures, we need to expand our analysis of vectorized types.
+	Notably, when we encounter e.g. a double, we should collect length information to see if the double is a vector of doubles or just a scalar.
+	This will feature in our next type system, L1, which we will get to momentarily.
 
@@
 
-	\item similarly with NULLs, if we encounter a ...
-
-	\item indices?
+	\item similarly with NULLs, we're interested in rolling NAs in to all types.
+	As we mentioned, this is actually what the R runtime does, but a default type needed to be chosen for NAs appearing without context.
+	Subsuming NAs should help to reveal other interesting patterns.
+
+	\item dealing with this polymorphism will be more tricky.
+	We might need to pull class information in (to pick up on dates), and possibly make a signature judgment depending on which argument position has the signature (e.g.
+a \{character, X\} signature on a return).
+	We will come back to this later.
 
@@
 
-\AT{!!! The layout of this section will change drastically. !!!}
-\AT{Old stuff beyond.}
-
-\AT{Limitations of L0.}
-... but there are some clear limitations to this approach.  For
-example, \code{typeof} makes no distinction between vectors and scalars, as
-indeed there is no such distinction in R.  But that doesn't mean that there
-{\it couldn't} be a distinction, as an all-new runtime environment design
-for R might benefit from e.g. not needing to vectorize scalar values.
-
-To address these limitations, we augment the logical naive type system with
-more information which was alluded to in the discussion of our data
-collection methodology.  We will consider unit-length vectors to be scalars,
-collect list member type information to build an accurate list type, and
-promote NAs to their own type.  In addition, we collect attribute and class
-information to distinguish a type for data frames and matrices, as the
-\code{typeof} these is simply {\tt list} in R.  This type system is referred
-to as the {\it Fine-Grained Type System}, or FGTS.  An example of a signature in
-this type system can be found in Figure~\ref{fig:exFGTS}.
+There are some clear limitations to the L0 type system.
+As we mentioned, \code{typeof} makes no distinction between vectors and scalars, as indeed there is no such distinction in R. 
+But as we saw, we might benefit from such a distinction, and indeed an all-new runtime environment design for R might benefit from e.g. not needing to vectorize scalar values.
+
+%
+%
+%
+%
+\subsection{L1}
+
+To address these limitations, we augment the L0 with more fine-grained information.
+We will consider unit-length vectors to be scalars, collect list member type information to build an accurate list type, and promote NAs to their own type. 
+In addition, we collect attribute and class information to distinguish a type for data frames and matrices, as the \code{typeof} these is simply {\tt list} in R. 
+We refer to this as the L1 type system.  An example of a signature in this type system can be found in Figure~\ref{fig:exL1}.
 
@@
 \end{tabular}
-\caption{Example call and FGTS signature}\label{fig:exFGTS}\end{figure}
-
-Broadly, the first question we will address is: {\it how polymorphic is R?} 
-We will answer this in two parts, and first we will look at the number of monomorphic {\it arguments} across all of the packages we analyzed.
-
-%
-%
-%
-%
-\subsection{Argument Polymorphism}
-
-In this section, we will explore the polymorphicity of function argument positions and return values.
-Exploring argument polymorphism will help to motivate a design for whole function signatures.
-
-To begin, let's look at the {\it amount} of polymorphism by way of {\it number of polymorphic arguments}.
-The data can be found in Tables~\ref{tab:argcountsFGTS}~and~\ref{tab:funcountsFGTS}.
+\caption{Example call and L1 signature.}\label{fig:exL1}\end{figure}
+
+Let us begin as we did in L0 by looking at the amount of polymorphism in L1.
+The data can be found in Tables~\ref{tab:argcountsL1}~and~\ref{tab:funcountsL1}.
 \AT{TODO: have these tables be generated automatically}
@@
 \begin{table}[ht]
-\label{tab:argcountsFGTS}
+\label{tab:argcountsL1}
 \centering
@@
 \end{tabular}
-\caption{Account of {\it argument} polymorphism in the FGTS.}
+\caption{Account of {\it argument} polymorphism in L1.}
 \end{table}
@@
 \begin{table}[ht]
-\label{tab:funcountsFGTS}
+\label{tab:funcountsL1}
 \centering
@@
 \end{tabular}
-\caption{Account of {\it function} polymorphism in the FGTS.}
+\caption{Account of {\it function} polymorphism in L1.}
 \end{table}
 
-\AT{We should figure out the pipeline to get the notebook output into here.}
-
-In the tables, ``Full Monomorphic'' denotes monomorphism in type, class, and attribute.
+\AT{We should figure out the pipeline to get the notebook output into here automatically.}
+
 We see that roughly 9\% of all argument positions in R are polymorphic in type.
@@
 These numbers alone are insufficient, so we ask: {\it of the polymorphic arguments, what are the most common patterns?}
-The answer to that question can be found in Table~\ref{tab:toppolyFDTS}.
+The answer to that question can be found in Table~\ref{tab:toppolyL1}.
 
@@
 \begin{table}[ht]
-\label{tab:toppolyFDTS}
+\label{tab:toppolyL1}
 \centering
@@
 \AT{This should probably be discussed earlier.}
-Not pictured here, the most common polymorphic pattern in the FGTS is the {\tt scalar/X}, {\tt vector/X} pattern, but this is dealt with separately:
+Not pictured here, the most common polymorphic pattern in L1 is the {\tt scalar/X}, {\tt vector/X} pattern, but this is dealt with separately:
 If a function argument is observed to be both a unit-length and non-unit-length vector, then it's reasonable to assume that the argument is intended to be a vector, so we collapse the distinction in such cases.
 
-In Table~\ref{tab:toppolyFDTS}, we see the prominence of {\tt double} and {\tt integer} polymorphism.
+In Table~\ref{tab:toppolyL1}, we see the prominence of {\tt double} and {\tt integer} polymorphism.
 In many dynamic languages, primitive double and integer types can easily stand-in for each other as coercion between them is easy.
@@
 Another clear pattern in Table~\ref{tab:toppolyFDTS} is that of NULL, X.
-In many programming languages, NULLs inhabit every type, and this is not the case in our FGTS.
-The FGTS reflects many of R's implementation details, and this distinction is no exception:
+In many programming languages, NULLs inhabit every type, and this is not the case in our L1.
+L1 reflects many of R's implementation details, and this distinction is no exception:
 NULLs are represented as \AT{fill in} in the R runtime.
@@
 
-Not pictured in Table~\ref{tab:toppolyFDTS} are NAs, which are ascribed the type {\tt raw\_NA} in FGTS; NAs in fact inhabit every type, though the default NA type is {\tt logical}.
+Not pictured in Table~\ref{tab:toppolyFDTS} are NAs, which are ascribed the type {\tt raw\_NA} in L1; NAs in fact inhabit every type, though the default NA type is {\tt logical}.
 Just as we fold NULL into all types, we will do the same with NAs.
@@
 %
-%\item {\it The Fine-Grained Type System (FGTS)}: this type systems includes a number of relevant additions.
+%\item {\it The Fine-Grained Type System (L1)}: this type systems includes a number of relevant additions.
 %Chiefly, it distinguishes between vectors and scalars, as it is conceivable that boxing scalars as single-element vectors might be overkill.
@@
 %
-%Of course, one could conceive of a more fine-grained type system than FGTS, but it's unclear that we would gain much in doing so.
+%Of course, one could conceive of a more fine-grained type system than L1, but it's unclear that we would gain much in doing so.
 %Indeed, there is a space to explore in slightly less fine-grained type systems, which make modifications including:
@@
 %similarly, NA is an inhabitant of all types, though NA standing alone has type {\tt logical}.
-%FGTS distinguishes such NAs with the {\tt raw\_NA} type, and if instead we consider NA to inhabit all types we may \AT{win}.
+%L1 distinguishes such NAs with the {\tt raw\_NA} type, and if instead we consider NA to inhabit all types we may \AT{win}.
 %

### 2019-03-31 paper/main.tex
ok

@@
   stringstyle=\color{Gray},
-  numbers=left
+ % numbers=left
 }
@@
 \newcommand{\DATAPKGSPERC}{1.5\%\xspace}
+\newcommand{\METAARGCOUNT}{7,051\xspace}
 
@@
 
-In this section, we will detail our methodology for collecting data.  
-Basically, our aim is to observe arguments and return values of function calls, and from these generalize possible type signatures for the called functions. 
-To this end, we build on an open source tool called \genthat whose purpose is to generate unit tests for R libraries~\cite{issta18}.  
-\genthat achieves this by synthesizing unit tests from recorded function argument and return values, comparing the test against existing ones to avoid generating tests which do not increase code coverage.
-To suit our purposes, we change the existing tool in two main ways: 
-We record argument \emph{shapes} rather than values, and given the limited number of shapes this allowed us to avoid having to run \genthat's code coverage minimization phase. 
-Both changes were beneficial for scalability, allowing us to trace far more calls than was reported in the \genthat work~\cite{issta18}.
-
-\subsection{Example}
-
-To illustrate our approach, consider the following script which adds a double to an integer, and then creates a matrix from a vector, finally adding them together to get a new matrix.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-> 1 + 2L                        # add double and integer
+In this section, we detail our methodology for collecting data.  Our aim is
+to observe arguments and return values of function calls, and from these
+generalize possible type signatures for the called functions.  We base our
+infrastructure on an open source tool called \genthat whose purpose is to
+generate unit tests for R libraries~\cite{issta18}.  \genthat achieves this
+by synthesizing unit tests from recorded function argument and return
+values, comparing the test against existing ones to avoid generating tests
+which do not increase code coverage.  To suit our purposes, we change the
+existing tool in two main ways: we record \emph{shapes} rather than values,
+and ignore the code coverage optimization phase.  Both changes are
+beneficial for scalability, allowing us to trace far more calls.
+
+To illustrate our approach, consider the following script which adds a
+double to an integer, and then creates a matrix from a vector, finally
+adding them together to get a new matrix.
+
+\begin{figure}[!hb]
+\begin{tabular}{ll}\begin{minipage}{5cm}
+{\small\begin{lstlisting}[style=R]
+> 1 + 2L                     
 [1] 3
-> x <- c(1,2,3,4)               # save double vector
-> attr(x,"dim") <- c(2,2)       # assign to "dim" attribute
-> x + c(1,2)                    # add double 3 to each matrix element matrix
+> x <- c(1,2,3,4)            
+> attr(x,"dim") <- c(2,2)    
+> x + c(1,2)                 
      [,1] [,2]
@@
 [2,]    4    6
-\end{lstlisting}}\caption{Example script.}\label{example}\end{figure}
+\end{lstlisting}}
+\end{minipage} & 
+\begin{minipage}{8cm}\small
+\begin{tabular}{rl}
+\tt `+` &\tt D[1], I[1] -> D[1] \\
+\tt c& D[1], D[1], D[1], D[1] -> D[4]\\
+\tt `<-` &\tt S, D[3] -> D[3]\\
+\tt c & \tt D[1], D[1] -> D[2]\\
+\tt `attr<-1`&\tt  S, C, D[2] -> \\
+\tt c &\tt D[1], D[1] -> D[2]\\
+\tt `+` &\tt D[4]{dim=D[2]}, D[2] -> D[4]{dim=D[2]}
+\end{tabular}
+\end{minipage}
+\end{tabular}
+\caption{Example script and recorded signature}\label{example}\end{figure}
 
 The functions being called here are \k{`+`}, \k{`<-`}, \k{c} and
-\k{`attr<-`}.  Some functions syntactic support \AT{?}, addition is infix, and the
-assignment to an attribute is written mixfix. The shapes we would expect to
-record are as follows:
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-`+`: D[1], I[1] -> D[1] # (args: double size 1, integer size 1, return: double size 1)
-c: D[1], D[1], D[1], D[1] -> D[4]
-`<-`: S, D[3] -> D[3]
-c: D[1], D[1] -> D[2]
-`attr<-1`:  S, C, D[2] -> 
-c: D[1], D[1] -> D[2]
-`+`: D[4]{dim=D[2]}, D[2] -> D[4]{dim=D[2]}
-\end{lstlisting}}\caption{Recorded shapes}\label{shapes}\end{figure}
-
+\k{`attr<-`}.  The shapes we would expect to record are as follows:
 In the above we abreviate types, \k{double}, \k{integer}, \k{symbol} and
@@
 
-In R all arguments are passed as promises, and unused arguments will be unevaluated.
-Our tracer does still collect information on these arguments, as whether or not they were used doesn't change the fact that they were in fact passed. 
-In some cases, arguments aren't even intended to be evaluated.
-For example, consider the {\tt magrittr} package, which implements the Tidyverse's pipe functions.
-Refer to Figure~\ref{fig:magpipeex} for a translation of a call to {\tt magrittr}'s pipe ({\tt \%>\%}) function.
+In R all arguments are passed as promises, and unused arguments will be
+unevaluated.  Our tracer does still collect information on these arguments,
+as whether or not they were used doesn't change the fact that they were in
+fact passed.  In some cases, arguments aren't even intended to be evaluated.
+For example, consider the {\tt magrittr} package, which implements the
+Tidyverse's pipe functions.  Refer to Figure~\ref{fig:magpipeex} for a
+translation of a call to {\tt magrittr}'s pipe ({\tt \%>\%}) function.
+
 \begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
@@
 \end{lstlisting}}\caption{Example and rough translation of \%$>$\% call.}\label{fig:magpipeex}\end{figure}
-Our tracer evaluates function arguments, and here \code{rhs = mul_by(3)}, but that's erroneous code (you can't call {\tt mul\_by} without supplying two arguments).
-In this situation, our tracer will still register the entire call, but flag the {\tt rhs} argument to indicate to later stages of the analysis that there was an issue evaluating that argument.
+
+Our tracer evaluates function arguments, and here \code{rhs = mul_by(3)},
+but that's erroneous code (you can't call {\tt mul\_by} without supplying
+two arguments).  In this situation, our tracer will still register the
+entire call, but flag the {\tt rhs} argument to indicate to later stages of
+the analysis that there was an issue evaluating that argument.
 
@@
 
+\AT{Maybe it's best to ignore attributes and classes for now?}
+
 %
@@
 	For example, \code{c(1, 2)[1.9] == 1}.
+	This distinction is mainly just relevant to the implementation, as integers and doubles are stored and dealt with differently in the runtime but programmers only see the distinction when printing.
 
@@
 	As we have established that an annotation for an optional argument is useful, this will allow us to reveal other interesting patterns.
+	\AT{If we feel the need to go into more details on NULL or NA, I have the old paragraphs commented out in the source.}
 
@@
 
-There are some clear limitations to the L0 type system.
+There are some clear limitations to the L0 type system that we should address before working through any of these other issues
 As we mentioned, \code{typeof} makes no distinction between vectors and scalars, as indeed there is no such distinction in R. 
 But as we saw, we might benefit from such a distinction, and indeed an all-new runtime environment design for R might benefit from e.g. not needing to vectorize scalar values.
+This is one of the many modifications we make, all of which are thoroughly explored in the following section.
+
+%In terms of programmer behavior, we observe a pattern of optional arguments throughout several R packages, wherein programmers specify a default argument value which is sometimes NULL.
+%Consider the snippet in Figure~\ref{fig:optnull} for an example.
+%\AT{TODO put that figure here}
+%The distinction between NULL and other types is not particularly useful to the programmer, so we propose to also collapse this distinction.
+%Not pictured in Table~\ref{tab:toppolyFDTS} are NAs, which are ascribed the type {\tt raw\_NA} in L1; NAs in fact inhabit every type, though the default NA type is {\tt logical}.
+%Just as we fold NULL into all types, we will do the same with NAs.
 
@@
 In addition, we collect attribute and class information to distinguish a type for data frames and matrices, as the \code{typeof} these is simply {\tt list} in R. 
-We refer to this as the L1 type system.  An example of a signature in this type system can be found in Figure~\ref{fig:exL1}.
+We refer to this as the L1 type system. 
+An example of a signature in this type system can be found in Figure~\ref{fig:exL1}.
 
@@
 \end{lstlisting}}
+
 \begin{tabular}{@{}r|l|l|l@{}}\hline
@@
 We see that roughly 9\% of all argument positions in R are polymorphic in type.
-The numbers for attribute and class polymorphism are included for
-completeness, but the important number is that of type polymorphism (as our
-method of constructing types sometimes factors in class and attribute).
-Specifically for classes, often the class and \code{typeof} a value will
-line up (refer to Section~\ref{sec:rlang}).
-
-As for functions, roughly 22\% are polymorphic in type, or put differently 22\% have at least one polymorphic argument or a polymorphic return.
+The numbers for attribute and class polymorphism are included for completeness, but the important number is that of type polymorphism (as our method of constructing types sometimes factors in class and attribute).
+Specifically for classes, often the class and \code{typeof} a value will line up (refer to Section~\ref{sec:rlang}).
+
+As for functions, roughly 23\% are polymorphic in type, or put differently 23\% have at least one polymorphic argument or a polymorphic return.
 The discrepancy between this and the polymorphicity of arguments is explained by the tendency of some functions to have {\it many} arguments.
@@
 \end{tabular}
-\caption{Top polymorphic argument signatures in the FDTS.}
+\caption{Top polymorphic argument signatures in the L1.}
 \end{table}
@@
 
-In Table~\ref{tab:toppolyL1}, we see the prominence of {\tt double} and {\tt integer} polymorphism.
-In many dynamic languages, primitive double and integer types can easily stand-in for each other as coercion between them is easy.
-In R, the distinction is mainly just relevant to the implementation, as integers and doubles are stored and dealt with differently in the runtime but programmers only see the distinction when printing.
-We address this by creating a new type, {\it real}, encompassing both.
-The expected coercion is evident throughout R.
-Refer to the snippet in Figure~\ref{fig:realex} for details.
+In Table~\ref{tab:toppolyL1}, we again see the prominence of {\tt double} and {\tt integer} polymorphism.
+As discussed previously, we will address this by creating a new {\it real} type encompassing both.
+Refer back to the snippet in Figure~\ref{fig:realex} for details.
 
@@
 As it happens, over 40\% of monomorphic arguments with primitive vectorized types are always scalar (so, 40\% of {\it monomorphic} occurrences of {\tt scalar/X} or {\tt vector/X} are scalar).
-This is important as it shows that there is value in a scalar annotation, which could be used to communicate to an R runtime that vectorization of a value is not necessary, which may in turn lead to a performance improvement.
-
-Another clear pattern in Table~\ref{tab:toppolyFDTS} is that of NULL, X.
-In many programming languages, NULLs inhabit every type, and this is not the case in our L1.
-L1 reflects many of R's implementation details, and this distinction is no exception:
-NULLs are represented as \AT{fill in} in the R runtime.
-
-In terms of programmer behavior, we observe a pattern of optional arguments throughout several R packages, wherein programmers specify a default argument value which is sometimes NULL.
-Consider the snippet in Figure~\ref{fig:optnull} for an example.
-\AT{TODO put that figure here}
-The distinction between NULL and other types is not particularly useful to the programmer, so we propose to also collapse this distinction.
-
-Not pictured in Table~\ref{tab:toppolyFDTS} are NAs, which are ascribed the type {\tt raw\_NA} in L1; NAs in fact inhabit every type, though the default NA type is {\tt logical}.
-Just as we fold NULL into all types, we will do the same with NAs.
+This shows that there is value in a scalar annotation, which could be used to communicate to an R runtime that vectorization of a value is not necessary, which may in turn lead to a performance improvement.
+
+Another clear pattern in Table~\ref{tab:toppolyL1} is that of NULL, X.
+As we discussed, we will roll NULL (and raw NAs) into other types.
 
@@
 ...
+
+%
+%
+%
+%
+\subsection{Metaprogramming}
+
+\AT{Will probably change.}
+
+Tidyverse owes some of its success to the new syntax defined by some of the included packages.
+One such example is the pipe operator \%$>$\% that we saw in the last section, and indeed this function was highly exercised by many other packages \AT{get counts}.
+We also grappled with the unfortunate truth that arguments to \%$>$\% can't always be evaluated, which causes some annoyances in the analysis.
+
+We would like to know how often packages contain functions designed for metaprogramming.
+As it happens, when arguments fail to evaluate during our analysis even as functions execute successfully, we are actually dealing with instances of metaprogramming.
+Some functions, such as those in the magrittr package, will take one (or more) of their arguments and use them as an AST, performing surgery on the expression to create some new expression that they evaluate.
+Sometimes, these arguments intended for metaprogramming {\it will} successfully evaluate in our tracer, but this is merely chance.
+For example, consider the code snippet in Figure~\ref{fig:metaexworks}.
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+mul_by_w_defaults <- function(x=1, y=1) x * y
+
+4 %>% mul_by_w_defaults(2) == `%>%`(lhs = 4, rhs = mul_by(3)) 
+\end{lstlisting}}\caption{Dates in R.}\label{fig:metaexworks}\end{figure}
+
+In this example, our analysis will successfully evaluate the {\tt rhs} argument
 

### 2019-03-31 notebooks/paper_L123_results_section.Rmd
some more results prose, metaprogramming discussion might have to change with new run results

@@
 # remove "errors"
+# TODO: at a later stage, we should show results about the errors
 los <- lapply(los, function(lofun) lapply(lofun, remove_err_df))
@@
 ```
+
+3. Extras. What about errors? How often do we have metaprogramming.
+```{r}
+meta_los <- lapply(los, function(lof) lapply(lof, annihilate_errors_df))
+```

### 2019-03-31 paper/main.tex
some more results prose, metaprogramming discussion might have to change with new run results

@@
   stringstyle=\color{Gray},
- % numbers=left
+  numbers=left
 }
@@
 
-In this section, we detail our methodology for collecting data.  Our aim is
-to observe arguments and return values of function calls, and from these
-generalize possible type signatures for the called functions.  We base our
-infrastructure on an open source tool called \genthat whose purpose is to
-generate unit tests for R libraries~\cite{issta18}.  \genthat achieves this
-by synthesizing unit tests from recorded function argument and return
-values, comparing the test against existing ones to avoid generating tests
-which do not increase code coverage.  To suit our purposes, we change the
-existing tool in two main ways: we record \emph{shapes} rather than values,
-and ignore the code coverage optimization phase.  Both changes are
-beneficial for scalability, allowing us to trace far more calls.
-
-To illustrate our approach, consider the following script which adds a
-double to an integer, and then creates a matrix from a vector, finally
-adding them together to get a new matrix.
-
-\begin{figure}[!hb]
-\begin{tabular}{ll}\begin{minipage}{5cm}
-{\small\begin{lstlisting}[style=R]
-> 1 + 2L                     
+In this section, we will detail our methodology for collecting data.  
+Basically, our aim is to observe arguments and return values of function calls, and from these generalize possible type signatures for the called functions. 
+To this end, we build on an open source tool called \genthat whose purpose is to generate unit tests for R libraries~\cite{issta18}.  
+\genthat achieves this by synthesizing unit tests from recorded function argument and return values, comparing the test against existing ones to avoid generating tests which do not increase code coverage.
+To suit our purposes, we change the existing tool in two main ways: 
+We record argument \emph{shapes} rather than values, and given the limited number of shapes this allowed us to avoid having to run \genthat's code coverage minimization phase. 
+Both changes were beneficial for scalability, allowing us to trace far more calls than was reported in the \genthat work~\cite{issta18}.
+
+\subsection{Example}
+
+To illustrate our approach, consider the following script which adds a double to an integer, and then creates a matrix from a vector, finally adding them together to get a new matrix.
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+> 1 + 2L                        # add double and integer
 [1] 3
-> x <- c(1,2,3,4)            
-> attr(x,"dim") <- c(2,2)    
-> x + c(1,2)                 
+> x <- c(1,2,3,4)               # save double vector
+> attr(x,"dim") <- c(2,2)       # assign to "dim" attribute
+> x + c(1,2)                    # add double 3 to each matrix element matrix
      [,1] [,2]
@@
 [2,]    4    6
-\end{lstlisting}}
-\end{minipage} & 
-\begin{minipage}{8cm}\small
-\begin{tabular}{rl}
-\tt `+` &\tt D[1], I[1] -> D[1] \\
-\tt c& D[1], D[1], D[1], D[1] -> D[4]\\
-\tt `<-` &\tt S, D[3] -> D[3]\\
-\tt c & \tt D[1], D[1] -> D[2]\\
-\tt `attr<-1`&\tt  S, C, D[2] -> \\
-\tt c &\tt D[1], D[1] -> D[2]\\
-\tt `+` &\tt D[4]{dim=D[2]}, D[2] -> D[4]{dim=D[2]}
-\end{tabular}
-\end{minipage}
-\end{tabular}
-\caption{Example script and recorded signature}\label{example}\end{figure}
+\end{lstlisting}}\caption{Example script.}\label{example}\end{figure}
 
 The functions being called here are \k{`+`}, \k{`<-`}, \k{c} and
-\k{`attr<-`}.  The shapes we would expect to record are as follows:
+\k{`attr<-`}.  Some functions syntactic support \AT{?}, addition is infix, and the
+assignment to an attribute is written mixfix. The shapes we would expect to
+record are as follows:
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+`+`: D[1], I[1] -> D[1] # (args: double size 1, integer size 1, return: double size 1)
+c: D[1], D[1], D[1], D[1] -> D[4]
+`<-`: S, D[3] -> D[3]
+c: D[1], D[1] -> D[2]
+`attr<-1`:  S, C, D[2] -> 
+c: D[1], D[1] -> D[2]
+`+`: D[4]{dim=D[2]}, D[2] -> D[4]{dim=D[2]}
+\end{lstlisting}}\caption{Recorded shapes}\label{shapes}\end{figure}
+
 In the above we abreviate types, \k{double}, \k{integer}, \k{symbol} and
@@
 
-In R all arguments are passed as promises, and unused arguments will be
-unevaluated.  Our tracer does still collect information on these arguments,
-as whether or not they were used doesn't change the fact that they were in
-fact passed.  In some cases, arguments aren't even intended to be evaluated.
-For example, consider the {\tt magrittr} package, which implements the
-Tidyverse's pipe functions.  Refer to Figure~\ref{fig:magpipeex} for a
-translation of a call to {\tt magrittr}'s pipe ({\tt \%>\%}) function.
-
+In R all arguments are passed as promises, and unused arguments will be unevaluated.
+Our tracer does still collect information on these arguments, as whether or not they were used doesn't change the fact that they were in fact passed. 
+In some cases, arguments aren't even intended to be evaluated.
+For example, consider the {\tt magrittr} package, which implements the Tidyverse's pipe functions.
+Refer to Figure~\ref{fig:magpipeex} for a translation of a call to {\tt magrittr}'s pipe ({\tt \%>\%}) function.
 \begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
@@
 \end{lstlisting}}\caption{Example and rough translation of \%$>$\% call.}\label{fig:magpipeex}\end{figure}
-
-Our tracer evaluates function arguments, and here \code{rhs = mul_by(3)},
-but that's erroneous code (you can't call {\tt mul\_by} without supplying
-two arguments).  In this situation, our tracer will still register the
-entire call, but flag the {\tt rhs} argument to indicate to later stages of
-the analysis that there was an issue evaluating that argument.
+Our tracer evaluates function arguments, and here \code{rhs = mul_by(3)}, but that's erroneous code (you can't call {\tt mul\_by} without supplying two arguments).
+In this situation, our tracer will still register the entire call, but flag the {\tt rhs} argument to indicate to later stages of the analysis that there was an issue evaluating that argument.
 

### 2019-04-01 notebooks/paper_L123_results_section.Rmd
more prose

@@
 
-3. Extras. What about errors? How often do we have metaprogramming.
+3. Extra. What about errors? How often do we have metaprogramming.
 ```{r}
+# This can change depending on how we address metaprogramming.
 meta_los <- lapply(los, function(lof) lapply(lof, annihilate_errors_df))
 ```
+
+Step 4. Translate to L3?
+What is L3?
+We want class info and attribute info, on some level.
+Let's do classes first.
+```{r}
+# get rid of classes we have dealt with.
+los <- lapply(los, function(lof) lapply(lof, class_to_primitive_df))
+```
+
+How do we want to deal with the classes?
+Should they become a type, if they are not primitive?
+Let's look at class poly counts maybe.
+```{r}
+L3_c_basic_df <- build_arg_count_df_from_lopkgs(los)
+
+L3_poly_class_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_class_args))
+L3_poly_class_args <- L3_poly_class_args[order(-L3_poly_class_args$count),] # sort by count
+L3_poly_class_args$perc <- round(L3_poly_class_args$count/sum(L3_poly_class_args$count) * 100, 2)
+```
+
+All the numbers are low, so there's no clear pattern, which makes sense.
+How best to incorporate this...
+```{r}
+
+```

### 2019-04-01 paper/main.tex
more prose

--- 2019-03-31 paper/main.tex
@@
 }
+
 \newcommand{\code}[1]{\lstinline|#1|\xspace}
-
 \newcommand{\genthat}{{\sc Genthat}\xspace}
@@
   Monomorphic in Type & 834K & 91.23 \\ 
-  Monomorphic in Class & 827K & 90.42 \\ 
-  Monomorphic in Attribute Pattern & 825K & 90.27 \\ 
+%  Monomorphic in Class & 827K & 90.42 \\ 
+%  Monomorphic in Attribute Pattern & 825K & 90.27 \\ 
   Total Seen & 914K & --- \\ 
@@
   Monomorphic in Type & 167K & 77.25 \\ 
-  Monomorphic in Class & 163K & 75.24 \\ 
-  Monomorphic in Attribute Pattern & 161K & 74.37 \\ 
+%  Monomorphic in Class & 163K & 75.24 \\ 
+%  Monomorphic in Attribute Pattern & 161K & 74.37 \\ 
   Total Seen & 216K & --- \\ 
@@
 
-We see that roughly 9\% of all argument positions in R are polymorphic in type.
-The numbers for attribute and class polymorphism are included for completeness, but the important number is that of type polymorphism (as our method of constructing types sometimes factors in class and attribute).
-Specifically for classes, often the class and \code{typeof} a value will line up (refer to Section~\ref{sec:rlang}).
+We see that roughly 9\% of all argument positions in L1 are polymorphic in type, as opposed to 7.5\% in L0.
+This small difference is perhaps surprising, as we did define a number of additional types and might expect for there to be a lot more polymorphism.
+This is explained by the fact that drawing the distinctions added a lot of polymorphism to already polymorphic arguments.
+\AT{We can get numbers on this.}
 
 As for functions, roughly 23\% are polymorphic in type, or put differently 23\% have at least one polymorphic argument or a polymorphic return.
-The discrepancy between this and the polymorphicity of arguments is explained by the tendency of some functions to have {\it many} arguments.
-
-These numbers alone are insufficient, so we ask: {\it of the polymorphic arguments, what are the most common patterns?}
+Again, the small difference between L0 and L1 in this regard is explained by a tendency to make already polymorphic functions more polymorphic.
+
+As before, these numbers alone are insufficient, so we ask again: {\it of the polymorphic arguments, what are the most common patterns?}
 The answer to that question can be found in Table~\ref{tab:toppolyL1}.
@@
 
-\AT{This should probably be discussed earlier.}
-Not pictured here, the most common polymorphic pattern in L1 is the {\tt scalar/X}, {\tt vector/X} pattern, but this is dealt with separately:
-If a function argument is observed to be both a unit-length and non-unit-length vector, then it's reasonable to assume that the argument is intended to be a vector, so we collapse the distinction in such cases.
+We would like to make one important note before digesting the data in Table~\ref{tab:toppolyL1}.
+We drew a distinction between scalars and vectors of the same type, and that distinction would be the single largest source of polymorphism in L1 had we not dealt with it appropriately.
+We note that an argument which exhibits a signature including e.g.
+\{scalar/double, vector/double\} really must be an argument designed to take vectors, as recall that unit length vectors appear as scalars by our definition.
+We have already collapsed this distinction.
 
@@
 Refer back to the snippet in Figure~\ref{fig:realex} for details.
+Also, another clear pattern in Table~\ref{tab:toppolyL1} is that of NULL, X.
+As we discussed, we will roll NULL (and raw NAs) into other types.
+
+%
+%
+\subsubsection{Scalars}
 
 As we make the distinction between vectors and scalars, one other interesting question that arises is: {\it how many arguments are strictly scalar?}
-As it happens, over 40\% of monomorphic arguments with primitive vectorized types are always scalar (so, 40\% of {\it monomorphic} occurrences of {\tt scalar/X} or {\tt vector/X} are scalar).
+As it happens, over 40\% of monomorphic arguments with primitive vectorized types are always scalar (so, 40\% of {\it monomorphic} occurrences of {\tt scalar/X} and {\tt vector/X} are scalar).
 This shows that there is value in a scalar annotation, which could be used to communicate to an R runtime that vectorization of a value is not necessary, which may in turn lead to a performance improvement.
 
-Another clear pattern in Table~\ref{tab:toppolyL1} is that of NULL, X.
-As we discussed, we will roll NULL (and raw NAs) into other types.
-
-The next type system design we will consider is ... .
-The number of monomorphic arguments and functions can be found in Tables~\ref{tab:argcountsWHATTS} and~\ref{tab:funcountsWHATTS} respectively.
+At this point, we have seen a number of reasonable polymorphic patterns that arise in R, both in L0 and L1.
+The next type system design we consider will deal with these, and we refer to it as L$1^{+}$.
+The number of monomorphic arguments and functions can be found in Tables~\ref{tab:argcountsL1p} and~\ref{tab:funcountsL1p} respectively.
 
@@
 \begin{table}[ht]
-\label{tab:argcountsWHATTS}
+\label{tab:argcountsL1p}
 \centering
@@
 \end{tabular}
-\caption{Account of {\it argument} polymorphism in the NEW TYPE SYSTEM.}
+\caption{Account of {\it argument} polymorphism in L$1^{+}$.}
 \end{table}
@@
 \begin{table}[ht]
-\label{tab:funcountsWHATTS}
+\label{tab:funcountsL1p}
 \centering
@@
 \end{tabular}
-\caption{Account of {\it function} polymorphism in the NEW TYPE SYSTEM.}
+\caption{Account of {\it function} polymorphism in L$1^{+}$.}
 \end{table}
@@
 To get an idea of what more we can do, let's see what the most frequent remaining polymorphic signatures are.
-See Table~\ref{tab:toppolyWHATTS}.
+See Table~\ref{tab:toppolyL1p}.
 
@@
 \begin{table}[ht]
-\label{tab:toppolyWHATTS}
+\label{tab:toppolyL1p}
 \centering
@@
 \end{tabular}
-\caption{Top 10 argument signatures in the NEW TYPE SYSTEM.}
+\caption{Top 10 argument signatures in L$1^{+}$.}
 \end{table}
@@
 \AT{We should figure out how we want to deal with this polymorphism.}
+
+\AT{list<list> might be objects?}
 
@@
 That said, translating between the data types is straightforward.
+
+We have seen that once we have done away with trivial sources of polymorphism, a number of interesting patterns arise.
+In many of these cases, we can appeal to other sources of reflection in R to expand our understanding.
+For example, to deal with the possibility of dates, we can appeal to the \code{class} of values, and to gain more information on matrices we might be interested in looking at attributes.
+The next type system addresses these concerns.
+
+%
+%
+%
+%
+\subsection{L2}
+
+\AT{Eyy time for L2.}
 

### 2019-04-01 paper/main.tex
some type name macros

@@
 
+% Macros for type names?
+\newcommand{\reals}{$real$\xspace}
+\newcommand{\realv}{$\overline{real}$\xspace}
+\newcommand{\chars}{$char$\xspace}
+\newcommand{\charv}{$\overline{char}$\xspace}
+\newcommand{\logis}{$logical$\xspace}
+\newcommand{\logiv}{$\overline{logical}$\xspace}
+\newcommand{\dbls}{$double$\xspace}
+\newcommand{\dblv}{$\overline{double}$\xspace}
+\newcommand{\ints}{$integer$\xspace}
+\newcommand{\intv}{$\overline{integer}$\xspace}
+
 \setcopyright{none}
@@
 used to capture most of the interesting cases.
+
 \end{abstract}
@@
 
-	\item if the analysis encounters \code{ typeof(x) == "double"}, it will look at the length of {\tt x} to determine if {\tt x} is a scalar or a vector, generating the annotation {\tt scalar/double} or {\tt vector/double} accordingly.
+	\item if the analysis encounters \code{ typeof(x) == "double"}, it will look at the length of {\tt x} to determine if {\tt x} is a scalar or a vector, generating the annotation {\tt \dbls} or {\tt \dblv} accordingly.
 	This is also true for integer, complex, raw, logical, and character {\tt typeof}s (i.e., the primitive vectorized data types).
@@
   }
-\end{lstlisting}}\caption{Example of a data.frame/(vector/double) argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}
+\end{lstlisting}}\caption{Example of a data.frame/(\dblv) argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}
 
@@
     ...
-\end{lstlisting}}\caption{Example of matrix/(vector/double) argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}
+\end{lstlisting}}\caption{Example of matrix/(\dblv) argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}
 
@@
 \bf Argument & \bf Type & \bf Class &\bf Attributes \\
-x & \tt scalar/double & \tt numeric & \tt \{\} \\
-y & \tt vector/double & \tt numeric & \tt \{\} \\
-retv & \tt list<double> & \tt list & \tt \{example: scalar/character\}  
+x & \dbls & numeric & \{\} \\
+y & \dblv & numeric & \{\} \\
+retv & list<double> & list & \{\xspace example: \chars\}  
 \end{tabular}
@@
   \hline
-  vector/double, vector/integer & 7742 & 9.66 \\ 
-  scalar/double, scalar/integer & 7201 & 8.99 \\ 
-  matrix, vector/double & 5384 & 6.72 \\ 
-  NULL, scalar/character & 2325 & 2.90 \\ 
+  \dblv, \intv & 7742 & 9.66 \\ 
+  \dbls, \ints & 7201 & 8.99 \\ 
+  matrix, \dblv & 5384 & 6.72 \\ 
+  NULL, \chars & 2325 & 2.90 \\ 
   list$<$any$>$, list$<$double$>$ & 2231 & 2.78 \\ 
   data.frame, matrix & 2116 & 2.64 \\ 
-  NULL, scalar/double & 2082 & 2.60 \\ 
-  NULL, vector/double & 2069 & 2.58 \\ 
+  NULL, \dbls & 2082 & 2.60 \\ 
+  NULL, \dblv & 2069 & 2.58 \\ 
   list$<$any$>$, list$<$list$>$ & 1432 & 1.79 \\ 
-  NULL, vector/character & 1380 & 1.72 \\
+  NULL, \charv & 1380 & 1.72 \\
      \hline
@@
 We note that an argument which exhibits a signature including e.g.
-\{scalar/double, vector/double\} really must be an argument designed to take vectors, as recall that unit length vectors appear as scalars by our definition.
+\{\dbls, \dblv\} really must be an argument designed to take vectors, as recall that unit length vectors appear as scalars by our definition.
 We have already collapsed this distinction.
@@
   \hline
-  matrix, vector/real & 6360 & 13.94 \\ 
+  matrix, \realv & 6360 & 13.94 \\ 
   list$<$any$>$, list$<$real$>$ & 2799 & 6.13 \\ 
-  vector/character, vector/real & 2774 & 6.08 \\ 
-  scalar/character, scalar/real & 2552 & 5.59 \\ 
+  \charv, \realv & 2774 & 6.08 \\ 
+  \chars, \reals & 2552 & 5.59 \\ 
   data.frame, matrix & 2167 & 4.75 \\ 
@@
   list$<$any$>$, list$<$character$>$ & 1214 & 2.66 \\ 
-  data.frame, vector/real & 1046 & 2.29 \\ 
-  scalar/logical, scalar/real & 920 & 2.02 \\ 
-  scalar/character, vector/real & 868 & 1.90 \\
+  data.frame, \realv & 1046 & 2.29 \\ 
+  \logis, \reals & 920 & 2.02 \\ 
+  \chars, \realv & 868 & 1.90 \\
    \hline
@@
 A few of the top polymorphic signatures stand out.
-First, the matrix/(vector/real) signature is interesting: \AT{as we showed in our polymorphism examples section}, converting between vectors and matrices is very simple, and indeed numeric vectors are mathematically {\tt 1 x n} matrices.
+First, the matrix/\realv signature is interesting: \AT{as we showed in our polymorphism examples section}, converting between vectors and matrices is very simple, and indeed numeric vectors are mathematically {\tt 1 x n} matrices.
 In fact, in R matrices can be used identically to vectors in many cases, and we refer the reader to Figure~\ref{fig:matasvec} for examples.

### 2019-04-01 paper/main.tex
cleaned up more of L1 prose, working on data and prose for L2

@@
 To get an idea of what kinds of polymorphism occur, we will look at the most commonly occurring polymorphic argument signatures.
-Table~\ref{tab:L1toppoly} has the top polymorphic signatures for L0.
+Table~\ref{tab:L0toppoly} has the top polymorphic signatures for L0.
 
@@
 \begin{table}[ht]
-\label{tab:L1toppoly}
+\label{tab:L0toppoly}
 \centering
@@
 
-Some entries in the Table~\ref{tab:L1toppoly} are clear (e.g., double, integer polymorphism is unsurprising), others are strange (e.g., character, double), but in fact they all have some explanation.
+Some entries in the Table~\ref{tab:L0toppoly} are clear (e.g., double, integer polymorphism is unsurprising), others are strange (e.g., character, double), but in fact they all have some explanation.
 
@@
 \end{tabular}
-\caption{Top polymorphic argument signatures in the L1.}
+\caption{Top polymorphic argument signatures in L1.}
 \end{table}
@@
 A few of the top polymorphic signatures stand out.
-First, the matrix/\realv signature is interesting: \AT{as we showed in our polymorphism examples section}, converting between vectors and matrices is very simple, and indeed numeric vectors are mathematically {\tt 1 x n} matrices.
+First, the matrix/\realv signature seems interesting, converting between vectors and matrices is very simple, and indeed numeric vectors are mathematically {\tt 1 x n} matrices.
 In fact, in R matrices can be used identically to vectors in many cases, and we refer the reader to Figure~\ref{fig:matasvec} for examples.
@@
 
+This similarity is reminiscent of another pattern which was more clear in Table~\ref{tab:L0toppoly}; that of list/vector polymorphism.
+Transforming between vectors and matrices is just as easy as transforming between lists and vectors is simple.
 \AT{We should figure out how we want to deal with this polymorphism.}
 
-\AT{list<list> might be objects?}
-
-Another significant source of apparently strange polymorphism is the {\tt character}, {\tt real} polymorphism.
+Another pattern that we can shed some light on the list$<$any$>$/list$<$X$>$ pattern.
+These data points are an artifact of R's three object systems: S3, S4, and R5.
+An S3 object is any value with a ``class'' attribute, and indeed many of these tend to be lists.
+For example, an S3-style point might be a list with an entry for each dimension in the point, with a ``class'' attribute set to ``Point''.
+Our next type system, L2, will account for these.
+
+Another significant source of apparently strange polymorphism is the character, real polymorphism.
 Recall Figure~\ref{fig:chardbl}, which laid out an example of this polymorphic signature: lists and vectors can be accessed by strings as well as by numerics.
@@
 
-Another option is a date.
+Another possible explanation for character, real polymorphism is the date data type.
 In R, dates are often represented as either strings or numbers, and many packages make use of dates.
-Dates (sometimes) manifest themselves in the {\tt class} of a value, for examples refer to Figure~\ref{fig:date}.
+Dates (sometimes) manifest themselves in the class of a value, for examples refer to Figure~\ref{fig:date}.
 \begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
@@
 
-This figure reveals an unfortunate truth that date ``data type'' in R is not standardized.
-That said, translating between the data types is straightforward.
+This figure reveals an unfortunate truth that date data type in R is not standardized.
+That said, as before with matrices and vectors, and vectors and lists, translating between the data types is straightforward.
+Once again, this polymorphism can be captured with classes, which is the focus of our next discussion.
 

### 2019-04-02 paper/main.tex
changes all over

@@
 
+\newcommand{\attr}[2]{\ensuremath{#1\,({\mathit #2})}\xspace}
+\renewcommand{\to}{\ensuremath{\rightarrow}\xspace}
+\newcommand{\D}{\ensuremath{\small\vec{\mathtt D}}\xspace} % Double
+\newcommand{\I}{\ensuremath{\small\vec{\mathtt I}}\xspace} % Integer
+\renewcommand{\C}{\ensuremath{\small\vec{\mathtt C}}\xspace} % Character
+\renewcommand{\L}{\ensuremath{\small\vec{\mathtt L}}\xspace} % Logical
+\newcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
+\newcommand{\X}{\ensuremath{\small\vec{\mathtt X}}\xspace} % Complex
+\newcommand{\Y}{\ensuremath{\small\vec{\mathtt Y}}\xspace} % Symbol
+\newcommand{\sY}{\ensuremath{\small{\mathtt Y}}\xspace} % Symbol
+\newcommand{\sS}{\ensuremath{\small{\mathtt S}}\xspace} % S4
+\newcommand{\sF}{\ensuremath{\small{\mathtt F}}\xspace} % Closure
+\newcommand{\sE}{\ensuremath{\small{\mathtt E}}\xspace} % Env
+\renewcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
+\newcommand{\sN}{\ensuremath{\small{\mathtt N}}\xspace}     % Null
+\renewcommand{\l}{\ensuremath{\small\underline{\mathtt ?}}\xspace}     % List
+\newcommand{\sD}{\ensuremath{\small{\mathtt D}}\xspace} % Double
+\newcommand{\sI}{\ensuremath{\small{\mathtt I}}\xspace} % Integer
+\newcommand{\sC}{\ensuremath{\small{\mathtt C}}\xspace} % Character
+\newcommand{\sL}{\ensuremath{\small{\mathtt L}}\xspace} % Logical
+\newcommand{\sX}{\ensuremath{\small{\mathtt X}}\xspace} % Complex
+\newcommand{\sR}{\ensuremath{\small{\mathtt R}}\xspace} % Raw
+\newcommand{\ANY}{\ensuremath{\small{\mathtt ?}}\xspace}     % Any
+\newcommand{\lT}[1]{\ensuremath{\small\underline{\mathtt #1}}\xspace}     % Any
+
 \begin{abstract}
@@
 \begin{wrapfigure}{r}{6.1cm}
-\footnotesize\begin{tabular}{l|l@{}}\hline
-\multicolumn{2}{l}{\bf Vectorized data types:}  \\\hline
-\k{logical}  & vector of boolean values\\
-\k{integer}   & vector of 32 bit integer values\\
-\k{double} & vector of 64 bit floating points\\
-\k{complex} & vector of complex values\\
-\k{character} & vector of strings values\\
-\k{raw} & vector of bytes\\
-\k{list} & vector of values of any type\\\hline
-\multicolumn{2}{l}{\bf Scalar data types:}\\\hline
-\k{NULL}  &  singleton null value\\
-\k{S4}    &  instance of a S4 class \\
-\k{closure} & a function with its environment\\
-\k{environment} & a mapping from symbol to value \\\hline
-\multicolumn{2}{l}{\bf Implementation data types:}\\\hline
-\multicolumn{2}{l}{\k{special},
+\footnotesize\begin{tabular}{l|c|l@{}}\hline
+\multicolumn{3}{l}{\bf Vectorized data types:}  \\\hline
+\k{logical}   & \L & vector of boolean values\\
+\k{integer}   & \I & vector of 32 bit integer values\\
+\k{double}    & \D & vector of 64 bit floating points\\
+\k{complex}   & \X & vector of complex values\\
+\k{character} & \C & vector of strings values\\
+\k{raw}       & \R & vector of bytes\\
+\k{list}      & \l & vector of values of any type\\\hline
+\multicolumn{3}{l}{\bf Scalar data types:}\\\hline
+\k{NULL}      & \sN &  singleton null value\\
+\k{S4}        & \sS &  instance of a S4 class \\
+\k{closure}   & \sF & a function with its environment\\
+\k{environment}&\sE &  a mapping from symbol to value \\\hline
+\multicolumn{3}{l}{\bf Implementation data types:}\\\hline
+\multicolumn{3}{l}{\k{special},
 \k{builtin},
-\k{symbol},
+\k{symbol} (\sY),
 \k{pairlist},
 \k{promise}}\\
-\multicolumn{2}{l}{
+\multicolumn{3}{l}{
 \k{language},
@@
 }\\
-\multicolumn{2}{l}{
+\multicolumn{3}{l}{
 \k{externalprt},
@@
 \begin{tabular}{rl}
-\tt `+` &\tt D[1], I[1] -> D[1] \\
-\tt c& D[1], D[1], D[1], D[1] -> D[4]\\
-\tt `<-` &\tt S, D[3] -> D[3]\\
-\tt c & \tt D[1], D[1] -> D[2]\\
-\tt `attr<-1`&\tt  S, C, D[2] -> \\
-\tt c &\tt D[1], D[1] -> D[2]\\
-\tt `+` &\tt D[4]{dim=D[2]}, D[2] -> D[4]{dim=D[2]}
+\tt `+`: &\tt \sD \sI \to \sD \\
+\tt c:& \sD \sD \sD \sD \to \D\\
+\tt `<-`: &\tt \sY \D \to \D\\
+\tt c: & \tt \sD \sD \to \D\\
+\tt `attr<-`:&\tt  \sY \C \D \to \\
+\tt c: &\tt \sD \sD \to \D\\
+\tt `+`: &\tt \attr\D{dim=\D} \D \to \attr\D{dim=\D}
 \end{tabular}
@@
 
-\section{Stuff}
-
-We will need a notion of polymorphism, and thus a notion of type, when undertaking this analysis. 
-R is unityped, and the only notion of polymorphism in the language is its use of dynamic dispatch on the class of an argument.  
-For our purposes, we will define a notion in type roughly in line with R's \code{typeof} reflection function, in some cases augmented by information obtained by \code{class} and \code{attributes} reflection.
-We start by defining what it means for an {\it argument} to be polymorphic.
+\section{Types for R}
+
+We will need a notion of polymorphism, and thus a notion of type, when
+undertaking this analysis.  R is unityped, and the only notion of
+polymorphism in the language is its use of dynamic dispatch on the class of
+an argument.  For our purposes, we will define a notion in type roughly in
+line with R's \code{typeof} reflection function, in some cases augmented by
+information obtained by \code{class} and \code{attributes} reflection.  We
+start by defining what it means for an {\it argument} to be polymorphic.
 
@@
   least two distinct types according to our definition of type.
-
-   \item an argument is {\it polymorphic in class} if it has been called with at least two distinct classes according to the \code{class} function.
-
+\item an argument is {\it polymorphic in class} if it has been called with
+  at least two distinct classes according to the \code{class} function.
 \item attributes are a bit more tricky.  We can't simply count the number of
@@
   at least two distinct attribute patterns.
-
 \end{itemize}
 
-The main ingredient of our types is the {\tt typeof} a value, and there is an important distinction to be made between the result of \code{typeof} and that of \code{class} and \code{attributes}.
-In R, the \code{typeof} a value cannot change, and this is not the case with class and attributes.  
-As mentioned in \AT{an earlier section}, the class of values can be easily redefined by assigning to the value's class attribute, and similarly attributes can be dynamically added, removed, or changed.
-
-While we do consider some class and attribute information when constructing our types, it is also valuable to consider them entirely separate from types.
-Users are free to define their own classes, and arbitrarily add and remove attributes, and interesting patterns might arise in considering them.
-That said, not attribute patterns and classes are necessarily interesting, and indeed some of R's built-in data structures use them to some extend.
-For one, data frames have a {\tt data.frame} class (explicitly on the class attribute).
-Other standard patterns are:
+The main ingredient of our types is the {\tt typeof} a value, and there is
+an important distinction to be made between the result of \code{typeof} and
+that of \code{class} and \code{attributes}.  In R, the \code{typeof} a value
+cannot change, and this is not the case with class and attributes.  As
+mentioned in \AT{an earlier section}, the class of values can be easily
+redefined by assigning to the value's class attribute, and similarly
+attributes can be dynamically added, removed, or changed.
+
+While we do consider some class and attribute information when constructing
+our types, it is also valuable to consider them entirely separate from
+types.  Users are free to define their own classes, and arbitrarily add and
+remove attributes, and interesting patterns might arise in considering them.
+That said, not attribute patterns and classes are necessarily interesting,
+and indeed some of R's built-in data structures use them to some extend.
+For one, data frames have a {\tt data.frame} class (explicitly on the class
+attribute).  Other standard patterns are:
 
 \begin{itemize}
-
 \item data frames have three attributes: {\tt names} for column names, {\tt
@@
   "data.frame"} upon creation).
-
 \item matrices have at least a {\tt dim} attribute with matrix dimensions,
-  and optionally a {\tt dimnames} attribute for dimension names.
-Curiously, the {\tt dim} attribute is all that is needed to define a matrix (i.e., a list with a {\tt dim} attribute {\it is} a matrix).
-
+  and optionally a {\tt dimnames} attribute for dimension names. Curiously,
+  the {\tt dim} attribute is all that is needed to define a matrix (i.e., a
+  list with a {\tt dim} attribute {\it is} a matrix).
 \item lists and vectors may have a {\tt names} attribute, which assigns
   names to locations.
-
 \item to ease dealing with large sorted sets of non-numeric values, R offers
@@
   factored to \texttt{"factor"}.
-
 \end{itemize}
@@
 
-%
-%
-%
-%
 \subsection{Granularity of Type Information}
 
-When constructing a type for a value, we begin with the {\tt typeof} the value.
-This alone paints an incomplete picture of the value, though, especially when it comes to vectors (recall that \code{typeof(5) == typeof(c(5, 5)) == "double"}).
-In some cases, the class or attributes of a value might enlighten us, but in many cases we can be a little more precise.
+When constructing a type for a value, we begin with the {\tt typeof} the
+value.  This alone paints an incomplete picture of the value, though,
+especially when it comes to vectors (recall that \code{typeof(5) ==
+  typeof(c(5, 5)) == "double"}).  In some cases, the class or attributes of
+a value might enlighten us, but in many cases we can be a little more
+precise.
 
@@
 \begin{itemize}
-
-	\item if the analysis encounters \code{ typeof(x) == "double"}, it will look at the length of {\tt x} to determine if {\tt x} is a scalar or a vector, generating the annotation {\tt \dbls} or {\tt \dblv} accordingly.
-	This is also true for integer, complex, raw, logical, and character {\tt typeof}s (i.e., the primitive vectorized data types).
-	A notable imperfection here is is that unit length vectors will appear as scalars, so later in our analysis we ensure that argument signatures containing both scalar/X and vector/X are collapsed down to contain only vector/X.
-
-	\item if the analysis encounters \code{typeof(x) == "list"}, we collect type information on all list elements in order to ascribe a general type to the contents of the list.
-	To avoid undue slowdowns, we only collect the \code{typeof} the contents, and ascribe type {\tt any} if two or more {\tt typeof}s are observed.
-
-	\item if the analysis encounters \code{NA}, we ascribe a unique {\tt raw\_NA} type.
-	In R, \code{NA} inhabits all types, but \code{typeof(NA) == "logical"}.
-	The analysis deals with these {\tt raw\_NA}s at a later stage.
-
-	\item if the analysis encounters a matrix (i.e. the value has matrix class and has a dims and optionally dimnames attribute), we ascribe the type {\tt matrix}.
-
-	\item similarly, if it encounters a data frame (i.e., class {\tt data.frame} with appropriate attributes), we ascribe the type {\tt data.frame}.
-
+\item if the analysis encounters \code{ typeof(x) == "double"}, it will look
+  at the length of {\tt x} to determine if {\tt x} is a scalar or a vector,
+  generating the annotation {\tt \dbls} or {\tt \dblv} accordingly.  This is
+  also true for integer, complex, raw, logical, and character {\tt typeof}s
+  (i.e., the primitive vectorized data types).  A notable imperfection here
+  is is that unit length vectors will appear as scalars, so later in our
+  analysis we ensure that argument signatures containing both scalar/X and
+  vector/X are collapsed down to contain only vector/X.
+
+\item if the analysis encounters \code{typeof(x) == "list"}, we collect type
+  information on all list elements in order to ascribe a general type to the
+  contents of the list.  To avoid undue slowdowns, we only collect the
+  \code{typeof} the contents, and ascribe type {\tt any} if two or more {\tt
+    typeof}s are observed.
+
+\item if the analysis encounters \code{NA}, we ascribe a unique {\tt
+  raw\_NA} type.  In R, \code{NA} inhabits all types, but \code{typeof(NA)
+  == "logical"}.  The analysis deals with these {\tt raw\_NA}s at a later
+  stage.
+
+\item if the analysis encounters a matrix (i.e. the value has matrix class
+  and has a dims and optionally dimnames attribute), we ascribe the type
+  {\tt matrix}.
+
+\item similarly, if it encounters a data frame (i.e., class {\tt data.frame}
+  with appropriate attributes), we ascribe the type {\tt data.frame}.
 \end{itemize}
 
-%
-%
-%
 %
@@
 
-\AT{Putting examples of polymorphism in this section, we can talk about them and where they should go.}
-
-First, Figure~\ref{fig:realex} shows an example of how integers and doubles are very compatible.
-\AT{TODO add matrix (dimensions) to the example.}
+First, Figure~\ref{fig:realex} shows an example of how integers and doubles
+are very compatible.  \AT{TODO add matrix (dimensions) to the example.}
 
@@
 
-Figure~\ref{fig:optnull} shows an example of a function with an optional argument.
-In the function, {\tt x} is a sorted vector of income values, and {\tt w} is a vector of weights, optionally NULL.
-The function computes fractional ranks required for the computation of some coefficient implemented by the package.
-We see that if {\tt w} is NULL, then default unit weights are generated.
+Figure~\ref{fig:optnull} shows an example of a function with an optional
+argument.  In the function, {\tt x} is a sorted vector of income values, and
+{\tt w} is a vector of weights, optionally NULL.  The function computes
+fractional ranks required for the computation of some coefficient
+implemented by the package.  We see that if {\tt w} is NULL, then default
+unit weights are generated.
 
@@
 
-Figure~\ref{fig:listvec} shows an example of a function with list and vector polymorphism.
-The function takes a list or vector {\tt point} and a data frame {\tt polyg} representing a polygon.
-In it, we see that {\tt point} can be either a list or a vector, and the function code casts the argument to a double vector.
+Figure~\ref{fig:listvec} shows an example of a function with list and vector
+polymorphism.  The function takes a list or vector {\tt point} and a data
+frame {\tt polyg} representing a polygon.  In it, we see that {\tt point}
+can be either a list or a vector, and the function code casts the argument
+to a double vector.
 
@@
 
-Figure~\ref{fig:charclos} shows an example of a function with character and function polymorphism.
-\AT{Bluh.}
-At a high level, this function calculates point estimates for an {\tt angmcmc} object (specific to some packages).
-The caller specifies {\tt fn}, either a function or a function name which will be evaluated on the object samples to estimate parameters.
+Figure~\ref{fig:charclos} shows an example of a function with character and
+function polymorphism.  \AT{Bluh.}  At a high level, this function
+calculates point estimates for an {\tt angmcmc} object (specific to some
+packages).  The caller specifies {\tt fn}, either a function or a function
+name which will be evaluated on the object samples to estimate parameters.
 The call to lookup the function name is on line 10.
@@
     ...
-    if (is.character(fn)) {
-        if (fn == "MODE" | fn == "MAP") {
-            do_MAP <- TRUE
-        }
+    if (is.character(fn)) 
+        if (fn == "MODE" | fn == "MAP")   do_MAP <- TRUE
         else {
@@
         }
-    }
-    ...
 \end{lstlisting}}\caption{Example of char/closure argument (from {\tt BAMBI} package).}\label{fig:charclos}\end{figure}
 
-\AT{I'll explain this better, maybe?}
-In Figure~\ref{fig:dfdbl}, we see a function that, among other things, takes in some data ({\tt dat}) and a character string ({\tt spss}) specifying either {\tt "in"} or {\tt "out"}.
-Now, depending on the value of {\tt spss}, {\tt dat} will either be a data frame or a vector of doubles.
+\AT{I'll explain this better, maybe?}  In Figure~\ref{fig:dfdbl}, we see a
+function that, among other things, takes in some data ({\tt dat}) and a
+character string ({\tt spss}) specifying either {\tt "in"} or {\tt "out"}.
+Now, depending on the value of {\tt spss}, {\tt dat} will either be a data
+frame or a vector of doubles.
 
@@
   if(identical(spss, "out")) {
-    if(!is.vector(dat)) {
-        stop(simpleError("...'dat' must be a vector!"))
-      }
+    if(!is.vector(dat))  stop(simpleError("...'dat' must be a vector!"))
     ...
@@
 
-In Figure~\ref{fig:chardbl}, we see a function which takes in a list ({\tt network}), a vector of indices of that list ({\tt fixIndices}), and a vector of values ({\tt values}).
-The locations specified by {\tt fixIndicies} in {\tt network} are updated with {\tt values}.
-Here, {\tt fixIndicies} has been observed to be either a character or double vector:
-In R, list indices are typically doubles, but can be characters (if the list or vector being indexed has a {\tt names} attribute).
+In Figure~\ref{fig:chardbl}, we see a function which takes in a list ({\tt
+  network}), a vector of indices of that list ({\tt fixIndices}), and a
+vector of values ({\tt values}).  The locations specified by {\tt
+  fixIndicies} in {\tt network} are updated with {\tt values}.  Here, {\tt
+  fixIndicies} has been observed to be either a character or double vector:
+In R, list indices are typically doubles, but can be characters (if the list
+or vector being indexed has a {\tt names} attribute).
 
@@
   ...
-}
-\end{lstlisting}}\caption{Example of character/double argument (from {\tt BoolNet} package).}\label{fig:chardbl}\end{figure}
-
-In Figure~\ref{fig:matvec}, we see that the function can take in a vector, but immediately transforms it (transpose, with {\tt t}) into a matrix.
-As an idea, matrix/vector polymorphism seems sensible, as mathematically vectors are matrices.
-R echoes this by making conversion between the two ``types'' easy:
-\code{as.vector(m)} flattens a matrix \code{m} into a vector (e.g., \code{as.vector(matrix(2, 2, 2)) == c(2, 2, 2, 2)}), and \code{as.matrix(v)} builds a {\tt length(v) x 1} matrix (e.g., \code{as.matrix(c(1, 2)) == matrix(data=c(1, 2))}).
-\AT{TODO fix line break in code at end of last sentence}
+\end{lstlisting}}\caption{Example of character/double argument (from {\tt BoolNet} package).}\label{fig:chardbl}\end{figure}%$
+
+In Figure~\ref{fig:matvec}, we see that the function can take in a vector,
+but immediately transforms it (transpose, with {\tt t}) into a matrix.  As
+an idea, matrix/vector polymorphism seems sensible, as mathematically
+vectors are matrices.  R echoes this by making conversion between the two
+``types'' easy: \code{as.vector(m)} flattens a matrix \code{m} into a vector
+(e.g., \code{as.vector(matrix(2, 2, 2)) == c(2, 2, 2, 2)}), and
+\code{as.matrix(v)} builds a {\tt length(v) x 1} matrix (e.g.,
+\code{as.matrix(c(1, 2)) == matrix(data=c(1, 2))}).  \AT{TODO fix line break
+  in code at end of last sentence}
 
@@
 tee <- function (x, theta, D1, D2, phi)  {
-    if (is.vector(x)) {
-        x <- t(x)
-    }
+    if (is.vector(x)) x <- t(x)
     ...
@@
 %
-%
-%
-%
 \section{Results}\label{sec:results}
@@
 
-The logical ``baseline'' for the design of a static type system for R is promotion of the language's dynamic types (in the sense of \code{typeof}) to static types; we call this type system L0.
-Such a type system exactly reflects the types of values in the R runtime environment, as the \code{typeof} an object is in direct correspondence with the runtime type tag of the value.
-
-In this section, we will begin by breaking down our corpus of data with L0 in mind.
-From this perspective, we will see how current R programmers see types in the language, and we will leverage our newfound insight to design a type system which should be relevant to them.
-
-Without further ado, let's look at the data, first focusing on a broad overview of the polymorphism in L0.
-Figure~\ref{fig:L1argcounts} shows the polymorphicity of argument positions in R function.
-From there, we build up a picture of the polymorphicity of functions as a whole, and the data for this function polymorphism can be found in Figure~\ref{fig:L1funcounts}.
+The logical ``baseline'' for the design of a static type system for R is
+promotion of the language's dynamic types (in the sense of \code{typeof}) to
+static types; we call this type system L0.  Such a type system exactly
+reflects the types of values in the R runtime environment, as the
+\code{typeof} an object is in direct correspondence with the runtime type
+tag of the value.
+
+In this section, we will begin by breaking down our corpus of data with L0
+in mind.  From this perspective, we will see how current R programmers see
+types in the language, and we will leverage our newfound insight to design a
+type system which should be relevant to them.
+
+Without further ado, let's look at the data, first focusing on a broad
+overview of the polymorphism in L0.  Figure~\ref{fig:L1argcounts} shows the
+polymorphicity of argument positions in R function.  From there, we build up
+a picture of the polymorphicity of functions as a whole, and the data for
+this function polymorphism can be found in Figure~\ref{fig:L1funcounts}.
 
@@
 
-From Figure~\ref{fig:L1argcounts}, we glean that argument positions aren't very polymorphic in L0.
-An argument which is "Monomorphic in Type" is monomorphic w.r.t. \code{typeof}, in that it was only ever observed to have one \code{typeof}.
-Similarly, an argument is "Monomorphic in Class" if it is only ever called with one {\tt class}.
-As we said earlier, attributes are a little more complicated, and an argument which is "Monomorphic in Attribute Pattern" was only ever observed to have one attribute pattern (and that pattern may itself contain many attributes).
+From Figure~\ref{fig:L1argcounts}, we glean that argument positions aren't
+very polymorphic in L0.  An argument which is "Monomorphic in Type" is
+monomorphic w.r.t. \code{typeof}, in that it was only ever observed to have
+one \code{typeof}.  Similarly, an argument is "Monomorphic in Class" if it
+is only ever called with one {\tt class}.  As we said earlier, attributes
+are a little more complicated, and an argument which is "Monomorphic in
+Attribute Pattern" was only ever observed to have one attribute pattern (and
+that pattern may itself contain many attributes).
 
@@
 % Fri Mar 29 18:13:17 2019
-\begin{table}[ht]
-\label{tab:L0toppoly}
-\centering
-\begin{tabular}{lrr}
-  \hline
-type & count & perc \\ 
-  \hline
-  double, integer & 19,799 & 28.73 \\ 
-  double, list & 5,324 & 7.72 \\ 
-  double, NULL & 5,152 & 7.48 \\ 
-  character, NULL & 3,682 & 5.34 \\ 
-  double, logical & 3,388 & 4.92 \\ 
-  character, double & 2,842 & 4.12 \\ 
-  list, NULL & 2,799 & 4.06 \\ 
-  character, list & 1,836 & 2.66 \\ 
-  character, integer & 1,663 & 2.41 \\ 
-  character, double, integer & 1,362 & 1.98 \\ 
-   \hline
+\begin{table}[ht]\label{tab:L0toppoly}\centering
+\begin{tabular}{lrr}  \hline
+type & count & perc \\\hline\\[-3mm] 
+  \D, \I & 19,799 & 28.73 \\[1mm] 
+  \D, \l & 5,324 & 7.72 \\ [1mm] 
+  \D, \sN & 5,152 & 7.48 \\ [1mm] 
+  \C, \sN & 3,682 & 5.34 \\ [1mm] 
+  \D, \L & 3,388 & 4.92 \\ [1mm] 
+  \C, \D & 2,842 & 4.12 \\ [1mm] 
+  \l, \sN & 2,799 & 4.06 \\ [1mm] 
+  \C, \l & 1,836 & 2.66 \\ [1mm] 
+  \C, \I & 1,663 & 2.41 \\ [1mm] 
+  \C, \I, \D & 1,362 & 1.98 \\[1mm]   \hline
 \end{tabular}
@@
 
-Some entries in the Table~\ref{tab:L0toppoly} are clear (e.g., double, integer polymorphism is unsurprising), others are strange (e.g., character, double), but in fact they all have some explanation.
+Some entries in the Table~\ref{tab:L0toppoly} are clear (e.g., double,
+integer polymorphism is unsurprising), others are strange (e.g., character,
+double), but in fact they all have some explanation.
 
 \begin{enumerate}
-
-	\item \{double, integer\}: in R, doubles and integers are effectively interchangeable, as casting from one to the other is straightforward and automatically performed when appropriate.
-	For example, \code{c(1, 2)[1.9] == 1}.
-	This distinction is mainly just relevant to the implementation, as integers and doubles are stored and dealt with differently in the runtime but programmers only see the distinction when printing.
-
-	\item \{double, list\} and \{character, list\}: at first glance, these type signatures are suspect.
-	Why would so many programmers want to pass a double or a list to some function?
-	The key insight here is that doubles and characters are {\it vectorized} in R, meaning that a double is really a vector of doubles.
-	Thus, these seemingly strange signatures are revealed to be entirely reasonable: these signatures are likely to describe arguments which are either vectors or lists.
-
-	\item \{X, NULL\}: in most languages, NULL is an inhabitant of all types, and we see that this is not the case in R.
-	That said, there is a pattern of many functions having default or optional arguments, and in many of these cases NULL is the default value.
-
-	\item \{double, logical\}: the first values of type logical that come to mind are TRUE and FALSE, which does little to explain this strange signature.
-	Indeed, the key here is an implementation detail of R's: recall that NAs inhabit every type.
-	As it happens, the default type of NA is logical (e.g., \code{typeof(NA) == "logical"}), and this is likely what is being seen here.
-	We will control for this by distinguishing NAs with their own type, {\tt raw\_NA}, but we will get to that in time.
-
-	\item \{character, double\}, \{character, integer\} and \{character, double, integer\}: yet more strange signatures.
-	These signatures are a little more subtle to explain.
-	For one, indexing lists and vectors can be done with either strings (i.e., to access a named index) or with numbers.
-	Another option is dates: R has many ways to present dates, and strings and numbers feature among them.
-	Finally, many \AT{get numbers} functions in R return a string if an error occurs during execution.
+\item \{\D, \I\}: in R, doubles and integers are effectively
+  interchangeable, as casting from one to the other is straightforward and
+  automatically performed when appropriate.  For example, \code{c(1, 2)[1.9]
+    == 1}.  This distinction is mainly just relevant to the implementation,
+  as integers and doubles are stored and dealt with differently in the
+  runtime but programmers only see the distinction when printing.
+
+\item \{\D, \l\} and \{\C, \l\}: at first glance, these type signatures are
+  suspect.  Why would so many programmers want to pass a double or a list to
+  some function?  The key insight here is that doubles and characters are
+  {\it vectorized} in R, meaning that a double is really a vector of
+  doubles.  Thus, these seemingly strange signatures are revealed to be
+  entirely reasonable: these signatures are likely to describe arguments
+  which are either vectors or lists.
+
+\item \{\ANY,\sN\}: in most languages, NULL is an inhabitant of all types,
+  and we see that this is not the case in R.  That said, there is a pattern
+  of many functions having default or optional arguments, and in many of
+  these cases NULL is the default value.
+
+\item \{\D,\L\}: the first values of type logical that come to mind are TRUE
+  and FALSE, which does little to explain this strange signature.  Indeed,
+  the key here is an implementation detail of R's: recall that NAs inhabit
+  every type.  As it happens, the default type of NA is logical (e.g.,
+  \code{typeof(NA) == "logical"}), and this is likely what is being seen
+  here.  We will control for this by distinguishing NAs with their own type,
+  {\tt raw\_NA}, but we will get to that in time.
+
+\item \{\C,\D\}, \{\C,\I\} and \{character, double, integer\}: yet more
+  strange signatures.  These signatures are a little more subtle to explain.
+  For one, indexing lists and vectors can be done with either strings (i.e.,
+  to access a named index) or with numbers.  Another option is dates: R has
+  many ways to present dates, and strings and numbers feature among them.
+  Finally, many \AT{get numbers} functions in R return a string if an error
+  occurs during execution.
 

### 2019-04-02 paper/main.tex
merging paper

--- 2019-04-01 paper/main.tex
@@
 \newcommand{\METAARGCOUNT}{7,051\xspace}
+\newcommand{\NUMPOLYTYPEMONOCLASS}{1,357\xspace}
 
@@
 	As we have established that an annotation for an optional argument is useful, this will allow us to reveal other interesting patterns.
-	\AT{If we feel the need to go into more details on NULL or NA, I have the old paragraphs commented out in the source.}
 
@@
 This is one of the many modifications we make, all of which are thoroughly explored in the following section.
-
-%In terms of programmer behavior, we observe a pattern of optional arguments throughout several R packages, wherein programmers specify a default argument value which is sometimes NULL.
-%Consider the snippet in Figure~\ref{fig:optnull} for an example.
-%\AT{TODO put that figure here}
-%The distinction between NULL and other types is not particularly useful to the programmer, so we propose to also collapse this distinction.
-%Not pictured in Table~\ref{tab:toppolyFDTS} are NAs, which are ascribed the type {\tt raw\_NA} in L1; NAs in fact inhabit every type, though the default NA type is {\tt logical}.
-%Just as we fold NULL into all types, we will do the same with NAs.
-
 %
@@
   Full Monomorphic & 781K & 85.47 \\
-  Monomorphic in Type & 867K & 95.01 \\
+  Monomorphic in Type & 869K & 95.01 \\
 %  Monomorphic in Class & 831K & 90.31 \\ 
@@
   Full Monomorphic & 143K & 66.04 \\ 
-  Monomorphic in Type & 184K & 85.30 \\
+  Monomorphic in Type & 185K & 85.30 \\
 %  Monomorphic in Class & 163K & 75.13 \\ 
@@
 
-\AT{Eyy time for L2.}
-
-%
-%
-%
-%
+In L2, we aim to incorporate R's other reflective capabilities to address some of the more subtle instances of polymorphism.
+
+We observe that the act of ascribing a class to a value in R carries the intent of the programmer.
+Now, not all classes are user-defined, and R has a number of built-in or otherwise standard classes, ranging from truly primitive classes such as ``numeric'' to the more complex ``data.frame''.
+Thus, before considering the insights provided by the class of values, we transform each of R's standard classes into a ``primitive'' class, which will serve to highlight user-defined classes and those particular to some packages.
+
+\AT{I don't know the best way to approach this subsection, so I'm just gonna start looking at some stuff and see where it leads.}
+
+First, we consider the purpose of classes in R.
+The
+
+As before, we start with the counts, see Table~\ref{tab:classcountsL2}.
+
+% latex table generated in R 3.5.1 by xtable 1.8-3 package
+% Mon Apr  1 13:02:08 2019
+\begin{table}[ht]
+\label{tab:classcountsL2}
+\centering
+\begin{tabular}{lrr}
+  \hline
+Type & Count & Percentage \\ 
+  \hline
+Full Monomorphic & 777 & 84.96 \\ 
+%  Monomorphic in Type & 869.00 & 95.01 \\ 
+  Monomorphic in Class & 850 & 92.94 \\ 
+%  Monomorphic in Attribute Pattern & 825.00 & 90.27 \\ 
+  Total Seen & 914 & --- \\ 
+   \hline
+\end{tabular}
+\caption{Account of class-based polymorphism in L2.}
+\end{table}
+
+Top signatures in Table~\ref{tab:classpolyL2}.
+
+% latex table generated in R 3.5.1 by xtable 1.8-3 package
+% Mon Apr  1 14:48:55 2019
+\begin{table}[ht]
+\label{tab:classpolyL2}
+\centering
+\begin{tabular}{lrr}
+  \hline
+Class & Count & Precentage \\ 
+  \hline
+primitive, tbl, tbl\_df & 1796 & 2.78 \\ 
+  array, primitive & 1379 & 2.14 \\ 
+  data.table, primitive & 777 & 1.20 \\ 
+  POSIXct, POSIXt & 716 & 1.11 \\ 
+  formula, primitive & 639 & 0.99 \\ 
+  xts, zoo & 588 & 0.91 \\ 
+  glm, lm & 494 & 0.77 \\ 
+  primitive, ts & 432 & 0.67 \\ 
+  mts, primitive, ts & 384 & 0.59 \\ 
+  huxtable, primitive & 318 & 0.49 \\ 
+   \hline
+\end{tabular}
+\caption{Top 10 signatures for class-based polymorphism in L2.}
+\end{table}
+
+Given the low percentages overall, we see that there are no clear patterns here.
+This data does highlight some common classes across R packages:
+tbl and tbl\_df are from the dplyr package, as classes for data objects.
+data.table is the primary data type in the data.table package, again for data objects.
+POSIXct and POSIXt are for dates.
+
+To see how we might use class information, we should see how often a monomorphic class signature co-occurs with a polymorphic type signature.
+We count \NUMPOLYTYPEMONOCLASS argument positions which match this criteria.
+
+% latex table generated in R 3.5.1 by xtable 1.8-3 package
+% Mon Apr  1 16:50:22 2019
+\begin{table}[ht]
+\centering
+\begin{tabular}{llrl}
+  \hline
+  Type Sig. & Class Sig. & Count & Percentage \\ 
+  \hline
+  list$<$any$>$, list$<$character$>$ & roxy\_block & 38 & 2.80 \\ 
+  list$<$any$>$, list$<$list$>$ & Rd & 25 & 1.84 \\ 
+  list$<$any$>$, list$<$real$>$ & TSdata & 25 & 1.84 \\ 
+  list$<$any$>$, list$<$list$>$ & xml\_nodeset & 24 & 1.77 \\ 
+  list$<$any$>$, list$<$list$>$, list$<$real$>$ & dendrogram & 24 & 1.77 \\ 
+  list$<$any$>$, list$<$real$>$ & phylo & 23 & 1.69 \\ 
+  list$<$any$>$, list$<$list$>$, list$<$real$>$, scalar/real & dendrogram & 21 & 1.55 \\ 
+  list$<$any$>$, list$<$language$>$ & uneval & 18 & 1.33 \\ 
+  list$<$any$>$, list$<$list$>$ & dendrogram & 18 & 1.33 \\ 
+  list$<$any$>$, list$<$list$>$ & lazy\_dots & 17 & 1.25 \\ 
+   \hline
+\end{tabular}
+\end{table}
+
+\AT{I don't think this table is telling us much...}
+
+Top attributes, maybe?
+
+% latex table generated in R 3.5.1 by xtable 1.8-3 package
+% Mon Apr  1 19:21:13 2019
+\begin{table}[ht]
+\centering
+\begin{tabular}{lrl}
+  \hline
+attr & count & perc \\ 
+  \hline
+\{\}, \{\xspace names: \charv\} & 9274.00 & 10.43 \\ 
+  \{\}, \{\xspace names: \chars\} & 4900.00 & 5.51 \\ 
+  \{\}, \{\xspace dim: \intv\} & 4758.00 & 5.35 \\ 
+  \{\xspace names: \chars\}, \{\xspace names: \charv\} & 3457.00 & 3.89 \\ 
+  \{\}, \{\xspace names: \chars\}, \{\xspace names: \charv\} & 2415.00 & 2.72 \\ 
+  \{\xspace dim: \intv, dimnames:list$<$any$>$ \}, \{\xspace dim:\intv\} & 2264.00 & 2.55 \\ 
+  \{\}, \{\xspace srcref: \intv\} & 2037.00 & 2.29 \\ 
+  \{\xspace dim: \intv, dimnames: list$<$character$>$\}, \{ dim:\intv\} & 1734.00 & 1.95 \\ 
+  \{\xspace class: \chars, names: \charv, row.names:\charv\}, \{class:\chars, names:\charv, row.names:\intv\} & 1464.00 & 1.65 \\ 
+  \{\xspace class: \chars, names: \charv\}, \{ class:\charv, names:\charv\} & 1277.00 & 1.44 \\ 
+   \hline
+\end{tabular}
+\caption{Top attribute patterns. \AT{TODO fix this up}}
+\end{table}\charv
+
+A lot of the polymorphism that we see is because we haven't propagated our changes through the attribute types.
+Will fix this.
+
+%
+%
+%
+%
+\subsection{R's Most Polymorphic Functions}
+
 \AT{We might want to discuss some of the most polymorphic functions in the language.}
-\subsection{R's Most Polymorphic Functions}
 

### 2019-04-02 paper/main.tex
changes all over

@@
 
-\newcommand{\attr}[2]{\ensuremath{#1\,({\mathit #2})}\xspace}
+\newcommand{\attr}[2]{\ensuremath{#1_{\mathtt{#2}}}\xspace}
 \renewcommand{\to}{\ensuremath{\rightarrow}\xspace}
@@
 In R all arguments are passed as promises, and unused arguments will be
-unevaluated.  Our tracer does still collect information on these arguments,
-as whether or not they were used doesn't change the fact that they were in
-fact passed.  In some cases, arguments aren't even intended to be evaluated.
-For example, consider the {\tt magrittr} package, which implements the
-Tidyverse's pipe functions.  Refer to Figure~\ref{fig:magpipeex} for a
-translation of a call to {\tt magrittr}'s pipe ({\tt \%>\%}) function.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-mul_by <- function(x, y) { x * y }
-
-4 %>% mul_by(3) === `%>%`(lhs = 4, rhs = mul_by(3)) === mul_by(4, 3)
-\end{lstlisting}}\caption{Example and rough translation of \%$>$\% call.}\label{fig:magpipeex}\end{figure}
-
-Our tracer evaluates function arguments, and here \code{rhs = mul_by(3)},
-but that's erroneous code (you can't call {\tt mul\_by} without supplying
-two arguments).  In this situation, our tracer will still register the
-entire call, but flag the {\tt rhs} argument to indicate to later stages of
-the analysis that there was an issue evaluating that argument.
-
-For each function of each package, we consolidate all recorded observations
-into a function signature. The details of how consolidation is parameterized
-by the definition of a potential type system for R and will be discussed in
-the next section.
+unevaluated.  Our tracer does not collect information on arguments that were
+not evaluated as they may not be valid expressions.  Consider the {\tt
+  magrittr} package, which implements the pipe function that performs a kind
+of currying.
+
+{\small\begin{lstlisting}[style=R]
+mul <- function(x,y)  x*y
+4 %>% mul(3) 
+\end{lstlisting}}
+
+The call to the pipe desugars to \code{`\%>\%`(lhs=4, rhs=mul(3))} where the
+\code{rhs} is not a valid expression. It is treated as code that will
+manipulated to add one argument (the \code{lhs}).
+
+To be as specific as possible, our analysis collects shape information about
+arguments and return values. The details of the information we collect follows.
+
+\begin{itemize}
+\item If the analysis encounters a primitive vector, say
+  \code{typeof(x)=="double"}, it records the length of {\tt x} to determine
+  if it can be treated as a scalar.
+\item For a list, \code{typeof(x)=="list"}, shape information on all
+  elements is used to describe the list's shape.  To avoid undue slowdowns,
+  we only collect the content's \code{typeof}, and ascribe \ANY (any) if the
+  elements are not of the same kind.k
+\item If the analysis encounters a scalar \code{NA}, we ascribe a unique
+  NULL type.  In R, \code{NA} inhabits all types, but for a scalar NA,
+  \code{typeof(NA)=="logical"}. Scalar NAs are used as uninitialized values.
+\item For a matrix (i.e. the value has matrix class and has a dims and
+  optionally dimnames attribute), we ascribe the shape \attr{\tt T}{mat}
+  where {\tt T} is the primitive element type.
+\item For a data frame (i.e., class {\tt data.frame} with appropriate
+  attributes), we ascribe the shape \attr{\l}{df}.
+\end{itemize}
+
+The infrastructure memoizes emitted signatures to avoid writing multiple
+identical signatures for the same function.
 
@@
 
-We will need a notion of polymorphism, and thus a notion of type, when
-undertaking this analysis.  R is unityped, and the only notion of
-polymorphism in the language is its use of dynamic dispatch on the class of
-an argument.  For our purposes, we will define a notion in type roughly in
-line with R's \code{typeof} reflection function, in some cases augmented by
-information obtained by \code{class} and \code{attributes} reflection.  We
-start by defining what it means for an {\it argument} to be polymorphic.
+For a meaningful notion of polymorphism, it is necessary to settle on a set
+of basic types for R. There are several possible choices which lead to
+increasingly more expressive type systems. The simplest notion of types
+aligns with R basic types. Richers notions incorporate attributes and the
+various object systems built on top of those. For concreteness we will pick
+out three different sets of types.
 
 \begin{itemize}
-\item an argument is {\it polymorphic in type} if it has been called with at
-  least two distinct types according to our definition of type.
-\item an argument is {\it polymorphic in class} if it has been called with
-  at least two distinct classes according to the \code{class} function.
-\item attributes are a bit more tricky.  We can't simply count the number of
-  attributes that have inhabited an argument, since values can contain
-  arbitrarily many attributes (and having several attributes isn't
-  polymorphism---think records).  Instead, we define the {\it attribute
-    pattern} of a value to be the set of all attribute name and type pairs
-  in the attribute list of a value (obtained via \code{attributes}).  And so
-  an argument is {\it polymorphic in attribute} if it has been called with
-  at least two distinct attribute patterns.
+\item {\bf L0}: This level includes vectors of primitive types (integers \I,
+  double \D, character \C, logical \L, raw \R, complex \X), lists of any
+  type (\l), and scalar values (environment \sE, closures \sF, symbols \sS
+  and null \sN). We add the type \ANY for any type.
+\item {\bf L1}: The second level enhances {\bf L0} with scalar values for
+  primitive types (\sI, \sD, \sC, \sL, \sR, and \sX), type parametric lists
+  (written \lT{T} for any type {\tt T}), data frames \attr{\l}{df}, and
+  matrices \attr{\vec{\tt T}}{mat}.
+\item {\bf L2}: The third level adds attributes lists...
 \end{itemize}
 
-The main ingredient of our types is the {\tt typeof} a value, and there is
-an important distinction to be made between the result of \code{typeof} and
-that of \code{class} and \code{attributes}.  In R, the \code{typeof} a value
-cannot change, and this is not the case with class and attributes.  As
-mentioned in \AT{an earlier section}, the class of values can be easily
-redefined by assigning to the value's class attribute, and similarly
-attributes can be dynamically added, removed, or changed.
+In the following we say that an argument is {\it polymorphic in type at
+  level L} if it has been called with at least two values that belong to
+distinct types at level \emph{L}. This definition extends naturally to
+classes, data frames, and matrices (as these are simple attributes).
+Generalized attribute polymorphism is a bit trickier as values can have a
+large and changing set of attributes.
 
@@
 
-\subsection{Granularity of Type Information}
-
-When constructing a type for a value, we begin with the {\tt typeof} the
-value.  This alone paints an incomplete picture of the value, though,
-especially when it comes to vectors (recall that \code{typeof(5) ==
-  typeof(c(5, 5)) == "double"}).  In some cases, the class or attributes of
-a value might enlighten us, but in many cases we can be a little more
-precise.
-
-To be as specific as possible, our analysis collects additional information
-to create a more specific type than the result of any of R's reflection
-functions.  The details of the information we collect follows.
-
-\begin{itemize}
-\item if the analysis encounters \code{ typeof(x) == "double"}, it will look
-  at the length of {\tt x} to determine if {\tt x} is a scalar or a vector,
-  generating the annotation {\tt \dbls} or {\tt \dblv} accordingly.  This is
-  also true for integer, complex, raw, logical, and character {\tt typeof}s
-  (i.e., the primitive vectorized data types).  A notable imperfection here
-  is is that unit length vectors will appear as scalars, so later in our
-  analysis we ensure that argument signatures containing both scalar/X and
-  vector/X are collapsed down to contain only vector/X.
-
-\item if the analysis encounters \code{typeof(x) == "list"}, we collect type
-  information on all list elements in order to ascribe a general type to the
-  contents of the list.  To avoid undue slowdowns, we only collect the
-  \code{typeof} the contents, and ascribe type {\tt any} if two or more {\tt
-    typeof}s are observed.
-
-\item if the analysis encounters \code{NA}, we ascribe a unique {\tt
-  raw\_NA} type.  In R, \code{NA} inhabits all types, but \code{typeof(NA)
-  == "logical"}.  The analysis deals with these {\tt raw\_NA}s at a later
-  stage.
-
-\item if the analysis encounters a matrix (i.e. the value has matrix class
-  and has a dims and optionally dimnames attribute), we ascribe the type
-  {\tt matrix}.
-
-\item similarly, if it encounters a data frame (i.e., class {\tt data.frame}
-  with appropriate attributes), we ascribe the type {\tt data.frame}.
-\end{itemize}
-
-%
-%
 %

### 2019-04-02 paper/main.tex
changes all over

@@
 increasingly more expressive type systems. The simplest notion of types
-aligns with R basic types. Richers notions incorporate attributes and the
-various object systems built on top of those. For concreteness we will pick
-out three different sets of types.
-
-\begin{itemize}
-\item {\bf L0}: This level includes vectors of primitive types (integers \I,
-  double \D, character \C, logical \L, raw \R, complex \X), lists of any
-  type (\l), and scalar values (environment \sE, closures \sF, symbols \sS
-  and null \sN). We add the type \ANY for any type.
-\item {\bf L1}: The second level enhances {\bf L0} with scalar values for
-  primitive types (\sI, \sD, \sC, \sL, \sR, and \sX), type parametric lists
-  (written \lT{T} for any type {\tt T}), data frames \attr{\l}{df}, and
-  matrices \attr{\vec{\tt T}}{mat}.
-\item {\bf L2}: The third level adds attributes lists...
-\end{itemize}
+aligns with R's builtin types. Richers notions incorporate attributes and
+the various object systems built on top of those. For concreteness we will
+pick out three different sets of types.
 
@@
 large and changing set of attributes.
+
+\subsection{L0 Types}
+
+The logical ``baseline'' for the design of a type system for R is to adopt
+the language's dynamic types (in the sense of \code{typeof}).  We call this
+L0.  The types reflects the types of values in the R runtime environment, as
+the \code{typeof} an object is in direct correspondence with the runtime
+type tag of the value.
+
+Thus L0 includes the following types: vectors of primitive types (integers
+\I, double \D, character \C, logical \L, raw \R, complex \X), lists of any
+type (\l), and scalar values (environment \sE, closures \sF, symbols \sS and
+null \sN). We add the type \ANY to stand in for any type.
+
+We also include a notion of subtyping that follows from the conversion rules
+of the language.  All types are subtypes of \ANY, thus {\tt T}<:\ANY.
+Moreover, logicals are subtypes of integers, which are subtypes of doubles,
+\L <: \I <: \D.
+
+\subsection{L1 Types}
+
+L0 is limited in expressivity. To address its limitations, we augment it
+scalars, parametrized lists.  In addition, we collect attribute and class
+information to distinguish a type for data frames and matrices, as the
+\code{typeof} these is simply {\tt list} in R.
+
+The second level enhances {\bf L0} with scalar values for primitive types
+(\sI, \sD, \sC, \sL, \sR, and \sX), type parametric lists (written \lT{T}
+for any type {\tt T}), data frames \attr{\l}{df}, and matrices
+\attr{\vec{\tt T}}{mat}.
+
+\subsection{L2 Types}
+ {\bf L2}: The third level adds attributes lists...
 
@@
 
-These naturally-occurring attribute patterns and classes can be separately
-accounted for to highlight user-defined patterns and behaviors.  This will
-be made clear in Section~\ref{sec:results}.
-
-%
-\section{Examples of Polymorphism}
-\label{sec:polyex}
-
-First, Figure~\ref{fig:realex} shows an example of how integers and doubles
-are very compatible.  \AT{TODO add matrix (dimensions) to the example.}
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-> 5L + 1L # => 6L, an integer
-> 5L + 1.2 # => 6.2, a double (5L coerced to 5.0)
-> c(10, 20, 30)[1.2] # => 10, 1.2 coerced to 1L
-\end{lstlisting}}\caption{Example of {\it real} type usage.}\label{fig:realex}\end{figure}
-
-Figure~\ref{fig:optnull} shows an example of a function with an optional
-argument.  In the function, {\tt x} is a sorted vector of income values, and
-{\tt w} is a vector of weights, optionally NULL.  The function computes
-fractional ranks required for the computation of some coefficient
-implemented by the package.  We see that if {\tt w} is NULL, then default
-unit weights are generated.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-frac.ranks <- function(x, w=NULL) {
-  if (is.null(w)) w <- rep(1, length(x)) # if no weights passed, take all weights = 1
-  ...
-\end{lstlisting}}\caption{Example of optional argument (from {\tt acid} package).}\label{fig:optnull}\end{figure}
-
-Figure~\ref{fig:listvec} shows an example of a function with list and vector
-polymorphism.  The function takes a list or vector {\tt point} and a data
-frame {\tt polyg} representing a polygon.  In it, we see that {\tt point}
-can be either a list or a vector, and the function code casts the argument
-to a double vector.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-is.point.inside <- function (point, polyg) {
-    p <- as.numeric(point) # as.numeric(list(1, 2)) => c(1, 2)
-    ...
-\end{lstlisting}}\caption{Example of list/vector argument (from {\tt bivrp} package).}\label{fig:listvec}\end{figure}
-
-Figure~\ref{fig:charclos} shows an example of a function with character and
-function polymorphism.  \AT{Bluh.}  At a high level, this function
-calculates point estimates for an {\tt angmcmc} object (specific to some
-packages).  The caller specifies {\tt fn}, either a function or a function
-name which will be evaluated on the object samples to estimate parameters.
-The call to lookup the function name is on line 10.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-pointest <- function (object, fn = mean, par.name, comp.label, chain.no,  ...) {
-    ...
-    if (is.character(fn)) 
-        if (fn == "MODE" | fn == "MAP")   do_MAP <- TRUE
-        else {
-            do_MAP <- FALSE
-            fn <- match.fun(fn) # looks up the function name
-        }
-\end{lstlisting}}\caption{Example of char/closure argument (from {\tt BAMBI} package).}\label{fig:charclos}\end{figure}
-
-\AT{I'll explain this better, maybe?}  In Figure~\ref{fig:dfdbl}, we see a
-function that, among other things, takes in some data ({\tt dat}) and a
-character string ({\tt spss}) specifying either {\tt "in"} or {\tt "out"}.
-Now, depending on the value of {\tt spss}, {\tt dat} will either be a data
-frame or a vector of doubles.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-nret.translator <- function(dat, spss="out", ...) {
-  ...
-  if(identical(spss, "out")) {
-    if(!is.vector(dat))  stop(simpleError("...'dat' must be a vector!"))
-    ...
-  } else {
-    ...
-    # here, dat is data.frame
-    items.idx <- items.idx[order(names(dat[, items.idx]))]
-  }
-\end{lstlisting}}\caption{Example of a data.frame/(\dblv) argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}
-
-In Figure~\ref{fig:chardbl}, we see a function which takes in a list ({\tt
-  network}), a vector of indices of that list ({\tt fixIndices}), and a
-vector of values ({\tt values}).  The locations specified by {\tt
-  fixIndicies} in {\tt network} are updated with {\tt values}.  Here, {\tt
-  fixIndicies} has been observed to be either a character or double vector:
-In R, list indices are typically doubles, but can be characters (if the list
-or vector being indexed has a {\tt names} attribute).
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-fixGenes <- function (network, fixIndices, values) {
-  ...
-  network$fixed[fixIndices] <- as.integer(values)
-  ...
-\end{lstlisting}}\caption{Example of character/double argument (from {\tt BoolNet} package).}\label{fig:chardbl}\end{figure}%$
-
-In Figure~\ref{fig:matvec}, we see that the function can take in a vector,
-but immediately transforms it (transpose, with {\tt t}) into a matrix.  As
-an idea, matrix/vector polymorphism seems sensible, as mathematically
-vectors are matrices.  R echoes this by making conversion between the two
-``types'' easy: \code{as.vector(m)} flattens a matrix \code{m} into a vector
-(e.g., \code{as.vector(matrix(2, 2, 2)) == c(2, 2, 2, 2)}), and
-\code{as.matrix(v)} builds a {\tt length(v) x 1} matrix (e.g.,
-\code{as.matrix(c(1, 2)) == matrix(data=c(1, 2))}).  \AT{TODO fix line break
-  in code at end of last sentence}
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-tee <- function (x, theta, D1, D2, phi)  {
-    if (is.vector(x)) x <- t(x)
-    ...
-\end{lstlisting}}\caption{Example of matrix/(\dblv) argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}
-
-%
-%
-%
 \section{Results}\label{sec:results}
 
-Having built an understanding of our data collection and analysis pipeline, we will turn our attention to the results of our analysis.
-
-\AT{Maybe it's best to ignore attributes and classes for now?}
-
-%
-%
-%
-%
+Having built an understanding of our data collection and analysis pipeline,
+we will turn our attention to the results of our analysis.
+
 \subsection{L0}
-
-The logical ``baseline'' for the design of a static type system for R is
-promotion of the language's dynamic types (in the sense of \code{typeof}) to
-static types; we call this type system L0.  Such a type system exactly
-reflects the types of values in the R runtime environment, as the
-\code{typeof} an object is in direct correspondence with the runtime type
-tag of the value.
 
@@
 \begin{enumerate}
-
-	\item consolidating doubles and integers into a single real-number type would account for all of this polymorphism.
-
-	\item to get an idea of how often vectors and lists co-occur in signatures, we need to expand our analysis of vectorized types.
-	Notably, when we encounter e.g. a double, we should collect length information to see if the double is a vector of doubles or just a scalar.
-	This will feature in our next type system, L1, which we will get to momentarily.
-
-	\item if we roll NULLs into every type, we can cleanly deal with these X, NULL signatures.
-	As we have established that an annotation for an optional argument is useful, this will allow us to reveal other interesting patterns.
-
+\item consolidating doubles and integers into a single real-number type
+  would account for all of this polymorphism.
+\item to get an idea of how often vectors and lists co-occur in signatures,
+  we need to expand our analysis of vectorized types.  Notably, when we
+  encounter e.g. a double, we should collect length information to see if
+  the double is a vector of doubles or just a scalar.  This will feature in
+  our next type system, L1, which we will get to momentarily.
+\item if we roll NULLs into every type, we can cleanly deal with these X,
+  NULL signatures.  As we have established that an annotation for an
+  optional argument is useful, this will allow us to reveal other
+  interesting patterns.
 \item similarly with NULLs, we're interested in rolling NAs in to all types.
-	As we mentioned, this is actually what the R runtime does, but a default type needed to be chosen for NAs appearing without context.
-	Subsuming NAs should help to reveal other interesting patterns.
-
-	\item dealing with this polymorphism will be more tricky.
-	We might need to pull class information in (to pick up on dates), and possibly make a signature judgment depending on which argument position has the signature (e.g.
-a \{character, X\} signature on a return).
-	We will come back to this later.
+  As we mentioned, this is actually what the R runtime does, but a default
+  type needed to be chosen for NAs appearing without context.  Subsuming NAs
+  should help to reveal other interesting patterns.
+\item dealing with this polymorphism will be more tricky.  We might need to
+  pull class information in (to pick up on dates), and possibly make a
+  signature judgment depending on which argument position has the signature
+  (e.g. a \{character, X\} signature on a return).  We will come back to
+  this later.
 
@@
 
-There are some clear limitations to the L0 type system that we should address before working through any of these other issues
-As we mentioned, \code{typeof} makes no distinction between vectors and scalars, as indeed there is no such distinction in R. 
-But as we saw, we might benefit from such a distinction, and indeed an all-new runtime environment design for R might benefit from e.g. not needing to vectorize scalar values.
-This is one of the many modifications we make, all of which are thoroughly explored in the following section.
+There are some clear limitations to the L0 type system that we should
+address before working through any of these other issues As we mentioned,
+\code{typeof} makes no distinction between vectors and scalars, as indeed
+there is no such distinction in R.  But as we saw, we might benefit from
+such a distinction, and indeed an all-new runtime environment design for R
+might benefit from e.g. not needing to vectorize scalar values.  This is one
+of the many modifications we make, all of which are thoroughly explored in
+the following section.
 %
@@
 
-To address these limitations, we augment the L0 with more fine-grained information.
-We will consider unit-length vectors to be scalars, collect list member type information to build an accurate list type, and promote NAs to their own type. 
-In addition, we collect attribute and class information to distinguish a type for data frames and matrices, as the \code{typeof} these is simply {\tt list} in R. 
-We refer to this as the L1 type system. 
-An example of a signature in this type system can be found in Figure~\ref{fig:exL1}.
+An example of a signature in this type system can be found in
+Figure~\ref{fig:exL1}.
 
@@
 %
+\section{Examples of Polymorphism}
+\label{sec:polyex}
+
+First, Figure~\ref{fig:realex} shows an example of how integers and doubles
+are very compatible.  \AT{TODO add matrix (dimensions) to the example.}
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+> 5L + 1L # => 6L, an integer
+> 5L + 1.2 # => 6.2, a double (5L coerced to 5.0)
+> c(10, 20, 30)[1.2] # => 10, 1.2 coerced to 1L
+\end{lstlisting}}\caption{Example of {\it real} type usage.}\label{fig:realex}\end{figure}
+
+Figure~\ref{fig:optnull} shows an example of a function with an optional
+argument.  In the function, {\tt x} is a sorted vector of income values, and
+{\tt w} is a vector of weights, optionally NULL.  The function computes
+fractional ranks required for the computation of some coefficient
+implemented by the package.  We see that if {\tt w} is NULL, then default
+unit weights are generated.
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+frac.ranks <- function(x, w=NULL) {
+  if (is.null(w)) w <- rep(1, length(x)) # if no weights passed, take all weights = 1
+  ...
+\end{lstlisting}}\caption{Example of optional argument (from {\tt acid} package).}\label{fig:optnull}\end{figure}
+
+Figure~\ref{fig:listvec} shows an example of a function with list and vector
+polymorphism.  The function takes a list or vector {\tt point} and a data
+frame {\tt polyg} representing a polygon.  In it, we see that {\tt point}
+can be either a list or a vector, and the function code casts the argument
+to a double vector.
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+is.point.inside <- function (point, polyg) {
+    p <- as.numeric(point) # as.numeric(list(1, 2)) => c(1, 2)
+    ...
+\end{lstlisting}}\caption{Example of list/vector argument (from {\tt bivrp} package).}\label{fig:listvec}\end{figure}
+
+Figure~\ref{fig:charclos} shows an example of a function with character and
+function polymorphism.  \AT{Bluh.}  At a high level, this function
+calculates point estimates for an {\tt angmcmc} object (specific to some
+packages).  The caller specifies {\tt fn}, either a function or a function
+name which will be evaluated on the object samples to estimate parameters.
+The call to lookup the function name is on line 10.
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+pointest <- function (object, fn = mean, par.name, comp.label, chain.no,  ...) {
+    ...
+    if (is.character(fn)) 
+        if (fn == "MODE" | fn == "MAP")   do_MAP <- TRUE
+        else {
+            do_MAP <- FALSE
+            fn <- match.fun(fn) # looks up the function name
+        }
+\end{lstlisting}}\caption{Example of char/closure argument (from {\tt BAMBI} package).}\label{fig:charclos}\end{figure}
+
+\AT{I'll explain this better, maybe?}  In Figure~\ref{fig:dfdbl}, we see a
+function that, among other things, takes in some data ({\tt dat}) and a
+character string ({\tt spss}) specifying either {\tt "in"} or {\tt "out"}.
+Now, depending on the value of {\tt spss}, {\tt dat} will either be a data
+frame or a vector of doubles.
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+nret.translator <- function(dat, spss="out", ...) {
+  ...
+  if(identical(spss, "out")) {
+    if(!is.vector(dat))  stop(simpleError("...'dat' must be a vector!"))
+    ...
+  } else {
+    ...
+    # here, dat is data.frame
+    items.idx <- items.idx[order(names(dat[, items.idx]))]
+  }
+\end{lstlisting}}\caption{Example of a data.frame/(\dblv) argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}
+
+In Figure~\ref{fig:chardbl}, we see a function which takes in a list ({\tt
+  network}), a vector of indices of that list ({\tt fixIndices}), and a
+vector of values ({\tt values}).  The locations specified by {\tt
+  fixIndicies} in {\tt network} are updated with {\tt values}.  Here, {\tt
+  fixIndicies} has been observed to be either a character or double vector:
+In R, list indices are typically doubles, but can be characters (if the list
+or vector being indexed has a {\tt names} attribute).
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+fixGenes <- function (network, fixIndices, values) {
+  ...
+  network$fixed[fixIndices] <- as.integer(values)
+  ...
+\end{lstlisting}}\caption{Example of character/double argument (from {\tt BoolNet} package).}\label{fig:chardbl}\end{figure}%$
+
+In Figure~\ref{fig:matvec}, we see that the function can take in a vector,
+but immediately transforms it (transpose, with {\tt t}) into a matrix.  As
+an idea, matrix/vector polymorphism seems sensible, as mathematically
+vectors are matrices.  R echoes this by making conversion between the two
+``types'' easy: \code{as.vector(m)} flattens a matrix \code{m} into a vector
+(e.g., \code{as.vector(matrix(2, 2, 2)) == c(2, 2, 2, 2)}), and
+\code{as.matrix(v)} builds a {\tt length(v) x 1} matrix (e.g.,
+\code{as.matrix(c(1, 2)) == matrix(data=c(1, 2))}).  \AT{TODO fix line break
+  in code at end of last sentence}
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+tee <- function (x, theta, D1, D2, phi)  {
+    if (is.vector(x)) x <- t(x)
+    ...
+\end{lstlisting}}\caption{Example of matrix/(\dblv) argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}
+
+%
+%
+%
 %
@@
 \bibliography{bib/biblio,bib/jv,bib/r,bib/new}
-
-%%
-%%
-%%
-%%
-%\subsection{Types in R}
-%
-%\AT{I'll pull from this section when appropriate, as the discussion of the type systems becomes relevant.}
-%
-%%
-%%
-%\subsubsection{What are possible type systems?}
-%
-%Great question.
-%Some things:
-%
-%\begin{itemize}
-%
-%\item {\it The Obvious Type System (OTS)}: this type system corresponds directly with the type information that one can fetch using the {\tt typeof}, {\tt attributes}, and {\tt class} functions.
-%In this system, there is no distinction between vectors and scalars, and lists of differently-typed elements (e.g., a list of integers vs.
-a list of strings), and there is a distinction between integers and doubles, and the type of NULL and all other types.
-%This type system isn't particularly informative to a programmer, though it does reflect the treatment of values in the R runtime, for the {\tt typeof} a value is the type that the value has according to the R internals.
-%
-%\item {\it The Fine-Grained Type System (L1)}: this type systems includes a number of relevant additions.
-%Chiefly, it distinguishes between vectors and scalars, as it is conceivable that boxing scalars as single-element vectors might be overkill.
-%In addition, list types are parameterized over the types of their elements (e.g., list<integer> vs. list<character>).
-%Finally, NAs ``in the wild'' are given the type {\tt raw\_NA}, as {\tt typeof(NA) = logical} in R.
-%
-%\item {\it The Null-and-NA-Free Type System (NNTS)}: most languages consider NULLs to inhabit all types, and R does so for NAs \AT{elaborate}.
-%This gives us ... .
-%
-%\end{itemize}
-%
-%Of course, one could conceive of a more fine-grained type system than L1, but it's unclear that we would gain much in doing so.
-%Indeed, there is a space to explore in slightly less fine-grained type systems, which make modifications including:
-%
-%\begin{itemize}
-%
-%\item {\bf collapsing vectors and scalars}: this yields a type system slightly closer to OTS.
-%
-%\item {\bf rolling NULL into other types}: in most languages, NULL is an inhabitant of all types, and this adjustment seems interesting.
-%Note that this goes beyond even OTS, as in R {\tt typeof(NULL) = NULL}.
-%
-%\item {\bf rolling NA into other types}: \AT{i should try this}
-%similarly, NA is an inhabitant of all types, though NA standing alone has type {\tt logical}.
-%L1 distinguishes such NAs with the {\tt raw\_NA} type, and if instead we consider NA to inhabit all types we may \AT{win}.
-%
-%\item {\bf combining integers and doubles}: as in most languages, doubles and ints can stand-in for each other in nearly every scenario.
-%Even if not, it is trivial to convert from one to the other, particularly converting ints to doubles.
-%Aside, in R a double can be used to index a list, and R will floor it when performing the lookup.
-%
-%\end{itemize}
-%\subsection{Usage Patterns}
-%
-%In this section, we will discuss function usage patterns which arose in our analysis.
-%We will start by looking at the morphicity of functions.
-%
-%\subsubsection{Function Argument and Return Morphicity}
-%
-%First and foremost, we would like to know how often R programmers create polymorphic functions.
-%Recall that we define a polymorphic function to be a function with at least one polymorphic argument, or a polymorphic return.
-%We will turn our attention now to the data in \todo{Figure}, and go through each entry in the table.
-%
-%Here, we see that the vast majority of function arguments are indeed monomorphic.
-%Monomorphic arguments are easy to annotate, as the {\tt typeof} the arguments is exactly the most precise annotation we could give, at least in terms of \textit{type}.
-%That said, types alone don't always paint the whole picture:
-%recall that \textit{attributes} are an R language feature which allows programmers to stick metadata onto values.
-%So in which circumstances \textit{do} types paint the whole picture?
-%
-%The second data point in \textbf{TODO FIGURE} indicates that a
-%\isit{majority} of function arguments are monomorphic in type \textit{and}
-%have no attributes.  These represent arguments which are truly trivial to
-%annotate, as the {\tt typeof} an argument perfectly describes the usage of
-%that argument.  That said, some attributes arise naturally in R: For
-%instance, names in a named list (e.g. {\tt x} and {\tt y} in {\tt list(x=0,
-%  y=0)}) appear as an attribute on the value.  In
-%Section~\ref{sec:method:attributes}, we outlined these naturally-occurring
-%attribute patterns, and the third data point in \textbf{TODO FIGURE} shows
-%that a nontrivial amount of functions are monomorphic in type with said
-%natural attribute patterns.
-%
-%Another facet of type information in R is in the {\tt class} attribute.
-%Recall that values have a \textit{class} in addition to a type: For example,
-%a {\tt data.frame} has type {\tt list} and class {\tt data.frame}.
-%\todo{Plug classes earlier} R has a number of built-in classes, such as
-%\todo{X, Y, and Z}, but users are free to redefine the class of any value at
-%runtime, and easily define new classes.  The next data point in
-%\todo{Figure} shows that user-defined classes don't appear altogether often,
-%though they do indeed feature.  \AT{In a following section, we will discuss
-%  how the usage of these classes manifests itself.}
-%
-%\AT{Separate section for functions? Right now, this is at argument granularity.}
-%
-%\subsubsection{Type Signatures}
-%
-%In the last section, we presented a high-level overview of the morphicity ... .
-%
-%\subsubsection{Attribute Signatures}
-%
-%Recall that attributes are a way for programmers to store metadata on values in R.
-%What are the common attribute patterns?
-%And how often is the attribute pattern polymorphic?
-%
-%\subsubsection{Takeaways}
-%
-%\begin{itemize}
-%    \item the vast majority of arguments are monomorphic in type;
-%    \item of those, over 60\% have no attribute information;
-%    \item of the 40\% with attribute information, roughly 1/2 have fairly
-%      simple attributes corresponding to base R constructs (named lists and
-%      vectors, matrices, and data frames);
-%    \item now, of those arguments which are polymorphic, a sizable chunk (well over half) have defensible signatures (e.g., double and character for named list indexing, double and integer for obvious reasons, etc.).
-%\end{itemize}
-%
-%In short, it looks like R (package) programmers are reasonable.  I'd
-%conjecture that a lot of the polymorphism (e.g., double and list) is coming
-%from how easy it is to use either type in a given situation (e.g.,
-%converting from vector of doubles to list or vice versa is simple).
-
 \end{document}

### 2019-04-02 paper/main.tex
merging

--- 2019-04-02 paper/main.tex
@@
 % Macros for type names?
-\newcommand{\reals}{$real$\xspace}
-\newcommand{\realv}{$\overline{real}$\xspace}
-\newcommand{\chars}{$char$\xspace}
-\newcommand{\charv}{$\overline{char}$\xspace}
-\newcommand{\logis}{$logical$\xspace}
-\newcommand{\logiv}{$\overline{logical}$\xspace}
-\newcommand{\dbls}{$double$\xspace}
-\newcommand{\dblv}{$\overline{double}$\xspace}
-\newcommand{\ints}{$integer$\xspace}
-\newcommand{\intv}{$\overline{integer}$\xspace}
+\newcommand{\reals}{$\mathtt {\rm I\!R}$\xspace}
+\newcommand{\realv}{$\overline{\rm I\!R}$\xspace}
 
@@
   at the length of {\tt x} to determine if {\tt x} is a scalar or a vector,
-  generating the annotation {\tt \dbls} or {\tt \dblv} accordingly.  This is
+  generating the annotation {\tt \sD} or {\tt \D} accordingly.  This is
   also true for integer, complex, raw, logical, and character {\tt typeof}s
@@
   }
-\end{lstlisting}}\caption{Example of a data.frame/(\dblv) argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}
+\end{lstlisting}}\caption{Example of a data.frame/(\D) argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}
 
@@
     ...
-\end{lstlisting}}\caption{Example of matrix/(\dblv) argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}
+\end{lstlisting}}\caption{Example of matrix/(\D) argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}
 
@@
 \bf Argument & \bf Type & \bf Class &\bf Attributes \\
-x & \dbls & numeric & \{\} \\
-y & \dblv & numeric & \{\} \\
-retv & list<double> & list & \{\xspace example: \chars\}  
+x & \sD & numeric & \{\} \\
+y & \D & numeric & \{\} \\
+retv & list<double> & list & \{\xspace example: \sC\}  
 \end{tabular}
@@
   \hline
-  \dblv, \intv & 7742 & 9.66 \\ 
-  \dbls, \ints & 7201 & 8.99 \\ 
-  matrix, \dblv & 5384 & 6.72 \\ 
-  NULL, \chars & 2325 & 2.90 \\ 
+  \D, \I & 7742 & 9.66 \\ 
+  \sD, \sI & 7201 & 8.99 \\ 
+  matrix, \D & 5384 & 6.72 \\ 
+  NULL, \sC & 2325 & 2.90 \\ 
   list$<$any$>$, list$<$double$>$ & 2231 & 2.78 \\ 
   data.frame, matrix & 2116 & 2.64 \\ 
-  NULL, \dbls & 2082 & 2.60 \\ 
-  NULL, \dblv & 2069 & 2.58 \\ 
+  NULL, \sD & 2082 & 2.60 \\ 
+  NULL, \D & 2069 & 2.58 \\ 
   list$<$any$>$, list$<$list$>$ & 1432 & 1.79 \\ 
-  NULL, \charv & 1380 & 1.72 \\
+  NULL, \C & 1380 & 1.72 \\
      \hline
@@
 We note that an argument which exhibits a signature including e.g.
-\{\dbls, \dblv\} really must be an argument designed to take vectors, as recall that unit length vectors appear as scalars by our definition.
+\{\sD, \D\} really must be an argument designed to take vectors, as recall that unit length vectors appear as scalars by our definition.
 We have already collapsed this distinction.
@@
   list$<$any$>$, list$<$real$>$ & 2799 & 6.13 \\ 
-  \charv, \realv & 2774 & 6.08 \\ 
-  \chars, \reals & 2552 & 5.59 \\ 
+  \C, \realv & 2774 & 6.08 \\ 
+  \sC, \reals & 2552 & 5.59 \\ 
   data.frame, matrix & 2167 & 4.75 \\ 
@@
   data.frame, \realv & 1046 & 2.29 \\ 
-  \logis, \reals & 920 & 2.02 \\ 
-  \chars, \realv & 868 & 1.90 \\
+  \sL, \reals & 920 & 2.02 \\ 
+  \sC, \realv & 868 & 1.90 \\
    \hline
@@
 
-In L2, we aim to incorporate R's other reflective capabilities to address some of the more subtle instances of polymorphism.
-
-We observe that the act of ascribing a class to a value in R carries the intent of the programmer.
-Now, not all classes are user-defined, and R has a number of built-in or otherwise standard classes, ranging from truly primitive classes such as ``numeric'' to the more complex ``data.frame''.
-Thus, before considering the insights provided by the class of values, we transform each of R's standard classes into a ``primitive'' class, which will serve to highlight user-defined classes and those particular to some packages.
-
-\AT{I don't know the best way to approach this subsection, so I'm just gonna start looking at some stuff and see where it leads.}
-
-First, we consider the purpose of classes in R.
-The
-
-As before, we start with the counts, see Table~\ref{tab:classcountsL2}.
+In L2, we aim to incorporate R's other reflective capabilities to address some of the more subtle instances of polymorphism that we have observed.
+
+First, we reiterate that there are a finite number of results from the {\tt typeof} function, and this is not the case for \code{class} and \code{attributes}.
+Classes and attributes can be arbitrarily defined by the user, and users create their own classes and assign their own attributes frequently.
+Unlike types (in the sense of \code{typeof}), the number of possible classes and attributes is arbitrary.
+
+Of course, not all classes are user-defined, and R has a number of built-in or otherwise standard classes, ranging from truly primitive classes such as ``numeric'' to the more complex ``data.frame''.
+We have already used these when building our more refined types, and so we transform each of R's standard classes into a ``primitive'' class, which will minimize their impact and thus highlight user-defined classes.
+
+As for attributes, again some patterns arise naturally, for instance all matrices have a ``dims'' attribute (indeed, it's this attribute that makes a matrix a matrix).
+\AT{This sounds like it's implying that we should do something about that, but I'm not sure if that's a good idea.}
+In principle, attributes are name, value pairs, but to keep things simple we will reduce these to name, type pairs, so the name of an attribute and the type that it was observed to have.
+We will show an L0 type for attributes.
+
+To see how often arguments are polymorphic with respect to class and attributes, we refer to Table~\ref{tab:classcountsL2}.
+\AT{Showing full monomorphic all the time might be a bit confusing?}
 
@@
   \hline
-Full Monomorphic & 777 & 84.96 \\ 
+Full Monomorphic & 783K & 85.59 \\ 
 %  Monomorphic in Type & 869.00 & 95.01 \\ 
-  Monomorphic in Class & 850 & 92.94 \\ 
-%  Monomorphic in Attribute Pattern & 825.00 & 90.27 \\ 
-  Total Seen & 914 & --- \\ 
-   \hline
-\end{tabular}
-\caption{Account of class-based polymorphism in L2.}
+  Monomorphic in Class & 850K & 92.94 \\ 
+  Monomorphic in Attribute Pattern & 836K & 91.41 \\ 
+  Total Seen & 914K & --- \\ 
+   \hline
+\end{tabular}
+\caption{Account of class- and attribute-based polymorphism in L2.}
 \end{table}
 
-Top signatures in Table~\ref{tab:classpolyL2}.
+The numbers here are in line with the type-based polymorphism of L1.
+As before, to build up a better sense of what kind of polymorphism we're dealing with, we extract some of the most common signatures.
+We begin with classes, and refer to Table~\ref{tab:classpolyL2}.
 
@@
   \hline
-Class & Count & Precentage \\ 
+Class & Count & Percentage \\ 
   \hline
@@
   formula, primitive & 639 & 0.99 \\ 
-  xts, zoo & 588 & 0.91 \\ 
-  glm, lm & 494 & 0.77 \\ 
-  primitive, ts & 432 & 0.67 \\ 
-  mts, primitive, ts & 384 & 0.59 \\ 
-  huxtable, primitive & 318 & 0.49 \\ 
-   \hline
-\end{tabular}
-\caption{Top 10 signatures for class-based polymorphism in L2.}
+%  xts, zoo & 588 & 0.91 \\ 
+%  glm, lm & 494 & 0.77 \\ 
+%  primitive, ts & 432 & 0.67 \\ 
+%  mts, primitive, ts & 384 & 0.59 \\ 
+%  huxtable, primitive & 318 & 0.49 \\ 
+   \hline
+\end{tabular}
+\caption{Top 5 class signatures for class-based polymorphism in L2.}
 \end{table}
 
-Given the low percentages overall, we see that there are no clear patterns here.
-This data does highlight some common classes across R packages:
-tbl and tbl\_df are from the dplyr package, as classes for data objects.
-data.table is the primary data type in the data.table package, again for data objects.
-POSIXct and POSIXt are for dates.
+The first thing to note is the low percentages overall: unlike with types in L0, L1, or L1+, there are no major contributors for class-based polymorphism.
+For instance, the single highest source of polymorphism here is due to the ``tbl'' and ``tbl\_df'' classes (of the dplyr package).
+For what it's worth, these are classes ascribed to dplyr data objects (i.e., spruced-up data frames).
+The next largest contributor is the array class, which is a class for N>2-dimensional matrices (part of the core R language).
+\AT{Maybe I should also turn arrays into primitive class?}
 
@@
   list$<$any$>$, list$<$list$>$, list$<$real$>$ & dendrogram & 24 & 1.77 \\ 
-  list$<$any$>$, list$<$real$>$ & phylo & 23 & 1.69 \\ 
-  list$<$any$>$, list$<$list$>$, list$<$real$>$, scalar/real & dendrogram & 21 & 1.55 \\ 
-  list$<$any$>$, list$<$language$>$ & uneval & 18 & 1.33 \\ 
-  list$<$any$>$, list$<$list$>$ & dendrogram & 18 & 1.33 \\ 
-  list$<$any$>$, list$<$list$>$ & lazy\_dots & 17 & 1.25 \\ 
-   \hline
-\end{tabular}
+%  list$<$any$>$, list$<$real$>$ & phylo & 23 & 1.69 \\ 
+%  list$<$any$>$, list$<$list$>$, list$<$real$>$, scalar/real & dendrogram & 21 & 1.55 \\ 
+%  list$<$any$>$, list$<$language$>$ & uneval & 18 & 1.33 \\ 
+%  list$<$any$>$, list$<$list$>$ & dendrogram & 18 & 1.33 \\ 
+%  list$<$any$>$, list$<$list$>$ & lazy\_dots & 17 & 1.25 \\ 
+   \hline
+\end{tabular}
+\caption{Top instances of type-polymorphic but class-monomorphic signatures.}
 \end{table}
@@
 % latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Mon Apr  1 19:21:13 2019
+% Mon Apr  1 23:06:59 2019
 \begin{table}[ht]
@@
   \hline
-\{\}, \{\xspace names: \charv\} & 9274.00 & 10.43 \\ 
-  \{\}, \{\xspace names: \chars\} & 4900.00 & 5.51 \\ 
-  \{\}, \{\xspace dim: \intv\} & 4758.00 & 5.35 \\ 
-  \{\xspace names: \chars\}, \{\xspace names: \charv\} & 3457.00 & 3.89 \\ 
-  \{\}, \{\xspace names: \chars\}, \{\xspace names: \charv\} & 2415.00 & 2.72 \\ 
-  \{\xspace dim: \intv, dimnames:list$<$any$>$ \}, \{\xspace dim:\intv\} & 2264.00 & 2.55 \\ 
-  \{\}, \{\xspace srcref: \intv\} & 2037.00 & 2.29 \\ 
-  \{\xspace dim: \intv, dimnames: list$<$character$>$\}, \{ dim:\intv\} & 1734.00 & 1.95 \\ 
-  \{\xspace class: \chars, names: \charv, row.names:\charv\}, \{class:\chars, names:\charv, row.names:\intv\} & 1464.00 & 1.65 \\ 
-  \{\xspace class: \chars, names: \charv\}, \{ class:\charv, names:\charv\} & 1277.00 & 1.44 \\ 
-   \hline
-\end{tabular}
-\caption{Top attribute patterns. \AT{TODO fix this up}}
-\end{table}\charv
+\{\}, \{names: \C\} & 16589 & 21.11 \\ 
+  \{\}, \{dim: \I\} & 5010 & 6.38 \\ 
+  \{dim:\I,dimnames:list$<$any$>$\}, \{dim:\I\} & 2264 & 2.88 \\ 
+  \{class:\C,names:\C,row.names:\C\}, \{class:\C,names:\C,row.names:\I\} & 2102 & 2.68 \\ 
+  \{\}, \{srcref:\I\} & 2037 & 2.59 \\ 
+  \{dim:\I,dimnames:list$<$\C$>$\}, \{dim:\I\} & 1740 & 2.21 \\ 
+  \{\}, \{class:\C\} & 1673 & 2.13 \\ 
+  \{\}, \{class:\C,levels:\C\} & 1437 & 1.83 \\ 
+  \{\}, \{class:\C,names:\C\} & 1324 & 1.69 \\ 
+  \{\}, \{class:\C,names:\C,row.names:\I\} & 1251 & 1.59 \\ 
+   \hline
+\end{tabular}
+\end{table}
 

### 2019-04-02 paper/main.tex

--- 2019-04-02 paper/main.tex
@@
 \bf Argument & \bf Type & \bf Class &\bf Attributes \\
-x & \dbls & numeric & \{\} \\
-y & \dblv & numeric & \{\} \\
-retv & list<double> & list & \{\xspace example: \chars\}  
+x & \sD & numeric & \{\} \\
+y & \D & numeric & \{\} \\
+retv & list<double> & list & \{\xspace example: \sC\}
 \end{tabular}
@@
   \hline
-  \dblv, \intv & 7742 & 9.66 \\ 
-  \dbls, \ints & 7201 & 8.99 \\ 
-  matrix, \dblv & 5384 & 6.72 \\ 
-  NULL, \chars & 2325 & 2.90 \\ 
+  \D, \I & 7742 & 9.66 \\
+  \sD, \sI & 7201 & 8.99 \\
+  matrix, \D & 5384 & 6.72 \\
+  NULL, \sC & 2325 & 2.90 \\
   list$<$any$>$, list$<$double$>$ & 2231 & 2.78 \\
   data.frame, matrix & 2116 & 2.64 \\
-  NULL, \dbls & 2082 & 2.60 \\ 
-  NULL, \dblv & 2069 & 2.58 \\ 
+  NULL, \sD & 2082 & 2.60 \\
+  NULL, \D & 2069 & 2.58 \\
   list$<$any$>$, list$<$list$>$ & 1432 & 1.79 \\
-  NULL, \charv & 1380 & 1.72 \\
+  NULL, \C & 1380 & 1.72 \\
      \hline
@@
   list$<$any$>$, list$<$real$>$ & 2799 & 6.13 \\
-  \charv, \realv & 2774 & 6.08 \\ 
-  \chars, \reals & 2552 & 5.59 \\ 
+  \C, \realv & 2774 & 6.08 \\
+  \sC, \reals & 2552 & 5.59 \\
   data.frame, matrix & 2167 & 4.75 \\
@@
   data.frame, \realv & 1046 & 2.29 \\
-  \logis, \reals & 920 & 2.02 \\ 
-  \chars, \realv & 868 & 1.90 \\
+  \sL, \reals & 920 & 2.02 \\
+  \sC, \realv & 868 & 1.90 \\
    \hline
@@
   \hline
-Full Monomorphic & 777 & 84.96 \\ 
+Full Monomorphic & 783K & 85.59 \\
 %  Monomorphic in Type & 869.00 & 95.01 \\
-  Monomorphic in Class & 850 & 92.94 \\ 
-%  Monomorphic in Attribute Pattern & 825.00 & 90.27 \\ 
-  Total Seen & 914 & --- \\ 
-   \hline
-\end{tabular}
-\caption{Account of class-based polymorphism in L2.}
+  Monomorphic in Class & 850K & 92.94 \\
+  Monomorphic in Attribute Pattern & 836K & 91.41 \\
+  Total Seen & 914K & --- \\
+   \hline
+\end{tabular}
+\caption{Account of class- and attribute-based polymorphism in L2.}
 \end{table}
 
-Top signatures in Table~\ref{tab:classpolyL2}.
+The numbers here are in line with the type-based polymorphism of L1.
+As before, to build up a better sense of what kind of polymorphism we're dealing with, we extract some of the most common signatures.
+We begin with classes, and refer to Table~\ref{tab:classpolyL2}.
 
@@
   \hline
-Class & Count & Precentage \\ 
+Class & Count & Percentage \\
   \hline
@@
   formula, primitive & 639 & 0.99 \\
-  xts, zoo & 588 & 0.91 \\ 
-  glm, lm & 494 & 0.77 \\ 
-  primitive, ts & 432 & 0.67 \\ 
-  mts, primitive, ts & 384 & 0.59 \\ 
-  huxtable, primitive & 318 & 0.49 \\ 
-   \hline
-\end{tabular}
-\caption{Top 10 signatures for class-based polymorphism in L2.}
+%  xts, zoo & 588 & 0.91 \\
+%  glm, lm & 494 & 0.77 \\
+%  primitive, ts & 432 & 0.67 \\
+%  mts, primitive, ts & 384 & 0.59 \\
+%  huxtable, primitive & 318 & 0.49 \\
+   \hline
+\end{tabular}
+\caption{Top 5 class signatures for class-based polymorphism in L2.}
 \end{table}
 
-Given the low percentages overall, we see that there are no clear patterns here.
-This data does highlight some common classes across R packages:
-tbl and tbl\_df are from the dplyr package, as classes for data objects.
-data.table is the primary data type in the data.table package, again for data objects.
-POSIXct and POSIXt are for dates.
+The first thing to note is the low percentages overall: unlike with types in L0, L1, or L1+, there are no major contributors for class-based polymorphism.
+For instance, the single highest source of polymorphism here is due to the ``tbl'' and ``tbl\_df'' classes (of the dplyr package).
+For what it's worth, these are classes ascribed to dplyr data objects (i.e., spruced-up data frames).
+The next largest contributor is the array class, which is a class for N>2-dimensional matrices (part of the core R language).
+\AT{Maybe I should also turn arrays into primitive class?}
 
@@
   list$<$any$>$, list$<$list$>$, list$<$real$>$ & dendrogram & 24 & 1.77 \\
-  list$<$any$>$, list$<$real$>$ & phylo & 23 & 1.69 \\ 
-  list$<$any$>$, list$<$list$>$, list$<$real$>$, scalar/real & dendrogram & 21 & 1.55 \\ 
-  list$<$any$>$, list$<$language$>$ & uneval & 18 & 1.33 \\ 
-  list$<$any$>$, list$<$list$>$ & dendrogram & 18 & 1.33 \\ 
-  list$<$any$>$, list$<$list$>$ & lazy\_dots & 17 & 1.25 \\ 
-   \hline
-\end{tabular}
+%  list$<$any$>$, list$<$real$>$ & phylo & 23 & 1.69 \\
+%  list$<$any$>$, list$<$list$>$, list$<$real$>$, scalar/real & dendrogram & 21 & 1.55 \\
+%  list$<$any$>$, list$<$language$>$ & uneval & 18 & 1.33 \\
+%  list$<$any$>$, list$<$list$>$ & dendrogram & 18 & 1.33 \\
+%  list$<$any$>$, list$<$list$>$ & lazy\_dots & 17 & 1.25 \\
+   \hline
+\end{tabular}
+\caption{Top instances of type-polymorphic but class-monomorphic signatures.}
 \end{table}
@@
   \hline
-\{\}, \{\xspace names: \charv\} & 9274.00 & 10.43 \\ 
-  \{\}, \{\xspace names: \chars\} & 4900.00 & 5.51 \\ 
-  \{\}, \{\xspace dim: \intv\} & 4758.00 & 5.35 \\ 
-  \{\xspace names: \chars\}, \{\xspace names: \charv\} & 3457.00 & 3.89 \\ 
-  \{\}, \{\xspace names: \chars\}, \{\xspace names: \charv\} & 2415.00 & 2.72 \\ 
-  \{\xspace dim: \intv, dimnames:list$<$any$>$ \}, \{\xspace dim:\intv\} & 2264.00 & 2.55 \\ 
-  \{\}, \{\xspace srcref: \intv\} & 2037.00 & 2.29 \\ 
-  \{\xspace dim: \intv, dimnames: list$<$character$>$\}, \{ dim:\intv\} & 1734.00 & 1.95 \\ 
-  \{\xspace class: \chars, names: \charv, row.names:\charv\}, \{class:\chars, names:\charv, row.names:\intv\} & 1464.00 & 1.65 \\ 
-  \{\xspace class: \chars, names: \charv\}, \{ class:\charv, names:\charv\} & 1277.00 & 1.44 \\ 
-   \hline
-\end{tabular}
-\caption{Top attribute patterns. \AT{TODO fix this up}}
-\end{table}\charv
+\{\}, \{names: \C\} & 16589 & 21.11 \\
+  \{\}, \{dim: \I\} & 5010 & 6.38 \\
+  \{dim:\I,dimnames:list$<$any$>$\}, \{dim:\I\} & 2264 & 2.88 \\
+  \{class:\C,names:\C,row.names:\C\}, \{class:\C,names:\C,row.names:\I\} & 2102 & 2.68 \\
+  \{\}, \{srcref:\I\} & 2037 & 2.59 \\
+  \{dim:\I,dimnames:list$<$\C$>$\}, \{dim:\I\} & 1740 & 2.21 \\
+  \{\}, \{class:\C\} & 1673 & 2.13 \\
+  \{\}, \{class:\C,levels:\C\} & 1437 & 1.83 \\
+  \{\}, \{class:\C,names:\C\} & 1324 & 1.69 \\
+  \{\}, \{class:\C,names:\C,row.names:\I\} & 1251 & 1.59 \\
+   \hline
+\end{tabular}
+\end{table}
 

### 2019-04-02 notebooks/paper_L123_results_section.Rmd
got rid of leftover macro and fixed compile error, also some new code

@@
 
-Step 4. Translate to L3?
+Step 4: Translate to L1+.
+
+In L1+, we will collapse some obviously collapsible distinctions (e.g., reals).
+```{r}
+los <- lapply(los, function(s) lapply(s, fold_NULL_and_NA_into_other_types_df))
+
+los <- lapply(los, function(lof) {lapply(lof, function(df) {change_type_systems_df(df, type_map_sv_to_real)})})
+
+los <- lapply(los, function(lofuns) {lapply(lofuns, combine_scalar_vector_where_appropriate)})
+
+null_na_fold_real_arg_counts <- build_arg_count_df_from_lopkgs(los)
+null_na_fold_real_fun_counts <- build_fun_count_df_from_lopkgs(los)
+```
+
+TODO oops scalar vector thing.
+
+Write counts.
+```{r}
+xtable(null_na_fold_real_arg_counts) %>% print(include.rownames=F)
+xtable(null_na_fold_real_fun_counts) %>% print(include.rownames=F)
+```
+
+Breakdown by argument signature.
+```{r}
+poly_type_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_type_args))
+poly_type_args <- poly_type_args[order(-poly_type_args$count),] # sort by count
+poly_type_args$perc <- round(poly_type_args$count/sum(poly_type_args$count) * 100, 2)
+```
+
+Print top 10.
+```{r}
+format_df_for_print(poly_type_args[1:10,]) -> put_me
+put_me
+xtable(put_me) %>% print(include.rownames=F)
+```
+
+Step 5. Translate to L3?
 What is L3?
@@
 ```{r}
+xtable(L3_c_basic_df) %>% print(include.rownames=F)
+```
 
+```{r}
+format_df_for_print(L3_poly_class_args[1:10,]) -> put_me
+put_me
+xtable(put_me) %>% print(include.rownames=F)
 ```
+
+Let's find out when class is monomorphic and type is polymorphic.
+```{r}
+num_poly_type_not_poly_class <- apply_count_to_lopkg_c(los, args_count_all_poly_type_mono_class)
+```
+
+Are there patterns?
+```{r}
+lo_pt_mc <- reduce_extracted_lopkgs_to_one_new(apply_extract_to_lopkg(los, extract_poly_type_args_w_mono_class))
+lo_pt_mc <- lo_pt_mc[order(-lo_pt_mc$count),] # sort by count
+lo_pt_mc$perc <- round(lo_pt_mc$count/sum(lo_pt_mc$count) * 100, 2)
+
+format_df_for_print_new(lo_pt_mc[1:10,]) -> put_me
+put_me
+xtable(put_me) %>% print(include.rownames=F)
+```
+
+How about attributes?
+```{r}
+los_a <- lapply(los, function(lofun) lapply(lofun, function(df) convert_df_attr_to_ts(df, type_map_T0_to_r)))
+
+L3_poly_attr_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los_a, extract_poly_attr_args))
+L3_poly_attr_args <- L3_poly_attr_args[order(-L3_poly_attr_args$count),] # sort by count
+L3_poly_attr_args$perc <- round(L3_poly_attr_args$count/sum(L3_poly_attr_args$count) * 100, 2)
+
+format_df_for_print_new(L3_poly_attr_args[1:10,]) -> put_me
+put_me
+xtable(put_me) %>% print(include.rownames=F)
+```

### 2019-04-02 paper/main.tex
got rid of leftover macro and fixed compile error, also some new code

@@
   }
-\end{lstlisting}}\caption{Example of a data.frame/(\dblv) argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}
-
-In Figure~\ref{fig:chardbl}, we see a function which takes in a list ({\tt
-  network}), a vector of indices of that list ({\tt fixIndices}), and a
-vector of values ({\tt values}).  The locations specified by {\tt
-  fixIndicies} in {\tt network} are updated with {\tt values}.  Here, {\tt
-  fixIndicies} has been observed to be either a character or double vector:
+\end{lstlisting}}\caption{Example of a data.frame/(\D) argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}
+
+In Figure~\ref{fig:chardbl}, we see a function which takes in a list ({\tt network}), a vector of indices of that list ({\tt fixIndices}), and a
+vector of values ({\tt values}).
+The locations specified by {\tt fixIndicies} in {\tt network} are updated with {\tt values}.
+Here, {\tt fixIndicies} has been observed to be either a character or double vector:
 In R, list indices are typically doubles, but can be characters (if the list

### 2019-04-02 paper/main.tex
a bit more prose in results

@@
   formula, primitive & 639 & 0.99 \\
-%  xts, zoo & 588 & 0.91 \\
-%  glm, lm & 494 & 0.77 \\
-%  primitive, ts & 432 & 0.67 \\
-%  mts, primitive, ts & 384 & 0.59 \\
-%  huxtable, primitive & 318 & 0.49 \\
    \hline
@@
 To see how we might use class information, we should see how often a monomorphic class signature co-occurs with a polymorphic type signature.
-We count \NUMPOLYTYPEMONOCLASS argument positions which match this criteria.
-
-% latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Mon Apr  1 16:50:22 2019
-\begin{table}[ht]
-\centering
-\begin{tabular}{llrl}
-  \hline
-  Type Sig. & Class Sig. & Count & Percentage \\
-  \hline
-  list$<$any$>$, list$<$character$>$ & roxy\_block & 38 & 2.80 \\
-  list$<$any$>$, list$<$list$>$ & Rd & 25 & 1.84 \\
-  list$<$any$>$, list$<$real$>$ & TSdata & 25 & 1.84 \\
-  list$<$any$>$, list$<$list$>$ & xml\_nodeset & 24 & 1.77 \\
-  list$<$any$>$, list$<$list$>$, list$<$real$>$ & dendrogram & 24 & 1.77 \\
-%  list$<$any$>$, list$<$real$>$ & phylo & 23 & 1.69 \\
-%  list$<$any$>$, list$<$list$>$, list$<$real$>$, scalar/real & dendrogram & 21 & 1.55 \\
-%  list$<$any$>$, list$<$language$>$ & uneval & 18 & 1.33 \\
-%  list$<$any$>$, list$<$list$>$ & dendrogram & 18 & 1.33 \\
-%  list$<$any$>$, list$<$list$>$ & lazy\_dots & 17 & 1.25 \\
-   \hline
-\end{tabular}
-\caption{Top instances of type-polymorphic but class-monomorphic signatures.}
-\end{table}
-
-\AT{I don't think this table is telling us much...}
-
-Top attributes, maybe?
+We count \NUMPOLYTYPEMONOCLASS argument positions which match this criteria, which certainly doesn't explain the much higher number of list polymorphism in L1+.
+
+In our analysis, we did not encounter any substantial patterns with class-based polymorphism.
+Class is nonetheless important in R, as the dynamic dispatcher uses class information to select the correct method to call.
+In short, a small optional annotation indicating the class of an argument may well be enough.
+
+Let us now turn our attention to attributes.
+The top contributors for attribute-based polymorphism can be found in Table~\ref{tab:topattrpoly}.
 
@@
 \begin{table}[ht]
+\label{tab:topattrpoly}
 \centering
@@
   \hline
-attr & count & perc \\
+Attribute Signature & Count & Percentage \\
   \hline
@@
 
-A lot of the polymorphism that we see is because we haven't propagated our changes through the attribute types.
-Will fix this.
+First, the largest source of attribute-based polymorphism is in optional list names.
+The names attribute on a vector or list allows indexing with name rather than numeric position.
+Another significant source of this polymorphism is matrices, which can be seen in all the ``dim'' and ``dimnames'' attributes.
 

### 2019-04-02 paper/main.tex
small todoo list for data tmr
minor

@@
 \I, double \D, character \C, logical \L, raw \R, complex \X), lists of any
-type (\l), and scalar values (environment \sE, closures \sF, symbols \sS and
-null \sN). We add the type \ANY to stand in for any type.
+type (\l), and scalar values (environment \sE, closures \sF, symbols \sY, S4
+object \sS, and null \sN). We add the type \ANY to stand in for any type.
 
@@
 
-\section{Results}\label{sec:results}
-
-Having built an understanding of our data collection and analysis pipeline,
-we will turn our attention to the results of our analysis.
-
-\subsection{L0}
-
-In this section, we will begin by breaking down our corpus of data with L0
-in mind.  From this perspective, we will see how current R programmers see
-types in the language, and we will leverage our newfound insight to design a
-type system which should be relevant to them.
-
-Without further ado, let's look at the data, first focusing on a broad
-overview of the polymorphism in L0.  Figure~\ref{fig:L1argcounts} shows the
-polymorphicity of argument positions in R function.  From there, we build up
-a picture of the polymorphicity of functions as a whole, and the data for
-this function polymorphism can be found in Figure~\ref{fig:L1funcounts}.
+\section{Analysis Results}\label{sec:results}
+
+This section presents our results for each of the levels we have considered.
+
+\subsection{L0 polymorphism}
+
+At L0, the type system is aligned with the types of the core language.  We
+have observed over 173K monomorphic functions; this accounts for 80\% of our
+corpus.  In terms of argument positions, 92\% of arguments are monomorphic.
+
+Figure~\ref{fig:L1argcounts} shows the polymorphicity of argument positions
+ in R function.  From there, we build up a picture of the polymorphicity of
+ functions as a whole, and the data for this function polymorphism can be
+ found in Figure~\ref{fig:L1funcounts}.
 
@@
 % Fri Mar 29 18:10:23 2019
-\begin{table}[ht]
-\label{fig:L1argcounts}
-\centering
-\begin{tabular}{lrr}
-  \hline
-Type & Count & Percentage \\
-  \hline
+\begin{table}[ht]\label{fig:L1argcounts}\centering\begin{tabular}{lrr}  \hline
+Type                   & Count & Percentage \\  \hline
 Full Monomorphic       & 784K & 85.76 \\
@@
 
-As for the polymorphism of functions as a whole, we glean from Figure~\ref{fig:L1funcounts} that a sizeable number of functions are polymorphic.
-Functions are said to be polymorphic if at least one of their arguments or return are polymorphic.
-\AT{Say more?}
-
-At present, we will only concern ourselves with \code{typeof} polymorphism, and we will later worry about classes and attributes.
-To get an idea of what kinds of polymorphism occur, we will look at the most commonly occurring polymorphic argument signatures.
-Table~\ref{tab:L0toppoly} has the top polymorphic signatures for L0.
+As for the polymorphism of functions as a whole, we glean from
+Figure~\ref{fig:L1funcounts} that a sizeable number of functions are
+polymorphic.  Functions are said to be polymorphic if at least one of their
+arguments or return are polymorphic.  \AT{Say more?}
+
+At present, we will only concern ourselves with \code{typeof} polymorphism,
+and we will later worry about classes and attributes.  To get an idea of
+what kinds of polymorphism occur, we will look at the most commonly
+occurring polymorphic argument signatures.  Table~\ref{tab:L0toppoly} has
+the top polymorphic signatures for L0.
 
@@
 
+\AT{TODO number of colocated primitive/other classes. Could use to suggest some optional class annotation.
+Maybe grab full class poly to see if there's something big? Without collapse to primitives.}
+
 The first thing to note is the low percentages overall: unlike with types in L0, L1, or L1+, there are no major contributors for class-based polymorphism.
@@
 Another significant source of this polymorphism is matrices, which can be seen in all the ``dim'' and ``dimnames'' attributes.
+
+\AT{TODO control for the common cases?}
 

### 2019-04-02 notebooks/paper_L123_results_section.Rmd
adjusted l0, looking at l1, matrices next, will update discussion in results to reflect new components of the TS

@@
 ```{r}
+# raw_NA -> NULL
+los <- lapply(los, function(lofun) lapply(lofun, make_NAs_into_NULLs))
+
 los <- change_type_systems_lopkg(los, type_map_T0_to_r, unparam=T)
+
+# L <: I <: D
+los <- lapply(los, function(lofun) lapply(lofun, logical_int_double_subtype_df))
 ```
@@
 ```{r}
-L1_c_basic_df <- build_arg_count_df_from_lopkgs(los)
+L0_c_basic_df <- build_arg_count_df_from_lopkgs(los)
 
-L1_c_f_basic_df <- build_fun_count_df_from_lopkgs(los)
+L0_c_f_basic_df <- build_fun_count_df_from_lopkgs(los)
 ```
@@
 ```{r}
-xtable(L1_c_basic_df) %>% print(include.rownames=F)
-xtable(L1_c_f_basic_df) %>% print(include.rownames=F)
+xtable(L0_c_basic_df) %>% print(include.rownames=F)
+xtable(L0_c_f_basic_df) %>% print(include.rownames=F)
 ```
@@
 ```{r}
-L1_poly_type_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_type_args))
-L1_poly_type_args <- L1_poly_type_args[order(-L1_poly_type_args$count),] # sort by count
-L1_poly_type_args$perc <- round(L1_poly_type_args$count/sum(L1_poly_type_args$count) * 100, 2)
+L0_poly_type_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_type_args))
+L0_poly_type_args <- L0_poly_type_args[order(-L0_poly_type_args$count),] # sort by count
+L0_poly_type_args$perc <- round(L0_poly_type_args$count/sum(L0_poly_type_args$count) * 100, 2)
 ```
@@
 ```{r}
-format_df_for_print(L1_poly_type_args[1:10,]) %>%
+format_df_for_print(L0_poly_type_args[1:10,]) %>%
 xtable %>% print(include.rownames=F)
@@
 
-TODO oops scalar vector thing.
-
 Write counts.
@@
 los <- lapply(los, function(lof) lapply(lof, class_to_primitive_df))
+
+los_a <- lapply(los, function(lofun) lapply(lofun, function(df) convert_df_attr_to_ts(df, type_map_T0_to_r)))
 ```
@@
 ```{r}
-L3_c_basic_df <- build_arg_count_df_from_lopkgs(los)
+L3_c_basic_df <- build_arg_count_df_from_lopkgs(los_a)
 
@@
 ```{r}
-num_poly_type_not_poly_class <- apply_count_to_lopkg_c(los, args_count_all_poly_type_mono_class)
+# num_poly_type_not_poly_class <- apply_count_to_lopkg_c(los, args_count_all_poly_type_mono_class)
 ```
@@
 ```{r}
-lo_pt_mc <- reduce_extracted_lopkgs_to_one_new(apply_extract_to_lopkg(los, extract_poly_type_args_w_mono_class))
-lo_pt_mc <- lo_pt_mc[order(-lo_pt_mc$count),] # sort by count
-lo_pt_mc$perc <- round(lo_pt_mc$count/sum(lo_pt_mc$count) * 100, 2)
-
-format_df_for_print_new(lo_pt_mc[1:10,]) -> put_me
-put_me
-xtable(put_me) %>% print(include.rownames=F)
+# lo_pt_mc <- reduce_extracted_lopkgs_to_one_new(apply_extract_to_lopkg(los, extract_poly_type_args_w_mono_class))
+# lo_pt_mc <- lo_pt_mc[order(-lo_pt_mc$count),] # sort by count
+# lo_pt_mc$perc <- round(lo_pt_mc$count/sum(lo_pt_mc$count) * 100, 2)
+# 
+# format_df_for_print_new(lo_pt_mc[1:10,]) -> put_me
+# put_me
+# xtable(put_me) %>% print(include.rownames=F)
 ```
@@
 ```{r}
-los_a <- lapply(los, function(lofun) lapply(lofun, function(df) convert_df_attr_to_ts(df, type_map_T0_to_r)))
-
 L3_poly_attr_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los_a, extract_poly_attr_args))

### 2019-04-02 paper/main.tex
adjusted l0, looking at l1, matrices next, will update discussion in results to reflect new components of the TS

@@
 
-Figure~\ref{fig:L1argcounts} shows the polymorphicity of argument positions
+Table~\ref{tab:L0argcounts} shows the polymorphicity of argument positions
  in R function.  From there, we build up a picture of the polymorphicity of
  functions as a whole, and the data for this function polymorphism can be
- found in Figure~\ref{fig:L1funcounts}.
-
-% latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Fri Mar 29 18:10:23 2019
-\begin{table}[ht]\label{fig:L1argcounts}\centering\begin{tabular}{lrr}  \hline
+ found in Figure~\ref{tab:L0funcounts}.
+
+\begin{table}[ht]\label{tab:L0argcounts}\centering\begin{tabular}{lrr}  \hline
 Type                   & Count & Percentage \\  \hline
-Full Monomorphic       & 784K & 85.76 \\
-  Monomorphic in Type & 845K & 92.46 \\
-  Monomorphic in Class & 827K & 90.41 \\
-  Monomorphic in Attribute Pattern & 825K & 90.27 \\
+%  Full Monomorphic & 785K & 85.87 \\ 
+  Monomorphic in Type & 866K & 94.76 \\ 
+%  Monomorphic in Class & 827K & 90.41 \\ 
+%  Monomorphic in Attribute Pattern & 825K & 90.27 \\ 
   Total Seen & 914K & --- \\ 
@@
 \end{tabular}
-\caption{typeof type system (L0) (arguments)}
+\caption{Argument position polymorphism in L0.}
 \end{table}
-% latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Fri Mar 29 18:10:23 2019
+
 \begin{table}[ht]
-\label{fig:L1funcounts}
+\label{tab:L0funcounts}
 \centering
@@
   \hline
-Full Monomorphic & 144K & 66.54 \\
-  Monomorphic in Type & 173K & 80.16 \\
-  Monomorphic in Class & 163K & 75.21 \\
-  Monomorphic in Attribute Pattern & 161K & 74.37 \\
+%  Full Monomorphic & 145K & 66.77 \\ 
+  Monomorphic in Type & 186K & 85.78 \\ 
+%  Monomorphic in Class & 163K & 75.21 \\
+%  Monomorphic in Attribute Pattern & 161K & 74.37 \\
   Total Seen & 216K & --- \\
@@
 \end{tabular}
-\caption{typeof type system (L0) (functions)}
+\caption{Function polymorphism in L0.}
 \end{table}
 
-From Figure~\ref{fig:L1argcounts}, we glean that argument positions aren't
+From Table~\ref{tab:L0argcounts}, we glean that argument positions aren't
 very polymorphic in L0.  An argument which is "Monomorphic in Type" is
@@
 As for the polymorphism of functions as a whole, we glean from
-Figure~\ref{fig:L1funcounts} that a sizeable number of functions are
+Figure~\ref{tab:L0funcounts} that a sizeable number of functions are
 polymorphic.  Functions are said to be polymorphic if at least one of their
@@
 
-% latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Fri Mar 29 18:13:17 2019
 \begin{table}[ht]\label{tab:L0toppoly}\centering
 \begin{tabular}{lrr}  \hline
-type & count & perc \\\hline\\[-3mm]
-  \D, \I & 19,799 & 28.73 \\[1mm]
-  \D, \l & 5,324 & 7.72 \\ [1mm]
-  \D, \sN & 5,152 & 7.48 \\ [1mm]
-  \C, \sN & 3,682 & 5.34 \\ [1mm]
-  \D, \L & 3,388 & 4.92 \\ [1mm]
-  \C, \D & 2,842 & 4.12 \\ [1mm]
-  \l, \sN & 2,799 & 4.06 \\ [1mm]
-  \C, \l & 1,836 & 2.66 \\ [1mm]
-  \C, \I & 1,663 & 2.41 \\ [1mm]
-  \C, \I, \D & 1,362 & 1.98 \\[1mm]   \hline
+type & count & perc \\ 
+\hline
+  \D, \sN & 8301 & 17.33 \\ 
+  \D, \l & 6291 & 13.13 \\ 
+  \C, \D & 4338 & 9.06 \\ 
+  \C, \sN & 4162 & 8.69 \\ 
+  \l, \sN & 2912 & 6.08 \\ 
+  \C, \l & 1836 & 3.83 \\ 
+  \C, \I & 1719 & 3.59 \\ 
+  \C, \D, \sN & 1057 & 2.21 \\ 
+  \D, error & 906 & 1.89 \\ 
+  \I, \sN & 797 & 1.66 \\   \hline
 \end{tabular}
@@
   \hline
-  Full Monomorphic & 781K & 85.37 \\
+%  Full Monomorphic & 781K & 85.37 \\
   Monomorphic in Type & 834K & 91.23 \\
@@
   \hline
-  Full Monomorphic & 143K & 65.88 \\
+%  Full Monomorphic & 143K & 65.88 \\
   Monomorphic in Type & 167K & 77.25 \\

### 2019-04-02 paper/main.tex
levels

--- 2019-04-02 paper/main.tex
@@
 \newcommand{\attr}[2]{\ensuremath{#1_{\mathtt{#2}}}\xspace}
+\newcommand{\attrclass}[3]{\ensuremath{#1^{\mathtt{#3}}_{\mathtt{#2}}}\xspace}
 \renewcommand{\to}{\ensuremath{\rightarrow}\xspace}
@@
 The logical ``baseline'' for the design of a type system for R is to adopt
-the language's dynamic types (in the sense of \code{typeof}).  We call this
-L0.  The types reflects the types of values in the R runtime environment, as
-the \code{typeof} an object is in direct correspondence with the runtime
-type tag of the value.
-
-Thus L0 includes the following types: vectors of primitive types (integers
+the language's builtin types.  Types in this system, which we call {\bf L0}
+reflect the types of values in the runtime environment, the \code{typeof} an
+object is in direct correspondence with the runtime type tag of the value.
+Thus we include the following types: vectors of primitive types (integers
 \I, double \D, character \C, logical \L, raw \R, complex \X), lists of any
 type (\l), and scalar values (environment \sE, closures \sF, symbols \sY, S4
-object \sS, and null \sN). We add the type \ANY to stand in for any type.
-
-We also include a notion of subtyping that follows from the conversion rules
-of the language.  All types are subtypes of \ANY, thus {\tt T}<:\ANY.
-Moreover, logicals are subtypes of integers, which are subtypes of doubles,
-\L <: \I <: \D.
+object \sS, and null \sN). We add one type stand in for unevaluated values,
+the type any (written \ANY). In {\bf L0} subtyping follows from the
+conversion rules of the language.  Logicals are subtypes of integers, which
+are subtypes of doubles, \L <: \I <: \D.  All types are subtypes of \ANY,
+thus {\tt T}<:\ANY for an type {\tt T}.
 
@@
 
-L0 is limited in expressivity. To address its limitations, we augment it
-scalars, parametrized lists.  In addition, we collect attribute and class
-information to distinguish a type for data frames and matrices, as the
-\code{typeof} these is simply {\tt list} in R.
-
-The second level enhances {\bf L0} with scalar values for primitive types
-(\sI, \sD, \sC, \sL, \sR, and \sX), type parametric lists (written \lT{T}
-for any type {\tt T}), data frames \attr{\l}{df}, and matrices
-\attr{\vec{\tt T}}{mat}.
+{\bf L0} is limited in expressivity as it supports polymoprhism for list in
+a trivial way, it does not address some of the popular constructed data
+types such as matrices and data frames, and cannot differentiate between
+scalar values and vectors.  {\bf L1} types build on and extend the previous
+level with with scalar values for primitive types (\sI, \sD, \sC, \sL, \sR,
+and \sX), type parametric lists (written \lT{T} for any type {\tt T}), data
+frames (written \attr{\l}{df} as they are built up from lists of type any),
+and matrices (\attr{\vec{\tt T}}{mat} where {\tt T} is a primitive type).
+Subtyping is extended such that {\tt T}<:{$\vec{\tt T}$} for any primitive
+type {\tt T}. There is no subtyping relation between lists of different
+types.
 
 \subsection{L2 Types}
- {\bf L2}: The third level adds attributes lists...
-
-While we do consider some class and attribute information when constructing
-our types, it is also valuable to consider them entirely separate from
-types.  Users are free to define their own classes, and arbitrarily add and
-remove attributes, and interesting patterns might arise in considering them.
-That said, not attribute patterns and classes are necessarily interesting,
-and indeed some of R's built-in data structures use them to some extend.
-For one, data frames have a {\tt data.frame} class (explicitly on the class
-attribute).  Other standard patterns are:
-
-\begin{itemize}
-\item data frames have three attributes: {\tt names} for column names, {\tt
-  row.names} for row names, and {\tt class} (which is instantiated to {\tt
-  "data.frame"} upon creation).
-\item matrices have at least a {\tt dim} attribute with matrix dimensions,
-  and optionally a {\tt dimnames} attribute for dimension names. Curiously,
-  the {\tt dim} attribute is all that is needed to define a matrix (i.e., a
-  list with a {\tt dim} attribute {\it is} a matrix).
-\item lists and vectors may have a {\tt names} attribute, which assigns
-  names to locations.
-\item to ease dealing with large sorted sets of non-numeric values, R offers
-  the \code{factor} and \code{levels} functions.  \code{factor(x,
-    levels=someOrder)} adds a {\tt "levels"} attribute to \code{x}, which
-  specifies an ordering for \code{x} according to the \code{sortedOrder}
-  parameter.  \code{factor} also changes the class of the parameter being
-  factored to \texttt{"factor"}.
-\end{itemize}
+
+{\bf L2} extends the previous system with more support for attributes.
+While there is a finite number of results from the {\tt typeof} function,
+this is not the case for \code{class} and \code{attributes}.  Classes and
+attributes are user-defined.  Of course, not all classes are user-defined,
+and R has a number of built-in or otherwise standard classes, ranging from
+truly primitive classes such as {\tt numeric} to the more complex {\tt
+  data.frame}.  We have already used the latter in {\bf L1}. We will ignore
+those classes and thus highlight user-defined ones.  As for attributes,
+again some patterns arise naturally, for instance all matrices have a {\tt
+  dims} attribute (indeed, it's this attribute that makes a matrix a
+matrix).  Attributes are name-value pairs, to keep things simple we reduce
+these to names. The only attribute that we treat specially is the {\tt
+  class} for those we will recurse and capture the set of values defined by
+the user (these are the class names for the object).  Thus, we extend {\bf
+  L1} with attribute lists and classes: \attrclass{T}{C}{N} where {\tt T}
+can be any L1 type other than data frame or matrix, {\tt C} and {\tt N}
+stands for a list names.  The subtype relation is extended with
+\attrclass{T}{C}{N}<:\attrclass{T}{C'}{D'} if {\tt C} $\subseteq$ {\tt C'}
+and {\tt N}$\subseteq${\tt N'}.
 
@@
 Type                   & Count & Percentage \\  \hline
-Full Monomorphic       & 784K & 85.76 \\
   Monomorphic in Type & 845K & 92.46 \\
-  Monomorphic in Class & 827K & 90.41 \\
-  Monomorphic in Attribute Pattern & 825K & 90.27 \\
   Total Seen & 914K & --- \\
@@
   \hline
-Full Monomorphic & 144K & 66.54 \\
   Monomorphic in Type & 173K & 80.16 \\
-  Monomorphic in Class & 163K & 75.21 \\
-  Monomorphic in Attribute Pattern & 161K & 74.37 \\
   Total Seen & 216K & --- \\
@@
   \hline
-  Full Monomorphic & 143K & 65.88 \\
   Monomorphic in Type & 167K & 77.25 \\
-%  Monomorphic in Class & 163K & 75.24 \\
-%  Monomorphic in Attribute Pattern & 161K & 74.37 \\
   Total Seen & 216K & --- \\
@@
 
-We see that roughly 9\% of all argument positions in L1 are polymorphic in type, as opposed to 7.5\% in L0.
-This small difference is perhaps surprising, as we did define a number of additional types and might expect for there to be a lot more polymorphism.
-This is explained by the fact that drawing the distinctions added a lot of polymorphism to already polymorphic arguments.
-\AT{We can get numbers on this.}
-
-As for functions, roughly 23\% are polymorphic in type, or put differently 23\% have at least one polymorphic argument or a polymorphic return.
-Again, the small difference between L0 and L1 in this regard is explained by a tendency to make already polymorphic functions more polymorphic.
+We see that roughly 9\% of all argument positions in L1 are polymorphic in
+type, as opposed to 7.5\% in L0.  This small difference is perhaps
+surprising, as we did define a number of additional types and might expect
+for there to be a lot more polymorphism.  This is explained by the fact that
+drawing the distinctions added a lot of polymorphism to already polymorphic
+arguments.  \AT{We can get numbers on this.}
+
+As for functions, roughly 23\% are polymorphic in type, or put differently
+23\% have at least one polymorphic argument or a polymorphic return.  Again,
+the small difference between L0 and L1 in this regard is explained by a
+tendency to make already polymorphic functions more polymorphic.
 
@@
 
-We would like to make one important note before digesting the data in Table~\ref{tab:toppolyL1}.
-We drew a distinction between scalars and vectors of the same type, and that distinction would be the single largest source of polymorphism in L1 had we not dealt with it appropriately.
-We note that an argument which exhibits a signature including e.g.
-\{\sD, \D\} really must be an argument designed to take vectors, as recall that unit length vectors appear as scalars by our definition.
-We have already collapsed this distinction.
+We would like to make one important note before digesting the data in
+Table~\ref{tab:toppolyL1}.  We drew a distinction between scalars and
+vectors of the same type, and that distinction would be the single largest
+source of polymorphism in L1 had we not dealt with it appropriately.  We
+note that an argument which exhibits a signature including e.g. \{\sD, \D\}
+really must be an argument designed to take vectors, as recall that unit
+length vectors appear as scalars by our definition.  We have already
+collapsed this distinction.
 
@@
 
-As we make the distinction between vectors and scalars, one other interesting question that arises is: {\it how many arguments are strictly scalar?}
-As it happens, over 40\% of monomorphic arguments with primitive vectorized types are always scalar (so, 40\% of {\it monomorphic} occurrences of {\tt scalar/X} and {\tt vector/X} are scalar).
-This shows that there is value in a scalar annotation, which could be used to communicate to an R runtime that vectorization of a value is not necessary, which may in turn lead to a performance improvement.
+As we make the distinction between vectors and scalars, one other
+interesting question that arises is: {\it how many arguments are strictly
+  scalar?}  As it happens, over 40\% of monomorphic arguments with primitive
+vectorized types are always scalar (so, 40\% of {\it monomorphic}
+occurrences of {\tt scalar/X} and {\tt vector/X} are scalar).  This shows
+that there is value in a scalar annotation, which could be used to
+communicate to an R runtime that vectorization of a value is not necessary,
+which may in turn lead to a performance improvement.
 
@@
   \hline
-  Full Monomorphic & 781K & 85.47 \\
   Monomorphic in Type & 869K & 95.01 \\
@@
   \hline
-  Full Monomorphic & 143K & 66.04 \\
   Monomorphic in Type & 185K & 85.30 \\
@@
 \subsection{L2}
-
-In L2, we aim to incorporate R's other reflective capabilities to address some of the more subtle instances of polymorphism that we have observed.
-
-First, we reiterate that there are a finite number of results from the {\tt typeof} function, and this is not the case for \code{class} and \code{attributes}.
-Classes and attributes can be arbitrarily defined by the user, and users create their own classes and assign their own attributes frequently.
-Unlike types (in the sense of \code{typeof}), the number of possible classes and attributes is arbitrary.
-
-Of course, not all classes are user-defined, and R has a number of built-in or otherwise standard classes, ranging from truly primitive classes such as ``numeric'' to the more complex ``data.frame''.
-We have already used these when building our more refined types, and so we transform each of R's standard classes into a ``primitive'' class, which will minimize their impact and thus highlight user-defined classes.
-
-As for attributes, again some patterns arise naturally, for instance all matrices have a ``dims'' attribute (indeed, it's this attribute that makes a matrix a matrix).
-\AT{This sounds like it's implying that we should do something about that, but I'm not sure if that's a good idea.}
-In principle, attributes are name, value pairs, but to keep things simple we will reduce these to name, type pairs, so the name of an attribute and the type that it was observed to have.
-We will show an L0 type for attributes.
 

### 2019-04-02 notebooks/paper_L123_results_section.Rmd

@@
 # L <: I <: D
-los <- lapply(los, function(lofun) lapply(lofun, logical_int_double_subtype_df))
+los <- lapply(los, function(lofun) lapply(lofun, function(df) logical_int_double_subtype_df(logical_int_double_subclass_df(df))))
+
+# TODO: error -> ?
+los <- lapply(los, function(lofun) lapply(lofun, map_error_to_any))
 ```
@@
 
+DEBUG: Classes
+```{r}
+L0_poly_class_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_class_args))
+L0_poly_class_args <- L0_poly_class_args[order(-L0_poly_class_args$count),] # sort by count
+L0_poly_class_args$perc <- round(L0_poly_class_args$count/sum(L0_poly_class_args$count) * 100, 1)
+```
+```
+
 Step 2.2: Top.
@@
 L0_poly_type_args <- L0_poly_type_args[order(-L0_poly_type_args$count),] # sort by count
-L0_poly_type_args$perc <- round(L0_poly_type_args$count/sum(L0_poly_type_args$count) * 100, 2)
+L0_poly_type_args$perc <- round(L0_poly_type_args$count/sum(L0_poly_type_args$count) * 100, 1)
 ```
@@
 ```{r}
-format_df_for_print(L0_poly_type_args[1:10,]) %>%
+format_df_for_print(L0_poly_type_args[1:20,]) %>%
+xtable %>% print(include.rownames=F)
+```
+
+Get the distribution of arguments in the monomorphic case.
+```{r}
+L0_mono_type_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_mono_type_args))
+L0_mono_type_args <- L0_mono_type_args[order(-L0_mono_type_args$count),] # sort by count
+L0_mono_type_args$perc <- round(L0_mono_type_args$count/sum(L0_mono_type_args$count) * 100, 1)
+
+format_df_for_print(L0_mono_type_args[1:20,]) %>%
 xtable %>% print(include.rownames=F)
@@
 ```{r}
-format_df_for_print(poly_type_args[1:10,]) -> put_me
+format_df_for_print(poly_type_args[1:20,]) -> put_me
 put_me
@@
 ```{r}
-format_df_for_print(L3_poly_class_args[1:10,]) -> put_me
+format_df_for_print(L3_poly_class_args[1:20,]) -> put_me
 put_me

### 2019-04-02 paper/main.tex

@@
 
-Figure~\ref{fig:L1argcounts} shows the polymorphicity of argument positions
+Table~\ref{tab:L0argcounts} shows the polymorphicity of argument positions
  in R function.  From there, we build up a picture of the polymorphicity of
  functions as a whole, and the data for this function polymorphism can be
- found in Figure~\ref{fig:L1funcounts}.
-
-% latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Fri Mar 29 18:10:23 2019
-\begin{table}[ht]\label{fig:L1argcounts}\centering\begin{tabular}{lrr}  \hline
+ found in Figure~\ref{tab:L0funcounts}.
+
+\begin{table}[ht]\label{tab:L0argcounts}\centering\begin{tabular}{lrr}  \hline
 Type                   & Count & Percentage \\  \hline
@@
 
-% latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Fri Mar 29 18:13:17 2019
 \begin{table}[ht]\label{tab:L0toppoly}\centering
 \begin{tabular}{lrr}  \hline
-type & count & perc \\\hline\\[-3mm]
-  \D, \I & 19,799 & 28.73 \\[1mm]
-  \D, \l & 5,324 & 7.72 \\ [1mm]
-  \D, \sN & 5,152 & 7.48 \\ [1mm]
-  \C, \sN & 3,682 & 5.34 \\ [1mm]
-  \D, \L & 3,388 & 4.92 \\ [1mm]
-  \C, \D & 2,842 & 4.12 \\ [1mm]
-  \l, \sN & 2,799 & 4.06 \\ [1mm]
-  \C, \l & 1,836 & 2.66 \\ [1mm]
-  \C, \I & 1,663 & 2.41 \\ [1mm]
-  \C, \I, \D & 1,362 & 1.98 \\[1mm]   \hline
+type & count & perc \\
+\hline
+  \D, \sN & 8301 & 17.33 \\
+  \D, \l & 6291 & 13.13 \\
+  \C, \D & 4338 & 9.06 \\
+  \C, \sN & 4162 & 8.69 \\
+  \l, \sN & 2912 & 6.08 \\
+  \C, \l & 1836 & 3.83 \\
+  \C, \I & 1719 & 3.59 \\
+  \C, \D, \sN & 1057 & 2.21 \\
+  \D, error & 906 & 1.89 \\
+  \I, \sN & 797 & 1.66 \\   \hline
 \end{tabular}

### 2019-04-02 paper/main.tex
a bit on L2

--- 2019-04-02 paper/main.tex
@@
 
-L0 is limited in expressivity. To address its limitations, we augment it
-scalars, parametrized lists.  In addition, we collect attribute and class
+L0 is limited in expressivity. To address its limitations, we augment it with
+scalars and parametrized lists.  In addition, we collect attribute and class
 information to distinguish a type for data frames and matrices, as the
@@
 \subsection{L2 Types}
- {\bf L2}: The third level adds attributes lists...
-
-While we do consider some class and attribute information when constructing
-our types, it is also valuable to consider them entirely separate from
-types.  Users are free to define their own classes, and arbitrarily add and
-remove attributes, and interesting patterns might arise in considering them.
-That said, not attribute patterns and classes are necessarily interesting,
-and indeed some of R's built-in data structures use them to some extend.
-For one, data frames have a {\tt data.frame} class (explicitly on the class
-attribute).  Other standard patterns are:
+
+Though more expressive than L0, L1 only incorporates a limited amount of
+class and attribute information. For instance, L1 says nothing about user-defined
+classes and non-standard attributes. The standard attribute patterns in R are:
 
@@
   the {\tt dim} attribute is all that is needed to define a matrix (i.e., a
-  list with a {\tt dim} attribute {\it is} a matrix).
+  vector with a {\tt dim} attribute {\it is} a matrix).
 \item lists and vectors may have a {\tt names} attribute, which assigns
@@
 
+All of the possible \code{typeof}s in correspond to a class, so we must control
+for these as well.
+
+Now, recall that the attributes of a value is a list of name, value pairs of metadata
+on the value accessible through the \code{attributes} function. For our purposes,
+we will only consider the names of attributes, and not values nor their types. This
+avoids expensive recursion, and including type information in attribute patterns 
+creates a lot of noise in the data that hides other interesting patterns.
+
 \section{Analysis Results}\label{sec:results}
@@
 
-At L0, the type system is aligned with the types of the core language.  We
-have observed over 173K monomorphic functions; this accounts for 80\% of our
-corpus.  In terms of argument positions, 92\% of arguments are monomorphic.
+At L0, the type system is mostly aligned with the types of the core language.  We
+have observed over 186K monomorphic functions; this accounts for 86\% of our
+corpus.  In terms of argument positions, 95\% of arguments are monomorphic.
 
@@
  functions as a whole, and the data for this function polymorphism can be
- found in Figure~\ref{tab:L0funcounts}.
+ found in Table~\ref{tab:L0funcounts}.
 
@@
 %  Full Monomorphic & 785K & 85.87 \\ 
-  Monomorphic in Type & 866K & 94.76 \\ 
+  Monomorphic in Type & 870K & 95.2\% \\ 
 %  Monomorphic in Class & 827K & 90.41 \\ 
@@
 %  Full Monomorphic & 145K & 66.77 \\ 
-  Monomorphic in Type & 186K & 85.78 \\ 
+  Monomorphic in Type & 187K & 86.4\% \\ 
 %  Monomorphic in Class & 163K & 75.21 \\
@@
 
-From Table~\ref{tab:L0argcounts}, we glean that argument positions aren't
-very polymorphic in L0.  An argument which is "Monomorphic in Type" is
-monomorphic w.r.t. \code{typeof}, in that it was only ever observed to have
-one \code{typeof}.  Similarly, an argument is "Monomorphic in Class" if it
-is only ever called with one {\tt class}.  As we said earlier, attributes
-are a little more complicated, and an argument which is "Monomorphic in
-Attribute Pattern" was only ever observed to have one attribute pattern (and
-that pattern may itself contain many attributes).
-
-As for the polymorphism of functions as a whole, we glean from
-Figure~\ref{tab:L0funcounts} that a sizeable number of functions are
-polymorphic.  Functions are said to be polymorphic if at least one of their
-arguments or return are polymorphic.  \AT{Say more?}
-
-At present, we will only concern ourselves with \code{typeof} polymorphism,
-and we will later worry about classes and attributes.  To get an idea of
-what kinds of polymorphism occur, we will look at the most commonly
-occurring polymorphic argument signatures.  Table~\ref{tab:L0toppoly} has
-the top polymorphic signatures for L0.
+Tables~\ref{tab:L0argcounts} and~\ref{tab:L0funcounts} show that the overall amount of polymorphism in R is quite low, but that doesn't quite tell the whole story.
+To get an idea of what kinds of polymorphism occur, we will look at the most commonly occurring polymorphic argument signatures.  
+Table~\ref{tab:L0toppoly} has the top polymorphic signatures for L0.
 
@@
 
-Some entries in the Table~\ref{tab:L0toppoly} are clear (e.g., double,
-integer polymorphism is unsurprising), others are strange (e.g., character,
-double), but in fact they all have some explanation.
-
-\begin{enumerate}
-\item \{\D, \I\}: in R, doubles and integers are effectively
-  interchangeable, as casting from one to the other is straightforward and
-  automatically performed when appropriate.  For example, \code{c(1, 2)[1.9]
-    == 1}.  This distinction is mainly just relevant to the implementation,
-  as integers and doubles are stored and dealt with differently in the
-  runtime but programmers only see the distinction when printing.
-
-\item \{\D, \l\} and \{\C, \l\}: at first glance, these type signatures are
-  suspect.  Why would so many programmers want to pass a double or a list to
-  some function?  The key insight here is that doubles and characters are
-  {\it vectorized} in R, meaning that a double is really a vector of
-  doubles.  Thus, these seemingly strange signatures are revealed to be
-  entirely reasonable: these signatures are likely to describe arguments
-  which are either vectors or lists.
-
-\item \{\ANY,\sN\}: in most languages, NULL is an inhabitant of all types,
-  and we see that this is not the case in R.  That said, there is a pattern
-  of many functions having default or optional arguments, and in many of
-  these cases NULL is the default value.
-
-\item \{\D,\L\}: the first values of type logical that come to mind are TRUE
-  and FALSE, which does little to explain this strange signature.  Indeed,
-  the key here is an implementation detail of R's: recall that NAs inhabit
-  every type.  As it happens, the default type of NA is logical (e.g.,
-  \code{typeof(NA) == "logical"}), and this is likely what is being seen
-  here.  We will control for this by distinguishing NAs with their own type,
-  {\tt raw\_NA}, but we will get to that in time.
-
-\item \{\C,\D\}, \{\C,\I\} and \{character, double, integer\}: yet more
-  strange signatures.  These signatures are a little more subtle to explain.
-  For one, indexing lists and vectors can be done with either strings (i.e.,
-  to access a named index) or with numbers.  Another option is dates: R has
-  many ways to present dates, and strings and numbers feature among them.
-  Finally, many \AT{get numbers} functions in R return a string if an error
-  occurs during execution.
-
-\end{enumerate}
-
-How will we deal with this?
-
-\begin{enumerate}
-\item consolidating doubles and integers into a single real-number type
-  would account for all of this polymorphism.
-\item to get an idea of how often vectors and lists co-occur in signatures,
-  we need to expand our analysis of vectorized types.  Notably, when we
-  encounter e.g. a double, we should collect length information to see if
-  the double is a vector of doubles or just a scalar.  This will feature in
-  our next type system, L1, which we will get to momentarily.
-\item if we roll NULLs into every type, we can cleanly deal with these X,
-  NULL signatures.  As we have established that an annotation for an
-  optional argument is useful, this will allow us to reveal other
-  interesting patterns.
-\item similarly with NULLs, we're interested in rolling NAs in to all types.
-  As we mentioned, this is actually what the R runtime does, but a default
-  type needed to be chosen for NAs appearing without context.  Subsuming NAs
-  should help to reveal other interesting patterns.
-\item dealing with this polymorphism will be more tricky.  We might need to
-  pull class information in (to pick up on dates), and possibly make a
-  signature judgment depending on which argument position has the signature
-  (e.g. a \{character, X\} signature on a return).  We will come back to
-  this later.
-
-\end{enumerate}
+\AT{TODO explain results in table.}
+
+%Some entries in the Table~\ref{tab:L0toppoly} are clear (e.g., double,
+%integer polymorphism is unsurprising), others are strange (e.g., character,
+%double), but in fact they all have some explanation.
+
+%\begin{enumerate}
+%\item \{\D, \I\}: in R, doubles and integers are effectively
+%  interchangeable, as casting from one to the other is straightforward and
+%  automatically performed when appropriate.  For example, \code{c(1, 2)[1.9]
+%    == 1}.  This distinction is mainly just relevant to the implementation,
+%  as integers and doubles are stored and dealt with differently in the
+%  runtime but programmers only see the distinction when printing.
+%
+%\item \{\D, \l\} and \{\C, \l\}: at first glance, these type signatures are
+%  suspect.  Why would so many programmers want to pass a double or a list to
+%  some function?  The key insight here is that doubles and characters are
+%  {\it vectorized} in R, meaning that a double is really a vector of
+%  doubles.  Thus, these seemingly strange signatures are revealed to be
+%  entirely reasonable: these signatures are likely to describe arguments
+%  which are either vectors or lists.
+%
+%\item \{\ANY,\sN\}: in most languages, NULL is an inhabitant of all types,
+%  and we see that this is not the case in R.  That said, there is a pattern
+%  of many functions having default or optional arguments, and in many of
+%  these cases NULL is the default value.
+%
+%\item \{\D,\L\}: the first values of type logical that come to mind are TRUE
+%  and FALSE, which does little to explain this strange signature.  Indeed,
+%  the key here is an implementation detail of R's: recall that NAs inhabit
+%  every type.  As it happens, the default type of NA is logical (e.g.,
+%  \code{typeof(NA) == "logical"}), and this is likely what is being seen
+%  here.  We will control for this by distinguishing NAs with their own type,
+%  {\tt raw\_NA}, but we will get to that in time.
+%
+%\item \{\C,\D\}, \{\C,\I\} and \{character, double, integer\}: yet more
+%  strange signatures.  These signatures are a little more subtle to explain.
+%  For one, indexing lists and vectors can be done with either strings (i.e.,
+%  to access a named index) or with numbers.  Another option is dates: R has
+%  many ways to present dates, and strings and numbers feature among them.
+%  Finally, many \AT{get numbers} functions in R return a string if an error
+%  occurs during execution.
+%
+%\end{enumerate}
+%
+%How will we deal with this?
+%
+%\begin{enumerate}
+%\item consolidating doubles and integers into a single real-number type
+%  would account for all of this polymorphism.
+%\item to get an idea of how often vectors and lists co-occur in signatures,
+%  we need to expand our analysis of vectorized types.  Notably, when we
+%  encounter e.g. a double, we should collect length information to see if
+%  the double is a vector of doubles or just a scalar.  This will feature in
+%  our next type system, L1, which we will get to momentarily.
+%\item if we roll NULLs into every type, we can cleanly deal with these X,
+%  NULL signatures.  As we have established that an annotation for an
+%  optional argument is useful, this will allow us to reveal other
+%  interesting patterns.
+%\item similarly with NULLs, we're interested in rolling NAs in to all types.
+%  As we mentioned, this is actually what the R runtime does, but a default
+%  type needed to be chosen for NAs appearing without context.  Subsuming NAs
+%  should help to reveal other interesting patterns.
+%\item dealing with this polymorphism will be more tricky.  We might need to
+%  pull class information in (to pick up on dates), and possibly make a
+%  signature judgment depending on which argument position has the signature
+%  (e.g. a \{character, X\} signature on a return).  We will come back to
+%  this later.
+%
+%\end{enumerate}
 
 There are some clear limitations to the L0 type system that we should
-address before working through any of these other issues As we mentioned,
+address before working through any of these other issues. As we mentioned,
 \code{typeof} makes no distinction between vectors and scalars, as indeed
@@
 such a distinction, and indeed an all-new runtime environment design for R
-might benefit from e.g. not needing to vectorize scalar values.  This is one
-of the many modifications we make, all of which are thoroughly explored in
+might benefit from e.g. not needing to vectorize scalar values.  This is only one
+of the modifications we make, all of which are thoroughly explored in
 the following section.
@@
 \subsection{L1}
+
+As we mentioned in \AT{an earlier section}, and as was revealed in the previous section,
+L0 is somewhat limited in its expressivity. To address this, we develop L1, which espouses
+a finer-grained approach to types. Major modifications from L0 include the differentiation of
+scalars and vectors, the parameterization of list types over element types, and the use of
+attribute and class information to distinguish types for data frames and matrices.
 
@@
 y & \D & numeric & \{\} \\
-retv & list<double> & list & \{\xspace example: \sC\}
+retv & \lT{D} & list & \{\xspace example: \sC\}
 \end{tabular}

### 2019-04-02 paper/main.tex

@@
 
-L0 is limited in expressivity. To address its limitations, we augment it with
-scalars and parametrized lists.  In addition, we collect attribute and class
-information to distinguish a type for data frames and matrices, as the
-\code{typeof} these is simply {\tt list} in R.
-
-The second level enhances {\bf L0} with scalar values for primitive types
-(\sI, \sD, \sC, \sL, \sR, and \sX), type parametric lists (written \lT{T}
-for any type {\tt T}), data frames \attr{\l}{df}, and matrices
-\attr{\vec{\tt T}}{mat}.
+{\bf L0} is limited in expressivity as it supports polymoprhism for list in
+a trivial way, it does not address some of the popular constructed data
+types such as matrices and data frames, and cannot differentiate between
+scalar values and vectors.  {\bf L1} types build on and extend the previous
+level with with scalar values for primitive types (\sI, \sD, \sC, \sL, \sR,
+and \sX), type parametric lists (written \lT{T} for any type {\tt T}), data
+frames (written \attr{\l}{df} as they are built up from lists of type any),
+and matrices (\attr{\vec{\tt T}}{mat} where {\tt T} is a primitive type).
+Subtyping is extended such that {\tt T}<:{$\vec{\tt T}$} for any primitive
+type {\tt T}. There is no subtyping relation between lists of different
+types.
 
@@
 
-Though more expressive than L0, L1 only incorporates a limited amount of
-class and attribute information. For instance, L1 says nothing about user-defined
-classes and non-standard attributes. The standard attribute patterns in R are:
-
-\begin{itemize}
-\item data frames have three attributes: {\tt names} for column names, {\tt
-  row.names} for row names, and {\tt class} (which is instantiated to {\tt
-  "data.frame"} upon creation).
-\item matrices have at least a {\tt dim} attribute with matrix dimensions,
-  and optionally a {\tt dimnames} attribute for dimension names. Curiously,
-  the {\tt dim} attribute is all that is needed to define a matrix (i.e., a
-  vector with a {\tt dim} attribute {\it is} a matrix).
-\item lists and vectors may have a {\tt names} attribute, which assigns
-  names to locations.
-\item to ease dealing with large sorted sets of non-numeric values, R offers
-  the \code{factor} and \code{levels} functions.  \code{factor(x,
-    levels=someOrder)} adds a {\tt "levels"} attribute to \code{x}, which
-  specifies an ordering for \code{x} according to the \code{sortedOrder}
-  parameter.  \code{factor} also changes the class of the parameter being
-  factored to \texttt{"factor"}.
-\end{itemize}
+{\bf L2} extends the previous system with more support for attributes.
+While there is a finite number of results from the {\tt typeof} function,
+this is not the case for \code{class} and \code{attributes}.  Classes and
+attributes are user-defined.  Of course, not all classes are user-defined,
+and R has a number of built-in or otherwise standard classes, ranging from
+truly primitive classes such as {\tt numeric} to the more complex {\tt
+  data.frame}.  We have already used the latter in {\bf L1}. We will ignore
+those classes and thus highlight user-defined ones.  As for attributes,
+again some patterns arise naturally, for instance all matrices have a {\tt
+  dims} attribute (indeed, it's this attribute that makes a matrix a
+matrix).  Attributes are name-value pairs, to keep things simple we reduce
+these to names. The only attribute that we treat specially is the {\tt
+  class} for those we will recurse and capture the set of values defined by
+the user (these are the class names for the object).  Thus, we extend {\bf
+  L1} with attribute lists and classes: \attrclass{T}{C}{N} where {\tt T}
+can be any L1 type other than data frame or matrix, {\tt C} and {\tt N}
+stands for a list names.  The subtype relation is extended with
+\attrclass{T}{C}{N}<:\attrclass{T}{C'}{D'} if {\tt C} $\subseteq$ {\tt C'}
+and {\tt N}$\subseteq${\tt N'}.
 
@@
 Type                   & Count & Percentage \\  \hline
-%  Full Monomorphic & 785K & 85.87 \\ 
-  Monomorphic in Type & 870K & 95.2\% \\ 
-%  Monomorphic in Class & 827K & 90.41 \\ 
-%  Monomorphic in Attribute Pattern & 825K & 90.27 \\ 
+  Monomorphic in Type & 845K & 92.46 \\
   Total Seen & 914K & --- \\
@@
   \hline
-%  Full Monomorphic & 143K & 65.88 \\
   Monomorphic in Type & 167K & 77.25 \\
-%  Monomorphic in Class & 163K & 75.24 \\
-%  Monomorphic in Attribute Pattern & 161K & 74.37 \\
   Total Seen & 216K & --- \\

### 2019-04-02 paper/main.tex
collision

@@
 
-All of the possible \code{typeof}s in correspond to a class, so we must control
-for these as well.
-
-Now, recall that the attributes of a value is a list of name, value pairs of metadata
-on the value accessible through the \code{attributes} function. For our purposes,
-we will only consider the names of attributes, and not values nor their types. This
-avoids expensive recursion, and including type information in attribute patterns
-creates a lot of noise in the data that hides other interesting patterns.
-
 \section{Analysis Results}\label{sec:results}

### 2019-04-03 notebooks/jan.Rmd
minor

--- 2019-03-28 notebooks/paper_results_section.Rmd
@@
 ```{r, echo=FALSE}
-los <- list.files("../data/full_sigs", full.names=TRUE)
-names <- list.files("../data/full_sigs")
-names(los) <- substr(names, 1, nchar(names)-4)
-los <- lapply(los, readRDS)
-los <- remove_empties(los)
-# annoying strange error that I should fix pops up
+df <- list.files("../data/full_sigs", full.names=TRUE)
 
-# There's at least one transformation that we're always gonna want to apply, namly the vector/scalar
-# collapse where both are present in a signature.
-los <- lapply(los, function(lofuns) {lapply(lofuns, combine_scalar_vector_where_appropriate)})
-# Figure out which ones are matrices. Remove if the analysis picks this up (and we rerun).
-los <- lapply(los, function(lofun) lapply(lofun, distinguish_matrices_df))
 ```
-
-Get the counts.
-```{r}
-# argument counts
-c_basic_df <- build_arg_count_df_from_lopkgs(los)
-
-xtable(c_basic_df) %>% print(include.rownames=F)
-```
-
-```{r}
-# function counts
-c_f_basic_df <- build_fun_count_df_from_lopkgs(los)
-
-xtable(c_f_basic_df) %>% print(include.rownames=F)
-```
-
-Top signatures, with printing.
-This is how you get the top sigs.
-```{r}
-poly_type_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_type_args))
-poly_type_args <- poly_type_args[order(-poly_type_args$count),] # sort by count
-poly_type_args$perc <- round(poly_type_args$count/sum(poly_type_args$count) * 100, 2)
-```
-
-Print top 10.
-```{r}
-format_df_for_print(poly_type_args[1:10,]) -> put_me
-put_me
-xtable(put_me) %>% print(include.rownames=F)
-```
-
-Now, having noted that:
-(1) ints and doubles are basically interchangable, and
-(2) null and NA should be a part of every type,
-while also noting that together these make up a huge portion of the polymorphism, we transform.
-```{r}
-los <- lapply(los, function(s) lapply(s, fold_NULL_and_NA_into_other_types_df))
-null_na_fold_arg_counts <- build_arg_count_df_from_lopkgs(los)
-null_na_fold_fun_counts <- build_fun_count_df_from_lopkgs(los)
-
-los <- lapply(los, function(lof) {lapply(lof, function(df) {change_type_systems_df(df, type_map_sv_to_real)})})
-null_na_fold_real_arg_counts <- build_arg_count_df_from_lopkgs(los)
-null_na_fold_real_fun_counts <- build_fun_count_df_from_lopkgs(los)
-```
-
-Generate...
-```{r}
-xtable(null_na_fold_real_arg_counts) %>% print(include.rownames=F)
-xtable(null_na_fold_real_fun_counts) %>% print(include.rownames=F)
-```
-
-Top sig time.
-```{r}
-poly_type_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_type_args))
-poly_type_args <- poly_type_args[order(-poly_type_args$count),] # sort by count
-poly_type_args$perc <- round(poly_type_args$count/sum(poly_type_args$count) * 100, 2)
-```
-
-Print top 10.
-```{r}
-format_df_for_print(poly_type_args[1:10,]) -> put_me
-put_me
-xtable(put_me) %>% print(include.rownames=F)
-```
-
-Get the biggest signature.
-```{r}
-big <- get_biggest_function(los, metric=size_of_signature_type)
-
-# it's %>%
-```

### 2019-04-03 notebooks/data_builder.Rmd
L1 translate code and merging

---
title: "Data Builder"
output: html_notebook
---

```{r}
source("../code/data_processing.R")
```

```{r}
# load the data
psp <- read_rds("../data/per_sig_partial.RDS")

# get rid of empties
# they are due to packages having no data, this may get resolved as the prl3 run finishes
psp <- psp[!psp$pkg == "no_data",]

# do this otherwise there are lists
psp <- lists_to_chars_sig_df(psp)
```

```{r}
# first, turn specials, closures, and builtins to functions
psp <- transform_new_df_w_TS_map(psp, type_map_to_fun)
```

```{r}
# make it T0
psp_L0 <- transform_new_df_w_TS_map(psp, type_map_T0_to_r)
```

```{r}
# write it to file for Java processing
write_csv(psp_L0, "../data/partial_L0.csv")
```

```{r}
# read in new one
# DONT FORGET TO GET RID OF }"" lol
psp_L0 <- read_csv("../data/partial_L0_subtype.csv")

# remove SUBTYPE rows
psp_L0 <- psp_L0[!psp_L0$pkg == "SUBTYPE",]
```

```{r}
# write L1 (which is just psp) to file for Java
write_csv(psp, "../data/partial_L1.csv")

# go to terminal, gzip it for Java
# DONT FORGET TO GET RID OF }"" lol
```

```{r}
# read it back
psp <- read_csv("../data/partial_L1_subtype.csv")

# remove SUBTYPE rows
psp <- psp[!psp$pkg == "SUBTYPE",]
```

### 2019-04-04 notebooks/data_builder.Rmd
minimized type names, added support for L1 (see Main.java for flag)

@@
 # read in new one
-# DONT FORGET TO GET RID OF }"" lol
 psp_L0 <- read_csv("../data/partial_L0_subtype.csv")
@@
 # go to terminal, gzip it for Java
-# DONT FORGET TO GET RID OF }"" lol
 ```

### 2019-04-04 paper/main.tex
new data and some code
csvs with counts, and fix weird error in paper

@@
 
+%
+%
 \subsection{L1 Types}
@@
 \end{lstlisting}}\caption{Example of character/double argument (from {\tt BoolNet} package).}\label{fig:chardbl}\end{figure}%$
-
 In Figure~\ref{fig:matvec}, we see that the function can take in a vector,
@@
   in code at end of last sentence}
-
 \begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
@@
 \end{lstlisting}}\caption{Example of matrix/(\D) argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}
-
 %
@@
 \section{Related Work}
-
 \AT{Not sure what the best place for this is, but we should discuss these papers somewhere.}
-
 The idea of measuring polymorphism via some program analysis isn't new.
 For instance, some work~\cite{aakerblom2015measuring} ... .
-
 More generally, there is an existing literature on analyzing usage patterns
@@
 analyzed in some work~\cite{callau2011howdevelopers}, ...
-
 Other analysis work~\cite{milojkovic2017duck} ... .
-
 \section{Conclusions and Future Work}
-
 \bibliographystyle{boilerplate/ACM-Reference-Format}

### 2019-04-04 notebooks/data_builder.Rmd
new pipeline is happening, dealing with the data. final numbers and adjusted discussion will have to be done tomorrow

@@
 ```
+
+```{r}
+# collapse
+# make sure the right function is uncommented in Java
+
+psp_col <- read_csv("../data/partial_L1_collapsed.csv.gz")
+psp_L0_col <- read_csv("../data/partial_L0_collapsed.csv.gz")
+```
+
+```{r}
+# arg counts
+# again, Java
+L0_arg_sig_c <- read_csv("../data/partial_L0_arg_sig_counts.csv")
+L1_arg_sig_c <- read_csv("../data/partial_L1_arg_sig_counts.csv")
+
+# sort
+L0_arg_sig_c[order(-L0_arg_sig_c$count),] -> L0_arg_sig_c
+L1_arg_sig_c[order(-L1_arg_sig_c$count),] -> L1_arg_sig_c
+
+# rm []
+L0_arg_sig_c[1,"count"] <- 0
+L1_arg_sig_c[1,"count"] <- 0
+
+# re sort
+L0_arg_sig_c[order(-L0_arg_sig_c$count),] -> L0_arg_sig_c
+L1_arg_sig_c[order(-L1_arg_sig_c$count),] -> L1_arg_sig_c
+
+# percentages
+L0_arg_sig_c$perc_tot <- round(L0_arg_sig_c$count / sum(L0_arg_sig_c$count) * 100, 2)
+L1_arg_sig_c$perc_tot <- round(L1_arg_sig_c$count / sum(L1_arg_sig_c$count) * 100, 2)
+
+# collapse the obvious signatures
+L0_arg_sig_c <- new_df_count_collape_arg_sigs(L0_arg_sig_c, "L0")
+L1_arg_sig_c <- new_df_count_collape_arg_sigs(L1_arg_sig_c, "L1")
+
+# poly arg counts only
+L0_poly_sig <- L0_arg_sig_c[L0_arg_sig_c$arg_sig %>% strsplit(split=",") %>% lapply(function(x) length(x) != 1) %>% unlist, ]
+L1_poly_sig <- L1_arg_sig_c[L1_arg_sig_c$arg_sig %>% strsplit(split=",") %>% lapply(function(x) length(x) != 1) %>% unlist, ]
+
+# display
+L0_poly_sig[1:10,] %>% ready_print_count_fun %>% xtable %>% print(include.rownames=F)
+L1_poly_sig[1:10,] %>% ready_print_count_fun %>% xtable %>% print(include.rownames=F)
+```
+
+```{r}
+# all pckg (w/ dups)
+pf_L0 <- psp_L0 %>% mutate(pkgfun = paste(pkg, fun, sep="::")) %>% select(pkgfun)
+pf_L1 <- psp %>% mutate(pkgfun = paste(pkg, fun, sep="::")) %>% select(pkgfun)
+
+# tot fun counts
+pf_L0 %>% unique %>% nrow -> tot_L0_funs
+pf_L1 %>% unique %>% nrow -> tot_L1_funs
+
+# poly fun counts
+pf_L0[duplicated(pf_L0),] %>% unique -> poly_L0_funs
+pf_L1[duplicated(pf_L1),] %>% unique -> poly_L1_funs
+
+poly_L0_funs_c <- poly_L0_funs %>% nrow
+poly_L1_funs_c <- poly_L1_funs %>% nrow
+
+# mono fun count (sanity check)
+setdiff(pf_L0 %>% unique, poly_L0_funs) %>% nrow -> mono_L0_funs_c
+setdiff(pf_L1 %>% unique, poly_L1_funs) %>% nrow -> mono_L1_funs_c
+
+data.frame(`Signature Type`=c("Monomorphic", "Total Seen"), Count=paste0(c(mono_L0_funs_c/1000, tot_L0_funs/1000) %>% as.integer, "K"), Percentage=c(paste0(round(mono_L0_funs_c/tot_L0_funs * 100, 1), "%"), "---")) %>% xtable %>% print(include.rownames=F)
+```
+
+```{r}
+# arg poly count
+
+# L0
+data.frame(`Signature Type`=c("Monomorphic", "Total Seen"), Count=c(L0_arg_sig_c$count %>% sum - L0_poly_sig$count %>% sum, L0_arg_sig_c$count %>% sum), Percentage=c(paste0(round(100 * (1 - (L0_poly_sig$count %>% sum / L0_arg_sig_c$count %>% sum)), 1), "%"), "---")) %>% xtable %>% print(include.rownames=F)
+
+```

### 2019-04-04 paper/main.tex
new pipeline is happening, dealing with the data. final numbers and adjusted discussion will have to be done tomorrow

@@
 
-\subsection{L0 polymorphism}
+\AT{NOTE: I am updating this with new numbers as we get them.
+The fixed tracer is still running, I should be able to fill in final numbers by tomorrow night.}
+
+\subsection{L0}
 
@@
 have observed over 186K monomorphic functions; this accounts for 86\% of our
-corpus.  In terms of argument positions, 95\% of arguments are monomorphic.
-
-Table~\ref{tab:L0argcounts} shows the polymorphicity of argument positions
- in R function.  From there, we build up a picture of the polymorphicity of
- functions as a whole, and the data for this function polymorphism can be
- found in Table~\ref{tab:L0funcounts}.
-
-\begin{table}[ht]\label{tab:L0argcounts}\centering\begin{tabular}{lrr}  \hline
-Type                   & Count & Percentage \\  \hline
-  Monomorphic in Type & 845K & 92.46 \\
-  Total Seen & 914K & --- \\
-   \hline
-\end{tabular}
-\caption{Argument position polymorphism in L0.}
+corpus.
+
+Across all of the arguments that we've observed, the vast majority are monomorphic.
+Table~\ref{tab:L0top10arg} shows the breakdown of the 10 most frequently occurring argument signatures.
+\begin{table}[ht]
+\label{tab:L0top10arg}
+\centering
+\begin{tabular}{lrr}
+  \hline
+ Argument Signature & Count & Percentage \\ 
+  \hline
+  \D & 69806 & 26.3\% \\ 
+  \l\ & 55311 & 20.8\% \\ 
+  \C & 27789 & 10.5\% \\ 
+  \L & 26230 & 9.9\% \\ 
+  \sF & 16806 & 6.3\% \\ 
+  \sN & 13976 & 5.3\% \\ 
+  unevaled & 9125 & 3.4\% \\ 
+  \I & 8971 & 3.4\% \\ 
+  \sS & 4748 & 1.8\% \\ 
+  \D, \I & 3256 & 1.2\% \\ 
+   \hline
+\end{tabular}
+\caption{Most frequent argument signatures.}
 \end{table}
+
+We see that doubles are the most common argument type, which is not entirely surprising as doubles are the backbone of any numeric data.
+Next is lists, which seems reasonable given that lists are often used to store the aforementioned data.
+Recall that in L0, data frames will appear as lists.
+As for ``unevaled'', first recall the discussion that all function arguments are wrapped in promises in R from \AT{where?}.
+The ``unevaled'' signature appears whenever an argument was provided but unused, and one might wonder why such a signature might be so common.
+As it happens, many R functions metaprogram on some of their arguments, and this metaprogramming does not force the argument promise, so it will appear as ``unevaled''.
+The magrittr package's pipe \%$>$\% function is an example of this.
+\AT{Move magrittr discussion up.}
+
+One of our chief goals is to generate useful function signatures.
+After tracing and capturing unique signatures from function call sites, \AT{recall} that we collapse some of the more obvious ones together.
+After this, we are curious to know: {\it how many functions have more than one signature?}
+Table~\ref{tab:L0funcounts} has the answer.
 
@@
 \centering
-\begin{tabular}{lrr}
-  \hline
-Type & Count & Percentage \\
-  \hline
-%  Full Monomorphic & 145K & 66.77 \\
-  Monomorphic in Type & 187K & 86.4\% \\
-%  Monomorphic in Class & 163K & 75.21 \\
-%  Monomorphic in Attribute Pattern & 161K & 74.37 \\
-  Total Seen & 216K & --- \\
+\begin{tabular}{llr}
+  \hline
+Signature Type & Count & Percentage \\ 
+  \hline
+Monomorphic & 46K & 83\% \\ 
+  Total Seen & 56K & --- \\ 
    \hline
@@
 \caption{Function polymorphism in L0.}
+\end{table}
+
+Here, we see that the vast majority of functions are monomorphic in L0.
+To choosing how to express the polymorphicity of a function, there are two main options one might consider:
+First, having several separate whole-function signatures, e.g.
+\D $\rightarrow$ \D and \C $\rightarrow$ \D, or alternatively expressing the polymorphism at the argument level with a union type, e.g.
+(\D $\cup$ \C) $\rightarrow$ \C.
+Earlier in Table~\ref{tab:L0top10arg} we saw the most frequently occurring argument signatures.
+To get an idea of how expressing polymorphism at the argument level would look, we ask: {\it what are the most frequently occurring polymorphic argument signatures?}
+Table~\ref{tab:L0toppoly} has the answer.
+
+\begin{table}[ht]\label{tab:L0toppoly}\centering
+\begin{tabular}{lrr}  \hline
+Signature & Count & \% of All Signatures \\
+\hline
+  \D, \l & 2799 & 1\% \\ 
+  \D, \sN & 2290 & 0.9\% \\ 
+  \C, \sN & 1527 & 0.6\% \\ 
+  \l, \sN & 1193 & 0.4\% \\ 
+  \C, \D & 858 & 0.3\% \\ 
+  \D, \sF & 413 & 0.2\% \\ 
+  \C, \I & 382 & 0.1\% \\ 
+  \I, \sN & 378 & 0.1\% \\ 
+  \C, \l & 371 & 0.1\% \\ 
+  \C, \sF & 334 & 0.1\% \\ 
+\end{tabular}
+\caption{Top polymorphic argument signatures for the L0 type system.}
+\end{table}
+
+Let's unpack some of the signatures in Table~\ref{tab:L0toppoly}:
+
+\begin{itemize}
+
+	\item \D, \l (and \C, \l): L0 doesn't parameterize list types, and it's very likely that this signature describes vectors of doubles and lists of doubles.
+	Coercing vectors to lists is trivial in R, and indeed many of the functions with such an argument signature do.
+	The story is similar with vectors of characters and lists of characters.
+
+	\item $\square$, \sN: there are a few instances of this ``null'' polymorphism.
+	The frequency of these signatures indicates a tendency to create ``optional arguments'', where some argument's default value might be NULL.
+	This pattern is quite frequent in R code, \AT{and there are a number of examples sprinkled throughout the paper.}
+
+	\item \C, \D and \C, \I: these are interesting patterns.
+	In R, lists can always be indexed with numerics (\L, \D, \I) {\it and also} characters.
+	When a list is indexed by character, the character is matched against the element names in the list and the corresponding element is selected.
+	If there is no match, the lookup returns NULL.
+	Another possible explanation for these signatures are dates, which can be represented by numerics or characters and the value usually has one of the date classes.
+	L0 doesn't deal with classes, which are one of the topics of L2.
+
+	\item \D, \sF: this signature appears more strange than it is.
+	Some arguments with this signature are either parameters for a function, or the function itself already constructed.
+	Another examples, such as the {\tt br} argument of the {\tt rtree} function in the {\tt ape} package, are parameters for a complex data generation process, or a function to be used during the generation.
+	It would appear that this pattern is used to make complex simulation, modelling, or data generation functions more flexible.
+
+	\item \C, \sF: another puzzling signature, though its explanation is surprisingly straightforward:
+	In R, programmers are free to look functions up {\it by name}, and indeed the frequently encountered {\tt UseMethod} function for dynamic class-based function dispatch is one such example.
+	In short, this signature usually represents a function or a function name.
+
+\end{itemize}
+
+Now, we have some understanding of argument polymorphism in L0.
+The next question we address is: {\it just how pervasive is argument polymorphism?}.
+Refer to Table~\ref{tab:L0argcounts}.
+
+\begin{table}[ht]\label{tab:L0argcounts}\centering
+\begin{tabular}{lrl}
+  \hline
+Signature.Type & Count & Percentage \\ 
+  \hline
+Monomorphic & 250K & 94\% \\ 
+  Total Seen & 266K & --- \\ 
+   \hline
+\end{tabular}
+\caption{Argument position polymorphism in L0.}
 \end{table}
@@
 Table~\ref{tab:L0toppoly} has the top polymorphic signatures for L0.
-
-\begin{table}[ht]\label{tab:L0toppoly}\centering
-\begin{tabular}{lrr}  \hline
-type & count & perc \\
-\hline
-  \D, \sN & 8301 & 17.33 \\
-  \D, \l & 6291 & 13.13 \\
-  \C, \D & 4338 & 9.06 \\
-  \C, \sN & 4162 & 8.69 \\
-  \l, \sN & 2912 & 6.08 \\
-  \C, \l & 1836 & 3.83 \\
-  \C, \I & 1719 & 3.59 \\
-  \C, \D, \sN & 1057 & 2.21 \\
-  \D, error & 906 & 1.89 \\
-  \I, \sN & 797 & 1.66 \\   \hline
-\end{tabular}
-\caption{Top polymorphic signatures for the L0 type system.}
-\end{table}
 

### 2019-04-04 notebooks/data_builder.Rmd
prose in results up to parity, working out L2 now. will also get plots and data for function-level polymorphism

@@
 
-# display
+# display all args 
+L0_arg_sig_c[1:10,] %>% ready_print_count_fun %>% xtable %>% print(include.rownames=F)
+L1_arg_sig_c[1:10,] %>% ready_print_count_fun %>% xtable %>% print(include.rownames=F)
+
+# display poly
 L0_poly_sig[1:10,] %>% ready_print_count_fun %>% xtable %>% print(include.rownames=F)
@@
 data.frame(`Signature Type`=c("Monomorphic", "Total Seen"), Count=paste0(c(mono_L0_funs_c/1000, tot_L0_funs/1000) %>% as.integer, "K"), Percentage=c(paste0(round(mono_L0_funs_c/tot_L0_funs * 100, 1), "%"), "---")) %>% xtable %>% print(include.rownames=F)
+
+data.frame(`Signature Type`=c("Monomorphic", "Total Seen"), Count=paste0(c(mono_L1_funs_c/1000, tot_L1_funs/1000) %>% as.integer, "K"), Percentage=c(paste0(round(mono_L1_funs_c/tot_L1_funs * 100, 1), "%"), "---")) %>% xtable %>% print(include.rownames=F)
 ```
@@
 
+# L1
+data.frame(`Signature Type`=c("Monomorphic", "Total Seen"), Count=c(L1_arg_sig_c$count %>% sum - L1_poly_sig$count %>% sum, L1_arg_sig_c$count %>% sum), Percentage=c(paste0(round(100 * (1 - (L1_poly_sig$count %>% sum / L1_arg_sig_c$count %>% sum)), 1), "%"), "---")) %>% xtable %>% print(include.rownames=F)
+
 ```
+
+```{r}
+prim_scal <- c("sD", "sL", "sC", "sX", "sI", "sR")
+# compute scalars 
+L1_sc_c <- L1_arg_sig_c[unlist(lapply(L1_arg_sig_c$arg_sig, function(s) nchar(s) == 2 && substr(s, 1, 1) == "s")), ]
+L1_sc_c <- L1_sc_c[L1_sc_c$arg_sig %in% prim_scal,]
+num_scal <- sum(L1_sc_c$count)
+
+percent_scalar_mono <- round( 100 * num_scal / sum(L1_arg_sig_c$count), 1)
+```

### 2019-04-04 paper/main.tex
prose in results up to parity, working out L2 now. will also get plots and data for function-level polymorphism

@@
 \newcommand{\NUMPOLYTYPEMONOCLASS}{1,357\xspace}
+\newcommand{\PERCSCALARMONO}{37.6\%\xspace}
 
@@
 \newcommand{\ANY}{\ensuremath{\small{\mathtt ?}}\xspace}     % Any
-\newcommand{\lT}[1]{\ensuremath{\small\underline{\mathtt #1}}\xspace}     % Any
+\newcommand{\lT}[1]{\ensuremath{\small\underline{\mathtt #1}}\xspace}     % list<T>
+\newcommand{\M}[1]{\ensuremath{\attr{\vec{\tt #1}}{mat}}\xspace}     % matrix
+\newcommand{\df}{\ensuremath{\attr{\l}{df}}\xspace}     % data.frame
 
@@
 \subsection{L0 Types}
+\label{sec:L0def}
 
@@
 \subsection{L1 Types}
-
-{\bf L0} is limited in expressivity as it supports polymoprhism for list in
+\label{sec:L1def}
+
+{\bf L0} is limited in expressivity as it supports polymorphism for list in
 a trivial way, it does not address some of the popular constructed data
@@
 \subsection{L2 Types}
+\label{sec:L2def}
 
@@
 \end{tabular}
-\caption{Most frequent argument signatures.}
+\caption{Most frequent argument signatures in L0.}
 \end{table}
@@
 	Another examples, such as the {\tt br} argument of the {\tt rtree} function in the {\tt ape} package, are parameters for a complex data generation process, or a function to be used during the generation.
+	\AT{Do we want to put the fun def and arg documentation here?}
 	It would appear that this pattern is used to make complex simulation, modelling, or data generation functions more flexible.
@@
 Now, we have some understanding of argument polymorphism in L0.
-The next question we address is: {\it just how pervasive is argument polymorphism?}.
+The next question we address is: {\it just how pervasive is this argument polymorphism?}
 Refer to Table~\ref{tab:L0argcounts}.
@@
   \hline
-Signature.Type & Count & Percentage \\ 
+Signature Type & Count & Percentage \\ 
   \hline
@@
 
-Tables~\ref{tab:L0argcounts} and~\ref{tab:L0funcounts} show that the overall amount of polymorphism in R is quite low, but that doesn't quite tell the whole story.
-To get an idea of what kinds of polymorphism occur, we will look at the most commonly occurring polymorphic argument signatures.
-Table~\ref{tab:L0toppoly} has the top polymorphic signatures for L0.
-
-\AT{TODO explain results in table.}
-
-%Some entries in the Table~\ref{tab:L0toppoly} are clear (e.g., double,
-%integer polymorphism is unsurprising), others are strange (e.g., character,
-%double), but in fact they all have some explanation.
-
-%\begin{enumerate}
-%\item \{\D, \I\}: in R, doubles and integers are effectively
-%  interchangeable, as casting from one to the other is straightforward and
-%  automatically performed when appropriate.  For example, \code{c(1, 2)[1.9]
-%    == 1}.  This distinction is mainly just relevant to the implementation,
-%  as integers and doubles are stored and dealt with differently in the
-%  runtime but programmers only see the distinction when printing.
-%
-%\item \{\D, \l\} and \{\C, \l\}: at first glance, these type signatures are
-%  suspect.  Why would so many programmers want to pass a double or a list to
-%  some function?  The key insight here is that doubles and characters are
-%  {\it vectorized} in R, meaning that a double is really a vector of
-%  doubles.  Thus, these seemingly strange signatures are revealed to be
-%  entirely reasonable: these signatures are likely to describe arguments
-%  which are either vectors or lists.
-%
-%\item \{\ANY,\sN\}: in most languages, NULL is an inhabitant of all types,
-%  and we see that this is not the case in R.  That said, there is a pattern
-%  of many functions having default or optional arguments, and in many of
-%  these cases NULL is the default value.
-%
-%\item \{\D,\L\}: the first values of type logical that come to mind are TRUE
-%  and FALSE, which does little to explain this strange signature.  Indeed,
-%  the key here is an implementation detail of R's: recall that NAs inhabit
-%  every type.  As it happens, the default type of NA is logical (e.g.,
-%  \code{typeof(NA) == "logical"}), and this is likely what is being seen
-%  here.  We will control for this by distinguishing NAs with their own type,
-%  {\tt raw\_NA}, but we will get to that in time.
-%
-%\item \{\C,\D\}, \{\C,\I\} and \{character, double, integer\}: yet more
-%  strange signatures.  These signatures are a little more subtle to explain.
-%  For one, indexing lists and vectors can be done with either strings (i.e.,
-%  to access a named index) or with numbers.  Another option is dates: R has
-%  many ways to present dates, and strings and numbers feature among them.
-%  Finally, many \AT{get numbers} functions in R return a string if an error
-%  occurs during execution.
-%
-%\end{enumerate}
-%
-%How will we deal with this?
-%
-%\begin{enumerate}
-%\item consolidating doubles and integers into a single real-number type
-%  would account for all of this polymorphism.
-%\item to get an idea of how often vectors and lists co-occur in signatures,
-%  we need to expand our analysis of vectorized types.  Notably, when we
-%  encounter e.g. a double, we should collect length information to see if
-%  the double is a vector of doubles or just a scalar.  This will feature in
-%  our next type system, L1, which we will get to momentarily.
-%\item if we roll NULLs into every type, we can cleanly deal with these X,
-%  NULL signatures.  As we have established that an annotation for an
-%  optional argument is useful, this will allow us to reveal other
-%  interesting patterns.
-%\item similarly with NULLs, we're interested in rolling NAs in to all types.
-%  As we mentioned, this is actually what the R runtime does, but a default
-%  type needed to be chosen for NAs appearing without context.  Subsuming NAs
-%  should help to reveal other interesting patterns.
-%\item dealing with this polymorphism will be more tricky.  We might need to
-%  pull class information in (to pick up on dates), and possibly make a
-%  signature judgment depending on which argument position has the signature
-%  (e.g. a \{character, X\} signature on a return).  We will come back to
-%  this later.
-%
-%\end{enumerate}
-
-There are some clear limitations to the L0 type system that we should
-address before working through any of these other issues. As we mentioned,
-\code{typeof} makes no distinction between vectors and scalars, as indeed
-there is no such distinction in R.  But as we saw, we might benefit from
-such a distinction, and indeed an all-new runtime environment design for R
-might benefit from e.g. not needing to vectorize scalar values.  This is only one
-of the modifications we make, all of which are thoroughly explored in
-the following section.
+From Table~\ref{tab:L0argcounts} we see that, with our definition of subtyping on L0 and other simplifications we made, an overwhelming amount of argument polymorphism is accounted for.
+We also saw in Table~\ref{tab:L0toppoly} that each of the most frequently occurring argument signatures had some explanation.
+That said, L0 is not without limitations.
+
+The main issues with L0 stem from a lack of adequate information, as the type system doesn't capture precise list types, and does not distinguish scalars from vectors.
+The next type system we will explore, L1, tackles these issues by collecting more information to build more precise types.
+While the L1 types are less ``in-tune'' with current R implementations, one can imagine designing an all-new R runtime which may benefit from the more specific information in L1, e.g.
+by not needing to vectorize scalar values, or designing a static analysis which uses precise list types to give programmers better information in an IDE.
+L1 is discussed next.
+
+\AT{TODO do function signatures maybe? Maybe just an example that we can run with.
+Plots also.}
+
 %
@@
 
-As we mentioned in \AT{an earlier section}, and as was revealed in the previous section,
-L0 is somewhat limited in its expressivity. To address this, we develop L1, which espouses
-a finer-grained approach to types. Major modifications from L0 include the differentiation of
-scalars and vectors, the parameterization of list types over element types, and the use of
-attribute and class information to distinguish types for data frames and matrices.
+As we mentioned in Section~\ref{sec:L1def}, L0 is somewhat limited in its expressivity. 
+To address this, we develop L1, which espouses a finer-grained approach to types. 
+Major modifications from L0 include:
+
+\begin{enumerate}
+	\item the differentiation of scalars and vectors;
+	\item the parameterization of list types over element types;
+	\item the use of attribute and class information to distinguish types for data frames and matrices, and;
+	\item subtyping of \sN (i.e., $\forall T$, \sN $<:$ $T$).
+\end{enumerate}
 
@@
 \end{lstlisting}}
-
-\begin{tabular}{@{}r|l|l|l@{}}\hline
-\bf Argument & \bf Type & \bf Class &\bf Attributes \\
-x & \sD & numeric & \{\} \\
-y & \D & numeric & \{\} \\
-retv & \lT{D} & list & \{\xspace example: \sC\}
+\caption{Call signature in L1: x: \sD, y: \D $\rightarrow$ \lT{D}.}\label{fig:exL1}\end{figure}
+
+Let us begin as we did in L0 by looking at the breakdown of argument signatures in L1.
+Consult Table~\ref{tab:L1top10arg}.
+\begin{table}[ht]
+\label{tab:L1top10arg}
+\centering
+\begin{tabular}{lrr}
+  \hline
+ Argument Signature & Count & Percentage \\ 
+  \hline
+  \sD & 48815 & 18.4\% \\ 
+  \D & 27367 & 10.3\% \\ 
+  \sL & 24284 & 9.2\% \\ 
+  \sC & 22025 & 8.3\% \\ 
+  \l & 20062 & 7.6\% \\ 
+  \sF & 18480 & 7\% \\ 
+  \M{D} & 14315 & 5.4\% \\ 
+  \sN & 14016 & 5.3\% \\ 
+  unevaled & 9093 & 3.4\% \\ 
+  \C & 7731 & 2.9\% \\
+   \hline
 \end{tabular}
-\caption{Example call and L1 signature.}\label{fig:exL1}\end{figure}
-
-Let us begin as we did in L0 by looking at the amount of polymorphism in L1.
-The data can be found in Tables~\ref{tab:argcountsL1}~and~\ref{tab:funcountsL1}.
-\AT{TODO: have these tables be generated automatically}
-
-% latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Fri Mar 22 18:03:30 2019
+\caption{Most frequent argument signatures in L1.}
+\end{table}
+
+From Table~\ref{tab:L1top10arg}, we see that all of the most frequent argument signatures are monomorphic even though we increased the precision of the types.
+We see that scalars actually figure quite prominently, which is encouraging, supporting our suggestion of an explicit scalar annotation.
+We also see that matrices (\M{D}) are quite popular, a testament to the versatility of the data structure, as they can easily be converted to vectors and data frames.
+
+Now that we have a picture of how arguments look in L1, let's see: {\it how many L1 functions have more than one type signature?}
+The answer can be found in Table~\ref{tab:funcountsL1}.
+
+\begin{table}[ht]
+\label{tab:funcountsL1}
+\centering
+\begin{tabular}{lll}
+  \hline
+Signature Type & Count & Percentage \\ 
+  \hline
+Monomorphic & 44K & 79.6\% \\ 
+  Total Seen & 56K & --- \\ 
+   \hline
+\end{tabular}
+\caption{Account of {\it function} polymorphism in L1.}
+\end{table}
+
+We see that we increased the polymorphism with our transition to L1, despite the monomorphic nature of the most frequently appearing argument signatures.
+As we were in L0, we are curious to know how argument-level polymorphism looks in L1.
+Refer to Table~\ref{tab:toppolyL1}.
+
+\begin{table}[ht]
+\label{tab:toppolyL1}
+\centering
+\begin{tabular}{lrl}
+  \hline
+Signature & Count & \% of All Signatures \\ 
+  \hline
+  \D, \M{D} & 1440 & 0.5\% \\ 
+  \l, \lT{D} & 525 & 0.2\% \\ 
+  \M{D}, \M{I} & 383 & 0.1\% \\ 
+  \sC, \sD & 382 & 0.1\% \\ 
+  \I, \sD & 370 & 0.1\% \\ 
+  \l, \lT{list} & 337 & 0.1\% \\ 
+  \df, \M{D} & 305 & 0.1\% \\ 
+  \C, \D & 294 & 0.1\% \\ 
+  \sD, \sF & 290 & 0.1\% \\ 
+  \sC, \sF & 273 & 0.1\% \\
+   \hline
+\end{tabular}
+\caption{Top polymorphic argument signatures in L1.}
+\end{table}
+
+So what do these signatures mean?
+
+\begin{itemize}
+	\item \D, \M{D}: vectors are just one-dimensional matrices.
+	Indeed, the only thing separating a vector from a matrix in R is a {\tt dims} attribute.
+
+	\item \l, \lT{D}: this hearkens back to L0, where we discussed how this very same signature was explained by the list being a list of doubles.
+	So why is it still here?
+	Well, first note that there are far less of these signatures in L1 than in L0, so clearly we captured {\it something}.
+	Next, note the method with which we parameterize lists: we ascribe some type $T$ to a list type \lT{T} if and only if {\it all} list elements have {\it exactly} type {\it T}---in our analysis, we have encountered lists with integers as well as doubles, and unfortunately those currently appear as \l.
+	This could be addressed by doing some subtyping computation while exploring list elements.
+
+	\item \M{D}, \M{I}: this signature is reasonable, as any computation in R on \M{D} would work on \M{I} and vice versa.
+
+	\item \sC, \sD and \C, \D: we discussed this in L0---this signature likely captures an index or a date.
+
+	\item \I, \sD: subtyping between primitives does not hold between scalars and vectors (i.e., \I $\not<:$ \sD) even though it's likely that the programmer intended for the argument to be a vector of the supertype (here, these arguments were likely intended to be of type \D).
+
+	\item \l, \lT{list}: here we have a list of any and a list of lists.
+	One explanation for this is that lists of lists sometimes contain NULL as an element, and this would show up with type \l, again as we don't do any subtyping while computing precise list types.
+	Another explanation is the way R handles object-orientation, which we discuss shortly in Section~\ref{sec:S3S4R5}.
+
+	\item \df, \M{D}: in R, it's easy to transform a matrix into a data frame, and this argument is likely a result of that:
+	Simply calling \code{as.data.frame} on a matrix is enough.
+
+	\item \sD, \sF and \sC, \sF: these were thoroughly discussed in the context of L0.
+
+\end{itemize}
+
+With an understanding of argument polymorphism in L1, we wonder: {\it how many argument positions are polymorphic in L1?}
+Refer to Table~\ref{tab:argcountsL1}.
+
 \begin{table}[ht]
@@
 \centering
-\begin{tabular}{lrr}
-  \hline
- Argument Polymorphism & Count & Percentage \\
-  \hline
-%  Full Monomorphic & 781K & 85.37 \\
-  Monomorphic in Type & 834K & 91.23 \\
-%  Monomorphic in Class & 827K & 90.42 \\
-%  Monomorphic in Attribute Pattern & 825K & 90.27 \\
-  Total Seen & 914K & --- \\
+\begin{tabular}{lrl}
+  \hline
+Signature Type & Count & Percentage \\ 
+  \hline
+  Monomorphic & 252K & 94.9\% \\ 
+  Total Seen & 265K & --- \\ 
    \hline
 \end{tabular}
-\caption{Account of {\it argument} polymorphism in L1.}
+\caption{Argument position polymorphism in L1.}
 \end{table}
 
-% latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Fri Mar 22 18:10:46 2019
-\begin{table}[ht]
-\label{tab:funcountsL1}
-\centering
-\begin{tabular}{lrr}
-  \hline
- Function Polymorphism & Count & Percentage \\
-  \hline
-  Monomorphic in Type & 167K & 77.25 \\
-  Total Seen & 216K & --- \\
-   \hline
-\end{tabular}
-\caption{Account of {\it function} polymorphism in L1.}
-\end{table}
-
-\AT{We should figure out the pipeline to get the notebook output into here automatically.}
-
-We see that roughly 9\% of all argument positions in L1 are polymorphic in
-type, as opposed to 7.5\% in L0.  This small difference is perhaps
-surprising, as we did define a number of additional types and might expect
-for there to be a lot more polymorphism.  This is explained by the fact that
-drawing the distinctions added a lot of polymorphism to already polymorphic
-arguments.  \AT{We can get numbers on this.}
-
-As for functions, roughly 23\% are polymorphic in type, or put differently
-23\% have at least one polymorphic argument or a polymorphic return.  Again,
-the small difference between L0 and L1 in this regard is explained by a
-tendency to make already polymorphic functions more polymorphic.
-
-As before, these numbers alone are insufficient, so we ask again: {\it of the polymorphic arguments, what are the most common patterns?}
-The answer to that question can be found in Table~\ref{tab:toppolyL1}.
-
-% latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Fri Mar 22 21:27:37 2019
-\begin{table}[ht]
-\label{tab:toppolyL1}
-\centering
-\begin{tabular}{lrr}
-  \hline
- Polymorphic Type Signature & \# Occurrences & \% of Total Polymorphism \\
-  \hline
-  \D, \I & 7742 & 9.66 \\
-  \sD, \sI & 7201 & 8.99 \\
-  matrix, \D & 5384 & 6.72 \\
-  NULL, \sC & 2325 & 2.90 \\
-  list$<$any$>$, list$<$double$>$ & 2231 & 2.78 \\
-  data.frame, matrix & 2116 & 2.64 \\
-  NULL, \sD & 2082 & 2.60 \\
-  NULL, \D & 2069 & 2.58 \\
-  list$<$any$>$, list$<$list$>$ & 1432 & 1.79 \\
-  NULL, \C & 1380 & 1.72 \\
-     \hline
-\end{tabular}
-\caption{Top polymorphic argument signatures in L1.}
-\end{table}
-
-We would like to make one important note before digesting the data in
-Table~\ref{tab:toppolyL1}.  We drew a distinction between scalars and
-vectors of the same type, and that distinction would be the single largest
-source of polymorphism in L1 had we not dealt with it appropriately.  We
-note that an argument which exhibits a signature including e.g. \{\sD, \D\}
-really must be an argument designed to take vectors, as recall that unit
-length vectors appear as scalars by our definition.  We have already
-collapsed this distinction.
-
-In Table~\ref{tab:toppolyL1}, we again see the prominence of {\tt double} and {\tt integer} polymorphism.
-As discussed previously, we will address this by creating a new {\it real} type encompassing both.
-Refer back to the snippet in Figure~\ref{fig:realex} for details.
-Also, another clear pattern in Table~\ref{tab:toppolyL1} is that of NULL, X.
-As we discussed, we will roll NULL (and raw NAs) into other types.
+In spite of drastically increasing the precision of our types, we are still happily only dealing with a small amount of polymorphism.
+This suggests that with a good notion of subtyping, precise type annotations won't cause an undue increase in the polymorphism of the language.
+
+\AT{Something more about functions. Plots...}
+
+Before presenting L2, we will digress slightly and explore some of the ramifications of our transition from L0 to L1.
 
@@
 \subsubsection{Scalars}
-
-As we make the distinction between vectors and scalars, one other
-interesting question that arises is: {\it how many arguments are strictly
-  scalar?}  As it happens, over 40\% of monomorphic arguments with primitive
-vectorized types are always scalar (so, 40\% of {\it monomorphic}
-occurrences of {\tt scalar/X} and {\tt vector/X} are scalar).  This shows
-that there is value in a scalar annotation, which could be used to
-communicate to an R runtime that vectorization of a value is not necessary,
-which may in turn lead to a performance improvement.
-
-At this point, we have seen a number of reasonable polymorphic patterns that arise in R, both in L0 and L1.
-The next type system design we consider will deal with these, and we refer to it as L$1^{+}$.
-The number of monomorphic arguments and functions can be found in Tables~\ref{tab:argcountsL1p} and~\ref{tab:funcountsL1p} respectively.
-
-% latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Fri Mar 22 18:03:30 2019
-\begin{table}[ht]
-\label{tab:argcountsL1p}
-\centering
-\begin{tabular}{lrr}
-  \hline
- Argument Polymorphism & Count & Percentage \\
-  \hline
-  Monomorphic in Type & 869K & 95.01 \\
-%  Monomorphic in Class & 831K & 90.31 \\
-%  Monomorphic in Attribute Pattern & 830K & 90.25 \\
-  Total Seen & 914K & --- \\
-     \hline
-\end{tabular}
-\caption{Account of {\it argument} polymorphism in L$1^{+}$.}
-\end{table}
-
-% latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Fri Mar 22 18:10:46 2019
-\begin{table}[ht]
-\label{tab:funcountsL1p}
-\centering
-\begin{tabular}{lrr}
-  \hline
- Function Polymorphism & Count & Percentage \\
-  \hline
-  Monomorphic in Type & 185K & 85.30 \\
-%  Monomorphic in Class & 163K & 75.13 \\
-%  Monomorphic in Attribute Pattern & 161K & 74.31 \\
-  Total Seen & 216K & --- \\
-   \hline
-\end{tabular}
-\caption{Account of {\it function} polymorphism in L$1^{+}$.}
-\end{table}
-
-We had some big wins here, with only ~5\% of argument polymorphism unaccounted for.
-To get an idea of what more we can do, let's see what the most frequent remaining polymorphic signatures are.
-See Table~\ref{tab:toppolyL1p}.
-
-% latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Tue Mar 26 17:46:45 2019
-\begin{table}[ht]
-\label{tab:toppolyL1p}
-\centering
-\begin{tabular}{lrr}
-  \hline
-Polymorphic Type Signature & \# Occurrences & \% of Total Polymorphism \\
-  \hline
-  matrix, \realv & 6360 & 13.94 \\
-  list$<$any$>$, list$<$real$>$ & 2799 & 6.13 \\
-  \C, \realv & 2774 & 6.08 \\
-  \sC, \reals & 2552 & 5.59 \\
-  data.frame, matrix & 2167 & 4.75 \\
-  list$<$any$>$, list$<$list$>$ & 1496 & 3.28 \\
-  list$<$any$>$, list$<$character$>$ & 1214 & 2.66 \\
-  data.frame, \realv & 1046 & 2.29 \\
-  \sL, \reals & 920 & 2.02 \\
-  \sC, \realv & 868 & 1.90 \\
-   \hline
-\end{tabular}
-\caption{Top 10 argument signatures in L$1^{+}$.}
-\end{table}
-
-A few of the top polymorphic signatures stand out.
-First, the matrix/\realv signature seems interesting, converting between vectors and matrices is very simple, and indeed numeric vectors are mathematically {\tt 1 x n} matrices.
-In fact, in R matrices can be used identically to vectors in many cases, and we refer the reader to Figure~\ref{fig:matasvec} for examples.
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-v <- c(1, 2, 3, 4)
-ma <- matrix(data=v, nrow=4, ncol=1)
-v[2] == m[2] == m[2, 1]
-
-mb <- matrix(data=v, nrow=2, ncol=2)
-v[3] == m[3] == m[1, 2]
-\end{lstlisting}}\caption{Using a matrix like a vector.}\label{fig:matasvec}\end{figure}
-
-This similarity is reminiscent of another pattern which was more clear in Table~\ref{tab:L0toppoly}; that of list/vector polymorphism.
-Transforming between vectors and matrices is just as easy as transforming between lists and vectors is simple.
-\AT{We should figure out how we want to deal with this polymorphism.}
-
-Another pattern that we can shed some light on the list$<$any$>$/list$<$X$>$ pattern.
-These data points are an artifact of R's three object systems: S3, S4, and R5.
+\label{sec:scalars}
+
+Our suggestion to add an explicit scalar type is not unfounded, though it is not compatible with current R implementations.
+We saw in this section that strictly scalar values appear quite frequently, making up some of the most frequently occurring argument signatures.
+In fact, \PERCSCALARMONO of arguments have a strictly scalar monomorphic signature.
+This suggests that such a distinction would more accurately capture current use of the language, and would also be beneficial to future implementations which can perform optimizations based on the type (e.g., not vectorizing scalar-only values).
+
+%
+%
+\subsubsection{Object-Oriented Programming in R}
+\label{sec:S3S4R5}
+
+There is something more to be said about the \l type, namely that it is sometimes linked to (one of) R's takes on object-oriented programming.
+R has three object systems: S3, S4, and R5, and S3 objects will often manifest themselves with this signatures in L1.
 An S3 object is any value with a ``class'' attribute, and indeed many of these tend to be lists.
-For example, an S3-style point might be a list with an entry for each dimension in the point, with a ``class'' attribute set to ``Point''.
-Our next type system, L2, will account for these.
-
-Another significant source of apparently strange polymorphism is the character, real polymorphism.
-Recall Figure~\ref{fig:chardbl}, which laid out an example of this polymorphic signature: lists and vectors can be accessed by strings as well as by numerics.
-This is a somewhat unusual example of polymorphism, and our argument depends on there being perhaps a list or vector passed alongside the {\tt character}- or {\tt real}-typed argument.
-So, just how often is this the case?
-Well, as it happens, character/real polymorphism coincides with some list-flavored type in \INDEXCOINCIDENCE instances (\INDEXYPERC of occurrences).
-
-Another possible explanation for character, real polymorphism is the date data type.
-In R, dates are often represented as either strings or numbers, and many packages make use of dates.
+For example, an S3-style ``Point'' might be a list with an entry for each dimension in the point, with a ``class'' attribute set to ``Point''.
+When these S3 objects have contents with a variety of types, their L1 type will appear as \l, which leaves something to be desired:
+Our next type system, L2, will remedy this.
+
+%
+%
+\subsubsection{\C, \I, \D Polymorphism}
+Another significant source of apparently strange polymorphism is the character, integer, double polymorphism.
+Figure~\ref{fig:chardbl} lays out an example of this polymorphic signature, showing that lists and vectors can be accessed by strings as well as by numerics.
+One could conceive of an ``index'' type to capture this, but there are certain limitations to such a type, as such an ``index'' cannot be incremented as it could be a string.
+An even more sophisticated dependent type system could capture this, but it's far from clear if such a sophisticated type system would be for a language like R.
+
+One other possible explanation for this polymorphism is the ``date'' data type.
+In R, dates are often represented as either strings or numbers, and they are widely used.
 Dates (sometimes) manifest themselves in the class of a value, for examples refer to Figure~\ref{fig:date}.
@@
 
-This figure reveals an unfortunate truth that date data type in R is not standardized.
-That said, as before with matrices and vectors, and vectors and lists, translating between the data types is straightforward.
-Once again, this polymorphism can be captured with classes, which is the focus of our next discussion.
-
-We have seen that once we have done away with trivial sources of polymorphism, a number of interesting patterns arise.
-In many of these cases, we can appeal to other sources of reflection in R to expand our understanding.
-For example, to deal with the possibility of dates, we can appeal to the \code{class} of values, and to gain more information on matrices we might be interested in looking at attributes.
-The next type system addresses these concerns.
+Unfortunately, the truth is that dates in R are not standardized.
+That said, translating between date representations is straightforward.
+As most dates are classes, L2 will be able to capture their use.
+
+Throughout this section, we have seen that even though we substantially increased the precision of our type annotations, we have not increased the amount of polymorphism that we observe.
+The transition from L0 to L1 revealed some interesting patterns that L1 cannot quite capture---for these, we will need to appeal to the class and attributes of values.
+The next type system we consider is L2, and is designed to achieve this.
 
@@
 \end{document}
+
+%
+% L1+ Stuff? We don't need this anymore.
+% 
+% 
+%
+%At this point, we have seen a number of reasonable polymorphic patterns that arise in R, both in L0 and L1.
+%The next type system design we consider will deal with these, and we refer to it as L$1^{+}$.
+%The number of monomorphic arguments and functions can be found in Tables~\ref{tab:argcountsL1p} and~\ref{tab:funcountsL1p} respectively.
+%
+%% latex table generated in R 3.5.1 by xtable 1.8-3 package
+%% Fri Mar 22 18:03:30 2019
+%\begin{table}[ht]
+%\label{tab:argcountsL1p}
+%\centering
+%\begin{tabular}{lrr}
+%  \hline
+% Argument Polymorphism & Count & Percentage \\
+%  \hline
+%  Monomorphic in Type & 869K & 95.01 \\
+%%  Monomorphic in Class & 831K & 90.31 \\
+%%  Monomorphic in Attribute Pattern & 830K & 90.25 \\
+%  Total Seen & 914K & --- \\
+%     \hline
+%\end{tabular}
+%\caption{Account of {\it argument} polymorphism in L$1^{+}$.}
+%\end{table}
+%
+%% latex table generated in R 3.5.1 by xtable 1.8-3 package
+%% Fri Mar 22 18:10:46 2019
+%\begin{table}[ht]
+%\label{tab:funcountsL1p}
+%\centering
+%\begin{tabular}{lrr}
+%  \hline
+% Function Polymorphism & Count & Percentage \\
+%  \hline
+%  Monomorphic in Type & 185K & 85.30 \\
+%%  Monomorphic in Class & 163K & 75.13 \\
+%%  Monomorphic in Attribute Pattern & 161K & 74.31 \\
+%  Total Seen & 216K & --- \\
+%   \hline
+%\end{tabular}
+%\caption{Account of {\it function} polymorphism in L$1^{+}$.}
+%\end{table}
+%
+%We had some big wins here, with only ~5\% of argument polymorphism unaccounted for.
+%To get an idea of what more we can do, let's see what the most frequent remaining polymorphic signatures are.
+%See Table~\ref{tab:toppolyL1p}.
+%
+%% latex table generated in R 3.5.1 by xtable 1.8-3 package
+%% Tue Mar 26 17:46:45 2019
+%\begin{table}[ht]
+%\label{tab:toppolyL1p}
+%\centering
+%\begin{tabular}{lrr}
+%  \hline
+%Polymorphic Type Signature & \# Occurrences & \% of Total Polymorphism \\
+%  \hline
+%  matrix, \realv & 6360 & 13.94 \\
+%  list$<$any$>$, list$<$real$>$ & 2799 & 6.13 \\
+%  \C, \realv & 2774 & 6.08 \\
+%  \sC, \reals & 2552 & 5.59 \\
+%  data.frame, matrix & 2167 & 4.75 \\
+%  list$<$any$>$, list$<$list$>$ & 1496 & 3.28 \\
+%  list$<$any$>$, list$<$character$>$ & 1214 & 2.66 \\
+%  data.frame, \realv & 1046 & 2.29 \\
+%  \sL, \reals & 920 & 2.02 \\
+%  \sC, \realv & 868 & 1.90 \\
+%   \hline
+%\end{tabular}
+%\caption{Top 10 argument signatures in L$1^{+}$.}
+%\end{table}
+%
+%A few of the top polymorphic signatures stand out.
+%First, the matrix/\realv signature seems interesting, converting between vectors and matrices is very simple, and indeed numeric vectors are mathematically {\tt 1 x n} matrices.
+%In fact, in R matrices can be used identically to vectors in many cases, and we refer the reader to Figure~\ref{fig:matasvec} for examples.
+%\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+%v <- c(1, 2, 3, 4)
+%ma <- matrix(data=v, nrow=4, ncol=1)
+%v[2] == m[2] == m[2, 1]
+%
+%mb <- matrix(data=v, nrow=2, ncol=2)
+%v[3] == m[3] == m[1, 2]
+%\end{lstlisting}}\caption{Using a matrix like a vector.}\label{fig:matasvec}\end{figure}
+%
+%This similarity is reminiscent of another pattern which was more clear in Table~\ref{tab:L0toppoly}; that of list/vector polymorphism.
+%Transforming between vectors and matrices is just as easy as transforming between lists and vectors is simple.
+%\AT{We should figure out how we want to deal with this polymorphism.}

### 2019-04-05 notebooks/data_builder.Rmd
some edits, lots of code, can generate all of the results up to where im at. working out the last bits of L2 and then plots.

@@
 ```
+
+```{r}
+# L2 Time
+psp_L2 <- read_csv("../data/partial_L1_L2.csv.gz")
+psp_L2 <- psp_L2[!psp_L2$pkg == "SUBTYPE",]
+
+# L2 collapsed?
+what <- read_csv("../data/partial_L1_sanitized_class_collapsed.csv.gz")
+```
+
+```{r}
+# arg sig counts?
+L2_arg_sig_c <- read_csv("../data/partial_L1_sanitized_class_arg_sig_counts_L2.csv.gz")
+
+# sort
+L2_arg_sig_c[order(-L2_arg_sig_c$count),] -> L2_arg_sig_c
+
+# rm []/[]/[]
+L2_arg_sig_c[1,"count"] <- 0
+
+# resort
+L2_arg_sig_c[order(-L2_arg_sig_c$count),] -> L2_arg_sig_c
+
+# %
+L2_arg_sig_c$perc_tot <- round(L2_arg_sig_c$count / sum(L2_arg_sig_c$count) * 100, 1)
+
+# collapse obvious subtypes
+# gets rid of stuff in the first box
+L2_arg_sig_c <- new_df_count_collape_arg_sigs_L2(L2_arg_sig_c)
+
+# what counts do we want?
+# split typeclassattr into type class attr
+L2_split <- split_L2_df(L2_arg_sig_c)
+
+# display some
+L2_split[1:10,] %>% ready_print_count_fun_L2 %>% xtable %>% print(include.rownames=F)
+
+L2_split %>% filter(Class != "") %>% slice(1:10) %>% as.data.frame %>% ready_print_count_fun_L2 %>% xtable %>% print(include.rownames=F)
+```

### 2019-04-05 paper/main.tex
some edits, lots of code, can generate all of the results up to where im at. working out the last bits of L2 and then plots.

@@
 
+%
+%
+%
+%
+%
+%
 \section{Analysis Results}\label{sec:results}
@@
 
-At L0, the type system is mostly aligned with the types of the core language.  We
+At {\bf L0}, the type system is mostly aligned with the types of the core language.  We
 have observed over 186K monomorphic functions; this accounts for 86\% of our
@@
 \end{tabular}
-\caption{Most frequent argument signatures in L0.}
+\caption{Most frequent argument signatures in {\bf L0}.}
 \end{table}
@@
 Next is lists, which seems reasonable given that lists are often used to store the aforementioned data.
-Recall that in L0, data frames will appear as lists.
+Recall that in {\bf L0}, data frames will appear as lists.
 As for ``unevaled'', first recall the discussion that all function arguments are wrapped in promises in R from \AT{where?}.
@@
 Another significant source of apparently strange polymorphism is the character, integer, double polymorphism.
-Figure~\ref{fig:chardbl} lays out an example of this polymorphic signature, showing that lists and vectors can be accessed by strings as well as by numerics.
+Figure~\ref{fig:chardbl} lays out an example of this polymorphic signature, showing that lists and vectors can be accessed by strings as well as numerics.
 One could conceive of an ``index'' type to capture this, but there are certain limitations to such a type, as such an ``index'' cannot be incremented as it could be a string.
@@
 
-Unfortunately, the truth is that dates in R are not standardized.
+Unfortunately, the format of dates in R is not standardized.
 That said, translating between date representations is straightforward.
@@
 \subsection{L2}
+
+{\bf L2} works class and attribute information into its type signatures.
+As per the discussion in Section~\ref{sec:L2def}, we omit builtin classes in order to highlight those which are user-defined.
+
+Now, {\bf L2} builds directly off of {\bf L1}, and as such at the argument level the type-based polymorphism shouldn't change.
+As in {\bf L0} and {\bf L1}, we will begin by showing a breakdown of the top argument signatures in Table~\ref{tab:L2top10arg}.
+
+\begin{table}[ht]
+\label{tab:L2top10arg}
+\centering
+\begin{tabular}{lllrl}
+  \hline
+Type & Class & Attr Ptrn & Count & \% of All Signatures \\ 
+  \hline
+  \sD &  & \{\} & 47069 & 17.6\% \\ 
+  \sL &  & \{\} & 24203 & 9\% \\ 
+  \sC &  & \{\} & 21437 & 8\% \\ 
+  \D &  & \{\} & 20928 & 7.8\% \\ 
+  \sF &  & \{\} & 14585 & 5.5\% \\ 
+  \sN &  & \{\} & 14070 & 5.3\% \\ 
+  unevaled &  & \{\} & 9116 & 3.4\% \\ 
+  \C &  & \{\} & 6949 & 2.6\% \\ 
+  \M{D} &  & \{dim\} & 6612 & 2.4\% \\ 
+  \l &  & \{names\} & 5936 & 2.2\% \\ 
+   \hline
+\end{tabular}
+\caption{Most frequent argument signatures in {\bf L2}.}
+\end{table}
+
+As you can see, the most frequent signatures in {\bf L1} are represented here.
+The inclusion of class and attribute information has changed the counts slightly.
+One tidbit of interest here is that monomorphic named lists are more common than their unnamed counterparts.
+The lack of classes is explained by our elimination of builtin classes, and it's doubtful that a user-defined class would be so pervasive as to feature in this breakdown.
+On that note, we wonder: {\it what are most frequent argument signatures with an interesting class?}
+Table~\ref{tab:L2top10argwclass} has the answer.
+
+% latex table generated in R 3.5.1 by xtable 1.8-3 package
+% Fri Apr  5 02:38:12 2019
+\begin{table}[ht]
+\label{tab:L2top10argwclass}
+\centering
+\begin{tabular}{lllrl}
+  \hline
+Type & Class & Attr Ptrn & Count & \% of All Signatures \\ 
+  \hline
+  \sE & environment & \{\} & 704 & 0.3\% \\ 
+  language & formula & \{.Environment\} & 544 & 0.2\% \\ 
+  \D & array & \{dim\} & 395 & 0.1\% \\ 
+  externalptr & externalptr & \{\} & 301 & 0.1\% \\ 
+  \sF & standardGeneric & \{default,...,srcref,valueClass\} & 255 & 0.1\% \\ 
+  \df & tbl\_df,tbl & \{names,row.names\} & 239 & 0.1\% \\ 
+  \sF & standardGeneric & \{default,...,valueClass\} & 239 & 0.1\% \\ 
+  \l & phylo & \{names,order\} & 199 & 0.1\% \\ 
+  \l & lm & \{names\} & 175 & 0.1\% \\ 
+  \l & lvm & \{names\} & 157 & 0.1\% \\ 
+   \hline
+\end{tabular}
+\caption{Most frequent argument signatures with class in {\bf L2}.}
+\end{table}
+
+Here, we've made a number of small discoveries.
+We see environments, which are essentially dignified lists (i.e., they are lists with a parent list, and are used to implement scoping).
+\AT{Possibly explain some of these? Dunno.}
+
+\AT{Alexi is here.}
 

### 2019-04-05 paper/main.tex
discussed a table in results

@@
 
-Here, we've made a number of small discoveries.
-We see environments, which are essentially dignified lists (i.e., they are lists with a parent list, and are used to implement scoping).
-\AT{Possibly explain some of these? Dunno.}
+Table~\ref{tab:L2top10argwclass} does indeed showcase some interesting classes.
+The three builtin classes which we have not yet seen are environment, formula, and array.
+An environments is a list which has a references to a parent environment (another list with environment class).
+A formula is a \AT{small explanation}.
+An array is a generalization of a matrix in R, where an array can have more than 2 dimensions.
+Each of the other classes pictured here are defined in some package, and are not part of the base language.
+An interesting fact about this table is that all of the pictures signatures are monomorphic.
+
+%Here, we've made a number of small discoveries.
+%We see environments, which are essentially dignified lists (i.e., they are lists with a parent list, and are used to implement scoping).
+%\AT{Possibly explain some of these? Dunno.}
 

### 2019-04-05 notebooks/data_builder.Rmd
Top tallies in results done, plots coming very soon

--- 2019-04-04 notebooks/data_builder.Rmd
@@
 ```
+
+```{r}
+# L2 Time
+psp_L2 <- read_csv("../data/partial_L1_L2.csv.gz")
+psp_L2 <- psp_L2[!psp_L2$pkg == "SUBTYPE",]
+
+# L2 collapsed?
+# what <- read_csv("../data/partial_L1_sanitized_class_collapsed.csv.gz")
+what <- read_csv("../data/partial_L1_collapsed_L2_new.csv.gz")
+```
+
+```{r}
+# ...
+list_of_L2_sig_c <- read_csv("../data/partial_L1_arg_sig_counts.csv.gz")
+
+list_of_L2_sig_c[order(-list_of_L2_sig_c$count),] -> list_of_L2_sig_c
+
+list_of_L2_sig_c[1,"count"] <- 0
+
+list_of_L2_sig_c[order(-list_of_L2_sig_c$count),] -> list_of_L2_sig_c
+
+list_of_L2_sig_c$perc_tot <- round(list_of_L2_sig_c$count / sum(list_of_L2_sig_c$count) * 100, 1)
+
+list_of_L2_sig_c <- new_df_count_collape_arg_sigs_L2_actual(list_of_L2_sig_c)
+
+list_of_L2_sig_c[1:10,] %>% ready_print_count_fun %>% xtable %>% print(include.rownames=F)
+
+# poly arg counts only
+L2_poly_sig <- list_of_L2_sig_c[list_of_L2_sig_c$arg_sig %>% strsplit(split="/") %>% lapply(function(x) length(x) != 3) %>% unlist, ]
+
+# remove error 
+L2_poly_sig[1, "count"] <- 0
+L2_poly_sig[1, "perc_tot"] <- 0
+
+# display poly
+L2_poly_sig[2:11,] %>% ready_print_count_fun %>% xtable %>% print(include.rownames=F)
+
+# Arg Poly Count L2
+data.frame(`Signature Type`=c("Monomorphic", "Total Seen"), Count=c(list_of_L2_sig_c$count %>% sum - L2_poly_sig$count %>% sum, list_of_L2_sig_c$count %>% sum), Percentage=c(paste0(round(100 * (1 - (L2_poly_sig$count %>% sum / list_of_L2_sig_c$count %>% sum)), 1), "%"), "---")) %>% xtable %>% print(include.rownames=F)
+```
+
+```{r}
+# arg sig counts?
+# L2_arg_sig_c <- read_csv("../data/partial_L1_sanitized_class_arg_sig_counts_L2.csv.gz")
+# 
+# # sort
+# L2_arg_sig_c[order(-L2_arg_sig_c$count),] -> L2_arg_sig_c
+# 
+# # rm []/[]/[]
+# L2_arg_sig_c[1,"count"] <- 0
+# 
+# # resort
+# L2_arg_sig_c[order(-L2_arg_sig_c$count),] -> L2_arg_sig_c
+# 
+# # %
+# L2_arg_sig_c$perc_tot <- round(L2_arg_sig_c$count / sum(L2_arg_sig_c$count) * 100, 1)
+# 
+# # collapse obvious subtypes
+# # gets rid of stuff in the first box
+# L2_arg_sig_c <- new_df_count_collape_arg_sigs_L2(L2_arg_sig_c)
+# 
+# # what counts do we want?
+# # split typeclassattr into type class attr
+# L2_split <- split_L2_df(L2_arg_sig_c)
+# 
+# # group class
+# L2_split$Class <- L2_split$Class %>% as.character %>% strsplit(split=", ") %>% sapply(function(x) sapply(x, function(y) paste0("[", paste(y, collapse=", "), "]"))) %>% sapply(function(s) paste(s, collapse=", "))
+# 
+# # display top
+# L2_split[1:10,] %>% ready_print_count_fun_L2 %>% xtable %>% print(include.rownames=F)
+# L2_split %>% filter(Class != "") %>% slice(1:10) %>% as.data.frame %>% ready_print_count_fun_L2 %>% xtable %>% print(include.rownames=F)
+# 
+# L2_poly_ind <- sapply(as.character(L2_split$Type), function(s) length(unlist(strsplit(s, split=", "))) > 1) |
+#                sapply(as.character(L2_split$Class), function(s) length(unlist(strsplit(s, split=", "))) > 1) |
+#                sapply(as.character(L2_split$Attr), function(s) length(unlist(strsplit(s, split=", "))) > 1)
+# L2_poly_sig <- L2_split[L2_poly_ind, ]
+# 
+# # display
+# head(L2_poly_sig, 10) %>% ready_print_count_fun_L2 %>% xtable %>% print(include.rownames=F)
+```

### 2019-04-05 paper/main.tex
Top tallies in results done, plots coming very soon

--- 2019-04-05 paper/main.tex
@@
 
-Now, {\bf L2} builds directly off of {\bf L1}, and as such at the argument level the type-based polymorphism shouldn't change.
 As in {\bf L0} and {\bf L1}, we will begin by showing a breakdown of the top argument signatures in Table~\ref{tab:L2top10arg}.
@@
 \centering
-\begin{tabular}{lllrl}
-  \hline
-Type & Class & Attr Ptrn & Count & \% of All Signatures \\ 
-  \hline
-  \sD &  & \{\} & 47069 & 17.6\% \\ 
-  \sL &  & \{\} & 24203 & 9\% \\ 
-  \sC &  & \{\} & 21437 & 8\% \\ 
-  \D &  & \{\} & 20928 & 7.8\% \\ 
-  \sF &  & \{\} & 14585 & 5.5\% \\ 
-  \sN &  & \{\} & 14070 & 5.3\% \\ 
-  unevaled &  & \{\} & 9116 & 3.4\% \\ 
-  \C &  & \{\} & 6949 & 2.6\% \\ 
-  \M{D} &  & \{dim\} & 6612 & 2.4\% \\ 
-  \l &  & \{names\} & 5936 & 2.2\% \\ 
+\begin{tabular}{lrl}
+  \hline
+Signature & Count & \% of All Signatures \\ 
+  \hline
+  \sD & 46938 & 17.6\% \\ 
+  \sL & 24113 & 9\% \\ 
+  \sC & 21387 & 8\% \\ 
+  \D & 20900 & 7.8\% \\ 
+  \sF & 14556 & 5.5\% \\ 
+  \sN & 14016 & 5.3\% \\ 
+  unevaled & 9093 & 3.4\% \\ 
+  \attrclass{\D}{}{dim} & 6974 & 2.5\% \\ 
+  \C& 6939 & 2.6\% \\ 
+  \attrclass{\l}{}{names} & 6137 & 2.2\% \\ 
    \hline
@@
 The lack of classes is explained by our elimination of builtin classes, and it's doubtful that a user-defined class would be so pervasive as to feature in this breakdown.
-On that note, we wonder: {\it what are most frequent argument signatures with an interesting class?}
-Table~\ref{tab:L2top10argwclass} has the answer.
+%On that note, we wonder: {\it what are most frequent argument signatures with an interesting class?}
+%Table~\ref{tab:L2top10argwclass} has the answer.
+
+%% latex table generated in R 3.5.1 by xtable 1.8-3 package
+%% Fri Apr  5 02:38:12 2019
+%\begin{table}[ht]
+%\label{tab:L2top10argwclass}
+%\centering
+%\begin{tabular}{lllrl}
+%  \hline
+%Type & Class & Attr Ptrn & Count & \% of All Signatures \\ 
+%  \hline
+%  \sE & environment & \{\} & 704 & 0.3\% \\ 
+%  language & formula & \{.Environment\} & 544 & 0.2\% \\ 
+%  \D & array & \{dim\} & 395 & 0.1\% \\ 
+%  externalptr & externalptr & \{\} & 301 & 0.1\% \\ 
+%  \sF & standardGeneric & \{default,...,srcref,valueClass\} & 255 & 0.1\% \\ 
+%  \df & tbl\_df,tbl & \{names,row.names\} & 239 & 0.1\% \\ 
+%  \sF & standardGeneric & \{default,...,valueClass\} & 239 & 0.1\% \\ 
+%  \l & phylo & \{names,order\} & 199 & 0.1\% \\ 
+%  \l & lm & \{names\} & 175 & 0.1\% \\ 
+%  \l & lvm & \{names\} & 157 & 0.1\% \\ 
+%   \hline
+%\end{tabular}
+%\caption{Most frequent argument signatures with class in {\bf L2}.}
+%\end{table}
+%
+%Table~\ref{tab:L2top10argwclass} does indeed showcase some interesting classes.
+%The three builtin classes which we have not yet seen are environment, formula, and array.
+%An environments is a list which has a references to a parent environment (another list with environment class).
+%A formula is a \AT{small explanation}.
+%An array is a generalization of a matrix in R, where an array can have more than 2 dimensions.
+%Each of the other classes pictured here are defined in some package, and are not part of the base language.
+%An interesting fact about this table is that all of the pictures signatures are monomorphic.
+
+Now that we have a sense for argument-level type signatures in {\bf L2}, we will see just how many functions are polymorphic.
+Now, {\bf L2} builds directly off of {\bf L1}, and as such at the argument level the type-based polymorphism shouldn't change.
+For this account of polymorphism, we deem an argument to be polymorphic if it was polymorphic in type, class, or attribute pattern.
+Consider Table~\ref{tab:funcountsL2}.
+\AT{This is where the per fun plot should go, will do.}
+
+\begin{table}[ht]
+\label{tab:funcountsL2}
+\centering
+\begin{tabular}{lll}
+  \hline
+Signature Type & Count & Percentage \\ 
+  \hline
+  Monomorphic & 44K & 79.6\% \\ 
+  Total Seen & 56K & --- \\ 
+   \hline
+\end{tabular}
+\caption{Account of {\it function} polymorphism in {\bf L2}.}
+\end{table}
+
+\AT{Connect, talk about increase/decrease in poly.}
+
+Having built a sense for how polymorphic whole-function signatures are in {\bf L2}, we will turn our attention to polymorphic arguments.
+Given the added class and attribute information, {\it what do the more frequent polymorphic argument signatures look like?}
+Table~\ref{tab:toppolyL2} has the breakdown.
 
 % latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Fri Apr  5 02:38:12 2019
+% Fri Apr  5 09:50:59 2019
 \begin{table}[ht]
-\label{tab:L2top10argwclass}
-\centering
-\begin{tabular}{lllrl}
-  \hline
-Type & Class & Attr Ptrn & Count & \% of All Signatures \\ 
-  \hline
-  \sE & environment & \{\} & 704 & 0.3\% \\ 
-  language & formula & \{.Environment\} & 544 & 0.2\% \\ 
-  \D & array & \{dim\} & 395 & 0.1\% \\ 
-  externalptr & externalptr & \{\} & 301 & 0.1\% \\ 
-  \sF & standardGeneric & \{default,...,srcref,valueClass\} & 255 & 0.1\% \\ 
-  \df & tbl\_df,tbl & \{names,row.names\} & 239 & 0.1\% \\ 
-  \sF & standardGeneric & \{default,...,valueClass\} & 239 & 0.1\% \\ 
-  \l & phylo & \{names,order\} & 199 & 0.1\% \\ 
-  \l & lm & \{names\} & 175 & 0.1\% \\ 
-  \l & lvm & \{names\} & 157 & 0.1\% \\ 
-   \hline
-\end{tabular}
-\caption{Most frequent argument signatures with class in {\bf L2}.}
-\end{table}
-
-Here, we've made a number of small discoveries.
-We see environments, which are essentially dignified lists (i.e., they are lists with a parent list, and are used to implement scoping).
-\AT{Possibly explain some of these? Dunno.}
-
-\AT{Alexi is here.}
-
-To see how often arguments are polymorphic with respect to class and attributes, we refer to Table~\ref{tab:classcountsL2}.
-\AT{Showing full monomorphic all the time might be a bit confusing?}
-
-% latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Mon Apr  1 13:02:08 2019
-\begin{table}[ht]
-\label{tab:classcountsL2}
-\centering
-\begin{tabular}{lrr}
-  \hline
-Type & Count & Percentage \\
-  \hline
-Full Monomorphic & 783K & 85.59 \\
-%  Monomorphic in Type & 869.00 & 95.01 \\
-  Monomorphic in Class & 850K & 92.94 \\
-  Monomorphic in Attribute Pattern & 836K & 91.41 \\
-  Total Seen & 914K & --- \\
-   \hline
-\end{tabular}
-\caption{Account of class- and attribute-based polymorphism in L2.}
-\end{table}
-
-The numbers here are in line with the type-based polymorphism of L1.
-As before, to build up a better sense of what kind of polymorphism we're dealing with, we extract some of the most common signatures.
-We begin with classes, and refer to Table~\ref{tab:classpolyL2}.
-
-% latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Mon Apr  1 14:48:55 2019
-\begin{table}[ht]
-\label{tab:classpolyL2}
-\centering
-\begin{tabular}{lrr}
-  \hline
-Class & Count & Percentage \\
-  \hline
-  primitive, tbl, tbl\_df & 1796 & 2.78 \\
-  array, primitive & 1379 & 2.14 \\
-  data.table, primitive & 777 & 1.20 \\
-  POSIXct, POSIXt & 716 & 1.11 \\
-  formula, primitive & 639 & 0.99 \\
-   \hline
-\end{tabular}
-\caption{Top 5 class signatures for class-based polymorphism in L2.}
-\end{table}
-
-\AT{TODO number of colocated primitive/other classes. Could use to suggest some optional class annotation.
-Maybe grab full class poly to see if there's something big? Without collapse to primitives.}
-
-The first thing to note is the low percentages overall: unlike with types in L0, L1, or L1+, there are no major contributors for class-based polymorphism.
-For instance, the single highest source of polymorphism here is due to the ``tbl'' and ``tbl\_df'' classes (of the dplyr package).
-For what it's worth, these are classes ascribed to dplyr data objects (i.e., spruced-up data frames).
-The next largest contributor is the array class, which is a class for N>2-dimensional matrices (part of the core R language).
-\AT{Maybe I should also turn arrays into primitive class?}
-
-To see how we might use class information, we should see how often a monomorphic class signature co-occurs with a polymorphic type signature.
-We count \NUMPOLYTYPEMONOCLASS argument positions which match this criteria, which certainly doesn't explain the much higher number of list polymorphism in L1+.
-
-In our analysis, we did not encounter any substantial patterns with class-based polymorphism.
-Class is nonetheless important in R, as the dynamic dispatcher uses class information to select the correct method to call.
-In short, a small optional annotation indicating the class of an argument may well be enough.
-
-Let us now turn our attention to attributes.
-The top contributors for attribute-based polymorphism can be found in Table~\ref{tab:topattrpoly}.
-
-% latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Mon Apr  1 23:06:59 2019
-\begin{table}[ht]
-\label{tab:topattrpoly}
+\label{tab:toppolyL2}
 \centering
@@
   \hline
-Attribute Signature & Count & Percentage \\
-  \hline
-\{\}, \{names: \C\} & 16589 & 21.11 \\
-  \{\}, \{dim: \I\} & 5010 & 6.38 \\
-  \{dim:\I,dimnames:list$<$any$>$\}, \{dim:\I\} & 2264 & 2.88 \\
-  \{class:\C,names:\C,row.names:\C\}, \{class:\C,names:\C,row.names:\I\} & 2102 & 2.68 \\
-  \{\}, \{srcref:\I\} & 2037 & 2.59 \\
-  \{dim:\I,dimnames:list$<$\C$>$\}, \{dim:\I\} & 1740 & 2.21 \\
-  \{\}, \{class:\C\} & 1673 & 2.13 \\
-  \{\}, \{class:\C,levels:\C\} & 1437 & 1.83 \\
-  \{\}, \{class:\C,names:\C\} & 1324 & 1.69 \\
-  \{\}, \{class:\C,names:\C,row.names:\I\} & 1251 & 1.59 \\
+Signature & Count & \% of All Signatures \\ 
+  \hline
+  \D, \attrclass{\D}{}{dim} & 553 & 0.2\% \\ 
+  \sC, \sD & 336 & 0.1\% \\ 
+  \I, \sD & 320 & 0.1\% \\ 
+  \attrclass{\l}{}{names}, \attrclass{\D}{}{names} & 269 & 0.1\% \\ 
+  \D, \attrclass{\D}{}{dim, dimnames} & 218 & 0\% \\ 
+  \sD, \sF & 173 & 0\% \\ 
+  \attrclass{\l}{}{names, row.names}, \attrclass{\D}{}{dim, dimnames} & 167 & 0\% \\ 
+  \attrclass{\D}{}{names}, \attrclass{\D}{}{dim, dimnames} & 155 & 0\% \\ 
+  \C, \I & 149 & 0\% \\ 
    \hline
 \end{tabular}
+\caption{Top polymorphic argument signatures in {\bf L2}.}
 \end{table}
 
-First, the largest source of attribute-based polymorphism is in optional list names.
-The names attribute on a vector or list allows indexing with name rather than numeric position.
-Another significant source of this polymorphism is matrices, which can be seen in all the ``dim'' and ``dimnames'' attributes.
-
-\AT{TODO control for the common cases?}
+\AT{Will have new numbers by tonight.}
+
+When comparing this table with Table~\ref{tab:toppolyL1}, we see that the distribution of signature frequencies is more uniform, as attributes and classes differentiated some signatures which were not distinct in {\bf L1}.
+
+Now that we have seen the most frequently appearing argument signatures, we ask: {\it how many argument positions are polymorphic?}.
+Table~\ref{tbl:argcountsL2} has the answer.
+
+\begin{table}[ht]
+\label{tab:argcountsL2}
+\centering
+\begin{tabular}{lll}
+  \hline
+Signature Type & Count & Percentage \\ 
+  \hline
+  Monomorphic & 248K & 93.4\% \\ 
+  Total Seen & 265K & --- \\ 
+   \hline
+\end{tabular}
+\caption{Account of {\it argument} polymorphism in {\bf L2}.}
+\end{table}
+
+Surprisingly, even with the added precision of classes and attributes, there is still a relatively small amount of polymorphism in {\bf L2}.
+Indeed, our notion of subtying on classes is rather weak, and one extension would be to include class hierarchies in the subtyping of classes;
+this would further reduce the amount of polymorphism.
+
+%% latex table generated in R 3.5.1 by xtable 1.8-3 package
+%% Mon Apr  1 14:48:55 2019
+%\begin{table}[ht]
+%\label{tab:classpolyL2}
+%\centering
+%\begin{tabular}{lrr}
+%  \hline
+%Class & Count & Percentage \\
+%  \hline
+%  primitive, tbl, tbl\_df & 1796 & 2.78 \\
+%  array, primitive & 1379 & 2.14 \\
+%  data.table, primitive & 777 & 1.20 \\
+%  POSIXct, POSIXt & 716 & 1.11 \\
+%  formula, primitive & 639 & 0.99 \\
+%   \hline
+%\end{tabular}
+%\caption{Top 5 class signatures for class-based polymorphism in L2.}
+%\end{table}
+%
+%The first thing to note is the low percentages overall: unlike with types in L0, L1, or L1+, there are no major contributors for class-based polymorphism.
+%For instance, the single highest source of polymorphism here is due to the ``tbl'' and ``tbl\_df'' classes (of the dplyr package).
+%For what it's worth, these are classes ascribed to dplyr data objects (i.e., spruced-up data frames).
+%The next largest contributor is the array class, which is a class for N>2-dimensional matrices (part of the core R language).
+%\AT{Maybe I should also turn arrays into primitive class?}
+%
+%In our analysis, we did not encounter any substantial patterns with class-based polymorphism.
+%Class is nonetheless important in R, as the dynamic dispatcher uses class information to select the correct method to call.
+%In short, a small optional annotation indicating the class of an argument may well be enough.
+%
+%First, the largest source of attribute-based polymorphism is in optional list names.
+%The names attribute on a vector or list allows indexing with name rather than numeric position.
+%Another significant source of this polymorphism is matrices, which can be seen in all the ``dim'' and ``dimnames'' attributes.
 

### 2019-04-05 notebooks/data_builder.Rmd
last plot, modified discussion, still plugging away at it

@@
 ---
+
+WORKFLOW
+0. build data on prl3
+  0.1. reduce traces
+  0.2. make_new_df \forall the packages
+  0.3. rbindlist them all
+1. get data from prl3 and do the notebook up to 2.
 
@@
 psp <- transform_new_df_w_TS_map(psp, type_map_to_fun)
+
+# any other transformations?
 ```
@@
 ```
+
+WORKFLOW
+2. write the partials out to csv for java
+3. in java, call doIt (remember -L=L0 for L0, -L=L1 for L1) ALSO NOTE FILE NAMES
 
@@
 data.frame(`Signature Type`=c("Monomorphic", "Total Seen"), Count=paste0(c(mono_L1_funs_c/1000, tot_L1_funs/1000) %>% as.integer, "K"), Percentage=c(paste0(round(mono_L1_funs_c/tot_L1_funs * 100, 1), "%"), "---")) %>% xtable %>% print(include.rownames=F)
+
+# actually want plots, next
+```
+
+```{r}
+# Actually, want per fun plots
+pf_L0 %>% mutate(num_sigs = 1) %>% group_by(pkgfun) %>% summarize(num_sigs=sum(num_sigs)) -> plot_L0
+pf_L1 %>% mutate(num_sigs = 1) %>% group_by(pkgfun) %>% summarize(num_sigs=sum(num_sigs)) -> plot_L1
+
+plot_L0 %>% arrange(num_sigs) %>% mutate(pkgfun_o = factor(pkgfun, pkgfun)) -> plot_L0_df
+plot_L1 %>% arrange(num_sigs) %>% mutate(pkgfun_o = factor(pkgfun, pkgfun)) -> plot_L1_df
+
+# if L0 has no data entry, get rid of it
+if (plot_L0_df[plot_L0_df$pkgfun == "no_data::no_data",] %>% nrow == 1)
+  plot_L0_df <- plot_L0_df[!plot_L0_df$pkgfun == "no_data::no_data",]
+
+L0_perc_poly <- round(100 * nrow(plot_L0_df[plot_L0_df$num_sigs > 1, ] )/nrow(plot_L0_df), 1)
+L1_perc_poly <- round(100 * nrow(plot_L1_df[plot_L1_df$num_sigs > 1, ] )/nrow(plot_L1_df), 1)
+
+# maybe only plot those with more than 1 sig?
+plot_L0_df %>%
+  ggplot(aes(x=pkgfun_o, y=num_sigs, group=1)) +
+  geom_line() + labs(title = "Number of Function Signatures per Function (L0)",
+                                    x = "",
+                                    y = "Number of L0 Signatures") + theme(axis.text.x = element_blank())
+
+ggsave("../paper/plots/L0_by_fun.pdf",width=10,height=4)
+
+plot_L1_df %>%
+  ggplot(aes(x=pkgfun_o, y=num_sigs, group=1)) +
+  geom_line() + labs(title = "Number of Function Signatures per Function (L1)",
+                                    x = "",
+                                    y = "Number of L1 Signatures") + theme(axis.text.x = element_blank())
+
+ggsave("../paper/plots/L1_by_fun.pdf",width=10,height=4)
 ```
@@
 psp_L2 <- psp_L2[!psp_L2$pkg == "SUBTYPE",]
+```
 
+```{r}
+# Plot it
+pf_L2 <- psp_L2 %>% mutate(pkgfun = paste(pkg, fun, sep="::")) %>% select(pkgfun)
+
+pf_L2 %>% mutate(num_sigs = 1) %>% group_by(pkgfun) %>% summarize(num_sigs=sum(num_sigs)) -> plot_L2
+
+plot_L2 %>% arrange(num_sigs) %>% mutate(pkgfun_o = factor(pkgfun, pkgfun)) -> plot_L2_df
+
+L2_perc_poly <- round(100 * nrow(plot_L2_df[plot_L2_df$num_sigs > 1, ] )/nrow(plot_L2_df), 1)
+
+plot_L2_df %>%
+  ggplot(aes(x=pkgfun_o, y=num_sigs, group=1)) +
+  geom_line() + labs(title = "Number of Function Signatures per Function (L2)",
+                                    x = "",
+                                    y = "Number of L2 Signatures") + theme(axis.text.x = element_blank())
+
+ggsave("../paper/plots/L2_by_fun.pdf",width=10,height=4)
+```
+
+```{r}
 # L2 collapsed?

### 2019-04-05 paper/main.tex
2/3 plots
last plot, modified discussion, still plugging away at it

@@
 \newcommand{\PERCSCALARMONO}{37.6\%\xspace}
+
+\newcommand{\LZEROPERCPOLY}{17\%\xspace}
+\newcommand{\LONEPERCPOLY}{20.4\%\xspace}
+\newcommand{\LTWOPERCPOLY}{23\%\xspace}
 
@@
   \sN & 13976 & 5.3\% \\ 
-  unevaled & 9125 & 3.4\% \\ 
+  \ANY & 9125 & 3.4\% \\ 
   \I & 8971 & 3.4\% \\ 
@@
 Recall that in {\bf L0}, data frames will appear as lists.
-As for ``unevaled'', first recall the discussion that all function arguments are wrapped in promises in R from \AT{where?}.
-The ``unevaled'' signature appears whenever an argument was provided but unused, and one might wonder why such a signature might be so common.
-As it happens, many R functions metaprogram on some of their arguments, and this metaprogramming does not force the argument promise, so it will appear as ``unevaled''.
-The magrittr package's pipe \%$>$\% function is an example of this.
-\AT{Move magrittr discussion up.}
+The \ANY type is not an R type per se: we ascribe such a type to arguments which were {\it always} unevaluated across all calls.
+As it happens, many R functions metaprogram on some of their arguments, and this metaprogramming does not force the argument promise, and if this metaprogramming is always performed in every recorded call we will ascribe the \ANY type, as we cannot say what the argument {\it is} without possibly forcing a promise.
+The magrittr package's pipe \%$>$\% function is an example of a function which performs a lot of metaprogramming.
 
@@
 After this, we are curious to know: {\it how many functions have more than one signature?}
-Table~\ref{tab:L0funcounts} has the answer.
-
-\begin{table}[ht]
-\label{tab:L0funcounts}
-\centering
-\begin{tabular}{llr}
-  \hline
-Signature Type & Count & Percentage \\ 
-  \hline
-Monomorphic & 46K & 83\% \\ 
-  Total Seen & 56K & --- \\ 
-   \hline
-\end{tabular}
-\caption{Function polymorphism in L0.}
-\end{table}
+Figure~\ref{fig:L0funcounts} has the answer with a breakdown by function.
+
+\begin{figure}[htbp]\begin{center}
+\includegraphics[width=.9\textwidth]{L0_by_fun}
+\caption{Record of Function Polymorphism in {\bf L0}. \LZEROPERCPOLY of functions are polymorphic.}
+\label{fig:L0funcounts}\end{center}
+\end{figure}
 
 Here, we see that the vast majority of functions are monomorphic in L0.
-To choosing how to express the polymorphicity of a function, there are two main options one might consider:
+Indeed, there are a few functions with a disproportionately large number of signatures.
+One such function is the magrittr package's pipe \%$>$\%:
+
+In choosing how to express the polymorphicity of a function, there are two main options one might consider:
 First, having several separate whole-function signatures, e.g.
@@
   \sN & 14016 & 5.3\% \\ 
-  unevaled & 9093 & 3.4\% \\ 
+  \ANY & 9093 & 3.4\% \\ 
   \C & 7731 & 2.9\% \\
@@
 Now that we have a picture of how arguments look in L1, let's see: {\it how many L1 functions have more than one type signature?}
-The answer can be found in Table~\ref{tab:funcountsL1}.
-
-\begin{table}[ht]
-\label{tab:funcountsL1}
-\centering
-\begin{tabular}{lll}
-  \hline
-Signature Type & Count & Percentage \\ 
-  \hline
-Monomorphic & 44K & 79.6\% \\ 
-  Total Seen & 56K & --- \\ 
-   \hline
-\end{tabular}
-\caption{Account of {\it function} polymorphism in L1.}
-\end{table}
+The answer can be found in Figure~\ref{fig:L1funcounts}.
+
+\begin{figure}[htbp]\begin{center}
+\includegraphics[width=.9\textwidth]{L1_by_fun}
+\caption{Record of Function Polymorphism in {\bf L1}. \LONEPERCPOLY of functions are polymorphic.}
+\label{fig:L1funcounts}\end{center}
+\end{figure}
 
@@
 The lack of classes is explained by our elimination of builtin classes, and it's doubtful that a user-defined class would be so pervasive as to feature in this breakdown.
-%On that note, we wonder: {\it what are most frequent argument signatures with an interesting class?}
-%Table~\ref{tab:L2top10argwclass} has the answer.
-
-%% latex table generated in R 3.5.1 by xtable 1.8-3 package
-%% Fri Apr  5 02:38:12 2019
-%\begin{table}[ht]
-%\label{tab:L2top10argwclass}
-%\centering
-%\begin{tabular}{lllrl}
-%  \hline
-%Type & Class & Attr Ptrn & Count & \% of All Signatures \\ 
-%  \hline
-%  \sE & environment & \{\} & 704 & 0.3\% \\ 
-%  language & formula & \{.Environment\} & 544 & 0.2\% \\ 
-%  \D & array & \{dim\} & 395 & 0.1\% \\ 
-%  externalptr & externalptr & \{\} & 301 & 0.1\% \\ 
-%  \sF & standardGeneric & \{default,...,srcref,valueClass\} & 255 & 0.1\% \\ 
-%  \df & tbl\_df,tbl & \{names,row.names\} & 239 & 0.1\% \\ 
-%  \sF & standardGeneric & \{default,...,valueClass\} & 239 & 0.1\% \\ 
-%  \l & phylo & \{names,order\} & 199 & 0.1\% \\ 
-%  \l & lm & \{names\} & 175 & 0.1\% \\ 
-%  \l & lvm & \{names\} & 157 & 0.1\% \\ 
-%   \hline
-%\end{tabular}
-%\caption{Most frequent argument signatures with class in {\bf L2}.}
-%\end{table}
-%
-%Table~\ref{tab:L2top10argwclass} does indeed showcase some interesting classes.
-%The three builtin classes which we have not yet seen are environment, formula, and array.
-%An environments is a list which has a references to a parent environment (another list with environment class).
-%A formula is a \AT{small explanation}.
-%An array is a generalization of a matrix in R, where an array can have more than 2 dimensions.
-%Each of the other classes pictured here are defined in some package, and are not part of the base language.
-%An interesting fact about this table is that all of the pictures signatures are monomorphic.
 
@@
 For this account of polymorphism, we deem an argument to be polymorphic if it was polymorphic in type, class, or attribute pattern.
-Consider Table~\ref{tab:funcountsL2}.
-\AT{This is where the per fun plot should go, will do.}
-
-\begin{table}[ht]
-\label{tab:funcountsL2}
-\centering
-\begin{tabular}{lll}
-  \hline
-Signature Type & Count & Percentage \\ 
-  \hline
-  Monomorphic & 44K & 79.6\% \\ 
-  Total Seen & 56K & --- \\ 
-   \hline
-\end{tabular}
-\caption{Account of {\it function} polymorphism in {\bf L2}.}
-\end{table}
-
-\AT{Connect, talk about increase/decrease in poly.}
+Consider Figure~\ref{fig:L2funcounts}.
+
+\begin{figure}[htbp]\begin{center}
+\includegraphics[width=.9\textwidth]{L2_by_fun}
+\caption{Record of Function Polymorphism in {\bf L2}. \LTWOPERCPOLY of functions are polymorphic.}
+\label{fig:L2funcounts}\end{center}
+\end{figure}
 
@@
 
-\AT{Will probably change.}
+\AT{Make into subsection at the end of analysis? Or in global results, as a takeaway.}
 

### 2019-04-05 notebooks/data_builder.Rmd
notebook w data

@@
 # read in new one
-psp_L0 <- read_csv("../data/partial_L0_subtype.csv")
-
+psp_L0 <- read_csv("../data/partial_L0_subtype.csv.gz")
+p
 # remove SUBTYPE rows
@@
 # read it back
-psp <- read_csv("../data/partial_L1_subtype.csv")
+psp <- read_csv("../data/partial_L1_subtype.csv.gz")
 

### 2019-04-05 paper/main.tex
edits

@@
 \subsection{Implementation}
+\label{sec:Impl}
 
@@
 This section presents our results for each of the levels we have considered.
-
-\AT{NOTE: I am updating this with new numbers as we get them.
-The fixed tracer is still running, I should be able to fill in final numbers by tomorrow night.}
+\AT{New numbers will be available tonight (Apr 5).}
+\AT{Will use that to fill in some blanks.}
 
@@
 At {\bf L0}, the type system is mostly aligned with the types of the core language.  We
-have observed over 186K monomorphic functions; this accounts for 86\% of our
+have observed over \AT{how many?} monomorphic functions; this accounts for \AT{what \%?} of our
 corpus.
@@
 As it happens, many R functions metaprogram on some of their arguments, and this metaprogramming does not force the argument promise, and if this metaprogramming is always performed in every recorded call we will ascribe the \ANY type, as we cannot say what the argument {\it is} without possibly forcing a promise.
-The magrittr package's pipe \%$>$\% function is an example of a function which performs a lot of metaprogramming.
+The magrittr package's pipe \%$>$\% function is an example of a function which performs a lot of metaprogramming; we will discuss it in a little more detail shortly.
 
@@
 Here, we see that the vast majority of functions are monomorphic in L0.
-Indeed, there are a few functions with a disproportionately large number of signatures.
-One such function is the magrittr package's pipe \%$>$\%:
+That said, there are a few functions with a disproportionately large number of signatures.
+One such function is the magrittr package's pipe \%$>$\%, which is used to implement a kind of currying (as per discussion in Section~\ref{sec:Impl}).
+Many of the metaprogramming functions in the Tidyverse are quite sophisticated, and are designed with the explicit goal of being very usable.
+This usability maps to polymorphism in our analysis, so it's no surprise that a Tidyverse function features so prominently here.
+
+Another function with many signatures is the data.table package's [ function, which is overloaded for objects with data.table class.
+This function has 16 formal arguments, and call configurations vary widely.
+Essentially, [ is {\it the} main function for data.tables, and performs very different tasks based on which arguments are provided.
+Two wildly different uses are:
+\code{DT[i, j, by]}, where columns \code{i} and \code{j} are accessed and grouped by \code{by}, and \code{DT[x == "some" & y > 42]}, which selects all rows who's \code{x} column is the string \code{"some"} and who's \code{y} column is greater than \code{42}.
+Given that, we find it unsurprising that the function is so polymorphic.
 
@@
 (\D $\cup$ \C) $\rightarrow$ \C.
-Earlier in Table~\ref{tab:L0top10arg} we saw the most frequently occurring argument signatures.
-To get an idea of how expressing polymorphism at the argument level would look, we ask: {\it what are the most frequently occurring polymorphic argument signatures?}
-Table~\ref{tab:L0toppoly} has the answer.
+The former is referred to as {\it per-function} or {\it function-level polymorphism}, and the latter as {\it per-argument} or {\it argument-level polymorphism}.
+Earlier in Table~\ref{tab:L0top10arg} we saw the most frequently occurring argument-level signatures, and Figure~\ref{fig:L0funcounts} broke our corpus down by function-level polymorphism.
+
+To get an idea of the most common forms of argument-level polymorphism, we refer toTable~\ref{tab:L0toppoly}.
 
@@
 	\item \D, \sF: this signature appears more strange than it is.
-	Some arguments with this signature are either parameters for a function, or the function itself already constructed.
-	Another examples, such as the {\tt br} argument of the {\tt rtree} function in the {\tt ape} package, are parameters for a complex data generation process, or a function to be used during the generation.
-	\AT{Do we want to put the fun def and arg documentation here?}
-	It would appear that this pattern is used to make complex simulation, modelling, or data generation functions more flexible.
-
-	\item \C, \sF: another puzzling signature, though its explanation is surprisingly straightforward:
+	Many models, processes, simulations, etc. that are prevalent throughout R are are {\it parameterized}.
+	Some arguments with the \D, \sF signature are either parameters for a function (\D), or the function itself already constructed (\sF).
+	Others, such as the {\tt br} argument of the {\tt rtree} function in the {\tt ape} package, are either parameters for complex data generation processes (\D), or a function to be used during the generation (\sF).
+	It would appear that this pattern is used to make complex simulation, modelling, or data generation functions more flexible for programmers.
+
+	\item \C, \sF: yet another puzzling signature, though its explanation is surprisingly straightforward:
 	In R, programmers are free to look functions up {\it by name}, and indeed the frequently encountered {\tt UseMethod} function for dynamic class-based function dispatch is one such example.
@@
 
-Now, we have some understanding of argument polymorphism in L0.
+We have established and discussed some of the most common polymorphic attribute signatures, and have built up an understanding of argument-level polymorphism in the {\bf L0} type system.
+This matches almost exactly with the language as it is today, as {\bf L0} is in (nearly) 1:1 correspondence with how types are currently represented in the language.
 The next question we address is: {\it just how pervasive is this argument polymorphism?}

### 2019-04-05 notebooks/data_builder.Rmd
more work on the end of results, looking at other interesting data points to flesh out

@@
 ```{r}
+# packages with the most polymorphism
+# which TS to use for this?
+
+num_funs_by_pkg <- psp_L0 %>% group_by(pkg, fun) %>% select(pkg, fun) %>% distinct(fun) %>% mutate(count=1) %>% group_by(pkg) %>% summarize(count=sum(count))
+
+num_sigs_by_pkg <- psp_L0 %>% mutate(sig_count = 1) %>% group_by(pkg) %>% summarize(sig_count=sum(sig_count))
+
+poly_count_normalized <- inner_join(num_funs_by_pkg, num_sigs_by_pkg, by="pkg") %>% mutate(poly_norm=sig_count/count)
+
+poly_count_normalized %>% arrange(-poly_norm) %>% slice(2:6)
+
+# print
+
+```
+
+Old stuff.
+```{r}
 # arg sig counts?

### 2019-04-05 paper/main.tex
more work on the end of results, looking at other interesting data points to flesh out

@@
 
-\AT{Something more about functions. Plots...}
-
 Before presenting L2, we will digress slightly and explore some of the ramifications of our transition from L0 to L1.
@@
 Surprisingly, even with the added precision of classes and attributes, there is still a relatively small amount of polymorphism in {\bf L2}.
-Indeed, our notion of subtying on classes is rather weak, and one extension would be to include class hierarchies in the subtyping of classes;
+Indeed, our notion of subtyping on classes is rather weak, and one extension would be to include class hierarchies in the subtyping of classes;
 this would further reduce the amount of polymorphism.
 
-%% latex table generated in R 3.5.1 by xtable 1.8-3 package
-%% Mon Apr  1 14:48:55 2019
-%\begin{table}[ht]
-%\label{tab:classpolyL2}
-%\centering
-%\begin{tabular}{lrr}
-%  \hline
-%Class & Count & Percentage \\
-%  \hline
-%  primitive, tbl, tbl\_df & 1796 & 2.78 \\
-%  array, primitive & 1379 & 2.14 \\
-%  data.table, primitive & 777 & 1.20 \\
-%  POSIXct, POSIXt & 716 & 1.11 \\
-%  formula, primitive & 639 & 0.99 \\
-%   \hline
-%\end{tabular}
-%\caption{Top 5 class signatures for class-based polymorphism in L2.}
-%\end{table}
-%
-%The first thing to note is the low percentages overall: unlike with types in L0, L1, or L1+, there are no major contributors for class-based polymorphism.
-%For instance, the single highest source of polymorphism here is due to the ``tbl'' and ``tbl\_df'' classes (of the dplyr package).
-%For what it's worth, these are classes ascribed to dplyr data objects (i.e., spruced-up data frames).
-%The next largest contributor is the array class, which is a class for N>2-dimensional matrices (part of the core R language).
-%\AT{Maybe I should also turn arrays into primitive class?}
-%
-%In our analysis, we did not encounter any substantial patterns with class-based polymorphism.
-%Class is nonetheless important in R, as the dynamic dispatcher uses class information to select the correct method to call.
-%In short, a small optional annotation indicating the class of an argument may well be enough.
-%
-%First, the largest source of attribute-based polymorphism is in optional list names.
-%The names attribute on a vector or list allows indexing with name rather than numeric position.
-%Another significant source of this polymorphism is matrices, which can be seen in all the ``dim'' and ``dimnames'' attributes.
-
-%
-%
-%
-%
-\subsection{R's Most Polymorphic Functions}
-
-\AT{We might want to discuss some of the most polymorphic functions in the language.}
-
-In this section, our goal is to shed some light on the most polymorphic functions in R.
-Table~\ref{tab:bigpolyfuns} shows the 10 functions with the biggest recorded signatures.
-
-% latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Thu Mar 28 16:01:37 2019
-\begin{table}[ht]
-\label{tab:bigpolyfuns}
-\centering
-\begin{tabular}{llrl}
-  \hline
-Package & Function & Size & Description \\
-  \hline
-magrittr & \%$>$\% & 23.00 & Tidyverse pipe function \\
-  magrittr & freduce & 17.67 & Apply a list of functions sequentially \\
-  rlang & captureArgInfo & 16.50 & Capture arg info \\
-  rlang & enquo & 16.00 & Create quosure \\
-  rlang & get\_expr & 15.00 & Extract expression \\
-  data.table & replace\_dot\_alias & 13.50 & Deals with '.' in passed expr \\
-  tibble & strip\_dim & 13.00 & Removes 'dim' attribute \\
-  data.table & copy & 11.00 & Create copy of passed value \\
-  htmlwidgets & shouldEval & 11.00 & Eval passed value depending on tag \\
-  magrittr & \%$<$$>$\% & 11.00 & Like \%$>$\% but assigns \\
-   \hline
-\end{tabular}
-\caption{10 most polymorphic functions.}
-\end{table}
-
-The size metric we use is roughly the size of each individual argument signature, divided by the number of function arguments.
-The division is performed to avoid unfairly favouring functions with many arguments.
-
-Which packages are the most polymorphic?
-Figure~\ref{tab:packagepolysize} has the answer.
+\AT{Conclude L2}
+
+%
+%
+%
+%
+\subsection{Notable Patterns and Observations}
+
+Besides exploring the patterns highlighted by our various type systems, we can address some interesting and tangentially related questions about the usage of types in R.
+
+%
+%
+%
+%
+\subsubsection{R's Most Polymorphic Functions}
+
+Given the data that we've collected, it's natural to wonder what the most polymorphic packages are.
+Indeed, we have shed some light on some of the most polymorphic {\it functions}, but what about the packages that these functions belong to?
+What is their purpose, and why are they so polymorphic?
+Figure~\ref{tab:packagepolysize} has some answers.
 
@@
 \centering
-\begin{tabular}{lrl}
-  \hline
-Package & Total Size & Description \\
-  \hline
-  magrittr & 4.16 & Tidyverse pipe suite \\
-  types & 3.25 & Simple type annotations for R \\
-  r6extended & 3.00 & Utilities extending barebones R6 package \\
-  utils & 2.93 & Utilities for R developers \\
-  zeallot & 2.85 & Metaprogramming package for assignment \\
-  listviewer & 2.72 & R list visualizer \\
-  proto & 2.60 & Prototype object-based programming \\
-  apercu & 2.49 & Display a short view of a complex R value (list, vector, etc.) \\
-  pipeR & 2.46 & Another pipe suite \\
-  rethinker & 2.44 & RethinkDB client for R \\
+\begin{tabular}{lrrr}
+  \hline
+  Package & \# of Functions &\# of Signatures & Description \\ 
+  \hline
+  knitLatex & 4 & 41 & \LaTeX \xspace generator from R data \\ 
+  magrittr & 17 & 126 & Tidyverse pipe suite \\ 
+  proto & 5 & 30 & Prototype object-based programming \\ 
+  gnlm & 10 & 53 & ?? \\ 
+  Formula & 17 & 82 & ?? \\ 
    \hline
 \end{tabular}
-\caption{10 most polymorphic packages.}
+\caption{5 most polymorphic packages.}
 \end{table}
 
-\AT{Draft}
-
-The most polymorphic function in the corpus of code we analyzed is {\tt magrittr}'s {\tt \%>\%} (pipe) function.
-{\tt magrittr} is part of the Tidyverse, and the pipe function is frequently used to chain function calls.
-The fact that the pipe is the most polymorphic function is not shocking, as its goal is simply to take the right hand argument and apply it to the left hand argument, and this in a sense irrespective of type.
-
-...
+Now, what makes these so polymorphic?
+\AT{maybe first score functions, penalizing big ones, and then consolidate?}
+
+\begin{enumerate}
+
+	\item knitLatex: two functions are largely responsible for this level of polymorphism: sTab, which produces \LaTeX \xspace supertabular environments, and lTab, which produces \LaTeX \xspace longtable environments.
+	These functions both have over 15 arguments, and there are a large number of recorded configurations, with many of the arguments being optional.
+
+	\item magrittr: we discussed magrittr earlier in the context of its pipe \%$>$\% function. 
+	It's not surprising that the package as a whole is quite polymorphic, as it implements {\it several} different pipe functions, e.g.
+\%$<>$\% which pipes as normal and then assigns to the left-hand argument.
+
+	\iitem proto:
+
+\end{enumerate}
+
+%  magrittr & 4.16 &  \\
+%  types & 3.25 & Simple type annotations for R \\
+%  r6extended & 3.00 & Utilities extending barebones R6 package \\
+%  utils & 2.93 & Utilities for R developers \\
+%  zeallot & 2.85 & Metaprogramming package for assignment \\
+%  listviewer & 2.72 & R list visualizer \\
+%  proto & 2.60 &  \\
+%  apercu & 2.49 & Display a short view of a complex R value (list, vector, etc.) \\
+%  pipeR & 2.46 & Another pipe suite \\
+%  rethinker & 2.44 & RethinkDB client for R \\
+
+%% latex table generated in R 3.5.1 by xtable 1.8-3 package
+%% Thu Mar 28 16:01:37 2019
+%\begin{table}[ht]
+%\label{tab:bigpolyfuns}
+%\centering
+%\begin{tabular}{llrl}
+%  \hline
+%Package & Function & Size & Description \\
+%  \hline
+%  magrittr & \%$>$\% & 23.00 & Tidyverse pipe function \\
+%  magrittr & freduce & 17.67 & Apply a list of functions sequentially \\
+%  rlang & captureArgInfo & 16.50 & Capture arg info \\
+%  rlang & enquo & 16.00 & Create quosure \\
+%  rlang & get\_expr & 15.00 & Extract expression \\
+%  data.table & replace\_dot\_alias & 13.50 & Deals with '.' in passed expr \\
+%  tibble & strip\_dim & 13.00 & Removes 'dim' attribute \\
+%  data.table & copy & 11.00 & Create copy of passed value \\
+%  htmlwidgets & shouldEval & 11.00 & Eval passed value depending on tag \\
+%  magrittr & \%$<$$>$\% & 11.00 & Like \%$>$\% but assigns \\
+%   \hline
+%\end{tabular}
+%\caption{10 most polymorphic functions.}
+%\end{table}
 

### 2019-04-05 paper/main.tex
more prose and some edits

@@
 
-\title{A Large-scale Study of Polymorphism in R}
+\title{A Large-Scale Study of Polymorphism in R}
 
@@
 \newcommand{\CLOC}{9,373,542\xspace}
-\newcommand{\YEARS}{?\xspace}
+\newcommand{\YEARS}{20\xspace}
 \newcommand{\INDEXCOINCIDENCE}{3,499\xspace}
@@
 code spread among some \PACKAGES packages, written over a
-period of over \YEARS years by \PROGRAMMERS programmers.  We perform a dynamic
+period of over \YEARS years by thousands of programmers.  We perform a dynamic
 analysis, leveraging tests and use-cases, to determine the level of
@@
 Subtyping is extended such that {\tt T}<:{$\vec{\tt T}$} for any primitive
-type {\tt T}. There is no subtyping relation between lists of different
+type {\tt T}. The subtype relation is undefined for lists of different
 types.
@@
 the user (these are the class names for the object).  Thus, we extend {\bf
-  L1} with attribute lists and classes: \attrclass{T}{C}{N} where {\tt T}
-can be any L1 type other than data frame or matrix, {\tt C} and {\tt N}
-stands for a list names.  The subtype relation is extended with
-\attrclass{T}{C}{N}<:\attrclass{T}{C'}{D'} if {\tt C} $\subseteq$ {\tt C'}
-and {\tt N}$\subseteq${\tt N'}.
+  L1} with attribute lists and classes: \attrclass{T}{C}{A} where {\tt T}
+can be any L1 type other than data frame or matrix, and {\tt C} and {\tt A}
+are lists of classes and attribute names respectively.  The subtype relation is defined as:
+\attrclass{T}{C}{A} $<:_{\mathbf{L2}}$ \xspace \attrclass{\dot T}{C'}{A'} if 
+$ T <:_{\mathbf{L1}} \dot T$, $C \subseteq C'$, and $A \subseteq A'$.
 
@@
 As per the discussion in Section~\ref{sec:L2def}, we omit builtin classes in order to highlight those which are user-defined.
+{\bf L2} differs from {\bf L1} in the following ways:
+
+\begin{enumerate}
+
+	\item where {\bf L1} and {\bf L0} only considered classes and attributes when ascribing matrix or data frame types, {\bf L2} considers classes and attributes to be an integral part of the type system.
+	Annotations in {\bf L2} include classes and attribute names when they are present.
+
+	\item recall that subtyping between two {\bf L2} types is defined as
+
+\end{enumerate}
 
@@
   \sN & 14016 & 5.3\% \\ 
-  unevaled & 9093 & 3.4\% \\ 
+  \ANY & 9093 & 3.4\% \\ 
   \attrclass{\D}{}{dim} & 6974 & 2.5\% \\ 
@@
 
-Surprisingly, even with the added precision of classes and attributes, there is still a relatively small amount of polymorphism in {\bf L2}.
-Indeed, our notion of subtyping on classes is rather weak, and one extension would be to include class hierarchies in the subtyping of classes;
-this would further reduce the amount of polymorphism.
-
-\AT{Conclude L2}
+Surprisingly, even with the added precision of classes and attributes, there is still a relatively small amount of argument-level polymorphism in {\bf L2}.
+Our notion of subtyping on classes is rather weak, and one extension would be to include class hierarchies in the subtyping of classes;
+this would further reduce the amount of polymorphism in this type system.
 
@@
 Figure~\ref{tab:packagepolysize} has some answers.
+Note that we are in the context of {\bf L0}, here, to highlight current language patterns.
 
@@
   proto & 5 & 30 & Prototype object-based programming \\ 
-  gnlm & 10 & 53 & ?? \\ 
+  gnlm & 10 & 53 & Regression fit package \\ 
   Formula & 17 & 82 & ?? \\ 
@@
 \end{tabular}
-\caption{5 most polymorphic packages.}
+\caption{5 most polymorphic packages (in {\bf L0}).}
 \end{table}
@@
 Now, what makes these so polymorphic?
-\AT{maybe first score functions, penalizing big ones, and then consolidate?}
+
+\AT{TODO if time: fix to score per fun, then tally per package.}
 
@@
 
-	\iitem proto:
+	\item proto: most of the polymorphism in this package comes from its implementation of assignment ({\tt \$<-}) for proto class objects.
+	It's a small, 3 argument function which is exceedingly polymorphic, with over 20 unique call signatures.
+	As theoretically any R object can have a proto class, it stands to reason that this function ought to be quite polymorphic indeed.
+
+	\item
+
+	\item
 
 \end{enumerate}
+
+It's highly improbable that a static type system designed for an untyped language will be able to capture all of the language's existing code (without resorting to an \ANY type or abusing unions).
+It's also unclear how useful a type system would be if it {\it did} capture everything.
+Indeed, many of the packages discussed here are really designed to take in anything for particular arguments (e.g., left-hand side of \%$>$\%), and restricting these functions should not be the goal of any static type system designed after a language has matured.
 
@@
 %
-\subsection{Metaprogramming}
-
-\AT{Make into subsection at the end of analysis? Or in global results, as a takeaway.}
-
-Tidyverse owes some of its success to the new syntax defined by some of the included packages.
-One such example is the pipe operator \%$>$\% that we saw in the last section, and indeed this function was highly exercised by many other packages \AT{get counts}.
-We also grappled with the unfortunate truth that arguments to \%$>$\% can't always be evaluated, which causes some annoyances in the analysis.
-
-We would like to know how often packages contain functions designed for metaprogramming.
-As it happens, when arguments fail to evaluate during our analysis even as functions execute successfully, we are actually dealing with instances of metaprogramming.
-Some functions, such as those in the magrittr package, will take one (or more) of their arguments and use them as an AST, performing surgery on the expression to create some new expression that they evaluate.
-Sometimes, these arguments intended for metaprogramming {\it will} successfully evaluate in our tracer, but this is merely chance.
-For example, consider the code snippet in Figure~\ref{fig:metaexworks}.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-mul_by_w_defaults <- function(x=1, y=1) x * y
-
-4 %>% mul_by_w_defaults(2) == `%>%`(lhs = 4, rhs = mul_by(3))
-\end{lstlisting}}\caption{Dates in R.}\label{fig:metaexworks}\end{figure}
-
-In this example, our analysis will successfully evaluate the {\tt rhs} argument
+\subsubsection{Metaprogramming and Unevaluated Arguments}
+
+Our analysis uncovered a non-negligible amount of argument positions which were {\it never} evaluated across all observed function calls.
+This is indicative of a desire to metaprogram arguments and otherwise deconstruct their AST, as the \%$>$\% function does.
+This is an incredibly powerful language feature, and may well be at the source of R's popularity, if in a roundabout way.
+Tidyverse, by far the most popular package suite in R, owes some of its success to the new syntax defined by some of the included packages, and it's this ability to metaprogram arguments which allowed Tidyverse to define its unique syntax.
+An example of that syntax is \%$>$\%, and indeed this function was highly exercised by over 100 other packages.
+
+%
+%
+%
+%
+%
+%
+\section{Discussion}
+
+\AT{Short section to recap the analysis results, discuss trends, consolidate our suggestions, talk about argument vs. function poly.}
 
@@
 \subsection{Function Polymorphism}
+
+\AT{Retool this into possible meta discussion after results}
 

### 2019-04-05 paper/main.tex
method fixes and edits

@@
 by synthesizing unit tests from recorded function argument and return
-values, comparing the test against existing ones to avoid generating tests
+values, comparing candidate tests against existing ones to avoid generating tests
 which do not increase code coverage.  To suit our purposes, we change the
@@
 \tt c: &\tt \sD \sD \to \D\\
-\tt `+`: &\tt \attr\D{dim=\D} \D \to \attr\D{dim=\D}
+\tt `+`: &\tt \attr\D{dim} \D \to \attr\D{dim}
 \end{tabular}
@@
 The functions being called here are \k{`+`}, \k{`<-`}, \k{c} and
-\k{`attr<-`}.  The shapes we would expect to record are as follows:
-In the above we abreviate types, \k{double}, \k{integer}, \k{symbol} and
-\k{character}. For vectorized types we record their length and for all types
-we record attributes with some of their values. Looking at the signatures
-observed for addition, it is clear that the function is polymorphic as it
-starts with the addition of two scalar numbers of different types, and then
-adds a matrix to a vector returning a matrix.
+\k{`attr<-`}, and the shapes we expect to record are given to the right of the figure.
+We abbreviate the types \k{double}, \k{integer}, \k{symbol} and \k{character} to \sD, \sI, \sY, and \sC for simplicity.
+\D, \I, \Y, and \C denote vectors, whereas \sD, \sI, \sY, and \sC denote scalars.
+We record the lengths of all vectorized-typed values, and for all types we record their classes and attribute names. 
+Looking at the signatures observed for addition, it is clear that the function is polymorphic: it starts with the addition of two scalar numbers of different types, and later adds a matrix to a vector returning a matrix.
 
@@
   For \code{list} values, an extra bit of analysis is performed to record
-  the element type.
+  the element types.
 \item {\bf Writing:} Unique signatures are recorded to file with information
@@
 
-Our recording mechanism does not remember the order in which arguments were
-passed, nor does it record which arguments were not passed (and for which
-default values were used). If an argument was not passed, and no default
-value was specified, we record it as \emph{missing}. In R, trying to use
-such a missing value results in an error. For practical reasons, we do not
-record the contents of environments. These can be used as hash tables and
-may be big and are quite likely different from one another.
-
-In R all arguments are passed as promises, and unused arguments will be
-unevaluated.  Our tracer does not collect information on arguments that were
-not evaluated as they may not be valid expressions.  Consider the {\tt
-  magrittr} package, which implements the pipe function that performs a kind
-of currying.
+In R, all arguments are passed as promises, and unused arguments will be unevaluated. 
+Our recording mechanism captures the order in which arguments are passed, can figure out which arguments were {\it not} provided (recording them as {\it missing} if the missing argument has no default value), and does {\it not} force unforced promises, recording unforced promises as {\it unevaled}.
+In R, trying to use missing values results in an error and thus an exception, triggering our recording phase: arguments which cause errors in this way are recorded as {\it error}.
+For practical reasons, we do not record the contents of environments:
+These can be used as hash tables, may be large, and are quite likely different from one another.
+
+Unevaluated arguments in R are sometimes {\it carefully} used by their functions to implement some form of metaprogramming.
+Consider the {\tt magrittr} package, which implements a pipe function \%$>$\% that implements a flavor of function currying:
 
@@
 \code{rhs} is not a valid expression. It is treated as code that will
-manipulated to add one argument (the \code{lhs}).
-
-To be as specific as possible, our analysis collects shape information about
-arguments and return values. The details of the information we collect follows.
+manipulated to add one argument (the \code{lhs}), resulting in the function
+actually performing a call of the form \code{mul(4, 3)}.
+
+As we alluded to at the beginning of this section, our analysis collects shape information about
+arguments and return values in order to be as specific as possible. The details of the information we collect follows.
+Note that the first step of any type collection in our analysis begins with \code{typeof}.
 
@@
   we only collect the content's \code{typeof}, and ascribe \ANY (any) if the
-  elements are not of the same kind.k
+  elements are not of the same kind.
 \item If the analysis encounters a scalar \code{NA}, we ascribe a unique
@@
   \code{typeof(NA)=="logical"}. Scalar NAs are used as uninitialized values.
-\item For a matrix (i.e. the value has matrix class and has a dims and
-  optionally dimnames attribute), we ascribe the shape \attr{\tt T}{mat}
+\item For a matrix, \code{typeof(m)=="some\_primitive\_type"}, and\code{m} will have matrix class and also a dims and
+  optionally dimnames attribute. Here, we ascribe the shape \attr{\tt T}{mat}
   where {\tt T} is the primitive element type.
-\item For a data frame (i.e., class {\tt data.frame} with appropriate
-  attributes), we ascribe the shape \attr{\l}{df}.
+\item For a data frame, \code{typeof(df)=="list"}, and \code{df} will have class {\tt data.frame}, and {\tt names} and {\tt row.names} attributes.
+ To reflect this, we ascribe the shape \attr{\l}{df}.
 \end{itemize}
@@
 increasingly more expressive type systems. The simplest notion of types
-aligns with R's builtin types. Richers notions incorporate attributes and
+aligns with R's builtin types. Richer notions incorporate attributes and
 the various object systems built on top of those. For concreteness we will

### 2019-04-05 paper/main.tex
edits to types section, with better correctness

@@
 
-For a meaningful notion of polymorphism, it is necessary to settle on a set
-of basic types for R. There are several possible choices which lead to
-increasingly more expressive type systems. The simplest notion of types
-aligns with R's builtin types. Richer notions incorporate attributes and
-the various object systems built on top of those. For concreteness we will
-pick out three different sets of types.
-
-In the following we say that an argument is {\it polymorphic in type at
-  level L} if it has been called with at least two values that belong to
-distinct types at level \emph{L}. This definition extends naturally to
-classes, data frames, and matrices (as these are simple attributes).
-Generalized attribute polymorphism is a bit trickier as values can have a
-large and changing set of attributes.
+Our aim is to capture the polymorphism in R, and to do so we must define a meaningful set of basic types for the language.
+There are several possible choices, with a variety of levels of expressivity and complexity.
+For concreteness, we will pick out three different sets of types.
+To start simple, we define a set of types which closely align with R's builtin types, and later define richer notions of type which incorporate attributes, classes, and capture the various object systems of R.
+
+Throughout this paper, we will say that an argument is {\it polymorphic at} \textbf{\emph{LX}} (read: level X) if it has been called with at least two values that belong to distinct types at {\bf LX}.
+\textbf{L2} will incorporate class and attribute lists, which are slightly trickier to ``type'', and we defer discussion of this until Section~\ref{sec:L2def}.
 
@@
 
-The logical ``baseline'' for the design of a type system for R is to adopt
-the language's builtin types.  Types in this system, which we call {\bf L0}
-reflect the types of values in the runtime environment, the \code{typeof} an
-object is in direct correspondence with the runtime type tag of the value.
-Thus we include the following types: vectors of primitive types (integers
+The logical ``baseline'' for the design of a type system for R is to adopt the language's builtin types.
+We dub these {\bf L0} types, and they are primarily built on the \code{typeof} values, which reflects the implementation of the type in the runtime environment.
+More specifically, the \code{typeof} a value is simply the runtime type tag of the value.
+Given that, {\bf L0} includes the following types: vectors of primitive types (integers
 \I, double \D, character \C, logical \L, raw \R, complex \X), lists of any
 type (\l), and scalar values (environment \sE, closures \sF, symbols \sY, S4
-object \sS, and null \sN). We add one type stand in for unevaluated values,
-the type any (written \ANY). In {\bf L0} subtyping follows from the
-conversion rules of the language.  Logicals are subtypes of integers, which
-are subtypes of doubles, \L <: \I <: \D.  All types are subtypes of \ANY,
+object \sS, and null \sN). We omit short forms for some of R's more esoteric types (e.g., {\tt environment}s).
+ We add the any (written \ANY)  type to stand in for unevaluated values
+Subtyping in {\bf L0} follows from the conversion rules of the language:  Logicals are subtypes of integers, which
+are themselves subtypes of doubles (so \L <: \I <: \D).  All types are subtypes of \ANY,
 thus {\tt T}<:\ANY for an type {\tt T}.
@@
 
-{\bf L0} is limited in expressivity as it supports polymorphism for list in
-a trivial way, it does not address some of the popular constructed data
-types such as matrices and data frames, and cannot differentiate between
-scalar values and vectors.  {\bf L1} types build on and extend the previous
-level with with scalar values for primitive types (\sI, \sD, \sC, \sL, \sR,
-and \sX), type parametric lists (written \lT{T} for any type {\tt T}), data
-frames (written \attr{\l}{df} as they are built up from lists of type any),
-and matrices (\attr{\vec{\tt T}}{mat} where {\tt T} is a primitive type).
+{\bf L0} is limited in expressivity as it (1) supports polymorphism for list in
+a trivial way, (2) does not address some of the popular constructed data
+types such as matrices and data frames, and (3) cannot differentiate between
+scalar values and vectors.  
+{\bf L1} types build on and extend {\bf L0} with with scalar values for primitive types (\sI, \sD, \sC, \sL, \sR, and \sX), type-parametric lists (written \lT{T} for any type {\tt T}), data frames (written \attr{\l}{df} as they are built up from lists of type any), and matrices (\attr{\vec{\tt T}}{mat} where {\tt T} is a primitive type).
 Subtyping is extended such that {\tt T}<:{$\vec{\tt T}$} for any primitive
-type {\tt T}. The subtype relation is undefined for lists of different
-types.
+type {\tt T}. The subtype relation is undefined for lists of different types, as well as for matrices and data frames.
 
@@
 
-{\bf L2} extends the previous system with more support for attributes.
-While there is a finite number of results from the {\tt typeof} function,
-this is not the case for \code{class} and \code{attributes}.  Classes and
-attributes are user-defined.  Of course, not all classes are user-defined,
-and R has a number of built-in or otherwise standard classes, ranging from
-truly primitive classes such as {\tt numeric} to the more complex {\tt
-  data.frame}.  We have already used the latter in {\bf L1}. We will ignore
-those classes and thus highlight user-defined ones.  As for attributes,
-again some patterns arise naturally, for instance all matrices have a {\tt
-  dims} attribute (indeed, it's this attribute that makes a matrix a
-matrix).  Attributes are name-value pairs, to keep things simple we reduce
-these to names. The only attribute that we treat specially is the {\tt
-  class} for those we will recurse and capture the set of values defined by
-the user (these are the class names for the object).  Thus, we extend {\bf
-  L1} with attribute lists and classes: \attrclass{T}{C}{A} where {\tt T}
-can be any L1 type other than data frame or matrix, and {\tt C} and {\tt A}
-are lists of classes and attribute names respectively.  The subtype relation is defined as:
-\attrclass{T}{C}{A} $<:_{\mathbf{L2}}$ \xspace \attrclass{\dot T}{C'}{A'} if 
-$ T <:_{\mathbf{L1}} \dot T$, $C \subseteq C'$, and $A \subseteq A'$.
+{\bf L2} extends the previous system with support for classes and attributes.
+While there are a finite number of results from the \code{typeof} function, this is {\it not} the case for \code{class} and \code{attributes}, as classes and attributes {\it can} be user-defined. 
+In terms of classes, R has a number of built-in or otherwise standard classes, ranging from truly primitive classes such as {\tt numeric} to the more complex {\tt data.frame}. 
+{\bf L1} already incorporates these builtin classes, namely through its data frame and matrix types, so we will ignore those classes and thus highlight those which are user-defined.  
+As for attributes, again some patterns arise naturally in R, e.g.
+all matrices have a {\tt dims} attribute, but even these are somewhat interesting, as a programmer could in theory set their own {\it dims} attribute on an arbitrary value.
+The only attribute we ignore is the {\tt class} attribute, as we collect the class of values separately (and the {\tt class} attribute, if present, is always equal to the class of a value).
+Attributes are name-value pairs, and to keep things simple we consider only attribute names.
+
+In sum, we extend {\bf L1} with attribute lists and classes: \attrclass{T}{C}{A} where {\tt T} can be any L1 type other than data frame or matrix, and {\tt C} and {\tt A} are lists of classes and attribute names respectively.
+The subtype relation is defined as: \attrclass{T}{C}{A} $<:_{\mathbf{L2}}$ \xspace \attrclass{\dot T}{C'}{A'} if $ T <:_{\mathbf{L1}} \dot T$ (i.e.
+are subtypes in the sense of {\bf L1}), $C \subseteq C'$, and $A \subseteq A'$.
 

### 2019-04-05 paper/main.tex
moved poly examples, filled in and edited descriptions

@@
 
+Before presenting the results of our analysis, we will quickly survey examples of real, polymorphic R functions that we have analyzed, to build up a global sense of how types manifest themselves in current R code.
+
+%
+%
+%
+%
+%
+%
+\section{Examples of Polymorphism}
+\label{sec:polyex}
+
+In this section, we will present an eclectic mix of R functions which exhibit polymorphic behavior at {\bf L0}.
+
+We begin by referring the reader to Figure~\ref{fig:realex}, which showcases the compatibility of integers and doubles.
+Coercion from integer to double is performed automatically where appropriate (e.g., here when a double and integer are added together), and likewise coercion from double to integer is automatically performed when integers are strictly required (e.g., here when doubles are being used to index a vector or list).
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+> 5L + 1L # => 6L, an integer
+> 5L + 1.2 # => 6.2, a double (5L coerced to 5.0)
+> c(10, 20, 30)[1.2] # => 10, 1.2 coerced to 1L
+\end{lstlisting}}\caption{Example of \sD, \sI type usage.}\label{fig:realex}\end{figure}
+
+Figure~\ref{fig:optnull} shows an example of a function with an optional NULL argument. 
+In the function, {\tt x} is a sorted vector of income values, and {\tt w} is a vector of weights with default value NULL.
+The function computes fractional ranks required in other package functions.
+We see that if {\tt w} is NULL, then default unit weights are generated.
+This pattern of default NULL (or NA) arguments is very prevalent in R.
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+frac.ranks <- function(x, w=NULL) {
+  if (is.null(w)) w <- rep(1, length(x)) # if no weights passed, take all weights = 1
+  ...
+\end{lstlisting}}\caption{Example of optional argument (from {\tt acid} package).}\label{fig:optnull}\end{figure}
+
+Figure~\ref{fig:listvec} shows an example of a function with list and vector
+polymorphism.  The function takes a list or vector {\tt point} and a data
+frame {\tt polyg} representing a polygon.  In it, we see that {\tt point} is explicitly converted
+to a vector (via the \code{as.numeric} function) regardless of its type.
+This showcases the ease with which lists and vectors may be converted in R (the corresponding vector-to-list function is \code{as.list}).
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+is.point.inside <- function (point, polyg) {
+    p <- as.numeric(point) # as.numeric(list(1, 2)) => c(1, 2)
+    ...
+\end{lstlisting}}\caption{Example of \l/\D argument (from {\tt bivrp} package).}\label{fig:listvec}\end{figure}
+
+Figure~\ref{fig:charclos} shows an example of a function with character and
+function polymorphism.  At a high level, this function
+calculates point estimates for an {\tt angmcmc} object (specific to some
+packages).  The caller specifies {\tt fn}, which is either a function {\it or} the name of a function which will be evaluated on the object samples to estimate parameters.
+If {\tt fn} is a string, the function lookup is performed by \code{match.fun}.
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+pointest <- function (object, fn = mean, par.name, comp.label, chain.no,  ...) {
+    ...
+    if (is.character(fn))
+        if (fn == "MODE" | fn == "MAP")   do_MAP <- TRUE
+        else {
+            do_MAP <- FALSE
+            fn <- match.fun(fn) # looks up the function name
+        }
+\end{lstlisting}}\caption{Example of \sC/\sF argument (from {\tt BAMBI} package).}\label{fig:charclos}\end{figure}
+
+In Figure~\ref{fig:dfdbl}, we see a function that, among other things, takes in some data ({\tt dat}) and a
+character string ({\tt spss}) specifying either {\tt "in"} or {\tt "out"}.
+Here, the type of {\tt dat} is {\it dependent} on the value of {\tt spss}:
+if {\tt spss == "in"}, {\tt dat} must be a data frame, and otherwise if {\tt spss == "out"}, {\tt dat} must be a vector.
+Our analysis does not capture this granularity of dependency, as we only collect type information.
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+nret.translator <- function(dat, spss="out", ...) {
+  ...
+  if(identical(spss, "out")) {
+    if(!is.vector(dat))  stop(simpleError("...'dat' must be a vector!"))
+    ...
+  } else {
+    ...
+    # here, dat is data.frame
+    items.idx <- items.idx[order(names(dat[, items.idx]))]
+  }
+\end{lstlisting}}\caption{Example of a \df/\D argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}
+
+In Figure~\ref{fig:chardbl}, we see a function which takes in a list ({\tt network}), a vector of indices of that list ({\tt fixIndices}), and a
+vector of values ({\tt values}).
+The locations specified by {\tt fixIndicies} in {\tt network} are updated with {\tt values}.
+Here, {\tt fixIndicies} has been observed to be either a character or double vector:
+In R, list indices are typically doubles, but can be characters (if the list
+or vector being indexed has a {\tt names} attribute).
+Note that type level {\bf L2} can capture the presence of a {\tt names} attribute.
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+fixGenes <- function (network, fixIndices, values) {
+  ...
+  network$fixed[fixIndices] <- as.integer(values)
+  ...
+\end{lstlisting}}\caption{Example of \C/\D argument (from {\tt BoolNet} package).}\label{fig:chardbl}\end{figure}
+
+In Figure~\ref{fig:matvec}, we see that a function which can take in a vector,
+but immediately transforms it (transpose, with {\tt t}) into a matrix.  As
+an idea, matrix/vector polymorphism seems sensible, as mathematically
+vectors {\it are} one-dimensional matrices.  R echoes this by making conversion between the two
+types easy: \code{as.vector(m)} flattens a matrix \code{m} into a vector
+(e.g., \code{as.vector(matrix(2, 2, 2)) == c(2, 2, 2, 2)}), and
+\code{as.matrix(v)} builds a {\tt length(v) x 1} matrix (e.g.,
+\code{as.matrix(c(1, 2)) == matrix(data=c(1, 2))}).  \AT{TODO fix line break
+  in code at end of last sentence}
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+tee <- function (x, theta, D1, D2, phi)  {
+    if (is.vector(x)) x <- t(x)
+    ...
+\end{lstlisting}}\caption{Example of matrix/(\D) argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}
+
+Throughout this short section, we have shown real examples of polymorphic R functions.
+Our goal is to capture these instances of polymorphism across a large subset of existing R code, and to highlight and discuss patterns which arise.
+We present the results of our analysis in the following section.
+
 %
@@
 	The frequency of these signatures indicates a tendency to create ``optional arguments'', where some argument's default value might be NULL.
-	This pattern is quite frequent in R code, \AT{and there are a number of examples sprinkled throughout the paper.}
+	This pattern is quite frequent in R code, and an example can be found in Figure~\ref{fig:optnull}.
 
@@
 %
-\section{Examples of Polymorphism}
-\label{sec:polyex}
-
-First, Figure~\ref{fig:realex} shows an example of how integers and doubles
-are very compatible.  \AT{TODO add matrix (dimensions) to the example.}
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-> 5L + 1L # => 6L, an integer
-> 5L + 1.2 # => 6.2, a double (5L coerced to 5.0)
-> c(10, 20, 30)[1.2] # => 10, 1.2 coerced to 1L
-\end{lstlisting}}\caption{Example of {\it real} type usage.}\label{fig:realex}\end{figure}
-
-Figure~\ref{fig:optnull} shows an example of a function with an optional
-argument.  In the function, {\tt x} is a sorted vector of income values, and
-{\tt w} is a vector of weights, optionally NULL.  The function computes
-fractional ranks required for the computation of some coefficient
-implemented by the package.  We see that if {\tt w} is NULL, then default
-unit weights are generated.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-frac.ranks <- function(x, w=NULL) {
-  if (is.null(w)) w <- rep(1, length(x)) # if no weights passed, take all weights = 1
-  ...
-\end{lstlisting}}\caption{Example of optional argument (from {\tt acid} package).}\label{fig:optnull}\end{figure}
-
-Figure~\ref{fig:listvec} shows an example of a function with list and vector
-polymorphism.  The function takes a list or vector {\tt point} and a data
-frame {\tt polyg} representing a polygon.  In it, we see that {\tt point}
-can be either a list or a vector, and the function code casts the argument
-to a double vector.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-is.point.inside <- function (point, polyg) {
-    p <- as.numeric(point) # as.numeric(list(1, 2)) => c(1, 2)
-    ...
-\end{lstlisting}}\caption{Example of list/vector argument (from {\tt bivrp} package).}\label{fig:listvec}\end{figure}
-
-Figure~\ref{fig:charclos} shows an example of a function with character and
-function polymorphism.  \AT{Bluh.}  At a high level, this function
-calculates point estimates for an {\tt angmcmc} object (specific to some
-packages).  The caller specifies {\tt fn}, either a function or a function
-name which will be evaluated on the object samples to estimate parameters.
-The call to lookup the function name is on line 10.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-pointest <- function (object, fn = mean, par.name, comp.label, chain.no,  ...) {
-    ...
-    if (is.character(fn))
-        if (fn == "MODE" | fn == "MAP")   do_MAP <- TRUE
-        else {
-            do_MAP <- FALSE
-            fn <- match.fun(fn) # looks up the function name
-        }
-\end{lstlisting}}\caption{Example of char/closure argument (from {\tt BAMBI} package).}\label{fig:charclos}\end{figure}
-
-\AT{I'll explain this better, maybe?}  In Figure~\ref{fig:dfdbl}, we see a
-function that, among other things, takes in some data ({\tt dat}) and a
-character string ({\tt spss}) specifying either {\tt "in"} or {\tt "out"}.
-Now, depending on the value of {\tt spss}, {\tt dat} will either be a data
-frame or a vector of doubles.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-nret.translator <- function(dat, spss="out", ...) {
-  ...
-  if(identical(spss, "out")) {
-    if(!is.vector(dat))  stop(simpleError("...'dat' must be a vector!"))
-    ...
-  } else {
-    ...
-    # here, dat is data.frame
-    items.idx <- items.idx[order(names(dat[, items.idx]))]
-  }
-\end{lstlisting}}\caption{Example of a data.frame/(\D) argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}
-
-In Figure~\ref{fig:chardbl}, we see a function which takes in a list ({\tt network}), a vector of indices of that list ({\tt fixIndices}), and a
-vector of values ({\tt values}).
-The locations specified by {\tt fixIndicies} in {\tt network} are updated with {\tt values}.
-Here, {\tt fixIndicies} has been observed to be either a character or double vector:
-In R, list indices are typically doubles, but can be characters (if the list
-or vector being indexed has a {\tt names} attribute).
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-fixGenes <- function (network, fixIndices, values) {
-  ...
-  network$fixed[fixIndices] <- as.integer(values)
-  ...
-\end{lstlisting}}\caption{Example of character/double argument (from {\tt BoolNet} package).}\label{fig:chardbl}\end{figure}%$
-In Figure~\ref{fig:matvec}, we see that the function can take in a vector,
-but immediately transforms it (transpose, with {\tt t}) into a matrix.  As
-an idea, matrix/vector polymorphism seems sensible, as mathematically
-vectors are matrices.  R echoes this by making conversion between the two
-``types'' easy: \code{as.vector(m)} flattens a matrix \code{m} into a vector
-(e.g., \code{as.vector(matrix(2, 2, 2)) == c(2, 2, 2, 2)}), and
-\code{as.matrix(v)} builds a {\tt length(v) x 1} matrix (e.g.,
-\code{as.matrix(c(1, 2)) == matrix(data=c(1, 2))}).  \AT{TODO fix line break
-  in code at end of last sentence}
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-tee <- function (x, theta, D1, D2, phi)  {
-    if (is.vector(x)) x <- t(x)
-    ...
-\end{lstlisting}}\caption{Example of matrix/(\D) argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}
-%
-%
-%
 %

### 2019-04-05 paper/main.tex
analysis intro

@@
     ...
-\end{lstlisting}}\caption{Example of matrix/(\D) argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}
+\end{lstlisting}}\caption{Example of \M{D}/\D argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}
 
@@
 
-This section presents our results for each of the levels we have considered.
-\AT{New numbers will be available tonight (Apr 5).}
-\AT{Will use that to fill in some blanks.}
+In this section, we will present the results of our analysis at each of our type levels, describing the types of polymorphism that we have observed and shedding light on the frequency of polymorphic patterns in R.
+
+Before we begin, we define the kinds of polymorphism we will consider.
+In choosing how to express the polymorphicity of a function, there are two main options:
+First, having several separate whole-function signatures, e.g.
+\D $\rightarrow$ \D and \C $\rightarrow$ \D, or alternatively expressing the polymorphism at the argument level with a union type, e.g.
+(\D $\cup$ \C) $\rightarrow$ \C.
+We refer to the former as {\it per-function} or {\it function-level polymorphism}, and to the latter as {\it per-argument} or {\it argument-level polymorphism}.
+
+We believe that both of these notions of polymorphism are valuable for different reasons.
+Per-argument polymorphism highlights the type-based behavior of function arguments themselves, and helps to describe the type-based behavior of all R values.
+For instance, we observe that many arguments have (\sC, \sF) as a signature, from which we deduce that many R programmers are taking advantage of functions such as \code{match.fun("fun\_name")} to call functions by name.
+On the other hand, per-function polymorphism highlights how R programmers interact with functions as a whole, and allows us to discern more subtle patterns.
+For example, observing two function signatures such as \D \to \D and \sF \to \sF may lead us to ascribe a more useful and precise type for the function than if we had simply collapsed the argument and return types (e.g., $X$ \to $X$ vs.
+\D $\cup$ \sF \to \D $\cup$ \sF).
+
+The layout of this section is as follows:
+In order, we present analysis results for {\bf L0}, {\bf L1}, and {\bf L2}, speaking to both per-argument and per-function polymorphism at each level.
+Where appropriate, we digress to highlight interesting tidbits of tangential information which help us understand the style of code that R programmers write.
+We begin with {\bf L0}.
 
@@
 Given that, we find it unsurprising that the function is so polymorphic.
-
-In choosing how to express the polymorphicity of a function, there are two main options one might consider:
-First, having several separate whole-function signatures, e.g.
-\D $\rightarrow$ \D and \C $\rightarrow$ \D, or alternatively expressing the polymorphism at the argument level with a union type, e.g.
-(\D $\cup$ \C) $\rightarrow$ \C.
-The former is referred to as {\it per-function} or {\it function-level polymorphism}, and the latter as {\it per-argument} or {\it argument-level polymorphism}.
-Earlier in Table~\ref{tab:L0top10arg} we saw the most frequently occurring argument-level signatures, and Figure~\ref{fig:L0funcounts} broke our corpus down by function-level polymorphism.
 

### 2019-04-05 paper/main.tex
automatic subtying discussion

@@
 
+\subsection{Automatic Subtyping}
+\label{sec:autosub}
+
+When analyzing our data, we perform a certain amount of subtyping automatically to minimize the footprint of obviously equivalent signatures.
+This subtyping is performed at the function level of granularity.
+For example, if a function is observed to have two signatures \sD \to \df to \D \to \df, we collapse the signature to \D \to \df as \sD $<:$ \D and \df $<:$ \df.
+We would not collapse \sD, \I \to \df and \D, \sI \to \df, as \I $\not <:$ \sI and \D $\not<:$ \sD.
+
 Before presenting the results of our analysis, we will quickly survey examples of real, polymorphic R functions that we have analyzed, to build up a global sense of how types manifest themselves in current R code.

### 2019-04-05 paper/main.tex
working on discussion, have some new numbers that i will deal with soon, so the final results should be up in a few hours

@@
 One of our chief goals is to generate useful function signatures.
-After tracing and capturing unique signatures from function call sites, \AT{recall} that we collapse some of the more obvious ones together.
+After tracing and capturing unique signatures from function call sites, recall from Section~\ref{sec:autosub} that we collapse some of the more obvious ones together.
 After this, we are curious to know: {\it how many functions have more than one signature?}
@@
 
-\AT{TODO do function signatures maybe? Maybe just an example that we can run with.
-Plots also.}
+\AT{TODO simple running example}
 
@@
 \end{table}
-
-\AT{Will have new numbers by tonight.}
 
@@
 
+Throughout Section~\ref{sec:results}, we have thoroughly explored the trends and patterns of polymorphism observed in our analysis of a large subset of existing R code.
+By now, we have an understanding of how each of our type levels captures the polymorphic patterns that exist in R.
+
+\AT{DRAFT}
+
+In this section, we aim to comment globally on the things we learned during our analysis, and propose a type system which is reasonable given its coverage of existing code.
+
+We will look at argument-level polymorphism, and then function level polymorphism.
+
+%
+%
+%
+%
+\subsection{Argument-Level Polymorphisms}
+
+From our analysis, we discovered a number of interesting patterns at the argument level.
+To summarize:
+
+\begin{itemize}
+
+	\item {\bf scalars}: according to our analysis, many instances of vectorized values are always scalar.
+	This is strong evidence that an annotation denoting a scalar value would be useful to R programmers, and further such an annotation could be used by future implementations to statically determine when not to vectorize a vectorizable type.
+
+	\item {\bf missing}: our analysis has found a number of instances of argument positions remaining unevaluated across all calls to a function.
+	A small annotation, such as a tag on an argument position, would capture this behavior and could be used to warn programmers if they try to use the unevaluated value, and also warn them if an operation might evaluate it accidentally.
+	Techniques to avoid forcing promises in R are not well understood by most programmers, and such an annotation would be useful to assist them in avoiding unwanted forcing of promises.
+
+	\item {\bf classes}: in transitioning from {\bf L1} to {\bf L2}, we found that the prevalence of \l, \lT{X} polymorphism decreased significantly.
+	In many cases, these lists share a class, and so the signature which was prominent in {\bf L1} was broken down into several less prominent {\bf L2} signatures with added class information.
+	Classes could also be used to capture the usage of dates (two of the standard date formats have class ``Date'' and ``POSIXct, POSIXt''), and could help package writers enforce usage of package functions.
+
+\end{itemize}
+
+Generating the code to dynamically enforce these signatures is also possible, with varying levels of complexity.
+Simple checks include checking the class of a value, or checking for the presence of certain attributes.
+R makes it easy to exit a function prematurely, and in fact many functions already exhibit this behavior.
+For example: \AT{example}.
+
+We also saw that, despite being far more specific and descriptive than previous levels, the {\bf L2} type system didn't bring about an undue increase in the amount of overall polymorphism in the language.
+This signifies that {\bf L2} captures the vast majority of already existing polymorphic patterns, and very few functions would need to resort to an \ANY annotation, and indeed many arguments are really \ANY-typed (e.g., \%$>$\% from magrittr and the prototype assignment operator {\tt \$<-} from proto).
+
 %
@@
 
-\AT{Retool this into possible meta discussion after results}
-
-Now, we will turn our attention to whole function polymorphism.
-
-This is different because we can now look at argument polymorphism across several arguments, like we did when thinking about character/real polymorphism.
-
-We're interested to see what kinds of ``meta-annotations'' we can make at the function level.
+\AT{Working through this now.}
+
+Now we wonder: {\it what kinds of function-level annotations are useful?}
+
 For instance, should we link argument types with return types?
@@
 # [2] x: {double -> double, x: list<double> -> list<double>}
-\end{lstlisting}}\caption{Possible function annotations.}\label{fig:funanno}\end{figure}
-\AT{Say more about this?}
+\end{lstlisting}}\caption{Simple identity function.}\label{fig:funanno}\end{figure}
 
@@
 %
-\subsection{Coverage}
-
-\AT{What is the coverage of these new annotations?}
+%\subsection{Coverage}
+%
+%\AT{Manual check of some functions.}
 

### 2019-04-06 paper/main.tex
touched up discussion of per fun poly

@@
 
-\AT{Working through this now.}
-
-Now we wonder: {\it what kinds of function-level annotations are useful?}
-
-For instance, should we link argument types with return types?
-Consider Figure~\ref{fig:funanno}.
+We saw that argument-level signatures and polymorphism are insufficient to capture some interesting patterns at the function level.
+For instance, consider Figure~\ref{fig:funanno}.
+
 \begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
@@
 
-Harkening back to our discussion of character/real polymorphism, we mentioned that we might want to consider such a type signature to be a list or vector index only if a list or vector was present as another argument.
-Now, this may be moving well beyond the purview of R programmers, but perhaps there is value in such a scenario, particularly if we include even {\it more} information in a type system.
-For example, say we have a function which takes in a 2D point list.
-There are a plethora of useful annotations, here, and of varying degrees of granularity.
-We might want simply a "list" annotation, or we might want a "list<double>" annotation to be more specific.
-In addition, though, we might want to know if the list has a names attribute, yielding "list<double>@names".
-Then, perhaps it would be useful to know what the names were, so "list<double>@names\{x, y\}".
-
-Now, depending on the granularity of individual argument annotations, we might want to link certain arguments together.
-Consider the code snippet in Figure~\ref{fig:indexof}.
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-access <- function(l, i) l[[i]]
-
-L <- list(a=1, b=2, c=3)
-access(L, 2) # => 2
-access(L, "c") # => 3
-
-# possible annotations:
-# [1] l: list, i: {character, real} -> real
-# [2] l: list<double>, i: {character, real} -> real
-# [3] l: list<X>, i: {character, real} -> X
-# [4] l: list<double>@names, i: index -> double
-# [5] l: list<double>@names{a, b, c}, i: index@indexOf(l) -> double
-\end{lstlisting}}\caption{Linking annotations.}\label{fig:indexof}\end{figure}
-The annotations could be used to generate code to check for compliance, like a contract.
+In certain cases, language implementations could benefit from knowing the return type of a function when the function is fed certain arguments.
+In Figure~\ref{fig:funanno}, we can assign the function a more sophisticated signature if we don't simply collapse the observed argument signatures element-wise.
+This gets more interesting with attributes, as we might want to annotate a function to communicate that it adds an attribute to whatever the list of attributes of its input value.
+
+One drawback of function-level polymorphism is that functions with many arguments tend to be sensitive to small changes in argument types, which inflates the number of unique signatures observed.
+We saw this in Figure~\ref{fig:L2funcounts}, where the distribution of signatures per function was roughly in line with those from Figures~\ref{fig:L0funcounts} and~\ref{fig:L1funcounts} but the peak (i.e., the function with the most signatures) was far more pronounced.
+That said, overall the number of polymorphic functions did not increase by too large a margin considering the added expressivity of {\bf L2}'s types.
 

### 2019-04-06 paper/main.tex
have new data, will process now. also put some stuff in where the conclusion should be

@@
 %
-%
+%  ls trace_export_promises > promise_packages_FINAL.txt
 %
@@
 \section{Discussion}
-
-\AT{Short section to recap the analysis results, discuss trends, consolidate our suggestions, talk about argument vs. function poly.}
 
@@
 
-\AT{DRAFT}
-
-In this section, we aim to comment globally on the things we learned during our analysis, and propose a type system which is reasonable given its coverage of existing code.
-
+In this section, we aim to comment globally on the things we learned during our analysis, and propose a type system for R which is supported by empirical evidence that it captures the patterns of existing R code.
 We will look at argument-level polymorphism, and then function level polymorphism.
@@
 
+To determine which kinds of whole function signatures would be useful, a more sophisticated analysis is required, which would need to develop a correlation between function arguments and return types.
+There may also be other interesting per-function trends which our current analysis does not pick up on.
+We aim to make our data available for this purpose.
+
 %
@@
 \section{Related Work}
-\AT{Not sure what the best place for this is, but we should discuss these papers somewhere.}
+
 The idea of measuring polymorphism via some program analysis isn't new.
 For instance, some work~\cite{aakerblom2015measuring} ... .
+
 More generally, there is an existing literature on analyzing usage patterns
-of language features.  For instance, the dynamic features of Smalltalk were
-analyzed in some work~\cite{callau2011howdevelopers}, ...
+of language features.  
+For instance, the dynamic features of Smalltalk were analyzed in some work~\cite{callau2011howdevelopers}, ...
+
 Other analysis work~\cite{milojkovic2017duck} ... .
+
+%
+%. nohup parallel -j55 -a promise_reduce_me.txt Rscript mar_new_processing_files/promise_processed_reduced_to_new_form.R &
+%
+%
+%
+%
 \section{Conclusions and Future Work}
+
+Ours is a community which strives to improve the lives of programmers, and understanding programmers is key to ensure that our vision aligns with their needs.
+Increasingly, programming is becoming accessible to a wider audience, with many domain-specific languages being developed to address the concerns of specialists such as analysts, mathematicians, and data scientists.
+Designing systems and languages to meet the specific needs of these domain experts can prove to be challenging, as our own intuitions as (often) classically-trained programmers may fail us, and in many cases they know their own needs better than we do.
+
+One particularly interesting language which falls into this category, and indeed the object of this work, is R.
+R is a language made by statisticians, for statisticians. 
+The language itself is fascinating: it is a vectorized, dynamically
+typed, lazy, and functional, with limited side-effects, extensive
+reflective facilities and retrofitted object-oriented programming support.
+Understanding how these disparate and sometimes adversarial features interact is critical to improving or otherwise further developing the language.
+
+In our work, we built up an understanding of R by analyzing the usage of types across a large corpus of R code.
+With this knowledge, we highlighted some interesting patterns of type usage, and developed a number of type systems with differing levels of granularity to describe and capture the use of types in the language.
+This task was complicated by the many notions of type in R, as well as the lack of non-package R code, but we persevered:
+leveraging the \genthat tool, we analyzed tests, examples, and vignettes of said packages as a close approximation for how these packages would in fact be used by programmers.
+Our final type system, {\bf L2}, consolidated information from 3 of R's main reflective functions (\code{typeof}, \code{class}, and \code{attributes}), as well as some extra information about certain values (e.g.
+the type of list elements, and the length of vectorized types) to be as precise as reasonable.
+
+We found that a type system such as {\bf L2} captures an overwhelming majority (>90\%) of argument-level type usage, and in it a vast majority ($\approx$80\%) of functions were monomorphic.
+Among other interesting data points, we discovered that many argument positions with a vectorized type in R are monomorphic {\it and strictly scalar}, suggesting that types or annotations for scalar-only values is useful.
+
+\AT{Need a somewhat stronger conclusion.}
+
+%
+%
+%
 \bibliographystyle{boilerplate/ACM-Reference-Format}
@@
 \end{document}
-
-%
-% L1+ Stuff? We don't need this anymore.
-% 
-% 
-%
-%At this point, we have seen a number of reasonable polymorphic patterns that arise in R, both in L0 and L1.
-%The next type system design we consider will deal with these, and we refer to it as L$1^{+}$.
-%The number of monomorphic arguments and functions can be found in Tables~\ref{tab:argcountsL1p} and~\ref{tab:funcountsL1p} respectively.
-%
-%% latex table generated in R 3.5.1 by xtable 1.8-3 package
-%% Fri Mar 22 18:03:30 2019
-%\begin{table}[ht]
-%\label{tab:argcountsL1p}
-%\centering
-%\begin{tabular}{lrr}
-%  \hline
-% Argument Polymorphism & Count & Percentage \\
-%  \hline
-%  Monomorphic in Type & 869K & 95.01 \\
-%%  Monomorphic in Class & 831K & 90.31 \\
-%%  Monomorphic in Attribute Pattern & 830K & 90.25 \\
-%  Total Seen & 914K & --- \\
-%     \hline
-%\end{tabular}
-%\caption{Account of {\it argument} polymorphism in L$1^{+}$.}
-%\end{table}
-%
-%% latex table generated in R 3.5.1 by xtable 1.8-3 package
-%% Fri Mar 22 18:10:46 2019
-%\begin{table}[ht]
-%\label{tab:funcountsL1p}
-%\centering
-%\begin{tabular}{lrr}
-%  \hline
-% Function Polymorphism & Count & Percentage \\
-%  \hline
-%  Monomorphic in Type & 185K & 85.30 \\
-%%  Monomorphic in Class & 163K & 75.13 \\
-%%  Monomorphic in Attribute Pattern & 161K & 74.31 \\
-%  Total Seen & 216K & --- \\
-%   \hline
-%\end{tabular}
-%\caption{Account of {\it function} polymorphism in L$1^{+}$.}
-%\end{table}
-%
-%We had some big wins here, with only ~5\% of argument polymorphism unaccounted for.
-%To get an idea of what more we can do, let's see what the most frequent remaining polymorphic signatures are.
-%See Table~\ref{tab:toppolyL1p}.
-%
-%% latex table generated in R 3.5.1 by xtable 1.8-3 package
-%% Tue Mar 26 17:46:45 2019
-%\begin{table}[ht]
-%\label{tab:toppolyL1p}
-%\centering
-%\begin{tabular}{lrr}
-%  \hline
-%Polymorphic Type Signature & \# Occurrences & \% of Total Polymorphism \\
-%  \hline
-%  matrix, \realv & 6360 & 13.94 \\
-%  list$<$any$>$, list$<$real$>$ & 2799 & 6.13 \\
-%  \C, \realv & 2774 & 6.08 \\
-%  \sC, \reals & 2552 & 5.59 \\
-%  data.frame, matrix & 2167 & 4.75 \\
-%  list$<$any$>$, list$<$list$>$ & 1496 & 3.28 \\
-%  list$<$any$>$, list$<$character$>$ & 1214 & 2.66 \\
-%  data.frame, \realv & 1046 & 2.29 \\
-%  \sL, \reals & 920 & 2.02 \\
-%  \sC, \realv & 868 & 1.90 \\
-%   \hline
-%\end{tabular}
-%\caption{Top 10 argument signatures in L$1^{+}$.}
-%\end{table}
-%
-%A few of the top polymorphic signatures stand out.
-%First, the matrix/\realv signature seems interesting, converting between vectors and matrices is very simple, and indeed numeric vectors are mathematically {\tt 1 x n} matrices.
-%In fact, in R matrices can be used identically to vectors in many cases, and we refer the reader to Figure~\ref{fig:matasvec} for examples.
-%\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-%v <- c(1, 2, 3, 4)
-%ma <- matrix(data=v, nrow=4, ncol=1)
-%v[2] == m[2] == m[2, 1]
-%
-%mb <- matrix(data=v, nrow=2, ncol=2)
-%v[3] == m[3] == m[1, 2]
-%\end{lstlisting}}\caption{Using a matrix like a vector.}\label{fig:matasvec}\end{figure}
-%
-%This similarity is reminiscent of another pattern which was more clear in Table~\ref{tab:L0toppoly}; that of list/vector polymorphism.
-%Transforming between vectors and matrices is just as easy as transforming between lists and vectors is simple.
-%\AT{We should figure out how we want to deal with this polymorphism.}

### 2019-04-06 notebooks/data_builder.Rmd
update

@@
 # load the data
-psp <- read_rds("../data/per_sig_partial.RDS")
+psp <- read_rds("../data/promise_big_final.RDS")
+
+psp <- as.tibble(psp)
 
@@
 psp <- psp[!psp$pkg == "no_data",]
+
+# first, turn specials, closures, and builtins to functions
+psp <- transform_new_df_w_TS_map(psp, type_map_to_fun)
 
@@
 ```{r}
-# first, turn specials, closures, and builtins to functions
-psp <- transform_new_df_w_TS_map(psp, type_map_to_fun)
-
-# any other transformations?
-```
-
-```{r}
 # make it T0
-psp_L0 <- transform_new_df_w_TS_map(psp, type_map_T0_to_r)
+psp_L0 <- transform_new_df_w_TS_map(psp, type_map_T0_to_r, unparam = T)
+psp_L0 <- lists_to_chars_sig_df(psp_L0)
 ```
@@
 # write it to file for Java processing
-write_csv(psp_L0, "../data/partial_L0.csv")
+write_csv(psp_L0, "../data/L0.csv")
 ```
@@
 # read in new one
-psp_L0 <- read_csv("../data/partial_L0_subtype.csv.gz")
-p
+psp_L0 <- read_csv("../data/L0_subtype.csv.gz")
+
 # remove SUBTYPE rows
@@
 # write L1 (which is just psp) to file for Java
-write_csv(psp, "../data/partial_L1.csv")
+write_csv(psp, "../data/L1.csv")
 
@@
 # read it back
-psp <- read_csv("../data/partial_L1_subtype.csv.gz")
+psp <- read_csv("../data/L1_subtype.csv.gz")
 
@@
 
-psp_col <- read_csv("../data/partial_L1_collapsed.csv.gz")
-psp_L0_col <- read_csv("../data/partial_L0_collapsed.csv.gz")
+psp_L0_col <- read_csv("../data/L0_collapsed.csv.gz")
+psp_col <- read_csv("../data/L1_collapsed.csv.gz")
+
 ```
@@
 # again, Java
-L0_arg_sig_c <- read_csv("../data/partial_L0_arg_sig_counts.csv")
-L1_arg_sig_c <- read_csv("../data/partial_L1_arg_sig_counts.csv")
+L0_arg_sig_c <- read_csv("../data/L0_arg_sig_counts.csv.gz")
+L1_arg_sig_c <- read_csv("../data/L1_arg_sig_counts.csv.gz")
 
@@
 # rm []
-L0_arg_sig_c[1,"count"] <- 0
-L1_arg_sig_c[1,"count"] <- 0
+# L0_arg_sig_c[1,"count"] <- 0
+# L1_arg_sig_c[1,"count"] <- 0
+L0_arg_sig_c <- L0_arg_sig_c[2:nrow(L0_arg_sig_c), ]
+L1_arg_sig_c <- L1_arg_sig_c[2:nrow(L1_arg_sig_c), ]
 
 # re sort
-L0_arg_sig_c[order(-L0_arg_sig_c$count),] -> L0_arg_sig_c
-L1_arg_sig_c[order(-L1_arg_sig_c$count),] -> L1_arg_sig_c
+# L0_arg_sig_c[order(-L0_arg_sig_c$count),] -> L0_arg_sig_c
+# L1_arg_sig_c[order(-L1_arg_sig_c$count),] -> L1_arg_sig_c
 
@@
 
+# if want top 10
+plot_L0_df %>% tail
+
 plot_L1_df %>%
@@
 ggsave("../paper/plots/L1_by_fun.pdf",width=10,height=4)
+
+plot_L1_df %>% tail
 ```
@@
 # L2 Time
-psp_L2 <- read_csv("../data/partial_L1_L2.csv.gz")
+psp_L2 <- read_csv("../data/L1_L2.csv.gz")
 psp_L2 <- psp_L2[!psp_L2$pkg == "SUBTYPE",]
@@
 
+# gonna remove 2 outliers and talk about them separately
+plot_L2_df <- plot_L2_df[1:(nrow(plot_L2_df)-2), ]
+
+# THE L2 PLOT
+
 plot_L2_df %>%
@@
 ggsave("../paper/plots/L2_by_fun.pdf",width=10,height=4)
+
+plot_L2_df %>% tail
 ```
@@
 # ...
-list_of_L2_sig_c <- read_csv("../data/partial_L1_arg_sig_counts.csv.gz")
+list_of_L2_sig_c <- read_csv("../data/L1_arg_sig_counts_L2.csv.gz")
 
@@
 
-list_of_L2_sig_c$perc_tot <- round(list_of_L2_sig_c$count / sum(list_of_L2_sig_c$count) * 100, 1)
+# remove last
+list_of_L2_sig_c <- list_of_L2_sig_c[1:nrow(list_of_L2_sig_c)-1,]
+
+list_of_L2_sig_c$perc_tot <- round(list_of_L2_sig_c$count / sum(list_of_L2_sig_c$count) * 100, 2)
 
@@
 # Arg Poly Count L2
-data.frame(`Signature Type`=c("Monomorphic", "Total Seen"), Count=c(list_of_L2_sig_c$count %>% sum - L2_poly_sig$count %>% sum, list_of_L2_sig_c$count %>% sum), Percentage=c(paste0(round(100 * (1 - (L2_poly_sig$count %>% sum / list_of_L2_sig_c$count %>% sum)), 1), "%"), "---")) %>% xtable %>% print(include.rownames=F)
+data.frame(`Signature Type`=c("Monomorphic", "Total Seen"), Count=c(list_of_L2_sig_c$count %>% sum - L2_poly_sig$count %>% sum, list_of_L2_sig_c$count %>% sum), Percentage=c(paste0(round(100 * (1 - (L2_poly_sig$count %>% sum / list_of_L2_sig_c$count %>% sum)), 2), "%"), "---")) %>% xtable %>% print(include.rownames=F)
 ```
@@
 
-poly_count_normalized %>% arrange(-poly_norm) %>% slice(2:6)
+poly_count_normalized %>% arrange(-poly_norm) %>% slice(1:10) %>% xtable %>% print(include.rownames=F)
 

### 2019-04-06 paper/main.tex
update

@@
 \newcommand{\NUMPOLYTYPEMONOCLASS}{1,357\xspace}
-\newcommand{\PERCSCALARMONO}{37.6\%\xspace}
-
-\newcommand{\LZEROPERCPOLY}{17\%\xspace}
-\newcommand{\LONEPERCPOLY}{20.4\%\xspace}
-\newcommand{\LTWOPERCPOLY}{23\%\xspace}
+\newcommand{\PERCSCALARMONO}{38.2\%\xspace}
+
+\newcommand{\LZEROPERCPOLY}{19\%\xspace}
+\newcommand{\LONEPERCPOLY}{18.1\%\xspace}
+\newcommand{\LTWOPERCPOLY}{21.3\%\xspace}
 
@@
   \hline
-  \D & 69806 & 26.3\% \\ 
-  \l\ & 55311 & 20.8\% \\ 
-  \C & 27789 & 10.5\% \\ 
-  \L & 26230 & 9.9\% \\ 
-  \sF & 16806 & 6.3\% \\ 
-  \sN & 13976 & 5.3\% \\ 
-  \ANY & 9125 & 3.4\% \\ 
-  \I & 8971 & 3.4\% \\ 
-  \sS & 4748 & 1.8\% \\ 
-  \D, \I & 3256 & 1.2\% \\ 
+  \D & 269K & 29.7\% \\ 
+  \l & 193K & 21.3\% \\ 
+  \C & 96K & 10.6\% \\ 
+  \L & 83K & 9.2\% \\ 
+  \sF & 60K & 6.6\% \\ 
+  \sN & 47K & 5.2\% \\ 
+  \ANY & 31K & 3.4\% \\ 
+  \I & 30K & 3.2\% \\ 
+  \sS & 16K & 1.7\% \\ 
+  \D, \l & 10K & 1.1\% \\ 
+   \hline
    \hline
@@
 The magrittr package's pipe \%$>$\% function is an example of a function which performs a lot of metaprogramming; we will discuss it in a little more detail shortly.
-
-One of our chief goals is to generate useful function signatures.
-After tracing and capturing unique signatures from function call sites, recall from Section~\ref{sec:autosub} that we collapse some of the more obvious ones together.
-After this, we are curious to know: {\it how many functions have more than one signature?}
-Figure~\ref{fig:L0funcounts} has the answer with a breakdown by function.
+Interestingly, a polymorphic attribute signature \D, \l is among the top 10 overall---this is the vector and list polymorphism (shown in Figure~\ref{fig:listvec}) made manifest.
+
+Now that we've seen the most common attribute-level signatures, we will turn our attention to function-level polymorphism.
+Indeed, one of our goals is to generate useful function signatures, so we ask: {\it how many functions have more than one recorded signature?}
+Figure~\ref{fig:L0funcounts} has the answer, with a breakdown by function.
 
@@
 That said, there are a few functions with a disproportionately large number of signatures.
-One such function is the magrittr package's pipe \%$>$\%, which is used to implement a kind of currying (as per discussion in Section~\ref{sec:Impl}).
+One such function is the magrittr package's pipe \%$>$\%, which is used to implement a kind of currying (as per discussion in Section~\ref{sec:Impl})---here, \%$>$\% has nearly 100 unique signatures!
 Many of the metaprogramming functions in the Tidyverse are quite sophisticated, and are designed with the explicit goal of being very usable.
@@
 
-Another function with many signatures is the data.table package's [ function, which is overloaded for objects with data.table class.
+Another function with many signatures is the data.table package's [ function, which is a reimplementation of the square bracket operator for data.tables.
 This function has 16 formal arguments, and call configurations vary widely.
@@
 
-To get an idea of the most common forms of argument-level polymorphism, we refer toTable~\ref{tab:L0toppoly}.
+Having had a taste for function-level polymorphism, we will now explore common patterns in argument-level polymorphism.
+For this, refer to Table~\ref{tab:L0toppoly}.
 
@@
 \hline
-  \D, \l & 2799 & 1\% \\ 
-  \D, \sN & 2290 & 0.9\% \\ 
-  \C, \sN & 1527 & 0.6\% \\ 
-  \l, \sN & 1193 & 0.4\% \\ 
-  \C, \D & 858 & 0.3\% \\ 
-  \D, \sF & 413 & 0.2\% \\ 
-  \C, \I & 382 & 0.1\% \\ 
-  \I, \sN & 378 & 0.1\% \\ 
-  \C, \l & 371 & 0.1\% \\ 
-  \C, \sF & 334 & 0.1\% \\ 
+  \D, \l & 9566 & 1.1\% \\ 
+  \D, \sN & 8801 & 1\% \\ 
+  \C, \sN & 5590 & 0.6\% \\ 
+  \l, \sN & 5016 & 0.6\% \\ 
+  \C, \D & 2887 & 0.3\% \\ 
+  \D, \sF & 1608 & 0.2\% \\ 
+  \l, \sF & 1345 & 0.1\% \\ 
+  \C, \l & 1283 & 0.1\% \\ 
+  \C, \sF & 1168 & 0.1\% \\ 
+  \C, \I & 1104 & 0.1\% \\ 
+   \hline
 \end{tabular}
@@
 	\item \D, \sF: this signature appears more strange than it is.
-	Many models, processes, simulations, etc. that are prevalent throughout R are are {\it parameterized}.
-	Some arguments with the \D, \sF signature are either parameters for a function (\D), or the function itself already constructed (\sF).
+	Many models, processes, simulations, etc. that are prevalent throughout R have parameters to fine-tune their behavior.
+	This pattern manifests itself here: some arguments with the \D, \sF signature are either parameters for a function (\D), or the function itself already constructed (\sF).
 	Others, such as the {\tt br} argument of the {\tt rtree} function in the {\tt ape} package, are either parameters for complex data generation processes (\D), or a function to be used during the generation (\sF).
-	It would appear that this pattern is used to make complex simulation, modelling, or data generation functions more flexible for programmers.
+	It would appear that this pattern is used to make complex simulation, modelling, or data generation functions more flexible for programmers by allowing them to supply parameters or the parameterized function itself.
 
@@
   \hline
-  Monomorphic & 250K & 94\% \\ 
-  Total Seen & 266K & --- \\ 
+Monomorphic & 852K & 94\% \\ 
+  Total Seen & 906K & --- \\ 
    \hline
@@
 We also saw in Table~\ref{tab:L0toppoly} that each of the most frequently occurring argument signatures had some explanation.
-That said, L0 is not without limitations.
-
-The main issues with L0 stem from a lack of adequate information, as the type system doesn't capture precise list types, and does not distinguish scalars from vectors.
+
+That said, L0 is not without limitations, and the main issues stem from a lack of adequate information, as the type system doesn't capture precise list types, and does not distinguish scalars from vectors.
 The next type system we will explore, L1, tackles these issues by collecting more information to build more precise types.
@@
   \hline
-  \sD & 48815 & 18.4\% \\ 
-  \D & 27367 & 10.3\% \\ 
-  \sL & 24284 & 9.2\% \\ 
-  \sC & 22025 & 8.3\% \\ 
-  \l & 20062 & 7.6\% \\ 
-  \sF & 18480 & 7\% \\ 
-  \M{D} & 14315 & 5.4\% \\ 
-  \sN & 14016 & 5.3\% \\ 
-  \ANY & 9093 & 3.4\% \\ 
-  \C & 7731 & 2.9\% \\
+  \sD & 179K & 19.7\% \\ 
+  \D & 97K & 10.7\% \\ 
+  \sL & 77K & 8.6\% \\ 
+  \sC & 75K & 8.3\% \\ 
+  \l & 67K & 7.4\% \\ 
+  \sF & 61K & 6.7\% \\ 
+  \sN & 50K & 5.5\% \\ 
+  \M{D} & 49K & 5.5\% \\ 
+  \ANY & 31K & 3.4\% \\ 
+  \df & 29K & 3.2\% \\ 
    \hline
@@
 From Table~\ref{tab:L1top10arg}, we see that all of the most frequent argument signatures are monomorphic even though we increased the precision of the types.
-We see that scalars actually figure quite prominently, which is encouraging, supporting our suggestion of an explicit scalar annotation.
+We see that scalars actually figure quite prominently, which is encouraging: this supports an idea that an explicit scalar annotation or type may be useful {\it for programmers}---indeed, such an annotation would be helpful for compiler optimizations (e.g., scalar-only values need not have vectors allocated to store them).
 We also see that matrices (\M{D}) are quite popular, a testament to the versatility of the data structure, as they can easily be converted to vectors and data frames.
@@
 
-We see that we increased the polymorphism with our transition to L1, despite the monomorphic nature of the most frequently appearing argument signatures.
+We see that we barely increased the polymorphism with our transition to L1, despite the far richer type information.
+The biggest outlier here is indeed \%$>$\%, as was the case in {\bf L0}.
+Otherwise, the shape is very similar.
+In short: even though {\L1} types are more expressive, the breakdown of per-function polymorphism is largely unchanged from {\bf L0}.
+
 As we were in L0, we are curious to know how argument-level polymorphism looks in L1.
@@
   \hline
-  \D, \M{D} & 1440 & 0.5\% \\ 
-  \l, \lT{D} & 525 & 0.2\% \\ 
-  \M{D}, \M{I} & 383 & 0.1\% \\ 
-  \sC, \sD & 382 & 0.1\% \\ 
-  \I, \sD & 370 & 0.1\% \\ 
-  \l, \lT{list} & 337 & 0.1\% \\ 
-  \df, \M{D} & 305 & 0.1\% \\ 
-  \C, \D & 294 & 0.1\% \\ 
-  \sD, \sF & 290 & 0.1\% \\ 
-  \sC, \sF & 273 & 0.1\% \\
+  \D, \M{D} & 4848 & 0.5\% \\ 
+  \l, \lT{D} & 1843 & 0.2\% \\ 
+  \sC, \sD & 1590 & 0.2\% \\ 
+  \M{D}, \M{I} & 1502 & 0.2\% \\ 
+  \l, \lT{list} & 1199 & 0.1\% \\ 
+  \df, \M{D} & 1180 & 0.1\% \\ 
+  \C, \D & 1019 & 0.1\% \\ 
+  \sD, \sF & 1005 & 0.1\% \\ 
+  \I, \sD & 997 & 0.1\% \\ 
+  \sC, \sF & 991 & 0.1\% \\ 
    \hline
@@
 
+	\item \sC, \sD and \C, \D: we discussed this in L0---this signature likely captures an index or a date.
+
 	\item \M{D}, \M{I}: this signature is reasonable, as any computation in R on \M{D} would work on \M{I} and vice versa.
-
-	\item \sC, \sD and \C, \D: we discussed this in L0---this signature likely captures an index or a date.
-
-	\item \I, \sD: subtyping between primitives does not hold between scalars and vectors (i.e., \I $\not<:$ \sD) even though it's likely that the programmer intended for the argument to be a vector of the supertype (here, these arguments were likely intended to be of type \D).
 
@@
 
-	\item \sD, \sF and \sC, \sF: these were thoroughly discussed in the context of L0.
+	\item \sD, \sF and \sC, \sF: these were discussed in the context of L0.
+
+	\item \I, \sD: subtyping between primitives does not hold between scalars and vectors (i.e., \I $\not<:$ \sD) even though it's likely that the programmer intended for the argument to be a vector of the supertype (here, these arguments were likely intended to be of type \D).
 
@@
   \hline
-  Monomorphic & 252K & 94.9\% \\ 
-  Total Seen & 265K & --- \\ 
+Monomorphic & 864K & 95.3\% \\ 
+  Total Seen & 906K & --- \\ 
    \hline
@@
 
-In spite of drastically increasing the precision of our types, we are still happily only dealing with a small amount of polymorphism.
+In spite of drastically increasing the precision of our types, we have actually {\it reduced} the amount of polymorphism.
 This suggests that with a good notion of subtyping, precise type annotations won't cause an undue increase in the polymorphism of the language.
@@
 
-Our suggestion to add an explicit scalar type is not unfounded, though it is not compatible with current R implementations.
+It seems that there is a place for a scalar annotation in R.
 We saw in this section that strictly scalar values appear quite frequently, making up some of the most frequently occurring argument signatures.
@@
 Throughout this section, we have seen that even though we substantially increased the precision of our type annotations, we have not increased the amount of polymorphism that we observe.
-The transition from L0 to L1 revealed some interesting patterns that L1 cannot quite capture---for these, we will need to appeal to the class and attributes of values.
-The next type system we consider is L2, and is designed to achieve this.
+The transition from L0 to L1 revealed some interesting patterns that L1 cannot quite capture---for these, we will need to appeal to the class and attributes of values, which are incorporated into {\bf L2}, discussed next.
 
@@
   \hline
-  \sD & 46938 & 17.6\% \\ 
-  \sL & 24113 & 9\% \\ 
-  \sC & 21387 & 8\% \\ 
-  \D & 20900 & 7.8\% \\ 
-  \sF & 14556 & 5.5\% \\ 
-  \sN & 14016 & 5.3\% \\ 
-  \ANY & 9093 & 3.4\% \\ 
-  \attrclass{\D}{}{dim} & 6974 & 2.5\% \\ 
-  \C& 6939 & 2.6\% \\ 
-  \attrclass{\l}{}{names} & 6137 & 2.2\% \\ 
+ \sD & 172K & 19\% \\ 
+  \sL & 77K & 8.5\% \\ 
+  \D & 77K & 8.4\% \\ 
+  \sC & 73K & 8\% \\ 
+  \sN & 50K & 5.5\% \\ 
+  \sF & 45K & 5\% \\ 
+  \ANY & 31K & 3.4\% \\ 
+  \attrclass{\D}{}{dim} & 25K & 2.6\% \\ 
+  \C & 25K & 2.6\% \\ 
+   \attrclass{\l}{}{names, row.names}  & 21K & 2.2\% \\ 
    \hline
@@
 The inclusion of class and attribute information has changed the counts slightly.
-One tidbit of interest here is that monomorphic named lists are more common than their unnamed counterparts.
-The lack of classes is explained by our elimination of builtin classes, and it's doubtful that a user-defined class would be so pervasive as to feature in this breakdown.
+Not pictures here, but interesting: named lists are more frequent than their unnamed counterparts.
+No classes are represented here since (1) we eliminated builtin classes, as we've used them to construct types already, and (2) most other classes are specific to some package or set of packages---it's doubtful that such a class would be so pervasive as to feature in this breakdown.
 
 Now that we have a sense for argument-level type signatures in {\bf L2}, we will see just how many functions are polymorphic.
-Now, {\bf L2} builds directly off of {\bf L1}, and as such at the argument level the type-based polymorphism shouldn't change.
-For this account of polymorphism, we deem an argument to be polymorphic if it was polymorphic in type, class, or attribute pattern.
 Consider Figure~\ref{fig:L2funcounts}.
@@
 \end{figure}
+
+Not pictured here are two outlier values: the summary and print functions of the base R package.
+Summary is a generic function which is used to summarize various types of data, and print is used to convert values to string format and print them to the console.
+These functions really {\it should} be outliers, as they are really operating over any type.
+They feature so prominently here since many of the values which flowed into them had some amount of attribute or class information.
+
+Overall, the amount of polymorphism only slightly increased, even though we added two new dimensions of type information.
+Again, far richer types do not increase the amount of polymorphism by significant amounts, perhaps a testament to the monomorphic nature of many R functions.
 
@@
   \hline
-  \D, \attrclass{\D}{}{dim} & 553 & 0.2\% \\ 
-  \sC, \sD & 336 & 0.1\% \\ 
-  \I, \sD & 320 & 0.1\% \\ 
-  \attrclass{\l}{}{names}, \attrclass{\D}{}{names} & 269 & 0.1\% \\ 
-  \D, \attrclass{\D}{}{dim, dimnames} & 218 & 0\% \\ 
-  \sD, \sF & 173 & 0\% \\ 
-  \attrclass{\l}{}{names, row.names}, \attrclass{\D}{}{dim, dimnames} & 167 & 0\% \\ 
-  \attrclass{\D}{}{names}, \attrclass{\D}{}{dim, dimnames} & 155 & 0\% \\ 
-  \C, \I & 149 & 0\% \\ 
+  \D, \attrclass{\D}{}{dim} & 1942 & 0.2\% \\ 
+  \sC, \sD & 1431 & 0.2\% \\ 
+  \attrclass{\l}{}{names}, \attrclass{\D}{}{names}  & 1046 & 0.1\% \\ 
+  \I, \sD & 855 & 0.1\% \\ 
+  \D, \attrclass{\D}{}{dim, dimnames} & 810 & 0\% \\ 
+  \attrclass{\l}{}{names, row.names}, \attrclass{\D}{}{dim, dimnames} & 708 & 0.1\% \\ 
+  \C, \D & 672 & 0\% \\ 
+   \attrclass{\D}{}{names}, \attrclass{\D}{}{dim, dimnames} & 591 & 0\% \\ 
+  \attrclass{\D}{}{dim}, \attrclass{\I}{}{dim}  & 445 & 0\% \\ 
+  \sD, \attrclass{\sF}{}{srcref} & 445 & 0\% \\ 
    \hline
@@
 
-When comparing this table with Table~\ref{tab:toppolyL1}, we see that the distribution of signature frequencies is more uniform, as attributes and classes differentiated some signatures which were not distinct in {\bf L1}.
+When comparing this table with Table~\ref{tab:toppolyL1}, we see that by-and-large the same types are represented.
+We note the {\tt names} attributes on the lists in row 3: these named lists are, in general, more common than their unnamed counterparts.
+In the 10th signature, we see a function with attribute {\tt srcref}: functions with this attribute are for dealing with open files in the context of the srcfile set of functions in the base package.
+As for why such a function would share an argument position with scalar double, there are other representations of open files, and numbers feature among them.
 
@@
   \hline
-  Monomorphic & 248K & 93.4\% \\ 
-  Total Seen & 265K & --- \\ 
+  Monomorphic & 851K & 93.9\% \\ 
+  Total Seen & 906K & --- \\ 
    \hline
@@
   \hline
-  knitLatex & 4 & 41 & \LaTeX \xspace generator from R data \\ 
-  magrittr & 17 & 126 & Tidyverse pipe suite \\ 
-  proto & 5 & 30 & Prototype object-based programming \\ 
-  gnlm & 10 & 53 & Regression fit package \\ 
-  Formula & 17 & 82 & ?? \\ 
+  magrittr & 20 & 225 & Tidyverse pipe suite \\ 
+  oshka & 4 & 38 & quoted expression expansion \\ 
+  scatterD3 & 1 & 9 & create D3 (JS) scatter plots from R \\ 
+  yaml & 4 & 33 & convert R to YAML \\ 
+  knitLatex & 4 & 32 & \LaTeX \xspace generator from R data \\ 
    \hline
 \end{tabular}
-\caption{5 most polymorphic packages (in {\bf L0}).}
+\caption{5 of the most polymorphic packages (in {\bf L0}).}
 \end{table}
@@
 Now, what makes these so polymorphic?
-
-\AT{TODO if time: fix to score per fun, then tally per package.}
-
+The first package, magrittr, is one that we discussed earlier in the context of its pipe \%$>$\% function. 
+It's not surprising that the package as a whole is quite polymorphic, as it implements {\it several} different pipe functions, e.g.
+\%$<>$\% which pipes as normal and then assigns to the left-hand argument.
+In fact, all but knitLatex implement some form of metaprogramming:
+oshka meta
 \begin{enumerate}
@@
 	These functions both have over 15 arguments, and there are a large number of recorded configurations, with many of the arguments being optional.
-
-	\item magrittr: we discussed magrittr earlier in the context of its pipe \%$>$\% function. 
-	It's not surprising that the package as a whole is quite polymorphic, as it implements {\it several} different pipe functions, e.g.
-\%$<>$\% which pipes as normal and then assigns to the left-hand argument.
-
-	\item proto: most of the polymorphism in this package comes from its implementation of assignment ({\tt \$<-}) for proto class objects.
-	It's a small, 3 argument function which is exceedingly polymorphic, with over 20 unique call signatures.
-	As theoretically any R object can have a proto class, it stands to reason that this function ought to be quite polymorphic indeed.
 
@@
 
-To determine which kinds of whole function signatures would be useful, a more sophisticated analysis is required, which would need to develop a correlation between function arguments and return types.
+Another note regarding function signatures is the {\tt ...} function argument.
+Functions with such an argument can be passed an arbitrary number of extra arguments, see Figure~\ref{fig:dotdotdot} for a small example.
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+f <- function(x, ...) {
+  g(x, ...)
+}
+
+f(42, z=2, y=4) # in f, will perform g(42, z=2, y=4)
+
+\end{lstlisting}}\caption{Simple ... example.}\label{fig:dotdotdot}\end{figure}
+
+Our analysis ignores these extra arguments, instead flagging functions which have them.
+If building an annotation for such a function, the ... argument is always ascribed the \l type.
+
+Ultimately, to determine which kinds of whole function signatures would be useful, a more sophisticated analysis is required, which would need to develop a correlation between function arguments and return types.
 There may also be other interesting per-function trends which our current analysis does not pick up on.
@@
 \section{Conclusions and Future Work}
+
+OUTLINE
 

### 2019-04-06 paper/main.tex
smol update

@@
 In fact, all but knitLatex implement some form of metaprogramming:
-oshka meta
-\begin{enumerate}
-
-	\item knitLatex: two functions are largely responsible for this level of polymorphism: sTab, which produces \LaTeX \xspace supertabular environments, and lTab, which produces \LaTeX \xspace longtable environments.
-	These functions both have over 15 arguments, and there are a large number of recorded configurations, with many of the arguments being optional.
-
-	\item
-
-	\item
-
-\end{enumerate}
+oshka metaprograms in the same style as Tidyverse, scatterD3 metaprograms on plot objects, and yaml converts R expressions to YAML expressions.
+This is a trend we have observed: {\it the most polymorphic functions in R are metaprogramming their arguments}.
 

### 2019-04-06 paper/main.tex
conclusion WIP

@@
 In fact, all but knitLatex implement some form of metaprogramming:
-oshka metaprograms in the same style as Tidyverse, scatterD3 metaprograms on plot objects, and yaml converts R expressions to YAML expressions.
+the oshka package metaprograms in the same style as Tidyverse, the scatterD3 package metaprograms on plot objects, and the yaml package converts R expressions to YAML expressions.
 This is a trend we have observed: {\it the most polymorphic functions in R are metaprogramming their arguments}.
@@
 It's also unclear how useful a type system would be if it {\it did} capture everything.
-Indeed, many of the packages discussed here are really designed to take in anything for particular arguments (e.g., left-hand side of \%$>$\%), and restricting these functions should not be the goal of any static type system designed after a language has matured.
+Indeed, many of the packages discussed here are really designed with an \ANY-typed argument in mind (e.g., left-hand side of \%$>$\%), and restricting these functions should not be the goal of any static type system designed for a language after it has matured.
 
@@
 %
-%. nohup parallel -j55 -a promise_reduce_me.txt Rscript mar_new_processing_files/promise_processed_reduced_to_new_form.R &
+%.
 %
@@
 
-OUTLINE
-
-Ours is a community which strives to improve the lives of programmers, and understanding programmers is key to ensure that our vision aligns with their needs.
-Increasingly, programming is becoming accessible to a wider audience, with many domain-specific languages being developed to address the concerns of specialists such as analysts, mathematicians, and data scientists.
-Designing systems and languages to meet the specific needs of these domain experts can prove to be challenging, as our own intuitions as (often) classically-trained programmers may fail us, and in many cases they know their own needs better than we do.
-
-One particularly interesting language which falls into this category, and indeed the object of this work, is R.
-R is a language made by statisticians, for statisticians. 
-The language itself is fascinating: it is a vectorized, dynamically
-typed, lazy, and functional, with limited side-effects, extensive
-reflective facilities and retrofitted object-oriented programming support.
-Understanding how these disparate and sometimes adversarial features interact is critical to improving or otherwise further developing the language.
+Truly understanding how programmers use a programming language is critical to building usable and effective language features.
+When working on a general purpose language such as JavaScript, we as programmers can appeal to our own intuitions, as many of us have had at least some interaction with the language.
+This intuition does not extend to domain-specific languages (DSLs), which are not really designed for members of our community.
+DSLs are gaining in popularity as more and more domain-experts, who may not have a traditional programming background, are exposed to the practice of programming, and the needs of these experts differ from our own.
+
+%Ours is a community which strives to improve the lives of programmers, and understanding programmers is key to ensure that our vision aligns with their needs.
+%Increasingly, programming is becoming accessible to a wider audience, with many domain-specific languages being developed to address the concerns of specialists such as analysts, mathematicians, and data scientists.
+%Designing systems and languages to meet the specific needs of these domain experts can prove to be challenging, as our own intuitions as (often) classically-trained programmers may fail us, and in many cases they know their own needs better than we do.
+%
+%One particularly interesting language which falls into this category, and indeed the object of this work, is R.
+%R is a language made by statisticians, for statisticians. 
+%The language itself is fascinating: it is a vectorized, dynamically
+%typed, lazy, and functional, with limited side-effects, extensive
+%reflective facilities and retrofitted object-oriented programming support.
+%Understanding how these disparate and sometimes adversarial features interact is critical to improving or otherwise further developing the language.
 

### 2019-04-06 notebooks/data_builder.Rmd
just dumping current state of paper

@@
 # maybe only plot those with more than 1 sig?
-plot_L0_df %>%
-  ggplot(aes(x=pkgfun_o, y=num_sigs, group=1)) +
-  geom_line() + labs(title = "Number of Function Signatures per Function (L0)",
-                                    x = "",
-                                    y = "Number of L0 Signatures") + theme(axis.text.x = element_blank())
+plot_L0_df %>% # filter(num_sigs > 1) %>%
+  ggplot(aes(x=num_sigs)) +
+  geom_histogram(binwidth = 0.5) + labs(title = "Number of Function Signatures per Function (L0)",
+                                    x = "Number of Unique Signatures",
+                                    y = "Number of Functions") + scale_x_sqrt(breaks=c(1, 2, 4, 8, 16, 32, 64)) + scale_y_sqrt(breaks=c(1, 10, 100, 1000, 10000, 100000), labels=c("1", "10", "100", "1K", "10K", "100K"))
+
+# plot_L0_df %>%
+#   ggplot(aes(x=pkgfun_o, y=num_sigs, group=1)) +
+#   geom_line() + labs(title = "Number of Function Signatures per Function (L0)",
+#                                     x = "",
+#                                     y = "Number of L0 Signatures") + theme(axis.text.x = element_blank())
 
@@
 
-plot_L1_df %>%
-  ggplot(aes(x=pkgfun_o, y=num_sigs, group=1)) +
-  geom_line() + labs(title = "Number of Function Signatures per Function (L1)",
-                                    x = "",
-                                    y = "Number of L1 Signatures") + theme(axis.text.x = element_blank())
+plot_L1_df %>% # filter(num_sigs > 1) %>%
+  ggplot(aes(x=num_sigs)) +
+  geom_histogram(binwidth = 0.5) + labs(title = "Number of Function Signatures per Function (L1)",
+                                    x = "Number of Unique Signatures",
+                                    y = "Number of Functions") + scale_x_sqrt(breaks=c(1, 2, 4, 8, 16, 32, 64, 128, 256)) + scale_y_sqrt(breaks=c(1, 10, 100, 1000, 10000, 100000), labels=c("1", "10", "100", "1K", "10K", "100K"))
 
@@
 
-plot_L2_df %>%
-  ggplot(aes(x=pkgfun_o, y=num_sigs, group=1)) +
-  geom_line() + labs(title = "Number of Function Signatures per Function (L2)",
-                                    x = "",
-                                    y = "Number of L2 Signatures") + theme(axis.text.x = element_blank())
+plot_L2_df %>% # filter(num_sigs > 1) %>%
+  ggplot(aes(x=num_sigs)) +
+  geom_histogram(binwidth = 0.5) + labs(title = "Number of Function Signatures per Function (L2)",
+                                    x = "Number of Unique Signatures",
+                                    y = "Number of Functions") + scale_x_sqrt(breaks=c(1, 2, 4, 8, 16, 32, 64, 128, 256, 512)) + scale_y_sqrt(breaks=c(1, 10, 100, 1000, 10000, 100000), labels=c("1", "10", "100", "1K", "10K", "100K"))
 
@@
 
-poly_count_normalized %>% arrange(-poly_norm) %>% slice(1:10) %>% xtable %>% print(include.rownames=F)
+poly_count_normalized %>% arrange(-poly_norm) %>% filter(count>1) %>% slice(1:10) %>% xtable %>% print(include.rownames=F)
 

### 2019-04-06 paper/main.tex
just dumping current state of paper

@@
 
-Unevaluated arguments in R are sometimes {\it carefully} used by their functions to implement some form of metaprogramming.
+Unevaluated arguments in R are sometimes {\it carefully} used by their functions to implement some form of meta-programming.
 Consider the {\tt magrittr} package, which implements a pipe function \%$>$\% that implements a flavor of function currying:
@@
 
-At {\bf L0}, the type system is mostly aligned with the types of the core language.  We
-have observed over \AT{how many?} monomorphic functions; this accounts for \AT{what \%?} of our
-corpus.
+At {\bf L0}, the type system is mostly aligned with the types of the core language.
 
@@
 The \ANY type is not an R type per se: we ascribe such a type to arguments which were {\it always} unevaluated across all calls.
-As it happens, many R functions metaprogram on some of their arguments, and this metaprogramming does not force the argument promise, and if this metaprogramming is always performed in every recorded call we will ascribe the \ANY type, as we cannot say what the argument {\it is} without possibly forcing a promise.
-The magrittr package's pipe \%$>$\% function is an example of a function which performs a lot of metaprogramming; we will discuss it in a little more detail shortly.
+As it happens, many R functions meta-program on some of their arguments, and this meta-programming does not force the argument promise, and if this meta-programming is always performed in every recorded call we will ascribe the \ANY type, as we cannot say what the argument {\it is} without possibly forcing a promise.
+The magrittr package's pipe \%$>$\% function is an example of a function which performs a lot of meta-programming; we will discuss it in a little more detail shortly.
 Interestingly, a polymorphic attribute signature \D, \l is among the top 10 overall---this is the vector and list polymorphism (shown in Figure~\ref{fig:listvec}) made manifest.
@@
 One such function is the magrittr package's pipe \%$>$\%, which is used to implement a kind of currying (as per discussion in Section~\ref{sec:Impl})---here, \%$>$\% has nearly 100 unique signatures!
-Many of the metaprogramming functions in the Tidyverse are quite sophisticated, and are designed with the explicit goal of being very usable.
+Many of the meta-programming functions in the Tidyverse are quite sophisticated, and are designed with the explicit goal of being very usable.
 This usability maps to polymorphism in our analysis, so it's no surprise that a Tidyverse function features so prominently here.
@@
 
-\AT{TODO simple running example}
-
 %
@@
 Summary is a generic function which is used to summarize various types of data, and print is used to convert values to string format and print them to the console.
-These functions really {\it should} be outliers, as they are really operating over any type.
-They feature so prominently here since many of the values which flowed into them had some amount of attribute or class information.
+These functions have over 1,000 unique signatures each, which is within reason as they are really operating over any type.
+They feature so prominently in {\bf L2} (and not in {\bf L1} or {\bf L0}) since many of the values which flowed into them had some amount of attribute or class information, and indeed these functions are very frequently exercised by other packages which may implement package-specific classes.
 
@@
 \%$<>$\% which pipes as normal and then assigns to the left-hand argument.
-In fact, all but knitLatex implement some form of metaprogramming:
-the oshka package metaprograms in the same style as Tidyverse, the scatterD3 package metaprograms on plot objects, and the yaml package converts R expressions to YAML expressions.
-This is a trend we have observed: {\it the most polymorphic functions in R are metaprogramming their arguments}.
+In fact, all but knitLatex implement some form of meta-programming:
+the oshka package meta-programs in the same style as Tidyverse, the scatterD3 package meta-programs on plot objects, and the yaml package converts R expressions to YAML expressions.
+This is a trend we have observed: {\it the most polymorphic functions in R are meta-programming their arguments}.
 
@@
 In this section, we aim to comment globally on the things we learned during our analysis, and propose a type system for R which is supported by empirical evidence that it captures the patterns of existing R code.
-We will look at argument-level polymorphism, and then function level polymorphism.
-
-%
-%
-%
-%
-\subsection{Argument-Level Polymorphisms}
-
-From our analysis, we discovered a number of interesting patterns at the argument level.
-To summarize:
-
-\begin{itemize}
-
-	\item {\bf scalars}: according to our analysis, many instances of vectorized values are always scalar.
-	This is strong evidence that an annotation denoting a scalar value would be useful to R programmers, and further such an annotation could be used by future implementations to statically determine when not to vectorize a vectorizable type.
-
-	\item {\bf missing}: our analysis has found a number of instances of argument positions remaining unevaluated across all calls to a function.
-	A small annotation, such as a tag on an argument position, would capture this behavior and could be used to warn programmers if they try to use the unevaluated value, and also warn them if an operation might evaluate it accidentally.
-	Techniques to avoid forcing promises in R are not well understood by most programmers, and such an annotation would be useful to assist them in avoiding unwanted forcing of promises.
-
-	\item {\bf classes}: in transitioning from {\bf L1} to {\bf L2}, we found that the prevalence of \l, \lT{X} polymorphism decreased significantly.
-	In many cases, these lists share a class, and so the signature which was prominent in {\bf L1} was broken down into several less prominent {\bf L2} signatures with added class information.
-	Classes could also be used to capture the usage of dates (two of the standard date formats have class ``Date'' and ``POSIXct, POSIXt''), and could help package writers enforce usage of package functions.
-
-\end{itemize}
-
-Generating the code to dynamically enforce these signatures is also possible, with varying levels of complexity.
-Simple checks include checking the class of a value, or checking for the presence of certain attributes.
-R makes it easy to exit a function prematurely, and in fact many functions already exhibit this behavior.
-For example: \AT{example}.
-
-We also saw that, despite being far more specific and descriptive than previous levels, the {\bf L2} type system didn't bring about an undue increase in the amount of overall polymorphism in the language.
-This signifies that {\bf L2} captures the vast majority of already existing polymorphic patterns, and very few functions would need to resort to an \ANY annotation, and indeed many arguments are really \ANY-typed (e.g., \%$>$\% from magrittr and the prototype assignment operator {\tt \$<-} from proto).
+We will first look at function level polymorphism before moving on to argument-level polymorphism.
 
@@
 
-We saw that argument-level signatures and polymorphism are insufficient to capture some interesting patterns at the function level.
+There are some function-level polymorphic patterns that cannot be captures solely with polymorphic function arguments.
 For instance, consider Figure~\ref{fig:funanno}.
@@
 # [1] x: {double, list<double>} -> {double, list<double>}
-# [2] x: {double -> double, x: list<double> -> list<double>}
+# [2] x: {double -> double, list<double> -> list<double>}
 \end{lstlisting}}\caption{Simple identity function.}\label{fig:funanno}\end{figure}
 
-In certain cases, language implementations could benefit from knowing the return type of a function when the function is fed certain arguments.
-In Figure~\ref{fig:funanno}, we can assign the function a more sophisticated signature if we don't simply collapse the observed argument signatures element-wise.
+Sophisticated dependent type signatures such as these could benefit language implementations in a myriad of ways.
+For instance, a language implementation could perform type-based optimization in the context of a function return simply by knowing the function argument type.
+In Figure~\ref{fig:funanno}, we can assign the function a more sophisticated signature ({\tt [2]}) if we don't simply collapse the observed argument signatures element-wise ({\tt [1]}).
 This gets more interesting with attributes, as we might want to annotate a function to communicate that it adds an attribute to whatever the list of attributes of its input value.
 
-One drawback of function-level polymorphism is that functions with many arguments tend to be sensitive to small changes in argument types, which inflates the number of unique signatures observed.
-We saw this in Figure~\ref{fig:L2funcounts}, where the distribution of signatures per function was roughly in line with those from Figures~\ref{fig:L0funcounts} and~\ref{fig:L1funcounts} but the peak (i.e., the function with the most signatures) was far more pronounced.
-That said, overall the number of polymorphic functions did not increase by too large a margin considering the added expressivity of {\bf L2}'s types.
-
-Another note regarding function signatures is the {\tt ...} function argument.
+A clear drawback to function-level polymorphism is that functions with many arguments tend to be sensitive to small changes in argument types, which inflates the number of unique signatures observed.
+Even so, overall the number of polymorphic functions did not increase by too large a margin when adding and specializing types (from {\bf L0} to {\bf L1}, and from {\bf L1} to {\bf L2})---for the most part, monomorphic functions in {\bf L0} stayed monomorphic throughout.
+
+One tidbit about function signatures that we have not thoroughly discussed is the {\tt ...} function argument.
 Functions with such an argument can be passed an arbitrary number of extra arguments, see Figure~\ref{fig:dotdotdot} for a small example.
@@
 
-Our analysis ignores these extra arguments, instead flagging functions which have them.
-If building an annotation for such a function, the ... argument is always ascribed the \l type.
+Our analysis ignores these extra arguments, instead it simply flags functions which have a {\tt ...} in their definition.
+If building an annotation for such a function, the ... argument is always ascribed the \l type, as it can indeed be inhabited by anything.
 
 Ultimately, to determine which kinds of whole function signatures would be useful, a more sophisticated analysis is required, which would need to develop a correlation between function arguments and return types.
-There may also be other interesting per-function trends which our current analysis does not pick up on.
-We aim to make our data available for this purpose.
+There may also be other interesting per-function trends which our current analysis does not pick up on that are hidden in our data.
+
+%
+%
+%
+%
+\subsection{Argument-Level Polymorphism}
+
+From our analysis, we discovered a number of interesting patterns at the argument level.
+To summarize:
+
+\begin{itemize}
+
+	\item {\bf scalars}: according to our analysis, many instances of vectorized values are always scalar.
+	This is strong evidence that {\it an annotation denoting a scalar value would be useful to R programmers}, and further such an annotation could be used by future implementations to statically determine when not to vectorize a vectorizable type.
+
+	\item {\bf missing}: our analysis has found a number of instances of argument positions remaining unevaluated across all calls to a function.
+	{\it A small annotation, such as a tag on an argument position, would capture this behavior} and could be used to warn programmers if they try to use the unevaluated value, and also warn them if an operation might evaluate it accidentally.
+	Techniques to avoid forcing promises in R are not well understood by most programmers, and such an annotation would be useful to assist them in avoiding unwanted forcing of promises.
+
+	\item {\bf classes}: in transitioning from {\bf L1} to {\bf L2}, we found that the prevalence of \l, \lT{X} polymorphism decreased significantly.
+	In many cases, these lists share a class, and so the signature which was prominent in {\bf L1} was broken down into several less prominent {\bf L2} signatures with added class information.
+	Classes could also be used to capture the usage of dates (two of the standard date formats have class ``Date'' and ``POSIXct, POSIXt''), and could help package writers enforce usage of package functions.
+	R's dynamic dispatch system is class-based, so including annotations for these would help programmers determine which functions are being dispatched for which values.
+
+\end{itemize}
+
+Generating the code to dynamically enforce these signatures is also possible, with varying levels of complexity.
+Simple checks include checking the class of a value, or checking for the presence of certain attributes.
+R makes it easy to exit a function prematurely, and in fact many functions already exhibit this behavior.
+For example, see Figure~\ref{fig:stopifnot}.
+In it, the function exits if the {\tt encoding} argument is provided, and is not a character.
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+is_character <- function(x, n = NULL, encoding = NULL) {
+  ...
+  if (!is_null(encoding)) {
+    stopifnot(typeof(encoding) == "character")
+    ...
+  }
+  ...
+}
+\end{lstlisting}}\caption{Real example of premature function exit.}\label{fig:stopifnot}\end{figure}
+
+We also saw that, despite being far more specific and descriptive than previous levels, the {\bf L2} type system didn't bring about an undue increase in the amount of overall polymorphism in the language.
+This signifies that {\bf L2} captures the vast majority of already existing polymorphic patterns, and very few functions would need to resort to an \ANY annotation, and indeed many arguments are really \ANY-typed (e.g., \%$>$\% from magrittr and the prototype assignment operator {\tt \$<-} from proto).
+
+Based on our analysis of the data, we conclude that {\it a type system such as} \textbf{\emph{L2}} {\it captures the vast majority of argument and function usage in the language}.
+{\bf L2} also contains a wealth of useful information about values that could be used for optimizations and IDE tooling.
 
@@
 
+The type system we propose in this work could be incorporated into R as a {\it gradual type system}.
+Gradual type systems~\cite{SiekTaha06} aim to bridge the gap between dynamic and static typing by allowing programmers to gradually add type annotations to their programs.
+Gradually-typed languages are becoming increasingly popular, with languages such as TypeScript~\cite{typescript13} becoming more mainstream.
+...
+
 %
@@
 Truly understanding how programmers use a programming language is critical to building usable and effective language features.
-When working on a general purpose language such as JavaScript, we as programmers can appeal to our own intuitions, as many of us have had at least some interaction with the language.
+When working on a general purpose languages (such as JavaScript), we as programmers can appeal to our own intuitions, as many of us have had at least some interaction with said languages.
 This intuition does not extend to domain-specific languages (DSLs), which are not really designed for members of our community.
-DSLs are gaining in popularity as more and more domain-experts, who may not have a traditional programming background, are exposed to the practice of programming, and the needs of these experts differ from our own.
-
-%Ours is a community which strives to improve the lives of programmers, and understanding programmers is key to ensure that our vision aligns with their needs.
-%Increasingly, programming is becoming accessible to a wider audience, with many domain-specific languages being developed to address the concerns of specialists such as analysts, mathematicians, and data scientists.
-%Designing systems and languages to meet the specific needs of these domain experts can prove to be challenging, as our own intuitions as (often) classically-trained programmers may fail us, and in many cases they know their own needs better than we do.
-%
-%One particularly interesting language which falls into this category, and indeed the object of this work, is R.
-%R is a language made by statisticians, for statisticians. 
-%The language itself is fascinating: it is a vectorized, dynamically
-%typed, lazy, and functional, with limited side-effects, extensive
-%reflective facilities and retrofitted object-oriented programming support.
-%Understanding how these disparate and sometimes adversarial features interact is critical to improving or otherwise further developing the language.
-
-In our work, we built up an understanding of R by analyzing the usage of types across a large corpus of R code.
-With this knowledge, we highlighted some interesting patterns of type usage, and developed a number of type systems with differing levels of granularity to describe and capture the use of types in the language.
-This task was complicated by the many notions of type in R, as well as the lack of non-package R code, but we persevered:
-leveraging the \genthat tool, we analyzed tests, examples, and vignettes of said packages as a close approximation for how these packages would in fact be used by programmers.
-Our final type system, {\bf L2}, consolidated information from 3 of R's main reflective functions (\code{typeof}, \code{class}, and \code{attributes}), as well as some extra information about certain values (e.g.
+As more and more domain-experts, who may not have a traditional programming background, are exposed to the practice of programming, DSLs are developed to suit their needs.
+
+The R programming language is an example of this.
+R was developed to foster an easy and interactive programming experience, and the language is not very restrictive and highly expressive.
+Unfortunately, these liberties are not without a cost: debugging R code is challenging, and given how easy it is to do anything in the language, it's no surprise that it's easy to make a mistake.
+Further, R is a very slow, and the poor performance is quickly becomes prohibitive if programmers are not careful when coding analysis for large data sets.
+
+Types are a perfect way to both make the language easier to debug and make it run faster.
+For debugging, some level of static type enforcement would give programmers more direct feedback if they are misusing values.
+As for performance, type information is crucial for many optimizations performed by JITs and other language implementations.
+If we're interested in developing an easily adoptable type system for this language, we must ensure that it captures how existing code is being used.
+That is precisely what we have accomplished in this work.
+
+Throughout this paper, we have built up an understanding of types and polymorphism in R by analyzing the usage of types across a large corpus of R code.
+With this knowledge, we highlighted some interesting patterns of type usage in the language, and developed a number of type systems with differing levels of granularity to describe and capture the use of types by programmers in existing code.
+Leveraging the \genthat~\cite{issta18} tool, we analyzed tests, examples, and vignettes of over ten thousand packages on CRAN (over \PERCENTCRAN of available packages), and traced over 15 million lines of R code.
+Analyzing package examples mitigates the lack of end-user R code,
+
+\AT{Need something more, stuff commented in source.}
+
+%Our final type system, {\bf L2}, consolidated information from 3 of R's main reflective functions (\code{typeof}, \code{class}, and \code{attributes}), as well as some extra information about certain values (e.g.
 the type of list elements, and the length of vectorized types) to be as precise as reasonable.
-
-We found that a type system such as {\bf L2} captures an overwhelming majority (>90\%) of argument-level type usage, and in it a vast majority ($\approx$80\%) of functions were monomorphic.
-Among other interesting data points, we discovered that many argument positions with a vectorized type in R are monomorphic {\it and strictly scalar}, suggesting that types or annotations for scalar-only values is useful.
-
-\AT{Need a somewhat stronger conclusion.}
+%We found that a type system such as {\bf L2} captures an overwhelming majority (>90\%) of argument-level type usage, and in it a vast majority ($\approx$80\%) of functions were monomorphic.
+%Among other interesting data points, we discovered that many argument positions with a vectorized type in R are monomorphic {\it and strictly scalar}, suggesting that types or annotations for scalar-only values is useful.
 

### 2019-04-06 paper/main.tex
tables

@@
 \newcommand{\sN}{\ensuremath{\small{\mathtt N}}\xspace}     % Null
+%\renewcommand{\l}{\ensuremath{\small L<{\mathtt ?}>}\xspace}     % List
 \renewcommand{\l}{\ensuremath{\small\underline{\mathtt ?}}\xspace}     % List
@@
 question: \emph{what could a static type system for R look like?}  To answer
-that question we study the polymorphism that is present in \RLOC lines of R
+it, we study the polymorphism that is present in \RLOC lines of R
 code spread among some \PACKAGES packages, written over a
@@
 polymorphism that is present in the code. We do this for several potential
-notions of types. Our result suggest that polymorphism is important in some
+notions of types. Our results suggest that polymorphism is important in some
 key parts of the system but that relatively simple type annotations could be
@@
 
-R and its ancestor S are languages designed, implemented and maintained by
+R and its ancestor S are languages designed, implemented, and maintained by
 statisticians. Originally they were designed as glue languages, languages
@@
 
-This papers is a data-driven study of what a type system for the R language
-could look like. Longer term, our intention is to propose changes to
-language, but for any changes to be accepted by the user community, they
+This paper is a data-driven study of what a type system for the R language
+could look like. Our intention is to propose changes to the
+language, and for any changes to be accepted by the user community they
 must clearly benefit the language without endangering backwards
@@
   test generation tool named \genthat.
-\item Manual analysis of 100 functions to validate the dynamic analysis
-  results.
+\item A set of type annotations that captures how programmers use types
+in existing R code.
 \end{itemize}
@@
 functions are evaluated only when needed, they are bundled in so-called
-promises which package the original expression (as an AST), its environment
+promises which package the original expression (as an abstract syntax tree, or AST), its environment
 as well as the result of evaluating the expression. Promises can be
@@
 function. Over the years, programmers have found the need for a richer type
-structure and have added attributes. The best way to think of attributes is
+structure and have added {\it attributes}. The best way to think of attributes is
 as an optional map from name to values that can be attached to any object.
@@
 types easy: \code{as.vector(m)} flattens a matrix \code{m} into a vector
-(e.g., \code{as.vector(matrix(2, 2, 2)) == c(2, 2, 2, 2)}), and
+(e.g., \code{as.vector(matrix(1, 2))} is equivalent to \code{c(1, 1)}), and
 \code{as.matrix(v)} builds a {\tt length(v) x 1} matrix (e.g.,
-\code{as.matrix(c(1, 2)) == matrix(data=c(1, 2))}).  \AT{TODO fix line break
-  in code at end of last sentence}
+\code{as.matrix(c(1, 2))} is equivalent to \code{matrix(data=c(1, 2))}).
 
@@
 Table~\ref{tab:L0top10arg} shows the breakdown of the 10 most frequently occurring argument signatures.
+
 \begin{table}[ht]
-\label{tab:L0top10arg}
 \centering
@@
 \caption{Most frequent argument signatures in {\bf L0}.}
+\label{tab:L0top10arg}
 \end{table}
@@
 
-\begin{table}[ht]\label{tab:L0toppoly}\centering
+\begin{table}[ht]\centering
 \begin{tabular}{lrr}  \hline
@@
 \caption{Top polymorphic argument signatures for the L0 type system.}
+\label{tab:L0toppoly}
 \end{table}
@@
 
-\begin{table}[ht]\label{tab:L0argcounts}\centering
+\begin{table}[ht]\centering
 \begin{tabular}{lrl}
@@
 \caption{Argument position polymorphism in L0.}
+\label{tab:L0argcounts}
 \end{table}
@@
 \begin{table}[ht]
-\label{tab:L1top10arg}
 \centering
@@
 \caption{Most frequent argument signatures in L1.}
+\label{tab:L1top10arg}
 \end{table}
@@
 \begin{table}[ht]
-\label{tab:toppolyL1}
 \centering
@@
 \caption{Top polymorphic argument signatures in L1.}
+\label{tab:toppolyL1}
 \end{table}
@@
 \begin{table}[ht]
-\label{tab:argcountsL1}
 \centering
@@
 \caption{Argument position polymorphism in L1.}
+\label{tab:argcountsL1}
 \end{table}
@@
 \begin{table}[ht]
-\label{tab:L2top10arg}
 \centering
@@
 \caption{Most frequent argument signatures in {\bf L2}.}
+\label{tab:L2top10arg}
 \end{table}
@@
 \begin{table}[ht]
-\label{tab:toppolyL2}
 \centering
@@
 \caption{Top polymorphic argument signatures in {\bf L2}.}
+\label{tab:toppolyL2}
 \end{table}
@@
 Now that we have seen the most frequently appearing argument signatures, we ask: {\it how many argument positions are polymorphic?}.
-Table~\ref{tbl:argcountsL2} has the answer.
+Table~\ref{tab:argcountsL2} has the answer.
 
 \begin{table}[ht]
-\label{tab:argcountsL2}
 \centering
@@
 \caption{Account of {\it argument} polymorphism in {\bf L2}.}
+\label{tab:argcountsL2}
 \end{table}
@@
 \begin{table}[ht]
-\label{tab:packagepolysize}
 \centering
@@
 \caption{5 of the most polymorphic packages (in {\bf L0}).}
+\label{tab:packagepolysize}
 \end{table}
@@
 
-%  magrittr & 4.16 &  \\
-%  types & 3.25 & Simple type annotations for R \\
-%  r6extended & 3.00 & Utilities extending barebones R6 package \\
-%  utils & 2.93 & Utilities for R developers \\
-%  zeallot & 2.85 & Metaprogramming package for assignment \\
-%  listviewer & 2.72 & R list visualizer \\
-%  proto & 2.60 &  \\
-%  apercu & 2.49 & Display a short view of a complex R value (list, vector, etc.) \\
-%  pipeR & 2.46 & Another pipe suite \\
-%  rethinker & 2.44 & RethinkDB client for R \\
-
-%% latex table generated in R 3.5.1 by xtable 1.8-3 package
-%% Thu Mar 28 16:01:37 2019
-%\begin{table}[ht]
-%\label{tab:bigpolyfuns}
-%\centering
-%\begin{tabular}{llrl}
-%  \hline
-%Package & Function & Size & Description \\
-%  \hline
-%  magrittr & \%$>$\% & 23.00 & Tidyverse pipe function \\
-%  magrittr & freduce & 17.67 & Apply a list of functions sequentially \\
-%  rlang & captureArgInfo & 16.50 & Capture arg info \\
-%  rlang & enquo & 16.00 & Create quosure \\
-%  rlang & get\_expr & 15.00 & Extract expression \\
-%  data.table & replace\_dot\_alias & 13.50 & Deals with '.' in passed expr \\
-%  tibble & strip\_dim & 13.00 & Removes 'dim' attribute \\
-%  data.table & copy & 11.00 & Create copy of passed value \\
-%  htmlwidgets & shouldEval & 11.00 & Eval passed value depending on tag \\
-%  magrittr & \%$<$$>$\% & 11.00 & Like \%$>$\% but assigns \\
-%   \hline
-%\end{tabular}
-%\caption{10 most polymorphic functions.}
-%\end{table}
-
 %
@@
 %
-%  ls trace_export_promises > promise_packages_FINAL.txt
+%  
 %
@@
 Our analysis ignores these extra arguments, instead it simply flags functions which have a {\tt ...} in their definition.
-If building an annotation for such a function, the ... argument is always ascribed the \l type, as it can indeed be inhabited by anything.
+If building an annotation for such a function, the {\tt ...} argument is always ascribed the \l type, as it can indeed be inhabited by anything.
 
@@
 The idea of measuring polymorphism via some program analysis isn't new.
-For instance, some work~\cite{aakerblom2015measuring} ... .
-
-More generally, there is an existing literature on analyzing usage patterns
-of language features.  
-For instance, the dynamic features of Smalltalk were analyzed in some work~\cite{callau2011howdevelopers}, ...
+For instance, some work~\cite{aakerblom2015measuring} measures the polymorphism of Python programs, but with a slightly different goal than ours: authors seek to assess the need for expressiveness in type systems that are retrofit onto dynamic languages.
+Interestingly, they draw some similar conclusions to us: they find that monomorphism is far more frequent than polymorphism, and even find that typically very few functions are very highly polymorphic.
+
+More generally, there is an existing literature on analyzing usage patterns of language features.  
+For instance, the dynamic features of Smalltalk were analyzed in some work~\cite{callau2011howdevelopers},
 
@@
 Unfortunately, these liberties are not without a cost: debugging R code is challenging, and given how easy it is to do anything in the language, it's no surprise that it's easy to make a mistake.
-Further, R is a very slow, and the poor performance is quickly becomes prohibitive if programmers are not careful when coding analysis for large data sets.
-
-Types are a perfect way to both make the language easier to debug and make it run faster.
+Further, R is very slow, and the poor performance quickly becomes prohibitive if programmers are not careful when coding analysis for large data sets.
+
+Types are a perfect way to both make a language easier to debug and make it run faster.
 For debugging, some level of static type enforcement would give programmers more direct feedback if they are misusing values.
 As for performance, type information is crucial for many optimizations performed by JITs and other language implementations.
-If we're interested in developing an easily adoptable type system for this language, we must ensure that it captures how existing code is being used.
+If we're interested in developing an easily adoptable type system for R language, we must ensure that it captures how existing code is being used.
 That is precisely what we have accomplished in this work.
@@
 With this knowledge, we highlighted some interesting patterns of type usage in the language, and developed a number of type systems with differing levels of granularity to describe and capture the use of types by programmers in existing code.
-Leveraging the \genthat~\cite{issta18} tool, we analyzed tests, examples, and vignettes of over ten thousand packages on CRAN (over \PERCENTCRAN of available packages), and traced over 15 million lines of R code.
-Analyzing package examples mitigates the lack of end-user R code,
-
-\AT{Need something more, stuff commented in source.}
-
-%Our final type system, {\bf L2}, consolidated information from 3 of R's main reflective functions (\code{typeof}, \code{class}, and \code{attributes}), as well as some extra information about certain values (e.g.
-the type of list elements, and the length of vectorized types) to be as precise as reasonable.
-%We found that a type system such as {\bf L2} captures an overwhelming majority (>90\%) of argument-level type usage, and in it a vast majority ($\approx$80\%) of functions were monomorphic.
-%Among other interesting data points, we discovered that many argument positions with a vectorized type in R are monomorphic {\it and strictly scalar}, suggesting that types or annotations for scalar-only values is useful.
+We analyzed the tests, examples, and vignettes of over ten thousand packages on CRAN (over \PERCENTCRAN of available packages), and traced over 15 million lines of R code.
+
+We believe that a type system akin to {\bf L2} is well suited for R, as it is expressive enough to capture current use of the language without being so specific as to make everything polymorphic.
+In fact, the levels of polymorphism observed were rather low overall, with only 6\% of function arguments and 21\% of functions being polymorphic.
+Importantly, {\bf L2} carries enough information to be useful to R implementations and IDEs:
+language implementations would surely benefit from statically-available type information, and IDEs could use the rich type and attribute information to give programmers immediate feedback during the programming process.
 
@@
 \bibliographystyle{boilerplate/ACM-Reference-Format}
-\bibliography{bib/biblio,bib/jv,bib/r,bib/new}
+\bibliography{bib/biblio,bib/jv,bib/r,bib/new,bib/gradual}
 \end{document}

### 2019-04-06 paper/main.tex
small changes, edits, still working on related work

@@
 \begin{figure}[!th]{\footnotesize\begin{tabular}{@{}r||l|r|r|r|r|r@{}}\hline
-\bf Package & \bf Description & \bf R LOC &\bf C LOC &\bf Scripts & \bf Calls Observed & \bf Calls Recorded \\
+\bf Package & \bf Description & \bf R LOC &\bf C LOC &\bf Scripts & \bf Calls Observed & \bf Signatures Recorded \\
 \hline
@@
 corresponds to either one use-case or a set of unit tests. We print the
-number of function calls that were observed by our infastructure, and the
-number of unique signatures that were recorded.
+number of function calls that were observed by our infrastructure, and the
+number of unique call signatures that were recorded.
 
@@
 maximum is 19 million.  The median number of recorded signatures is 16 and
-the maxium is 8,422. These numbers are skewed by a number of scripts doing
+the maximum is 8,422. These numbers are skewed by a number of scripts doing
 very few calls before plunging into C code.
@@
 \newpage
-\section{Methods}
+\section{Methodology}
 
@@
 More generally, there is an existing literature on analyzing usage patterns of language features.  
-For instance, the dynamic features of Smalltalk were analyzed in some work~\cite{callau2011howdevelopers},
+For instance, the dynamic features of Smalltalk were analyzed in some work~\cite{callau2011howdevelopers}, which aimed to see how often the highly dynamic capabilities of Smalltalk were used and in which kinds of projects.
 
@@
 %
-\section{Conclusions and Future Work}
+\section{Conclusion}
 

### 2019-04-06 paper/main.tex
related

@@
   ...
-\end{lstlisting}}\caption{Example of \C/\D argument (from {\tt BoolNet} package).}\label{fig:chardbl}\end{figure}
+\end{lstlisting}}\caption{Example of \C/\D argument (from {\tt BoolNet} package).}\label{fig:chardbl}\end{figure}%$
 
@@
 
-The idea of measuring polymorphism via some program analysis isn't new.
-For instance, some work~\cite{aakerblom2015measuring} measures the polymorphism of Python programs, but with a slightly different goal than ours: authors seek to assess the need for expressiveness in type systems that are retrofit onto dynamic languages.
-Interestingly, they draw some similar conclusions to us: they find that monomorphism is far more frequent than polymorphism, and even find that typically very few functions are very highly polymorphic.
-
-More generally, there is an existing literature on analyzing usage patterns of language features.  
-For instance, the dynamic features of Smalltalk were analyzed in some work~\cite{callau2011howdevelopers}, which aimed to see how often the highly dynamic capabilities of Smalltalk were used and in which kinds of projects.
-
-Other analysis work~\cite{milojkovic2017duck} ... .
-
-The type system we propose in this work could be incorporated into R as a {\it gradual type system}.
-Gradual type systems~\cite{SiekTaha06} aim to bridge the gap between dynamic and static typing by allowing programmers to gradually add type annotations to their programs.
-Gradually-typed languages are becoming increasingly popular, with languages such as TypeScript~\cite{typescript13} becoming more mainstream.
-...
+The idea of measuring polymorphism via some program analysis is not new.
+For instance, \citep{aakerblom2015measuring} measured the polymorphism of
+Python programs. Their goals were related to our, they wanted to assess the
+need for expressiveness in type systems that are retrofitted onto dynamic
+languages.  Interestingly, they drew similar conclusions: monomorphism is
+far more frequent than polymorphism, and very few functions are very highly
+polymorphic.  More generally, the dynamic features of Smalltalk were
+analyzed by~\citep{callau2011howdevelopers} and duck typing was studied
+by~\citep{milojkovic2017duck}.  Adding type systems to untyped language was
+proposed by~\citep{SiekTaha06} who aim to bridge the gap between dynamic and
+static typing.  Gradually-typed languages such as
+TypeScript~\cite{typescript13} are pointing the way to what we could do
+for R.
 

### 2019-04-06 paper/main.tex
merge

--- 2019-04-06 paper/main.tex
@@
 
-In R, all arguments are passed as promises, and unused arguments will be unevaluated. 
-Our recording mechanism captures the order in which arguments are passed, can figure out which arguments were {\it not} provided (recording them as {\it missing} if the missing argument has no default value), and does {\it not} force unforced promises, recording unforced promises as {\it unevaled}.
+In R, all arguments are passed as promises, and unused arguments will remain unevaluated promises, which we record as {\it unevaled}. 
+Our recording mechanism captures the order in which arguments are passed, can figure out which arguments were {\it not} provided (recording them as {\it missing} if the missing argument has no default value), and does {\it not} force unforced promises.
 In R, trying to use missing values results in an error and thus an exception, triggering our recording phase: arguments which cause errors in this way are recorded as {\it error}.
@@
 For instance, the dynamic features of Smalltalk were analyzed in some work~\cite{callau2011howdevelopers}, which aimed to see how often the highly dynamic capabilities of Smalltalk were used and in which kinds of projects.
+Other work~\cite{pldi10a} explores how the dynamic features of JavaScript are used.
+An interesting finding of theirs is that
 

### 2019-04-06 paper/main.tex
conclusion

--- 2019-04-06 paper/main.tex
@@
 
-Truly understanding how programmers use a programming language is critical to building usable and effective language features.
-When working on a general purpose languages (such as JavaScript), we as programmers can appeal to our own intuitions, as many of us have had at least some interaction with said languages.
-This intuition does not extend to domain-specific languages (DSLs), which are not really designed for members of our community.
-As more and more domain-experts, who may not have a traditional programming background, are exposed to the practice of programming, DSLs are developed to suit their needs.
-
-The R programming language is an example of this.
-R was developed to foster an easy and interactive programming experience, and the language is not very restrictive and highly expressive.
-Unfortunately, these liberties are not without a cost: debugging R code is challenging, and given how easy it is to do anything in the language, it's no surprise that it's easy to make a mistake.
-Further, R is very slow, and the poor performance quickly becomes prohibitive if programmers are not careful when coding analysis for large data sets.
-
-Types are a perfect way to both make a language easier to debug and make it run faster.
-For debugging, some level of static type enforcement would give programmers more direct feedback if they are misusing values.
-As for performance, type information is crucial for many optimizations performed by JITs and other language implementations.
-If we're interested in developing an easily adoptable type system for R language, we must ensure that it captures how existing code is being used.
-That is precisely what we have accomplished in this work.
-
-Throughout this paper, we have built up an understanding of types and polymorphism in R by analyzing the usage of types across a large corpus of R code.
-With this knowledge, we highlighted some interesting patterns of type usage in the language, and developed a number of type systems with differing levels of granularity to describe and capture the use of types by programmers in existing code.
-We analyzed the tests, examples, and vignettes of over ten thousand packages on CRAN (over \PERCENTCRAN of available packages), and traced over 15 million lines of R code.
-
-We believe that a type system akin to {\bf L2} is well suited for R, as it is expressive enough to capture current use of the language without being so specific as to make everything polymorphic.
-In fact, the levels of polymorphism observed were rather low overall, with only 6\% of function arguments and 21\% of functions being polymorphic.
-Importantly, {\bf L2} carries enough information to be useful to R implementations and IDEs:
-language implementations would surely benefit from statically-available type information, and IDEs could use the rich type and attribute information to give programmers immediate feedback during the programming process.
+Truly understanding how programmers use a programming language is critical
+to building usable and effective language features.  When working on a
+general purpose languages (such as JavaScript), we as programmers can appeal
+to our own intuitions, as many of us have had at least some interaction with
+said languages.  This intuition does not extend to domain-specific languages
+(DSLs), which are not really designed for members of our community.
+
+The R programming language is an example of this.  R was developed to foster
+an easy and interactive programming experience. For this reason R is very
+flexible and expressive.  Unfortunately, these liberties come at a cost:
+debugging R code is challenging. Further, performance quickly becomes
+prohibitive if programmers are not careful.
+
+Types can make a language easier to debug and run faster.  For debugging,
+some level of static type enforcement would give programmers more direct
+feedback.  As for performance, type information enables optimizations. If we
+are interested in developing an easily adoptable type system for R, we must
+ensure that it captures how existing code is being used.
+
+Throughout this paper, we have built up an understanding of types and
+polymorphism in R by analyzing the usage of types across a large corpus of
+code.  We highlighted some patterns of type usage, and developed a number of
+type systems with differing levels of granularity. We analyzed the tests,
+examples, and vignettes of over ten thousand packages on CRAN (over
+\PERCENTCRAN of available packages), and traced over 15 million lines of R
+code.
+
+We believe that a type system akin to {\bf L2} is well suited for R, as it
+is expressive enough to capture current use of the language without being so
+specific as to make everything polymorphic.  In fact, the levels of
+polymorphism observed were rather low overall, with only 6\% of function
+arguments and 21\% of functions being polymorphic.  Importantly, {\bf L2}
+carries enough information to be useful to R implementations and IDEs:
+language implementations would surely benefit from statically-available type
+information, and IDEs could use the rich type and attribute information to
+give programmers immediate feedback during the programming process.
 

### 2019-04-06 paper/main.tex

--- 2019-04-06 paper/main.tex
@@
 function polymorphism.  At a high level, this function
-calculates point estimates for an {\tt angmcmc} object (specific to some
-packages).  The caller specifies {\tt fn}, which is either a function {\it or} the name of a function which will be evaluated on the object samples to estimate parameters.
+calculates point estimates for some object.
+The caller specifies {\tt fn}, which is either a function {\it or} the name of a function which will be evaluated on the object samples to estimate parameters.
 If {\tt fn} is a string, the function lookup is performed by \code{match.fun}.
@@
 Other work~\cite{pldi10a} explores how the dynamic features of JavaScript are used.
-An interesting finding of theirs is that
-
-Other analysis work~\cite{milojkovic2017duck} ... .
+An interesting finding of theirs is that only 81\% of calls are monomorphic, which is very similar to what we observed in R.
+Other analysis work~\cite{milojkovic2017duck} looks into the usage patterns of duck typing.
 
@@
 Gradually-typed languages are becoming increasingly popular, with languages such as TypeScript~\cite{typescript13} becoming more mainstream.
-...
+An eventual implementation of a type system for R will surely be inspired by this work.
 

### 2019-04-06 paper/main.tex
examples

@@
 by synthesizing unit tests from recorded function argument and return
-values, comparing candidate tests against existing ones to avoid generating tests
-which do not increase code coverage.  To suit our purposes, we change the
-existing tool in two main ways: we record \emph{shapes} rather than values,
-and ignore the code coverage optimization phase.  Both changes are
+values, comparing candidate tests against existing ones to avoid generating
+tests which do not increase code coverage.  To suit our purposes, we change
+the existing tool in two main ways: we record \emph{shapes} rather than
+values, and ignore the code coverage optimization phase.  Both changes are
 beneficial for scalability, allowing us to trace far more calls.
@@
 
-Our aim is to capture the polymorphism in R, and to do so we must define a meaningful set of basic types for the language.
-There are several possible choices, with a variety of levels of expressivity and complexity.
-For concreteness, we will pick out three different sets of types.
-To start simple, we define a set of types which closely align with R's builtin types, and later define richer notions of type which incorporate attributes, classes, and capture the various object systems of R.
-
-Throughout this paper, we will say that an argument is {\it polymorphic at} \textbf{\emph{LX}} (read: level X) if it has been called with at least two values that belong to distinct types at {\bf LX}.
-\textbf{L2} will incorporate class and attribute lists, which are slightly trickier to ``type'', and we defer discussion of this until Section~\ref{sec:L2def}.
+Our aim is to capture the polymorphism in R, and to do so we must define a
+meaningful set of basic types for the language. Also, we are interested in
+finding a type system that could be adopted, for this we aim to investigate
+the simplest possible solution.  There are several possible choices, with a
+variety of levels of expressivity and complexity.  For concreteness, we will
+pick out three different sets of types.  To start simple, we define a set of
+types which closely align with R's builtin types, and later define richer
+notions of type which incorporate attributes, classes, and capture the
+various object systems of R.
+
+Throughout this paper, we will say that an argument is {\it polymorphic at}
+\textbf{\emph{LX}} (read: level X) if it has been called with at least two
+values that belong to distinct types at {\bf LX}.  \textbf{L2} will
+incorporate class and attribute lists, which are slightly trickier to
+``type'', and we defer discussion of this until Section~\ref{sec:L2def}.
 
@@
 
-The logical ``baseline'' for the design of a type system for R is to adopt the language's builtin types.
-We dub these {\bf L0} types, and they are primarily built on the \code{typeof} values, which reflects the implementation of the type in the runtime environment.
-More specifically, the \code{typeof} a value is simply the runtime type tag of the value.
-Given that, {\bf L0} includes the following types: vectors of primitive types (integers
-\I, double \D, character \C, logical \L, raw \R, complex \X), lists of any
-type (\l), and scalar values (environment \sE, closures \sF, symbols \sY, S4
-object \sS, and null \sN). We omit short forms for some of R's more esoteric types (e.g., {\tt environment}s).
- We add the any (written \ANY)  type to stand in for unevaluated values
-Subtyping in {\bf L0} follows from the conversion rules of the language:  Logicals are subtypes of integers, which
-are themselves subtypes of doubles (so \L <: \I <: \D).  All types are subtypes of \ANY,
-thus {\tt T}<:\ANY for an type {\tt T}.
-
-%
+The logical ``baseline'' for the design of a type system for R is to adopt
+the language's builtin types.  We dub these {\bf L0} types, and they are
+primarily built on the \code{typeof} values, which reflects the
+implementation of the type in the runtime environment.  More specifically,
+the \code{typeof} a value is simply the runtime type tag of the value.
+Given that, {\bf L0} includes the following types: vectors of primitive
+types (integers \I, double \D, character \C, logical \L, raw \R, complex
+\X), lists of any type (\l), and scalar values (environment \sE, closures
+\sF, symbols \sY, S4 object \sS, and null \sN). We omit short forms for some
+of R's more esoteric types (e.g., {\tt environment}s).  We add the any
+(written \ANY) type to stand in for unevaluated values Subtyping in {\bf L0}
+follows from the conversion rules of the language: Logicals are subtypes of
+integers, which are themselves subtypes of doubles (so \L <: \I <: \D).  All
+types are subtypes of \ANY, thus {\tt T}<:\ANY for an type {\tt T}.
+
 %
@@
 
-{\bf L0} is limited in expressivity as it (1) supports polymorphism for list in
-a trivial way, (2) does not address some of the popular constructed data
-types such as matrices and data frames, and (3) cannot differentiate between
-scalar values and vectors.  
-{\bf L1} types build on and extend {\bf L0} with with scalar values for primitive types (\sI, \sD, \sC, \sL, \sR, and \sX), type-parametric lists (written \lT{T} for any type {\tt T}), data frames (written \attr{\l}{df} as they are built up from lists of type any), and matrices (\attr{\vec{\tt T}}{mat} where {\tt T} is a primitive type).
+{\bf L0} is limited in expressivity as it (1) supports polymorphism for
+lists in a trivial way, (2) does not address some of the popular constructed
+data types such as matrices and data frames, and (3) cannot differentiate
+between scalar values and vectors.  {\bf L1} types build on and extend {\bf
+  L0} with with scalar values for primitive types (\sI, \sD, \sC, \sL, \sR,
+and \sX), type-parametric lists (written \lT{T} for any type {\tt T}), data
+frames (written \attr{\l}{df} as they are built up from lists of type any),
+and matrices (\attr{\vec{\tt T}}{mat} where {\tt T} is a primitive type).
 Subtyping is extended such that {\tt T}<:{$\vec{\tt T}$} for any primitive
-type {\tt T}. The subtype relation is undefined for lists of different types, as well as for matrices and data frames.
-
-\subsection{L2 Types}
-\label{sec:L2def}
-
-{\bf L2} extends the previous system with support for classes and attributes.
-While there are a finite number of results from the \code{typeof} function, this is {\it not} the case for \code{class} and \code{attributes}, as classes and attributes {\it can} be user-defined. 
-In terms of classes, R has a number of built-in or otherwise standard classes, ranging from truly primitive classes such as {\tt numeric} to the more complex {\tt data.frame}. 
-{\bf L1} already incorporates these builtin classes, namely through its data frame and matrix types, so we will ignore those classes and thus highlight those which are user-defined.  
-As for attributes, again some patterns arise naturally in R, e.g.
-all matrices have a {\tt dims} attribute, but even these are somewhat interesting, as a programmer could in theory set their own {\it dims} attribute on an arbitrary value.
-The only attribute we ignore is the {\tt class} attribute, as we collect the class of values separately (and the {\tt class} attribute, if present, is always equal to the class of a value).
-Attributes are name-value pairs, and to keep things simple we consider only attribute names.
-
-In sum, we extend {\bf L1} with attribute lists and classes: \attrclass{T}{C}{A} where {\tt T} can be any L1 type other than data frame or matrix, and {\tt C} and {\tt A} are lists of classes and attribute names respectively.
-The subtype relation is defined as: \attrclass{T}{C}{A} $<:_{\mathbf{L2}}$ \xspace \attrclass{\dot T}{C'}{A'} if $ T <:_{\mathbf{L1}} \dot T$ (i.e.
-are subtypes in the sense of {\bf L1}), $C \subseteq C'$, and $A \subseteq A'$.
-
-\subsection{Automatic Subtyping}
-\label{sec:autosub}
-
-When analyzing our data, we perform a certain amount of subtyping automatically to minimize the footprint of obviously equivalent signatures.
-This subtyping is performed at the function level of granularity.
-For example, if a function is observed to have two signatures \sD \to \df to \D \to \df, we collapse the signature to \D \to \df as \sD $<:$ \D and \df $<:$ \df.
-We would not collapse \sD, \I \to \df and \D, \sI \to \df, as \I $\not <:$ \sI and \D $\not<:$ \sD.
-
-Before presenting the results of our analysis, we will quickly survey examples of real, polymorphic R functions that we have analyzed, to build up a global sense of how types manifest themselves in current R code.
-
-%
-%
-%
-%
-%
-%
-\section{Examples of Polymorphism}
-\label{sec:polyex}
-
-In this section, we will present an eclectic mix of R functions which exhibit polymorphic behavior at {\bf L0}.
-
-We begin by referring the reader to Figure~\ref{fig:realex}, which showcases the compatibility of integers and doubles.
-Coercion from integer to double is performed automatically where appropriate (e.g., here when a double and integer are added together), and likewise coercion from double to integer is automatically performed when integers are strictly required (e.g., here when doubles are being used to index a vector or list).
+type {\tt T}. The subtype relation is undefined for lists of different
+types, as well as for matrices and data frames.
+
+\subsection{L2 Types}\label{sec:L2def}
+
+{\bf L2} extends the previous system with support for classes and
+attributes.  While there are a finite number of results from the
+\code{typeof} function, this is {\it not} the case for \code{class} and
+\code{attributes}, as classes and attributes {\it can} be user-defined.  In
+terms of classes, R has a number of built-in or otherwise standard classes,
+ranging from truly primitive classes such as {\tt numeric} to the more
+complex {\tt data.frame}.  {\bf L1} already incorporates these builtin
+classes, namely through its data frame and matrix types, so we will ignore
+those classes and thus highlight those which are user-defined.  As for
+attributes, again some patterns arise naturally in R, e.g. all matrices have
+a {\tt dims} attribute, but even these are somewhat interesting, as a
+programmer could in theory set their own {\it dims} attribute on an
+arbitrary value.  The only attribute we ignore is the {\tt class} attribute,
+as we collect the class of values separately (and the {\tt class} attribute,
+if present, is always equal to the class of a value).  Attributes are
+name-value pairs, and to keep things simple we consider only attribute
+names.
+
+In sum, we extend {\bf L1} with attribute lists and classes:
+\attrclass{T}{C}{A} where {\tt T} can be any L1 type other than data frame
+or matrix, and {\tt C} and {\tt A} are lists of classes and attribute names
+respectively.  The subtype relation is defined as: \attrclass{T}{C}{A}
+$<:_{\mathbf{L2}}$ \xspace \attrclass{S}{C'}{A'} if $ T
+<:_{\mathbf{L1}} S$ (i.e. are subtypes in the sense of {\bf L1}), $C
+\subseteq C'$, and $A \subseteq A'$.
+
+\subsection{Simplification by Subtyping}\label{sec:autosub}
+
+When analyzing our data, we perform a certain amount of simplifications
+automatically to minimize the footprint of equivalent signatures.  This
+subtyping is performed at the function level of granularity.  For example,
+if a function is observed to have two signatures \sD \to \df to \D \to \df,
+we collapse the signature to \D \to \df as \sD $<:$ \D and \df $<:$ \df.  We
+would not collapse \sD, \I \to \df and \D, \sI \to \df, as \I $\not <:$ \sI
+and \D $\not<:$ \sD.
+
+%
+%
+%
+\section{Examples of Polymorphism}\label{sec:polyex}
+
+Before presenting the results of our analysis, we  quickly survey
+examples of real, polymorphic R functions that we have analyzed, to build up
+a global sense of how types manifest themselves in current R code.
+
+We begin by referring the reader to Figure~\ref{fig:realex}, which showcases
+the compatibility of integers and doubles.  Coercion from integer to double
+is performed automatically where appropriate, and likewise coercion from
+double to integer is automatically performed when integers are strictly
+required.
 
@@
 > c(10, 20, 30)[1.2] # => 10, 1.2 coerced to 1L
-\end{lstlisting}}\caption{Example of \sD, \sI type usage.}\label{fig:realex}\end{figure}
-
-Figure~\ref{fig:optnull} shows an example of a function with an optional NULL argument. 
-In the function, {\tt x} is a sorted vector of income values, and {\tt w} is a vector of weights with default value NULL.
-The function computes fractional ranks required in other package functions.
-We see that if {\tt w} is NULL, then default unit weights are generated.
-This pattern of default NULL (or NA) arguments is very prevalent in R.
+\end{lstlisting}}\caption{Example of \sD, \sI type usage}\label{fig:realex}\end{figure}
+
+Figure~\ref{fig:optnull} shows a function with an optional NULL argument.
+In the function, {\tt x} is a sorted vector, and {\tt w} a vector with
+default value NULL.  If {\tt w} is NULL, then default unit weights are
+generated.  This pattern of default NULL (or NA) arguments is prevalent.
 
@@
   ...
-\end{lstlisting}}\caption{Example of optional argument (from {\tt acid} package).}\label{fig:optnull}\end{figure}
-
-Figure~\ref{fig:listvec} shows an example of a function with list and vector
-polymorphism.  The function takes a list or vector {\tt point} and a data
-frame {\tt polyg} representing a polygon.  In it, we see that {\tt point} is explicitly converted
-to a vector (via the \code{as.numeric} function) regardless of its type.
-This showcases the ease with which lists and vectors may be converted in R (the corresponding vector-to-list function is \code{as.list}).
+\end{lstlisting}}\caption{Optional argument ({\tt acid} package)}\label{fig:optnull}\end{figure}
+
+Figure~\ref{fig:listvec} shows a function with list and vector polymorphism.
+It takes a list or a vector {\tt point} and a data frame {\tt polyg}.  The
+{\tt point} is explicitly converted to a vector (by \code{as.numeric})
+regardless of type.  This showcases the ease of going from lists to vectors
+(and conversely).
 
@@
     ...
-\end{lstlisting}}\caption{Example of \l/\D argument (from {\tt bivrp} package).}\label{fig:listvec}\end{figure}
-
-Figure~\ref{fig:charclos} shows an example of a function with character and
-function polymorphism.  At a high level, this function
-calculates point estimates for some object.
-The caller specifies {\tt fn}, which is either a function {\it or} the name of a function which will be evaluated on the object samples to estimate parameters.
-If {\tt fn} is a string, the function lookup is performed by \code{match.fun}.
+\end{lstlisting}}\caption{List and vector polymorphism ({\tt bivrp} package)}\label{fig:listvec}\end{figure}
+
+Figure~\ref{fig:charclos} is a function with character and function
+polymorphism.  The caller specifies {\tt fn}, which is either a function
+{\it or} the name of a function which will be looked up by string.
 
@@
 pointest <- function (object, fn = mean, par.name, comp.label, chain.no,  ...) {
-    ...
     if (is.character(fn))
@@
         }
-\end{lstlisting}}\caption{Example of \sC/\sF argument (from {\tt BAMBI} package).}\label{fig:charclos}\end{figure}
-
-In Figure~\ref{fig:dfdbl}, we see a function that, among other things, takes in some data ({\tt dat}) and a
-character string ({\tt spss}) specifying either {\tt "in"} or {\tt "out"}.
-Here, the type of {\tt dat} is {\it dependent} on the value of {\tt spss}:
-if {\tt spss == "in"}, {\tt dat} must be a data frame, and otherwise if {\tt spss == "out"}, {\tt dat} must be a vector.
-Our analysis does not capture this granularity of dependency, as we only collect type information.
+\end{lstlisting}}\caption{Character and function polymorphism ({\tt BAMBI} package)}\label{fig:charclos}\end{figure}
+
+In Figure~\ref{fig:dfdbl}, we see a function that, among other things, takes
+in some data ({\tt dat}) and a character string ({\tt spss}) specifying
+either {\tt "in"} or {\tt "out"}.  Here, the type of {\tt dat} is {\it
+  dependent} on the value of {\tt spss}: if {\tt spss == "in"}, {\tt dat}
+must be a data frame, and otherwise if {\tt spss == "out"}, {\tt dat} must
+be a vector.  Our analysis does not capture this granularity of dependency,
+as we only collect type information.
 
 \begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-nret.translator <- function(dat, spss="out", ...) {
-  ...
+nret.translator <- function(dat, spss="out", ...) {  ...
   if(identical(spss, "out")) {
-    if(!is.vector(dat))  stop(simpleError("...'dat' must be a vector!"))
-    ...
+    if(!is.vector(dat))  stop(simpleError("...'dat' must be a vector!")) ...
   } else {
     ...
-    # here, dat is data.frame
     items.idx <- items.idx[order(names(dat[, items.idx]))]
@@
 
-In Figure~\ref{fig:chardbl}, we see a function which takes in a list ({\tt network}), a vector of indices of that list ({\tt fixIndices}), and a
-vector of values ({\tt values}).
-The locations specified by {\tt fixIndicies} in {\tt network} are updated with {\tt values}.
-Here, {\tt fixIndicies} has been observed to be either a character or double vector:
+In Figure~\ref{fig:chardbl}, we see a function which takes in a list ({\tt
+  network}), a vector of indices of that list ({\tt fixIndices}), and a
+vector of values ({\tt values}).  The locations specified by {\tt
+  fixIndicies} in {\tt network} are updated with {\tt values}.  Here, {\tt
+  fixIndicies} has been observed to be either a character or double vector:
 In R, list indices are typically doubles, but can be characters (if the list
-or vector being indexed has a {\tt names} attribute).
-Note that type level {\bf L2} can capture the presence of a {\tt names} attribute.
+or vector being indexed has a {\tt names} attribute).  Note that type level
+{\bf L2} can capture the presence of a {\tt names} attribute.
 

### 2019-04-06 paper/main.tex
daylight savings

--- 2019-04-06 paper/main.tex
@@
 \newcommand{\sN}{\ensuremath{\small{\mathtt N}}\xspace}     % Null
-%\renewcommand{\l}{\ensuremath{\small L<{\mathtt ?}>}\xspace}     % List
+%\renewcommand{\l}{\ensuremath{\small L<?>}\xspace}     % List
 \renewcommand{\l}{\ensuremath{\small\underline{\mathtt ?}}\xspace}     % List
@@
     ...
-\end{lstlisting}}\caption{Example of \l/\D argument (from {\tt bivrp} package).}\label{fig:listvec}\end{figure}
+\end{lstlisting}}\caption{Example of \l or \D (list or vector of doubles) argument (from {\tt bivrp} package).}\label{fig:listvec}\end{figure}
 
@@
 From Table~\ref{tab:L1top10arg}, we see that all of the most frequent argument signatures are monomorphic even though we increased the precision of the types.
-We see that scalars actually figure quite prominently, which is encouraging: this supports an idea that an explicit scalar annotation or type may be useful {\it for programmers}---indeed, such an annotation would be helpful for compiler optimizations (e.g., scalar-only values need not have vectors allocated to store them).
+We see that scalars actually feature quite prominently, which is encouraging: this supports an idea that an explicit scalar annotation or type may be useful {\it for programmers}---indeed, such an annotation would also be helpful for compiler optimizations (e.g., scalar-only values need not have vectors allocated to store them).
 We also see that matrices (\M{D}) are quite popular, a testament to the versatility of the data structure, as they can easily be converted to vectors and data frames.
@@
 Our analysis uncovered a non-negligible amount of argument positions which were {\it never} evaluated across all observed function calls.
-This is indicative of a desire to metaprogram arguments and otherwise deconstruct their AST, as the \%$>$\% function does.
+This is indicative of a desire to meta-program arguments and otherwise deconstruct their AST, as the \%$>$\% function does.
 This is an incredibly powerful language feature, and may well be at the source of R's popularity, if in a roundabout way.
-Tidyverse, by far the most popular package suite in R, owes some of its success to the new syntax defined by some of the included packages, and it's this ability to metaprogram arguments which allowed Tidyverse to define its unique syntax.
+Tidyverse, by far the most popular package suite in R, owes some of its success to the new syntax defined by some of the included packages, and it's this ability to meta-program arguments which allowed Tidyverse to define its unique syntax.
 An example of that syntax is \%$>$\%, and indeed this function was highly exercised by over 100 other packages.

### 2019-04-06 paper/main.tex

--- 2019-04-06 paper/main.tex
@@
 \newcommand{\ANY}{\ensuremath{\small{\mathtt ?}}\xspace}     % Any
-\newcommand{\lT}[1]{\ensuremath{\small\underline{\mathtt #1}}\xspace}     % list<T>
+\newcommand{\lT}[1]{\ensuremath{\small\underline{\mathtt{#1}}}\xspace}     % list<T>
 \newcommand{\M}[1]{\ensuremath{\attr{\vec{\tt #1}}{mat}}\xspace}     % matrix
@@
 polymorphism.  The caller specifies {\tt fn}, which is either a function
-{\it or} the name of a function which will be looked up by string.
+{\it or} the name of a function which will be looked up by name (a string).
 
@@
   \M{D}, \M{I} & 1502 & 0.2\% \\ 
-  \l, \lT{list} & 1199 & 0.1\% \\ 
+%  \l, \lT{list} & 1199 & 0.1\% \\ 
   \df, \M{D} & 1180 & 0.1\% \\ 
@@
 
-	\item \l, \lT{list}: here we have a list of any and a list of lists.
-	One explanation for this is that lists of lists sometimes contain NULL as an element, and this would show up with type \l, again as we don't do any subtyping while computing precise list types.
-	Another explanation is the way R handles object-orientation, which we discuss shortly in Section~\ref{sec:S3S4R5}.
+%	\item \l, \lT{list}: here we have a list of any and a list of lists.
+%	One explanation for this is that lists of lists sometimes contain NULL as an element, and this would show up with type \l, again as we don't do any subtyping while computing precise list types.
+%	Another explanation is the way R handles object-orientation, which we discuss shortly in Section~\ref{sec:S3S4R5}.
 

### 2019-04-06 paper/main.tex
stuff

@@
 
-In this section, we will present the results of our analysis at each of our type levels, describing the types of polymorphism that we have observed and shedding light on the frequency of polymorphic patterns in R.
-
-Before we begin, we define the kinds of polymorphism we will consider.
-In choosing how to express the polymorphicity of a function, there are two main options:
-First, having several separate whole-function signatures, e.g.
-\D $\rightarrow$ \D and \C $\rightarrow$ \D, or alternatively expressing the polymorphism at the argument level with a union type, e.g.
-(\D $\cup$ \C) $\rightarrow$ \C.
-We refer to the former as {\it per-function} or {\it function-level polymorphism}, and to the latter as {\it per-argument} or {\it argument-level polymorphism}.
-
-We believe that both of these notions of polymorphism are valuable for different reasons.
-Per-argument polymorphism highlights the type-based behavior of function arguments themselves, and helps to describe the type-based behavior of all R values.
-For instance, we observe that many arguments have (\sC, \sF) as a signature, from which we deduce that many R programmers are taking advantage of functions such as \code{match.fun("fun\_name")} to call functions by name.
-On the other hand, per-function polymorphism highlights how R programmers interact with functions as a whole, and allows us to discern more subtle patterns.
-For example, observing two function signatures such as \D \to \D and \sF \to \sF may lead us to ascribe a more useful and precise type for the function than if we had simply collapsed the argument and return types (e.g., $X$ \to $X$ vs.
-\D $\cup$ \sF \to \D $\cup$ \sF).
-
-The layout of this section is as follows:
-In order, we present analysis results for {\bf L0}, {\bf L1}, and {\bf L2}, speaking to both per-argument and per-function polymorphism at each level.
-Where appropriate, we digress to highlight interesting tidbits of tangential information which help us understand the style of code that R programmers write.
-We begin with {\bf L0}.
+This section,  presents the results of our analysis at each of our type levels, describing the types of polymorphism that we have observed and shedding light on the frequency of polymorphic patterns in R.
+
+Before we begin, we define the kinds of polymorphism we will consider.  In
+choosing how to express the polymorphism of a function, we consider two
+simple options: First, having the union of signatures, e.g. \D $\rightarrow$
+\D and \C $\rightarrow$ \D, or a union types for each argument, e.g. (\D
+$\cup$ \C) $\rightarrow$ \C.  We refer to the former as {\it function-level}
+and the latter as {\it argument-level polymorphism}.
+
+Per-argument polymorphism highlights the type-based behavior of function
+arguments themselves, and helps to describe the type-based behavior of all R
+values.  For instance, we observe that many arguments have (\sC, \sF) as a
+signature, from which we deduce that many R programmers are taking advantage
+of functions such as \code{match.fun("fun\_name")} to call functions by
+name.  On the other hand, function-level polymorphism highlights how R
+programmers interact with functions as a whole, and allows us to discern
+more subtle patterns.  For example, observing two function signatures such
+as \D \to \D and \sF \to \sF may lead us to ascribe a more useful and
+precise type for the function than if we had simply collapsed the argument
+and return types (e.g., $X$ \to $X$ vs. \D $\cup$ \sF \to \D $\cup$ \sF).
 
@@
 
-At {\bf L0}, the type system is mostly aligned with the types of the core language.
-
-Across all of the arguments that we've observed, the vast majority are monomorphic.
-Table~\ref{tab:L0top10arg} shows the breakdown of the 10 most frequently occurring argument signatures.
+At {\bf L0}, the type system is mostly aligned with the types of the core
+language. Across all of the arguments that we have observed, the vast
+majority are monomorphic.  Table~\ref{tab:L0top10arg} shows the breakdown of
+the 10 most frequently occurring argument signatures.
 
@@
 
-We see that doubles are the most common argument type, which is not entirely surprising as doubles are the backbone of any numeric data.
-Next is lists, which seems reasonable given that lists are often used to store the aforementioned data.
-Recall that in {\bf L0}, data frames will appear as lists.
-The \ANY type is not an R type per se: we ascribe such a type to arguments which were {\it always} unevaluated across all calls.
-As it happens, many R functions meta-program on some of their arguments, and this meta-programming does not force the argument promise, and if this meta-programming is always performed in every recorded call we will ascribe the \ANY type, as we cannot say what the argument {\it is} without possibly forcing a promise.
-The magrittr package's pipe \%$>$\% function is an example of a function which performs a lot of meta-programming; we will discuss it in a little more detail shortly.
-Interestingly, a polymorphic attribute signature \D, \l is among the top 10 overall---this is the vector and list polymorphism (shown in Figure~\ref{fig:listvec}) made manifest.
-
-Now that we've seen the most common attribute-level signatures, we will turn our attention to function-level polymorphism.
-Indeed, one of our goals is to generate useful function signatures, so we ask: {\it how many functions have more than one recorded signature?}
+Doubles are the most common argument type, which is not entirely surprising
+as doubles are the backbone of most statistical functions.  Next is lists,
+which is reasonable as lists are used to build data frames and other widely
+used data structures.  The \ANY type is not an R type per se: we ascribe
+such a type to arguments which were {\it always} unevaluated across all
+calls.  As it happens, many R functions meta-program on some of their
+arguments, and this meta-programming does not force the argument promise,
+and if this meta-programming is always performed in every recorded call we
+will ascribe the \ANY type, as we cannot say what the argument {\it is}
+without possibly forcing a promise.  The magrittr package's pipe \%$>$\%
+function is an example of a function which does meta-programming; we will
+discuss it shortly.  Interestingly, a polymorphic attribute signature \D, \l
+is among the top 10 overall---this is the vector and list polymorphism made
+manifest.
+
+We turn our attention to function-level polymorphism.  Indeed, one of our
+goals is to generate useful function signatures, so we ask: {\it how many
+  functions have more than one recorded signature?}
 Figure~\ref{fig:L0funcounts} has the answer, with a breakdown by function.
+Most functions are monomorphic.  That said, a few functions have
+disproportionately large number of signatures.  One such function is the
+magrittr package's pipe \%$>$\%, which is used to implement a kind of
+currying. It has nearly 100 unique signatures!
 
@@
 \includegraphics[width=.9\textwidth]{L0_by_fun}
-\caption{Record of Function Polymorphism in {\bf L0}. \LZEROPERCPOLY of functions are polymorphic.}
+\caption{Record of Function Polymorphism in {\bf L0}; \LZEROPERCPOLY of
+  functions are polymorphic}
 \label{fig:L0funcounts}\end{center}
@@
 
-Here, we see that the vast majority of functions are monomorphic in L0.
-That said, there are a few functions with a disproportionately large number of signatures.
-One such function is the magrittr package's pipe \%$>$\%, which is used to implement a kind of currying (as per discussion in Section~\ref{sec:Impl})---here, \%$>$\% has nearly 100 unique signatures!
-Many of the meta-programming functions in the Tidyverse are quite sophisticated, and are designed with the explicit goal of being very usable.
-This usability maps to polymorphism in our analysis, so it's no surprise that a Tidyverse function features so prominently here.
-
-Another function with many signatures is the data.table package's [ function, which is a reimplementation of the square bracket operator for data.tables.
-This function has 16 formal arguments, and call configurations vary widely.
-Essentially, [ is {\it the} main function for data.tables, and performs very different tasks based on which arguments are provided.
-Two wildly different uses are:
-\code{DT[i, j, by]}, where columns \code{i} and \code{j} are accessed and grouped by \code{by}, and \code{DT[x == "some" & y > 42]}, which selects all rows who's \code{x} column is the string \code{"some"} and who's \code{y} column is greater than \code{42}.
-Given that, we find it unsurprising that the function is so polymorphic.
-
-Having had a taste for function-level polymorphism, we will now explore common patterns in argument-level polymorphism.
-For this, refer to Table~\ref{tab:L0toppoly}.
+Another function with many signatures is the data.table package's \code{[}
+  function, which is a reimplementation of the square bracket operator for
+  selecting values in a data.table.  This function has 16 formal arguments,
+  and call configurations vary widely.  Essentially, the function is {\it
+    the} main function for data.tables, and performs very different tasks
+  based on which arguments are provided.  Two wildly different uses are:
+  \code{dtT[i,j,by]}, where columns \code{i} and \code{j} are accessed and
+  grouped by \code{by}, and \code{dt[x=="some" & y>42]}, which selects
+  all rows who's \code{x} column is the string \code{"some"} and who's
+  \code{y} column is greater than \code{42}.  Given that, we find it
+  unsurprising that the function is so polymorphic.
 
@@
 
+Having had a taste for function-level polymorphism, we will now explore
+common patterns in argument-level polymorphism.  For this, refer to
+Table~\ref{tab:L0toppoly}.
 Let's unpack some of the signatures in Table~\ref{tab:L0toppoly}:
@@
 \begin{itemize}
-
 \item \D, \l (and \C, \l): L0 doesn't parameterize list types, and it's very likely that this signature describes vectors of doubles and lists of doubles.
@@
 	In short, this signature usually represents a function or a function name.
-
 \end{itemize}
 
-We have established and discussed some of the most common polymorphic attribute signatures, and have built up an understanding of argument-level polymorphism in the {\bf L0} type system.
-This matches almost exactly with the language as it is today, as {\bf L0} is in (nearly) 1:1 correspondence with how types are currently represented in the language.
-The next question we address is: {\it just how pervasive is this argument polymorphism?}
-Refer to Table~\ref{tab:L0argcounts}.
-
-\begin{table}[ht]\centering
-\begin{tabular}{lrl}
-  \hline
-Signature Type & Count & Percentage \\ 
-  \hline
-Monomorphic & 852K & 94\% \\ 
-  Total Seen & 906K & --- \\ 
-   \hline
-\end{tabular}
-\caption{Argument position polymorphism in L0.}
-\label{tab:L0argcounts}
-\end{table}
-
-From Table~\ref{tab:L0argcounts} we see that, with our definition of subtyping on L0 and other simplifications we made, an overwhelming amount of argument polymorphism is accounted for.
-We also saw in Table~\ref{tab:L0toppoly} that each of the most frequently occurring argument signatures had some explanation.
-
-That said, L0 is not without limitations, and the main issues stem from a lack of adequate information, as the type system doesn't capture precise list types, and does not distinguish scalars from vectors.
-The next type system we will explore, L1, tackles these issues by collecting more information to build more precise types.
-While the L1 types are less ``in-tune'' with current R implementations, one can imagine designing an all-new R runtime which may benefit from the more specific information in L1, e.g.
-by not needing to vectorize scalar values, or designing a static analysis which uses precise list types to give programmers better information in an IDE.
-L1 is discussed next.
-
-%
-%
+We have established and discussed some of the most common polymorphic
+attribute signatures, and have built up an understanding of argument-level
+polymorphism in the {\bf L0} type system.  This matches almost exactly with
+the language as it is today, as {\bf L0} is in (nearly) 1:1 correspondence
+with how types are currently represented in the language.  The next question
+we address is: {\it just how pervasive is this argument polymorphism?}
+We find that 94\% of arguments are monomorphic.
+
+From Table~\ref{tab:L0argcounts} we see that, with our definition of
+subtyping on L0 and other simplifications we made, an overwhelming amount of
+argument polymorphism is accounted for.  We also saw in
+Table~\ref{tab:L0toppoly} that each of the most frequently occurring
+argument signatures had some explanation.
+
+That said, L0 is not without limitations, and the main issues stem from a
+lack of adequate information, as the type system doesn't capture precise
+list types, and does not distinguish scalars from vectors.
+
 %
@@
 
-As we mentioned in Section~\ref{sec:L1def}, L0 is somewhat limited in its expressivity. 
-To address this, we develop L1, which espouses a finer-grained approach to types. 
-Major modifications from L0 include:
+L1 espouses a finer-grained approach to types.  Major modifications from L0
+include:
 
@@
 
-Let us begin as we did in L0 by looking at the breakdown of argument signatures in L1.
-Consult Table~\ref{tab:L1top10arg}.
+Let us begin as we did in L0 by looking at the breakdown of argument
+signatures in L1.  Consult Table~\ref{tab:L1top10arg}.
 \begin{table}[ht]
@@
 
-From Table~\ref{tab:L1top10arg}, we see that all of the most frequent argument signatures are monomorphic even though we increased the precision of the types.
-We see that scalars actually feature quite prominently, which is encouraging: this supports an idea that an explicit scalar annotation or type may be useful {\it for programmers}---indeed, such an annotation would also be helpful for compiler optimizations (e.g., scalar-only values need not have vectors allocated to store them).
-We also see that matrices (\M{D}) are quite popular, a testament to the versatility of the data structure, as they can easily be converted to vectors and data frames.
+From Table~\ref{tab:L1top10arg}, we see that all of the most frequent
+argument signatures are monomorphic even though we increased the precision
+of the types.  We see that scalars actually figure quite prominently, which
+is encouraging: this supports an idea that an explicit scalar annotation or
+type may be useful {\it for programmers}---indeed, such an annotation would
+be helpful for compiler optimizations (e.g., scalar-only values need not
+have vectors allocated to store them).  We also see that matrices (\M{D})
+are quite popular, a testament to the versatility of the data structure, as
+they can easily be converted to vectors and data frames.
 
@@
 
-We see that we barely increased the polymorphism with our transition to L1, despite the far richer type information.
-The biggest outlier here is indeed \%$>$\%, as was the case in {\bf L0}.
-Otherwise, the shape is very similar.
-In short: even though {\L1} types are more expressive, the breakdown of per-function polymorphism is largely unchanged from {\bf L0}.
-
-As we were in L0, we are curious to know how argument-level polymorphism looks in L1.
-Refer to Table~\ref{tab:toppolyL1}.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{lrl}
-  \hline
-Signature & Count & \% of All Signatures \\ 
-  \hline
+We see that we barely increased the polymorphism with our transition to L1,
+despite the far richer type information.  The biggest outlier here is indeed
+\%$>$\%, as was the case in {\bf L0}.  Otherwise, the shape is very similar.
+In short: even though {\L1} types are more expressive, the breakdown of
+per-function polymorphism is largely unchanged from {\bf L0}.
+
+As we were in L0, we are curious to know how argument-level polymorphism
+looks in L1.  Refer to Table~\ref{tab:toppolyL1}.
+
+\begin{table}[ht]\centering\begin{tabular}{lrl}  \hline
+Signature & Count & \% of All Signatures \\   \hline
   \D, \M{D} & 4848 & 0.5\% \\ 
@@
   \I, \sD & 997 & 0.1\% \\ 
-  \sC, \sF & 991 & 0.1\% \\ 
-   \hline
+  \sC, \sF & 991 & 0.1\% \\    \hline
 \end{tabular}
-\caption{Top polymorphic argument signatures in L1.}
-\label{tab:toppolyL1}
+\caption{Top polymorphic argument signatures in L1.}\label{tab:toppolyL1}
 \end{table}
@@
 
-With an understanding of argument polymorphism in L1, we wonder: {\it how many argument positions are polymorphic in L1?}
-Refer to Table~\ref{tab:argcountsL1}.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{lrl}
-  \hline
-Signature Type & Count & Percentage \\ 
-  \hline
-Monomorphic & 864K & 95.3\% \\ 
-  Total Seen & 906K & --- \\ 
-   \hline
-\end{tabular}
-\caption{Argument position polymorphism in L1.}
-\label{tab:argcountsL1}
-\end{table}
-
-In spite of drastically increasing the precision of our types, we have actually {\it reduced} the amount of polymorphism.
-This suggests that with a good notion of subtyping, precise type annotations won't cause an undue increase in the polymorphism of the language.
+With an understanding of argument polymorphism in L1, we wonder: {\it how
+  many argument positions are polymorphic in L1?}  95.3\% of arguments are
+monomorphic.  In spite of drastically increasing the precision of our types,
+we have actually {\it reduced} the amount of polymorphism.  This suggests
+that with a good notion of subtyping, precise type annotations won't cause
+an undue increase in the polymorphism of the language.
 
@@
 %
-\subsubsection{Scalars}
-\label{sec:scalars}
-
-It seems that there is a place for a scalar annotation in R.
-We saw in this section that strictly scalar values appear quite frequently, making up some of the most frequently occurring argument signatures.
-In fact, \PERCSCALARMONO of arguments have a strictly scalar monomorphic signature.
-This suggests that such a distinction would more accurately capture current use of the language, and would also be beneficial to future implementations which can perform optimizations based on the type (e.g., not vectorizing scalar-only values).
+\subsubsection{Scalars}\label{sec:scalars}
+
+It seems that there is a place for a scalar annotation in R.  We saw in this
+section that strictly scalar values appear quite frequently, making up some
+of the most frequently occurring argument signatures.  In fact,
+\PERCSCALARMONO of arguments have a strictly scalar monomorphic signature.
+This suggests that such a distinction would more accurately capture current
+use of the language, and would also be beneficial to future implementations
+which can perform optimizations based on the type (e.g., not vectorizing
+scalar-only values).
 
@@
 
-There is something more to be said about the \l type, namely that it is sometimes linked to (one of) R's takes on object-oriented programming.
-R has three object systems: S3, S4, and R5, and S3 objects will often manifest themselves with this signatures in L1.
-An S3 object is any value with a ``class'' attribute, and indeed many of these tend to be lists.
-For example, an S3-style ``Point'' might be a list with an entry for each dimension in the point, with a ``class'' attribute set to ``Point''.
-When these S3 objects have contents with a variety of types, their L1 type will appear as \l, which leaves something to be desired:
-Our next type system, L2, will remedy this.
+There is something more to be said about the \l type, namely that it is
+sometimes linked to (one of) R's takes on object-oriented programming.  R
+has three object systems: S3, S4, and R5, and S3 objects will often manifest
+themselves with this signatures in L1.  An S3 object is any value with a
+``class'' attribute, and indeed many of these tend to be lists.  For
+example, an S3-style ``Point'' might be a list with an entry for each
+dimension in the point, with a ``class'' attribute set to ``Point''.  When
+these S3 objects have contents with a variety of types, their L1 type will
+appear as \l, which leaves something to be desired: Our next type system,
+L2, will remedy this.
 
@@
 \subsubsection{\C, \I, \D Polymorphism}
-Another significant source of apparently strange polymorphism is the character, integer, double polymorphism.
-Figure~\ref{fig:chardbl} lays out an example of this polymorphic signature, showing that lists and vectors can be accessed by strings as well as numerics.
-One could conceive of an ``index'' type to capture this, but there are certain limitations to such a type, as such an ``index'' cannot be incremented as it could be a string.
-An even more sophisticated dependent type system could capture this, but it's far from clear if such a sophisticated type system would be for a language like R.
+Another significant source of apparently strange polymorphism is the
+character, integer, double polymorphism.  Figure~\ref{fig:chardbl} lays out
+an example of this polymorphic signature, showing that lists and vectors can
+be accessed by strings as well as numerics.  One could conceive of an
+``index'' type to capture this, but there are certain limitations to such a
+type, as such an ``index'' cannot be incremented as it could be a string.
+An even more sophisticated dependent type system could capture this, but
+it's far from clear if such a sophisticated type system would be for a
+language like R.
 
@@
 
-Throughout this section, we have seen that even though we substantially increased the precision of our type annotations, we have not increased the amount of polymorphism that we observe.
-The transition from L0 to L1 revealed some interesting patterns that L1 cannot quite capture---for these, we will need to appeal to the class and attributes of values, which are incorporated into {\bf L2}, discussed next.
+Throughout this section, we have seen that even though we substantially
+increased the precision of our type annotations, we have not increased the
+amount of polymorphism that we observe.  The transition from L0 to L1
+revealed some interesting patterns that L1 cannot quite capture---for these,
+we will need to appeal to the class and attributes of values, which are
+incorporated into {\bf L2}, discussed next.
 

### 2019-04-06 paper/main.tex
merging

--- 2019-04-06 paper/main.tex
@@
 dynamically checked at runtime. The starting point for our work is the
-question: \emph{what could a static type system for R look like?}  To answer
+question: \emph{what could a static type system for R look like?}  In order to answer
 it, we study the polymorphism that is present in \RLOC lines of R

### 2019-04-06 paper/main.tex
final

@@
 language we are working with as well as the real-world needs it
-answers. Often, we, as researchers, can appeal to our intuition as many
-languages are intended to be general purpose and appeal to users with some
-computer science training. Unfortunately, these intuitions don't always
+answers. Often, we as researchers can appeal to our intuition, as many
+languages are intended to be general purpose and attract users with some
+computer science training. Unfortunately, these intuitions do not always
 apply to domain-specific languages, languages designed for and by a specific
@@
 Many of the design decisions that gave us R were intended to foster an
-interactive, exploratory, programming style. This includes, to name a few,
+interactive, exploratory, programming style. These include, to name a few,
 the lack of type annotations on variables and functions, the ability to use
 syntactic shortcut, and the automatic conversion between data types.  While
-these choices have led to a language that is surprisingly easy to use by
-beginners --many data science programs do not teach the language itself but
-simply introduce some of its key libraries-- they have also created a
+these choices have led to a language that is surprisingly easy to use for
+beginners---many data science programs do not teach the language itself but
+simply introduce some of its key libraries---they have also created a
 language where almost all computations yield a numeric result and where
@@
 
-We propose to capture the degree of polymorphism present in R by the means
+We propose to capture the degree of polymorphism present in R by means
 of a dynamic analysis of widely used libraries. For each function call we
@@
 our conclusions are only as good as the coverage of the possibly function
-behaviors. Previous work~\cite{issta18}, reported that running all the
+behaviors. Previous work~\cite{issta18} reported that running all the
 scripts that come with CRAN packages gives, on average, 68\% test coverage.
@@
 leveraged for meta-programming as it is possible to retrieve the text of a
-promise and evaluate that in a different environment.
+promise and evaluate it in a different environment.
 
@@
 results, in notebooks. Few notebooks are publicly shared, and when they are,
-the data isn't. For this reason our analysis focuses on packages which
+the data is not. For this reason our analysis focuses on packages which
 bundle reusable units of R code with documentation, sample data and
@@
 is used to implement performance critical portions of the code. The majority
-of package, 8,375 to be precise, have no C or C++ code.  The remaining 3,078
-packages, have a median 572 lines of C and the largest package has 385,839
+of packages, 8,375 to be precise, have no C or C++ code.  The remaining 3,078
+packages have a median 572 lines of C and the largest package has 385,839
 lines of C.
@@
 For each package, we extracted all executable code snippets from
-documentation, vignettes and tests and ran them independently recording all
+documentation, vignettes and tests and ran them independently, recording all
 calls to R functions.  It is noteworthy that in order to run the scripts in
@@
 script execution, coupled with the fact that some scripts take inordinate
-amounts of time to run, this is why we limited our analysis of any given
+amounts of time to run, we limited our analysis of any one
 package to one hour.
@@
 \begin{itemize}
-\item \D, \l (and \C, \l): L0 doesn't parameterize list types, and it's very likely that this signature describes vectors of doubles and lists of doubles.
+\item \D, \l (and \C, \l): L0 does not parameterize list types, and it is very likely that this signature describes vectors of doubles and lists of doubles.
 	Coercing vectors to lists is trivial in R, and indeed many of the functions with such an argument signature do.
@@
 	Another possible explanation for these signatures are dates, which can be represented by numerics or characters and the value usually has one of the date classes.
-	L0 doesn't deal with classes, which are one of the topics of L2.
+	L0 does not deal with classes, which are one of the topics of L2.
 
@@
 That said, L0 is not without limitations, and the main issues stem from a
-lack of adequate information, as the type system doesn't capture precise
+lack of adequate information, as the type system does not capture precise
 list types, and does not distinguish scalars from vectors.
@@
 %	\item \l, \lT{list}: here we have a list of any and a list of lists.
-%	One explanation for this is that lists of lists sometimes contain NULL as an element, and this would show up with type \l, again as we don't do any subtyping while computing precise list types.
+%	One explanation for this is that lists of lists sometimes contain NULL as an element, and this would show up with type \l, again as we do not do any subtyping while computing precise list types.
 %	Another explanation is the way R handles object-orientation, which we discuss shortly in Section~\ref{sec:S3S4R5}.
 
-	\item \df, \M{D}: in R, it's easy to transform a matrix into a data frame, and this argument is likely a result of that:
+	\item \df, \M{D}: in R, it is easy to transform a matrix into a data frame, and this argument is likely a result of that:
 	Simply calling \code{as.data.frame} on a matrix is enough.
@@
 
-	\item \I, \sD: subtyping between primitives does not hold between scalars and vectors (i.e., \I $\not<:$ \sD) even though it's likely that the programmer intended for the argument to be a vector of the supertype (here, these arguments were likely intended to be of type \D).
+	\item \I, \sD: subtyping between primitives does not hold between scalars and vectors (i.e., \I $\not<:$ \sD) even though it is likely that the programmer intended for the argument to be a vector of the supertype (here, these arguments were likely intended to be of type \D).
 
@@
 we have actually {\it reduced} the amount of polymorphism.  This suggests
-that with a good notion of subtyping, precise type annotations won't cause
+that with a good notion of subtyping, precise type annotations will not cause
 an undue increase in the polymorphism of the language.
@@
 An even more sophisticated dependent type system could capture this, but
-it's far from clear if such a sophisticated type system would be for a
+it is far from clear if such a sophisticated type system would be for a
 language like R.
@@
 Not pictures here, but interesting: named lists are more frequent than their unnamed counterparts.
-No classes are represented here since (1) we eliminated builtin classes, as we've used them to construct types already, and (2) most other classes are specific to some package or set of packages---it's doubtful that such a class would be so pervasive as to feature in this breakdown.
+No classes are represented here since (1) we eliminated builtin classes, as we've used them to construct types already, and (2) most other classes are specific to some package or set of packages---it is doubtful that such a class would be so pervasive as to feature in this breakdown.
 
@@
 
-Given the data that we've collected, it's natural to wonder what the most polymorphic packages are.
+Given the data that we've collected, it is natural to wonder what the most polymorphic packages are.
 Indeed, we have shed some light on some of the most polymorphic {\it functions}, but what about the packages that these functions belong to?
@@
 The first package, magrittr, is one that we discussed earlier in the context of its pipe \%$>$\% function. 
-It's not surprising that the package as a whole is quite polymorphic, as it implements {\it several} different pipe functions, e.g.
+It is not surprising that the package as a whole is quite polymorphic, as it implements {\it several} different pipe functions, e.g.
 \%$<>$\% which pipes as normal and then assigns to the left-hand argument.
@@
 
-It's highly improbable that a static type system designed for an untyped language will be able to capture all of the language's existing code (without resorting to an \ANY type or abusing unions).
-It's also unclear how useful a type system would be if it {\it did} capture everything.
+It is highly improbable that a static type system designed for an untyped language will be able to capture all of the language's existing code (without resorting to an \ANY type or abusing unions).
+It is also unclear how useful a type system would be if it {\it did} capture everything.
 Indeed, many of the packages discussed here are really designed with an \ANY-typed argument in mind (e.g., left-hand side of \%$>$\%), and restricting these functions should not be the goal of any static type system designed for a language after it has matured.
@@
 This is an incredibly powerful language feature, and may well be at the source of R's popularity, if in a roundabout way.
-Tidyverse, by far the most popular package suite in R, owes some of its success to the new syntax defined by some of the included packages, and it's this ability to meta-program arguments which allowed Tidyverse to define its unique syntax.
+Tidyverse, by far the most popular package suite in R, owes some of its success to the new syntax defined by some of the included packages, and it is this ability to meta-program arguments which allowed Tidyverse to define its unique syntax.
 An example of that syntax is \%$>$\%, and indeed this function was highly exercised by over 100 other packages.
@@
 For instance, a language implementation could perform type-based optimization in the context of a function return simply by knowing the function argument type.
-In Figure~\ref{fig:funanno}, we can assign the function a more sophisticated signature ({\tt [2]}) if we don't simply collapse the observed argument signatures element-wise ({\tt [1]}).
+In Figure~\ref{fig:funanno}, we can assign the function a more sophisticated signature ({\tt [2]}) if we do not simply collapse the observed argument signatures element-wise ({\tt [1]}).
 This gets more interesting with attributes, as we might want to annotate a function to communicate that it adds an attribute to whatever the list of attributes of its input value.
@@
 
-Generating the code to dynamically enforce these signatures is also possible, with varying levels of complexity.
+Generating the code to dynamically enforce these signatures is also possible.
 Simple checks include checking the class of a value, or checking for the presence of certain attributes.
@@
 
-We also saw that, despite being far more specific and descriptive than previous levels, the {\bf L2} type system didn't bring about an undue increase in the amount of overall polymorphism in the language.
+We also saw that, despite being far more specific and descriptive than previous levels, the {\bf L2} type system did not bring about an undue increase in the amount of overall polymorphism in the language.
 This signifies that {\bf L2} captures the vast majority of already existing polymorphic patterns, and very few functions would need to resort to an \ANY annotation, and indeed many arguments are really \ANY-typed (e.g., \%$>$\% from magrittr and the prototype assignment operator {\tt \$<-} from proto).
@@
 
-The idea of measuring polymorphism via some program analysis isn't new.
+The idea of measuring polymorphism via some program analysis is not new.
 For instance, some work~\cite{aakerblom2015measuring} measures the polymorphism of Python programs, but with a slightly different goal than ours: authors seek to assess the need for expressiveness in type systems that are retrofit onto dynamic languages.

### 2019-04-24 notebooks/rplus.Rmd
r plus, and updated java

---
title: "Tracing R Base: +"
output: html_notebook
---

Preamble.

```{r}
source("../code/data_processing.R")
```

In this notebook, you'll find our results from tracing part of the R base package.

Unfortunately, tracing base isn't going to be quite so straightforward.
The first and most relevant issue is that a number of the base package's functions are primitives (i.e. builtins) and are not amenable to the Genthat framework.
I needed to transform the + function into a closure in order to get this to shake out, and needed to make a number of light modifications to Genthat altogether.

For the record, this is what I needed to do to get + to trace:

```{r}
# unlock binding so we can write to it
# unlockBinding(as.symbol("+"), baseenv())
# This makes + into a closure, so we can deal with it
# assign("+", function(e1, e2) .Primitive("+")(e1, e2), envir=baseenv())
# Actually decorate the base env functions
# genthat::decorate_function(`+`, env=baseenv())
```

The above takes place in the genthat-tracing-site-file.

To load the data, run:

```{r}
# technically in L2
base_df <- readRDS("../data/some_plus_examples.RDS")

# for L0
base_df_L0 <- transform_new_df_w_TS_map(base_df, type_map_T0_to_r, unparam = T)

# just types
base_df_just_types <- base_df %>% select(arg1_t, arg2_t, ret_t)
```

### 2019-04-25 notebooks/corpus_stats_and_plots.Rmd
light modifications to some data files, new notebook, and new data

@@
   ggplot(aes(y=rec, x=obs)) + geom_point(size=0.1) + 
-    labs( x = "Observed", y = "Recorded") +
+    labs( x = "Calls Observed", y = "Signatures Recorded") +
     scale_x_continuous(breaks = c( 1, 2, 3, 4, 5, 6, 7),

### 2019-04-25 notebooks/refactor_test.Rmd
light modifications to some data files, new notebook, and new data

---
title: "Test Changes to Java Code"
output: html_notebook
---

This notebook is just to test the changes that were made to the Java code to make it more reasonable.
The main change was making Signatures contain a List of stuff, rather than just a bunch of fields.
It's easier to iterate through the fields this way, but a bit slower.

Ok. We ran Main with doIt uncommented.
What does L0_subtype look like?

```{r}
source("../code/data_processing.R")
```

```{r}
psp_L0 <- read_csv("../data/L0_subtype.csv.gz")
# ok that seems fine, try something else

psp_L0_col <- read_csv("../data/L0_collapsed.csv.gz")
# seems ok

L0_arg_sig_c <- read_csv("../data/L0_arg_sig_counts.csv.gz")
# seems fine

L0_sanitized <- read_csv("../data/L0_sanitized_class.csv.gz")
# yeah

psp_L2 <- read_csv("../data/L1_L2.csv.gz")
# yeah seems ok

list_of_L2_sig_c <- read_csv("../data/L1_arg_sig_counts_L2.csv.gz")
# yeah

```

Ok, things seem to have worked out.

### 2019-04-25 notebooks/rplus.Rmd
light modifications to some data files, new notebook, and new data

@@
 base_df_L0 <- transform_new_df_w_TS_map(base_df, type_map_T0_to_r, unparam = T)
+base_df_L0 <- unlist_new_df(base_df_L0)
 
@@
 ```
+
+```{r}
+# pass things over to java
+# dont forget to tar
+
+# save L0
+write_csv(base_df_L0, "../data/plus_80_L0.csv")
+# save L1
+write_csv(base_df, "../data/plus_80_L1.csv")
+
+# do the java
+
+# read subtype
+subtype_df_L0 <- read_csv("../data/plus_80_L0_subtype.csv.gz")
+subtype_df_L0 <- subtype_df_L0[!subtype_df_L0$pkg == "SUBTYPE",]
+subtype_df_L1 <- read_csv("../data/plus_80_L1_subtype.csv.gz")
+subtype_df_L1 <- subtype_df_L1[!subtype_df_L1$pkg == "SUBTYPE",]
+subtype_df_L2 <- read_csv("../data/plus_80_L1_L2.csv.gz")
+subtype_df_L2 <- subtype_df_L2[!subtype_df_L2$pkg == "SUBTYPE",]
+```
+
+```{r}
+# just get +
+
+subtype_df_L0_p <- filter(subtype_df_L0, fun == "+")
+subtype_df_L1_p <- filter(subtype_df_L1, fun == "+")
+subtype_df_L2_p <- filter(subtype_df_L2, fun == "+")
+
+```

### 2019-04-25 notebooks/rplus.Rmd
small fix

@@
 
-subtype_df_L0_p <- filter(subtype_df_L0, fun == "+")
-subtype_df_L1_p <- filter(subtype_df_L1, fun == "+")
-subtype_df_L2_p <- filter(subtype_df_L2, fun == "+")
+subtype_df_L0_p <- filter(subtype_df_L0, fun == "+") %>% salient_only_one_fun
+subtype_df_L1_p <- filter(subtype_df_L1, fun == "+") %>% salient_only_one_fun
+subtype_df_L2_p <- filter(subtype_df_L2, fun == "+") %>% salient_only_one_fun(just_types=F)
 

### 2019-05-05 notebooks/rplus.Rmd
fixed optionals for brackets, and pushing data for that too

@@
 # technically in L2
-base_df <- readRDS("../data/some_plus_examples.RDS")
+# base_df <- readRDS("../data/some_plus_examples.RDS")
+# base_df <- readRDS("../data/base_with_plus_and_brack.RDS")
+base_df <- readRDS("../data/base_with_brack_fixed.RDS")
 
@@
 
+# just get [
+
+subtype_df_L0_b <- filter(subtype_df_L0, fun == "[") %>% salient_only_one_fun(just_types = F)
+subtype_df_L1_b <- filter(subtype_df_L1, fun == "[") %>% salient_only_one_fun(just_types = F)
+subtype_df_L2_b <- filter(subtype_df_L2, fun == "[") %>% salient_only_one_fun(just_types=F)
+
 ```
+
+] is somewhat interesting.
+There were some issues with ].
+The number of arguments that it takes is somewhat related to the dimensions of the thing being indexed.
+For example, in data frames you can have a call to ] with 3 arguments: the data frame first, followed by the row and then column numbers (optionally).
+For lists, it's always just 2 arguments.
+Not sure how ... best to deal with this.
+Right now, I just have it take 2 arguments: the thing being accessed, and an index variable.
+
+There is perhaps room for a more sophisticateed analysis here, looking at the domensions of the first argument.
+I don't think it'll be too challenging, but again, do we want to go that far?
+I think that would be useful for some IDE tool or something: for instance, telling the programmer that they aren't using the thing with the right amount of dimensions.
+Then again, you can totally do that and it's completely fine...
+
+TODO: What package has the call with 1 argument?

### 2019-05-05 notebooks/rplus.Rmd
some new stuff

@@
 
-] is somewhat interesting.
-There were some issues with ].
+[ is somewhat interesting.
+There were some issues with [.
 The number of arguments that it takes is somewhat related to the dimensions of the thing being indexed.
-For example, in data frames you can have a call to ] with 3 arguments: the data frame first, followed by the row and then column numbers (optionally).
+For example, in data frames you can have a call to [ with 3 arguments: the data frame first, followed by the row and then column numbers (optionally).
 For lists, it's always just 2 arguments.
@@
 I think that would be useful for some IDE tool or something: for instance, telling the programmer that they aren't using the thing with the right amount of dimensions.
-Then again, you can totally do that and it's completely fine...
+Then again, you can totally do that and it's completely fine, you just might not get what you expect.
 
-TODO: What package has the call with 1 argument?
+Cool:
+src/main/names.c contains ~ function headers for builtin and special functions (minus types).
+Notably, it includes the arity of the functions.

### 2019-06-12 paper/rebuttal.txt
response

#5 A Large-Scale Study of Polymorphism in R

We thank the reviewers for their comments.

TAKEAWAYS

The R language is highly dynamic, and we set out to see if R code was as dynamic as the language allows. Our data suggest that while dynamism is prevalent, a majority of functions have a simple type signature.  We put forward three sets of potential type annotations for R. L0, a language of types which is closely related to R's existing runtime types, L1, which is designed to capture many of the polymorphic patterns that were observed in L0, and finally L2, which incorporates attributes and classes.

We will make our framework and data set available: our data contains over 460K call signature recordings from over 190K functions across over 11K packages.

CORPUS

Little end-user code is available online. Based on our experience, end-users do not use R as a general purpose language, they glue library calls together. Thus, we focus on analyzing libraries. Luckily, libraries are shipped with tests, examples and vignettes which showcase how they are used.

PARAMETRIC POLYMORPHISM

We have not included parametric polymorphism in the proposed type annotation languages because our intuition suggests that it may not help as much as in languages with user-defined types, and it may be more difficult to get adopted. We will consider it in a revision.

In case this paper were accepted, our plans for changes are as follows:

Typographical errors will be fixed.
We will add a section on parametric polymorphism in Sec.5.
We will add a section comparing our choices with those taken in other gradual language designs (e.g. TypeScript) in Sec.8.
Clarify and emphasize the takeaways of our study.

# Review #5A C/X

> The present paper, however, does not quite answer this question. The take-away message
that I read from this paper are: 1) L2 is an obvious candidate for classifying types, 2) in that 80% of the functions are monomorphic, and 3) 95% of the arguments are monomorphic (always instantiated by the same type). Yet, I am not quite sure what can be concluded from these points.

R highly dynamic and domain specific (data analysis), our hope was to find that actual uses of R are not as dynamic because the domain does not require it. This is what argument/function monomorphism tells us. For most arguments a simple type is enough. From the point of view of proposing a language of types that can be adopted by the community, we believe L1 is our best bet. While L2 is more expressive, it leads to more complex signatures (which we know to be a turn off) and its benefits are not as obvious (L1 gives the most direct performance benefits).

> One conclusion is that an R programmer would (in principle) be better off annotating with a type any function argument for which there exists an obvious type. The present study shows that in numerous place it would be straightforward to do so. (This remark would probably apply equally well to many other programming languages, btw.)

We agree with the last point -- our work quantifies that bit of common wisdom in one specific language.

> Yet, a proposed "type system" should strive to provide more than just type annotation for obvious monomorphic types. And, with that respect, the present paper does not make any explicit proposal.

True. We are not proposing a type system, at least not yet -- but rather a type annotation language that is sufficiently rich to express most observed function signature compactly. That type annotation language will drive the design of an actual (gradual) type system. Knowing our audience, our goal was the simplest language not the one that would be the most expressive.

> Even without going as far as proposing a full-blown type system, I would have found it interesting to evaluate how the handling in polymorphism in existing type systems for other languages (e.g. Typescript) could, or could not, be applied to handle the form of polymorphisms most commonly exploited in R code.

We will add a discussion in Sec.8.

> Another potential criticism is that, as the author rightfully acknowledge, there is a significant bias in the study due to the fact that the code base considered consists of reusable libraries, which differ quite significantly in flavor from what is commonly found in end-user R code. Thus, all measured figures must be taken with a grain of salt.

We have reason to believe (but no proof) that the code is representative. In our limited experience, R end-users do not define functions or write data types, they have in fact no CS training. So, our working hypothesis is that the majority of R “user” code is calls to libraries. We agree with the above grain of salt, but betting it is only a grain and not the whole shaker.

> Question: what about types of the form:   'a matrix -> 'a   (where 'a could be integer or double or complex) I do not understand how the types and shapes considered in the study cover such form of polymorphism.

The annotation language we have explored does not support bounded parametric polymorphism. Given the paucity of types in R, the proposals we have put on the table seem sufficient. More broadly, parametric polymorphism is tricky when the code performs explicit type tests.

>Typos:

Thanks.

# Review #5B A/X

> Can you measure what fraction of the 15 million lines of code are actually executed by the test suites you use, or otherwise discuss its coverage?   Please comment on this in the response.

In the ISSTA’18 paper by Krikava, the same set up with tests, vignettes, and examples for a corpus of 1.5K R packages exercised, on average, 68% of package code.

> 117: Explain 'c'.

‘c’ is a function for building vectors. c(1, 2, 3) is the vector [1, 2, 3].

> 187: This example doesn't explain what a data frame in general is, and later we need to know.

A data frame is a list of lists where each row represents an observation, and each column represents an observed characteristic. E.g., a data frame for students has a row for each person, and a column for their grades, etc.

> 236: It would be more useful to show the C data as a separate cumulative distribution, not a bar per package, especially as the real bars are presumably much less than one pixel wide.

We wanted to show the relationship between lines of R and of C code. We will revisit.

> 262: Are these signatures exactly the Fig. 2 argument+return types of the call?  Or do they include attributes, or attribute values, or maybe even vector lengths?  For functions with very large numbers of recorded signatures, can you explain why?  I see later you say this in Section 4, but it's needed here to understand the relationship between the number of function calls and the number of recorded signatures. Re-order the two parts?

The “unique signatures” are indeed unique function signatures (including all arg/ret shapes), with the shape defined in Sec.4. We will reorder. Functions with many recorded signatures typically have many (> 10) arguments, or simply have essentially any-typed arguments (e.g. the %>% pipe function accepts and returns anything). For details see Sec.7 (e.g. 591-594, 611-617).

> 400: Does "We omit short forms for..." just mean you don't have an abbreviated syntax, or are these "esoteric" types not in L0 at all?

They are in L0, we just didn’t shorten their names.

> 406: It looks as if L1 doesn't include any parametric polymorphism, e.g. to identify functions that are called on matching T and list(T) types, but it's not totally clear.  This is worth some discussion up-front, explaining why you only consider subtype polymorphism (if that's what you do).  I would have thought it was easy enough to try anti-unifying the observed signatures.

We held back from looking at parametric polymorphism, because our analysis cannot ensure that a function would indeed work for all possible types (unless it witnessed all types). This is related to reflection and run-time type checking in R, programmers test types of arguments and, for example, stop execution---our framework does not analyze the code, and thus cannot rule this out. Beyond that, we are rather confident that the R community would not accept a type system that they think of as “fancy” (if they accept any type system), and parametric polymorphism is likely “fancy” in their view. We will add a discussion elaborating on this in Sec.5.

> In the same vein, it would be useful to discuss explicitly whether you're doing any approximation, eg taking least upper bounds in your subtype orders.  If, based on your observations, you give an L0/L1/L2 type for some function, what assurance is there that the function will  work for any argument with that type?  If types exactly capture observations, then we'd know at least there is a test case, but if they approximate (which presumably they have to), we know less. Please discuss this in the response.

Each signature is based on an observation which had that exact signature. When taking a LUB, such as A = LUB(B,C), B and C are observed and A is always either B or C. Thus we never generalize to a type that has not been observed.

> 435: This gives an example, but it doesn't say what you actually do just "a certain amount".  Do you discard all observed types that are subsumed by another observed type?

Noted. We do discard subsumed signatures when simplifying.

> 442: Throughout this section, could you give the L0/L1/L2 typings for these examples?

Good suggestion, we will.

> 448: 402 says I<:D, but here it looks like D<:I too ?   And what does "strictly required" mean?

We will adjust the wording here. Indeed, I <: D. Line 448 is not an indication that D <: I, only that the bracket function [ is equipped to coerce doubles to integers. “Strictly required” here means: [ needs an integer argument, and will coerce doubles to integers.  (And the coercion is a runtime operation, and not subtyping.)

> 525: do you mean c(1,2) not c(1,1)?  And vectors can be regarded as matrices in two different ways (1xN vs Nx1)...

Yes, good catch. And indeed they can.

> 548: I can't make sense of this.

We will fix the wording in this paragraph.

> 560: are you counting \underbar{?} as monomorphic here?  That's confusing.

In L0, the only list type is “list”, and indeed we do not differentiate between all possible elements.

> 808: "would be" what?

Bad edit.  It’s unclear if a dependent type system would be well-received by R programmers.

> 837: "Is defined as" what?

Bad edit, this was intended to repeat from 5.3: for T1 <: T2, they must be subtypes in the sense of L1, and also the attribute and class lists of T1 are subsets of the attribute and class lists of T2.

> 840: It would be better to make these three tables somehow comparable without paging around.

Point taken. We will group them together.

> 1073-1075: I don't find this persuasive.  The "overall polymorphism" numbers are dominated by the monomorphic functions in all cases, so why is this evidence that "L2 captures the vast majority of existing polymorphism patterns"?  It seems quite possible that parametric (or maybe even dependent) types would capture existing patterns much better - I don't see that your data rules that out. Please discuss this in the response.

Given a set of observed signatures for each function, the “goodness” of an annotation language (L0-2) is determined by the tradeoff between compactness of the representation and precision of the types described. So a type is “better” if it capture more precisely the set of values and is more compact. Here we are thinking of ease of use (compactness) and benefits for compilation/verification (precision).

Adding parametric polymorphism to one of our systems will improve the compactness of type terms but come at a cost in complexity (and thus decrease the likelihood of adoption).

Also parametric polymorphism may be less useful in R where users do not define new types, and most of the values of interest are vectors of some basic types.

Lastly the presence of runtime type tests complicates life.

> 1084 This anonymous ("some work", etc.) style of discussing related work reads poorly.

Point taken.

# Review #5C D/Y

> Points against:  - I am not sure I agree with the authors' understanding of "polymorphism". This term is so fundamental to the work, it needs a crisp unambiguous definition early on in the paper. For that reason, throughout the paper, it remained unclear to me what exactly the study wanted to measure. I was hoping the analysis results would clarify this, but it did not. How were individual function signatures aggregated?

Our working definition of polymorphism is the “ad hoc” polymorphism attributed to Strachey [1967] and defined “as a kind of polymorphism in which polymorphic functions can be applied to arguments of different types, because a polymorphic function can denote a number of distinct and potentially heterogeneous implementations depending on the type of argument(s) to which it is applied.” [Wiki]

> What is an "argument signature" in Table 1–4? Why is `C, F` a polymorphic signature?

An argument signature here is a list of all the types that have inhabited a particular argument. E.g., if we observe two calls to a function: C->D and F->D, the signature for its argument is (C, F).  It is polymorphic because it has been called with both a character and a function.

> Why is the identity function `function(x) x` said to require a "sophisticated dependent type" rather than "polymorphic function arguments?

Typo.

>  I fail to see what ultimately learned from this study. There are a certain amount of polymorphic functions, and most functions are not used polymorphically. Ok, so what now? It would have helped to make the goals and hypotheses explicit.

This study is a step towards creating a full-fledged type system for R. We have established just how pervasive polymorphism is in R, through the lens of 3 potential type annotation languages. Not only do we know how many polymorphic functions are out there, we have brought forward the common usage patterns of these polymorphic functions. This information has allowed us to devise and subsequently evaluate candidate sets of types for which embody existing programming paradigms.

>  I find it problematic that the paper mixes the empirical study of types in R with the design of a specific R type system. The problem is that the study does _not_ generate knowledge about R itself but about the specific notion of type proposed in this paper. Thus, the empirical results do not inform the design of the type system, but instead try to confirm the design. That's a fine approach too, but the paper is sold as a study of R code.

We have struggled with this ourselves. The reviewer has a point. The reason we chose the paper’s approach is that since R does not have a type system it is difficult to present data about polymorphism. Polymorphism is only meaningful with a notion of type. So, we chose three type annotation languages and tried to fit them to the data.  Evidently, if the R community picks a different set of annotations, we will be able to fit that proposal too.

>  Corpus selection: Are the test cases and use cases of packages really a reliable source of representative R code? I would expect most package use cases demonstrate and focus on the functionality of a single package. I wouldn't be surprised if the use of more packages in the same program increases the likelihood of polymorphic function invocations, because different packages might use slightly different data representations.

We believe that they are representative. Anecdotally, based on looking at notebooks stored on public repositories and our own code, we believe that most “user” code merely glues libraries together. They don’t define many functions or any new data types.  The package tests are more like use-cases and show representative uses of the package. So they are not a bad starting point.

> The raw data should be made available.

All of our data and code is open source. We did not link it because we ran out of time to anonymize the code and find a place where to store all of the data. Our intent is to have it ready for artifact evaluation in case our paper is accepted.

> Section 4, Fig 6: It seems to me standard functions like `+`, `c`, and `<-` should be excluded from the study. In other languages these constructs would be primitive syntactic forms with an exceptional high polymorphic usage.

In R, all functions can be redefined.  So there is no guarantee that + is standard.
This said, we want to know how polymorphic these “worst case” scenarios looked like.

> Section 6: I did not understand the point of this section. I also did not understand what about these examples was polymorphic exactly. For example, how is the code in Fig 8 polymorphic? Wouldn't it permit a standard monomorphic type annotation `w: int vec`? Maybe the code snippets in Section 6 can be used to motivate this work directly after Section 2?

Fig. 8 is polymorphic as it takes either the value NULL or a vector.  NULL is not a “subtype” of vector, it is an unrelated value of a different kind. Sometimes, R libraries use the value NA for the same purpose, and in other cases they may pick other sentinels.

> Section 7, line 548ff: I couldn't make sense of this paragraph.

We will fix the wording in this paragraph.

> Table1: What is an argument signature?

An argument signature here is the types that have inhabited a particular argument. E.g., if we observe two calls to a function: D->D and I->D, the signature for its argument is (D, I).

> Sec 7.1, "each of the most frequently occurring argument signatures had some explanation": more like a guess, at best a potential explanation. Such interpretation of the data should be clearly separated from the data itself.

Point taken.

> Sec 7.4, "the most polymorphic functions in R are meta-programming their arguments." What does that mean "meta-programming their arguments"?

In R, arguments are passed in a lazy fashion, thus a function can take any of its arguments, and obtain the source code of that argument and its environment, then the function can modify that code and that environment before executing it. This is how, for instance, control structures can be implemented.

> Sec 8.2, "missing": Isn't this also an artifact of the corpus being too small?

We have seen R functions with up to 80 arguments. It is not too surprising that some arguments may not be evaluated. Some programs have dead code, R has dead arguments.

> Fig 21: What is the point of this code?

It shows one implementation of ad hoc type checking in user code that causes the program to stop if the argument is neither NULL or Character.

### 2019-07-30 paper/main.tex
light modifications to some data files, new notebook, and new data
new format

@@
-\documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[sigconf,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 \usepackage{booktabs,listings,xspace,wrapfig}
@@
 
+\section{ICSE TODO}
+
+For one, the page limit is way less.
+This is probably fine. 
+I took out the graphs for now, maybe we`ll just want one giant page for all the figures?
+We can figure out where to place them.
+
+We want to beef up the contribution.
+Some low hanging fruit include a call-site analysis, and developing a metric to judge whether we trust a suggested annotation. 
+This might involve a rerun, for the call site stuff.
+We should think of anything else we want in the rerun before I spend the cycles to do it.
+
+Broadly, I think we should focus on the study and what we learned from it.
+There are some fluffy sections (like 8.2) that can probably be removed.
+
+We probably want something to generate code from the type annotations we have. 
+I'm working on a prototype.
+
+Better justify the experiment.
+Why is looking at tests, examples, vignettes ok?
+
+Address parametric polymorphism.
+
 \section{Introduction}
@@
 
-\newpage  %%Leave here
+% \newpage
 
@@
 
+\begin{comment}
+
 \begin{figure}[!b]\begin{center}
@@
 \end{figure}
+
+\end{comment}
 
@@
 values, the types of these function call are often similar. Over the entire
-corpus, we can see the relation between observed and recorded calls in
+corpus, we can see the relation between observed calls and recorded signatures in
 Figure~\ref{recorded}.  The median number of observed calls is 82 and
@@
 very few calls before plunging into C code.
+
+\begin{comment}
 
@@
 
-\newpage
+\end{comment}
+
+% \newpage
 \section{Methodology}
@@
 
+\begin{comment}
+
 \begin{figure}[htbp]\begin{center}

### 2019-09-23 paper/main.tex
new vers

--- 2019-04-06 paper/main.tex
@@
-\documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 \usepackage{booktabs,listings,xspace,wrapfig}
@@
 
-% Macros for type names?
-\newcommand{\reals}{$\mathtt {\rm I\!R}$\xspace}
-\newcommand{\realv}{$\overline{\rm I\!R}$\xspace}
-
 \setcopyright{none}
-%\setcopyright{acmcopyright}%\setcopyright{acmlicensed}
-%\acmDOI{10.475/123_4}
-%\acmConference[OOPSLAs]{Woodstock conference}{July 1997}{El Paso, Texas USA}
-%\acmYear{1997}%\copyrightyear{2016}%\acmPrice{15.00}
 \begin{document}
 
-\title{A Large-Scale Study of Polymorphism in R}
-
-\newcommand{\PACKAGES}{11,463\xspace}
-\newcommand{\PROGRAMMERS}{?\xspace}
-\newcommand{\PERCENTCRAN}{83\%\xspace}
-\newcommand{\CRANTOTAL}{13,841\xspace}
-\newcommand{\RLOC}{15,050,267\xspace}
-\newcommand{\CLOC}{9,373,542\xspace}
-\newcommand{\YEARS}{20\xspace}
-\newcommand{\INDEXCOINCIDENCE}{3,499\xspace}
-\newcommand{\TOTALINDEXY}{6,561\xspace}
-\newcommand{\INDEXYPERC}{53\%\xspace}		% TODO Not an accurate count.
-\newcommand{\DATAPKGS}{206\xspace}
-\newcommand{\DATAPKGSPERC}{1.5\%\xspace}
-\newcommand{\METAARGCOUNT}{7,051\xspace}
-\newcommand{\NUMPOLYTYPEMONOCLASS}{1,357\xspace}
-\newcommand{\PERCSCALARMONO}{38.2\%\xspace}
-
-\newcommand{\LZEROPERCPOLY}{19\%\xspace}
-\newcommand{\LONEPERCPOLY}{18.1\%\xspace}
-\newcommand{\LTWOPERCPOLY}{21.3\%\xspace}
-
-\newcommand{\attr}[2]{\ensuremath{#1_{\mathtt{#2}}}\xspace}
-\newcommand{\attrclass}[3]{\ensuremath{#1^{\mathtt{#3}}_{\mathtt{#2}}}\xspace}
-\renewcommand{\to}{\ensuremath{\rightarrow}\xspace}
-\newcommand{\D}{\ensuremath{\small\vec{\mathtt D}}\xspace} % Double
-\newcommand{\I}{\ensuremath{\small\vec{\mathtt I}}\xspace} % Integer
-\renewcommand{\C}{\ensuremath{\small\vec{\mathtt C}}\xspace} % Character
-\renewcommand{\L}{\ensuremath{\small\vec{\mathtt L}}\xspace} % Logical
-\newcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
-\newcommand{\X}{\ensuremath{\small\vec{\mathtt X}}\xspace} % Complex
-\newcommand{\Y}{\ensuremath{\small\vec{\mathtt Y}}\xspace} % Symbol
-\newcommand{\sY}{\ensuremath{\small{\mathtt Y}}\xspace} % Symbol
-\newcommand{\sS}{\ensuremath{\small{\mathtt S}}\xspace} % S4
-\newcommand{\sF}{\ensuremath{\small{\mathtt F}}\xspace} % Closure
-\newcommand{\sE}{\ensuremath{\small{\mathtt E}}\xspace} % Env
-\renewcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
-\newcommand{\sN}{\ensuremath{\small{\mathtt N}}\xspace}     % Null
-%\renewcommand{\l}{\ensuremath{\small L<?>}\xspace}     % List
-\renewcommand{\l}{\ensuremath{\small\underline{\mathtt ?}}\xspace}     % List
-\newcommand{\sD}{\ensuremath{\small{\mathtt D}}\xspace} % Double
-\newcommand{\sI}{\ensuremath{\small{\mathtt I}}\xspace} % Integer
-\newcommand{\sC}{\ensuremath{\small{\mathtt C}}\xspace} % Character
-\newcommand{\sL}{\ensuremath{\small{\mathtt L}}\xspace} % Logical
-\newcommand{\sX}{\ensuremath{\small{\mathtt X}}\xspace} % Complex
-\newcommand{\sR}{\ensuremath{\small{\mathtt R}}\xspace} % Raw
-\newcommand{\ANY}{\ensuremath{\small{\mathtt ?}}\xspace}     % Any
-\newcommand{\lT}[1]{\ensuremath{\small\underline{\mathtt{#1}}}\xspace}     % list<T>
-\newcommand{\M}[1]{\ensuremath{\attr{\vec{\tt #1}}{mat}}\xspace}     % matrix
-\newcommand{\df}{\ensuremath{\attr{\l}{df}}\xspace}     % data.frame
+\title{A Type System for R: Empirically Designed and Evaluated}
 
 \begin{abstract}
-The R programming language is widely used in a variety of scientific domains
-for tasks related to data science. The language was designed to favor an
-interactive style of programming with minimal syntactic and conceptual
-overhead. This design is well suited to support interactive data analysis,
-but is not well suited to generating performant code or catching programming
-errors.  In particular, R has no type annotations and all operations are
-dynamically checked at runtime. The starting point for our work is the
-question: \emph{what could a static type system for R look like?}  In order to answer
-it, we study the polymorphism that is present in \RLOC lines of R
-code spread among some \PACKAGES packages, written over a
-period of over \YEARS years by thousands of programmers.  We perform a dynamic
-analysis, leveraging tests and use-cases, to determine the level of
-polymorphism that is present in the code. We do this for several potential
-notions of types. Our results suggest that polymorphism is important in some
-key parts of the system but that relatively simple type annotations could be
-used to capture most of the interesting cases.
+
+\AT{PLDI TODO:}
+
+\begin{itemize}
+
+\item complete propagatr to generate constraints
+
+\item feed existing data through resulting constraints
+
+\item finish the annotation and contract generation code
+
+\item write the paper
+
+\end{itemize}
 
@@
 \maketitle
-
-\section{Introduction}
-
-Our community builds, improves, and reasons about programming languages.  To
-make design decisions that benefit most users, we need to understand the
-language we are working with as well as the real-world needs it
-answers. Often, we as researchers can appeal to our intuition, as many
-languages are intended to be general purpose and attract users with some
-computer science training. Unfortunately, these intuitions do not always
-apply to domain-specific languages, languages designed for and by a specific
-group of users to solve very specific needs. This is the case of the data
-science language R.
-
-R and its ancestor S are languages designed, implemented, and maintained by
-statisticians. Originally they were designed as glue languages, languages
-that would allow to read data into vectors and call statistical routines
-written in Fortran. Over three decades, the languages became widely used
-across many fields of science and in industry for data analysis and data
-visualization; with time additional features were added.  Modern R, as a
-linguistic object of study, is fascinating. It is a vectorized, dynamically
-typed, lazy functional language with limited side-effects, extensive
-reflective facilities and retrofitted object-oriented programming support.
-
-Many of the design decisions that gave us R were intended to foster an
-interactive, exploratory, programming style. These include, to name a few,
-the lack of type annotations on variables and functions, the ability to use
-syntactic shortcut, and the automatic conversion between data types.  While
-these choices have led to a language that is surprisingly easy to use for
-beginners---many data science programs do not teach the language itself but
-simply introduce some of its key libraries---they have also created a
-language where almost all computations yield a numeric result and where
-errors can go undetected.
-
-One way to increase assurance in the results obtained when using R would be
-to add type annotations to functions and variable declarations. These
-annotations could then be used, either statically or (more likely)
-dynamically, to catch mismatches between expected and provided data values.
-The nature of R is such that it is unlikely to be ever fully statically
-checked, furthermore end users may not be willing to write types when
-carrying out exploratory programming tasks. So, we are looking for an
-optional type system that would allow us to capture as much of behavior of
-library functions as possible while remaining easy to understand for
-end-users and library developers alike.
-
-This paper is a data-driven study of what a type system for the R language
-could look like. Our intention is to propose changes to the
-language, and for any changes to be accepted by the user community they
-must clearly benefit the language without endangering backwards
-compatibility. Our goal is thus to find a compromise between simplicity and
-usefulness; the proposed type system should cover most common programming
-idioms while remaining easy to use. In order to do this, we need to
-understand the degree of polymorphism present in R code, that is to say, how
-programmers leverage the dynamic nature of R to write code that can accept
-arguments of different types.  This understanding will drive our design.
-
-We propose to capture the degree of polymorphism present in R by means
-of a dynamic analysis of widely used libraries. For each function call we
-can record the types of its arguments and of its return value. This allows
-us to observe how many different combination of types are accepted by any
-given function. Unlike many other languages, R has a carefully curated
-software repository called CRAN. To be deposited in CRAN, a package must
-come with sample dataset, tests and executable use-cases. As part of normal
-operations these tests are run regularly and failing packages are removed.
-This allowed us to have access to \PACKAGES libraries and about an order of
-magnitude more runnable scripts that exercise those libraries.
-
-The contributions of this paper are thus as follows:
-\begin{itemize}
-\item A large-scale analysis of the polymorphism present in function
-  signatures of \PACKAGES widely used and actively maintained R packages.
-\item A tracing and analysis pipeline that extends a previously published
-  test generation tool named \genthat.
-\item A set of type annotations that captures how programmers use types
-in existing R code.
-\end{itemize}
-
-One threat to validity of our work is that we rely on dynamic analysis, so
-our conclusions are only as good as the coverage of the possibly function
-behaviors. Previous work~\cite{issta18} reported that running all the
-scripts that come with CRAN packages gives, on average, 68\% test coverage.
-We attempted to mitigate the threat coming from the fact that only part of
-the code is being exercised by manual analysis. It would be reasonable to
-ask for confirmation of the data by static analysis of the code, but sound
-static analysis of R is difficult because of the extensive use of reflective
-features such as \code{eval} and of the ability to redefine the meaning of
-operators such as \code{+} and \code{if}.  Another threat to validity is
-that we only have access to code that has been deposited in the CRAN
-repository. While this may bias our findings towards code written to be
-reusable and, possibly, better engineered than typical user code. This is
-also the code that would most benefit from type annotations.
-
-\newpage  %%Leave here
-
-\section{The R Programming Language}\label{sec:rlang}
-
-Over the last decade, the R Project has become a key tool for implementing
-sophisticated data analysis algorithms in fields ranging from Computational
-Biology~\cite{R05} to Political Science~\cite{R:Keele:2008}. At the heart of
-the R project is a \emph{vectorized, dynamic, lazy, functional,
-  object-oriented} programming language with a rather unusual combination of
-features~\cite{ecoop12} designed to ease learning by non-programmer and
-enable rapid development of new statistical methods.  The language, commonly
-referred to as R was designed in 1993 by Ross Ihaka and Robert
-Gentleman~\cite{R96} as a successor to S~\cite{S88}.  First released in
-1995, under a GNU license, R rapidly became the lingua franca for
-statistical data analysis. Today, there are over 13,000 R packages available
-from repositories such as CRAN and Bioconductor.  With 55 R user groups
-world-wide, Smith~\cite{eco11} estimates that there are over 2 million
-end-users.
-
-As an introduction to R, consider the code snippet in Fig.~\ref{sample} from
-a top-level interaction where the user defines a function \code{normSum}
-that accepts vectors of integers, logicals, doubles and complex values and
-normalizes the vector with respect to its sum and rounds the results. The
-function definition does not require type annotations, and all operations
-transparently work on vectors of any length and different types.
-
-\begin{figure}[!hb]{\small
-\begin{lstlisting}[style=R]
-> normSum <- function( m )  round( m / sum(m), 2)
-> normSum(c(1L,3L,6L))
-[1] 0.1 0.3 0.6
-> normSum(c(1.1,3.3,6.6))
-[1] 0.1 0.3 0.6
-> normSum(c(1.6,3.3,6.1))
-[1] 0.15 0.30 0.55
-> normSum(complex(r=rnorm(3),i=rnorm(3)))
-[1] 0.49+0.21i 0.30-0.18i 0.22-0.03i
-\end{lstlisting}}
-\caption{Sample R code}\label{sample}
-\end{figure}
-
-In R, function can be called with named parameters, R support variable
-argument lists, and arguments can have default values. Putting all of these
-together consider the following declaration:
-
-\begin{lstlisting}[style=R]
-f <- function(x, ..., y=3) x + y
-\end{lstlisting}
-
-\noindent
-Function \k{f} can be called with a single argument \code{f(3)}, with named
-argument \code{f(y=4,x=2)} and with a variable number of arguments,
-\code{f(1,2,3,4,y=5)}, all of these calls will return \code{6}.
-
-R has a number of features that are not crucial to the present
-discussion. We will mention some of them here for completeness.  In R, data
-structures are reference counted and have copy-on-write semantics, thus the
-assignment \code{x[12]<-3} results in an update to a copy of \code{x} unless
-the reference count on that object is 1.  This semantics gives R a
-functional flavor while allowing updating in place within loops (the first
-update copies, subsequent updates are performed on the copy). Arguments to
-functions are evaluated only when needed, they are bundled in so-called
-promises which package the original expression (as an abstract syntax tree, or AST), its environment
-as well as the result of evaluating the expression. Promises can be
-leveraged for meta-programming as it is possible to retrieve the text of a
-promise and evaluate it in a different environment.
-
-\subsection{Types of Data}
-
-Before attempting to define a type system for R, we should understand the
-different kinds of values that programs operate on.  As we will see
-different notions of type may emerge depending on how granular we want to
-be.
-
-\renewcommand{\k}[1]{{\tt #1}\xspace}
-
-R has one builtin notion of type that can be queried by the \k{typeof}
-function. Over the years, programmers have found the need for a richer type
-structure and have added {\it attributes}. The best way to think of attributes is
-as an optional map from name to values that can be attached to any object.
-Attributes are used to encode various type structures. They can be queried
-with functions such as \k{attributes} and \k{class}.
-
-\begin{wrapfigure}{r}{6.1cm}
-\footnotesize\begin{tabular}{l|c|l@{}}\hline
-\multicolumn{3}{l}{\bf Vectorized data types:}  \\\hline
-\k{logical}   & \L & vector of boolean values\\
-\k{integer}   & \I & vector of 32 bit integer values\\
-\k{double}    & \D & vector of 64 bit floating points\\
-\k{complex}   & \X & vector of complex values\\
-\k{character} & \C & vector of strings values\\
-\k{raw}       & \R & vector of bytes\\
-\k{list}      & \l & vector of values of any type\\\hline
-\multicolumn{3}{l}{\bf Scalar data types:}\\\hline
-\k{NULL}      & \sN &  singleton null value\\
-\k{S4}        & \sS &  instance of a S4 class \\
-\k{closure}   & \sF & a function with its environment\\
-\k{environment}&\sE &  a mapping from symbol to value \\\hline
-\multicolumn{3}{l}{\bf Implementation data types:}\\\hline
-\multicolumn{3}{l}{\k{special},
-\k{builtin},
-\k{symbol} (\sY),
-\k{pairlist},
-\k{promise}}\\
-\multicolumn{3}{l}{
-\k{language},
-\k{char},
-\k{...},
-\k{any},
-\k{expression},
-}\\
-\multicolumn{3}{l}{
-\k{externalprt},
-\k{bytecode},
-\k{weakref}}\\\hline
-\end{tabular}\caption{Builtin Types}\label{types}\end{wrapfigure}
-
-Figure~\ref{types} lists all of the builtin types that are provided by the
-language. They are the possible return values of \k{typeof}. There is no
-intrinsic notion of subtyping in R. But, in many context a \k{logical} will
-convert to \k{integer}, and an \k{integer} will convert to \k{double}.  Some
-off conversion can occur in corner cases, such as \k{1<"2"} holds and
-\k{c(1,2)[1.6]} returns the first element of the vector, as the double is
-converted to an integer. R does not distinguish between scalars and vectors
-(they are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
-  typeof(c(5,5))} \code{ == "double"}. Finally all vectorized data types have a
-distinguished missing value denoted by \code{NA}. The default type of
-\code{NA} is \k{logical}. We can see that \code{typeof(NA)=="logical"}, but
-NA inhabits every type: \code{typeof(c(1,NA)[2])=="double"}.
-
-With one exception all vectorized data types are monomorphic, the exception
-is the \k{list} type which can hold values of any other type including
-\k{list}. For all monomorphic data types, attempting to store a value of a
-different type will cause a conversion. Either the value is converted to the
-type of vector, or the vector is converted to the type of the value.
-
-Scalar data types include the distinguished \k{NULL} value, which is also of
-type \k{NULL}, instance of classes written using the S4 object system,
-closures and environments.  The implementation of R has a number of other
-types that are mostly not used by user code, they are listed in
-Figure~\ref{types} for reference.
-
-The addition of attributes lets programmers extend the set of types by
-tagging data with user-defined attributes. For example, one could define a
-vector of four values, \code{x<-c(1,2,3,4)} and then attach the attribute
-\k{dim} with a pair of numbers as value: \code{attr(x,"dim")<-c(2,2)}.  From
-that point, arithmetic functions will treat \k{x} as a 2x2 matrix. Another
-attribute that can be set is the \k{class}.  This attribute can be bound to
-a list of class names. For instance, \code{class(x)<-"human"}, set the class
-of \k{x} to be \k{human}.  Attributes are thus used for object-oriented
-programming. The S3 object system support single dispatch on the class of
-the first argument of a function, whereas the S4 object system allows
-multiple dispatch (on all arguments). Some of the most widely used data
-type, such as data frames, leverage attributes. A data frame, for instance,
-is a list of vectors with a class and a column name attribute.
-
-\paragraph{Summary.} The most common values in R computations are vectorized
-types. R programs do not have a way to constrain values to be scalar.
-\k{NULL} is sometimes used to represent the case when no value is
-available. \k{NA} is used within vector to represent missing observations.
-Attributes can decorate values and are used as building blocks for
-object-oriented programming. A potential type system for R could focus only
-on the builtin types, if one wanted to strive for simplicity, or it could
-try to capture attributes at the risk of increased complexity.
-
-\newpage
-\section{Corpus}\label{sec:corpus}
-
-In this section, we present our dataset. The R language aims to accommodate
-data analysts; their workflows start with data import, followed by cleaning,
-and then by steps of modeling, transformation and visualization. Often, the
-code of these analysis pipelines resides, together with the data and
-results, in notebooks. Few notebooks are publicly shared, and when they are,
-the data is not. For this reason our analysis focuses on packages which
-bundle reusable units of R code with documentation, sample data and
-use-cases.
-
-We focus on packages hosted on the \emph{Comprehensive R Archive Network} or
-CRAN.  With over 13,000 packages, CRAN is the largest repository of software
-written in R. It is experiencing sustained growth with an average of size
-new packages a day~\cite{LIgges2017}.  Unlike sites like GitHub, CRAN is a
-\emph{curated} collection: A package is only accepted to CRAN if it abides
-by a number of well-formedness rules.  Most relevant for our purposes,
-packages must have data, examples, vignettes and tests, all of which must
-successfully run. From our perspective this means that each package in CRAN
-comes with several executable scripts that exercise some of its
-functionality.  Notable exceptions to this rule are packages only containing
-data, which have no runnable code but are referenced by other packages.
-Only \DATAPKGS packages had no executable code, accounting for \DATAPKGSPERC
-of CRAN.
-
-The corpus used in this paper is a subset of CRAN. We retained packages that
-could be run by our infrastructure in less than one hour. This corpus
-consists of \PACKAGES packages, accounting for some \PERCENTCRAN of all
-packages.  These packages have a total of \RLOC lines of R code and \CLOC
-lines of C code. Figure~\ref{allcloc} shows a per-package breakdown of the
-size of each package sorted by increasing numbers of lines of R. The figure
-suggests that there is little correlation between the C/C++ parts and how
-many lines of R a package contains. The median size of a package is 541
-lines of R code and the largest package has 86K LOC. Typically, C/C++ code
-is used to implement performance critical portions of the code. The majority
-of packages, 8,375 to be precise, have no C or C++ code.  The remaining 3,078
-packages have a median 572 lines of C and the largest package has 385,839
-lines of C.
-
-\begin{figure}[!b]\begin{center}
-\includegraphics[width=.9\textwidth]{linesofrandccode}
-\caption{Lines of code (log scal); for each package, R is above 0, and C/C++
-  below}\label{allcloc}\end{center}
-\end{figure}
-
-For each package, we extracted all executable code snippets from
-documentation, vignettes and tests and ran them independently, recording all
-calls to R functions.  It is noteworthy that in order to run the scripts in
-one package, it is often necessary to load a number of other packages.
-In~\cite{issta18}, the authors estimated code coverage to be around 68\%
-when including reverse dependencies.  As our infrastructure adds overhead to
-script execution, coupled with the fact that some scripts take inordinate
-amounts of time to run, we limited our analysis of any one
-package to one hour.
-
-% https://www.r-pkg.org/downloaded
-\begin{figure}[!th]{\footnotesize\begin{tabular}{@{}r||l|r|r|r|r|r@{}}\hline
-\bf Package & \bf Description & \bf R LOC &\bf C LOC &\bf Scripts & \bf Calls Observed & \bf Signatures Recorded \\
-\hline
-\tt Rcpp  & Seamless C++ integration & 2.2K & 4.2K & 25 & 55K & 340 \\
-\tt rlang & Functions for 'Tidyverse' & 7.0K & 6.1K & 122 & 3,924K & 8,422 \\
-\tt glue  & Interpreted string literals & 0.3K & 0.3K & 8 & 4K & 145 \\
-\tt tibble & Simple data frames & 2.0K & 0.3K & 16 & 1,332K & 6,367 \\
-\tt stringi &  String processing & 1.5K & 515K & 64 & 923K & 873 \\
-\tt ggplot2 & Data visualisations & 14K & 0 & 130 & 153K & 4,608 \\
-\tt dplyr  &  Data manipulation & 4.5K & 4.7K & 78 & 233K & 3,099 \\
-\tt pillar & Formatting for columns & 1.4K & 0 & 13 & 803K & 1,514 \\
-\tt R6 & Classes w. ref. semantics & 0.7K & 0 & 2 & 1K & 330 \\
-\tt stringr & String operations & 0.5K & 0 & 32 & 1,764K & 534 \\
-\end{tabular}}\caption{10 Most Downloaded Packages.}\label{most}
-\end{figure}
-
-Figure~\ref{most} shows the ten most downloaded CRAN packages.  For each
-one, we list how many lines of R and C/C++ the packages contains.  We show
-the number of scripts that could be extracted from the package. Each script
-corresponds to either one use-case or a set of unit tests. We print the
-number of function calls that were observed by our infrastructure, and the
-number of unique call signatures that were recorded.
-
-Our infrastructure only retains unique argument/return combinations. Thus,
-while we observe large number of functions being called with different
-values, the types of these function call are often similar. Over the entire
-corpus, we can see the relation between observed and recorded calls in
-Figure~\ref{recorded}.  The median number of observed calls is 82 and
-maximum is 19 million.  The median number of recorded signatures is 16 and
-the maximum is 8,422. These numbers are skewed by a number of scripts doing
-very few calls before plunging into C code.
-
-\begin{figure}[htbp]\begin{center}
-\includegraphics[width=.9\textwidth]{recordsbypkg}
-\caption{Numbers of Recorded Function Invocations in the Analyzed Corpus}
-\label{recorded}\end{center}
-\end{figure}
-
-\newpage
-\section{Methodology}
-
-In this section, we detail our methodology for collecting data.  Our aim is
-to observe arguments and return values of function calls, and from these
-generalize possible type signatures for the called functions.  We base our
-infrastructure on an open source tool called \genthat whose purpose is to
-generate unit tests for R libraries~\cite{issta18}.  \genthat achieves this
-by synthesizing unit tests from recorded function argument and return
-values, comparing candidate tests against existing ones to avoid generating
-tests which do not increase code coverage.  To suit our purposes, we change
-the existing tool in two main ways: we record \emph{shapes} rather than
-values, and ignore the code coverage optimization phase.  Both changes are
-beneficial for scalability, allowing us to trace far more calls.
-
-To illustrate our approach, consider the following script which adds a
-double to an integer, and then creates a matrix from a vector, finally
-adding them together to get a new matrix.
-
-\begin{figure}[!hb]
-\begin{tabular}{ll}\begin{minipage}{5cm}
-{\small\begin{lstlisting}[style=R]
-> 1 + 2L
-[1] 3
-> x <- c(1,2,3,4)
-> attr(x,"dim") <- c(2,2)
-> x + c(1,2)
-     [,1] [,2]
-[1,]    2    4
-[2,]    4    6
-\end{lstlisting}}
-\end{minipage} &
-\begin{minipage}{8cm}\small
-\begin{tabular}{rl}
-\tt `+`: &\tt \sD \sI \to \sD \\
-\tt c:& \sD \sD \sD \sD \to \D\\
-\tt `<-`: &\tt \sY \D \to \D\\
-\tt c: & \tt \sD \sD \to \D\\
-\tt `attr<-`:&\tt  \sY \C \D \to \\
-\tt c: &\tt \sD \sD \to \D\\
-\tt `+`: &\tt \attr\D{dim} \D \to \attr\D{dim}
-\end{tabular}
-\end{minipage}
-\end{tabular}
-\caption{Example script and recorded signature}\label{example}\end{figure}
-
-The functions being called here are \k{`+`}, \k{`<-`}, \k{c} and
-\k{`attr<-`}, and the shapes we expect to record are given to the right of the figure.
-We abbreviate the types \k{double}, \k{integer}, \k{symbol} and \k{character} to \sD, \sI, \sY, and \sC for simplicity.
-\D, \I, \Y, and \C denote vectors, whereas \sD, \sI, \sY, and \sC denote scalars.
-We record the lengths of all vectorized-typed values, and for all types we record their classes and attribute names. 
-Looking at the signatures observed for addition, it is clear that the function is polymorphic: it starts with the addition of two scalar numbers of different types, and later adds a matrix to a vector returning a matrix.
-
-\subsection{Implementation}
-\label{sec:Impl}
-
-A high-level description of the workflow of our tool for one package
-retrieved from CRAN is as follows:
-
-\begin{enumerate}
-\item {\bf Exec generation:} All runnable code in the package is extracted
-  from its tests, examples and vignettes. The code snippets are combined
-  into a single file.
-\item {\bf Installation:} All packages that are required for execution of
-  the current package are downloaded and installed.
-\item {\bf Instrumentation:} As code is loaded into the R, every function
-  definition is instrumented with an \code{on-exit} hook which is invoked
-  when the function returns either normally or through an exception.
-\item {\bf Recording:} When a hook is called, arguments and return value are
-  inspected. We record \k{typeof}, \k{class} and \k{attributes} recursively.
-  For \code{list} values, an extra bit of analysis is performed to record
-  the element types.
-\item {\bf Writing:} Unique signatures are recorded to file with information
-  about which package triggered the recording.
-\end{enumerate}
-
-In R, all arguments are passed as promises, and unused arguments will remain unevaluated promises, which we record as {\it unevaled}. 
-Our recording mechanism captures the order in which arguments are passed, can figure out which arguments were {\it not} provided (recording them as {\it missing} if the missing argument has no default value), and does {\it not} force unforced promises.
-In R, trying to use missing values results in an error and thus an exception, triggering our recording phase: arguments which cause errors in this way are recorded as {\it error}.
-For practical reasons, we do not record the contents of environments:
-These can be used as hash tables, may be large, and are quite likely different from one another.
-
-Unevaluated arguments in R are sometimes {\it carefully} used by their functions to implement some form of meta-programming.
-Consider the {\tt magrittr} package, which implements a pipe function \%$>$\% that implements a flavor of function currying:
-
-{\small\begin{lstlisting}[style=R]
-mul <- function(x,y)  x*y
-4 %>% mul(3)
-\end{lstlisting}}
-
-The call to the pipe desugars to \code{`\%>\%`(lhs=4, rhs=mul(3))} where the
-\code{rhs} is not a valid expression. It is treated as code that will
-manipulated to add one argument (the \code{lhs}), resulting in the function
-actually performing a call of the form \code{mul(4, 3)}.
-
-As we alluded to at the beginning of this section, our analysis collects shape information about
-arguments and return values in order to be as specific as possible. The details of the information we collect follows.
-Note that the first step of any type collection in our analysis begins with \code{typeof}.
-
-\begin{itemize}
-\item If the analysis encounters a primitive vector, say
-  \code{typeof(x)=="double"}, it records the length of {\tt x} to determine
-  if it can be treated as a scalar.
-\item For a list, \code{typeof(x)=="list"}, shape information on all
-  elements is used to describe the list's shape.  To avoid undue slowdowns,
-  we only collect the content's \code{typeof}, and ascribe \ANY (any) if the
-  elements are not of the same kind.
-\item If the analysis encounters a scalar \code{NA}, we ascribe a unique
-  NULL type.  In R, \code{NA} inhabits all types, but for a scalar NA,
-  \code{typeof(NA)=="logical"}. Scalar NAs are used as uninitialized values.
-\item For a matrix, \code{typeof(m)=="some\_primitive\_type"}, and\code{m} will have matrix class and also a dims and
-  optionally dimnames attribute. Here, we ascribe the shape \attr{\tt T}{mat}
-  where {\tt T} is the primitive element type.
-\item For a data frame, \code{typeof(df)=="list"}, and \code{df} will have class {\tt data.frame}, and {\tt names} and {\tt row.names} attributes.
- To reflect this, we ascribe the shape \attr{\l}{df}.
-\end{itemize}
-
-The infrastructure memoizes emitted signatures to avoid writing multiple
-identical signatures for the same function.
-
-\section{Types for R}
-
-Our aim is to capture the polymorphism in R, and to do so we must define a
-meaningful set of basic types for the language. Also, we are interested in
-finding a type system that could be adopted, for this we aim to investigate
-the simplest possible solution.  There are several possible choices, with a
-variety of levels of expressivity and complexity.  For concreteness, we will
-pick out three different sets of types.  To start simple, we define a set of
-types which closely align with R's builtin types, and later define richer
-notions of type which incorporate attributes, classes, and capture the
-various object systems of R.
-
-Throughout this paper, we will say that an argument is {\it polymorphic at}
-\textbf{\emph{LX}} (read: level X) if it has been called with at least two
-values that belong to distinct types at {\bf LX}.  \textbf{L2} will
-incorporate class and attribute lists, which are slightly trickier to
-``type'', and we defer discussion of this until Section~\ref{sec:L2def}.
-
-\subsection{L0 Types}
-\label{sec:L0def}
-
-The logical ``baseline'' for the design of a type system for R is to adopt
-the language's builtin types.  We dub these {\bf L0} types, and they are
-primarily built on the \code{typeof} values, which reflects the
-implementation of the type in the runtime environment.  More specifically,
-the \code{typeof} a value is simply the runtime type tag of the value.
-Given that, {\bf L0} includes the following types: vectors of primitive
-types (integers \I, double \D, character \C, logical \L, raw \R, complex
-\X), lists of any type (\l), and scalar values (environment \sE, closures
-\sF, symbols \sY, S4 object \sS, and null \sN). We omit short forms for some
-of R's more esoteric types (e.g., {\tt environment}s).  We add the any
-(written \ANY) type to stand in for unevaluated values Subtyping in {\bf L0}
-follows from the conversion rules of the language: Logicals are subtypes of
-integers, which are themselves subtypes of doubles (so \L <: \I <: \D).  All
-types are subtypes of \ANY, thus {\tt T}<:\ANY for an type {\tt T}.
-
-%
-\subsection{L1 Types}
-\label{sec:L1def}
-
-{\bf L0} is limited in expressivity as it (1) supports polymorphism for
-lists in a trivial way, (2) does not address some of the popular constructed
-data types such as matrices and data frames, and (3) cannot differentiate
-between scalar values and vectors.  {\bf L1} types build on and extend {\bf
-  L0} with with scalar values for primitive types (\sI, \sD, \sC, \sL, \sR,
-and \sX), type-parametric lists (written \lT{T} for any type {\tt T}), data
-frames (written \attr{\l}{df} as they are built up from lists of type any),
-and matrices (\attr{\vec{\tt T}}{mat} where {\tt T} is a primitive type).
-Subtyping is extended such that {\tt T}<:{$\vec{\tt T}$} for any primitive
-type {\tt T}. The subtype relation is undefined for lists of different
-types, as well as for matrices and data frames.
-
-\subsection{L2 Types}\label{sec:L2def}
-
-{\bf L2} extends the previous system with support for classes and
-attributes.  While there are a finite number of results from the
-\code{typeof} function, this is {\it not} the case for \code{class} and
-\code{attributes}, as classes and attributes {\it can} be user-defined.  In
-terms of classes, R has a number of built-in or otherwise standard classes,
-ranging from truly primitive classes such as {\tt numeric} to the more
-complex {\tt data.frame}.  {\bf L1} already incorporates these builtin
-classes, namely through its data frame and matrix types, so we will ignore
-those classes and thus highlight those which are user-defined.  As for
-attributes, again some patterns arise naturally in R, e.g. all matrices have
-a {\tt dims} attribute, but even these are somewhat interesting, as a
-programmer could in theory set their own {\it dims} attribute on an
-arbitrary value.  The only attribute we ignore is the {\tt class} attribute,
-as we collect the class of values separately (and the {\tt class} attribute,
-if present, is always equal to the class of a value).  Attributes are
-name-value pairs, and to keep things simple we consider only attribute
-names.
-
-In sum, we extend {\bf L1} with attribute lists and classes:
-\attrclass{T}{C}{A} where {\tt T} can be any L1 type other than data frame
-or matrix, and {\tt C} and {\tt A} are lists of classes and attribute names
-respectively.  The subtype relation is defined as: \attrclass{T}{C}{A}
-$<:_{\mathbf{L2}}$ \xspace \attrclass{S}{C'}{A'} if $ T
-<:_{\mathbf{L1}} S$ (i.e. are subtypes in the sense of {\bf L1}), $C
-\subseteq C'$, and $A \subseteq A'$.
-
-\subsection{Simplification by Subtyping}\label{sec:autosub}
-
-When analyzing our data, we perform a certain amount of simplifications
-automatically to minimize the footprint of equivalent signatures.  This
-subtyping is performed at the function level of granularity.  For example,
-if a function is observed to have two signatures \sD \to \df to \D \to \df,
-we collapse the signature to \D \to \df as \sD $<:$ \D and \df $<:$ \df.  We
-would not collapse \sD, \I \to \df and \D, \sI \to \df, as \I $\not <:$ \sI
-and \D $\not<:$ \sD.
-
-%
-%
-%
-\section{Examples of Polymorphism}\label{sec:polyex}
-
-Before presenting the results of our analysis, we  quickly survey
-examples of real, polymorphic R functions that we have analyzed, to build up
-a global sense of how types manifest themselves in current R code.
-
-We begin by referring the reader to Figure~\ref{fig:realex}, which showcases
-the compatibility of integers and doubles.  Coercion from integer to double
-is performed automatically where appropriate, and likewise coercion from
-double to integer is automatically performed when integers are strictly
-required.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-> 5L + 1L # => 6L, an integer
-> 5L + 1.2 # => 6.2, a double (5L coerced to 5.0)
-> c(10, 20, 30)[1.2] # => 10, 1.2 coerced to 1L
-\end{lstlisting}}\caption{Example of \sD, \sI type usage}\label{fig:realex}\end{figure}
-
-Figure~\ref{fig:optnull} shows a function with an optional NULL argument.
-In the function, {\tt x} is a sorted vector, and {\tt w} a vector with
-default value NULL.  If {\tt w} is NULL, then default unit weights are
-generated.  This pattern of default NULL (or NA) arguments is prevalent.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-frac.ranks <- function(x, w=NULL) {
-  if (is.null(w)) w <- rep(1, length(x)) # if no weights passed, take all weights = 1
-  ...
-\end{lstlisting}}\caption{Optional argument ({\tt acid} package)}\label{fig:optnull}\end{figure}
-
-Figure~\ref{fig:listvec} shows a function with list and vector polymorphism.
-It takes a list or a vector {\tt point} and a data frame {\tt polyg}.  The
-{\tt point} is explicitly converted to a vector (by \code{as.numeric})
-regardless of type.  This showcases the ease of going from lists to vectors
-(and conversely).
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-is.point.inside <- function (point, polyg) {
-    p <- as.numeric(point) # as.numeric(list(1, 2)) => c(1, 2)
-    ...
-\end{lstlisting}}\caption{List and vector polymorphism ({\tt bivrp} package)}\label{fig:listvec}\end{figure}
-
-Figure~\ref{fig:charclos} is a function with character and function
-polymorphism.  The caller specifies {\tt fn}, which is either a function
-{\it or} the name of a function which will be looked up by name (a string).
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-pointest <- function (object, fn = mean, par.name, comp.label, chain.no,  ...) {
-    if (is.character(fn))
-        if (fn == "MODE" | fn == "MAP")   do_MAP <- TRUE
-        else {
-            do_MAP <- FALSE
-            fn <- match.fun(fn) # looks up the function name
-        }
-\end{lstlisting}}\caption{Character and function polymorphism ({\tt BAMBI} package)}\label{fig:charclos}\end{figure}
-
-In Figure~\ref{fig:dfdbl}, we see a function that, among other things, takes
-in some data ({\tt dat}) and a character string ({\tt spss}) specifying
-either {\tt "in"} or {\tt "out"}.  Here, the type of {\tt dat} is {\it
-  dependent} on the value of {\tt spss}: if {\tt spss == "in"}, {\tt dat}
-must be a data frame, and otherwise if {\tt spss == "out"}, {\tt dat} must
-be a vector.  Our analysis does not capture this granularity of dependency,
-as we only collect type information.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-nret.translator <- function(dat, spss="out", ...) {  ...
-  if(identical(spss, "out")) {
-    if(!is.vector(dat))  stop(simpleError("...'dat' must be a vector!")) ...
-  } else {
-    ...
-    items.idx <- items.idx[order(names(dat[, items.idx]))]
-  }
-\end{lstlisting}}\caption{Example of a \df/\D argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}
-
-In Figure~\ref{fig:chardbl}, we see a function which takes in a list ({\tt
-  network}), a vector of indices of that list ({\tt fixIndices}), and a
-vector of values ({\tt values}).  The locations specified by {\tt
-  fixIndicies} in {\tt network} are updated with {\tt values}.  Here, {\tt
-  fixIndicies} has been observed to be either a character or double vector:
-In R, list indices are typically doubles, but can be characters (if the list
-or vector being indexed has a {\tt names} attribute).  Note that type level
-{\bf L2} can capture the presence of a {\tt names} attribute.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-fixGenes <- function (network, fixIndices, values) {
-  ...
-  network$fixed[fixIndices] <- as.integer(values)
-  ...
-\end{lstlisting}}\caption{Example of \C/\D argument (from {\tt BoolNet} package).}\label{fig:chardbl}\end{figure}%$
-
-In Figure~\ref{fig:matvec}, we see that a function which can take in a vector,
-but immediately transforms it (transpose, with {\tt t}) into a matrix.  As
-an idea, matrix/vector polymorphism seems sensible, as mathematically
-vectors {\it are} one-dimensional matrices.  R echoes this by making conversion between the two
-types easy: \code{as.vector(m)} flattens a matrix \code{m} into a vector
-(e.g., \code{as.vector(matrix(1, 2))} is equivalent to \code{c(1, 1)}), and
-\code{as.matrix(v)} builds a {\tt length(v) x 1} matrix (e.g.,
-\code{as.matrix(c(1, 2))} is equivalent to \code{matrix(data=c(1, 2))}).
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-tee <- function (x, theta, D1, D2, phi)  {
-    if (is.vector(x)) x <- t(x)
-    ...
-\end{lstlisting}}\caption{Example of \M{D}/\D argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}
-
-Throughout this short section, we have shown real examples of polymorphic R functions.
-Our goal is to capture these instances of polymorphism across a large subset of existing R code, and to highlight and discuss patterns which arise.
-We present the results of our analysis in the following section.
 
@@
 %
-\section{Analysis Results}\label{sec:results}
-
-This section,  presents the results of our analysis at each of our type levels, describing the types of polymorphism that we have observed and shedding light on the frequency of polymorphic patterns in R.
-
-Before we begin, we define the kinds of polymorphism we will consider.  In
-choosing how to express the polymorphism of a function, we consider two
-simple options: First, having the union of signatures, e.g. \D $\rightarrow$
-\D and \C $\rightarrow$ \D, or a union types for each argument, e.g. (\D
-$\cup$ \C) $\rightarrow$ \C.  We refer to the former as {\it function-level}
-and the latter as {\it argument-level polymorphism}.
-
-Per-argument polymorphism highlights the type-based behavior of function
-arguments themselves, and helps to describe the type-based behavior of all R
-values.  For instance, we observe that many arguments have (\sC, \sF) as a
-signature, from which we deduce that many R programmers are taking advantage
-of functions such as \code{match.fun("fun\_name")} to call functions by
-name.  On the other hand, function-level polymorphism highlights how R
-programmers interact with functions as a whole, and allows us to discern
-more subtle patterns.  For example, observing two function signatures such
-as \D \to \D and \sF \to \sF may lead us to ascribe a more useful and
-precise type for the function than if we had simply collapsed the argument
-and return types (e.g., $X$ \to $X$ vs. \D $\cup$ \sF \to \D $\cup$ \sF).
-
-\subsection{L0}
-
-At {\bf L0}, the type system is mostly aligned with the types of the core
-language. Across all of the arguments that we have observed, the vast
-majority are monomorphic.  Table~\ref{tab:L0top10arg} shows the breakdown of
-the 10 most frequently occurring argument signatures.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{lrr}
-  \hline
- Argument Signature & Count & Percentage \\ 
-  \hline
-  \D & 269K & 29.7\% \\ 
-  \l & 193K & 21.3\% \\ 
-  \C & 96K & 10.6\% \\ 
-  \L & 83K & 9.2\% \\ 
-  \sF & 60K & 6.6\% \\ 
-  \sN & 47K & 5.2\% \\ 
-  \ANY & 31K & 3.4\% \\ 
-  \I & 30K & 3.2\% \\ 
-  \sS & 16K & 1.7\% \\ 
-  \D, \l & 10K & 1.1\% \\ 
-   \hline
-   \hline
-\end{tabular}
-\caption{Most frequent argument signatures in {\bf L0}.}
-\label{tab:L0top10arg}
-\end{table}
-
-Doubles are the most common argument type, which is not entirely surprising
-as doubles are the backbone of most statistical functions.  Next is lists,
-which is reasonable as lists are used to build data frames and other widely
-used data structures.  The \ANY type is not an R type per se: we ascribe
-such a type to arguments which were {\it always} unevaluated across all
-calls.  As it happens, many R functions meta-program on some of their
-arguments, and this meta-programming does not force the argument promise,
-and if this meta-programming is always performed in every recorded call we
-will ascribe the \ANY type, as we cannot say what the argument {\it is}
-without possibly forcing a promise.  The magrittr package's pipe \%$>$\%
-function is an example of a function which does meta-programming; we will
-discuss it shortly.  Interestingly, a polymorphic attribute signature \D, \l
-is among the top 10 overall---this is the vector and list polymorphism made
-manifest.
-
-We turn our attention to function-level polymorphism.  Indeed, one of our
-goals is to generate useful function signatures, so we ask: {\it how many
-  functions have more than one recorded signature?}
-Figure~\ref{fig:L0funcounts} has the answer, with a breakdown by function.
-Most functions are monomorphic.  That said, a few functions have
-disproportionately large number of signatures.  One such function is the
-magrittr package's pipe \%$>$\%, which is used to implement a kind of
-currying. It has nearly 100 unique signatures!
-
-\begin{figure}[htbp]\begin{center}
-\includegraphics[width=.9\textwidth]{L0_by_fun}
-\caption{Record of Function Polymorphism in {\bf L0}; \LZEROPERCPOLY of
-  functions are polymorphic}
-\label{fig:L0funcounts}\end{center}
-\end{figure}
-
-Another function with many signatures is the data.table package's \code{[}
-  function, which is a reimplementation of the square bracket operator for
-  selecting values in a data.table.  This function has 16 formal arguments,
-  and call configurations vary widely.  Essentially, the function is {\it
-    the} main function for data.tables, and performs very different tasks
-  based on which arguments are provided.  Two wildly different uses are:
-  \code{dtT[i,j,by]}, where columns \code{i} and \code{j} are accessed and
-  grouped by \code{by}, and \code{dt[x=="some" & y>42]}, which selects
-  all rows who's \code{x} column is the string \code{"some"} and who's
-  \code{y} column is greater than \code{42}.  Given that, we find it
-  unsurprising that the function is so polymorphic.
-
-\begin{table}[ht]\centering
-\begin{tabular}{lrr}  \hline
-Signature & Count & \% of All Signatures \\
-\hline
-  \D, \l & 9566 & 1.1\% \\ 
-  \D, \sN & 8801 & 1\% \\ 
-  \C, \sN & 5590 & 0.6\% \\ 
-  \l, \sN & 5016 & 0.6\% \\ 
-  \C, \D & 2887 & 0.3\% \\ 
-  \D, \sF & 1608 & 0.2\% \\ 
-  \l, \sF & 1345 & 0.1\% \\ 
-  \C, \l & 1283 & 0.1\% \\ 
-  \C, \sF & 1168 & 0.1\% \\ 
-  \C, \I & 1104 & 0.1\% \\ 
-   \hline
-\end{tabular}
-\caption{Top polymorphic argument signatures for the L0 type system.}
-\label{tab:L0toppoly}
-\end{table}
-
-Having had a taste for function-level polymorphism, we will now explore
-common patterns in argument-level polymorphism.  For this, refer to
-Table~\ref{tab:L0toppoly}.
-Let's unpack some of the signatures in Table~\ref{tab:L0toppoly}:
+\section{Introduction}
 
 \begin{itemize}
-\item \D, \l (and \C, \l): L0 does not parameterize list types, and it is very likely that this signature describes vectors of doubles and lists of doubles.
-	Coercing vectors to lists is trivial in R, and indeed many of the functions with such an argument signature do.
-	The story is similar with vectors of characters and lists of characters.
-
-	\item $\square$, \sN: there are a few instances of this ``null'' polymorphism.
-	The frequency of these signatures indicates a tendency to create ``optional arguments'', where some argument's default value might be NULL.
-	This pattern is quite frequent in R code, and an example can be found in Figure~\ref{fig:optnull}.
-
-	\item \C, \D and \C, \I: these are interesting patterns.
-	In R, lists can always be indexed with numerics (\L, \D, \I) {\it and also} characters.
-	When a list is indexed by character, the character is matched against the element names in the list and the corresponding element is selected.
-	If there is no match, the lookup returns NULL.
-	Another possible explanation for these signatures are dates, which can be represented by numerics or characters and the value usually has one of the date classes.
-	L0 does not deal with classes, which are one of the topics of L2.
-
-	\item \D, \sF: this signature appears more strange than it is.
-	Many models, processes, simulations, etc. that are prevalent throughout R have parameters to fine-tune their behavior.
-	This pattern manifests itself here: some arguments with the \D, \sF signature are either parameters for a function (\D), or the function itself already constructed (\sF).
-	Others, such as the {\tt br} argument of the {\tt rtree} function in the {\tt ape} package, are either parameters for complex data generation processes (\D), or a function to be used during the generation (\sF).
-	It would appear that this pattern is used to make complex simulation, modelling, or data generation functions more flexible for programmers by allowing them to supply parameters or the parameterized function itself.
-
-	\item \C, \sF: yet another puzzling signature, though its explanation is surprisingly straightforward:
-	In R, programmers are free to look functions up {\it by name}, and indeed the frequently encountered {\tt UseMethod} function for dynamic class-based function dispatch is one such example.
-	In short, this signature usually represents a function or a function name.
+  \item Adding types to dynamic languages is increasingly common.
+  \item Types help write better code.
+  \item We want to design a type system for R that is based on the patterns that we observe in current use of the language. 
+  \item We also present a tool which takes a given package and synthesizes type signatures for package functions.
+  \item This tool is tested by synthesizing annotations from package tests, examples, and vignettes, and running the code of the package's reverse dependencies (i.e.
+clients) to validate the type signatures.
+  \item We find X, Y, Z.
 \end{itemize}
 
-We have established and discussed some of the most common polymorphic
-attribute signatures, and have built up an understanding of argument-level
-polymorphism in the {\bf L0} type system.  This matches almost exactly with
-the language as it is today, as {\bf L0} is in (nearly) 1:1 correspondence
-with how types are currently represented in the language.  The next question
-we address is: {\it just how pervasive is this argument polymorphism?}
-We find that 94\% of arguments are monomorphic.
-
-From Table~\ref{tab:L0argcounts} we see that, with our definition of
-subtyping on L0 and other simplifications we made, an overwhelming amount of
-argument polymorphism is accounted for.  We also saw in
-Table~\ref{tab:L0toppoly} that each of the most frequently occurring
-argument signatures had some explanation.
-
-That said, L0 is not without limitations, and the main issues stem from a
-lack of adequate information, as the type system does not capture precise
-list types, and does not distinguish scalars from vectors.
-
-%
-%
-\subsection{L1}
-
-L1 espouses a finer-grained approach to types.  Major modifications from L0
-include:
-
-\begin{enumerate}
-	\item the differentiation of scalars and vectors;
-	\item the parameterization of list types over element types;
-	\item the use of attribute and class information to distinguish types for data frames and matrices, and;
-	\item subtyping of \sN (i.e., $\forall T$, \sN $<:$ $T$).
-\end{enumerate}
-
-An example of a signature in this type system can be found in
-Figure~\ref{fig:exL1}.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-add_make_list <- function(x, y) {
-  l <- as.list(x + y)
-  attr(l, "example") <- "this is an attribute"
-  l
-}
-add_make_list(2, c(3, 1)) # => list(5, 3)
-\end{lstlisting}}
-\caption{Call signature in L1: x: \sD, y: \D $\rightarrow$ \lT{D}.}\label{fig:exL1}\end{figure}
-
-Let us begin as we did in L0 by looking at the breakdown of argument
-signatures in L1.  Consult Table~\ref{tab:L1top10arg}.
-\begin{table}[ht]
-\centering
-\begin{tabular}{lrr}
-  \hline
- Argument Signature & Count & Percentage \\ 
-  \hline
-  \sD & 179K & 19.7\% \\ 
-  \D & 97K & 10.7\% \\ 
-  \sL & 77K & 8.6\% \\ 
-  \sC & 75K & 8.3\% \\ 
-  \l & 67K & 7.4\% \\ 
-  \sF & 61K & 6.7\% \\ 
-  \sN & 50K & 5.5\% \\ 
-  \M{D} & 49K & 5.5\% \\ 
-  \ANY & 31K & 3.4\% \\ 
-  \df & 29K & 3.2\% \\ 
-   \hline
-\end{tabular}
-\caption{Most frequent argument signatures in L1.}
-\label{tab:L1top10arg}
-\end{table}
-
-From Table~\ref{tab:L1top10arg}, we see that all of the most frequent
-argument signatures are monomorphic even though we increased the precision
-of the types.  We see that scalars actually figure quite prominently, which
-is encouraging: this supports an idea that an explicit scalar annotation or
-type may be useful {\it for programmers}---indeed, such an annotation would
-be helpful for compiler optimizations (e.g., scalar-only values need not
-have vectors allocated to store them).  We also see that matrices (\M{D})
-are quite popular, a testament to the versatility of the data structure, as
-they can easily be converted to vectors and data frames.
-
-Now that we have a picture of how arguments look in L1, let's see: {\it how many L1 functions have more than one type signature?}
-The answer can be found in Figure~\ref{fig:L1funcounts}.
-
-\begin{figure}[htbp]\begin{center}
-\includegraphics[width=.9\textwidth]{L1_by_fun}
-\caption{Record of Function Polymorphism in {\bf L1}. \LONEPERCPOLY of functions are polymorphic.}
-\label{fig:L1funcounts}\end{center}
-\end{figure}
-
-We see that we barely increased the polymorphism with our transition to L1,
-despite the far richer type information.  The biggest outlier here is indeed
-\%$>$\%, as was the case in {\bf L0}.  Otherwise, the shape is very similar.
-In short: even though {\L1} types are more expressive, the breakdown of
-per-function polymorphism is largely unchanged from {\bf L0}.
-
-As we were in L0, we are curious to know how argument-level polymorphism
-looks in L1.  Refer to Table~\ref{tab:toppolyL1}.
-
-\begin{table}[ht]\centering\begin{tabular}{lrl}  \hline
-Signature & Count & \% of All Signatures \\   \hline
-  \D, \M{D} & 4848 & 0.5\% \\ 
-  \l, \lT{D} & 1843 & 0.2\% \\ 
-  \sC, \sD & 1590 & 0.2\% \\ 
-  \M{D}, \M{I} & 1502 & 0.2\% \\ 
-%  \l, \lT{list} & 1199 & 0.1\% \\ 
-  \df, \M{D} & 1180 & 0.1\% \\ 
-  \C, \D & 1019 & 0.1\% \\ 
-  \sD, \sF & 1005 & 0.1\% \\ 
-  \I, \sD & 997 & 0.1\% \\ 
-  \sC, \sF & 991 & 0.1\% \\    \hline
-\end{tabular}
-\caption{Top polymorphic argument signatures in L1.}\label{tab:toppolyL1}
-\end{table}
-
-So what do these signatures mean?
+The contributions of this paper are:
 
 \begin{itemize}
-	\item \D, \M{D}: vectors are just one-dimensional matrices.
-	Indeed, the only thing separating a vector from a matrix in R is a {\tt dims} attribute.
 
-	\item \l, \lT{D}: this hearkens back to L0, where we discussed how this very same signature was explained by the list being a list of doubles.
-	So why is it still here?
-	Well, first note that there are far less of these signatures in L1 than in L0, so clearly we captured {\it something}.
-	Next, note the method with which we parameterize lists: we ascribe some type $T$ to a list type \lT{T} if and only if {\it all} list elements have {\it exactly} type {\it T}---in our analysis, we have encountered lists with integers as well as doubles, and unfortunately those currently appear as \l.
-	This could be addressed by doing some subtyping computation while exploring list elements.
-
-	\item \sC, \sD and \C, \D: we discussed this in L0---this signature likely captures an index or a date.
-
-	\item \M{D}, \M{I}: this signature is reasonable, as any computation in R on \M{D} would work on \M{I} and vice versa.
-
-%	\item \l, \lT{list}: here we have a list of any and a list of lists.
-%	One explanation for this is that lists of lists sometimes contain NULL as an element, and this would show up with type \l, again as we do not do any subtyping while computing precise list types.
-%	Another explanation is the way R handles object-orientation, which we discuss shortly in Section~\ref{sec:S3S4R5}.
-
-	\item \df, \M{D}: in R, it is easy to transform a matrix into a data frame, and this argument is likely a result of that:
-	Simply calling \code{as.data.frame} on a matrix is enough.
-
-	\item \sD, \sF and \sC, \sF: these were discussed in the context of L0.
-
-	\item \I, \sD: subtyping between primitives does not hold between scalars and vectors (i.e., \I $\not<:$ \sD) even though it is likely that the programmer intended for the argument to be a vector of the supertype (here, these arguments were likely intended to be of type \D).
+\item our analysis data;
+\item a type system based on this data;
+\item a tool to infer types from package code, and;
+\item an evaluation of the tool on \AT{all of} CRAN.
 
 \end{itemize}
-
-With an understanding of argument polymorphism in L1, we wonder: {\it how
-  many argument positions are polymorphic in L1?}  95.3\% of arguments are
-monomorphic.  In spite of drastically increasing the precision of our types,
-we have actually {\it reduced} the amount of polymorphism.  This suggests
-that with a good notion of subtyping, precise type annotations will not cause
-an undue increase in the polymorphism of the language.
-
-Before presenting L2, we will digress slightly and explore some of the ramifications of our transition from L0 to L1.
-
-%
-%
-\subsubsection{Scalars}\label{sec:scalars}
-
-It seems that there is a place for a scalar annotation in R.  We saw in this
-section that strictly scalar values appear quite frequently, making up some
-of the most frequently occurring argument signatures.  In fact,
-\PERCSCALARMONO of arguments have a strictly scalar monomorphic signature.
-This suggests that such a distinction would more accurately capture current
-use of the language, and would also be beneficial to future implementations
-which can perform optimizations based on the type (e.g., not vectorizing
-scalar-only values).
-
-%
-%
-\subsubsection{Object-Oriented Programming in R}
-\label{sec:S3S4R5}
-
-There is something more to be said about the \l type, namely that it is
-sometimes linked to (one of) R's takes on object-oriented programming.  R
-has three object systems: S3, S4, and R5, and S3 objects will often manifest
-themselves with this signatures in L1.  An S3 object is any value with a
-``class'' attribute, and indeed many of these tend to be lists.  For
-example, an S3-style ``Point'' might be a list with an entry for each
-dimension in the point, with a ``class'' attribute set to ``Point''.  When
-these S3 objects have contents with a variety of types, their L1 type will
-appear as \l, which leaves something to be desired: Our next type system,
-L2, will remedy this.
-
-%
-%
-\subsubsection{\C, \I, \D Polymorphism}
-Another significant source of apparently strange polymorphism is the
-character, integer, double polymorphism.  Figure~\ref{fig:chardbl} lays out
-an example of this polymorphic signature, showing that lists and vectors can
-be accessed by strings as well as numerics.  One could conceive of an
-``index'' type to capture this, but there are certain limitations to such a
-type, as such an ``index'' cannot be incremented as it could be a string.
-An even more sophisticated dependent type system could capture this, but
-it is far from clear if such a sophisticated type system would be for a
-language like R.
-
-One other possible explanation for this polymorphism is the ``date'' data type.
-In R, dates are often represented as either strings or numbers, and they are widely used.
-Dates (sometimes) manifest themselves in the class of a value, for examples refer to Figure~\ref{fig:date}.
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-class(date()) # => character
-class(Sys.Date()) # => Date
-class(as.POSIXlt(Sys.Date())) # => POSIXlt, POSIXt
-as.integer(as.Date(1, origin = "1900-01-01"))
-\end{lstlisting}}\caption{Dates in R.}\label{fig:date}\end{figure}
-
-Unfortunately, the format of dates in R is not standardized.
-That said, translating between date representations is straightforward.
-As most dates are classes, L2 will be able to capture their use.
-
-Throughout this section, we have seen that even though we substantially
-increased the precision of our type annotations, we have not increased the
-amount of polymorphism that we observe.  The transition from L0 to L1
-revealed some interesting patterns that L1 cannot quite capture---for these,
-we will need to appeal to the class and attributes of values, which are
-incorporated into {\bf L2}, discussed next.
-
-%
-%
-%
-%
-\subsection{L2}
-
-{\bf L2} works class and attribute information into its type signatures.
-As per the discussion in Section~\ref{sec:L2def}, we omit builtin classes in order to highlight those which are user-defined.
-{\bf L2} differs from {\bf L1} in the following ways:
-
-\begin{enumerate}
-
-	\item where {\bf L1} and {\bf L0} only considered classes and attributes when ascribing matrix or data frame types, {\bf L2} considers classes and attributes to be an integral part of the type system.
-	Annotations in {\bf L2} include classes and attribute names when they are present.
-
-	\item recall that subtyping between two {\bf L2} types is defined as
-
-\end{enumerate}
-
-As in {\bf L0} and {\bf L1}, we will begin by showing a breakdown of the top argument signatures in Table~\ref{tab:L2top10arg}.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{lrl}
-  \hline
-Signature & Count & \% of All Signatures \\ 
-  \hline
- \sD & 172K & 19\% \\ 
-  \sL & 77K & 8.5\% \\ 
-  \D & 77K & 8.4\% \\ 
-  \sC & 73K & 8\% \\ 
-  \sN & 50K & 5.5\% \\ 
-  \sF & 45K & 5\% \\ 
-  \ANY & 31K & 3.4\% \\ 
-  \attrclass{\D}{}{dim} & 25K & 2.6\% \\ 
-  \C & 25K & 2.6\% \\ 
-   \attrclass{\l}{}{names, row.names}  & 21K & 2.2\% \\ 
-   \hline
-\end{tabular}
-\caption{Most frequent argument signatures in {\bf L2}.}
-\label{tab:L2top10arg}
-\end{table}
-
-As you can see, the most frequent signatures in {\bf L1} are represented here.
-The inclusion of class and attribute information has changed the counts slightly.
-Not pictures here, but interesting: named lists are more frequent than their unnamed counterparts.
-No classes are represented here since (1) we eliminated builtin classes, as we've used them to construct types already, and (2) most other classes are specific to some package or set of packages---it is doubtful that such a class would be so pervasive as to feature in this breakdown.
-
-Now that we have a sense for argument-level type signatures in {\bf L2}, we will see just how many functions are polymorphic.
-Consider Figure~\ref{fig:L2funcounts}.
-
-\begin{figure}[htbp]\begin{center}
-\includegraphics[width=.9\textwidth]{L2_by_fun}
-\caption{Record of Function Polymorphism in {\bf L2}. \LTWOPERCPOLY of functions are polymorphic.}
-\label{fig:L2funcounts}\end{center}
-\end{figure}
-
-Not pictured here are two outlier values: the summary and print functions of the base R package.
-Summary is a generic function which is used to summarize various types of data, and print is used to convert values to string format and print them to the console.
-These functions have over 1,000 unique signatures each, which is within reason as they are really operating over any type.
-They feature so prominently in {\bf L2} (and not in {\bf L1} or {\bf L0}) since many of the values which flowed into them had some amount of attribute or class information, and indeed these functions are very frequently exercised by other packages which may implement package-specific classes.
-
-Overall, the amount of polymorphism only slightly increased, even though we added two new dimensions of type information.
-Again, far richer types do not increase the amount of polymorphism by significant amounts, perhaps a testament to the monomorphic nature of many R functions.
-
-Having built a sense for how polymorphic whole-function signatures are in {\bf L2}, we will turn our attention to polymorphic arguments.
-Given the added class and attribute information, {\it what do the more frequent polymorphic argument signatures look like?}
-Table~\ref{tab:toppolyL2} has the breakdown.
-
-% latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Fri Apr  5 09:50:59 2019
-\begin{table}[ht]
-\centering
-\begin{tabular}{lrl}
-  \hline
-Signature & Count & \% of All Signatures \\ 
-  \hline
-  \D, \attrclass{\D}{}{dim} & 1942 & 0.2\% \\ 
-  \sC, \sD & 1431 & 0.2\% \\ 
-  \attrclass{\l}{}{names}, \attrclass{\D}{}{names}  & 1046 & 0.1\% \\ 
-  \I, \sD & 855 & 0.1\% \\ 
-  \D, \attrclass{\D}{}{dim, dimnames} & 810 & 0\% \\ 
-  \attrclass{\l}{}{names, row.names}, \attrclass{\D}{}{dim, dimnames} & 708 & 0.1\% \\ 
-  \C, \D & 672 & 0\% \\ 
-   \attrclass{\D}{}{names}, \attrclass{\D}{}{dim, dimnames} & 591 & 0\% \\ 
-  \attrclass{\D}{}{dim}, \attrclass{\I}{}{dim}  & 445 & 0\% \\ 
-  \sD, \attrclass{\sF}{}{srcref} & 445 & 0\% \\ 
-   \hline
-\end{tabular}
-\caption{Top polymorphic argument signatures in {\bf L2}.}
-\label{tab:toppolyL2}
-\end{table}
-
-When comparing this table with Table~\ref{tab:toppolyL1}, we see that by-and-large the same types are represented.
-We note the {\tt names} attributes on the lists in row 3: these named lists are, in general, more common than their unnamed counterparts.
-In the 10th signature, we see a function with attribute {\tt srcref}: functions with this attribute are for dealing with open files in the context of the srcfile set of functions in the base package.
-As for why such a function would share an argument position with scalar double, there are other representations of open files, and numbers feature among them.
-
-Now that we have seen the most frequently appearing argument signatures, we ask: {\it how many argument positions are polymorphic?}.
-Table~\ref{tab:argcountsL2} has the answer.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{lll}
-  \hline
-Signature Type & Count & Percentage \\ 
-  \hline
-  Monomorphic & 851K & 93.9\% \\ 
-  Total Seen & 906K & --- \\ 
-   \hline
-\end{tabular}
-\caption{Account of {\it argument} polymorphism in {\bf L2}.}
-\label{tab:argcountsL2}
-\end{table}
-
-Surprisingly, even with the added precision of classes and attributes, there is still a relatively small amount of argument-level polymorphism in {\bf L2}.
-Our notion of subtyping on classes is rather weak, and one extension would be to include class hierarchies in the subtyping of classes;
-this would further reduce the amount of polymorphism in this type system.
-
-%
-%
-%
-%
-\subsection{Notable Patterns and Observations}
-
-Besides exploring the patterns highlighted by our various type systems, we can address some interesting and tangentially related questions about the usage of types in R.
-
-%
-%
-%
-%
-\subsubsection{R's Most Polymorphic Functions}
-
-Given the data that we've collected, it is natural to wonder what the most polymorphic packages are.
-Indeed, we have shed some light on some of the most polymorphic {\it functions}, but what about the packages that these functions belong to?
-What is their purpose, and why are they so polymorphic?
-Figure~\ref{tab:packagepolysize} has some answers.
-Note that we are in the context of {\bf L0}, here, to highlight current language patterns.
-
-% latex table generated in R 3.5.1 by xtable 1.8-3 package
-% Thu Mar 28 18:15:26 2019
-\begin{table}[ht]
-\centering
-\begin{tabular}{lrrr}
-  \hline
-  Package & \# of Functions &\# of Signatures & Description \\ 
-  \hline
-  magrittr & 20 & 225 & Tidyverse pipe suite \\ 
-  oshka & 4 & 38 & quoted expression expansion \\ 
-  scatterD3 & 1 & 9 & create D3 (JS) scatter plots from R \\ 
-  yaml & 4 & 33 & convert R to YAML \\ 
-  knitLatex & 4 & 32 & \LaTeX \xspace generator from R data \\ 
-   \hline
-\end{tabular}
-\caption{5 of the most polymorphic packages (in {\bf L0}).}
-\label{tab:packagepolysize}
-\end{table}
-
-Now, what makes these so polymorphic?
-The first package, magrittr, is one that we discussed earlier in the context of its pipe \%$>$\% function. 
-It is not surprising that the package as a whole is quite polymorphic, as it implements {\it several} different pipe functions, e.g.
-\%$<>$\% which pipes as normal and then assigns to the left-hand argument.
-In fact, all but knitLatex implement some form of meta-programming:
-the oshka package meta-programs in the same style as Tidyverse, the scatterD3 package meta-programs on plot objects, and the yaml package converts R expressions to YAML expressions.
-This is a trend we have observed: {\it the most polymorphic functions in R are meta-programming their arguments}.
-
-It is highly improbable that a static type system designed for an untyped language will be able to capture all of the language's existing code (without resorting to an \ANY type or abusing unions).
-It is also unclear how useful a type system would be if it {\it did} capture everything.
-Indeed, many of the packages discussed here are really designed with an \ANY-typed argument in mind (e.g., left-hand side of \%$>$\%), and restricting these functions should not be the goal of any static type system designed for a language after it has matured.
-
-%
-%
-%
-%
-\subsubsection{Metaprogramming and Unevaluated Arguments}
-
-Our analysis uncovered a non-negligible amount of argument positions which were {\it never} evaluated across all observed function calls.
-This is indicative of a desire to meta-program arguments and otherwise deconstruct their AST, as the \%$>$\% function does.
-This is an incredibly powerful language feature, and may well be at the source of R's popularity, if in a roundabout way.
-Tidyverse, by far the most popular package suite in R, owes some of its success to the new syntax defined by some of the included packages, and it is this ability to meta-program arguments which allowed Tidyverse to define its unique syntax.
-An example of that syntax is \%$>$\%, and indeed this function was highly exercised by over 100 other packages.
 
@@
 %
-\section{Discussion}
+\section{The R Programming Language}
+\label{sec:rlang}
 
-Throughout Section~\ref{sec:results}, we have thoroughly explored the trends and patterns of polymorphism observed in our analysis of a large subset of existing R code.
-By now, we have an understanding of how each of our type levels captures the polymorphic patterns that exist in R.
-
-In this section, we aim to comment globally on the things we learned during our analysis, and propose a type system for R which is supported by empirical evidence that it captures the patterns of existing R code.
-We will first look at function level polymorphism before moving on to argument-level polymorphism.
+\AT{This can probably be copied from the old version.
+Is this somewhere we want to talk about R-dyntrace?}
 
@@
 %
-\subsection{Function Polymorphism}
+%
+%
+\section{The Corpus}
+\label{sec:corpus}
 
-There are some function-level polymorphic patterns that cannot be captures solely with polymorphic function arguments.
-For instance, consider Figure~\ref{fig:funanno}.
+\AT{Probably talk about CRAN, and give some stats on the corpus that we worked with. 
+This section probably doesn't have to be very long.}
 
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-iden <- function(x) x;
+\AT{Fit this in here:}
+Unfortunately, there is not a whole lot of end-user R code available for analysis; typical R programs are effectively data pipelines, and many R users elect not to share their data, possibly because the data is proprietary or somehow confidential (e.g., census data), and these data pipeline programs are not very useful without the underlying data.
+To alleviate this, we focused on the analysis of package test, example, and vignette code.
 
-# observed calls:
-f(2); # x: double -> double
-f(list(1, 2, 3)) # x: list<double> -> list<double>
+\AT{How do we want to argue that our data is valid and representative?}
+End-user R scripts are hard to come by.
+R is a data science language first-and-foremost, and so R programs are typically data analyses.
+The nature of data is that, in many scenarios, it is confidential; imagine a government sharing sensitive information publicly, or a company divulging customer data.
+For this reason, it is rare for R programmers to share their code, and even if they did, the code would not be particularly useful without the data underlying it.
 
-# possible annotations:
-# [1] x: {double, list<double>} -> {double, list<double>}
-# [2] x: {double -> double, list<double> -> list<double>}
-\end{lstlisting}}\caption{Simple identity function.}\label{fig:funanno}\end{figure}
+Separately, idiomatic usage of R is to appeal to libraries to implement the main functionality of an analysis.
+\AT{Probably want some good evidence of this.}
+R packages are mainly sourced from the Comprehensive R Archive Network (CRAN), a large curated repository of R packages (libraries).
+Of particular relevance to this work is a requirement that all packages should ship with tests, examples, and vignette programs to showcase how packages are intended to be used---indeed, these serve as a starting point for potential package clients.
+We can leverage these programs to build up a picture of how programmers interact with packages, and thus of how programmers write R code for their own analyses.
 
-Sophisticated dependent type signatures such as these could benefit language implementations in a myriad of ways.
-For instance, a language implementation could perform type-based optimization in the context of a function return simply by knowing the function argument type.
-In Figure~\ref{fig:funanno}, we can assign the function a more sophisticated signature ({\tt [2]}) if we do not simply collapse the observed argument signatures element-wise ({\tt [1]}).
-This gets more interesting with attributes, as we might want to annotate a function to communicate that it adds an attribute to whatever the list of attributes of its input value.
-
-A clear drawback to function-level polymorphism is that functions with many arguments tend to be sensitive to small changes in argument types, which inflates the number of unique signatures observed.
-Even so, overall the number of polymorphic functions did not increase by too large a margin when adding and specializing types (from {\bf L0} to {\bf L1}, and from {\bf L1} to {\bf L2})---for the most part, monomorphic functions in {\bf L0} stayed monomorphic throughout.
-
-One tidbit about function signatures that we have not thoroughly discussed is the {\tt ...} function argument.
-Functions with such an argument can be passed an arbitrary number of extra arguments, see Figure~\ref{fig:dotdotdot} for a small example.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-f <- function(x, ...) {
-  g(x, ...)
-}
-
-f(42, z=2, y=4) # in f, will perform g(42, z=2, y=4)
-
-\end{lstlisting}}\caption{Simple ... example.}\label{fig:dotdotdot}\end{figure}
-
-Our analysis ignores these extra arguments, instead it simply flags functions which have a {\tt ...} in their definition.
-If building an annotation for such a function, the {\tt ...} argument is always ascribed the \l type, as it can indeed be inhabited by anything.
-
-Ultimately, to determine which kinds of whole function signatures would be useful, a more sophisticated analysis is required, which would need to develop a correlation between function arguments and return types.
-There may also be other interesting per-function trends which our current analysis does not pick up on that are hidden in our data.
+\AT{I think that talking avout how ``this is as good as it gets guys'' might be a bit of a defeatist way of looking at it.
+But then again, I'm not sure what a good balance would be here.}
 
@@
 %
-\subsection{Argument-Level Polymorphism}
+%
+%
+\section{Overview}
 
-From our analysis, we discovered a number of interesting patterns at the argument level.
-To summarize:
+\AT{Somewhere, we need to overview the contribution, talking about the type system, about the tool, and about the data.}
 
-\begin{itemize}
+Broadly, the contributions of this work can be divided into two.
 
-	\item {\bf scalars}: according to our analysis, many instances of vectorized values are always scalar.
-	This is strong evidence that {\it an annotation denoting a scalar value would be useful to R programmers}, and further such an annotation could be used by future implementations to statically determine when not to vectorize a vectorizable type.
+First, we present a {\it large-scale analysis} of the type-use programming practices in R.
+As mentioned in Section~\ref{sec:corpus}, we focused our efforts on package test, example, and vignette code.
+Our tracing framework collects type information of function arguments and return values, and the precise details are elaborated in Section~\ref{sec:methodology}.
 
-	\item {\bf missing}: our analysis has found a number of instances of argument positions remaining unevaluated across all calls to a function.
-	{\it A small annotation, such as a tag on an argument position, would capture this behavior} and could be used to warn programmers if they try to use the unevaluated value, and also warn them if an operation might evaluate it accidentally.
-	Techniques to avoid forcing promises in R are not well understood by most programmers, and such an annotation would be useful to assist them in avoiding unwanted forcing of promises.
+Unfortunately, in spite of the requirements imposed on CRAN packages, some package functions are not thoroughly tested and are poorly exercised.
+Previous work in a similar vein (\AT{Genthat}) reported that their tool achieved roughly 70\% coverage of package code, which is good but far from perfect (and our tool \AT{is liable to achieve the same coverage}).
+In order to further improve our data, we perform an additional phase of type propagation, where we use known type signatures of inner functions to improve the signatures of outer functions.
+Essentially, we build a graph of the flows of argument values throughout a function, and infer constraints from that graph and feed our data into it; the precise details will be discussed further in Section~\ref{sec:methodology}.
+The data itself will be presented in Section~\ref{sec:results}.
 
-	\item {\bf classes}: in transitioning from {\bf L1} to {\bf L2}, we found that the prevalence of \l, \lT{X} polymorphism decreased significantly.
-	In many cases, these lists share a class, and so the signature which was prominent in {\bf L1} was broken down into several less prominent {\bf L2} signatures with added class information.
-	Classes could also be used to capture the usage of dates (two of the standard date formats have class ``Date'' and ``POSIXct, POSIXt''), and could help package writers enforce usage of package functions.
-	R's dynamic dispatch system is class-based, so including annotations for these would help programmers determine which functions are being dispatched for which values.
+Besides helping us understand the R language, we also used the data to inform the {\it design of a type system} for the language.
+Based on the common patterns observed in the analysis, we devised a type system for the language that \AT{maximizes the ``monomorphicity'' in the data} while capturing many of the idioms we observed.
+For example, we often ran across function arguments which were exclusively inhabited by unit-length vectors of some primitive type; this led us to add a scalar version of primitive types to the type system, and include a rule akin to scalar\_primitive <: vector\_primitive to ensure that unit-length vectors are treated as such when there is evidence that the function is intended to accept vectors.
+This will be elaborated further in Section~\ref{sec:results}.
 
-\end{itemize}
-
-Generating the code to dynamically enforce these signatures is also possible.
-Simple checks include checking the class of a value, or checking for the presence of certain attributes.
-R makes it easy to exit a function prematurely, and in fact many functions already exhibit this behavior.
-For example, see Figure~\ref{fig:stopifnot}.
-In it, the function exits if the {\tt encoding} argument is provided, and is not a character.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-is_character <- function(x, n = NULL, encoding = NULL) {
-  ...
-  if (!is_null(encoding)) {
-    stopifnot(typeof(encoding) == "character")
-    ...
-  }
-  ...
-}
-\end{lstlisting}}\caption{Real example of premature function exit.}\label{fig:stopifnot}\end{figure}
-
-We also saw that, despite being far more specific and descriptive than previous levels, the {\bf L2} type system did not bring about an undue increase in the amount of overall polymorphism in the language.
-This signifies that {\bf L2} captures the vast majority of already existing polymorphic patterns, and very few functions would need to resort to an \ANY annotation, and indeed many arguments are really \ANY-typed (e.g., \%$>$\% from magrittr and the prototype assignment operator {\tt \$<-} from proto).
-
-Based on our analysis of the data, we conclude that {\it a type system such as} \textbf{\emph{L2}} {\it captures the vast majority of argument and function usage in the language}.
-{\bf L2} also contains a wealth of useful information about values that could be used for optimizations and IDE tooling.
+Now, in order to {\it validate} this type system, we implemented a framework for annotating R programs, as well as a tool which infers types from package example code, inserts lightweight runtime contracts into the code, and runs the example code in the package's reverse dependencies to try to unearth type errors.
+The annotations are lightweight as they are only run {\it the first time} an argument is used, if it is used at all.
+Details on and the results of the validation are available in Section~\ref{sec:results}.
+Details on the process itself will be given in the next section.
 
@@
 %
-%\subsection{Coverage}
+\subsection{Annotations}
+\label{subsec:annotations}
+
+We extended the R language to support annotations on function arguments and returns.
+This involved reserving syntax for the annotations, and modifying the R parser to parse and accept them.
+\AT{What other stuff do we want to add to this? Other than presenting the syntax and stuff.}
+The syntax allows programmers to specify the basic type, class, attributes, length, and parameterized types of arguments.
+Also, we support type variables so as to allow programmers to specify large but common type signatures one time only.
+
+To illustrate the annotation syntax, consider the code in Figure~\ref{fig:annotation-syntax}.
+\AT{TODO: This figure.}
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+my_fun <- function(@{int} x) {
+  ...
+}
+\end{lstlisting}}\caption{Type annotation syntax.}\label{fig:annotation-syntax}\end{figure}
+
 %
-%\AT{Manual check of some functions.}
+%
+%
+%
+%
+%
+\section{Methodology}
+\label{sec:methodology}
 
+\AT{I think there are two distinct methodologies.}
+
+\AT{First, we had to come up with the type system, and it partly based on things that we saw from our analysis.}
+
+\AT{Second, we should talk about the evaluation of the tool.}
+
+%
+%
+%
+%
+\subsection{Typr}
+
+\AT{Are we going to change this? To make it run off of R-dyntrace?}
+
+The first tool in our pipeline is Typr.
+The goal of Typr is to gather type information on function invocations, and it is a modification of the Genthat~\cite{issta18} tool.
+Genthat records all function invocations and builds a standalone trace to allow for independent reproduction of the call. 
+We modified Genthat to instead record a wealth of information on argument and return values.
+Typr records:
+
+\begin{itemize}
+\item the results of calling {\tt typeof}, {\tt class}, and {\tt attributes} on the argument;
+\item the length (if the argument has a length);
+\item if the argument is a list or data.frame we additionally reflect on the types of argument members;
+\item if the argument is an environment, we reflect on the bindings in the environment.
+\end{itemize}
+
+The precise details of the approach are as follows:
+
+\begin{enumerate}
+\item for every package that is to be run, all functions (including those inherited from dependencies) are instrumented with on.exit hooks, code which runs when the function is finished executing.
+The purpose of waiting for function exit is to ensure that argument promises are not forced, and argument values which are unused are recorded as such, as this indicates that they are likely used for some metaprogramming within the function.
+\item package tests, examples, and vignettes are run, and decorated function invocations trigger the on.exit hook, which records argument and return value type information.
+\item each new function trace (i.e., each call with a new call signature) is saved.
+\end{enumerate}
+
+\AT{At what point do we introduce the type system?
+I was gonna mention subtyping rules in a hypothetical item 4 where we consolidate trace info.
+Maybe we should just talk about how we looked at this data, and then add a new phase after?}
+
+%
+%
+%
+%
+\subsection{Propagatr}
+
+\AT{Explaining this succinctly will be important.}
+After collecting and consolidating type information for function calls, we use our Propagatr tool to propagate known type information between functions that are used in similar circumstances.
+In essence, our tool runs over the tests, examples, and vignettes, much like our Typer tool, but instead of collecting type information it tracks values as they flow through code and builds a graph connecting use-sites of values to the argument positions that they are drawn from.
+\AT{Man, this sucks.}
+
+For example, consider the code snippet in Figure~\ref{fig:propagatr-example}.
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+f <- function(x, y) {
+  z <- x
+  g(z, x)
+}
+
+f(0, 42)
+\end{lstlisting}}\caption{Propagatr example.}\label{fig:propagatr-example}\end{figure}
+
+Propagatr tracks the following data flow:
+
+\begin{itemize}
+\item \AT{This might be better achieved as a graph, or something.}
+\item the assignment of x to z on line 2 is actually a function call in R; think of it as {\tt assign(x, z)}.
+The flow of data here is from x to z, and thus in the remainder of the function both variables are known to share the same value.
+\item The call to g(z, x) allows us to deduce that g takes two arguments {\it of the same type}. 
+\end{itemize}
+
+An important purpose of the Propagatr tool is to allow us to infer type signatures for poorly-exercised functions.
+As a (somewhat degenerate) example, consider the code in Figure~\ref{fig:propagatr-degenerate}.
+\AT{maybe shove this example somewhere to save space}
+
+\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
+add_wrapper <- function(a, b) { a + b }
+\end{lstlisting}}\caption{Degenerate Propagatr example.}\label{fig:propagatr-degenerate}\end{figure}
+
+Even if {\tt add\_wrapper} is rarely exercised, Propagatr deduces that it should have the same signature as R's builtin addition operator.
+
+The precise details of the tool follow.
+\AT{Pseudocode? Something like that? A formal description of the algorithm.}
+
+%
+%
+%
+%
+\subsection{Infr}
+
+Our Infr tool leverages our type system to test the effectiveness of package test, example, and vignette code at covering the possible use-cases of package functions.
+
+The approach is as follows:
+
+\begin{itemize}
+\item Typr and Propagatr are run on the test, example, and vignette code of a particular package.
+This serves to infer an annotation for all package functions.
+\item Infr turns these annotations into dynamically-checked contracts on functions:
+The contracts check that function argument and return values adhere to the type of their annotation.
+These contracts are inserted directly into the promise code of the values, as will be elaborated in Section~\ref{subsubsec:promise}.
+\item Then, the test, example, and vignette code for all reverse package dependencies (i.e., for all other packages which have the current package as a dependency) is run, and any type errors that are caught are recorded.
+\end{itemize}
+
+\AT{Do we want to do something special about data frames? We sort of already infer their type from the dynamic analysis.}
+
+%
+%
+\subsubsection{I Promise I'll Do the Check!}
+\label{subsubsec:promise}
+
+As mentioned in Section~\ref{sec:rlang}, arguments to R functions are wrapped in promises whose code is only executed if the argument is used.
+To insert checks that do not disrupt these semantics, we insert our checks {\it directly into the code of the promise}, which is done in C++ leveraging the Rcpp framework.
+
+Essentially, the value of the e.g. argument promise is replaced by code which checks for compliance with the argument's annotation.
+If the check fails, a call to R's stop function is triggered, yielding an informative error message (which includes the value itself, the function, package, and type that the value failed to adhere to).
+If successful, the checking code is removed (by reassigning to the argument within the function), so subsequent accesses to the argument are fast.
+\AT{Can return values be wrapped in the same way?}
+
+%
+%
+%
+%
+%
+%
+\section{Results}
+\label{sec:results}
+
+\AT{What should go in here?}
+
+\AT{There's the data itself, and our breakdown of it.
+There's also the results of validating the types and the tool against the reverse dependencies.}
+
+%
+%
 %
@@
 
+\AT{Look at the stuff we put in the old version of the paper.}
+
+%
+%
+%
+%
+\subsection{Studies of Dynamic Languages}
+
 The idea of measuring polymorphism via some program analysis is not new.
 For instance, some work~\cite{aakerblom2015measuring} measures the polymorphism of Python programs, but with a slightly different goal than ours: authors seek to assess the need for expressiveness in type systems that are retrofit onto dynamic languages.
-Interestingly, they draw some similar conclusions to us: they find that monomorphism is far more frequent than polymorphism, and even find that typically very few functions are very highly polymorphic.
+Interestingly, they find that 4\% of function call sites are polymorphic, which is considerably less than what we have found with our analysis of R.
+In essence, it would appear that R is far more polymorphic in Python.
 
@@
 An interesting finding of theirs is that only 81\% of calls are monomorphic, which is very similar to what we observed in R.
-Other analysis work~\cite{milojkovic2017duck} looks into the usage patterns of duck typing.
+This 80/20 split between monomorphism and polymorphism is echoed in other analysis work on the usage patterns of duck typing in Smalltalk~\cite{milojkovic2017duck}, which finds that ~20\% of functions are ``duck-typed'' (\AT{read: polymorphic}) in Smalltalk.
+\AT{It's notable that this 20\% shows up very often.}
 
-The type system we propose in this work could be incorporated into R as a {\it gradual type system}.
-Gradual type systems~\cite{SiekTaha06} aim to bridge the gap between dynamic and static typing by allowing programmers to gradually add type annotations to their programs.
-Gradually-typed languages are becoming increasingly popular, with languages such as TypeScript~\cite{typescript13} becoming more mainstream.
-An eventual implementation of a type system for R will surely be inspired by this work.
+%
+%
+%
+%
+\subsection{Synthesizing Types from Dynamic Analysis}
+
+In this work, we have retrofit a type system onto the R programming language.
+We based many of the decisions on which types and which rules to include on the results of our analysis.
+This shares some similarity with work on so-called {\it trace typing}~\cite{andreasen2016trace}.
+
+%
+%
+%
+%
+\subsection{Type Systems for Dynamic Languages}
+
+The type system introduced in this work is a {\it gradual type system}.
 
@@
 
-Truly understanding how programmers use a programming language is critical
-to building usable and effective language features.  When working on a
-general purpose languages (such as JavaScript), we as programmers can appeal
-to our own intuitions, as many of us have had at least some interaction with
-said languages.  This intuition does not extend to domain-specific languages
-(DSLs), which are not really designed for members of our community.
-
-The R programming language is an example of this.  R was developed to foster
-an easy and interactive programming experience. For this reason R is very
-flexible and expressive.  Unfortunately, these liberties come at a cost:
-debugging R code is challenging. Further, performance quickly becomes
-prohibitive if programmers are not careful.
-
-Types can make a language easier to debug and run faster.  For debugging,
-some level of static type enforcement would give programmers more direct
-feedback.  As for performance, type information enables optimizations. If we
-are interested in developing an easily adoptable type system for R, we must
-ensure that it captures how existing code is being used.
-
-Throughout this paper, we have built up an understanding of types and
-polymorphism in R by analyzing the usage of types across a large corpus of
-code.  We highlighted some patterns of type usage, and developed a number of
-type systems with differing levels of granularity. We analyzed the tests,
-examples, and vignettes of over ten thousand packages on CRAN (over
-\PERCENTCRAN of available packages), and traced over 15 million lines of R
-code.
-
-We believe that a type system akin to {\bf L2} is well suited for R, as it
-is expressive enough to capture current use of the language without being so
-specific as to make everything polymorphic.  In fact, the levels of
-polymorphism observed were rather low overall, with only 6\% of function
-arguments and 21\% of functions being polymorphic.  Importantly, {\bf L2}
-carries enough information to be useful to R implementations and IDEs:
-language implementations would surely benefit from statically-available type
-information, and IDEs could use the rich type and attribute information to
-give programmers immediate feedback during the programming process.
-
 %
@@
 \bibliographystyle{boilerplate/ACM-Reference-Format}
-\bibliography{bib/biblio,bib/jv,bib/r,bib/new,bib/gradual}
+\bibliography{bib/biblio,bib/jv,bib/r,bib/new,bib/gradual,bib/types4R}
 \end{document}

### 2019-09-23 paper/old_version.tex
new vers

\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\usepackage{booktabs,listings,xspace,wrapfig}
\lstset{language=R}
\usepackage{my_style}
\definecolor{LightGray}{rgb}{.95,.95,.95}
\definecolor{Gray}{rgb}{.3,.3,.3}
\definecolor{DarkGray}{rgb}{.5,.5,.5}

\graphicspath{ {./plots/} }

\lstset{ %
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
  framexleftmargin=-1mm,
  framexrightmargin=-1mm,
  tabsize=2,
  belowskip=0pt,
  basicstyle=\small\ttfamily,
  backgroundcolor=\color{LightGray},
  emphstyle=\sffamily,
  keywordstyle=\bfseries,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray},
 % numbers=left
}

\lstdefinestyle{R}{ %
  language=R,
  deletekeywords={env, equal, c, runif, trace, args},
  breaklines=true
}

\lstdefinestyle{Rin}{ %
  style=R,
  numberstyle=none,
  basicstyle=\normalsize\ttfamily,
  breaklines=false
}

\newcommand{\code}[1]{\lstinline|#1|\xspace}
\newcommand{\genthat}{{\sc Genthat}\xspace}

% Macros for type names?
\newcommand{\reals}{$\mathtt {\rm I\!R}$\xspace}
\newcommand{\realv}{$\overline{\rm I\!R}$\xspace}

\setcopyright{none}
%\setcopyright{acmcopyright}%\setcopyright{acmlicensed}
%\acmDOI{10.475/123_4}
%\acmConference[OOPSLAs]{Woodstock conference}{July 1997}{El Paso, Texas USA}
%\acmYear{1997}%\copyrightyear{2016}%\acmPrice{15.00}
\begin{document}

\title{A Type System for R: Empirically Designed and Evaluated}

\newcommand{\PACKAGES}{11,463\xspace}
\newcommand{\PROGRAMMERS}{?\xspace}
\newcommand{\PERCENTCRAN}{83\%\xspace}
\newcommand{\CRANTOTAL}{13,841\xspace}
\newcommand{\RLOC}{15,050,267\xspace}
\newcommand{\CLOC}{9,373,542\xspace}
\newcommand{\YEARS}{20\xspace}
\newcommand{\INDEXCOINCIDENCE}{3,499\xspace}
\newcommand{\TOTALINDEXY}{6,561\xspace}
\newcommand{\INDEXYPERC}{53\%\xspace}		% TODO Not an accurate count.
\newcommand{\DATAPKGS}{206\xspace}
\newcommand{\DATAPKGSPERC}{1.5\%\xspace}
\newcommand{\METAARGCOUNT}{7,051\xspace}
\newcommand{\NUMPOLYTYPEMONOCLASS}{1,357\xspace}
\newcommand{\PERCSCALARMONO}{38.2\%\xspace}

\newcommand{\LZEROPERCPOLY}{19\%\xspace}
\newcommand{\LONEPERCPOLY}{18.1\%\xspace}
\newcommand{\LTWOPERCPOLY}{21.3\%\xspace}

\newcommand{\attr}[2]{\ensuremath{#1_{\mathtt{#2}}}\xspace}
\newcommand{\attrclass}[3]{\ensuremath{#1^{\mathtt{#3}}_{\mathtt{#2}}}\xspace}
\renewcommand{\to}{\ensuremath{\rightarrow}\xspace}
\newcommand{\D}{\ensuremath{\small\vec{\mathtt D}}\xspace} % Double
\newcommand{\I}{\ensuremath{\small\vec{\mathtt I}}\xspace} % Integer
\renewcommand{\C}{\ensuremath{\small\vec{\mathtt C}}\xspace} % Character
\renewcommand{\L}{\ensuremath{\small\vec{\mathtt L}}\xspace} % Logical
\newcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\X}{\ensuremath{\small\vec{\mathtt X}}\xspace} % Complex
\newcommand{\Y}{\ensuremath{\small\vec{\mathtt Y}}\xspace} % Symbol
\newcommand{\sY}{\ensuremath{\small{\mathtt Y}}\xspace} % Symbol
\newcommand{\sS}{\ensuremath{\small{\mathtt S}}\xspace} % S4
\newcommand{\sF}{\ensuremath{\small{\mathtt F}}\xspace} % Closure
\newcommand{\sE}{\ensuremath{\small{\mathtt E}}\xspace} % Env
\renewcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\sN}{\ensuremath{\small{\mathtt N}}\xspace}     % Null
%\renewcommand{\l}{\ensuremath{\small L<?>}\xspace}     % List
\renewcommand{\l}{\ensuremath{\small\underline{\mathtt ?}}\xspace}     % List
\newcommand{\sD}{\ensuremath{\small{\mathtt D}}\xspace} % Double
\newcommand{\sI}{\ensuremath{\small{\mathtt I}}\xspace} % Integer
\newcommand{\sC}{\ensuremath{\small{\mathtt C}}\xspace} % Character
\newcommand{\sL}{\ensuremath{\small{\mathtt L}}\xspace} % Logical
\newcommand{\sX}{\ensuremath{\small{\mathtt X}}\xspace} % Complex
\newcommand{\sR}{\ensuremath{\small{\mathtt R}}\xspace} % Raw
\newcommand{\ANY}{\ensuremath{\small{\mathtt ?}}\xspace}     % Any
\newcommand{\lT}[1]{\ensuremath{\small\underline{\mathtt{#1}}}\xspace}     % list<T>
\newcommand{\M}[1]{\ensuremath{\attr{\vec{\tt #1}}{mat}}\xspace}     % matrix
\newcommand{\df}{\ensuremath{\attr{\l}{df}}\xspace}     % data.frame

\begin{abstract}
The R programming language is widely used in a variety of scientific domains
for tasks related to data science. The language was designed to favor an
interactive style of programming with minimal syntactic and conceptual
overhead. This design is well suited to support interactive data analysis,
but is not well suited to generating performant code or catching programming
errors.  In particular, R has no type annotations and all operations are
dynamically checked at runtime. The starting point for our work is the
question: \emph{what could a static type system for R look like?}  In order to answer
it, we study the polymorphism that is present in \RLOC lines of R
code spread among some \PACKAGES packages, written over a
period of over \YEARS years by thousands of programmers.  We perform a dynamic
analysis, leveraging tests and use-cases, to determine the level of
polymorphism that is present in the code. We do this for several potential
notions of types. Our results suggest that polymorphism is important in some
key parts of the system but that relatively simple type annotations could be
used to capture most of the interesting cases.

\end{abstract}

\maketitle

\section{PLDI TODO}

For one, the page limit is way less (12).
This is probably fine. 
I took out the graphs for now, maybe we`ll just want one giant page for all the figures?
We can figure out where to place them.

We want to beef up the contribution.
For this, we should:
\begin{itemize}
\item generate annotations, and turn those annotations to contracts, and check the contracts.
\item try to propagate information from inner functions to outer ones.
\end{itemize}

Refocus on: we analyze nearly all of the existing, readily-available code base for R, and from that analysis we develop a type system for the language.
Can abstract: analyze code for a language to develop a type system for that language.
We use R cause of the tests.

Remove fluff.

Address parametric polymorphism.

\AT{PLDI Layout:}

\begin{itemize}

\item Introduction

\item The R Language: R intro, types, introduce tracers

\item The Corpus: CRAN, talk about the code we looked at

\item Methodology: what is the methodology for developing our types, and of testing the packages?

\item Results: what are the results?

\item Related Work: plug related work

\item Conclusion

\end{itemize}

\section{Introduction}

Our community builds, improves, and reasons about programming languages.  To
make design decisions that benefit most users, we need to understand the
language we are working with as well as the real-world needs it
answers. Often, we as researchers can appeal to our intuition, as many
languages are intended to be general purpose and attract users with some
computer science training. Unfortunately, these intuitions do not always
apply to domain-specific languages, languages designed for and by a specific
group of users to solve very specific needs. This is the case of the data
science language R.

R and its ancestor S are languages designed, implemented, and maintained by
statisticians. Originally they were designed as glue languages, languages
that would allow users to read data into vectors and call statistical routines
written in Fortran. Over three decades, the languages became widely used
across many fields of science and in industry for data analysis and data
visualization; with time additional features were added.  Modern R, as a
linguistic object of study, is fascinating. It is a vectorized, dynamically
typed, lazy functional language with limited side-effects, extensive
reflective facilities and retrofitted object-oriented programming support.

Many of the design decisions that gave us R were intended to foster an
interactive, exploratory, programming style. These include, to name a few,
the lack of type annotations on variables and functions, the ability to use
syntactic shortcut, and the automatic conversion between data types.  While
these choices have led to a language that is surprisingly easy to use for
beginners---many data science programs do not teach the language itself but
simply introduce some of its key libraries---they have also created a
language where almost all computations yield a numeric result and where
errors can go undetected.

One way to increase assurance in the results obtained when using R would be
to add type annotations to functions and variable declarations. These
annotations could then be used, either statically or (more likely)
dynamically, to catch mismatches between expected and provided data values.
The nature of R is such that it is unlikely to be ever fully statically
checked, furthermore end users may not be willing to write types when
carrying out exploratory programming tasks. So, we are looking for an
optional type system that would allow us to capture as much of behavior of
library functions as possible while remaining easy to understand for
end-users and library developers alike.

This paper is a data-driven study of what a type system for the R language
could look like. Our intention is to propose changes to the
language, and for any changes to be accepted by the user community they
must clearly benefit the language without endangering backwards
compatibility. Our goal is thus to find a compromise between simplicity and
usefulness; the proposed type system should cover most common programming
idioms while remaining easy to use. In order to do this, we need to
understand the degree of polymorphism present in R code, that is to say, how
programmers leverage the dynamic nature of R to write code that can accept
arguments of different types.  This understanding will drive our design.

We propose to capture the degree of polymorphism present in R by means
of a dynamic analysis of widely used libraries. For each function call we
can record the types of its arguments and of its return value. This allows
us to observe how many different combination of types are accepted by any
given function. Unlike many other languages, R has a carefully curated
software repository called CRAN. To be deposited in CRAN, a package must
come with sample dataset, tests and executable use-cases. As part of normal
operations these tests are run regularly and failing packages are removed.
This allowed us to have access to \PACKAGES libraries and about an order of
magnitude more runnable scripts that exercise those libraries.

The contributions of this paper are thus as follows:
\begin{itemize}
\item A large-scale analysis of the polymorphism present in function
  signatures of \PACKAGES widely used and actively maintained R packages.
\item A tracing and analysis pipeline that extends a previously published
  test generation tool named \genthat.
\item A type system based on the evidence gathered, fully implemented with
  lightweight, non-intrusive contracts to perform runtime type-checking.
\end{itemize}

One threat to validity of our work is that we rely on dynamic analysis, so
our conclusions are only as good as the coverage of the possibly function
behaviors. Previous work~\cite{issta18} reported that running all the
scripts that come with CRAN packages gives, on average, 68\% test coverage.
We attempted to mitigate the threat coming from the fact that only part of
the code is being exercised by manual analysis. It would be reasonable to
ask for confirmation of the data by static analysis of the code, but sound
static analysis of R is difficult because of the extensive use of reflective
features such as \code{eval} and of the ability to redefine the meaning of
operators such as \code{+} and \code{if}.  Another threat to validity is
that we only have access to code that has been deposited in the CRAN
repository. While this may bias our findings towards code written to be
reusable and, possibly, better engineered than typical user code. This is
also the code that would most benefit from type annotations.

% \newpage

\section{The R Programming Language}\label{sec:rlang}

Over the last decade, the R Project has become a key tool for implementing
sophisticated data analysis algorithms in fields ranging from Computational
Biology~\cite{R05} to Political Science~\cite{R:Keele:2008}. At the heart of
the R project is a \emph{vectorized, dynamic, lazy, functional,
  object-oriented} programming language with a rather unusual combination of
features~\cite{ecoop12} designed to ease learning by non-programmer and
enable rapid development of new statistical methods.  The language, commonly
referred to as R was designed in 1993 by Ross Ihaka and Robert
Gentleman~\cite{R96} as a successor to S~\cite{S88}.  First released in
1995, under a GNU license, R rapidly became the lingua franca for
statistical data analysis. Today, there are over 13,000 R packages available
from repositories such as CRAN and Bioconductor.  With 55 R user groups
world-wide, Smith~\cite{eco11} estimates that there are over 2 million
end-users.

As an introduction to R, consider the code snippet in Fig.~\ref{sample} from
a top-level interaction where the user defines a function \code{normSum}
that accepts vectors of integers, logicals, doubles and complex values and
normalizes the vector with respect to its sum and rounds the results. The
function definition does not require type annotations, and all operations
transparently work on vectors of any length and different types.

\begin{figure}[!hb]{\small
\begin{lstlisting}[style=R]
> normSum <- function( m )  round( m / sum(m), 2)
> normSum(c(1L,3L,6L))
[1] 0.1 0.3 0.6
> normSum(c(1.1,3.3,6.6))
[1] 0.1 0.3 0.6
> normSum(c(1.6,3.3,6.1))
[1] 0.15 0.30 0.55
> normSum(complex(r=rnorm(3),i=rnorm(3)))
[1] 0.49+0.21i 0.30-0.18i 0.22-0.03i
\end{lstlisting}}
\caption{Sample R code}\label{sample}
\end{figure}

In R, function can be called with named parameters, R support variable
argument lists, and arguments can have default values. Putting all of these
together consider the following declaration:

\begin{lstlisting}[style=R]
f <- function(x, ..., y=3) x + y
\end{lstlisting}

\noindent
Function \k{f} can be called with a single argument \code{f(3)}, with named
argument \code{f(y=4,x=2)} and with a variable number of arguments,
\code{f(1,2,3,4,y=5)}, all of these calls will return \code{6}.

R has a number of features that are not crucial to the present
discussion. We will mention some of them here for completeness.  In R, data
structures are reference counted and have copy-on-write semantics, thus the
assignment \code{x[12]<-3} results in an update to a copy of \code{x} unless
the reference count on that object is 1.  This semantics gives R a
functional flavor while allowing updating in place within loops (the first
update copies, subsequent updates are performed on the copy). Arguments to
functions are evaluated only when needed, they are bundled in so-called
promises which package the original expression (as an abstract syntax tree, or AST), its environment
as well as the result of evaluating the expression. Promises can be
leveraged for meta-programming as it is possible to retrieve the text of a
promise and evaluate it in a different environment.

\subsection{Types of Data}

Before attempting to define a type system for R, we should understand the
different kinds of values that programs operate on.  As we will see
different notions of type may emerge depending on how granular we want to
be.

\renewcommand{\k}[1]{{\tt #1}\xspace}

R has one builtin notion of type that can be queried by the \k{typeof}
function. Over the years, programmers have found the need for a richer type
structure and have added {\it attributes}. The best way to think of attributes is
as an optional map from name to values that can be attached to any object.
Attributes are used to encode various type structures. They can be queried
with functions such as \k{attributes} and \k{class}.

\begin{wrapfigure}{r}{6.1cm}
\footnotesize\begin{tabular}{l|c|l@{}}\hline
\multicolumn{3}{l}{\bf Vectorized data types:}  \\\hline
\k{logical}   & \L & vector of boolean values\\
\k{integer}   & \I & vector of 32 bit integer values\\
\k{double}    & \D & vector of 64 bit floating points\\
\k{complex}   & \X & vector of complex values\\
\k{character} & \C & vector of strings values\\
\k{raw}       & \R & vector of bytes\\
\k{list}      & \l & vector of values of any type\\\hline
\multicolumn{3}{l}{\bf Scalar data types:}\\\hline
\k{NULL}      & \sN &  singleton null value\\
\k{S4}        & \sS &  instance of a S4 class \\
\k{closure}   & \sF & a function with its environment\\
\k{environment}&\sE &  a mapping from symbol to value \\\hline
\multicolumn{3}{l}{\bf Implementation data types:}\\\hline
\multicolumn{3}{l}{\k{special},
\k{builtin},
\k{symbol} (\sY),
\k{pairlist},
\k{promise}}\\
\multicolumn{3}{l}{
\k{language},
\k{char},
\k{...},
\k{any},
\k{expression},
}\\
\multicolumn{3}{l}{
\k{externalprt},
\k{bytecode},
\k{weakref}}\\\hline
\end{tabular}\caption{Builtin Types}\label{types}\end{wrapfigure}

Figure~\ref{types} lists all of the builtin types that are provided by the
language. They are the possible return values of \k{typeof}. There is no
intrinsic notion of subtyping in R. But, in many context a \k{logical} will
convert to \k{integer}, and an \k{integer} will convert to \k{double}.  Some
off conversion can occur in corner cases, such as \k{1<"2"} holds and
\k{c(1,2)[1.6]} returns the first element of the vector, as the double is
converted to an integer. R does not distinguish between scalars and vectors
(they are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
  typeof(c(5,5))} \code{ == "double"}. Finally all vectorized data types have a
distinguished missing value denoted by \code{NA}. The default type of
\code{NA} is \k{logical}. We can see that \code{typeof(NA)=="logical"}, but
NA inhabits every type: \code{typeof(c(1,NA)[2])=="double"}.

With one exception all vectorized data types are monomorphic, the exception
is the \k{list} type which can hold values of any other type including
\k{list}. For all monomorphic data types, attempting to store a value of a
different type will cause a conversion. Either the value is converted to the
type of vector, or the vector is converted to the type of the value.

Scalar data types include the distinguished \k{NULL} value, which is also of
type \k{NULL}, instance of classes written using the S4 object system,
closures and environments.  The implementation of R has a number of other
types that are mostly not used by user code, they are listed in
Figure~\ref{types} for reference.

The addition of attributes lets programmers extend the set of types by
tagging data with user-defined attributes. For example, one could define a
vector of four values, \code{x<-c(1,2,3,4)} and then attach the attribute
\k{dim} with a pair of numbers as value: \code{attr(x,"dim")<-c(2,2)}.  From
that point, arithmetic functions will treat \k{x} as a 2x2 matrix. Another
attribute that can be set is the \k{class}.  This attribute can be bound to
a list of class names. For instance, \code{class(x)<-"human"}, set the class
of \k{x} to be \k{human}.  Attributes are thus used for object-oriented
programming. The S3 object system support single dispatch on the class of
the first argument of a function, whereas the S4 object system allows
multiple dispatch (on all arguments). Some of the most widely used data
type, such as data frames, leverage attributes. A data frame, for instance,
is a list of vectors with a class and a column name attribute.

\paragraph{Summary.} The most common values in R computations are vectorized
types. R programs do not have a way to constrain values to be scalar.
\k{NULL} is sometimes used to represent the case when no value is
available. \k{NA} is used within vector to represent missing observations.
Attributes can decorate values and are used as building blocks for
object-oriented programming. A potential type system for R could focus only
on the builtin types, if one wanted to strive for simplicity, or it could
try to capture attributes at the risk of increased complexity.

\newpage
\section{Corpus}\label{sec:corpus}

In this section, we present our dataset. The R language aims to accommodate
data analysts; their workflows start with data import, followed by cleaning,
and then by steps of modeling, transformation and visualization. Often, the
code of these analysis pipelines resides, together with the data and
results, in notebooks. Few notebooks are publicly shared, and when they are,
the data is not. For this reason our analysis focuses on packages which
bundle reusable units of R code with documentation, sample data and
use-cases.

We focus on packages hosted on the \emph{Comprehensive R Archive Network} or
CRAN.  With over 13,000 packages, CRAN is the largest repository of software
written in R. It is experiencing sustained growth with an average of size
new packages a day~\cite{LIgges2017}.  Unlike sites like GitHub, CRAN is a
\emph{curated} collection: A package is only accepted to CRAN if it abides
by a number of well-formedness rules.  Most relevant for our purposes,
packages must have data, examples, vignettes and tests, all of which must
successfully run. From our perspective this means that each package in CRAN
comes with several executable scripts that exercise some of its
functionality.  Notable exceptions to this rule are packages only containing
data, which have no runnable code but are referenced by other packages.
Only \DATAPKGS packages had no executable code, accounting for \DATAPKGSPERC
of CRAN.

The corpus used in this paper is a subset of CRAN. We retained packages that
could be run by our infrastructure in less than one hour. This corpus
consists of \PACKAGES packages, accounting for some \PERCENTCRAN of all
packages.  These packages have a total of \RLOC lines of R code and \CLOC
lines of C code. Figure~\ref{allcloc} shows a per-package breakdown of the
size of each package sorted by increasing numbers of lines of R. The figure
suggests that there is little correlation between the C/C++ parts and how
many lines of R a package contains. The median size of a package is 541
lines of R code and the largest package has 86K LOC. Typically, C/C++ code
is used to implement performance critical portions of the code. The majority
of packages, 8,375 to be precise, have no C or C++ code.  The remaining 3,078
packages have a median 572 lines of C and the largest package has 385,839
lines of C.

\begin{comment}

\begin{figure}[!b]\begin{center}
\includegraphics[width=.9\textwidth]{linesofrandccode}
\caption{Lines of code (log scal); for each package, R is above 0, and C/C++
  below}\label{allcloc}\end{center}
\end{figure}

\end{comment}

For each package, we extracted all executable code snippets from
documentation, vignettes and tests and ran them independently, recording all
calls to R functions.  It is noteworthy that in order to run the scripts in
one package, it is often necessary to load a number of other packages.
In~\cite{issta18}, the authors estimated code coverage to be around 68\%
when including reverse dependencies.  As our infrastructure adds overhead to
script execution, coupled with the fact that some scripts take inordinate
amounts of time to run, we limited our analysis of any one
package to one hour.

% https://www.r-pkg.org/downloaded
\begin{figure}[!th]{\footnotesize\begin{tabular}{@{}r||l|r|r|r|r|r@{}}\hline
\bf Package & \bf Description & \bf R LOC &\bf C LOC &\bf Scripts & \bf Calls Observed & \bf Signatures Recorded \\
\hline
\tt Rcpp  & Seamless C++ integration & 2.2K & 4.2K & 25 & 55K & 340 \\
\tt rlang & Functions for 'Tidyverse' & 7.0K & 6.1K & 122 & 3,924K & 8,422 \\
\tt glue  & Interpreted string literals & 0.3K & 0.3K & 8 & 4K & 145 \\
\tt tibble & Simple data frames & 2.0K & 0.3K & 16 & 1,332K & 6,367 \\
\tt stringi &  String processing & 1.5K & 515K & 64 & 923K & 873 \\
\tt ggplot2 & Data visualisations & 14K & 0 & 130 & 153K & 4,608 \\
\tt dplyr  &  Data manipulation & 4.5K & 4.7K & 78 & 233K & 3,099 \\
\tt pillar & Formatting for columns & 1.4K & 0 & 13 & 803K & 1,514 \\
\tt R6 & Classes w. ref. semantics & 0.7K & 0 & 2 & 1K & 330 \\
\tt stringr & String operations & 0.5K & 0 & 32 & 1,764K & 534 \\
\end{tabular}}\caption{10 Most Downloaded Packages.}\label{most}
\end{figure}

Figure~\ref{most} shows the ten most downloaded CRAN packages.  For each
one, we list how many lines of R and C/C++ the packages contains.  We show
the number of scripts that could be extracted from the package. Each script
corresponds to either one use-case or a set of unit tests. We print the
number of function calls that were observed by our infrastructure, and the
number of unique call signatures that were recorded.

Our infrastructure only retains unique argument/return combinations. Thus,
while we observe large number of functions being called with different
values, the types of these function call are often similar. Over the entire
corpus, we can see the relation between observed calls and recorded signatures in
Figure~\ref{recorded}.  The median number of observed calls is 82 and
maximum is 19 million.  The median number of recorded signatures is 16 and
the maximum is 8,422. These numbers are skewed by a number of scripts doing
very few calls before plunging into C code.

\begin{comment}

\begin{figure}[htbp]\begin{center}
\includegraphics[width=.9\textwidth]{recordsbypkg}
\caption{Numbers of Recorded Function Invocations in the Analyzed Corpus}
\label{recorded}\end{center}
\end{figure}

\end{comment}

% \newpage
\section{Methodology}

\subsection{PLDI Changes}

We need to make some changes for this submission, as there are now two distinct methodologies.

First, to {\it develop the type system}, we analyzed function calls in package code, including considering package reverse dependencies (i.e. clients of the package) to develop complete type signatures.
\AT{Do we want to do this?} We additionally propagate type information for inner functions to outer ones by connecting known signatures of inner functions to outer functions. 
We previously justify how the R ecosystem is particularly well-suited for this, as package code is curated and packages ship with a suite to showcase package use.

Separately, to {\it test our annotations}, we run our inference tool on package code (incl. aforementioned suite) to synthesize type signatures, turn those signatures into annotations, and turn those annotations into contracts on package functions.
Then, we run the example suites of all client packages to test if the annotations were effective.

\subsection{Old}

In this section, we detail our methodology for collecting data.  Our aim is
to observe arguments and return values of function calls, and from these
generalize possible type signatures for the called functions.  We base our
infrastructure on an open source tool called \genthat whose purpose is to
generate unit tests for R libraries~\cite{issta18}.  \genthat achieves this
by synthesizing unit tests from recorded function argument and return
values, comparing candidate tests against existing ones to avoid generating
tests which do not increase code coverage.  To suit our purposes, we change
the existing tool in two main ways: we record \emph{shapes} rather than
values, and ignore the code coverage optimization phase.  Both changes are
beneficial for scalability, allowing us to trace far more calls.

To illustrate our approach, consider the following script which adds a
double to an integer, and then creates a matrix from a vector, finally
adding them together to get a new matrix.

\begin{figure}[!hb]
\begin{tabular}{ll}\begin{minipage}{5cm}
{\small\begin{lstlisting}[style=R]
> 1 + 2L
[1] 3
> x <- c(1,2,3,4)
> attr(x,"dim") <- c(2,2)
> x + c(1,2)
     [,1] [,2]
[1,]    2    4
[2,]    4    6
\end{lstlisting}}
\end{minipage} &
\begin{minipage}{8cm}\small
\begin{tabular}{rl}
\tt `+`: &\tt \sD \sI \to \sD \\
\tt c:& \sD \sD \sD \sD \to \D\\
\tt `<-`: &\tt \sY \D \to \D\\
\tt c: & \tt \sD \sD \to \D\\
\tt `attr<-`:&\tt  \sY \C \D \to \\
\tt c: &\tt \sD \sD \to \D\\
\tt `+`: &\tt \attr\D{dim} \D \to \attr\D{dim}
\end{tabular}
\end{minipage}
\end{tabular}
\caption{Example script and recorded signature}\label{example}\end{figure}

The functions being called here are \k{`+`}, \k{`<-`}, \k{c} and
\k{`attr<-`}, and the shapes we expect to record are given to the right of the figure.
We abbreviate the types \k{double}, \k{integer}, \k{symbol} and \k{character} to \sD, \sI, \sY, and \sC for simplicity.
\D, \I, \Y, and \C denote vectors, whereas \sD, \sI, \sY, and \sC denote scalars.
We record the lengths of all vectorized-typed values, and for all types we record their classes and attribute names. 
Looking at the signatures observed for addition, it is clear that the function is polymorphic: it starts with the addition of two scalar numbers of different types, and later adds a matrix to a vector returning a matrix.

\subsection{Implementation}
\label{sec:Impl}

A high-level description of the workflow of our tool for one package
retrieved from CRAN is as follows:

\begin{enumerate}
\item {\bf Exec generation:} All runnable code in the package is extracted
  from its tests, examples and vignettes. The code snippets are combined
  into a single file.
\item {\bf Installation:} All packages that are required for execution of
  the current package are downloaded and installed.
\item {\bf Instrumentation:} As code is loaded into the R, every function
  definition is instrumented with an \code{on-exit} hook which is invoked
  when the function returns either normally or through an exception.
\item {\bf Recording:} When a hook is called, arguments and return value are
  inspected. We record \k{typeof}, \k{class} and \k{attributes} recursively.
  For \code{list} values, an extra bit of analysis is performed to record
  the element types.
\item {\bf Writing:} Unique signatures are recorded to file with information
  about which package triggered the recording.
\end{enumerate}

In R, all arguments are passed as promises, and unused arguments will remain unevaluated promises, which we record as {\it unevaled}. 
Our recording mechanism captures the order in which arguments are passed, can figure out which arguments were {\it not} provided (recording them as {\it missing} if the missing argument has no default value), and does {\it not} force unforced promises.
In R, trying to use missing values results in an error and thus an exception, triggering our recording phase: arguments which cause errors in this way are recorded as {\it error}.
For practical reasons, we do not record the contents of environments:
These can be used as hash tables, may be large, and are quite likely different from one another.

Unevaluated arguments in R are sometimes {\it carefully} used by their functions to implement some form of meta-programming.
Consider the {\tt magrittr} package, which implements a pipe function \%$>$\% that implements a flavor of function currying:

{\small\begin{lstlisting}[style=R]
mul <- function(x,y)  x*y
4 %>% mul(3)
\end{lstlisting}}

The call to the pipe desugars to \code{`\%>\%`(lhs=4, rhs=mul(3))} where the
\code{rhs} is not a valid expression. It is treated as code that will
manipulated to add one argument (the \code{lhs}), resulting in the function
actually performing a call of the form \code{mul(4, 3)}.

As we alluded to at the beginning of this section, our analysis collects shape information about
arguments and return values in order to be as specific as possible. The details of the information we collect follows.
Note that the first step of any type collection in our analysis begins with \code{typeof}.

\begin{itemize}
\item If the analysis encounters a primitive vector, say
  \code{typeof(x)=="double"}, it records the length of {\tt x} to determine
  if it can be treated as a scalar.
\item For a list, \code{typeof(x)=="list"}, shape information on all
  elements is used to describe the list's shape.  To avoid undue slowdowns,
  we only collect the content's \code{typeof}, and ascribe \ANY (any) if the
  elements are not of the same kind.
\item If the analysis encounters a scalar \code{NA}, we ascribe a unique
  NULL type.  In R, \code{NA} inhabits all types, but for a scalar NA,
  \code{typeof(NA)=="logical"}. Scalar NAs are used as uninitialized values.
\item For a matrix, \code{typeof(m)=="some\_primitive\_type"}, and\code{m} will have matrix class and also a dims and
  optionally dimnames attribute. Here, we ascribe the shape \attr{\tt T}{mat}
  where {\tt T} is the primitive element type.
\item For a data frame, \code{typeof(df)=="list"}, and \code{df} will have class {\tt data.frame}, and {\tt names} and {\tt row.names} attributes.
 To reflect this, we ascribe the shape \attr{\l}{df}.
\end{itemize}

The infrastructure memoizes emitted signatures to avoid writing multiple
identical signatures for the same function.

\section{Types for R}

Our aim is to capture the polymorphism in R, and to do so we must define a
meaningful set of basic types for the language. Also, we are interested in
finding a type system that could be adopted, for this we aim to investigate
the simplest possible solution.  There are several possible choices, with a
variety of levels of expressivity and complexity.  For concreteness, we will
pick out three different sets of types.  To start simple, we define a set of
types which closely align with R's builtin types, and later define richer
notions of type which incorporate attributes, classes, and capture the
various object systems of R.

Throughout this paper, we will say that an argument is {\it polymorphic at}
\textbf{\emph{LX}} (read: level X) if it has been called with at least two
values that belong to distinct types at {\bf LX}.  \textbf{L2} will
incorporate class and attribute lists, which are slightly trickier to
``type'', and we defer discussion of this until Section~\ref{sec:L2def}.

\subsection{L0 Types}
\label{sec:L0def}

The logical ``baseline'' for the design of a type system for R is to adopt
the language's builtin types.  We dub these {\bf L0} types, and they are
primarily built on the \code{typeof} values, which reflects the
implementation of the type in the runtime environment.  More specifically,
the \code{typeof} a value is simply the runtime type tag of the value.
Given that, {\bf L0} includes the following types: vectors of primitive
types (integers \I, double \D, character \C, logical \L, raw \R, complex
\X), lists of any type (\l), and scalar values (environment \sE, closures
\sF, symbols \sY, S4 object \sS, and null \sN). We omit short forms for some
of R's more esoteric types (e.g., {\tt environment}s).  We add the any
(written \ANY) type to stand in for unevaluated values Subtyping in {\bf L0}
follows from the conversion rules of the language: Logicals are subtypes of
integers, which are themselves subtypes of doubles (so \L <: \I <: \D).  All
types are subtypes of \ANY, thus {\tt T}<:\ANY for an type {\tt T}.

%
\subsection{L1 Types}
\label{sec:L1def}

{\bf L0} is limited in expressivity as it (1) supports polymorphism for
lists in a trivial way, (2) does not address some of the popular constructed
data types such as matrices and data frames, and (3) cannot differentiate
between scalar values and vectors.  {\bf L1} types build on and extend {\bf
  L0} with with scalar values for primitive types (\sI, \sD, \sC, \sL, \sR,
and \sX), type-parametric lists (written \lT{T} for any type {\tt T}), data
frames (written \attr{\l}{df} as they are built up from lists of type any),
and matrices (\attr{\vec{\tt T}}{mat} where {\tt T} is a primitive type).
Subtyping is extended such that {\tt T}<:{$\vec{\tt T}$} for any primitive
type {\tt T}. The subtype relation is undefined for lists of different
types, as well as for matrices and data frames.

\subsection{L2 Types}\label{sec:L2def}

{\bf L2} extends the previous system with support for classes and
attributes.  While there are a finite number of results from the
\code{typeof} function, this is {\it not} the case for \code{class} and
\code{attributes}, as classes and attributes {\it can} be user-defined.  In
terms of classes, R has a number of built-in or otherwise standard classes,
ranging from truly primitive classes such as {\tt numeric} to the more
complex {\tt data.frame}.  {\bf L1} already incorporates these builtin
classes, namely through its data frame and matrix types, so we will ignore
those classes and thus highlight those which are user-defined.  As for
attributes, again some patterns arise naturally in R, e.g. all matrices have
a {\tt dims} attribute, but even these are somewhat interesting, as a
programmer could in theory set their own {\it dims} attribute on an
arbitrary value.  The only attribute we ignore is the {\tt class} attribute,
as we collect the class of values separately (and the {\tt class} attribute,
if present, is always equal to the class of a value).  Attributes are
name-value pairs, and to keep things simple we consider only attribute
names.

In sum, we extend {\bf L1} with attribute lists and classes:
\attrclass{T}{C}{A} where {\tt T} can be any L1 type other than data frame
or matrix, and {\tt C} and {\tt A} are lists of classes and attribute names
respectively.  The subtype relation is defined as: \attrclass{T}{C}{A}
$<:_{\mathbf{L2}}$ \xspace \attrclass{S}{C'}{A'} if $ T
<:_{\mathbf{L1}} S$ (i.e. are subtypes in the sense of {\bf L1}), $C
\subseteq C'$, and $A \subseteq A'$.

\subsection{Simplification by Subtyping}\label{sec:autosub}

When analyzing our data, we perform a certain amount of simplifications
automatically to minimize the footprint of equivalent signatures.  This
subtyping is performed at the function level of granularity.  For example,
if a function is observed to have two signatures \sD \to \df to \D \to \df,
we collapse the signature to \D \to \df as \sD $<:$ \D and \df $<:$ \df.  We
would not collapse \sD, \I \to \df and \D, \sI \to \df, as \I $\not <:$ \sI
and \D $\not<:$ \sD.

%
%
%
\section{Examples of Polymorphism}\label{sec:polyex}

Before presenting the results of our analysis, we  quickly survey
examples of real, polymorphic R functions that we have analyzed, to build up
a global sense of how types manifest themselves in current R code.

We begin by referring the reader to Figure~\ref{fig:realex}, which showcases
the compatibility of integers and doubles.  Coercion from integer to double
is performed automatically where appropriate, and likewise coercion from
double to integer is automatically performed when integers are strictly
required.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
> 5L + 1L # => 6L, an integer
> 5L + 1.2 # => 6.2, a double (5L coerced to 5.0)
> c(10, 20, 30)[1.2] # => 10, 1.2 coerced to 1L
\end{lstlisting}}\caption{Example of \sD, \sI type usage}\label{fig:realex}\end{figure}

Figure~\ref{fig:optnull} shows a function with an optional NULL argument.
In the function, {\tt x} is a sorted vector, and {\tt w} a vector with
default value NULL.  If {\tt w} is NULL, then default unit weights are
generated.  This pattern of default NULL (or NA) arguments is prevalent.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
frac.ranks <- function(x, w=NULL) {
  if (is.null(w)) w <- rep(1, length(x)) # if no weights passed, take all weights = 1
  ...
\end{lstlisting}}\caption{Optional argument ({\tt acid} package)}\label{fig:optnull}\end{figure}

Figure~\ref{fig:listvec} shows a function with list and vector polymorphism.
It takes a list or a vector {\tt point} and a data frame {\tt polyg}.  The
{\tt point} is explicitly converted to a vector (by \code{as.numeric})
regardless of type.  This showcases the ease of going from lists to vectors
(and conversely).

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
is.point.inside <- function (point, polyg) {
    p <- as.numeric(point) # as.numeric(list(1, 2)) => c(1, 2)
    ...
\end{lstlisting}}\caption{List and vector polymorphism ({\tt bivrp} package)}\label{fig:listvec}\end{figure}

Figure~\ref{fig:charclos} is a function with character and function
polymorphism.  The caller specifies {\tt fn}, which is either a function
{\it or} the name of a function which will be looked up by name (a string).

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
pointest <- function (object, fn = mean, par.name, comp.label, chain.no,  ...) {
    if (is.character(fn))
        if (fn == "MODE" | fn == "MAP")   do_MAP <- TRUE
        else {
            do_MAP <- FALSE
            fn <- match.fun(fn) # looks up the function name
        }
\end{lstlisting}}\caption{Character and function polymorphism ({\tt BAMBI} package)}\label{fig:charclos}\end{figure}

In Figure~\ref{fig:dfdbl}, we see a function that, among other things, takes
in some data ({\tt dat}) and a character string ({\tt spss}) specifying
either {\tt "in"} or {\tt "out"}.  Here, the type of {\tt dat} is {\it
  dependent} on the value of {\tt spss}: if {\tt spss == "in"}, {\tt dat}
must be a data frame, and otherwise if {\tt spss == "out"}, {\tt dat} must
be a vector.  Our analysis does not capture this granularity of dependency,
as we only collect type information.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
nret.translator <- function(dat, spss="out", ...) {  ...
  if(identical(spss, "out")) {
    if(!is.vector(dat))  stop(simpleError("...'dat' must be a vector!")) ...
  } else {
    ...
    items.idx <- items.idx[order(names(dat[, items.idx]))]
  }
\end{lstlisting}}\caption{Example of a \df/\D argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}

In Figure~\ref{fig:chardbl}, we see a function which takes in a list ({\tt
  network}), a vector of indices of that list ({\tt fixIndices}), and a
vector of values ({\tt values}).  The locations specified by {\tt
  fixIndicies} in {\tt network} are updated with {\tt values}.  Here, {\tt
  fixIndicies} has been observed to be either a character or double vector:
In R, list indices are typically doubles, but can be characters (if the list
or vector being indexed has a {\tt names} attribute).  Note that type level
{\bf L2} can capture the presence of a {\tt names} attribute.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
fixGenes <- function (network, fixIndices, values) {
  ...
  network$fixed[fixIndices] <- as.integer(values)
  ...
\end{lstlisting}}\caption{Example of \C/\D argument (from {\tt BoolNet} package).}\label{fig:chardbl}\end{figure}%$

In Figure~\ref{fig:matvec}, we see that a function which can take in a vector,
but immediately transforms it (transpose, with {\tt t}) into a matrix.  As
an idea, matrix/vector polymorphism seems sensible, as mathematically
vectors {\it are} one-dimensional matrices.  R echoes this by making conversion between the two
types easy: \code{as.vector(m)} flattens a matrix \code{m} into a vector
(e.g., \code{as.vector(matrix(1, 2))} is equivalent to \code{c(1, 1)}), and
\code{as.matrix(v)} builds a {\tt length(v) x 1} matrix (e.g.,
\code{as.matrix(c(1, 2))} is equivalent to \code{matrix(data=c(1, 2))}).

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
tee <- function (x, theta, D1, D2, phi)  {
    if (is.vector(x)) x <- t(x)
    ...
\end{lstlisting}}\caption{Example of \M{D}/\D argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}

Throughout this short section, we have shown real examples of polymorphic R functions.
Our goal is to capture these instances of polymorphism across a large subset of existing R code, and to highlight and discuss patterns which arise.
We present the results of our analysis in the following section.

%
%
%
%
%
%
\section{Analysis Results}\label{sec:results}

This section,  presents the results of our analysis at each of our type levels, describing the types of polymorphism that we have observed and shedding light on the frequency of polymorphic patterns in R.

Before we begin, we define the kinds of polymorphism we will consider.  In
choosing how to express the polymorphism of a function, we consider two
simple options: First, having the union of signatures, e.g. \D $\rightarrow$
\D and \C $\rightarrow$ \D, or a union types for each argument, e.g. (\D
$\cup$ \C) $\rightarrow$ \C.  We refer to the former as {\it function-level}
and the latter as {\it argument-level polymorphism}.

Per-argument polymorphism highlights the type-based behavior of function
arguments themselves, and helps to describe the type-based behavior of all R
values.  For instance, we observe that many arguments have (\sC, \sF) as a
signature, from which we deduce that many R programmers are taking advantage
of functions such as \code{match.fun("fun\_name")} to call functions by
name.  On the other hand, function-level polymorphism highlights how R
programmers interact with functions as a whole, and allows us to discern
more subtle patterns.  For example, observing two function signatures such
as \D \to \D and \sF \to \sF may lead us to ascribe a more useful and
precise type for the function than if we had simply collapsed the argument
and return types (e.g., $X$ \to $X$ vs. \D $\cup$ \sF \to \D $\cup$ \sF).

\subsection{L0}

At {\bf L0}, the type system is mostly aligned with the types of the core
language. Across all of the arguments that we have observed, the vast
majority are monomorphic.  Table~\ref{tab:L0top10arg} shows the breakdown of
the 10 most frequently occurring argument signatures.

\begin{table}[ht]
\centering
\begin{tabular}{lrr}
  \hline
 Argument Signature & Count & Percentage \\ 
  \hline
  \D & 269K & 29.7\% \\ 
  \l & 193K & 21.3\% \\ 
  \C & 96K & 10.6\% \\ 
  \L & 83K & 9.2\% \\ 
  \sF & 60K & 6.6\% \\ 
  \sN & 47K & 5.2\% \\ 
  \ANY & 31K & 3.4\% \\ 
  \I & 30K & 3.2\% \\ 
  \sS & 16K & 1.7\% \\ 
  \D, \l & 10K & 1.1\% \\ 
   \hline
   \hline
\end{tabular}
\caption{Most frequent argument signatures in {\bf L0}.}
\label{tab:L0top10arg}
\end{table}

Doubles are the most common argument type, which is not entirely surprising
as doubles are the backbone of most statistical functions.  Next is lists,
which is reasonable as lists are used to build data frames and other widely
used data structures.  The \ANY type is not an R type per se: we ascribe
such a type to arguments which were {\it always} unevaluated across all
calls.  As it happens, many R functions meta-program on some of their
arguments, and this meta-programming does not force the argument promise,
and if this meta-programming is always performed in every recorded call we
will ascribe the \ANY type, as we cannot say what the argument {\it is}
without possibly forcing a promise.  The magrittr package's pipe \%$>$\%
function is an example of a function which does meta-programming; we will
discuss it shortly.  Interestingly, a polymorphic attribute signature \D, \l
is among the top 10 overall---this is the vector and list polymorphism made
manifest.

We turn our attention to function-level polymorphism.  Indeed, one of our
goals is to generate useful function signatures, so we ask: {\it how many
  functions have more than one recorded signature?}
Figure~\ref{fig:L0funcounts} has the answer, with a breakdown by function.
Most functions are monomorphic.  That said, a few functions have
disproportionately large number of signatures.  One such function is the
magrittr package's pipe \%$>$\%, which is used to implement a kind of
currying. It has nearly 100 unique signatures!

\begin{comment}

\begin{figure}[htbp]\begin{center}
\includegraphics[width=.9\textwidth]{L0_by_fun}
\caption{Record of Function Polymorphism in {\bf L0}; \LZEROPERCPOLY of
  functions are polymorphic}
\label{fig:L0funcounts}\end{center}
\end{figure}

\end{comment}

Another function with many signatures is the data.table package's \code{[}
  function, which is a reimplementation of the square bracket operator for
  selecting values in a data.table.  This function has 16 formal arguments,
  and call configurations vary widely.  Essentially, the function is {\it
    the} main function for data.tables, and performs very different tasks
  based on which arguments are provided.  Two wildly different uses are:
  \code{dtT[i,j,by]}, where columns \code{i} and \code{j} are accessed and
  grouped by \code{by}, and \code{dt[x=="some" & y>42]}, which selects
  all rows who's \code{x} column is the string \code{"some"} and who's
  \code{y} column is greater than \code{42}.  Given that, we find it
  unsurprising that the function is so polymorphic.

\begin{table}[ht]\centering
\begin{tabular}{lrr}  \hline
Signature & Count & \% of All Signatures \\
\hline
  \D, \l & 9566 & 1.1\% \\ 
  \D, \sN & 8801 & 1\% \\ 
  \C, \sN & 5590 & 0.6\% \\ 
  \l, \sN & 5016 & 0.6\% \\ 
  \C, \D & 2887 & 0.3\% \\ 
  \D, \sF & 1608 & 0.2\% \\ 
  \l, \sF & 1345 & 0.1\% \\ 
  \C, \l & 1283 & 0.1\% \\ 
  \C, \sF & 1168 & 0.1\% \\ 
  \C, \I & 1104 & 0.1\% \\ 
   \hline
\end{tabular}
\caption{Top polymorphic argument signatures for the L0 type system.}
\label{tab:L0toppoly}
\end{table}

Having had a taste for function-level polymorphism, we will now explore
common patterns in argument-level polymorphism.  For this, refer to
Table~\ref{tab:L0toppoly}.
Let's unpack some of the signatures in Table~\ref{tab:L0toppoly}:

\begin{itemize}
\item \D, \l (and \C, \l): L0 does not parameterize list types, and it is very likely that this signature describes vectors of doubles and lists of doubles.
	Coercing vectors to lists is trivial in R, and indeed many of the functions with such an argument signature do.
	The story is similar with vectors of characters and lists of characters.

	\item $\square$, \sN: there are a few instances of this ``null'' polymorphism.
	The frequency of these signatures indicates a tendency to create ``optional arguments'', where some argument's default value might be NULL.
	This pattern is quite frequent in R code, and an example can be found in Figure~\ref{fig:optnull}.

	\item \C, \D and \C, \I: these are interesting patterns.
	In R, lists can always be indexed with numerics (\L, \D, \I) {\it and also} characters.
	When a list is indexed by character, the character is matched against the element names in the list and the corresponding element is selected.
	If there is no match, the lookup returns NULL.
	Another possible explanation for these signatures are dates, which can be represented by numerics or characters and the value usually has one of the date classes.
	L0 does not deal with classes, which are one of the topics of L2.

	\item \D, \sF: this signature appears more strange than it is.
	Many models, processes, simulations, etc. that are prevalent throughout R have parameters to fine-tune their behavior.
	This pattern manifests itself here: some arguments with the \D, \sF signature are either parameters for a function (\D), or the function itself already constructed (\sF).
	Others, such as the {\tt br} argument of the {\tt rtree} function in the {\tt ape} package, are either parameters for complex data generation processes (\D), or a function to be used during the generation (\sF).
	It would appear that this pattern is used to make complex simulation, modelling, or data generation functions more flexible for programmers by allowing them to supply parameters or the parameterized function itself.

	\item \C, \sF: yet another puzzling signature, though its explanation is surprisingly straightforward:
	In R, programmers are free to look functions up {\it by name}, and indeed the frequently encountered {\tt UseMethod} function for dynamic class-based function dispatch is one such example.
	In short, this signature usually represents a function or a function name.
\end{itemize}

We have established and discussed some of the most common polymorphic
attribute signatures, and have built up an understanding of argument-level
polymorphism in the {\bf L0} type system.  This matches almost exactly with
the language as it is today, as {\bf L0} is in (nearly) 1:1 correspondence
with how types are currently represented in the language.  The next question
we address is: {\it just how pervasive is this argument polymorphism?}
We find that 94\% of arguments are monomorphic.

From Table~\ref{tab:L0argcounts} we see that, with our definition of
subtyping on L0 and other simplifications we made, an overwhelming amount of
argument polymorphism is accounted for.  We also saw in
Table~\ref{tab:L0toppoly} that each of the most frequently occurring
argument signatures had some explanation.

That said, L0 is not without limitations, and the main issues stem from a
lack of adequate information, as the type system does not capture precise
list types, and does not distinguish scalars from vectors.

%
%
\subsection{L1}

L1 espouses a finer-grained approach to types.  Major modifications from L0
include:

\begin{enumerate}
	\item the differentiation of scalars and vectors;
	\item the parameterization of list types over element types;
	\item the use of attribute and class information to distinguish types for data frames and matrices, and;
	\item subtyping of \sN (i.e., $\forall T$, \sN $<:$ $T$).
\end{enumerate}

An example of a signature in this type system can be found in
Figure~\ref{fig:exL1}.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
add_make_list <- function(x, y) {
  l <- as.list(x + y)
  attr(l, "example") <- "this is an attribute"
  l
}
add_make_list(2, c(3, 1)) # => list(5, 3)
\end{lstlisting}}
\caption{Call signature in L1: x: \sD, y: \D $\rightarrow$ \lT{D}.}\label{fig:exL1}\end{figure}

Let us begin as we did in L0 by looking at the breakdown of argument
signatures in L1.  Consult Table~\ref{tab:L1top10arg}.
\begin{table}[ht]
\centering
\begin{tabular}{lrr}
  \hline
 Argument Signature & Count & Percentage \\ 
  \hline
  \sD & 179K & 19.7\% \\ 
  \D & 97K & 10.7\% \\ 
  \sL & 77K & 8.6\% \\ 
  \sC & 75K & 8.3\% \\ 
  \l & 67K & 7.4\% \\ 
  \sF & 61K & 6.7\% \\ 
  \sN & 50K & 5.5\% \\ 
  \M{D} & 49K & 5.5\% \\ 
  \ANY & 31K & 3.4\% \\ 
  \df & 29K & 3.2\% \\ 
   \hline
\end{tabular}
\caption{Most frequent argument signatures in L1.}
\label{tab:L1top10arg}
\end{table}

From Table~\ref{tab:L1top10arg}, we see that all of the most frequent
argument signatures are monomorphic even though we increased the precision
of the types.  We see that scalars actually figure quite prominently, which
is encouraging: this supports an idea that an explicit scalar annotation or
type may be useful {\it for programmers}---indeed, such an annotation would
be helpful for compiler optimizations (e.g., scalar-only values need not
have vectors allocated to store them).  We also see that matrices (\M{D})
are quite popular, a testament to the versatility of the data structure, as
they can easily be converted to vectors and data frames.

Now that we have a picture of how arguments look in L1, let's see: {\it how many L1 functions have more than one type signature?}
The answer can be found in Figure~\ref{fig:L1funcounts}.

\begin{comment}

\begin{figure}[htbp]\begin{center}
\includegraphics[width=.9\textwidth]{L1_by_fun}
\caption{Record of Function Polymorphism in {\bf L1}. \LONEPERCPOLY of functions are polymorphic.}
\label{fig:L1funcounts}\end{center}
\end{figure}

\end{comment}

We see that we barely increased the polymorphism with our transition to L1,
despite the far richer type information.  The biggest outlier here is indeed
\%$>$\%, as was the case in {\bf L0}.  Otherwise, the shape is very similar.
In short: even though {\L1} types are more expressive, the breakdown of
per-function polymorphism is largely unchanged from {\bf L0}.

As we were in L0, we are curious to know how argument-level polymorphism
looks in L1.  Refer to Table~\ref{tab:toppolyL1}.

\begin{table}[ht]\centering\begin{tabular}{lrl}  \hline
Signature & Count & \% of All Signatures \\   \hline
  \D, \M{D} & 4848 & 0.5\% \\ 
  \l, \lT{D} & 1843 & 0.2\% \\ 
  \sC, \sD & 1590 & 0.2\% \\ 
  \M{D}, \M{I} & 1502 & 0.2\% \\ 
%  \l, \lT{list} & 1199 & 0.1\% \\ 
  \df, \M{D} & 1180 & 0.1\% \\ 
  \C, \D & 1019 & 0.1\% \\ 
  \sD, \sF & 1005 & 0.1\% \\ 
  \I, \sD & 997 & 0.1\% \\ 
  \sC, \sF & 991 & 0.1\% \\    \hline
\end{tabular}
\caption{Top polymorphic argument signatures in L1.}\label{tab:toppolyL1}
\end{table}

So what do these signatures mean?

\begin{itemize}
	\item \D, \M{D}: vectors are just one-dimensional matrices.
	Indeed, the only thing separating a vector from a matrix in R is a {\tt dims} attribute.

	\item \l, \lT{D}: this hearkens back to L0, where we discussed how this very same signature was explained by the list being a list of doubles.
	So why is it still here?
	Well, first note that there are far less of these signatures in L1 than in L0, so clearly we captured {\it something}.
	Next, note the method with which we parameterize lists: we ascribe some type $T$ to a list type \lT{T} if and only if {\it all} list elements have {\it exactly} type {\it T}---in our analysis, we have encountered lists with integers as well as doubles, and unfortunately those currently appear as \l.
	This could be addressed by doing some subtyping computation while exploring list elements.

	\item \sC, \sD and \C, \D: we discussed this in L0---this signature likely captures an index or a date.

	\item \M{D}, \M{I}: this signature is reasonable, as any computation in R on \M{D} would work on \M{I} and vice versa.

%	\item \l, \lT{list}: here we have a list of any and a list of lists.
%	One explanation for this is that lists of lists sometimes contain NULL as an element, and this would show up with type \l, again as we do not do any subtyping while computing precise list types.
%	Another explanation is the way R handles object-orientation, which we discuss shortly in Section~\ref{sec:S3S4R5}.

	\item \df, \M{D}: in R, it is easy to transform a matrix into a data frame, and this argument is likely a result of that:
	Simply calling \code{as.data.frame} on a matrix is enough.

	\item \sD, \sF and \sC, \sF: these were discussed in the context of L0.

	\item \I, \sD: subtyping between primitives does not hold between scalars and vectors (i.e., \I $\not<:$ \sD) even though it is likely that the programmer intended for the argument to be a vector of the supertype (here, these arguments were likely intended to be of type \D).

\end{itemize}

With an understanding of argument polymorphism in L1, we wonder: {\it how
  many argument positions are polymorphic in L1?}  95.3\% of arguments are
monomorphic.  In spite of drastically increasing the precision of our types,
we have actually {\it reduced} the amount of polymorphism.  This suggests
that with a good notion of subtyping, precise type annotations will not cause
an undue increase in the polymorphism of the language.

Before presenting L2, we will digress slightly and explore some of the ramifications of our transition from L0 to L1.

%
%
\subsubsection{Scalars}\label{sec:scalars}

It seems that there is a place for a scalar annotation in R.  We saw in this
section that strictly scalar values appear quite frequently, making up some
of the most frequently occurring argument signatures.  In fact,
\PERCSCALARMONO of arguments have a strictly scalar monomorphic signature.
This suggests that such a distinction would more accurately capture current
use of the language, and would also be beneficial to future implementations
which can perform optimizations based on the type (e.g., not vectorizing
scalar-only values).

%
%
\subsubsection{Object-Oriented Programming in R}
\label{sec:S3S4R5}

There is something more to be said about the \l type, namely that it is
sometimes linked to (one of) R's takes on object-oriented programming.  R
has three object systems: S3, S4, and R5, and S3 objects will often manifest
themselves with this signatures in L1.  An S3 object is any value with a
``class'' attribute, and indeed many of these tend to be lists.  For
example, an S3-style ``Point'' might be a list with an entry for each
dimension in the point, with a ``class'' attribute set to ``Point''.  When
these S3 objects have contents with a variety of types, their L1 type will
appear as \l, which leaves something to be desired: Our next type system,
L2, will remedy this.

%
%
\subsubsection{\C, \I, \D Polymorphism}
Another significant source of apparently strange polymorphism is the
character, integer, double polymorphism.  Figure~\ref{fig:chardbl} lays out
an example of this polymorphic signature, showing that lists and vectors can
be accessed by strings as well as numerics.  One could conceive of an
``index'' type to capture this, but there are certain limitations to such a
type, as such an ``index'' cannot be incremented as it could be a string.
An even more sophisticated dependent type system could capture this, but
it is far from clear if such a sophisticated type system would be for a
language like R.

One other possible explanation for this polymorphism is the ``date'' data type.
In R, dates are often represented as either strings or numbers, and they are widely used.
Dates (sometimes) manifest themselves in the class of a value, for examples refer to Figure~\ref{fig:date}.
\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
class(date()) # => character
class(Sys.Date()) # => Date
class(as.POSIXlt(Sys.Date())) # => POSIXlt, POSIXt
as.integer(as.Date(1, origin = "1900-01-01"))
\end{lstlisting}}\caption{Dates in R.}\label{fig:date}\end{figure}

Unfortunately, the format of dates in R is not standardized.
That said, translating between date representations is straightforward.
As most dates are classes, L2 will be able to capture their use.

Throughout this section, we have seen that even though we substantially
increased the precision of our type annotations, we have not increased the
amount of polymorphism that we observe.  The transition from L0 to L1
revealed some interesting patterns that L1 cannot quite capture---for these,
we will need to appeal to the class and attributes of values, which are
incorporated into {\bf L2}, discussed next.

%
%
%
%
\subsection{L2}

{\bf L2} works class and attribute information into its type signatures.
As per the discussion in Section~\ref{sec:L2def}, we omit builtin classes in order to highlight those which are user-defined.
{\bf L2} differs from {\bf L1} in the following ways:

\begin{enumerate}

	\item where {\bf L1} and {\bf L0} only considered classes and attributes when ascribing matrix or data frame types, {\bf L2} considers classes and attributes to be an integral part of the type system.
	Annotations in {\bf L2} include classes and attribute names when they are present.

	\item recall that subtyping between two {\bf L2} types is defined as

\end{enumerate}

As in {\bf L0} and {\bf L1}, we will begin by showing a breakdown of the top argument signatures in Table~\ref{tab:L2top10arg}.

\begin{table}[ht]
\centering
\begin{tabular}{lrl}
  \hline
Signature & Count & \% of All Signatures \\ 
  \hline
 \sD & 172K & 19\% \\ 
  \sL & 77K & 8.5\% \\ 
  \D & 77K & 8.4\% \\ 
  \sC & 73K & 8\% \\ 
  \sN & 50K & 5.5\% \\ 
  \sF & 45K & 5\% \\ 
  \ANY & 31K & 3.4\% \\ 
  \attrclass{\D}{}{dim} & 25K & 2.6\% \\ 
  \C & 25K & 2.6\% \\ 
   \attrclass{\l}{}{names, row.names}  & 21K & 2.2\% \\ 
   \hline
\end{tabular}
\caption{Most frequent argument signatures in {\bf L2}.}
\label{tab:L2top10arg}
\end{table}

As you can see, the most frequent signatures in {\bf L1} are represented here.
The inclusion of class and attribute information has changed the counts slightly.
Not pictures here, but interesting: named lists are more frequent than their unnamed counterparts.
No classes are represented here since (1) we eliminated builtin classes, as we've used them to construct types already, and (2) most other classes are specific to some package or set of packages---it is doubtful that such a class would be so pervasive as to feature in this breakdown.

Now that we have a sense for argument-level type signatures in {\bf L2}, we will see just how many functions are polymorphic.
Consider Figure~\ref{fig:L2funcounts}.

\begin{comment}

\begin{figure}[htbp]\begin{center}
\includegraphics[width=.9\textwidth]{L2_by_fun}
\caption{Record of Function Polymorphism in {\bf L2}. \LTWOPERCPOLY of functions are polymorphic.}
\label{fig:L2funcounts}\end{center}
\end{figure}

\end{comment}

Not pictured here are two outlier values: the summary and print functions of the base R package.
Summary is a generic function which is used to summarize various types of data, and print is used to convert values to string format and print them to the console.
These functions have over 1,000 unique signatures each, which is within reason as they are really operating over any type.
They feature so prominently in {\bf L2} (and not in {\bf L1} or {\bf L0}) since many of the values which flowed into them had some amount of attribute or class information, and indeed these functions are very frequently exercised by other packages which may implement package-specific classes.

Overall, the amount of polymorphism only slightly increased, even though we added two new dimensions of type information.
Again, far richer types do not increase the amount of polymorphism by significant amounts, perhaps a testament to the monomorphic nature of many R functions.

Having built a sense for how polymorphic whole-function signatures are in {\bf L2}, we will turn our attention to polymorphic arguments.
Given the added class and attribute information, {\it what do the more frequent polymorphic argument signatures look like?}
Table~\ref{tab:toppolyL2} has the breakdown.

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Apr  5 09:50:59 2019
\begin{table}[ht]
\centering
\begin{tabular}{lrl}
  \hline
Signature & Count & \% of All Signatures \\ 
  \hline
  \D, \attrclass{\D}{}{dim} & 1942 & 0.2\% \\ 
  \sC, \sD & 1431 & 0.2\% \\ 
  \attrclass{\l}{}{names}, \attrclass{\D}{}{names}  & 1046 & 0.1\% \\ 
  \I, \sD & 855 & 0.1\% \\ 
  \D, \attrclass{\D}{}{dim, dimnames} & 810 & 0\% \\ 
  \attrclass{\l}{}{names, row.names}, \attrclass{\D}{}{dim, dimnames} & 708 & 0.1\% \\ 
  \C, \D & 672 & 0\% \\ 
   \attrclass{\D}{}{names}, \attrclass{\D}{}{dim, dimnames} & 591 & 0\% \\ 
  \attrclass{\D}{}{dim}, \attrclass{\I}{}{dim}  & 445 & 0\% \\ 
  \sD, \attrclass{\sF}{}{srcref} & 445 & 0\% \\ 
   \hline
\end{tabular}
\caption{Top polymorphic argument signatures in {\bf L2}.}
\label{tab:toppolyL2}
\end{table}

When comparing this table with Table~\ref{tab:toppolyL1}, we see that by-and-large the same types are represented.
We note the {\tt names} attributes on the lists in row 3: these named lists are, in general, more common than their unnamed counterparts.
In the 10th signature, we see a function with attribute {\tt srcref}: functions with this attribute are for dealing with open files in the context of the srcfile set of functions in the base package.
As for why such a function would share an argument position with scalar double, there are other representations of open files, and numbers feature among them.

Now that we have seen the most frequently appearing argument signatures, we ask: {\it how many argument positions are polymorphic?}.
Table~\ref{tab:argcountsL2} has the answer.

\begin{table}[ht]
\centering
\begin{tabular}{lll}
  \hline
Signature Type & Count & Percentage \\ 
  \hline
  Monomorphic & 851K & 93.9\% \\ 
  Total Seen & 906K & --- \\ 
   \hline
\end{tabular}
\caption{Account of {\it argument} polymorphism in {\bf L2}.}
\label{tab:argcountsL2}
\end{table}

Surprisingly, even with the added precision of classes and attributes, there is still a relatively small amount of argument-level polymorphism in {\bf L2}.
Our notion of subtyping on classes is rather weak, and one extension would be to include class hierarchies in the subtyping of classes;
this would further reduce the amount of polymorphism in this type system.

%
%
%
%
\subsection{Notable Patterns and Observations}

Besides exploring the patterns highlighted by our various type systems, we can address some interesting and tangentially related questions about the usage of types in R.

%
%
%
%
\subsubsection{R's Most Polymorphic Functions}

Given the data that we've collected, it is natural to wonder what the most polymorphic packages are.
Indeed, we have shed some light on some of the most polymorphic {\it functions}, but what about the packages that these functions belong to?
What is their purpose, and why are they so polymorphic?
Figure~\ref{tab:packagepolysize} has some answers.
Note that we are in the context of {\bf L0}, here, to highlight current language patterns.

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Thu Mar 28 18:15:26 2019
\begin{table}[ht]
\centering
\begin{tabular}{lrrr}
  \hline
  Package & \# of Functions &\# of Signatures & Description \\ 
  \hline
  magrittr & 20 & 225 & Tidyverse pipe suite \\ 
  oshka & 4 & 38 & quoted expression expansion \\ 
  scatterD3 & 1 & 9 & create D3 (JS) scatter plots from R \\ 
  yaml & 4 & 33 & convert R to YAML \\ 
  knitLatex & 4 & 32 & \LaTeX \xspace generator from R data \\ 
   \hline
\end{tabular}
\caption{5 of the most polymorphic packages (in {\bf L0}).}
\label{tab:packagepolysize}
\end{table}

Now, what makes these so polymorphic?
The first package, magrittr, is one that we discussed earlier in the context of its pipe \%$>$\% function. 
It is not surprising that the package as a whole is quite polymorphic, as it implements {\it several} different pipe functions, e.g. \%$<>$\% which pipes as normal and then assigns to the left-hand argument.
In fact, all but knitLatex implement some form of meta-programming:
the oshka package meta-programs in the same style as Tidyverse, the scatterD3 package meta-programs on plot objects, and the yaml package converts R expressions to YAML expressions.
This is a trend we have observed: {\it the most polymorphic functions in R are meta-programming their arguments}.

It is highly improbable that a static type system designed for an untyped language will be able to capture all of the language's existing code (without resorting to an \ANY type or abusing unions).
It is also unclear how useful a type system would be if it {\it did} capture everything.
Indeed, many of the packages discussed here are really designed with an \ANY-typed argument in mind (e.g., left-hand side of \%$>$\%), and restricting these functions should not be the goal of any static type system designed for a language after it has matured.

%
%
%
%
\subsubsection{Metaprogramming and Unevaluated Arguments}

Our analysis uncovered a non-negligible amount of argument positions which were {\it never} evaluated across all observed function calls.
This is indicative of a desire to meta-program arguments and otherwise deconstruct their AST, as the \%$>$\% function does.
This is an incredibly powerful language feature, and may well be at the source of R's popularity, if in a roundabout way.
Tidyverse, by far the most popular package suite in R, owes some of its success to the new syntax defined by some of the included packages, and it is this ability to meta-program arguments which allowed Tidyverse to define its unique syntax.
An example of that syntax is \%$>$\%, and indeed this function was highly exercised by over 100 other packages.

%
%  
%
%
%
%
\section{Discussion}

Throughout Section~\ref{sec:results}, we have thoroughly explored the trends and patterns of polymorphism observed in our analysis of a large subset of existing R code.
By now, we have an understanding of how each of our type levels captures the polymorphic patterns that exist in R.

In this section, we aim to comment globally on the things we learned during our analysis, and propose a type system for R which is supported by empirical evidence that it captures the patterns of existing R code.
We will first look at function level polymorphism before moving on to argument-level polymorphism.

%
%
%
%
\subsection{Function Polymorphism}

There are some function-level polymorphic patterns that cannot be captures solely with polymorphic function arguments.
For instance, consider Figure~\ref{fig:funanno}.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
iden <- function(x) x;

# observed calls:
f(2); # x: double -> double
f(list(1, 2, 3)) # x: list<double> -> list<double>

# possible annotations:
# [1] x: {double, list<double>} -> {double, list<double>}
# [2] x: {double -> double, list<double> -> list<double>}
\end{lstlisting}}\caption{Simple identity function.}\label{fig:funanno}\end{figure}

Sophisticated dependent type signatures such as these could benefit language implementations in a myriad of ways.
For instance, a language implementation could perform type-based optimization in the context of a function return simply by knowing the function argument type.
In Figure~\ref{fig:funanno}, we can assign the function a more sophisticated signature ({\tt [2]}) if we do not simply collapse the observed argument signatures element-wise ({\tt [1]}).
This gets more interesting with attributes, as we might want to annotate a function to communicate that it adds an attribute to whatever the list of attributes of its input value.

A clear drawback to function-level polymorphism is that functions with many arguments tend to be sensitive to small changes in argument types, which inflates the number of unique signatures observed.
Even so, overall the number of polymorphic functions did not increase by too large a margin when adding and specializing types (from {\bf L0} to {\bf L1}, and from {\bf L1} to {\bf L2})---for the most part, monomorphic functions in {\bf L0} stayed monomorphic throughout.

One tidbit about function signatures that we have not thoroughly discussed is the {\tt ...} function argument.
Functions with such an argument can be passed an arbitrary number of extra arguments, see Figure~\ref{fig:dotdotdot} for a small example.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
f <- function(x, ...) {
  g(x, ...)
}

f(42, z=2, y=4) # in f, will perform g(42, z=2, y=4)

\end{lstlisting}}\caption{Simple ... example.}\label{fig:dotdotdot}\end{figure}

Our analysis ignores these extra arguments, instead it simply flags functions which have a {\tt ...} in their definition.
If building an annotation for such a function, the {\tt ...} argument is always ascribed the \l type, as it can indeed be inhabited by anything.

Ultimately, to determine which kinds of whole function signatures would be useful, a more sophisticated analysis is required, which would need to develop a correlation between function arguments and return types.
There may also be other interesting per-function trends which our current analysis does not pick up on that are hidden in our data.

%
%
%
%
\subsection{Argument-Level Polymorphism}

From our analysis, we discovered a number of interesting patterns at the argument level.
To summarize:

\begin{itemize}

	\item {\bf scalars}: according to our analysis, many instances of vectorized values are always scalar.
	This is strong evidence that {\it an annotation denoting a scalar value would be useful to R programmers}, and further such an annotation could be used by future implementations to statically determine when not to vectorize a vectorizable type.

	\item {\bf missing}: our analysis has found a number of instances of argument positions remaining unevaluated across all calls to a function.
	{\it A small annotation, such as a tag on an argument position, would capture this behavior} and could be used to warn programmers if they try to use the unevaluated value, and also warn them if an operation might evaluate it accidentally.
	Techniques to avoid forcing promises in R are not well understood by most programmers, and such an annotation would be useful to assist them in avoiding unwanted forcing of promises.

	\item {\bf classes}: in transitioning from {\bf L1} to {\bf L2}, we found that the prevalence of \l, \lT{X} polymorphism decreased significantly.
	In many cases, these lists share a class, and so the signature which was prominent in {\bf L1} was broken down into several less prominent {\bf L2} signatures with added class information.
	Classes could also be used to capture the usage of dates (two of the standard date formats have class ``Date'' and ``POSIXct, POSIXt''), and could help package writers enforce usage of package functions.
	R's dynamic dispatch system is class-based, so including annotations for these would help programmers determine which functions are being dispatched for which values.

\end{itemize}

Generating the code to dynamically enforce these signatures is also possible.
Simple checks include checking the class of a value, or checking for the presence of certain attributes.
R makes it easy to exit a function prematurely, and in fact many functions already exhibit this behavior.
For example, see Figure~\ref{fig:stopifnot}.
In it, the function exits if the {\tt encoding} argument is provided, and is not a character.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
is_character <- function(x, n = NULL, encoding = NULL) {
  ...
  if (!is_null(encoding)) {
    stopifnot(typeof(encoding) == "character")
    ...
  }
  ...
}
\end{lstlisting}}\caption{Real example of premature function exit.}\label{fig:stopifnot}\end{figure}

We also saw that, despite being far more specific and descriptive than previous levels, the {\bf L2} type system did not bring about an undue increase in the amount of overall polymorphism in the language.
This signifies that {\bf L2} captures the vast majority of already existing polymorphic patterns, and very few functions would need to resort to an \ANY annotation, and indeed many arguments are really \ANY-typed (e.g., \%$>$\% from magrittr and the prototype assignment operator {\tt \$<-} from proto).

Based on our analysis of the data, we conclude that {\it a type system such as} \textbf{\emph{L2}} {\it captures the vast majority of argument and function usage in the language}.
{\bf L2} also contains a wealth of useful information about values that could be used for optimizations and IDE tooling.

%
%
%
%
%\subsection{Coverage}
%
%\AT{Manual check of some functions.}

%
%
%
%
\section{Related Work}

The idea of measuring polymorphism via some program analysis is not new.
For instance, some work~\cite{aakerblom2015measuring} measures the polymorphism of Python programs, but with a slightly different goal than ours: authors seek to assess the need for expressiveness in type systems that are retrofit onto dynamic languages.
Interestingly, they draw some similar conclusions to us: they find that monomorphism is far more frequent than polymorphism, and even find that typically very few functions are very highly polymorphic.

More generally, there is an existing literature on analyzing usage patterns of language features.  
For instance, the dynamic features of Smalltalk were analyzed in some work~\cite{callau2011howdevelopers}, which aimed to see how often the highly dynamic capabilities of Smalltalk were used and in which kinds of projects.
Other work~\cite{pldi10a} explores how the dynamic features of JavaScript are used.
An interesting finding of theirs is that only 81\% of calls are monomorphic, which is very similar to what we observed in R.
Other analysis work~\cite{milojkovic2017duck} looks into the usage patterns of duck typing.

The type system we propose in this work could be incorporated into R as a {\it gradual type system}.
Gradual type systems~\cite{SiekTaha06} aim to bridge the gap between dynamic and static typing by allowing programmers to gradually add type annotations to their programs.
Gradually-typed languages are becoming increasingly popular, with languages such as TypeScript~\cite{typescript13} becoming more mainstream.
An eventual implementation of a type system for R will surely be inspired by this work.

%
%.
%
%
%
%
\section{Conclusion}

Truly understanding how programmers use a programming language is critical
to building usable and effective language features.  When working on a
general purpose languages (such as JavaScript), we as programmers can appeal
to our own intuitions, as many of us have had at least some interaction with
said languages.  This intuition does not extend to domain-specific languages
(DSLs), which are not really designed for members of our community.

The R programming language is an example of this.  R was developed to foster
an easy and interactive programming experience. For this reason R is very
flexible and expressive.  Unfortunately, these liberties come at a cost:
debugging R code is challenging. Further, performance quickly becomes
prohibitive if programmers are not careful.

Types can make a language easier to debug and run faster.  For debugging,
some level of static type enforcement would give programmers more direct
feedback.  As for performance, type information enables optimizations. If we
are interested in developing an easily adoptable type system for R, we must
ensure that it captures how existing code is being used.

Throughout this paper, we have built up an understanding of types and
polymorphism in R by analyzing the usage of types across a large corpus of
code.  We highlighted some patterns of type usage, and developed a number of
type systems with differing levels of granularity. We analyzed the tests,
examples, and vignettes of over ten thousand packages on CRAN (over
\PERCENTCRAN of available packages), and traced over 15 million lines of R
code.

We believe that a type system akin to {\bf L2} is well suited for R, as it
is expressive enough to capture current use of the language without being so
specific as to make everything polymorphic.  In fact, the levels of
polymorphism observed were rather low overall, with only 6\% of function
arguments and 21\% of functions being polymorphic.  Importantly, {\bf L2}
carries enough information to be useful to R implementations and IDEs:
language implementations would surely benefit from statically-available type
information, and IDEs could use the rich type and attribute information to
give programmers immediate feedback during the programming process.

%
%
%
\bibliographystyle{boilerplate/ACM-Reference-Format}
\bibliography{bib/biblio,bib/jv,bib/r,bib/new,bib/gradual}
\end{document}

### 2020-01-24 paper/main.tex
2020/01/24 update, cleaned old data, added some readmes, and some new data

@@
 
-Now, in order to {\it validate} this type system, we implemented a framework for annotating R programs, as well as a tool which infers types from package example code, inserts lightweight runtime contracts into the code, and runs the example code in the package's reverse dependencies to try to unearth type errors.
+To make our type system concrete, we developed a tool to infer the types of package functions, and decorate package code with the appropriate annotations.
+This tool is also able to insert lightweight runtime contracts into the code to ensure dynamic adherence to the annotated type.
 The annotations are lightweight as they are only run {\it the first time} an argument is used, if it is used at all.
+Our tool allows programmers to quickly generate annotations for their functions from their own existing tests, and edit them as they see fit.
+
+% Now, in order to {\it validate} this type system, we implemented a framework for annotating R programs, as well as a tool which infers types from package example code, inserts lightweight runtime contracts into the code, and runs the example code in the package's reverse dependencies to try to unearth type errors.
+To help validate our tools and type system, we perform an experiment wherein we analyze package test, example, and vignette code, generate annotations (and contracts) based on the inferred types, and then run the tests, examples, and vignettes of the package's reverse dependencies to try to unearth type errors.
 Details on and the results of the validation are available in Section~\ref{sec:results}.
@@
 
-\AT{Are we going to change this? To make it run off of R-dyntrace?}
+\AT{TODO change this section to use the Typr approach?}
 
-The first tool in our pipeline is Typr.
-The goal of Typr is to gather type information on function invocations, and it is a modification of the Genthat~\cite{issta18} tool.
-Genthat records all function invocations and builds a standalone trace to allow for independent reproduction of the call. 
-We modified Genthat to instead record a wealth of information on argument and return values.
+The first tool in our pipeline is Typr, a tool designed to record the types of function arguments and return values.
+Typr is implemented in the R-dyntrace framework.
 Typr records:
@@
 \begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-f <- function(x, y) {
+f <- function(x) {
   z <- x
@@
 
-f(0, 42)
+f(0)
 \end{lstlisting}}\caption{Propagatr example.}\label{fig:propagatr-example}\end{figure}
@@
 
+\AT{Description of how the constraints look.}
+Once all of the constraints are generated, we use the {\tt ifscs} tool written in Haskell; it is a tool built to solve set constraints (or inclusion constraints).
+Consider again the program in Figure~\ref{fig:propagatr-example}.
+Propagatr will observe the following flows:
+
+\begin{itemize}
+
+\item $f_{0}$: $z$, $g_{1}$
+\item $f_{R}$: $g_{R}$
+\item $z$: $g_0$ \AT{we can phrase this as a return constraint?}
+
+\end{itemize}
+
+We will want to generate the following constraints:
+
+\begin{itemize}
+
+\item $f_{0} \subseteq z$
+\item $f_{0} \subseteq g_{1}$
+\item $f_{R} \supseteq g_{R}$
+\item $z \supseteq g_0$ \AT{we can phrase this as a return constraint?}
+
+\end{itemize}
+
+\AT{CHECK co/contra...} We gave covariant constraints for function arguments, and contravariant constraints for function returns.
+That is to say, function argument types are subsets of their usage in the function body, while (the true type of) a function return is a superset of the types of all possible returns in the function. 
+In order to work the dynamic analysis results into the constraint system, we add the following atomic constraints:
+\AT{maybe we want a slightly more complicated example...}
+
+\begin{itemize}
+\item \{integer\} $\subseteq f_{0}$
+\end{itemize}
+
+Intuitively, we create a set of constraints which state that the signatures obtained by the dynamic analysis are a subset of the true signatures.
+Crucially, we take the dynamic analysis results for the {\it base} package to be a ground truth; we have faith in these signatures as the base package is incredibly highly exercised, with over \AT{X} calls to base package functions.
+
 %
@@
 
-Our Infr tool leverages our type system to test the effectiveness of package test, example, and vignette code at covering the possible use-cases of package functions.
+\AT{Might be worthwhile to discuss how we implement basically a ``contract type system'', and the Infr tool consolidates Typr and Propagatr to generate the runtime contracts.}
+\AT{We can expand this section to be more broadly useful to R programmers, but it depends on how we want to pitch this.}
+
+% Our Infr tool leverages our type system to test the effectiveness of package test, example, and vignette code at covering the possible use-cases of package functions.
+Our Infr tool consolidates Typr and Propagatr into a tool that R package writers can use to both infer types for their functions and generate lightweight contracts to dynamically ensure adherence to function type signatures.
 
@@
 These contracts are inserted directly into the promise code of the values, as will be elaborated in Section~\ref{subsubsec:promise}.
-\item Then, the test, example, and vignette code for all reverse package dependencies (i.e., for all other packages which have the current package as a dependency) is run, and any type errors that are caught are recorded.
+% \item Then, the test, example, and vignette code for all reverse package dependencies (i.e., for all other packages which have the current package as a dependency) is run, and any type errors that are caught are recorded.
 \end{itemize}
 
-\AT{Do we want to do something special about data frames? We sort of already infer their type from the dynamic analysis.}
+\AT{...}
+We additionally devised an experiment to test our type system and simultaneously test the ``type coverage'' of R package example code.
+In our experiment, we run Infr on a package's tests, examples, and vignettes, and generate annotations and contracts for package functions based on that.
+Then, we run the tests, examples, and vignettes of {\it all package clients}, to see if the clients of the package adhered to the inferred type specification.
 
@@
 \AT{What should go in here?}
+\AT{We have to give the type system somewhere, also!}
+
+%
+%
+%
+%
+\subsection{The Data}
+\label{sec:results_data}
 
@@
 %
+\subsection{The Experiment}
+\label{sec:results_experiment}
+
+\AT{We should also talk about the experiment itself, and the results we find here.}
+
+%
+%
+%
+%
 %
@@
 \section{Related Work}
-
-\AT{Look at the stuff we put in the old version of the paper.}
 

### 2020-01-24 wip_paper/main.tex
2020/01/24 update, cleaned old data, added some readmes, and some new data

\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\usepackage{booktabs,listings,xspace,wrapfig}
\lstset{language=R}
\usepackage{my_style}
\definecolor{LightGray}{rgb}{.95,.95,.95}
\definecolor{Gray}{rgb}{.3,.3,.3}
\definecolor{DarkGray}{rgb}{.5,.5,.5}

\graphicspath{ {./plots/} }

\lstset{ %
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
  framexleftmargin=-1mm,
  framexrightmargin=-1mm,
  tabsize=2,
  belowskip=0pt,
  basicstyle=\small\ttfamily,
  backgroundcolor=\color{LightGray},
  emphstyle=\sffamily,
  keywordstyle=\bfseries,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray},
 % numbers=left
}

\lstdefinestyle{R}{ %
  language=R,
  deletekeywords={env, equal, c, runif, trace, args},
  breaklines=true
}

\lstdefinestyle{Rin}{ %
  style=R,
  numberstyle=none,
  basicstyle=\normalsize\ttfamily,
  breaklines=false
}

\newcommand{\code}[1]{\lstinline|#1|\xspace}
\newcommand{\genthat}{{\sc Genthat}\xspace}

% Macros for type names?
\newcommand{\reals}{$\mathtt {\rm I\!R}$\xspace}
\newcommand{\realv}{$\overline{\rm I\!R}$\xspace}

\setcopyright{none}
%\setcopyright{acmcopyright}%\setcopyright{acmlicensed}
%\acmDOI{10.475/123_4}
%\acmConference[OOPSLAs]{Woodstock conference}{July 1997}{El Paso, Texas USA}
%\acmYear{1997}%\copyrightyear{2016}%\acmPrice{15.00}
\begin{document}

\title{Towards a Type System for R}

\begin{abstract}

R is a crazy language.
\AT{Briefly describe why R is crazy.}
In spite of this, R is widely used.
We would like to build a full-fledged type system for the language {\it that programmers will want to use}.
Not only would this would serve to increase the assurance of R code, it would also open the door to type-based program optimizations.

\end{abstract}

\maketitle

%
%
%
%
\section{Introduction}

%
%
%
%
\section{The R Programming Language}

R~\cite{R_man} is ... .

%
%
%
%
\section{Why Types?}

We want to add types to R for two main reasons.

First, types increase the assurance that programmers have in their code.
If using a static type system, type-related errors can be caught statically, which communicates potential bugs to programmers at compile time.
If using a dynamic type system, type-related errors are caught at runtime, which may well abort the program, but at least does not allow for values to be quietly misused.

When we increase assurance in a programming language, we thereby increase assurance in the products of that language.
For instance, one might have more faith in a web server written in TypeScript than one written in JavaScript, since TypeScript at least catches static type errors in annotated code.
So what are the products of R?
R is a language primarily used by statisticians and data scientists to perform data analysis. 
The conclusions drawn from these analyses go on to inform decisions in broader organizations:
For instance, census data analysis results go on to inform policymaking decisions.
It is critical that we can trust the software used to draw these conclusions.

Second, type information can be used by just-in-time compilers (JIT compilers, or JITs) to perform program optimizations at runtime.
\AT{This one goes without saying, not sure how much to elaborate here.}

%
%
%
%
\section{Concerns}

Designing a type system for R is no small task.

%
%
\subsection{Type System Granularity and Complexity}

When designing a type system for an existing, dynamic language, we have the freedom to decide what granularity of type information we want to express.
In data-oriented languages, such as R, choosing the level of granularity is not so straightforward.

For instance, consider vectors in R.
Primitive types in R are vectorized, meaning that even scalars are actually vectors (e.g., the scalar 1 is implemented as a unit-length vector).
That said, in many other languages, a distinction is made between scalars and vectors, and this distinction is coupled with some performance benefit.
When designing a type system for R, we might be interested in distinguishing vectors and scalars.

Going a step further, perhaps it would be valuable to include the dimensions of data in a vector type.
R already implements different functionality for operators acting on vectors depending on the lengths of the vectors.
For example, consider the following code snippet:

\begin{lstlisting}
c(1, 2) + c(1, 2) # => c(2, 4)
c(1, 2, 3) + c(1, 2) # => c(2, 4, 4)
c(1, 2, 3, 4) + c(1, 2). # => c(2, 4, 4, 6)
\end{lstlisting}

On vectorized primitives, the $+$ operator will duplicate the second argument until it matches the length of the first.
The 3rd line above illustrates this clearly: the second argument to $+$ there becomes {\tt c(1, 2, 1, 2)}.
This is useful functionality, and indeed users of R are familiar with it, but it does introduce possible sources of error if used incorrectly.
For instance, imagine if we were working with some data, and we wanted to normalize the data w.r.t. a known vector of values.
Consider:

\begin{lstlisting}
norm <- function(data) {
  data / c(1, 2, 4)
}

norm( c( 5, 4, 8)) # => c(5, 2, 2)
norm( c( 5, 2)) # => c(5, 1, 0.25)
\end{lstlisting}

In the first call to {\tt norm}, we see that the argument to {\tt norm} was divided element-wise by the vector defined in the function itself.
In the second call, we see how R automatically pads vectors when the lengths don't quite match: here, {\tt c(5, 2)} essentially becomes {\tt c(5, 2, 1)}.
In general, this is rather harmless, but when we consider R's use case as a data analysis tool, this represents {\it generating arbitrary data}: 
The value inserted by the division operator is really a data point fabricated by the implementation of the function.
In a sense, off-by-one errors can easily go unnoticed in R programs.

All that said, adding lengths to types may not be desirable.
The complexity of the type system and associated annotations would be far greater, and there's no telling if users would subscribe to the restrictions imposed by e.g. statically encoding lengths.
Many of these errors could be caught by runtime checks, for instance with an explicit length check in the {\tt norm} function above, but it is as easy to write those checks as it is to forget to put them.
Further, designing annotations to allow users to express types in the language is complicated by the lexical complexity of R itself; in short, there is not a lot of wiggle room in the parser.
It is critical that we balance complexity and granularity with usability.
This is discussed next.

%
%
\subsection{Ensuring User Engagement}

If one of our goals is to increase our confidence in R programs, we need to ensure that language users will want to use our types.

R users represent an interesting point in this design space.
R is an old language, and R users are not your average programmers in the sense that they are likely not software engineers, developers, or computer scientists.
The best way to understand how they interact with the language is through observation and analysis of the programs that they have written.
But even that is no small feat, as R programs are not published the same way that the products of many other languages are on e.g. GitHub.
This last point will be discussed further in Section~\ref{sec:practicalTS}.

Ultimately, we need to balance the complexity of the type system with needs of R programmers.
Perhaps a dependent type system would be useful and capable of capturing large swaths of existing R paradigms, but these types are difficult to express and may be too burdensome for R users. 
On the other hand, differentiating scalars and vectors may not reflect existing R implementations, but perhaps R users would be able to quickly adapt to such a distinction.
The key to adoption is to strike this balance, and conceive of a type system which captures everything that R programmers would want to capture. 
We have made some progress towards this, discussed further in Section~\ref{sec:plan}.

%
%
\subsection{Types for Big Data Objects}

A unique aspect of R is the data frame, which serves as the cornerstone of nearly all analyses written in R.
As we mentioned, the process of a typical R program is to load data, modify and/or transform it, and perform an analysis.
Typically, imported data is a {\it data frame}: an R data type which is effectively a list of lists, where each row represents an observation, and each column represented a thing that was observed. 
These objects underpin data analysis in R, and understanding the shape of your data is critical to successful data analysis.

As data frames are of utmost importance, it stands to reason that we might like to create some type for them.
How specific that type should be, however, is unclear: data frames are often quite large, and writing a type for such a thing would be tedious and perhaps not altogether informative if the data frame is used in simple ways.
If the only operation performed on the data frame is to normalize some column and plot it, then types are not supremely helpful.
But what if, during the modification and transformation phase of a data pipeline, the analyst is working with several data frames and joining and collating them to create new data frames?
Many of these operations are dependent on the types of some columns, and could fail in various ways if the column types are not compatible.
These errors would be mitigated by (relatively) sophisticated data frame types.

%
%
\subsection{The Burden of Annotations}

A nontrivial barrier to building types into R is the need for annotations.
For one, the syntax of the language is tricky to extend, as many characters already serve a purpose.
This makes designing a satisfying syntax for types nontrivial.
Further, as the type system becomes more complex, so too do the annotations:
R is rich in runtime type information already, and statically encoding that via annotations can quickly become burdensome.
For instance, if we would like to express that an argument to a function is a list, with some class, and certain attributes, we have over 3 things to annotate onto a single function argument.
Ultimately, friendly-looking annotations will be attractive to users, and making the annotations easy to write and minimizing the burden of annotating is critical to adoption.

%
%
%
%
\section{Our Plan}
\label{sec:plan}

%
%
\subsection{A Practical Type System}
\label{sec:practicalTS}

End-user R code is typically some small script (called a {\it notebook}) which imports some data and modifies it in some way before performing some analysis.
In many cases, the data being fed to the script is proprietary (e.g. customer data) or confidential (e.g. census data), which results in an inability to publish the script, even if the analyst were so inclined.
In sum, getting our hands on R user code can be tricky.

Thankfully, all is not lost: the Comprehensive R Archive Network (CRAN) is a repository of publicly available R package (i.e. library) code.
Packages made available through CRAN must adhere to some criteria; among them, a guarantee that the package must be accompanied by some examples, tests, and/or vignettes to showcase package functionality.
These examples are about as close as we can come to end-user R code, and serve as a great starting point for any analysis we undertake.

%
%
%
%
\section{Related Work}

There is a wealth of literature on analyzing language usage patterns.
For instance, \ref{}

%
%
%
\bibliographystyle{boilerplate/ACM-Reference-Format}
\bibliography{bib/biblio,bib/jv,bib/r,bib/new,bib/gradual}
\end{document}

### 2020-01-24 data/genthat_extracted_code/AutoDeskR/examples/getToken.Rd.R

@@ -1,18 +1,0 @@
-library(AutoDeskR)
-
-
-### Name: getToken
-### Title: Get a 2-Legged Token for Authentication.
-### Aliases: getToken
-
-### ** Examples
-
-## Not run: 
-##D # Get a 2-legged token with the "data:read" and "data:write" scopes
-##D resp <- getToken(id = Sys.getenv("client_id"), secret = Sys.getenv("client_secret"),
-##D            scope = "data:write data:read")
-##D myToken <- resp$content$access_token
-## End(Not run)
-
-
-

### 2020-01-24 data/genthat_extracted_code/BaylorEdPsych/examples/LittleMCAR.Rd.R

@@ -1,15 +1,0 @@
-library(BaylorEdPsych)
-
-
-### Name: LittleMCAR
-### Title: Little's missing completely at random (MCAR) test
-### Aliases: LittleMCAR
-### Keywords: Missing Little's Chi-Square Missing Completely At Random MCAR
-
-### ** Examples
-
-data(EndersTable1_1)
-LittleMCAR(EndersTable1_1)
-
-
-

### 2020-01-24 data/genthat_extracted_code/BoolFilter/examples/BKF.Rd.R

@@ -1,26 +1,0 @@
-library(BoolFilter)
-
-
-### Name: BKF
-### Title: Boolean Kalman Filter
-### Aliases: BKF
-
-### ** Examples
-
-data(p53net_DNAdsb0)
-
-obsModel = list(type = 'NB', 
-                   s = 10.875, 
-                  mu = 0.01, 
-               delta = c(2, 2, 2, 2), 
-                 phi = c(3, 3, 3, 3))
-
-#Simulate a network with Negative-Binomial observation model
-data <- simulateNetwork(p53net_DNAdsb0, n.data = 100, p = 0.02, obsModel)
-          
-#Derive the optimal estimate of the network using a BKF approach
-Results <- BKF(data$Y, p53net_DNAdsb0, p = 0.02, obsModel)
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/DEVis/examples/create_dir_struct.Rd.R

@@ -1,44 +1,0 @@
-library(DEVis)
-
-
-### Name: create_dir_struct
-### Title: Initialize the directory structure for automatically storing and
-###   structuring results.
-### Aliases: create_dir_struct
-### Keywords: directory initialization organization output
-###   project-management
-
-### ** Examples
-
-## Not run: 
-##D 
-##D #Initialize a DEVis directory structure at a base directory.
-##D create_dir_struct(base_dir="/Users/adam/projects/ebola/")
-##D 
-##D This command will create the following folders:
-##D /Users/adam/projects/ebola/
-##D   /results/
-##D     /DE/
-##D       /boxplot/
-##D       /counts/
-##D       /data/
-##D       /density_plots/
-##D       /divergence/
-##D       /heatmaps/
-##D       /profile_plots/
-##D       /series_plots/
-##D       /volcano/
-##D    /dendrograms/
-##D    /geneplots/
-##D    /group_stats/
-##D    /MDS/
-##D      /standard/
-##D      /hulls/
-##D    /sample_distance/
-##D      /euclidian/
-##D      /poisson/
-##D 
-## End(Not run)
-
-
-

### 2020-01-24 data/genthat_extracted_code/DPpackage/examples/nodal.Rd.R

@@ -1,42 +1,0 @@
-library(DPpackage)
-
-
-### Name: nodal
-### Title: Nodal Involvement Data
-### Aliases: nodal
-### Keywords: datasets
-
-### ** Examples
-
-## Not run: 
-##D     # Data
-##D       data(nodal)
-##D       attach(nodal)
-##D       lacid<-log(acid)
-##D 
-##D     # Initial state
-##D       state <- NULL
-##D 
-##D     # MCMC parameters
-##D       nburn<-5000
-##D       nsave<-10000
-##D       nskip<-10
-##D       ndisplay<-100
-##D       mcmc <- list(nburn=nburn,nsave=nsave,nskip=nskip,ndisplay=ndisplay,
-##D                    tune=1.1)
-##D 
-##D 
-##D     # Prior distribution
-##D       prior <- list(alpha=1,beta0=c(0,rep(0.75,5)),
-##D                     Sbeta0=diag(c(100,rep(25,5)),6))
-##D 
-##D     # Fit the model
-##D       fit1 <- DPbinary(ssln~age+lacid+xray+size+grade,prior=prior,mcmc=mcmc,
-##D                        state=state,status=TRUE) 
-##D       fit1
-##D 
-## End(Not run)
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/DtD/examples/merton_ll.Rd.R

@@ -1,41 +1,0 @@
-library(DtD)
-
-
-### Name: merton_ll
-### Title: Compute Log-Likelihood of Merton Model
-### Aliases: merton_ll
-
-### ** Examples
-
-# we get the same if we call `optim` as follows. The former is faster and is
-# recommended
-set.seed(4648394)
-sims <- BS_sim(
-  vol = .1, mu = .05, dt = .1, V_0 = 100, T. = 1, D = rep(80, 20), r = .01)
-
-r1 <- with(
-  sims, BS_fit(S = S, D = D, T. = T, r = r, time = time, method = "mle",
-               eps = 1e-8, vol_start = .2))
-
-r2 <- optim(c(mu = 0, log_vol = log(.2)), function(par)
-  -with(
-    sims, merton_ll(S = S, D = D, T. = T, r = r, time = time,
-                    mu = par["mu"], vol = exp(par["log_vol"]))))
-
-all.equal(r1$n_iter, unname(r2$counts[1]))
-all.equal(r1$ests[1], r2$par[1])
-all.equal(r1$ests[2], exp(r2$par[2]), check.attributes = FALSE)
-
-# the log-likelihood integrates to one as it should though likely not the
-# most stable way to test this
-ll <- integrate(
-  function(x) sapply(x, function(S)
-    exp(merton_ll(
-      S = c(1, S), D = .8, T. = 3, r = .01, dt = 1/250, vol = .2,
-      mu = .05))),
-  lower = 1e-4, upper = 6)
-stopifnot(isTRUE(all.equal(ll$value, 1, tolerance = 1e-5)))
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/FixSeqMTP/examples/FSFDR.indept.p.adjust.Rd.R

@@ -1,22 +1,0 @@
-library(FixSeqMTP)
-
-
-### Name: FSFDR.indept.p.adjust
-### Title: Adjusted P-values for Fixed Sequence FDR Controlling Procedure
-###   under Independence
-### Aliases: FSFDR.indept.p.adjust
-
-### ** Examples
-
-## generate a pre-ordered pvalue vector for 50 hypotheses, where 80% are true nulls
-set.seed(1234); m <- 50; pi0 <- 0.8; m0 <- m*pi0; m1 <- m-m0
-mu <- c(4*0.9^(1:m1), rep(0,m0))
-Zstat <- rnorm(n = m, mean = mu)
-Pval <- 1-pnorm(Zstat)
-## conventional fixed sequence procedure
-FSFDR.indept.p.adjust(p = Pval, alpha = 0.05)
-## generalized fixed sequence procedure allowing stop at 5th acceptance
-FSFDR.indept.p.adjust(p = Pval, alpha = 0.05, k=5)
-
-
-

### 2020-01-24 data/genthat_extracted_code/GISTools/examples/jitter.points.Rd.R

@@ -1,15 +1,0 @@
-library(GISTools)
-
-
-### Name: Computational Inference from Point Data
-### Title: Bootstrap and Kernel Bootstrap from Points
-### Aliases: jitter.points bstrap.points
-
-### ** Examples
-
-data(newhaven)
-plot(blocks)
-for (i in 1:20) plot(jitter.points(breach,150),add=TRUE,pch=1,col='red') 
-
-
-

### 2020-01-24 data/genthat_extracted_code/GUILDS/examples/generate.Guilds.Rd.R

@@ -1,13 +1,0 @@
-library(GUILDS)
-
-
-### Name: generate.Guilds
-### Title: Generate Artificial data under the GUILDS model
-### Aliases: generate.Guilds
-
-### ** Examples
-
-generate.Guilds(theta=200,alpha_x = 0.005, alpha_y = 0.001,J=20000);
-
-
-

### 2020-01-24 data/genthat_extracted_code/GeneClusterNet/examples/GeneClusterInterp.Rd.R

@@ -1,21 +1,0 @@
-library(GeneClusterNet)
-
-
-### Name: GeneClusterInterp
-### Title: Interpolating gene expression measurements
-### Aliases: GeneClusterInterp
-
-### ** Examples
-
- # load the package 
- library(GeneClusterNet)
- set.seed(1234)
- data(mExpression)
- Sample=mExpression[sample(1:nrow(mExpression),50,replace=FALSE),]
- LOPCoefficient =GeneCluster (Sample, times=c(1:18), NumberOfCluster=3,orderLOP=5)$ LOPCoefficient
- GeneClusterInterp (LOPCoefficient, OriginalTime=c(1:18),outLen=20)
-
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/IDPmisc/examples/grid.control.Rd.R

@@ -1,14 +1,0 @@
-library(IDPmisc)
-
-
-### Name: grid.control
-### Title: Auxilary for Controlling the Grid Appearance of a Rose Plot
-### Aliases: grid.control
-### Keywords: dplot
-
-### ** Examples
-
-grid.control()
-
-
-

### 2020-01-24 data/genthat_extracted_code/LncFinder/examples/svm_tune.Rd.R

@@ -1,40 +1,0 @@
-library(LncFinder)
-
-
-### Name: svm_tune
-### Title: Parameter Tuning of SVM
-### Aliases: svm_tune
-
-### ** Examples
-
-## Not run: 
-##D data(demo_DNA.seq)
-##D Seqs <- demo_DNA.seq
-##D 
-##D positive_data <- extract_features(Seqs[1:5], label = "NonCoding",
-##D                                   SS.features = FALSE, format = "DNA",
-##D                                   frequencies.file = "human",
-##D                                   parallel.cores = 2)
-##D 
-##D negative_data <- extract_features(Seqs[6:10], label = "Coding",
-##D                                   SS.features = FALSE, format = "DNA",
-##D                                   frequencies.file = "human",
-##D                                   parallel.cores = 2)
-##D 
-##D my_dataset <- rbind(positive_data, negative_data)
-##D 
-##D ### Or use our data "demo_dataset"
-##D data(demo_dataset)
-##D my_dataset <- demo_dataset
-##D 
-##D optimal_parameter <- svm_tune(my_dataset, positive.class = "NonCoding",
-##D                               folds.num = 2, seed = 1,
-##D                               gamma.range = (2 ^ seq(-5, 0, 2)),
-##D                               cost.range = c(1, 8, 16),
-##D                               return.model = FALSE, parallel.core = 2)
-##D 
-##D ### Users can set return.model = TRUE to return the best model.
-## End(Not run)
-
-
-

### 2020-01-24 data/genthat_extracted_code/MPsychoR/examples/Bergh.Rd.R

@@ -1,15 +1,0 @@
-library(MPsychoR)
-
-
-### Name: Bergh
-### Title: Generalized Prejudice Dataset
-### Aliases: Bergh
-### Keywords: datasets
-
-### ** Examples
-
-data("Bergh")
-str(Bergh)
-
-
-

### 2020-01-24 data/genthat_extracted_code/NCA/examples/nca_analysis.Rd.R

@@ -1,71 +1,0 @@
-library(NCA)
-
-
-### Name: nca_analysis
-### Title: Run NCA analyses on a data set
-### Aliases: nca_analysis
-### Keywords: functions
-
-### ** Examples
-
-# Load the data
-data(nca.example)
-data <- nca.example
-
-# Basic NCA analysis, with independent variables in the first 2 columns
-# and the dependent variable in the third column
-model <- nca_analysis(data, c(1, 2), 3)
-
-# Use nca_output to show the summaries (see nca_output documentation for more options)
-nca_output(model)
-
-# Columns can be selected by name as well
-model <- nca_analysis(data, c('Individualism', 'Risk taking'), 'Innovation performance')
-
-# Define the ceiling techniques via the ceilings parameter, see 'ceilings' for all types
-model <- nca_analysis(data, c(1, 2), 3, ceilings=c('ce_fdh', 'ce_vrs'))
-# These are the available ceiling techniques
-print(ceilings)
-
-# For using the upper right corner(s), 'flip' the x variables
-model <- nca_analysis(data, c(1, 2), 3, flip.x=TRUE)
-# It is also possible to flip a single x variable
-model <- nca_analysis(data, c(1, 2), 3, flip.x=c(TRUE, FALSE))
-# Flip the y variable if the lower corners need analysing
-model <- nca_analysis(data, c(1, 2), 3, flip.x=c(TRUE, FALSE), flip.y=TRUE)
-
-# Use a theoretical scope instead of the (calculated) empirical scope
-model <- nca_analysis(data, c(1, 2), 3, scope=c(0, 120, 0, 240))
-
-# By default the bottleneck tables use percentages of the range for the x and y values.
-# Using the percentage of the max value is also possible
-model <- nca_analysis(data, c(1, 2), 3, bottleneck.y='percentage.max')
-# Use the actual values, in this case the x-value
-model <- nca_analysis(data, c(1, 2), 3, bottleneck.x='actual')
-# Use percentile, in this case for the y-values
-model <- nca_analysis(data, c(1, 2), 3, bottleneck.y='percentile')
-# Any combination is possible
-model <- nca_analysis(data, c(1, 2), 3, bottleneck.x='actual', bottleneck.y='percentile')
-
-# The number of steps is adjustible via the steps parameter
-model <- nca_analysis(data, c(1, 2), 3, steps=20)
-# Or via the step.size parameter, this ignores the steps parameter
[…]

### 2020-01-24 data/genthat_extracted_code/NISTunits/examples/NISTsqrMileUSsurveyTOsqrMeter.Rd.R

@@ -1,14 +1,0 @@
-library(NISTunits)
-
-
-### Name: NISTsqrMileUSsurveyTOsqrMeter
-### Title: Convert square mile 7 to square meter
-### Aliases: NISTsqrMileUSsurveyTOsqrMeter
-### Keywords: programming
-
-### ** Examples
-
-NISTsqrMileUSsurveyTOsqrMeter(10)
-
-
-

### 2020-01-24 data/genthat_extracted_code/NSM3/examples/pUmbrPU.Rd.R

@@ -1,19 +1,0 @@
-library(NSM3)
-
-
-### Name: pUmbrPU
-### Title: Mack-Wolfe Peak Unknown
-### Aliases: pUmbrPU
-### Keywords: Mack-Wolfe Peak Unknown
-
-### ** Examples
-
-##Hollander-Wolfe-Chicken Example 6.4 Learning Comprehension and Age
-wechsler<-list("16-19"=c(8.62,9.94,10.06),"20-34"=c(9.85,10.43,11.31),"35-54"=c(9.98,10.69,11.40),
-"55-69"=c(9.12,9.89,10.57),"70+"=c(4.80,9.18,9.27))
-
-#pUmbrPU(wechsler,method="Monte Carlo",n.mc=20000)
-pUmbrPU(wechsler,method="Monte Carlo",n.mc=1000)
-
-
-

### 2020-01-24 data/genthat_extracted_code/OptSig/examples/R.OLS.Rd.R

@@ -1,20 +1,0 @@
-library(OptSig)
-
-
-### Name: R.OLS
-### Title: Restricted OLS estimation and F-test
-### Aliases: R.OLS
-### Keywords: htest
-
-### ** Examples
-
-data(data1)
-# Define Y and X
-y=data1$lnoutput; x=cbind(data1$lncapital,data1$lnlabor)
-# Restriction matrices to test for constant returns to scale
-Rmat=matrix(c(0,1,1),nrow=1); rvec=matrix(1,nrow=1)
-# Model Estimation and F-test
-M=R.OLS(y,x,Rmat,rvec) 
-
-
-

### 2020-01-24 data/genthat_extracted_code/PASWR/examples/GD.Rd.R

@@ -1,23 +1,0 @@
-library(PASWR)
-
-
-### Name: GD
-### Title: Times Until Failure
-### Aliases: GD
-### Keywords: datasets
-
-### ** Examples
-
-with(data = GD, 
-hist(attf,prob=TRUE))
-with(data = GD, 
-lines(density(attf)))
-# Trellis Approach
-histogram(~attf, data = GD, type="density",
-panel = function(x, ...) {
-panel.histogram(x, ...)
-panel.densityplot(x, col = "blue", plot.points=TRUE, lwd=2)
-} )
-
-
-

### 2020-01-24 data/genthat_extracted_code/PPRL/examples/CreateBF.Rd.R

@@ -1,22 +1,0 @@
-library(PPRL)
-
-
-### Name: CreateBF
-### Title: Bloom Filter Encoding
-### Aliases: CreateBF BF Bloomfilter 'Bloom Filter'
-
-### ** Examples
-
-# Load test data
-testFile <- file.path(path.package("PPRL"), "extdata/testdata.csv")
-testData <- read.csv(testFile, head = FALSE, sep = "\t",
-  colClasses = "character")
-
-# Encode data
-BF <- CreateBF(ID = testData$V1, data = testData$V7,
-  k = 20, padding = 1, q = 2, l = 1000,
-  password = "(H]$6Uh*-Z204q")
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/PointFore/examples/constant.Rd.R

@@ -1,19 +1,0 @@
-library(PointFore)
-
-
-### Name: constant
-### Title: Constant specification model
-### Aliases: constant
-
-### ** Examples
-
-# the returned level does not depend on the state variable
-constant(0,.5)
-constant(1,.5)
-
-# if theta is not in the unit interval, the constant specification model forces it to be so
-constant(0, 2)
-constant(0, -1)
-
-
-

### 2020-01-24 data/genthat_extracted_code/ProfileLikelihood/examples/profilelike.lme.Rd.R

@@ -1,16 +1,0 @@
-library(ProfileLikelihood)
-
-
-### Name: profilelike.lme
-### Title: Profile Likelihood for Linear Mixed-Effects Models
-### Aliases: profilelike.lme
-
-### ** Examples
-
-xx <- profilelike.lme(formula = yield ~ endpoint, random = ~ 1 | id, 
-		correlation=corAR1(form = ~ 1 | id), data=Gasoline, subject="Sample", 
-		profile.theta="vapor", method="ML", lo.theta=1, hi.theta=5, length=500, round=2)
-profilelike.plot(theta=xx$theta, profile.lik.norm=xx$profile.lik.norm, round=4)
-
-
-

### 2020-01-24 data/genthat_extracted_code/PtProcess/examples/gif.Rd.R

@@ -1,34 +1,0 @@
-library(PtProcess)
-
-
-### Name: gif
-### Title: General Notes on Ground Intensity Functions
-### Aliases: gif
-### Keywords: documentation
-
-### ** Examples
-
-#  Ogata's Data: ground intensity function
-#  evaluate lambda_g(t) at certain times
-
-data(Ogata)
-
-p <- c(0.02, 70.77, 0.47, 0.002, 1.25)
-times <- sort(c(seq(0, 800, 0.5), Ogata$time))
-TT <- c(0, 800)
-
-plot(times, log(etas_gif(Ogata, times, params=p)), type="l",
-     ylab=expression(paste(log, " ", lambda[g](t))),
-     xlab=expression(t), xlim=TT)
-
-#  Evaluate the integral
-#   The first form below is where the arguments are in their
-#   default positions, the 2nd is where they are not, hence
-#   their names must be specified
-
-print(etas_gif(Ogata, NULL, p, TT))
-#  or
-print(etas_gif(Ogata, params=p, TT=TT))
-
-
-

### 2020-01-24 data/genthat_extracted_code/R0/examples/sensitivity.analysis.Rd.R

@@ -1,61 +1,0 @@
-library(R0)
-
-
-### Name: sensitivity.analysis
-### Title: Sensitivity analysis of basic reproduction ratio to begin/end
-###   dates
-### Aliases: sensitivity.analysis
-
-### ** Examples
-#Loading package
-library(R0)
-
-## Data is taken from the paper by Nishiura for key transmission parameters of an institutional
-## outbreak during 1918 influenza pandemic in Germany)
-data(Germany.1918)
-
-## For this exemple, we use the exact same call as for the internal sensitivity analysis function
-
-## sa.type = "GT"
-
-## Here we will test GT with means of 1 to 5, each time with SD constant (1)
-## GT and SD can be either fixed value or vectors of values
-## Actual value in simulations may differ, as they are adapted according to the distribution type
-tmp<-sensitivity.analysis(sa.type="GT", incid=Germany.1918, GT.type="gamma", GT.mean=seq(1,5,1), 
-                          GT.sd.range=1, begin=1, end=27, est.method="EG")
-
-## Results are stored in a matrix, each line dedicated to a (mean,sd) couple
-plot(x=tmp[,"GT.Mean"], xlab="mean GT (days)", y=tmp[,"R"], ylim=c(1.2, 2.1), ylab="R0 (95% CI)", 
-     type="p", pch=19, col="black", main="Sensitivity of R0 to mean GT")
-arrows(x0=as.numeric(tmp[,"GT.Mean"]), y0=as.numeric(tmp[,"CI.lower"]), 
-       y1=as.numeric(tmp[,"CI.upper"]), angle=90, code=3, col="black", length=0.05)
-## One could tweak this example to change sorting of values (per mean, or per standard deviation)
-## eg: 'x=tmp[,c('GT.Mean')]' could become 'x=tmp[,c('GT.SD')]'
-
-
-## sa.type="time"
-
-mGT<-generation.time("gamma", c(2.6,1))
-sen=sensitivity.analysis(sa.type="time", incid=Germany.1918, GT=mGT, begin=1:15, end=16:30, 
-                         est.method="EG")
-# ...
-# Warning message:
-# If 'begin' and 'end' overlap, cases where begin >= end are skipped.
-# These cases often return Rsquared = 1 and are thus ignored.
-## A list with different estimates of reproduction ratio, exponential growth rate and 95%CI 
-## wtih different pairs of begin and end dates in form of data frame is returned.
-## If method is "EG", results will include growth rate and deviance R-squared measure
[…]

### 2020-01-24 data/genthat_extracted_code/RMark/examples/crdms.Rd.R

@@ -1,85 +1,0 @@
-library(RMark)
-
-
-### Name: crdms
-### Title: Example data for Closed Robust Design Multistrata
-### Aliases: crdms
-### Keywords: datasets
-
-### ** Examples
-
-## No test: 
-# This example is excluded from testing to reduce package check time
-#Script to simulate the MSCRD 
-#example of 15.7.1 from the MARK
-#book
-#created by AJP 21 Dec 2010
-
-
-#convert .inp data - only needed to create crdms
-#ch.data<-convert.inp("rd_simple1.inp")
-data(crdms)
-#set time intervals
-#4 primary periods each with 3 secondary occasions
-t.int<-c(rep(c(0,0,1),3),c(0,0))
-
-#process data for RMark
-crdms.data<-process.data(crdms,model="CRDMS",time.interval=t.int,
-			strata.labels=c("1","U"))
-#change Psi parameters that are obtained by subtraction
-crdms.ddl<-make.design.data(crdms.data,
-		parameters=list(Psi=list(subtract.stratum=c("1","1"))))
-
-#create grouping index for unobserved p and c (i.e., always zero)
-up=as.numeric(row.names(crdms.ddl$p[crdms.ddl$p$stratum=="U",]))
-
-#create grouping index to fix Psi for unobs to unbos at time 1
-#this isn't necessary but it allows this Psi to be fixed to a value
-#that can be flagged and not erroneously interpreted
-Psiuu1=as.numeric(row.names(crdms.ddl$Psi[crdms.ddl$Psi$stratum=="U"&
-			crdms.ddl$Psi$time==1,]))
-
-#create dummy variable for constraining last Psi in Markovian model
-#variable is called ctime for constrained time
-crdms.ddl$Psi$ctime=crdms.ddl$Psi$time
-crdms.ddl$Psi$ctime[crdms.ddl$Psi$time==3]=2
-do_example=function()
-{
-#Initial assumptions
-S.dot=list(formula=~1)  #S equal for both states and constant over time
-p.session=list(formula=~session, share=TRUE,  #p=c varies with session 
-		fixed=list(index=up,value=0)) #p set to zero for unobs
-
-#Model 1 - Markovian movement
-Psi.markov=list(formula=~ctime+stratum,
-		fixed=list(index=Psiuu1,value=9e-99)) #9e-99 is a flag
[…]

### 2020-01-24 data/genthat_extracted_code/RankResponse/examples/rank.btmm.Rd.R

@@ -1,23 +1,0 @@
-library(RankResponse)
-
-
-### Name: rank.btmm
-### Title: Rank responses based on the Bradley-Terry model with the MM
-###   method
-### Aliases: rank.btmm
-
-### ** Examples
-
-## This is an example to rank three responses in a multiple response question when
-## the number of respondents is 1000. In this example, we do not use a real data,
-## but generate data in the first three lines.
-A <-sample.int(2,1000,replace=TRUE,prob=c(0.37,0.63))-1
-B <-sample.int(2,1000,replace=TRUE,prob=c(0.65,0.35))-1
-C <-sample.int(2,1000,replace=TRUE,prob=c(0.5,0.5))-1
-D <-cbind(A,B,C)
-data <-matrix(D,nrow=1000,ncol=3)
-## or upload the true data
-rank.btmm(data)
-
-
-

### 2020-01-24 data/genthat_extracted_code/SampleSizeProportions/examples/propdiff.alc.Rd.R

@@ -1,13 +1,0 @@
-library(SampleSizeProportions)
-
-
-### Name: propdiff.alc
-### Title: Bayesian sample size determination for the difference between
-###   two binomial proportions using the Average Length Criterion
-### Aliases: propdiff.alc
-### Keywords: design
-
-### ** Examples
-propdiff.alc(len=0.05, c1=3, d1=11, c2=11, d2=54)
-
-

### 2020-01-24 data/genthat_extracted_code/SimTimeVar/examples/closest.Rd.R

@@ -1,13 +1,0 @@
-library(SimTimeVar)
-
-
-### Name: closest
-### Title: Return closest value
-### Aliases: closest
-
-### ** Examples
-
-closest( x = 5, candidates = c(-3, 8, 25) )
-
-
-

### 2020-01-24 data/genthat_extracted_code/StatDA/examples/cor.sign.Rd.R

@@ -1,17 +1,0 @@
-library(StatDA)
-
-
-### Name: cor.sign
-### Title: Correlation Matrix
-### Aliases: cor.sign
-### Keywords: univar multivariate
-
-### ** Examples
-
-data(chorizon)
-x=chorizon[,c("Ca","Cu","Mg","Na","P","Sr","Zn")]
-
-cor.sign(log10(x),method="spearman")
-
-
-

### 2020-01-24 data/genthat_extracted_code/SuperGauss/tests/test-Snorm.hess.R

@@ -1,99 +1,0 @@
-library(SuperGauss)
-library(mvtnorm)
-library(numDeriv)
-source("SuperGauss-test-functions.R")
-context("Hessian")
-
-# ok construct a simple example
-# acf(t, theta): sigma^2 * exp(t, alpha)
-# mu(t, theta) = alpha * t + log(gamma + t)
-# theta = (alpha, gamma, sigma)
-
-acf.fun <- function(tseq, theta) {
-  theta[3]^2 * exp(-theta[1]*tseq)
-}
-mu.fun <- function(tseq, theta) {
-  theta[1] * tseq + log(theta[2] + tseq)
-}
-dacf.fun <- function(tseq, theta) {
-  jacobian(acf.fun, theta, tseq = tseq)
-}
-dmu.fun <- function(tseq, theta) {
-  jacobian(mu.fun, theta, tseq = tseq)
-}
-d2acf.fun <- function(tseq, theta) {
-  H <- vapply(tseq, function(tt) {
-    hessian(function(th) acf.fun(tt, th), theta)
-  }, matrix(0, length(theta), length(theta)))
-  aperm(H, dim = c(3, 1, 2))
-}
-d2mu.fun <- function(tseq, theta) {
-  H <- vapply(tseq, function(tt) {
-    hessian(function(th) mu.fun(tt, th), theta)
-  }, matrix(0, length(theta), length(theta)))
-  aperm(H, dim = c(3, 1, 2))
-}
-
-test_that("Snorm.hess and numerical deriv. of dmvnorm agree", {
-  case.par <- expand.grid(N = sample(10:30, 10, replace = TRUE),
-                          miss = c("none", "dacf", "dmu", "mu"))
-  for(ii in 1:nrow(case.par)) {
-    N <- case.par$N[ii]
-    miss <- as.character(case.par$miss[ii])
-    theta0 <- runif(3, min = .75, max = 1.25)
-    dT <- runif(1, min = .75, max = 1.25)
-    tseq <- (1:N-1)*dT
-    if(miss == "none") {
-      loglik.mvn <- function(theta, X, tseq) {
-        dmvnorm(x = X, mean = mu.fun(tseq, theta),
-                sigma = toeplitz(acf.fun(tseq, theta)), log = TRUE)
-      }
-      X <- rmvnorm(n = 1, mean = mu.fun(tseq, theta0),
[…]

### 2020-01-24 data/genthat_extracted_code/VGAM/examples/skellamUC.Rd.R

@@ -1,19 +1,0 @@
-library(VGAM)
-
-
-### Name: Skellam
-### Title: The Skellam Distribution
-### Aliases: Skellam dskellam rskellam
-### Keywords: distribution
-
-### ** Examples
-
-## Not run: 
-##D  mu1 <- 1; mu2 <- 2; x <- (-7):7
-##D plot(x, dskellam(x, mu1, mu2), type = "h", las = 1, col = "blue",
-##D      main = paste("Density of Skellam distribution with mu1 = ", mu1,
-##D                   " and mu2 = ", mu2, sep = "")) 
-## End(Not run)
-
-
-

### 2020-01-24 data/genthat_extracted_code/WaterML/examples/GetVariables.Rd.R

@@ -1,14 +1,0 @@
-library(WaterML)
-
-
-### Name: GetVariables
-### Title: GetVariables
-### Aliases: GetVariables
-### Keywords: WaterML
-
-### ** Examples
-
-GetVariables("http://worldwater.byu.edu/app/index.php/rushvalley/services/cuahsi_1_1.asmx?WSDL")
-
-
-

### 2020-01-24 data/genthat_extracted_code/XiMpLe/examples/XMLGetters-methods.Rd.R

@@ -1,42 +1,0 @@
-library(XiMpLe)
-
-
-### Name: XMLName
-### Title: Getter/setter methods for S4 objects of XiMpLe XML classes
-### Aliases: XMLName XMLName,XiMpLe.node-method XMLName,-methods XMLName<-
-###   XMLName<-,XiMpLe.node-method XMLName<-,-methods XMLAttrs
-###   XMLAttrs,XiMpLe.node-method XMLAttrs,-methods XMLAttrs<-
-###   XMLAttrs<-,XiMpLe.node-method XMLAttrs<-,-methods XMLChildren
-###   XMLChildren,XiMpLe.node-method XMLChildren,-methods
-###   XMLChildren,XiMpLe.doc-method XMLChildren,-methods XMLChildren<-
-###   XMLChildren<-,XiMpLe.node-method XMLChildren<-,-methods
-###   XMLChildren<-,XiMpLe.doc-method XMLChildren<-,-methods XMLValue
-###   XMLValue,XiMpLe.node-method XMLValue,-methods XMLValue<-
-###   XMLValue<-,XiMpLe.node-method XMLValue<-,-methods XMLFile
-###   XMLFile,XiMpLe.doc-method XMLFile,-methods XMLFile<-
-###   XMLFile<-,XiMpLe.doc-method XMLFile<-,-methods XMLDecl
-###   XMLDecl,XiMpLe.doc-method XMLDecl,-methods XMLDecl<-
-###   XMLDecl<-,XiMpLe.doc-method XMLDecl<-,-methods XMLDTD
-###   XMLDTD,XiMpLe.doc-method XMLDTD,-methods XMLDTD<-
-###   XMLDTD<-,XiMpLe.doc-method XMLDTD<-,-methods XMLScan
-###   XMLScan,XiMpLe.node-method XMLScan,-methods XMLScan,XiMpLe.doc-method
-###   XMLScan<- XMLScan<-,XiMpLe.node-method XMLScan<-,-methods
-###   XMLScan<-,XiMpLe.doc-method XMLScanDeep
-###   XMLScanDeep,XiMpLe.node-method XMLScanDeep,-methods
-###   XMLScanDeep,XiMpLe.doc-method
-### Keywords: methods
-
-### ** Examples
-
-xmlTestNode <- XMLNode("foo", XMLNode("testchild"))
-XMLName(xmlTestNode) # returns "foo"
-XMLName(xmlTestNode) <- "bar"
[…]

### 2020-01-24 data/genthat_extracted_code/appnn/examples/plot.Rd.R

@@ -1,17 +1,0 @@
-library(appnn)
-
-
-### Name: plot
-### Title: Plots generation of the amyloidogenicity propensity predicted
-###   values per amino acid residue.
-### Aliases: plot.appnn
-### Keywords: ~kwd1 ~kwd2
-
-### ** Examples
-
-sequences <- c('STVIIE','KKSSTT','KYSTVI')
-predictions <- appnn(sequences)
-plot(predictions,c(1,2,3))
-
-
-

### 2020-01-24 data/genthat_extracted_code/assist/examples/Thin.rd.R

@@ -1,16 +1,0 @@
-library(assist)
-
-
-### Name: Thin
-### Title: Calculate Reproducing Kernels for Thin Plate Splines
-### Aliases: tp.pseudo tp tp.linear
-### Keywords: file
-
-### ** Examples
-
-data(acid)
-## Not run: tp.pseudo(list(acid$x1, acid$x2))
-## Not run: tp.pseud0(list(acid$x1, acid$x2), order=3)
-
-
-

### 2020-01-24 data/genthat_extracted_code/assocInd/examples/iieSimHWI.Rd.R

@@ -1,45 +1,0 @@
-library(assocInd)
-
-
-### Name: iieSimHWI
-### Title: Simulate HWI with individual identification error
-### Aliases: iieSimHWI
-
-### ** Examples
-
-
-	# Set a real association index
-	aAB <- 0.5
-
-	# Create a range of errors
-	e <- seq(0,0.8,0.01)
-	
-	# Replicate N times
-	replicates <- 100  # small number used to save computation time
-	
-	# Create a blank storage matrices
-	assocStrength <- matrix(NA,nrow=replicates,ncol=length(e))
-	inCIs <- matrix(NA,nrow=replicates,ncol=length(e))
-	
-	# Loop through repeating N times for each error value
-	for (i in 1:length(e)) { 
-		for (j in 1:replicates) {
-			out <- iieSimHWI(aAB,e[i],20)
-			assocStrength[j,i] <- out[1]
-			inCIs[j,i] <- out[2]
-		}
-	}
-	
-	# Plot the results
-	par(mfrow=c(1,2))
-	plot(e,colMeans(assocStrength, na.rm=TRUE), pch=20, ylim=c(0,1), ylab="Simulated HWI")
-	CIs <- apply(assocStrength,2,quantile,c(0.025,0.975),na.rm=TRUE)
-	arrows(e,CIs[1,],e,CIs[2,],len=0.1,code=3,angle=90)
-	abline(h=0.5,col="red")
-	
-	plot(e,colMeans(inCIs, na.rm=TRUE), pch=20, ylim=c(0,1), ylab="Percent of times in CIs")
-	abline(h=0.95, col="red")
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/augmentedRCBD/examples/freqdist.augmentedRCBD.Rd.R

@@ -1,51 +1,0 @@
-library(augmentedRCBD)
-
-
-### Name: freqdist.augmentedRCBD
-### Title: Plot Frequency Distribution from 'augmentedRCBD' Output
-### Aliases: freqdist.augmentedRCBD
-
-### ** Examples
-
-# Example data
-blk <- c(rep(1,7),rep(2,6),rep(3,7))
-trt <- c(1, 2, 3, 4, 7, 11, 12, 1, 2, 3, 4, 5, 9, 1, 2, 3, 4, 8, 6, 10)
-y1 <- c(92, 79, 87, 81, 96, 89, 82, 79, 81, 81, 91, 79, 78, 83, 77, 78, 78,
-        70, 75, 74)
-y2 <- c(258, 224, 238, 278, 347, 300, 289, 260, 220, 237, 227, 281, 311, 250,
-        240, 268, 287, 226, 395, 450)
-data <- data.frame(blk, trt, y1, y2)
-# Convert block and treatment to factors
-data$blk <- as.factor(data$blk)
-data$trt <- as.factor(data$trt)
-# Results for variable y1
-out1 <- augmentedRCBD(data$blk, data$trt, data$y1, method.comp = "lsd",
-                      alpha = 0.05, group = TRUE, console = TRUE)
-# Results for variable y2
-out2 <- augmentedRCBD(data$blk, data$trt, data$y2, method.comp = "lsd",
-                     alpha = 0.05, group = TRUE, console = TRUE)
-
-# Frequency distribution plots
-freq1 <- freqdist.augmentedRCBD(out1, xlab = "Trait 1")
-class(freq1)
-plot(freq1)
-freq2 <- freqdist.augmentedRCBD(out2, xlab = "Trait 2")
-plot(freq2)
-
-# Change check colours
-colset <- c("red3", "green4", "purple3", "darkorange3")
-freq1 <- freqdist.augmentedRCBD(out1, xlab = "Trait 1", check.col = colset)
-plot(freq1)
-freq2 <- freqdist.augmentedRCBD(out2, xlab = "Trait 2", check.col = colset)
-plot(freq2)
-
-# Without checks highlighted
-freq1 <- freqdist.augmentedRCBD(out1, xlab = "Trait 1",
-                                highlight.check = FALSE)
-plot(freq1)
-freq2 <- freqdist.augmentedRCBD(out2, xlab = "Trait 2",
-                                highlight.check = FALSE)
-plot(freq2)
-
-
-

### 2020-01-24 data/genthat_extracted_code/aurelius/examples/avro_array.Rd.R

@@ -1,14 +1,0 @@
-library(aurelius)
-
-
-### Name: avro_array
-### Title: avro_array
-### Aliases: avro_array
-
-### ** Examples
-
-avro_array(avro_int)
-avro_array(avro_string)
-
-
-

### 2020-01-24 data/genthat_extracted_code/binneR/tests/test-detectInfusionScans.R

@@ -1,12 +1,0 @@
-
-context('detectInfusionScans')
-
-test_that('detectInfusionScans works',{
-	file <- list.files(system.file('mzML',package = 'binneR'),
-										 full.names = TRUE)[1]
-	
-	scans <- detectInfusionScans(file)
-	
-	expect_true(class(scans) == 'integer')
-	expect_true(identical(scans,as.integer(c(6:18))))
-})

### 2020-01-24 data/genthat_extracted_code/bio3d/examples/get.seq.Rd.R

@@ -1,21 +1,0 @@
-library(bio3d)
-
-
-### Name: get.seq
-### Title: Download FASTA Sequence Files
-### Aliases: get.seq
-### Keywords: IO utilities
-
-### ** Examples
-
-## Not run: 
-##D ## Sequence identifiers (GI or PDB codes e.g. from blast.pdb etc.)
-##D get.seq( c("P01112", "Q61411", "P20171") )
-##D 
-##D #aa <-get.seq( c("4q21", "5p21") )
-##D #aa$id
-##D #aa$ali
-## End(Not run)
-
-
-

### 2020-01-24 data/genthat_extracted_code/biomod2/examples/calculate.stat.Rd.R

@@ -1,24 +1,0 @@
-library(biomod2)
-
-
-### Name: calculate.stat
-### Title: Calculate evaluation metrics based on a misclassification table
-### Aliases: calculate.stat
-### Keywords: models, formula, options
-
-### ** Examples
-
-  a <- sample(c(0,1),100, replace=TRUE)
-  b <- sample(c(0,1),100, replace=TRUE)
-  
-  miscTab_aa <- table(a,a)
-  miscTab_ab <- table(a,b)
-  
-  # perfect score
-  calculate.stat( miscTab_aa, stat='TSS')
-  # random score
-  calculate.stat( miscTab_ab, stat='TSS')
-  
-
-
-

### 2020-01-24 data/genthat_extracted_code/colorscience/examples/RGB2LMS.Rd.R

@@ -1,15 +1,0 @@
-library(colorscience)
-
-
-### Name: RGB2LMS
-### Title: Convert RGB coordinates to LMS
-### Aliases: RGB2LMS
-### Keywords: datasets
-
-### ** Examples
-
-RGB<-c(124,63,78)
-RGB2LMS(RGB)
-
-
-

### 2020-01-24 data/genthat_extracted_code/ctl/examples/ctl.print.ctlscan.Rd.R

@@ -1,14 +1,0 @@
-library(ctl)
-
-
-### Name: print.CTLscan
-### Title: Print the results of a single phenotype CTL scan
-### Aliases: print.CTLscan
-### Keywords: print
-
-### ** Examples
-
-  #TODO
-
-
-

### 2020-01-24 data/genthat_extracted_code/darts/examples/drawHeatmap.Rd.R

@@ -1,18 +1,0 @@
-library(darts)
-
-
-### Name: drawHeatmap
-### Title: Heatmap of Expected Scores
-### Aliases: drawHeatmap
-
-### ** Examples
-
-# This will take a couple of seconds
-e = simpleExpScores(s=0)
-
-# This is what your heatmap would look like if you had
-# perfect accuracy
-drawHeatmap(e)
-
-
-

### 2020-01-24 data/genthat_extracted_code/emld/examples/as_xml.Rd.R

@@ -1,19 +1,0 @@
-library(emld)
-
-
-### Name: as_xml
-### Title: Coerce an emld object into XML (EML's standard format)
-### Aliases: as_xml
-
-### ** Examples
-
-f <- system.file("extdata/example.xml", package = "emld")
-emld <- as_emld(f)
-xml <- as_xml(emld)
-
-## can also write directly to a file:
-xml_file <- tempfile()
-as_xml(emld, xml_file)
-
-
-

### 2020-01-24 data/genthat_extracted_code/enrichvs/examples/matplot_enrichment_curve.Rd.R

@@ -1,16 +1,0 @@
-library(enrichvs)
-
-
-### Name: matplot_enrichment_curve
-### Title: Function to plot multiple enrichment curves for virtual
-###   screening
-### Aliases: matplot_enrichment_curve
-
-### ** Examples
-
-x <- matrix(rnorm(5000), ncol=5)  # random 5 scores for 1000 compounds
-y <- c(rep(1,50), rep(0,950))     # activity labels for "x"
-matplot_enrichment_curve(x, y)
-
-
-

### 2020-01-24 data/genthat_extracted_code/extremeStat/examples/plotLfit.Rd.R

@@ -1,15 +1,0 @@
-library(extremeStat)
-
-
-### Name: plotLfit
-### Title: Plot distributions fitted with L-moments
-### Aliases: plotLfit
-### Keywords: distribution hplot
-
-### ** Examples
-
- # See distLfit
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/fabricatr/examples/correlate.Rd.R

@@ -1,24 +1,0 @@
-library(fabricatr)
-
-
-### Name: correlate
-### Title: Perform generation of a correlated random variable.
-### Aliases: correlate
-
-### ** Examples
-
-
-# Generate a variable of interest
-exam_score <- pmin(100, rnorm(n = 100, mean = 80, sd = 10))
-
-# Generate a correlated variable using fabricatr variable generation
-scholarship_offers <- correlate(given = exam_score, rho = 0.7,
-                                draw_count, mean = 3)
-
-# Generate a correlated variable using base R distributions
-final_grade <- pmax(100, correlate(given = exam_score, rho = 0.7,
-                                   rnorm, mean = 80, sd = 10))
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/fungible/examples/RnpdMAP.Rd.R

@@ -1,118 +1,0 @@
-library(fungible)
-
-
-### Name: RnpdMAP
-### Title: Generate Random NPD R matrices from a user-supplied population R
-### Aliases: RnpdMAP
-### Keywords: fungible
-
-### ** Examples
-
-library(MASS)
-
-Nvar = 20
-Nfac = 4
-NSubj = 600
-Seed = 123    
-
-set.seed(Seed)
-
-## Generate a vector of classical item difficulties
-p <- runif(Nvar)
-
-cat("\nClassical Item Difficulties:\n")
-
-print(rbind(1:Nvar,round(p,2)) )
-
-summary(p)
-
-
-## Convert item difficulties to quantiles
-b <- qnorm(p)
-
-
-## fnc to compute root mean squared standard deviation
-RMSD <- function(A, B){
-  sqrt(mean( ( A[lower.tri(A, diag = FALSE)] - B[lower.tri(B, diag = FALSE)] )^2))
-}
-
-
-## Generate vector of eigenvalues with clear factor structure
-  L <- eigGen(nDimensions = Nvar, 
-            nMajorFactors = Nfac, 
-            PrcntMajor = .60, 
-            threshold  = .50)
-          
-
-## Generate a population R matrix with the eigenvalues in L
-  Rpop <- rGivens(eigs = L)$R
-  
-## Generate continuous data that will reproduce Rpop (exactly)
-  X <- mvrnorm(n = NSubj, mu = rep(0, Nvar), 
-               Sigma = Rpop, empirical = TRUE)
-               
-if( any(colSums(X) == 0) ){
-  stop("One or more variables have zero variance. Generate a new data set.")               
- }
- 
-## Cut X at thresholds given in b to produce binary data U
-  U <- matrix(0, nrow(X), ncol(X))
-  for(j in 1:Nvar){
-    U[X[,j] <= b[j],j] <- 1
-  }
-  
-## Compute tetrachoric correlations
-  Rtet <- tetcor(U, Smooth = FALSE, PRINT = TRUE)$r
-  # Calculate eigenvalues of tetrachoric R matrix
-  Ltet <- eigen(Rtet)$values
-  
-  if(Ltet[Nvar] >= 0) stop("Rtet is P(S)D")
-  
-## Simulate NPD R matrix with minimum eigenvalue equal to 
-  # min(Ltet)
[…]

### 2020-01-24 data/genthat_extracted_code/geozoo/examples/torus.Rd.R

@@ -1,16 +1,0 @@
-library(geozoo)
-
-
-### Name: torus
-### Title: Torus
-### Aliases: torus
-### Keywords: dynamic
-
-### ** Examples
-
-## Generates a torus
-torus(p = 3, n = 1000)
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/ggalt/examples/geom_lollipop.Rd.R

@@ -1,18 +1,0 @@
-library(ggalt)
-
-
-### Name: geom_lollipop
-### Title: Lollipop charts
-### Aliases: geom_lollipop
-
-### ** Examples
-
-df <- data.frame(trt=LETTERS[1:10],
-                 value=seq(100, 10, by=-10))
-
-ggplot(df, aes(trt, value)) + geom_lollipop()
-
-ggplot(df, aes(value, trt)) + geom_lollipop(horizontal=TRUE)
-
-
-

### 2020-01-24 data/genthat_extracted_code/ggvis/examples/left_right.Rd.R

@@ -1,17 +1,0 @@
-library(ggvis)
-
-
-### Name: left_right
-### Title: Interactive inputs bound to arrow keys.
-### Aliases: left_right up_down
-
-### ** Examples
-
-size <- left_right(1, 801, value = 51, step = 50)
-opacity <- up_down(0, 1, value = 0.9, step = 0.05)
-
-mtcars %>% ggvis(~mpg, ~wt, size := size, opacity := opacity) %>%
-  layer_points()
-
-
-

### 2020-01-24 data/genthat_extracted_code/googleVis/examples/gvisPieChart.Rd.R

@@ -1,32 +1,0 @@
-library(googleVis)
-
-
-### Name: gvisPieChart
-### Title: Google Pie Chart with R
-### Aliases: gvisPieChart
-### Keywords: iplot
-
-### ** Examples
-
-
-## Please note that by default the googleVis plot command
-## will open a browser window and requires an internet
-## connection to display the visualisation.
-
-Pie1 <- gvisPieChart(CityPopularity)
-plot(Pie1)
-
-## Doughnut chart - a pie with a hole
-Pie2 <- gvisPieChart(CityPopularity, options=list(
-                    slices="{4: {offset: 0.2}, 0: {offset: 0.3}}",
-                    title='City popularity',
-                    legend='none',
-                    pieSliceText='label',
-                    pieHole=0.5))
-plot(Pie2)
-
-
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/gsDesign/examples/gsProbability.Rd.R

@@ -1,49 +1,0 @@
-library(gsDesign)
-
-
-### Name: gsProbability
-### Title: 2.2: Boundary Crossing Probabilities
-### Aliases: gsProbability print.gsProbability
-### Keywords: design
-
-### ** Examples
-
-# making a gsDesign object first may be easiest...
-x <- gsDesign()
-
-# take a look at it
-x
-
-# default plot for gsDesign object shows boundaries
-plot(x)
-
-# plottype=2 shows boundary crossing probabilities
-plot(x, plottype=2)
-
-# now add boundary crossing probabilities and 
-# expected sample size for more theta values
-y <- gsProbability(d=x, theta=x$delta*seq(0, 2, .25))
-class(y)
-
-# note that "y" below is equivalent to print(y) and
-# print.gsProbability(y)
-y
-
-# the plot does not change from before since this is a
-# gsDesign object; note that theta/delta is on x axis
-plot(y, plottype=2)
-
-# now let's see what happens with a gsProbability object
-z <- gsProbability(k=3, a=x$lower$bound, b=x$upper$bound, 
-    n.I=x$n.I, theta=x$delta*seq(0, 2, .25))
-
-# with the above form,  the results is a gsProbability object
-class(z)
-z
-
-# default plottype is now 2
-# this is the same range for theta, but plot now has theta on x axis
-plot(z)
-
-
-

### 2020-01-24 data/genthat_extracted_code/hrbrthemes/vignettes/why_hrbrthemes.R

@@ -1,17 +1,0 @@
-## ----echo=FALSE, fig.width=8---------------------------------------------
-library(ggplot2)
-library(gridExtra)
-
-ggplot(mtcars, aes(wt, mpg)) + 
-  geom_point() +
-  labs(x="Weight (tons)", y="Efficiency (MPG)",
-       title="Lorem ipsum dolor sit amet",
-       subtitle="Mauris ultricies sem volutpat magna venenatis") -> gg
-
-grid.arrange(
-  gg + theme_minimal(base_family="serif") + theme(plot.margin=margin(t=20, b=20, l=15, r=30)),
-  gg + theme_minimal(base_family="sans") + theme(plot.margin=margin(t=20, b=20, l=30, r=15)),
-  ncol=2
-)
-
-

### 2020-01-24 data/genthat_extracted_code/hutils/examples/select_which.Rd.R

@@ -1,17 +1,0 @@
-library(hutils)
-
-
-### Name: select_which
-### Title: Select columns satisfying a condition
-### Aliases: select_which
-
-### ** Examples
-
-library(data.table)
-DT <- data.table(x = 1:5,
-                 y = letters[1:5],
-                 AB = c(NA, TRUE, FALSE))
-select_which(DT, anyNA, .and.dots = "y")
-
-
-

### 2020-01-24 data/genthat_extracted_code/hypergate/examples/channels_contributions.Rd.R

@@ -1,18 +1,0 @@
-library(hypergate)
-
-
-### Name: channels_contributions
-### Title: channels_contributions
-### Aliases: channels_contributions
-
-### ** Examples
-
-data(Samusik_01_subset)
-xp=Samusik_01_subset$xp_src[,Samusik_01_subset$regular_channels]
-gate_vector=Samusik_01_subset$labels
-hg=hypergate(xp=xp,gate_vector=gate_vector,level=23,delta_add=0)
-contribs=channels_contributions(gate=hg,xp=xp,gate_vector=gate_vector,level=23,beta=1)
-contribs
-
-
-

### 2020-01-24 data/genthat_extracted_code/igraph/examples/make_empty_graph.Rd.R

@@ -1,14 +1,0 @@
-library(igraph)
-
-
-### Name: make_empty_graph
-### Title: A graph with no edges
-### Aliases: make_empty_graph graph.empty empty_graph
-
-### ** Examples
-
-make_empty_graph(n = 10)
-make_empty_graph(n = 5, directed = FALSE)
-
-
-

### 2020-01-24 data/genthat_extracted_code/inlmisc/examples/AddColorKey.Rd.R

@@ -1,25 +1,0 @@
-library(inlmisc)
-
-
-### Name: AddColorKey
-### Title: Add Color Key to Plot
-### Aliases: AddColorKey
-### Keywords: hplot
-
-### ** Examples
-
-op <- par(mfrow = c(7, 1), omi = c(1, 1, 1, 1), mar = c(2, 3, 2, 3))
-AddColorKey(breaks = 0:10, explanation = "Example description of data variable.")
-AddColorKey(breaks = 0:1000, at = pretty(0:1000))
-AddColorKey(breaks = c(0, 1, 2, 4, 8, 16))
-breaks <- c(pi * 10^(-5:5))
-AddColorKey(breaks = breaks, log = TRUE)
-AddColorKey(breaks = breaks, at = breaks[as.logical(seq_along(breaks) %% 2)],
-            scipen = NULL, log = TRUE)
-AddColorKey(is.categorical = TRUE, labels = LETTERS[1:5])
-AddColorKey(is.categorical = TRUE, col = GetColors(5, scheme = "bright"))
-par(op)
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/kdecopula/examples/fitted.kdecopula.Rd.R

@@ -1,17 +1,0 @@
-library(kdecopula)
-
-
-### Name: fitted.kdecopula
-### Title: Extract fitted values from a 'kdecop()' fits.
-### Aliases: fitted.kdecopula
-
-### ** Examples
-
-data(wdbc)
-udat <- apply(wdbc[, -1], 2, function(x) rank(x) / (length(x) + 1))
-fit <- kdecop(udat[, 5:6])
-
-all.equal(fitted(fit), predict(fit, fit$udata))
-
-
-

### 2020-01-24 data/genthat_extracted_code/landscapemetrics/examples/lsm_l_siei.Rd.R

@@ -1,15 +1,0 @@
-library(landscapemetrics)
-
-
-### Name: lsm_l_siei
-### Title: SIEI (landscape level)
-### Aliases: lsm_l_siei lsm_l_siei.RasterLayer lsm_l_siei.RasterStack
-###   lsm_l_siei.RasterBrick lsm_l_siei.stars lsm_l_siei.list
-
-### ** Examples
-
-lsm_l_siei(landscape)
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/lemon/examples/brackets.Rd.R

@@ -1,26 +1,0 @@
-library(lemon)
-
-
-### Name: brackets_horizontal
-### Title: Axis brackets instead of axis ticks and lines
-### Aliases: brackets_horizontal brackets_horisontal brackets_vertical
-
-### ** Examples
-
-library(ggplot2)
-p <- ggplot(mpg, aes(as.factor(cyl), hwy, colour=class)) +
-  geom_point(position=position_jitter(width=0.3)) +
-  theme_bw() +
-  theme(panel.border = element_blank(), axis.line = element_line())
-p
-
-p <- p + coord_flex_cart(bottom=brackets_horizontal(length=unit(0.08, 'npc')))
-p
-# However getting the correct width is a matter of tweaking either length or
-# position_jitter...
-
-# A further adjustment,
-p + theme(panel.grid.major.x = element_blank())
-
-
-

### 2020-01-24 data/genthat_extracted_code/lexicon/examples/hash_grady_pos.Rd.R

@@ -1,21 +1,0 @@
-library(lexicon)
-
-
-### Name: hash_grady_pos
-### Title: Grady Ward's Moby Parts of Speech
-### Aliases: hash_grady_pos grady_pos_feature
-### Keywords: datasets
-
-### ** Examples
-
-## Not run: 
-##D library(data.table)
-##D 
-##D hash_grady_pos <- grady_pos_feature(hash_grady_pos)
-##D hash_grady_pos['dog']
-##D hash_grady_pos[primary == TRUE, ]
-##D hash_grady_pos[primary == TRUE & space == FALSE, ]
-## End(Not run)
-
-
-

### 2020-01-24 data/genthat_extracted_code/lordif/examples/probgpcm.Rd.R

@@ -1,14 +1,0 @@
-library(lordif)
-
-
-### Name: probgpcm
-### Title: calculates item response probabilities according to GPCM
-### Aliases: probgpcm
-### Keywords: ~kwd1 ~kwd2
-
-### ** Examples
-
-probgrm(seq(-4,4,.1), 1.5, c(-1.2,0.5,1.5))
-
-
-

### 2020-01-24 data/genthat_extracted_code/manymodelr/examples/select_col.Rd.R

@@ -1,19 +1,0 @@
-library(manymodelr)
-
-
-### Name: select_col
-### Title: A convenient selector gadget
-### Aliases: select_col
-
-### ** Examples
-
-select_col(iris,Petal.Length,Sepal.Length,Species,Petal.Width)
-# A pipe friendly example
-## Not run: 
-##D library(dplyr)
-##D as_tibble(iris) %>%
-##D select_col(Species)
-## End(Not run)
-
-
-

### 2020-01-24 data/genthat_extracted_code/marg/examples/rsm.families.Rd.R

@@ -1,17 +1,0 @@
-library(marg)
-
-
-### Name: rsm.families
-### Title: Generate a RSM Family Object
-### Aliases: rsm.families extreme Huber logistic logWeibull student
-### Keywords: classes methods
-
-### ** Examples
-
-student(df = 3) ## generates Student's t error distribution with 3 d.f.
-## Not run: 
-##D rsm(formula = value, data = value, family = extreme)
-## End(Not run)
-
-
-

### 2020-01-24 data/genthat_extracted_code/marinespeed/examples/get_folds.Rd.R

@@ -1,24 +1,0 @@
-library(marinespeed)
-
-
-### Name: get_folds
-### Title: Get folds
-### Aliases: get_folds
-
-### ** Examples
-
-## Not run: 
-##D folds <- get_folds("random")
-##D 
-##D abalistes <- "Abalistes stellatus"
-##D occ <- get_occurrences(abalistes)
-##D bg <- get_background("random")
-##D 
-##D occ_train <- kfold_data(abalistes, occ, folds$species, k=1, training=TRUE)
-##D occ_test <- kfold_data(abalistes, occ, folds$species, k=1, training=FALSE)
-##D bg_train <- kfold_data(abalistes, bg, folds$background, k=1, training=TRUE)
-##D bg_test <- kfold_data(abalistes, bg, folds$background, k=1, training=FALSE)
-## End(Not run)
-
-
-

### 2020-01-24 data/genthat_extracted_code/mclust/examples/defaultPrior.Rd.R

@@ -1,32 +1,0 @@
-library(mclust)
-
-
-### Name: defaultPrior
-### Title: Default conjugate prior for Gaussian mixtures.
-### Aliases: defaultPrior
-### Keywords: cluster
-
-### ** Examples
-
-# default prior
-irisBIC <- mclustBIC(iris[,-5], prior = priorControl())
-summary(irisBIC, iris[,-5])
-
-# equivalent to previous example
-irisBIC <- mclustBIC(iris[,-5], 
-                     prior = priorControl(functionName = "defaultPrior"))
-summary(irisBIC, iris[,-5])
-
-# no prior on the mean; default prior on variance
-irisBIC <- mclustBIC(iris[,-5], prior = priorControl(shrinkage = 0))
-summary(irisBIC, iris[,-5])
-
-# equivalent to previous example
-irisBIC <- mclustBIC(iris[,-5], prior =
-                     priorControl(functionName="defaultPrior", shrinkage=0))
-summary(irisBIC, iris[,-5])
-
-defaultPrior( iris[-5], G = 3, modelName = "VVV")
-
-
-

### 2020-01-24 data/genthat_extracted_code/mi/examples/censored-continuous.Rd.R

@@ -1,26 +1,0 @@
-library(mi)
-
-
-### Name: censored-continuous-class
-### Title: The "censored-continuous" Class, the "truncated-continuous"
-###   Class and Inherited Classes
-### Aliases: truncated-continuous-class truncated-continuous
-###   FF_truncated-continuous-class FN_truncated-continuous-class
-###   NF_truncated-continuous-class NN_truncated-continuous-class
-###   censored-continuous-class censored-continuous
-###   FF_censored-continuous-class FN_censored-continuous-class
-###   NF_censored-continuous-class NN_censored-continuous-class
-### Keywords: classes DirectedTowardDevelopeRs
-
-### ** Examples
-
-# STEP 0: GET DATA
-data(CHAIN, package = "mi")
-
-# STEP 0.5 CREATE A missing_variable (you never need to actually do this)
-#log_virus <- missing_variable(CHAIN$log_virus, type = "NN_censored-continuous", 
-#                              lower = 0, upper = Inf)
-#show(log_virus)
-
-
-

### 2020-01-24 data/genthat_extracted_code/mistat/examples/FAILTIME.Rd.R

@@ -1,22 +1,0 @@
-library(mistat)
-
-
-### Name: FAILTIME
-### Title: Failure Times
-### Aliases: FAILTIME
-### Keywords: datasets
-
-### ** Examples
-
-data(FAILTIME)
-
-library(survival)
-
-SuRe <- survreg(
-  Surv(time=FAILTIME) ~ 1 , 
-  dist = "exponential")
-
-summary(SuRe)
-
-
-

### 2020-01-24 data/genthat_extracted_code/mlmRev/examples/guImmun.Rd.R

@@ -1,15 +1,0 @@
-library(mlmRev)
-
-
-### Name: guImmun
-### Title: Immunization in Guatemala
-### Aliases: guImmun
-### Keywords: datasets
-
-### ** Examples
-
-data(guImmun)
-summary(guImmun)
-
-
-

### 2020-01-24 data/genthat_extracted_code/oce/examples/swAbsoluteSalinity.Rd.R

@@ -1,17 +1,0 @@
-library(oce)
-
-
-### Name: swAbsoluteSalinity
-### Title: Seawater absolute salinity, in GSW formulation
-### Aliases: swAbsoluteSalinity
-
-### ** Examples
-
-## Not run: 
-##D sa <- swAbsoluteSalinity(35.5, 300, 260, 16)
-##D stopifnot(abs(35.671358392019094 - sa) < 00.000000000000010)
-## End(Not run)
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/phangorn/examples/splitsNetwork.Rd.R

@@ -1,21 +1,0 @@
-library(phangorn)
-
-
-### Name: splitsNetwork
-### Title: Phylogenetic Network
-### Aliases: splitsNetwork
-### Keywords: cluster
-
-### ** Examples
-
-
-data(yeast)
-dm <- dist.ml(yeast)
-fit <- splitsNetwork(dm)
-net <- as.networx(fit)
-plot(net, "2D")
-write.nexus.splits(fit)
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/phylolm/examples/rbinTrait.Rd.R

@@ -1,16 +1,0 @@
-library(phylolm)
-
-
-### Name: rbinTrait
-### Title: Binary trait simulation
-### Aliases: rbinTrait
-
-### ** Examples
-
-tre = rtree(50)
-x = rTrait(n=1,phy=tre)
-X = cbind(rep(1,50),x)
-y = rbinTrait(n=1, phy=tre, beta=c(-1,0.5), alpha=1, X=X)
-
-
-

### 2020-01-24 data/genthat_extracted_code/phylosim/examples/setRate.ToleranceSubstitution.Rd.R

@@ -1,25 +1,0 @@
-library(phylosim)
-
-
-### Name: setRate.ToleranceSubstitution
-### Title: Set an unscaled rate for an event from a ToleranceSubstitution
-###   object
-### Aliases: setRate.ToleranceSubstitution ToleranceSubstitution.setRate
-###   setRate,ToleranceSubstitution-method
-
-### ** Examples
-
-	# create a ToleranceSubstitution object
-	# provide an Alphabet object and the rates
-	p<-ToleranceSubstitution(alphabet=BinaryAlphabet(), rate.list=list("1->0"=1,"0->1"=1))
-	# set the unscaled rate by event name
-	setRate(p,"0->1",2)
-	# get the unscaled rate of "0->1" by name
-	getRate(p,"0->1")
-	# set the unscaled rate by states
-	setRate(p,"0->1",0.5)
-	# get the unscaled rate of "0->1" by states
-	getRate(p,from="0",to="1")
- 
-
-

### 2020-01-24 data/genthat_extracted_code/phylotools/examples/get.phylip.name.Rd.R

@@ -1,24 +1,0 @@
-library(phylotools)
-
-
-### Name: get.phylip.name
-### Title: get the names of sequences from a PHYLIP file
-### Aliases: get.phylip.name
-### Keywords: phylip
-
-### ** Examples
-
-
-  cat("6 22",
-  "seq_1    --TTACAAATTGACTTATTATA",
-  "seq_2    GATTACAAATTGACTTATTATA",
-  "seq_3    GATTACAAATTGACTTATTATA",
-  "seq_5    GATTACAAATTGACTTATTATA",
-  "seq_8    GATTACAAATTGACTTATTATA",
-  "seq_10   ---TACAAATTGAATTATTATA",
-  file = "matk.phy", sep = "\n")
-get.phylip.name("matk.phy")
-unlink("matk.phy")
-
-
-

### 2020-01-24 data/genthat_extracted_code/pomp/examples/dacca.Rd.R

@@ -1,20 +1,0 @@
-library(pomp)
-
-
-### Name: dacca
-### Title: Model of cholera transmission for historic Bengal.
-### Aliases: dacca
-### Keywords: models datasets
-
-### ** Examples
-
-pompExample(dacca)
-plot(dacca)
-#MLEs on the natural scale
-coef(dacca)
-#MLEs on the transformed scale
-coef(dacca,transform=TRUE)
-plot(simulate(dacca))
-
-
-

### 2020-01-24 data/genthat_extracted_code/psychmeta/examples/var_error_u.Rd.R

@@ -1,14 +1,0 @@
-library(psychmeta)
-
-
-### Name: var_error_u
-### Title: Estimate the error variance of u ratios
-### Aliases: var_error_u
-
-### ** Examples
-
-var_error_u(u = .8, ni = 100, na = 200)
-var_error_u(u = .8, ni = 100, na = NA)
-
-
-

### 2020-01-24 data/genthat_extracted_code/rebmix/examples/RCLS.chunk.Rd.R

@@ -1,27 +1,0 @@
-library(rebmix)
-
-
-### Name: chunk-methods
-### Title: Extracts Chunk from Train and Test Datasets
-### Aliases: chunk chunk-methods chunk,RCLS.chunk-method
-###   show,RCLS.chunk-method
-### Keywords: classification
-
-### ** Examples
-
-data("iris")
-
-# Split dataset into train (75%) and test (25%) subsets.
-
-set.seed(5)
-
-Iris <- split(p = 0.75, Dataset = iris, class = 5)
-
-# Extract chunk from train and test datasets.
-
-Iris14 <- chunk(x = Iris, variables = c(1,4))
-
-Iris14
-
-
-

### 2020-01-24 data/genthat_extracted_code/replicationInterval/examples/ri.d.Rd.R

@@ -1,14 +1,0 @@
-library(replicationInterval)
-
-
-### Name: ri.d
-### Title: _d_-value (i.e., standardized mean difference) Replication
-###   Interval
-### Aliases: ri.d
-
-### ** Examples
-
-ri.d(d=.65,n1=50,n2=50,rep.n1=100,rep.n2=100)
-
-
-

### 2020-01-24 data/genthat_extracted_code/robis/examples/dataset.Rd.R

@@ -1,15 +1,0 @@
-library(robis)
-
-
-### Name: dataset
-### Title: Fetch dataset metadata.
-### Aliases: dataset
-
-### ** Examples
-
-datasets <- dataset(id = 1)
-datasets <- dataset(q = "MICROBIS")
-datasets <- dataset(nodeid = c(0,1))
-
-
-

### 2020-01-24 data/genthat_extracted_code/rpf/examples/rpf.rparam.Rd.R

@@ -1,17 +1,0 @@
-library(rpf)
-
-
-### Name: rpf.rparam
-### Title: Generates item parameters
-### Aliases: rpf.rparam rpf.rparam,rpf.1dim.drm-method
-###   rpf.rparam,rpf.mdim.drm-method rpf.rparam,rpf.1dim.graded-method
-###   rpf.rparam,rpf.mdim.graded-method rpf.rparam,rpf.mdim.nrm-method
-###   rpf.rparam,rpf.mdim.mcm-method rpf.rparam,rpf.1dim.lmp-method
-
-### ** Examples
-
-i1 <- rpf.drm()
-rpf.rparam(i1)
-
-
-

### 2020-01-24 data/genthat_extracted_code/rportfolios/examples/rshortonly.test.Rd.R

@@ -1,21 +1,0 @@
-library(rportfolios)
-
-
-### Name: rshortonly.test
-### Title: Generate random short only portfolios
-### Aliases: rshortonly.test
-### Keywords: math
-
-### ** Examples
-
-###
-### generate 100 short only portfolios of 30 investments with 30 non zero positions
-###
-x.result <- rshortonly.test( 100, 30 )
-###
-### generate 100 short only portfolios of 30 investments with 10 non zero positions
-###
-x.result <- rshortonly.test( 100, 30, 10 )
-
-
-

### 2020-01-24 data/genthat_extracted_code/rriskDistributions/examples/fit.cont.Rd.R

@@ -1,29 +1,0 @@
-library(rriskDistributions)
-
-
-### Name: fit.cont
-### Title: GUI for fitting continuous distributions to given data
-### Aliases: fit.cont
-### Keywords: gui
-
-### ** Examples
-
-## Not run: 
-##D   if ( class(tcltk::tclRequire("Tktable")) == "tclObj" ) {
-##D     res1 <- fit.cont(data2fit = rgamma(374, 4, 0.08))
-##D     res1
-##D 
-##D     res2 <- fit.cont(data2fit = rbeta(300, shape1 = 1, shape2 = 2))
-##D     res2
-##D 
-##D     res3 <- fit.cont(data2fit = mc2d::rtriang(300, min = 1, mode = 3, max = 10))
-##D     res3
-##D 
-##D     res4 <- fit.cont(data2fit = stats::rnorm(300))
-##D     res4
-##D   }
-## End(Not run)
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/samplingVarEst/examples/VE_Jk_CBS_HT_Total_Hajek.rd.R

@@ -1,27 +1,0 @@
-library(samplingVarEst)
-
-
-### Name: VE.Jk.CBS.HT.Total.Hajek
-### Title: The Campbell-Berger-Skinner unequal probability jackknife
-###   variance estimator for the Hajek (1971) estimator of a total
-###   (Horvitz-Thompson form)
-### Aliases: VE.Jk.CBS.HT.Total.Hajek
-### Keywords: variance estimation total
-
-### ** Examples
-
-data(oaxaca)                                 #Loads the Oaxaca municipalities dataset
-pik.U  <- Pk.PropNorm.U(373, oaxaca$HOMES00) #Reconstructs the 1st order incl. probs.
-s      <- oaxaca$sHOMES00                    #Defines the sample to be used
-N      <- dim(oaxaca)[1]                     #Defines the population size
-y1     <- oaxaca$POP10                       #Defines the variable of interest y1
-y2     <- oaxaca$POPMAL10                    #Defines the variable of interest y2
-#This approximation is only suitable for large-entropy sampling designs
-pikl.s <- Pkl.Hajek.s(pik.U[s==1])           #Approx. 2nd order incl. probs. from s
-#Computes the var. est. of the Hajek total point estimator using y1
-VE.Jk.CBS.HT.Total.Hajek(y1[s==1], pik.U[s==1], pikl.s, N)
-#Computes the var. est. of the Hajek total point estimator using y2
-VE.Jk.CBS.HT.Total.Hajek(y2[s==1], pik.U[s==1], pikl.s, N)
-
-
-

### 2020-01-24 data/genthat_extracted_code/sars/examples/cole_sim.Rd.R

@@ -1,14 +1,0 @@
-library(sars)
-
-
-### Name: cole_sim
-### Title: A simulated species-site abundance matrix with site areas
-### Aliases: cole_sim
-### Keywords: datasets
-
-### ** Examples
-
-data(cole_sim)
-
-
-

### 2020-01-24 data/genthat_extracted_code/sbtools/examples/item_upload_create.Rd.R

@@ -1,18 +1,0 @@
-library(sbtools)
-
-
-### Name: item_upload_create
-### Title: Upload file(s) and create a new item
-### Aliases: item_upload_create
-
-### ** Examples
-
-## Not run: 
-##D # You'll need a parent id for a folder/item
-##D ## here, using your highest level parent folder
-##D file <- system.file("examples", "books.json", package = "sbtools")
-##D item_upload_create(user_id(), file)
-## End(Not run)
-
-
-

### 2020-01-24 data/genthat_extracted_code/sddpack/examples/sddpack-package.Rd.R

@@ -1,15 +1,0 @@
-library(sddpack)
-
-
-### Name: sddpack-package
-### Title: Semidiscrete Decomposition
-### Aliases: sddpack-package sddpack
-### Keywords: package
-
-### ** Examples
-
-	A = matrix(rnorm(100), nrow=10)
-	sdd(A)
-
-
-

### 2020-01-24 data/genthat_extracted_code/secr/examples/region.N.Rd.R

@@ -1,29 +1,0 @@
-library(secr)
-
-
-### Name: region.N
-### Title: Population Size
-### Aliases: region.N region.N.secr region.N.secrlist 'population size'
-### Keywords: models
-
-### ** Examples
-
-
-region.N(secrdemo.0)
-
-## Not run: 
-##D ## a couple more routine examples
-##D region.N(secrdemo.CL)
-##D region.N(ovenbird.model.D)
-##D 
-##D ## region defined as vector polygon
-##D ## retain and plot region mask
-##D temp <- region.N(possum.model.0, possumarea, spacing = 40,
-##D     keep.region = TRUE)
-##D temp
-##D plot (attr(temp, "region"), type = "l")
-## End(Not run)
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/seleniumPipes/tests/setup.R

@@ -1,20 +1,0 @@
-library(testthat)
-if(is.null(getOption("seleniumPipes_selOptions"))){
-  stop("Please set test options")
-}
-remDr <- do.call(remoteDr, getOption("seleniumPipes_selOptions"))
-if(identical(getOption("seleniumPipes_SL"), TRUE)){
-  options(seleniumPipes_sauceID = remDr$sessionInfo$webdriver.remote.sessionid)
-}
-remDr %>% setTimeout(milliseconds = 10000) # set page load timeout to 3 secs
-remDr %>% setTimeout(type = "implicit", milliseconds = 5000) # wait 5 secs for elements to load
-rdBrowser <- remDr$sessionInfo$browserName
-if(rdBrowser %in% c("iPhone", "iPad", "safari")){
-  htmlSrc <- "myname.local:3000"
-}else{
-  # add somethin like 127.0.0.1	myname.local to /etc/hosts or equivalent
-  htmlSrc <- "localhost:3000"
-}
-loadPage <- function(pgStr){
-  paste0("http://", file.path(htmlSrc, paste0(pgStr, ".html")))
-}

### 2020-01-24 data/genthat_extracted_code/sitar/examples/codeplot.Rd.R

@@ -1,22 +1,0 @@
-library(sitar)
-
-
-### Name: codeplot
-### Title: Plot and zap velocity outliers in growth curves
-### Aliases: codeplot zapvelout zapvelout
-
-### ** Examples
-
-
-## identify outliers
-outliers <- velout(age, height, id, heights, limit=2)
-
-## plot outliers with code 4 or 6
-codeplot(outliers, icode=c(4,6))
-
-## set the 8 outliers missing
-newheights <- zapvelout(outliers, icode=6)
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/sjlabelled/examples/get_label.Rd.R

@@ -1,50 +1,0 @@
-library(sjlabelled)
-
-
-### Name: get_label
-### Title: Retrieve variable label(s) of labelled data
-### Aliases: get_label
-
-### ** Examples
-
-# import SPSS data set
-# mydat <- read_spss("my_spss_data.sav", enc="UTF-8")
-
-# retrieve variable labels
-# mydat.var <- get_label(mydat)
-
-# retrieve value labels
-# mydat.val <- get_labels(mydat)
-
-data(efc)
-
-# get variable lable
-get_label(efc$e42dep)
-
-# alternative way
-get_label(efc)["e42dep"]
-
-# 'get_label()' also works within pipe-chains
-efc %>% get_label(e42dep, e16sex)
-
-# set default values
-get_label(mtcars, mpg, cyl, def.value = "no var labels")
-
-# simple barplot
-barplot(table(efc$e42dep))
-# get value labels to annotate barplot
-barplot(table(efc$e42dep),
-        names.arg = get_labels(efc$e42dep),
-        main = get_label(efc$e42dep))
-
-# get labels from multiple variables
-get_label(list(efc$e42dep, efc$e16sex, efc$e15relat))
-
-# use case conversion for human-readable labels
-data(iris)
-get_label(iris, def.value = colnames(iris))
-get_label(iris, def.value = colnames(iris), case = "parsed")
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/sppmix/examples/GetDensityValues.Rd.R

@@ -1,26 +1,0 @@
-library(sppmix)
-
-
-### Name: GetDensityValues
-### Title: Retrieve density values
-### Aliases: GetDensityValues
-
-### ** Examples
-
-## No test: 
-# create the true mixture intensity surface
-truesurf =normmix(ps=c(.2, .6,.2), mus=list(c(0.3, 0.3), c(0.7, 0.7),
- c(0.5, 0.5)),sigmas=list(.01*diag(2), .01*diag(2), .01*diag(2)),
- lambda=100,win=spatstat::square(1))
-plot(truesurf)
-# generate the point pattern, truncate=TRUE by default
-genPP=rsppmix(truesurf,truncate=FALSE)
-fit=est_mix_damcmc(pp = genPP, m = 3)
-allvals=GetDensityValues(fit)
-MAPest=GetMAPEst(fit,vals=allvals)
-plot(MAPest,main="IPPP intensity surface of MAP estimates")
-## End(No test)
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/stm/tests/test-prepare.R

@@ -1,14 +1,0 @@
-context("Prepare text")
-
-test_that("textProcessor and prepDocuments correctly subset metadata ", {
-  data("gadarian")
-
-  txtOut <- textProcessor(documents = gadarian$open.ended.response,
-                          metadata = data.frame(MetaID = gadarian$MetaID),
-                          sparselevel = .8)
-  expect_equal(nrow(txtOut$meta), length(txtOut$documents))
-
-  prepped <- prepDocuments(txtOut$documents, txtOut$vocab,
-                           txtOut$meta, upper.thresh = 100)
-  expect_equal(nrow(prepped$meta), length(prepped$documents))
-})

### 2020-01-24 data/genthat_extracted_code/taxize/examples/get_iucn.Rd.R

@@ -1,25 +1,0 @@
-library(taxize)
-
-
-### Name: get_iucn
-### Title: Get a IUCN Redlist taxon
-### Aliases: get_iucn as.iucn as.iucn.iucn as.iucn.character as.iucn.list
-###   as.iucn.numeric as.iucn.data.frame as.data.frame.iucn
-
-### ** Examples
-
-## Not run: 
-##D get_iucn(x = "Branta canadensis")
-##D get_iucn(x = "Branta bernicla")
-##D get_iucn(x = "Panthera uncia")
-##D 
-##D # as coercion
-##D as.iucn(22732)
-##D as.iucn("22732")
-##D (res <- as.iucn(c(22679946, 22732, 22679935)))
-##D data.frame(res)
-##D as.iucn(data.frame(res))
-## End(Not run)
-
-
-

### 2020-01-24 data/genthat_extracted_code/tsibble/examples/new-data.Rd.R

@@ -1,23 +1,0 @@
-library(tsibble)
-
-
-### Name: new_data
-### Title: New tsibble data and append new observations to a tsibble
-### Aliases: new_data new_data.tbl_ts append_row append_case
-
-### ** Examples
-
-new_data(pedestrian)
-new_data(pedestrian, keep_all = TRUE)
-new_data(pedestrian, n = 3)
-tsbl <- tsibble(
-  date = rep(as.Date("2017-01-01") + 0:2, each = 2), 
-  group = rep(letters[1:2], 3),
-  value = rnorm(6),
-  key = id(group)
-)
-append_row(tsbl)
-append_row(tsbl, n = 2)
-
-
-

### 2020-01-24 data/genthat_extracted_code/tuber/examples/list_caption_tracks.Rd.R

@@ -1,18 +1,0 @@
-library(tuber)
-
-
-### Name: list_caption_tracks
-### Title: List Captions of a Video
-### Aliases: list_caption_tracks
-
-### ** Examples
-
-## Not run: 
-##D 
-##D # Set API token via yt_oauth() first
-##D 
-##D list_caption_tracks(video_id = "yJXTXN4xrI8")
-## End(Not run)
-
-
-

### 2020-01-24 data/genthat_extracted_code/vdiffr/tests/setup-mock.R

@@ -1,38 +1,0 @@
-
-subset_results <- function(results, file, test) {
-  subset <- purrr::keep(results, function(result) {
-    result$file == file && result$test == test
-  })
-  subset[[1]]$results
-}
-
-
-mock_pkg_dir <- create_mock_pkg()
-mock_test_dir <- file.path(mock_pkg_dir, "tests", "testthat")
-
-vars <- c(
-  VDIFFR_LOG_PATH = file.path(mock_pkg_dir, "tests", "vdiffr.Rout.fail"),
-  NOT_CRAN = "true"
-)
-withr::with_envvar(vars, {
-  test_results <- testthat::test_dir(mock_test_dir, reporter = "silent")
-})
-
-mock_cases_outputs <- purrr::quietly(purrr::safely(collect_cases))(mock_pkg_dir)
-
-quietly_out <- mock_cases_outputs$result
-if (inherits(quietly_out$error, "condition")) {
-  cat("While collecting testing cases:\n")
-  stop(quietly_out$error)
-}
-
-log_path <- file.path(mock_pkg_dir, "tests", "vdiffr.Rout.fail")
-if (file.exists(log_path)) {
-  copied <- file.copy(log_path, "../mock.Rout.fail", overwrite = TRUE)
-  if (!copied) {
-    warning("Failed to copy mock package log")
-  }
-}
-
-mock_cases <- quietly_out$result
-validate_cases(mock_cases)

### 2020-01-24 data/genthat_extracted_code/weathermetrics/examples/humidity.to.dewpoint.Rd.R

@@ -1,21 +1,0 @@
-library(weathermetrics)
-
-
-### Name: humidity.to.dewpoint
-### Title: Calculate dew point temperature.
-### Aliases: humidity.to.dewpoint
-
-### ** Examples
-
-# Calculate dew point temperature from relative humidity and
-# air temperature.
-
-data(newhaven)
-newhaven$DP <- humidity.to.dewpoint(t = newhaven$TemperatureF,
-                                    rh = newhaven$Relative.Humidity,
-                                    temperature.metric = 'fahrenheit')
-newhaven
-
-
-
-

### 2020-01-24 data/genthat_extracted_code/wiqid/tests/test-Bnormal.R

@@ -1,72 +1,0 @@
-
-# test_that code for Bnormal
-
-# library(testthat)
-require(wiqid)
-
-context("Bayesian normal models")
-
-test_that("Bnormal gives same answers",  {
-  # Generate data
-  set.seed(123)
-  x <- rnorm(10, 1, 0.15)
-  # With default (flat) priors:
-  Bout <- Bnormal(x)
-  expect_that(class(Bout), equals(c("Bwiqid", "data.frame")))
-  expect_that(dim(Bout), equals(c(30000, 2)))
-  expect_that(names(Bout), equals(c("mu", "sigma")))
-  expect_that(attr(Bout, "header"),
-    equals("Model fitted in R with a Gibbs sampler"))
-  expect_that(attr(Bout, "n.chains"), equals(3))
-  expect_equal(as.character(attr(Bout, "call")), c("Bnormal", "x"))
-  expect_equivalent(round(attr(Bout, "n.eff")), c(28942, 23869))
-  expect_equivalent(round(attr(Bout, "Rhat"), 3), c(1.001, 1.000))
-
-  expect_equivalent(round(colMeans(Bout), 4), c(1.0112, 0.1563))
-  expect_equivalent(round(c(hdi(Bout)), 4), c(0.9089, 1.1118, 0.0901, 0.2406))
-  xx <- x * 1000
-  expect_warning(Bout <- Bnormal(xx, priors=list(muMean=0, muSD=10)), # silly prior
-      "Sample mean is outside the prior range")
-  set.seed(123)
-  Bout <- Bnormal(x, priors=list(muMean=0, muSD=10)) # informative prior for mu
-  expect_equivalent(round(colMeans(Bout), 4), c(1.0112, 0.1563))
-  expect_equivalent(round(c(hdi(Bout)), 4), c(0.9088, 1.1117, 0.0901, 0.2406))
-  set.seed(234)
-  Bout <- Bnormal(xx, priors=list(muMean=1000))  # sensible priors
-  expect_equivalent(round(colMeans(Bout), 1), c(1011.3, 156.5))
-  expect_equivalent(round(c(hdi(Bout)), 1), c(908.6, 1112.8,   91.4,  242.8))
-})
-
-test_that("Bnormal2 gives same answers",  {
-  # Generate data
-  set.seed(123)
-  x <- rnorm(10, 1, 0.15)
-  dummy <- capture.output(Bout <- Bnormal2(x, seed=345))  # default prior
[…]

### 2020-01-24 data/genthat_extracted_code/xpose4/examples/export.graph.par.Rd.R

@@ -1,27 +1,0 @@
-library(xpose4)
-
-
-### Name: export.graph.par
-### Title: Exports Xpose graphics settings to a file.
-### Aliases: export.graph.par xpose.write xpose.write
-### Keywords: methods
-
-### ** Examples
-
-
-## Not run: 
-##D ## xpdb5 is an Xpose data object
-##D ## We expect to find the required NONMEM run and table files for run
-##D ## 5 in the current working directory
-##D xpdb5 <- xpose.data(5)
-##D 
-##D ## For a filename prompt
-##D export.graph.par(xpdb5)
-##D 
-##D ## Command-line driven
-##D xpose.write(xpdb5, "c:/XposeSettings/mytheme.ini")
-## End(Not run)
-
-
-
-

### 2020-01-24 notebooks/data_showcase.Rmd
updated notebooks (by moving old ones into subdir), added new one, new code

---
title: "Data Showcase"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

The intent of this notebook is to showcase our current data analysis pipeline, and show some of the queries we indend to run.

```{r results='hide'}
source("../code/data_analysis_lib.R")
```

Load the data as `df`. `small_subset.csv` is around 50MB, so it should be manageable.

```{r results='hide'}
# This seems to be a fine size. We can run some easy stuff on it.
# The data is trimmed to only 20 arguments.
#
#
# 
# If the header isn't saved on the data:
# but you know how many arguments (or columns) are in the data, use
# either generate_header_for_num_args(num_args) or generate_col_names_for_count(num_columns)
df <- read_csv("../data/small_subset.csv")

# For some reason, the 
```

Now, here are some examples of the things that we might be interested in doing:

```{r}
# Get all `+` call traces:
#
# filter(fun_name == "+") returns all rows where the fun_name column is equal to "+".
#
# quick_trim_na_cols gets rid of columns entirely comprised of NAs. Useful when the function you're interested in
# doesn't have that many arguments, like `+`.
df %>% filter(fun_name == "+") %>% quick_trim_na_cols
```

Dealing with S3 dispatch:

```{r}
# How many function calls used the S3 dispatch mechanism?
df %>% filter(dispatch == "S3") %>% select(count) %>% sum -> num_S3

# You can also use some functions from the data_processing_lib imported at the top of this file.
# To get a breakdown of the dispatch usage in df, try:
df %>% breakdown_dispatch

# Takeaways:

# Breakdown by package:
df %>% breakdown_dispatch_by_pkg

# Takeaways:

# What classes were dispatched on?
# Note: Some of them might be {}, meaning no class. This is because whenever you call an S3 method, regardless of
#       the class of the first argument, it tries to dispatch. So we have cases where the class is empty ("{}")
#       which end up calling the generic (name.default) dispatch target.
df %>% breakdown_S3_dispatch_by_class

# Takeaways: most dispatch is done on data.frames, and POSIXct/POSIXt, then factors, then tibbles.
```

We also might be interested in instances of functions which have arguments which have multiple classes.

```{r}
# get_multi_classes adds columns of the form arg_cX_c, which has the # of classes in argument X of the function.
# 
df_multi <- df %>% get_multi_classes

# Breakdown by package::function:
df_multi %>% group_by(package, fun_name) %>% summarize
```

Will look into some of these and report results.

### 2020-01-24 notebooks/old_notebooks/paper_L123_results_section.Rmd
updated notebooks (by moving old ones into subdir), added new one, new code

--- 2019-04-02 notebooks/paper_L123_results_section.Rmd
@@
 
-# TODO: error -> ?
+# error -> ?
 los <- lapply(los, function(lofun) lapply(lofun, map_error_to_any))
+
+# ?? Attribute name, type to name only
+los <- lapply(los, function(lofun) lapply(lofun, attr_name_type_to_name_df))
 ```
@@
 xtable(L0_c_f_basic_df) %>% print(include.rownames=F)
-```
-
-DEBUG: Classes
-```{r}
-L0_poly_class_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_class_args))
-L0_poly_class_args <- L0_poly_class_args[order(-L0_poly_class_args$count),] # sort by count
-L0_poly_class_args$perc <- round(L0_poly_class_args$count/sum(L0_poly_class_args$count) * 100, 1)
-```
 ```

### 2020-01-24 notebooks/old_notebooks/paper_results_section.Rmd
updated notebooks (by moving old ones into subdir), added new one, new code

--- 2019-03-28 notebooks/paper_results_section.Rmd
@@
 los <- lapply(los, function(lofun) lapply(lofun, distinguish_matrices_df))
+# remove "errors"
+los <- lapply(los, function(lofun) lapply(lofun, remove_err_df))
 ```
@@
 los <- lapply(los, function(s) lapply(s, fold_NULL_and_NA_into_other_types_df))
-null_na_fold_arg_counts <- build_arg_count_df_from_lopkgs(los)
-null_na_fold_fun_counts <- build_fun_count_df_from_lopkgs(los)
+# null_na_fold_arg_counts <- build_arg_count_df_from_lopkgs(los)
+# null_na_fold_fun_counts <- build_fun_count_df_from_lopkgs(los)
 
@@
 
+Coindcidence stuff.
+```{r}
+index_list_coincide <- get_fun_sigs_matching_pred_from_lopkg(los, does_df_have_coinciding_types_index_and_list_or_vector)
+
+# how many index-y things are there in here?
+poly_type_args_i <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(list(index_list_coincide), extract_poly_type_args))
+poly_type_args_i <- poly_type_args_i[order(-poly_type_args_i$count),] # sort by count
+
+# in the reduced one:
+sum_indexy_in_lopoly(poly_type_args_i)
+
+# in full poly:
+sum_indexy_in_lopoly(poly_type_args)
+
+# perc
+sum_indexy_in_lopoly(poly_type_args_i) / sum_indexy_in_lopoly(poly_type_args)
+```
+
 Get the biggest signatures.
@@
 
-top_10_big_df %>% xtable
+top_10_big_df %>% xtable %>% print(include.rownames=F)
 
 ```
+
+Get biggest packages.
+```{r}
+packages_by_size <- get_biggest_packages(los, metric = size_of_signature_type)
+
+top_10_size_df <- data.frame(
+  pkg = names(packages_by_size[1:10]),
+  size = packages_by_size[1:10]
+)
+
+rownames(top_10_size_df) <- NULL
+
+top_10_size_df %>% xtable %>% print(include.rownames=F)
+```

### 2020-03-13 notebooks/data_showcase.Rmd
updated outline and some code

@@
 title: "Data Showcase"
-output: html_notebook
+output:
+  pdf_document: default
+  html_notebook: default
 editor_options:
@@
 ```{r results='hide'}
+library(tidyverse)
+library(tidyr)
+
 source("../code/data_analysis_lib.R")
@@
 
+Note that above, S3 calls with no class (represented by `{}`) end up calling the `functionName.default` method.
+
 We also might be interested in instances of functions which have arguments which have multiple classes.
@@
 # Breakdown by package::function:
-df_multi %>% group_by(package, fun_name) %>% summarize
+df_multi %>% group_by(package, fun_name) %>% summarize(count = sum(count)) %>% arrange(-count)
 ```
@@
 Will look into some of these and report results.
+These are functions which take arguments which have multiple classes.
+Need to figure something out about the FUN thing, which I believe is a function which is passed as an argument to a function.
+
+Find instances where classes co-occur with types:
+
+```{r}
+df %>% get_type_class_match -> df_types_class
+
+df_types_class %>% filter(class != "{}")
+```
+
+And instances where classes co-occur with attributes:
+
+```{r}
+df %>% get_class_attr_match -> df_class_attr
+
+df_class_attr %>% filter(class != "{}" & attributes != "{}")
+```
+
+To tackle polymorphism, we need to settle on a notion of type.
+At this stage, we have three separate sources of type information: the typeof (with extra fixings), the class, and the attributes.
+There's a function called `quick_simplify_types` (spelling?) which reduces the typeof++ column to a very simple type, but in truth we want to incorporate the class into the type when appropriate.
+
+```{r}
+# To get rid of dimensions and tags:
+df %>% quick_simplify_types -> df_simpl
+
+# The type_hash is no longer relevant, here.
+df_simpl %>% select(-type_hash) -> df_simpl
+
+# Might be worth grouping to reduce.
+df_simpl %>% group_by_at(setdiff(names(df_simpl), "count")) %>% summarize(count = sum(count)) -> df_gr
+```
+
+What other data do we want in the paper?
+
+One thing that's missing here: the call graph business. This we implemented in Haskell, but that might need some work.

### 2020-03-13 paper/main.tex
updated outline and some code

--- 2019-09-23 paper/main.tex
@@
-\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-\usepackage{booktabs,listings,xspace,wrapfig}
-\lstset{language=R}
+%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
+\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+%% For double-blind review submission, w/ CCS and ACM Reference
+%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
+%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
+%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+%% For single-blind review submission, w/ CCS and ACM Reference
+%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
+%% For final camera-ready submission, w/ required CCS and ACM Reference
+%\documentclass[acmsmall]{acmart}\settopmatter{}
+
+%% Journal information
+%% Supplied to authors by publisher for camera-ready submission;
+%% use defaults for review submission.
+\acmJournal{PACMPL}
+\acmVolume{1}
+\acmNumber{OOPSLA} % CONF = POPL or ICFP or OOPSLA
+\acmArticle{1}
+\acmYear{2018}
+\acmMonth{1}
+\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
+\startPage{1}
+
+%% Copyright information
+%% Supplied to authors (based on authors' rights management selection;
+%% see authors.acm.org) by publisher for camera-ready submission;
+%% use 'none' for review submission.
+\setcopyright{none}
+%\setcopyright{acmcopyright}
+%\setcopyright{acmlicensed}
+%\setcopyright{rightsretained}
+%\copyrightyear{2018}           %% If different from \acmYear
+
+%% Bibliography style
+\bibliographystyle{ACM-Reference-Format}
+%% Citation style
+%% Note: author/year citations are required for papers published as an
+%% issue of PACMPL.
+\citestyle{acmauthoryear}   %% For author/year citations
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Note: Authors migrating a paper from PACMPL format to traditional
+%% SIGPLAN proceedings format must update the '\documentclass' and
+%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%% Some recommended packages.
+\usepackage{booktabs}   %% For formal tables:
+                        %% http://ctan.org/pkg/booktabs
+\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
+                        %% http://ctan.org/pkg/subcaption
 \usepackage{my_style}
-\definecolor{LightGray}{rgb}{.95,.95,.95}
-\definecolor{Gray}{rgb}{.3,.3,.3}
-\definecolor{DarkGray}{rgb}{.5,.5,.5}
 
-\graphicspath{ {./plots/} }
-
-\lstset{ %
-  columns=flexible,
-  captionpos=b,
-  frame=single,
-  framerule=0pt,
-  framexleftmargin=-1mm,
-  framexrightmargin=-1mm,
-  tabsize=2,
-  belowskip=0pt,
-  basicstyle=\small\ttfamily,
-  backgroundcolor=\color{LightGray},
-  emphstyle=\sffamily,
-  keywordstyle=\bfseries,
-  commentstyle=\color{Gray}\em,
-  stringstyle=\color{Gray},
- % numbers=left
-}
-
-\lstdefinestyle{R}{ %
-  language=R,
-  deletekeywords={env, equal, c, runif, trace, args},
-  breaklines=true
-}
-
-\lstdefinestyle{Rin}{ %
-  style=R,
-  numberstyle=none,
-  basicstyle=\normalsize\ttfamily,
-  breaklines=false
-}
-
-\newcommand{\code}[1]{\lstinline|#1|\xspace}
-\newcommand{\genthat}{{\sc Genthat}\xspace}
-
-\setcopyright{none}
 \begin{document}
 
-\title{A Type System for R: Empirically Designed and Evaluated}
+%% Title information
+\title[]{Types for R, Empirically Designed and Evaluated}         %% [Short Title] is optional;
+                                        %% when present, will be used in
+                                        %% header instead of Full Title.
+% \titlenote{}             %% \titlenote is optional;
+                                        %% can be repeated if necessary;
+                                        %% contents suppressed with 'anonymous'
+% \subtitle{}                     %% \subtitle is optional
+% \subtitlenote{}       %% \subtitlenote is optional;
+                                        %% can be repeated if necessary;
+                                        %% contents suppressed with 'anonymous'
 
+%% Author information
+%% Contents and number of authors suppressed with 'anonymous'.
+%% Each author should be introduced by \author, followed by
+%% \authornote (optional), \orcid (optional), \affiliation, and
+%% \email.
+%% An author may have multiple affiliations and/or emails; repeat the
+%% appropriate command.
+%% Many elements are not rendered, but should be provided for metadata
+%% extraction tools.
+
+%% Author with single affiliation.
+\author{First1 Last1}
+\authornote{with author1 note}          %% \authornote is optional;
+                                        %% can be repeated if necessary
+\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+\affiliation{
+  \position{Position1}
+  \department{Department1}              %% \department is recommended
+  \institution{Institution1}            %% \institution is required
+  \streetaddress{Street1 Address1}
+  \city{City1}
+  \state{State1}
+  \postcode{Post-Code1}
+  \country{Country1}                    %% \country is recommended
+}
+\email{first1.last1@inst1.edu}          %% \email is recommended
+
+%% Author with two affiliations and emails.
+\author{First2 Last2}
+\authornote{with author2 note}          %% \authornote is optional;
+                                        %% can be repeated if necessary
+\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+\affiliation{
+  \position{Position2a}
+  \department{Department2a}             %% \department is recommended
+  \institution{Institution2a}           %% \institution is required
+  \streetaddress{Street2a Address2a}
+  \city{City2a}
+  \state{State2a}
+  \postcode{Post-Code2a}
+  \country{Country2a}                   %% \country is recommended
+}
+\email{first2.last2@inst2a.com}         %% \email is recommended
+\affiliation{
+  \position{Position2b}
+  \department{Department2b}             %% \department is recommended
+  \institution{Institution2b}           %% \institution is required
+  \streetaddress{Street3b Address2b}
+  \city{City2b}
+  \state{State2b}
+  \postcode{Post-Code2b}
+  \country{Country2b}                   %% \country is recommended
+}
+\email{first2.last2@inst2b.org}         %% \email is recommended
+
+%% Abstract
+%% Note: \begin{abstract}...\end{abstract} environment must come
+%% before \maketitle command
 \begin{abstract}
-
-\AT{PLDI TODO:}
-
-\begin{itemize}
-
-\item complete propagatr to generate constraints
-
-\item feed existing data through resulting constraints
-
-\item finish the annotation and contract generation code
-
-\item write the paper
-
-\end{itemize}
 
@@
 
+%% 2012 ACM Computing Classification System (CSS) concepts
+%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10011007.10011006.10011008</concept_id>
+<concept_desc>Software and its engineering~General programming languages</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10003456.10003457.10003521.10003525</concept_id>
+<concept_desc>Social and professional topics~History of programming languages</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering~General programming languages}
+\ccsdesc[300]{Social and professional topics~History of programming languages}
+%% End of generated code
+
+%% Keywords
+%% comma separated list
+\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
+
+%% \maketitle
+%% Note: \maketitle command must come after title commands, author
+%% commands, abstract environment, Computing Classification System
+%% environment and commands, and keywords command.
 \maketitle
 
-%
-%
-%
-%
-%
-%
 \section{Introduction}
 
-\begin{itemize}
-  \item Adding types to dynamic languages is increasingly common.
-  \item Types help write better code.
-  \item We want to design a type system for R that is based on the patterns that we observe in current use of the language. 
-  \item We also present a tool which takes a given package and synthesizes type signatures for package functions.
-  \item This tool is tested by synthesizing annotations from package tests, examples, and vignettes, and running the code of the package's reverse dependencies (i.e.
-clients) to validate the type signatures.
-  \item We find X, Y, Z.
-\end{itemize}
+The story that we're aiming for:
+Designing type systems for established languages can be a challenging.
+We aim to approach this problem by essentially reverse-engineering a type system from a corpora analysis of a representative subset of R code.
+We present a simple type system that captures the vast majority of usage of R --- it is lightweight, and \AT{should provide enough information to be useful to a compiler}.
 
-The contributions of this paper are:
+To evaluate our type system, we present a type-checker that is aware of R's promise semantics. 
+In R, function arguments are wrapped in promises such that they are not evaluated until they are needed.
+Our tool deals with these promises by inserting a check which fires only when the promise is forced, at which point the value is checked for compliance with the type.
 
-\begin{itemize}
+\AT{I can grab some of the intro Jan wrote for the last paper, rework it, and put it here.}
 
-\item our analysis data;
-\item a type system based on this data;
-\item a tool to infer types from package code, and;
-\item an evaluation of the tool on \AT{all of} CRAN.
+\section{Background}
 
-\end{itemize}
+\AT{I think a lot of the background we can pull from an older version of the paper.
+I think both are milling around this directory.}
 
-%
-%
-%
-%
-%
-%
-\section{The R Programming Language}
-\label{sec:rlang}
+\subsection{The R Programming Language}
 
-\AT{This can probably be copied from the old version.
-Is this somewhere we want to talk about R-dyntrace?}
+\AT{Introduce the language, and give some context about its user base.}
 
-%
-%
-%
-%
-%
-%
-\section{The Corpus}
-\label{sec:corpus}
+\subsection{R-Dyntrace}
 
-\AT{Probably talk about CRAN, and give some stats on the corpus that we worked with. 
-This section probably doesn't have to be very long.}
+\AT{Worth talking a bit about the analysis framework?}
 
-\AT{Fit this in here:}
-Unfortunately, there is not a whole lot of end-user R code available for analysis; typical R programs are effectively data pipelines, and many R users elect not to share their data, possibly because the data is proprietary or somehow confidential (e.g., census data), and these data pipeline programs are not very useful without the underlying data.
-To alleviate this, we focused on the analysis of package test, example, and vignette code.
+\subsection{The Corpus}
 
-\AT{How do we want to argue that our data is valid and representative?}
-End-user R scripts are hard to come by.
-R is a data science language first-and-foremost, and so R programs are typically data analyses.
-The nature of data is that, in many scenarios, it is confidential; imagine a government sharing sensitive information publicly, or a company divulging customer data.
-For this reason, it is rare for R programmers to share their code, and even if they did, the code would not be particularly useful without the data underlying it.
+\AT{This section needs to be tight.
+Reviewers were skeptical of CRAN and using the test and example code.
+We should convince them otherwise.}
 
-Separately, idiomatic usage of R is to appeal to libraries to implement the main functionality of an analysis.
-\AT{Probably want some good evidence of this.}
-R packages are mainly sourced from the Comprehensive R Archive Network (CRAN), a large curated repository of R packages (libraries).
-Of particular relevance to this work is a requirement that all packages should ship with tests, examples, and vignette programs to showcase how packages are intended to be used---indeed, these serve as a starting point for potential package clients.
-We can leverage these programs to build up a picture of how programmers interact with packages, and thus of how programmers write R code for their own analyses.
+\section{Methodology}
 
-\AT{I think that talking avout how ``this is as good as it gets guys'' might be a bit of a defeatist way of looking at it.
-But then again, I'm not sure what a good balance would be here.}
+\subsection{Initial Analysis}
 
-%
-%
-%
-%
-%
-%
-\section{Overview}
+\AT{Lay out the analysis, spell out the data we collect, how we reduce the data.
+Not sure how much detail to go into here.
+Will have had to have talked about the analysis framework in the background.}
 
-\AT{Somewhere, we need to overview the contribution, talking about the type system, about the tool, and about the data.}
+\subsection{Implementation}
 
-Broadly, the contributions of this work can be divided into two.
+\AT{Talk about how the type system was implemented, in terms of the checks, how they work, how they are performed by the runtime, etc.}
 
-First, we present a {\it large-scale analysis} of the type-use programming practices in R.
-As mentioned in Section~\ref{sec:corpus}, we focused our efforts on package test, example, and vignette code.
-Our tracing framework collects type information of function arguments and return values, and the precise details are elaborated in Section~\ref{sec:methodology}.
+\subsection{Evaluation}
 
-Unfortunately, in spite of the requirements imposed on CRAN packages, some package functions are not thoroughly tested and are poorly exercised.
-Previous work in a similar vein (\AT{Genthat}) reported that their tool achieved roughly 70\% coverage of package code, which is good but far from perfect (and our tool \AT{is liable to achieve the same coverage}).
-In order to further improve our data, we perform an additional phase of type propagation, where we use known type signatures of inner functions to improve the signatures of outer functions.
-Essentially, we build a graph of the flows of argument values throughout a function, and infer constraints from that graph and feed our data into it; the precise details will be discussed further in Section~\ref{sec:methodology}.
-The data itself will be presented in Section~\ref{sec:results}.
+\AT{Talk about Aviral's work on evaluation.}
 
-Besides helping us understand the R language, we also used the data to inform the {\it design of a type system} for the language.
-Based on the common patterns observed in the analysis, we devised a type system for the language that \AT{maximizes the ``monomorphicity'' in the data} while capturing many of the idioms we observed.
-For example, we often ran across function arguments which were exclusively inhabited by unit-length vectors of some primitive type; this led us to add a scalar version of primitive types to the type system, and include a rule akin to scalar\_primitive <: vector\_primitive to ensure that unit-length vectors are treated as such when there is evidence that the function is intended to accept vectors.
-This will be elaborated further in Section~\ref{sec:results}.
+\section{Type System Design}
 
-Now, in order to {\it validate} this type system, we implemented a framework for annotating R programs, as well as a tool which infers types from package example code, inserts lightweight runtime contracts into the code, and runs the example code in the package's reverse dependencies to try to unearth type errors.
-The annotations are lightweight as they are only run {\it the first time} an argument is used, if it is used at all.
-Details on and the results of the validation are available in Section~\ref{sec:results}.
-Details on the process itself will be given in the next section.
+\AT{Lay out the type system.
+Ideally, for every interaction we propose, we should be able to point to an analysis that supports our claim.
+Maybe we don't always have to do this, especially for some of the simple, obvious types, but any departures from basic R functionality should be supported by the data.}
 
-%
-%
-%
-%
-\subsection{Annotations}
-\label{subsec:annotations}
+\section{Evaluation Results}
 
-We extended the R language to support annotations on function arguments and returns.
-This involved reserving syntax for the annotations, and modifying the R parser to parse and accept them.
-\AT{What other stuff do we want to add to this? Other than presenting the syntax and stuff.}
-The syntax allows programmers to specify the basic type, class, attributes, length, and parameterized types of arguments.
-Also, we support type variables so as to allow programmers to specify large but common type signatures one time only.
+\AT{The decisions we made when coming up with the types will be validated in the last section.
+This section is for determining how effective the type system is.
+We have two evaluation strategies currently: comparing against reverse dependencies, and comparing against the Kaggle programs.}
 
-To illustrate the annotation syntax, consider the code in Figure~\ref{fig:annotation-syntax}.
-\AT{TODO: This figure.}
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-my_fun <- function(@{int} x) {
-  ...
-}
-\end{lstlisting}}\caption{Type annotation syntax.}\label{fig:annotation-syntax}\end{figure}
-
-%
-%
-%
-%
-%
-%
-\section{Methodology}
-\label{sec:methodology}
-
-\AT{I think there are two distinct methodologies.}
-
-\AT{First, we had to come up with the type system, and it partly based on things that we saw from our analysis.}
-
-\AT{Second, we should talk about the evaluation of the tool.}
-
-%
-%
-%
-%
-\subsection{Typr}
-
-\AT{Are we going to change this? To make it run off of R-dyntrace?}
-
-The first tool in our pipeline is Typr.
-The goal of Typr is to gather type information on function invocations, and it is a modification of the Genthat~\cite{issta18} tool.
-Genthat records all function invocations and builds a standalone trace to allow for independent reproduction of the call. 
-We modified Genthat to instead record a wealth of information on argument and return values.
-Typr records:
-
-\begin{itemize}
-\item the results of calling {\tt typeof}, {\tt class}, and {\tt attributes} on the argument;
-\item the length (if the argument has a length);
-\item if the argument is a list or data.frame we additionally reflect on the types of argument members;
-\item if the argument is an environment, we reflect on the bindings in the environment.
-\end{itemize}
-
-The precise details of the approach are as follows:
-
-\begin{enumerate}
-\item for every package that is to be run, all functions (including those inherited from dependencies) are instrumented with on.exit hooks, code which runs when the function is finished executing.
-The purpose of waiting for function exit is to ensure that argument promises are not forced, and argument values which are unused are recorded as such, as this indicates that they are likely used for some metaprogramming within the function.
-\item package tests, examples, and vignettes are run, and decorated function invocations trigger the on.exit hook, which records argument and return value type information.
-\item each new function trace (i.e., each call with a new call signature) is saved.
-\end{enumerate}
-
-\AT{At what point do we introduce the type system?
-I was gonna mention subtyping rules in a hypothetical item 4 where we consolidate trace info.
-Maybe we should just talk about how we looked at this data, and then add a new phase after?}
-
-%
-%
-%
-%
-\subsection{Propagatr}
-
-\AT{Explaining this succinctly will be important.}
-After collecting and consolidating type information for function calls, we use our Propagatr tool to propagate known type information between functions that are used in similar circumstances.
-In essence, our tool runs over the tests, examples, and vignettes, much like our Typer tool, but instead of collecting type information it tracks values as they flow through code and builds a graph connecting use-sites of values to the argument positions that they are drawn from.
-\AT{Man, this sucks.}
-
-For example, consider the code snippet in Figure~\ref{fig:propagatr-example}.
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-f <- function(x, y) {
-  z <- x
-  g(z, x)
-}
-
-f(0, 42)
-\end{lstlisting}}\caption{Propagatr example.}\label{fig:propagatr-example}\end{figure}
-
-Propagatr tracks the following data flow:
-
-\begin{itemize}
-\item \AT{This might be better achieved as a graph, or something.}
-\item the assignment of x to z on line 2 is actually a function call in R; think of it as {\tt assign(x, z)}.
-The flow of data here is from x to z, and thus in the remainder of the function both variables are known to share the same value.
-\item The call to g(z, x) allows us to deduce that g takes two arguments {\it of the same type}. 
-\end{itemize}
-
-An important purpose of the Propagatr tool is to allow us to infer type signatures for poorly-exercised functions.
-As a (somewhat degenerate) example, consider the code in Figure~\ref{fig:propagatr-degenerate}.
-\AT{maybe shove this example somewhere to save space}
-
-\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
-add_wrapper <- function(a, b) { a + b }
-\end{lstlisting}}\caption{Degenerate Propagatr example.}\label{fig:propagatr-degenerate}\end{figure}
-
-Even if {\tt add\_wrapper} is rarely exercised, Propagatr deduces that it should have the same signature as R's builtin addition operator.
-
-The precise details of the tool follow.
-\AT{Pseudocode? Something like that? A formal description of the algorithm.}
-
-%
-%
-%
-%
-\subsection{Infr}
-
-Our Infr tool leverages our type system to test the effectiveness of package test, example, and vignette code at covering the possible use-cases of package functions.
-
-The approach is as follows:
-
-\begin{itemize}
-\item Typr and Propagatr are run on the test, example, and vignette code of a particular package.
-This serves to infer an annotation for all package functions.
-\item Infr turns these annotations into dynamically-checked contracts on functions:
-The contracts check that function argument and return values adhere to the type of their annotation.
-These contracts are inserted directly into the promise code of the values, as will be elaborated in Section~\ref{subsubsec:promise}.
-\item Then, the test, example, and vignette code for all reverse package dependencies (i.e., for all other packages which have the current package as a dependency) is run, and any type errors that are caught are recorded.
-\end{itemize}
-
-\AT{Do we want to do something special about data frames? We sort of already infer their type from the dynamic analysis.}
-
-%
-%
-\subsubsection{I Promise I'll Do the Check!}
-\label{subsubsec:promise}
-
-As mentioned in Section~\ref{sec:rlang}, arguments to R functions are wrapped in promises whose code is only executed if the argument is used.
-To insert checks that do not disrupt these semantics, we insert our checks {\it directly into the code of the promise}, which is done in C++ leveraging the Rcpp framework.
-
-Essentially, the value of the e.g. argument promise is replaced by code which checks for compliance with the argument's annotation.
-If the check fails, a call to R's stop function is triggered, yielding an informative error message (which includes the value itself, the function, package, and type that the value failed to adhere to).
-If successful, the checking code is removed (by reassigning to the argument within the function), so subsequent accesses to the argument are fast.
-\AT{Can return values be wrapped in the same way?}
-
-%
-%
-%
-%
-%
-%
-\section{Results}
-\label{sec:results}
-
-\AT{What should go in here?}
-
-\AT{There's the data itself, and our breakdown of it.
-There's also the results of validating the types and the tool against the reverse dependencies.}
-
-%
-%
-%
-%
-%
-%
 \section{Related Work}
 
-\AT{Look at the stuff we put in the old version of the paper.}
+\AT{Good to mention the related projects, e.g. how this was undertaken in other languages, like Ruby.
+There's a lot of literature in that space I think, I've read some of it but not all.
+We also have some related work in the old paper, I'll grab it and put it here.}
 
-%
-%
-%
-%
-\subsection{Studies of Dynamic Languages}
-
-The idea of measuring polymorphism via some program analysis is not new.
-For instance, some work~\cite{aakerblom2015measuring} measures the polymorphism of Python programs, but with a slightly different goal than ours: authors seek to assess the need for expressiveness in type systems that are retrofit onto dynamic languages.
-Interestingly, they find that 4\% of function call sites are polymorphic, which is considerably less than what we have found with our analysis of R.
-In essence, it would appear that R is far more polymorphic in Python.
-
-More generally, there is an existing literature on analyzing usage patterns of language features.  
-For instance, the dynamic features of Smalltalk were analyzed in some work~\cite{callau2011howdevelopers}, which aimed to see how often the highly dynamic capabilities of Smalltalk were used and in which kinds of projects.
-Other work~\cite{pldi10a} explores how the dynamic features of JavaScript are used.
-An interesting finding of theirs is that only 81\% of calls are monomorphic, which is very similar to what we observed in R.
-This 80/20 split between monomorphism and polymorphism is echoed in other analysis work on the usage patterns of duck typing in Smalltalk~\cite{milojkovic2017duck}, which finds that ~20\% of functions are ``duck-typed'' (\AT{read: polymorphic}) in Smalltalk.
-\AT{It's notable that this 20\% shows up very often.}
-
-%
-%
-%
-%
-\subsection{Synthesizing Types from Dynamic Analysis}
-
-In this work, we have retrofit a type system onto the R programming language.
-We based many of the decisions on which types and which rules to include on the results of our analysis.
-This shares some similarity with work on so-called {\it trace typing}~\cite{andreasen2016trace}.
-
-%
-%
-%
-%
-\subsection{Type Systems for Dynamic Languages}
-
-The type system introduced in this work is a {\it gradual type system}.
-
-%
-%.
-%
-%
-%
-%
 \section{Conclusion}
 
-%
-%
-%
-\bibliographystyle{boilerplate/ACM-Reference-Format}
-\bibliography{bib/biblio,bib/jv,bib/r,bib/new,bib/gradual,bib/types4R}
+\ldots
+
+\subsection{Threats to Validity}
+
+\ldots
+
+\subsection{Future Work}
+
+\ldots
+
+%% Acknowledgments
+\begin{acks}                            %% acks environment is optional
+                                        %% contents suppressed with 'anonymous'
+  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
+  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
+  %% acknowledge financial support and will be used by metadata
+  %% extraction tools.
+  This material is based upon work supported by the
+  \grantsponsor{GS100000001}{National Science
+    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
+  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
+  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
+  conclusions or recommendations expressed in this material are those
+  of the author and do not necessarily reflect the views of the
+  National Science Foundation.
+\end{acks}
+
+%% Bibliography
+%\bibliography{bibfile}
+
+%% Appendix
+\appendix
+\section{Appendix}
+
+Text of appendix \ldots
+
 \end{document}

### 2020-03-17 paper/main.tex
working on analysis details section

@@
 \usepackage{my_style}
+
+\usepackage{listings}
+
+\lstset{ 
+  language=R,                     % the language of the code
+  basicstyle=\tiny\ttfamily, % the size of the fonts that are used for the code
+  numbers=left,                   % where to put the line-numbers
+  numberstyle=\tiny\color{blue},  % the style that is used for the line-numbers
+  stepnumber=1,                   % the step between two line-numbers. If it is 1, each line
+                                  % will be numbered
+  numbersep=5pt,                  % how far the line-numbers are from the code
+  backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
+  showspaces=false,               % show spaces adding particular underscores
+  showstringspaces=false,         % underline spaces within strings
+  showtabs=false,                 % show tabs within strings adding particular underscores
+  frame=single,                   % adds a frame around the code
+  rulecolor=\color{black},        % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. commens (green here))
+  tabsize=2,                      % sets default tabsize to 2 spaces
+  captionpos=b,                   % sets the caption-position to bottom
+  breaklines=true,                % sets automatic line breaking
+  breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
+  keywordstyle=\color{teal},      % keyword style
+  commentstyle=\color{olive},   % comment style
+  stringstyle=\color{magenta}      % string literal style
+}
+
+\newcommand{\code}[1]{\text{\lstinline[basicstyle=\ttfamily\smaller]~#1~}}
 
@@
 
-\subsection{R-Dyntrace}
+\begin{lstlisting}
+hello <-  "world"
+paste(c("this", "is", "some", "code"))
+\end{lstlisting}
+
+\subsection{R-dyntrace}
+\label{sec:r-dyntrace}
 
@@
 
+\AT{Some sentences that might be useful:}
+We selected packages that have both \AT{a good amount of} reverse dependencies to ensure their broad usefulness, as well \AT{sufficiently high} code and branch coverage of test, example, and vignette programs to ensure that our analysis built up an accurate picture of function use.
+
+%
+%
+%
+%
+%
+%
 \section{Methodology}
 
+%
+%
+%
+%
 \subsection{Initial Analysis}
 
-\AT{Lay out the analysis, spell out the data we collect, how we reduce the data.
-Not sure how much detail to go into here.
-Will have had to have talked about the analysis framework in the background.}
+%
+%
+\subsubsection{Overview}
+
+In order to ensure that our type system design aligns with the day-to-day usage of R, we performed a corpus analysis of some of the most widely used R packages.
+We built our dynamic analysis in the R-dyntrace \AT{cite something?} dynamic analysis framework, which is itself built on a modified version of R targeting R version 3.5.0---we refer the reader to Section~\ref{sec:r-dyntrace} for more details on R-dyntrace.
+
+The general idea of the analysis is to intercept calls to functions, builtins, and specials, and collect information on function arguments and returns.
+The nature of R is such that all function arguments are wrapped in promises which are only evaluated when the value is actually used.
+This could cause issues in a dynamic analysis, as careless use of function arguments can force them early, which can cause unindented side-effects (as these promises are not necessarily pure).
+Our analysis deals with this by inspecting the {\it value} slot of these promise objects, which does not force the promise. 
+If the value slot contains another promise, we look into its value slot, and so forth until we find a value (or not! in which case, the value is a promise).
+
+Once the value is obtained in this way, we collect the following high-level information about it:
+\AT{TODO: need to make the inline code macro and the R listing, and change the tts in here}
+
+\begin{itemize}
+\item the result of calling the {\tt typeof} function on the value; in other words, the type that the value has according to R's runtime;
+\item its class, which is a list of string class names;
+\item its attributes, a list of metadata attached to the value.
+\end{itemize}
+
+Now, depending on the ``type'' of the value, we collect further information:
+If the value is \ldots
+
+\begin{itemize}
+\item one of R's primitive types, we collect its dimension (and if it is a matrix, its dimensions), and if it has a names attribute we collect the names as well;
+\item a list, we recursively collect the types of list elements;
+\item a list {\it with a names attribute}, we collect the names as well as the types of the slots that those names refer to (data.frames fall into this category);
+\item a closure, we simply tag it with a UID computed from the code of the closure as well as its name, which can be later cross-referenced with the analysis results to get a type for the function.
+\end{itemize}
+
+Some further miscellaneous information we collect includes:
+
+\begin{itemize}
+\item lists which do not contain any NULL elements are qualified with a NULL-free tag; 
+\item similarly, vectors which contain no NAs are qualified with a NA-free tag;
+\item promises which are never forced throughout function execution are tagged as being {\it metaprogrammed}---this is a common practice in the {\tt tidyverse} suite of packages;
+\item formal arguments (arguments which are named in the function definition) which are not passed are recorded as {\it missing}---these represent optional arguments.
+\end{itemize}
+
+Now that we've seen the high-level picture of our approach, we will elaborate on some key details on our use of the R-dyntrace framework, and some of the finer points of the analysis.
+
+%
+%
+\subsubsection{Select Details}
+
+As mentioned, our analysis is built on the R-dyntrace framework.
+R-dyntrace is essentially a very efficient instrumentation framework for R, which is built on a modified version of the R runtime.
+Fortunately, our dynamic type analysis only needs to be run once, ever---from the analysis results, we synthesize function signatures which are passed along to later type-checking tools which do not rely on a heavily modified runtime.
+
+We primarily rely on 6 R-dyntrace callbacks: \texttt{closure\_entry}, \texttt{closure\_exit}, \texttt{builtin\_entry}, \texttt{builtin\_exit}, \texttt{special\_entry}, and \texttt{special\_exit}.
+These callbacks fire when closures, builtin, and special functions are entered and exited.
+In the \texttt{*\_entry} variants, we do some bookkeeping work: the analysis is notified that a closure/builtin/special has been entered by pushing the call onto a stack.
+Further, the type of dispatch is recorded: \AT{recall} that there are multiple ways that R performs dynamic dispatch, based on the class of function arguments.
+We collect dispatch information in order to analyze dispatch usage patterns.
+
+The main work is done in \texttt{*\_exit} variants.
+As arguments to functions in R are wrapped in promises, we delay our reflection to function exit time. 
+\AT{Actually, why do we do this?}
+In R, promises are essentially objects with value and expression slots.
+When a promise is created for a particular expression, that expression is merely put into the expression slot of a promise object, and a special {\it unbound} value is loaded into the value slot.
+When the promise is forced, the expression is evaluated in the forcing context, and the value is stored in the value slot of the promise.
+So to get the type of a function argument, we check the value slot of the argument promise---perhaps recursively if the value itself is a promise.
+If we reach the unbound value, we know the promise was never forced, and thus the argument was unused.
+
+To construct a ``type'' for each argument, we make use of R's C FFI and use low-level machinery to collect type tags and attributes from the R runtime.
+At this level, all R values are {\tt SEXP}s with a type tag that can be obtained using the {\tt TYPEOF} function from the C FFI.
+In certain circumstances, such as when {\tt TYPEOF} indicates that a value is a primitive, we perform additional reflection e.g.
+on the length (with {\tt LENGTH}) and also vector element names (through a call to {\tt getAttrib(the\_value, R\_NamesSymbol)}).
 

### 2020-04-01 paper/main.tex
adjusted discussion to reflect new type collection strategy

@@
 This could cause issues in a dynamic analysis, as careless use of function arguments can force them early, which can cause unindented side-effects (as these promises are not necessarily pure).
-Our analysis deals with this by inspecting the {\it value} slot of these promise objects, which does not force the promise. 
-If the value slot contains another promise, we look into its value slot, and so forth until we find a value (or not! in which case, the value is a promise).
+Our analysis deals with this by only inspecting the value of an argument when the argument is forced, and if it is never forced we make note of that by ascribing the type {\tt ???}.
+
+% Our analysis deals with this by inspecting the {\it value} slot of these promise objects, which does not force the promise. 
+% If the value slot contains another promise, we look into its value slot, and so forth until we find a value (or not! in which case, the value is a promise).
 
@@
 As mentioned, our analysis is built on the R-dyntrace framework.
-R-dyntrace is essentially a very efficient instrumentation framework for R, which is built on a modified version of the R runtime.
-Fortunately, our dynamic type analysis only needs to be run once, ever---from the analysis results, we synthesize function signatures which are passed along to later type-checking tools which do not rely on a heavily modified runtime.
+R-dyntrace is a very efficient instrumentation framework for R, which is built on a modified version of the R runtime.
+Our dynamic type analysis only needs to be run once, ever---from the analysis results, we synthesize function signatures which are passed along to later type-checking tools which do not rely on a heavily modified runtime.
 
-We primarily rely on 6 R-dyntrace callbacks: \texttt{closure\_entry}, \texttt{closure\_exit}, \texttt{builtin\_entry}, \texttt{builtin\_exit}, \texttt{special\_entry}, and \texttt{special\_exit}.
-These callbacks fire when closures, builtin, and special functions are entered and exited.
-In the \texttt{*\_entry} variants, we do some bookkeeping work: the analysis is notified that a closure/builtin/special has been entered by pushing the call onto a stack.
-Further, the type of dispatch is recorded: \AT{recall} that there are multiple ways that R performs dynamic dispatch, based on the class of function arguments.
+We primarily rely on 8 R-dyntrace callbacks: \texttt{closure\_entry}, \texttt{closure\_exit}, \texttt{builtin\_entry}, \texttt{builtin\_exit}, \texttt{special\_entry},\texttt{special\_exit}, \texttt{promise\_force\_entry}, and \texttt{promise\_force\_exit}.
+These callbacks fire when closures, builtin, and special functions are entered and exited, and before and after the forcing of promises.
+The function-related callbacks are used mainly for bookkeeping: the analysis is notified that a closure/builtin/special has been entered/exited by pushing/popping the call onto a stack.
+The calls themselves store a {\it call trace} object, where we store the type information that we collect.
+Further, the type of dispatch is recorded on the {\tt \_entry} variants: \AT{recall} that there are multiple ways that R performs dynamic dispatch, based on the class of function arguments.
 We collect dispatch information in order to analyze dispatch usage patterns.
 
-The main work is done in \texttt{*\_exit} variants.
-As arguments to functions in R are wrapped in promises, we delay our reflection to function exit time. 
-\AT{Actually, why do we do this?}
+The main work is done in {\tt promise\_force\_exit}.
+As arguments to functions in R are wrapped in promises, we delay our reflection to promise force time.
 In R, promises are essentially objects with value and expression slots.
@@
 So to get the type of a function argument, we check the value slot of the argument promise---perhaps recursively if the value itself is a promise.
-If we reach the unbound value, we know the promise was never forced, and thus the argument was unused.
+On function {\tt \_exit} variants, we deal with situations where an argument was never used (and thus never forced): these show up as missing slots in the call trace object, and we ascribe them the type {\tt ???}.
 

### 2020-05-06 paper/main.tex
pushing current paper version, with comments. im laying out the narrative surrounding the type design decisions

@@
 
-\usepackage{listings}
+\usepackage{listings, wrapfig}
 
@@
 \newcommand{\code}[1]{\text{\lstinline[basicstyle=\ttfamily\smaller]~#1~}}
+
+% Macros for type names from the old paper.
+\newcommand{\attr}[2]{\ensuremath{#1_{\mathtt{#2}}}\xspace}
+\newcommand{\attrclass}[3]{\ensuremath{#1^{\mathtt{#3}}_{\mathtt{#2}}}\xspace}
+\renewcommand{\to}{\ensuremath{\rightarrow}\xspace}
+\newcommand{\D}{\ensuremath{\small\vec{\mathtt D}}\xspace} % Double
+\newcommand{\I}{\ensuremath{\small\vec{\mathtt I}}\xspace} % Integer
+\renewcommand{\C}{\ensuremath{\small\vec{\mathtt C}}\xspace} % Character
+\renewcommand{\L}{\ensuremath{\small\vec{\mathtt L}}\xspace} % Logical
+\newcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
+\newcommand{\X}{\ensuremath{\small\vec{\mathtt X}}\xspace} % Complex
+\newcommand{\Y}{\ensuremath{\small\vec{\mathtt Y}}\xspace} % Symbol
+\newcommand{\sY}{\ensuremath{\small{\mathtt Y}}\xspace} % Symbol
+\newcommand{\sS}{\ensuremath{\small{\mathtt S}}\xspace} % S4
+\newcommand{\sF}{\ensuremath{\small{\mathtt F}}\xspace} % Closure
+\newcommand{\sE}{\ensuremath{\small{\mathtt E}}\xspace} % Env
+\renewcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
+\newcommand{\sN}{\ensuremath{\small{\mathtt N}}\xspace}     % Null
+%\renewcommand{\l}{\ensuremath{\small L<?>}\xspace}     % List
+\renewcommand{\l}{\ensuremath{\small\underline{\mathtt ?}}\xspace}     % List
+\newcommand{\sD}{\ensuremath{\small{\mathtt D}}\xspace} % Double
+\newcommand{\sI}{\ensuremath{\small{\mathtt I}}\xspace} % Integer
+\newcommand{\sC}{\ensuremath{\small{\mathtt C}}\xspace} % Character
+\newcommand{\sL}{\ensuremath{\small{\mathtt L}}\xspace} % Logical
+\newcommand{\sX}{\ensuremath{\small{\mathtt X}}\xspace} % Complex
+\newcommand{\sR}{\ensuremath{\small{\mathtt R}}\xspace} % Raw
+\newcommand{\ANY}{\ensuremath{\small{\mathtt ?}}\xspace}     % Any
+\newcommand{\lT}[1]{\ensuremath{\small\underline{\mathtt{#1}}}\xspace}     % list<T>
+\newcommand{\M}[1]{\ensuremath{\attr{\vec{\tt #1}}{mat}}\xspace}     % matrix
+\newcommand{\df}{\ensuremath{\attr{\l}{df}}\xspace}     % data.frame
 
@@
 
+%
+%
+%
+%
+%
+%
 \section{Background}
 
-\AT{I think a lot of the background we can pull from an older version of the paper.
-I think both are milling around this directory.}
+In this section, we will introduce the R programming language, as well as discuss the dynamic analysis framework we used as part of our initial analysis.
 
+%
+%
+%
+%
 \subsection{The R Programming Language}
 
-\AT{Introduce the language, and give some context about its user base.}
+Over the last decade, the R Project has become a key tool for implementing
+sophisticated data analysis algorithms in fields ranging from Computational
+Biology~\cite{R05} to Political Science~\cite{R:Keele:2008}. At the heart of
+the R project is a \emph{vectorized, dynamic, lazy, functional,
+  object-oriented} programming language with a rather unusual combination of
+features~\cite{ecoop12} designed to ease learning by non-programmer and
+enable rapid development of new statistical methods.  The language, commonly
+referred to as R was designed in 1993 by Ross Ihaka and Robert
+Gentleman~\cite{R96} as a successor to S~\cite{S88}.  First released in
+1995, under a GNU license, R rapidly became the lingua franca for
+statistical data analysis. Today, there are over 13,000 R packages available
+from repositories such as CRAN and Bioconductor.  With 55 R user groups
+world-wide, Smith~\cite{eco11} estimates that there are over 2 million
+end-users.
+
+As an introduction to R, consider the code snippet in Fig.~\ref{sample} from
+a top-level interaction where the user defines a function \code{normSum}
+that accepts vectors of integers, logicals, doubles and complex values and
+normalizes the vector with respect to its sum and rounds the results. The
+function definition does not require type annotations, and all operations
+transparently work on vectors of any length and different types.
+
+\begin{figure}[!hb]{\small
+\begin{lstlisting}
+> normSum <- function( m )  round( m / sum(m), 2)
+> normSum(c(1L,3L,6L))
+[1] 0.1 0.3 0.6
+> normSum(c(1.1,3.3,6.6))
+[1] 0.1 0.3 0.6
+> normSum(c(1.6,3.3,6.1))
+[1] 0.15 0.30 0.55
+> normSum(complex(r=rnorm(3),i=rnorm(3)))
+[1] 0.49+0.21i 0.30-0.18i 0.22-0.03i
+\end{lstlisting}}
+\caption{Sample R code}\label{sample}
+\end{figure}
+
+In R, function can be called with named parameters, R support variable
+argument lists, and arguments can have default values. Putting all of these
+together consider the following declaration:
 
 \begin{lstlisting}
-hello <-  "world"
-paste(c("this", "is", "some", "code"))
+f <- function(x, ..., y=3) x + y
 \end{lstlisting}
 
+\noindent
+Function \k{f} can be called with a single argument \code{f(3)}, with named
+argument \code{f(y=4,x=2)} and with a variable number of arguments,
+\code{f(1,2,3,4,y=5)}, all of these calls will return \code{6}.
+
+R has a number of features that are not crucial to the present
+discussion. We will mention some of them here for completeness.  In R, data
+structures are reference counted and have copy-on-write semantics, thus the
+assignment \code{x[12]<-3} results in an update to a copy of \code{x} unless
+the reference count on that object is 1.  This semantics gives R a
+functional flavor while allowing updating in place within loops (the first
+update copies, subsequent updates are performed on the copy). Arguments to
+functions are evaluated only when needed, they are bundled in so-called
+promises which package the original expression (as an abstract syntax tree, or AST), its environment
+as well as the result of evaluating the expression. Promises can be
+leveraged for meta-programming as it is possible to retrieve the text of a
+promise and evaluate it in a different environment.
+
+\subsubsection{Types of Data}
+
+Before attempting to define a type system for R, we should understand the
+different kinds of values that programs operate on.  As we will see
+different notions of type may emerge depending on how granular we want to
+be.
+
+\renewcommand{\k}[1]{{\tt #1}\xspace}
+
+R has one builtin notion of type that can be queried by the \k{typeof}
+function. Over the years, programmers have found the need for a richer type
+structure and have added {\it attributes}. The best way to think of attributes is
+as an optional map from name to values that can be attached to any object.
+Attributes are used to encode various type structures. They can be queried
+with functions such as \k{attributes} and \k{class}.
+
+\begin{wrapfigure}{r}{6.1cm}
+\footnotesize\begin{tabular}{l|c|l@{}}\hline
+\multicolumn{3}{l}{\bf Vectorized data types:}  \\\hline
+\k{logical}   & \L & vector of boolean values\\
+\k{integer}   & \I & vector of 32 bit integer values\\
+\k{double}    & \D & vector of 64 bit floating points\\
+\k{complex}   & \X & vector of complex values\\
+\k{character} & \C & vector of strings values\\
+\k{raw}       & \R & vector of bytes\\
+\k{list}      & \l & vector of values of any type\\\hline
+\multicolumn{3}{l}{\bf Scalar data types:}\\\hline
+\k{NULL}      & \sN &  singleton null value\\
+\k{S4}        & \sS &  instance of a S4 class \\
+\k{closure}   & \sF & a function with its environment\\
+\k{environment}&\sE &  a mapping from symbol to value \\\hline
+\multicolumn{3}{l}{\bf Implementation data types:}\\\hline
+\multicolumn{3}{l}{\k{special},
+\k{builtin},
+\k{symbol} (\sY),
+\k{pairlist},
+\k{promise}}\\
+\multicolumn{3}{l}{
+\k{language},
+\k{char},
+\k{...},
+\k{any},
+\k{expression},
+}\\
+\multicolumn{3}{l}{
+\k{externalprt},
+\k{bytecode},
+\k{weakref}}\\\hline
+\end{tabular}\caption{Builtin Types}\label{types}\end{wrapfigure}
+
+Figure~\ref{types} lists all of the builtin types that are provided by the
+language. They are the possible return values of \k{typeof}. There is no
+intrinsic notion of subtyping in R. But, in many context a \k{logical} will
+convert to \k{integer}, and an \k{integer} will convert to \k{double}.  Some
+off conversion can occur in corner cases, such as \k{1<"2"} holds and
+\k{c(1,2)[1.6]} returns the first element of the vector, as the double is
+converted to an integer. R does not distinguish between scalars and vectors
+(they are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
+  typeof(c(5,5))} \code{ == "double"}. Finally all vectorized data types have a
+distinguished missing value denoted by \code{NA}. The default type of
+\code{NA} is \k{logical}. We can see that \code{typeof(NA)=="logical"}, but
+NA inhabits every type: \code{typeof(c(1,NA)[2])=="double"}.
+
+With one exception all vectorized data types are monomorphic, the exception
+is the \k{list} type which can hold values of any other type including
+\k{list}. For all monomorphic data types, attempting to store a value of a
+different type will cause a conversion. Either the value is converted to the
+type of vector, or the vector is converted to the type of the value.
+
+Scalar data types include the distinguished \k{NULL} value, which is also of
+type \k{NULL}, instance of classes written using the S4 object system,
+closures and environments.  The implementation of R has a number of other
+types that are mostly not used by user code, they are listed in
+Figure~\ref{types} for reference.
+
+The addition of attributes lets programmers extend the set of types by
+tagging data with user-defined attributes. For example, one could define a
+vector of four values, \code{x<-c(1,2,3,4)} and then attach the attribute
+\k{dim} with a pair of numbers as value: \code{attr(x,"dim")<-c(2,2)}.  From
+that point, arithmetic functions will treat \k{x} as a 2x2 matrix. Another
+attribute that can be set is the \k{class}.  This attribute can be bound to
+a list of class names. For instance, \code{class(x)<-"human"}, set the class
+of \k{x} to be \k{human}.  Attributes are thus used for object-oriented
+programming. The S3 object system support single dispatch on the class of
+the first argument of a function, whereas the S4 object system allows
+multiple dispatch (on all arguments). Some of the most widely used data
+type, such as data frames, leverage attributes. A data frame, for instance,
+is a list of vectors with a class and a column name attribute.
+
+\paragraph{Summary.} The most common values in R computations are vectorized
+types. R programs do not have a way to constrain values to be scalar.
+\k{NULL} is sometimes used to represent the case when no value is
+available. \k{NA} is used within vector to represent missing observations.
+Attributes can decorate values and are used as building blocks for
+object-oriented programming. A potential type system for R could focus only
+on the builtin types, if one wanted to strive for simplicity, or it could
+try to capture attributes at the risk of increased complexity.
+
+%
+%
+%
+%
 \subsection{R-dyntrace}
@@
 
-\AT{Worth talking a bit about the analysis framework?}
+\AT{Leaving this one for Aviral.}
 
+R-dyntrace is an efficient dynamic analysis framework for R.
+
+%
+%
+%
+%
 \subsection{The Corpus}
@@
 We selected packages that have both \AT{a good amount of} reverse dependencies to ensure their broad usefulness, as well \AT{sufficiently high} code and branch coverage of test, example, and vignette programs to ensure that our analysis built up an accurate picture of function use.
+
+%
+%
+\subsubsection{Cross-Validation Against Kaggle}
+
+Kaggle is a website offering an environment for hosting Jupyter notebooks, and importantly provides access to a large number of open-source data sets and R and Python code.
+This solved a major issue with R programs hosted on e.g. Github, where often the data is not made available.
+Our aim is to run R programs hosted on Kaggle to validate the type signatures we generated for popular R packages.
+
+\AT{Anything else to say on Kaggle, here? We will probably talk more about the corpus we analyzed later.}
 
@@
 
-\AT{Talk about Aviral's work on evaluation.}
+\AT{Talk about the evaluation.}
+
+\AT{Ok, so we haven't discussed the story for the results, so I'll just write some disjointed sections, discussing design decisions, and backing them up with the data.
+This should give us an idea of what we want to say, and we can decide how to organize everything once we have stuff written.
+Some of the things I'll be pulling from the data are: how many functions are higher-order? How many times are structs present? How many times do classes supercede other types?
+Basically, we want to use this section to turn the simplicity of our type system from a weakness into a strength. 
+The simplicity {\it is} a strength, we just need to make sure that's obvious to the reviewer.}
 
@@
 %% Bibliography
-%\bibliography{bibfile}
+\bibliography{bib/biblio,bib/jv,bib/r,bib/new,bib/gradual}
 

### 2020-05-06 evaluation/R/evaluate-type-analysis.Rmd
Add code for type coverage evaluation and type analysis and Rmd files

---
title: Evaluation of Types
author: Aviral Goel
output:
  html_document:
    theme: cerulean

params:
  type_analysis_datafile: ../data/type-analysis/merged/merged.csv
---

```{r setup, echo=FALSE, eval=TRUE}
#    toc: true
#    toc_float: true
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(DT))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(tibble))

type_analysis_data <- 
    read_csv(params$type_analysis_datafile,
             col_names = TRUE,
             col_types = cols(col_character(),
                              col_character(),
                              col_integer(),
                              col_integer(),
                              col_integer(),
                              col_character(),
                              col_integer()))

to_percentage <- function(proportions) {
    str_c(round(100 * proportions, 2), "%")
}
```

# High-level Summary

```{r high-level, echo=FALSE, eval=TRUE}
total_package_count <- nrow(distinct(type_analysis_data, package_name))
total_function_count <- nrow(distinct(type_analysis_data, package_name, function_name))
```

**Packages**: `r total_package_count`

**Functions**: `r total_function_count`

In a function declaration of the form:

`type f (<int | chr | lgl, dbl, lgl> => lgl | chr) | (<dbl | chr, int, raw> => lgl | chr)`

There are two outer alternatives and 3 inner alternatives for parameter 1.

# Outer Alternatives

Now, we look at the number of alternatives in the function type.

```{r outer_alternative_count, eval=TRUE, echo=FALSE}
outer_alternative_distribution <-
   type_analysis_data %>%
   group_by(package_name, function_name) %>%
   summarize(outer_alternative_count = length(unique(outer_alternative))) %>%
   ungroup() %>%
   group_by(outer_alternative_count) %>%
   summarize(function_count = n(), package_count = length(unique(package_name))) %>%
   ungroup()

datatable(outer_alternative_distribution)
```

# Inner Alternatives

## Parameters

Now, we look at the number of alternatives in the function argument type.

```{r argument_inner_alternative_count, eval=TRUE, echo=FALSE}
argument_inner_alternative_count <-
   type_analysis_data %>%
   dplyr::filter(parameter_position != -1L) %>%
   group_by(package_name, function_name, parameter_position) %>%
   summarize(inner_alternative_count = length(unique(inner_alternative))) %>%
   ungroup() %>%
   arrange(desc(inner_alternative_count))

datatable(argument_inner_alternative_count)
```

## Return

Now, we look at the number of alternatives in the function return type.

```{r result_inner_alternative_count, eval=TRUE, echo=FALSE}
result_inner_alternative_count <-
   type_analysis_data %>%
   dplyr::filter(parameter_position == -1L) %>%
   group_by(package_name, function_name) %>%
   summarize(inner_alternative_count = length(unique(inner_alternative))) %>%
   ungroup() %>%
   arrange(desc(inner_alternative_count))

result_inner_alternative_count %>%
dplyr::filter(inner_alternative_count != 1) %>%
datatable(caption="Number of union alternatives in return type")

result_inner_alternative_count_graph_data <-
result_inner_alternative_count %>%
group_by(inner_alternative_count) %>%
summarize(function_count = n()) %>%
ungroup()

result_inner_alternative_count %>%
mutate(inner_alternative_count = ceiling(inner_alternative_count / 5) * 5) %>%
group_by(inner_alternative_count) %>%
summarize(function_count = n(), package_count = length(unique(package_name))) %>%
ungroup() %>%
mutate(function_percentage = to_percentage(function_count / total_function_count),
       package_percentage = to_percentage(package_count / total_package_count)) %>%
arrange(desc(inner_alternative_count)) %>%
mutate(inner_alternative_count = str_c(inner_alternative_count - 4, 
                                       "-", 
                                       inner_alternative_count)) %>%
select(package_count, package_percentage, function_count, function_percentage, inner_alternative_count) %>%
datatable(caption="Distribution of return type alternatives")

explicit_cases <-
result_inner_alternative_count_graph_data %>%
filter(inner_alternative_count <= 5)

implicit_cases <-
result_inner_alternative_count_graph_data %>%
filter(inner_alternative_count > 5) %>%
summarize(function_count = length(function_count),
          inner_alternative_count = "> 5")

result_inner_alternative_count_graph_data <-
    explicit_cases %>%
    mutate(inner_alternative_count = as.character(inner_alternative_count)) %>%
    rbind(implicit_cases)

result_inner_alternative_count_graph_data %>%
ggplot(aes(inner_alternative_count, function_count)) +
geom_col() +
scale_x_discrete(limits = c(1:5, "> 5")) #%>%
#coord_flip()

```

### 2020-05-06 paper/main.tex
initial discussion of primitive types

@@
 \subsubsection{Types of Data}
+\label{subsubsec:backgroundtypes}
 
@@
 
+%
+%
+%
+%
+%
+%
 \section{Type System Design}
+\label{sec:typesystemdesign}
 
-\AT{Lay out the type system.
-Ideally, for every interaction we propose, we should be able to point to an analysis that supports our claim.
-Maybe we don't always have to do this, especially for some of the simple, obvious types, but any departures from basic R functionality should be supported by the data.}
+Our corpora analysis reported \AT{number} call traces for over \AT{number} functions.
+We distilled these call traces into function types which captured the dynamic behaviour of the functions.
+We emphasize that this was an iterative process: initial analysis suggested a certain set of types, which we tried and implemented, and further analysis suggested improvements, and so on.
+In this section, we will present the final type system that we designed, which we believe captures the essence of R.
+Throughout, we will touch on design decisions and weigh our choices against alternatives, speaking to the strengths and weaknesses of our chosen design.
 
+%
+%
+%
+%
+\subsection{Basic Types}
+\label{subsec:basictypes}
+
+As discussed in Section~\ref{subsubsec:backgroundtypes}, the workhorse types in R are the vectorized primitives: integers, doubles, complex numbers, logicals (booleans), characters, and raws (bytes).
+These are {\it vectorized} in that they are always considered to be vectors by the R runtime: for instance, even ``scalar'' numbers are considered to be unit-length vectors to R.
+These primitive vectors are homogeneous, in that a vector of doubles contains only doubles.
+
+Related to these primitive types is one of R's notions of ``nullness'': for each of the primitives, R distinguishes a special ``NA'' value which represents missing data.
+There is an NA for each of the primitive types (i.e. there is a double NA, integer NA, etc.).
+Thus, even homogeneous vectors can represent missing data with their appropriate NA value.
+
+As with everything in R, there is nuance in even the simplest cases.
+We will now discuss our design decisions relating to the vectorized primitives.
+
+%
+%
+\subsubsection{Possibly-NA Primitives}
+
+A common primitive usage pattern is related to the presence of NAs: \AT{often}, programmers explicitly check for NAs in primitive vectors, possibly sanitizing them if they are present.
+For instance, consider the code in \AT{do an NA example}.
+NAs complicate practically every computation using primitive vectors, a fact that many programmers appear to be acutely aware of.
+
+%
+%
+\subsubsection{Scalar Primitives}
+
+\AT{Maybe we combine the next 3 subsections into one on dimensionality? There are overlapping arguments.}
+Initial analyses of our data revealed that programmers \AT{often} use scalars (as explicitly as they can, anyway).
+For example, consider the code in \AT{do a scalar example}.
+
+We found that scalars occurred more often than vectors, and chose the scalar type syntax as the default.
+The syntax for primitive scalars and vectors is as follows:
+
+\AT{syntax...}
+
+%
+%
+\subsubsection{Forgoing Matrices}
+
+Another layer of complexity on vectorized primitives is the presence of ``matrices'', which are \AT{more detail}.
+To R, a vector is a single-dimensional matrix, and R coerces vectors to matrices where appropriate.
+That said, if all vectors are matrices, then any computation that is valid on a matrix should also be valid on a vector.
+Restrictions on matrix-types arguments amount to restrictions on the dimensions of the argument, which suggests a dependent type system.
+We will touch on this shortly.
+\AT{I can try to come up with some numbers, here. I'm open to a discussion on how best to word this.}
+
+%
+%
+\subsubsection{Forgoing Dimensionality}
+
+\AT{Will revisit to tighten discussion, need to think about best way to phrase this.
+Perhaps "dependent type system bad" is enough?}
+Including data dimensions in types would make our type system a dependent type system, which we believe is too complex to retrofit onto a dynamic data science language such as R.
+The simplest way to include dimensions would be to explicitly specify the dimensions of an argument (e.g.
+specifying that a function should always be called with a double vector of length 10), which is not particularly useful, and a pattern which \AT{rarely appeared in our analysis}.
+
+%
+%
+%
+%
+\subsection{Lists}
+
+%
+%
+%
+%
+%
+%
 \section{Evaluation Results}

### 2020-05-07 paper/main.tex
Add description of contractr

@@
 
+\subsection{contractr}
+\label{sec:contractr}
+contractr is an R package that implements our type system for the R package ecosystem.
+It adds contracts to package functions for type-checking function arguments and results.
+On contract failure, it reports a warning message with the package name, function name, argument name and position, expected type (according to the type declaration), actual type (of the value observed), and a stack trace for the called function.
+
+contractr is implemented in 434 R LOC and 2358 C++ LOC.
+It has been designed and tested with GNU R-3.5.0 but it also works with recent R releases.
+The package has been hardened with a battery of 400 test cases and battle-tested with sanity checking for the 500 packages for which we generate type signatures and 12,000 packages against which we evaluate them.
+
+contractr works by modifying function definitions to insert calls to our type-checking function.
+When the modified function is called, our type-checking function gets executed first.
+The type-checker is invoked on each call to the function on all of its arguments and return value.
+There are two interesting challenges that need to be tackled to make this work reliably, without surprises for the end-users.
+Firstly, we need to correctly type-check arguments.
+R is a non-strict language (citation to my lazy paper); when a function is called, instead of getting values as arguments which can be immediately type-checked, we get unevaluated code thunks.
+We can force the evaluation of these thunks to obtain values for type-checking but that will not preserve the language’s semantics.
+This will lead to incorrect results and surprises for the end-users.
+To preserve the original semantics, we modify the thunks to call our type-checker on the value obtained when they are forced, along with the context necessary to associate the thunk to the function and its parameter (for failure messages).
+This works, but with one wrinkle.
+GNU R has a bytecode compiler that can sometimes optimize away thunks, in those cases, we don’t get thunks as arguments but values that can be immediately type-checked.
+However, we still can’t type-check the value and issue failure messages because that would violate the semantics.
+Non-strict semantics require that any computation related to the argument happens on the first use of the argument.
+If this argument was never used, it would be incorrect to type-check it.
+To make this work correctly, we explicitly wrap the argument value in a thunk with a call to the type-checker as in the previous case.
+Secondly, we need to correctly type-check function return values.
+In R, the result of function evaluation is the result of evaluating the last expression in its body; so a return call is absent from most function definitions.
+So there are many potential subexpressions which need to be type-checked aginst the return type.
+For this, we register our type-checker to be called on the return value in a function exit hook.
+This hook is executed in the function’s environment after it has finished executing.
+While this works, we also have to contend with yet another wrinkle, R interpreter can perform longjumps in its C implementation, which causes R functions on the stack to be discarded.
+When they are discarded, their exit hooks are called, which in our case calls the type-checker.
+But, these functions are in the middle of an active computation, so they don’t have a return value to type-check.
+We deal with this problem by allocating a unique object which serves as the return value for calls that are discarded.
+Our exit hook does not call the type-checker if this unique value happens to be the return value of the function.
+
+Type declarations can be made available to contractr in three ways: as part of its internal database of types (which already has type signatures for our corpus of 500 packages), as part of a designated TYPEDECLARATION file supplied by the package author and installed along with the package, or through a user-level API function insert\_contract that lets the user insert contract to a custom function with the type declaration provided as a string argument.
+Through this package, we intend to provide a clear path for the R developers to add types to untyped package code. Hence, we have paid a lot of attention to usability in its design. 
+Firstly, the package enables type-checking for all packages with type signatures, automatically, on importing, offering a seamless and hassle-free experience for both package developers and end-users.
+A mere library(contractr) is enough to insert contracts, both, in packages pre-loaded in the user’s workspace and in packages that will be loaded eventually.
+For this, contractr relies heavily on R’s reflective and dynamic capabilities.
+Upon loading contractr scans all the pre-loaded packages in the user’s workspace and inserts contracts in functions for which type signatures are available.
+For all other packages installed on user’s machine but not yet loaded in the workspace, contractr sets up package load hooks.
+The hooks get executed when these packages get loaded and insert contracts to the package functions.
+The ensures that users do not fiddle manually with the contractr API to enable type-checking for packages loaded eventually.
+Furthermore, contractr automatically removes contracts from all the package functions and restores them to their original state when it is unloaded, an operation that is rarely performed but is possible, nonetheless.
+Secondly, we enable package authors to supply type declarations to contractr for their package functions, without requiring any modification to package code.
+This is accomplished by providing type declarations inside roxygen2 function comment blocks.
+roxygen2 is an R package that enables authors to add documentation to R functions in the form of organized plaintext sections which are automatically exported to R’s latex style custom documentation format.
+roxygen2 is used by … R packages in CRAN.
+It also provides an extension API for processing metadata in custom documentation tag sections.
+A package can register a method to be called when this tag is encountered.
+The metadata for that tag section is made available to the registrant before package build and installation steps.
+contractr uses this API to implement a @type tag.
+Our roxygen2 hook extracts type declarations from these sections from all the documented functions of the package and stores then in a TYPEDECLARATION file inside the package folder.
+This files is copied as is on package installation and automatically picked up contractr when the two packges are loaded.
+Here is an example of how a roxygen code block looks like with our type declaration:
+
+This technique enables the function and its type signature to coexist next to each other, where they are more likely to remain synchronized.
+It also enables package authors to use our type system without modifying their code.
+Thirdly, we provide a very expressive API and package switches to the users, covering a variety of uses cases.
+During development, developers can explicitly insert contracts by supplying the type declaration as a string argument to the insert\_contract API function.
+Contracts can be removed by calling remove\_contract on a function.
+Contracts can be selectively enabled or disabled for code blocks by wrapping them in calls to ignore\_contracts and capture\_contracts functions.
+This enables developers and users to selectively type check sections of code.
+Furthermore, these functions return a data frame that contains all the information about failed and successful contract assertions in the wrapped code blocks.
+Multiple type-checking failures can introduce a lot of noise in the program output.
+To alleviate this problem, users can set a flag, contractr.severity, to “silence” which suppresses the failure messages.
+The contract assertions can then be obtained as a data frame through a call to get\_contracts at any program point.
+Setting this flag to “error” will turn the type-checking failure warnings to errors.
+
 %

### 2020-05-07 paper/main.tex
Reformat text and add citations

@@
 \label{sec:contractr}
-contractr is an R package that implements our type system for the R package ecosystem.
-It adds contracts to package functions for type-checking function arguments and results.
-On contract failure, it reports a warning message with the package name, function name, argument name and position, expected type (according to the type declaration), actual type (of the value observed), and a stack trace for the called function.
+contractr is an R package that implements our type system for the R package
+ecosystem. It adds contracts to package functions for type-checking function
+arguments and results. On contract failure, it reports a warning message with
+the package name, function name, argument name and position, expected type
+(according to the type declaration), actual type (of the value observed), and a
+stack trace for the called function.
 
-contractr is implemented in 434 R LOC and 2358 C++ LOC.
-It has been designed and tested with GNU R-3.5.0 but it also works with recent R releases.
-The package has been hardened with a battery of 400 test cases and battle-tested with sanity checking for the 500 packages for which we generate type signatures and 12,000 packages against which we evaluate them.
+contractr is implemented in 434 R LOC and 2358 C++ LOC. It has been designed and
+tested with GNU R-3.5.0 but it also works with recent R releases. The package
+has been hardened with a battery of 400 test cases and battle-tested with sanity
+checking for the 500 packages for which we generate type signatures and 12,000
+packages against which we evaluate them.
 
-contractr works by modifying function definitions to insert calls to our type-checking function.
-When the modified function is called, our type-checking function gets executed first.
-The type-checker is invoked on each call to the function on all of its arguments and return value.
-There are two interesting challenges that need to be tackled to make this work reliably, without surprises for the end-users.
-Firstly, we need to correctly type-check arguments.
-R is a non-strict language (citation to my lazy paper); when a function is called, instead of getting values as arguments which can be immediately type-checked, we get unevaluated code thunks.
-We can force the evaluation of these thunks to obtain values for type-checking but that will not preserve the language’s semantics.
-This will lead to incorrect results and surprises for the end-users.
-To preserve the original semantics, we modify the thunks to call our type-checker on the value obtained when they are forced, along with the context necessary to associate the thunk to the function and its parameter (for failure messages).
-This works, but with one wrinkle.
-GNU R has a bytecode compiler that can sometimes optimize away thunks, in those cases, we don’t get thunks as arguments but values that can be immediately type-checked.
-However, we still can’t type-check the value and issue failure messages because that would violate the semantics.
-Non-strict semantics require that any computation related to the argument happens on the first use of the argument.
-If this argument was never used, it would be incorrect to type-check it.
-To make this work correctly, we explicitly wrap the argument value in a thunk with a call to the type-checker as in the previous case.
-Secondly, we need to correctly type-check function return values.
-In R, the result of function evaluation is the result of evaluating the last expression in its body; so a return call is absent from most function definitions.
-So there are many potential subexpressions which need to be type-checked aginst the return type.
-For this, we register our type-checker to be called on the return value in a function exit hook.
-This hook is executed in the function’s environment after it has finished executing.
-While this works, we also have to contend with yet another wrinkle, R interpreter can perform longjumps in its C implementation, which causes R functions on the stack to be discarded.
-When they are discarded, their exit hooks are called, which in our case calls the type-checker.
-But, these functions are in the middle of an active computation, so they don’t have a return value to type-check.
-We deal with this problem by allocating a unique object which serves as the return value for calls that are discarded.
-Our exit hook does not call the type-checker if this unique value happens to be the return value of the function.
+contractr works by modifying function definitions to insert calls to our
+type-checking function. When the modified function is called, our type-checking
+function gets executed first.  The type-checker is invoked on each call to the
+function on all of its arguments and return value.
 
-Type declarations can be made available to contractr in three ways: as part of its internal database of types (which already has type signatures for our corpus of 500 packages), as part of a designated TYPEDECLARATION file supplied by the package author and installed along with the package, or through a user-level API function insert\_contract that lets the user insert contract to a custom function with the type declaration provided as a string argument.
-Through this package, we intend to provide a clear path for the R developers to add types to untyped package code. Hence, we have paid a lot of attention to usability in its design. 
-Firstly, the package enables type-checking for all packages with type signatures, automatically, on importing, offering a seamless and hassle-free experience for both package developers and end-users.
-A mere library(contractr) is enough to insert contracts, both, in packages pre-loaded in the user’s workspace and in packages that will be loaded eventually.
-For this, contractr relies heavily on R’s reflective and dynamic capabilities.
-Upon loading contractr scans all the pre-loaded packages in the user’s workspace and inserts contracts in functions for which type signatures are available.
-For all other packages installed on user’s machine but not yet loaded in the workspace, contractr sets up package load hooks.
-The hooks get executed when these packages get loaded and insert contracts to the package functions.
-The ensures that users do not fiddle manually with the contractr API to enable type-checking for packages loaded eventually.
-Furthermore, contractr automatically removes contracts from all the package functions and restores them to their original state when it is unloaded, an operation that is rarely performed but is possible, nonetheless.
-Secondly, we enable package authors to supply type declarations to contractr for their package functions, without requiring any modification to package code.
-This is accomplished by providing type declarations inside roxygen2 function comment blocks.
-roxygen2 is an R package that enables authors to add documentation to R functions in the form of organized plaintext sections which are automatically exported to R’s latex style custom documentation format.
-roxygen2 is used by … R packages in CRAN.
-It also provides an extension API for processing metadata in custom documentation tag sections.
-A package can register a method to be called when this tag is encountered.
-The metadata for that tag section is made available to the registrant before package build and installation steps.
-contractr uses this API to implement a @type tag.
-Our roxygen2 hook extracts type declarations from these sections from all the documented functions of the package and stores then in a TYPEDECLARATION file inside the package folder.
-This files is copied as is on package installation and automatically picked up contractr when the two packges are loaded.
+%
+%
+\subsubsection{Challenges}
+There are two interesting challenges that need to be tackled to make this work
+reliably, without surprises for the end-users. Firstly, we need to correctly
+type-check arguments. R is a non-strict language~\cite{oopsla19};
+when a function is called, instead of getting values as arguments which can be
+immediately type-checked, we get unevaluated code thunks. We can force the
+evaluation of these thunks to obtain values for type-checking but that will not
+preserve the language’s semantics. This will lead to incorrect results and
+surprises for the end-users. To preserve the original semantics, we modify the
+thunks to call our type-checker on the value obtained when they are forced,
+along with the context necessary to associate the thunk to the function and its
+parameter (for failure messages). This works, but with one wrinkle. GNU R has a
+bytecode compiler that can sometimes optimize away thunks, in those cases, we
+don’t get thunks as arguments but values that can be immediately type-checked.
+However, we still can’t type-check the value and issue failure messages because
+that would violate the semantics. Non-strict semantics require that any
+computation related to the argument happens on the first use of the argument. If
+this argument was never used, it would be incorrect to type-check it. To make
+this work correctly, we explicitly wrap the argument value in a thunk with a
+call to the type-checker as in the previous case.  Secondly, we need to
+correctly type-check function return values. In R, the result of function
+evaluation is the result of evaluating the last expression in its body; so a
+return call is absent from most function definitions. So there are many
+potential subexpressions which need to be type-checked aginst the return type.
+For this, we register our type-checker to be called on the return value in a
+function exit hook. This hook is executed in the function’s environment after it
+has finished executing. While this works, we also have to contend with yet
+another wrinkle, R interpreter can perform longjumps in its C implementation,
+which causes R functions on the stack to be discarded. When they are discarded,
+their exit hooks are called, which in our case calls the type-checker. But,
+these functions are in the middle of an active computation, so they don’t have a
+return value to type-check. We deal with this problem by allocating a unique
+object which serves as the return value for calls that are discarded. Our exit
+hook does not call the type-checker if this unique value happens to be the
+return value of the function.
+
+%
+%
+\subsubsection{Type Declarations}
+Type declarations can be made available to contractr in three ways: as part of
+its internal database of types (which already has type signatures for our corpus
+of 500 packages), as part of a designated TYPEDECLARATION file supplied by the
+package author and installed along with the package, or through a user-level API
+function insert\_contract that lets the user insert contract to a custom
+function with the type declaration provided as a string argument.
+
+%
+%
+\subsubsection{Usability}
+Through this package, we intend to provide a clear path for the R developers to
+add types to untyped package code. Hence, we have paid a lot of attention to
+usability in its design.
+Firstly, the package enables type-checking for all packages with type
+signatures, automatically, on importing, offering a seamless and hassle-free
+experience for both package developers and end-users. A mere library(contractr)
+is enough to insert contracts, both, in packages pre-loaded in the user’s
+workspace and in packages that will be loaded eventually. For this, contractr
+relies heavily on R’s reflective and dynamic capabilities. Upon loading
+contractr scans all the pre-loaded packages in the user’s workspace and inserts
+contracts in functions for which type signatures are available. For all other
+packages installed on user’s machine but not yet loaded in the workspace,
+contractr sets up package load hooks. The hooks get executed when these packages
+get loaded and insert contracts to the package functions. The ensures that users
+do not fiddle manually with the contractr API to enable type-checking for
+packages loaded eventually. Furthermore, contractr automatically removes
+contracts from all the package functions and restores them to their original
+state when it is unloaded, an operation that is rarely performed but is
+possible, nonetheless.
+Secondly, we enable package authors to supply type declarations to contractr for
+their package functions, without requiring any modification to package code.
+This is accomplished by providing type declarations inside roxygen2 function
+comment blocks. roxygen2 is an R package that enables authors to add
+documentation to R functions in the form of organized plaintext sections which
+are automatically exported to R’s latex style custom documentation format.
+roxygen2 is used by … R packages in CRAN. It also provides an extension API for
+processing metadata in custom documentation tag sections. A package can register
+a method to be called when this tag is encountered. The metadata for that tag
+section is made available to the registrant before package build and
+installation steps. contractr uses this API to implement a @type tag. Our
+roxygen2 hook extracts type declarations from these sections from all the
+documented functions of the package and stores then in a TYPEDECLARATION file
+inside the package folder. This files is copied as is on package installation
+and automatically picked up contractr when the two packges are loaded.
 Here is an example of how a roxygen code block looks like with our type declaration:
 
-This technique enables the function and its type signature to coexist next to each other, where they are more likely to remain synchronized.
-It also enables package authors to use our type system without modifying their code.
-Thirdly, we provide a very expressive API and package switches to the users, covering a variety of uses cases.
-During development, developers can explicitly insert contracts by supplying the type declaration as a string argument to the insert\_contract API function.
-Contracts can be removed by calling remove\_contract on a function.
-Contracts can be selectively enabled or disabled for code blocks by wrapping them in calls to ignore\_contracts and capture\_contracts functions.
-This enables developers and users to selectively type check sections of code.
-Furthermore, these functions return a data frame that contains all the information about failed and successful contract assertions in the wrapped code blocks.
-Multiple type-checking failures can introduce a lot of noise in the program output.
-To alleviate this problem, users can set a flag, contractr.severity, to “silence” which suppresses the failure messages.
-The contract assertions can then be obtained as a data frame through a call to get\_contracts at any program point.
-Setting this flag to “error” will turn the type-checking failure warnings to errors.
+This technique enables the function and its type signature to coexist next to
+each other, where they are more likely to remain synchronized. It also enables
+package authors to use our type system without modifying their code.
+Thirdly, we provide a very expressive API and package switches to the users,
+covering a variety of uses cases. During development, developers can explicitly
+insert contracts by supplying the type declaration as a string argument to the
+insert\_contract API function. Contracts can be removed by calling
+remove\_contract on a function. Contracts can be selectively enabled or disabled
+for code blocks by wrapping them in calls to ignore\_contracts and
+capture\_contracts functions. This enables developers and users to selectively
+type check sections of code. Furthermore, these functions return a data frame
+that contains all the information about failed and successful contract
+assertions in the wrapped code blocks. Multiple type-checking failures can
+introduce a lot of noise in the program output. To alleviate this problem, users
+can set a flag, contractr.severity, to “silence” which suppresses the failure
+messages. The contract assertions can then be obtained as a data frame through a
+call to get\_contracts at any program point. Setting this flag to “error” will
+turn the type-checking failure warnings to errors.
 
@@
 In order to ensure that our type system design aligns with the day-to-day usage of R, we performed a corpus analysis of some of the most widely used R packages.
-We built our dynamic analysis in the R-dyntrace \AT{cite something?} dynamic analysis framework, which is itself built on a modified version of R targeting R version 3.5.0---we refer the reader to Section~\ref{sec:r-dyntrace} for more details on R-dyntrace.
+We built our dynamic analysis in the R-dyntrace~\cite{oopsla19} dynamic analysis framework, which is itself built on a modified version of R targeting R version 3.5.0---we refer the reader to Section~\ref{sec:r-dyntrace} for more details on R-dyntrace.
 

### 2020-05-07 paper/main.tex
Fix contractr introduction

@@
 \label{sec:contractr}
-contractr is an R package that implements our type system for the R package
-ecosystem. It adds contracts to package functions for type-checking function
-arguments and results. On contract failure, it reports a warning message with
-the package name, function name, argument name and position, expected type
-(according to the type declaration), actual type (of the value observed), and a
-stack trace for the called function.
+\emph{contractr} is an R package that implements our type system for the R
+package ecosystem. It adds contracts to package functions for type-checking
+arguments and results on function calls. On contract failure, it reports a type
+mismatch warning with the package name, function name, argument name, argument
+position, expected type signature, actual type signature of the value observed,
+and a stack trace for the called function. \emph{contractr} works by modifying
+function definition to insert a call to the type-checking function. When the
+modified function is called, the type-checking function is invoked on the
+modified function's arguments and return value.
 
-contractr is implemented in 434 R LOC and 2358 C++ LOC. It has been designed and
-tested with GNU R-3.5.0 but it also works with recent R releases. The package
-has been hardened with a battery of 400 test cases and battle-tested with sanity
-checking for the 500 packages for which we generate type signatures and 12,000
-packages against which we evaluate them.
-
-contractr works by modifying function definitions to insert calls to our
-type-checking function. When the modified function is called, our type-checking
-function gets executed first.  The type-checker is invoked on each call to the
-function on all of its arguments and return value.
+\emph{contractr} is implemented in 434 R LOC and 2358 C++ LOC. The core
+implementation is in C++ to keep the runtime overhead low. It been designed and
+tested with GNU R-3.5.0 but it also works with recent versions of R. It has been
+hardened with a battery of 400 test cases. We have used it extensively during
+the course of this work; firstly, for sanity checking of XXX type signatures
+generated for the XXX packages during the development phase, and secondly, for
+assessing the quality of these type signatures on XXX packages during the
+evaluation phase.
 

### 2020-05-07 paper/main.tex
Fix quotes

@@
 evaluation of these thunks to obtain values for type-checking but that will not
-preserve the language’s semantics. This will lead to incorrect results and
+preserve the language's semantics. This will lead to incorrect results and
 surprises for the end-users. To preserve the original semantics, we modify the
@@
 bytecode compiler that can sometimes optimize away thunks, in those cases, we
-don’t get thunks as arguments but values that can be immediately type-checked.
-However, we still can’t type-check the value and issue failure messages because
+don't get thunks as arguments but values that can be immediately type-checked.
+However, we still can't type-check the value and issue failure messages because
 that would violate the semantics. Non-strict semantics require that any
@@
 For this, we register our type-checker to be called on the return value in a
-function exit hook. This hook is executed in the function’s environment after it
+function exit hook. This hook is executed in the function's environment after it
 has finished executing. While this works, we also have to contend with yet
@@
 their exit hooks are called, which in our case calls the type-checker. But,
-these functions are in the middle of an active computation, so they don’t have a
+these functions are in the middle of an active computation, so they don't have a
 return value to type-check. We deal with this problem by allocating a unique
@@
 experience for both package developers and end-users. A mere library(contractr)
-is enough to insert contracts, both, in packages pre-loaded in the user’s
+is enough to insert contracts, both, in packages pre-loaded in the user's
 workspace and in packages that will be loaded eventually. For this, contractr
-relies heavily on R’s reflective and dynamic capabilities. Upon loading
-contractr scans all the pre-loaded packages in the user’s workspace and inserts
+relies heavily on R's reflective and dynamic capabilities. Upon loading
+contractr scans all the pre-loaded packages in the user's workspace and inserts
 contracts in functions for which type signatures are available. For all other
-packages installed on user’s machine but not yet loaded in the workspace,
+packages installed on user's machine but not yet loaded in the workspace,
 contractr sets up package load hooks. The hooks get executed when these packages
@@
 documentation to R functions in the form of organized plaintext sections which
-are automatically exported to R’s latex style custom documentation format.
+are automatically exported to R's latex style custom documentation format.
 roxygen2 is used by … R packages in CRAN. It also provides an extension API for

### 2020-05-07 paper/main.tex
Fix challenge description for contractr

@@
 %
-\subsubsection{Challenges}
-There are two interesting challenges that need to be tackled to make this work
-reliably, without surprises for the end-users. Firstly, we need to correctly
-type-check arguments. R is a non-strict language~\cite{oopsla19};
-when a function is called, instead of getting values as arguments which can be
-immediately type-checked, we get unevaluated code thunks. We can force the
-evaluation of these thunks to obtain values for type-checking but that will not
-preserve the language's semantics. This will lead to incorrect results and
-surprises for the end-users. To preserve the original semantics, we modify the
-thunks to call our type-checker on the value obtained when they are forced,
-along with the context necessary to associate the thunk to the function and its
-parameter (for failure messages). This works, but with one wrinkle. GNU R has a
-bytecode compiler that can sometimes optimize away thunks, in those cases, we
-don't get thunks as arguments but values that can be immediately type-checked.
-However, we still can't type-check the value and issue failure messages because
-that would violate the semantics. Non-strict semantics require that any
-computation related to the argument happens on the first use of the argument. If
-this argument was never used, it would be incorrect to type-check it. To make
-this work correctly, we explicitly wrap the argument value in a thunk with a
-call to the type-checker as in the previous case.  Secondly, we need to
-correctly type-check function return values. In R, the result of function
-evaluation is the result of evaluating the last expression in its body; so a
-return call is absent from most function definitions. So there are many
-potential subexpressions which need to be type-checked aginst the return type.
-For this, we register our type-checker to be called on the return value in a
-function exit hook. This hook is executed in the function's environment after it
-has finished executing. While this works, we also have to contend with yet
-another wrinkle, R interpreter can perform longjumps in its C implementation,
-which causes R functions on the stack to be discarded. When they are discarded,
-their exit hooks are called, which in our case calls the type-checker. But,
-these functions are in the middle of an active computation, so they don't have a
-return value to type-check. We deal with this problem by allocating a unique
-object which serves as the return value for calls that are discarded. Our exit
-hook does not call the type-checker if this unique value happens to be the
-return value of the function.
+\subsubsection{Type Declarations}
+Type declarations can be made available to \emph{contractr} in three ways: as
+part of its internal database of types (which already has type signatures for
+our corpus of 500 packages), as part of a designated \emph{TYPEDECLARATION} file
+supplied by the package author and installed along with the package, or through
+a user-level API function insert\_contract that lets the user insert contract to
+a custom function with the type declaration provided as a string argument.
 
@@
 %
-\subsubsection{Type Declarations}
-Type declarations can be made available to contractr in three ways: as part of
-its internal database of types (which already has type signatures for our corpus
-of 500 packages), as part of a designated TYPEDECLARATION file supplied by the
-package author and installed along with the package, or through a user-level API
-function insert\_contract that lets the user insert contract to a custom
-function with the type declaration provided as a string argument.
+\subsubsection{Challenges}
+Retrofitting a robust and efficient type-system as an external package in R has
+been a significant undertaking. We had to contend with the complex design
+choices of R to make it work without surprises for the end-users. We describe
+two key challenges that we faced in the development of \emph{contractr}.
+Firstly, we faced the problem of type-checking arguments in a non-strict
+language while retaining the non-strict semantics. When a function is called in
+R, parameters are bound to unevaluated code thunks called
+promises~\cite{oopsla19}, instead of values which can be immediately
+type-checked. Naively forcing promises at function call to obtain a value for
+type-checking would violate the language semantics, leading to incorrect
+results. To preserve the original semantics, \emph{contractr} modifies the
+promise by wrapping the unevaluated promise expression in a call to its
+type-checker along with the context necessary to associate the promise to the
+function and its parameter (for failure messages). The contract checking happens
+when the promise is evaluated, either inside the function or many calls deep.
+This works, but with one wrinkle. GNU R has a bytecode compiler that can
+sometimes optimize away promises; in those cases, \emph{contractr} receives
+values that can be immediately type-checked. Non-strict semantics dictate that
+any computation related to the argument should happen on the first use of the
+argument. So type-checking at this point and issuing an error message would
+violate the non-strict semantics; if this argument is never used, it would be
+incorrect to type-check it at the beginning of call. To make this work,
+\emph{contractr} mutates the function parameter binding to a promise that wraps
+the argument value in a call to the type-checker, as in the previous case.
+Secondly, we faced the problem of type-checking function return value. In R, the
+result of a function call is the result of evaluating the last expression in the
+function body; so an explicit return call is absent from most function
+definitions. Thus, there could be many potential sub-expressions in the function
+body which need to be type-checked aginst the return type. To address this,
+\emph{contractr} registers the type-checker to be called on the return value
+through a function exit hook. This hook is executed in the function call's
+environment after it has executed. While this works, we have to contend with yet
+another wrinkle, R interpreter can perform longjumps in its C implementation,
+which causes active R function calls on the stack to be discarded. When they are
+discarded, their exit hooks are called, which in this case calls the registered
+type-checker. But, these functions are in the middle of an active computation,
+so they don't have a return value to type-check. \emph{contractr} deals with
+this problem by allocating a unique sentinel object which serves as the return
+value for calls that are discarded. The function call exit hook does not call
+the type-checker if this unique value happens to be the return value of the
+call.
 
@@
 add types to untyped package code. Hence, we have paid a lot of attention to
-usability in its design.
-Firstly, the package enables type-checking for all packages with type
-signatures, automatically, on importing, offering a seamless and hassle-free
-experience for both package developers and end-users. A mere library(contractr)
-is enough to insert contracts, both, in packages pre-loaded in the user's
-workspace and in packages that will be loaded eventually. For this, contractr
-relies heavily on R's reflective and dynamic capabilities. Upon loading
-contractr scans all the pre-loaded packages in the user's workspace and inserts
-contracts in functions for which type signatures are available. For all other
-packages installed on user's machine but not yet loaded in the workspace,
-contractr sets up package load hooks. The hooks get executed when these packages
-get loaded and insert contracts to the package functions. The ensures that users
-do not fiddle manually with the contractr API to enable type-checking for
-packages loaded eventually. Furthermore, contractr automatically removes
-contracts from all the package functions and restores them to their original
-state when it is unloaded, an operation that is rarely performed but is
-possible, nonetheless.
-Secondly, we enable package authors to supply type declarations to contractr for
-their package functions, without requiring any modification to package code.
-This is accomplished by providing type declarations inside roxygen2 function
-comment blocks. roxygen2 is an R package that enables authors to add
-documentation to R functions in the form of organized plaintext sections which
-are automatically exported to R's latex style custom documentation format.
+usability in its design. Firstly, the package enables type-checking for all
+packages with type signatures, automatically, on importing, offering a seamless
+and hassle-free experience for both package developers and end-users. A mere
+library(\emph{contractr}) is enough to insert contracts, both, in packages
+pre-loaded in the user's workspace and in packages that will be loaded
+eventually. For this, \emph{contractr} relies heavily on R's reflective and
+dynamic capabilities. Upon loading \emph{contractr} scans all the pre-loaded
+packages in the user's workspace and inserts contracts in functions for which
+type signatures are available. For all other packages installed on user's
+machine but not yet loaded in the workspace, contractr sets up package load
+hooks. The hooks get executed when these packages get loaded and insert
+contracts to the package functions. The ensures that users do not fiddle
+manually with the \emph{contractr} API to enable type-checking for packages
+loaded eventually. Furthermore, \emph{contractr} automatically removes contracts
+from all the package functions and restores them to their original state when it
+is unloaded, an operation that is rarely performed but is possible, nonetheless.
+Secondly, we enable package authors to supply type declarations to
+\emph{contractr} for their package functions, without requiring any modification
+to package code. This is accomplished by providing type declarations inside
+roxygen2 function comment blocks. roxygen2 is an R package that enables authors
+to add documentation to R functions in the form of organized plaintext sections
+which are automatically exported to R's latex style custom documentation format.
 roxygen2 is used by … R packages in CRAN. It also provides an extension API for
@@
 section is made available to the registrant before package build and
-installation steps. contractr uses this API to implement a @type tag. Our
+installation steps. \emph{contractr} uses this API to implement a @type tag. Our
 roxygen2 hook extracts type declarations from these sections from all the
@@
 inside the package folder. This files is copied as is on package installation
-and automatically picked up contractr when the two packges are loaded.
-Here is an example of how a roxygen code block looks like with our type declaration:
+and automatically picked up \emph{contractr} when the two packges are loaded.
+Here is an example of how a roxygen code block looks like with our type
+declaration:
 
@@
 each other, where they are more likely to remain synchronized. It also enables
-package authors to use our type system without modifying their code.
-Thirdly, we provide a very expressive API and package switches to the users,
-covering a variety of uses cases. During development, developers can explicitly
-insert contracts by supplying the type declaration as a string argument to the
+package authors to use our type system without modifying their code. Thirdly, we
+provide a very expressive API and package switches to the users, covering a
+variety of uses cases. During development, developers can explicitly insert
+contracts by supplying the type declaration as a string argument to the
 insert\_contract API function. Contracts can be removed by calling
@@
 introduce a lot of noise in the program output. To alleviate this problem, users
-can set a flag, contractr.severity, to “silence” which suppresses the failure
-messages. The contract assertions can then be obtained as a data frame through a
-call to get\_contracts at any program point. Setting this flag to “error” will
-turn the type-checking failure warnings to errors.
+can set a flag, \emph{contractr}.severity, to “silence” which suppresses the
+failure messages. The contract assertions can then be obtained as a data frame
+through a call to get\_contracts at any program point. Setting this flag to
+“error” will turn the type-checking failure warnings to errors.
 

### 2020-05-07 paper/main.tex
initial list discussion

@@
 
+\AT{I'll revisit this section to adjust for the final details, as the analysis has changed since this was written.}
+
 %
@@
 
+\AT{List rehash.}
+Separate from the vectorized primitives in R are lists:
+Whereas vectors are homogeneous and restricted to primitive types, lists can contain arbitrarily-typed data.
+R even builds on lists to create a more complex data structure known as the {\it data frame}, an essential data structure to R's functionality \AT{maybe too aggressively worded, but we can back this up with data, e.g.
+how many functions take data frames?}.
+
+\AT{Outline concerns for typing.}
+We are faced with several possible designs for list types.
+Chief concerns include the heterogeneity of list elements, and the presence of a \code{names} attribute, allowing list elements to be accessed in a struct-like manner.
+\AT{Rarely?}, programmers even write code expecting lists of a particular pre-defined length.
+The next sections highlight some important design decisions for our list types.
+
+\AT{Dig into concerns for typing.}
+
+%
+%
+\subsubsection{Forgoing Tuples, or Short, Constant Length Lists}
+
+We briefly entertained the possibility of including a \code{tuple} type, which would have looked like e.g. \code{tuple<int, int>} for a list of two integers.
+When configured to detect tuples, our analysis did indeed pick up \AT{a number}, but post-processing and simplification revealed that in \AT{many} cases, they would be subsumed by co-occurring \code{list} types.
+In other words, it just so happened that functions were being called with short lists.
+Of the cases that tuples were not subsumed by lists, \AT{often} there were \AT{tuples of different lengths}, implying that the shape of the argument was not essential.
+\AT{Wording...}
+
+%
+%
+\subsubsection{Forgoing Structs}
+
+%
+%
+\subsubsection{Forgoing Data Frames}
+
 %

### 2020-05-07 paper/main.tex
Improve usability discussion

@@
 \newcommand{\df}{\ensuremath{\attr{\l}{df}}\xspace}     % data.frame
+
+\newcommand{\contractr}{\emph{contractr}\xspace} % contractr
+\newcommand{\roxygen}{\emph{roxygen2}\xspace} % roxygen
 
@@
 \label{sec:contractr}
-\emph{contractr} is an R package that implements our type system for the R
+\contractr is an R package that implements our type system for the R
 package ecosystem. It adds contracts to package functions for type-checking
@@
 position, expected type signature, actual type signature of the value observed,
-and a stack trace for the called function. \emph{contractr} works by modifying
+and a stack trace for the called function. \contractr works by modifying
 function definition to insert a call to the type-checking function. When the
@@
 
-\emph{contractr} is implemented in 434 R LOC and 2358 C++ LOC. The core
+\contractr is implemented in 434 R LOC and 2358 C++ LOC. The core
 implementation is in C++ to keep the runtime overhead low. It been designed and
@@
 \subsubsection{Type Declarations}
-Type declarations can be made available to \emph{contractr} in three ways: as
+Type declarations can be made available to \contractr in three ways: as
 part of its internal database of types (which already has type signatures for
@@
 choices of R to make it work without surprises for the end-users. We describe
-two key challenges that we faced in the development of \emph{contractr}.
+two key challenges that we faced in the development of \contractr.
 Firstly, we faced the problem of type-checking arguments in a non-strict
@@
 type-checking would violate the language semantics, leading to incorrect
-results. To preserve the original semantics, \emph{contractr} modifies the
+results. To preserve the original semantics, \contractr modifies the
 promise by wrapping the unevaluated promise expression in a call to its
@@
 This works, but with one wrinkle. GNU R has a bytecode compiler that can
-sometimes optimize away promises; in those cases, \emph{contractr} receives
+sometimes optimize away promises; in those cases, \contractr receives
 values that can be immediately type-checked. Non-strict semantics dictate that
@@
 incorrect to type-check it at the beginning of call. To make this work,
-\emph{contractr} mutates the function parameter binding to a promise that wraps
+\contractr mutates the function parameter binding to a promise that wraps
 the argument value in a call to the type-checker, as in the previous case.
@@
 body which need to be type-checked aginst the return type. To address this,
-\emph{contractr} registers the type-checker to be called on the return value
+\contractr registers the type-checker to be called on the return value
 through a function exit hook. This hook is executed in the function call's
@@
 type-checker. But, these functions are in the middle of an active computation,
-so they don't have a return value to type-check. \emph{contractr} deals with
+so they don't have a return value to type-check. \contractr deals with
 this problem by allocating a unique sentinel object which serves as the return
@@
 \subsubsection{Usability}
-Through this package, we intend to provide a clear path for the R developers to
-add types to untyped package code. Hence, we have paid a lot of attention to
-usability in its design. Firstly, the package enables type-checking for all
-packages with type signatures, automatically, on importing, offering a seamless
-and hassle-free experience for both package developers and end-users. A mere
-library(\emph{contractr}) is enough to insert contracts, both, in packages
-pre-loaded in the user's workspace and in packages that will be loaded
-eventually. For this, \emph{contractr} relies heavily on R's reflective and
-dynamic capabilities. Upon loading \emph{contractr} scans all the pre-loaded
-packages in the user's workspace and inserts contracts in functions for which
-type signatures are available. For all other packages installed on user's
-machine but not yet loaded in the workspace, contractr sets up package load
-hooks. The hooks get executed when these packages get loaded and insert
-contracts to the package functions. The ensures that users do not fiddle
-manually with the \emph{contractr} API to enable type-checking for packages
-loaded eventually. Furthermore, \emph{contractr} automatically removes contracts
+With \contractr, we aim to provide a hassle-free path to the R developers for
+adding types to untyped package code. Hence, we have paid a lot of attention
+towards usability in its design. We discuss three such usability features.
+\begin{itemize}
+
+\item \contractr enables automatic type-checking on importing without
+  any user intervention. Just running \code{library(contractr)} R code is enough
+  to insert contracts, both, in packages pre-loaded in the user's workspace, and,
+  in packages that will be loaded eventually. To achieve this automation,
+  \contractr relies heavily on R's reflective and dynamic capabilities. On
+  loading, \contractr scans all the pre-loaded packages in the user's workspace
+  and inserts contracts in functions for which type signatures are available. For
+  all other packages installed on user's machine but not yet loaded in the
+  workspace, \contractr sets up package load hooks which are executed by R when
+  those packages are loaded. The hook registered by \contractr insert contracts to
+  those package's functions. Thus, developers and users do not have to call the
+  \contractr API to enable type-checking for packages at any point of their
+  programming workflow. Furthermore, \contractr automatically removes contracts
   from all the package functions and restores them to their original state when it
-is unloaded, an operation that is rarely performed but is possible, nonetheless.
-Secondly, we enable package authors to supply type declarations to
-\emph{contractr} for their package functions, without requiring any modification
-to package code. This is accomplished by providing type declarations inside
-roxygen2 function comment blocks. roxygen2 is an R package that enables authors
-to add documentation to R functions in the form of organized plaintext sections
-which are automatically exported to R's latex style custom documentation format.
-roxygen2 is used by … R packages in CRAN. It also provides an extension API for
-processing metadata in custom documentation tag sections. A package can register
-a method to be called when this tag is encountered. The metadata for that tag
-section is made available to the registrant before package build and
-installation steps. \emph{contractr} uses this API to implement a @type tag. Our
-roxygen2 hook extracts type declarations from these sections from all the
-documented functions of the package and stores then in a TYPEDECLARATION file
-inside the package folder. This files is copied as is on package installation
-and automatically picked up \emph{contractr} when the two packges are loaded.
-Here is an example of how a roxygen code block looks like with our type
-declaration:
+  is unloaded, an operation that is rarely performed, but useful in interactive
+  settings.
 
-This technique enables the function and its type signature to coexist next to
-each other, where they are more likely to remain synchronized. It also enables
-package authors to use our type system without modifying their code. Thirdly, we
-provide a very expressive API and package switches to the users, covering a
-variety of uses cases. During development, developers can explicitly insert
-contracts by supplying the type declaration as a string argument to the
-insert\_contract API function. Contracts can be removed by calling
-remove\_contract on a function. Contracts can be selectively enabled or disabled
-for code blocks by wrapping them in calls to ignore\_contracts and
-capture\_contracts functions. This enables developers and users to selectively
-type check sections of code. Furthermore, these functions return a data frame
-that contains all the information about failed and successful contract
-assertions in the wrapped code blocks. Multiple type-checking failures can
-introduce a lot of noise in the program output. To alleviate this problem, users
-can set a flag, \emph{contractr}.severity, to “silence” which suppresses the
-failure messages. The contract assertions can then be obtained as a data frame
-through a call to get\_contracts at any program point. Setting this flag to
-“error” will turn the type-checking failure warnings to errors.
+\item \contractr enables package authors to supply type declarations for their
+  package without requiring package code modifications. Type declarations can be
+  written alongside a \code{@type} section inside \roxygen function comment
+  blocks. \roxygen is an R package that enables authors to add documentation to
+  R functions in the form of organized plaintext sections which are
+  automatically exported to R's latex style custom documentation format.
+  \roxygen is a widely adopted package, used by over XXX R packages in CRAN. It
+  provides an extension API for adding custom documentation tag sections in
+  comment blocks. The sections are processed by \roxygen before package build
+  step and a registered method for each custom tag section is invoked with the
+  section data and documented object. \contractr uses this mechanism to register
+  a hook for a custom \code{@type} tag. This hook parses the type declarations
+  from the \code{@type} sections, extracts the function name from the documented
+  function definition, and stores them in a TYPEDECLARATION file inside the
+  package folder. This files is copied verbatim on package installation and
+  picked up \contractr when the package is loaded. Here is an example of how a
+  roxygen code block looks like with our type declaration:
+
+  This feature enables a function and its type signature to coexist next to each
+  other, where they are more likely to remain synchronized. Furthermore, the
+  \contractr hook can also add the function's type declaration to its
+  documentation, which seamlessly integrate our type system with the existing R
+  tooling.
+
+\item \contractr provides a very expressive API to the users, covering a variety
+  of uses cases. During interactive development, developers can explicitly
+  insert contracts by supplying the type declaration as a string argument to the
+  \code{insert\_contract} API function. Conversely, contracts can be removed by
+  calling the \code{remove\_contract} API on functions. Contracts can be
+  selectively enabled or disabled for code blocks by wrapping them in calls to
+  \code{ignore\_contracts} and \code{capture\_contracts} functions. This enables
+  selective type-checking of code section during the development phase.
+  Furthermore, these functions return a data frame that contains all the
+  information about failed and successful contract assertions in the wrapped
+  code blocks. This is useful for post-hoc investigation. Multiple type-checking
+  failures can introduce a lot of noise in the program output. To alleviate this
+  problem, \code{contractr.severity} option can be set to \code{'silence'}.
+  This suppresses failure messages but \contractr still performs type-checking whose
+  results can be explicitly obtained as a data frame from the
+  \code{get\_contracts} API function. R suppresses printing of warnings on the
+  terminal when a function issues too many warnings. This can hide type-checking
+  failure messages. Setting \code{contractr.severity} flag to \code{'error'}
+  turns the type-checking failure warnings to errors, which halts the program at
+  the point of contract failure.
+
+\end{itemize}
+
+While \contractr's primary logic has been implemented in C++, we have not
+observed a single segfault during our use of the package, either during the
+development and sanity-checking phase, or during the evaluation phase.
+We conclude the discussion of \contractr by drawing reader's attention to the
+engineering effort it takes to package experimental ideas in the form of stable
+and easy-to-use tools. R has an eclectic mix of features and design oddities
+that make this even harder.
 

### 2020-05-07 paper/main.tex
Itemize contractr challenges

@@
 two key challenges that we faced in the development of \contractr.
-Firstly, we faced the problem of type-checking arguments in a non-strict
+\begin{itemize}
+\item Firstly, we faced the problem of type-checking arguments in a non-strict
   language while retaining the non-strict semantics. When a function is called in
@@
   the argument value in a call to the type-checker, as in the previous case.
-Secondly, we faced the problem of type-checking function return value. In R, the
+\item Secondly, we faced the problem of type-checking function return value. In R, the
   result of a function call is the result of evaluating the last expression in the
@@
   call.
-
+\end{itemize}
 %

### 2020-05-08 evaluation/R/monkey-patching.Rmd
Add analysis Rmds

---
title: "Monkey patching"
output: html_document
---

```{r setup, include=FALSE}
library(fs)
library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source("paths.R")
```

## Fix parallel.log when running with `--resume-failed`

This will only keep the last run in the duplicated elements.

```{r}
parallel_log <- read_tsv(PACKAGE_COVERAGE_LOG)
count(parallel_log, Exitval)
updated_parallel_log <- group_by(parallel_log, Seq) %>% top_n(1, Starttime) %>% ungroup()
count(updated_parallel_log, Exitval)
write_tsv(updated_parallel_log, PACKAGE_COVERAGE_LOG)
```

## Fix package-coverage error code

```{r load coverage data}
df_coverage_run <- read_csv(PACKAGE_COVERAGE_RUN)
df_coverage <- read_csv(PACKAGE_COVERAGE)
```

```{r}
df_coverage_summary <- group_by(df_coverage, package) %>% summarise(success=sum(is.na(error)))

df_updated_coverage_run <- 
  df_coverage_run %>% 
  left_join(df_coverage_summary, by=c("job"="package")) %>% 
  mutate(exitval=case_when(
    exitval == 0 & success == 4 ~ 0,
    exitval == 0 & success != 4 ~ 1,
    TRUE ~ exitval
  )) %>%
  select(-success)
```

```{r}
count(df_updated_coverage_run, exitval)
```

```{r}
write_csv(df_updated_coverage_run, PACKAGE_COVERAGE_LOG)
```

```{r}
group_by(parallel_log, Seq) %>% top_n(1, -Starttime) %>% ungroup() %>% count(Exitval)
```

```{r}
parallel_log <- read_tsv(PACKAGE_COVERAGE_LOG)
updated_parallel_log <- left_join(parallel_log, select(df_coverage_run, Seq=seq, Exitval=exitval), by="Seq") %>%
  mutate(Exitval.x=Exitval.y) %>%
  rename(Exitval=Exitval.x) %>%
  select(-Exitval.y)
write_tsv(updated_parallel_log, PACKAGE_COVERAGE_LOG)
```

### 2020-05-08 evaluation/R/package-corpus.Rmd
Add analysis Rmds

---
title: "Package corpus"
output: html_document
editor_options: 
  chunk_output_type: console
params:
  min_revdeps: 5
  min_coverage: .65
  which_dependencies:
    - Depends
    - Imports
---

```{r setup, include=FALSE}
library(cranlogs)
library(DT)
library(fs)
library(lubridate)
library(purrr)
library(tidyverse)
library(tools)
library(withr)

knitr::opts_chunk$set(echo = TRUE)

source("inc/paths.R")
source("inc/setup.R")
source("inc/latextags.R")

options(repos=CRAN_LOCAL_MIRROR)

create_tags(path(TAGS_DIR, "package-corpus.tex"), prefix="PC", default=TRUE)
```

## Loading data

### CRAN

We start with all CRAN packages that we have mirrored locally using rsync.

```{r cran packages}
pkgs_cran_raw <- available.packages(contriburl = str_c("file://", path(CRAN_MIRROR_DIR, "src", "contrib")))
pkgs_cran <- tibble(package=pkgs_cran_raw[,1])
```

### Installed packages

We tried to instal all of the `r fmt(nrow(pkgs_cran))` these packages.
Some packages failed to install, in most cases this is because of missing native dependencies.

```{r find install packages}
pkgs_installed_raw <- installed.packages(LIB_DIR)
pkgs_installed <- tibble(package=pkgs_installed_raw[,1])
```

### Loadable packages

We managed to install `r fmt(nrow(pkgs_installed))` packages.
Next is to chech how many of the installed packages can actually be loaded.
We do that by running the `package-load.R` task:

```
./rapr/inst/on-each-package.sh ./rapr/inst/tasks/package-load.R 
env-rscript ./rapr/inst/process-csv-result.R run/package-load
```

This changes wether a package can be loaded using `library(package)` call.

```{r loadable packages}
pkgs_loadable <- 
  read_parallel_run_csv(PACKAGE_LOAD_RUN) %>%
  filter(exitval==0) %>%
  rename(package=job)
```

### Package metadata

Next we pull in some package metadata.

```{r load package sloc}
pkgs_sloc_raw <- 
  read_csv(
    PACKAGE_SLOC,
    col_types=cols(
      package = col_character(),
      path = col_character(),
      files = col_double(),
      language = col_character(),
      blank = col_double(),
      comment = col_double(),
      code = col_double()
    )
  )

pkgs_sloc <- 
  pkgs_sloc_raw %>%
  filter(endsWith(path, "/R") | endsWith(path, "/src")) %>%
  select(-blank, -comment, -files, -path) %>%
  mutate(language=case_when(
    language == "R" ~ "package_r_code",
    language == "C" ~ "package_native_code",
    language == "C/C++ Header" ~ "package_native_code",
    language == "C++" ~ "package_native_code",
    startsWith(language, "Fortran") ~ "package_native_code",
    TRUE ~ as.character(NA)
  )) %>%
  filter(!is.na(language)) %>%
  group_by(package, language) %>%
  summarise(code=sum(code)) %>%
  spread(key=language, value=code, fill=0) %>%
  mutate(package_code=package_native_code + package_r_code)
```

### Package runnable code

Next to the code itself, R packages have examples, tests and vignettes.
These can include snippets of R code which calls the package functions.
We extract this code and run it to narrow down the list of packages that can be considered into the final corpus.

```{r runnable code metadata}
pkgs_runnable_code_raw <- 
  read_csv(
    PACKAGE_RUNNABLE_CODE_METADATA,
    col_types=cols(
      package = col_character(),
      path = col_character(),
      files = col_double(),
      language = col_character(),
      blank = col_double(),
      comment = col_double(),
      code = col_double()
    )
  )

pkgs_runnable_code <- 
  pkgs_runnable_code_raw %>%
  filter(language=="R") %>%
  select(package, path, code, files) %>%
  pivot_wider(
    names_from=path, 
    values_from=c(code, files), 
    values_fill=list(code=0, files=0)
  ) %>%
  # we have old version of tidyr, names_glue is missing, so have to rename manually
  rename(
    runnable_code_examples=code_examples,
    runnable_code_tests=code_tests,
    runnable_code_vignettes=code_vignettes,
    runnable_files_examples=files_examples,
    runnable_files_tests=files_tests,
    runnable_files_vignettes=files_vignettes,
  ) %>%
  mutate(
    runnable_code=runnable_code_examples+runnable_code_tests+runnable_code_vignettes,
    runnable_files=runnable_files_examples+runnable_files_tests+runnable_files_vignettes
  ) %>%
  select(
    package,
    runnable_code,
    runnable_files,
    everything()
  )
```

### Package coverage

```{r load coverage data}
pkgs_coverage <- 
  read_csv(
    PACKAGE_COVERAGE,
    col_types=cols(
      package = col_character(),
      type = col_character(),
      coverage_expression = col_double(),
      coverage_line = col_double(),
      error = col_logical()
    )
  ) %>%
  filter(type=="all", is.na(error), !is.na(coverage_expression)) %>% 
  select(-type, -error, coverage_expr=coverage_expression) %>%
  mutate(coverage_expr=coverage_expr/100, coverage_line=coverage_line/100)
```

### Package runs

```{r package runs}
pkgs_runs_raw <-
  read_csv(
    PACKAGE_RUN,
    col_types=cols(
      package = col_character(),
      file = col_character(),
      type = col_character(),
      time = col_double(),
      error = col_character(),
      output = col_character()
    )
  )
pkgs_runs <-
  pkgs_runs_raw %>%
  group_by(package) %>%
  summarise(runnable_success=sum(is.na(error)), runnable_error=sum(!is.na(error)), runnable_time=sum(time, na.rm=TRUE)) %>%
  mutate(runnable_rate=runnable_success/(runnable_success+runnable_error))
```

### Package functions

```{r load functions}
pkgs_functions <- 
  read_csv(
    PACKAGE_FUNCTIONS,
    col_types=cols(
      package = col_character(),
      fun = col_character(),
      num_params = col_double(),
      has_elipsis = col_logical(),
      exported = col_logical()
    )
  ) %>%
  count(package, exported) %>%
  mutate(type=ifelse(exported, "funs_public", "funs_private")) %>%
  select(-exported) %>%
  spread(key=type, value=n, fill=0) %>%
  mutate(funs=funs_public+funs_private)
```

### Package reverse dependencies

There are two functions `package_dependencies` and `dependsOnPkgs`.
Not sure what is the difference, but the `dependsOnPkgs` is very slow.

```{r compute reverse dependencies}
pkgs_revdeps_raw <- withr::with_libpaths(LIB_DIR, {
  package_dependencies(
    pkgs_installed$package,
    db=pkgs_installed_raw,
    which=params$which_dependencies,
    reverse=TRUE,
    recursive=FALSE
  )
}) %>% imap_dfr(~tibble(package=.y, revdep=.x))

pkgs_revdeps <- count(pkgs_revdeps_raw, package) %>% rename(revdeps=n)
```

### Package asserts

```{r runtime checks}
pkgs_asserts <- 
  read_csv(
    PACKAGE_ASSERTS,
    col_types=cols(
      package = col_character(),
      fun_name = col_character(),
      file = col_character(),
      line1 = col_double(),
      col1 = col_double(),
      line2 = col_double(),
      col2 = col_double(),
      assert = col_character(),
      args = col_character()
    )
  ) %>%
  count(package, fun_name) %>%
  group_by(package) %>%
  summarise(funs_with_assert=length(unique(fun_name)), asserts=sum(n))
```

### Join

```{r joining data}
pkgs <-
  pkgs_cran %>%
  left_join(mutate(pkgs_installed, installed=TRUE), by="package") %>% mutate(installed=replace_na(installed, FALSE)) %>%
  left_join(transmute(pkgs_loadable, package, loadable=TRUE), by="package") %>% mutate(loadable=replace_na(loadable, FALSE)) %>%
  left_join(pkgs_sloc, by="package") %>%
  left_join(pkgs_runnable_code, by="package") %>%
  left_join(pkgs_coverage, by="package") %>%
  left_join(pkgs_runs, by="package") %>%
  left_join(pkgs_functions, by="package") %>%
  left_join(pkgs_revdeps, by="package") %>%
  left_join(pkgs_asserts, by="package") 
```

### Sanity checks

#### Packages that are loaded, but not installed?

```{r}
filter(pkgs, !installed && loadable) %>%
  select(package, installed, loadable)
```

#### Packages that do not have any extracted runnable code, yet they have some runs or coverage?

This can actually be still OK, perhaps the package has some code in its `.onLoad` or `.onAttach` hooks.

```{r}
filter(pkgs, runnable_code==0, coverage_expr > 0) %>%
  select(package, package_code, runnable_code, runnable_files, starts_with("coverage"))
```

### Curiosity

* Packages with no R code (this is perfectly fine):

```{r}
filter(pkgs, package_r_code==0)
```

* Packages with no code at all (also fine - just data packages):

```{r}
filter(pkgs, package_code==0)
```

### Overview

#### Table

```{r summary}
summary_corpus <- function(df, prefix) {
  overview_table(
    r(paste(prefix, "cran"), df),
    r(paste(prefix, "installed"), filter(df, installed)),
    r(paste(prefix, "loadable"), filter(df, loadable)),

    r(paste(prefix, "with runnable code"), filter(df, runnable_code > 0)),
    r(paste(prefix, "runnable code"), sum(df$runnable_code, na.rm=T)),
    r(paste(prefix, "examples code"), sum(df$runnable_code_examples, na.rm=T)),
    r(paste(prefix, "tests code"), sum(df$runnable_code_tests, na.rm=T)),
    r(paste(prefix, "vignettes code"), sum(df$runnable_code_vignettes , na.rm=T)),

    r(paste(prefix, "with coverage"), filter(df, coverage_expr > 0)),
    r(paste(prefix, "mean expr coverage"), percent(mean(df$coverage_expr, na.rm=T))),
    r(paste(prefix, "median expr coverage"), percent(median(df$coverage_expr, na.rm=T))),

    r(paste(prefix, "with R code"), filter(df, package_r_code > 0)),
    r(paste(prefix, "R code"), sum(df$package_r_code, na.rm=T)),
    r(paste(prefix, "with native code"), filter(df, package_native_code > 0)),
    r(paste(prefix, "native code"), sum(df$package_native_code, na.rm=T)),

    r(paste(prefix, "w asserts"), filter(df, asserts > 0)),
    r(paste(prefix, "asserts"), sum(df$asserts, na.rm=T)),
    r(paste(prefix, "funs with asserts"), sum(df$funs_with_assert, na.rm=T)),

    r(paste(prefix, "functions"), sum(df$funs, na.rm=T)),
    r(paste(prefix, "public functions"), sum(df$funs_public, na.rm=T)),
    r(paste(prefix, "private functions"), sum(df$funs_private, na.rm=T)),

    r(paste(prefix, "mean revdes"), mean(df$revdeps, na.rm=T)),
    r(paste(prefix, "median revdes"), median(df$revdeps, na.rm=T)),

    r(paste(prefix, "runtime"), as.duration(sum(df$runnable_time, na.rm=T))),
    r(paste(prefix, "mean runtime"), as.duration(mean(df$runnable_time, na.rm=T))),
    r(paste(prefix, "median runtime"), as.duration(median(df$runnable_time, na.rm=T)))
  )
}

summary_corpus(pkgs, "pkgs")

write_csv(pkgs, ALL_PACKAGES_DETAILS)
```

#### Plots

```{r considered packages}
pkgs_considered <- filter(pkgs, loadable, package_r_code > 0, coverage_expr > 0)
```

```{r plot revdeps histogram}
pkgs_considered %>%
  ggplot(aes(x=revdeps)) + 
    geom_histogram(bins=50) +
    scale_x_log10(labels=scales::comma) +
    labs(x="# reverse dependencies (log)")
```

```{r plot coverage histogram}
pkgs_considered %>%
  ggplot(aes(x=coverage_expr)) + 
    geom_histogram() +
    scale_x_continuous(labels=scales::percent) +
    labs(x="Expression coverage")
```

```{r plot all considered packages}
plot_all <- function(df) {
  coverage_mean <- mean(df$coverage_expr)
  coverage_median <- median(df$coverage_expr)
  revdeps_mean <- mean(df$revdeps)
  revdeps_median <- median(df$revdeps)

  df %>%
    mutate(
      outlier=package_code > 5e4,
      label=if_else(outlier, str_glue("{package}"), as.character(NA))
    ) %>%
    ggplot(
      aes(
        x=coverage_expr, 
        y=revdeps,
        label=label,
        size=package_code
      )
    ) +
    geom_point(alpha=.6) +
    geom_text(size=3, check_overlap = T, vjust=1.5, na.rm = TRUE) +

    geom_hline(aes(yintercept=revdeps_mean), linetype="dashed", color="black", size=0.2) +
    geom_vline(aes(xintercept=coverage_mean), linetype="dashed", color="black", size=0.2) +
    geom_vline(aes(xintercept=params$min_coverage),linetype="solid", color="red", size=0.4) +
    geom_hline(aes(yintercept=params$min_revdeps),linetype="solid", color="red", size=0.4) +

    scale_x_continuous(labels = scales::percent) + 
    scale_y_log10(labels = scales::comma) +
    scale_size(range = c(.1, 5), labels = scales::comma) +
      theme(
      legend.position=c(0.2, 0.85),
      legend.box="horizontal",
      legend.box.background = element_rect(fill="white", size=0.1)
    ) +
    labs(
      x="Code coverage", 
      y="Number of reverse dependencies (log)", 
      size="Code size",
      linetype=""
    ) +
    guides(
      size=guide_legend(order=1),
      color=guide_legend(order=2),
      linetype=F
    )
}

plot_all(pkgs_considered)
ggsave(CORPUS_PLOT)
```

## Project corpus

We only consider packages that have at least `r params$min_revdeps` reverse dependencies and `r params$min_coverage*100`% expression coverage.

```{r project corpus}
corpus <- 
  filter(
    pkgs, 
    loadable, 
    runnable_code > 0, 
    revdeps >= params$min_revdeps, 
    package_r_code > 0, 
    coverage_expr >= params$min_coverage
  )

write_lines(corpus$package, CORPUS_FILE)
write_csv(corpus, CORPUS_DETAILS)
```

### Sanity check

TODO

### Overview

#### Table

```{r corpus summary}
summary_corpus(corpus, "corpus")

write_csv(corpus, CORPUS_DETAILS)
write_lines(corpus$package, CORPUS_FILE)
```

#### Plot

```{r plot corpus}
plot_corpus <- function(df, code_outlier=5e3) {
  coverage_mean <- mean(df$coverage_expr)
  coverage_median <- median(df$coverage_expr)
  revdeps_mean <- mean(df$revdeps)
  revdeps_median <- median(df$revdeps)

  df %>%
    mutate(
      outlier=package_code > code_outlier,
      label=if_else(outlier, str_glue("{package}"), as.character(NA))
    ) %>%
    ggplot(
      aes(
        x=coverage_expr, 
        y=revdeps,
        label=label,
        size=package_code
      )
    ) +
    geom_point(alpha=.6) +
    geom_text(size=3, check_overlap = T, vjust=1.5, na.rm = TRUE) +

    geom_hline(aes(yintercept=revdeps_mean), linetype="dashed", color="red", size=0.2) +
    geom_vline(aes(xintercept=coverage_mean), linetype="dashed", color="red", size=0.2) +

    scale_x_continuous(limits = c(params$min_coverage, 1), labels = scales::percent) + 
    scale_y_log10(labels = scales::comma) +
    scale_size(range = c(.1, 3), labels = scales::comma) +
      theme(
      legend.position=c(0.2, 0.85),
      legend.box="horizontal",
      legend.box.background = element_rect(fill="white", size=0.1)
    ) +
    labs(
      x="Code coverage", 
      y="Number of reverse dependencies (log)", 
      size="Code size",
      linetype=""
    ) +
    guides(
      size=guide_legend(order=1),
      color=guide_legend(order=2),
      linetype=F
    )
}

plot_corpus(corpus)
```

### List of reverse dependencies

```{r list of reverse depenencies}
corpus_revdeps <- pkgs_revdeps_raw %>% semi_join(corpus, by="package") %>% distinct(revdep)
write_lines(corpus_revdeps, CORPUS_REVDEPS)
```

### List of corpus packages

```{r list of packages}
datatable(
  corpus, 
  filter="top", 
  options=list(
    pageLength=50,
    autoWidth=TRUE
  )
)
```

## Appendix: Most downloaded CRAN packages

```{r top cran downloads}
df_top_cran <- cran_top_downloads(when="last-month", count=100)
```

The cranlogs allow us to get 100 most downloaded packages over the last month.
How many of these popular packages do we have in out corpus?

```{r}
missed_top_cran_packages <- setdiff(df_top_cran$package, corpus$package)
```

```{r}
filter(df_top_cran, package %in% missed_top_cran_packages) %>%
  select(-from, -to) %>%
  datatable()
```

The main reason is that we use R 3.5. Some key dependencies such as Hmisc are no longer available.

### 2020-05-08 evaluation/R/runtime-checks.Rmd
Add analysis Rmds

---
title: "Runtime checks"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(DT)
library(fs)
library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)
```

```{r paths, include=FALSE}
source("inc/paths.R")
```

## Setup

The `r RUNTIME_CHECKS` file contains information about the use of assertions across packages.
To generate it, run the following:

```
 ./rapr/inst/on-each-package.sh ./rapr/inst/tasks/find-runtime-type-checks.R
```

or (with a `PACKAGES` variable set to restrict the package selection)

```
PACKAGES=packages-typer-35.txt ./rapr/inst/on-each-package.sh ./rapr/inst/tasks/find-runtime-type-checks.R
```

and then

```
env-rscript ./rapr/inst/process-csv-result.R run/find-runtime-type-checks runtime-checks.csv
```

to generate the CSV file.

## Data

```{r load run}
run <- read_csv(RUNTIME_CHECKS_RUN)
run <- semi_join(run, CORPUS_PACKAGES_DF, by=c("job"="package"))
```

```{r check run packages}
#stopifnot(nrow(run) == nrow(CORPUS_PACKAGES_DF))
```

```{r load runtime checks}
runtime_checks <- read_csv(RUNTIME_CHECKS)
runtime_checks <- semi_join(runtime_checks, CORPUS_PACKAGES_DF, by="package")
```

## Analysis

### By package

```{r}
per_package <- count(runtime_checks, package)
```

```{r}
per_package %>%
  arrange(desc(n)) %>%
  datatable()
```

```{r}
ggplot(per_package, aes(n)) + geom_histogram(binwidth = 1)
```

### By package function

```{r}
per_package_fun <- count(runtime_checks, package, fun_name)
```

```{r}
per_package_fun %>%
  arrange(desc(n)) %>%
  datatable()
```

```{r}
ggplot(per_package_fun, aes(n)) + geom_histogram(binwidth = 1)
```

### By asertion function

```{r}
per_assert <- count(runtime_checks, assert)
```

```{r}
per_assert %>%
  arrange(desc(n)) %>%
  datatable()
```

## Raw data

```{r}
datatable(
  runtime_checks, 
  filter="top", 
  options=list(
    pageLength=50,
    autoWidth=TRUE
  )
)
```

### 2020-05-08 evaluation/R/validation.Rmd
Add analysis Rmds

---
title: "Data and Validation"
output: html_notebook
---

Setup (this grabs Alexi's data analysis script. Ask him to move it if you're gonna use the notebook.)

```{r setup}
source("/mnt/home/pldi/scripts/support_scripts/data_analysis_lib.R")
```

Load the data.

```{r}
# df <- read_csv("/mnt/arraySSD/alexi/2020_03_14_candidate_reduced_20_simpl/all.csv")
df <- read_csv("/mnt/arraySSD/alexi/2020_03_14_candidate_reduced_20_simpl/all_twelfth.csv")
```

Now, onto the validation.
For this section, I'll include the queries that were used to validate the decisions we made when designing the type system.
The end goal will be to generate some latex to be included by the paper which should have the numbers, that way the paper is totally reproducible.
First, we get the data frame which counts instances of types.

```{r}
df_c <- df %>% quick_get_type_counts_with_row_nums
df_c <- df_c %>% ungroup
```

Let's start by breaking down the counts for tuples, lists, and structs.

```{r}
num_lists <- df_c %>% filter(substr(type, 1, 4) == "list") %>% select(count) %>% sum
# We don't have tuples per se, actually.
# num_tuples <- df_c %>% filter(substr(type, 1, 5) == "tuple") %>% select(count) %>% sum
# Data frames?
num_data_frame <- df_c %>% filter(substr(type, 1, 4) == "data") %>% select(count) %>% sum
num_structs <- df_c %>% filter(substr(type, 1, 6) == "struct") %>% select(count) %>% sum

total_listy <- num_lists + num_data_frame + num_structs

summary_listy <- data.frame(type = c("list", "data.frame", "struct"), count = c(num_lists, num_data_frame, num_structs), percentage = round(c(num_lists, num_data_frame, num_structs) / total_listy * 100, 2))

summary_listy
```

The above counts treat data.frames and structs differently, though they should be the same.
We also don't do tuples, they don't seem particularly relevant.

Let's grab counts for usage of scalar-variants of primitive types.

```{r}
df_c_just_primitives <- df_c %>% filter(substr(type, 1, 5) %in% substr(PRIM_TYPES, 1, 5))

num_scalars_no_names <- df_c_just_primitives %>% filter(type %in% paste0(PRIM_TYPES, "@NA-free")) %>% select(count) %>% sum
num_scalars_names <- df_c_just_primitives %>% from_counts_df_extract_dim_and_names %>% filter(num_names == 1) %>% select(count) %>% sum
num_vec_no_names <- df_c_just_primitives %>% filter(type %in% paste0(PRIM_TYPES, "[]") | type %in% paste0(PRIM_TYPES, "[]@NA-free")) %>% select(count) %>% sum
num_mat <- df_c_just_primitives %>% filter(type %in% paste0(PRIM_TYPES, "[,]") | type %in% paste0(PRIM_TYPES, "[,]@NA-free")) %>% select(count) %>% sum
num_vec_names <- df_c_just_primitives %>% from_counts_df_extract_dim_and_names %>% filter(num_names > 1) %>% select(count) %>% sum

total_prims <- df_c_just_primitives %>% select(count) %>% sum

summary_primitives <- data.frame(type = c("nameless scalar", "named scalar", "nameless vector", "named vector", "matrices"), count = c(num_scalars_no_names, num_scalars_names, num_vec_no_names, num_vec_names, num_mat), percentage = round( c(num_scalars_no_names, num_scalars_names, num_vec_no_names, num_vec_names, num_mat) / total_prims * 100, 2)) %>% as_tibble

summary_primitives
```

TODO: Look into why the percentages don't add up to 1. Maybe we missed something?

Next, let's try for ...

...

We can also synthesize type signatures from the data.
Let's try it:

First, an abstraction:

```{r}
get_fun_types_for_package <- function(df, pkg) {
  # First, filter out for a particular package.
  df_p <- df %>% filter(package == pkg)

  # What are the function names?
  p_fun_names <- df_p %>% select(fun_name) %>% unique %>% unlist %>% unname
  p_fun_names <- p_fun_names[!is.na(p_fun_names)]

  cat("p_fun_names: ")
  cat(p_fun_names, collapse=" ")
  cat("\n")

  # Let's build a type for each of these.
  names(p_fun_names) <- p_fun_names
  map(p_fun_names, function(x) {
    cat("processing ", x, "...\n");
    fn_id <- df_p %>% filter(fun_name == x) %>% select(fun_id) %>% unique %>% unlist %>% unname
    paste0(fn_id, ": ", get_type_of_function(df_p, pkg, x))
  })
}
```

Do it for some packages.
These are long-running, we split up the computation.

For dplyr:

```{r}
dplyr_types <- get_fun_types_for_package(df, "dplyr")
```

For the abind package:

```{r}
abind_types <- get_fun_types_for_package(df, "abind")
```

For the alluvial package:

```{r}
alluvial_types <- get_fun_types_for_package(df, "alluvial")
```

For the bayesplot package:

```{r}
bayesplot_types <- get_fun_types_for_package(df, "bayesplot")
```

```{r}
# Don't run this.
base_types <- get_fun_types_for_package(df, "base")
```

### 2020-05-08 paper/main.tex
initial class discussion

@@
 
+Another pattern that we observed was the use of consistently-named lists.
+In R, ``fields'' of named lists can be accessed using the \code{$} operator, such as:
+\begin{lstlisting}
+p <- list(x=1, y=2)
+p$x + p$y
+\end{lstlisting}
+We thought to capture this type of behaviour with a {\it struct} type, such as \code{struct<x: integer, y: integer>}.
+Initial analyses suggested that structs were very common, but further inspection revealed that they failed to tell the full story.
+For instance, \AT{many} structs were merely representing list-typed values with a {\it class}, and we could capture these types more succinctly with a class-based type.
+Further, it is common in R example code to use some built-in data sets, such as \AT{mention some}.
+These caused undue noise, as even with our struct name unification strategy, poorly tested functions appeared to operate on e.g. \AT{funny struct type from the previous data set}.
+Ultimately, we found that sticking to list types and adding a class-based was the wisest course of action.
+We will discuss class types shortly.
+
 %
@@
 \subsubsection{Forgoing Data Frames}
+
+Speaking of classes, an important classful value in R is the data frame.
+Data frames underpin nearly all idiomatic use of the language, and even popular \code{tibble} (from the \code{tidyverse} ecosystem) and \code{data.table} (from the \code{data.table} package) build on the essential type.
+We entertained the possibility of including data frame types, but \AT{found them to be clunky? Question mark?}
+Consider: \AT{example of why data.frame types would be bad atm.}
+
+Data frames do have a type, through the \code{class<data.frame>} type, and future work will expand on this type, discussed further in Section~\ref{sec:futurework}.
+
+%
+%
+%
+%
+\subsection{Classes in R}
+
+\AT{Fill in.}
+
+%
+%
+\subsubsection{Class Types}
+
+\AT{We have classes.}
+
+%
+%
+\subsubsection{Multi-Classes}
+
+\AT{We support multiple classes on values.}
+
+%
+%
+\subsubsection{Forgoing R's Object Systems}
+
+\AT{Roughly: too complicated. We have rudimentary class types for S3, and an S4 type for S4.
+Future work will explore further.}
 
@@
 \section{Evaluation Results}
+\label{sec:evaluation}
 
@@
 
+%
+%
+%
+%
+%
+%
 \section{Related Work}
+\label{sec:relatedwork}
 
@@
 
+%
+%
+%
+%
+%
+%
 \section{Conclusion}
@@
 
+%
+%
+%
+%
 \subsection{Threats to Validity}
+\label{sec:threatstovalidity}
 
@@
 
+%
+%
+%
+%
 \subsection{Future Work}
+\label{sec:futurework}
 

### 2020-05-09 paper/main.tex
disparate labels, and comments
cleanup

@@
-%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
 \documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For double-blind review submission, w/ CCS and ACM Reference
-%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
-%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For single-blind review submission, w/ CCS and ACM Reference
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
-%% For final camera-ready submission, w/ required CCS and ACM Reference
-%\documentclass[acmsmall]{acmart}\settopmatter{}
-
-%% Journal information
-%% Supplied to authors by publisher for camera-ready submission;
-%% use defaults for review submission.
 \acmJournal{PACMPL}
@@
 \startPage{1}
+\setcopyright{none}
+\bibliographystyle{ACM-Reference-Format}
+\citestyle{acmauthoryear}   %% For author/year citations
+\usepackage{my_style}
+\usepackage{listings, wrapfig,xspace}
 
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission;
-%% use 'none' for review submission.
-\setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2018}           %% If different from \acmYear
-
-%% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%% Note: author/year citations are required for papers published as an
-%% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Note: Authors migrating a paper from PACMPL format to traditional
-%% SIGPLAN proceedings format must update the '\documentclass' and
-%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Some recommended packages.
-\usepackage{booktabs}   %% For formal tables:
-                        %% http://ctan.org/pkg/booktabs
-\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
-                        %% http://ctan.org/pkg/subcaption
-\usepackage{my_style}
-
-\usepackage{listings, wrapfig}
-
-\lstset{ 
-  language=R,                     % the language of the code
-  basicstyle=\tiny\ttfamily, % the size of the fonts that are used for the code
-  numbers=left,                   % where to put the line-numbers
-  numberstyle=\tiny\color{blue},  % the style that is used for the line-numbers
-  stepnumber=1,                   % the step between two line-numbers. If it is 1, each line
-                                  % will be numbered
-  numbersep=5pt,                  % how far the line-numbers are from the code
-  backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
-  showspaces=false,               % show spaces adding particular underscores
-  showstringspaces=false,         % underline spaces within strings
-  showtabs=false,                 % show tabs within strings adding particular underscores
-  frame=single,                   % adds a frame around the code
-  rulecolor=\color{black},        % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. commens (green here))
-  tabsize=2,                      % sets default tabsize to 2 spaces
-  captionpos=b,                   % sets the caption-position to bottom
-  breaklines=true,                % sets automatic line breaking
-  breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
-  keywordstyle=\color{teal},      % keyword style
-  commentstyle=\color{olive},   % comment style
-  stringstyle=\color{magenta}      % string literal style
+\lstset{language=R}
+\definecolor{LightGray}{rgb}{.92,.92,.92}
+\definecolor{Gray}{rgb}{.3,.3,.3}
+\definecolor{DarkGray}{rgb}{.5,.5,.5}
+\lstset{ %
+  columns=flexible,
+  captionpos=b,
+  frame=single,
+  framerule=0pt,
+  tabsize=2,
+  belowskip=0.5em,
+  backgroundcolor=\color{LightGray},
+  basicstyle=\small\ttfamily,
+  emphstyle=,
+  keywordstyle=,
+  commentstyle=\color{Gray}\em,
+  stringstyle=\color{Gray}
+}
+\lstdefinestyle{R}{ %
+  language=R,
+  morekeywords={assign, delayedAssign},
+  deletekeywords={env, equal, c, runif, trace, args, exp, t, all},
+  breaklines=true
+}
+\lstdefinestyle{Rin}{ %
+  style=R,
+  breaklines=false
 }
 
-\newcommand{\code}[1]{\text{\lstinline[basicstyle=\ttfamily\smaller]~#1~}}
+\newcommand{\eg}{\emph{e.g.},\xspace}
+\newcommand{\ie}{\emph{i.e.},\xspace}
 
+\newcommand{\PIR}{\textsf{PIR}\xspace}
+\newcommand\pirI[1]{\mathtt{#1}}
+\renewcommand{\c}[1]{{\lstinline[style=Rin]!#1!}\xspace}
+\newcommand{\code}[1]{{\lstinline[style=Rin]!#1!}\xspace}
 % Macros for type names from the old paper.
@@
 \begin{document}
-
-%% Title information
-\title[]{Types for R, Empirically Designed and Evaluated}         %% [Short Title] is optional;
-                                        %% when present, will be used in
-                                        %% header instead of Full Title.
-% \titlenote{}             %% \titlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
-% \subtitle{}                     %% \subtitle is optional
-% \subtitlenote{}       %% \subtitlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
-
-%% Author information
-%% Contents and number of authors suppressed with 'anonymous'.
-%% Each author should be introduced by \author, followed by
-%% \authornote (optional), \orcid (optional), \affiliation, and
-%% \email.
-%% An author may have multiple affiliations and/or emails; repeat the
-%% appropriate command.
-%% Many elements are not rendered, but should be provided for metadata
-%% extraction tools.
-
-%% Author with single affiliation.
-\author{First1 Last1}
-\authornote{with author1 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
-\affiliation{
-  \position{Position1}
-  \department{Department1}              %% \department is recommended
-  \institution{Institution1}            %% \institution is required
-  \streetaddress{Street1 Address1}
-  \city{City1}
-  \state{State1}
-  \postcode{Post-Code1}
-  \country{Country1}                    %% \country is recommended
-}
-\email{first1.last1@inst1.edu}          %% \email is recommended
-
-%% Author with two affiliations and emails.
-\author{First2 Last2}
-\authornote{with author2 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
-\affiliation{
-  \position{Position2a}
-  \department{Department2a}             %% \department is recommended
-  \institution{Institution2a}           %% \institution is required
-  \streetaddress{Street2a Address2a}
-  \city{City2a}
-  \state{State2a}
-  \postcode{Post-Code2a}
-  \country{Country2a}                   %% \country is recommended
-}
-\email{first2.last2@inst2a.com}         %% \email is recommended
-\affiliation{
-  \position{Position2b}
-  \department{Department2b}             %% \department is recommended
-  \institution{Institution2b}           %% \institution is required
-  \streetaddress{Street3b Address2b}
-  \city{City2b}
-  \state{State2b}
-  \postcode{Post-Code2b}
-  \country{Country2b}                   %% \country is recommended
-}
-\email{first2.last2@inst2b.org}         %% \email is recommended
-
-%% Abstract
-%% Note: \begin{abstract}...\end{abstract} environment must come
-%% before \maketitle command
+\title{Types for R, Empirically Designed and Evaluated}
 \begin{abstract}
@@
 \end{abstract}
-
-%% 2012 ACM Computing Classification System (CSS) concepts
-%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
-\begin{CCSXML}
-<ccs2012>
-<concept>
-<concept_id>10011007.10011006.10011008</concept_id>
-<concept_desc>Software and its engineering~General programming languages</concept_desc>
-<concept_significance>500</concept_significance>
-</concept>
-<concept>
-<concept_id>10003456.10003457.10003521.10003525</concept_id>
-<concept_desc>Social and professional topics~History of programming languages</concept_desc>
-<concept_significance>300</concept_significance>
-</concept>
-</ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Software and its engineering~General programming languages}
-\ccsdesc[300]{Social and professional topics~History of programming languages}
-%% End of generated code
-
-%% Keywords
-%% comma separated list
-\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
-
-%% \maketitle
-%% Note: \maketitle command must come after title commands, author
-%% commands, abstract environment, Computing Classification System
-%% environment and commands, and keywords command.
 \maketitle
@@
 \label{sec:contractr}
+
+\AT{I think we should move this to the Methodology section, which we should maybe rename to implementation. Thoughts?}
+
 \contractr is an R package that implements our type system for the R
@@
 \subsection{The Corpus}
+\label{sec:corpus}
 
@@
 In order to ensure that our type system design aligns with the day-to-day usage of R, we performed a corpus analysis of some of the most widely used R packages.
+The corpus is discussed in detail in Section~\ref{sec:corpus}.
 We built our dynamic analysis in the R-dyntrace~\cite{oopsla19} dynamic analysis framework, which is itself built on a modified version of R targeting R version 3.5.0---we refer the reader to Section~\ref{sec:r-dyntrace} for more details on R-dyntrace.
@@
 \subsection{Threats to Validity}
-\label{sec:threatstovalidity}
 
-\ldots
+\subsection{Future Work}
 
-%
-%
-%
-%
-\subsection{Future Work}
-\label{sec:futurework}
-
-\ldots
-
-%% Acknowledgments
-\begin{acks}                            %% acks environment is optional
-                                        %% contents suppressed with 'anonymous'
-  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
-  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
-  %% acknowledge financial support and will be used by metadata
-  %% extraction tools.
-  This material is based upon work supported by the
-  \grantsponsor{GS100000001}{National Science
-    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-  conclusions or recommendations expressed in this material are those
-  of the author and do not necessarily reflect the views of the
-  National Science Foundation.
-\end{acks}
-
-%% Bibliography
 \bibliography{bib/biblio,bib/jv,bib/r,bib/new,bib/gradual}
 
-%% Appendix
+\end{document}
 \appendix
 \section{Appendix}
-
-Text of appendix \ldots
-
-\end{document}

### 2020-05-09 paper/main.tex
merge

--- 2020-05-08 paper/main.tex
@@
 In order to ensure that our type system design aligns with the day-to-day usage of R, we performed a corpus analysis of some of the most widely used R packages.
-We built our dynamic analysis in the R-dyntrace~\cite{oopsla19} dynamic analysis framework, which is itself built on a modified version of R targeting R version 3.5.0---we refer the reader to Section~\ref{sec:r-dyntrace} for more details on R-dyntrace.
+The corpus is discussed in detail in Section~\ref{sec:corpus}.
+We built our dynamic analysis in the R-dyntrace~\cite{oopsla19} dynamic analysis framework, which is itself built on a modified version of R targeting R version 3.5.0---we refer the reader to Section~\ref{sec:r-dyntrace} for more details on the framework.
 
-The general idea of the analysis is to intercept calls to functions, builtins, and specials, and collect information on function arguments and returns.
+The goal of our analysis is to discover argument types for functions, builtins, and specials.
+The naive approach is to intercept calls to functions, builtins, and specials, recording function argument types, but recall that R wraps arguments in promises (Section \AT{what section}).
+To circumvent this obstacle, we intercept the forcing of promises, identify the function call that they have been forced in, collect the type of the value, and add it to the call trace associated with the current function call.
+Once a function, builtin, or special completes execution, we collect information about the return value, add it to the call trace, and save the call trace to be written out later. 
+Particular attention has been paid to the numerous corner cases present in the R runtime, which we discuss later in this section.
+
 The nature of R is such that all function arguments are wrapped in promises which are only evaluated when the value is actually used.
 This could cause issues in a dynamic analysis, as careless use of function arguments can force them early, which can cause unindented side-effects (as these promises are not necessarily pure).
-Our analysis deals with this by only inspecting the value of an argument when the argument is forced, and if it is never forced we make note of that by ascribing the type {\tt ???}.
-
-% Our analysis deals with this by inspecting the {\it value} slot of these promise objects, which does not force the promise. 
-% If the value slot contains another promise, we look into its value slot, and so forth until we find a value (or not! in which case, the value is a promise).
 

### 2020-05-09 paper/main.tex

--- 2020-05-09 paper/main.tex
@@
 The corpus is discussed in detail in Section~\ref{sec:corpus}.
-We built our dynamic analysis in the R-dyntrace~\cite{oopsla19} dynamic analysis framework, which is itself built on a modified version of R targeting R version 3.5.0---we refer the reader to Section~\ref{sec:r-dyntrace} for more details on R-dyntrace.
+We built our dynamic analysis in the R-dyntrace~\cite{oopsla19} dynamic analysis framework, which is itself built on a modified version of R targeting R version 3.5.0---we refer the reader to Section~\ref{sec:r-dyntrace} for more details on the framework.
 
-The general idea of the analysis is to intercept calls to functions, builtins, and specials, and collect information on function arguments and returns.
+The goal of our analysis is to discover argument types for functions, builtins, and specials.
+The naive approach is to intercept calls to functions, builtins, and specials, recording function argument types, but recall that R wraps arguments in promises (Section \AT{what section}).
+To circumvent this obstacle, we intercept the forcing of promises, identify the function call that they have been forced in, collect the type of the value, and add it to the call trace associated with the current function call.
+Once a function, builtin, or special completes execution, we collect information about the return value, add it to the call trace, and save the call trace to be written out later. 
+Particular attention has been paid to the numerous corner cases present in the R runtime, which we discuss later in this section.
+
 The nature of R is such that all function arguments are wrapped in promises which are only evaluated when the value is actually used.
 This could cause issues in a dynamic analysis, as careless use of function arguments can force them early, which can cause unindented side-effects (as these promises are not necessarily pure).
-Our analysis deals with this by only inspecting the value of an argument when the argument is forced, and if it is never forced we make note of that by ascribing the type {\tt ???}.
-
-% Our analysis deals with this by inspecting the {\it value} slot of these promise objects, which does not force the promise. 
-% If the value slot contains another promise, we look into its value slot, and so forth until we find a value (or not! in which case, the value is a promise).
 

### 2020-05-09 paper/main.tex
initial class discussion

@@
 \subsection{The R Programming Language}
+\label{sec:R}
 
@@
 
-\AT{Fill in.}
+As discussed in Section~\ref{sec:R}, the R runtime has more than one notion of type.
+The {\it class} (rather, the {\it classes}, as values can have multiple!) of function arguments are used by R to {\it dispatch} the correct function call for the argument \AT{$\leftarrow$ wording}.
+This is best served with an example:
+\AT{Example, unless we talked about this earlier, in which case refer reader back.}
 
@@
 
-\AT{We have classes.}
+The class of a value suggest a shape for the value, and we felt class to be a more succinct way of capturing struct-like types.
+
+\AT{We used classes to stand-in for structs.
+The discussion here is naturally tied with the discussion of structs, so we need to make sure to divide the discussion appropriately.}
 
@@
 
-\AT{We support multiple classes on values.}
+As R does, our type system supports multiple classes on values, and the type of a classful value is simply a list of all of its classes.
+
+\AT{What more to say here? Discussing occurrences of multi-classes?}
 
@@
 
-\AT{Roughly: too complicated. We have rudimentary class types for S3, and an S4 type for S4.
-Future work will explore further.}
+R has multiple, disparate object systems, with the language providing direct support for S3, S4, and R5.
+Thus far, we've mainly discussed S3 and its dispatch mechanism, and indeed our class types roughly capture the S3 ``types'' of values.
+We will not discuss R5 as it is still under development.
+
+For S4, we have an \code{s4} type to denote S4 values, though we did not elaborate on this type.
+We stuck to simple S4 types since doing anything more would be immensely complicated, and outside of the scope of this work.
+For instance, the mechanics of S4 dispatch are more complex than for S3 (S4 dispatches on the class of {\it all} arguments), and users can define their own class hierarchies that we would need to incorporate in our type analysis and contract checking frameworks.
+Further, \AT{we found limited use of S4 during our analysis}, with S4 types being heavily used in a select subset of packages (\AT{such as ...}), some of which make heavy use of R's metaprogramming capabilities.
+Coming up with a type system that accounts for all of these factors, and consolidates multiple object-orientation frameworks, is an interesting problem in and of itself, and we aim to extend our type system to account for this in future work.
+For now, S3 types suffice: S3 is the most common object system in R, with no sophisticated hierarchies, and its semantics are simple enough to be captured by our proposed design.
 

### 2020-05-09 evaluation/R/package-corpus.Rmd
Update corpus analysis - revdeps, roxygen, renaming

@@
 
-create_tags(path(TAGS_DIR, "package-corpus.tex"), prefix="PC", default=TRUE)
+create_tags(path(TAGS_DIR, "package-corpus.tex"), prefix="", default=TRUE)
 ```
@@
 
+### Package roxygen
+
+```{r}
+pkgs_roxygen <-
+  read_csv(
+    PACKAGE_USES_ROXYGEN,
+    col_types=cols(
+      package = col_character(),
+      uses_roxygen = col_logical()
+    )
+  )
+```
+
 ### Join
@@
   left_join(pkgs_revdeps, by="package") %>%
-  left_join(pkgs_asserts, by="package") 
+  left_join(pkgs_asserts, by="package") %>%
+  left_join(pkgs_roxygen, by="package")
 ```
@@
     r(paste(prefix, "mean runtime"), as.duration(mean(df$runnable_time, na.rm=T))),
-    r(paste(prefix, "median runtime"), as.duration(median(df$runnable_time, na.rm=T)))
+    r(paste(prefix, "median runtime"), as.duration(median(df$runnable_time, na.rm=T))),
+
+    r(paste(prefix, "using roxygen"), sum(df$uses_roxygen, na.rm=T))
   )
@@
 ```{r list of reverse depenencies}
-corpus_revdeps <- pkgs_revdeps_raw %>% semi_join(corpus, by="package") %>% distinct(revdep)
-write_lines(corpus_revdeps, CORPUS_REVDEPS)
+corpus_revdeps <- pkgs_revdeps_raw %>% semi_join(corpus, by="package")
+write_csv(corpus_revdeps, CORPUS_REVDEPS)
+write_lines(unique(corpus_revdeps$revdep), CORPUS_REVDEPS_UNIQUE)
 ```

### 2020-05-09 paper/main.tex
pushing partial update to initial tracer section

@@
 \newcommand{\roxygen}{\emph{roxygen2}\xspace} % roxygen
+\newcommand{\typetracer}{\emph{typetracer}\xspace} % typetracer
 
@@
   definitions. Thus, there could be many potential sub-expressions in the function
-  body which need to be type-checked aginst the return type. To address this,
+  body which need to be type-checked against the return type. To address this,
   \contractr registers the type-checker to be called on the return value
@@
 %
-\subsection{Initial Analysis}
+\subsection{typetracer and Initial Analysis}
 
-\AT{I'll revisit this section to adjust for the final details, as the analysis has changed since this was written.}
+\AT{So, we understand that we iterated on our type system design.
+Subsequently, the analysis has undergone many iterations, but essentially it collects the most detailed information possible.
+Do we want to mention all of that information, and how we ignore it later?
+Or do we just want to focus on the relevant information for our final types?}
+
+To help inform the design of our type system, and to ensure that it aligns with the day-to-day usage of R, we performed a corpus analysis of some of the most widely used R packages.
+The corpus is discussed in detail in Section~\ref{sec:corpus}.
+We built \typetracer, our dynamic analysis, in the R-dyntrace~\cite{oopsla19} dynamic analysis framework which is itself built on a modified version of R targeting R version 3.5.0---we refer the reader to Section~\ref{sec:r-dyntrace} for more details on the framework.
 
@@
 \subsubsection{Overview}
+The goal of our analysis is to discover argument types for functions, builtins, and specials.
+Naively intercepting function calls and recording argument types is incorrect, as recall that R wraps arguments in promises (Section \AT{what section}).
+To circumvent this obstacle, we instead intercept the forcing of promises: when a promise is forced, we identify the surrounding call, collect the type of the value, and add it to the call trace associated with the current function call.
+Once function execution completes, we collect information about the return value, add it to the call trace, and save the completed call trace to be written out later. 
+\AT{This paragraph overlaps with the initial discussion of \contractr. We need to make sure that we don't inadvertently repeat ourselves.}
 
-In order to ensure that our type system design aligns with the day-to-day usage of R, we performed a corpus analysis of some of the most widely used R packages.
-The corpus is discussed in detail in Section~\ref{sec:corpus}.
-We built our dynamic analysis in the R-dyntrace~\cite{oopsla19} dynamic analysis framework, which is itself built on a modified version of R targeting R version 3.5.0---we refer the reader to Section~\ref{sec:r-dyntrace} for more details on the framework.
-
-The goal of our analysis is to discover argument types for functions, builtins, and specials.
-The naive approach is to intercept calls to functions, builtins, and specials, recording function argument types, but recall that R wraps arguments in promises (Section \AT{what section}).
-To circumvent this obstacle, we intercept the forcing of promises, identify the function call that they have been forced in, collect the type of the value, and add it to the call trace associated with the current function call.
-Once a function, builtin, or special completes execution, we collect information about the return value, add it to the call trace, and save the call trace to be written out later. 
-Particular attention has been paid to the numerous corner cases present in the R runtime, which we discuss later in this section.
-
-The nature of R is such that all function arguments are wrapped in promises which are only evaluated when the value is actually used.
-This could cause issues in a dynamic analysis, as careless use of function arguments can force them early, which can cause unindented side-effects (as these promises are not necessarily pure).
-
-Once the value is obtained in this way, we collect the following high-level information about it:
-\AT{TODO: need to make the inline code macro and the R listing, and change the tts in here}
+%
+%
+\subsubsection{Type Information}
+Once a value is obtained, either on function return or when an argument promise is forced, we probe the value for the following information:
 
 \begin{itemize}
-\item the result of calling the {\tt typeof} function on the value; in other words, the type that the value has according to R's runtime;
-\item its class, which is a list of string class names;
+\item the value's type tag according to R's runtime. R's runtime implements type tags as a typedef, e.g. \code{INTSXP} for integers, and, confusingly, \code{VECSXP} for lists;
+\item its class, which is a list of string class names.
+Typically, the class of a value is held in an attribute with name ``class'', though some classes are blessed by the R runtime, such as matrix and array classes;
 \item its attributes, a list of metadata attached to the value.
@@
 
-Now, depending on the ``type'' of the value, we collect further information:
-If the value is \ldots
+Now, depending on the value's type tag, we collect further information:
 
 \begin{itemize}
-\item one of R's primitive types, we collect its dimension (and if it is a matrix, its dimensions), and if it has a names attribute we collect the names as well;
-\item a list, we recursively collect the types of list elements;
-\item a list {\it with a names attribute}, we collect the names as well as the types of the slots that those names refer to (data.frames fall into this category);
-\item a closure, we simply tag it with a UID computed from the code of the closure as well as its name, which can be later cross-referenced with the analysis results to get a type for the function.
+\item For R's primitive types, we collect its dimension, and if it has a names attribute we collect the names as well;
+\item For lists, we recursively collect the types of list elements;
+\item For lists {\it with a names attribute}, we collect the names as well as the types of the slots that those names refer to (data.frames fall into this category);
+\item For promises, we carefully dig into the promise value {\it without forcing it}, and collect information about the innermost value.
+If the innermost value is the unbound value, then the promise is an as-of-yet unevaluated expression.
+In this case, we do our best to guess the value of the unevaluated expression: if the expression is not a language expression, we collect information about it, otherwise we can say nothing without evaluating it and ascribe it type ``any''.
+\footnote{\AT{I think this is confusing, reword.}}
 \end{itemize}
 
-Some further miscellaneous information we collect includes:
+For vectors (and lists), we additionally iterate through the elements, checking if the values are NA (resp.
+NULL); if none are NA (resp.
+NULL), we ascribe the NA-free (resp.
+NULL-free) tag to the value.
+Arguments that are part of a function's varargs (denoted \ldots in function definitions) are tagged as such and ignored, as varargs can contain anything (and their type will be equivalent to and any-typed list).
+
+Now that we've seen the high-level picture of our approach, we will elaborate on some key details and discuss particular challenges that we faced.
+
+%
+%
+\subsubsection{Challenges}
+
+We grappled with a number of corner cases related to R's non-strict semantics.
 
 \begin{itemize}
-\item lists which do not contain any NULL elements are qualified with a NULL-free tag; 
-\item similarly, vectors which contain no NAs are qualified with a NA-free tag;
-\item promises which are never forced throughout function execution are tagged as being {\it metaprogrammed}---this is a common practice in the {\tt tidyverse} suite of packages;
-\item formal arguments (arguments which are named in the function definition) which are not passed are recorded as {\it missing}---these represent optional arguments.
+
+\item First, we wanted to ensure that \typetracer captured type for unevaluated arguments. 
+Not every argument of a function needs to be used during every function execution, and the approach we discussed above relies on argument promises being forced (and, consequently, on arguments being used).
+There are a few reasons for this.
+
+One possibility is that the argument was simply unused given the control flow of this particular execution (e.g. if the argument is only needed in one branch of an \code{if}).
+This mainly occurs when arguments have default values for some parameters that are not critical to every execution oft he function.
+In these cases, we still want to collect the type of the default value.
+
+Another possibility is that a value is used {\it without being forced}, as is the case when values are ``metaprogrammed'': \AT{describe, code example.}
+
+To tackle this, we make an {\it initial guess} of function argument types on function entry.
+This collects types for argument default values, and collects an ``unused'' type for unevaluated expressions.
+If an argument promise is later forced, we simply update the recorded type for the argument.
+
+\item Second, we wanted to account for formal arguments (arguments which are named in the function definition) that were {\it not} passed to the function.
+Put simply, a function was not called with all of its arguments, and no default value was specified for missing arguments.
+In this case, we record a ``missing'' type for the argument.
+Later, when we build types during a post-processing phase, we combine
+
 \end{itemize}
-
-Now that we've seen the high-level picture of our approach, we will elaborate on some key details on our use of the R-dyntrace framework, and some of the finer points of the analysis.
 

### 2020-05-09 paper/main.tex
typetracer discussion, onto the type inference discussion

@@
 \subsubsection{Challenges}
+\label{subsec:typetracer-challenges}
 
@@
 In this case, we record a ``missing'' type for the argument.
-Later, when we build types during a post-processing phase, we combine
+
+There are two obvious ways to deal with missing arguments: type them as bot, or as top.
+As we are performing a dynamic analysis, based on package test code, we conservatively type these arguments as top, or \code{any}.
+It is impossible to know dynamically what all possible function behaviours are, \AT{I'm still not convinced about this.}
+
+\item Third, we had to contend with R's C implementation sometimes performing longjumps, which discards current R calls from the call stack.
+In order to ensure that call traces are not lost when a longjump occurs, \typetracer intercepts the unwinding process initiated by the longjump and mimics the functionality of the functions being exited.
+The R-dyntrace framework supplies the return value during the unwinding process (if there is indeed such a value), so we can capture it, analyze it, and include it in the call trace.
+In certain situations, however, no return value can be obtained (e.g.
+a longjump occurred as part of some error handling), and then we simply record a ``jumped'' return value, which we can deal with in post-processing.
+These longjumps occur quite often, particularly when S3 dispatch occurs, as the R implementation optimizes returns from dispatched functions by initiating a longjump.
 
 \end{itemize}
+
+Having outlined these challenges, we will now explore the implementation of \typetracer in more detail.
 
@@
 These callbacks fire when closures, builtin, and special functions are entered and exited, and before and after the forcing of promises.
-The function-related callbacks are used mainly for bookkeeping: the analysis is notified that a closure/builtin/special has been entered/exited by pushing/popping the call onto a stack.
-The calls themselves store a {\it call trace} object, where we store the type information that we collect.
+The function-related callbacks are used mainly for bookkeeping: the analysis is notified that a closure, builtin, or special has been entered/exited by pushing/popping the call onto a stack.
+The calls themselves store a {\it call trace} object, where we store the type information that we collect, populated with initial guesses of argument types (as discussed in Section~\ref{subsec:typetracer-challenges}).
 Further, the type of dispatch is recorded on the {\tt \_entry} variants: \AT{recall} that there are multiple ways that R performs dynamic dispatch, based on the class of function arguments.
@@
 In R, promises are essentially objects with value and expression slots.
-When a promise is created for a particular expression, that expression is merely put into the expression slot of a promise object, and a special {\it unbound} value is loaded into the value slot.
+When a promise is created for a particular expression, that expression is merely put into the expression slot of a promise object, and a special {\it unbound} value is loaded into the value slot (unless some default value is specified, wherein it is loaded into the value slot---this is the case with default argument values).
 When the promise is forced, the expression is evaluated in the forcing context, and the value is stored in the value slot of the promise.
 So to get the type of a function argument, we check the value slot of the argument promise---perhaps recursively if the value itself is a promise.
-On function {\tt \_exit} variants, we deal with situations where an argument was never used (and thus never forced): these show up as missing slots in the call trace object, and we ascribe them the type {\tt ???}.
 
-To construct a ``type'' for each argument, we make use of R's C FFI and use low-level machinery to collect type tags and attributes from the R runtime.
-At this level, all R values are {\tt SEXP}s with a type tag that can be obtained using the {\tt TYPEOF} function from the C FFI.
-In certain circumstances, such as when {\tt TYPEOF} indicates that a value is a primitive, we perform additional reflection e.g.
-on the length (with {\tt LENGTH}) and also vector element names (through a call to {\tt getAttrib(the\_value, R\_NamesSymbol)}).
+To construct a type for each argument, we make use of R's C FFI and use low-level machinery to collect type tags and attributes from the R runtime.
+The types that we provide to users are constructed during post-processing, and rely on the detailed information made available by these low-level reflection mechanisms.
 
-\subsection{Implementation}
-
-\AT{Talk about how the type system was implemented, in terms of the checks, how they work, how they are performed by the runtime, etc.}
-
-\subsection{Evaluation}
-
-\AT{Talk about the evaluation.}
-
-\AT{Ok, so we haven't discussed the story for the results, so I'll just write some disjointed sections, discussing design decisions, and backing them up with the data.
-This should give us an idea of what we want to say, and we can decide how to organize everything once we have stuff written.
-Some of the things I'll be pulling from the data are: how many functions are higher-order? How many times are structs present? How many times do classes supercede other types?
-Basically, we want to use this section to turn the simplicity of our type system from a weakness into a strength. 
-The simplicity {\it is} a strength, we just need to make sure that's obvious to the reviewer.}
+%
+%
+%
+%
+\subsection{Post-Processing and Type Ascription}
 

### 2020-05-09 evaluation/R/corpus-analysis.Rmd
Split corpus creation and analysis

--- 2020-05-09 evaluation/R/package-corpus.Rmd
@@
 ---
-title: "Package corpus"
+title: "Package Corpus Analysis"
 output: html_document
@@
 library(tidyverse)
-library(tools)
-library(withr)
 
@@
 
-create_tags(path(TAGS_DIR, "package-corpus.tex"), prefix="", default=TRUE)
+create_tags(path(TAGS_DIR, "corpus.tex"), prefix="", default=TRUE)
 ```
 
-## Loading data
+## Load data
 
-### CRAN
-
-We start with all CRAN packages that we have mirrored locally using rsync.
-
-```{r cran packages}
-pkgs_cran_raw <- available.packages(contriburl = str_c("file://", path(CRAN_MIRROR_DIR, "src", "contrib")))
-pkgs_cran <- tibble(package=pkgs_cran_raw[,1])
-```
-
-### Installed packages
-
-We tried to instal all of the `r fmt(nrow(pkgs_cran))` these packages.
-Some packages failed to install, in most cases this is because of missing native dependencies.
-
-```{r find install packages}
-pkgs_installed_raw <- installed.packages(LIB_DIR)
-pkgs_installed <- tibble(package=pkgs_installed_raw[,1])
-```
-
-### Loadable packages
-
-We managed to install `r fmt(nrow(pkgs_installed))` packages.
-Next is to chech how many of the installed packages can actually be loaded.
-We do that by running the `package-load.R` task:
-
-```
-./rapr/inst/on-each-package.sh ./rapr/inst/tasks/package-load.R 
-env-rscript ./rapr/inst/process-csv-result.R run/package-load
-```
-
-This changes wether a package can be loaded using `library(package)` call.
-
-```{r loadable packages}
-pkgs_loadable <- 
-  read_parallel_run_csv(PACKAGE_LOAD_RUN) %>%
-  filter(exitval==0) %>%
-  rename(package=job)
-```
-
-### Package metadata
-
-Next we pull in some package metadata.
-
-```{r load package sloc}
-pkgs_sloc_raw <- 
-  read_csv(
-    PACKAGE_SLOC,
+```{r}
+pkgs <- read_csv(
+  ALL_PACKAGES_DETAILS,
   col_types=cols(
     package = col_character(),
-      path = col_character(),
-      files = col_double(),
-      language = col_character(),
-      blank = col_double(),
-      comment = col_double(),
-      code = col_double()
-    )
-  )
-
-pkgs_sloc <- 
-  pkgs_sloc_raw %>%
-  filter(endsWith(path, "/R") | endsWith(path, "/src")) %>%
-  select(-blank, -comment, -files, -path) %>%
-  mutate(language=case_when(
-    language == "R" ~ "package_r_code",
-    language == "C" ~ "package_native_code",
-    language == "C/C++ Header" ~ "package_native_code",
-    language == "C++" ~ "package_native_code",
-    startsWith(language, "Fortran") ~ "package_native_code",
-    TRUE ~ as.character(NA)
-  )) %>%
-  filter(!is.na(language)) %>%
-  group_by(package, language) %>%
-  summarise(code=sum(code)) %>%
-  spread(key=language, value=code, fill=0) %>%
-  mutate(package_code=package_native_code + package_r_code)
-```
-
-### Package runnable code
-
-Next to the code itself, R packages have examples, tests and vignettes.
-These can include snippets of R code which calls the package functions.
-We extract this code and run it to narrow down the list of packages that can be considered into the final corpus.
-
-```{r runnable code metadata}
-pkgs_runnable_code_raw <- 
-  read_csv(
-    PACKAGE_RUNNABLE_CODE_METADATA,
-    col_types=cols(
-      package = col_character(),
-      path = col_character(),
-      files = col_double(),
-      language = col_character(),
-      blank = col_double(),
-      comment = col_double(),
-      code = col_double()
-    )
-  )
-
-pkgs_runnable_code <- 
-  pkgs_runnable_code_raw %>%
-  filter(language=="R") %>%
-  select(package, path, code, files) %>%
-  pivot_wider(
-    names_from=path, 
-    values_from=c(code, files), 
-    values_fill=list(code=0, files=0)
-  ) %>%
-  # we have old version of tidyr, names_glue is missing, so have to rename manually
-  rename(
-    runnable_code_examples=code_examples,
-    runnable_code_tests=code_tests,
-    runnable_code_vignettes=code_vignettes,
-    runnable_files_examples=files_examples,
-    runnable_files_tests=files_tests,
-    runnable_files_vignettes=files_vignettes,
-  ) %>%
-  mutate(
-    runnable_code=runnable_code_examples+runnable_code_tests+runnable_code_vignettes,
-    runnable_files=runnable_files_examples+runnable_files_tests+runnable_files_vignettes
-  ) %>%
-  select(
-    package,
-    runnable_code,
-    runnable_files,
-    everything()
-  )
-```
-
-### Package coverage
-
-```{r load coverage data}
-pkgs_coverage <- 
-  read_csv(
-    PACKAGE_COVERAGE,
-    col_types=cols(
-      package = col_character(),
-      type = col_character(),
-      coverage_expression = col_double(),
+    installed = col_logical(),
+    loadable = col_logical(),
+    package_native_code = col_double(),
+    package_r_code = col_double(),
+    package_code = col_double(),
+    runnable_code = col_double(),
+    runnable_files = col_double(),
+    runnable_code_examples = col_double(),
+    runnable_code_vignettes = col_double(),
+    runnable_code_tests = col_double(),
+    runnable_files_examples = col_double(),
+    runnable_files_vignettes = col_double(),
+    runnable_files_tests = col_double(),
+    coverage_expr = col_double(),
     coverage_line = col_double(),
-      error = col_logical()
-    )
-  ) %>%
-  filter(type=="all", is.na(error), !is.na(coverage_expression)) %>% 
-  select(-type, -error, coverage_expr=coverage_expression) %>%
-  mutate(coverage_expr=coverage_expr/100, coverage_line=coverage_line/100)
-```
-
-### Package runs
-
-```{r package runs}
-pkgs_runs_raw <-
-  read_csv(
-    PACKAGE_RUN,
-    col_types=cols(
-      package = col_character(),
-      file = col_character(),
-      type = col_character(),
-      time = col_double(),
-      error = col_character(),
-      output = col_character()
-    )
-  )
-pkgs_runs <-
-  pkgs_runs_raw %>%
-  group_by(package) %>%
-  summarise(runnable_success=sum(is.na(error)), runnable_error=sum(!is.na(error)), runnable_time=sum(time, na.rm=TRUE)) %>%
-  mutate(runnable_rate=runnable_success/(runnable_success+runnable_error))
-```
-
-### Package functions
-
-```{r load functions}
-pkgs_functions <- 
-  read_csv(
-    PACKAGE_FUNCTIONS,
-    col_types=cols(
-      package = col_character(),
-      fun = col_character(),
-      num_params = col_double(),
-      has_elipsis = col_logical(),
-      exported = col_logical()
-    )
-  ) %>%
-  count(package, exported) %>%
-  mutate(type=ifelse(exported, "funs_public", "funs_private")) %>%
-  select(-exported) %>%
-  spread(key=type, value=n, fill=0) %>%
-  mutate(funs=funs_public+funs_private)
-```
-
-### Package reverse dependencies
-
-There are two functions `package_dependencies` and `dependsOnPkgs`.
-Not sure what is the difference, but the `dependsOnPkgs` is very slow.
-
-```{r compute reverse dependencies}
-pkgs_revdeps_raw <- withr::with_libpaths(LIB_DIR, {
-  package_dependencies(
-    pkgs_installed$package,
-    db=pkgs_installed_raw,
-    which=params$which_dependencies,
-    reverse=TRUE,
-    recursive=FALSE
-  )
-}) %>% imap_dfr(~tibble(package=.y, revdep=.x))
-
-pkgs_revdeps <- count(pkgs_revdeps_raw, package) %>% rename(revdeps=n)
-```
-
-### Package asserts
-
-```{r runtime checks}
-pkgs_asserts <- 
-  read_csv(
-    PACKAGE_ASSERTS,
-    col_types=cols(
-      package = col_character(),
-      fun_name = col_character(),
-      file = col_character(),
-      line1 = col_double(),
-      col1 = col_double(),
-      line2 = col_double(),
-      col2 = col_double(),
-      assert = col_character(),
-      args = col_character()
-    )
-  ) %>%
-  count(package, fun_name) %>%
-  group_by(package) %>%
-  summarise(funs_with_assert=length(unique(fun_name)), asserts=sum(n))
-```
-
-### Package roxygen
-
-```{r}
-pkgs_roxygen <-
-  read_csv(
-    PACKAGE_USES_ROXYGEN,
-    col_types=cols(
-      package = col_character(),
+    runnable_success = col_double(),
+    runnable_error = col_double(),
+    runnable_time = col_double(),
+    runnable_rate = col_double(),
+    funs_private = col_double(),
+    funs_public = col_double(),
+    funs = col_double(),
+    revdeps = col_double(),
+    funs_with_assert = col_double(),
+    asserts = col_double(),
     uses_roxygen = col_logical()
@@
 )
-```
-
-### Join
-
-```{r joining data}
-pkgs <-
-  pkgs_cran %>%
-  left_join(mutate(pkgs_installed, installed=TRUE), by="package") %>% mutate(installed=replace_na(installed, FALSE)) %>%
-  left_join(transmute(pkgs_loadable, package, loadable=TRUE), by="package") %>% mutate(loadable=replace_na(loadable, FALSE)) %>%
-  left_join(pkgs_sloc, by="package") %>%
-  left_join(pkgs_runnable_code, by="package") %>%
-  left_join(pkgs_coverage, by="package") %>%
-  left_join(pkgs_runs, by="package") %>%
-  left_join(pkgs_functions, by="package") %>%
-  left_join(pkgs_revdeps, by="package") %>%
-  left_join(pkgs_asserts, by="package") %>%
-  left_join(pkgs_roxygen, by="package")
-```
-
-### Sanity checks
-
-#### Packages that are loaded, but not installed?
-
-```{r}
-filter(pkgs, !installed && loadable) %>%
-  select(package, installed, loadable)
-```
-
-#### Packages that do not have any extracted runnable code, yet they have some runs or coverage?
-
-This can actually be still OK, perhaps the package has some code in its `.onLoad` or `.onAttach` hooks.
-
-```{r}
-filter(pkgs, runnable_code==0, coverage_expr > 0) %>%
-  select(package, package_code, runnable_code, runnable_files, starts_with("coverage"))
-```
-
-### Curiosity
-
-* Packages with no R code (this is perfectly fine):
-
-```{r}
-filter(pkgs, package_r_code==0)
-```
-
-* Packages with no code at all (also fine - just data packages):
-
-```{r}
-filter(pkgs, package_code==0)
 ```
@@
 summary_corpus(pkgs, "pkgs")
-
-write_csv(pkgs, ALL_PACKAGES_DETAILS)
 ```

### 2020-05-09 evaluation/R/corpus.Rmd
Split corpus creation and analysis

--- 2020-05-09 evaluation/R/package-corpus.Rmd
@@
 ---
-title: "Package corpus"
+title: "Corpus"
 output: html_document
@@
 params:
-  min_revdeps: 5
-  min_coverage: .65
   which_dependencies:
@@
 options(repos=CRAN_LOCAL_MIRROR)
+```
 
-create_tags(path(TAGS_DIR, "package-corpus.tex"), prefix="", default=TRUE)
-```
+Tis Rmd file builds the project corpus.
 
@@
 
+### Output
+
+```{r}
+write_csv(pkgs, ALL_PACKAGES_DETAILS)
+write_lines(pkgs$package, ALL_PACKAGES_FILE)
+```
+
 ### Sanity checks
@@
 ```
-
-### Overview
-
-#### Table
-
-```{r summary}
-summary_corpus <- function(df, prefix) {
-  overview_table(
-    r(paste(prefix, "cran"), df),
-    r(paste(prefix, "installed"), filter(df, installed)),
-    r(paste(prefix, "loadable"), filter(df, loadable)),
-
-    r(paste(prefix, "with runnable code"), filter(df, runnable_code > 0)),
-    r(paste(prefix, "runnable code"), sum(df$runnable_code, na.rm=T)),
-    r(paste(prefix, "examples code"), sum(df$runnable_code_examples, na.rm=T)),
-    r(paste(prefix, "tests code"), sum(df$runnable_code_tests, na.rm=T)),
-    r(paste(prefix, "vignettes code"), sum(df$runnable_code_vignettes , na.rm=T)),
-
-    r(paste(prefix, "with coverage"), filter(df, coverage_expr > 0)),
-    r(paste(prefix, "mean expr coverage"), percent(mean(df$coverage_expr, na.rm=T))),
-    r(paste(prefix, "median expr coverage"), percent(median(df$coverage_expr, na.rm=T))),
-
-    r(paste(prefix, "with R code"), filter(df, package_r_code > 0)),
-    r(paste(prefix, "R code"), sum(df$package_r_code, na.rm=T)),
-    r(paste(prefix, "with native code"), filter(df, package_native_code > 0)),
-    r(paste(prefix, "native code"), sum(df$package_native_code, na.rm=T)),
-
-    r(paste(prefix, "w asserts"), filter(df, asserts > 0)),
-    r(paste(prefix, "asserts"), sum(df$asserts, na.rm=T)),
-    r(paste(prefix, "funs with asserts"), sum(df$funs_with_assert, na.rm=T)),
-
-    r(paste(prefix, "functions"), sum(df$funs, na.rm=T)),
-    r(paste(prefix, "public functions"), sum(df$funs_public, na.rm=T)),
-    r(paste(prefix, "private functions"), sum(df$funs_private, na.rm=T)),
-
-    r(paste(prefix, "mean revdes"), mean(df$revdeps, na.rm=T)),
-    r(paste(prefix, "median revdes"), median(df$revdeps, na.rm=T)),
-
-    r(paste(prefix, "runtime"), as.duration(sum(df$runnable_time, na.rm=T))),
-    r(paste(prefix, "mean runtime"), as.duration(mean(df$runnable_time, na.rm=T))),
-    r(paste(prefix, "median runtime"), as.duration(median(df$runnable_time, na.rm=T))),
-
-    r(paste(prefix, "using roxygen"), sum(df$uses_roxygen, na.rm=T))
-  )
-}
-
-summary_corpus(pkgs, "pkgs")
-
-write_csv(pkgs, ALL_PACKAGES_DETAILS)
-```
-
-#### Plots
-
-```{r considered packages}
-pkgs_considered <- filter(pkgs, loadable, package_r_code > 0, coverage_expr > 0)
-```
-
-```{r plot revdeps histogram}
-pkgs_considered %>%
-  ggplot(aes(x=revdeps)) + 
-    geom_histogram(bins=50) +
-    scale_x_log10(labels=scales::comma) +
-    labs(x="# reverse dependencies (log)")
-```
-
-```{r plot coverage histogram}
-pkgs_considered %>%
-  ggplot(aes(x=coverage_expr)) + 
-    geom_histogram() +
-    scale_x_continuous(labels=scales::percent) +
-    labs(x="Expression coverage")
-```
-
-```{r plot all considered packages}
-plot_all <- function(df) {
-  coverage_mean <- mean(df$coverage_expr)
-  coverage_median <- median(df$coverage_expr)
-  revdeps_mean <- mean(df$revdeps)
-  revdeps_median <- median(df$revdeps)
-
-  df %>%
-    mutate(
-      outlier=package_code > 5e4,
-      label=if_else(outlier, str_glue("{package}"), as.character(NA))
-    ) %>%
-    ggplot(
-      aes(
-        x=coverage_expr, 
-        y=revdeps,
-        label=label,
-        size=package_code
-      )
-    ) +
-    geom_point(alpha=.6) +
-    geom_text(size=3, check_overlap = T, vjust=1.5, na.rm = TRUE) +
-
-    geom_hline(aes(yintercept=revdeps_mean), linetype="dashed", color="black", size=0.2) +
-    geom_vline(aes(xintercept=coverage_mean), linetype="dashed", color="black", size=0.2) +
-    geom_vline(aes(xintercept=params$min_coverage),linetype="solid", color="red", size=0.4) +
-    geom_hline(aes(yintercept=params$min_revdeps),linetype="solid", color="red", size=0.4) +
-
-    scale_x_continuous(labels = scales::percent) + 
-    scale_y_log10(labels = scales::comma) +
-    scale_size(range = c(.1, 5), labels = scales::comma) +
-      theme(
-      legend.position=c(0.2, 0.85),
-      legend.box="horizontal",
-      legend.box.background = element_rect(fill="white", size=0.1)
-    ) +
-    labs(
-      x="Code coverage", 
-      y="Number of reverse dependencies (log)", 
-      size="Code size",
-      linetype=""
-    ) +
-    guides(
-      size=guide_legend(order=1),
-      color=guide_legend(order=2),
-      linetype=F
-    )
-}
-
-plot_all(pkgs_considered)
-ggsave(CORPUS_PLOT)
-```
-
-## Project corpus
-
-We only consider packages that have at least `r params$min_revdeps` reverse dependencies and `r params$min_coverage*100`% expression coverage.
-
-```{r project corpus}
-corpus <- 
-  filter(
-    pkgs, 
-    loadable, 
-    runnable_code > 0, 
-    revdeps >= params$min_revdeps, 
-    package_r_code > 0, 
-    coverage_expr >= params$min_coverage
-  )
-
-write_lines(corpus$package, CORPUS_FILE)
-write_csv(corpus, CORPUS_DETAILS)
-```
-
-### Sanity check
-
-TODO
-
-### Overview
-
-#### Table
-
-```{r corpus summary}
-summary_corpus(corpus, "corpus")
-
-write_csv(corpus, CORPUS_DETAILS)
-write_lines(corpus$package, CORPUS_FILE)
-```
-
-#### Plot
-
-```{r plot corpus}
-plot_corpus <- function(df, code_outlier=5e3) {
-  coverage_mean <- mean(df$coverage_expr)
-  coverage_median <- median(df$coverage_expr)
-  revdeps_mean <- mean(df$revdeps)
-  revdeps_median <- median(df$revdeps)
-
-  df %>%
-    mutate(
-      outlier=package_code > code_outlier,
-      label=if_else(outlier, str_glue("{package}"), as.character(NA))
-    ) %>%
-    ggplot(
-      aes(
-        x=coverage_expr, 
-        y=revdeps,
-        label=label,
-        size=package_code
-      )
-    ) +
-    geom_point(alpha=.6) +
-    geom_text(size=3, check_overlap = T, vjust=1.5, na.rm = TRUE) +
-
-    geom_hline(aes(yintercept=revdeps_mean), linetype="dashed", color="red", size=0.2) +
-    geom_vline(aes(xintercept=coverage_mean), linetype="dashed", color="red", size=0.2) +
-
-    scale_x_continuous(limits = c(params$min_coverage, 1), labels = scales::percent) + 
-    scale_y_log10(labels = scales::comma) +
-    scale_size(range = c(.1, 3), labels = scales::comma) +
-      theme(
-      legend.position=c(0.2, 0.85),
-      legend.box="horizontal",
-      legend.box.background = element_rect(fill="white", size=0.1)
-    ) +
-    labs(
-      x="Code coverage", 
-      y="Number of reverse dependencies (log)", 
-      size="Code size",
-      linetype=""
-    ) +
-    guides(
-      size=guide_legend(order=1),
-      color=guide_legend(order=2),
-      linetype=F
-    )
-}
-
-plot_corpus(corpus)
-```
-
-### List of reverse dependencies
-
-```{r list of reverse depenencies}
-corpus_revdeps <- pkgs_revdeps_raw %>% semi_join(corpus, by="package")
-write_csv(corpus_revdeps, CORPUS_REVDEPS)
-write_lines(unique(corpus_revdeps$revdep), CORPUS_REVDEPS_UNIQUE)
-```
-
-### List of corpus packages
-
-```{r list of packages}
-datatable(
-  corpus, 
-  filter="top", 
-  options=list(
-    pageLength=50,
-    autoWidth=TRUE
-  )
-)
-```
-
-## Appendix: Most downloaded CRAN packages
-
-```{r top cran downloads}
-df_top_cran <- cran_top_downloads(when="last-month", count=100)
-```
-
-The cranlogs allow us to get 100 most downloaded packages over the last month.
-How many of these popular packages do we have in out corpus?
-
-```{r}
-missed_top_cran_packages <- setdiff(df_top_cran$package, corpus$package)
-```
-
-```{r}
-filter(df_top_cran, package %in% missed_top_cran_packages) %>%
-  select(-from, -to) %>%
-  datatable()
-```
-
-The main reason is that we use R 3.5. Some key dependencies such as Hmisc are no longer available.

### 2020-05-09 evaluation/R/corpus-analysis.Rmd
Draft of 2

@@
 ---
-title: "Package Corpus Analysis"
+title: "Corpus Analysis"
 output: html_document
@@
   min_coverage: .65
-  which_dependencies:
-    - Depends
-    - Imports
+  package_size_outlier: 5000
 ---
@@
 knitr::opts_chunk$set(echo = TRUE)
+
+Sys.setenv(
+  TYPER_BASE_DIR=system2(
+    "git", 
+    c("rev-parse", "--show-toplevel"), 
+    stdout=TRUE
+  )
+)
 
@@
 
-### Overview
+## Overview
 
-#### Table
+### Thresholds
+
+```{r}
+overview_table(
+  r("threshold code coverage", percent(params$min_coverage)),
+  r("threshold revdeps", params$min_revdeps),
+  r("package size outier", params$package_size_outlier)
+)
+```
+
+### Table
 
@@
 
-summary_corpus(pkgs, "pkgs")
+summary_corpus(pkgs, "All")
 ```
 
-#### Plots
+### Plots
 
@@
 plot_all(pkgs_considered)
-ggsave(CORPUS_PLOT)
 ```
@@
 ```{r plot corpus}
-plot_corpus <- function(df, code_outlier=5e3) {
+plot_corpus <- function(df, code_outlier) {
   coverage_mean <- mean(df$coverage_expr)
@@
 
-plot_corpus(corpus)
-```
-
-### List of reverse dependencies
-
-```{r list of reverse depenencies}
-corpus_revdeps <- pkgs_revdeps_raw %>% semi_join(corpus, by="package")
-write_csv(corpus_revdeps, CORPUS_REVDEPS)
-write_lines(unique(corpus_revdeps$revdep), CORPUS_REVDEPS_UNIQUE)
+plot_corpus(corpus, params$package_size_outlier)
+ggsave(CORPUS_PLOT)
 ```

### 2020-05-09 paper/main.tex
Draft of 2

@@
 \usepackage{listings, wrapfig,xspace}
+\usepackage{paralist}
 
@@
 \newcommand{\ie}{\emph{i.e.},\xspace}
+\newcommand{\cf}{\emph{cf.}\xspace}
 
@@
 \newcommand{\typetracer}{\emph{typetracer}\xspace} % typetracer
+\newcommand{\rdt}{\emph{R-dyntrace}\xspace}
+\newcommand{\covr}{\emph{covr}\xspace}
+
+\input{corpus.tex}
 
@@
 %
-\subsection{The Corpus}
-\label{sec:corpus}
+\section{Project Corpus}\label{sec:corpus}
 
-\AT{This section needs to be tight.
-Reviewers were skeptical of CRAN and using the test and example code.
-We should convince them otherwise.}
+For this paper we have selected \CorpusLoadable packages consisting of
+\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native code
+(C/C++/Fortran).
+\FK{Add more details}
+Figure~\ref{fig:corpus} shows all the packages, the size of the dots reflects
+the size of the project in terms of both R and native lines of code\footnote{
+Lines of source code reported excludes comments and blank lines, counted by \emph{cloc}, \cf \url{https://github.com/AlDanial/cloc}}, the x-axis
+indicates the expression code coverage in percents and the y-axis gives the
+number of reverse dependencies in log scale. Dotted lines indicate means.
+Packages with over \PackageSizeOutierRnd lines of code are annotated.
 
-\AT{Some sentences that might be useful:}
-We selected packages that have both \AT{a good amount of} reverse dependencies to ensure their broad usefulness, as well \AT{sufficiently high} code and branch coverage of test, example, and vignette programs to ensure that our analysis built up an accurate picture of function use.
+\begin{figure}[!h]  \centering
+  \includegraphics[width=.8\linewidth]{plots/corpus.pdf}
+  \caption{Corpus overview}\label{fig:corpus}
+\end{figure}
 
-%
-%
-\subsubsection{Cross-Validation Against Kaggle}
+These packages come from the Comprehensive R Achive Network
+(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of R
+code with over \AllCranRnd packages\footnote{CRAN receives about 6 new package
+  submissions a day~\cite{Ligges2017}} containing over \AllRCodeRnd and
+\AllNativeCodeRnd R and native code respectively. Unlike other open code
+repositories like GitHub, CRAN is a curated repository. Each submitted package
+must abide to a number of well-formedness rules that are automatically checked
+asserting certain quality. Most relevant for this work is that all of the
+runnable code (including code snippets from examples and vignettes) is tested
+and only a successfully running package is admitted in the archive.
 
-Kaggle is a website offering an environment for hosting Jupyter notebooks, and importantly provides access to a large number of open-source data sets and R and Python code.
-This solved a major issue with R programs hosted on e.g. Github, where often the data is not made available.
-Our aim is to run R programs hosted on Kaggle to validate the type signatures we generated for popular R packages.
+We have downloaded and installed all available CRAN packages. Out of the
+\AllCranRnd packages, we managed to install \AllLoadableRnd. The main
+reason for this is that \rdt is based on R 3.5.0 and some of the packages are no
+longer compatible with it. Some packages also require extra native dependencies
+which were not present on our servers.
+We have two criteria for including a
+package into the corpus:
+\begin{inparaenum}[(1)]
+\item have runnable code that covers significant part of the package source
+code from which type signatures could be inferred, and
+\item have some reverse dependencies that allows us to evaluate the inferred
+types on runnable code from these dependent packages.
+\end{inparaenum}
+The concrete thresholds were set to at least \ThresholdCodeCoverage of expression
+coverage and at minimum \ThresholdRevdeps reverse dependencies.
 
-\AT{Anything else to say on Kaggle, here? We will probably talk more about the corpus we analyzed later.}
+The code coverage was computed for each package using
+\covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R tool.
+The reverse package dependencies were extracted from the package metadata using
+builtin function.
+
+% \AT{This section needs to be tight.
+% Reviewers were skeptical of CRAN and using the test and example code.
+% We should convince them otherwise.}
+
+% \AT{Some sentences that might be useful:}
+% We selected packages that have both \AT{a good amount of} reverse dependencies to ensure their broad usefulness, as well \AT{sufficiently high} code and branch coverage of test, example, and vignette programs to ensure that our analysis built up an accurate picture of function use.
+
+% %
+% %
+% \subsubsection{Cross-Validation Against Kaggle}
+
+% Kaggle is a website offering an environment for hosting Jupyter notebooks, and importantly provides access to a large number of open-source data sets and R and Python code.
+% This solved a major issue with R programs hosted on e.g. Github, where often the data is not made available.
+% Our aim is to run R programs hosted on Kaggle to validate the type signatures we generated for popular R packages.
+
+% \AT{Anything else to say on Kaggle, here? We will probably talk more about the corpus we analyzed later.}
 

### 2020-05-09 paper/main.tex
finished type ascription and consolidiation section. evaluation results will help to inform the rest of the narrative

--- 2020-05-09 paper/main.tex
@@
 
+The type information we collect is far more detailed than the types we actually generate, and so the next stage of our inference process is to distill the rich type information into simpler, manageable types.
+For each unique observed call trace, we translate the representation of the type of each of the function arguments and returns according to the design of our type system:
+for instance, we might have determined that a function returned a vector of integer with 4 elements, with names, that was free of NA values, in which case we would write out \code{integer[]} as the type.
+
+Once types have been simplified in this way, for each function we move to consolidate its multiple call traces into a single function type.
+We do this by combining the observed types of each argument into one type (possibly a union, mind), according to subtyping rules that are part of our design.
+The precise subtyping rules we used will be discussed further in Section~\ref{sec:typesystemdesign}.
+
+While our analysis collects more detailed information than necessary for type generation, the information proved useful for building up a usage profile of the language.
+Through it, we were able to determine the best translation rules (i.e., determining the type system we are translating to) through an iterative process, involving analyzing usage patterns that arose in the data and making design decisions based on them, revisiting the data as necessary.
+In the following section, we will discuss these decisions in more detail, highlighting the reasoning behind our type system design.
+
 %

### 2020-05-10 paper/main.tex
clean up
intro wip

@@
 \begin{document}
-\title{Types for R, Empirically Designed and Evaluated}
+\title{Extracting Function Signature from R Program Traces}
+
 \begin{abstract}
-
+  This paper documents the first steps towards retrofitting a type system
+  onto the R programming language. We introduce an extension of R with
+  function signatures. To evaluate the fit of our design to the language, we
+  start by automatically extracting the signatures for 500 widely used
+  packages. In a second step we use the reverse dependencies of these
+  packages to validate that these signatures are accurate.
 \end{abstract}
@@
 
-The story that we're aiming for:
-Designing type systems for established languages can be a challenging.
-We aim to approach this problem by essentially reverse-engineering a type system from a corpora analysis of a representative subset of R code.
-We present a simple type system that captures the vast majority of usage of R --- it is lightweight, and \AT{should provide enough information to be useful to a compiler}.
+How does one retrofit a type system to an existing language? Three succesful
+models are Hack, Typed Racket and TypeScript.
 
-To evaluate our type system, we present a type-checker that is aware of R's promise semantics. 
-In R, function arguments are wrapped in promises such that they are not evaluated until they are needed.
-Our tool deals with these promises by inserting a check which fires only when the promise is forced, at which point the value is checked for compliance with the type.
+The story that we're aiming for: Designing type systems for established
+languages can be a challenging.  We aim to approach this problem by
+essentially reverse-engineering a type system from a corpora analysis of a
+representative subset of R code.  We present a simple type system that
+captures the vast majority of usage of R --- it is lightweight, and
+\AT{should provide enough information to be useful to a compiler}.
 
-\AT{I can grab some of the intro Jan wrote for the last paper, rework it, and put it here.}
+To evaluate our type system, we present a type-checker that is aware of R's
+promise semantics.  In R, function arguments are wrapped in promises such
+that they are not evaluated until they are needed.  Our tool deals with
+these promises by inserting a check which fires only when the promise is
+forced, at which point the value is checked for compliance with the type.
 
-%
-%
-%
 %

### 2020-05-11 paper/main.tex
Add type grammar

@@
 
+\subsection{Type Grammar}
+\label{subsec:typegrammar}
+
+\begin{figure}          % Types
+    \noindent           % Collapse for readability.
+    \centering
+    \begin{minipage}{.45\linewidth}
+    \begin{flushleft}
+    % Types
+      $ \begin{array}{lclr}
+    D & ::= & {\bf type} \; ID \; T {\bf ;}  & \text{\it type declaration}\\
+    T & ::= & {\bf null} & \text{\it unit type} \\
+      & |   & {\bf any} & \text{\it top type} \\
+      & |   & {\bf environment} & \text{\it environment type} \\
+      & |   & {\bf expression} & \text{\it expression type} \\
+      & |   & {\bf language} & \text{\it language type} \\
+      & |   & {\bf symbol} & \text{\it symbol type} \\
+      & |   & {\bf externalptr} & \text{\it externalptr type} \\
+      & |   & {\bf pairlist} & \text{\it pairlist type} \\
+      & |   & {\bf s4} & \text{\it S4 type} \\
+      & |   & {\bf weakref} & \text{\it weakref type} \\
+      & |   & S & \text{\it scalar type} \\
+      & |   & V & \text{\it vector type} \\
+      & |   & \left( T \right) & \text{\it group type} \\
+      & |   & T_1 \; {\bf |} \; T_2 & \text{\it union type} \\
+      & |   & \langle A_1 , ..., A_n\rangle \Rightarrow T & \text{\it function type} \\
+      & |   & {\bf struct}\langle F_1, ..., F_n \rangle & \text{\it struct type} \\
+      & |   & {\bf list} \langle T \rangle & \text{\it list type}\\
+      & |   & {\bf class}\langle ID_1, ..., ID_n \rangle & \text{\it class type}\vspace{5pt}\\
+    \end{array} $
+    \end{flushleft}
+    \end{minipage}
+    \hfill
+    \begin{minipage}{.45\linewidth}
+    \begin{flushright}
+    $ \begin{array}{lclr}
+    % Argument Types
+    A & ::= & T & \text{\it argument type} \\
+      & |   & {\bf ...} \vspace{5pt} \\
+    % Field Types
+    F & ::= & ID : T & \text{\it struct field}\vspace{5pt}\\
+    % Scalar Types
+    S & ::= & B & \text{\it scalar types} \\
+      & |   & N \vspace{5pt}\\
+    % Vector Types
+    V & ::= & S \lbrack \rbrack & \text{\it vector types} \vspace{5pt}\\
+    % Base Types
+    B & ::= & {\bf int} & \text{\it base types} \\
+      & |   & {\bf chr} \\
+      & |   & {\bf dbl} \\
+      & |   & {\bf lgl} \\
+      & |   & {\bf clx} \\
+      & |   & {\bf raw} \vspace{5pt}\\
+    % NA Types
+    N & ::= & {\bf \string^int} & \text{\it na types} \\
+      & |   & {\bf \string^chr} \\
+      & |   & {\bf \string^dbl} \\
+      & |   & {\bf \string^lgl} \\
+      & |   & {\bf \string^clx} \\
+\end{array} $
+    \end{flushright}
+    \end{minipage}
+    \caption{The R type system.}
+    \label{fig:types}
+\end{figure}
+
 %

### 2020-05-11 evaluation/R/corpus-analysis.Rmd

@@
 pkgs <- read_csv(
-  ALL_PACKAGES_DETAILS,
+  CRAN_PACKAGES_DETAILS,
   col_types=cols(
@@
     loadable = col_logical(),
+    version = col_character(),
+    title = col_character(),
+    size = col_double(),
     package_native_code = col_double(),
@@
 
+### Reverse dependencies
+
+```{r}
+pkgs_revdeps_raw <- read_csv(
+  CRAN_REVDEPS, 
+  col_types=cols(
+    package = col_character(),
+    revdep = col_character()
+  )
+)
+corpus_revdeps <- pkgs_revdeps_raw %>% semi_join(corpus, by="package")
+write_csv(pkgs_revdeps_raw, CORPUS_REVDEPS)
+write_lines(unique(corpus_revdeps$revdep), CORPUS_REVDEPS_UNIQUE)
+```
+
 ### Sanity check

### 2020-05-11 evaluation/R/corpus.Rmd
Draft of 2

@@
 
-Next we pull in some package metadata.
+Next we pull in some package metadata:
+
+```{r}
+pkgs_metadata <-
+  read_csv(
+    PACKAGE_METADATA,
+    col_types=cols(
+      package = col_character(),
+      package_name = col_character(),
+      version = col_character(),
+      title = col_character(),
+      size = col_double()
+    )
+  ) %>% select(
+    package,
+    version,
+    title,
+    size
+  )
+```
+
+and size:
 
@@
   left_join(transmute(pkgs_loadable, package, loadable=TRUE), by="package") %>% mutate(loadable=replace_na(loadable, FALSE)) %>%
+  left_join(pkgs_metadata, by="package") %>%
   left_join(pkgs_sloc, by="package") %>%
@@
 ```{r}
-write_csv(pkgs, ALL_PACKAGES_DETAILS)
-write_lines(pkgs$package, ALL_PACKAGES_FILE)
+write_csv(pkgs, CRAN_PACKAGES_DETAILS)
+write_lines(pkgs$package, CRAN_PACKAGES_FILE)
+write_csv(pkgs_revdeps_raw, CRAN_REVDEPS)
 ```

### 2020-05-11 paper/main.tex

@@
 \begin{itemize}
-\item Firstly, we faced the problem of type-checking arguments in a non-strict
-  language while retaining the non-strict semantics. When a function is called in
-  R, parameters are bound to unevaluated code thunks called
-  promises~\cite{oopsla19}, instead of values which can be immediately
-  type-checked. Naively forcing promises at function call to obtain a value for
-  type-checking would violate the language semantics, leading to incorrect
-  results. To preserve the original semantics, \contractr modifies the
-  promise by wrapping the unevaluated promise expression in a call to its
-  type-checker along with the context necessary to associate the promise to the
-  function and its parameter (for failure messages). The contract checking happens
-  when the promise is evaluated, either inside the function or many calls deep.
-  This works, but with one wrinkle. GNU R has a bytecode compiler that can
-  sometimes optimize away promises; in those cases, \contractr receives
-  values that can be immediately type-checked. Non-strict semantics dictate that
-  any computation related to the argument should happen on the first use of the
-  argument. So type-checking at this point and issuing an error message would
-  violate the non-strict semantics; if this argument is never used, it would be
-  incorrect to type-check it at the beginning of call. To make this work,
-  \contractr mutates the function parameter binding to a promise that wraps
-  the argument value in a call to the type-checker, as in the previous case.
-\item Secondly, we faced the problem of type-checking function return value. In R, the
-  result of a function call is the result of evaluating the last expression in the
-  function body; so an explicit return call is absent from most function
-  definitions. Thus, there could be many potential sub-expressions in the function
-  body which need to be type-checked against the return type. To address this,
-  \contractr registers the type-checker to be called on the return value
-  through a function exit hook. This hook is executed in the function call's
-  environment after it has executed. While this works, we have to contend with yet
-  another wrinkle, R interpreter can perform longjumps in its C implementation,
-  which causes active R function calls on the stack to be discarded. When they are
-  discarded, their exit hooks are called, which in this case calls the registered
-  type-checker. But, these functions are in the middle of an active computation,
-  so they don't have a return value to type-check. \contractr deals with
-  this problem by allocating a unique sentinel object which serves as the return
-  value for calls that are discarded. The function call exit hook does not call
-  the type-checker if this unique value happens to be the return value of the
-  call.
+\item Firstly, we faced the problem of type-checking arguments in a
+  non-strict language while retaining the non-strict semantics. When a
+  function is called in R, parameters are bound to unevaluated code thunks
+  called promises~\cite{oopsla19}, instead of values which can be
+  immediately type-checked. Naively forcing promises at function call to
+  obtain a value for type-checking would violate the language semantics,
+  leading to incorrect results. To preserve the original semantics,
+  \contractr modifies the promise by wrapping the unevaluated promise
+  expression in a call to its type-checker along with the context necessary
+  to associate the promise to the function and its parameter (for failure
+  messages). The contract checking happens when the promise is evaluated,
+  either inside the function or many calls deep.  This works, but with one
+  wrinkle. GNU R has a bytecode compiler that can sometimes optimize away
+  promises; in those cases, \contractr receives values that can be
+  immediately type-checked. Non-strict semantics dictate that any
+  computation related to the argument should happen on the first use of the
+  argument. So type-checking at this point and issuing an error message
+  would violate the non-strict semantics; if this argument is never used, it
+  would be incorrect to type-check it at the beginning of call. To make this
+  work, \contractr mutates the function parameter binding to a promise that
+  wraps the argument value in a call to the type-checker, as in the previous
+  case.
+\item Secondly, we faced the problem of type-checking function return
+  value. In R, the result of a function call is the result of evaluating the
+  last expression in the function body; so an explicit return call is absent
+  from most function definitions. Thus, there could be many potential
+  sub-expressions in the function body which need to be type-checked against
+  the return type. To address this, \contractr registers the type-checker to
+  be called on the return value through a function exit hook. This hook is
+  executed in the function call's environment after it has executed. While
+  this works, we have to contend with yet another wrinkle, R interpreter can
+  perform longjumps in its C implementation, which causes active R function
+  calls on the stack to be discarded. When they are discarded, their exit
+  hooks are called, which in this case calls the registered
+  type-checker. But, these functions are in the middle of an active
+  computation, so they don't have a return value to type-check. \contractr
+  deals with this problem by allocating a unique sentinel object which
+  serves as the return value for calls that are discarded. The function call
+  exit hook does not call the type-checker if this unique value happens to
+  be the return value of the call.
 \end{itemize}
@@
 \subsubsection{Usability}
-With \contractr, we aim to provide a hassle-free path to the R developers for
-adding types to untyped package code. Hence, we have paid a lot of attention
-towards usability in its design. We discuss three such usability features.
+With \contractr, we aim to provide a hassle-free path to the R developers
+for adding types to untyped package code. Hence, we have paid a lot of
+attention towards usability in its design. We discuss three such usability
+features.
 \begin{itemize}
 
-\item \contractr enables automatic type-checking on importing without
-  any user intervention. Just running \code{library(contractr)} R code is enough
-  to insert contracts, both, in packages pre-loaded in the user's workspace, and,
-  in packages that will be loaded eventually. To achieve this automation,
-  \contractr relies heavily on R's reflective and dynamic capabilities. On
-  loading, \contractr scans all the pre-loaded packages in the user's workspace
-  and inserts contracts in functions for which type signatures are available. For
-  all other packages installed on user's machine but not yet loaded in the
-  workspace, \contractr sets up package load hooks which are executed by R when
-  those packages are loaded. The hook registered by \contractr insert contracts to
-  those package's functions. Thus, developers and users do not have to call the
-  \contractr API to enable type-checking for packages at any point of their
-  programming workflow. Furthermore, \contractr automatically removes contracts
-  from all the package functions and restores them to their original state when it
-  is unloaded, an operation that is rarely performed, but useful in interactive
+\item \contractr enables automatic type-checking on importing without any
+  user intervention. Just running \code{library(contractr)} R code is enough
+  to insert contracts, both, in packages pre-loaded in the user's workspace,
+  and, in packages that will be loaded eventually. To achieve this
+  automation, \contractr relies heavily on R's reflective and dynamic
+  capabilities. On loading, \contractr scans all the pre-loaded packages in
+  the user's workspace and inserts contracts in functions for which type
+  signatures are available. For all other packages installed on user's
+  machine but not yet loaded in the workspace, \contractr sets up package
+  load hooks which are executed by R when those packages are loaded. The
+  hook registered by \contractr insert contracts to those package's
+  functions. Thus, developers and users do not have to call the \contractr
+  API to enable type-checking for packages at any point of their programming
+  workflow. Furthermore, \contractr automatically removes contracts from all
+  the package functions and restores them to their original state when it is
+  unloaded, an operation that is rarely performed, but useful in interactive
   settings.
 
-\item \contractr enables package authors to supply type declarations for their
-  package without requiring package code modifications. Type declarations can be
-  written alongside a \code{@type} section inside \roxygen function comment
-  blocks. \roxygen is an R package that enables authors to add documentation to
-  R functions in the form of organized plaintext sections which are
-  automatically exported to R's latex style custom documentation format.
-  \roxygen is a widely adopted package, used by over XXX R packages in CRAN. It
-  provides an extension API for adding custom documentation tag sections in
-  comment blocks. The sections are processed by \roxygen before package build
-  step and a registered method for each custom tag section is invoked with the
-  section data and documented object. \contractr uses this mechanism to register
-  a hook for a custom \code{@type} tag. This hook parses the type declarations
-  from the \code{@type} sections, extracts the function name from the documented
+\item \contractr enables package authors to supply type declarations for
+  their package without requiring package code modifications. Type
+  declarations can be written alongside a \code{@type} section inside
+  \roxygen function comment blocks. \roxygen is an R package that enables
+  authors to add documentation to R functions in the form of organized
+  plaintext sections which are automatically exported to R's latex style
+  custom documentation format.  \roxygen is a widely adopted package, used
+  by over XXX R packages in CRAN. It provides an extension API for adding
+  custom documentation tag sections in comment blocks. The sections are
+  processed by \roxygen before package build step and a registered method
+  for each custom tag section is invoked with the section data and
+  documented object. \contractr uses this mechanism to register a hook for a
+  custom \code{@type} tag. This hook parses the type declarations from the
+  \code{@type} sections, extracts the function name from the documented
   function definition, and stores them in a TYPEDECLARATION file inside the
   package folder. This files is copied verbatim on package installation and
-  picked up \contractr when the package is loaded. Here is an example of how a
-  roxygen code block looks like with our type declaration:
+  picked up \contractr when the package is loaded. Here is an example of how
+  a roxygen code block looks like with our type declaration:
 
@@
 
-\item \contractr provides a very expressive API to the users, covering a variety
-  of uses cases. During interactive development, developers can explicitly
-  insert contracts by supplying the type declaration as a string argument to the
-  \code{insert\_contract} API function. Conversely, contracts can be removed by
-  calling the \code{remove\_contract} API on functions. Contracts can be
-  selectively enabled or disabled for code blocks by wrapping them in calls to
-  \code{ignore\_contracts} and \code{capture\_contracts} functions. This enables
-  selective type-checking of code section during the development phase.
-  Furthermore, these functions return a data frame that contains all the
-  information about failed and successful contract assertions in the wrapped
-  code blocks. This is useful for post-hoc investigation. Multiple type-checking
-  failures can introduce a lot of noise in the program output. To alleviate this
-  problem, \code{contractr.severity} option can be set to \code{'silence'}.
-  This suppresses failure messages but \contractr still performs type-checking whose
-  results can be explicitly obtained as a data frame from the
-  \code{get\_contracts} API function. R suppresses printing of warnings on the
-  terminal when a function issues too many warnings. This can hide type-checking
-  failure messages. Setting \code{contractr.severity} flag to \code{'error'}
-  turns the type-checking failure warnings to errors, which halts the program at
-  the point of contract failure.
+\item \contractr provides a very expressive API to the users, covering a
+  variety of uses cases. During interactive development, developers can
+  explicitly insert contracts by supplying the type declaration as a string
+  argument to the \code{insert\_contract} API function. Conversely,
+  contracts can be removed by calling the \code{remove\_contract} API on
+  functions. Contracts can be selectively enabled or disabled for code
+  blocks by wrapping them in calls to \code{ignore\_contracts} and
+  \code{capture\_contracts} functions. This enables selective type-checking
+  of code section during the development phase.  Furthermore, these
+  functions return a data frame that contains all the information about
+  failed and successful contract assertions in the wrapped code blocks. This
+  is useful for post-hoc investigation. Multiple type-checking failures can
+  introduce a lot of noise in the program output. To alleviate this problem,
+  \code{contractr.severity} option can be set to \code{'silence'}.  This
+  suppresses failure messages but \contractr still performs type-checking
+  whose results can be explicitly obtained as a data frame from the
+  \code{get\_contracts} API function. R suppresses printing of warnings on
+  the terminal when a function issues too many warnings. This can hide
+  type-checking failure messages. Setting \code{contractr.severity} flag to
+  \code{'error'} turns the type-checking failure warnings to errors, which
+  halts the program at the point of contract failure.
 
@@
 For this paper we have selected \CorpusLoadable packages consisting of
-\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native code
-(C/C++/Fortran).
-\FK{Add more details}
-Figure~\ref{fig:corpus} shows all the packages, the size of the dots reflects
-the size of the project in terms of both R and native lines of code\footnote{
-Lines of source code reported excludes comments and blank lines, counted by \emph{cloc}, \cf \url{https://github.com/AlDanial/cloc}}, the x-axis
-indicates the expression code coverage in percents and the y-axis gives the
-number of reverse dependencies in log scale. Dotted lines indicate means.
-Packages with over \PackageSizeOutierRnd lines of code are annotated.
+\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native
+code (C/C++/Fortran).  \FK{Add more details} Figure~\ref{fig:corpus} shows
+all the packages, the size of the dots reflects the size of the project in
+terms of both R and native lines of code\footnote{ Lines of source code
+  reported excludes comments and blank lines, counted by \emph{cloc}, \cf
+  \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the
+expression code coverage in percents and the y-axis gives the number of
+reverse dependencies in log scale. Dotted lines indicate means.  Packages
+with over \PackageSizeOutierRnd lines of code are annotated.
 
@@
 These packages come from the Comprehensive R Achive Network
-(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of R
-code with over \AllCranRnd packages\footnote{CRAN receives about 6 new package
-  submissions a day~\cite{Ligges2017}} containing over \AllRCodeRnd and
-\AllNativeCodeRnd R and native code respectively. Unlike other open code
-repositories like GitHub, CRAN is a curated repository. Each submitted package
-must abide to a number of well-formedness rules that are automatically checked
-asserting certain quality. Most relevant for this work is that all of the
-runnable code (including code snippets from examples and vignettes) is tested
-and only a successfully running package is admitted in the archive.
+(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of
+R code with over \AllCranRnd packages\footnote{CRAN receives about 6 new
+  package submissions a day~\cite{Ligges2017}} containing over \AllRCodeRnd
+and \AllNativeCodeRnd R and native code respectively. Unlike other open code
+repositories like GitHub, CRAN is a curated repository. Each submitted
+package must abide to a number of well-formedness rules that are
+automatically checked asserting certain quality. Most relevant for this work
+is that all of the runnable code (including code snippets from examples and
+vignettes) is tested and only a successfully running package is admitted in
+the archive.
 
 We have downloaded and installed all available CRAN packages. Out of the
-\AllCranRnd packages, we managed to install \AllLoadableRnd. The main
-reason for this is that \rdt is based on R 3.5.0 and some of the packages are no
-longer compatible with it. Some packages also require extra native dependencies
-which were not present on our servers.
-We have two criteria for including a
-package into the corpus:
+\AllCranRnd packages, we managed to install \AllLoadableRnd. The main reason
+for this is that \rdt is based on R 3.5.0 and some of the packages are no
+longer compatible with it. Some packages also require extra native
+dependencies which were not present on our servers.  We have two criteria
+for including a package into the corpus:
 \begin{inparaenum}[(1)]
@@
 \end{inparaenum}
-The concrete thresholds were set to at least \ThresholdCodeCoverage of expression
-coverage and at minimum \ThresholdRevdeps reverse dependencies.
+The concrete thresholds were set to at least \ThresholdCodeCoverage of
+expression coverage and at minimum \ThresholdRevdeps reverse dependencies.
 
 The code coverage was computed for each package using
-\covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R tool.
-The reverse package dependencies were extracted from the package metadata using
-builtin function.
-
-% \AT{This section needs to be tight.
-% Reviewers were skeptical of CRAN and using the test and example code.
-% We should convince them otherwise.}
-
-% \AT{Some sentences that might be useful:}
-% We selected packages that have both \AT{a good amount of} reverse dependencies to ensure their broad usefulness, as well \AT{sufficiently high} code and branch coverage of test, example, and vignette programs to ensure that our analysis built up an accurate picture of function use.
+\covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R
+tool.  The reverse package dependencies were extracted from the package
+metadata using builtin function.
 
@@
 \AT{So, we understand that we iterated on our type system design.
-Subsequently, the analysis has undergone many iterations, but essentially it collects the most detailed information possible.
-Do we want to mention all of that information, and how we ignore it later?
-Or do we just want to focus on the relevant information for our final types?}
-
-To help inform the design of our type system, and to ensure that it aligns with the day-to-day usage of R, we performed a corpus analysis of some of the most widely used R packages.
-The corpus is discussed in detail in Section~\ref{sec:corpus}.
-We built \typetracer, our dynamic analysis, in the R-dyntrace~\cite{oopsla19} dynamic analysis framework which is itself built on a modified version of R targeting R version 3.5.0---we refer the reader to Section~\ref{sec:r-dyntrace} for more details on the framework.
+  Subsequently, the analysis has undergone many iterations, but essentially
+  it collects the most detailed information possible.  Do we want to mention
+  all of that information, and how we ignore it later?  Or do we just want
+  to focus on the relevant information for our final types?}
+
+To help inform the design of our type system, and to ensure that it aligns
+with the day-to-day usage of R, we performed a corpus analysis of some of
+the most widely used R packages.  The corpus is discussed in detail in
+Section~\ref{sec:corpus}.  We built \typetracer, our dynamic analysis, in
+the R-dyntrace~\cite{oopsla19} dynamic analysis framework which is itself
+built on a modified version of R targeting R version 3.5.0---we refer the
+reader to Section~\ref{sec:r-dyntrace} for more details on the framework.
 
@@
 \subsubsection{Overview}
-The goal of our analysis is to discover argument types for functions, builtins, and specials.
-Naively intercepting function calls and recording argument types is incorrect, as recall that R wraps arguments in promises (Section \AT{what section}).
-To circumvent this obstacle, we instead intercept the forcing of promises: when a promise is forced, we identify the surrounding call, collect the type of the value, and add it to the call trace associated with the current function call.
-Once function execution completes, we collect information about the return value, add it to the call trace, and save the completed call trace to be written out later. 
-\AT{This paragraph overlaps with the initial discussion of \contractr. We need to make sure that we don't inadvertently repeat ourselves.}
+The goal of our analysis is to discover argument types for functions,
+builtins, and specials.  Naively intercepting function calls and recording
+argument types is incorrect, as recall that R wraps arguments in promises
+(Section \AT{what section}).  To circumvent this obstacle, we instead
+intercept the forcing of promises: when a promise is forced, we identify the
+surrounding call, collect the type of the value, and add it to the call
+trace associated with the current function call.  Once function execution
+completes, we collect information about the return value, add it to the call
+trace, and save the completed call trace to be written out later.  \AT{This
+  paragraph overlaps with the initial discussion of \contractr. We need to
+  make sure that we don't inadvertently repeat ourselves.}
 
@@
 \subsubsection{Type Information}
-Once a value is obtained, either on function return or when an argument promise is forced, we probe the value for the following information:
+Once a value is obtained, either on function return or when an argument
+promise is forced, we probe the value for the following information:
 
 \begin{itemize}
-\item the value's type tag according to R's runtime. R's runtime implements type tags as a typedef, e.g. \code{INTSXP} for integers, and, confusingly, \code{VECSXP} for lists;
-\item its class, which is a list of string class names.
-Typically, the class of a value is held in an attribute with name ``class'', though some classes are blessed by the R runtime, such as matrix and array classes;
+\item the value's type tag according to R's runtime. R's runtime implements
+  type tags as a typedef, e.g. \code{INTSXP} for integers, and, confusingly,
+  \code{VECSXP} for lists;
+\item its class, which is a list of string class names. Typically, the class
+  of a value is held in an attribute with name ``class'', though some
+  classes are blessed by the R runtime, such as matrix and array classes;
 \item its attributes, a list of metadata attached to the value.
@@
 \begin{itemize}
-\item For R's primitive types, we collect its dimension, and if it has a names attribute we collect the names as well;
+\item For R's primitive types, we collect its dimension, and if it has a
+  names attribute we collect the names as well;
 \item For lists, we recursively collect the types of list elements;
 \item For lists {\it with a names attribute}, we collect the names as well as the types of the slots that those names refer to (data.frames fall into this category);
-\item For promises, we carefully dig into the promise value {\it without forcing it}, and collect information about the innermost value.
-If the innermost value is the unbound value, then the promise is an as-of-yet unevaluated expression.
-In this case, we do our best to guess the value of the unevaluated expression: if the expression is not a language expression, we collect information about it, otherwise we can say nothing without evaluating it and ascribe it type ``any''.
-\footnote{\AT{I think this is confusing, reword.}}
+\item For promises, we carefully dig into the promise value {\it without
+  forcing it}, and collect information about the innermost value. If the
+  innermost value is the unbound value, then the promise is an as-of-yet
+  unevaluated expression. In this case, we do our best to guess the value of
+  the unevaluated expression: if the expression is not a language
+  expression, we collect information about it, otherwise we can say nothing
+  without evaluating it and ascribe it type ``any''. \footnote{\AT{I think
+      this is confusing, reword.}}
 \end{itemize}
 
-For vectors (and lists), we additionally iterate through the elements, checking if the values are NA (resp.
-NULL); if none are NA (resp.
-NULL), we ascribe the NA-free (resp.
-NULL-free) tag to the value.
-Arguments that are part of a function's varargs (denoted \ldots in function definitions) are tagged as such and ignored, as varargs can contain anything (and their type will be equivalent to and any-typed list).
+For vectors (and lists), we additionally iterate through the elements,
+checking if the values are NA (resp. NULL); if none are NA (resp. NULL), we
+ascribe the NA-free (resp. NULL-free) tag to the value.  Arguments that are
+part of a function's varargs (denoted \ldots in function definitions) are
+tagged as such and ignored, as varargs can contain anything (and their type
+will be equivalent to and any-typed list).
 
@@
 
-\item First, we wanted to ensure that \typetracer captured type for unevaluated arguments. 
-Not every argument of a function needs to be used during every function execution, and the approach we discussed above relies on argument promises being forced (and, consequently, on arguments being used).
-There are a few reasons for this.
-
-One possibility is that the argument was simply unused given the control flow of this particular execution (e.g. if the argument is only needed in one branch of an \code{if}).
-This mainly occurs when arguments have default values for some parameters that are not critical to every execution oft he function.
-In these cases, we still want to collect the type of the default value.
-
-Another possibility is that a value is used {\it without being forced}, as is the case when values are ``metaprogrammed'': \AT{describe, code example.}
-
-To tackle this, we make an {\it initial guess} of function argument types on function entry.
-This collects types for argument default values, and collects an ``unused'' type for unevaluated expressions.
-If an argument promise is later forced, we simply update the recorded type for the argument.
-
-\item Second, we wanted to account for formal arguments (arguments which are named in the function definition) that were {\it not} passed to the function.
-Put simply, a function was not called with all of its arguments, and no default value was specified for missing arguments.
-In this case, we record a ``missing'' type for the argument.
-
-There are two obvious ways to deal with missing arguments: type them as bot, or as top.
-As we are performing a dynamic analysis, based on package test code, we conservatively type these arguments as top, or \code{any}.
-It is impossible to know dynamically what all possible function behaviours are, \AT{I'm still not convinced about this.}
-
-\item Third, we had to contend with R's C implementation sometimes performing longjumps, which discards current R calls from the call stack.
-In order to ensure that call traces are not lost when a longjump occurs, \typetracer intercepts the unwinding process initiated by the longjump and mimics the functionality of the functions being exited.
-The R-dyntrace framework supplies the return value during the unwinding process (if there is indeed such a value), so we can capture it, analyze it, and include it in the call trace.
-In certain situations, however, no return value can be obtained (e.g.
-a longjump occurred as part of some error handling), and then we simply record a ``jumped'' return value, which we can deal with in post-processing.
-These longjumps occur quite often, particularly when S3 dispatch occurs, as the R implementation optimizes returns from dispatched functions by initiating a longjump.
+\item First, we wanted to ensure that \typetracer captured type for
+  unevaluated arguments.  Not every argument of a function needs to be used
+  during every function execution, and the approach we discussed above
+  relies on argument promises being forced (and, consequently, on arguments
+  being used).  There are a few reasons for this.
+
+One possibility is that the argument was simply unused given the control
+flow of this particular execution (e.g. if the argument is only needed in
+one branch of an \code{if}).  This mainly occurs when arguments have default
+values for some parameters that are not critical to every execution oft he
+function.  In these cases, we still want to collect the type of the default
+value.
+
+Another possibility is that a value is used {\it without being forced}, as
+is the case when values are ``metaprogrammed'': \AT{describe, code example.}
+
+To tackle this, we make an {\it initial guess} of function argument types on
+function entry. This collects types for argument default values, and
+collects an ``unused'' type for unevaluated expressions.  If an argument
+promise is later forced, we simply update the recorded type for the
+argument.
+
+\item Second, we wanted to account for formal arguments (arguments which are
+  named in the function definition) that were {\it not} passed to the
+  function.  Put simply, a function was not called with all of its
+  arguments, and no default value was specified for missing arguments.  In
+  this case, we record a ``missing'' type for the argument.
+
+There are two obvious ways to deal with missing arguments: type them as bot,
+or as top.  As we are performing a dynamic analysis, based on package test
+code, we conservatively type these arguments as top, or \code{any}.  It is
+impossible to know dynamically what all possible function behaviours are,
+\AT{I'm still not convinced about this.}
+
+\item Third, we had to contend with R's C implementation sometimes
+  performing longjumps, which discards current R calls from the call stack.
+  In order to ensure that call traces are not lost when a longjump occurs,
+  \typetracer intercepts the unwinding process initiated by the longjump and
+  mimics the functionality of the functions being exited.  The R-dyntrace
+  framework supplies the return value during the unwinding process (if there
+  is indeed such a value), so we can capture it, analyze it, and include it
+  in the call trace.  In certain situations, however, no return value can be
+  obtained (e.g. a longjump occurred as part of some error handling), and
+  then we simply record a ``jumped'' return value, which we can deal with in
+  post-processing.  These longjumps occur quite often, particularly when S3
+  dispatch occurs, as the R implementation optimizes returns from dispatched
+  functions by initiating a longjump.
 
@@
 
-As mentioned, our analysis is built on the R-dyntrace framework.
-R-dyntrace is a very efficient instrumentation framework for R, which is built on a modified version of the R runtime.
-Our dynamic type analysis only needs to be run once, ever---from the analysis results, we synthesize function signatures which are passed along to later type-checking tools which do not rely on a heavily modified runtime.
-
-We primarily rely on 8 R-dyntrace callbacks: \texttt{closure\_entry}, \texttt{closure\_exit}, \texttt{builtin\_entry}, \texttt{builtin\_exit}, \texttt{special\_entry},\texttt{special\_exit}, \texttt{promise\_force\_entry}, and \texttt{promise\_force\_exit}.
-These callbacks fire when closures, builtin, and special functions are entered and exited, and before and after the forcing of promises.
-The function-related callbacks are used mainly for bookkeeping: the analysis is notified that a closure, builtin, or special has been entered/exited by pushing/popping the call onto a stack.
-The calls themselves store a {\it call trace} object, where we store the type information that we collect, populated with initial guesses of argument types (as discussed in Section~\ref{subsec:typetracer-challenges}).
-Further, the type of dispatch is recorded on the {\tt \_entry} variants: \AT{recall} that there are multiple ways that R performs dynamic dispatch, based on the class of function arguments.
-We collect dispatch information in order to analyze dispatch usage patterns.
-
-The main work is done in {\tt promise\_force\_exit}.
-As arguments to functions in R are wrapped in promises, we delay our reflection to promise force time.
-In R, promises are essentially objects with value and expression slots.
-When a promise is created for a particular expression, that expression is merely put into the expression slot of a promise object, and a special {\it unbound} value is loaded into the value slot (unless some default value is specified, wherein it is loaded into the value slot---this is the case with default argument values).
-When the promise is forced, the expression is evaluated in the forcing context, and the value is stored in the value slot of the promise.
-So to get the type of a function argument, we check the value slot of the argument promise---perhaps recursively if the value itself is a promise.
-
-To construct a type for each argument, we make use of R's C FFI and use low-level machinery to collect type tags and attributes from the R runtime.
-The types that we provide to users are constructed during post-processing, and rely on the detailed information made available by these low-level reflection mechanisms.
+As mentioned, our analysis is built on the R-dyntrace framework.  R-dyntrace
+is a very efficient instrumentation framework for R, which is built on a
+modified version of the R runtime.  Our dynamic type analysis only needs to
+be run once, ever---from the analysis results, we synthesize function
+signatures which are passed along to later type-checking tools which do not
+rely on a heavily modified runtime.
+
+We primarily rely on 8 R-dyntrace callbacks: \texttt{closure\_entry},
+\texttt{closure\_exit}, \texttt{builtin\_entry}, \texttt{builtin\_exit},
+\texttt{special\_entry},\texttt{special\_exit},
+\texttt{promise\_force\_entry}, and \texttt{promise\_force\_exit}.  These
+callbacks fire when closures, builtin, and special functions are entered and
+exited, and before and after the forcing of promises.  The function-related
+callbacks are used mainly for bookkeeping: the analysis is notified that a
+closure, builtin, or special has been entered/exited by pushing/popping the
+call onto a stack.  The calls themselves store a {\it call trace} object,
+where we store the type information that we collect, populated with initial
+guesses of argument types (as discussed in
+Section~\ref{subsec:typetracer-challenges}).  Further, the type of dispatch
+is recorded on the {\tt \_entry} variants: \AT{recall} that there are
+multiple ways that R performs dynamic dispatch, based on the class of
+function arguments.  We collect dispatch information in order to analyze
+dispatch usage patterns.
+
+The main work is done in {\tt promise\_force\_exit}.  As arguments to
+functions in R are wrapped in promises, we delay our reflection to promise
+force time.  In R, promises are essentially objects with value and
+expression slots.  When a promise is created for a particular expression,
+that expression is merely put into the expression slot of a promise object,
+and a special {\it unbound} value is loaded into the value slot (unless some
+default value is specified, wherein it is loaded into the value slot---this
+is the case with default argument values).  When the promise is forced, the
+expression is evaluated in the forcing context, and the value is stored in
+the value slot of the promise.  So to get the type of a function argument,
+we check the value slot of the argument promise---perhaps recursively if the
+value itself is a promise.
+
+To construct a type for each argument, we make use of R's C FFI and use
+low-level machinery to collect type tags and attributes from the R runtime.
+The types that we provide to users are constructed during post-processing,
+and rely on the detailed information made available by these low-level
+reflection mechanisms.
 
@@
 
-The type information we collect is far more detailed than the types we actually generate, and so the next stage of our inference process is to distill the rich type information into simpler, manageable types.
-For each unique observed call trace, we translate the representation of the type of each of the function arguments and returns according to the design of our type system:
-for instance, we might have determined that a function returned a vector of integer with 4 elements, with names, that was free of NA values, in which case we would write out \code{integer[]} as the type.
-
-Once types have been simplified in this way, for each function we move to consolidate its multiple call traces into a single function type.
-We do this by combining the observed types of each argument into one type (possibly a union, mind), according to subtyping rules that are part of our design.
-The precise subtyping rules we used will be discussed further in Section~\ref{sec:typesystemdesign}.
-
-While our analysis collects more detailed information than necessary for type generation, the information proved useful for building up a usage profile of the language.
-Through it, we were able to determine the best translation rules (i.e., determining the type system we are translating to) through an iterative process, involving analyzing usage patterns that arose in the data and making design decisions based on them, revisiting the data as necessary.
-In the following section, we will discuss these decisions in more detail, highlighting the reasoning behind our type system design.
+The type information we collect is far more detailed than the types we
+actually generate, and so the next stage of our inference process is to
+distill the rich type information into simpler, manageable types.  For each
+unique observed call trace, we translate the representation of the type of
+each of the function arguments and returns according to the design of our
+type system: for instance, we might have determined that a function returned
+a vector of integer with 4 elements, with names, that was free of NA values,
+in which case we would write out \code{integer[]} as the type.
+
+Once types have been simplified in this way, for each function we move to
+consolidate its multiple call traces into a single function type.  We do
+this by combining the observed types of each argument into one type
+(possibly a union, mind), according to subtyping rules that are part of our
+design.  The precise subtyping rules we used will be discussed further in
+Section~\ref{sec:typesystemdesign}.
+
+While our analysis collects more detailed information than necessary for
+type generation, the information proved useful for building up a usage
+profile of the language. Through it, we were able to determine the best
+translation rules (i.e., determining the type system we are translating to)
+through an iterative process, involving analyzing usage patterns that arose
+in the data and making design decisions based on them, revisiting the data
+as necessary.  In the following section, we will discuss these decisions in
+more detail, highlighting the reasoning behind our type system design.
 
@@
 
-Our corpora analysis reported \AT{number} call traces for over \AT{number} functions.
-We distilled these call traces into function types which captured the dynamic behaviour of the functions.
-We emphasize that this was an iterative process: initial analysis suggested a certain set of types, which we tried and implemented, and further analysis suggested improvements, and so on.
-In this section, we will present the final type system that we designed, which we believe captures the essence of R.
-Throughout, we will touch on design decisions and weigh our choices against alternatives, speaking to the strengths and weaknesses of our chosen design.
+Our corpora analysis reported \AT{number} call traces for over \AT{number}
+functions.  We distilled these call traces into function types which
+captured the dynamic behaviour of the functions.  We emphasize that this was
+an iterative process: initial analysis suggested a certain set of types,
+which we tried and implemented, and further analysis suggested improvements,
+and so on.  In this section, we will present the final type system that we
+designed, which we believe captures the essence of R.  Throughout, we will
+touch on design decisions and weigh our choices against alternatives,
+speaking to the strengths and weaknesses of our chosen design.
 
@@
 
-As discussed in Section~\ref{subsubsec:backgroundtypes}, the workhorse types in R are the vectorized primitives: integers, doubles, complex numbers, logicals (booleans), characters, and raws (bytes).
-These are {\it vectorized} in that they are always considered to be vectors by the R runtime: for instance, even ``scalar'' numbers are considered to be unit-length vectors to R.
-These primitive vectors are homogeneous, in that a vector of doubles contains only doubles.
-
-Related to these primitive types is one of R's notions of ``nullness'': for each of the primitives, R distinguishes a special ``NA'' value which represents missing data.
-There is an NA for each of the primitive types (i.e. there is a double NA, integer NA, etc.).
-Thus, even homogeneous vectors can represent missing data with their appropriate NA value.
+As discussed in Section~\ref{subsubsec:backgroundtypes}, the workhorse types
+in R are the vectorized primitives: integers, doubles, complex numbers,
+logicals (booleans), characters, and raws (bytes).  These are {\it
+  vectorized} in that they are always considered to be vectors by the R
+runtime: for instance, even ``scalar'' numbers are considered to be
+unit-length vectors to R.  These primitive vectors are homogeneous, in that
+a vector of doubles contains only doubles.
+
+Related to these primitive types is one of R's notions of ``nullness'': for
+each of the primitives, R distinguishes a special ``NA'' value which
+represents missing data.  There is an NA for each of the primitive types
+(i.e. there is a double NA, integer NA, etc.).  Thus, even homogeneous
+vectors can represent missing data with their appropriate NA value.
 
@@
 
-A common primitive usage pattern is related to the presence of NAs: \AT{often}, programmers explicitly check for NAs in primitive vectors, possibly sanitizing them if they are present.
-For instance, consider the code in \AT{do an NA example}.
-NAs complicate practically every computation using primitive vectors, a fact that many programmers appear to be acutely aware of.
+A common primitive usage pattern is related to the presence of NAs:
+\AT{often}, programmers explicitly check for NAs in primitive vectors,
+possibly sanitizing them if they are present.  For instance, consider the
+code in \AT{do an NA example}.  NAs complicate practically every computation
+using primitive vectors, a fact that many programmers appear to be acutely
+aware of.
 
@@
 
-\AT{Maybe we combine the next 3 subsections into one on dimensionality? There are overlapping arguments.}
-Initial analyses of our data revealed that programmers \AT{often} use scalars (as explicitly as they can, anyway).
+\AT{Maybe we combine the next 3 subsections into one on dimensionality?
+  There are overlapping arguments.}  Initial analyses of our data revealed
+that programmers \AT{often} use scalars (as explicitly as they can, anyway).
 For example, consider the code in \AT{do a scalar example}.
 
-We found that scalars occurred more often than vectors, and chose the scalar type syntax as the default.
-The syntax for primitive scalars and vectors is as follows:
+We found that scalars occurred more often than vectors, and chose the scalar
+type syntax as the default.  The syntax for primitive scalars and vectors is
+as follows:
 
@@
 
-Another layer of complexity on vectorized primitives is the presence of ``matrices'', which are \AT{more detail}.
-To R, a vector is a single-dimensional matrix, and R coerces vectors to matrices where appropriate.
-That said, if all vectors are matrices, then any computation that is valid on a matrix should also be valid on a vector.
-Restrictions on matrix-types arguments amount to restrictions on the dimensions of the argument, which suggests a dependent type system.
-We will touch on this shortly.
-\AT{I can try to come up with some numbers, here. I'm open to a discussion on how best to word this.}
+Another layer of complexity on vectorized primitives is the presence of
+``matrices'', which are \AT{more detail}.  To R, a vector is a
+single-dimensional matrix, and R coerces vectors to matrices where
+appropriate.  That said, if all vectors are matrices, then any computation
+that is valid on a matrix should also be valid on a vector.  Restrictions on
+matrix-types arguments amount to restrictions on the dimensions of the
+argument, which suggests a dependent type system.  We will touch on this
+shortly.  \AT{I can try to come up with some numbers, here. I'm open to a
+  discussion on how best to word this.}
 
@@
 
-\AT{Will revisit to tighten discussion, need to think about best way to phrase this.
-Perhaps "dependent type system bad" is enough?}
-Including data dimensions in types would make our type system a dependent type system, which we believe is too complex to retrofit onto a dynamic data science language such as R.
-The simplest way to include dimensions would be to explicitly specify the dimensions of an argument (e.g.
-specifying that a function should always be called with a double vector of length 10), which is not particularly useful, and a pattern which \AT{rarely appeared in our analysis}.
+\AT{Will revisit to tighten discussion, need to think about best way to
+  phrase this.  Perhaps "dependent type system bad" is enough?}  Including
+data dimensions in types would make our type system a dependent type system,
+which we believe is too complex to retrofit onto a dynamic data science
+language such as R.  The simplest way to include dimensions would be to
+explicitly specify the dimensions of an argument (e.g. specifying that a
+function should always be called with a double vector of length 10), which
+is not particularly useful, and a pattern which \AT{rarely appeared in our
+  analysis}.
 
@@
 
-\AT{List rehash.}
-Separate from the vectorized primitives in R are lists:
-Whereas vectors are homogeneous and restricted to primitive types, lists can contain arbitrarily-typed data.
-R even builds on lists to create a more complex data structure known as the {\it data frame}, an essential data structure to R's functionality \AT{maybe too aggressively worded, but we can back this up with data, e.g.
-how many functions take data frames?}.
-
-\AT{Outline concerns for typing.}
-We are faced with several possible designs for list types.
-Chief concerns include the heterogeneity of list elements, and the presence of a \code{names} attribute, allowing list elements to be accessed in a struct-like manner.
-\AT{Rarely?}, programmers even write code expecting lists of a particular pre-defined length.
-The next sections highlight some important design decisions for our list types.
+\AT{List rehash.}  Separate from the vectorized primitives in R are lists:
+Whereas vectors are homogeneous and restricted to primitive types, lists can
+contain arbitrarily-typed data.  R even builds on lists to create a more
+complex data structure known as the {\it data frame}, an essential data
+structure to R's functionality \AT{maybe too aggressively worded, but we can
+  back this up with data, e.g. how many functions take data frames?}.
+
+\AT{Outline concerns for typing.}  We are faced with several possible
+designs for list types.  Chief concerns include the heterogeneity of list
+elements, and the presence of a \code{names} attribute, allowing list
+elements to be accessed in a struct-like manner.  \AT{Rarely?}, programmers
+even write code expecting lists of a particular pre-defined length.  The
+next sections highlight some important design decisions for our list types.
 
@@
 
-We briefly entertained the possibility of including a \code{tuple} type, which would have looked like e.g. \code{tuple<int, int>} for a list of two integers.
-When configured to detect tuples, our analysis did indeed pick up \AT{a number}, but post-processing and simplification revealed that in \AT{many} cases, they would be subsumed by co-occurring \code{list} types.
-In other words, it just so happened that functions were being called with short lists.
-Of the cases that tuples were not subsumed by lists, \AT{often} there were \AT{tuples of different lengths}, implying that the shape of the argument was not essential.
-\AT{Wording...}
+We briefly entertained the possibility of including a \code{tuple} type,
+which would have looked like e.g. \code{tuple<int, int>} for a list of two
+integers.  When configured to detect tuples, our analysis did indeed pick up
+\AT{a number}, but post-processing and simplification revealed that in
+\AT{many} cases, they would be subsumed by co-occurring \code{list} types.
+In other words, it just so happened that functions were being called with
+short lists.  Of the cases that tuples were not subsumed by lists,
+\AT{often} there were \AT{tuples of different lengths}, implying that the
+shape of the argument was not essential.  \AT{Wording...}
 
@@
 Another pattern that we observed was the use of consistently-named lists.
-In R, ``fields'' of named lists can be accessed using the \code{$} operator, such as:
+In R, ``fields'' of named lists can be accessed using the \code{$} operator,
+such as:
 \begin{lstlisting}
@@
 \end{lstlisting}
-We thought to capture this type of behaviour with a {\it struct} type, such as \code{struct<x: integer, y: integer>}.
-Initial analyses suggested that structs were very common, but further inspection revealed that they failed to tell the full story.
-For instance, \AT{many} structs were merely representing list-typed values with a {\it class}, and we could capture these types more succinctly with a class-based type.
-Further, it is common in R example code to use some built-in data sets, such as \AT{mention some}.
-These caused undue noise, as even with our struct name unification strategy, poorly tested functions appeared to operate on e.g. \AT{funny struct type from the previous data set}.
-Ultimately, we found that sticking to list types and adding a class-based was the wisest course of action.
-We will discuss class types shortly.
+We thought to capture this type of behaviour with a {\it struct} type, such
+as \code{struct<x: integer, y: integer>}.  Initial analyses suggested that
+structs were very common, but further inspection revealed that they failed
+to tell the full story.  For instance, \AT{many} structs were merely
+representing list-typed values with a {\it class}, and we could capture
+these types more succinctly with a class-based type.  Further, it is common
+in R example code to use some built-in data sets, such as \AT{mention some}.
+These caused undue noise, as even with our struct name unification strategy,
+poorly tested functions appeared to operate on e.g. \AT{funny struct type
+  from the previous data set}.  Ultimately, we found that sticking to list
+types and adding a class-based was the wisest course of action.  We will
+discuss class types shortly.
 
@@
 Speaking of classes, an important classful value in R is the data frame.
-Data frames underpin nearly all idiomatic use of the language, and even popular \code{tibble} (from the \code{tidyverse} ecosystem) and \code{data.table} (from the \code{data.table} package) build on the essential type.
-We entertained the possibility of including data frame types, but \AT{found them to be clunky? Question mark?}
-Consider: \AT{example of why data.frame types would be bad atm.}
-
-Data frames do have a type, through the \code{class<data.frame>} type, and future work will expand on this type, discussed further in Section~\ref{sec:futurework}.
+Data frames underpin nearly all idiomatic use of the language, and even
+popular \code{tibble} (from the \code{tidyverse} ecosystem) and
+\code{data.table} (from the \code{data.table} package) build on the
+essential type.  We entertained the possibility of including data frame
+types, but \AT{found them to be clunky? Question mark?}  Consider:
+\AT{example of why data.frame types would be bad atm.}
+
+Data frames do have a type, through the \code{class<data.frame>} type, and
+future work will expand on this type, discussed further in
+Section~\ref{sec:futurework}.
 
@@
 
-As discussed in Section~\ref{sec:R}, the R runtime has more than one notion of type.
-The {\it class} (rather, the {\it classes}, as values can have multiple!) of function arguments are used by R to {\it dispatch} the correct function call for the argument \AT{$\leftarrow$ wording}.
-This is best served with an example:
-\AT{Example, unless we talked about this earlier, in which case refer reader back.}
+As discussed in Section~\ref{sec:R}, the R runtime has more than one notion
+of type.  The {\it class} (rather, the {\it classes}, as values can have
+multiple!) of function arguments are used by R to {\it dispatch} the correct
+function call for the argument \AT{$\leftarrow$ wording}.  This is best
+served with an example: \AT{Example, unless we talked about this earlier, in
+  which case refer reader back.}
 
@@
 
-\AT{We used classes to stand-in for structs.
-The discussion here is naturally tied with the discussion of structs, so we need to make sure to divide the discussion appropriately.}
+\AT{We used classes to stand-in for structs.  The discussion here is
+  naturally tied with the discussion of structs, so we need to make sure to
+  divide the discussion appropriately.}
 
@@
 
-As R does, our type system supports multiple classes on values, and the type of a classful value is simply a list of all of its classes.
+As R does, our type system supports multiple classes on values, and the type
+of a classful value is simply a list of all of its classes.
 
@@
 
-R has multiple, disparate object systems, with the language providing direct support for S3, S4, and R5.
-Thus far, we've mainly discussed S3 and its dispatch mechanism, and indeed our class types roughly capture the S3 ``types'' of values.
-We will not discuss R5 as it is still under development.
-
-For S4, we have an \code{s4} type to denote S4 values, though we did not elaborate on this type.
-We stuck to simple S4 types since doing anything more would be immensely complicated, and outside of the scope of this work.
-For instance, the mechanics of S4 dispatch are more complex than for S3 (S4 dispatches on the class of {\it all} arguments), and users can define their own class hierarchies that we would need to incorporate in our type analysis and contract checking frameworks.
-Further, \AT{we found limited use of S4 during our analysis}, with S4 types being heavily used in a select subset of packages (\AT{such as ...}), some of which make heavy use of R's metaprogramming capabilities.
-Coming up with a type system that accounts for all of these factors, and consolidates multiple object-orientation frameworks, is an interesting problem in and of itself, and we aim to extend our type system to account for this in future work.
-For now, S3 types suffice: S3 is the most common object system in R, with no sophisticated hierarchies, and its semantics are simple enough to be captured by our proposed design.
+R has multiple, disparate object systems, with the language providing direct
+support for S3, S4, and R5.  Thus far, we've mainly discussed S3 and its
+dispatch mechanism, and indeed our class types roughly capture the S3
+``types'' of values.  We will not discuss R5 as it is still under
+development.
+
+For S4, we have an \code{s4} type to denote S4 values, though we did not
+elaborate on this type.  We stuck to simple S4 types since doing anything
+more would be immensely complicated, and outside of the scope of this work.
+For instance, the mechanics of S4 dispatch are more complex than for S3 (S4
+dispatches on the class of {\it all} arguments), and users can define their
+own class hierarchies that we would need to incorporate in our type analysis
+and contract checking frameworks.  Further, \AT{we found limited use of S4
+  during our analysis}, with S4 types being heavily used in a select subset
+of packages (\AT{such as ...}), some of which make heavy use of R's
+metaprogramming capabilities.  Coming up with a type system that accounts
+for all of these factors, and consolidates multiple object-orientation
+frameworks, is an interesting problem in and of itself, and we aim to extend
+our type system to account for this in future work.  For now, S3 types
+suffice: S3 is the most common object system in R, with no sophisticated
+hierarchies, and its semantics are simple enough to be captured by our
+proposed design.
 
@@
 
-\AT{The decisions we made when coming up with the types will be validated in the last section.
-This section is for determining how effective the type system is.
-We have two evaluation strategies currently: comparing against reverse dependencies, and comparing against the Kaggle programs.}
+\AT{The decisions we made when coming up with the types will be validated in
+  the last section.  This section is for determining how effective the type
+  system is.  We have two evaluation strategies currently: comparing against
+  reverse dependencies, and comparing against the Kaggle programs.}
 
@@
 
-\AT{Good to mention the related projects, e.g. how this was undertaken in other languages, like Ruby.
-There's a lot of literature in that space I think, I've read some of it but not all.
-We also have some related work in the old paper, I'll grab it and put it here.}
+\AT{Good to mention the related projects, e.g. how this was undertaken in
+  other languages, like Ruby.  There's a lot of literature in that space I
+  think, I've read some of it but not all.  We also have some related work
+  in the old paper, I'll grab it and put it here.}
 

### 2020-05-12 paper/main.tex
beginning evaluation section. lots to fill in, but some initial discussion is now present

@@
 
-\AT{The decisions we made when coming up with the types will be validated in
-  the last section.  This section is for determining how effective the type
-  system is.  We have two evaluation strategies currently: comparing against
-  reverse dependencies, and comparing against the Kaggle programs.}
+We contended with multiple type language designs, and in this section we will explore in more detail how we compared our candidate designs.
+We will also discuss a large-scale evaluation of our final set of type annotations.
+
+%
+%
+%
+%
+\subsection{Comparative Analyses}
+
+\AT{We compared different designs: with/out structs and tuples, with/out classes, with/out primitive subtyping. We looked at type sizes and determined that e.g. \code{? T} is a good type.}
+
+%
+%
+\subsubsection{Analysis of Polymorphism}
+
+\AT{How many arguments, functions are polymorphic? How many packages? Are there pathologically polymorphic packages? Why didn't you handle those?}
+
+%
+%
+\subsubsection{Use of \code{any}}
+
+\AT{When is any introduced? Why? How does it affect the results? What if missing/unused isn't any, do the results change substantially?}
+
+%
+%
+%
+%
+\subsection{Large-Scale Evaluation}
+
+Now that we have settled on a language design, we endeavour to evaluate it against a representative subset of existing R code.
+To this end, we took the detailed type information we collected from \AT{corpus} popular R packages, inferred types for the functions exposed by those packages, and ran the test, example, and vignette code of each of their reverse dependencies.
+
+%
+%
+\subsubsection{Experiment Setup}
+
+\AT{What is the design of the experiment? The setup? How much code was ran? What does this evaluate? Why does it mean anything for the annotation framework?}
+
+\typetracer is capable of inferring type signatures based on the call traces it has collected.
+These call traces were consolidated into a single function type for each package function:
+\typetracer employs unification strategies discussed in Section~\ref{sec:typesystemdesign} to keep the size of signatures in check and only generates an \code{any} type under extenuating circumstances: when inner list types got too big (we decided to cap the union type in lists to 5), when missing and unused arguments were recored, and finally when a longjump caused function execution to abort prematurely.
+Finally, we loaded each of these inferred types into \contractr, and ran all example code from the reverse dependencies of these popular packages.
+
+%
+%
+\subsubsection{Results}
+
+\AT{How many failing assertions?
+How many functions had these failed assertions?
+Which packages?
+How was the coverage?
+What are patterns in the actual types of the failed assertions?
+In the expected types?
+How many times did the code trigger the contracts?}
+
+Overall, we found that only \AT{2\%} of contract assertions failed.
+\AT{What does this indicate? One thought:}
+This indicates two things: firstly, the corpus of packages we selected is quite well-tested, and much of the intended function behaviour is exemplified in the package example code. 
+Secondly, and more importantly, this success indicates that our analysis framework is able to generate function types which accurately capture the behaviour of the function. 
+\AT{It kind of feels like these need to simultaneously be true. Maybe we pull in the fact that coverage is good?}
+
+Of the \AT{2\%} of failing contract assertions, we picked up on a few significant patterns.
+Firstly, we found that \AT{10\%} of the failing assertions were due to \code{double}-typed values being passed to function arguments expecting an \code{integer}.
+While we cannot say for sure that {\it all} of these cases could be handled by coercion, it is worth noting that coercion from doubles to integers is exceedingly common in R: the practice is not codified in the language spec, but many of R's internal C functions explicitly perform this coercion.
+\AT{We can put an example, e.g. nrow(df) is an integer but 5 is not. Also list access coerces the double to an int.}
+
+Next, \AT{30\%} of failing assertions were due to vector-typed primitive values being passed to function arguments expecting a \code{class<matrix>}.
+This case exemplifies something we have touched on earlier, namely that vectors can be readily considered to be unidimensional matrices.
+\AT{Is this coercion automatic? I've seen it mentioned often in the documentation.}
+
+Separately, we investigated the effect of running reverse dependency code on the code coverage of the packages in our corpus.
+We found that \AT{in many cases}.
+the reverse dependencies increased the code coverage (this aligns with findings in \AT{cite genthat}), and this was especially true for the functions for which contract assertions failed.
 

### 2020-05-12 paper/main.tex
possibly NA primitives

@@
 
-% %
-% %
-% \subsubsection{Cross-Validation Against Kaggle}
-
-% Kaggle is a website offering an environment for hosting Jupyter notebooks, and importantly provides access to a large number of open-source data sets and R and Python code.
-% This solved a major issue with R programs hosted on e.g. Github, where often the data is not made available.
-% Our aim is to run R programs hosted on Kaggle to validate the type signatures we generated for popular R packages.
-
-% \AT{Anything else to say on Kaggle, here? We will probably talk more about the corpus we analyzed later.}
-
 %
@@
 \subsection{typetracer and Initial Analysis}
-
-\AT{So, we understand that we iterated on our type system design.
-  Subsequently, the analysis has undergone many iterations, but essentially
-  it collects the most detailed information possible.  Do we want to mention
-  all of that information, and how we ignore it later?  Or do we just want
-  to focus on the relevant information for our final types?}
 
@@
 %
-\section{Type System Design}
+\section{Type Language Design}
 \label{sec:typesystemdesign}
@@
 an iterative process: initial analysis suggested a certain set of types,
-which we tried and implemented, and further analysis suggested improvements,
-and so on.  In this section, we will present the final type system that we
-designed, which we believe captures the essence of R.  Throughout, we will
+which we tried and implemented, and reanalysis suggested improvements,
+and so on.  In this section, we will present the final type annotation language design, 
+which we believe captures the essence of R.  Throughout, we will
 touch on design decisions and weigh our choices against alternatives,
 speaking to the strengths and weaknesses of our chosen design.
-
-\subsection{Type Grammar}
+We begin with basic, primitive types.
+
+%
+%
+%
+%
+\subsection{Basic Types}
+\label{subsec:basictypes}
+
+As discussed in Section~\ref{subsubsec:backgroundtypes}, the workhorse types
+in R are the vectorized primitives: integers, doubles, complex numbers,
+logicals (booleans), characters, and raws (bytes).  These are {\it
+  vectorized} in that they are always considered to be vectors by the R
+runtime: for instance, even ``scalar'' numbers are considered to be
+unit-length vectors to R.  These primitive vectors are homogeneous, in that
+a vector of doubles contains only doubles.
+
+Related to these primitive types is one of R's notions of ``nullness'': for
+each of the primitives, R distinguishes a special ``NA'' value which
+represents missing data.  There is an NA for each of the primitive types
+(i.e. there is a double NA, integer NA, etc.).  Thus, even homogeneous
+vectors can represent missing data with their appropriate NA value.
+
+As with everything in R, there is nuance in even the simplest cases.
+We will now discuss our design decisions relating to the vectorized primitives.
+
+%
+%
+\subsubsection{Possibly-NA Primitives}
+
+A common primitive usage pattern is related to the presence of NAs:
+\AT{often}, programmers explicitly check for NAs in primitive vectors,
+possibly sanitizing them if they are present.  NAs complicate practically every computation
+using primitive vectors, a fact that many programmers appear to be acutely
+aware of. For instance, consider the code in Figure~\ref{fig:na-example}.
+
+\begin{figure}[htbp]
+\begin{center}
+
+\begin{lstlisting}
+binom.profile <- function(x, n, ...) {
+  xn <- cbind(x = x, n = n)
+  ok <- !is.na(xn[, 1]) & !is.na(xn[, 2])
+  x <- xn[ok, "x"]
+  n <- xn[ok, "n"]
+  ...
+\end{lstlisting}
+
+\caption{NA checking example from the \code{binom} package.}
+\label{fig:na-example}
+\end{center}
+\end{figure}
+
+The \code{binom} package is a popular package \AT{do we want e.g.
+\# revdeps?} for computing confidence intervals for binomial experiments, and the \code{binom.profile} function does so using the profile likelihood, the details of which are not relevant here.
+Here, we highlighted a relevant snippet of code, which appears at the beginning of the function: the programmer first binds the vectors into a matrix (line 2), with one column for each of \code{x} and \code{n}, then finds rows where both columns are not NA (line 3), then extracts non-NA values and stores them into \code{x} and \code{n} respectively (lines 4-5).
+Importantly, this process is entirely quiet, and the user of \code{binom.profile} will not be notified if their data happens to contain NAs, and the NAs will be silently ignored. 
+In reality, the presence of NAs in data could signify some sort of data corruption, and quietly ignoring issues in data could lead users (read: statisticians) to draw flawed conclusions from erroneous data.
+
+To combat this, we introduce a modifier on primitive types: the type \code{^T} signifies that primitive type \code{T} might contain an NA value, while \code{T} signifies that \code{T} is NA-free. 
+If the \code{binom} developers were to annotate arguments \code{x} and \code{n} with such an NA-free type, \contractr will notify users when \code{binom.profile} is called with NA-full values, rather than just quietly ignoring them.
+\AT{Gather usage statistics.}
+
+%
+%
+\subsubsection{Scalar Primitives}
+
+\AT{Maybe we combine the next 3 subsections into one on dimensionality?
+  There are overlapping arguments.}  Initial analyses of our data revealed
+that programmers \AT{often} use scalars (as explicitly as they can, anyway).
+For example, consider the code in \AT{do a scalar example}.
+
+We found that scalars occurred more often than vectors, and chose the scalar
+type syntax as the default.  The syntax for primitive scalars and vectors is
+as follows:
+
+\AT{syntax...}
+
+%
+%
+\subsubsection{Forgoing Matrices}
+
+Another layer of complexity on vectorized primitives is the presence of
+``matrices'', which are \AT{more detail}.  To R, a vector is a
+single-dimensional matrix, and R coerces vectors to matrices where
+appropriate.  That said, if all vectors are matrices, then any computation
+that is valid on a matrix should also be valid on a vector.  Restrictions on
+matrix-types arguments amount to restrictions on the dimensions of the
+argument, which suggests a dependent type system.  We will touch on this
+shortly.  \AT{I can try to come up with some numbers, here. I'm open to a
+  discussion on how best to word this.}
+
+%
+%
+\subsubsection{Forgoing Dimensionality}
+
+\AT{Will revisit to tighten discussion, need to think about best way to
+  phrase this.  Perhaps "dependent type system bad" is enough?}  Including
+data dimensions in types would make our type system a dependent type system,
+which we believe is too complex to retrofit onto a dynamic data science
+language such as R.  The simplest way to include dimensions would be to
+explicitly specify the dimensions of an argument (e.g. specifying that a
+function should always be called with a double vector of length 10), which
+is not particularly useful, and a pattern which \AT{rarely appeared in our
+  analysis}.
+
+%
+%
+%
+%
+\subsection{Lists}
+
+\AT{List rehash.}  Separate from the vectorized primitives in R are lists:
+Whereas vectors are homogeneous and restricted to primitive types, lists can
+contain arbitrarily-typed data.  R even builds on lists to create a more
+complex data structure known as the {\it data frame}, an essential data
+structure to R's functionality \AT{maybe too aggressively worded, but we can
+  back this up with data, e.g. how many functions take data frames?}.
+
+\AT{Outline concerns for typing.}  We are faced with several possible
+designs for list types.  Chief concerns include the heterogeneity of list
+elements, and the presence of a \code{names} attribute, allowing list
+elements to be accessed in a struct-like manner.  \AT{Rarely?}, programmers
+even write code expecting lists of a particular pre-defined length.  The
+next sections highlight some important design decisions for our list types.
+
+\AT{Dig into concerns for typing.}
+
+%
+%
+\subsubsection{Forgoing Tuples, or Short, Constant Length Lists}
+
+We briefly entertained the possibility of including a \code{tuple} type,
+which would have looked like e.g. \code{tuple<int, int>} for a list of two
+integers.  When configured to detect tuples, our analysis did indeed pick up
+\AT{a number}, but post-processing and simplification revealed that in
+\AT{many} cases, they would be subsumed by co-occurring \code{list} types.
+In other words, it just so happened that functions were being called with
+short lists.  Of the cases that tuples were not subsumed by lists,
+\AT{often} there were \AT{tuples of different lengths}, implying that the
+shape of the argument was not essential.  \AT{Wording...}
+
+%
+%
+\subsubsection{Forgoing Structs}
+
+Another pattern that we observed was the use of consistently-named lists.
+In R, ``fields'' of named lists can be accessed using the \code{$} operator,
+such as:
+\begin{lstlisting}
+p <- list(x=1, y=2)
+p$x + p$y
+\end{lstlisting}
+We thought to capture this type of behaviour with a {\it struct} type, such
+as \code{struct<x: integer, y: integer>}.  Initial analyses suggested that
+structs were very common, but further inspection revealed that they failed
+to tell the full story.  For instance, \AT{many} structs were merely
+representing list-typed values with a {\it class}, and we could capture
+these types more succinctly with a class-based type.  Further, it is common
+in R example code to use some built-in data sets, such as \AT{mention some}.
+These caused undue noise, as even with our struct name unification strategy,
+poorly tested functions appeared to operate on e.g. \AT{funny struct type
+  from the previous data set}.  Ultimately, we found that sticking to list
+types and adding a class-based was the wisest course of action.  We will
+discuss class types shortly.
+
+%
+%
+\subsubsection{Forgoing Data Frames}
+
+Speaking of classes, an important classful value in R is the data frame.
+Data frames underpin nearly all idiomatic use of the language, and even
+popular \code{tibble} (from the \code{tidyverse} ecosystem) and
+\code{data.table} (from the \code{data.table} package) build on the
+essential type.  We entertained the possibility of including data frame
+types, but \AT{found them to be clunky? Question mark?}  Consider:
+\AT{example of why data.frame types would be bad atm.}
+
+Data frames do have a type, through the \code{class<data.frame>} type, and
+future work will expand on this type, discussed further in
+Section~\ref{sec:futurework}.
+
+%
+%
+%
+%
+\subsection{Classes in R}
+
+As discussed in Section~\ref{sec:R}, the R runtime has more than one notion
+of type.  The {\it class} (rather, the {\it classes}, as values can have
+multiple!) of function arguments are used by R to {\it dispatch} the correct
+function call for the argument \AT{$\leftarrow$ wording}.  This is best
+served with an example: \AT{Example, unless we talked about this earlier, in
+  which case refer reader back.}
+
+%
+%
+\subsubsection{Class Types}
+
+The class of a value suggest a shape for the value, and we felt class to be a more succinct way of capturing struct-like types.
+
+\AT{We used classes to stand-in for structs.  The discussion here is
+  naturally tied with the discussion of structs, so we need to make sure to
+  divide the discussion appropriately.}
+
+%
+%
+\subsubsection{Multi-Classes}
+
+As R does, our type system supports multiple classes on values, and the type
+of a classful value is simply a list of all of its classes.
+
+\AT{What more to say here? Discussing occurrences of multi-classes?}
+
+%
+%
+\subsubsection{Forgoing R's Object Systems}
+
+R has multiple, disparate object systems, with the language providing direct
+support for S3, S4, and R5.  Thus far, we've mainly discussed S3 and its
+dispatch mechanism, and indeed our class types roughly capture the S3
+``types'' of values.  We will not discuss R5 as it is still under
+development.
+
+For S4, we have an \code{s4} type to denote S4 values, though we did not
+elaborate on this type.  We stuck to simple S4 types since doing anything
+more would be immensely complicated, and outside of the scope of this work.
+For instance, the mechanics of S4 dispatch are more complex than for S3 (S4
+dispatches on the class of {\it all} arguments), and users can define their
+own class hierarchies that we would need to incorporate in our type analysis
+and contract checking frameworks.  Further, \AT{we found limited use of S4
+  during our analysis}, with S4 types being heavily used in a select subset
+of packages (\AT{such as ...}), some of which make heavy use of R's
+metaprogramming capabilities.  Coming up with a type system that accounts
+for all of these factors, and consolidates multiple object-orientation
+frameworks, is an interesting problem in and of itself, and we aim to extend
+our type system to account for this in future work.  For now, S3 types
+suffice: S3 is the most common object system in R, with no sophisticated
+hierarchies, and its semantics are simple enough to be captured by our
+proposed design.
+
+%
+%
+%
+%
+\subsection{Final Type Grammar}
 \label{subsec:typegrammar}
@@
 %
-\subsection{Basic Types}
-\label{subsec:basictypes}
-
-As discussed in Section~\ref{subsubsec:backgroundtypes}, the workhorse types
-in R are the vectorized primitives: integers, doubles, complex numbers,
-logicals (booleans), characters, and raws (bytes).  These are {\it
-  vectorized} in that they are always considered to be vectors by the R
-runtime: for instance, even ``scalar'' numbers are considered to be
-unit-length vectors to R.  These primitive vectors are homogeneous, in that
-a vector of doubles contains only doubles.
-
-Related to these primitive types is one of R's notions of ``nullness'': for
-each of the primitives, R distinguishes a special ``NA'' value which
-represents missing data.  There is an NA for each of the primitive types
-(i.e. there is a double NA, integer NA, etc.).  Thus, even homogeneous
-vectors can represent missing data with their appropriate NA value.
-
-As with everything in R, there is nuance in even the simplest cases.
-We will now discuss our design decisions relating to the vectorized primitives.
-
-%
-%
-\subsubsection{Possibly-NA Primitives}
-
-A common primitive usage pattern is related to the presence of NAs:
-\AT{often}, programmers explicitly check for NAs in primitive vectors,
-possibly sanitizing them if they are present.  For instance, consider the
-code in \AT{do an NA example}.  NAs complicate practically every computation
-using primitive vectors, a fact that many programmers appear to be acutely
-aware of.
-
-%
-%
-\subsubsection{Scalar Primitives}
-
-\AT{Maybe we combine the next 3 subsections into one on dimensionality?
-  There are overlapping arguments.}  Initial analyses of our data revealed
-that programmers \AT{often} use scalars (as explicitly as they can, anyway).
-For example, consider the code in \AT{do a scalar example}.
-
-We found that scalars occurred more often than vectors, and chose the scalar
-type syntax as the default.  The syntax for primitive scalars and vectors is
-as follows:
-
-\AT{syntax...}
-
-%
-%
-\subsubsection{Forgoing Matrices}
-
-Another layer of complexity on vectorized primitives is the presence of
-``matrices'', which are \AT{more detail}.  To R, a vector is a
-single-dimensional matrix, and R coerces vectors to matrices where
-appropriate.  That said, if all vectors are matrices, then any computation
-that is valid on a matrix should also be valid on a vector.  Restrictions on
-matrix-types arguments amount to restrictions on the dimensions of the
-argument, which suggests a dependent type system.  We will touch on this
-shortly.  \AT{I can try to come up with some numbers, here. I'm open to a
-  discussion on how best to word this.}
-
-%
-%
-\subsubsection{Forgoing Dimensionality}
-
-\AT{Will revisit to tighten discussion, need to think about best way to
-  phrase this.  Perhaps "dependent type system bad" is enough?}  Including
-data dimensions in types would make our type system a dependent type system,
-which we believe is too complex to retrofit onto a dynamic data science
-language such as R.  The simplest way to include dimensions would be to
-explicitly specify the dimensions of an argument (e.g. specifying that a
-function should always be called with a double vector of length 10), which
-is not particularly useful, and a pattern which \AT{rarely appeared in our
-  analysis}.
-
-%
-%
-%
-%
-\subsection{Lists}
-
-\AT{List rehash.}  Separate from the vectorized primitives in R are lists:
-Whereas vectors are homogeneous and restricted to primitive types, lists can
-contain arbitrarily-typed data.  R even builds on lists to create a more
-complex data structure known as the {\it data frame}, an essential data
-structure to R's functionality \AT{maybe too aggressively worded, but we can
-  back this up with data, e.g. how many functions take data frames?}.
-
-\AT{Outline concerns for typing.}  We are faced with several possible
-designs for list types.  Chief concerns include the heterogeneity of list
-elements, and the presence of a \code{names} attribute, allowing list
-elements to be accessed in a struct-like manner.  \AT{Rarely?}, programmers
-even write code expecting lists of a particular pre-defined length.  The
-next sections highlight some important design decisions for our list types.
-
-\AT{Dig into concerns for typing.}
-
-%
-%
-\subsubsection{Forgoing Tuples, or Short, Constant Length Lists}
-
-We briefly entertained the possibility of including a \code{tuple} type,
-which would have looked like e.g. \code{tuple<int, int>} for a list of two
-integers.  When configured to detect tuples, our analysis did indeed pick up
-\AT{a number}, but post-processing and simplification revealed that in
-\AT{many} cases, they would be subsumed by co-occurring \code{list} types.
-In other words, it just so happened that functions were being called with
-short lists.  Of the cases that tuples were not subsumed by lists,
-\AT{often} there were \AT{tuples of different lengths}, implying that the
-shape of the argument was not essential.  \AT{Wording...}
-
-%
-%
-\subsubsection{Forgoing Structs}
-
-Another pattern that we observed was the use of consistently-named lists.
-In R, ``fields'' of named lists can be accessed using the \code{$} operator,
-such as:
-\begin{lstlisting}
-p <- list(x=1, y=2)
-p$x + p$y
-\end{lstlisting}
-We thought to capture this type of behaviour with a {\it struct} type, such
-as \code{struct<x: integer, y: integer>}.  Initial analyses suggested that
-structs were very common, but further inspection revealed that they failed
-to tell the full story.  For instance, \AT{many} structs were merely
-representing list-typed values with a {\it class}, and we could capture
-these types more succinctly with a class-based type.  Further, it is common
-in R example code to use some built-in data sets, such as \AT{mention some}.
-These caused undue noise, as even with our struct name unification strategy,
-poorly tested functions appeared to operate on e.g. \AT{funny struct type
-  from the previous data set}.  Ultimately, we found that sticking to list
-types and adding a class-based was the wisest course of action.  We will
-discuss class types shortly.
-
-%
-%
-\subsubsection{Forgoing Data Frames}
-
-Speaking of classes, an important classful value in R is the data frame.
-Data frames underpin nearly all idiomatic use of the language, and even
-popular \code{tibble} (from the \code{tidyverse} ecosystem) and
-\code{data.table} (from the \code{data.table} package) build on the
-essential type.  We entertained the possibility of including data frame
-types, but \AT{found them to be clunky? Question mark?}  Consider:
-\AT{example of why data.frame types would be bad atm.}
-
-Data frames do have a type, through the \code{class<data.frame>} type, and
-future work will expand on this type, discussed further in
-Section~\ref{sec:futurework}.
-
-%
-%
-%
-%
-\subsection{Classes in R}
-
-As discussed in Section~\ref{sec:R}, the R runtime has more than one notion
-of type.  The {\it class} (rather, the {\it classes}, as values can have
-multiple!) of function arguments are used by R to {\it dispatch} the correct
-function call for the argument \AT{$\leftarrow$ wording}.  This is best
-served with an example: \AT{Example, unless we talked about this earlier, in
-  which case refer reader back.}
-
-%
-%
-\subsubsection{Class Types}
-
-The class of a value suggest a shape for the value, and we felt class to be a more succinct way of capturing struct-like types.
-
-\AT{We used classes to stand-in for structs.  The discussion here is
-  naturally tied with the discussion of structs, so we need to make sure to
-  divide the discussion appropriately.}
-
-%
-%
-\subsubsection{Multi-Classes}
-
-As R does, our type system supports multiple classes on values, and the type
-of a classful value is simply a list of all of its classes.
-
-\AT{What more to say here? Discussing occurrences of multi-classes?}
-
-%
-%
-\subsubsection{Forgoing R's Object Systems}
-
-R has multiple, disparate object systems, with the language providing direct
-support for S3, S4, and R5.  Thus far, we've mainly discussed S3 and its
-dispatch mechanism, and indeed our class types roughly capture the S3
-``types'' of values.  We will not discuss R5 as it is still under
-development.
-
-For S4, we have an \code{s4} type to denote S4 values, though we did not
-elaborate on this type.  We stuck to simple S4 types since doing anything
-more would be immensely complicated, and outside of the scope of this work.
-For instance, the mechanics of S4 dispatch are more complex than for S3 (S4
-dispatches on the class of {\it all} arguments), and users can define their
-own class hierarchies that we would need to incorporate in our type analysis
-and contract checking frameworks.  Further, \AT{we found limited use of S4
-  during our analysis}, with S4 types being heavily used in a select subset
-of packages (\AT{such as ...}), some of which make heavy use of R's
-metaprogramming capabilities.  Coming up with a type system that accounts
-for all of these factors, and consolidates multiple object-orientation
-frameworks, is an interesting problem in and of itself, and we aim to extend
-our type system to account for this in future work.  For now, S3 types
-suffice: S3 is the most common object system in R, with no sophisticated
-hierarchies, and its semantics are simple enough to be captured by our
-proposed design.
-
-%
-%
-%
-%
 %

### 2020-05-12 paper/main.tex
scalars vs vectors

@@
 
+Note that when combining call traces involving \code{^T} and \code{T} types, we subsume \code{T} types into \code{^T} types: if NA-free and NA-full vectors are passed to an argument, then we say that the argument can handle NA-full vectors.
+This is equivalent to \code{T <: ^T}.
+
 %
@@
 
-\AT{Maybe we combine the next 3 subsections into one on dimensionality?
-  There are overlapping arguments.}  Initial analyses of our data revealed
-that programmers \AT{often} use scalars (as explicitly as they can, anyway).
-For example, consider the code in \AT{do a scalar example}.
-
-We found that scalars occurred more often than vectors, and chose the scalar
-type syntax as the default.  The syntax for primitive scalars and vectors is
-as follows:
-
-\AT{syntax...}
+Initial analyses of our data revealed that programmers \AT{often} use scalars (as explicitly as they can, anyway), and often do high-level dimensionality checks on their data.
+To illustrate, consider the code in Figure~\ref{fig:scalar-vector-example}.
+
+\begin{figure}[htbp]
+\begin{center}
+
+\begin{lstlisting}
+hankel.matrix <- function( n, x ) { 
+### arguments
+### n = a positive integer value for the order of the Hankel matrix
+### x = an order 2 * n + 1 vector of numeric values
+    ...
+    if ( n != trunc( n ) )
+        stop( "argument n ix not an integer" )
+    if ( !is.vector( x ) )
+        stop( "argument x is not a vector" )
+    m <- length( x )
+    if ( m < n )
+        stop( "length of argument x is less than n" )
+    ...
+\end{lstlisting}
+
+\caption{Scalar/vector example from the \code{matrixcalc} package.}
+\label{fig:scalar-vector-example}
+\end{center}
+\end{figure}
+
+The \code{hankel.matrix} function takes two arguments (as described in the comment in the code), and returns a Hankel matrix, which essentially is a matrix where the skew-diagonals have a desirable property.
+The function body has a number of interesting checks:
+on lines 6-7, \code{n} undergoes an integer ``type check'';
+on lines 8-9, \code{x} undergoes a vector ``type check'';
+on lines 11-12, \code{n} is assumed to be a scalar (if it were a vector, R would generate a warning, as \code{m < n} would return a vector of logicals obtained from checking if \code{m} were less than each element of \code{n} \AT{do we want to put the warning here?}).
+These checks exemplify behaviour we would like to capture with our annotation language, as we would like to differentiate between scalars and vectors in addition to checking for the basic type of a value (e.g.
+integer vs double),
+
+To that end, we introduce the following types into our type language:
+for some primitive type \code{T}, the user may modify \code{T} with a modifier \code{[]} to indicate that it is a vector of \code{T}.
+Perhaps surprisingly, we found that scalars occurred more often than vectors, and chose the scalar type syntax as the default. 
+The NA-free modifier works as expected: \code{^T[]} indicates a vector of \code{T} that is free of NAs, while \code{^T} indicates a scalar which is not NA.
 

### 2020-05-12 paper/main.tex
matrix discussion

@@
   stringstyle=\color{Gray},
-  numbers=left
+  numbers=left,
+  showstringspaces=false
 }
@@
 %
-\subsubsection{Forgoing Matrices}
-
-Another layer of complexity on vectorized primitives is the presence of
-``matrices'', which are \AT{more detail}.  To R, a vector is a
-single-dimensional matrix, and R coerces vectors to matrices where
-appropriate.  That said, if all vectors are matrices, then any computation
-that is valid on a matrix should also be valid on a vector.  Restrictions on
-matrix-types arguments amount to restrictions on the dimensions of the
-argument, which suggests a dependent type system.  We will touch on this
-shortly.  \AT{I can try to come up with some numbers, here. I'm open to a
-  discussion on how best to word this.}
+\subsubsection{Matrices}
+
+On the subject of matrices, they represent another layer of complexity on top of vectorized primitives.
+Internally, matrices are simply vectors with class ``matrix'' and a \code{dims} attribute indicating matrix dimensions, and many functions automatically coerce vectors to matrices when appropriate (while not codified in the language semantics, many internal functions perform the coercion).
+Relying on automatic coercion is unwise, and many functions check to ensure that their arguments are matrices, particularly those in packages which rely on matrices.
+For instance, consider the code in Figure~\ref{fig:matrix-example}.
+
+\begin{figure}[htbp]
+\begin{center}
+
+\begin{lstlisting}
+rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL,
+                             na.rm = FALSE, ...) {
+  # - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+  # Validate arguments
+  # - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+  # Argument 'x':
+  if (!is.matrix(x)) {
+    .Defunct(msg = sprintf("Argument 'x' is of class %s, but should be a matrix. 
+    The use of a %s is not supported, the correctness of the result is not guaranteed. 
+    Please update your code accordingly.", sQuote(class(x)[1]), sQuote(class(x)[1])))
+  }
+  ...
+\end{lstlisting}
+
+\caption{Matrix example from the \code{matrixStats} package.}
+\label{fig:matrix-example}
+\end{center}
+\end{figure}
+
+The \code{rowWeightedMeans} function calculated the weighted means of rows or columns of a matrix \code{x}.
+This function exhibits good practice in type checking arguments, rightly producing a message to the user indicating that an explicit matrix should be passed.
+
+To capture cases like this, we include a \code{class<matrix>} type in our type annotation language.
+If \code{rowWeightedMeans} were outfitted with such a type, \contractr can perform the check present in this code automatically.
+That said, we don't' intend to subsume all validation code, a topic we discuss next.
 

### 2020-05-12 paper/main.tex
lists and relationship between list and vectors

@@
 If \code{rowWeightedMeans} were outfitted with such a type, \contractr can perform the check present in this code automatically.
-That said, we don't' intend to subsume all validation code, a topic we discuss next.
+That said, we don't intend to subsume all argument validation code, discussed next.
 
@@
 
-\AT{Will revisit to tighten discussion, need to think about best way to
-  phrase this.  Perhaps "dependent type system bad" is enough?}  Including
-data dimensions in types would make our type system a dependent type system,
-which we believe is too complex to retrofit onto a dynamic data science
-language such as R.  The simplest way to include dimensions would be to
-explicitly specify the dimensions of an argument (e.g. specifying that a
-function should always be called with a double vector of length 10), which
-is not particularly useful, and a pattern which \AT{rarely appeared in our
-  analysis}.
+Given the discussion of scalars, vectors, and matrices, it might be tempting to consider factoring dimensionality into types.
+The primary drawback of this would be rendering our type annotation language into a language of dependent types, which we believe too complex to retrofit onto a dynamic data science language such as R.
+This does mean our types cannot subsume all validation code, such as some later code from the aforementioned \code{rowWeightedMeans} function, shown in Figure~\ref{fig:general-validation-example}.
+
+\begin{figure}[htbp]
+\begin{center}
+
+\begin{lstlisting}
+rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL,
+                             na.rm = FALSE, ...) {
+  ...
+  n <- ncol(x)
+  if (length(w) != n) {
+    stop("The length of argument 'w' is does not match the number of column in 'x': ", 
+    length(w), " != ", n)
+  }
+  ...
+\end{lstlisting}
+
+\caption{General dimensionality check example from the \code{matrixStats} package.}
+\label{fig:general-validation-example}
+\end{center}
+\end{figure}
+
+Here, the length of the weights vector \code{w} is being checked for equality with the number of columns in the matrix \code{x}. 
+Ultimately, we decided to leave these kinds of checks up to the programmer.
+\AT{I feel like we might want a better argument to not make our type language dependent (better than lol pls no dependent types), but I could be wrong.
+Maybe reviewers would be ok with a simple argument.}
 
@@
 
-\AT{List rehash.}  Separate from the vectorized primitives in R are lists:
-Whereas vectors are homogeneous and restricted to primitive types, lists can
-contain arbitrarily-typed data.  R even builds on lists to create a more
-complex data structure known as the {\it data frame}, an essential data
-structure to R's functionality \AT{maybe too aggressively worded, but we can
-  back this up with data, e.g. how many functions take data frames?}.
-
-\AT{Outline concerns for typing.}  We are faced with several possible
-designs for list types.  Chief concerns include the heterogeneity of list
-elements, and the presence of a \code{names} attribute, allowing list
-elements to be accessed in a struct-like manner.  \AT{Rarely?}, programmers
-even write code expecting lists of a particular pre-defined length.  The
-next sections highlight some important design decisions for our list types.
-
-\AT{Dig into concerns for typing.}
+Separate from the vectorized primitives in R are lists:
+Whereas vectors are homogeneous and restricted to primitive types, lists can contain arbitrarily-typed data.  
+Further, lists and vectors have a different syntax for accessing elements, where lists are indexed using \code{[[} and vectors using \code{[}.
+R even builds on lists to create a more complex data structure known as the {\it data frame}, an essential data structure to R's functionality.
+
+We are faced with several possible designs for list types.  
+Chief concerns include the heterogeneity of list elements, and the presence of a \code{names} attribute, allowing list elements to be accessed in a struct-like manner.  
+\AT{Rarely?}, programmers even write code expecting lists of a particular pre-defined length.  
+The next sections highlight some important design decisions for our list types.
+
+%
+%
+\subsubsection{Relationship Between Lists and Vectors}
+
+The ease with which vectors may be transformed into lists is perhaps indicative of a relationship between the two.
+In R, all vectors can be converted to lists with a simple call to \code{as.list}, and lists to vectors with \code{unlist}, though care must be taken when unlisting a non-homogeneous list, as coercions will be performed where possible, and the function will simply do nothing if coercion is impossible.
+That said, lists and vectors have different access syntax, so vectors cannot stand-in for lists without the \code{as.list} call above.
+
+In sum, we opted not to include a subtyping relationship between vectors and lists, keeping both types separate.
+In practice, \AT{how often do \code{T[] | list<T>} types occur?}.
+During our evaluation we also found it rare that programmers would attempt to pass list-typed values to vector-typed arguments and vice versa, and we will discuss this further in Section~\ref{sec:evaluation}.
 

### 2020-05-13 paper/main.tex
structs

@@
 
-We briefly entertained the possibility of including a \code{tuple} type,
-which would have looked like e.g. \code{tuple<int, int>} for a list of two
-integers.  When configured to detect tuples, our analysis did indeed pick up
-\AT{a number}, but post-processing and simplification revealed that in
-\AT{many} cases, they would be subsumed by co-occurring \code{list} types.
-In other words, it just so happened that functions were being called with
-short lists.  Of the cases that tuples were not subsumed by lists,
-\AT{often} there were \AT{tuples of different lengths}, implying that the
-shape of the argument was not essential.  \AT{Wording...}
-
-%
-%
-\subsubsection{Forgoing Structs}
+We briefly entertained the possibility of including a \code{tuple} type, which would have looked like e.g. \code{tuple<int, int>} for a list of two integers.  
+When configured to detect tuples, our analysis did indeed pick up \AT{a number}, but post-processing and simplification revealed that in \AT{many} cases, they would be subsumed by co-occurring \code{list} types.
+In other words, more often than not it just so happened that functions were being called with short lists.  
+Of the cases that tuples were not subsumed by lists, \AT{often} there were \AT{tuples of different lengths}, implying that the shape of the argument was not essential.
+
+\AT{I feel like this is the least important of the design decisions we made, so I'm leaving it for the end, we might not even want it.}
+
+%
+%
+\subsubsection{Structs}
 
 Another pattern that we observed was the use of consistently-named lists.
-In R, ``fields'' of named lists can be accessed using the \code{$} operator,
-such as:
+In R, ``fields'' of named lists can be accessed using the \code{$} operator, as illustrated in the code snippet in Figure~\ref{fig:struct-ex}.
+
+\begin{figure}[htbp]
+\begin{center}
+
 \begin{lstlisting}
-p <- list(x=1, y=2)
-p$x + p$y
+calculate.neuralnet <- function (data, model.list, hidden, rep,
+            startweights, exclude, constant.weights) {
+    result <- generate.startweights(model.list, hidden, startweights,
+                                    rep, exclude, constant.weights)
+    weights <- result$weights
+    exclude <- result$exclude
+    ...
 \end{lstlisting}
-We thought to capture this type of behaviour with a {\it struct} type, such
-as \code{struct<x: integer, y: integer>}.  Initial analyses suggested that
-structs were very common, but further inspection revealed that they failed
-to tell the full story.  For instance, \AT{many} structs were merely
-representing list-typed values with a {\it class}, and we could capture
-these types more succinctly with a class-based type.  Further, it is common
-in R example code to use some built-in data sets, such as \AT{mention some}.
-These caused undue noise, as even with our struct name unification strategy,
-poorly tested functions appeared to operate on e.g. \AT{funny struct type
-  from the previous data set}.  Ultimately, we found that sticking to list
-types and adding a class-based was the wisest course of action.  We will
-discuss class types shortly.
+
+\caption{Struct example from the \code{neuralnet} package.}
+\label{fig:general-validation-example}
+\end{center}
+\end{figure}
+
+In the code snippet, the function \code{generate.startweights} return a list with (at least) two named elements: \code{weights} and \code{exclude}, which are accessed through \code{result$weights} and \code{result$exclude} respectively.
+We include a struct type in our language to capture this behaviour: we would encode the return type of \code{generate.startweights} as e.g. \code{struct<x: double[], y: logical[]>}.  
+We also thought to encode data frames as structs, as they are essentially lists with named columns of homogeneous values, but it was difficult to draw any meaningful conclusions from the data as unfortunately structs introduced a lot of noise into our analysis.
+
+It is common for R example code to use some built-in data sets, such as \code{mtcars}, a data set extracted from a Motor Trends US magazine in 1974.
+These data sets are used to show off some package functionality with ``real data'', and when \typetracer is configured to detect structs we saw a marked increase in misrepresentative types.
+We implemented a unification strategy for struct types, removing them if they co-occurred with list types, and turning them into lists if two co-occurring struct types shared no common names, but even so we found ourselves with function signatures that were full of noise introduced by the struct names.
+
+We still believe struct types to be useful to R programmers, and so our language supports the type and \contractr will check them, though we configured \typetracer to ignore them as they made the data substantially less useful.
+On the positive side, we found that \AT{get the number, but often} structs-typed values had a {\it class}, and so we decided to include class-based types.
+Classes are discussed in more detail in the next section, though we will quickly discuss data frames before focusing on them.
 

### 2020-05-13 evaluation/R/evaluate-type-analysis.Rmd
Add static type analysis and html document

@@
 title: Evaluation of Types
-author: Aviral Goel
 output:
@@
     theme: cerulean
-
+    toc: true
+    toc_float: true
 params:
@@
 ```{r setup, echo=FALSE, eval=TRUE}
-#    toc: true
-#    toc_float: true
+
 suppressPackageStartupMessages(library(stringr))
@@
 
+infer_kind <- function(type) {
+    if(type %in% c("integer", "double", "character", "logical", "complex", "raw")) {
+        "scalar"
+    }
+    else if(type %in% c("^integer", "^double", "^character", "^logical", "^complex")) {
+        "scalar-na"
+    }
+    else if(type %in% c("integer[]", "double[]", "character[]", "logical[]", "complex[]", "raw[]")) {
+        "vector"
+    }
+    else if(type %in% c("^integer[]", "^double[]", "^character[]", "^logical[]", "^complex[]")) {
+        "vector-na"
+    }
+    else if(str_detect(type, "^list")) {
+        "list"
+    }
+    else if(str_detect(type, "^struct")) {
+        "struct"
+    }
+    else if(str_detect(type, "^class")) {
+        "class"
+    }
+    else if(type == "any") {
+        "any"
+    }
+    else if(type == "null") {
+        "null"
+    }
+    else if(type == "...") {
+        "..."
+    }
+    else {
+        "other"
+    }
+}
+
+type_analysis_data <-
+   type_analysis_data %>%
+   mutate(kind = map_chr(type, infer_kind))
+
 to_percentage <- function(proportions) {
@@
 
-# High-level Summary
+# Incompatible Functions
+
+```{r incompatible-functions, echo=FALSE, eval=TRUE}
+type_compatibility <-
+    type_analysis_data %>%
+    group_by(package_name, function_name, outer_alternative) %>%
+    summarize(parameter_count = length(unique(parameter_position))) %>%
+    ungroup() %>%
+    group_by(package_name, function_name) %>%
+    mutate(compatible = length(unique(parameter_count)) == 1) %>%
+    ungroup()
+
+incompatible_functions <-
+    type_compatibility %>%
+    filter(!compatible) %>%
+    select(package_name, function_name) %>%
+    distinct()
+
+datatable(incompatible_functions)
+```
+
+There are `r nrow(distinct(incompatible_functions, package_name, function_name))` incompatible functions. 
+We remove them from the rest of the analysis.
+
+```{r remove-incompatible-functions, echo=FALSE, eval=TRUE}
+type_analysis_data <-
+    type_analysis_data %>%
+    dplyr::filter(!(paste0(package_name, c("::"), function_name) %in%
+                    paste0(incompatible_functions$package_name,
+                           "::",
+                           incompatible_functions$function_name)))
+```
 
@@
 
-In a function declaration of the form:
+# Corpus
 
-`type f (<int | chr | lgl, dbl, lgl> => lgl | chr) | (<dbl | chr, int, raw> => lgl | chr)`
+## Function Frequency
 
-There are two outer alternatives and 3 inner alternatives for parameter 1.
+```{r function-frequency, eval=TRUE, echo=FALSE}
+package_function_count <-
+    type_analysis_data %>%
+    group_by(package_name) %>%
+    summarize(function_count = length(unique(function_name))) %>%
+    ungroup()
 
-# Outer Alternatives
+datatable(package_function_count)
 
-Now, we look at the number of alternatives in the function type.
+function_frequency <-
+package_function_count %>%
+mutate(function_count = as.integer(5 * ceiling(function_count / 5))) %>%
+group_by(function_count) %>%
+summarize(package_count = n()) %>%
+ungroup() %>%
+mutate(package_perc = round(100 * package_count / sum(package_count), 2)) %>%
+mutate(package_cumperc = round(100 * cumsum(package_count) / total_package_count, 2))
 
-```{r outer_alternative_count, eval=TRUE, echo=FALSE}
-outer_alternative_distribution <-
+datatable(function_frequency)
+
+function_frequency %>%
+ggplot(aes(function_count, package_count)) +
+geom_col() +
+scale_y_continuous(
+    "Packages", 
+    sec.axis = sec_axis(~ 100 * . / total_package_count, name = "Packages %")
+  )
+```
+
+## Parameter Frequency
+
+```{r parameter-frequency, eval=TRUE, echo=FALSE}
+function_parameter_count <-
     type_analysis_data %>%
     group_by(package_name, function_name) %>%
-   summarize(outer_alternative_count = length(unique(outer_alternative))) %>%
-   ungroup() %>%
-   group_by(outer_alternative_count) %>%
-   summarize(function_count = n(), package_count = length(unique(package_name))) %>%
+    ## subtract one for return position (-1)
+    summarize(parameter_count = length(unique(parameter_position)) - 1) %>%
     ungroup()
 
-datatable(outer_alternative_distribution)
+datatable(function_parameter_count)
+
+parameter_frequency <-
+function_parameter_count %>%
+#mutate(parameter_count = as.integer(5 * ceiling(parameter_count / 5))) %>%
+group_by(parameter_count) %>%
+summarize(function_count = n()) %>%
+ungroup() %>%
+mutate(function_perc = round(100 * function_count / total_function_count, 2)) %>%
+mutate(function_cumperc = round(100 * cumsum(function_count) / total_function_count, 2))
+
+datatable(parameter_frequency)
+
+parameter_frequency %>%
+ggplot(aes(parameter_count, function_count)) +
+geom_col() +
+scale_y_continuous(
+    "Function", 
+    sec.axis = sec_axis(~ 100 * . / total_function_count, name = "Function %")
+  )
 ```
 
-# Inner Alternatives
+# Monomorphism
 
-## Parameters
+```{r monomorphism, type-data, eval=TRUE,echo=FALSE}
+monomorphic_data <-
+type_analysis_data %>%
+group_by(package_name, function_name, parameter_position) %>%
+summarize(monomorphic = n() == 1) %>%
+ungroup()
 
-Now, we look at the number of alternatives in the function argument type.
+monomorphic_function_data <-
+monomorphic_data %>%
+group_by(package_name, function_name) %>%
+summarize(monomorphic = all(monomorphic)) %>%
+ungroup()
 
-```{r argument_inner_alternative_count, eval=TRUE, echo=FALSE}
-argument_inner_alternative_count <-
-   type_analysis_data %>%
-   dplyr::filter(parameter_position != -1L) %>%
-   group_by(package_name, function_name, parameter_position) %>%
-   summarize(inner_alternative_count = length(unique(inner_alternative))) %>%
+datatable(monomorphic_function_data)
+
+monomorphic_function_count <- sum(monomorphic_function_data$monomorphic)
+polymorphic_function_count <- sum(!monomorphic_function_data$monomorphic)
+
+monomorphic_package_data <-
+monomorphic_function_data %>%
+group_by(package_name) %>%
+summarize(monomorphism = round(100 * as.double(sum(monomorphic)) / length(monomorphic)),
+          monomorphic = all(monomorphic)) %>%
 ungroup() %>%
-   arrange(desc(inner_alternative_count))
+arrange(desc(monomorphism))
 
-datatable(argument_inner_alternative_count)
+datatable(monomorphic_package_data)
+
+monomorphic_package_count <- sum(monomorphic_package_data$monomorphic)
+polymorphic_package_count <- sum(!monomorphic_package_data$monomorphic)
+
+monomorphism_package_data <-
+monomorphic_package_data %>%
+group_by(monomorphism) %>%
+summarize(package_count = n()) %>%
+ungroup() %>%
+arrange(desc(monomorphism)) %>%
+mutate(package_cumsum = round(100 * cumsum(package_count) / total_package_count, 2))
+
+datatable(monomorphism_package_data)
+
+monomorphism_package_data %>%
+ggplot(aes(monomorphism, package_count)) +
+geom_col() +
+scale_y_continuous(
+    "Package", 
+    sec.axis = sec_axis(~ 100 * . / total_package_count, name = "Package %")
+  )
 ```
 
-## Return
+There are `r monomorphic_function_count` (`r round(100 * monomorphic_function_count/total_function_count, 2)`%) monomorphic functions.
+There are `r monomorphic_package_count` (`r round(100 * monomorphic_package_count/total_package_count, 2)`%) monomorphic packages.
 
-Now, we look at the number of alternatives in the function return type.
+There are `r polymorphic_function_count` (`r round(100 * polymorphic_function_count/total_function_count, 2)`%) polymorphic functions.
+There are `r polymorphic_package_count` (`r round(100 * polymorphic_package_count/total_package_count, 2)`%) polymorphic packages.
 
-```{r result_inner_alternative_count, eval=TRUE, echo=FALSE}
-result_inner_alternative_count <-
+# Type Distribution
+
+```{r type-distribution, type-data, eval=TRUE, echo=FALSE}
+type_data <-
 type_analysis_data %>%
-   dplyr::filter(parameter_position == -1L) %>%
-   group_by(package_name, function_name) %>%
-   summarize(inner_alternative_count = length(unique(inner_alternative))) %>%
+select(kind, type) %>%
+group_by(kind, type) %>%
+summarize(type_count = n()) %>%
+mutate(kind_count = sum(type_count)) %>%
 ungroup() %>%
-   arrange(desc(inner_alternative_count))
+arrange(desc(kind_count, type_count)) %>%
+mutate(kind_perc = round(100 * kind_count / sum(kind_count), 2),
+       type_perc = round(100 * type_count / sum(type_count), 2)) %>%
+mutate(kind_cumperc = round(100 * cumsum(kind_count) / sum(kind_count), 2),
+       type_cumperc = round(100 * cumsum(type_count) / sum(type_count), 2))
 
-result_inner_alternative_count %>%
-dplyr::filter(inner_alternative_count != 1) %>%
-datatable(caption="Number of union alternatives in return type")
+datatable(type_data)
 
-result_inner_alternative_count_graph_data <-
-result_inner_alternative_count %>%
-group_by(inner_alternative_count) %>%
-summarize(function_count = n()) %>%
+kind_count <-
+type_analysis_data %>%
+select(kind) %>%
+group_by(kind) %>%
+summarize(kind_count = n()) %>%
+ungroup() %>%
+arrange(desc(kind_count)) %>%
+mutate(kind_perc = round(100 * kind_count / sum(kind_count), 2)) %>%
+mutate(kind_cumperc = round(100 * cumsum(kind_count) / sum(kind_count), 2))
+
+datatable(kind_count)
+
+type_count <-
+type_analysis_data %>%
+select(type) %>%
+group_by(type) %>%
+summarize(type_count = n()) %>%
+ungroup() %>%
+arrange(desc(type_count)) %>%
+mutate(type_perc = round(100 * type_count / sum(type_count), 2)) %>%
+mutate(type_cumperc = round(100 * cumsum(type_count) / sum(type_count), 2))
+
+datatable(type_count)
+```
+
+## Scalar Types
+
+```{r scalar-type, eval=TRUE, echo=FALSE}
+scalar_data <-
+type_analysis_data %>%
+filter(kind == "scalar") %>%
+group_by(type) %>%
+summarize(count = n()) %>%
+ungroup() %>%
+arrange(desc(count)) %>%
+mutate(percentage = round(100 * count / sum(count), 2),
+       cumulative_percentage = round(100 * cumsum(count) / sum(count), 2))
+
+datatable(scalar_data)
+```
+
+## Scalar-na Types
+
+```{r scalar-na-type, eval=TRUE, echo=FALSE}
+scalar_na_data <-
+type_analysis_data %>%
+filter(kind == "scalar-na") %>%
+group_by(type) %>%
+summarize(count = n()) %>%
+ungroup() %>%
+arrange(desc(count)) %>%
+mutate(percentage = round(100 * count / sum(count), 2),
+       cumulative_percentage = round(100 * cumsum(count) / sum(count), 2))
+
+datatable(scalar_na_data)
+```
+
+## Vector Types
+
+```{r vector-type, eval=TRUE, echo=FALSE}
+vector_data <-
+type_analysis_data %>%
+filter(kind == "vector") %>%
+group_by(type) %>%
+summarize(count = n()) %>%
+ungroup() %>%
+arrange(desc(count)) %>%
+mutate(percentage = round(100 * count / sum(count), 2),
+       cumulative_percentage = round(100 * cumsum(count) / sum(count), 2))
+
+datatable(vector_data)
+```
+
+## Vector-na Types
+
+```{r vector-na-type, eval=TRUE, echo=FALSE}
+vector_na_data <-
+type_analysis_data %>%
+filter(kind == "vector-na") %>%
+group_by(type) %>%
+summarize(count = n()) %>%
+ungroup() %>%
+arrange(desc(count)) %>%
+mutate(percentage = round(100 * count / sum(count), 2),
+       cumulative_percentage = round(100 * cumsum(count) / sum(count), 2))
+
+datatable(vector_na_data)
+```
+
+## Class Types
+
+```{r class-type, eval=TRUE, echo=FALSE}
+class_data <-
+type_analysis_data %>%
+filter(kind == "class") %>%
+group_by(type) %>%
+summarize(count = n()) %>%
+ungroup() %>%
+arrange(desc(count)) %>%
+mutate(percentage = round(100 * count / sum(count), 2),
+       cumulative_percentage = round(100 * cumsum(count) / sum(count), 2))
+
+datatable(class_data)
+```
+
+## List Types
+
+```{r list-type, eval=TRUE, echo=FALSE}
+list_data <-
+type_analysis_data %>%
+filter(kind == "list") %>%
+group_by(type) %>%
+summarize(count = n()) %>%
+ungroup() %>%
+arrange(desc(count)) %>%
+mutate(percentage = round(100 * count / sum(count), 2),
+       cumulative_percentage = round(100 * cumsum(count) / sum(count), 2))
+
+datatable(list_data)
+```
+
+## Struct Types
+
+```{r struct-type, eval=TRUE, echo=FALSE}
+struct_data <-
+type_analysis_data %>%
+filter(kind == "struct") %>%
+group_by(type) %>%
+summarize(count = n()) %>%
+ungroup() %>%
+arrange(desc(count)) %>%
+mutate(percentage = round(100 * count / sum(count), 2),
+       cumulative_percentage = round(100 * cumsum(count) / sum(count), 2))
+
+datatable(struct_data)
+```
+
+## Other Types
+
+```{r other-type, eval=TRUE, echo=FALSE}
+other_data <-
+type_analysis_data %>%
+filter(kind == "other") %>%
+group_by(type) %>%
+summarize(count = n()) %>%
+ungroup() %>%
+arrange(desc(count)) %>%
+mutate(percentage = round(100 * count / sum(count), 2),
+       cumulative_percentage = round(100 * cumsum(count) / sum(count), 2))
+
+datatable(other_data)
+```
+
+## Union Types
+
+```{r union-type, eval=TRUE, echo=FALSE}
+union_data <-
+type_analysis_data %>%
+group_by(package_name, function_name, parameter_position) %>%
+mutate(alternative_count = length(paste0(outer_alternative, "-", inner_alternative))) %>%
+ungroup()
+# %>%
+#filter(alternative_count != 1)
+
+union_frequency <-
+union_data %>%
+distinct(package_name, function_name, parameter_position, alternative_count) %>%
+group_by(alternative_count) %>%
+summarize(parameter_count = n()) %>%
+ungroup() %>%
+arrange(desc(parameter_count)) %>%
+mutate(percentage = round(100 * parameter_count / sum(parameter_count), 2),
+       cumulative_percentage = round(100 * cumsum(parameter_count) / sum(parameter_count), 2))
+
+datatable(union_frequency)
+
+union_types <-
+union_data %>%
+group_by(package_name, function_name, parameter_position, alternative_count) %>%
+summarize(type = paste0(sort(type), collapse=" | ")) %>%
+ungroup() %>%
+group_by(alternative_count, type) %>%
+summarize(count = n()) %>%
+ungroup() %>%
+group_by(alternative_count) %>%
+arrange(desc(count)) %>%
+ungroup() %>%
+group_by(alternative_count) %>%
+mutate(percentage = round(100 * count / sum(count), 2),
+       cumsum_percentage = round(100 * cumsum(count) / sum(count), 2)) %>%
+ungroup() %>%
+group_by(alternative_count) %>%
+slice(1:20) %>%
 ungroup()
 
-result_inner_alternative_count %>%
-mutate(inner_alternative_count = ceiling(inner_alternative_count / 5) * 5) %>%
-group_by(inner_alternative_count) %>%
-summarize(function_count = n(), package_count = length(unique(package_name))) %>%
-ungroup() %>%
-mutate(function_percentage = to_percentage(function_count / total_function_count),
-       package_percentage = to_percentage(package_count / total_package_count)) %>%
-arrange(desc(inner_alternative_count)) %>%
-mutate(inner_alternative_count = str_c(inner_alternative_count - 4, 
-                                       "-", 
-                                       inner_alternative_count)) %>%
-select(package_count, package_percentage, function_count, function_percentage, inner_alternative_count) %>%
-datatable(caption="Distribution of return type alternatives")
-
-explicit_cases <-
-result_inner_alternative_count_graph_data %>%
-filter(inner_alternative_count <= 5)
-
-implicit_cases <-
-result_inner_alternative_count_graph_data %>%
-filter(inner_alternative_count > 5) %>%
-summarize(function_count = length(function_count),
-          inner_alternative_count = "> 5")
-
-result_inner_alternative_count_graph_data <-
-    explicit_cases %>%
-    mutate(inner_alternative_count = as.character(inner_alternative_count)) %>%
-    rbind(implicit_cases)
-
-result_inner_alternative_count_graph_data %>%
-ggplot(aes(inner_alternative_count, function_count)) +
-geom_col() +
-scale_x_discrete(limits = c(1:5, "> 5")) #%>%
-#coord_flip()
-
+datatable(union_types)
 ```

### 2020-05-13 paper/main.tex
data.frames

@@
 \caption{Struct example from the \code{neuralnet} package.}
-\label{fig:general-validation-example}
+\label{fig:struct-ex}
 \end{center}
@@
 %
-\subsubsection{Forgoing Data Frames}
-
-Speaking of classes, an important classful value in R is the data frame.
-Data frames underpin nearly all idiomatic use of the language, and even
-popular \code{tibble} (from the \code{tidyverse} ecosystem) and
-\code{data.table} (from the \code{data.table} package) build on the
-essential type.  We entertained the possibility of including data frame
-types, but \AT{found them to be clunky? Question mark?}  Consider:
-\AT{example of why data.frame types would be bad atm.}
-
-Data frames do have a type, through the \code{class<data.frame>} type, and
-future work will expand on this type, discussed further in
-Section~\ref{sec:futurework}.
+\subsubsection{Data Frames}
+
+One of the most popular classes in R is the \code{data.frame} class \AT{we have numbers here, grab em}.
+Data frames and the derivative \code{tibble} (from the \code{tidyverse} ecosystem) and \code{data.table} (from the \code{data.table} package) types underpin a huge amount of data analysis in R.
+
+One way to deal with data frames is through the struct type, with a named field for each column of the data frame, but as mentioned previously structs introduced undue noise into \typetracer's analysis results.
+Further complicated issues for data frames is that many functions built to operate on them operate in a name-agnostic way.
+For instance, the \code{tidyverse} package ecosystem allows programmers to pass column names to functions which operate on their data frames.
+In base R, idiomatic data frame use is to use string column names to select rows from the frame (unless only a single column is of interest, wherein the \code{$} syntax is appropriate). 
+We include some examples in Figure~\ref{fig:data-frames-bad} to illustrate the complexities.
+
+\begin{figure}[htbp]
+\begin{center}
+
+\begin{lstlisting}
+# Let's put some examples showing that data frames can get out of hand, and that
+# structs are inadequate.
+\end{lstlisting}
+
+\caption{Select examples of idiomatic \code{data.frame} use.}
+\label{fig:data-frames-bad}
+\end{center}
+\end{figure}
+
+Doing justice to this type will be an endeavour in and of itself, requiring reconciliation of base R's \code{data.frame}, \code{tidyverse}'s \code{tibble}, and \code{data.table}'s \code{data.table}.
+\AT{We could talk about how metaprogramming is pervasive in dplyr, and how we would need to capture that, too, which is a huge undertaking.}
+It will also require proper treatment of function types, and perhaps even parametric types, which dynamic analyses such as \typetracer are unable to discern soundly.
+\AT{General words to convince the reader that data frames are, in fact, hard, and worthy of their own paper.}
+
+In the context of our type annotation language, we will ascribe the type \code{class<data.frame>} to \code{data.frame}-typed values.
+A more in-depth discussion of class types follows.
 

### 2020-05-13 paper/main.tex
class discussion revisit

@@
 
-As discussed in Section~\ref{sec:R}, the R runtime has more than one notion
-of type.  The {\it class} (rather, the {\it classes}, as values can have
-multiple!) of function arguments are used by R to {\it dispatch} the correct
-function call for the argument \AT{$\leftarrow$ wording}.  This is best
-served with an example: \AT{Example, unless we talked about this earlier, in
-  which case refer reader back.}
-
-%
-%
-\subsubsection{Class Types}
-
-The class of a value suggest a shape for the value, and we felt class to be a more succinct way of capturing struct-like types.
-
-\AT{We used classes to stand-in for structs.  The discussion here is
-  naturally tied with the discussion of structs, so we need to make sure to
-  divide the discussion appropriately.}
+As discussed in Section~\ref{sec:R}, the R runtime has more than one notion of type.  
+The classes of function arguments are used by R as part of its dispatch mechanism, the most prevalent of which (\AT{we have that number, too!}) is S3 dispatch, in which R dispatches on the (S3) class of the first argument to a function, selecting the appropriate method to call for the class.
+To quickly illustrate, consider the code snippet in Figure~\ref{fig:dispatch-ex}.
+
+\begin{figure}[htbp]
+\begin{center}
+
+\begin{lstlisting}
+# Quick example of S3 dispatch.
+\end{lstlisting}
+
+\caption{S3 dispatch in R.}
+\label{fig:dispatch-ex}
+\end{center}
+\end{figure}
+
+Classes are very prevalent in R \AT{number}, and so we include a \code{class<class-name>} ... \AT{todo}
+\typetracer was outfit to report the class of values, but some care needed to be taken so as to not overreport class types, as {\it all} values have a class in R.
+The main classes we were interested in were matrix, array (higher dimensional matrices), and data.frames, in addition to any user-defined classes.
+In many cases \AT{number?}, values with a user-defined class are in reality lists with named elements, or structs.
+Including class types, and having \typetracer report them in place of structs, significantly reduced the noise in the analysis and allowed us to gather more focused data.
+To get a sense for this difference, consider the code in Figure~\ref{fig:class-vs-struct}.
+
+\begin{figure}[htbp]
+\begin{center}
+
+\begin{lstlisting}
+# Example of gross struct that turns into class.
+\end{lstlisting}
+
+\caption{S3 dispatch in R.}
+\label{fig:class-vs-struct}
+\end{center}
+\end{figure}
+
+When set to ignore classes, \typetracer reports \AT{type} as the type of the function, as compared with \AT{other type} when configured to report classes.
 

### 2020-05-13 paper/main.tex
S4 revisited, comments about parametric to be elaborated

@@
 and so on.  In this section, we will present the final type annotation language design, 
-which we believe captures the essence of R.  Throughout, we will
+which we believe captures the essence of R.  We will end the section with a formal type 
+language grammar, and leading up to that we will
 touch on design decisions and weigh our choices against alternatives,
@@
 
-As R does, our type system supports multiple classes on values, and the type
-of a classful value is simply a list of all of its classes.
-
-\AT{What more to say here? Discussing occurrences of multi-classes?}
+A value can have multiple classes in R.
+This is common practice in the \code{dplyr} package, which extends the functionality of data frames: a \code{data.frame} that passes through \code{dplyr} functions gains two additional classes (\code{tbl} and \code{tbl_df}), bringing the total to at least 3, in addition to whichever other classes that data frame happened to have.
+Note that when looking to perform S3 dispatch, R will try to match {\it any} of the classes of the first argument to a dispatched function, so each of these classes are important.
+
+We found these so-called multi-classes to be surprisingly common, making up \AT{occurrence} of the occurrences of class-typed values.
+One possible reason for this is that values can gain classes quite easily (e.g. the \code{dplyr} example discussed above), somewhat polluting the output of \typetracer.
+Nevertheless, we found \AT{number} of occurrences of monomorphic multi-class arguments, meaning that the function was always called with a multi-class value in that argument position.
+
+In the end, we support these multi-class types by allowing \code{class<...>} types to contain a list of class names, so for instance a \code{dplyr} \code{tibble} would have type \code{class<data.frame, tbl, tbl_df>} in our type language, and when performing related contract assertions \contractr will check that the value has every class that make up its type.
 
@@
 dispatch mechanism, and indeed our class types roughly capture the S3
-``types'' of values.  We will not discuss R5 as it is still under
+classes of values.  We will not discuss R5 as it is still under
 development.
 
-For S4, we have an \code{s4} type to denote S4 values, though we did not
-elaborate on this type.  We stuck to simple S4 types since doing anything
-more would be immensely complicated, and outside of the scope of this work.
-For instance, the mechanics of S4 dispatch are more complex than for S3 (S4
-dispatches on the class of {\it all} arguments), and users can define their
-own class hierarchies that we would need to incorporate in our type analysis
-and contract checking frameworks.  Further, \AT{we found limited use of S4
-  during our analysis}, with S4 types being heavily used in a select subset
-of packages (\AT{such as ...}), some of which make heavy use of R's
-metaprogramming capabilities.  Coming up with a type system that accounts
-for all of these factors, and consolidates multiple object-orientation
-frameworks, is an interesting problem in and of itself, and we aim to extend
-our type system to account for this in future work.  For now, S3 types
-suffice: S3 is the most common object system in R, with no sophisticated
-hierarchies, and its semantics are simple enough to be captured by our
-proposed design.
+For S4, we have an \code{s4} type to denote S4 values, though we did not elaborate on this type, and in practice S4 values have a class, and thus a \code{class<...>} type.  
+We stuck to simple class-based types for S4 since doing anything more would be immensely complicated, and outside of the scope of this work.
+For instance, the mechanics of S4 dispatch are more complex than for S3 (S4 dispatches on the class of {\it all} arguments), and users can define their own class hierarchies that we would need to incorporate in our type analysis and contract checking frameworks.
+Further, \AT{we found limited use of S4 during our analysis}, with S4 types being heavily used in a select subset of packages (\AT{such as ...}), some of which make heavy use of R's metaprogramming capabilities.
+Coming up with a type system that accounts for all of these factors and consolidates multiple object-orientation frameworks in a single language design is an interesting problem in and of itself, and we aim to extend our type annotation framework to account for this in future work.
+For now, S3 types suffice: S3 is the most common object system in R, with no sophisticated hierarchies, and its semantics are simple enough to be captured by our proposed design.
+
+%
+%
+%
+%
+\subsection{Parametric Types and Higher-Order Function Types}
+
+\AT{Ok, high level:
+We don't do higher-order function types, beyond any to any.
+Typetracer detected them once, but the data was way too noisy.
+We found that mostly these correlated with functions where parametric polymorphism would be suitable, though there's no way to say that from the data, and only manual inspection revealed it.
+We could implement some heuristic parametric polymorphism detector, but that would be way unsound.
+Ultimately, we felt that higher-order function types and parametric polymorphism were related, and they didn't occur frequently enough across all of R.
+Some packages are heavy users, but those packages are heavily metaprogrammed and effectively implement DSLs.
+We will revisit this issue, but it's not something that we can pick up with \typetracer as it is, and doing justice to these types will require significant effort.}
+
+In R, functions are first-class values, and they can be passed as function arguments.
 

### 2020-05-13 paper/main.tex
parametric and higher order fun initial discussion

@@
 %
-\subsection{Parametric Types and Higher-Order Function Types}
-
-\AT{Ok, high level:
-We don't do higher-order function types, beyond any to any.
-Typetracer detected them once, but the data was way too noisy.
-We found that mostly these correlated with functions where parametric polymorphism would be suitable, though there's no way to say that from the data, and only manual inspection revealed it.
-We could implement some heuristic parametric polymorphism detector, but that would be way unsound.
-Ultimately, we felt that higher-order function types and parametric polymorphism were related, and they didn't occur frequently enough across all of R.
-Some packages are heavy users, but those packages are heavily metaprogrammed and effectively implement DSLs.
-We will revisit this issue, but it's not something that we can pick up with \typetracer as it is, and doing justice to these types will require significant effort.}
-
-In R, functions are first-class values, and they can be passed as function arguments.
+\subsection{Parametric and Higher-Order Function Types}
+
+Functions are first-class values in R.
+
+\typetracer found \AT{number} functions to be higher-order, having at least one function type among their arguments.
+They are markedly more common in a subset of packages, such as \AT{such as...}.
+By-and-large, these higher-order functions are similar to well-known higher-order functions like map, filter, and fold.
+
+The truth is that higher-order function types are best served by {\it parametric types} and parametric polymorphism, wherein type variables are used within a type declaration to specify type-based relationships between function arguments.
+Unfortunately, the nature of dynamic analysis is such that \typetracer cannot soundly detect parametric polymorphism, as without a static analysis of the function body it is impossible to say if function execution will halt given certain argument types, and we saw some examples of this in Figures~\ref{fig:scalar-vector-example}~and~\ref{fig:matrix-example}.
+Without parametric types, higher-order function types become noisy and fail to tell the complete picture.
+
+Ultimately, we configured \typetracer to report only top-level function types (i.e., detect if a value is a function or not).
+\AT{The data suggests that} most of the use of higher-order functions is through map-, filter-, and apply-like functions, whose behaviour is best captured with parametric polymorphism.
+Other common uses include higher-order functions operating on data frame columns (in packages such as \code{dplyr}), which would necessitate a more in-depth treatment of data frame types, which as we discussed is a huge endeavour in and of itself.
+We aim to extend our type annotation framework with parametric, higher-order, and detailed data frame types together in future work.
+
+%
+%
+%
+%
+\subsection{Nullable Types}
 

### 2020-05-13 paper/main.tex
environments, nullable, and started final synthesis/discussion of type grammar

@@
 %
+\subsubsection{Environments}
+
+Environments in R are essentially lists with reference semantics: mutating a value in an environment is performed in-place, and that mutation can thus be shared with other code holding a reference to the environment.
+They are used primarily by the R runtime to store variables (their primary function is that of an execution environment), though they also often used by programmers either to themselves reflect on local, previous, or global scope, or simply because of their reference semantics.
+Environments are comprised of name, value pairs, and each environment has a reference to its parent environment.
+
+We thought to include detailed environment types in our type language by way of allowing environment types to define names and types of values in the environment.
+Unfortunately, this led to an undue explosion in the ``size'' of our types: 
+often, environments had hundreds of values, particularly those which were obtained from R (by e.g. passing a reference to a local scope).
+These types were difficult to parse visually and did not feel informative.
+Further, environments \AT{did not occur often}, and giving them a richer type was not worth the increase in the visual complexity of our types.
+
+%
+%
 \subsubsection{Forgoing Tuples, or Short, Constant Length Lists}
@@
 \subsubsection{Structs}
+
+\AT{Do we want to talk about how list names are not unique?}
 
@@
 
+One final, common pattern that we would like to highlight is the co-occurrence of \code{NULL} with other types of values.
+In Section~\ref{subsec:basictypes} we discussed NA, and how NAs represent missing data for the primitive types, and more generally R provides a \code{NULL} value that programmers use to represent missing data for other types.
+Programmers often use \code{NULL} as a default value for function arguments, and the code snippet in Figure~\ref{fig:null} exemplifies this.
+
+\begin{figure}[htbp]
+\begin{center}
+
+\begin{lstlisting}
+# NULL default argument example. 
+\end{lstlisting}
+
+\caption{Idiomatic NULL usage in R.}
+\label{fig:null}
+\end{center}
+\end{figure}
+
+To capture this behaviour, we introduce a nullable type \code{? T} into our type language and contract checking framework.
+That way, programmers can succinctly ascribe types to null-by-default arguments (such as in the example in Figure~\ref{fig:null}).
+This also served to reduce the number of polymorphic arguments we detected:
+Without this type, \AT{X\%} of polymorphic argument positions were of the form \code{null | T}.
+Including a nullable type served to reduce observed polymorphism by \AT{Y\%}.
+
 %
@@
 \subsection{Final Type Grammar}
+
+Finally, Figure~\ref{fig:types} presents our final type language design, and the grammar of types that \contractr is equipped to handle.
+\contractr will look for a top-level type declaration file in a package, with function types having the form $type \; ID \; T$, where $ID$ is the function name and $T$ is a function type.
+These function types have the form $\langle A_1 , ..., A_n\rangle \Rightarrow T$, where each of the $A_i$ arguments are either some type $T_i$, or the vararg type $...$.
+Recall that functions with a vararg parameter (\code{...}) can be called with arbitrarily many values, and extra values are lumped together in the vararg $...$ value.
+Internally, the type $...$ is implemented as a ${\bf list} \langle any \rangle$ type.
+
+We did not discuss ${\bf expression}$, ${\bf language}$, ${\bf symbol}$, ${\bf externalptr}$, ${\bf pairlist}$, ${\bf S4}$, and ${\bf weakref}$ types throughout this section.
+They occur \AT{rarely}, and are by-and-large uninteresting to most R programmers.
+\AT{Fill in...}
+
 \label{subsec:typegrammar}
@@
       & |   & T_1 \; {\bf |} \; T_2 & \text{\it union type} \\
+      & |   & ? \; T & \text{\it nullable type} \\
       & |   & \langle A_1 , ..., A_n\rangle \Rightarrow T & \text{\it function type} \\
@@
     \end{minipage}
-    \caption{The R type system.}
+    \caption{The R type annotation language.}
     \label{fig:types}

### 2020-05-13 paper/main.tex
final type grammar

@@
 \subsubsection{Structs}
+\label{subsec:structs}
 
@@
 
-We did not discuss ${\bf expression}$, ${\bf language}$, ${\bf symbol}$, ${\bf externalptr}$, ${\bf pairlist}$, ${\bf S4}$, and ${\bf weakref}$ types throughout this section.
-They occur \AT{rarely}, and are by-and-large uninteresting to most R programmers.
-\AT{Fill in...}
+We have discussed most of these types throughout this section, with some notable exceptions.
+${\bf expression}$, ${\bf language}$, ${\bf symbol}$, ${\bf externalptr}$, ${\bf pairlist}$, ${\bf S4}$, and ${\bf weakref}$ types occur \AT{rarely}, and are by-and-large uninteresting to most R programmers.
+\AT{Fill in details of these basic types.}
+% $S$ and $V$ are scalar and vector primitives, with vectors taking the form $S[]$.
+% The scalar type $S$ can be either a base type $B$ (one of the 6 primitives in R), or an NA-able base type (roughly, $\string^B$).
+One other type we did not discuss is the union type $T_1 \, | \, T_2$, which is supported by \contractr and generated by \typetracer when it cannot find some way to consolidate argument types (for example, if an argument is observed to be ${\bf chr}$ and ${\bf int}$, \typetracer will ascribe type ${\bf chr} \, | \, {\bf int}$ to it).
+And while not generated by \typetracer (recall discussion in Section~\ref{subsec:structs}), user-defined struct types have the form ${\bf struct}\langle F_1, ..., F_n \rangle$, with fields $F_i$ being name $:$ type pairs.
 

### 2020-05-13 paper/main.tex

@@
 
-We contended with multiple type language designs, and in this section we will explore in more detail how we compared our candidate designs.
-We will also discuss a large-scale evaluation of our final set of type annotations.
-
-%
-%
-%
-%
-\subsection{Comparative Analyses}
-
-\AT{We compared different designs: with/out structs and tuples, with/out classes, with/out primitive subtyping. We looked at type sizes and determined that e.g. \code{? T} is a good type.}
+To evaluate the effectiveness of our type annotation language and contract checking framework, we conducted a large-scale experiment based on the packages we used to inform our language design.
+The core corpus of 412 packages represents some of the most widely used and thoroughly tested R libraries on CRAN.
+To conduct the evaluation in this section, we configured \typetracer to infer the types of each of the functions that are visible to package clients.
+\typetracer inferred these types based on the package text, example, and vignette code.
+This section is divided in two parts:
+First, we present general information about the generated types, and about the core analysis corpus, and next we discuss an evaluation of the inferred types against the reverse-dependencies (read: clients) of the corpus we analyzed.
+
+%
+%
+%
+%
+\subsection{Language Statistics}
+
+\AT{Let's talk about some of the data related to our types themselves.
+How many types?
+How big are they?
+How many are mono? Poly?
+Stuff like that.}
 
@@
 
+%
+%
+%
+%
 \subsection{Future Work}
+
+%
+%
+\subsubsection{Object Orientation in R}
+
+%
+%
+\subsubsection{Types for Data Frames}
+
+% ======================================
+%
+% End
+%
+%
 

### 2020-05-13 paper/main.tex
reverse dependency eval disc

--- 2020-05-13 paper/main.tex
@@
 
+\AT{Revisit this section since discussion of polymorphism in results.}
+
 Functions are first-class values in R.
-
 \typetracer found \AT{number} functions to be higher-order, having at least one function type among their arguments.
@@
 
-We contended with multiple type language designs, and in this section we will explore in more detail how we compared our candidate designs.
-We will also discuss a large-scale evaluation of our final set of type annotations.
-
-%
-%
-%
-%
-\subsection{Comparative Analyses}
-
-\AT{We compared different designs: with/out structs and tuples, with/out classes, with/out primitive subtyping. We looked at type sizes and determined that e.g. \code{? T} is a good type.}
-
-%
-%
-\subsubsection{Analysis of Polymorphism}
-
-\AT{How many arguments, functions are polymorphic? How many packages? Are there pathologically polymorphic packages? Why didn't you handle those?}
+To evaluate the effectiveness of our type annotation language and contract checking framework, we conducted a large-scale experiment based on the corpus of 412 packages we used to inform our language design.
+This corpus represents some of the most widely used and thoroughly tested R libraries on CRAN.
+
+We configured \typetracer to infer the types of each of the functions that are visible to package clients.
+Recall that \typetracer's primary function is to collect call signatures from observed function calls, and we ran \typetracer on the test, example, and vignette code of all packages in the aforementioned corpus.
+These call traces were consolidated into a single function type for each visible package function:
+\typetracer employs unification strategies discussed in Section~\ref{sec:typesystemdesign} to keep the size of signatures in check and only generates an \code{any} type when missing and unused arguments were recored, and when a longjump caused function execution to abort prematurely.
+The types generated by \typetracer will be analyzed in more depth in Section~\ref{subsec:stats}, and a large-scale evaluation of the quality of these types will be discussed in Section~\ref{subsec:eval}.
+
+%
+%
+%
+%
+\subsection{Language Statistics}
+\label{subsec:stats}
+
+We analyzed the type signatures generated by \typetracer in order to build up a profile of our type annotation language.
+We were interested in broad, high-level information, such as the distribution of generated types, the level of polymorphism observed in the types, and the distribution of sizes of union types.
+
+%
+%
+\subsubsection{Type Distribution}
+
+%
+%
+\subsubsection{Polymorphism}
+
+%
+%
+\subsubsection{Size of Union Types}
 
@@
 
-\AT{When is any introduced? Why? How does it affect the results? What if missing/unused isn't any, do the results change substantially?}
+%
+%
+\subsubsection{Select Examples}
 
@@
 \subsection{Large-Scale Evaluation}
-
-Now that we have settled on a language design, we endeavour to evaluate it against a representative subset of existing R code.
-To this end, we took the detailed type information we collected from \AT{corpus} popular R packages, inferred types for the functions exposed by those packages, and ran the test, example, and vignette code of each of their reverse dependencies.
-
-%
-%
-\subsubsection{Experiment Setup}
-
-\AT{What is the design of the experiment? The setup? How much code was ran? What does this evaluate? Why does it mean anything for the annotation framework?}
-
-\typetracer is capable of inferring type signatures based on the call traces it has collected.
-These call traces were consolidated into a single function type for each package function:
-\typetracer employs unification strategies discussed in Section~\ref{sec:typesystemdesign} to keep the size of signatures in check and only generates an \code{any} type under extenuating circumstances: when inner list types got too big (we decided to cap the union type in lists to 5), when missing and unused arguments were recored, and finally when a longjump caused function execution to abort prematurely.
-Finally, we loaded each of these inferred types into \contractr, and ran all example code from the reverse dependencies of these popular packages.
+\label{subsec:eval}
+
+To measure the effectiveness of our type annotation language, we designed an experiment pitting the types generated by a package's own testing code against the testing code of the package's clients.
+We collected failed contract assertions in order to analyze them and perhaps identify weaknesses in our type annotation language.
+
+%
+%
+\subsubsection{Experimental Design}
+
+As mentioned earlier in this section, we configured \typetracer to infer types for all client-visible functions for each of the 412 packages in our corpus.
+We loaded each of these inferred types into \contractr, and ran all example code from the reverse dependencies of these popular packages.
+\AT{As discussed in genthat}, running R code can be tricky, as runnable code may depend on e.g.
+data present in a hard-coded directory structure, and we employed similar strategies to ensure that we ran as much code as possible.
+\AT{Maybe Filip can write a few sentences on this, or write me a message and I can work it in.}
+
+\contractr was equipped to provide a list of all failed contract assertions, complete with diagnostic information (including a stack trace) to help us make sense of the failed assertion.
 
@@
 Overall, we found that only \AT{2\%} of contract assertions failed.
-\AT{What does this indicate? One thought:}
 This indicates two things: firstly, the corpus of packages we selected is quite well-tested, and much of the intended function behaviour is exemplified in the package example code. 
 Secondly, and more importantly, this success indicates that our analysis framework is able to generate function types which accurately capture the behaviour of the function. 
-\AT{It kind of feels like these need to simultaneously be true. Maybe we pull in the fact that coverage is good?}
-
 Of the \AT{2\%} of failing contract assertions, we picked up on a few significant patterns.
+
 Firstly, we found that \AT{10\%} of the failing assertions were due to \code{double}-typed values being passed to function arguments expecting an \code{integer}.
@@
 
+Of the remaining \~60\% of failed contract assertions, we found quite an interesting pattern: the top as-of-yet unexplained contract failure, making up roughly 7\% of total failed assertions, is an assertion checking a ${\bf character}$ against the expected type ${\bf class} \langle bignum \rangle \, | \, {\bf raw}[]$.
+
 Separately, we investigated the effect of running reverse dependency code on the code coverage of the packages in our corpus.
+Further investigation revealed that most failures were coming from the \code{openssl::base64_encode} function, whose code is available in Figure~\ref{fig:base64-encode}.
+
+\begin{figure}[htbp]
+\begin{center}
+
+\begin{lstlisting}
+# The type according to typetracer:
+#' @type <bin: class<`bignum`> | raw[], linebreaks: logical> => character
+base64_encode <- function (bin, linebreaks = FALSE) 
+{
+    if (is.character(bin)) {
+        bin <- charToRaw(paste(bin, collapse = ""))
+    }
+    stopifnot(is.raw(bin))
+    .Call(R_base64_encode, bin, as.logical(linebreaks))
+}
+\end{lstlisting}
+
+\caption{Example of reverse dependency increasing code coverage.}
+\label{fig:base64-encode}
+\end{center}
+\end{figure}
+
+Contract failures are on the first argument, \code{bin}.
+This function is a wrapper for a C function (\code{R_base64_encode}) which encodes the input \code{bin} in base64.
+It would appear that the code for this function was never tested with a ${\bf character}$-typed input, even though it {\it is} valid.
+This is an example of a rather common phenomenon wherein the code of a package's reverse dependencies increases the coverage of some of the original package's functions, exemplified here by the code on line 6 never having been run in \code{openssl}'s own tests.
+
 We found that \AT{in many cases}.

### 2020-05-14 paper/main.tex
some data in evaluation

@@
 
-We found that \AT{in many cases}.
-the reverse dependencies increased the code coverage (this aligns with findings in \AT{cite genthat}), and this was especially true for the functions for which contract assertions failed.
+To try to detect this pattern, we cross-referenced functions with failing contract checks against the code coverage for that function from the package's tests.
+For a given function, a coverage of X\% indicates that X\% of the code in the function body was executing by running package test, example, and vignette code.
+Of the remaining failed contracts, we found that \~28\% of them were from functions which had less than 100\% coverage, meaning that it is possible that \typetracer failed to synthesize the correct type for the function as not all possible executions were recorded.
 

### 2020-05-14 paper/main.tex
abstract

@@
 \begin{document}
-\title{Extracting Function Signature from R Program Traces}
+\title{Extracting Function Signatures from Program Traces}
+
+\newcommand{\NUMFUNCTIONS}{25,000\xspace}  %%% TODO auto-gen
+\newcommand{\NUMPACKAGES}{412\xspace}  %%% TODO auto-gen
+\newcommand{\YEARS}{20\xspace} %%TODO fix
 
 \begin{abstract}
-  This paper documents the first steps towards retrofitting a type system
-  onto the R programming language. We introduce an extension of R with
-  function signatures. To evaluate the fit of our design to the language, we
-  start by automatically extracting the signatures for 500 widely used
-  packages. In a second step we use the reverse dependencies of these
-  packages to validate that these signatures are accurate.
+  Retrofitting a type system onto an existing dynamic language which was not
+  designed to be statically typed is challenging. This paper takes the first
+  step in the direction of designing a type system for the R programming
+  language. We argue that R is particularly challenging due to its unique
+  combination of features and to the programming idioms that have developped
+  over the last two decades. As our starting point, we infer type signatures
+  for \NUMFUNCTION functions from \NUMPACKAGES packages amongst the most
+  widely used open source R libraries. The signatures use a minimal type
+  language consisting of basic types, vectors, classes and disjunctions. We
+  purposefully leave out notions such as subtyping, higher-order types, and
+  polymorphism. Our goal is to understand what benefits a simple lanugage
+  can yield.  We also implement a contract system that can use the inferred
+  types to validate their correctness. We use the reverse depedencies of the
+  packages to evaluate the benefits of typing.
 \end{abstract}

### 2020-05-14 paper/main.tex
merge

--- 2020-05-14 paper/main.tex
@@
 Of the remaining \~60\% of failed contract assertions, we found quite an interesting pattern: the top as-of-yet unexplained contract failure, making up roughly 7\% of total failed assertions, is an assertion checking a ${\bf character}$ against the expected type ${\bf class} \langle bignum \rangle \, | \, {\bf raw}[]$.
-
-Separately, we investigated the effect of running reverse dependency code on the code coverage of the packages in our corpus.
-Further investigation revealed that most failures were coming from the \code{openssl::base64_encode} function, whose code is available in Figure~\ref{fig:base64-encode}.
+Further investigation revealed that most of these failures were coming from the \code{openssl::base64_encode} function, whose code is available in Figure~\ref{fig:base64-encode}.
 
@@
 
+But what of the large amount (\~70\%) of assertion failures in functions with 100\% coverage?
+As it happens, many of these functions have very few lines of code, many having as few as a single line.
+The most interesting among these are functions which are the entry point for S3 dispatch (these functions typically have a single line of code, \code{UseMethod("InsertS3MethodNameHere")}).
+
+It turns out that the lion's share of the remaining polymorphism is related to S3 dispatch.
+S3 generics are naturally polymorphic, and we aim to define types to capture these cases, and implement a full-fledged object-oriented type system for R in future work.
+In the meantime, if we count out S3 methods, we find that only \AT{few} contract failures remain.
+
 %

### 2020-05-14 paper/main.tex
minor

@@
   over the last two decades. As our starting point, we infer type signatures
-  for \NUMFUNCTION functions from \NUMPACKAGES packages amongst the most
+  for \NUMFUNCTIONS functions from \NUMPACKAGES packages amongst the most
   widely used open source R libraries. The signatures use a minimal type
@@
 
-\AT{Leaving this one for Aviral.}
-
 R-dyntrace is an efficient dynamic analysis framework for R.
@@
 \label{sec:contractr}
-
-\AT{I think we should move this to the Methodology section, which we should maybe rename to implementation. Thoughts?}
 

### 2020-05-14 paper/main.tex
reorganizing

@@
 
+\AT{Leaving this one for Aviral.}
+
 R-dyntrace is an efficient dynamic analysis framework for R.
 
+%
+%
+%
+%
+\section{Project Corpus}\label{sec:corpus}
+
+For this paper we have selected \CorpusLoadable packages consisting of
+\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native
+code (C/C++/Fortran).  \FK{Add more details} Figure~\ref{fig:corpus} shows
+all the packages, the size of the dots reflects the size of the project in
+terms of both R and native lines of code\footnote{ Lines of source code
+  reported excludes comments and blank lines, counted by \emph{cloc}, \cf
+  \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the
+expression code coverage in percents and the y-axis gives the number of
+reverse dependencies in log scale. Dotted lines indicate means.  Packages
+with over \PackageSizeOutierRnd lines of code are annotated.
+
+\begin{figure}[!h]  \centering
+  \includegraphics[width=.8\linewidth]{plots/corpus.pdf}
+  \caption{Corpus overview}\label{fig:corpus}
+\end{figure}
+
+These packages come from the Comprehensive R Achive Network
+(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of
+R code with over \AllCranRnd packages\footnote{CRAN receives about 6 new
+  package submissions a day~\cite{Ligges2017}} containing over \AllRCodeRnd
+and \AllNativeCodeRnd R and native code respectively. Unlike other open code
+repositories like GitHub, CRAN is a curated repository. Each submitted
+package must abide to a number of well-formedness rules that are
+automatically checked asserting certain quality. Most relevant for this work
+is that all of the runnable code (including code snippets from examples and
+vignettes) is tested and only a successfully running package is admitted in
+the archive.
+
+We have downloaded and installed all available CRAN packages. Out of the
+\AllCranRnd packages, we managed to install \AllLoadableRnd. The main reason
+for this is that \rdt is based on R 3.5.0 and some of the packages are no
+longer compatible with it. Some packages also require extra native
+dependencies which were not present on our servers.  We have two criteria
+for including a package into the corpus:
+\begin{inparaenum}[(1)]
+\item have runnable code that covers significant part of the package source
+  code from which type signatures could be inferred, and
+\item have some reverse dependencies that allows us to evaluate the inferred
+types on runnable code from these dependent packages.
+\end{inparaenum}
+The concrete thresholds were set to at least \ThresholdCodeCoverage of
+expression coverage and at minimum \ThresholdRevdeps reverse dependencies.
+
+The code coverage was computed for each package using
+\covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R
+tool.  The reverse package dependencies were extracted from the package
+metadata using builtin function.
+
+%
+%
+%
+%
+%
+%
+\section{Methodology}
+
+In this section, we will describe our approach to data collection and analysis through \typetracer, a dynamic analysis built in the \rdt framework.
+We will then describe the process by which we distill the rich information we collect with \typetracer to more digestible ``types'', though we discuss the precise details of our type annotation language in Section~\ref{sec:types}.
+Finally, we discuss \contractr, a package built in R which supports our type language and can generate and check contracts for function types.
+
+%
+%
+%
+%
+\subsection{typetracer and Initial Analysis}
+
+To help inform the design of our type system, and to ensure that it aligns
+with the day-to-day usage of R, we performed a corpus analysis of some of
+the most widely used R packages.  The corpus is discussed in detail in
+Section~\ref{sec:corpus}.  We built \typetracer, our dynamic analysis, in
+the R-dyntrace~\cite{oopsla19} dynamic analysis framework which is itself
+built on a modified version of R targeting R version 3.5.0---we refer the
+reader to Section~\ref{sec:r-dyntrace} for more details on the framework.
+
+%
+%
+\subsubsection{Overview}
+The goal of our analysis is to discover argument types for functions,
+builtins, and specials.  Naively intercepting function calls and recording
+argument types is incorrect, as recall that R wraps arguments in promises
+(Section~\ref{sec:R}).  To circumvent this obstacle, we instead
+intercept the forcing of promises: when a promise is forced, we identify the
+surrounding call, collect the type of the value, and add it to the call
+trace associated with the current function call.  Once function execution
+completes, we collect information about the return value, add it to the call
+trace, and save the completed call trace to be written out later.
+%
+%
+\subsubsection{Type Information}
+Once a value is obtained, either on function return or when an argument
+promise is forced, we probe the value for the following information:
+
+\begin{itemize}
+\item the value's type tag according to R's runtime. R's runtime implements
+  type tags as a typedef, e.g. \code{INTSXP} for integers, and, confusingly,
+  \code{VECSXP} for lists;
+\item its class, which is a list of string class names. Typically, the class
+  of a value is held in an attribute with name ``class'', though some
+  classes are blessed by the R runtime, such as matrix and array classes;
+\item its attributes, a list of metadata attached to the value.
+\end{itemize}
+
+Now, depending on the value's type tag, we collect further information:
+
+\begin{itemize}
+\item For R's primitive types, we collect its dimension, and if it has a
+  names attribute we collect the names as well;
+\item For lists, we recursively collect the types of list elements;
+\item For lists {\it with a names attribute}, we collect the names as well as the types of the slots that those names refer to (data.frames fall into this category);
+\item For promises, we carefully dig into the promise value {\it without
+  forcing it}, and collect information about the innermost value. If the
+  innermost value is the unbound value, then the promise is an as-of-yet
+  unevaluated expression. In this case, we do our best to guess the value of
+  the unevaluated expression: if the expression is not a language
+  expression, we collect information about it, otherwise we can say nothing
+  without evaluating it and ascribe it type ``any''. \footnote{\AT{I think
+      this is confusing, reword.}}
+\end{itemize}
+
+For vectors (and lists), we additionally iterate through the elements,
+checking if the values are NA (resp. NULL); if none are NA (resp. NULL), we
+ascribe the NA-free (resp. NULL-free) tag to the value.  Arguments that are
+part of a function's varargs (denoted \ldots in function definitions) are
+tagged as such and ignored, as varargs can contain anything (and their type
+will be equivalent to and any-typed list).
+
+Now that we've seen the high-level picture of our approach, we will elaborate on some key details and discuss particular challenges that we faced.
+
+%
+%
+\subsubsection{Challenges}
+\label{subsec:typetracer-challenges}
+
+We grappled with a number of corner cases related to R's non-strict semantics.
+
+\begin{itemize}
+
+\item First, we wanted to ensure that \typetracer captured type for
+  unevaluated arguments.  Not every argument of a function needs to be used
+  during every function execution, and the approach we discussed above
+  relies on argument promises being forced.  There are a few reasons for this.
+
+One possibility is that the argument was simply unused given the control
+flow of this particular execution (e.g. if the argument is only needed in
+one branch of a conditional expression).  This mainly occurs when arguments have default
+values for some parameters that are not critical to every execution of the
+function.  In these cases, we still want to collect the type of the default
+value.
+
+Another possibility is that a value is used {\it without being forced}, as
+is the case when values are ``metaprogrammed''.
+This is common practice in e.g. the \code{tidyverse} ecosystem, as they use the metaprogramming capabilities of R to implement a DSL for data analysis.
+
+To tackle this, we make an {\it initial guess} of function argument types on
+function entry. This collects types for argument default values, and
+collects an ``unused'' type for unevaluated expressions.  If an argument
+promise is later forced, we simply update the recorded type for the
+argument.
+
+\item Second, we wanted to account for formal arguments (arguments which are
+  named in the function definition) that were {\it not} passed to the
+  function. This represents cases where a function was not called with all of its
+  arguments, and no default value was specified for missing arguments. 
+  We record a ``missing'' type for such argument.
+
+There are two obvious ways to deal with missing arguments: type them as bot
+or as top.  As we are performing a dynamic analysis based on package test
+code, we conservatively type these arguments as top, or \code{any}, as It is
+impossible to know dynamically what all possible function behaviours are.
+We discuss the prevalence of such any-typed arguments in Section~\ref{sec:evaluation}.
+
+\item Third, we had to contend with R's C implementation sometimes
+  performing longjumps, which discards current R calls from the call stack.
+  In order to ensure that call traces are not lost when a longjump occurs,
+  \typetracer intercepts the unwinding process initiated by the longjump and
+  mimics the functionality of the functions being exited.  The R-dyntrace
+  framework supplies the return value during the unwinding process (if there
+  is indeed such a value), so we can capture it, analyze it, and include it
+  in the call trace.  In certain situations, however, no return value can be
+  obtained (e.g. a longjump occurred as part of some error handling), and
+  then we simply record a ``jumped'' return value, which we can deal with in
+  post-processing.  
+  % These longjumps occur quite often, particularly when S3 dispatch occurs, as the R implementation optimizes returns from dispatched functions by initiating a longjump.
+
+\end{itemize}
+
+Having outlined these challenges, we will now explore the implementation of \typetracer in more detail.
+
+%
+%
+\subsubsection{Select Details}
+
+As mentioned, our analysis is built on the R-dyntrace framework.  R-dyntrace
+is a very efficient instrumentation framework for R, which is built on a
+modified version of the R runtime.  Our dynamic type analysis only needs to
+be run once, ever---from the analysis results, we synthesize function
+signatures which are passed along to later type-checking tools which do not
+rely on a heavily modified runtime.
+
+We primarily rely on 8 R-dyntrace callbacks: \texttt{closure\_entry},
+\texttt{closure\_exit}, \texttt{builtin\_entry}, \texttt{builtin\_exit},
+\texttt{special\_entry},\texttt{special\_exit},
+\texttt{promise\_force\_entry}, and \texttt{promise\_force\_exit}.  These
+callbacks fire when closures, builtin, and special functions are entered and
+exited, and before and after the forcing of promises.  The function-related
+callbacks are used mainly for bookkeeping: the analysis is notified that a
+closure, builtin, or special has been entered/exited by pushing/popping the
+call onto a stack.  The calls themselves store a {\it call trace} object,
+where we store the type information that we collect, populated with initial
+guesses of argument types (as discussed in
+Section~\ref{subsec:typetracer-challenges}).  Further, the type of dispatch
+is recorded on the {\tt \_entry} variants: recall that R can perform single or multiple dispatch on function arguments depending on their class.
+We collect dispatch information in order to analyze dispatch usage patterns.
+
+The main work is done in {\tt promise\_force\_exit}.  As arguments to
+functions in R are wrapped in promises, we delay our reflection to promise
+force time.  In R, promises are essentially objects with value and
+expression slots.  When a promise is created for a particular expression,
+that expression is merely put into the expression slot of a promise object,
+and a special {\it unbound} value is loaded into the value slot (unless some
+default value is specified, wherein it is loaded into the value slot---this
+is the case with default argument values).  When the promise is forced, the
+expression is evaluated in the forcing context, and the value is stored in
+the value slot of the promise.  So to get the type of a function argument,
+we check the value slot of the argument promise---perhaps recursively if the
+value itself is a promise.
+
+To construct a type for each argument, we make use of R's C FFI and use
+low-level machinery to collect type tags and attributes from the R runtime.
+The types that we provide to users are constructed during post-processing,
+and rely on the detailed information made available by these low-level
+reflection mechanisms.
+
+%
+%
+%
+%
+\subsection{Post-Processing and Type Ascription}
+
+The type information we collect is far more detailed than the types we
+actually generate, and so the next stage of our inference process is to
+distill the rich type information into simpler, manageable types.  For each
+unique observed call trace, we translate the representation of the type of
+each of the function arguments and returns according to the design of our
+type system: for instance, we might have determined that a function returned
+a vector of integer with 4 elements, with names, that was free of NA values,
+in which case we would write out \code{integer[]} as the type.
+
+Once types have been simplified in this way, for each function we move to
+consolidate its multiple call traces into a single function type.  We do
+this by combining the observed types of each argument into one type according to subtyping rules that are part of our
+design.  The precise subtyping rules we used will be discussed further in
+Section~\ref{sec:typesystemdesign}.
+If we cannot consolidate multiple observed argument types into a single type, we simply union the disparate types together.
+
+While our analysis collects more detailed information than necessary for type generation, the information proved useful for building up a usage profile of the language. 
+This profile informed our language design: based on the rich type information we collected, we were able to determine types that would be useful for capturing the observed behaviour.
+We discuss this iterative, data-driven language design process in Section~\ref{types}, highlighting the reasoning behind our type system design.
+
+%
+%
+%
+%
 \subsection{contractr}
@@
 that make this even harder.
-
-%
-%
-%
-%
-\section{Project Corpus}\label{sec:corpus}
-
-For this paper we have selected \CorpusLoadable packages consisting of
-\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native
-code (C/C++/Fortran).  \FK{Add more details} Figure~\ref{fig:corpus} shows
-all the packages, the size of the dots reflects the size of the project in
-terms of both R and native lines of code\footnote{ Lines of source code
-  reported excludes comments and blank lines, counted by \emph{cloc}, \cf
-  \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the
-expression code coverage in percents and the y-axis gives the number of
-reverse dependencies in log scale. Dotted lines indicate means.  Packages
-with over \PackageSizeOutierRnd lines of code are annotated.
-
-\begin{figure}[!h]  \centering
-  \includegraphics[width=.8\linewidth]{plots/corpus.pdf}
-  \caption{Corpus overview}\label{fig:corpus}
-\end{figure}
-
-These packages come from the Comprehensive R Achive Network
-(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of
-R code with over \AllCranRnd packages\footnote{CRAN receives about 6 new
-  package submissions a day~\cite{Ligges2017}} containing over \AllRCodeRnd
-and \AllNativeCodeRnd R and native code respectively. Unlike other open code
-repositories like GitHub, CRAN is a curated repository. Each submitted
-package must abide to a number of well-formedness rules that are
-automatically checked asserting certain quality. Most relevant for this work
-is that all of the runnable code (including code snippets from examples and
-vignettes) is tested and only a successfully running package is admitted in
-the archive.
-
-We have downloaded and installed all available CRAN packages. Out of the
-\AllCranRnd packages, we managed to install \AllLoadableRnd. The main reason
-for this is that \rdt is based on R 3.5.0 and some of the packages are no
-longer compatible with it. Some packages also require extra native
-dependencies which were not present on our servers.  We have two criteria
-for including a package into the corpus:
-\begin{inparaenum}[(1)]
-\item have runnable code that covers significant part of the package source
-  code from which type signatures could be inferred, and
-\item have some reverse dependencies that allows us to evaluate the inferred
-types on runnable code from these dependent packages.
-\end{inparaenum}
-The concrete thresholds were set to at least \ThresholdCodeCoverage of
-expression coverage and at minimum \ThresholdRevdeps reverse dependencies.
-
-The code coverage was computed for each package using
-\covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R
-tool.  The reverse package dependencies were extracted from the package
-metadata using builtin function.
-
-%
-%
-%
-%
-%
-%
-\section{Methodology}
-
-%
-%
-%
-%
-\subsection{typetracer and Initial Analysis}
-
-To help inform the design of our type system, and to ensure that it aligns
-with the day-to-day usage of R, we performed a corpus analysis of some of
-the most widely used R packages.  The corpus is discussed in detail in
-Section~\ref{sec:corpus}.  We built \typetracer, our dynamic analysis, in
-the R-dyntrace~\cite{oopsla19} dynamic analysis framework which is itself
-built on a modified version of R targeting R version 3.5.0---we refer the
-reader to Section~\ref{sec:r-dyntrace} for more details on the framework.
-
-%
-%
-\subsubsection{Overview}
-The goal of our analysis is to discover argument types for functions,
-builtins, and specials.  Naively intercepting function calls and recording
-argument types is incorrect, as recall that R wraps arguments in promises
-(Section \AT{what section}).  To circumvent this obstacle, we instead
-intercept the forcing of promises: when a promise is forced, we identify the
-surrounding call, collect the type of the value, and add it to the call
-trace associated with the current function call.  Once function execution
-completes, we collect information about the return value, add it to the call
-trace, and save the completed call trace to be written out later.  \AT{This
-  paragraph overlaps with the initial discussion of \contractr. We need to
-  make sure that we don't inadvertently repeat ourselves.}
-
-%
-%
-\subsubsection{Type Information}
-Once a value is obtained, either on function return or when an argument
-promise is forced, we probe the value for the following information:
-
-\begin{itemize}
-\item the value's type tag according to R's runtime. R's runtime implements
-  type tags as a typedef, e.g. \code{INTSXP} for integers, and, confusingly,
-  \code{VECSXP} for lists;
-\item its class, which is a list of string class names. Typically, the class
-  of a value is held in an attribute with name ``class'', though some
-  classes are blessed by the R runtime, such as matrix and array classes;
-\item its attributes, a list of metadata attached to the value.
-\end{itemize}
-
-Now, depending on the value's type tag, we collect further information:
-
-\begin{itemize}
-\item For R's primitive types, we collect its dimension, and if it has a
-  names attribute we collect the names as well;
-\item For lists, we recursively collect the types of list elements;
-\item For lists {\it with a names attribute}, we collect the names as well as the types of the slots that those names refer to (data.frames fall into this category);
-\item For promises, we carefully dig into the promise value {\it without
-  forcing it}, and collect information about the innermost value. If the
-  innermost value is the unbound value, then the promise is an as-of-yet
-  unevaluated expression. In this case, we do our best to guess the value of
-  the unevaluated expression: if the expression is not a language
-  expression, we collect information about it, otherwise we can say nothing
-  without evaluating it and ascribe it type ``any''. \footnote{\AT{I think
-      this is confusing, reword.}}
-\end{itemize}
-
-For vectors (and lists), we additionally iterate through the elements,
-checking if the values are NA (resp. NULL); if none are NA (resp. NULL), we
-ascribe the NA-free (resp. NULL-free) tag to the value.  Arguments that are
-part of a function's varargs (denoted \ldots in function definitions) are
-tagged as such and ignored, as varargs can contain anything (and their type
-will be equivalent to and any-typed list).
-
-Now that we've seen the high-level picture of our approach, we will elaborate on some key details and discuss particular challenges that we faced.
-
-%
-%
-\subsubsection{Challenges}
-\label{subsec:typetracer-challenges}
-
-We grappled with a number of corner cases related to R's non-strict semantics.
-
-\begin{itemize}
-
-\item First, we wanted to ensure that \typetracer captured type for
-  unevaluated arguments.  Not every argument of a function needs to be used
-  during every function execution, and the approach we discussed above
-  relies on argument promises being forced (and, consequently, on arguments
-  being used).  There are a few reasons for this.
-
-One possibility is that the argument was simply unused given the control
-flow of this particular execution (e.g. if the argument is only needed in
-one branch of an \code{if}).  This mainly occurs when arguments have default
-values for some parameters that are not critical to every execution oft he
-function.  In these cases, we still want to collect the type of the default
-value.
-
-Another possibility is that a value is used {\it without being forced}, as
-is the case when values are ``metaprogrammed'': \AT{describe, code example.}
-
-To tackle this, we make an {\it initial guess} of function argument types on
-function entry. This collects types for argument default values, and
-collects an ``unused'' type for unevaluated expressions.  If an argument
-promise is later forced, we simply update the recorded type for the
-argument.
-
-\item Second, we wanted to account for formal arguments (arguments which are
-  named in the function definition) that were {\it not} passed to the
-  function.  Put simply, a function was not called with all of its
-  arguments, and no default value was specified for missing arguments.  In
-  this case, we record a ``missing'' type for the argument.
-
-There are two obvious ways to deal with missing arguments: type them as bot,
-or as top.  As we are performing a dynamic analysis, based on package test
-code, we conservatively type these arguments as top, or \code{any}.  It is
-impossible to know dynamically what all possible function behaviours are,
-\AT{I'm still not convinced about this.}
-
-\item Third, we had to contend with R's C implementation sometimes
-  performing longjumps, which discards current R calls from the call stack.
-  In order to ensure that call traces are not lost when a longjump occurs,
-  \typetracer intercepts the unwinding process initiated by the longjump and
-  mimics the functionality of the functions being exited.  The R-dyntrace
-  framework supplies the return value during the unwinding process (if there
-  is indeed such a value), so we can capture it, analyze it, and include it
-  in the call trace.  In certain situations, however, no return value can be
-  obtained (e.g. a longjump occurred as part of some error handling), and
-  then we simply record a ``jumped'' return value, which we can deal with in
-  post-processing.  These longjumps occur quite often, particularly when S3
-  dispatch occurs, as the R implementation optimizes returns from dispatched
-  functions by initiating a longjump.
-
-\end{itemize}
-
-Having outlined these challenges, we will now explore the implementation of \typetracer in more detail.
-
-%
-%
-\subsubsection{Select Details}
-
-As mentioned, our analysis is built on the R-dyntrace framework.  R-dyntrace
-is a very efficient instrumentation framework for R, which is built on a
-modified version of the R runtime.  Our dynamic type analysis only needs to
-be run once, ever---from the analysis results, we synthesize function
-signatures which are passed along to later type-checking tools which do not
-rely on a heavily modified runtime.
-
-We primarily rely on 8 R-dyntrace callbacks: \texttt{closure\_entry},
-\texttt{closure\_exit}, \texttt{builtin\_entry}, \texttt{builtin\_exit},
-\texttt{special\_entry},\texttt{special\_exit},
-\texttt{promise\_force\_entry}, and \texttt{promise\_force\_exit}.  These
-callbacks fire when closures, builtin, and special functions are entered and
-exited, and before and after the forcing of promises.  The function-related
-callbacks are used mainly for bookkeeping: the analysis is notified that a
-closure, builtin, or special has been entered/exited by pushing/popping the
-call onto a stack.  The calls themselves store a {\it call trace} object,
-where we store the type information that we collect, populated with initial
-guesses of argument types (as discussed in
-Section~\ref{subsec:typetracer-challenges}).  Further, the type of dispatch
-is recorded on the {\tt \_entry} variants: \AT{recall} that there are
-multiple ways that R performs dynamic dispatch, based on the class of
-function arguments.  We collect dispatch information in order to analyze
-dispatch usage patterns.
-
-The main work is done in {\tt promise\_force\_exit}.  As arguments to
-functions in R are wrapped in promises, we delay our reflection to promise
-force time.  In R, promises are essentially objects with value and
-expression slots.  When a promise is created for a particular expression,
-that expression is merely put into the expression slot of a promise object,
-and a special {\it unbound} value is loaded into the value slot (unless some
-default value is specified, wherein it is loaded into the value slot---this
-is the case with default argument values).  When the promise is forced, the
-expression is evaluated in the forcing context, and the value is stored in
-the value slot of the promise.  So to get the type of a function argument,
-we check the value slot of the argument promise---perhaps recursively if the
-value itself is a promise.
-
-To construct a type for each argument, we make use of R's C FFI and use
-low-level machinery to collect type tags and attributes from the R runtime.
-The types that we provide to users are constructed during post-processing,
-and rely on the detailed information made available by these low-level
-reflection mechanisms.
-
-%
-%
-%
-%
-\subsection{Post-Processing and Type Ascription}
-
-The type information we collect is far more detailed than the types we
-actually generate, and so the next stage of our inference process is to
-distill the rich type information into simpler, manageable types.  For each
-unique observed call trace, we translate the representation of the type of
-each of the function arguments and returns according to the design of our
-type system: for instance, we might have determined that a function returned
-a vector of integer with 4 elements, with names, that was free of NA values,
-in which case we would write out \code{integer[]} as the type.
-
-Once types have been simplified in this way, for each function we move to
-consolidate its multiple call traces into a single function type.  We do
-this by combining the observed types of each argument into one type
-(possibly a union, mind), according to subtyping rules that are part of our
-design.  The precise subtyping rules we used will be discussed further in
-Section~\ref{sec:typesystemdesign}.
-
-While our analysis collects more detailed information than necessary for
-type generation, the information proved useful for building up a usage
-profile of the language. Through it, we were able to determine the best
-translation rules (i.e., determining the type system we are translating to)
-through an iterative process, involving analyzing usage patterns that arose
-in the data and making design decisions based on them, revisiting the data
-as necessary.  In the following section, we will discuss these decisions in
-more detail, highlighting the reasoning behind our type system design.
 
@@
 
+\AT{Include a figure? Histogram(s) of types?}
+
+In order to count the occurrences of each type, we subdivided them into {\it kinds}: 
+For our purposes, the {\it kind} of a type is equivalent to it's top-level type constructor, so all classes have kind ``class'', all lists have kind ``list'', and so on.
+We found that classes are the most common kind of type, accounting for roughly 31\% of types.
+\AT{Make sure to transform class<function> into any=>any.}
+In order, the most common classes are matrices (12\%), data.frames (7.5\%), formulas (2\%), factors (2\%), and tibbles (2\%).
+Roughly 25\% of classes are part of R's base language implementation, the others being defined as part of some package's functionality.
+
+Besides classes, scalars and vectors are the next most common kind, making up 41\% of types. with scalars making up 28.1\% of types and vectors 12.65\%.
+Nulls and lists follow at 8\% and 7\% respectively, and the vararg type \ldots makes up 6.5\% of arguments.
+This all totals up to over 90\% of types.
+We will discuss the any type in more detail in Section~\ref{subsec:any}.
+
+It would appear that our simple type annotations nicely account for the landscape of values in R.
+\AT{Not sure what to say about classes.}
+% Classes are a common kind of type in R, and for the purposes of this type annotation language allowing users to specify class names as part of class types is welcome: we saw that the breakdown of the class kind does not reveal many outlier classes,
+
 %
@@
 
+The polymorphism in R is best described by {\it ad hoc} polymorphism, wherein polymorphism is implemented by function and operator overloading (see S3 dispatch) rather than in some more fundamental way in the language design.
+For the purposes of this evaluation, we will say that an argument is polymorphic if its type is a union of two or more types, in other words the argument has been observed to take more than one type of value.
+
+\AT{We need better numbers on this.}
+We found that 15\% of argument positions were polymorphic, and that X\% of functions were polymorphic.
+
+Our type annotation language does not support sophisticated class typing, and much of the polymorphism in R is due to these classes, namely through the S3 dispatch mechanism. 
+In future work, we aim to more fully handle S3 and objects in R.
+For now, if we count out
+
 %
@@
 \subsubsection{Use of \code{any}}
+\label{subsec:any}
 

### 2020-05-14 paper/main.tex
minor

@@
 \item Firstly, we faced the problem of type-checking arguments in a
-  non-strict language while retaining the non-strict semantics. When a
+  non-strict language while retaining the non-strict semantics. Recall that when a
   function is called in R, parameters are bound to unevaluated code thunks

### 2020-05-14 paper/main.tex
minor

@@
 
-The \code{binom} package is a popular package \AT{do we want e.g.
-\# revdeps?} for computing confidence intervals for binomial experiments, and the \code{binom.profile} function does so using the profile likelihood, the details of which are not relevant here.
-Here, we highlighted a relevant snippet of code, which appears at the beginning of the function: the programmer first binds the vectors into a matrix (line 2), with one column for each of \code{x} and \code{n}, then finds rows where both columns are not NA (line 3), then extracts non-NA values and stores them into \code{x} and \code{n} respectively (lines 4-5).
+The \code{binom} package is a popular R package for computing confidence intervals for binomial experiments, and the \code{binom.profile} function does so using the profile likelihood, the details of which are not relevant here.
+This code snippet highlights a data sanitization pattern appearing at the beginning of the function: the programmer first binds the vectors into a matrix (line 2), with one column for each of \code{x} and \code{n}, then finds rows where both columns are not NA (line 3), then extracts non-NA values and stores them into \code{x} and \code{n} respectively (lines 4-5).
 Importantly, this process is entirely quiet, and the user of \code{binom.profile} will not be notified if their data happens to contain NAs, and the NAs will be silently ignored. 
@@
 To combat this, we introduce a modifier on primitive types: the type \code{^T} signifies that primitive type \code{T} might contain an NA value, while \code{T} signifies that \code{T} is NA-free. 
+The NA-free case has a shorter syntax as NA-free vectors are \AT{more common} than the alternative.
 If the \code{binom} developers were to annotate arguments \code{x} and \code{n} with such an NA-free type, \contractr will notify users when \code{binom.profile} is called with NA-full values, rather than just quietly ignoring them.
-\AT{Gather usage statistics.}
 
 Note that when combining call traces involving \code{^T} and \code{T} types, we subsume \code{T} types into \code{^T} types: if NA-free and NA-full vectors are passed to an argument, then we say that the argument can handle NA-full vectors.
-This is equivalent to \code{T <: ^T}.
+This is equivalent to saying \code{T <: ^T}.
 
@@
 on lines 8-9, \code{x} undergoes a vector ``type check'';
-on lines 11-12, \code{n} is assumed to be a scalar (if it were a vector, R would generate a warning, as \code{m < n} would return a vector of logicals obtained from checking if \code{m} were less than each element of \code{n} \AT{do we want to put the warning here?}).
-These checks exemplify behaviour we would like to capture with our annotation language, as we would like to differentiate between scalars and vectors in addition to checking for the basic type of a value (e.g.
-integer vs double),
-
+on line 11, \code{n} is assumed to be a scalar, as if it were a vector, the code \code{m < n} would essentially check \code{m < n[i]} for each element in \code{n}, returning a vector of comparison results, and the conditional of an if must be a scalar else R generates a warning.
+
+These checks exemplify functionality that we would like to include in our type annotation language, namely, we would like our type language to differentiate between scalars and vectors.
 To that end, we introduce the following types into our type language:
 for some primitive type \code{T}, the user may modify \code{T} with a modifier \code{[]} to indicate that it is a vector of \code{T}.
-Perhaps surprisingly, we found that scalars occurred more often than vectors, and chose the scalar type syntax as the default. 
+Perhaps surprisingly, we found that scalars \AT{occurred more frequently} than vectors, and for that reason we chose the shorter scalar type syntax as the default.
 The NA-free modifier works as expected: \code{^T[]} indicates a vector of \code{T} that is free of NAs, while \code{^T} indicates a scalar which is not NA.

### 2020-05-14 paper/main.tex
abstract

@@
 \begin{abstract}
-  Retrofitting a type system onto an existing dynamic language which was not
-  designed to be statically typed is challenging. This paper takes the first
-  step in the direction of designing a type system for the R programming
-  language. We argue that R is particularly challenging due to its unique
-  combination of features and to the programming idioms that have developped
-  over the last two decades. As our starting point, we infer type signatures
-  for \NUMFUNCTIONS functions from \NUMPACKAGES packages amongst the most
-  widely used open source R libraries. The signatures use a minimal type
-  language consisting of basic types, vectors, classes and disjunctions. We
-  purposefully leave out notions such as subtyping, higher-order types, and
-  polymorphism. Our goal is to understand what benefits a simple lanugage
-  can yield.  We also implement a contract system that can use the inferred
-  types to validate their correctness. We use the reverse depedencies of the
-  packages to evaluate the benefits of typing.
+The R programming language is widely used in a variety of domains for tasks
+related to data science. R was designed to favor an interactive style of
+programming with minimal syntactic and conceptual overhead. This design is
+well suited to interactive data analysis, but a bad firt for tools such as
+compilers or program analyzers which must generate performant code or catch
+programming errors.  In particular, R has no type annotations, its
+operations are dynamically checked at runtime. The starting point for our
+work are the twin questions, \emph{what expressive power is needed to
+  accurately type R code?} and \emph{which type system is the R community
+  willing to adopt?} Both questions are difficult to answer without actually
+designing a type system and attempting to convince users and package
+developers to try the proposed system.  The goal of this paper is to provide
+data that can feed into the design process. To this end, we perform a large
+corpus analysis with aim to gain insights in the degree of polymorphism
+exhibited by idiomatic R code and explore potential benefits that the R
+community could accrue from even a simple type system.  As a starting point,
+we infer type signatures for \NUMFUNCTIONS functions from \NUMPACKAGES
+packages among the most widely used open source R libraries. The signatures
+use a minimal type language with only basic types, vectors and classes.  We
+implement a contract checker that can be used to validate the inferred type
+signatures.
 \end{abstract}

### 2020-05-14 paper/main.tex
fixes

--- 2020-05-14 paper/main.tex
@@
 
-In R, function can be called with named parameters, R support variable
-argument lists, and arguments can have default values. Putting all of these
-together consider the following declaration:
+In R, functions can be called with named parameters, arguments can have default values, and R also support variable argument lists. 
+To illustrate all of this in a single example, consider:
 
@@
 R has one builtin notion of type that can be queried by the \k{typeof}
-function. Over the years, programmers have found the need for a richer type
-structure and have added {\it attributes}. The best way to think of attributes is
-as an optional map from name to values that can be attached to any object.
-Attributes are used to encode various type structures. They can be queried
-with functions such as \k{attributes} and \k{class}.
+function. Figure~\ref{types} lists all of the builtin types that are provided by the
+language. They are the possible return values of \k{typeof}. There is no
+intrinsic notion of subtyping in R. But, in many context a \k{logical} will
+convert to \k{integer}, and an \k{integer} will convert to \k{double}.  Some
+off conversion can occur in corner cases, such as \k{1<"2"} holds and
+\k{c(1,2)[1.6]} returns the first element of the vector, as the double is
+converted to an integer. R does not distinguish between scalars and vectors
+(they are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
+  typeof(c(5,5))} \code{ == "double"}. Finally all vectorized data types have a
+distinguished missing value denoted by \code{NA}. The default type of
+\code{NA} is \k{logical}. We can see that \code{typeof(NA)=="logical"}, but
+NA inhabits every type: \code{typeof(c(1,NA)[2])=="double"}.
 
@@
 
-Figure~\ref{types} lists all of the builtin types that are provided by the
-language. They are the possible return values of \k{typeof}. There is no
-intrinsic notion of subtyping in R. But, in many context a \k{logical} will
-convert to \k{integer}, and an \k{integer} will convert to \k{double}.  Some
-off conversion can occur in corner cases, such as \k{1<"2"} holds and
-\k{c(1,2)[1.6]} returns the first element of the vector, as the double is
-converted to an integer. R does not distinguish between scalars and vectors
-(they are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
-  typeof(c(5,5))} \code{ == "double"}. Finally all vectorized data types have a
-distinguished missing value denoted by \code{NA}. The default type of
-\code{NA} is \k{logical}. We can see that \code{typeof(NA)=="logical"}, but
-NA inhabits every type: \code{typeof(c(1,NA)[2])=="double"}.
-
 With one exception all vectorized data types are monomorphic, the exception
@@
 
+Over the years, programmers have found the need for a richer type
+structure and have added {\it attributes}. The best way to think of attributes is
+as an optional map from name to values that can be attached to any object.
+Attributes are used to encode various type structures. They can be queried
+with functions such as \k{attributes} and \k{class}.
 The addition of attributes lets programmers extend the set of types by
@@
 
-On the subject of matrices, they represent another layer of complexity on top of vectorized primitives.
-Internally, matrices are simply vectors with class ``matrix'' and a \code{dims} attribute indicating matrix dimensions, and many functions automatically coerce vectors to matrices when appropriate (while not codified in the language semantics, many internal functions perform the coercion).
-Relying on automatic coercion is unwise, and many functions check to ensure that their arguments are matrices, particularly those in packages which rely on matrices.
+Matrices represent another layer of complexity on top of vectorized primitives.
+Internally, matrices are simply vectors with class ``matrix'' and a \code{dims} attribute indicating matrix dimensions, and while not codified in the language semantics, many internal functions will coerce vectors to matrices automatically.
+Relying on automatic coercion is unwise, and many functions check to ensure that their arguments are matrices.
 For instance, consider the code in Figure~\ref{fig:matrix-example}.
@@
 
-The \code{rowWeightedMeans} function calculated the weighted means of rows or columns of a matrix \code{x}.
+The \code{rowWeightedMeans} function calculates the weighted means of rows or columns of a matrix \code{x}.
 This function exhibits good practice in type checking arguments, rightly producing a message to the user indicating that an explicit matrix should be passed.
@@
 To capture cases like this, we include a \code{class<matrix>} type in our type annotation language.
-If \code{rowWeightedMeans} were outfitted with such a type, \contractr can perform the check present in this code automatically.
-That said, we don't intend to subsume all argument validation code, discussed next.
+If \code{rowWeightedMeans} were outfitted with such a type, \contractr can perform the check present in this code automatically, notifying the users passing ill-typed arguments.
 
@@
 
-Given the discussion of scalars, vectors, and matrices, it might be tempting to consider factoring dimensionality into types.
-The primary drawback of this would be rendering our type annotation language into a language of dependent types, which we believe too complex to retrofit onto a dynamic data science language such as R.
-This does mean our types cannot subsume all validation code, such as some later code from the aforementioned \code{rowWeightedMeans} function, shown in Figure~\ref{fig:general-validation-example}.
+Thus far we've seen instances where we introduce types that subsume existing argument validation code, but our intention is not to subsume {\it all} such code.
+As an example of something we {\it do not} want to do, consider the code in Figure~\ref{fig:general-validation-example}.
 
@@
 
+This code is from later in the same \code{rowWeightedMeans} function discussed previously.
 Here, the length of the weights vector \code{w} is being checked for equality with the number of columns in the matrix \code{x}. 
+Specifying this relationship as a type amounts to {\it dependent typing}, which we believe too complex to retrofit onto a dynamic data science language such as R.  
 Ultimately, we decided to leave these kinds of checks up to the programmer.
-\AT{I feel like we might want a better argument to not make our type language dependent (better than lol pls no dependent types), but I could be wrong.
-Maybe reviewers would be ok with a simple argument.}
+\AT{Can we say more here?}
 
@@
 We are faced with several possible designs for list types.  
-Chief concerns include the heterogeneity of list elements, and the presence of a \code{names} attribute, allowing list elements to be accessed in a struct-like manner.  
-\AT{Rarely?}, programmers even write code expecting lists of a particular pre-defined length.  
+Chief concerns include the heterogeneity of list elements and the presence list element names, which allow list elements to be accessed in a struct-like manner.  
+% \AT{Rarely?}, programmers even write code expecting lists of a particular pre-defined length.  
 The next sections highlight some important design decisions for our list types.
@@
 \subsubsection{Relationship Between Lists and Vectors}
+
+\AT{Worth a code example?}
 
@@
 %
-\subsubsection{Forgoing Tuples, or Short, Constant Length Lists}
-
-We briefly entertained the possibility of including a \code{tuple} type, which would have looked like e.g. \code{tuple<int, int>} for a list of two integers.  
-When configured to detect tuples, our analysis did indeed pick up \AT{a number}, but post-processing and simplification revealed that in \AT{many} cases, they would be subsumed by co-occurring \code{list} types.
-In other words, more often than not it just so happened that functions were being called with short lists.  
-Of the cases that tuples were not subsumed by lists, \AT{often} there were \AT{tuples of different lengths}, implying that the shape of the argument was not essential.
-
-\AT{I feel like this is the least important of the design decisions we made, so I'm leaving it for the end, we might not even want it.}
+% \subsubsection{Forgoing Tuples, or Short, Constant Length Lists}
+
+% We briefly entertained the possibility of including a \code{tuple} type, which would have looked like e.g. \code{tuple<int, int>} for a list of two integers.  
+% When configured to detect tuples, our analysis did indeed pick up \AT{a number}, but post-processing and simplification revealed that in \AT{many} cases, they would be subsumed by co-occurring \code{list} types.
+% In other words, more often than not it just so happened that functions were being called with short lists.  
+% Of the cases that tuples were not subsumed by lists, \AT{often} there were \AT{tuples of different lengths}, implying that the shape of the argument was not essential.
+
+% \AT{I feel like this is the least important of the design decisions we made, so I'm leaving it for the end, we might not even want it.}
 
@@
 
-Another pattern that we observed was the use of consistently-named lists.
+We observed the use of consistently-named lists in \AT{a number of functions}.
 In R, ``fields'' of named lists can be accessed using the \code{$} operator, as illustrated in the code snippet in Figure~\ref{fig:struct-ex}.
@@
 \begin{lstlisting}
-calculate.neuralnet <- function (data, model.list, hidden, rep,
-            startweights, exclude, constant.weights) {
-    result <- generate.startweights(model.list, hidden, startweights,
-                                    rep, exclude, constant.weights)
-    weights <- result$weights
-    exclude <- result$exclude
-    ...
+agricolae::cv.model <- function(x) {
+  suma2 <- sum(x$residual^2)
+  gl <- x$df.residual
+  promedio <- mean(x$fitted.values)
+  return(sqrt(suma2/gl)*100/promedio)
+}
+
+cars_model <- lm(dist ~ speed, data = cars)
+cv.model(cars_model)
 \end{lstlisting}
 
-\caption{Struct example from the \code{neuralnet} package.}
+\caption{Struct example from the \code{agricolae} package.}
 \label{fig:struct-ex}
@@
 \end{figure}
+
+In the code snippet, the function \code{cv.model} takes an argument \code{x}, which we observed to always be a {\it linear model}, which is part of R's built-in statistics functionality.
+Internally, linear models are represented as lists with named elements, and one may access these named elements using the \code{$} syntax, as seen on lines 2-4 in the snippet.
 

### 2020-05-14 paper/main.tex
intro

@@
 \begin{document}
-\title{Extracting Function Signatures from Program Traces}
+\title{Designing Types for R, Empirically}
 
@@
 \newcommand{\NUMPACKAGES}{412\xspace}  %%% TODO auto-gen
+\newcommand{\PACKAGES}{412\xspace}  %%% TODO auto-gen
+\newcommand{\genthat}{genthat\xspace}  %%% TODO auto-gen
 \newcommand{\YEARS}{20\xspace} %%TODO fix
@@
 signatures.
-\end{abstract}
-\maketitle
+\end{abstract} \maketitle
 
 \section{Introduction}
+
+Our community builds, improves, and reasons about programming languages.  To
+make design decisions that benefit users, we need to understand our target
+language as well as the real-world needs it answers. Often, we can appeal to
+our intuition, as many languages are intended for general purpose
+programming tasks. Unfortunately, intuition may fail when looking at
+domain-specific languages as these languages designed for a specific group
+of users to solve very specific needs. This is the case of the data science
+language R.
+
+R and its ancestor S were designed, implemented, and maintained by
+statisticians. Originally they aimed to be glue languages for reading data
+and calling statistical routines written in Fortran. Over three decades they
+became widely used across many fields for data analysis and visualization.
+Modern R, as an object of study, is fascinating. It is a vectorized,
+dynamically typed, lazy functional language with limited side-effects,
+extensive reflective facilities and retrofitted object-oriented programming
+support.
+
+Many of the design decisions that gave us R were intended to foster an
+interactive, exploratory, programming style. These include, to name a few,
+the lack of type annotations, the ability to use syntactic shortcut, and the
+automatic conversion between data types.  While these choices have led to a
+language with a low barrier to entry---many data science educational
+programs do not teach R itself but simply introduce some of its key
+libraries---they have also created a language where errors can go
+undetected.
+
+Retrofitting a type system to the R programming language would increase our
+assurance in the result of data analysis. But, we are faced with two
+challenges. First, it is unclear what would be the \emph{right} type system
+for a language as baroque as R. For example, one of the most popular data
+type, the \code{data.frame}, is manipulated through reflective operations --
+a data frame is a table whose columns can be added or removed on the fly.
+Second, but just as crucially, designing a type system that will be adopted
+would require overcoming some prejudices and educating large numbers of
+users.
+
+xxx WIP xxxx
+
+This paper is a data-driven study of what a type system for the R language
+could look like. Our intention is to propose changes to the
+language, and for any changes to be accepted by the user community they
+must clearly benefit the language without endangering backwards
+compatibility. Our goal is thus to find a compromise between simplicity and
+usefulness; the proposed type system should cover most common programming
+idioms while remaining easy to use. In order to do this, we need to
+understand the degree of polymorphism present in R code, that is to say, how
+programmers leverage the dynamic nature of R to write code that can accept
+arguments of different types.  This understanding will drive our design.
+
+We propose to capture the degree of polymorphism present in R by means
+of a dynamic analysis of widely used libraries. For each function call we
+can record the types of its arguments and of its return value. This allows
+us to observe how many different combination of types are accepted by any
+given function. Unlike many other languages, R has a carefully curated
+software repository called CRAN. To be deposited in CRAN, a package must
+come with sample dataset, tests and executable use-cases. As part of normal
+operations these tests are run regularly and failing packages are removed.
+This allowed us to have access to \PACKAGES libraries and about an order of
+magnitude more runnable scripts that exercise those libraries.
+
+The contributions of this paper are thus as follows:
+\begin{itemize}
+\item A large-scale analysis of the polymorphism present in function
+  signatures of \PACKAGES widely used and actively maintained R packages.
+\item A tracing and analysis pipeline that extends a previously published
+  test generation tool named \genthat.
+\item A set of type annotations that captures how programmers use types
+in existing R code.
+\end{itemize}
+
+One threat to validity of our work is that we rely on dynamic analysis, so
+our conclusions are only as good as the coverage of the possibly function
+behaviors. Previous work~\cite{issta18} reported that running all the
+scripts that come with CRAN packages gives, on average, 68\% test coverage.
+We attempted to mitigate the threat coming from the fact that only part of
+the code is being exercised by manual analysis. It would be reasonable to
+ask for confirmation of the data by static analysis of the code, but sound
+static analysis of R is difficult because of the extensive use of reflective
+features such as \code{eval} and of the ability to redefine the meaning of
+operators such as \code{+} and \code{if}.  Another threat to validity is
+that we only have access to code that has been deposited in the CRAN
+repository. While this may bias our findings towards code written to be
+reusable and, possibly, better engineered than typical user code. This is
+also the code that would most benefit from type annotations.
 

### 2020-05-14 paper/main.tex
fixes fixes fixes

--- 2020-05-14 paper/main.tex
@@
 convert to \k{integer}, and an \k{integer} will convert to \k{double}.  Some
-off conversion can occur in corner cases, such as \k{1<"2"} holds and
+off conversions can occur in corner cases, such as \k{1<"2"} holds and
 \k{c(1,2)[1.6]} returns the first element of the vector, as the double is
@@
 Scalar data types include the distinguished \k{NULL} value, which is also of
-type \k{NULL}, instance of classes written using the S4 object system,
+type \k{NULL}, instance of classes written using the S4~\footnote{put footnote on S4 here} object system,
 closures and environments.  The implementation of R has a number of other
@@
 The goal of our analysis is to discover argument types for functions,
-builtins, and specials.  Naively intercepting function calls and recording
-argument types is incorrect, as recall that R wraps arguments in promises
-(Section~\ref{sec:R}).  To circumvent this obstacle, we instead
+builtins, and specials (R functions such as \{, \code{if}, etc.).  
+First, recall that R wraps arguments in promises (Section~\ref{sec:R}).
+Naively intercepting function calls and recording argument types would result in incomplete data:
+While we can dig into promises without forcing them, if the innermost value is an unevaluated expression we are unable to probe it without evaluating it, which may alter program behaviour.
+To circumvent this obstacle, we instead
 intercept the forcing of promises: when a promise is forced, we identify the
@@
 Once a value is obtained, either on function return or when an argument
-promise is forced, we probe the value for the following information:
+promise is forced, we can probe it without interfering with program behaviour.
+We query the value for the following information:
 
@@
 
-\item First, we wanted to ensure that \typetracer captured type for
-  unevaluated arguments.  Not every argument of a function needs to be used
-  during every function execution, and the approach we discussed above
-  relies on argument promises being forced.  There are a few reasons for this.
+\item
+
+First, we wanted to ensure that \typetracer captured types for unevaluated arguments.  
+Unfortunately, not every argument of a function is necessarily used during every function execution, meaning that some arguments may remain unevaluated (as they are never forced), and there are a few reasons for this.
 
@@
 
-cars_model <- lm(dist ~ speed, data = cars)
-cv.model(cars_model)
+data(sweetpotato)
+model<-aov(yield~virus, data=sweetpotato)
+cv.model(model)
 \end{lstlisting}
 
-\caption{Struct example from the \code{agricolae} package.}
+\caption{Struct example from a test in the \code{agricolae} package.}
 \label{fig:struct-ex}
@@
 Internally, linear models are represented as lists with named elements, and one may access these named elements using the \code{$} syntax, as seen on lines 2-4 in the snippet.
-
-In the code snippet, the function \code{generate.startweights} return a list with (at least) two named elements: \code{weights} and \code{exclude}, which are accessed through \code{result$weights} and \code{result$exclude} respectively.
-We include a struct type in our language to capture this behaviour: we would encode the return type of \code{generate.startweights} as e.g. \code{struct<x: double[], y: logical[]>}.  
-We also thought to encode data frames as structs, as they are essentially lists with named columns of homogeneous values, but it was difficult to draw any meaningful conclusions from the data as unfortunately structs introduced a lot of noise into our analysis.
-
-It is common for R example code to use some built-in data sets, such as \code{mtcars}, a data set extracted from a Motor Trends US magazine in 1974.
-These data sets are used to show off some package functionality with ``real data'', and when \typetracer is configured to detect structs we saw a marked increase in misrepresentative types.
+To describe these kinds of values, we included a struct type in our type system, allowing users to specify name, type pairs as part of list type definitions.
+
+This example also illustrates an issue we grappled with when dealing with structs.
+On line 8, we see the call \code{data(sweetpotato)}.
+It is quite common for R example code to use some built-in data sets, such as \code{sweetpotato}, to show off some package functionality with ``real data''.
+We found that the types of functions which use these built-in data sets (such as \code{aov}) became polluted with the names from the example sets.
+
+% We also thought to encode data frames as structs, as they are essentially lists with named columns of homogeneous values, but it was difficult to draw any meaningful conclusions from the data as unfortunately structs introduced a lot of noise into our analysis.
+
+When \typetracer is configured to detect structs we saw a marked increase in misrepresentative types.
 We implemented a unification strategy for struct types, removing them if they co-occurred with list types, and turning them into lists if two co-occurring struct types shared no common names, but even so we found ourselves with function signatures that were full of noise introduced by the struct names.

### 2020-05-14 paper/main.tex
intro

@@
 
-xxx WIP xxxx
-
 This paper is a data-driven study of what a type system for the R language
-could look like. Our intention is to propose changes to the
-language, and for any changes to be accepted by the user community they
-must clearly benefit the language without endangering backwards
-compatibility. Our goal is thus to find a compromise between simplicity and
-usefulness; the proposed type system should cover most common programming
-idioms while remaining easy to use. In order to do this, we need to
-understand the degree of polymorphism present in R code, that is to say, how
-programmers leverage the dynamic nature of R to write code that can accept
-arguments of different types.  This understanding will drive our design.
+could look like. Our intention is to eventually propose changes to the
+language, but we are aware that for any changes to be accepted by the user
+community they must clearly benefit the language without endangering
+backwards compatibility. Our goal is thus to find a compromise between
+simplicity and usefulness; any proposed type system should cover common
+programming idioms while remaining easy to learn and to use.
+
+In order to do this, we need to understand the degree of polymorphism
+present in R code, that is to say, how programmers leverage the dynamic
+nature of R to write code that can accept arguments of different types.
+This understanding will drive our design.
 
@@
 
-How does one retrofit a type system to an existing language? Three succesful
-models are Hack, Typed Racket and TypeScript.
-
-The story that we're aiming for: Designing type systems for established
-languages can be a challenging.  We aim to approach this problem by
-essentially reverse-engineering a type system from a corpora analysis of a
-representative subset of R code.  We present a simple type system that
-captures the vast majority of usage of R --- it is lightweight, and
-\AT{should provide enough information to be useful to a compiler}.
-
-To evaluate our type system, we present a type-checker that is aware of R's
-promise semantics.  In R, function arguments are wrapped in promises such
-that they are not evaluated until they are needed.  Our tool deals with
-these promises by inserting a check which fires only when the promise is
-forced, at which point the value is checked for compliance with the type.
-
-%
-%
-%
-\section{Background}
-
-In this section, we will introduce the R programming language, as well as discuss the dynamic analysis framework we used as part of our initial analysis.
-
-%
-%
-%
-%
-\subsection{The R Programming Language}
-\label{sec:R}
+%
+\section{Background} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+In this section, we will introduce the R programming language, as well as
+discuss the dynamic analysis framework we used as part of our initial
+analysis.
+
+%
+%
+\subsection{The R Programming Language}\label{sec:R}
 
@@
 
-\begin{figure}[!hb]{\small
-\begin{lstlisting}
+\begin{figure}[!hb]{\begin{lstlisting}
 > normSum <- function( m )  round( m / sum(m), 2)
@@
 
-In R, functions can be called with named parameters, arguments can have default values, and R also support variable argument lists. 
-To illustrate all of this in a single example, consider:
+In R, functions can be called with named parameters, arguments can have
+default values, and R also support variable argument lists.  To illustrate
+all of this in a single example, consider:
 
@@
 R has one builtin notion of type that can be queried by the \k{typeof}
-function. Figure~\ref{types} lists all of the builtin types that are provided by the
-language. They are the possible return values of \k{typeof}. There is no
-intrinsic notion of subtyping in R. But, in many context a \k{logical} will
-convert to \k{integer}, and an \k{integer} will convert to \k{double}.  Some
-off conversions can occur in corner cases, such as \k{1<"2"} holds and
-\k{c(1,2)[1.6]} returns the first element of the vector, as the double is
-converted to an integer. R does not distinguish between scalars and vectors
-(they are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
-  typeof(c(5,5))} \code{ == "double"}. Finally all vectorized data types have a
-distinguished missing value denoted by \code{NA}. The default type of
-\code{NA} is \k{logical}. We can see that \code{typeof(NA)=="logical"}, but
-NA inhabits every type: \code{typeof(c(1,NA)[2])=="double"}.
+function. Figure~\ref{types} lists all of the builtin types that are
+provided by the language. They are the possible return values of
+\k{typeof}. There is no intrinsic notion of subtyping in R. But, in many
+context a \k{logical} will convert to \k{integer}, and an \k{integer} will
+convert to \k{double}.  Some off conversions can occur in corner cases, such
+as \k{1<"2"} holds and \k{c(1,2)[1.6]} returns the first element of the
+vector, as the double is converted to an integer. R does not distinguish
+between scalars and vectors (they are all vectors), so \code{typeof(5) ==}
+\code{typeof(c(5)) == typeof(c(5,5))} \code{ == "double"}. Finally all
+vectorized data types have a distinguished missing value denoted by
+\code{NA}. The default type of \code{NA} is \k{logical}. We can see that
+\code{typeof(NA)=="logical"}, but NA inhabits every type:
+\code{typeof(c(1,NA)[2])=="double"}.
 
 \begin{wrapfigure}{r}{6.1cm}
-\footnotesize\begin{tabular}{l|c|l@{}}\hline
-\multicolumn{3}{l}{\bf Vectorized data types:}  \\\hline
-\k{logical}   & \L & vector of boolean values\\
-\k{integer}   & \I & vector of 32 bit integer values\\
-\k{double}    & \D & vector of 64 bit floating points\\
-\k{complex}   & \X & vector of complex values\\
-\k{character} & \C & vector of strings values\\
-\k{raw}       & \R & vector of bytes\\
-\k{list}      & \l & vector of values of any type\\\hline
-\multicolumn{3}{l}{\bf Scalar data types:}\\\hline
-\k{NULL}      & \sN &  singleton null value\\
-\k{S4}        & \sS &  instance of a S4 class \\
-\k{closure}   & \sF & a function with its environment\\
-\k{environment}&\sE &  a mapping from symbol to value \\\hline
-\multicolumn{3}{l}{\bf Implementation data types:}\\\hline
-\multicolumn{3}{l}{\k{special},
+\footnotesize\begin{tabular}{l|l@{}}\hline
+\multicolumn{2}{l}{\bf Vectorized data types:}  \\\hline
+\k{logical}   & vector of boolean values\\
+\k{integer}   & vector of 32 bit integer values\\
+\k{double}    & vector of 64 bit floating points\\
+\k{complex}   & vector of complex values\\
+\k{character} & vector of strings values\\
+\k{raw}       & vector of bytes\\
+\k{list}      & vector of values of any type\\\hline
+\multicolumn{2}{l}{\bf Scalar data types:}\\\hline
+\k{NULL}      &  singleton null value\\
+\k{S4}        &  instance of a S4 class \\
+\k{closure}   &  function with its environment\\
+\k{environment}& mapping from symbol to value \\\hline
+\multicolumn{2}{l}{\bf Implementation data types:}\\\hline
+\multicolumn{2}{l}{\k{special},
 \k{builtin},
-\k{symbol} (\sY),
+\k{symbol},
 \k{pairlist},
 \k{promise}}\\
-\multicolumn{3}{l}{
+\multicolumn{2}{l}{
 \k{language},
@@
 }\\
-\multicolumn{3}{l}{
+\multicolumn{2}{l}{
 \k{externalprt},
@@
 
-\paragraph{Summary.} The most common values in R computations are vectorized
-types. R programs do not have a way to constrain values to be scalar.
-\k{NULL} is sometimes used to represent the case when no value is
-available. \k{NA} is used within vector to represent missing observations.
-Attributes can decorate values and are used as building blocks for
-object-oriented programming. A potential type system for R could focus only
-on the builtin types, if one wanted to strive for simplicity, or it could
-try to capture attributes at the risk of increased complexity.
-
-%
-%
-%
-%
-\subsection{R-dyntrace}
-\label{sec:r-dyntrace}
-
-R-dyntrace is an efficient dynamic analysis framework for R.
-
-%
-%
 %
@@
 \CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native
-code (C/C++/Fortran).  \FK{Add more details} Figure~\ref{fig:corpus} shows
-all the packages, the size of the dots reflects the size of the project in
-terms of both R and native lines of code\footnote{ Lines of source code
-  reported excludes comments and blank lines, counted by \emph{cloc}, \cf
+code (C/C++/Fortran). Figure~\ref{fig:corpus} shows all the packages, the
+size of the dots reflects the size of the project in terms of both R and
+native lines of code\footnote{ Lines of source code reported excludes
+  comments and blank lines, counted by \emph{cloc}, \cf
   \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the
@@
 %
-%
-%
-%
 \section{Methodology}
 
-In this section, we will describe our approach to data collection and analysis through \typetracer, a dynamic analysis built in the \rdt framework.
-We will then describe the process by which we distill the rich information we collect with \typetracer to more digestible ``types'', though we discuss the precise details of our type annotation language in Section~\ref{sec:types}.
-Finally, we discuss \contractr, a package built in R which supports our type language and can generate and check contracts for function types.
-
-%
-%
+In this section, we will describe our approach to data collection and
+analysis through \typetracer, a dynamic analysis built in the \rdt
+framework.  We will then describe the process by which we distill the rich
+information we collect with \typetracer to more digestible ``types'', though
+we discuss the precise details of our type annotation language in
+Section~\ref{sec:types}.  Finally, we discuss \contractr, a package built in
+R which supports our type language and can generate and check contracts for
+function types.
+
 %
@@
 The goal of our analysis is to discover argument types for functions,
-builtins, and specials (R functions such as \{, \code{if}, etc.).  
-First, recall that R wraps arguments in promises (Section~\ref{sec:R}).
-Naively intercepting function calls and recording argument types would result in incomplete data:
-While we can dig into promises without forcing them, if the innermost value is an unevaluated expression we are unable to probe it without evaluating it, which may alter program behaviour.
-To circumvent this obstacle, we instead
-intercept the forcing of promises: when a promise is forced, we identify the
-surrounding call, collect the type of the value, and add it to the call
-trace associated with the current function call.  Once function execution
-completes, we collect information about the return value, add it to the call
-trace, and save the completed call trace to be written out later.
+builtins, and specials (R functions such as \{, \code{if}, etc.).  First,
+recall that R wraps arguments in promises (Section~\ref{sec:R}).  Naively
+intercepting function calls and recording argument types would result in
+incomplete data: While we can dig into promises without forcing them, if the
+innermost value is an unevaluated expression we are unable to probe it
+without evaluating it, which may alter program behaviour.  To circumvent
+this obstacle, we instead intercept the forcing of promises: when a promise
+is forced, we identify the surrounding call, collect the type of the value,
+and add it to the call trace associated with the current function call.
+Once function execution completes, we collect information about the return
+value, add it to the call trace, and save the completed call trace to be
+written out later.
 %
@@
 \item For lists, we recursively collect the types of list elements;
-\item For lists {\it with a names attribute}, we collect the names as well as the types of the slots that those names refer to (data.frames fall into this category);
+\item For lists {\it with a names attribute}, we collect the names as well
+  as the types of the slots that those names refer to (data.frames fall into
+  this category);
 \item For promises, we carefully dig into the promise value {\it without
@@
 
-\item
-
-First, we wanted to ensure that \typetracer captured types for unevaluated arguments.  
-Unfortunately, not every argument of a function is necessarily used during every function execution, meaning that some arguments may remain unevaluated (as they are never forced), and there are a few reasons for this.
+\item First, we wanted to ensure that \typetracer captured types for
+  unevaluated arguments.  Unfortunately, not every argument of a function is
+  necessarily used during every function execution, meaning that some
+  arguments may remain unevaluated (as they are never forced), and there are
+  a few reasons for this.
 
@@
   flow of this particular execution (e.g. if the argument is only needed in
-one branch of a conditional expression).  This mainly occurs when arguments have default
-values for some parameters that are not critical to every execution of the
-function.  In these cases, we still want to collect the type of the default
-value.
+  one branch of a conditional expression).  This mainly occurs when
+  arguments have default values for some parameters that are not critical to
+  every execution of the function.  In these cases, we still want to collect
+  the type of the default value.
 
   Another possibility is that a value is used {\it without being forced}, as
-is the case when values are ``metaprogrammed''.
-This is common practice in e.g. the \code{tidyverse} ecosystem, as they use the metaprogramming capabilities of R to implement a DSL for data analysis.
+  is the case when values are ``metaprogrammed''.  This is common practice
+  in e.g. the \code{tidyverse} ecosystem, as they use the metaprogramming
+  capabilities of R to implement a DSL for data analysis.
 
@@
 impossible to know dynamically what all possible function behaviours are.
-We discuss the prevalence of such any-typed arguments in Section~\ref{sec:evaluation}.
+We discuss the prevalence of such any-typed arguments in
+Section~\ref{sec:evaluation}.
 
@@
   post-processing.  
-  % These longjumps occur quite often, particularly when S3 dispatch occurs, as the R implementation optimizes returns from dispatched functions by initiating a longjump.
-
 \end{itemize}

### 2020-05-14 paper/main.tex
before refocus

--- 2020-05-14 paper/main.tex
@@
 is the case when values are ``metaprogrammed''.
-This is common practice in e.g. the \code{tidyverse} ecosystem, as they use the metaprogramming capabilities of R to implement a DSL for data analysis.
-
-To tackle this, we make an {\it initial guess} of function argument types on
-function entry. This collects types for argument default values, and
-collects an ``unused'' type for unevaluated expressions.  If an argument
+This is common practice in e.g.
+the \code{tidyverse}~\footnote{\AT{either talk about it here or in background}} ecosystem, as they use the metaprogramming capabilities of R to implement a DSL for data analysis.
+
+To tackle possibly unevaluated arguments, we make an {\it initial guess} of function argument types on
+function entry. Specifically, we collect types for argument default values, and
+ascribe an ``unused'' type for unevaluated expressions.  If an argument
 promise is later forced, we simply update the recorded type for the
@@
 \item Second, we wanted to account for formal arguments (arguments which are
-  named in the function definition) that were {\it not} passed to the
-  function. This represents cases where a function was not called with all of its
+  named in the function definition) which received no values when the function was called. This represents cases where a function was not called with all of its
   arguments, and no default value was specified for missing arguments. 
@@
 
-There are two obvious ways to deal with missing arguments: type them as bot
-or as top.  As we are performing a dynamic analysis based on package test
-code, we conservatively type these arguments as top, or \code{any}, as It is
+There are two obvious ways to deal with missing arguments: type them as $\bot$
+or as $\top$.  As we are performing a dynamic analysis based on package test
+code, we conservatively type these arguments as $\top$, or \code{any}, as It is
 impossible to know dynamically what all possible function behaviours are.
@@
 
-\item Third, we had to contend with R's C implementation sometimes
+\item Third, we had to contend with R's backend sometimes
   performing longjumps, which discards current R calls from the call stack.
@@
 is a very efficient instrumentation framework for R, which is built on a
-modified version of the R runtime.  Our dynamic type analysis only needs to
-be run once, ever---from the analysis results, we synthesize function
-signatures which are passed along to later type-checking tools which do not
-rely on a heavily modified runtime.
+modified version of the R runtime. 
+% Our dynamic type analysis only needs to be run once, ever---from the analysis results, we synthesize function signatures which are passed along to later type-checking tools which do not rely on a heavily modified runtime.
 
@@
 guesses of argument types (as discussed in
-Section~\ref{subsec:typetracer-challenges}).  Further, the type of dispatch
-is recorded on the {\tt \_entry} variants: recall that R can perform single or multiple dispatch on function arguments depending on their class.
-We collect dispatch information in order to analyze dispatch usage patterns.
-
-The main work is done in {\tt promise\_force\_exit}.  As arguments to
+Section~\ref{subsec:typetracer-challenges}).  
+Also, recall that R can perform single or multiple dispatch on function arguments depending on their class.
+We collect such dispatch information the {\tt \_entry} variants.
+
+The primary functionality of \typetracer is in {\tt promise\_force\_exit}.  As arguments to
 functions in R are wrapped in promises, we delay our reflection to promise
 force time.  In R, promises are essentially objects with value and
-expression slots.  When a promise is created for a particular expression,
+expression slots. \AT{Put this as part of R-dyntrace background:} When a promise is created for a particular expression,
 that expression is merely put into the expression slot of a promise object,
@@
 the value slot of the promise.  So to get the type of a function argument,
-we check the value slot of the argument promise---perhaps recursively if the
+we check the value slot of the argument promise---recursively if the
 value itself is a promise.
@@
 
+\AT{Maybe further discuss how we can create many candidate TS from the data.}
+
 The type information we collect is far more detailed than the types we
-actually generate, and so the next stage of our inference process is to
+actually generate (which we discuss in Section~\ref{sec:types}), and so the next stage of our inference process is to
 distill the rich type information into simpler, manageable types.  For each
@@
 
+Thus far, we have discussed \typetracer and our strategies to take the data from the dynamic analysis and generate function types.
+This work also presents a framework for specifying these types in R code, as well as generating contract checks for these types.
+We dub this tool \contractr, discussed next.
+
 %
@@
 \contractr is an R package that implements our type system for the R
-package ecosystem. It adds contracts to package functions for type-checking
+package ecosystem. For functions with type annotations, it adds contracts for type-checking
 arguments and results on function calls. On contract failure, it reports a type
@@
 assessing the quality of these type signatures on XXX packages during the
-evaluation phase.
+evaluation phase. We discuss this further in Section~\ref{sec:evaluation}.
 
@@
 
-This example also illustrates an issue we grappled with when dealing with structs.
-On line 8, we see the call \code{data(sweetpotato)}.
+Unfortunately, the inclusion of structs complicated our analysis by polluting the data, illustrated on line 8 with the call \code{data(sweetpotato)}.
 It is quite common for R example code to use some built-in data sets, such as \code{sweetpotato}, to show off some package functionality with ``real data''.
-We found that the types of functions which use these built-in data sets (such as \code{aov}) became polluted with the names from the example sets.
+When \typetracer is configured to detect structs we saw a marked increase in misrepresentative types, as the types of functions which use these data sets (such as \code{aov} in Figure~\ref{fig:struct-ex}) became polluted with the names from the example sets.
+We implemented a unification strategy for struct types, removing them if they co-occurred with list types, and turning them into lists if two co-occurring struct types shared no common names, but even so we found ourselves with function signatures that were full of noise introduced by the struct names.
+
+We still believe struct types to be useful to R programmers, and so our language supports the type and \contractr will check them, though we configured \typetracer to ignore them as they made the data substantially less useful.
+On the positive side, we found that \AT{get the number, but often} structs-typed values had a {\it class}, which helped motivate our inclusion of class-like types in our type annotation language.
+Classes are discussed in more detail in Section~\ref{subsec:classes}.
 
 % We also thought to encode data frames as structs, as they are essentially lists with named columns of homogeneous values, but it was difficult to draw any meaningful conclusions from the data as unfortunately structs introduced a lot of noise into our analysis.
-
-When \typetracer is configured to detect structs we saw a marked increase in misrepresentative types.
-We implemented a unification strategy for struct types, removing them if they co-occurred with list types, and turning them into lists if two co-occurring struct types shared no common names, but even so we found ourselves with function signatures that were full of noise introduced by the struct names.
-
-We still believe struct types to be useful to R programmers, and so our language supports the type and \contractr will check them, though we configured \typetracer to ignore them as they made the data substantially less useful.
-On the positive side, we found that \AT{get the number, but often} structs-typed values had a {\it class}, and so we decided to include class-based types.
-Classes are discussed in more detail in the next section, though we will quickly discuss data frames before focusing on them.
 
@@
 One way to deal with data frames is through the struct type, with a named field for each column of the data frame, but as mentioned previously structs introduced undue noise into \typetracer's analysis results.
-Further complicated issues for data frames is that many functions built to operate on them operate in a name-agnostic way.
+Further complicating data frames is that many functions built to operate on them operate in a name-agnostic way.
 For instance, the \code{tidyverse} package ecosystem allows programmers to pass column names to functions which operate on their data frames.
@@
 \begin{lstlisting}
-# Let's put some examples showing that data frames can get out of hand, and that
-# structs are inadequate.
+data(cars)
+
+# tidyverse: Filter rows where speed is > 20. 
+fast_cars_tidy <- filter(cars, speed > 20)
+
+# base R: Get all rows where the "speed" column has a value > 20.
+fast_cars_base <- cars[cars[, "speed"] > 20, ] # base example
+
 \end{lstlisting}
@@
 Doing justice to this type will be an endeavour in and of itself, requiring reconciliation of base R's \code{data.frame}, \code{tidyverse}'s \code{tibble}, and \code{data.table}'s \code{data.table}.
-\AT{We could talk about how metaprogramming is pervasive in dplyr, and how we would need to capture that, too, which is a huge undertaking.}
+Developing useful types for the \code{tidyverse} ecosystem may well involve
+
 It will also require proper treatment of function types, and perhaps even parametric types, which dynamic analyses such as \typetracer are unable to discern soundly.
@@
 \subsection{Classes in R}
+\label{subsec:classes}
 
@@
 %
-\subsubsection{Select Examples}
+\subsubsection{Subsumed Checks}
 

### 2020-05-14 paper/main.tex

@@
   flow of this particular execution (e.g. if the argument is only needed in
-one branch of a conditional expression).  This mainly occurs when arguments have default
-values for some parameters that are not critical to every execution of the
-function.  In these cases, we still want to collect the type of the default
-value.
+  one branch of a conditional expression).  This mainly occurs when
+  arguments have default values for some parameters that are not critical to
+  every execution of the function.  In these cases, we still want to collect
+  the type of the default value.
 

### 2020-05-14 paper/main.tex
evaluation organization
Add a few lines about R-dyntrace

@@
 In this section, we will describe our approach to data collection and
-analysis through \typetracer, a dynamic analysis built in the \rdt
+analysis through \typetracer, a dynamic analysis built in the \R-dyntrace
 framework.  We will then describe the process by which we distill the rich
@@
 function types.
+
+%
+\subsection{R-dyntrace}
+R-dyntrace is a framework for performing dynamic analysis of R. It was designed
+and used previously for a large-scale study of laziness in R~\cite{oopsla19} It
+is an instrumented R Virtual Machine based on GNU-R version 3.5.0. It exposes
+hooks from key points inside the R interpreter to which user defined callbacks
+can be attached for gathering information about the program state at those
+points. This information can be used to generate detailed program traces for
+online or offline analysis. R-dyntrace provides hooks for function entry and
+exit, S3/S4 dispatch, non-local jumps (because of longjumps in the R interpreter
+on exceptions and other non-local returns), promise creation and forcing,
+variable definition and mutation and garbage collection. garbage collection
+entry and exit, function entry and exit.
 
@@
 
+\AT{Aviral}
+
 We analyzed the type signatures generated by \typetracer in order to build up a profile of our type annotation language.
@@
 %
-\subsubsection{Subsumed Checks}
-
-%
-%
-%
-%
-\subsection{Large-Scale Evaluation}
+%
+%
+\subsection{Evaluation of Types from Traces}
 \label{subsec:eval}
+
+\AT{Me}
 
@@
 %
+\subsection{Usefulness of the Type Checking Framework}
+
+\AT{Filip}
+
+%
+%
+%
+%
 %

### 2020-05-14 paper/main.tex
minor eval

--- 2020-05-14 paper/main.tex
@@
 %
-\subsubsection{Subsumed Checks}
-
-%
-%
-%
-%
-\subsection{Large-Scale Evaluation}
+%
+%
+\subsection{Evaluation of Types from Traces}
 \label{subsec:eval}
 
-To measure the effectiveness of our type annotation language, we designed an experiment pitting the types generated by a package's own testing code against the testing code of the package's clients.
+\AT{Me}
+
+In the last section, we evaluated our type annotation language based on a set of function types that were automatically generated by \typetracer.
+Now, we're interested in knowing how representative our trace-generated types are of the functions they were generated for.
+To measure this, we conducted a large-scale experiment: for each package in the corpus discussed in Section~\ref{sec:corpus}, we ran \typetracer on its test, example, and vignette code, obtaining types for package functions, and then subsequently ran the test, example, and vignette code of each of the package's reverse dependencies (or clients).
 We collected failed contract assertions in order to analyze them and perhaps identify weaknesses in our type annotation language.
-
-%
-%
-\subsubsection{Experimental Design}
-
-As mentioned earlier in this section, we configured \typetracer to infer types for all client-visible functions for each of the 412 packages in our corpus.
-We loaded each of these inferred types into \contractr, and ran all example code from the reverse dependencies of these popular packages.
-\AT{As discussed in genthat}, running R code can be tricky, as runnable code may depend on e.g.
-data present in a hard-coded directory structure, and we employed similar strategies to ensure that we ran as much code as possible.
-\AT{Maybe Filip can write a few sentences on this, or write me a message and I can work it in.}
-
-\contractr was equipped to provide a list of all failed contract assertions, complete with diagnostic information (including a stack trace) to help us make sense of the failed assertion.
 

### 2020-05-14 paper/main.tex
intro

@@
 
-In order to do this, we need to understand the degree of polymorphism
-present in R code, that is to say, how programmers leverage the dynamic
-nature of R to write code that can accept arguments of different types.
-This understanding will drive our design.
-
-We propose to capture the degree of polymorphism present in R by means
-of a dynamic analysis of widely used libraries. For each function call we
-can record the types of its arguments and of its return value. This allows
-us to observe how many different combination of types are accepted by any
-given function. Unlike many other languages, R has a carefully curated
-software repository called CRAN. To be deposited in CRAN, a package must
-come with sample dataset, tests and executable use-cases. As part of normal
-operations these tests are run regularly and failing packages are removed.
-This allowed us to have access to \PACKAGES libraries and about an order of
-magnitude more runnable scripts that exercise those libraries.
-
-The contributions of this paper are thus as follows:
+This paper focuses on a simpler problem than an entire type system, instead,
+we limit the scope of our investigation to giving types to function
+signatures. In order to do this, we design a simple type language, one that
+matches the R data types but omits features such as parametric polymorphism
+and subtyping between user-defined data types. We then extract type
+signature from execution traces of a corpus of widely used libraries.  This
+allows us to see how far one can get with the simple type language and
+identify limitations of our this particular design. We validate the
+robustness of the extracted type signature by implementing a contact system
+that weaves the types around their respective functions, and use a large
+number of clients of the target packages for validation. The contract system
+can be used by the R community to experiment with our type signatures and to
+replace ad hoc error checking code.
+
+To sum up, our paper makes the following main contributions:
 \begin{itemize}
-\item A large-scale analysis of the polymorphism present in function
-  signatures of \PACKAGES widely used and actively maintained R packages.
-\item A tracing and analysis pipeline that extends a previously published
-  test generation tool named \genthat.
-\item A set of type annotations that captures how programmers use types
-in existing R code.
+\item We implemented tooling to automatically extract type signatures from R
+  functions and to intrument R functions with checks based on their declared
+  types. The tooling is robust and scalable to the entire R language.
+\item We carried out a large-scale analyis of corpus of \PACKAGES widely
+  used and actively maintained libraries to extract function type signatures
+  validated the robustness of the inferred type signatures against XXX
+  programs that use those functions.
+\item We report on the appropriateness and usefulness of a simpe type
+  language for the R programming language.
 \end{itemize}
 
-One threat to validity of our work is that we rely on dynamic analysis, so
-our conclusions are only as good as the coverage of the possibly function
-behaviors. Previous work~\cite{issta18} reported that running all the
-scripts that come with CRAN packages gives, on average, 68\% test coverage.
-We attempted to mitigate the threat coming from the fact that only part of
-the code is being exercised by manual analysis. It would be reasonable to
-ask for confirmation of the data by static analysis of the code, but sound
-static analysis of R is difficult because of the extensive use of reflective
-features such as \code{eval} and of the ability to redefine the meaning of
-operators such as \code{+} and \code{if}.  Another threat to validity is
-that we only have access to code that has been deposited in the CRAN
-repository. While this may bias our findings towards code written to be
-reusable and, possibly, better engineered than typical user code. This is
-also the code that would most benefit from type annotations.
-
+Our tools are open source and publicly available on GitHub (link anonymize),
+all results in this paper are reproducible and will be submitted for
+artifact evaluation should this paper be accepted.
 %
@@
 
-Over the years, programmers have found the need for a richer type
-structure and have added {\it attributes}. The best way to think of attributes is
-as an optional map from name to values that can be attached to any object.
+Over the years, programmers have found the need for a richer type structure
+and have added {\it attributes}. The best way to think of attributes is as
+an optional map from name to values that can be attached to any object.
 Attributes are used to encode various type structures. They can be queried
-with functions such as \k{attributes} and \k{class}.
-The addition of attributes lets programmers extend the set of types by
-tagging data with user-defined attributes. For example, one could define a
-vector of four values, \code{x<-c(1,2,3,4)} and then attach the attribute
-\k{dim} with a pair of numbers as value: \code{attr(x,"dim")<-c(2,2)}.  From
-that point, arithmetic functions will treat \k{x} as a 2x2 matrix. Another
-attribute that can be set is the \k{class}.  This attribute can be bound to
-a list of class names. For instance, \code{class(x)<-"human"}, set the class
-of \k{x} to be \k{human}.  Attributes are thus used for object-oriented
+with functions such as \k{attributes} and \k{class}.  The addition of
+attributes lets programmers extend the set of types by tagging data with
+user-defined attributes. For example, one could define a vector of four
+values, \code{x<-c(1,2,3,4)} and then attach the attribute \k{dim} with a
+pair of numbers as value: \code{attr(x,"dim")<-c(2,2)}.  From that point,
+arithmetic functions will treat \k{x} as a 2x2 matrix. Another attribute
+that can be set is the \k{class}.  This attribute can be bound to a list of
+class names. For instance, \code{class(x)<-"human"}, set the class of \k{x}
+to be \k{human}.  Attributes are thus used for object-oriented
 programming. The S3 object system support single dispatch on the class of
@@
 
-To tackle possibly unevaluated arguments, we make an {\it initial guess} of function argument types on
-function entry. Specifically, we collect types for argument default values, and
-ascribe an ``unused'' type for unevaluated expressions.  If an argument
-promise is later forced, we simply update the recorded type for the
-argument.
+To tackle possibly unevaluated arguments, we make an {\it initial guess} of
+function argument types on function entry. Specifically, we collect types
+for argument default values, and ascribe an ``unused'' type for unevaluated
+expressions.  If an argument promise is later forced, we simply update the
+recorded type for the argument.
 
 \item Second, we wanted to account for formal arguments (arguments which are
-  named in the function definition) which received no values when the function was called. This represents cases where a function was not called with all of its
-  arguments, and no default value was specified for missing arguments. 
-  We record a ``missing'' type for such argument.
+  named in the function definition) which received no values when the
+  function was called. This represents cases where a function was not called
+  with all of its arguments, and no default value was specified for missing
+  arguments.  We record a ``missing'' type for such argument.
 
@@
 modified version of the R runtime.
-% Our dynamic type analysis only needs to be run once, ever---from the analysis results, we synthesize function signatures which are passed along to later type-checking tools which do not rely on a heavily modified runtime.
 
@@
 guesses of argument types (as discussed in
-Section~\ref{subsec:typetracer-challenges}).  
-Also, recall that R can perform single or multiple dispatch on function arguments depending on their class.
-We collect such dispatch information the {\tt \_entry} variants.
-
-The primary functionality of \typetracer is in {\tt promise\_force\_exit}.  As arguments to
-functions in R are wrapped in promises, we delay our reflection to promise
-force time.  In R, promises are essentially objects with value and
-expression slots. \AT{Put this as part of R-dyntrace background:} When a promise is created for a particular expression,
-that expression is merely put into the expression slot of a promise object,
-and a special {\it unbound} value is loaded into the value slot (unless some
+Section~\ref{subsec:typetracer-challenges}).  Also, recall that R can
+perform single or multiple dispatch on function arguments depending on their
+class.  We collect such dispatch information the {\tt \_entry} variants.
+
+The primary functionality of \typetracer is in {\tt promise\_force\_exit}.
+As arguments to functions in R are wrapped in promises, we delay our
+reflection to promise force time.  In R, promises are essentially objects
+with value and expression slots. \AT{Put this as part of R-dyntrace
+  background:} When a promise is created for a particular expression, that
+expression is merely put into the expression slot of a promise object, and a
+special {\it unbound} value is loaded into the value slot (unless some
 default value is specified, wherein it is loaded into the value slot---this
@@
 the value slot of the promise.  So to get the type of a function argument,
-we check the value slot of the argument promise---recursively if the
-value itself is a promise.
+we check the value slot of the argument promise---recursively if the value
+itself is a promise.
 
@@
 
-\AT{Maybe further discuss how we can create many candidate TS from the data.}
-
 The type information we collect is far more detailed than the types we
-actually generate (which we discuss in Section~\ref{sec:types}), and so the next stage of our inference process is to
-distill the rich type information into simpler, manageable types.  For each
-unique observed call trace, we translate the representation of the type of
-each of the function arguments and returns according to the design of our
-type system: for instance, we might have determined that a function returned
-a vector of integer with 4 elements, with names, that was free of NA values,
-in which case we would write out \code{integer[]} as the type.
+actually generate (which we discuss in Section~\ref{sec:types}), and so the
+next stage of our inference process is to distill the rich type information
+into simpler, manageable types.  For each unique observed call trace, we
+translate the representation of the type of each of the function arguments
+and returns according to the design of our type system: for instance, we
+might have determined that a function returned a vector of integer with 4
+elements, with names, that was free of NA values, in which case we would
+write out \code{integer[]} as the type.
 
@@
 consolidate its multiple call traces into a single function type.  We do
-this by combining the observed types of each argument into one type according to subtyping rules that are part of our
-design.  The precise subtyping rules we used will be discussed further in
-Section~\ref{sec:typesystemdesign}.
-If we cannot consolidate multiple observed argument types into a single type, we simply union the disparate types together.
-
-While our analysis collects more detailed information than necessary for type generation, the information proved useful for building up a usage profile of the language. 
-This profile informed our language design: based on the rich type information we collected, we were able to determine types that would be useful for capturing the observed behaviour.
-We discuss this iterative, data-driven language design process in Section~\ref{types}, highlighting the reasoning behind our type system design.
-
-Thus far, we have discussed \typetracer and our strategies to take the data from the dynamic analysis and generate function types.
-This work also presents a framework for specifying these types in R code, as well as generating contract checks for these types.
-We dub this tool \contractr, discussed next.
-
-%
+this by combining the observed types of each argument into one type
+according to subtyping rules that are part of our design.  The precise
+subtyping rules we used will be discussed further in
+Section~\ref{sec:typesystemdesign}.  If we cannot consolidate multiple
+observed argument types into a single type, we simply union the disparate
+types together.
+
+While our analysis collects more detailed information than necessary for
+type generation, the information proved useful for building up a usage
+profile of the language.  This profile informed our language design: based
+on the rich type information we collected, we were able to determine types
+that would be useful for capturing the observed behaviour.  We discuss this
+iterative, data-driven language design process in Section~\ref{types},
+highlighting the reasoning behind our type system design.
+
+Thus far, we have discussed \typetracer and our strategies to take the data
+from the dynamic analysis and generate function types.  This work also
+presents a framework for specifying these types in R code, as well as
+generating contract checks for these types.  We dub this tool \contractr,
+discussed next.
+
 %
@@
 %
-%
-%
-%
 \section{Type Language Design}
@@
 an iterative process: initial analysis suggested a certain set of types,
-which we tried and implemented, and reanalysis suggested improvements,
-and so on.  In this section, we will present the final type annotation language design, 
-which we believe captures the essence of R.  We will end the section with a formal type 
-language grammar, and leading up to that we will
-touch on design decisions and weigh our choices against alternatives,
-speaking to the strengths and weaknesses of our chosen design.
-We begin with basic, primitive types.
+which we tried and implemented, and reanalysis suggested improvements, and
+so on.  In this section, we will present the final type annotation language
+design, which we believe captures the essence of R.  We will end the section
+with a formal type language grammar, and leading up to that we will touch on
+design decisions and weigh our choices against alternatives, speaking to the
+strengths and weaknesses of our chosen design.  We begin with basic,
+primitive types.
 
@@
 A common primitive usage pattern is related to the presence of NAs:
-\AT{often}, programmers explicitly check for NAs in primitive vectors,
-possibly sanitizing them if they are present.  NAs complicate practically every computation
-using primitive vectors, a fact that many programmers appear to be acutely
-aware of. For instance, consider the code in Figure~\ref{fig:na-example}.
+programmers explicitly check for NAs in primitive vectors, possibly
+sanitizing them if they are present.  NAs complicate practically every
+computation using primitive vectors, a fact that many programmers appear to
+be acutely aware of. For instance, consider the code in
+Figure~\ref{fig:na-example}.
 
@@
 
-The \code{binom} package is a popular R package for computing confidence intervals for binomial experiments, and the \code{binom.profile} function does so using the profile likelihood, the details of which are not relevant here.
-This code snippet highlights a data sanitization pattern appearing at the beginning of the function: the programmer first binds the vectors into a matrix (line 2), with one column for each of \code{x} and \code{n}, then finds rows where both columns are not NA (line 3), then extracts non-NA values and stores them into \code{x} and \code{n} respectively (lines 4-5).
-Importantly, this process is entirely quiet, and the user of \code{binom.profile} will not be notified if their data happens to contain NAs, and the NAs will be silently ignored. 
-In reality, the presence of NAs in data could signify some sort of data corruption, and quietly ignoring issues in data could lead users (read: statisticians) to draw flawed conclusions from erroneous data.
-
-To combat this, we introduce a modifier on primitive types: the type \code{^T} signifies that primitive type \code{T} might contain an NA value, while \code{T} signifies that \code{T} is NA-free. 
-The NA-free case has a shorter syntax as NA-free vectors are \AT{more common} than the alternative.
-If the \code{binom} developers were to annotate arguments \code{x} and \code{n} with such an NA-free type, \contractr will notify users when \code{binom.profile} is called with NA-full values, rather than just quietly ignoring them.
-
-Note that when combining call traces involving \code{^T} and \code{T} types, we subsume \code{T} types into \code{^T} types: if NA-free and NA-full vectors are passed to an argument, then we say that the argument can handle NA-full vectors.
-This is equivalent to saying \code{T <: ^T}.
+The \code{binom} package is a popular R package for computing confidence
+intervals for binomial experiments, and the \code{binom.profile} function
+does so using the profile likelihood, the details of which are not relevant
+here.  This code snippet highlights a data sanitization pattern appearing at
+the beginning of the function: the programmer first binds the vectors into a
+matrix (line 2), with one column for each of \code{x} and \code{n}, then
+finds rows where both columns are not NA (line 3), then extracts non-NA
+values and stores them into \code{x} and \code{n} respectively (lines 4-5).
+Importantly, this process is entirely quiet, and the user of
+\code{binom.profile} will not be notified if their data happens to contain
+NAs, and the NAs will be silently ignored.  In reality, the presence of NAs
+in data could signify some sort of data corruption, and quietly ignoring
+issues in data could lead users (read: statisticians) to draw flawed
+conclusions from erroneous data.
+
+To combat this, we introduce a modifier on primitive types: the type
+\code{^T} signifies that primitive type \code{T} might contain an NA value,
+while \code{T} signifies that \code{T} is NA-free.  The NA-free case has a
+shorter syntax as NA-free vectors are \AT{more common} than the alternative.
+If the \code{binom} developers were to annotate arguments \code{x} and
+\code{n} with such an NA-free type, \contractr will notify users when
+\code{binom.profile} is called with NA-full values, rather than just quietly
+ignoring them.
+
+Note that when combining call traces involving \code{^T} and \code{T} types,
+we subsume \code{T} types into \code{^T} types: if NA-free and NA-full
+vectors are passed to an argument, then we say that the argument can handle
+NA-full vectors.  This is equivalent to saying \code{T <: ^T}.
 
@@
 
-Initial analyses of our data revealed that programmers \AT{often} use scalars (as explicitly as they can, anyway), and often do high-level dimensionality checks on their data.
-To illustrate, consider the code in Figure~\ref{fig:scalar-vector-example}.
+Initial analyses of our data revealed that programmers \AT{often} use
+scalars (as explicitly as they can, anyway), and often do high-level
+dimensionality checks on their data.  To illustrate, consider the code in
+Figure~\ref{fig:scalar-vector-example}.
 
@@
 
-The \code{hankel.matrix} function takes two arguments (as described in the comment in the code), and returns a Hankel matrix, which essentially is a matrix where the skew-diagonals have a desirable property.
-The function body has a number of interesting checks:
-on lines 6-7, \code{n} undergoes an integer ``type check'';
-on lines 8-9, \code{x} undergoes a vector ``type check'';
-on line 11, \code{n} is assumed to be a scalar, as if it were a vector, the code \code{m < n} would essentially check \code{m < n[i]} for each element in \code{n}, returning a vector of comparison results, and the conditional of an if must be a scalar else R generates a warning.
-
-These checks exemplify functionality that we would like to include in our type annotation language, namely, we would like our type language to differentiate between scalars and vectors.
-To that end, we introduce the following types into our type language:
-for some primitive type \code{T}, the user may modify \code{T} with a modifier \code{[]} to indicate that it is a vector of \code{T}.
-Perhaps surprisingly, we found that scalars \AT{occurred more frequently} than vectors, and for that reason we chose the shorter scalar type syntax as the default.
-The NA-free modifier works as expected: \code{^T[]} indicates a vector of \code{T} that is free of NAs, while \code{^T} indicates a scalar which is not NA.
+The \code{hankel.matrix} function takes two arguments (as described in the
+comment in the code), and returns a Hankel matrix, which essentially is a
+matrix where the skew-diagonals have a desirable property.  The function
+body has a number of interesting checks: on lines 6-7, \code{n} undergoes an
+integer ``type check''; on lines 8-9, \code{x} undergoes a vector ``type
+check''; on line 11, \code{n} is assumed to be a scalar, as if it were a
+vector, the code \code{m < n} would essentially check \code{m < n[i]} for
+each element in \code{n}, returning a vector of comparison results, and the
+conditional of an if must be a scalar else R generates a warning.
+
+These checks exemplify functionality that we would like to include in our
+type annotation language, namely, we would like our type language to
+differentiate between scalars and vectors.  To that end, we introduce the
+following types into our type language: for some primitive type \code{T},
+the user may modify \code{T} with a modifier \code{[]} to indicate that it
+is a vector of \code{T}.  Perhaps surprisingly, we found that scalars
+\AT{occurred more frequently} than vectors, and for that reason we chose the
+shorter scalar type syntax as the default.  The NA-free modifier works as
+expected: \code{^T[]} indicates a vector of \code{T} that is free of NAs,
+while \code{^T} indicates a scalar which is not NA.
 
@@
 
-Matrices represent another layer of complexity on top of vectorized primitives.
-Internally, matrices are simply vectors with class ``matrix'' and a \code{dims} attribute indicating matrix dimensions, and while not codified in the language semantics, many internal functions will coerce vectors to matrices automatically.
-Relying on automatic coercion is unwise, and many functions check to ensure that their arguments are matrices.
-For instance, consider the code in Figure~\ref{fig:matrix-example}.
+Matrices represent another layer of complexity on top of vectorized
+primitives.  Internally, matrices are simply vectors with class ``matrix''
+and a \code{dims} attribute indicating matrix dimensions, and while not
+codified in the language semantics, many internal functions will coerce
+vectors to matrices automatically.  Relying on automatic coercion is unwise,
+and many functions check to ensure that their arguments are matrices.  For
+instance, consider the code in Figure~\ref{fig:matrix-example}.
 
@@
 
-The \code{rowWeightedMeans} function calculates the weighted means of rows or columns of a matrix \code{x}.
-This function exhibits good practice in type checking arguments, rightly producing a message to the user indicating that an explicit matrix should be passed.
-
-To capture cases like this, we include a \code{class<matrix>} type in our type annotation language.
-If \code{rowWeightedMeans} were outfitted with such a type, \contractr can perform the check present in this code automatically, notifying the users passing ill-typed arguments.
+The \code{rowWeightedMeans} function calculates the weighted means of rows
+or columns of a matrix \code{x}.  This function exhibits good practice in
+type checking arguments, rightly producing a message to the user indicating
+that an explicit matrix should be passed.
+
+To capture cases like this, we include a \code{class<matrix>} type in our
+type annotation language.  If \code{rowWeightedMeans} were outfitted with
+such a type, \contractr can perform the check present in this code
+automatically, notifying the users passing ill-typed arguments.
 
@@
 
-Thus far we've seen instances where we introduce types that subsume existing argument validation code, but our intention is not to subsume {\it all} such code.
-As an example of something we {\it do not} want to do, consider the code in Figure~\ref{fig:general-validation-example}.
+Thus far we've seen instances where we introduce types that subsume existing
+argument validation code, but our intention is not to subsume {\it all} such
+code.  As an example of something we {\it do not} want to do, consider the
+code in Figure~\ref{fig:general-validation-example}.
 
@@
 
-This code is from later in the same \code{rowWeightedMeans} function discussed previously.
-Here, the length of the weights vector \code{w} is being checked for equality with the number of columns in the matrix \code{x}. 
-Specifying this relationship as a type amounts to {\it dependent typing}, which we believe too complex to retrofit onto a dynamic data science language such as R.  
-Ultimately, we decided to leave these kinds of checks up to the programmer.
-\AT{Can we say more here?}
+This code is from later in the same \code{rowWeightedMeans} function
+discussed previously.  Here, the length of the weights vector \code{w} is
+being checked for equality with the number of columns in the matrix
+\code{x}.  Specifying this relationship as a type amounts to {\it dependent
+  typing}, which we believe too complex to retrofit onto a dynamic data
+science language such as R.  Ultimately, we decided to leave these kinds of
+checks up to the programmer.
 
@@
 
-Separate from the vectorized primitives in R are lists:
-Whereas vectors are homogeneous and restricted to primitive types, lists can contain arbitrarily-typed data.  
-Further, lists and vectors have a different syntax for accessing elements, where lists are indexed using \code{[[} and vectors using \code{[}.
-R even builds on lists to create a more complex data structure known as the {\it data frame}, an essential data structure to R's functionality.
-
-We are faced with several possible designs for list types.  
-Chief concerns include the heterogeneity of list elements and the presence list element names, which allow list elements to be accessed in a struct-like manner.  
-% \AT{Rarely?}, programmers even write code expecting lists of a particular pre-defined length.  
-The next sections highlight some important design decisions for our list types.
+Separate from the vectorized primitives in R are lists: Whereas vectors are
+homogeneous and restricted to primitive types, lists can contain
+arbitrarily-typed data.  Further, lists and vectors have a different syntax
+for accessing elements, where lists are indexed using \code{[[} and vectors
+    using \code{[}.  R even builds on lists to create a more complex data
+      structure known as the {\it data frame}, an essential data structure
+      to R's functionality.
+
+We are faced with several possible designs for list types.  Chief concerns
+include the heterogeneity of list elements and the presence list element
+names, which allow list elements to be accessed in a struct-like manner.
+The next sections highlight some important design decisions for our list
+types.
 
@@
 
-\AT{Worth a code example?}
-
-The ease with which vectors may be transformed into lists is perhaps indicative of a relationship between the two.
-In R, all vectors can be converted to lists with a simple call to \code{as.list}, and lists to vectors with \code{unlist}, though care must be taken when unlisting a non-homogeneous list, as coercions will be performed where possible, and the function will simply do nothing if coercion is impossible.
-That said, lists and vectors have different access syntax, so vectors cannot stand-in for lists without the \code{as.list} call above.
-
-In sum, we opted not to include a subtyping relationship between vectors and lists, keeping both types separate.
-In practice, \AT{how often do \code{T[] | list<T>} types occur?}.
-During our evaluation we also found it rare that programmers would attempt to pass list-typed values to vector-typed arguments and vice versa, and we will discuss this further in Section~\ref{sec:evaluation}.
+The ease with which vectors may be transformed into lists is perhaps
+indicative of a relationship between the two.  In R, all vectors can be
+converted to lists with a simple call to \code{as.list}, and lists to
+vectors with \code{unlist}, though care must be taken when unlisting a
+non-homogeneous list, as coercions will be performed where possible, and the
+function will simply do nothing if coercion is impossible.  That said, lists
+and vectors have different access syntax, so vectors cannot stand-in for
+lists without the \code{as.list} call above.
+
+In sum, we opted not to include a subtyping relationship between vectors and
+lists, keeping both types separate.  In practice, \AT{how often do \code{T[]
+    | list<T>} types occur?}.  During our evaluation we also found it rare
+that programmers would attempt to pass list-typed values to vector-typed
+arguments and vice versa, and we will discuss this further in
+Section~\ref{sec:evaluation}.
 
@@
 
-Environments in R are essentially lists with reference semantics: mutating a value in an environment is performed in-place, and that mutation can thus be shared with other code holding a reference to the environment.
-They are used primarily by the R runtime to store variables (their primary function is that of an execution environment), though they also often used by programmers either to themselves reflect on local, previous, or global scope, or simply because of their reference semantics.
-Environments are comprised of name, value pairs, and each environment has a reference to its parent environment.
-
-We thought to include detailed environment types in our type language by way of allowing environment types to define names and types of values in the environment.
-Unfortunately, this led to an undue explosion in the ``size'' of our types: 
-often, environments had hundreds of values, particularly those which were obtained from R (by e.g. passing a reference to a local scope).
+Environments in R are essentially lists with reference semantics: mutating a
+value in an environment is performed in-place, and that mutation can thus be
+shared with other code holding a reference to the environment.  They are
+used primarily by the R runtime to store variables (their primary function
+is that of an execution environment), though they also often used by
+programmers either to themselves reflect on local, previous, or global
+scope, or simply because of their reference semantics.  Environments are
+comprised of name, value pairs, and each environment has a reference to its
+parent environment.
+
+We thought to include detailed environment types in our type language by way
+of allowing environment types to define names and types of values in the
+environment.  Unfortunately, this led to an undue explosion in the ``size''
+of our types: often, environments had hundreds of values, particularly those
+which were obtained from R (by e.g. passing a reference to a local scope).
 These types were difficult to parse visually and did not feel informative.
-Further, environments \AT{did not occur often}, and giving them a richer type was not worth the increase in the visual complexity of our types.
-
-%
-%
-% \subsubsection{Forgoing Tuples, or Short, Constant Length Lists}
-
-% We briefly entertained the possibility of including a \code{tuple} type, which would have looked like e.g. \code{tuple<int, int>} for a list of two integers.  
-% When configured to detect tuples, our analysis did indeed pick up \AT{a number}, but post-processing and simplification revealed that in \AT{many} cases, they would be subsumed by co-occurring \code{list} types.
-% In other words, more often than not it just so happened that functions were being called with short lists.  
-% Of the cases that tuples were not subsumed by lists, \AT{often} there were \AT{tuples of different lengths}, implying that the shape of the argument was not essential.
-
-% \AT{I feel like this is the least important of the design decisions we made, so I'm leaving it for the end, we might not even want it.}
+Further, environments \AT{did not occur often}, and giving them a richer
+type was not worth the increase in the visual complexity of our types.
 
@@
 
-\AT{Do we want to talk about how list names are not unique?}
-
-We observed the use of consistently-named lists in \AT{a number of functions}.
-In R, ``fields'' of named lists can be accessed using the \code{$} operator, as illustrated in the code snippet in Figure~\ref{fig:struct-ex}.
+We observed the use of consistently-named lists in \AT{a number of
+  functions}.  In R, ``fields'' of named lists can be accessed using the
+\code{$} operator, as illustrated in the code snippet in
+Figure~\ref{fig:struct-ex}.
 
@@
 
-In the code snippet, the function \code{cv.model} takes an argument \code{x}, which we observed to always be a {\it linear model}, which is part of R's built-in statistics functionality.
-Internally, linear models are represented as lists with named elements, and one may access these named elements using the \code{$} syntax, as seen on lines 2-4 in the snippet.
-To describe these kinds of values, we included a struct type in our type system, allowing users to specify name, type pairs as part of list type definitions.
-
-Unfortunately, the inclusion of structs complicated our analysis by polluting the data, illustrated on line 8 with the call \code{data(sweetpotato)}.
-It is quite common for R example code to use some built-in data sets, such as \code{sweetpotato}, to show off some package functionality with ``real data''.
-When \typetracer is configured to detect structs we saw a marked increase in misrepresentative types, as the types of functions which use these data sets (such as \code{aov} in Figure~\ref{fig:struct-ex}) became polluted with the names from the example sets.
-We implemented a unification strategy for struct types, removing them if they co-occurred with list types, and turning them into lists if two co-occurring struct types shared no common names, but even so we found ourselves with function signatures that were full of noise introduced by the struct names.
-
-We still believe struct types to be useful to R programmers, and so our language supports the type and \contractr will check them, though we configured \typetracer to ignore them as they made the data substantially less useful.
-On the positive side, we found that \AT{get the number, but often} structs-typed values had a {\it class}, which helped motivate our inclusion of class-like types in our type annotation language.
-Classes are discussed in more detail in Section~\ref{subsec:classes}.
-
-% We also thought to encode data frames as structs, as they are essentially lists with named columns of homogeneous values, but it was difficult to draw any meaningful conclusions from the data as unfortunately structs introduced a lot of noise into our analysis.
-
-%
+In the code snippet, the function \code{cv.model} takes an argument
+\code{x}, which we observed to always be a {\it linear model}, which is part
+of R's built-in statistics functionality.  Internally, linear models are
+represented as lists with named elements, and one may access these named
+elements using the \code{$} syntax, as seen on lines 2-4 in the snippet.  To
+describe these kinds of values, we included a struct type in our type
+system, allowing users to specify name, type pairs as part of list type
+definitions.
+
+Unfortunately, the inclusion of structs complicated our analysis by
+polluting the data, illustrated on line 8 with the call
+\code{data(sweetpotato)}.  It is quite common for R example code to use some
+built-in data sets, such as \code{sweetpotato}, to show off some package
+functionality with ``real data''.  When \typetracer is configured to detect
+structs we saw a marked increase in misrepresentative types, as the types of
+functions which use these data sets (such as \code{aov} in
+Figure~\ref{fig:struct-ex}) became polluted with the names from the example
+sets.  We implemented a unification strategy for struct types, removing them
+if they co-occurred with list types, and turning them into lists if two
+co-occurring struct types shared no common names, but even so we found
+ourselves with function signatures that were full of noise introduced by the
+struct names.
+
+We still believe struct types to be useful to R programmers, and so our
+language supports the type and \contractr will check them, though we
+configured \typetracer to ignore them as they made the data substantially
+less useful.  On the positive side, we found that \AT{get the number, but
+  often} structs-typed values had a {\it class}, which helped motivate our
+inclusion of class-like types in our type annotation language.  Classes are
+discussed in more detail in Section~\ref{subsec:classes}.
+
 %
@@
 
-One of the most popular classes in R is the \code{data.frame} class \AT{we have numbers here, grab em}.
-Data frames and the derivative \code{tibble} (from the \code{tidyverse} ecosystem) and \code{data.table} (from the \code{data.table} package) types underpin a huge amount of data analysis in R.
-
-One way to deal with data frames is through the struct type, with a named field for each column of the data frame, but as mentioned previously structs introduced undue noise into \typetracer's analysis results.
-Further complicating data frames is that many functions built to operate on them operate in a name-agnostic way.
-For instance, the \code{tidyverse} package ecosystem allows programmers to pass column names to functions which operate on their data frames.
-In base R, idiomatic data frame use is to use string column names to select rows from the frame (unless only a single column is of interest, wherein the \code{$} syntax is appropriate). 
-We include some examples in Figure~\ref{fig:data-frames-bad} to illustrate the complexities.
+One of the most popular classes in R is the \code{data.frame} class \AT{we
+  have numbers here, grab em}.  Data frames and the derivative \code{tibble}
+(from the \code{tidyverse} ecosystem) and \code{data.table} (from the
+\code{data.table} package) types underpin a huge amount of data analysis in
+R.
+
+One way to deal with data frames is through the struct type, with a named
+field for each column of the data frame, but as mentioned previously structs
+introduced undue noise into \typetracer's analysis results.  Further
+complicating data frames is that many functions built to operate on them
+operate in a name-agnostic way.  For instance, the \code{tidyverse} package
+ecosystem allows programmers to pass column names to functions which operate
+on their data frames.  In base R, idiomatic data frame use is to use string
+column names to select rows from the frame (unless only a single column is
+of interest, wherein the \code{$} syntax is appropriate).  We include some
+examples in Figure~\ref{fig:data-frames-bad} to illustrate the complexities.
 
@@
 
-Doing justice to this type will be an endeavour in and of itself, requiring reconciliation of base R's \code{data.frame}, \code{tidyverse}'s \code{tibble}, and \code{data.table}'s \code{data.table}.
-Developing useful types for the \code{tidyverse} ecosystem may well involve
-
-It will also require proper treatment of function types, and perhaps even parametric types, which dynamic analyses such as \typetracer are unable to discern soundly.
-\AT{General words to convince the reader that data frames are, in fact, hard, and worthy of their own paper.}
+Doing justice to this type will be an endeavour in and of itself, requiring
+reconciliation of base R's \code{data.frame}, \code{tidyverse}'s
+\code{tibble}, and \code{data.table}'s \code{data.table}.  Developing useful
+types for the \code{tidyverse} ecosystem may well involve
+
+It will also require proper treatment of function types, and perhaps even
+parametric types, which dynamic analyses such as \typetracer are unable to
+discern soundly.
 
@@
 
-As discussed in Section~\ref{sec:R}, the R runtime has more than one notion of type.  
-The classes of function arguments are used by R as part of its dispatch mechanism, the most prevalent of which (\AT{we have that number, too!}) is S3 dispatch, in which R dispatches on the (S3) class of the first argument to a function, selecting the appropriate method to call for the class.
-To quickly illustrate, consider the code snippet in Figure~\ref{fig:dispatch-ex}.
+As discussed in Section~\ref{sec:R}, the R runtime has more than one notion
+of type.  The classes of function arguments are used by R as part of its
+dispatch mechanism, the most prevalent of which (\AT{we have that number,
+  too!}) is S3 dispatch, in which R dispatches on the (S3) class of the
+first argument to a function, selecting the appropriate method to call for
+the class.  To quickly illustrate, consider the code snippet in
+Figure~\ref{fig:dispatch-ex}.
 
@@
 
-Classes are very prevalent in R \AT{number}, and so we include a \code{class<class-name>} ... \AT{todo}
-\typetracer was outfit to report the class of values, but some care needed to be taken so as to not overreport class types, as {\it all} values have a class in R.
-The main classes we were interested in were matrix, array (higher dimensional matrices), and data.frames, in addition to any user-defined classes.
-In many cases \AT{number?}, values with a user-defined class are in reality lists with named elements, or structs.
-Including class types, and having \typetracer report them in place of structs, significantly reduced the noise in the analysis and allowed us to gather more focused data.
-To get a sense for this difference, consider the code in Figure~\ref{fig:class-vs-struct}.
+Classes are very prevalent in R \AT{number}, and so we include a
+\code{class<class-name>} ... \AT{todo} \typetracer was outfit to report the
+class of values, but some care needed to be taken so as to not overreport
+class types, as {\it all} values have a class in R.  The main classes we
+were interested in were matrix, array (higher dimensional matrices), and
+data.frames, in addition to any user-defined classes.  In many cases
+\AT{number?}, values with a user-defined class are in reality lists with
+named elements, or structs.  Including class types, and having \typetracer
+report them in place of structs, significantly reduced the noise in the
+analysis and allowed us to gather more focused data.  To get a sense for
+this difference, consider the code in Figure~\ref{fig:class-vs-struct}.
 
@@
 
-A value can have multiple classes in R.
-This is common practice in the \code{dplyr} package, which extends the functionality of data frames: a \code{data.frame} that passes through \code{dplyr} functions gains two additional classes (\code{tbl} and \code{tbl_df}), bringing the total to at least 3, in addition to whichever other classes that data frame happened to have.
-Note that when looking to perform S3 dispatch, R will try to match {\it any} of the classes of the first argument to a dispatched function, so each of these classes are important.
-
-We found these so-called multi-classes to be surprisingly common, making up \AT{occurrence} of the occurrences of class-typed values.
-One possible reason for this is that values can gain classes quite easily (e.g. the \code{dplyr} example discussed above), somewhat polluting the output of \typetracer.
-Nevertheless, we found \AT{number} of occurrences of monomorphic multi-class arguments, meaning that the function was always called with a multi-class value in that argument position.
-
-In the end, we support these multi-class types by allowing \code{class<...>} types to contain a list of class names, so for instance a \code{dplyr} \code{tibble} would have type \code{class<data.frame, tbl, tbl_df>} in our type language, and when performing related contract assertions \contractr will check that the value has every class that make up its type.
+A value can have multiple classes in R.  This is common practice in the
+\code{dplyr} package, which extends the functionality of data frames: a
+\code{data.frame} that passes through \code{dplyr} functions gains two
+additional classes (\code{tbl} and \code{tbl_df}), bringing the total to at
+least 3, in addition to whichever other classes that data frame happened to
+have.  Note that when looking to perform S3 dispatch, R will try to match
+{\it any} of the classes of the first argument to a dispatched function, so
+each of these classes are important.
+
+We found these so-called multi-classes to be surprisingly common, making up
+\AT{occurrence} of the occurrences of class-typed values.  One possible
+reason for this is that values can gain classes quite easily (e.g. the
+\code{dplyr} example discussed above), somewhat polluting the output of
+\typetracer.  Nevertheless, we found \AT{number} of occurrences of
+monomorphic multi-class arguments, meaning that the function was always
+called with a multi-class value in that argument position.
+
+In the end, we support these multi-class types by allowing \code{class<...>}
+types to contain a list of class names, so for instance a \code{dplyr}
+\code{tibble} would have type \code{class<data.frame, tbl, tbl_df>} in our
+type language, and when performing related contract assertions \contractr
+will check that the value has every class that make up its type.
 
@@
 dispatch mechanism, and indeed our class types roughly capture the S3
-classes of values.  We will not discuss R5 as it is still under
-development.
-
-For S4, we have an \code{s4} type to denote S4 values, though we did not elaborate on this type, and in practice S4 values have a class, and thus a \code{class<...>} type.  
-We stuck to simple class-based types for S4 since doing anything more would be immensely complicated, and outside of the scope of this work.
-For instance, the mechanics of S4 dispatch are more complex than for S3 (S4 dispatches on the class of {\it all} arguments), and users can define their own class hierarchies that we would need to incorporate in our type analysis and contract checking frameworks.
-Further, \AT{we found limited use of S4 during our analysis}, with S4 types being heavily used in a select subset of packages (\AT{such as ...}), some of which make heavy use of R's metaprogramming capabilities.
-Coming up with a type system that accounts for all of these factors and consolidates multiple object-orientation frameworks in a single language design is an interesting problem in and of itself, and we aim to extend our type annotation framework to account for this in future work.
-For now, S3 types suffice: S3 is the most common object system in R, with no sophisticated hierarchies, and its semantics are simple enough to be captured by our proposed design.
+classes of values.  We will not discuss R5 as it is still under development.
+
+For S4, we have an \code{s4} type to denote S4 values, though we did not
+elaborate on this type, and in practice S4 values have a class, and thus a
+\code{class<...>} type.  We stuck to simple class-based types for S4 since
+doing anything more would be immensely complicated, and outside of the scope
+of this work.  For instance, the mechanics of S4 dispatch are more complex
+than for S3 (S4 dispatches on the class of {\it all} arguments), and users
+can define their own class hierarchies that we would need to incorporate in
+our type analysis and contract checking frameworks.  Further, \AT{we found
+  limited use of S4 during our analysis}, with S4 types being heavily used
+in a select subset of packages (\AT{such as ...}), some of which make heavy
+use of R's metaprogramming capabilities.  Coming up with a type system that
+accounts for all of these factors and consolidates multiple
+object-orientation frameworks in a single language design is an interesting
+problem in and of itself, and we aim to extend our type annotation framework
+to account for this in future work.  For now, S3 types suffice: S3 is the
+most common object system in R, with no sophisticated hierarchies, and its
+semantics are simple enough to be captured by our proposed design.
 
@@
 
-\AT{Revisit this section since discussion of polymorphism in results.}
-
-Functions are first-class values in R.
-\typetracer found \AT{number} functions to be higher-order, having at least one function type among their arguments.
-They are markedly more common in a subset of packages, such as \AT{such as...}.
-By-and-large, these higher-order functions are similar to well-known higher-order functions like map, filter, and fold.
-
-The truth is that higher-order function types are best served by {\it parametric types} and parametric polymorphism, wherein type variables are used within a type declaration to specify type-based relationships between function arguments.
-Unfortunately, the nature of dynamic analysis is such that \typetracer cannot soundly detect parametric polymorphism, as without a static analysis of the function body it is impossible to say if function execution will halt given certain argument types, and we saw some examples of this in Figures~\ref{fig:scalar-vector-example}~and~\ref{fig:matrix-example}.
-Without parametric types, higher-order function types become noisy and fail to tell the complete picture.
-
-Ultimately, we configured \typetracer to report only top-level function types (i.e., detect if a value is a function or not).
-\AT{The data suggests that} most of the use of higher-order functions is through map-, filter-, and apply-like functions, whose behaviour is best captured with parametric polymorphism.
-Other common uses include higher-order functions operating on data frame columns (in packages such as \code{dplyr}), which would necessitate a more in-depth treatment of data frame types, which as we discussed is a huge endeavour in and of itself.
-We aim to extend our type annotation framework with parametric, higher-order, and detailed data frame types together in future work.
+Functions are first-class values in R.  \typetracer found \AT{number}
+functions to be higher-order, having at least one function type among their
+arguments.  They are markedly more common in a subset of packages, such as
+\AT{such as...}.  By-and-large, these higher-order functions are similar to
+well-known higher-order functions like map, filter, and fold.
+
+The truth is that higher-order function types are best served by {\it
+  parametric types} and parametric polymorphism, wherein type variables are
+used within a type declaration to specify type-based relationships between
+function arguments.  Unfortunately, the nature of dynamic analysis is such
+that \typetracer cannot soundly detect parametric polymorphism, as without a
+static analysis of the function body it is impossible to say if function
+execution will halt given certain argument types, and we saw some examples
+of this in
+Figures~\ref{fig:scalar-vector-example}~and~\ref{fig:matrix-example}.
+Without parametric types, higher-order function types become noisy and fail
+to tell the complete picture.
+
+Ultimately, we configured \typetracer to report only top-level function
+types (i.e., detect if a value is a function or not).  \AT{The data suggests
+  that} most of the use of higher-order functions is through map-, filter-,
+and apply-like functions, whose behaviour is best captured with parametric
+polymorphism.  Other common uses include higher-order functions operating on
+data frame columns (in packages such as \code{dplyr}), which would
+necessitate a more in-depth treatment of data frame types, which as we
+discussed is a huge endeavour in and of itself.  We aim to extend our type
+annotation framework with parametric, higher-order, and detailed data frame
+types together in future work.
 
@@
 
-One final, common pattern that we would like to highlight is the co-occurrence of \code{NULL} with other types of values.
-In Section~\ref{subsec:basictypes} we discussed NA, and how NAs represent missing data for the primitive types, and more generally R provides a \code{NULL} value that programmers use to represent missing data for other types.
-Programmers often use \code{NULL} as a default value for function arguments, and the code snippet in Figure~\ref{fig:null} exemplifies this.
+One final, common pattern that we would like to highlight is the
+co-occurrence of \code{NULL} with other types of values.  In
+Section~\ref{subsec:basictypes} we discussed NA, and how NAs represent
+missing data for the primitive types, and more generally R provides a
+\code{NULL} value that programmers use to represent missing data for other
+types.  Programmers often use \code{NULL} as a default value for function
+arguments, and the code snippet in Figure~\ref{fig:null} exemplifies this.
 
@@
 
-To capture this behaviour, we introduce a nullable type \code{? T} into our type language and contract checking framework.
-That way, programmers can succinctly ascribe types to null-by-default arguments (such as in the example in Figure~\ref{fig:null}).
-This also served to reduce the number of polymorphic arguments we detected:
-Without this type, \AT{X\%} of polymorphic argument positions were of the form \code{null | T}.
-Including a nullable type served to reduce observed polymorphism by \AT{Y\%}.
+To capture this behaviour, we introduce a nullable type \code{? T} into our
+type language and contract checking framework.  That way, programmers can
+succinctly ascribe types to null-by-default arguments (such as in the
+example in Figure~\ref{fig:null}).  This also served to reduce the number of
+polymorphic arguments we detected: Without this type, \AT{X\%} of
+polymorphic argument positions were of the form \code{null | T}.  Including
+a nullable type served to reduce observed polymorphism by \AT{Y\%}.
 
@@
 
-Finally, Figure~\ref{fig:types} presents our final type language design, and the grammar of types that \contractr is equipped to handle.
-\contractr will look for a top-level type declaration file in a package, with function types having the form $type \; ID \; T$, where $ID$ is the function name and $T$ is a function type.
-These function types have the form $\langle A_1 , ..., A_n\rangle \Rightarrow T$, where each of the $A_i$ arguments are either some type $T_i$, or the vararg type $...$.
-Recall that functions with a vararg parameter (\code{...}) can be called with arbitrarily many values, and extra values are lumped together in the vararg $...$ value.
-Internally, the type $...$ is implemented as a ${\bf list} \langle any \rangle$ type.
-
-We have discussed most of these types throughout this section, with some notable exceptions.
-${\bf expression}$, ${\bf language}$, ${\bf symbol}$, ${\bf externalptr}$, ${\bf pairlist}$, ${\bf S4}$, and ${\bf weakref}$ types occur \AT{rarely}, and are by-and-large uninteresting to most R programmers.
-\AT{Fill in details of these basic types.}
-% $S$ and $V$ are scalar and vector primitives, with vectors taking the form $S[]$.
-% The scalar type $S$ can be either a base type $B$ (one of the 6 primitives in R), or an NA-able base type (roughly, $\string^B$).
-One other type we did not discuss is the union type $T_1 \, | \, T_2$, which is supported by \contractr and generated by \typetracer when it cannot find some way to consolidate argument types (for example, if an argument is observed to be ${\bf chr}$ and ${\bf int}$, \typetracer will ascribe type ${\bf chr} \, | \, {\bf int}$ to it).
-And while not generated by \typetracer (recall discussion in Section~\ref{subsec:structs}), user-defined struct types have the form ${\bf struct}\langle F_1, ..., F_n \rangle$, with fields $F_i$ being name $:$ type pairs.
+Finally, Figure~\ref{fig:types} presents our final type language design, and
+the grammar of types that \contractr is equipped to handle.  \contractr will
+look for a top-level type declaration file in a package, with function types
+having the form $type \; ID \; T$, where $ID$ is the function name and $T$
+is a function type.  These function types have the form $\langle A_1 , ...,
+A_n\rangle \Rightarrow T$, where each of the $A_i$ arguments are either some
+type $T_i$, or the vararg type $...$.  Recall that functions with a vararg
+parameter (\code{...}) can be called with arbitrarily many values, and extra
+values are lumped together in the vararg $...$ value.  Internally, the type
+$...$ is implemented as a ${\bf list} \langle any \rangle$ type.
+
+We have discussed most of these types throughout this section, with some
+notable exceptions.  ${\bf expression}$, ${\bf language}$, ${\bf symbol}$,
+${\bf externalptr}$, ${\bf pairlist}$, ${\bf S4}$, and ${\bf weakref}$ types
+occur \AT{rarely}, and are by-and-large uninteresting to most R programmers.
+\AT{Fill in details of these basic types.}  One other type we did not
+discuss is the union type $T_1 \, | \, T_2$, which is supported by
+\contractr and generated by \typetracer when it cannot find some way to
+consolidate argument types (for example, if an argument is observed to be
+${\bf chr}$ and ${\bf int}$, \typetracer will ascribe type ${\bf chr} \, |
+\, {\bf int}$ to it).  And while not generated by \typetracer (recall
+discussion in Section~\ref{subsec:structs}), user-defined struct types have
+the form ${\bf struct}\langle F_1, ..., F_n \rangle$, with fields $F_i$
+being name $:$ type pairs.
 

### 2020-05-15 paper/main.tex
minor
RW

@@
 
-In this section, we will introduce the R programming language, as well as
-discuss the dynamic analysis framework we used as part of our initial
-analysis.
-
-%
-%
-\subsection{The R Programming Language}\label{sec:R}
+In this section, we will introduce related work and the R programming language.
+
+\subsection{Related Work}
+
+Dynamic programming languages such as Racket, JavaScript and PHP have been
+extended post-facto with various static type systems.  In each case, the
+type system was carefully engineered to match the salient characteristics of
+the host languages and to foster a particular programming style. For
+example, Typed Racket emphasizes functional programming and support the
+migration from untyped to fully typed code~\cite{tf-popl08},
+Hack~\cite{hack13} and TypeScript~\cite{BAT14} focus on typing
+object-oriented features of PHP and JavaScript, respectively. They allow to
+intersperse typed and untyped code in a fine-grained manner.
+
+But what if the design of the type system is unclear?  \citet{tip} propose
+an intriguing approach called trace typing. With trace typing, a new type
+system can be prototyped and evaluated by applying the type rule to
+execution traces of programs. While the appraoch has the limitation of
+dynamic analysis techniques, namely that the results are only as good at the
+coverage of the source code, it allows to quickly test new design and
+quantify how much of a code base can be type-checked.  Other approaches that
+infer types for dynamic analysis include the work of \citet{FurrAF2009} for
+Ruby.
+
+There is no previous work for the R programming language. We take
+inspiration in the above mentionned works but focus on adapting them to our
+target language.
+
+%
+\subsection{The R Programming Language}
 

### 2020-05-15 paper/main.tex
Add some text for first evaluation section

@@
 %
-\section{Evaluation Results}
+\section{Evaluation}
 \label{sec:evaluation}
@@
 
-\AT{Aviral}
+\subsection{Is our simple type language expressive enough to type R code?} Functions in
+R can support arguments of multiple types by reflecting on the type to perform
+appropriate conversions and computations. On the one hand, this form of ad-hoc
+polymorphism makes R functions expressive; on the other hand, it requires
+support for polymorphism in the typing scheme, making the type system complex. 
+Our system lacks explicit support for polymorphism. Instead, it supports union and any
+types. A union type is inferred when values of multiple types are observed at a
+function parameter and return position. any type is inferred when an argument is
+missing or unused. Furthermore, list element types can be inferred as any if
+they have more than 5 different types. Values with many different types can
+result in a large number of alternatives in the type inferred for that position,
+making it hard to read and unwieldy for end-users. To evaluate this
+''expressivity gap`' between our type inference scheme and R functions,
+we look at the number of different types in the parameter and return types of
+the XXX functions in our corpus.
+
+We observe that XXX% of the positions have a single type ascribed to them. XXX%
+have five or more types ascribed to them. Parameter PARAMNAME of function
+XXXXCOLONCOLONXXX has the maximum, XXX, types ascribed to it. This implies that
+most functions admit or return a uni-typed value.
+
+Figure XXX shows ten most common type categories that occur by themselves. We
+observe that three scalar types, double, logical, and character are most
+frequent. This is followed by XXX vararg parameter positions. Next, XX% of the
+parameter types are any because arguments at those positions were either missing
+or unused in at least one call to those functions. Together, these types amount
+to more than 50%. This is followed by double vector, the unit type, null,
+character vector, and classes matrix and data.frame respectively.
+
+Figure 1 shows the number of alternatives in the inferred type of function
+parameter and return values.
+
+XXX% of the parameter positions are monomorphic. 
+\typetracer observes XXX calls to XXX functions with XXX arguments. This
+translates to XXX parameters and XXX return types. 
+A parameter or return type is  from unioning all the distinct observed types of
+
+Figure 1 shows the distribution of parameter and return types for these
+functions.
+
+class is the most common type in our corpus, accounting for XXX% of all the
+inferred types. This is followed by an almost equal proportion,
+28%, scalar types and, surprisingly, only 13% vector types. Vectors are the
+workhorse of R, most primitive operations in R are vectorized and literals are
+treated as single element vectors. It is unexpected to observe twice as many
+scalars as vectors. It is even more unexpected to observe class as the most
+common type. class of an object is used for dynamic dispatch by the S3 object
+system in R. This suggests that usage of S3 is prevalent in our corpus.
+
+null types. Together, these four types make up for 80% of the observed types in
+our corpus.
+The other types, occurring only XXX% of the times include XXX environment types,
+XXXX expression types, XXXX pairlist types and lastly, XXX externalptr types.
 

### 2020-05-15 paper/main.tex
new text for dynamic eval

--- 2020-05-14 paper/main.tex
@@
 \usepackage{paralist}
+\usepackage{booktabs} % To thicken table lines
 
@@
 \newcommand{\YEARS}{20\xspace} %%TODO fix
+\newcommand{\PERCFAILEDASSERTIONS}{2\%\xspace}
+\newcommand{\NUMPKGSEVAL}{7000\xspace} % autogenerate
+\newcommand{\NUMASSERTIONS}{39210391\xspace} %autogenerate
+\newcommand{\PROPFUNSFAILEDCHECK}{21.21\%\xspace}
+\newcommand{\PROPFUNSFAILEDCHECKNOSTHREE}{15.48\%\xspace} 
+\newcommand{\PROPARGSFAILINGASSERTS}{3.53\%\xspace} % autogen....
 
@@
 
-This paper focuses on a simpler problem than an entire type system, instead,
-we limit the scope of our investigation to giving types to function
-signatures. In order to do this, we design a simple type language, one that
-matches the R data types but omits features such as parametric polymorphism
-and subtyping between user-defined data types. We then extract type
-signature from execution traces of a corpus of widely used libraries.  This
-allows us to see how far one can get with the simple type language and
-identify limitations of our this particular design. We validate the
-robustness of the extracted type signature by implementing a contact system
-that weaves the types around their respective functions, and use a large
-number of clients of the target packages for validation. The contract system
-can be used by the R community to experiment with our type signatures and to
-replace ad hoc error checking code.
-
-To sum up, our paper makes the following main contributions:
+In order to do this, we need to understand the degree of polymorphism
+present in R code, that is to say, how programmers leverage the dynamic
+nature of R to write code that can accept arguments of different types.
+This understanding will drive our design.
+
+We propose to capture the degree of polymorphism present in R by means
+of a dynamic analysis of widely used libraries. For each function call we
+can record the types of its arguments and of its return value. This allows
+us to observe how many different combination of types are accepted by any
+given function. Unlike many other languages, R has a carefully curated
+software repository called CRAN. To be deposited in CRAN, a package must
+come with sample dataset, tests and executable use-cases. As part of normal
+operations these tests are run regularly and failing packages are removed.
+This allowed us to have access to \PACKAGES libraries and about an order of
+magnitude more runnable scripts that exercise those libraries.
+
+The contributions of this paper are thus as follows:
 \begin{itemize}
-\item We implemented tooling to automatically extract type signatures from R
-  functions and to intrument R functions with checks based on their declared
-  types. The tooling is robust and scalable to the entire R language.
-\item We carried out a large-scale analyis of corpus of \PACKAGES widely
-  used and actively maintained libraries to extract function type signatures
-  validated the robustness of the inferred type signatures against XXX
-  programs that use those functions.
-\item We report on the appropriateness and usefulness of a simpe type
-  language for the R programming language.
+\item A large-scale analysis of the polymorphism present in function
+  signatures of \PACKAGES widely used and actively maintained R packages.
+\item A tracing and analysis pipeline that extends a previously published
+  test generation tool named \genthat.
+\item A set of type annotations that captures how programmers use types
+in existing R code.
 \end{itemize}
 
-Our tools are open source and publicly available on GitHub (link anonymize),
-all results in this paper are reproducible and will be submitted for
-artifact evaluation should this paper be accepted.
+One threat to validity of our work is that we rely on dynamic analysis, so
+our conclusions are only as good as the coverage of the possibly function
+behaviors. Previous work~\cite{issta18} reported that running all the
+scripts that come with CRAN packages gives, on average, 68\% test coverage.
+We attempted to mitigate the threat coming from the fact that only part of
+the code is being exercised by manual analysis. It would be reasonable to
+ask for confirmation of the data by static analysis of the code, but sound
+static analysis of R is difficult because of the extensive use of reflective
+features such as \code{eval} and of the ability to redefine the meaning of
+operators such as \code{+} and \code{if}.  Another threat to validity is
+that we only have access to code that has been deposited in the CRAN
+repository. While this may bias our findings towards code written to be
+reusable and, possibly, better engineered than typical user code. This is
+also the code that would most benefit from type annotations.
+
 %
@@
 
-Over the years, programmers have found the need for a richer type structure
-and have added {\it attributes}. The best way to think of attributes is as
-an optional map from name to values that can be attached to any object.
+Over the years, programmers have found the need for a richer type
+structure and have added {\it attributes}. The best way to think of attributes is
+as an optional map from name to values that can be attached to any object.
 Attributes are used to encode various type structures. They can be queried
-with functions such as \k{attributes} and \k{class}.  The addition of
-attributes lets programmers extend the set of types by tagging data with
-user-defined attributes. For example, one could define a vector of four
-values, \code{x<-c(1,2,3,4)} and then attach the attribute \k{dim} with a
-pair of numbers as value: \code{attr(x,"dim")<-c(2,2)}.  From that point,
-arithmetic functions will treat \k{x} as a 2x2 matrix. Another attribute
-that can be set is the \k{class}.  This attribute can be bound to a list of
-class names. For instance, \code{class(x)<-"human"}, set the class of \k{x}
-to be \k{human}.  Attributes are thus used for object-oriented
+with functions such as \k{attributes} and \k{class}.
+The addition of attributes lets programmers extend the set of types by
+tagging data with user-defined attributes. For example, one could define a
+vector of four values, \code{x<-c(1,2,3,4)} and then attach the attribute
+\k{dim} with a pair of numbers as value: \code{attr(x,"dim")<-c(2,2)}.  From
+that point, arithmetic functions will treat \k{x} as a 2x2 matrix. Another
+attribute that can be set is the \k{class}.  This attribute can be bound to
+a list of class names. For instance, \code{class(x)<-"human"}, set the class
+of \k{x} to be \k{human}.  Attributes are thus used for object-oriented
 programming. The S3 object system support single dispatch on the class of
@@
 
-To tackle possibly unevaluated arguments, we make an {\it initial guess} of
-function argument types on function entry. Specifically, we collect types
-for argument default values, and ascribe an ``unused'' type for unevaluated
-expressions.  If an argument promise is later forced, we simply update the
-recorded type for the argument.
+To tackle possibly unevaluated arguments, we make an {\it initial guess} of function argument types on
+function entry. Specifically, we collect types for argument default values, and
+ascribe an ``unused'' type for unevaluated expressions.  If an argument
+promise is later forced, we simply update the recorded type for the
+argument.
 
 \item Second, we wanted to account for formal arguments (arguments which are
-  named in the function definition) which received no values when the
-  function was called. This represents cases where a function was not called
-  with all of its arguments, and no default value was specified for missing
-  arguments.  We record a ``missing'' type for such argument.
+  named in the function definition) which received no values when the function was called. This represents cases where a function was not called with all of its
+  arguments, and no default value was specified for missing arguments. 
+  We record a ``missing'' type for such argument.
 
@@
 modified version of the R runtime. 
+% Our dynamic type analysis only needs to be run once, ever---from the analysis results, we synthesize function signatures which are passed along to later type-checking tools which do not rely on a heavily modified runtime.
 
@@
 guesses of argument types (as discussed in
-Section~\ref{subsec:typetracer-challenges}).  Also, recall that R can
-perform single or multiple dispatch on function arguments depending on their
-class.  We collect such dispatch information the {\tt \_entry} variants.
-
-The primary functionality of \typetracer is in {\tt promise\_force\_exit}.
-As arguments to functions in R are wrapped in promises, we delay our
-reflection to promise force time.  In R, promises are essentially objects
-with value and expression slots. \AT{Put this as part of R-dyntrace
-  background:} When a promise is created for a particular expression, that
-expression is merely put into the expression slot of a promise object, and a
-special {\it unbound} value is loaded into the value slot (unless some
+Section~\ref{subsec:typetracer-challenges}).  
+Also, recall that R can perform single or multiple dispatch on function arguments depending on their class.
+We collect such dispatch information the {\tt \_entry} variants.
+
+The primary functionality of \typetracer is in {\tt promise\_force\_exit}.  As arguments to
+functions in R are wrapped in promises, we delay our reflection to promise
+force time.  In R, promises are essentially objects with value and
+expression slots. \AT{Put this as part of R-dyntrace background:} When a promise is created for a particular expression,
+that expression is merely put into the expression slot of a promise object,
+and a special {\it unbound} value is loaded into the value slot (unless some
 default value is specified, wherein it is loaded into the value slot---this
@@
 the value slot of the promise.  So to get the type of a function argument,
-we check the value slot of the argument promise---recursively if the value
-itself is a promise.
+we check the value slot of the argument promise---recursively if the
+value itself is a promise.
 
@@
 
+\AT{Maybe further discuss how we can create many candidate TS from the data.}
+
 The type information we collect is far more detailed than the types we
-actually generate (which we discuss in Section~\ref{sec:types}), and so the
-next stage of our inference process is to distill the rich type information
-into simpler, manageable types.  For each unique observed call trace, we
-translate the representation of the type of each of the function arguments
-and returns according to the design of our type system: for instance, we
-might have determined that a function returned a vector of integer with 4
-elements, with names, that was free of NA values, in which case we would
-write out \code{integer[]} as the type.
+actually generate (which we discuss in Section~\ref{sec:types}), and so the next stage of our inference process is to
+distill the rich type information into simpler, manageable types.  For each
+unique observed call trace, we translate the representation of the type of
+each of the function arguments and returns according to the design of our
+type system: for instance, we might have determined that a function returned
+a vector of integer with 4 elements, with names, that was free of NA values,
+in which case we would write out \code{integer[]} as the type.
 
@@
 consolidate its multiple call traces into a single function type.  We do
-this by combining the observed types of each argument into one type
-according to subtyping rules that are part of our design.  The precise
-subtyping rules we used will be discussed further in
-Section~\ref{sec:typesystemdesign}.  If we cannot consolidate multiple
-observed argument types into a single type, we simply union the disparate
-types together.
-
-While our analysis collects more detailed information than necessary for
-type generation, the information proved useful for building up a usage
-profile of the language.  This profile informed our language design: based
-on the rich type information we collected, we were able to determine types
-that would be useful for capturing the observed behaviour.  We discuss this
-iterative, data-driven language design process in Section~\ref{types},
-highlighting the reasoning behind our type system design.
-
-Thus far, we have discussed \typetracer and our strategies to take the data
-from the dynamic analysis and generate function types.  This work also
-presents a framework for specifying these types in R code, as well as
-generating contract checks for these types.  We dub this tool \contractr,
-discussed next.
-
+this by combining the observed types of each argument into one type according to subtyping rules that are part of our
+design.  The precise subtyping rules we used will be discussed further in
+Section~\ref{sec:typesystemdesign}.
+If we cannot consolidate multiple observed argument types into a single type, we simply union the disparate types together.
+
+While our analysis collects more detailed information than necessary for type generation, the information proved useful for building up a usage profile of the language. 
+This profile informed our language design: based on the rich type information we collected, we were able to determine types that would be useful for capturing the observed behaviour.
+We discuss this iterative, data-driven language design process in Section~\ref{types}, highlighting the reasoning behind our type system design.
+
+Thus far, we have discussed \typetracer and our strategies to take the data from the dynamic analysis and generate function types.
+This work also presents a framework for specifying these types in R code, as well as generating contract checks for these types.
+We dub this tool \contractr, discussed next.
+
+%
 %
@@
 %
+%
+%
+%
 \section{Type Language Design}
@@
 an iterative process: initial analysis suggested a certain set of types,
-which we tried and implemented, and reanalysis suggested improvements, and
-so on.  In this section, we will present the final type annotation language
-design, which we believe captures the essence of R.  We will end the section
-with a formal type language grammar, and leading up to that we will touch on
-design decisions and weigh our choices against alternatives, speaking to the
-strengths and weaknesses of our chosen design.  We begin with basic,
-primitive types.
+which we tried and implemented, and reanalysis suggested improvements,
+and so on.  In this section, we will present the final type annotation language design, 
+which we believe captures the essence of R.  We will end the section with a formal type 
+language grammar, and leading up to that we will
+touch on design decisions and weigh our choices against alternatives,
+speaking to the strengths and weaknesses of our chosen design.
+We begin with basic, primitive types.
 
@@
 A common primitive usage pattern is related to the presence of NAs:
-programmers explicitly check for NAs in primitive vectors, possibly
-sanitizing them if they are present.  NAs complicate practically every
-computation using primitive vectors, a fact that many programmers appear to
-be acutely aware of. For instance, consider the code in
-Figure~\ref{fig:na-example}.
+\AT{often}, programmers explicitly check for NAs in primitive vectors,
+possibly sanitizing them if they are present.  NAs complicate practically every computation
+using primitive vectors, a fact that many programmers appear to be acutely
+aware of. For instance, consider the code in Figure~\ref{fig:na-example}.
 
@@
 
-The \code{binom} package is a popular R package for computing confidence
-intervals for binomial experiments, and the \code{binom.profile} function
-does so using the profile likelihood, the details of which are not relevant
-here.  This code snippet highlights a data sanitization pattern appearing at
-the beginning of the function: the programmer first binds the vectors into a
-matrix (line 2), with one column for each of \code{x} and \code{n}, then
-finds rows where both columns are not NA (line 3), then extracts non-NA
-values and stores them into \code{x} and \code{n} respectively (lines 4-5).
-Importantly, this process is entirely quiet, and the user of
-\code{binom.profile} will not be notified if their data happens to contain
-NAs, and the NAs will be silently ignored.  In reality, the presence of NAs
-in data could signify some sort of data corruption, and quietly ignoring
-issues in data could lead users (read: statisticians) to draw flawed
-conclusions from erroneous data.
-
-To combat this, we introduce a modifier on primitive types: the type
-\code{^T} signifies that primitive type \code{T} might contain an NA value,
-while \code{T} signifies that \code{T} is NA-free.  The NA-free case has a
-shorter syntax as NA-free vectors are \AT{more common} than the alternative.
-If the \code{binom} developers were to annotate arguments \code{x} and
-\code{n} with such an NA-free type, \contractr will notify users when
-\code{binom.profile} is called with NA-full values, rather than just quietly
-ignoring them.
-
-Note that when combining call traces involving \code{^T} and \code{T} types,
-we subsume \code{T} types into \code{^T} types: if NA-free and NA-full
-vectors are passed to an argument, then we say that the argument can handle
-NA-full vectors.  This is equivalent to saying \code{T <: ^T}.
+The \code{binom} package is a popular R package for computing confidence intervals for binomial experiments, and the \code{binom.profile} function does so using the profile likelihood, the details of which are not relevant here.
+This code snippet highlights a data sanitization pattern appearing at the beginning of the function: the programmer first binds the vectors into a matrix (line 2), with one column for each of \code{x} and \code{n}, then finds rows where both columns are not NA (line 3), then extracts non-NA values and stores them into \code{x} and \code{n} respectively (lines 4-5).
+Importantly, this process is entirely quiet, and the user of \code{binom.profile} will not be notified if their data happens to contain NAs, and the NAs will be silently ignored. 
+In reality, the presence of NAs in data could signify some sort of data corruption, and quietly ignoring issues in data could lead users (read: statisticians) to draw flawed conclusions from erroneous data.
+
+To combat this, we introduce a modifier on primitive types: the type \code{^T} signifies that primitive type \code{T} might contain an NA value, while \code{T} signifies that \code{T} is NA-free. 
+The NA-free case has a shorter syntax as NA-free vectors are \AT{more common} than the alternative.
+If the \code{binom} developers were to annotate arguments \code{x} and \code{n} with such an NA-free type, \contractr will notify users when \code{binom.profile} is called with NA-full values, rather than just quietly ignoring them.
+
+Note that when combining call traces involving \code{^T} and \code{T} types, we subsume \code{T} types into \code{^T} types: if NA-free and NA-full vectors are passed to an argument, then we say that the argument can handle NA-full vectors.
+This is equivalent to saying \code{T <: ^T}.
 
@@
 
-Initial analyses of our data revealed that programmers \AT{often} use
-scalars (as explicitly as they can, anyway), and often do high-level
-dimensionality checks on their data.  To illustrate, consider the code in
-Figure~\ref{fig:scalar-vector-example}.
+Initial analyses of our data revealed that programmers \AT{often} use scalars (as explicitly as they can, anyway), and often do high-level dimensionality checks on their data.
+To illustrate, consider the code in Figure~\ref{fig:scalar-vector-example}.
 
@@
 
-The \code{hankel.matrix} function takes two arguments (as described in the
-comment in the code), and returns a Hankel matrix, which essentially is a
-matrix where the skew-diagonals have a desirable property.  The function
-body has a number of interesting checks: on lines 6-7, \code{n} undergoes an
-integer ``type check''; on lines 8-9, \code{x} undergoes a vector ``type
-check''; on line 11, \code{n} is assumed to be a scalar, as if it were a
-vector, the code \code{m < n} would essentially check \code{m < n[i]} for
-each element in \code{n}, returning a vector of comparison results, and the
-conditional of an if must be a scalar else R generates a warning.
-
-These checks exemplify functionality that we would like to include in our
-type annotation language, namely, we would like our type language to
-differentiate between scalars and vectors.  To that end, we introduce the
-following types into our type language: for some primitive type \code{T},
-the user may modify \code{T} with a modifier \code{[]} to indicate that it
-is a vector of \code{T}.  Perhaps surprisingly, we found that scalars
-\AT{occurred more frequently} than vectors, and for that reason we chose the
-shorter scalar type syntax as the default.  The NA-free modifier works as
-expected: \code{^T[]} indicates a vector of \code{T} that is free of NAs,
-while \code{^T} indicates a scalar which is not NA.
+The \code{hankel.matrix} function takes two arguments (as described in the comment in the code), and returns a Hankel matrix, which essentially is a matrix where the skew-diagonals have a desirable property.
+The function body has a number of interesting checks:
+on lines 6-7, \code{n} undergoes an integer ``type check'';
+on lines 8-9, \code{x} undergoes a vector ``type check'';
+on line 11, \code{n} is assumed to be a scalar, as if it were a vector, the code \code{m < n} would essentially check \code{m < n[i]} for each element in \code{n}, returning a vector of comparison results, and the conditional of an if must be a scalar else R generates a warning.
+
+These checks exemplify functionality that we would like to include in our type annotation language, namely, we would like our type language to differentiate between scalars and vectors.
+To that end, we introduce the following types into our type language:
+for some primitive type \code{T}, the user may modify \code{T} with a modifier \code{[]} to indicate that it is a vector of \code{T}.
+Perhaps surprisingly, we found that scalars \AT{occurred more frequently} than vectors, and for that reason we chose the shorter scalar type syntax as the default.
+The NA-free modifier works as expected: \code{^T[]} indicates a vector of \code{T} that is free of NAs, while \code{^T} indicates a scalar which is not NA.
 
@@
 
-Matrices represent another layer of complexity on top of vectorized
-primitives.  Internally, matrices are simply vectors with class ``matrix''
-and a \code{dims} attribute indicating matrix dimensions, and while not
-codified in the language semantics, many internal functions will coerce
-vectors to matrices automatically.  Relying on automatic coercion is unwise,
-and many functions check to ensure that their arguments are matrices.  For
-instance, consider the code in Figure~\ref{fig:matrix-example}.
+Matrices represent another layer of complexity on top of vectorized primitives.
+Internally, matrices are simply vectors with class ``matrix'' and a \code{dims} attribute indicating matrix dimensions, and while not codified in the language semantics, many internal functions will coerce vectors to matrices automatically.
+Relying on automatic coercion is unwise, and many functions check to ensure that their arguments are matrices.
+For instance, consider the code in Figure~\ref{fig:matrix-example}.
 
@@
 
-The \code{rowWeightedMeans} function calculates the weighted means of rows
-or columns of a matrix \code{x}.  This function exhibits good practice in
-type checking arguments, rightly producing a message to the user indicating
-that an explicit matrix should be passed.
-
-To capture cases like this, we include a \code{class<matrix>} type in our
-type annotation language.  If \code{rowWeightedMeans} were outfitted with
-such a type, \contractr can perform the check present in this code
-automatically, notifying the users passing ill-typed arguments.
+The \code{rowWeightedMeans} function calculates the weighted means of rows or columns of a matrix \code{x}.
+This function exhibits good practice in type checking arguments, rightly producing a message to the user indicating that an explicit matrix should be passed.
+
+To capture cases like this, we include a \code{class<matrix>} type in our type annotation language.
+If \code{rowWeightedMeans} were outfitted with such a type, \contractr can perform the check present in this code automatically, notifying the users passing ill-typed arguments.
 
@@
 
-Thus far we've seen instances where we introduce types that subsume existing
-argument validation code, but our intention is not to subsume {\it all} such
-code.  As an example of something we {\it do not} want to do, consider the
-code in Figure~\ref{fig:general-validation-example}.
+Thus far we've seen instances where we introduce types that subsume existing argument validation code, but our intention is not to subsume {\it all} such code.
+As an example of something we {\it do not} want to do, consider the code in Figure~\ref{fig:general-validation-example}.
 
@@
 
-This code is from later in the same \code{rowWeightedMeans} function
-discussed previously.  Here, the length of the weights vector \code{w} is
-being checked for equality with the number of columns in the matrix
-\code{x}.  Specifying this relationship as a type amounts to {\it dependent
-  typing}, which we believe too complex to retrofit onto a dynamic data
-science language such as R.  Ultimately, we decided to leave these kinds of
-checks up to the programmer.
+This code is from later in the same \code{rowWeightedMeans} function discussed previously.
+Here, the length of the weights vector \code{w} is being checked for equality with the number of columns in the matrix \code{x}. 
+Specifying this relationship as a type amounts to {\it dependent typing}, which we believe too complex to retrofit onto a dynamic data science language such as R.  
+Ultimately, we decided to leave these kinds of checks up to the programmer.
+\AT{Can we say more here?}
 
@@
 
-Separate from the vectorized primitives in R are lists: Whereas vectors are
-homogeneous and restricted to primitive types, lists can contain
-arbitrarily-typed data.  Further, lists and vectors have a different syntax
-for accessing elements, where lists are indexed using \code{[[} and vectors
-    using \code{[}.  R even builds on lists to create a more complex data
-      structure known as the {\it data frame}, an essential data structure
-      to R's functionality.
-
-We are faced with several possible designs for list types.  Chief concerns
-include the heterogeneity of list elements and the presence list element
-names, which allow list elements to be accessed in a struct-like manner.
-The next sections highlight some important design decisions for our list
-types.
+Separate from the vectorized primitives in R are lists:
+Whereas vectors are homogeneous and restricted to primitive types, lists can contain arbitrarily-typed data.  
+Further, lists and vectors have a different syntax for accessing elements, where lists are indexed using \code{[[} and vectors using \code{[}.
+R even builds on lists to create a more complex data structure known as the {\it data frame}, an essential data structure to R's functionality.
+
+We are faced with several possible designs for list types.  
+Chief concerns include the heterogeneity of list elements and the presence list element names, which allow list elements to be accessed in a struct-like manner.  
+% \AT{Rarely?}, programmers even write code expecting lists of a particular pre-defined length.  
+The next sections highlight some important design decisions for our list types.
 
@@
 
-The ease with which vectors may be transformed into lists is perhaps
-indicative of a relationship between the two.  In R, all vectors can be
-converted to lists with a simple call to \code{as.list}, and lists to
-vectors with \code{unlist}, though care must be taken when unlisting a
-non-homogeneous list, as coercions will be performed where possible, and the
-function will simply do nothing if coercion is impossible.  That said, lists
-and vectors have different access syntax, so vectors cannot stand-in for
-lists without the \code{as.list} call above.
-
-In sum, we opted not to include a subtyping relationship between vectors and
-lists, keeping both types separate.  In practice, \AT{how often do \code{T[]
-    | list<T>} types occur?}.  During our evaluation we also found it rare
-that programmers would attempt to pass list-typed values to vector-typed
-arguments and vice versa, and we will discuss this further in
-Section~\ref{sec:evaluation}.
+\AT{Worth a code example?}
+
+The ease with which vectors may be transformed into lists is perhaps indicative of a relationship between the two.
+In R, all vectors can be converted to lists with a simple call to \code{as.list}, and lists to vectors with \code{unlist}, though care must be taken when unlisting a non-homogeneous list, as coercions will be performed where possible, and the function will simply do nothing if coercion is impossible.
+That said, lists and vectors have different access syntax, so vectors cannot stand-in for lists without the \code{as.list} call above.
+
+In sum, we opted not to include a subtyping relationship between vectors and lists, keeping both types separate.
+In practice, \AT{how often do \code{T[] | list<T>} types occur?}.
+During our evaluation we also found it rare that programmers would attempt to pass list-typed values to vector-typed arguments and vice versa, and we will discuss this further in Section~\ref{sec:evaluation}.
 
@@
 
-Environments in R are essentially lists with reference semantics: mutating a
-value in an environment is performed in-place, and that mutation can thus be
-shared with other code holding a reference to the environment.  They are
-used primarily by the R runtime to store variables (their primary function
-is that of an execution environment), though they also often used by
-programmers either to themselves reflect on local, previous, or global
-scope, or simply because of their reference semantics.  Environments are
-comprised of name, value pairs, and each environment has a reference to its
-parent environment.
-
-We thought to include detailed environment types in our type language by way
-of allowing environment types to define names and types of values in the
-environment.  Unfortunately, this led to an undue explosion in the ``size''
-of our types: often, environments had hundreds of values, particularly those
-which were obtained from R (by e.g. passing a reference to a local scope).
+Environments in R are essentially lists with reference semantics: mutating a value in an environment is performed in-place, and that mutation can thus be shared with other code holding a reference to the environment.
+They are used primarily by the R runtime to store variables (their primary function is that of an execution environment), though they also often used by programmers either to themselves reflect on local, previous, or global scope, or simply because of their reference semantics.
+Environments are comprised of name, value pairs, and each environment has a reference to its parent environment.
+
+We thought to include detailed environment types in our type language by way of allowing environment types to define names and types of values in the environment.
+Unfortunately, this led to an undue explosion in the ``size'' of our types: 
+often, environments had hundreds of values, particularly those which were obtained from R (by e.g. passing a reference to a local scope).
 These types were difficult to parse visually and did not feel informative.
-Further, environments \AT{did not occur often}, and giving them a richer
-type was not worth the increase in the visual complexity of our types.
+Further, environments \AT{did not occur often}, and giving them a richer type was not worth the increase in the visual complexity of our types.
+
+%
+%
+% \subsubsection{Forgoing Tuples, or Short, Constant Length Lists}
+
+% We briefly entertained the possibility of including a \code{tuple} type, which would have looked like e.g. \code{tuple<int, int>} for a list of two integers.  
+% When configured to detect tuples, our analysis did indeed pick up \AT{a number}, but post-processing and simplification revealed that in \AT{many} cases, they would be subsumed by co-occurring \code{list} types.
+% In other words, more often than not it just so happened that functions were being called with short lists.  
+% Of the cases that tuples were not subsumed by lists, \AT{often} there were \AT{tuples of different lengths}, implying that the shape of the argument was not essential.
+
+% \AT{I feel like this is the least important of the design decisions we made, so I'm leaving it for the end, we might not even want it.}
 
@@
 
-We observed the use of consistently-named lists in \AT{a number of
-  functions}.  In R, ``fields'' of named lists can be accessed using the
-\code{$} operator, as illustrated in the code snippet in
-Figure~\ref{fig:struct-ex}.
+\AT{Do we want to talk about how list names are not unique?}
+
+We observed the use of consistently-named lists in \AT{a number of functions}.
+In R, ``fields'' of named lists can be accessed using the \code{$} operator, as illustrated in the code snippet in Figure~\ref{fig:struct-ex}.
 
@@
 
-In the code snippet, the function \code{cv.model} takes an argument
-\code{x}, which we observed to always be a {\it linear model}, which is part
-of R's built-in statistics functionality.  Internally, linear models are
-represented as lists with named elements, and one may access these named
-elements using the \code{$} syntax, as seen on lines 2-4 in the snippet.  To
-describe these kinds of values, we included a struct type in our type
-system, allowing users to specify name, type pairs as part of list type
-definitions.
-
-Unfortunately, the inclusion of structs complicated our analysis by
-polluting the data, illustrated on line 8 with the call
-\code{data(sweetpotato)}.  It is quite common for R example code to use some
-built-in data sets, such as \code{sweetpotato}, to show off some package
-functionality with ``real data''.  When \typetracer is configured to detect
-structs we saw a marked increase in misrepresentative types, as the types of
-functions which use these data sets (such as \code{aov} in
-Figure~\ref{fig:struct-ex}) became polluted with the names from the example
-sets.  We implemented a unification strategy for struct types, removing them
-if they co-occurred with list types, and turning them into lists if two
-co-occurring struct types shared no common names, but even so we found
-ourselves with function signatures that were full of noise introduced by the
-struct names.
-
-We still believe struct types to be useful to R programmers, and so our
-language supports the type and \contractr will check them, though we
-configured \typetracer to ignore them as they made the data substantially
-less useful.  On the positive side, we found that \AT{get the number, but
-  often} structs-typed values had a {\it class}, which helped motivate our
-inclusion of class-like types in our type annotation language.  Classes are
-discussed in more detail in Section~\ref{subsec:classes}.
-
+In the code snippet, the function \code{cv.model} takes an argument \code{x}, which we observed to always be a {\it linear model}, which is part of R's built-in statistics functionality.
+Internally, linear models are represented as lists with named elements, and one may access these named elements using the \code{$} syntax, as seen on lines 2-4 in the snippet.
+To describe these kinds of values, we included a struct type in our type system, allowing users to specify name, type pairs as part of list type definitions.
+
+Unfortunately, the inclusion of structs complicated our analysis by polluting the data, illustrated on line 8 with the call \code{data(sweetpotato)}.
+It is quite common for R example code to use some built-in data sets, such as \code{sweetpotato}, to show off some package functionality with ``real data''.
+When \typetracer is configured to detect structs we saw a marked increase in misrepresentative types, as the types of functions which use these data sets (such as \code{aov} in Figure~\ref{fig:struct-ex}) became polluted with the names from the example sets.
+We implemented a unification strategy for struct types, removing them if they co-occurred with list types, and turning them into lists if two co-occurring struct types shared no common names, but even so we found ourselves with function signatures that were full of noise introduced by the struct names.
+
+We still believe struct types to be useful to R programmers, and so our language supports the type and \contractr will check them, though we configured \typetracer to ignore them as they made the data substantially less useful.
+On the positive side, we found that \AT{get the number, but often} structs-typed values had a {\it class}, which helped motivate our inclusion of class-like types in our type annotation language.
+Classes are discussed in more detail in Section~\ref{subsec:classes}.
+
+% We also thought to encode data frames as structs, as they are essentially lists with named columns of homogeneous values, but it was difficult to draw any meaningful conclusions from the data as unfortunately structs introduced a lot of noise into our analysis.
+
+%
 %
@@
 
-One of the most popular classes in R is the \code{data.frame} class \AT{we
-  have numbers here, grab em}.  Data frames and the derivative \code{tibble}
-(from the \code{tidyverse} ecosystem) and \code{data.table} (from the
-\code{data.table} package) types underpin a huge amount of data analysis in
-R.
-
-One way to deal with data frames is through the struct type, with a named
-field for each column of the data frame, but as mentioned previously structs
-introduced undue noise into \typetracer's analysis results.  Further
-complicating data frames is that many functions built to operate on them
-operate in a name-agnostic way.  For instance, the \code{tidyverse} package
-ecosystem allows programmers to pass column names to functions which operate
-on their data frames.  In base R, idiomatic data frame use is to use string
-column names to select rows from the frame (unless only a single column is
-of interest, wherein the \code{$} syntax is appropriate).  We include some
-examples in Figure~\ref{fig:data-frames-bad} to illustrate the complexities.
+One of the most popular classes in R is the \code{data.frame} class \AT{we have numbers here, grab em}.
+Data frames and the derivative \code{tibble} (from the \code{tidyverse} ecosystem) and \code{data.table} (from the \code{data.table} package) types underpin a huge amount of data analysis in R.
+
+One way to deal with data frames is through the struct type, with a named field for each column of the data frame, but as mentioned previously structs introduced undue noise into \typetracer's analysis results.
+Further complicating data frames is that many functions built to operate on them operate in a name-agnostic way.
+For instance, the \code{tidyverse} package ecosystem allows programmers to pass column names to functions which operate on their data frames.
+In base R, idiomatic data frame use is to use string column names to select rows from the frame (unless only a single column is of interest, wherein the \code{$} syntax is appropriate). 
+We include some examples in Figure~\ref{fig:data-frames-bad} to illustrate the complexities.
 
@@
 
-Doing justice to this type will be an endeavour in and of itself, requiring
-reconciliation of base R's \code{data.frame}, \code{tidyverse}'s
-\code{tibble}, and \code{data.table}'s \code{data.table}.  Developing useful
-types for the \code{tidyverse} ecosystem may well involve
-
-It will also require proper treatment of function types, and perhaps even
-parametric types, which dynamic analyses such as \typetracer are unable to
-discern soundly.
+Doing justice to this type will be an endeavour in and of itself, requiring reconciliation of base R's \code{data.frame}, \code{tidyverse}'s \code{tibble}, and \code{data.table}'s \code{data.table}.
+Developing useful types for the \code{tidyverse} ecosystem may well involve
+
+It will also require proper treatment of function types, and perhaps even parametric types, which dynamic analyses such as \typetracer are unable to discern soundly.
+\AT{General words to convince the reader that data frames are, in fact, hard, and worthy of their own paper.}
 
@@
 
-As discussed in Section~\ref{sec:R}, the R runtime has more than one notion
-of type.  The classes of function arguments are used by R as part of its
-dispatch mechanism, the most prevalent of which (\AT{we have that number,
-  too!}) is S3 dispatch, in which R dispatches on the (S3) class of the
-first argument to a function, selecting the appropriate method to call for
-the class.  To quickly illustrate, consider the code snippet in
-Figure~\ref{fig:dispatch-ex}.
+As discussed in Section~\ref{sec:R}, the R runtime has more than one notion of type.  
+The classes of function arguments are used by R as part of its dispatch mechanism, the most prevalent of which (\AT{we have that number, too!}) is S3 dispatch, in which R dispatches on the (S3) class of the first argument to a function, selecting the appropriate method to call for the class.
+To quickly illustrate, consider the code snippet in Figure~\ref{fig:dispatch-ex}.
 
@@
 
-Classes are very prevalent in R \AT{number}, and so we include a
-\code{class<class-name>} ... \AT{todo} \typetracer was outfit to report the
-class of values, but some care needed to be taken so as to not overreport
-class types, as {\it all} values have a class in R.  The main classes we
-were interested in were matrix, array (higher dimensional matrices), and
-data.frames, in addition to any user-defined classes.  In many cases
-\AT{number?}, values with a user-defined class are in reality lists with
-named elements, or structs.  Including class types, and having \typetracer
-report them in place of structs, significantly reduced the noise in the
-analysis and allowed us to gather more focused data.  To get a sense for
-this difference, consider the code in Figure~\ref{fig:class-vs-struct}.
+Classes are very prevalent in R \AT{number}, and so we include a \code{class<class-name>} ... \AT{todo}
+\typetracer was outfit to report the class of values, but some care needed to be taken so as to not overreport class types, as {\it all} values have a class in R.
+The main classes we were interested in were matrix, array (higher dimensional matrices), and data.frames, in addition to any user-defined classes.
+In many cases \AT{number?}, values with a user-defined class are in reality lists with named elements, or structs.
+Including class types, and having \typetracer report them in place of structs, significantly reduced the noise in the analysis and allowed us to gather more focused data.
+To get a sense for this difference, consider the code in Figure~\ref{fig:class-vs-struct}.
 
@@
 
-A value can have multiple classes in R.  This is common practice in the
-\code{dplyr} package, which extends the functionality of data frames: a
-\code{data.frame} that passes through \code{dplyr} functions gains two
-additional classes (\code{tbl} and \code{tbl_df}), bringing the total to at
-least 3, in addition to whichever other classes that data frame happened to
-have.  Note that when looking to perform S3 dispatch, R will try to match
-{\it any} of the classes of the first argument to a dispatched function, so
-each of these classes are important.
-
-We found these so-called multi-classes to be surprisingly common, making up
-\AT{occurrence} of the occurrences of class-typed values.  One possible
-reason for this is that values can gain classes quite easily (e.g. the
-\code{dplyr} example discussed above), somewhat polluting the output of
-\typetracer.  Nevertheless, we found \AT{number} of occurrences of
-monomorphic multi-class arguments, meaning that the function was always
-called with a multi-class value in that argument position.
-
-In the end, we support these multi-class types by allowing \code{class<...>}
-types to contain a list of class names, so for instance a \code{dplyr}
-\code{tibble} would have type \code{class<data.frame, tbl, tbl_df>} in our
-type language, and when performing related contract assertions \contractr
-will check that the value has every class that make up its type.
+A value can have multiple classes in R.
+This is common practice in the \code{dplyr} package, which extends the functionality of data frames: a \code{data.frame} that passes through \code{dplyr} functions gains two additional classes (\code{tbl} and \code{tbl_df}), bringing the total to at least 3, in addition to whichever other classes that data frame happened to have.
+Note that when looking to perform S3 dispatch, R will try to match {\it any} of the classes of the first argument to a dispatched function, so each of these classes are important.
+
+We found these so-called multi-classes to be surprisingly common, making up \AT{occurrence} of the occurrences of class-typed values.
+One possible reason for this is that values can gain classes quite easily (e.g. the \code{dplyr} example discussed above), somewhat polluting the output of \typetracer.
+Nevertheless, we found \AT{number} of occurrences of monomorphic multi-class arguments, meaning that the function was always called with a multi-class value in that argument position.
+
+In the end, we support these multi-class types by allowing \code{class<...>} types to contain a list of class names, so for instance a \code{dplyr} \code{tibble} would have type \code{class<data.frame, tbl, tbl_df>} in our type language, and when performing related contract assertions \contractr will check that the value has every class that make up its type.
 
@@
 dispatch mechanism, and indeed our class types roughly capture the S3
-classes of values.  We will not discuss R5 as it is still under development.
-
-For S4, we have an \code{s4} type to denote S4 values, though we did not
-elaborate on this type, and in practice S4 values have a class, and thus a
-\code{class<...>} type.  We stuck to simple class-based types for S4 since
-doing anything more would be immensely complicated, and outside of the scope
-of this work.  For instance, the mechanics of S4 dispatch are more complex
-than for S3 (S4 dispatches on the class of {\it all} arguments), and users
-can define their own class hierarchies that we would need to incorporate in
-our type analysis and contract checking frameworks.  Further, \AT{we found
-  limited use of S4 during our analysis}, with S4 types being heavily used
-in a select subset of packages (\AT{such as ...}), some of which make heavy
-use of R's metaprogramming capabilities.  Coming up with a type system that
-accounts for all of these factors and consolidates multiple
-object-orientation frameworks in a single language design is an interesting
-problem in and of itself, and we aim to extend our type annotation framework
-to account for this in future work.  For now, S3 types suffice: S3 is the
-most common object system in R, with no sophisticated hierarchies, and its
-semantics are simple enough to be captured by our proposed design.
+classes of values.  We will not discuss R5 as it is still under
+development.
+
+For S4, we have an \code{s4} type to denote S4 values, though we did not elaborate on this type, and in practice S4 values have a class, and thus a \code{class<...>} type.  
+We stuck to simple class-based types for S4 since doing anything more would be immensely complicated, and outside of the scope of this work.
+For instance, the mechanics of S4 dispatch are more complex than for S3 (S4 dispatches on the class of {\it all} arguments), and users can define their own class hierarchies that we would need to incorporate in our type analysis and contract checking frameworks.
+Further, \AT{we found limited use of S4 during our analysis}, with S4 types being heavily used in a select subset of packages (\AT{such as ...}), some of which make heavy use of R's metaprogramming capabilities.
+Coming up with a type system that accounts for all of these factors and consolidates multiple object-orientation frameworks in a single language design is an interesting problem in and of itself, and we aim to extend our type annotation framework to account for this in future work.
+For now, S3 types suffice: S3 is the most common object system in R, with no sophisticated hierarchies, and its semantics are simple enough to be captured by our proposed design.
 
@@
 
-Functions are first-class values in R.  \typetracer found \AT{number}
-functions to be higher-order, having at least one function type among their
-arguments.  They are markedly more common in a subset of packages, such as
-\AT{such as...}.  By-and-large, these higher-order functions are similar to
-well-known higher-order functions like map, filter, and fold.
-
-The truth is that higher-order function types are best served by {\it
-  parametric types} and parametric polymorphism, wherein type variables are
-used within a type declaration to specify type-based relationships between
-function arguments.  Unfortunately, the nature of dynamic analysis is such
-that \typetracer cannot soundly detect parametric polymorphism, as without a
-static analysis of the function body it is impossible to say if function
-execution will halt given certain argument types, and we saw some examples
-of this in
-Figures~\ref{fig:scalar-vector-example}~and~\ref{fig:matrix-example}.
-Without parametric types, higher-order function types become noisy and fail
-to tell the complete picture.
-
-Ultimately, we configured \typetracer to report only top-level function
-types (i.e., detect if a value is a function or not).  \AT{The data suggests
-  that} most of the use of higher-order functions is through map-, filter-,
-and apply-like functions, whose behaviour is best captured with parametric
-polymorphism.  Other common uses include higher-order functions operating on
-data frame columns (in packages such as \code{dplyr}), which would
-necessitate a more in-depth treatment of data frame types, which as we
-discussed is a huge endeavour in and of itself.  We aim to extend our type
-annotation framework with parametric, higher-order, and detailed data frame
-types together in future work.
+\AT{Revisit this section since discussion of polymorphism in results.}
+
+Functions are first-class values in R.
+\typetracer found \AT{number} functions to be higher-order, having at least one function type among their arguments.
+They are markedly more common in a subset of packages, such as \AT{such as...}.
+By-and-large, these higher-order functions are similar to well-known higher-order functions like map, filter, and fold.
+
+The truth is that higher-order function types are best served by {\it parametric types} and parametric polymorphism, wherein type variables are used within a type declaration to specify type-based relationships between function arguments.
+Unfortunately, the nature of dynamic analysis is such that \typetracer cannot soundly detect parametric polymorphism, as without a static analysis of the function body it is impossible to say if function execution will halt given certain argument types, and we saw some examples of this in Figures~\ref{fig:scalar-vector-example}~and~\ref{fig:matrix-example}.
+Without parametric types, higher-order function types become noisy and fail to tell the complete picture.
+
+Ultimately, we configured \typetracer to report only top-level function types (i.e., detect if a value is a function or not).
+\AT{The data suggests that} most of the use of higher-order functions is through map-, filter-, and apply-like functions, whose behaviour is best captured with parametric polymorphism.
+Other common uses include higher-order functions operating on data frame columns (in packages such as \code{dplyr}), which would necessitate a more in-depth treatment of data frame types, which as we discussed is a huge endeavour in and of itself.
+We aim to extend our type annotation framework with parametric, higher-order, and detailed data frame types together in future work.
 
@@
 
-One final, common pattern that we would like to highlight is the
-co-occurrence of \code{NULL} with other types of values.  In
-Section~\ref{subsec:basictypes} we discussed NA, and how NAs represent
-missing data for the primitive types, and more generally R provides a
-\code{NULL} value that programmers use to represent missing data for other
-types.  Programmers often use \code{NULL} as a default value for function
-arguments, and the code snippet in Figure~\ref{fig:null} exemplifies this.
+One final, common pattern that we would like to highlight is the co-occurrence of \code{NULL} with other types of values.
+In Section~\ref{subsec:basictypes} we discussed NA, and how NAs represent missing data for the primitive types, and more generally R provides a \code{NULL} value that programmers use to represent missing data for other types.
+Programmers often use \code{NULL} as a default value for function arguments, and the code snippet in Figure~\ref{fig:null} exemplifies this.
 
@@
 
-To capture this behaviour, we introduce a nullable type \code{? T} into our
-type language and contract checking framework.  That way, programmers can
-succinctly ascribe types to null-by-default arguments (such as in the
-example in Figure~\ref{fig:null}).  This also served to reduce the number of
-polymorphic arguments we detected: Without this type, \AT{X\%} of
-polymorphic argument positions were of the form \code{null | T}.  Including
-a nullable type served to reduce observed polymorphism by \AT{Y\%}.
+To capture this behaviour, we introduce a nullable type \code{? T} into our type language and contract checking framework.
+That way, programmers can succinctly ascribe types to null-by-default arguments (such as in the example in Figure~\ref{fig:null}).
+This also served to reduce the number of polymorphic arguments we detected:
+Without this type, \AT{X\%} of polymorphic argument positions were of the form \code{null | T}.
+Including a nullable type served to reduce observed polymorphism by \AT{Y\%}.
 
@@
 
-Finally, Figure~\ref{fig:types} presents our final type language design, and
-the grammar of types that \contractr is equipped to handle.  \contractr will
-look for a top-level type declaration file in a package, with function types
-having the form $type \; ID \; T$, where $ID$ is the function name and $T$
-is a function type.  These function types have the form $\langle A_1 , ...,
-A_n\rangle \Rightarrow T$, where each of the $A_i$ arguments are either some
-type $T_i$, or the vararg type $...$.  Recall that functions with a vararg
-parameter (\code{...}) can be called with arbitrarily many values, and extra
-values are lumped together in the vararg $...$ value.  Internally, the type
-$...$ is implemented as a ${\bf list} \langle any \rangle$ type.
-
-We have discussed most of these types throughout this section, with some
-notable exceptions.  ${\bf expression}$, ${\bf language}$, ${\bf symbol}$,
-${\bf externalptr}$, ${\bf pairlist}$, ${\bf S4}$, and ${\bf weakref}$ types
-occur \AT{rarely}, and are by-and-large uninteresting to most R programmers.
-\AT{Fill in details of these basic types.}  One other type we did not
-discuss is the union type $T_1 \, | \, T_2$, which is supported by
-\contractr and generated by \typetracer when it cannot find some way to
-consolidate argument types (for example, if an argument is observed to be
-${\bf chr}$ and ${\bf int}$, \typetracer will ascribe type ${\bf chr} \, |
-\, {\bf int}$ to it).  And while not generated by \typetracer (recall
-discussion in Section~\ref{subsec:structs}), user-defined struct types have
-the form ${\bf struct}\langle F_1, ..., F_n \rangle$, with fields $F_i$
-being name $:$ type pairs.
+Finally, Figure~\ref{fig:types} presents our final type language design, and the grammar of types that \contractr is equipped to handle.
+\contractr will look for a top-level type declaration file in a package, with function types having the form $type \; ID \; T$, where $ID$ is the function name and $T$ is a function type.
+These function types have the form $\langle A_1 , ..., A_n\rangle \Rightarrow T$, where each of the $A_i$ arguments are either some type $T_i$, or the vararg type $...$.
+Recall that functions with a vararg parameter (\code{...}) can be called with arbitrarily many values, and extra values are lumped together in the vararg $...$ value.
+Internally, the type $...$ is implemented as a ${\bf list} \langle any \rangle$ type.
+
+We have discussed most of these types throughout this section, with some notable exceptions.
+${\bf expression}$, ${\bf language}$, ${\bf symbol}$, ${\bf externalptr}$, ${\bf pairlist}$, ${\bf S4}$, and ${\bf weakref}$ types occur \AT{rarely}, and are by-and-large uninteresting to most R programmers.
+\AT{Fill in details of these basic types.}
+% $S$ and $V$ are scalar and vector primitives, with vectors taking the form $S[]$.
+% The scalar type $S$ can be either a base type $B$ (one of the 6 primitives in R), or an NA-able base type (roughly, $\string^B$).
+One other type we did not discuss is the union type $T_1 \, | \, T_2$, which is supported by \contractr and generated by \typetracer when it cannot find some way to consolidate argument types (for example, if an argument is observed to be ${\bf chr}$ and ${\bf int}$, \typetracer will ascribe type ${\bf chr} \, | \, {\bf int}$ to it).
+And while not generated by \typetracer (recall discussion in Section~\ref{subsec:structs}), user-defined struct types have the form ${\bf struct}\langle F_1, ..., F_n \rangle$, with fields $F_i$ being name $:$ type pairs.
 
@@
 %
-\subsubsection{Results}
-
-\AT{How many failing assertions?
-How many functions had these failed assertions?
-Which packages?
-How was the coverage?
-What are patterns in the actual types of the failed assertions?
-In the expected types?
-How many times did the code trigger the contracts?}
-
-Overall, we found that only \AT{2\%} of contract assertions failed.
-This indicates two things: firstly, the corpus of packages we selected is quite well-tested, and much of the intended function behaviour is exemplified in the package example code. 
-Secondly, and more importantly, this success indicates that our analysis framework is able to generate function types which accurately capture the behaviour of the function. 
-Of the \AT{2\%} of failing contract assertions, we picked up on a few significant patterns.
-
-Firstly, we found that \AT{10\%} of the failing assertions were due to \code{double}-typed values being passed to function arguments expecting an \code{integer}.
-While we cannot say for sure that {\it all} of these cases could be handled by coercion, it is worth noting that coercion from doubles to integers is exceedingly common in R: the practice is not codified in the language spec, but many of R's internal C functions explicitly perform this coercion.
-\AT{We can put an example, e.g. nrow(df) is an integer but 5 is not. Also list access coerces the double to an int.}
-
-Next, \AT{30\%} of failing assertions were due to vector-typed primitive values being passed to function arguments expecting a \code{class<matrix>}.
-This case exemplifies something we have touched on earlier, namely that vectors can be readily considered to be unidimensional matrices.
-\AT{Is this coercion automatic? I've seen it mentioned often in the documentation.}
-
-Of the remaining \~60\% of failed contract assertions, we found quite an interesting pattern: the top as-of-yet unexplained contract failure, making up roughly 7\% of total failed assertions, is an assertion checking a ${\bf character}$ against the expected type ${\bf class} \langle bignum \rangle \, | \, {\bf raw}[]$.
-Further investigation revealed that most of these failures were coming from the \code{openssl::base64_encode} function, whose code is available in Figure~\ref{fig:base64-encode}.
-
-\begin{figure}[htbp]
-\begin{center}
-
-\begin{lstlisting}
-# The type according to typetracer:
-#' @type <bin: class<`bignum`> | raw[], linebreaks: logical> => character
-base64_encode <- function (bin, linebreaks = FALSE) 
-{
-    if (is.character(bin)) {
-        bin <- charToRaw(paste(bin, collapse = ""))
-    }
-    stopifnot(is.raw(bin))
-    .Call(R_base64_encode, bin, as.logical(linebreaks))
-}
-\end{lstlisting}
-
-\caption{Example of reverse dependency increasing code coverage.}
-\label{fig:base64-encode}
-\end{center}
-\end{figure}
-
-Contract failures are on the first argument, \code{bin}.
-This function is a wrapper for a C function (\code{R_base64_encode}) which encodes the input \code{bin} in base64.
-It would appear that the code for this function was never tested with a ${\bf character}$-typed input, even though it {\it is} valid.
-This is an example of a rather common phenomenon wherein the code of a package's reverse dependencies increases the coverage of some of the original package's functions, exemplified here by the code on line 6 never having been run in \code{openssl}'s own tests.
-
-To try to detect this pattern, we cross-referenced functions with failing contract checks against the code coverage for that function from the package's tests.
-For a given function, a coverage of X\% indicates that X\% of the code in the function body was executing by running package test, example, and vignette code.
-Of the remaining failed contracts, we found that \~28\% of them were from functions which had less than 100\% coverage, meaning that it is possible that \typetracer failed to synthesize the correct type for the function as not all possible executions were recorded.
-
-But what of the large amount (\~70\%) of assertion failures in functions with 100\% coverage?
-As it happens, many of these functions have very few lines of code, many having as few as a single line.
-The most interesting among these are functions which are the entry point for S3 dispatch (these functions typically have a single line of code, \code{UseMethod("InsertS3MethodNameHere")}).
-
-It turns out that the lion's share of the remaining polymorphism is related to S3 dispatch.
-S3 generics are naturally polymorphic, and we aim to define types to capture these cases, and implement a full-fledged object-oriented type system for R in future work.
-In the meantime, if we count out S3 methods, we find that only \AT{few} contract failures remain.
+\subsubsection{Breakdown of Contract Assertion Failures}
+
+We ran our evaluation on \NUMPKGSEVAL packages and recorded \NUMASSERTIONS total assertions.
+Overall, we found that only \PERCFAILEDASSERTIONS of contract assertions failed.
+This number is low, and suggests that the type signatures generated by running \typetracer on our corpus are indeed representative, and capture intended function behaviour.
+
+We breakdown the \PERCFAILEDASSERTIONS of failed assertions by type in Table~\ref{tbl:fail-breakdown-0}.
+By far the most commonly failed assertion is one asserting that a value with ${\bf dbl}[]$ type has type ${\bf class}\langle matrix \rangle$.
+This was a deliberate design decision, as coercion from vectors to matrices is ad hoc at best, and surely not a practice codified in the language.
+In a similar vein, another popular failing assertion is checking if a ${\bf dbl}[]$ has type ${bf int}[]$, another case of commonly performed coercion.
+We did not include these types of coercions in our type annotation framework as programmers cannot rely on them, and it is not always the case that the coercions are safe to perform.
+
+\begin{table}
+% BEGIN Autogenerated
+
+\begin{tabular}{llrrr}
+\toprule
+Actual Type & Expected Type & \# Failed Assertions & \% Total & Cumulative \%\\
+\midrule
+double[] & class<`matrix`> & 371896 & 27.37 & 27.37\\
+double[] & integer[] & 117212 & 8.62 & 35.99\\
+character & class<`bignum`> | raw[] & 100100 & 7.37 & 43.36\\
+double & (integer | null) & 60570 & 4.46 & 47.81\\
+double[] & class<`timeSeries`> & 58872 & 4.33 & 52.15\\
+\addlinespace
+class<matrix> & class<`timeSeries`> & 55005 & 4.05 & 56.19\\
+double[] & double & 53338 & 3.92 & 60.12\\
+double & class<`data.frame`> & 32553 & 2.40 & 62.51\\
+double[] & class<`data.frame`> & 31576 & 2.32 & 64.84\\
+double & integer & 20586 & 1.51 & 66.35\\
+\bottomrule
+\end{tabular}
+% END Autogenerated
+\label{tbl:fail-breakdown-0}
+\caption{Top types of contract assertion failures.}
+\end{table}
+
+In addition to the raw number of failing contract checks, we were interested in how many functions had a parameter where a contract check failed.
+Overall, we found that \PROPFUNSFAILEDCHECK of functions had a failing contract check.
+Digging a little deeper, we noticed that many of these functions were performing S3 dispatch, which is part of R's object-orientation framework which we will tackle in future work.
+Removing those functions, we see that \PROPFUNSFAILEDCHECKNOSTHREE of functions exhibited failing contract checks.
+
+Turning our attention now to arguments, we found that only \PROPARGSFAILINGASSERTS of function parameters exhibited a failed assertion.
+This is promising, as by-and-large it appears that \typetracer inferred argument type signatures which were not violated during testing.
+But what do these failing assertions look like?
+Table~\ref{tbl:fail-breakdown-0} showed the raw results from counting runtime occurrences, but that data alone does not tell the full story, as some failures may be overrepresented if e.g.
+a failing contract assertion was in a loop.
+Table~\ref{tbl:fail-breakdown-1} breaks down the results from Table~\ref{tbl:fail-breakdown-0} by parameter position, folding away multiple identical failed contract assertions for the same parameter position.
+The first row of this table reads: a value of type ${\bf chr}[]$ was passed to 183 different argument positions expecting a value of type ${\bf chr}$, representing 2.05\% of all arguments where a contract check failed.
+
+The main takeaway from Table~\ref{tbl:fail-breakdown-1} is that there appear to be no pathological kinds of failing assertions: programmers are not disproportionately passing values of one type to arguments expecting values of another.
+\AT{I need to touch up the discussion of this table.}
+
+% While the raw number of failing checks is interesting, we were also curious to see if certain kinds of type checks were overrepresented in the data.
+% A breakdown of the top failing type checks by kind can be found in Table~\ref{tbl:fail-breakdown-2}.
+% % Note the last row in the table: this is a check that would succeed in a proper type system with subtyping for class types, as data frames are subclasses of tibbles (\code{tbl}), and we will revisit these scenarios in future work.
+
+ \begin{table}
+% BEGIN Autogenerated
+
+ \begin{tabular}{llrrr}
+\toprule
+Actual Type & Expected Type & \# Arguments & \% Total & Cumulative \%\\
+\midrule
+ character[] & character & 183 & 2.05 & 2.05\\
+ \textasciicircum{}logical & null & 158 & 1.77 & 3.81\\
+ character & null & 117 & 1.31 & 5.12\\
+ double & null & 107 & 1.20 & 6.32\\
+ double[] & double & 100 & 1.12 & 7.44\\
+ class<data.frame> & ${\bf class} \langle data.frame, tbl, tbl\_df \rangle$ & 63 & 0.70 & 8.14\\
+ \bottomrule
+\end{tabular}
+
+% END Autogenerated
+ \label{tbl:fail-breakdown-1}
+ \caption{Most common failing contract checks.
+Note that for this table, we collapsed multiple identical failed assertions into one, meaning that single arguments with multiple (identical) failed assertions are only counted once.}
+ \end{table}
 

### 2020-05-15 paper/main.tex
more tables, more discussion

@@
 \newcommand{\NUMPKGSEVAL}{7000\xspace} % autogenerate
-\newcommand{\NUMASSERTIONS}{39210391\xspace} %autogenerate
+\newcommand{\NUMASSERTIONS}{62,171,573\xspace} %autogenerate
 \newcommand{\PROPFUNSFAILEDCHECK}{21.21\%\xspace}
@@
 \newcommand{\PROPARGSFAILINGASSERTS}{3.53\%\xspace} % autogen....
+\newcommand{\PERCSUCCARG}{96.46\%\xspace}
+\newcommand{\PERCSUCCFUNNOSTHREE}{84.52\%\xspace}
+\newcommand{\PERCSUCCFUNS}{78.79\%\xspace}
 
@@
 %
-\subsection{Evaluation of Types from Traces}
+\subsection{RQ2: How good are type signatures generated from traces?}
 \label{subsec:eval}
@@
 %
-\subsubsection{Breakdown of Contract Assertion Failures}
+\subsubsection{Overall Results}
 
@@
 Overall, we found that only \PERCFAILEDASSERTIONS of contract assertions failed.
-This number is low, and suggests that the type signatures generated by running \typetracer on our corpus are indeed representative, and capture intended function behaviour.
-
-We breakdown the \PERCFAILEDASSERTIONS of failed assertions by type in Table~\ref{tbl:fail-breakdown-0}.
+Additionally, we found that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of functions types were never violated.
+The number of immaculate function types increases to \PERCSUCCFUNNOSTHREE if we count out S3 functions:
+These S3 functions represent user-defined polymorphism which we do not tackle in this work.
+Overall, these numbers are promising, and suggest that the type signatures generated by running \typetracer on our corpus are indeed representative, and capture intended function behaviour.
+
+%
+%
+\subsubsection{Breakdown of Contract Assertion Failures}
+
+We breakdown the failed assertions by type in Table~\ref{tbl:fail-breakdown-0}.
 By far the most commonly failed assertion is one asserting that a value with ${\bf dbl}[]$ type has type ${\bf class}\langle matrix \rangle$.
@@
 Removing those functions, we see that \PROPFUNSFAILEDCHECKNOSTHREE of functions exhibited failing contract checks.
+\AT{More on these functions?}
 
 Turning our attention now to arguments, we found that only \PROPARGSFAILINGASSERTS of function parameters exhibited a failed assertion.
-This is promising, as by-and-large it appears that \typetracer inferred argument type signatures which were not violated during testing.
-But what do these failing assertions look like?
 Table~\ref{tbl:fail-breakdown-0} showed the raw results from counting runtime occurrences, but that data alone does not tell the full story, as some failures may be overrepresented if e.g.
@@
 The main takeaway from Table~\ref{tbl:fail-breakdown-1} is that there appear to be no pathological kinds of failing assertions: programmers are not disproportionately passing values of one type to arguments expecting values of another.
-\AT{I need to touch up the discussion of this table.}
-
-% While the raw number of failing checks is interesting, we were also curious to see if certain kinds of type checks were overrepresented in the data.
-% A breakdown of the top failing type checks by kind can be found in Table~\ref{tbl:fail-breakdown-2}.
-% % Note the last row in the table: this is a check that would succeed in a proper type system with subtyping for class types, as data frames are subclasses of tibbles (\code{tbl}), and we will revisit these scenarios in future work.
+In other words, our type language appears to match with programmer intuitions: had we failed to account for a key, type-based assumption that R programmers make, we would see it at the top of the table.
 
@@
 %
+\subsubsection{Patterns in Failed Contract Assertions}
+
+While Table~\ref{tbl:fail-breakdown-1} revealed no glaring patterns in failed assertions, there are a few subtle issues that we will discuss.
+% First, \PERCFAILEDVECTOSCA of failed checks broken down by parameter position failed due to vectors being passed where scalars were expected.
+
+We observed an interesting corner case and quirk of R in looking at functions with values that default to NA.
+In R, recall that there is a different type of NA for each primitive, and when the literal value \code{NA} appears in code, R assumes it to be the logical NA.
+For the purposes of our analysis, this has the effect of introducing $\string^{\bf lgl}$ types where they (probably) don't belong, particularly when programmers fail to effectively test their code.
+For example, a function header like \code{function(x, y=NA)} will have $\string^{\bf lgl}$ as part of the type of parameter \code{y}.
+We tried to control for this in our analysis, and indeed $\string^{\bf lgl}$ is a subtype of all NA-able numeric types.
+
+%
+%
 %

### 2020-05-15 paper/main.tex
stuff

@@
 
-Over the last decade, the R Project has become a key tool for implementing
-sophisticated data analysis algorithms in fields ranging from Computational
-Biology~\cite{R05} to Political Science~\cite{R:Keele:2008}. At the heart of
-the R project is a \emph{vectorized, dynamic, lazy, functional,
-  object-oriented} programming language with a rather unusual combination of
+The R Project is a key tool for data analysis.  At the heart of the R
+project is a \emph{vectorized, dynamic, lazy, functional, object-oriented}
+programming language with a rather unusual combination of
 features~\cite{ecoop12} designed to ease learning by non-programmer and
-enable rapid development of new statistical methods.  The language, commonly
-referred to as R was designed in 1993 by Ross Ihaka and Robert
-Gentleman~\cite{R96} as a successor to S~\cite{S88}.  First released in
-1995, under a GNU license, R rapidly became the lingua franca for
-statistical data analysis. Today, there are over 13,000 R packages available
-from repositories such as CRAN and Bioconductor.  With 55 R user groups
-world-wide, Smith~\cite{eco11} estimates that there are over 2 million
-end-users.
-
-As an introduction to R, consider the code snippet in Fig.~\ref{sample} from
-a top-level interaction where the user defines a function \code{normSum}
-that accepts vectors of integers, logicals, doubles and complex values and
-normalizes the vector with respect to its sum and rounds the results. The
-function definition does not require type annotations, and all operations
-transparently work on vectors of any length and different types.
-
-\begin{figure}[!hb]{\begin{lstlisting}
-> normSum <- function( m )  round( m / sum(m), 2)
-> normSum(c(1L,3L,6L))
-[1] 0.1 0.3 0.6
-> normSum(c(1.1,3.3,6.6))
-[1] 0.1 0.3 0.6
-> normSum(c(1.6,3.3,6.1))
-[1] 0.15 0.30 0.55
-> normSum(complex(r=rnorm(3),i=rnorm(3)))
-[1] 0.49+0.21i 0.30-0.18i 0.22-0.03i
-\end{lstlisting}}
-\caption{Sample R code}\label{sample}
-\end{figure}
+enable rapid development of new statistical methods.  The language was
+designed~\citet{R96} as a successor to S~\cite{S88}.
 
@@
 functions are evaluated only when needed, they are bundled in so-called
-promises which package the original expression (as an abstract syntax tree, or AST), its environment
-as well as the result of evaluating the expression. Promises can be
-leveraged for meta-programming as it is possible to retrieve the text of a
-promise and evaluate it in a different environment.
+promises which package the original expression (as an abstract syntax tree,
+or AST), its environment as well as the result of evaluating the
+expression. Promises can be leveraged for meta-programming as it is possible
+to retrieve the text of a promise and evaluate it in a different
+environment.
 

### 2020-05-15 paper/main.tex
r overview

@@
   stringstyle=\color{Gray},
-  numbers=left,
+%  numbers=left,
   showstringspaces=false
@@
 }
+\renewcommand{\k}[1]{{\tt #1}\xspace}
 
@@
 
-%
-\subsection{The R Programming Language}
+\subsection{The R Programming Language} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-In R, functions can be called with named parameters, arguments can have
-default values, and R also support variable argument lists.  To illustrate
-all of this in a single example, consider:
+In this paper, we will attempt infer type signature for functions. We now
+introduce some relevant concepts.  Functions can be called with named
+parameters, arguments can have default values, and R also support variable
+argument lists.  To illustrate all of this in a single example, consider:
 
 \begin{lstlisting}
-f <- function(x, ..., y=3) x + y
+  f <- function(x, ..., y=if(z==0) 1, z=0) x + y + if (missing(...)) 0 else c(...)
 \end{lstlisting}
@@
 \noindent
-Function \k{f} can be called with a single argument \code{f(3)}, with named
-argument \code{f(y=4,x=2)} and with a variable number of arguments,
-\code{f(1,2,3,4,y=5)}, all of these calls will return \code{6}.
-
-R has a number of features that are not crucial to the present
-discussion. We will mention some of them here for completeness.  In R, data
-structures are reference counted and have copy-on-write semantics, thus the
-assignment \code{x[12]<-3} results in an update to a copy of \code{x} unless
-the reference count on that object is 1.  This semantics gives R a
-functional flavor while allowing updating in place within loops (the first
-update copies, subsequent updates are performed on the copy). Arguments to
-functions are evaluated only when needed, they are bundled in so-called
-promises which package the original expression (as an abstract syntax tree,
-or AST), its environment as well as the result of evaluating the
-expression. Promises can be leveraged for meta-programming as it is possible
-to retrieve the text of a promise and evaluate it in a different
-environment.
-
-\subsubsection{Types of Data}
-\label{subsubsec:backgroundtypes}
-
-Before attempting to define a type system for R, we should understand the
-different kinds of values that programs operate on.  As we will see
-different notions of type may emerge depending on how granular we want to
-be.
-
-\renewcommand{\k}[1]{{\tt #1}\xspace}
+This function four formal parameters, \k x, \k{...}, \k y and \k z. Argument
+\k x can be bound positionally or passed by name. The vararg argument,
+\k{...}, is always passed positionally. The remaining two arguments must be
+passed by name.  Arguments \k y and \k z have default values, in the case of
+\k z this is a constant, but \k{y}'s default value is an expression that
+depends on the value of \k z. The body of the function will add parameters
+\k x and \k y with either the scalar 0 or the result of concatenating the
+varargs into a primitive vector. The function \k{missing} tests if an
+argument was explicitly passed. The following are some valid invocations
+of \k f:
+
+\begin{lstlisting}
+ > f(1)    
+ [1] 2             % a double vector, y is 0, ... is missing
+ > f(1,2) 
+ [1] 4             % a double vector, y is 0, ... is 2
+ > f(1,2,3)
+ [1] 4 5           % a double vector, y is 0, ... is 2, 3
+ > f(2,3,x=1)
+ [1] 4 5           % a double vector, y is 0, ... is 2, 3
+ > f(x=1, y= 1)
+ [1] 2             % a double vector, y is 1, ... is missing
+ > f(x=1, z= 1)
+ numeric(0)        % a double vector of length 0, y is NULL
+ > f(1L,2L,y=1L)
+ [1] 4             % an integer vector, y is integer 1, ... is integer 2
+ > f(1, y=c(1,2))
+ [1] 1 2           % a double vetor, y is 1, 2, ... is missing
+\end{lstlisting}
+
+\noindent
+The above hints at the polymorphism of the language, \k f, may return a
+vector of integer or of double, the length of the vector depends on the
+length of the varargs and of \k x and \k y.  The language does not really
+differentiate between scalar and vectors.
+Some more exotic types that can be encountered include vectors of complex
+number and list of arbitrary types. Function \k f can be invoked with those
+as well.
+
+\begin{lstlisting}
+ > c1 <- complex(re=1, im=2)
+ > c2 <- complex(re=2, im=1)
+ > f(c1)
+ [1] 1             % a double vector, x is complex, y is double
+ > f(c1,y=c2)
+ [1] 2+1i          % a complex vector, x and y are complex
+ > l1 <- list(1,2)
+ > l2 <- list(c1,c2)
+ > f(l1) 
+ [1] 1             % a double vector, x is a list of doubles
+ > f(l1,y=l2)
+ [[1]]             % a list of complex, x is a list of doubles
+ [1] 1+2i          %                      y is a list of complex
+
+ [[2]]
+ [1] 2+1i
+\end{lstlisting}
 
@@
 context a \k{logical} will convert to \k{integer}, and an \k{integer} will
-convert to \k{double}.  Some off conversions can occur in corner cases, such
+convert to \k{double}.  Some odd conversions can occur in corner cases, such
 as \k{1<"2"} holds and \k{c(1,2)[1.6]} returns the first element of the
@@
 
-Scalar data types include the distinguished \k{NULL} value, which is also of
-type \k{NULL}, instance of classes written using the S4~\footnote{put footnote on S4 here} object system,
-closures and environments.  The implementation of R has a number of other
-types that are mostly not used by user code, they are listed in
-Figure~\ref{types} for reference.
-
 Over the years, programmers have found the need for a richer type structure
@@
 
-%
-%
-\section{Project Corpus}\label{sec:corpus}
+Scalar data types include the distinguished \k{NULL} value, which is also of
+ktype \k{NULL}, instance of classes written using the S4 object system,
+closures and environments.  The implementation of R has a number of other
+types that are mostly not used by user code, they are listed in
+Figure~\ref{types} for reference.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Project Corpus}\label{sec:corpus} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2020-05-15 paper/main.tex
3 wip

@@
 R has one builtin notion of type that can be queried by the \k{typeof}
-function. Figure~\ref{types} lists all of the builtin types that are
+function. Figure~\ref{rtypes} lists all of the builtin types that are
 provided by the language. They are the possible return values of
@@
 \k{weakref}}\\\hline
-\end{tabular}\caption{Builtin Types}\label{types}\end{wrapfigure}
+\end{tabular}\caption{Builtin Types}\label{rtypes}\end{wrapfigure}
 
@@
 closures and environments.  The implementation of R has a number of other
-types that are mostly not used by user code, they are listed in
-Figure~\ref{types} for reference.
+types listed in Figure~\ref{rtypes} for reference.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Design A Type Language for R}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\renewcommand{\null}{\texttt{\textbf{null}}\xspace}
+\newcommand{\any}{\texttt{\textbf{any}}\xspace}
+\newcommand{\environment}{\texttt{\textbf{environment}}\xspace}
+\newcommand{\expression}{\texttt{\textbf{expression}}\xspace}
+\newcommand{\Language}{\texttt{\textbf{language}}\xspace} %% Upper case!
+\newcommand{\externalptr}{\texttt{\textbf{externalptr}}\xspace}
+\renewcommand{\symbol}{\texttt{\textbf{symbol}}\xspace}
+\newcommand{\pairlist}{\texttt{\textbf{pairlist}}\xspace}
+\newcommand{\weakref}{\texttt{\textbf{weakref}}\xspace}
+\renewcommand{\int}{\texttt{\textbf{int}}\xspace}
+\newcommand{\chr}{\texttt{\textbf{chr}}\xspace}
+\newcommand{\dbl}{\texttt{\textbf{dbl}}\xspace}
+\newcommand{\lgl}{\texttt{\textbf{lgl}}\xspace}
+\newcommand{\clx}{\texttt{\textbf{clx}}\xspace}
+\newcommand{\raw}{\texttt{\textbf{raw}}\xspace}
+\newcommand{\VARGS}{\texttt{\textbf{...}}\xspace}
+\newcommand{\T}{\ensuremath{T}\xspace}
+\renewcommand{\S}{\ensuremath{S}\xspace}
+\newcommand{\V}{\ensuremath{V}\xspace}
+\newcommand{\A}{\ensuremath{A}\xspace}
+\newcommand{\ID}{\ensuremath{ID}\xspace}
+\newcommand{\F}{\ensuremath{F}\xspace}
+\newcommand{\B}{\ensuremath{B}\xspace}
+\newcommand{\FUN}[2]{\ensuremath{\langle #1 \rangle \rightarrow #2}\xspace}
+\newcommand{\STRUCT}[1]{\texttt{\textbf{struct}}\ensuremath{\langle #1\rangle}\xspace}
+\newcommand{\LIST}[1]{\texttt{\textbf{list}}\ensuremath{\langle #1\rangle}\xspace}
+\newcommand{\CLASS}[1]{\texttt{\textbf{class}}\ensuremath{\langle #1\rangle}\xspace}
+\newcommand{\VEC}[1]{#1\k{[]}\xspace}
+\newcommand{\NAVEC}[1]{\k{\string^}\!#1\k{[]}\xspace}
+
+This section sets out to propose a candidate design for a type language that
+can be used to describe the arguments and return values of functions in R.
+The goal is not to be the final design but rather a starting point for an
+iterative process.
+
+Fig.~\ref{types} presents our type language. An early design choice was to
+stay close to the R language and only depart in small, hopefully,
+non-controverwsial ways.  Functions types have the form \FUN{A_1, \dots,
+  A_n}\T where each $\A_i$ argument is either a type \T or \VARGS, a
+variable length argument list. The rest of this section details and
+motivates our design.
+
+\begin{figure}[!h] \noindent \small  \centering\begin{minipage}{.45\linewidth}
+\begin{tabular}{lclr}
+\T& ::= & \null         & \it unit type\\
+  & |   & \any          & \it top type\\
+  & |   & \environment  & \it environment type \\
+  & |   & \Language     & \it language type \\
+  & |   & \symbol       & \it symbol type \\
+  & |   & \S            & \it scalar type \\
+  & |   & \V            & \it vector type \\
+  & |   & \T~\k{|}\T    & \it union type \\
+  & |   & \k{?} \T      & \it nullable type \\
+  & |   & \FUN{A_1, ..., A_n}\T & \it function type \\
+  & |   & \STRUCT{F_1, ..., F_n} & \it struct type \\
+  & |   & \LIST{T}      & \it list type\\
+  & |   & \CLASS{ID_1, ..., ID_n}& \it class type\vspace{5pt}\\
+\end{tabular}\end{minipage}   \hfill
+\begin{minipage}{.45\linewidth}\begin{tabular}{lclr}
+\A & ::= & \T & \it arguments \\
+  & |   & \VARGS \vspace{5pt} \\
+\F & ::= & \ID\, :\,T & \it struct field\vspace{5pt}\\
+\V & ::= & ~\VEC\S &\\
+  & |   & \NAVEC\S & \it na vector types \vspace{5pt}\\
+\S & ::= & \int & \\
+  & |   & \chr \\
+  & |   & \dbl \\
+  & |   & \lgl \\
+  & |   & \clx \\
+  & |   & \raw \vspace{5pt}\\
+\end{tabular}\end{minipage}\caption{The R type language}\label{types}
+\end{figure}
+
+\subsection{Basic Types}
+
+The workhorse types in R are the vectorized primitives: integers, doubles,
+complex numbers, logicals (booleans), characters, and raws (bytes).  These
+are vectorized in that they are always considered to be vectors by the R
+runtime: for instance, even ``scalar'' numbers are considered to be
+unit-length vectors to R.  These primitive vectors are homogeneous, in that
+a vector of doubles contains only doubles.
+
+Related to these primitive types is one of R's notions of ``nullness'': for
+each of the primitives, R distinguishes a special ``NA'' value which
+represents missing data.  There is an NA for each of the primitive types
+(i.e. there is a double NA, integer NA, etc.).  Thus, even homogeneous
+vectors can represent missing data with their appropriate NA value.
+
+\subsubsection{Possibly-NA Primitives}
+
+A common primitive usage pattern is related to the presence of NAs:
+programmers explicitly check for NAs in primitive vectors, possibly
+sanitizing them if they are present.  NAs complicate practically every
+computation using primitive vectors, a fact that many programmers appear to
+be acutely aware of. For instance, consider the code in
+Figure~\ref{fig:na-example}.
+
+\begin{figure}[htbp]
+\begin{center}
+
+\begin{lstlisting}
+binom.profile <- function(x, n, ...) {
+  xn <- cbind(x = x, n = n)
+  ok <- !is.na(xn[, 1]) & !is.na(xn[, 2])
+  x <- xn[ok, "x"]
+  n <- xn[ok, "n"]
+  ...
+\end{lstlisting}
+
+\caption{NA checking example from the \code{binom} package.}
+\label{fig:na-example}
+\end{center}
+\end{figure}
+
+The \code{binom} package is a popular R package for computing confidence
+intervals for binomial experiments, and the \code{binom.profile} function
+does so using the profile likelihood, the details of which are not relevant
+here.  This code snippet highlights a data sanitization pattern appearing at
+the beginning of the function: the programmer first binds the vectors into a
+matrix (line 2), with one column for each of \code{x} and \code{n}, then
+finds rows where both columns are not NA (line 3), then extracts non-NA
+values and stores them into \code{x} and \code{n} respectively (lines 4-5).
+Importantly, this process is entirely quiet, and the user of
+\code{binom.profile} will not be notified if their data happens to contain
+NAs, and the NAs will be silently ignored.  In reality, the presence of NAs
+in data could signify some sort of data corruption, and quietly ignoring
+issues in data could lead users (read: statisticians) to draw flawed
+conclusions from erroneous data.
+
+To combat this, we introduce a modifier on primitive types: the type
+\code{^T} signifies that primitive type \code{T} might contain an NA value,
+while \code{T} signifies that \code{T} is NA-free.  The NA-free case has a
+shorter syntax as NA-free vectors are \AT{more common} than the alternative.
+If the \code{binom} developers were to annotate arguments \code{x} and
+\code{n} with such an NA-free type, \contractr will notify users when
+\code{binom.profile} is called with NA-full values, rather than just quietly
+ignoring them.
+
+Note that when combining call traces involving \code{^T} and \code{T} types,
+we subsume \code{T} types into \code{^T} types: if NA-free and NA-full
+vectors are passed to an argument, then we say that the argument can handle
+NA-full vectors.  This is equivalent to saying \code{T <: ^T}.
+
+%
+%
+\subsubsection{Scalar Primitives}
+
+scalars and often do high-level
+dimensionality checks on their data.  To illustrate, consider the code in
+Figure~\ref{fig:scalar-vector-example}.
+
+\begin{figure}[htbp]
+\begin{center}
+
+\begin{lstlisting}
+hankel.matrix <- function( n, x ) { 
+### arguments
+### n = a positive integer value for the order of the Hankel matrix
+### x = an order 2 * n + 1 vector of numeric values
+    ...
+    if ( n != trunc( n ) )
+        stop( "argument n ix not an integer" )
+    if ( !is.vector( x ) )
+        stop( "argument x is not a vector" )
+    m <- length( x )
+    if ( m < n )
+        stop( "length of argument x is less than n" )
+    ...
+\end{lstlisting}
+
+\caption{Scalar/vector example from the \code{matrixcalc} package.}
+\label{fig:scalar-vector-example}
+\end{center}
+\end{figure}
+
+The \code{hankel.matrix} function takes two arguments (as described in the
+comment in the code), and returns a Hankel matrix, which essentially is a
+matrix where the skew-diagonals have a desirable property.  The function
+body has a number of interesting checks: on lines 6-7, \code{n} undergoes an
+integer ``type check''; on lines 8-9, \code{x} undergoes a vector ``type
+check''; on line 11, \code{n} is assumed to be a scalar, as if it were a
+vector, the code \code{m < n} would essentially check \code{m < n[i]} for
+each element in \code{n}, returning a vector of comparison results, and the
+conditional of an if must be a scalar else R generates a warning.
+
+These checks exemplify functionality that we would like to include in our
+type annotation language, namely, we would like our type language to
+differentiate between scalars and vectors.  To that end, we introduce the
+following types into our type language: for some primitive type \code{T},
+the user may modify \code{T} with a modifier \code{[]} to indicate that it
+is a vector of \code{T}.  Perhaps surprisingly, we found that scalars
+\AT{occurred more frequently} than vectors, and for that reason we chose the
+shorter scalar type syntax as the default.  The NA-free modifier works as
+expected: \code{^T[]} indicates a vector of \code{T} that is free of NAs,
+while \code{^T} indicates a scalar which is not NA.
+
+%
+%
+\subsubsection{Matrices}
+
+Matrices represent another layer of complexity on top of vectorized
+primitives.  Internally, matrices are simply vectors with class ``matrix''
+and a \code{dims} attribute indicating matrix dimensions, and while not
+codified in the language semantics, many internal functions will coerce
+vectors to matrices automatically.  Relying on automatic coercion is unwise,
+and many functions check to ensure that their arguments are matrices.  For
+instance, consider the code in Figure~\ref{fig:matrix-example}.
+
+\begin{figure}[htbp]
+\begin{center}
+
+\begin{lstlisting}
+rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL,
+                             na.rm = FALSE, ...) {
+  # - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+  # Validate arguments
+  # - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+  # Argument 'x':
+  if (!is.matrix(x)) {
+    .Defunct(msg = sprintf("Argument 'x' is of class %s, but should be a matrix. 
+    The use of a %s is not supported, the correctness of the result is not guaranteed. 
+    Please update your code accordingly.", sQuote(class(x)[1]), sQuote(class(x)[1])))
+  }
+  ...
+\end{lstlisting}
+
+\caption{Matrix example from the \code{matrixStats} package.}
+\label{fig:matrix-example}
+\end{center}
+\end{figure}
+
+The \code{rowWeightedMeans} function calculates the weighted means of rows
+or columns of a matrix \code{x}.  This function exhibits good practice in
+type checking arguments, rightly producing a message to the user indicating
+that an explicit matrix should be passed.
+
+To capture cases like this, we include a \code{class<matrix>} type in our
+type annotation language.  If \code{rowWeightedMeans} were outfitted with
+such a type, \contractr can perform the check present in this code
+automatically, notifying the users passing ill-typed arguments.
+
+%
+%
+\subsubsection{Forgoing Dimensionality}
+
+Thus far we've seen instances where we introduce types that subsume existing
+argument validation code, but our intention is not to subsume {\it all} such
+code.  As an example of something we {\it do not} want to do, consider the
+code in Figure~\ref{fig:general-validation-example}.
+
+\begin{figure}[htbp]
+\begin{center}
+
+\begin{lstlisting}
+rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL,
+                             na.rm = FALSE, ...) {
+  ...
+  n <- ncol(x)
+  if (length(w) != n) {
+    stop("The length of argument 'w' is does not match the number of column in 'x': ", 
+    length(w), " != ", n)
+  }
+  ...
+\end{lstlisting}
+
+\caption{General dimensionality check example from the \code{matrixStats} package.}
+\label{fig:general-validation-example}
+\end{center}
+\end{figure}
+
+This code is from later in the same \code{rowWeightedMeans} function
+discussed previously.  Here, the length of the weights vector \code{w} is
+being checked for equality with the number of columns in the matrix
+\code{x}.  Specifying this relationship as a type amounts to {\it dependent
+  typing}, which we believe too complex to retrofit onto a dynamic data
+science language such as R.  Ultimately, we decided to leave these kinds of
+checks up to the programmer.
+
+%
+%
+%
+%
+\subsection{Lists}
+
+Separate from the vectorized primitives in R are lists: Whereas vectors are
+homogeneous and restricted to primitive types, lists can contain
+arbitrarily-typed data.  Further, lists and vectors have a different syntax
+for accessing elements, where lists are indexed using \code{[[} and vectors
+    using \code{[}.  R even builds on lists to create a more complex data
+      structure known as the {\it data frame}, an essential data structure
+      to R's functionality.
+
+We are faced with several possible designs for list types.  Chief concerns
+include the heterogeneity of list elements and the presence list element
+names, which allow list elements to be accessed in a struct-like manner.
+The next sections highlight some important design decisions for our list
+types.
+
+%
+%
+\subsubsection{Relationship Between Lists and Vectors}
+
+The ease with which vectors may be transformed into lists is perhaps
+indicative of a relationship between the two.  In R, all vectors can be
+converted to lists with a simple call to \code{as.list}, and lists to
+vectors with \code{unlist}, though care must be taken when unlisting a
+non-homogeneous list, as coercions will be performed where possible, and the
+function will simply do nothing if coercion is impossible.  That said, lists
+and vectors have different access syntax, so vectors cannot stand-in for
+lists without the \code{as.list} call above.
+
+In sum, we opted not to include a subtyping relationship between vectors and
+lists, keeping both types separate.  In practice, \AT{how often do \code{T[]
+    | list<T>} types occur?}.  During our evaluation we also found it rare
+that programmers would attempt to pass list-typed values to vector-typed
+arguments and vice versa, and we will discuss this further in
+Section~\ref{sec:evaluation}.
+
+%
+%
+\subsubsection{Environments}
+
+Environments in R are essentially lists with reference semantics: mutating a
+value in an environment is performed in-place, and that mutation can thus be
+shared with other code holding a reference to the environment.  They are
+used primarily by the R runtime to store variables (their primary function
+is that of an execution environment), though they also often used by
+programmers either to themselves reflect on local, previous, or global
+scope, or simply because of their reference semantics.  Environments are
+comprised of name, value pairs, and each environment has a reference to its
+parent environment.
+
+We thought to include detailed environment types in our type language by way
+of allowing environment types to define names and types of values in the
+environment.  Unfortunately, this led to an undue explosion in the ``size''
+of our types: often, environments had hundreds of values, particularly those
+which were obtained from R (by e.g. passing a reference to a local scope).
+These types were difficult to parse visually and did not feel informative.
+Further, environments \AT{did not occur often}, and giving them a richer
+type was not worth the increase in the visual complexity of our types.
+
+%
+%
+\subsubsection{Structs}
+\label{subsec:structs}
+
+We observed the use of consistently-named lists in \AT{a number of
+  functions}.  In R, ``fields'' of named lists can be accessed using the
+\code{$} operator, as illustrated in the code snippet in
+Figure~\ref{fig:struct-ex}.
+
+\begin{figure}[htbp]
+\begin{center}
+
+\begin{lstlisting}
+agricolae::cv.model <- function(x) {
+  suma2 <- sum(x$residual^2)
+  gl <- x$df.residual
+  promedio <- mean(x$fitted.values)
+  return(sqrt(suma2/gl)*100/promedio)
+}
+
+data(sweetpotato)
+model<-aov(yield~virus, data=sweetpotato)
+cv.model(model)
+\end{lstlisting}
+
+\caption{Struct example from a test in the \code{agricolae} package.}
+\label{fig:struct-ex}
+\end{center}
+\end{figure}
+
+In the code snippet, the function \code{cv.model} takes an argument
+\code{x}, which we observed to always be a {\it linear model}, which is part
+of R's built-in statistics functionality.  Internally, linear models are
+represented as lists with named elements, and one may access these named
+elements using the \code{$} syntax, as seen on lines 2-4 in the snippet.  To
+describe these kinds of values, we included a struct type in our type
+system, allowing users to specify name, type pairs as part of list type
+definitions.
+
+Unfortunately, the inclusion of structs complicated our analysis by
+polluting the data, illustrated on line 8 with the call
+\code{data(sweetpotato)}.  It is quite common for R example code to use some
+built-in data sets, such as \code{sweetpotato}, to show off some package
+functionality with ``real data''.  When \typetracer is configured to detect
+structs we saw a marked increase in misrepresentative types, as the types of
+functions which use these data sets (such as \code{aov} in
+Figure~\ref{fig:struct-ex}) became polluted with the names from the example
+sets.  We implemented a unification strategy for struct types, removing them
+if they co-occurred with list types, and turning them into lists if two
+co-occurring struct types shared no common names, but even so we found
+ourselves with function signatures that were full of noise introduced by the
+struct names.
+
+We still believe struct types to be useful to R programmers, and so our
+language supports the type and \contractr will check them, though we
+configured \typetracer to ignore them as they made the data substantially
+less useful.  On the positive side, we found that \AT{get the number, but
+  often} structs-typed values had a {\it class}, which helped motivate our
+inclusion of class-like types in our type annotation language.  Classes are
+discussed in more detail in Section~\ref{subsec:classes}.
+
+%
+\subsection{Classes}
+
+R has more than one notion of type.  The classes of function arguments are
+used by R as part of its dispatch mechanism, the most prevalent of which
+(\AT{we have that number, too!}) is S3 dispatch, in which R dispatches on
+the (S3) class of the first argument to a function, selecting the
+appropriate method to call for the class.
+
+Classes are very prevalent in R \AT{number}, and so we include a
+\code{class<class-name>} ... \AT{todo} \typetracer was outfit to report the
+class of values, but some care needed to be taken so as to not overreport
+class types, as {\it all} values have a class in R.  The main classes we
+were interested in were matrix, array (higher dimensional matrices), and
+data.frames, in addition to any user-defined classes.  In many cases
+\AT{number?}, values with a user-defined class are in reality lists with
+named elements, or structs.  Including class types, and having \typetracer
+report them in place of structs, significantly reduced the noise in the
+analysis and allowed us to gather more focused data.
+
+When set to ignore classes, \typetracer reports \AT{type} as the type of the
+function, as compared with \AT{other type} when configured to report
+classes.
+%
+A value can have multiple classes in R.  This is common practice in the
+\code{dplyr} package, which extends the functionality of data frames: a
+\code{data.frame} that passes through \code{dplyr} functions gains two
+additional classes (\code{tbl} and \code{tbl_df}), bringing the total to at
+least 3, in addition to whichever other classes that data frame happened to
+have.  Note that when looking to perform S3 dispatch, R will try to match
+{\it any} of the classes of the first argument to a dispatched function, so
+each of these classes are important.
+
+We found these so-called multi-classes to be surprisingly common, making up
+\AT{occurrence} of the occurrences of class-typed values.  One possible
+reason for this is that values can gain classes quite easily (e.g. the
+\code{dplyr} example discussed above), somewhat polluting the output of
+\typetracer.  Nevertheless, we found \AT{number} of occurrences of
+monomorphic multi-class arguments, meaning that the function was always
+called with a multi-class value in that argument position.
+
+In the end, we support these multi-class types by allowing \code{class<...>}
+types to contain a list of class names, so for instance a \code{dplyr}
+\code{tibble} would have type \code{class<data.frame, tbl, tbl_df>} in our
+type language, and when performing related contract assertions \contractr
+will check that the value has every class that make up its type.
+
+%
+\subsection{Nullable Types}
+
+One final, common pattern that we would like to highlight is the
+co-occurrence of \code{NULL} with other types of values.  In
+Section~\ref{subsec:basictypes} we discussed NA, and how NAs represent
+missing data for the primitive types, and more generally R provides a
+\code{NULL} value that programmers use to represent missing data for other
+types.  Programmers often use \code{NULL} as a default value for function
+arguments, and the code snippet in Figure~\ref{fig:null} exemplifies this.
+
+\begin{figure}[htbp]
+\begin{center}
+
+\begin{lstlisting}
+# NULL default argument example. 
+\end{lstlisting}
+
+\caption{Idiomatic NULL usage in R.}
+\label{fig:null}
+\end{center}
+\end{figure}
+
+To capture this behaviour, we introduce a nullable type \code{? T} into our
+type language and contract checking framework.  That way, programmers can
+succinctly ascribe types to null-by-default arguments (such as in the
+example in Figure~\ref{fig:null}).  This also served to reduce the number of
+polymorphic arguments we detected: Without this type, \AT{X\%} of
+polymorphic argument positions were of the form \code{null | T}.  Including
+a nullable type served to reduce observed polymorphism by \AT{Y\%}.
 

### 2020-05-15 evaluation/R/corpus-analysis.Rmd
Updates in corpus analysis

@@
 summary_corpus <- function(df, prefix) {
+  w_coverage <- filter(df, coverage_expr > 0)
+  w_revdeps <- filter(df, revdeps > 0)
+  w_runnable <- filter(df, !is.na(runnable_time))
+
   overview_table(
@@
 
-    r(paste(prefix, "with coverage"), filter(df, coverage_expr > 0)),
-    r(paste(prefix, "mean expr coverage"), percent(mean(df$coverage_expr, na.rm=T))),
-    r(paste(prefix, "median expr coverage"), percent(median(df$coverage_expr, na.rm=T))),
+    r(paste(prefix, "with coverage"), w_coverage),
+    r(paste(prefix, "mean expr coverage"), percent(mean(w_coverage))),
+    r(paste(prefix, "median expr coverage"), percent(median(w_coverage))),
 
@@
 
-    r(paste(prefix, "mean revdes"), mean(df$revdeps, na.rm=T)),
-    r(paste(prefix, "median revdes"), median(df$revdeps, na.rm=T)),
+    r(paste(prefix, "w revdes"), w_revdeps),
+    r(paste(prefix, "mean revdes"), mean(w_revdeps)),
+    r(paste(prefix, "median revdes"), median(w_revdeps)),
+    r(paste(prefix, "at least min revdep"), filter(w_revdeps, revdeps >= params$min_revdeps)),
 
-    r(paste(prefix, "runtime"), as.duration(sum(df$runnable_time, na.rm=T))),
-    r(paste(prefix, "mean runtime"), as.duration(mean(df$runnable_time, na.rm=T))),
-    r(paste(prefix, "median runtime"), as.duration(median(df$runnable_time, na.rm=T))),
+    r(paste(prefix, "runtime"), as.duration(sum(w_runnable$runnable_time))),
+    r(paste(prefix, "mean runtime"), as.duration(mean(w_runnable$runnable_time))),
+    r(paste(prefix, "median runtime"), as.duration(median(w_runnable$runnable_time))),
 
-    r(paste(prefix, "using roxygen"), sum(df$uses_roxygen, na.rm=T))
+    r(paste(prefix, "using roxygen"), sum(df$uses_roxygen, na.rm=T)),
   )

### 2020-05-15 paper/main.tex
3 wip

@@
 \newcommand{\raw}{\texttt{\textbf{raw}}\xspace}
+\newcommand{\NA}{\texttt{{NA}}\xspace}
 \newcommand{\VARGS}{\texttt{\textbf{...}}\xspace}
@@
 \begin{tabular}{lclr}
-\T& ::= & \null         & \it unit type\\
-  & |   & \any          & \it top type\\
+\T& ::= & \any          & \it top type\\
+  & |   & \null         & \it null type\\
   & |   & \environment  & \it environment type \\
@@
   & |   & \STRUCT{F_1, ..., F_n}   & \it struct type \\
-  & |   & \LIST{T}      & \it list type\\
-  & |   & \CLASS{ID_1, ..., ID_n}& \it class type\vspace{5pt}\\
+  & |   & \LIST\T                  & \it list type\\
+  & |   & \CLASS{ID_1, ..., ID_n}  & \it class type\vspace{5pt}\\
 \end{tabular}\end{minipage}   \hfill
@@
 \A & ::= & \T          & \it arguments \\
-  & |   & \VARGS \vspace{5pt} \\
-\F & ::= & \ID\, :\,T & \it struct field\vspace{5pt}\\
-\V & ::= & ~\VEC\S &\\
+   & |   & \VARGS      & \vspace{5pt} \\
+\F & ::= & \ID:~~\T & \it struct field\vspace{5pt}\\
+\V & ::= & ~\VEC\S     & \\
    & |   & \NAVEC\S    & \it na vector types \vspace{5pt}\\
@@
 
-\subsection{Basic Types}
-
-The workhorse types in R are the vectorized primitives: integers, doubles,
-complex numbers, logicals (booleans), characters, and raws (bytes).  These
-are vectorized in that they are always considered to be vectors by the R
-runtime: for instance, even ``scalar'' numbers are considered to be
-unit-length vectors to R.  These primitive vectors are homogeneous, in that
-a vector of doubles contains only doubles.
-
-Related to these primitive types is one of R's notions of ``nullness'': for
-each of the primitives, R distinguishes a special ``NA'' value which
-represents missing data.  There is an NA for each of the primitive types
-(i.e. there is a double NA, integer NA, etc.).  Thus, even homogeneous
-vectors can represent missing data with their appropriate NA value.
-
-\subsubsection{Possibly-NA Primitives}
-
-A common primitive usage pattern is related to the presence of NAs:
-programmers explicitly check for NAs in primitive vectors, possibly
-sanitizing them if they are present.  NAs complicate practically every
-computation using primitive vectors, a fact that many programmers appear to
-be acutely aware of. For instance, consider the code in
-Figure~\ref{fig:na-example}.
-
-\begin{figure}[htbp]
-\begin{center}
-
-\begin{lstlisting}
-binom.profile <- function(x, n, ...) {
-  xn <- cbind(x = x, n = n)
-  ok <- !is.na(xn[, 1]) & !is.na(xn[, 2])
-  x <- xn[ok, "x"]
-  n <- xn[ok, "n"]
-  ...
-\end{lstlisting}
-
-\caption{NA checking example from the \code{binom} package.}
-\label{fig:na-example}
-\end{center}
-\end{figure}
-
-The \code{binom} package is a popular R package for computing confidence
-intervals for binomial experiments, and the \code{binom.profile} function
-does so using the profile likelihood, the details of which are not relevant
-here.  This code snippet highlights a data sanitization pattern appearing at
-the beginning of the function: the programmer first binds the vectors into a
-matrix (line 2), with one column for each of \code{x} and \code{n}, then
-finds rows where both columns are not NA (line 3), then extracts non-NA
-values and stores them into \code{x} and \code{n} respectively (lines 4-5).
-Importantly, this process is entirely quiet, and the user of
-\code{binom.profile} will not be notified if their data happens to contain
-NAs, and the NAs will be silently ignored.  In reality, the presence of NAs
-in data could signify some sort of data corruption, and quietly ignoring
-issues in data could lead users (read: statisticians) to draw flawed
-conclusions from erroneous data.
-
-To combat this, we introduce a modifier on primitive types: the type
-\code{^T} signifies that primitive type \code{T} might contain an NA value,
-while \code{T} signifies that \code{T} is NA-free.  The NA-free case has a
-shorter syntax as NA-free vectors are \AT{more common} than the alternative.
-If the \code{binom} developers were to annotate arguments \code{x} and
-\code{n} with such an NA-free type, \contractr will notify users when
-\code{binom.profile} is called with NA-full values, rather than just quietly
-ignoring them.
-
-Note that when combining call traces involving \code{^T} and \code{T} types,
-we subsume \code{T} types into \code{^T} types: if NA-free and NA-full
-vectors are passed to an argument, then we say that the argument can handle
-NA-full vectors.  This is equivalent to saying \code{T <: ^T}.
-
-%
-%
-\subsubsection{Scalar Primitives}
-
-scalars and often do high-level
-dimensionality checks on their data.  To illustrate, consider the code in
-Figure~\ref{fig:scalar-vector-example}.
-
-\begin{figure}[htbp]
-\begin{center}
-
-\begin{lstlisting}
-hankel.matrix <- function( n, x ) { 
-### arguments
-### n = a positive integer value for the order of the Hankel matrix
-### x = an order 2 * n + 1 vector of numeric values
-    ...
-    if ( n != trunc( n ) )
-        stop( "argument n ix not an integer" )
-    if ( !is.vector( x ) )
-        stop( "argument x is not a vector" )
-    m <- length( x )
-    if ( m < n )
-        stop( "length of argument x is less than n" )
-    ...
-\end{lstlisting}
-
-\caption{Scalar/vector example from the \code{matrixcalc} package.}
-\label{fig:scalar-vector-example}
-\end{center}
-\end{figure}
-
-The \code{hankel.matrix} function takes two arguments (as described in the
-comment in the code), and returns a Hankel matrix, which essentially is a
-matrix where the skew-diagonals have a desirable property.  The function
-body has a number of interesting checks: on lines 6-7, \code{n} undergoes an
-integer ``type check''; on lines 8-9, \code{x} undergoes a vector ``type
-check''; on line 11, \code{n} is assumed to be a scalar, as if it were a
-vector, the code \code{m < n} would essentially check \code{m < n[i]} for
-each element in \code{n}, returning a vector of comparison results, and the
-conditional of an if must be a scalar else R generates a warning.
-
-These checks exemplify functionality that we would like to include in our
-type annotation language, namely, we would like our type language to
-differentiate between scalars and vectors.  To that end, we introduce the
-following types into our type language: for some primitive type \code{T},
-the user may modify \code{T} with a modifier \code{[]} to indicate that it
-is a vector of \code{T}.  Perhaps surprisingly, we found that scalars
-\AT{occurred more frequently} than vectors, and for that reason we chose the
-shorter scalar type syntax as the default.  The NA-free modifier works as
-expected: \code{^T[]} indicates a vector of \code{T} that is free of NAs,
-while \code{^T} indicates a scalar which is not NA.
-
-%
-%
-\subsubsection{Matrices}
-
-Matrices represent another layer of complexity on top of vectorized
-primitives.  Internally, matrices are simply vectors with class ``matrix''
-and a \code{dims} attribute indicating matrix dimensions, and while not
-codified in the language semantics, many internal functions will coerce
-vectors to matrices automatically.  Relying on automatic coercion is unwise,
-and many functions check to ensure that their arguments are matrices.  For
-instance, consider the code in Figure~\ref{fig:matrix-example}.
-
-\begin{figure}[htbp]
-\begin{center}
-
-\begin{lstlisting}
-rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL,
-                             na.rm = FALSE, ...) {
-  # - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-  # Validate arguments
-  # - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-  # Argument 'x':
-  if (!is.matrix(x)) {
-    .Defunct(msg = sprintf("Argument 'x' is of class %s, but should be a matrix. 
-    The use of a %s is not supported, the correctness of the result is not guaranteed. 
-    Please update your code accordingly.", sQuote(class(x)[1]), sQuote(class(x)[1])))
-  }
-  ...
-\end{lstlisting}
-
-\caption{Matrix example from the \code{matrixStats} package.}
-\label{fig:matrix-example}
-\end{center}
-\end{figure}
-
-The \code{rowWeightedMeans} function calculates the weighted means of rows
-or columns of a matrix \code{x}.  This function exhibits good practice in
-type checking arguments, rightly producing a message to the user indicating
-that an explicit matrix should be passed.
-
-To capture cases like this, we include a \code{class<matrix>} type in our
-type annotation language.  If \code{rowWeightedMeans} were outfitted with
-such a type, \contractr can perform the check present in this code
-automatically, notifying the users passing ill-typed arguments.
-
-%
-%
-\subsubsection{Forgoing Dimensionality}
-
-Thus far we've seen instances where we introduce types that subsume existing
-argument validation code, but our intention is not to subsume {\it all} such
-code.  As an example of something we {\it do not} want to do, consider the
-code in Figure~\ref{fig:general-validation-example}.
-
-\begin{figure}[htbp]
-\begin{center}
-
-\begin{lstlisting}
-rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL,
-                             na.rm = FALSE, ...) {
-  ...
-  n <- ncol(x)
-  if (length(w) != n) {
-    stop("The length of argument 'w' is does not match the number of column in 'x': ", 
-    length(w), " != ", n)
-  }
-  ...
-\end{lstlisting}
-
-\caption{General dimensionality check example from the \code{matrixStats} package.}
-\label{fig:general-validation-example}
-\end{center}
-\end{figure}
-
-This code is from later in the same \code{rowWeightedMeans} function
-discussed previously.  Here, the length of the weights vector \code{w} is
-being checked for equality with the number of columns in the matrix
-\code{x}.  Specifying this relationship as a type amounts to {\it dependent
-  typing}, which we believe too complex to retrofit onto a dynamic data
-science language such as R.  Ultimately, we decided to leave these kinds of
-checks up to the programmer.
-
-%
-%
-%
-%
-\subsection{Lists}
-
-Separate from the vectorized primitives in R are lists: Whereas vectors are
-homogeneous and restricted to primitive types, lists can contain
-arbitrarily-typed data.  Further, lists and vectors have a different syntax
-for accessing elements, where lists are indexed using \code{[[} and vectors
-    using \code{[}.  R even builds on lists to create a more complex data
-      structure known as the {\it data frame}, an essential data structure
-      to R's functionality.
-
-We are faced with several possible designs for list types.  Chief concerns
-include the heterogeneity of list elements and the presence list element
-names, which allow list elements to be accessed in a struct-like manner.
-The next sections highlight some important design decisions for our list
-types.
-
-%
-%
-\subsubsection{Relationship Between Lists and Vectors}
-
-The ease with which vectors may be transformed into lists is perhaps
-indicative of a relationship between the two.  In R, all vectors can be
-converted to lists with a simple call to \code{as.list}, and lists to
-vectors with \code{unlist}, though care must be taken when unlisting a
-non-homogeneous list, as coercions will be performed where possible, and the
-function will simply do nothing if coercion is impossible.  That said, lists
-and vectors have different access syntax, so vectors cannot stand-in for
-lists without the \code{as.list} call above.
-
-In sum, we opted not to include a subtyping relationship between vectors and
-lists, keeping both types separate.  In practice, \AT{how often do \code{T[]
-    | list<T>} types occur?}.  During our evaluation we also found it rare
-that programmers would attempt to pass list-typed values to vector-typed
-arguments and vice versa, and we will discuss this further in
-Section~\ref{sec:evaluation}.
-
-%
-%
-\subsubsection{Environments}
+\paragraph{Scalar vs. Vectors} Our type system distinguishes between
+vectors of length 1, and vectors of any dimensions. The six base types,
+integers, doubles, complex numbers, logicals (booleans), characters, and
+raws (bytes), can be either vectors (e.g., \VEC\int) or scalars (e.g.,
+\int).  It is worthwhile to note that a vector can happen to be of length 1
+and thus a scalar value is a subtype of the corresponding vector type.
+Vectors are homogeneous, in that a vector of doubles contains only doubles.
+
+\paragraph{Missing values} In R, a missing value is written
+\NA, each of basic types has its specific representation of \NA. Thus there
+is an \int missing observation as well as a double missing, and they are
+different.  The type system allows to distinguish between vectors that may
+contain missing observations (written \NAVEC\int) and those who are
+guarantee to be \NA-free (\VEC\int).  A \NA-free vector can be treated as a
+vector that may have missing observations.  We do not allow scalar values to
+be missings.
+
+\paragraph{Lists} Heterogeneous collections are implemented using lists.
+Lists and vectors are closely related. Any vector can be converted to a list
+with \code{as.list}, and lists to vectors with \code{unlist} (coercions may
+ensue). We support parametrization of the list type, \LIST\T, but for
+heterogenous lists this type will default to \any.
+
+\paragraph{Environments}
 
@@
 %
-\subsubsection{Structs}
-\label{subsec:structs}
+\paragraph{Structs}
 
@@
 %
-\subsection{Classes}
+\paragraph{Classes}
 
@@
 %
-\subsection{Nullable Types}
+\paragraph{Nullable Types}
 
@@
 arguments, and the code snippet in Figure~\ref{fig:null} exemplifies this.
-
-\begin{figure}[htbp]
-\begin{center}
-
-\begin{lstlisting}
-# NULL default argument example. 
-\end{lstlisting}
-
-\caption{Idiomatic NULL usage in R.}
-\label{fig:null}
-\end{center}
-\end{figure}
 

### 2020-05-15 evaluation/R/corpus.Rmd
Add additional data to corpus

@@
 
+#### Coverage details
+
+We have to manually add the header to the `coverage-details.csv` files.
+
+```{r process package coverage details}
+read_coverage_details <- function(file, package) {
+  read_csv(
+    file, 
+    col_names = c( 
+      "type",
+      "filename",
+      "functions",
+      "first_line",
+      "first_byte",
+      "last_line",
+      "last_byte",
+      "first_column",
+      "last_column",
+      "first_parsed",
+      "last_parsed",
+      "value"
+    ),
+    col_types=cols(
+      type = col_character(),
+      filename = col_character(),
+      functions = col_character(),
+      first_line = col_integer(),
+      first_byte = col_integer(),
+      last_line = col_integer(),
+      last_byte = col_integer(),
+      first_column = col_integer(),
+      last_column = col_integer(),
+      first_parsed = col_integer(),
+      last_parsed = col_integer(),
+      value = col_double() # some value seems to be too big to fit into an int
+    )
+  ) %>%
+    add_column(package, .before="type") %>%
+    rename(fun=functions)
+}
+
+if (file_exists(PACKAGE_COVERAGE_DETAILS)) {
+  coverage_details <- read_csv(
+    PACKAGE_COVERAGE_DETAILS,
+    col_types=cols(
+      package = col_character(),
+      type = col_character(),
+      filename = col_character(),
+      fun = col_character(),
+      first_line = col_double(),
+      first_byte = col_double(),
+      last_line = col_double(),
+      last_byte = col_double(),
+      first_column = col_double(),
+      last_column = col_double(),
+      first_parsed = col_double(),
+      last_parsed = col_double(),
+      value = col_double()
+    )
+  )
+} else {
+  coverage_details_files <- path(PACKAGE_COVERAGE_DIR, pkgs_coverage$package, "coverage-details.csv")
+  names(coverage_details_files) <- pkgs_coverage$package
+
+  stopifnot(all(file_exists(coverage_details_files)))
+
+  coverage_details <- imap_dfr(coverage_details_files, read_coverage_details)
+  write_csv(coverage_details, PACKAGE_COVERAGE_DETAILS)
+}
+```
+
+```{r function coverage}
+coverage_funs <- 
+  coverage_details %>%
+  filter(type=="all") %>%
+  group_by(package, fun) %>%
+  summarise(coverage=ifelse(n() > 0, sum(value > 0)/n(), NA))
+
+write_csv(coverage_funs, CRAN_COVERAGE_FUNCTIONS)
+```
+
 ### Package runs
@@
 ```{r load functions}
-pkgs_functions <- 
+pkgs_functions_raw <- 
   read_csv(
@@
       fun = col_character(),
-      num_params = col_double(),
-      has_elipsis = col_logical(),
-      exported = col_logical()
+      exported = col_logical(),
+      is_s3_dispatch = col_logical(),
+      is_s3_method = col_logical(),
+      params = col_character()
     )
-  ) %>%
+  )
+
+pkgs_functions_exported <-
+  pkgs_functions_raw %>%
   count(package, exported) %>%
-  mutate(type=ifelse(exported, "funs_public", "funs_private")) %>%
-  select(-exported) %>%
-  spread(key=type, value=n, fill=0) %>%
-  mutate(funs=funs_public+funs_private)
+  mutate(exported=ifelse(exported, "funs_public", "funs_private")) %>%
+  pivot_wider(
+    names_from=exported, 
+    values_from=n, 
+    values_fill=list(n=0)
+  )
+
+pkgs_functions_s3 <-
+  pkgs_functions_raw %>%
+  group_by(package) %>%
+  summarise(funs_s3_dispatch=sum(is_s3_dispatch), funs_s3_method=sum(is_s3_method))
+
+pkgs_functions <-
+  pkgs_functions_exported %>%
+  left_join(pkgs_functions_s3, by="package") %>%
+  mutate(
+    funs_s3_dispatch=replace_na(funs_s3_dispatch, 0),
+    funs_s3_method=replace_na(funs_s3_method, 0),
+    funs=funs_public+funs_private,
+    funs_s3=funs_s3_dispatch+funs_s3_method
+  )
+```
+
+### Package classes
+
+```{r load classes}
+pkgs_classes_raw <- 
+  read_csv(
+    PACKAGE_CLASSES,
+    col_types=cols(
+      package = col_character(),
+      class = col_character()
+    )
+  )
+
+pkgs_classes <- 
+  select(pkgs_loadable, package) %>%
+  left_join(count(pkgs_classes_raw, package), by="package") %>% 
+  rename(s3_classes=n) %>%
+  mutate(s3_classes=replace_na(s3_classes, 0))  
 ```
@@
 
-pkgs_revdeps <- count(pkgs_revdeps_raw, package) %>% rename(revdeps=n)
+pkgs_revdeps <- 
+  pkgs_revdeps_raw %>%
+  left_join(select(pkgs_runnable_code, package, runnable_code), by=c("revdep"="package")) %>%
+  group_by(package) %>%
+  summarise(revdeps=n(), revdeps_runnable_code=sum(runnable_code, na.rm=T))
 ```
@@
 ```{r runtime checks}
-pkgs_asserts <- 
+pkgs_asserts_raw <-
   read_csv(
@@
     )
-  ) %>%
+  )
+write_csv(pkgs_asserts_raw, CRAN_ASSERTS)
+
+pkgs_asserts <- 
+  pkgs_asserts_raw %>%
   count(package, fun_name) %>%
@@
   left_join(pkgs_functions, by="package") %>%
+  left_join(pkgs_classes, by="package") %>%
   left_join(pkgs_revdeps, by="package") %>%

### 2020-05-15 paper/main.tex
3

@@
   & |   & \FUN{A_1, ..., A_n}\T    & \it function type \\
-  & |   & \STRUCT{F_1, ..., F_n}   & \it struct type \\
   & |   & \LIST\T                  & \it list type\\
@@
    & |   & \VARGS      & \vspace{5pt} \\
-\F & ::= & \ID:~~\T & \it struct field\vspace{5pt}\\
 \V & ::= & ~\VEC\S     & \\
@@
 
+\paragraph{Nullables} 
+The type system has a \null type that is inhabited by a singleton
+\code{NULL} value. This value is often used as a sentinel or as a stand in
+for arguments that were not provided.  To capture this behavior, we
+introduce a nullable type \code{?}\T.  The difference between the \NA and
+\code{NULL} values is that \NA can be stored in primitive vectors.
+
 \paragraph{Lists} Heterogeneous collections are implemented using lists.
@@
 \paragraph{Environments}
-
-Environments in R are essentially lists with reference semantics: mutating a
-value in an environment is performed in-place, and that mutation can thus be
-shared with other code holding a reference to the environment.  They are
-used primarily by the R runtime to store variables (their primary function
-is that of an execution environment), though they also often used by
-programmers either to themselves reflect on local, previous, or global
-scope, or simply because of their reference semantics.  Environments are
-comprised of name, value pairs, and each environment has a reference to its
-parent environment.
-
-We thought to include detailed environment types in our type language by way
-of allowing environment types to define names and types of values in the
-environment.  Unfortunately, this led to an undue explosion in the ``size''
-of our types: often, environments had hundreds of values, particularly those
-which were obtained from R (by e.g. passing a reference to a local scope).
-These types were difficult to parse visually and did not feel informative.
-Further, environments \AT{did not occur often}, and giving them a richer
-type was not worth the increase in the visual complexity of our types.
-
-%
-%
-\paragraph{Structs}
-
-We observed the use of consistently-named lists in \AT{a number of
-  functions}.  In R, ``fields'' of named lists can be accessed using the
-\code{$} operator, as illustrated in the code snippet in
-Figure~\ref{fig:struct-ex}.
-
-\begin{figure}[htbp]
-\begin{center}
-
-\begin{lstlisting}
-agricolae::cv.model <- function(x) {
-  suma2 <- sum(x$residual^2)
-  gl <- x$df.residual
-  promedio <- mean(x$fitted.values)
-  return(sqrt(suma2/gl)*100/promedio)
-}
-
-data(sweetpotato)
-model<-aov(yield~virus, data=sweetpotato)
-cv.model(model)
-\end{lstlisting}
-
-\caption{Struct example from a test in the \code{agricolae} package.}
-\label{fig:struct-ex}
-\end{center}
-\end{figure}
-
-In the code snippet, the function \code{cv.model} takes an argument
-\code{x}, which we observed to always be a {\it linear model}, which is part
-of R's built-in statistics functionality.  Internally, linear models are
-represented as lists with named elements, and one may access these named
-elements using the \code{$} syntax, as seen on lines 2-4 in the snippet.  To
-describe these kinds of values, we included a struct type in our type
-system, allowing users to specify name, type pairs as part of list type
-definitions.
-
-Unfortunately, the inclusion of structs complicated our analysis by
-polluting the data, illustrated on line 8 with the call
-\code{data(sweetpotato)}.  It is quite common for R example code to use some
-built-in data sets, such as \code{sweetpotato}, to show off some package
-functionality with ``real data''.  When \typetracer is configured to detect
-structs we saw a marked increase in misrepresentative types, as the types of
-functions which use these data sets (such as \code{aov} in
-Figure~\ref{fig:struct-ex}) became polluted with the names from the example
-sets.  We implemented a unification strategy for struct types, removing them
-if they co-occurred with list types, and turning them into lists if two
-co-occurring struct types shared no common names, but even so we found
-ourselves with function signatures that were full of noise introduced by the
-struct names.
-
-We still believe struct types to be useful to R programmers, and so our
-language supports the type and \contractr will check them, though we
-configured \typetracer to ignore them as they made the data substantially
-less useful.  On the positive side, we found that \AT{get the number, but
-  often} structs-typed values had a {\it class}, which helped motivate our
-inclusion of class-like types in our type annotation language.  Classes are
-discussed in more detail in Section~\ref{subsec:classes}.
-
-%
+Environments are lists with reference semantics: mutating a value in an
+environment is performed in-place.  They are used to store variables and to
+escape from the copy-on-write semantics of other data types.  Environments
+are comprised of name, value pairs, and each environment has a reference to
+its parent environment.
+
 \paragraph{Classes} 
-
-R has more than one notion of type.  The classes of function arguments are
-used by R as part of its dispatch mechanism, the most prevalent of which
-(\AT{we have that number, too!}) is S3 dispatch, in which R dispatches on
-the (S3) class of the first argument to a function, selecting the
-appropriate method to call for the class.
-
-Classes are very prevalent in R \AT{number}, and so we include a
-\code{class<class-name>} ... \AT{todo} \typetracer was outfit to report the
-class of values, but some care needed to be taken so as to not overreport
-class types, as {\it all} values have a class in R.  The main classes we
-were interested in were matrix, array (higher dimensional matrices), and
-data.frames, in addition to any user-defined classes.  In many cases
-\AT{number?}, values with a user-defined class are in reality lists with
-named elements, or structs.  Including class types, and having \typetracer
-report them in place of structs, significantly reduced the noise in the
-analysis and allowed us to gather more focused data.
-
-When set to ignore classes, \typetracer reports \AT{type} as the type of the
-function, as compared with \AT{other type} when configured to report
-classes.
-%
-A value can have multiple classes in R.  This is common practice in the
-\code{dplyr} package, which extends the functionality of data frames: a
-\code{data.frame} that passes through \code{dplyr} functions gains two
-additional classes (\code{tbl} and \code{tbl_df}), bringing the total to at
-least 3, in addition to whichever other classes that data frame happened to
-have.  Note that when looking to perform S3 dispatch, R will try to match
-{\it any} of the classes of the first argument to a dispatched function, so
-each of these classes are important.
-
-We found these so-called multi-classes to be surprisingly common, making up
-\AT{occurrence} of the occurrences of class-typed values.  One possible
-reason for this is that values can gain classes quite easily (e.g. the
-\code{dplyr} example discussed above), somewhat polluting the output of
-\typetracer.  Nevertheless, we found \AT{number} of occurrences of
-monomorphic multi-class arguments, meaning that the function was always
-called with a multi-class value in that argument position.
-
-In the end, we support these multi-class types by allowing \code{class<...>}
-types to contain a list of class names, so for instance a \code{dplyr}
-\code{tibble} would have type \code{class<data.frame, tbl, tbl_df>} in our
-type language, and when performing related contract assertions \contractr
-will check that the value has every class that make up its type.
-
-%
-\paragraph{Nullable Types}
-
-One final, common pattern that we would like to highlight is the
-co-occurrence of \code{NULL} with other types of values.  In
-Section~\ref{subsec:basictypes} we discussed NA, and how NAs represent
-missing data for the primitive types, and more generally R provides a
-\code{NULL} value that programmers use to represent missing data for other
-types.  Programmers often use \code{NULL} as a default value for function
-arguments, and the code snippet in Figure~\ref{fig:null} exemplifies this.
-
-To capture this behaviour, we introduce a nullable type \code{? T} into our
-type language and contract checking framework.  That way, programmers can
-succinctly ascribe types to null-by-default arguments (such as in the
-example in Figure~\ref{fig:null}).  This also served to reduce the number of
-polymorphic arguments we detected: Without this type, \AT{X\%} of
-polymorphic argument positions were of the form \code{null | T}.  Including
-a nullable type served to reduce observed polymorphism by \AT{Y\%}.
+R has more than one notion of type. Values can be attributed, one important
+attribute is the class of a value.  A class is a list of names that are used
+to mimic object-oriented programming. The type system includes class types
+writte \CLASS{\ID_1,\dots,\ID_n} A value can have multiple classes.
+
+\paragraph{Unions} We support untagged unions of types written
+$\T_1 \k{|} \T_n$.
 
@@
 %
-\section{Type Language Design}
-\label{sec:typesystemdesign}
-
-Our corpora analysis reported \AT{number} call traces for over \AT{number}
-functions.  We distilled these call traces into function types which
-captured the dynamic behaviour of the functions.  We emphasize that this was
-an iterative process: initial analysis suggested a certain set of types,
-which we tried and implemented, and reanalysis suggested improvements, and
-so on.  In this section, we will present the final type annotation language
-design, which we believe captures the essence of R.  We will end the section
-with a formal type language grammar, and leading up to that we will touch on
-design decisions and weigh our choices against alternatives, speaking to the
-strengths and weaknesses of our chosen design.  We begin with basic,
-primitive types.
-
-%
-%
-%
-%
-\subsection{Basic Types}
-\label{subsec:basictypes}
-
-As discussed in Section~\ref{subsubsec:backgroundtypes}, the workhorse types
-in R are the vectorized primitives: integers, doubles, complex numbers,
-logicals (booleans), characters, and raws (bytes).  These are {\it
-  vectorized} in that they are always considered to be vectors by the R
-runtime: for instance, even ``scalar'' numbers are considered to be
-unit-length vectors to R.  These primitive vectors are homogeneous, in that
-a vector of doubles contains only doubles.
-
-Related to these primitive types is one of R's notions of ``nullness'': for
-each of the primitives, R distinguishes a special ``NA'' value which
-represents missing data.  There is an NA for each of the primitive types
-(i.e. there is a double NA, integer NA, etc.).  Thus, even homogeneous
-vectors can represent missing data with their appropriate NA value.
-
-As with everything in R, there is nuance in even the simplest cases.
-We will now discuss our design decisions relating to the vectorized primitives.
+\section{Left overs}
+
+SOME OF THIS STUFF CAN BE USED IN THE EVALUATION
 
@@
 %
-\subsection{Lists}
-
-Separate from the vectorized primitives in R are lists: Whereas vectors are
-homogeneous and restricted to primitive types, lists can contain
-arbitrarily-typed data.  Further, lists and vectors have a different syntax
-for accessing elements, where lists are indexed using \code{[[} and vectors
-    using \code{[}.  R even builds on lists to create a more complex data
-      structure known as the {\it data frame}, an essential data structure
-      to R's functionality.
-
-We are faced with several possible designs for list types.  Chief concerns
-include the heterogeneity of list elements and the presence list element
-names, which allow list elements to be accessed in a struct-like manner.
-The next sections highlight some important design decisions for our list
-types.
-
-%
 %
@@
 %
-%
-\subsubsection{Environments}
-
-Environments in R are essentially lists with reference semantics: mutating a
-value in an environment is performed in-place, and that mutation can thus be
-shared with other code holding a reference to the environment.  They are
-used primarily by the R runtime to store variables (their primary function
-is that of an execution environment), though they also often used by
-programmers either to themselves reflect on local, previous, or global
-scope, or simply because of their reference semantics.  Environments are
-comprised of name, value pairs, and each environment has a reference to its
-parent environment.
-
-We thought to include detailed environment types in our type language by way
-of allowing environment types to define names and types of values in the
-environment.  Unfortunately, this led to an undue explosion in the ``size''
-of our types: often, environments had hundreds of values, particularly those
-which were obtained from R (by e.g. passing a reference to a local scope).
-These types were difficult to parse visually and did not feel informative.
-Further, environments \AT{did not occur often}, and giving them a richer
-type was not worth the increase in the visual complexity of our types.
-
 %
@@
 %
-%
-%
-\subsection{Parametric and Higher-Order Function Types}
-
-Functions are first-class values in R.  \typetracer found \AT{number}
-functions to be higher-order, having at least one function type among their
-arguments.  They are markedly more common in a subset of packages, such as
-\AT{such as...}.  By-and-large, these higher-order functions are similar to
-well-known higher-order functions like map, filter, and fold.
-
-The truth is that higher-order function types are best served by {\it
-  parametric types} and parametric polymorphism, wherein type variables are
-used within a type declaration to specify type-based relationships between
-function arguments.  Unfortunately, the nature of dynamic analysis is such
-that \typetracer cannot soundly detect parametric polymorphism, as without a
-static analysis of the function body it is impossible to say if function
-execution will halt given certain argument types, and we saw some examples
-of this in
-Figures~\ref{fig:scalar-vector-example}~and~\ref{fig:matrix-example}.
-Without parametric types, higher-order function types become noisy and fail
-to tell the complete picture.
-
-Ultimately, we configured \typetracer to report only top-level function
-types (i.e., detect if a value is a function or not).  \AT{The data suggests
-  that} most of the use of higher-order functions is through map-, filter-,
-and apply-like functions, whose behaviour is best captured with parametric
-polymorphism.  Other common uses include higher-order functions operating on
-data frame columns (in packages such as \code{dplyr}), which would
-necessitate a more in-depth treatment of data frame types, which as we
-discussed is a huge endeavour in and of itself.  We aim to extend our type
-annotation framework with parametric, higher-order, and detailed data frame
-types together in future work.
-
-%
-%
-%
-%
-\subsection{Nullable Types}
-
-One final, common pattern that we would like to highlight is the
-co-occurrence of \code{NULL} with other types of values.  In
-Section~\ref{subsec:basictypes} we discussed NA, and how NAs represent
-missing data for the primitive types, and more generally R provides a
-\code{NULL} value that programmers use to represent missing data for other
-types.  Programmers often use \code{NULL} as a default value for function
-arguments, and the code snippet in Figure~\ref{fig:null} exemplifies this.
-
-\begin{figure}[htbp]
-\begin{center}
-
-\begin{lstlisting}
-# NULL default argument example. 
-\end{lstlisting}
-
-\caption{Idiomatic NULL usage in R.}
-\label{fig:null}
-\end{center}
-\end{figure}
-
-To capture this behaviour, we introduce a nullable type \code{? T} into our
-type language and contract checking framework.  That way, programmers can
-succinctly ascribe types to null-by-default arguments (such as in the
-example in Figure~\ref{fig:null}).  This also served to reduce the number of
-polymorphic arguments we detected: Without this type, \AT{X\%} of
-polymorphic argument positions were of the form \code{null | T}.  Including
-a nullable type served to reduce observed polymorphism by \AT{Y\%}.
-
-%
-%
-%
-%
-\subsection{Final Type Grammar}
-
-Finally, Figure~\ref{fig:types} presents our final type language design, and
-the grammar of types that \contractr is equipped to handle.  \contractr will
-look for a top-level type declaration file in a package, with function types
-having the form $type \; ID \; T$, where $ID$ is the function name and $T$
-is a function type.  These function types have the form $\langle A_1 , ...,
-A_n\rangle \Rightarrow T$, where each of the $A_i$ arguments are either some
-type $T_i$, or the vararg type $...$.  Recall that functions with a vararg
-parameter (\code{...}) can be called with arbitrarily many values, and extra
-values are lumped together in the vararg $...$ value.  Internally, the type
-$...$ is implemented as a ${\bf list} \langle any \rangle$ type.
-
-We have discussed most of these types throughout this section, with some
-notable exceptions.  ${\bf expression}$, ${\bf language}$, ${\bf symbol}$,
-${\bf externalptr}$, ${\bf pairlist}$, ${\bf S4}$, and ${\bf weakref}$ types
-occur \AT{rarely}, and are by-and-large uninteresting to most R programmers.
-\AT{Fill in details of these basic types.}  One other type we did not
-discuss is the union type $T_1 \, | \, T_2$, which is supported by
-\contractr and generated by \typetracer when it cannot find some way to
-consolidate argument types (for example, if an argument is observed to be
-${\bf chr}$ and ${\bf int}$, \typetracer will ascribe type ${\bf chr} \, |
-\, {\bf int}$ to it).  And while not generated by \typetracer (recall
-discussion in Section~\ref{subsec:structs}), user-defined struct types have
-the form ${\bf struct}\langle F_1, ..., F_n \rangle$, with fields $F_i$
-being name $:$ type pairs.
-
-\label{subsec:typegrammar}
-
-\begin{figure}          % Types
-    \noindent           % Collapse for readability.
-    \centering
-    \begin{minipage}{.45\linewidth}
-    \begin{flushleft}
-    % Types
-      $ \begin{array}{lclr}
-    D & ::= & {\bf type} \; ID \; T {\bf ;}  & \text{\it type declaration}\\
-    T & ::= & {\bf null} & \text{\it unit type} \\
-      & |   & {\bf any} & \text{\it top type} \\
-      & |   & {\bf environment} & \text{\it environment type} \\
-      & |   & {\bf expression} & \text{\it expression type} \\
-      & |   & {\bf language} & \text{\it language type} \\
-      & |   & {\bf symbol} & \text{\it symbol type} \\
-      & |   & {\bf externalptr} & \text{\it externalptr type} \\
-      & |   & {\bf pairlist} & \text{\it pairlist type} \\
-      & |   & {\bf s4} & \text{\it S4 type} \\
-      & |   & {\bf weakref} & \text{\it weakref type} \\
-      & |   & S & \text{\it scalar type} \\
-      & |   & V & \text{\it vector type} \\
-      & |   & \left( T \right) & \text{\it group type} \\
-      & |   & T_1 \; {\bf |} \; T_2 & \text{\it union type} \\
-      & |   & ? \; T & \text{\it nullable type} \\
-      & |   & \langle A_1 , ..., A_n\rangle \Rightarrow T & \text{\it function type} \\
-      & |   & {\bf struct}\langle F_1, ..., F_n \rangle & \text{\it struct type} \\
-      & |   & {\bf list} \langle T \rangle & \text{\it list type}\\
-      & |   & {\bf class}\langle ID_1, ..., ID_n \rangle & \text{\it class type}\vspace{5pt}\\
-    \end{array} $
-    \end{flushleft}
-    \end{minipage}
-    \hfill
-    \begin{minipage}{.45\linewidth}
-    \begin{flushright}
-    $ \begin{array}{lclr}
-    % Argument Types
-    A & ::= & T & \text{\it argument type} \\
-      & |   & {\bf ...} \vspace{5pt} \\
-    % Field Types
-    F & ::= & ID : T & \text{\it struct field}\vspace{5pt}\\
-    % Scalar Types
-    S & ::= & B & \text{\it scalar types} \\
-      & |   & N \vspace{5pt}\\
-    % Vector Types
-    V & ::= & S \lbrack \rbrack & \text{\it vector types} \vspace{5pt}\\
-    % Base Types
-    B & ::= & {\bf int} & \text{\it base types} \\
-      & |   & {\bf chr} \\
-      & |   & {\bf dbl} \\
-      & |   & {\bf lgl} \\
-      & |   & {\bf clx} \\
-      & |   & {\bf raw} \vspace{5pt}\\
-    % NA Types
-    N & ::= & {\bf \string^int} & \text{\it na types} \\
-      & |   & {\bf \string^chr} \\
-      & |   & {\bf \string^dbl} \\
-      & |   & {\bf \string^lgl} \\
-      & |   & {\bf \string^clx} \\
-\end{array} $
-    \end{flushright}
-    \end{minipage}
-    \caption{The R type annotation language.}
-    \label{fig:types}
-\end{figure}
-
-%
-%
-%
-%
-%
-%
 \section{Evaluation}
@@
 
-To evaluate the effectiveness of our type annotation language and contract checking framework, we conducted a large-scale experiment based on the corpus of 412 packages we used to inform our language design.
-This corpus represents some of the most widely used and thoroughly tested R libraries on CRAN.
-
-We configured \typetracer to infer the types of each of the functions that are visible to package clients.
-Recall that \typetracer's primary function is to collect call signatures from observed function calls, and we ran \typetracer on the test, example, and vignette code of all packages in the aforementioned corpus.
-These call traces were consolidated into a single function type for each visible package function:
-\typetracer employs unification strategies discussed in Section~\ref{sec:typesystemdesign} to keep the size of signatures in check and only generates an \code{any} type when missing and unused arguments were recored, and when a longjump caused function execution to abort prematurely.
-The types generated by \typetracer will be analyzed in more depth in Section~\ref{subsec:stats}, and a large-scale evaluation of the quality of these types will be discussed in Section~\ref{subsec:eval}.
+To evaluate the effectiveness of our type annotation language and contract
+checking framework, we conducted a large-scale experiment based on the
+corpus of 412 packages we used to inform our language design.  This corpus
+represents some of the most widely used and thoroughly tested R libraries on
+CRAN.
+
+We configured \typetracer to infer the types of each of the functions that
+are visible to package clients.  Recall that \typetracer's primary function
+is to collect call signatures from observed function calls, and we ran
+\typetracer on the test, example, and vignette code of all packages in the
+aforementioned corpus.  These call traces were consolidated into a single
+function type for each visible package function: \typetracer employs
+unification strategies discussed in Section~\ref{sec:typesystemdesign} to
+keep the size of signatures in check and only generates an \code{any} type
+when missing and unused arguments were recored, and when a longjump caused
+function execution to abort prematurely.  The types generated by \typetracer
+will be analyzed in more depth in Section~\ref{subsec:stats}, and a
+large-scale evaluation of the quality of these types will be discussed in
+Section~\ref{subsec:eval}.
 
@@
 \subsection{Is our simple type language expressive enough to type R code?} Functions in
-R can support arguments of multiple types by reflecting on the type to perform
-appropriate conversions and computations. On the one hand, this form of ad-hoc
-polymorphism makes R functions expressive; on the other hand, it requires
-support for polymorphism in the typing scheme, making the type system complex. 
-Our system lacks explicit support for polymorphism. Instead, it supports union and any
-types. A union type is inferred when values of multiple types are observed at a
-function parameter and return position. any type is inferred when an argument is
-missing or unused. Furthermore, list element types can be inferred as any if
-they have more than 5 different types. Values with many different types can
-result in a large number of alternatives in the type inferred for that position,
+R can support arguments of multiple types by reflecting on the type to
+perform appropriate conversions and computations. On the one hand, this form
+of ad-hoc polymorphism makes R functions expressive; on the other hand, it
+requires support for polymorphism in the typing scheme, making the type
+system complex.  Our system lacks explicit support for
+polymorphism. Instead, it supports union and any types. A union type is
+inferred when values of multiple types are observed at a function parameter
+and return position. any type is inferred when an argument is missing or
+unused. Furthermore, list element types can be inferred as any if they have
+more than 5 different types. Values with many different types can result in
+a large number of alternatives in the type inferred for that position,
 making it hard to read and unwieldy for end-users. To evaluate this
-''expressivity gap`' between our type inference scheme and R functions,
-we look at the number of different types in the parameter and return types of
+''expressivity gap`' between our type inference scheme and R functions, we
+look at the number of different types in the parameter and return types of
 the XXX functions in our corpus.
@@
 
-XXX% of the parameter positions are monomorphic. 
+XXX\% of the parameter positions are monomorphic. 
 \typetracer observes XXX calls to XXX functions with XXX arguments. This
@@
 
-We analyzed the type signatures generated by \typetracer in order to build up a profile of our type annotation language.
-We were interested in broad, high-level information, such as the distribution of generated types, the level of polymorphism observed in the types, and the distribution of sizes of union types.
+We analyzed the type signatures generated by \typetracer in order to build
+up a profile of our type annotation language.  We were interested in broad,
+high-level information, such as the distribution of generated types, the
+level of polymorphism observed in the types, and the distribution of sizes
+of union types.
 

### 2020-05-15 paper/main.tex
clenup

@@
 \newcommand{\any}{\texttt{\textbf{any}}\xspace}
-\newcommand{\environment}{\texttt{\textbf{environment}}\xspace}
+\newcommand{\environment}{\texttt{\textbf{envir}}\xspace}
 \newcommand{\expression}{\texttt{\textbf{expression}}\xspace}
-\newcommand{\Language}{\texttt{\textbf{language}}\xspace} %% Upper case!
+\newcommand{\Language}{\texttt{\textbf{lang}}\xspace} %% Upper case!
 \newcommand{\externalptr}{\texttt{\textbf{externalptr}}\xspace}
@@
   & |   & \environment  & \it environment type \\
-  & |   & \Language     & \it language type \\
-  & |   & \symbol       & \it symbol type \\
   & |   & \S            & \it scalar type \\
@@
 vector that may have missing observations.  We do not allow scalar values to
-be missings.
+be missings.  The type \raw does not allow {\NA}s.
 

### 2020-05-15 paper/main.tex
merging, and a few more words and macros

--- 2020-05-15 paper/main.tex
@@
 \newcommand{\YEARS}{20\xspace} %%TODO fix
-\newcommand{\PERCFAILEDASSERTIONS}{2\%\xspace}
-\newcommand{\NUMPKGSEVAL}{7000\xspace} % autogenerate
+\newcommand{\PERCFAILEDASSERTIONS}{2.19\%\xspace}
+\newcommand{\PERCFAILEDASSERTIONSWUNDEF}{2.23\%\xspace}
+\newcommand{\PERCASSERTIONSUNDEF}{0.04\%\xspace}
+\newcommand{\NUMPKGSEVAL}{7485\xspace} % autogenerate
 \newcommand{\NUMASSERTIONS}{62,171,573\xspace} %autogenerate
@@
 We ran our evaluation on \NUMPKGSEVAL packages and recorded \NUMASSERTIONS total assertions.
-Overall, we found that only \PERCFAILEDASSERTIONS of contract assertions failed.
+Overall, we found that only \PERCFAILEDASSERTIONSWUNDEF of contract assertions failed.
+We made a simplification in our analysis, capping the number of function arguments we generate types for to 20, and this resulted in only \PERCASSERTIONSUNDEF failed assertions, which suggests that the simplification we made did not greatly affect the results.
+Once we controlled for these, we found that \PERCFAILEDASSERTIONS of all assertions failed for reasons related to our generated types.
 Additionally, we found that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of functions types were never violated.
@@
 This was a deliberate design decision, as coercion from vectors to matrices is ad hoc at best, and surely not a practice codified in the language.
-In a similar vein, another popular failing assertion is checking if a ${\bf dbl}[]$ has type ${bf int}[]$, another case of commonly performed coercion.
+In a similar vein, another popular failing assertion is checking if a ${\bf dbl}[]$ has type ${\bf int}[]$, another case of commonly performed coercion.
 We did not include these types of coercions in our type annotation framework as programmers cannot rely on them, and it is not always the case that the coercions are safe to perform.
@@
 a failing contract assertion was in a loop.
+We were interested in a breakdown of type signature by parameter, and seeing how often those contracts were violated.
 Table~\ref{tbl:fail-breakdown-1} breaks down the results from Table~\ref{tbl:fail-breakdown-0} by parameter position, folding away multiple identical failed contract assertions for the same parameter position.

### 2020-05-15 paper/main.tex
tracer

@@
 \renewcommand{\to}{\ensuremath{\rightarrow}\xspace}
-\newcommand{\D}{\ensuremath{\small\vec{\mathtt D}}\xspace} % Double
-\newcommand{\I}{\ensuremath{\small\vec{\mathtt I}}\xspace} % Integer
-\renewcommand{\C}{\ensuremath{\small\vec{\mathtt C}}\xspace} % Character
-\renewcommand{\L}{\ensuremath{\small\vec{\mathtt L}}\xspace} % Logical
-\newcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
-\newcommand{\X}{\ensuremath{\small\vec{\mathtt X}}\xspace} % Complex
-\newcommand{\Y}{\ensuremath{\small\vec{\mathtt Y}}\xspace} % Symbol
-\newcommand{\sY}{\ensuremath{\small{\mathtt Y}}\xspace} % Symbol
-\newcommand{\sS}{\ensuremath{\small{\mathtt S}}\xspace} % S4
-\newcommand{\sF}{\ensuremath{\small{\mathtt F}}\xspace} % Closure
-\newcommand{\sE}{\ensuremath{\small{\mathtt E}}\xspace} % Env
-\renewcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
-\newcommand{\sN}{\ensuremath{\small{\mathtt N}}\xspace}     % Null
-%\renewcommand{\l}{\ensuremath{\small L<?>}\xspace}     % List
-\renewcommand{\l}{\ensuremath{\small\underline{\mathtt ?}}\xspace}     % List
-\newcommand{\sD}{\ensuremath{\small{\mathtt D}}\xspace} % Double
-\newcommand{\sI}{\ensuremath{\small{\mathtt I}}\xspace} % Integer
-\newcommand{\sC}{\ensuremath{\small{\mathtt C}}\xspace} % Character
-\newcommand{\sL}{\ensuremath{\small{\mathtt L}}\xspace} % Logical
-\newcommand{\sX}{\ensuremath{\small{\mathtt X}}\xspace} % Complex
-\newcommand{\sR}{\ensuremath{\small{\mathtt R}}\xspace} % Raw
-\newcommand{\ANY}{\ensuremath{\small{\mathtt ?}}\xspace}     % Any
-\newcommand{\lT}[1]{\ensuremath{\small\underline{\mathtt{#1}}}\xspace}     % list<T>
-\newcommand{\M}[1]{\ensuremath{\attr{\vec{\tt #1}}{mat}}\xspace}     % matrix
-\newcommand{\df}{\ensuremath{\attr{\l}{df}}\xspace}     % data.frame
-
-\newcommand{\contractr}{\emph{contractr}\xspace} % contractr
-\newcommand{\roxygen}{\emph{roxygen2}\xspace} % roxygen
-\newcommand{\typetracer}{\emph{typetracer}\xspace} % typetracer
-\newcommand{\rdt}{\emph{R-dyntrace}\xspace}
-\newcommand{\covr}{\emph{covr}\xspace}
+
+\renewcommand{\null}{\texttt{\textbf{null}}\xspace}
+\newcommand{\any}{\texttt{\textbf{any}}\xspace}
+\newcommand{\environment}{\texttt{\textbf{envir}}\xspace}
+\newcommand{\expression}{\texttt{\textbf{expression}}\xspace}
+\newcommand{\Language}{\texttt{\textbf{lang}}\xspace} %% Upper case!
+\newcommand{\externalptr}{\texttt{\textbf{externalptr}}\xspace}
+\renewcommand{\symbol}{\texttt{\textbf{symbol}}\xspace}
+\newcommand{\pairlist}{\texttt{\textbf{pairlist}}\xspace}
+\newcommand{\weakref}{\texttt{\textbf{weakref}}\xspace}
+\renewcommand{\int}{\texttt{\textbf{int}}\xspace}
+\newcommand{\chr}{\texttt{\textbf{chr}}\xspace}
+\newcommand{\dbl}{\texttt{\textbf{dbl}}\xspace}
+\newcommand{\lgl}{\texttt{\textbf{lgl}}\xspace}
+\newcommand{\clx}{\texttt{\textbf{clx}}\xspace}
+\newcommand{\raw}{\texttt{\textbf{raw}}\xspace}
+\newcommand{\NA}{\texttt{{NA}}\xspace}
+\newcommand{\VARGS}{\texttt{\textbf{...}}\xspace}
+\newcommand{\T}{\ensuremath{T}\xspace}
+\renewcommand{\S}{\ensuremath{S}\xspace}
+\newcommand{\V}{\ensuremath{V}\xspace}
+\newcommand{\A}{\ensuremath{A}\xspace}
+\newcommand{\ID}{\ensuremath{ID}\xspace}
+\newcommand{\F}{\ensuremath{F}\xspace}
+\newcommand{\B}{\ensuremath{B}\xspace}
+\newcommand{\FUN}[2]{\ensuremath{\langle #1 \rangle \rightarrow #2}\xspace}
+\newcommand{\STRUCT}[1]{\texttt{\textbf{struct}}\ensuremath{\langle #1\rangle}\xspace}
+\newcommand{\LIST}[1]{\texttt{\textbf{list}}\ensuremath{\langle #1\rangle}\xspace}
+\newcommand{\CLASS}[1]{\texttt{\textbf{class}}\ensuremath{\langle #1\rangle}\xspace}
+\newcommand{\VEC}[1]{#1\k{[]}\xspace}
+\newcommand{\NAVEC}[1]{\k{\string^}\!#1\k{[]}\xspace}
+
+\newcommand{\contractr}{{\sf ContractR}\xspace} % contractr
+\newcommand{\roxygen}{{\sf Roxygen2}\xspace} % roxygen
+\newcommand{\typetracer}{{\sf Typetracer}\xspace} % typetracer
+\newcommand{\rdt}{{\sf R-dyntrace}\xspace}
+\newcommand{\covr}{{\sf Covr}\xspace}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Design A Type Language for R}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\renewcommand{\null}{\texttt{\textbf{null}}\xspace}
-\newcommand{\any}{\texttt{\textbf{any}}\xspace}
-\newcommand{\environment}{\texttt{\textbf{envir}}\xspace}
-\newcommand{\expression}{\texttt{\textbf{expression}}\xspace}
-\newcommand{\Language}{\texttt{\textbf{lang}}\xspace} %% Upper case!
-\newcommand{\externalptr}{\texttt{\textbf{externalptr}}\xspace}
-\renewcommand{\symbol}{\texttt{\textbf{symbol}}\xspace}
-\newcommand{\pairlist}{\texttt{\textbf{pairlist}}\xspace}
-\newcommand{\weakref}{\texttt{\textbf{weakref}}\xspace}
-\renewcommand{\int}{\texttt{\textbf{int}}\xspace}
-\newcommand{\chr}{\texttt{\textbf{chr}}\xspace}
-\newcommand{\dbl}{\texttt{\textbf{dbl}}\xspace}
-\newcommand{\lgl}{\texttt{\textbf{lgl}}\xspace}
-\newcommand{\clx}{\texttt{\textbf{clx}}\xspace}
-\newcommand{\raw}{\texttt{\textbf{raw}}\xspace}
-\newcommand{\NA}{\texttt{{NA}}\xspace}
-\newcommand{\VARGS}{\texttt{\textbf{...}}\xspace}
-\newcommand{\T}{\ensuremath{T}\xspace}
-\renewcommand{\S}{\ensuremath{S}\xspace}
-\newcommand{\V}{\ensuremath{V}\xspace}
-\newcommand{\A}{\ensuremath{A}\xspace}
-\newcommand{\ID}{\ensuremath{ID}\xspace}
-\newcommand{\F}{\ensuremath{F}\xspace}
-\newcommand{\B}{\ensuremath{B}\xspace}
-\newcommand{\FUN}[2]{\ensuremath{\langle #1 \rangle \rightarrow #2}\xspace}
-\newcommand{\STRUCT}[1]{\texttt{\textbf{struct}}\ensuremath{\langle #1\rangle}\xspace}
-\newcommand{\LIST}[1]{\texttt{\textbf{list}}\ensuremath{\langle #1\rangle}\xspace}
-\newcommand{\CLASS}[1]{\texttt{\textbf{class}}\ensuremath{\langle #1\rangle}\xspace}
-\newcommand{\VEC}[1]{#1\k{[]}\xspace}
-\newcommand{\NAVEC}[1]{\k{\string^}\!#1\k{[]}\xspace}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Designing A Type Language for R}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+\paragraph{Unions} We support untagged unions of types written
+$\T_1 \k{|} \T_n$.
+
 \paragraph{Nullables} 
@@
 
+\paragraph{Classes} 
+R has more than one notion of type. Values can be attributed, one important
+attribute is the class of a value.  A class is a list of names that are used
+to mimic object-oriented programming. The type system includes class types
+written \CLASS{\ID_1,\dots,\ID_n}.
+
 \paragraph{Environments}
@@
 
-\paragraph{Classes} 
-R has more than one notion of type. Values can be attributed, one important
-attribute is the class of a value.  A class is a list of names that are used
-to mimic object-oriented programming. The type system includes class types
-writte \CLASS{\ID_1,\dots,\ID_n} A value can have multiple classes.
-
-\paragraph{Unions} We support untagged unions of types written
-$\T_1 \k{|} \T_n$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Project Corpus}\label{sec:corpus} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-For this paper we have selected \CorpusLoadable packages consisting of
-\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native
-code (C/C++/Fortran). Figure~\ref{fig:corpus} shows all the packages, the
-size of the dots reflects the size of the project in terms of both R and
-native lines of code\footnote{ Lines of source code reported excludes
-  comments and blank lines, counted by \emph{cloc}, \cf
-  \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the
-expression code coverage in percents and the y-axis gives the number of
-reverse dependencies in log scale. Dotted lines indicate means.  Packages
-with over \PackageSizeOutierRnd lines of code are annotated.
-
-\begin{figure}[!h]  \centering
-  \includegraphics[width=.8\linewidth]{plots/corpus.pdf}
-  \caption{Corpus overview}\label{fig:corpus}
-\end{figure}
-
-These packages come from the Comprehensive R Achive Network
-(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of
-R code with over \AllCranRnd packages\footnote{CRAN receives about 6 new
-  package submissions a day~\cite{Ligges2017}} containing over \AllRCodeRnd
-and \AllNativeCodeRnd R and native code respectively. Unlike other open code
-repositories like GitHub, CRAN is a curated repository. Each submitted
-package must abide to a number of well-formedness rules that are
-automatically checked asserting certain quality. Most relevant for this work
-is that all of the runnable code (including code snippets from examples and
-vignettes) is tested and only a successfully running package is admitted in
-the archive.
-
-We have downloaded and installed all available CRAN packages. Out of the
-\AllCranRnd packages, we managed to install \AllLoadableRnd. The main reason
-for this is that \rdt is based on R 3.5.0 and some of the packages are no
-longer compatible with it. Some packages also require extra native
-dependencies which were not present on our servers.  We have two criteria
-for including a package into the corpus:
-\begin{inparaenum}[(1)]
-\item have runnable code that covers significant part of the package source
-  code from which type signatures could be inferred, and
-\item have some reverse dependencies that allows us to evaluate the inferred
-  types on runnable code from these dependent packages.
-\end{inparaenum}
-The concrete thresholds were set to at least \ThresholdCodeCoverage of
-expression coverage and at minimum \ThresholdRevdeps reverse dependencies.
-
-The code coverage was computed for each package using
-\covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R
-tool.  The reverse package dependencies were extracted from the package
-metadata using builtin function.
-
-%
-%
-%
-\section{Methodology}
-
-In this section, we will describe our approach to data collection and
-analysis through \typetracer, a dynamic analysis built in the \R-dyntrace
-framework.  We will then describe the process by which we distill the rich
-information we collect with \typetracer to more digestible ``types'', though
-we discuss the precise details of our type annotation language in
-Section~\ref{sec:types}.  Finally, we discuss \contractr, a package built in
-R which supports our type language and can generate and check contracts for
-function types.
-
-%
-\subsection{R-dyntrace}
-R-dyntrace is a framework for performing dynamic analysis of R. It was designed
-and used previously for a large-scale study of laziness in R~\cite{oopsla19} It
-is an instrumented R Virtual Machine based on GNU-R version 3.5.0. It exposes
-hooks from key points inside the R interpreter to which user defined callbacks
-can be attached for gathering information about the program state at those
-points. This information can be used to generate detailed program traces for
-online or offline analysis. R-dyntrace provides hooks for function entry and
-exit, S3/S4 dispatch, non-local jumps (because of longjumps in the R interpreter
-on exceptions and other non-local returns), promise creation and forcing,
-variable definition and mutation and garbage collection. garbage collection
-entry and exit, function entry and exit.
-
-%
-%
-\subsection{typetracer and Initial Analysis}
-
-To help inform the design of our type system, and to ensure that it aligns
-with the day-to-day usage of R, we performed a corpus analysis of some of
-the most widely used R packages.  The corpus is discussed in detail in
-Section~\ref{sec:corpus}.  We built \typetracer, our dynamic analysis, in
-the R-dyntrace~\cite{oopsla19} dynamic analysis framework which is itself
-built on a modified version of R targeting R version 3.5.0---we refer the
-reader to Section~\ref{sec:r-dyntrace} for more details on the framework.
-
-%
-%
-\subsubsection{Overview}
+%
+%
+\section{Extracting and Checking Signatures}
+
+For this paper, we have built tooling to (\emph{a}) automate the extraction
+of raw type signatures from execution traces, (\emph{b}) infer type
+signatures from a set of raw types, and (\emph{c}) validate the inferred
+signatures by the means of contracts. This section presents each step
+of this pipeline.
+
+\subsection{Extracting raw type signature from traces}
+
+We implemented \typetracer an automated tool for extracting raw type
+signatures from execution traces of R programs. The goal of this tool is to
+output a tuple $\langle f, t_1, \dots, t_n, t\rangle$ for each function call
+during the execution of a program, where $f$ is an identifier for a
+function, $t_i$ are type-level summaries of the arguments and $t$ is a
+summary of the return value.
+
+While this task is seemingly simple, the details and their proverbial devil
+are surprisingly tricky to get right and to scale to long running programs.
+To avoid starting from scratch, our implementation reuses an open source
+dynamic analysis framework for R named \rdt~\cite{oopsla19} which consists
+of an instrumented R Virtual Machine based on GNU-R version 3.5.0. The
+framework exposes hooks at key points in the R interpreter to which user
+defined callbacks can be attached. These hooks include function entry and
+exit, method dispatch for the S3 and S4 object systems, the longjumps used
+by the interpreter to implement non-local exit, creation and forcing of
+promises (lazy evaluation), variable definition, value creation, mutation
+and garbage collection.
+
+\paragraph{Raw types}
+The type information output by the tool includes the \emph{type tag} of each
+value. The R internal types are translated to the type name of the proposed
+type system. The next bit of information output for each value is its
+\emph{class}.  The class is an optional list of names, it may be absent,
+and, in some cases, it may be implicit (i.e. the interpreter blesses some
+values with the \k{matrix} and \k{array} classes even when they have no
+attribute). Finally, for each value, the tool outputs the list of
+\emph{attribute names} (other than \k{class}).
+
+Depending on the value's type tag, the tool collects further information:
+
+\begin{itemize}
+\item For vectors, the number of dimensions and their names are output;
+\item For lists, a recursive traversal collects element types and
+  the name attributes of elements.  
+  TODO: HOW DO WE OUTPUT HETEROGENEOUS NAMES???
+\item For promises, we carefully dig into the promise value {\it without
+  forcing it}, and collect information about the innermost value. If the
+  innermost value is the unbound value, then the promise is an as-of-yet
+  unevaluated expression. In this case, we do our best to guess the value of
+  the unevaluated expression: if the expression is not a language
+  expression, we collect information about it, otherwise we can say nothing
+  without evaluating it and ascribe it type ``any''. 
+\end{itemize}
+
+For vectors (and lists), we additionally iterate through the elements,
+checking if the values are NA (resp. NULL); if none are NA (resp. NULL), we
+ascribe the NA-free (resp. NULL-free) tag to the value.  Arguments that are
+part of a function's varargs (denoted \ldots in function definitions) are
+tagged as such and ignored, as varargs can contain anything (and their type
+will be equivalent to and any-typed list).
+
 The goal of our analysis is to discover argument types for functions,
@@
 written out later.
-%
-%
-\subsubsection{Type Information}
-Once a value is obtained, either on function return or when an argument
-promise is forced, we can probe it without interfering with program behaviour.
-We query the value for the following information:
-
-\begin{itemize}
-\item the value's type tag according to R's runtime. R's runtime implements
-  type tags as a typedef, e.g. \code{INTSXP} for integers, and, confusingly,
-  \code{VECSXP} for lists;
-\item its class, which is a list of string class names. Typically, the class
-  of a value is held in an attribute with name ``class'', though some
-  classes are blessed by the R runtime, such as matrix and array classes;
-\item its attributes, a list of metadata attached to the value.
-\end{itemize}
-
-Now, depending on the value's type tag, we collect further information:
-
-\begin{itemize}
-\item For R's primitive types, we collect its dimension, and if it has a
-  names attribute we collect the names as well;
-\item For lists, we recursively collect the types of list elements;
-\item For lists {\it with a names attribute}, we collect the names as well
-  as the types of the slots that those names refer to (data.frames fall into
-  this category);
-\item For promises, we carefully dig into the promise value {\it without
-  forcing it}, and collect information about the innermost value. If the
-  innermost value is the unbound value, then the promise is an as-of-yet
-  unevaluated expression. In this case, we do our best to guess the value of
-  the unevaluated expression: if the expression is not a language
-  expression, we collect information about it, otherwise we can say nothing
-  without evaluating it and ascribe it type ``any''. \footnote{\AT{I think
-      this is confusing, reword.}}
-\end{itemize}
-
-For vectors (and lists), we additionally iterate through the elements,
-checking if the values are NA (resp. NULL); if none are NA (resp. NULL), we
-ascribe the NA-free (resp. NULL-free) tag to the value.  Arguments that are
-part of a function's varargs (denoted \ldots in function definitions) are
-tagged as such and ignored, as varargs can contain anything (and their type
-will be equivalent to and any-typed list).
-
-Now that we've seen the high-level picture of our approach, we will elaborate on some key details and discuss particular challenges that we faced.
 
@@
 Another possibility is that a value is used {\it without being forced}, as
-is the case when values are ``metaprogrammed''.
-This is common practice in e.g.
-the \code{tidyverse}~\footnote{\AT{either talk about it here or in background}} ecosystem, as they use the metaprogramming capabilities of R to implement a DSL for data analysis.
+is the case when values are ``metaprogrammed''.  This is common practice in
+e.g. the \code{tidyverse}~\footnote{\AT{either talk about it here or in
+    background}} ecosystem, as they use the metaprogramming capabilities of
+R to implement a DSL for data analysis.
 
@@
 
-%
-%
-%
-%
-\subsection{Post-Processing and Type Ascription}
+\subsection{Inferring type signatures from raw types}
 
@@
 %
-\subsection{contractr}
-\label{sec:contractr}
+\subsection{Checking types signatures with contracts}
 
@@
 %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Project Corpus}\label{sec:corpus} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+For this paper we have selected \CorpusLoadable packages consisting of
+\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native
+code (C/C++/Fortran). Figure~\ref{fig:corpus} shows all the packages, the
+size of the dots reflects the size of the project in terms of both R and
+native lines of code\footnote{ Lines of source code reported excludes
+  comments and blank lines, counted by \emph{cloc}, \cf
+  \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the
+expression code coverage in percents and the y-axis gives the number of
+reverse dependencies in log scale. Dotted lines indicate means.  Packages
+with over \PackageSizeOutierRnd lines of code are annotated.
+
+\begin{figure}[!h]  \centering
+  \includegraphics[width=.98\linewidth]{plots/corpus.pdf}
+  \caption{Corpus overview}\label{fig:corpus}
+\end{figure}
+
+These packages come from the Comprehensive R Achive Network
+(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of
+R code with over \AllCranRnd packages\footnote{CRAN receives about 6 new
+  package submissions a day~\cite{Ligges2017}} containing over \AllRCodeRnd
+and \AllNativeCodeRnd R and native code respectively. Unlike other open code
+repositories like GitHub, CRAN is a curated repository. Each submitted
+package must abide to a number of well-formedness rules that are
+automatically checked asserting certain quality. Most relevant for this work
+is that all of the runnable code (including code snippets from examples and
+vignettes) is tested and only a successfully running package is admitted in
+the archive.
+
+We have downloaded and installed all available CRAN packages. Out of the
+\AllCranRnd packages, we managed to install \AllLoadableRnd. The main reason
+for this is that \rdt is based on R 3.5.0 and some of the packages are no
+longer compatible with it. Some packages also require extra native
+dependencies which were not present on our servers.  We have two criteria
+for including a package into the corpus:
+\begin{inparaenum}[(1)]
+\item have runnable code that covers significant part of the package source
+  code from which type signatures could be inferred, and
+\item have some reverse dependencies that allows us to evaluate the inferred
+  types on runnable code from these dependent packages.
+\end{inparaenum}
+The concrete thresholds were set to at least \ThresholdCodeCoverage of
+expression coverage and at minimum \ThresholdRevdeps reverse dependencies.
+
+The code coverage was computed for each package using
+\covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R
+tool.  The reverse package dependencies were extracted from the package
+metadata using builtin function.
+
+%
 %

### 2020-05-15 paper/main.tex
merging

--- 2020-05-15 paper/main.tex
@@
 \newcommand{\PERCSUCCFUNS}{78.79\%\xspace}
+\newcommand{\PERCCALLSBADFUNSINTESTS}{5.38\%\xspace}
 
@@
 Removing those functions, we see that \PROPFUNSFAILEDCHECKNOSTHREE of functions exhibited failing contract checks.
-\AT{More on these functions?}
+These are functions which were under-tested, as calls to these functions represent only \PERCCALLSBADFUNSINTESTS of recorded calls during \typetracer's run on the core corpus to infer types.
 
@@
  \label{tbl:fail-breakdown-1}
- \caption{Most common failing contract checks.
+ \caption{Most common failing contract checks by type.
 Note that for this table, we collapsed multiple identical failed assertions into one, meaning that single arguments with multiple (identical) failed assertions are only counted once.}
@@
 %
-\subsubsection{Patterns in Failed Contract Assertions}
-
-While Table~\ref{tbl:fail-breakdown-1} revealed no glaring patterns in failed assertions, there are a few subtle issues that we will discuss.
+% \subsubsection{Patterns in Failed Contract Assertions}
+
+% While Table~\ref{tbl:fail-breakdown-1} revealed no glaring patterns in failed assertions, there are a few subtle issues that we will discuss.
 % First, \PERCFAILEDVECTOSCA of failed checks broken down by parameter position failed due to vectors being passed where scalars were expected.
 
-We observed an interesting corner case and quirk of R in looking at functions with values that default to NA.
-In R, recall that there is a different type of NA for each primitive, and when the literal value \code{NA} appears in code, R assumes it to be the logical NA.
-For the purposes of our analysis, this has the effect of introducing $\string^{\bf lgl}$ types where they (probably) don't belong, particularly when programmers fail to effectively test their code.
-For example, a function header like \code{function(x, y=NA)} will have $\string^{\bf lgl}$ as part of the type of parameter \code{y}.
-We tried to control for this in our analysis, and indeed $\string^{\bf lgl}$ is a subtype of all NA-able numeric types.
+% We observed an interesting corner case and quirk of R in looking at functions with values that default to NA.
+% In R, recall that there is a different type of NA for each primitive, and when the literal value \code{NA} appears in code, R assumes it to be the logical NA.
+% For the purposes of our analysis, this has the effect of introducing $\string^{\bf lgl}$ types where they (probably) don't belong, particularly when programmers fail to effectively test their code.
+% For example, a function header like \code{function(x, y=NA)} will have $\string^{\bf lgl}$ as part of the type of parameter \code{y}.
+% We tried to control for this in our analysis, and indeed $\string^{\bf lgl}$ is a subtype of all NA-able numeric types.
 

### 2020-05-15 evaluation/R/evaluate-type-analysis.Rmd
Update static type analysis
Add graph for unions

@@
   type_analysis_datafile: ../data/type-analysis/merged/merged.csv
+  graph_dirpath: ../../paper/plots/
 ---
@@
 suppressPackageStartupMessages(library(tibble))
+
+source("latex-macros.R")
+
+save_graph <- function(plot, filename, device = cairo_pdf, width=12, height=4, ...) {
+    ggsave(filename = file.path(params$graph_dirpath, filename), plot = plot, device = device, width=width, height=height, ...)
+    plot
+}
 
@@
 ```{r incompatible-functions, echo=FALSE, eval=TRUE}
+
+###type_compatibility <-
+###    type_analysis_data %>%
+###    group_by(package_name, function_name, outer_alternative) %>%
+###    summarize(parameter_count = length(unique(parameter_position))) %>%
+###    ungroup() %>%
+###    group_by(package_name, function_name) %>%
+###    mutate(compatible = length(unique(parameter_count)) == 1) %>%
+###    ungroup()
+
 type_compatibility <-
     type_analysis_data %>%
-    group_by(package_name, function_name, outer_alternative) %>%
-    summarize(parameter_count = length(unique(parameter_position))) %>%
-    ungroup() %>%
     group_by(package_name, function_name) %>%
-    mutate(compatible = length(unique(parameter_count)) == 1) %>%
+    summarize(compatible = length(unique(outer_alternative)) == 1) %>%
     ungroup()
@@
 arrange(desc(parameter_count)) %>%
-mutate(percentage = round(100 * parameter_count / sum(parameter_count), 2),
-       cumulative_percentage = round(100 * cumsum(parameter_count) / sum(parameter_count), 2))
+mutate(percentage = round(parameter_count / sum(parameter_count), 4),
+       cumulative_percentage = round(cumsum(parameter_count) / sum(parameter_count), 4))
 
 datatable(union_frequency)
+
+total_parameter_count <- sum(union_frequency$parameter_count)
+
+print(total_parameter_count)
+
+union_frequency_folded <-
+    union_frequency %>%
+    select(alternative_count, parameter_count, percentage) %>%
+    filter(alternative_count > 3) %>%
+    summarize(alternative_count = "> 3", parameter_count = sum(parameter_count)) %>%
+    mutate(percentage = round(as.double(parameter_count) / total_parameter_count, 3)) %>%
+    print()
+
+union_frequency_folded <-
+    union_frequency %>%
+    select(alternative_count, parameter_count, percentage) %>%
+    filter(alternative_count <= 3) %>%
+    mutate(alternative_count = as.character(alternative_count)) %>%
+    rbind(union_frequency_folded)
+
+datatable(union_frequency_folded)
+
+(union_frequency_folded %>%
+ggplot(aes(alternative_count, percentage)) +
+geom_col() +
+scale_y_continuous(sec.axis = sec_axis(~ . * total_parameter_count, labels = count_labels), labels = relative_labels) +
+labs(x="Types", y="Parameters") +
+scale_x_discrete(limits=c("> 3", 3, 2, 1)) +
+coord_flip() +
+theme(legend.title = element_blank(), legend.position = "top")) %>%
+save_graph("union_frequency.pdf")
 
@@
 datatable(union_types)
+
+single_union_types <-
+union_data %>%
+filter(alternative_count == 1) %>%
+group_by(kind) %>%
+summarize(count = n())%>%
+ungroup() %>%
+arrange(desc(count)) %>%
+mutate(percentage = round(100 * count / sum(count), 1),
+       cumsum_percentage = round(100 * cumsum(count) / sum(count), 1))
+
+datatable(single_union_types)
 ```

### 2020-05-15 paper/main.tex
Add graph for unions

--- 2020-05-15 paper/main.tex
@@
 system complex.  Our system lacks explicit support for
-polymorphism. Instead, it supports union and any types. A union type is
+polymorphism. Instead, it supports union, any and ... types. A union type is
 inferred when values of multiple types are observed at a function parameter
-and return position. any type is inferred when an argument is missing or
-unused. Furthermore, list element types can be inferred as any if they have
-more than 5 different types. Values with many different types can result in
+and return position, any type is inferred when an argument is missing or
+unused, and ... is inferred for vararg parameters. Furthermore, list element
+types can be inferred as any if they have more than 5 different types. Values with many different types can result in
 a large number of alternatives in the type inferred for that position,
@@
 the XXX functions in our corpus.
+
+\begin{figure}[!h]  \centering
+  \includegraphics[width=\linewidth]{plots/union_frequency.pdf}
+  \caption{Types inferred for each position}\label{fig:unionfreq}
+\end{figure}
 

### 2020-05-15 paper/main.tex
4.1 wip

@@
 The type information output by the tool includes the \emph{type tag} of each
-value. The R internal types are translated to the type name of the proposed
-type system. The next bit of information output for each value is its
-\emph{class}.  The class is an optional list of names, it may be absent,
-and, in some cases, it may be implicit (i.e. the interpreter blesses some
-values with the \k{matrix} and \k{array} classes even when they have no
-attribute). Finally, for each value, the tool outputs the list of
-\emph{attribute names} (other than \k{class}).
+value. The R internal types are translated to names in the proposed type
+system. The next bit of information output is the \emph{class} which is an
+optional list of names, it may be absent, and, in some cases, it may be
+implicit (i.e. the interpreter blesses some values with the \k{matrix} and
+\k{array} classes even when they have no attributes).
 
@@
 \begin{itemize}
-\item For vectors, the number of dimensions and their names are output;
-\item For lists, a recursive traversal collects element types and
-  the name attributes of elements.  
-  TODO: HOW DO WE OUTPUT HETEROGENEOUS NAMES???
-\item For promises, we carefully dig into the promise value {\it without
-  forcing it}, and collect information about the innermost value. If the
-  innermost value is the unbound value, then the promise is an as-of-yet
-  unevaluated expression. In this case, we do our best to guess the value of
-  the unevaluated expression: if the expression is not a language
-  expression, we collect information about it, otherwise we can say nothing
-  without evaluating it and ascribe it type ``any''. 
+\item For vectors, the number of dimensions and the present of \NA values.
+\item For lists, a recursive traversal collects element types.
+\item For promises, we attempt to establish the type of its result or output
+  \any.
 \end{itemize}
 
-For vectors (and lists), we additionally iterate through the elements,
-checking if the values are NA (resp. NULL); if none are NA (resp. NULL), we
-ascribe the NA-free (resp. NULL-free) tag to the value.  Arguments that are
-part of a function's varargs (denoted \ldots in function definitions) are
-tagged as such and ignored, as varargs can contain anything (and their type
-will be equivalent to and any-typed list).
+\noindent
+To obtain raw types, we make use of R's C FFI and use low-level machinery to
+collect type tags and attributes from the R runtime.  The types that we
+provide to users are constructed during post-processing, and rely on the
+detailed information made available by these low-level reflection
+mechanisms.
+
+\paragraph{Varargs}
+Arguments that are part of a function's varargs (denoted \VARGS) are
+ignored, as type \any is output for the entire varag construct.
+
+\paragraph{Promises}
 
@@
 
-%
-%
-\subsubsection{Challenges}
-\label{subsec:typetracer-challenges}
-
-We grappled with a number of corner cases related to R's non-strict semantics.
-
-\begin{itemize}
-
-\item First, we wanted to ensure that \typetracer captured types for
-  unevaluated arguments.  Unfortunately, not every argument of a function is
-  necessarily used during every function execution, meaning that some
-  arguments may remain unevaluated (as they are never forced), and there are
-  a few reasons for this.
+First, we wanted to ensure that \typetracer captured types for unevaluated
+arguments.  Unfortunately, not every argument of a function is necessarily
+used during every function execution, meaning that some arguments may remain
+unevaluated (as they are never forced), and there are a few reasons for
+this.
 
@@
 flow of this particular execution (e.g. if the argument is only needed in
-  one branch of a conditional expression).  This mainly occurs when
-  arguments have default values for some parameters that are not critical to
-  every execution of the function.  In these cases, we still want to collect
-  the type of the default value.
+one branch of a conditional expression).  This mainly occurs when arguments
+have default values for some parameters that are not critical to every
+execution of the function.  In these cases, we still want to collect the
+type of the default value.
 
@@
 
-\item Second, we wanted to account for formal arguments (arguments which are
-  named in the function definition) which received no values when the
-  function was called. This represents cases where a function was not called
-  with all of its arguments, and no default value was specified for missing
-  arguments.  We record a ``missing'' type for such argument.
+\paragraph{Missing arguments}
+Second, we wanted to account for formal arguments (arguments which are named
+in the function definition) which received no values when the function was
+called. This represents cases where a function was not called with all of
+its arguments, and no default value was specified for missing arguments.  We
+record a ``missing'' type for such argument.
 
@@
 
-\item Third, we had to contend with R's backend sometimes
-  performing longjumps, which discards current R calls from the call stack.
-  In order to ensure that call traces are not lost when a longjump occurs,
-  \typetracer intercepts the unwinding process initiated by the longjump and
-  mimics the functionality of the functions being exited.  The R-dyntrace
-  framework supplies the return value during the unwinding process (if there
-  is indeed such a value), so we can capture it, analyze it, and include it
-  in the call trace.  In certain situations, however, no return value can be
-  obtained (e.g. a longjump occurred as part of some error handling), and
-  then we simply record a ``jumped'' return value, which we can deal with in
-  post-processing.  
-\end{itemize}
-
-Having outlined these challenges, we will now explore the implementation of \typetracer in more detail.
-
-%
-%
-\subsubsection{Select Details}
-
-As mentioned, our analysis is built on the R-dyntrace framework.  R-dyntrace
-is a very efficient instrumentation framework for R, which is built on a
-modified version of the R runtime.
-
-We primarily rely on 8 R-dyntrace callbacks: \texttt{closure\_entry},
-\texttt{closure\_exit}, \texttt{builtin\_entry}, \texttt{builtin\_exit},
-\texttt{special\_entry},\texttt{special\_exit},
-\texttt{promise\_force\_entry}, and \texttt{promise\_force\_exit}.  These
-callbacks fire when closures, builtin, and special functions are entered and
-exited, and before and after the forcing of promises.  The function-related
-callbacks are used mainly for bookkeeping: the analysis is notified that a
-closure, builtin, or special has been entered/exited by pushing/popping the
-call onto a stack.  The calls themselves store a {\it call trace} object,
-where we store the type information that we collect, populated with initial
-guesses of argument types (as discussed in
-Section~\ref{subsec:typetracer-challenges}).  Also, recall that R can
-perform single or multiple dispatch on function arguments depending on their
-class.  We collect such dispatch information the {\tt \_entry} variants.
+\paragraph{Non-local returns}
+Third, we had to contend with R's backend sometimes performing longjumps,
+which discards current R calls from the call stack.  In order to ensure that
+call traces are not lost when a longjump occurs, \typetracer intercepts the
+unwinding process initiated by the longjump and mimics the functionality of
+the functions being exited.  The R-dyntrace framework supplies the return
+value during the unwinding process (if there is indeed such a value), so we
+can capture it, analyze it, and include it in the call trace.  In certain
+situations, however, no return value can be obtained (e.g. a longjump
+occurred as part of some error handling), and then we simply record a
+``jumped'' return value, which we can deal with in post-processing.
+
+\paragraph{Implementation details.}
+We primarily rely on eight callbacks: \k{closure\_entry}, \k{closure\_exit},
+\k{builtin\_entry}, \k{builtin\_exit}, \k{special\_entry},\k{special\_exit},
+\k{promise\_force\_entry}, and \k{promise\_force\_exit}.  The
+function-related callbacks are used mainly for bookkeeping: the analysis is
+notified that a construct has been entered by pushing the call onto a stack.
+The calls themselves store a trace object, it is that object that holds the
+type information. As R can perform single or multiple dispatch on function
+arguments depending on their class, the relevant information is kept by the
+{\tt \_entry} variants.
 
@@
 reflection to promise force time.  In R, promises are essentially objects
-with value and expression slots. \AT{Put this as part of R-dyntrace
-  background:} When a promise is created for a particular expression, that
-expression is merely put into the expression slot of a promise object, and a
-special {\it unbound} value is loaded into the value slot (unless some
-default value is specified, wherein it is loaded into the value slot---this
-is the case with default argument values).  When the promise is forced, the
-expression is evaluated in the forcing context, and the value is stored in
-the value slot of the promise.  So to get the type of a function argument,
-we check the value slot of the argument promise---recursively if the value
-itself is a promise.
-
-To construct a type for each argument, we make use of R's C FFI and use
-low-level machinery to collect type tags and attributes from the R runtime.
-The types that we provide to users are constructed during post-processing,
-and rely on the detailed information made available by these low-level
-reflection mechanisms.
+with value and expression slots.
+
+When a promise is created for a particular expression, that expression is
+merely put into the expression slot of a promise object, and a special {\it
+  unbound} value is loaded into the value slot (unless some default value is
+specified, wherein it is loaded into the value slot---this is the case with
+default argument values).  When the promise is forced, the expression is
+evaluated in the forcing context, and the value is stored in the value slot
+of the promise.  So to get the type of a function argument, we check the
+value slot of the argument promise---recursively if the value itself is a
+promise.
 
@@
 system complex.  Our system lacks explicit support for
-polymorphism. Instead, it supports union, any and ... types. A union type is
+polymorphism. Instead, it supports union and any types. A union type is
 inferred when values of multiple types are observed at a function parameter
-and return position, any type is inferred when an argument is missing or
-unused, and ... is inferred for vararg parameters. Furthermore, list element
-types can be inferred as any if they have more than 5 different types. Values with many different types can result in
+and return position. any type is inferred when an argument is missing or
+unused. Furthermore, list element types can be inferred as any if they have
+more than 5 different types. Values with many different types can result in
 a large number of alternatives in the type inferred for that position,
@@
 the XXX functions in our corpus.
-
-\begin{figure}[!h]  \centering
-  \includegraphics[width=\linewidth]{plots/union_frequency.pdf}
-  \caption{Types inferred for each position}\label{fig:unionfreq}
-\end{figure}
 

### 2020-05-15 paper/main.tex
merge

--- 2020-05-15 paper/main.tex
@@
 character & class<`bignum`> | raw[] & 100100 & 7.37 & 43.36\\
-double & (integer | null) & 60570 & 4.46 & 47.81\\
+double & integer | null & 60570 & 4.46 & 47.81\\
 double[] & class<`timeSeries`> & 58872 & 4.33 & 52.15\\
@@
 
- \begin{tabular}{llrrr}
-\toprule
-Actual Type & Expected Type & \# Arguments & \% Total & Cumulative \%\\
-\midrule
- character[] & character & 183 & 2.05 & 2.05\\
- \textasciicircum{}logical & null & 158 & 1.77 & 3.81\\
- character & null & 117 & 1.31 & 5.12\\
- double & null & 107 & 1.20 & 6.32\\
- double[] & double & 100 & 1.12 & 7.44\\
- class<data.frame> & ${\bf class} \langle data.frame, tbl, tbl\_df \rangle$ & 63 & 0.70 & 8.14\\
- \bottomrule
-\end{tabular}
-
 % END Autogenerated
  \label{tbl:fail-breakdown-1}
- \caption{Most common failing contract checks.
+ \caption{Most common failing contract checks by type.
 Note that for this table, we collapsed multiple identical failed assertions into one, meaning that single arguments with multiple (identical) failed assertions are only counted once.}

### 2020-05-15 paper/main.tex
4.1

--- 2020-05-15 paper/main.tex
@@
 \paragraph{Promises}
-
-The goal of our analysis is to discover argument types for functions,
-builtins, and specials (R functions such as \{, \code{if}, etc.).  First,
-recall that R wraps arguments in promises (Section~\ref{sec:R}).  Naively
-intercepting function calls and recording argument types would result in
-incomplete data: While we can dig into promises without forcing them, if the
-innermost value is an unevaluated expression we are unable to probe it
-without evaluating it, which may alter program behaviour.  To circumvent
-this obstacle, we instead intercept the forcing of promises: when a promise
-is forced, we identify the surrounding call, collect the type of the value,
-and add it to the call trace associated with the current function call.
-Once function execution completes, we collect information about the return
-value, add it to the call trace, and save the completed call trace to be
-written out later.
-
-First, we wanted to ensure that \typetracer captured types for unevaluated
-arguments.  Unfortunately, not every argument of a function is necessarily
-used during every function execution, meaning that some arguments may remain
-unevaluated (as they are never forced), and there are a few reasons for
-this.
-
-One possibility is that the argument was simply unused given the control
-flow of this particular execution (e.g. if the argument is only needed in
-one branch of a conditional expression).  This mainly occurs when arguments
-have default values for some parameters that are not critical to every
-execution of the function.  In these cases, we still want to collect the
-type of the default value.
-
-Another possibility is that a value is used {\it without being forced}, as
-is the case when values are ``metaprogrammed''.  This is common practice in
-e.g. the \code{tidyverse}~\footnote{\AT{either talk about it here or in
-    background}} ecosystem, as they use the metaprogramming capabilities of
-R to implement a DSL for data analysis.
-
-To tackle possibly unevaluated arguments, we make an {\it initial guess} of
-function argument types on function entry. Specifically, we collect types
-for argument default values, and ascribe an ``unused'' type for unevaluated
-expressions.  If an argument promise is later forced, we simply update the
-recorded type for the argument.
+The fact that arguments are evaluated lazily, i.e. expression are packing
+into promises and evaluated on first access, complicates the information
+gathering.  For example, some promises remain un-evaluated, and it would be
+erroneous to force them as they may cause side-effects that will affect the
+programs execution.  To deal with unevaluated arguments, we make an initial
+guess for each argument at fufnction entry.  If the promise is later forced,
+we simply update the recorded type for the argument.
 
 \paragraph{Missing arguments}
-Second, we wanted to account for formal arguments (arguments which are named
-in the function definition) which received no values when the function was
-called. This represents cases where a function was not called with all of
-its arguments, and no default value was specified for missing arguments.  We
-record a ``missing'' type for such argument.
-
-There are two obvious ways to deal with missing arguments: type them as $\bot$
-or as $\top$.  As we are performing a dynamic analysis based on package test
-code, we conservatively type these arguments as $\top$, or \code{any}, as It is
-impossible to know dynamically what all possible function behaviours are.
-We discuss the prevalence of such any-typed arguments in
-Section~\ref{sec:evaluation}.
+Parameters which receive no values when the function is called are termed
+missing. This occurs when a function was called with too few arguments and
+no default value was specified for missing arguments.  We record a
+\k{missing} type for such argument.  There are two obvious ways to deal with
+missing arguments: type them as \any or type them as some unit type.  As we
+are performing a dynamic analysis, we conservatively type them as \any.
 
 \paragraph{Non-local returns}
-Third, we had to contend with R's backend sometimes performing longjumps,
-which discards current R calls from the call stack.  In order to ensure that
-call traces are not lost when a longjump occurs, \typetracer intercepts the
-unwinding process initiated by the longjump and mimics the functionality of
-the functions being exited.  The R-dyntrace framework supplies the return
-value during the unwinding process (if there is indeed such a value), so we
-can capture it, analyze it, and include it in the call trace.  In certain
-situations, however, no return value can be obtained (e.g. a longjump
-occurred as part of some error handling), and then we simply record a
-``jumped'' return value, which we can deal with in post-processing.
+When a function exits with a longjump there is no return value to speak
+of. In order to ensure that call traces are valid when a longjump occurs, we
+intercept the unwinding process initiated by the longjump and mimics the
+functionality of the functions being exited. When no return value is present
+for a call, we record a special \k{jumped} return type.
 
@@
 We primarily rely on eight callbacks: \k{closure\_entry}, \k{closure\_exit},
-\k{builtin\_entry}, \k{builtin\_exit}, \k{special\_entry},\k{special\_exit},
-\k{promise\_force\_entry}, and \k{promise\_force\_exit}.  The
-function-related callbacks are used mainly for bookkeeping: the analysis is
-notified that a construct has been entered by pushing the call onto a stack.
-The calls themselves store a trace object, it is that object that holds the
-type information. As R can perform single or multiple dispatch on function
-arguments depending on their class, the relevant information is kept by the
-{\tt \_entry} variants.
-
-The primary functionality of \typetracer is in {\tt promise\_force\_exit}.
-As arguments to functions in R are wrapped in promises, we delay our
-reflection to promise force time.  In R, promises are essentially objects
-with value and expression slots.
-
-When a promise is created for a particular expression, that expression is
-merely put into the expression slot of a promise object, and a special {\it
-  unbound} value is loaded into the value slot (unless some default value is
-specified, wherein it is loaded into the value slot---this is the case with
-default argument values).  When the promise is forced, the expression is
-evaluated in the forcing context, and the value is stored in the value slot
-of the promise.  So to get the type of a function argument, we check the
-value slot of the argument promise---recursively if the value itself is a
-promise.
+\k{builtin\_entry}, \k{builtin\_exit}, \k{special\_entry},
+\k{special\_exit}, \k{promise\_force\_entry}, and
+\k{prom\-ise\_force\_exit}.  The function-related callbacks are used mainly
+for bookkeeping: the analysis is notified that a construct has been entered
+by pushing the call onto a stack.  The calls themselves store a trace
+object, it is that object that holds the type information. As R can perform
+single or multiple dispatch on function arguments depending on their class,
+the relevant information is kept by the {\tt \_entry} variants.
 

### 2020-05-15 paper/main.tex
minor
4.2 wip

@@
 \k{array} classes even when they have no attributes). 
-
 Depending on the value's type tag, the tool collects further information:
@@
 
-The type information we collect is far more detailed than the types we
-actually generate (which we discuss in Section~\ref{sec:types}), and so the
-next stage of our inference process is to distill the rich type information
-into simpler, manageable types.  For each unique observed call trace, we
-translate the representation of the type of each of the function arguments
-and returns according to the design of our type system: for instance, we
-might have determined that a function returned a vector of integer with 4
-elements, with names, that was free of NA values, in which case we would
-write out \code{integer[]} as the type.
-
-Once types have been simplified in this way, for each function we move to
-consolidate its multiple call traces into a single function type.  We do
-this by combining the observed types of each argument into one type
-according to subtyping rules that are part of our design.  The precise
-subtyping rules we used will be discussed further in
-Section~\ref{sec:typesystemdesign}.  If we cannot consolidate multiple
-observed argument types into a single type, we simply union the disparate
-types together.
-
-While our analysis collects more detailed information than necessary for
-type generation, the information proved useful for building up a usage
-profile of the language.  This profile informed our language design: based
-on the rich type information we collected, we were able to determine types
-that would be useful for capturing the observed behaviour.  We discuss this
-iterative, data-driven language design process in Section~\ref{types},
-highlighting the reasoning behind our type system design.
-
-Thus far, we have discussed \typetracer and our strategies to take the data
-from the dynamic analysis and generate function types.  This work also
-presents a framework for specifying these types in R code, as well as
-generating contract checks for these types.  We dub this tool \contractr,
-discussed next.
-
-%
-%
-%
+The output of the \typetracer tool consists of a set of tuples of row types,
+each representing a function call in some program's execution. The inference
+step consolidate the different tuples corresponding to a particular function
+definition across multiple programs and distills them into a single type
+signature.  The shape of the infered function signatures is:
+
+\medskip
+\FUN{ T_{1,1} \k{~|~} T_{1,i}, \dots, T_{n,1} \k{~|~} T_{n,j}}{T_1\k{~|~} T_k}
+\medskip
+
+In other words, we take the union of the types occurring at individual
+argument positions rather than an union of function types. Furthermore we
+apply some transformation on the types to keep the size of types in check.
+
+\begin{tabular}{rclll}
+  \T \k{~|~} \T  & $\Rightarrow$ &  \T &  \\
+  \T \k{~|~} $T'$ & $\Rightarrow$ &  \T & \it{iff} $T' <: T$  \\
+   $\null \k{~|~} \VEC{\S_1} \k{~|~}\dots \VEC{\S_n}$  & $\Rightarrow$ &
+    $ \NAVEC{\S_1} \k{~|~}\dots \NAVEC{\S_n}$ &  \\  
+\end{tabular}
+
+Assuming that type sequences can be reordered freely, redundant types can be
+deleted, types that are subsumed by subtyping can be removed, and nullable
+types can be introduced to replace \null.
+
 \subsection{Checking types signatures with contracts}
@@
 
-\AT{Filip}
-
-%
-%
-%
-%
-%
-%
-\section{Related Work}
-\label{sec:relatedwork}
-
-\AT{Good to mention the related projects, e.g. how this was undertaken in
-  other languages, like Ruby.  There's a lot of literature in that space I
-  think, I've read some of it but not all.  We also have some related work
-  in the old paper, I'll grab it and put it here.}
-
-%
-%
-%
-%
-%
-%
 \section{Conclusion}
@@
 
-%
-%
-%
-%
-\subsection{Threats to Validity}
-
-%
-%
-%
-%
-\subsection{Future Work}
-
-%
-%
-\subsubsection{Object Orientation in R}
-
-%
-%
-\subsubsection{Types for Data Frames}
-
-% ======================================
-%
-% End
-%
-%
-
 \bibliography{bib/biblio,bib/jv,bib/r,bib/new,bib/gradual}
-
 \end{document}
-\appendix
-\section{Appendix}

### 2020-05-15 paper/main.tex
draft of eval 2

--- 2020-05-15 paper/main.tex
@@
 \paragraph{Promises}
-The fact that arguments are evaluated lazily, i.e. expression are packing
-into promises and evaluated on first access, complicates the information
-gathering.  For example, some promises remain un-evaluated, and it would be
-erroneous to force them as they may cause side-effects that will affect the
-programs execution.  To deal with unevaluated arguments, we make an initial
-guess for each argument at fufnction entry.  If the promise is later forced,
-we simply update the recorded type for the argument.
+
+The goal of our analysis is to discover argument types for functions,
+builtins, and specials (R functions such as \{, \code{if}, etc.).  First,
+recall that R wraps arguments in promises (Section~\ref{sec:R}).  Naively
+intercepting function calls and recording argument types would result in
+incomplete data: While we can dig into promises without forcing them, if the
+innermost value is an unevaluated expression we are unable to probe it
+without evaluating it, which may alter program behaviour.  To circumvent
+this obstacle, we instead intercept the forcing of promises: when a promise
+is forced, we identify the surrounding call, collect the type of the value,
+and add it to the call trace associated with the current function call.
+Once function execution completes, we collect information about the return
+value, add it to the call trace, and save the completed call trace to be
+written out later.
+
+First, we wanted to ensure that \typetracer captured types for unevaluated
+arguments.  Unfortunately, not every argument of a function is necessarily
+used during every function execution, meaning that some arguments may remain
+unevaluated (as they are never forced), and there are a few reasons for
+this.
+
+One possibility is that the argument was simply unused given the control
+flow of this particular execution (e.g. if the argument is only needed in
+one branch of a conditional expression).  This mainly occurs when arguments
+have default values for some parameters that are not critical to every
+execution of the function.  In these cases, we still want to collect the
+type of the default value.
+
+Another possibility is that a value is used {\it without being forced}, as
+is the case when values are ``metaprogrammed''.  This is common practice in
+e.g. the \code{tidyverse}~\footnote{\AT{either talk about it here or in
+    background}} ecosystem, as they use the metaprogramming capabilities of
+R to implement a DSL for data analysis.
+
+To tackle possibly unevaluated arguments, we make an {\it initial guess} of
+function argument types on function entry. Specifically, we collect types
+for argument default values, and ascribe an ``unused'' type for unevaluated
+expressions.  If an argument promise is later forced, we simply update the
+recorded type for the argument.
 
 \paragraph{Missing arguments}
-Parameters which receive no values when the function is called are termed
-missing. This occurs when a function was called with too few arguments and
-no default value was specified for missing arguments.  We record a
-\k{missing} type for such argument.  There are two obvious ways to deal with
-missing arguments: type them as \any or type them as some unit type.  As we
-are performing a dynamic analysis, we conservatively type them as \any.
+Second, we wanted to account for formal arguments (arguments which are named
+in the function definition) which received no values when the function was
+called. This represents cases where a function was not called with all of
+its arguments, and no default value was specified for missing arguments.  We
+record a ``missing'' type for such argument.
+
+There are two obvious ways to deal with missing arguments: type them as $\bot$
+or as $\top$.  As we are performing a dynamic analysis based on package test
+code, we conservatively type these arguments as $\top$, or \code{any}, as It is
+impossible to know dynamically what all possible function behaviours are.
+We discuss the prevalence of such any-typed arguments in
+Section~\ref{sec:evaluation}.
 
 \paragraph{Non-local returns}
-When a function exits with a longjump there is no return value to speak
-of. In order to ensure that call traces are valid when a longjump occurs, we
-intercept the unwinding process initiated by the longjump and mimics the
-functionality of the functions being exited. When no return value is present
-for a call, we record a special \k{jumped} return type.
+Third, we had to contend with R's backend sometimes performing longjumps,
+which discards current R calls from the call stack.  In order to ensure that
+call traces are not lost when a longjump occurs, \typetracer intercepts the
+unwinding process initiated by the longjump and mimics the functionality of
+the functions being exited.  The R-dyntrace framework supplies the return
+value during the unwinding process (if there is indeed such a value), so we
+can capture it, analyze it, and include it in the call trace.  In certain
+situations, however, no return value can be obtained (e.g. a longjump
+occurred as part of some error handling), and then we simply record a
+``jumped'' return value, which we can deal with in post-processing.
 
@@
 We primarily rely on eight callbacks: \k{closure\_entry}, \k{closure\_exit},
-\k{builtin\_entry}, \k{builtin\_exit}, \k{special\_entry},
-\k{special\_exit}, \k{promise\_force\_entry}, and
-\k{prom\-ise\_force\_exit}.  The function-related callbacks are used mainly
-for bookkeeping: the analysis is notified that a construct has been entered
-by pushing the call onto a stack.  The calls themselves store a trace
-object, it is that object that holds the type information. As R can perform
-single or multiple dispatch on function arguments depending on their class,
-the relevant information is kept by the {\tt \_entry} variants.
+\k{builtin\_entry}, \k{builtin\_exit}, \k{special\_entry},\k{special\_exit},
+\k{promise\_force\_entry}, and \k{promise\_force\_exit}.  The
+function-related callbacks are used mainly for bookkeeping: the analysis is
+notified that a construct has been entered by pushing the call onto a stack.
+The calls themselves store a trace object, it is that object that holds the
+type information. As R can perform single or multiple dispatch on function
+arguments depending on their class, the relevant information is kept by the
+{\tt \_entry} variants.
+
+The primary functionality of \typetracer is in {\tt promise\_force\_exit}.
+As arguments to functions in R are wrapped in promises, we delay our
+reflection to promise force time.  In R, promises are essentially objects
+with value and expression slots.
+
+When a promise is created for a particular expression, that expression is
+merely put into the expression slot of a promise object, and a special {\it
+  unbound} value is loaded into the value slot (unless some default value is
+specified, wherein it is loaded into the value slot---this is the case with
+default argument values).  When the promise is forced, the expression is
+evaluated in the forcing context, and the value is stored in the value slot
+of the promise.  So to get the type of a function argument, we check the
+value slot of the argument promise---recursively if the value itself is a
+promise.
 
@@
 \toprule
-Actual Type & Expected Type & \# Failed Assertions & \% Total & Cumulative \%\\
+Passed & Arg Type & Occurrences & \% Total & Cumulative \%\\
 \midrule
@@
 character & class<`bignum`> | raw[] & 100100 & 7.37 & 43.36\\
-double & integer | null & 60570 & 4.46 & 47.81\\
+double & integer | null & 60578 & 4.46 & 47.81\\
 double[] & class<`timeSeries`> & 58872 & 4.33 & 52.15\\
@@
 a failing contract assertion was in a loop.
-We were interested in a breakdown of type signature by parameter, and seeing how often those contracts were violated.
-Table~\ref{tbl:fail-breakdown-1} breaks down the results from Table~\ref{tbl:fail-breakdown-0} by parameter position, folding away multiple identical failed contract assertions for the same parameter position.
-The first row of this table reads: a value of type ${\bf chr}[]$ was passed to 183 different argument positions expecting a value of type ${\bf chr}$, representing 2.05\% of all arguments where a contract check failed.
-
-The main takeaway from Table~\ref{tbl:fail-breakdown-1} is that there appear to be no pathological kinds of failing assertions: programmers are not disproportionately passing values of one type to arguments expecting values of another.
-In other words, our type language appears to match with programmer intuitions: had we failed to account for a key, type-based assumption that R programmers make, we would see it at the top of the table.
+We were interested in knowing for each of the most common violations in Table~\ref{tbl:fail-breakdown-0}, how many different arguments had that type, and how many of those exhibited the contract failure in question.
+Table~\ref{tbl:fail-breakdown-1} breaks down the results from Table~\ref{tbl:fail-breakdown-0} by unique parameter position, folding away multiple identical failed contract assertions for the same parameter position.
+The first row of this table reads: a value of type \code{double[]} was passed to 15 different function parameters expecting a matrix, of which there are 1394 in total: 1.08\% of matrix-typed parameters were passed double vector values.
+
+We see that even though the double vector and matrix issue was wildly prevalent in the raw, dynamic contract evaluation numbers, the number of actual function argument types that were violated is very small.
+The story is similar with the double and integer coercion we mentioned earlier: it represents many dynamic contract failures, but very few of the integer vector-typed arguments have their contracts violated by double vector-typed values.
+Rows 5 and 6 are interesting: we see that rather often arguments expecting time series data are passed double vector or matrix values.
+This is a quirk of the \code{timeSeries} package, whose functions often accept matrices and vectors, converting them to time series in an ad hoc manner.
+Note that the code coverage of the \code{timeSeries} tests, examples, and vignettes on the \code{timeSeries} package code is only 58.24\%, which is one possible explanation of why these contract failures are occurring: the types that \typetracer generates are only as good as the test code its run on.
 
@@
 
+\begin{tabular}{llrrr}
+\toprule
+Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
+\midrule
+double[] & class<`matrix`> & 15 & 1394 & 1.08\\
+double[] & integer[] & 44 & 657 & 6.70\\
+character & class<`bignum`> | raw[] & 1 & 3 & 33.33\\
+double & integer | null & 5 & 20 & 25.00\\
+double[] & class<`timeSeries`> & 19 & 35 & 54.29\\
+\addlinespace
+class<matrix> & class<`timeSeries`> & 19 & 35 & 54.29\\
+double[] & double & 100 & 5021 & 1.99\\
+double & class<`data.frame`> & 1 & 900 & 0.11\\
+double[] & class<`data.frame`> & 6 & 900 & 0.67\\
+double & integer & 48 & 529 & 9.07\\
+\bottomrule
+\end{tabular}
+
 % END Autogenerated
  \label{tbl:fail-breakdown-1}
- \caption{Most common failing contract checks by type.
-Note that for this table, we collapsed multiple identical failed assertions into one, meaning that single arguments with multiple (identical) failed assertions are only counted once.}
+ \caption{Results from Table~\ref{tbl:fail-breakdown-0}, broken down by occurrences of the expected type as a parameter type.}
  \end{table}
 
-%
-%
-% \subsubsection{Patterns in Failed Contract Assertions}
-
-% While Table~\ref{tbl:fail-breakdown-1} revealed no glaring patterns in failed assertions, there are a few subtle issues that we will discuss.
-% First, \PERCFAILEDVECTOSCA of failed checks broken down by parameter position failed due to vectors being passed where scalars were expected.
-
-% We observed an interesting corner case and quirk of R in looking at functions with values that default to NA.
-% In R, recall that there is a different type of NA for each primitive, and when the literal value \code{NA} appears in code, R assumes it to be the logical NA.
-% For the purposes of our analysis, this has the effect of introducing $\string^{\bf lgl}$ types where they (probably) don't belong, particularly when programmers fail to effectively test their code.
-% For example, a function header like \code{function(x, y=NA)} will have $\string^{\bf lgl}$ as part of the type of parameter \code{y}.
-% We tried to control for this in our analysis, and indeed $\string^{\bf lgl}$ is a subtype of all NA-able numeric types.
+Table~\ref{tbl:fail-breakdown-2} presents data on the most frequently violated contracts amongst the most frequently occurring argument types.
+We selected argument types which were in the 90th percentile of argument type occurrences, computed the most frequent type signature violations among them, and reported the most frequently violated contracts together with the type of the value that violated that contract.
+The first row of the table reads: 44 function arguments with integer vector type are passed double vectors instead, and 657 arguments have integer vector type.
+Had we failed to capture some key usage pattern of R with our type annotation framework, we would likely see it here, and we can see this in action if we consider Table~\ref{tbl:fail-breakdown-3}, which was obtained identically to Table~\ref{tbl:fail-breakdown-2} with a percentile of 80th percentile instead.
+The most frequent argument type violation pattern in that of data frames being passed to arguments expecting tibbles: this occurs in only 22\% of such arguments, and represents cases where tests did not adequately cover all valid function inputs.
+Separate from the issue of testing, we can capture this behaviour with user-defined subtyping, as data frames are subtypes of tibbles, but this necessitates a proper object-oriented type system for R which is outside the scope of this work.
+
+ \begin{table}
+% BEGIN Autogenerated
+
+\begin{tabular}{llrrr}
+\toprule
+Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
+\midrule
+double[] & integer[] & 44 & 657 & 6.70\\
+\textasciicircum{}logical[] & null & 158 & 2756 & 5.73\\
+character[] & character & 183 & 4102 & 4.46\\
+character & null & 117 & 2756 & 4.25\\
+double & null & 107 & 2756 & 3.88\\
+\bottomrule
+\end{tabular}
+% END Autogenerated
+ \label{tbl:fail-breakdown-2}
+ \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 90th percentile.}
+ \end{table}
+
+  \begin{table}
+% BEGIN Autogenerated
+
+\begin{tabular}{llrrr}
+\toprule
+Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
+\midrule
+class<data.frame> & class<`data.frame`, `tbl`, `tbl\_df`> & 63 & 284 & 22.18\\
+class<matrix> & class<`array`> & 13 & 70 & 18.57\\
+\textasciicircum{}logical[] & logical | null & 9 & 49 & 18.37\\
+character[] & list<character[]> & 20 & 109 & 18.35\\
+complex[] & complex & 7 & 42 & 16.67\\
+\bottomrule
+\end{tabular}
+% END Autogenerated
+ \label{tbl:fail-breakdown-3}
+ \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 80th percentile.}
+ \end{table}
+
+In sum, we believe that this evaluation shows that the type signatures we generate from traces are quite good.
+Only \PERCFAILEDASSERTIONSWUNDEF of contract assertions failed at runtime, representing failures in as few as \PROPARGSFAILINGASSERTS of argument types.
+Even though \PROPFUNSFAILEDCHECKNOSTHREE of functions had at least one argument type involved in a failing contract check, these functions are under-tested, representing only \PERCCALLSBADFUNSINTESTS of calls observed while inferring types.
 

### 2020-05-15 paper/main.tex
Add table for uni-typed values

@@
 system complex.  Our system lacks explicit support for
-polymorphism. Instead, it supports union, any and ... types. A union type is
+polymorphism. Instead, it supports union, \any and \VARGS types. A union type is
 inferred when values of multiple types are observed at a function parameter
-and return position, any type is inferred when an argument is missing or
-unused, and ... is inferred for vararg parameters. Furthermore, list element
-types can be inferred as any if they have more than 5 different types. Values with many different types can result in
+and return position, \any type is inferred when an argument is missing or
+unused, and \VARGS is inferred for variable length argument list positions. Furthermore, list element
+types can be inferred as \any if they have more than 5 different types. Values with many different types can result in
 a large number of alternatives in the type inferred for that position,
@@
 
-We observe that XXX% of the positions have a single type ascribed to them. XXX%
-have five or more types ascribed to them. Parameter PARAMNAME of function
-XXXXCOLONCOLONXXX has the maximum, XXX, types ascribed to it. This implies that
-most functions admit or return a uni-typed value.
+We observe that XXX of the positions have a single type ascribed to them, and,
+only XXX have four or more types ascribed to them. Parameter PARAMNAME of function
+XXXXCOLONCOLONXXX has the maximum, XXX, types ascribed to it.
+Hence, most functions admit or return a uni-typed value.
+
+Figure 1 provides a breakdown of the different type categories observed at
+uni-typed positions.
+
+\input{tables/uni-typed-breakdown.tex}
 

### 2020-05-15 paper/main.tex
small tweaks

--- 2020-05-15 paper/main.tex
@@
 
+\noindent
 In other words, we take the union of the types occurring at individual
@@
 
+\medskip
 \begin{tabular}{rclll}
@@
   \T \k{~|~} $T'$ & $\Rightarrow$ &  \T & \it{iff} $T' <: T$  \\
+  \LIST{\T} \k{~|~} $\LIST{T'}$ & $\Rightarrow$ &  \LIST\T & \it{iff} $T' <: T$  \\
    $\null \k{~|~} \VEC{\S_1} \k{~|~}\dots \VEC{\S_n}$  & $\Rightarrow$ &
@@
 \end{tabular}
-
-Assuming that type sequences can be reordered freely, redundant types can be
-deleted, types that are subsumed by subtyping can be removed, and nullable
-types can be introduced to replace \null.
+\medskip
+
+\noindent
+Assuming that type sequences can be reordered freely, we rewrite types to
+minimize their size by removing redundant types, types that are subsumed by
+subtyping, and \null types.
 
@@
 
-\contractr is an R package that implements our type system for the R
-package ecosystem. For functions with type annotations, it adds contracts for type-checking
-arguments and results on function calls. On contract failure, it reports a type
-mismatch warning with the package name, function name, argument name, argument
-position, expected type signature, actual type signature of the value observed,
-and a stack trace for the called function. \contractr works by modifying
-function definition to insert a call to the type-checking function. When the
-modified function is called, the type-checking function is invoked on the
-modified function's arguments and return value.
+\contractr is an R package that implements our type system for the R package
+ecosystem. For functions with type annotations, it adds contracts for
+type-checking arguments and results on function calls. On contract failure,
+it reports a type mismatch warning with the package name, function name,
+argument name, argument position, expected type signature, actual type
+signature of the value observed, and a stack trace for the called
+function. \contractr works by modifying function definition to insert a call
+to the type-checking function. When the modified function is called, the
+type-checking function is invoked on the modified function's arguments and
+return value.
 
@@
 \subsubsection{Challenges}
-Retrofitting a robust and efficient type-system as an external package in R has
-been a significant undertaking. We had to contend with the complex design
-choices of R to make it work without surprises for the end-users. We describe
-two key challenges that we faced in the development of \contractr.
-\begin{itemize}
-\item Firstly, we faced the problem of type-checking arguments in a
-  non-strict language while retaining the non-strict semantics. Recall that when a
+Retrofitting a robust and efficient type-system as an external package in R
+has been a significant undertaking. We had to contend with the complex
+design choices of R to make it work without surprises for the end-users. We
+describe two key challenges that we faced in the development of \contractr.
+
+Firstly, we faced the problem of type-checking arguments in a non-strict
+language while retaining the non-strict semantics. Recall that when a
 function is called in R, parameters are bound to unevaluated code thunks
-  called promises~\cite{oopsla19}, instead of values which can be
-  immediately type-checked. Naively forcing promises at function call to
-  obtain a value for type-checking would violate the language semantics,
-  leading to incorrect results. To preserve the original semantics,
-  \contractr modifies the promise by wrapping the unevaluated promise
-  expression in a call to its type-checker along with the context necessary
-  to associate the promise to the function and its parameter (for failure
-  messages). The contract checking happens when the promise is evaluated,
-  either inside the function or many calls deep.  This works, but with one
-  wrinkle. GNU R has a bytecode compiler that can sometimes optimize away
-  promises; in those cases, \contractr receives values that can be
-  immediately type-checked. Non-strict semantics dictate that any
+called promises~\cite{oopsla19}, instead of values which can be immediately
+type-checked. Naively forcing promises at function call to obtain a value
+for type-checking would violate the language semantics, leading to incorrect
+results. To preserve the original semantics, \contractr modifies the promise
+by wrapping the unevaluated promise expression in a call to its type-checker
+along with the context necessary to associate the promise to the function
+and its parameter (for failure messages). The contract checking happens when
+the promise is evaluated, either inside the function or many calls deep.
+This works, but with one wrinkle. GNU R has a bytecode compiler that can
+sometimes optimize away promises; in those cases, \contractr receives values
+that can be immediately type-checked. Non-strict semantics dictate that any
 computation related to the argument should happen on the first use of the
-  argument. So type-checking at this point and issuing an error message
-  would violate the non-strict semantics; if this argument is never used, it
-  would be incorrect to type-check it at the beginning of call. To make this
-  work, \contractr mutates the function parameter binding to a promise that
-  wraps the argument value in a call to the type-checker, as in the previous
-  case.
-\item Secondly, we faced the problem of type-checking function return
-  value. In R, the result of a function call is the result of evaluating the
-  last expression in the function body; so an explicit return call is absent
-  from most function definitions. Thus, there could be many potential
+argument. So type-checking at this point and issuing an error message would
+violate the non-strict semantics; if this argument is never used, it would
+be incorrect to type-check it at the beginning of call. To make this work,
+\contractr mutates the function parameter binding to a promise that wraps
+the argument value in a call to the type-checker, as in the previous case.
+
+Secondly, we faced the problem of type-checking function return value. In R,
+the result of a function call is the result of evaluating the last
+expression in the function body; so an explicit return call is absent from
+most function definitions. Thus, there could be many potential
 sub-expressions in the function body which need to be type-checked against
@@
 calls on the stack to be discarded. When they are discarded, their exit
-  hooks are called, which in this case calls the registered
-  type-checker. But, these functions are in the middle of an active
-  computation, so they don't have a return value to type-check. \contractr
-  deals with this problem by allocating a unique sentinel object which
-  serves as the return value for calls that are discarded. The function call
-  exit hook does not call the type-checker if this unique value happens to
-  be the return value of the call.
-\end{itemize}
-%
-%
+hooks are called, which in this case calls the registered type-checker. But,
+these functions are in the middle of an active computation, so they don't
+have a return value to type-check. \contractr deals with this problem by
+allocating a unique sentinel object which serves as the return value for
+calls that are discarded. The function call exit hook does not call the
+type-checker if this unique value happens to be the return value of the
+call.
+
 \subsubsection{Usability}
@@
 features.
-\begin{itemize}
-
-\item \contractr enables automatic type-checking on importing without any
-  user intervention. Just running \code{library(contractr)} R code is enough
-  to insert contracts, both, in packages pre-loaded in the user's workspace,
-  and, in packages that will be loaded eventually. To achieve this
-  automation, \contractr relies heavily on R's reflective and dynamic
-  capabilities. On loading, \contractr scans all the pre-loaded packages in
-  the user's workspace and inserts contracts in functions for which type
-  signatures are available. For all other packages installed on user's
-  machine but not yet loaded in the workspace, \contractr sets up package
-  load hooks which are executed by R when those packages are loaded. The
-  hook registered by \contractr insert contracts to those package's
-  functions. Thus, developers and users do not have to call the \contractr
-  API to enable type-checking for packages at any point of their programming
-  workflow. Furthermore, \contractr automatically removes contracts from all
-  the package functions and restores them to their original state when it is
-  unloaded, an operation that is rarely performed, but useful in interactive
-  settings.
-
-\item \contractr enables package authors to supply type declarations for
-  their package without requiring package code modifications. Type
-  declarations can be written alongside a \code{@type} section inside
-  \roxygen function comment blocks. \roxygen is an R package that enables
-  authors to add documentation to R functions in the form of organized
-  plaintext sections which are automatically exported to R's latex style
-  custom documentation format.  \roxygen is a widely adopted package, used
-  by over XXX R packages in CRAN. It provides an extension API for adding
-  custom documentation tag sections in comment blocks. The sections are
-  processed by \roxygen before package build step and a registered method
-  for each custom tag section is invoked with the section data and
-  documented object. \contractr uses this mechanism to register a hook for a
-  custom \code{@type} tag. This hook parses the type declarations from the
-  \code{@type} sections, extracts the function name from the documented
-  function definition, and stores them in a TYPEDECLARATION file inside the
-  package folder. This files is copied verbatim on package installation and
-  picked up \contractr when the package is loaded. Here is an example of how
-  a roxygen code block looks like with our type declaration:
-
-  This feature enables a function and its type signature to coexist next to each
-  other, where they are more likely to remain synchronized. Furthermore, the
-  \contractr hook can also add the function's type declaration to its
-  documentation, which seamlessly integrate our type system with the existing R
-  tooling.
-
-\item \contractr provides a very expressive API to the users, covering a
-  variety of uses cases. During interactive development, developers can
-  explicitly insert contracts by supplying the type declaration as a string
-  argument to the \code{insert\_contract} API function. Conversely,
-  contracts can be removed by calling the \code{remove\_contract} API on
-  functions. Contracts can be selectively enabled or disabled for code
-  blocks by wrapping them in calls to \code{ignore\_contracts} and
-  \code{capture\_contracts} functions. This enables selective type-checking
-  of code section during the development phase.  Furthermore, these
-  functions return a data frame that contains all the information about
-  failed and successful contract assertions in the wrapped code blocks. This
-  is useful for post-hoc investigation. Multiple type-checking failures can
-  introduce a lot of noise in the program output. To alleviate this problem,
-  \code{contractr.severity} option can be set to \code{'silence'}.  This
-  suppresses failure messages but \contractr still performs type-checking
-  whose results can be explicitly obtained as a data frame from the
-  \code{get\_contracts} API function. R suppresses printing of warnings on
-  the terminal when a function issues too many warnings. This can hide
-  type-checking failure messages. Setting \code{contractr.severity} flag to
-  \code{'error'} turns the type-checking failure warnings to errors, which
-  halts the program at the point of contract failure.
-
-\end{itemize}
+
+\contractr enables automatic type-checking on importing without any user
+intervention. Just running \code{library(contractr)} R code is enough to
+insert contracts, both, in packages pre-loaded in the user's workspace, and,
+in packages that will be loaded eventually. To achieve this automation,
+\contractr relies heavily on R's reflective and dynamic capabilities. On
+loading, \contractr scans all the pre-loaded packages in the user's
+workspace and inserts contracts in functions for which type signatures are
+available. For all other packages installed on user's machine but not yet
+loaded in the workspace, \contractr sets up package load hooks which are
+executed by R when those packages are loaded. The hook registered by
+\contractr insert contracts to those package's functions. Thus, developers
+and users do not have to call the \contractr API to enable type-checking for
+packages at any point of their programming workflow. Furthermore, \contractr
+automatically removes contracts from all the package functions and restores
+them to their original state when it is unloaded, an operation that is
+rarely performed, but useful in interactive settings.
+
+\contractr enables package authors to supply type declarations for their
+package without requiring package code modifications. Type declarations can
+be written alongside a \code{@type} section inside \roxygen function comment
+blocks. \roxygen is an R package that enables authors to add documentation
+to R functions in the form of organized plaintext sections which are
+automatically exported to R's latex style custom documentation format.
+\roxygen is a widely adopted package, used by over XXX R packages in
+CRAN. It provides an extension API for adding custom documentation tag
+sections in comment blocks. The sections are processed by \roxygen before
+package build step and a registered method for each custom tag section is
+invoked with the section data and documented object. \contractr uses this
+mechanism to register a hook for a custom \code{@type} tag. This hook parses
+the type declarations from the \code{@type} sections, extracts the function
+name from the documented function definition, and stores them in a
+TYPEDECLARATION file inside the package folder. This files is copied
+verbatim on package installation and picked up \contractr when the package
+is loaded. Here is an example of how a roxygen code block looks like with
+our type declaration:
+
+This feature enables a function and its type signature to coexist next to
+each other, where they are more likely to remain synchronized. Furthermore,
+the \contractr hook can also add the function's type declaration to its
+documentation, which seamlessly integrate our type system with the existing
+R tooling.
+
+\contractr provides a very expressive API to the users, covering a variety
+of uses cases. During interactive development, developers can explicitly
+insert contracts by supplying the type declaration as a string argument to
+the \code{insert\_contract} API function. Conversely, contracts can be
+removed by calling the \code{remove\_contract} API on functions. Contracts
+can be selectively enabled or disabled for code blocks by wrapping them in
+calls to \code{ignore\_contracts} and \code{capture\_contracts}
+functions. This enables selective type-checking of code section during the
+development phase.  Furthermore, these functions return a data frame that
+contains all the information about failed and successful contract assertions
+in the wrapped code blocks. This is useful for post-hoc
+investigation. Multiple type-checking failures can introduce a lot of noise
+in the program output. To alleviate this problem, \code{contractr.severity}
+option can be set to \code{'silence'}.  This suppresses failure messages but
+\contractr still performs type-checking whose results can be explicitly
+obtained as a data frame from the \code{get\_contracts} API function. R
+suppresses printing of warnings on the terminal when a function issues too
+many warnings. This can hide type-checking failure messages. Setting
+\code{contractr.severity} flag to \code{'error'} turns the type-checking
+failure warnings to errors, which halts the program at the point of contract
+failure.
 
@@
 \toprule
-Passed & Arg Type & Occurrences & \% Total & Cumulative \%\\
+Actual Type & Expected Type & \# Failed Assertions & \% Total & Cumulative \%\\
 \midrule
@@
 character & class<`bignum`> | raw[] & 100100 & 7.37 & 43.36\\
-double & integer | null & 60578 & 4.46 & 47.81\\
+double & integer | null & 60570 & 4.46 & 47.81\\
 double[] & class<`timeSeries`> & 58872 & 4.33 & 52.15\\
@@
 a failing contract assertion was in a loop.
-We were interested in knowing for each of the most common violations in Table~\ref{tbl:fail-breakdown-0}, how many different arguments had that type, and how many of those exhibited the contract failure in question.
-Table~\ref{tbl:fail-breakdown-1} breaks down the results from Table~\ref{tbl:fail-breakdown-0} by unique parameter position, folding away multiple identical failed contract assertions for the same parameter position.
-The first row of this table reads: a value of type \code{double[]} was passed to 15 different function parameters expecting a matrix, of which there are 1394 in total: 1.08\% of matrix-typed parameters were passed double vector values.
-
-We see that even though the double vector and matrix issue was wildly prevalent in the raw, dynamic contract evaluation numbers, the number of actual function argument types that were violated is very small.
-The story is similar with the double and integer coercion we mentioned earlier: it represents many dynamic contract failures, but very few of the integer vector-typed arguments have their contracts violated by double vector-typed values.
-Rows 5 and 6 are interesting: we see that rather often arguments expecting time series data are passed double vector or matrix values.
-This is a quirk of the \code{timeSeries} package, whose functions often accept matrices and vectors, converting them to time series in an ad hoc manner.
-Note that the code coverage of the \code{timeSeries} tests, examples, and vignettes on the \code{timeSeries} package code is only 58.24\%, which is one possible explanation of why these contract failures are occurring: the types that \typetracer generates are only as good as the test code its run on.
+We were interested in a breakdown of type signature by parameter, and seeing how often those contracts were violated.
+Table~\ref{tbl:fail-breakdown-1} breaks down the results from Table~\ref{tbl:fail-breakdown-0} by parameter position, folding away multiple identical failed contract assertions for the same parameter position.
+The first row of this table reads: a value of type ${\bf chr}[]$ was passed to 183 different argument positions expecting a value of type ${\bf chr}$, representing 2.05\% of all arguments where a contract check failed.
+
+The main takeaway from Table~\ref{tbl:fail-breakdown-1} is that there appear to be no pathological kinds of failing assertions: programmers are not disproportionately passing values of one type to arguments expecting values of another.
+In other words, our type language appears to match with programmer intuitions: had we failed to account for a key, type-based assumption that R programmers make, we would see it at the top of the table.
 
@@
 
-\begin{tabular}{llrrr}
-\toprule
-Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
-\midrule
-double[] & class<`matrix`> & 15 & 1394 & 1.08\\
-double[] & integer[] & 44 & 657 & 6.70\\
-character & class<`bignum`> | raw[] & 1 & 3 & 33.33\\
-double & integer | null & 5 & 20 & 25.00\\
-double[] & class<`timeSeries`> & 19 & 35 & 54.29\\
-\addlinespace
-class<matrix> & class<`timeSeries`> & 19 & 35 & 54.29\\
-double[] & double & 100 & 5021 & 1.99\\
-double & class<`data.frame`> & 1 & 900 & 0.11\\
-double[] & class<`data.frame`> & 6 & 900 & 0.67\\
-double & integer & 48 & 529 & 9.07\\
-\bottomrule
-\end{tabular}
-
 % END Autogenerated
  \label{tbl:fail-breakdown-1}
- \caption{Results from Table~\ref{tbl:fail-breakdown-0}, broken down by occurrences of the expected type as a parameter type.}
+ \caption{Most common failing contract checks by type.
+Note that for this table, we collapsed multiple identical failed assertions into one, meaning that single arguments with multiple (identical) failed assertions are only counted once.}
  \end{table}
 
-Table~\ref{tbl:fail-breakdown-2} presents data on the most frequently violated contracts amongst the most frequently occurring argument types.
-We selected argument types which were in the 90th percentile of argument type occurrences, computed the most frequent type signature violations among them, and reported the most frequently violated contracts together with the type of the value that violated that contract.
-The first row of the table reads: 44 function arguments with integer vector type are passed double vectors instead, and 657 arguments have integer vector type.
-Had we failed to capture some key usage pattern of R with our type annotation framework, we would likely see it here, and we can see this in action if we consider Table~\ref{tbl:fail-breakdown-3}, which was obtained identically to Table~\ref{tbl:fail-breakdown-2} with a percentile of 80th percentile instead.
-The most frequent argument type violation pattern in that of data frames being passed to arguments expecting tibbles: this occurs in only 22\% of such arguments, and represents cases where tests did not adequately cover all valid function inputs.
-Separate from the issue of testing, we can capture this behaviour with user-defined subtyping, as data frames are subtypes of tibbles, but this necessitates a proper object-oriented type system for R which is outside the scope of this work.
-
- \begin{table}
-% BEGIN Autogenerated
-
-\begin{tabular}{llrrr}
-\toprule
-Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
-\midrule
-double[] & integer[] & 44 & 657 & 6.70\\
-\textasciicircum{}logical[] & null & 158 & 2756 & 5.73\\
-character[] & character & 183 & 4102 & 4.46\\
-character & null & 117 & 2756 & 4.25\\
-double & null & 107 & 2756 & 3.88\\
-\bottomrule
-\end{tabular}
-% END Autogenerated
- \label{tbl:fail-breakdown-2}
- \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 90th percentile.}
- \end{table}
-
-  \begin{table}
-% BEGIN Autogenerated
-
-\begin{tabular}{llrrr}
-\toprule
-Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
-\midrule
-class<data.frame> & class<`data.frame`, `tbl`, `tbl\_df`> & 63 & 284 & 22.18\\
-class<matrix> & class<`array`> & 13 & 70 & 18.57\\
-\textasciicircum{}logical[] & logical | null & 9 & 49 & 18.37\\
-character[] & list<character[]> & 20 & 109 & 18.35\\
-complex[] & complex & 7 & 42 & 16.67\\
-\bottomrule
-\end{tabular}
-% END Autogenerated
- \label{tbl:fail-breakdown-3}
- \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 80th percentile.}
- \end{table}
-
-In sum, we believe that this evaluation shows that the type signatures we generate from traces are quite good.
-Only \PERCFAILEDASSERTIONSWUNDEF of contract assertions failed at runtime, representing failures in as few as \PROPARGSFAILINGASSERTS of argument types.
-Even though \PROPFUNSFAILEDCHECKNOSTHREE of functions had at least one argument type involved in a failing contract check, these functions are under-tested, representing only \PERCCALLSBADFUNSINTESTS of calls observed while inferring types.
+%
+%
+% \subsubsection{Patterns in Failed Contract Assertions}
+
+% While Table~\ref{tbl:fail-breakdown-1} revealed no glaring patterns in failed assertions, there are a few subtle issues that we will discuss.
+% First, \PERCFAILEDVECTOSCA of failed checks broken down by parameter position failed due to vectors being passed where scalars were expected.
+
+% We observed an interesting corner case and quirk of R in looking at functions with values that default to NA.
+% In R, recall that there is a different type of NA for each primitive, and when the literal value \code{NA} appears in code, R assumes it to be the logical NA.
+% For the purposes of our analysis, this has the effect of introducing $\string^{\bf lgl}$ types where they (probably) don't belong, particularly when programmers fail to effectively test their code.
+% For example, a function header like \code{function(x, y=NA)} will have $\string^{\bf lgl}$ as part of the type of parameter \code{y}.
+% We tried to control for this in our analysis, and indeed $\string^{\bf lgl}$ is a subtype of all NA-able numeric types.
 

### 2020-05-15 evaluation/R/evaluate-type-analysis.Rmd

@@
   graph_dirpath: ../../paper/plots/
+  table_dirpath: ../../paper/tables/
 ---
@@
 suppressPackageStartupMessages(library(tibble))
+suppressPackageStartupMessages(library(knitr))
+suppressPackageStartupMessages(library(kableExtra))
 
@@
 
-type_analysis_data <- 
-    read_csv(params$type_analysis_datafile,
-             col_names = TRUE,
-             col_types = cols(col_character(),
-                              col_character(),
-                              col_integer(),
-                              col_integer(),
-                              col_integer(),
-                              col_character(),
-                              col_integer()))
+save_table <- function(content, filename) {
+    write(content, file = file.path(params$table_dirpath, filename))
+}
+
+simplify_types <- function(type) {
+    if(type %in% c("any", "externalptr", "expression", "pairlist")) {
+        "any"
+    }
+    else {
+        type
+    }
+}
+
+is_polymorphic <- function(kind) {
+    kind %in% c("...", "any", "list")
+}
 
@@
     }
-    else if(type %in% c("^integer", "^double", "^character", "^logical", "^complex")) {
-        "scalar-na"
-    }
     else if(type %in% c("integer[]", "double[]", "character[]", "logical[]", "complex[]", "raw[]")) {
@@
     }
-    else if(type %in% c("^integer[]", "^double[]", "^character[]", "^logical[]", "^complex[]")) {
-        "vector-na"
+    else if(type %in% c("^integer[]", "^double[]", "^character[]", "^logical[]", "^complex[]", 
+                        "^integer", "^double", "^character", "^logical", "^complex")) {
+        "^vector"
     }
@@
     }
+    else if(type == "environment") {
+        "environment"
+    }
     else {
@@
 type_analysis_data <- 
+    read_csv(params$type_analysis_datafile,
+             col_names = TRUE,
+             col_types = cols(col_character(),
+                              col_character(),
+                              col_integer(),
+                              col_integer(),
+                              col_integer(),
+                              col_character(),
+                              col_integer()))
+
+type_analysis_data <-
     type_analysis_data %>%
+    mutate(type = map_chr(type, simplify_types)) %>%
     mutate(kind = map_chr(type, infer_kind))
@@
 group_by(package_name, function_name, parameter_position, alternative_count) %>%
-summarize(type = paste0(sort(type), collapse=" | ")) %>%
+summarize(kind = paste0(sort(kind), collapse=" | ")) %>%
 ungroup() %>%
-group_by(alternative_count, type) %>%
+group_by(alternative_count, kind) %>%
 summarize(count = n()) %>%
@@
 
+unityped_positions <-
+union_types %>% 
+filter(alternative_count == 1)
+
+polymorphic_unityped_rows <-
+unityped_positions %>%
+pull(kind) %>% 
+map_lgl(is_polymorphic)
+
+polymorphic_unityped_rows <- (1:nrow(unityped_positions))[polymorphic_unityped_rows]
+
+print(polymorphic_unityped_rows)
+
+unityped_positions %>%
+select("Types" = kind, "Parameter #" = count, "%" = percentage, "Cumulative %" = cumsum_percentage) %>%
+kable("latex", booktabs = T) %>%
+kable_styling(full_width = F, latex_options = c("scale_down")) %>%
+row_spec(polymorphic_unityped_rows, color = "black", background = "lightgray") %>%
+save_table("uni-typed-breakdown.tex")
+
 single_union_types <-

### 2020-05-15 paper/tables/uni-typed-breakdown.tex

\begin{table}[H]
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}{lrrr}
\toprule
Types & Parameter \# & \% & Cumulative \%\\
\midrule
scalar & 28045 & 35.80 & 35.80\\
class & 18510 & 23.63 & 59.43\\
vector & 10296 & 13.14 & 72.58\\
\rowcolor{lightgray}  \textcolor{black}{...} & \textcolor{black}{6611} & \textcolor{black}{8.44} & \textcolor{black}{81.02}\\
\rowcolor{lightgray}  \textcolor{black}{any} & \textcolor{black}{5993} & \textcolor{black}{7.65} & \textcolor{black}{88.67}\\
\addlinespace
null & 4795 & 6.12 & 94.79\\
\rowcolor{lightgray}  \textcolor{black}{list} & \textcolor{black}{3229} & \textcolor{black}{4.12} & \textcolor{black}{98.91}\\
\textasciicircum{}vector & 461 & 0.59 & 99.50\\
environment & 390 & 0.50 & 100.00\\
\bottomrule
\end{tabular}}
\end{table}

### 2020-05-15 paper/main.tex
4.3

@@
 
-\contractr is an R package that implements our type system for the R package
-ecosystem. For functions with type annotations, it adds contracts for
-type-checking arguments and results on function calls. On contract failure,
-it reports a type mismatch warning with the package name, function name,
-argument name, argument position, expected type signature, actual type
-signature of the value observed, and a stack trace for the called
-function. \contractr works by modifying function definition to insert a call
-to the type-checking function. When the modified function is called, the
-type-checking function is invoked on the modified function's arguments and
-return value.
-
-\contractr is implemented in 434 R LOC and 2358 C++ LOC. The core
-implementation is in C++ to keep the runtime overhead low. It been designed and
-tested with GNU R-3.5.0 but it also works with recent versions of R. It has been
-hardened with a battery of 400 test cases. We have used it extensively during
-the course of this work; firstly, for sanity checking of XXX type signatures
-generated for the XXX packages during the development phase, and secondly, for
-assessing the quality of these type signatures on XXX packages during the
-evaluation phase. We discuss this further in Section~\ref{sec:evaluation}.
-
-%
-%
-\subsubsection{Type Declarations}
-Type declarations can be made available to \contractr in three ways: as
-part of its internal database of types (which already has type signatures for
-our corpus of 500 packages), as part of a designated \emph{TYPEDECLARATION} file
-supplied by the package author and installed along with the package, or through
-a user-level API function insert\_contract that lets the user insert contract to
-a custom function with the type declaration provided as a string argument.
-
-%
-%
-\subsubsection{Challenges}
-Retrofitting a robust and efficient type-system as an external package in R
-has been a significant undertaking. We had to contend with the complex
-design choices of R to make it work without surprises for the end-users. We
-describe two key challenges that we faced in the development of \contractr.
+One way to validate the inferred type annotations is to check that different
+programs using one of the functions for which a type signature has been
+inferred respect the signature.  \contractr is an R package that allows to
+decorate function with assertions. We use it to insert type checks of the
+arguments and return values. As usual with R, this is not straightforward.
+
+While \contractr's primary logic has been implemented in C++ to reduce its
+runtime overhead, we have not observed a single segfault during our use of
+the package.  It has been tested with GNU R-3.5.0 and hardened with a
+battery of 400 test cases. It works by modifying function definition to
+insert a call to a type-checking function.
+
+In terms of useability, our checker is enabled automatically whenever a
+package is loaded in R. So, a simple invocation of \code{library(contractr)}
+causes contracts to be injected in all packages.  On loading, \contractr
+scans all packages in the user's workspace and inserts contracts in
+functions for which type signatures are available. Then \contractr sets up
+package load hooks which are executed when new packages are loaded.
+Furthermore, \contractr automatically removes contracts from all functions
+and restores them to their original state when it is unloaded.
+
+The type signatures can be provided externally, in a file, thus avoiding the
+need to change the source code of checked packages.  Type declarations can
+also be written alongside a \code{@type} section inside \roxygen function
+comment blocks. \roxygen is an R package that enables authors to add
+documentation. The documentation sections are processed by \roxygen before
+package build and a registered method for each custom tag section is
+invoked with the section data and documented object. \contractr uses this
+mechanism to register a hook for a custom \code{@type} tag.
+
+The \contractr API allows users to explicitly insert contracts during
+interactive development by supplying the type signature as a string.
+Contracts can be removed by calling \code{remove\_contract}.  Contracts can
+be selectively enabled or disabled for code blocks by wrapping them in calls
+to \code{ignore\_contracts} and \code{capture\_contracts}.  These functions
+return a data frame that contains all the information about failed and
+successful contract assertions in the wrapped code blocks.  Multiple
+failures can introduce noise; this is avoided by setting
+\code{contractr.severity} to \code{'silence'}.  \contractr still performs
+type-checking so that results can be obtained by calling
+\code{get\_contracts}. Setting \code{contractr.severity} flag to
+\code{'error'} turns the type-checking warnings to errors and
+halt the program.
 
@@
 call.
-
-\subsubsection{Usability}
-With \contractr, we aim to provide a hassle-free path to the R developers
-for adding types to untyped package code. Hence, we have paid a lot of
-attention towards usability in its design. We discuss three such usability
-features.
-
-\contractr enables automatic type-checking on importing without any user
-intervention. Just running \code{library(contractr)} R code is enough to
-insert contracts, both, in packages pre-loaded in the user's workspace, and,
-in packages that will be loaded eventually. To achieve this automation,
-\contractr relies heavily on R's reflective and dynamic capabilities. On
-loading, \contractr scans all the pre-loaded packages in the user's
-workspace and inserts contracts in functions for which type signatures are
-available. For all other packages installed on user's machine but not yet
-loaded in the workspace, \contractr sets up package load hooks which are
-executed by R when those packages are loaded. The hook registered by
-\contractr insert contracts to those package's functions. Thus, developers
-and users do not have to call the \contractr API to enable type-checking for
-packages at any point of their programming workflow. Furthermore, \contractr
-automatically removes contracts from all the package functions and restores
-them to their original state when it is unloaded, an operation that is
-rarely performed, but useful in interactive settings.
-
-\contractr enables package authors to supply type declarations for their
-package without requiring package code modifications. Type declarations can
-be written alongside a \code{@type} section inside \roxygen function comment
-blocks. \roxygen is an R package that enables authors to add documentation
-to R functions in the form of organized plaintext sections which are
-automatically exported to R's latex style custom documentation format.
-\roxygen is a widely adopted package, used by over XXX R packages in
-CRAN. It provides an extension API for adding custom documentation tag
-sections in comment blocks. The sections are processed by \roxygen before
-package build step and a registered method for each custom tag section is
-invoked with the section data and documented object. \contractr uses this
-mechanism to register a hook for a custom \code{@type} tag. This hook parses
-the type declarations from the \code{@type} sections, extracts the function
-name from the documented function definition, and stores them in a
-TYPEDECLARATION file inside the package folder. This files is copied
-verbatim on package installation and picked up \contractr when the package
-is loaded. Here is an example of how a roxygen code block looks like with
-our type declaration:
-
-This feature enables a function and its type signature to coexist next to
-each other, where they are more likely to remain synchronized. Furthermore,
-the \contractr hook can also add the function's type declaration to its
-documentation, which seamlessly integrate our type system with the existing
-R tooling.
-
-\contractr provides a very expressive API to the users, covering a variety
-of uses cases. During interactive development, developers can explicitly
-insert contracts by supplying the type declaration as a string argument to
-the \code{insert\_contract} API function. Conversely, contracts can be
-removed by calling the \code{remove\_contract} API on functions. Contracts
-can be selectively enabled or disabled for code blocks by wrapping them in
-calls to \code{ignore\_contracts} and \code{capture\_contracts}
-functions. This enables selective type-checking of code section during the
-development phase.  Furthermore, these functions return a data frame that
-contains all the information about failed and successful contract assertions
-in the wrapped code blocks. This is useful for post-hoc
-investigation. Multiple type-checking failures can introduce a lot of noise
-in the program output. To alleviate this problem, \code{contractr.severity}
-option can be set to \code{'silence'}.  This suppresses failure messages but
-\contractr still performs type-checking whose results can be explicitly
-obtained as a data frame from the \code{get\_contracts} API function. R
-suppresses printing of warnings on the terminal when a function issues too
-many warnings. This can hide type-checking failure messages. Setting
-\code{contractr.severity} flag to \code{'error'} turns the type-checking
-failure warnings to errors, which halts the program at the point of contract
-failure.
-
-While \contractr's primary logic has been implemented in C++, we have not
-observed a single segfault during our use of the package, either during the
-development and sanity-checking phase, or during the evaluation phase.
-We conclude the discussion of \contractr by drawing reader's attention to the
-engineering effort it takes to package experimental ideas in the form of stable
-and easy-to-use tools. R has an eclectic mix of features and design oddities
-that make this even harder.
 

### 2020-05-15 paper/main.tex
edits

--- 2020-05-15 paper/main.tex
@@
 \newcommand{\PERCCALLSBADFUNSINTESTS}{5.38\%\xspace}
+\newcommand{\TOTNUMSIGS}{21,730\xspace}
 
@@
 
-We implemented \typetracer an automated tool for extracting raw type
+We implemented \typetracer, an automated tool for extracting raw type
 signatures from execution traces of R programs. The goal of this tool is to
@@
 \paragraph{Promises}
-The fact that arguments are evaluated lazily, i.e. expression are packing
-into promises and evaluated on first access, complicates the information
+The fact that arguments are evaluated lazily (in that expression are packed
+into promises and evaluated on first access) complicates the information
 gathering.  For example, some promises remain un-evaluated, and it would be
@@
 programs execution.  To deal with unevaluated arguments, we make an initial
-guess for each argument at fufnction entry.  If the promise is later forced,
+guess for each argument at function entry.  If the promise is later forced,
 we simply update the recorded type for the argument.
@@
 of. In order to ensure that call traces are valid when a longjump occurs, we
-intercept the unwinding process initiated by the longjump and mimics the
+intercept the unwinding process initiated by the longjump and mimic the
 functionality of the functions being exited. When no return value is present
@@
 
-The output of the \typetracer tool consists of a set of tuples of row types,
+The output of the \typetracer tool consists of a set of tuples of raw types,
 each representing a function call in some program's execution. The inference
-step consolidate the different tuples corresponding to a particular function
+step consolidates the different tuples corresponding to a particular function
 definition across multiple programs and distills them into a single type
@@
 In other words, we take the union of the types occurring at individual
-argument positions rather than an union of function types. Furthermore we
+argument positions rather than an union of function types. Furthermore, we
 apply some transformation on the types to keep the size of types in check.
@@
 hardened with a battery of 400 test cases. We have used it extensively during
-the course of this work; firstly, for sanity checking of XXX type signatures
+the course of this work; firstly, for sanity checking of \TOTNUMSIGS type signatures
 generated for the XXX packages during the development phase, and secondly, for
@@
   \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the
-expression code coverage in percents and the y-axis gives the number of
+expression code coverage in percent and the y-axis gives the number of
 reverse dependencies in log scale. Dotted lines indicate means.  Packages

### 2020-05-15 evaluation/R/package-asserts.Rmd
Update corpus analysis - revdeps, roxygen, renaming
Add asserts analysis and data

--- 2020-05-08 evaluation/R/runtime-checks.Rmd
@@
 ---
-title: "Runtime checks"
+title: "Package asserts"
 output: html_document
@@
 knitr::opts_chunk$set(echo = TRUE)
-```
 
-```{r paths, include=FALSE}
 source("inc/paths.R")
+source("inc/setup.R")
+source("inc/latextags.R")
+
+options(repos=CRAN_LOCAL_MIRROR)
+
+create_tags(path(TAGS_DIR, "package-asserts.tex"), prefix="", default=TRUE)
 ```
@@
 
-The `r RUNTIME_CHECKS` file contains information about the use of assertions across packages.
+The `r CRAN_ASSERTS` file contains information about the use of assertions across packages.
 To generate it, run the following:
@@
 
-```{r load run}
-run <- read_csv(RUNTIME_CHECKS_RUN)
-run <- semi_join(run, CORPUS_PACKAGES_DF, by=c("job"="package"))
+```{r load corpus}
+corpus <- read_csv(CORPUS_DETAILS)
 ```
 
-```{r check run packages}
-#stopifnot(nrow(run) == nrow(CORPUS_PACKAGES_DF))
+```{r load asserts}
+cran_asserts <- read_csv(CRAN_ASSERTS)
+corpus_asserts <- semi_join(cran_asserts, corpus, by="package")
 ```
 
-```{r load runtime checks}
-runtime_checks <- read_csv(RUNTIME_CHECKS)
-runtime_checks <- semi_join(runtime_checks, CORPUS_PACKAGES_DF, by="package")
-```
-
-## Analysis
-
-### By package
+## Typeable and partially typeable
 
 ```{r}
-per_package <- count(runtime_checks, package)
+TYPE_FUNS <- c(
+  "is.array",
+  "is.atomic",
+  "is.call",
+  "is.character",
+  "is.complex",
+  "is.data.frame",
+  "is.data.table",
+  "is.double",
+  "is.environment",
+  "is.factor",
+  "is.function",
+  "is.integer",
+  "is.language",
+  "is.list",
+  "is.logical",
+  "is.matrix",
+  "is.name",
+  "is.numeric",
+  "is.primitive",
+  "is.symbol",
+  "is.raw",
+  "is.vector",
+
+  "is.na",
+  "is.null",
+
+  "inherits",
+  "is",
+  "extends",
+
+  # from desc package
+  "is_bool",
+  "is_call",
+  "is_character",
+  "is_formula",
+  "is_list",
+  "is_logical",
+  "is_null",
+  "is_string",
+  "is_scalar_character",
+  "is_scalar_logical",
+  "has_no_na",
+  "is_string_or_null"
+)
+
+AUX_FUNS <- c(
+  "stopifnot",
+  "|",
+  "||",
+  "&",
+  "&&",
+  "!",
+  "("
+)
+
+remove_type_asserts <- function(x, simplify=FALSE) {
+  ret <- if (!is.call(x)) {
+    x
+  } else {
+    fun <- as.character(x[[1]])
+    if (fun == "==" || fun == "!=") {
+      if (is.call(x[[2]]) && x[[2]][[1]] == "length" && (x[[3]] == 1 || x[[3]] == 1L)) {
+        NULL
+      } else {
+        x
+      }
+    } else if (any(fun %in% AUX_FUNS)) {
+      args <- map(x[-1], remove_type_asserts, simplify=simplify)
+      if (is.null(unlist(args))) {
+        NULL
+      } else if (any(sapply(args, is.null)) && simplify) {
+          NULL
+      } else {
+          x
+      }
+    } else if (any(fun %in% TYPE_FUNS)) {
+      NULL
+    } else {
+      x
+    }
+  }
+
+  if (is.list(ret)) {
+    ret[[1]]
+  } else {
+    ret
+  }
+}
+
+remove_type_asserts(quote(stopifnot(is.integer(x) && is.character2(x))))
+remove_type_asserts(quote(stopifnot(is.integer(x) && is.character2(x))), simplify = T)
+
+remove_type_asserts(quote(stopifnot(is.integer(x) | is.character(x))))
+remove_type_asserts(quote(stopifnot(length(x) == 1)))
+remove_type_asserts(quote(stopifnot(length(x) == 1L)))
 ```
@@
 ```{r}
-per_package %>%
-  arrange(desc(n)) %>%
-  datatable()
+remove_type_asserts_from_args <- function(args, simplify=FALSE) {
+  args %>%
+  map(~str_c("stopifnot(", ., ")")) %>%
+  map(~tryCatch(parse(text=.)[[1]], error=function(e) { message("Unable to parse: ",., " : ", e$message); NA })) %>%
+  map(~remove_type_asserts(., simplify=simplify)) %>%
+  map(~is.null(.))
+}
+
+remove_type_asserts_from_df <- function(df) {
+  df %>%
+  mutate(
+    typed=unlist(remove_type_asserts_from_args(args)),
+    partially_typed=unlist(remove_type_asserts_from_args(args, TRUE))
+  )
+}
+
+# df %>%
+#   filter(!partially_typed) %>%
+#   select(package, fun_name, args) %>%
+#  View()
 ```
@@
 ```{r}
-ggplot(per_package, aes(n)) + geom_histogram(binwidth = 1)
+summary_asserts <- function(df, prefix) {
+  df <- remove_type_asserts_from_df(df)
+  typed <- filter(df, typed)
+  partially_typed <- filter(df, partially_typed)
+
+  overview_table(
+    r(paste(prefix, "asserts"), df),
+    r(paste(prefix, "asserts in packages"), count(df, package)),
+    r(paste(prefix, "asserts in functions"), count(df, fun_name)),
+    r(paste(prefix, "stopifnot ratio"), ratio(filter(df, assert=="stopifnot"), df)),
+
+    r(paste(prefix, "typed asserts"), typed),
+    r(paste(prefix, "typed asserts ratio"), ratio(typed, df)),
+    r(paste(prefix, "typed asserts packages"), count(typed, package)),
+    r(paste(prefix, "typed asserts packages ratio"), ratio(count(typed, package), count(df, package))),
+    r(paste(prefix, "typed asserts functions"), count(typed, fun_name)),
+    r(paste(prefix, "typed asserts functions ration"), ratio(count(typed, fun_name), count(df, fun_name))),
+
+    r(paste(prefix, "partially typed asserts"), partially_typed),
+    r(paste(prefix, "partially typed asserts ratio"), ratio(partially_typed, df)),
+    r(paste(prefix, "partially typed asserts packages"), count(partially_typed, package)),
+    r(paste(prefix, "partially typed asserts packages ratio"), ratio(count(partially_typed, package), count(df, package))),
+    r(paste(prefix, "partially typed asserts functions"), count(partially_typed, fun_name)),
+    r(paste(prefix, "partially typed asserts functions ration"), ratio(count(partially_typed, fun_name), count(df, fun_name))),
+  )
+}
 ```
 
-### By package function
-
-```{r}
-per_package_fun <- count(runtime_checks, package, fun_name)
+```{r summary for cran}
+summary_asserts(cran_asserts, "cran")
 ```
 
-```{r}
-per_package_fun %>%
-  arrange(desc(n)) %>%
-  datatable()
+```{r summary for corpus}
+summary_asserts(corpus_asserts, "corpus")
 ```
-
-```{r}
-ggplot(per_package_fun, aes(n)) + geom_histogram(binwidth = 1)
-```
-
-### By asertion function
-
-```{r}
-per_assert <- count(runtime_checks, assert)
-```
-
-```{r}
-per_assert %>%
-  arrange(desc(n)) %>%
-  datatable()
-```
-
-## Raw data
-
-```{r}
-datatable(
-  runtime_checks, 
-  filter="top", 
-  options=list(
-    pageLength=50,
-    autoWidth=TRUE
-  )
-)
-```

### 2020-05-15 paper/main.tex
4.3

--- 2020-05-15 paper/main.tex
@@
 
-Firstly, we faced the problem of type-checking arguments in a non-strict
-language while retaining the non-strict semantics. Recall that when a
-function is called in R, parameters are bound to unevaluated code thunks
-called promises~\cite{oopsla19}, instead of values which can be immediately
-type-checked. Naively forcing promises at function call to obtain a value
-for type-checking would violate the language semantics, leading to incorrect
-results. To preserve the original semantics, \contractr modifies the promise
-by wrapping the unevaluated promise expression in a call to its type-checker
-along with the context necessary to associate the promise to the function
-and its parameter (for failure messages). The contract checking happens when
-the promise is evaluated, either inside the function or many calls deep.
-This works, but with one wrinkle. GNU R has a bytecode compiler that can
-sometimes optimize away promises; in those cases, \contractr receives values
-that can be immediately type-checked. Non-strict semantics dictate that any
-computation related to the argument should happen on the first use of the
-argument. So type-checking at this point and issuing an error message would
-violate the non-strict semantics; if this argument is never used, it would
-be incorrect to type-check it at the beginning of call. To make this work,
-\contractr mutates the function parameter binding to a promise that wraps
-the argument value in a call to the type-checker, as in the previous case.
-
-Secondly, we faced the problem of type-checking function return value. In R,
-the result of a function call is the result of evaluating the last
-expression in the function body; so an explicit return call is absent from
-most function definitions. Thus, there could be many potential
-sub-expressions in the function body which need to be type-checked against
-the return type. To address this, \contractr registers the type-checker to
-be called on the return value through a function exit hook. This hook is
-executed in the function call's environment after it has executed. While
-this works, we have to contend with yet another wrinkle, R interpreter can
-perform longjumps in its C implementation, which causes active R function
-calls on the stack to be discarded. When they are discarded, their exit
-hooks are called, which in this case calls the registered type-checker. But,
-these functions are in the middle of an active computation, so they don't
-have a return value to type-check. \contractr deals with this problem by
-allocating a unique sentinel object which serves as the return value for
-calls that are discarded. The function call exit hook does not call the
-type-checker if this unique value happens to be the return value of the
-call.
-
-%
+\paragraph{Checking the values} A number of properties can be checked
+by a simple tag check, namely wether a value is a: null, environment, vector
+or scalar. Other properties require an inspection of the contents of a
+value: the absence of \NA or the type of the elements of list. Union types
+require checking for each of the alternatives. We do not check higher-order
+functions in this version of the tool.
+
+When argument values are wrapped in promises (this is not always the case
+due to compiler optimizations), in order to retain the non-strict semantics
+of R, the expression held in the promise is wrapped in a call to the type
+checker, and type checking is delayed until the promise is forced. This
+leads to corner cases such that the type checking of a function may happen
+after that function has returned.
+
+Return values require care as well.  Functions return the last expression
+they evaluate, thus to avoid having to analyze the code of the called
+function the checker will register a callback on the exit hook. The hook is
+executed in the function call's environment. Another wrinkle is due to
+longjumps which causes active function calls on the stack to be
+discarded. When they are discarded, their exit hooks are called.  But, these
+function do not have a return value to type-check. \contractr deals with
+this problem by allocating a unique sentinel object which serves as the
+return value for calls that are discarded. The exit hook does not call the
+type-checker if it see the sentinel.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2020-05-15 evaluation/R/corpus.Rmd

@@
 
+```{r}
+pkgs_coverage_funs <-
+  filter(coverage_funs, coverage > 0) %>%
+  count(package) %>%
+  rename(funs_with_coverage=n)
+```
+
 ### Package runs
@@
 
+```{r}
+pkgs_revdeps_coverage <- 
+  read_csv(
+    PACKAGE_REVDEPS_COVERAGE,
+    col_types=cols(
+      package = col_character(),
+      revdep_coverage_expression = col_double(),
+      revdep_coverage_line = col_double(),
+      n_revdeps = col_double(),
+      n_ran_revdeps = col_double()
+    )
+  ) %>%
+  select(
+    package,
+    revdep_coverage_expr=revdep_coverage_expression,
+    revdep_coverage_line,
+    revdep_coverage_runs=n_ran_revdeps
+  )
+```
+
 ### Package asserts
@@
   left_join(pkgs_coverage, by="package") %>%
+  left_join(pkgs_coverage_funs, by="package") %>%
   left_join(pkgs_runs, by="package") %>%

### 2020-05-15 paper/main.tex
merge

--- 2020-05-15 paper/main.tex
@@
 
-\AT{Me}
-
-In the last section, we evaluated our type annotation language based on a set of function types that were automatically generated by \typetracer.
+In the last section, we mainly discussed the function types that \typetracer generated, focusing on analyzing the polymorphism present in the types.
 Now, we're interested in knowing how representative our trace-generated types are of the functions they were generated for.
-To measure this, we conducted a large-scale experiment: for each package in the corpus discussed in Section~\ref{sec:corpus}, we ran \typetracer on its test, example, and vignette code, obtaining types for package functions, and then subsequently ran the test, example, and vignette code of each of the package's reverse dependencies (or clients).
+To measure this, we conducted a large-scale experiment: for each package in the corpus discussed in Section~\ref{sec:corpus}, we ran \typetracer on its test, example, and vignette code, obtaining types for package functions, and then subsequently ran the test, example, and vignette code of each of the package's reverse dependencies.
 We collected failed contract assertions in order to analyze them and perhaps identify weaknesses in our type annotation language.
@@
 Overall, we found that only \PERCFAILEDASSERTIONSWUNDEF of contract assertions failed.
-We made a simplification in our analysis, capping the number of function arguments we generate types for to 20, and this resulted in only \PERCASSERTIONSUNDEF failed assertions, which suggests that the simplification we made did not greatly affect the results.
+We made a simplification in our analysis, capping the number of function arguments we generate types for to 20, and this resulted in only \PERCASSERTIONSUNDEF of assertions failing: this suggests that the simplification we made did not greatly affect the results.
 Once we controlled for these, we found that \PERCFAILEDASSERTIONS of all assertions failed for reasons related to our generated types.
@@
 We breakdown the failed assertions by type in Table~\ref{tbl:fail-breakdown-0}.
-By far the most commonly failed assertion is one asserting that a value with ${\bf dbl}[]$ type has type ${\bf class}\langle matrix \rangle$.
+By far the most commonly failed assertion is one asserting that a value with \VEC\dbl type has type \CLASS{matrix}.
 This was a deliberate design decision, as coercion from vectors to matrices is ad hoc at best, and surely not a practice codified in the language.
-In a similar vein, another popular failing assertion is checking if a ${\bf dbl}[]$ has type ${\bf int}[]$, another case of commonly performed coercion.
+In a similar vein, another popular failing assertion is checking if a \VEC\dbl has type \VEC\int, another case of commonly performed coercion.
 We did not include these types of coercions in our type annotation framework as programmers cannot rely on them, and it is not always the case that the coercions are safe to perform.
@@
 
-In addition to the raw number of failing contract checks, we were interested in how many functions had a parameter where a contract check failed.
-Overall, we found that \PROPFUNSFAILEDCHECK of functions had a failing contract check.
-Digging a little deeper, we noticed that many of these functions were performing S3 dispatch, which is part of R's object-orientation framework which we will tackle in future work.
+In addition to the raw number of failing contract checks, we were interested in how many functions had a parameter where a contract check failed, and overall we found this to be the case in \PROPFUNSFAILEDCHECK of functions.
+Digging a little deeper, we noticed that many of these functions were performing S3 dispatch, which is part of R's object-orientation framework: thus, these functions implement user-defined polymorphism which we do not handle in our type framework.
 Removing those functions, we see that \PROPFUNSFAILEDCHECKNOSTHREE of functions exhibited failing contract checks.

### 2020-05-15 evaluation/R/corpus.Rmd
Add revdeps code coverage into corpus

@@
   left_join(pkgs_revdeps, by="package") %>%
+  left_join(pkgs_revdeps_coverage, by="package") %>%
   left_join(pkgs_asserts, by="package") %>%

### 2020-05-15 evaluation/R/corpus-analysis.Rmd
Updated 5

@@
 ```{r}
-pkgs <- read_csv(
-  CRAN_PACKAGES_DETAILS,
-  col_types=cols(
-    package = col_character(),
-    installed = col_logical(),
-    loadable = col_logical(),
-    version = col_character(),
-    title = col_character(),
-    size = col_double(),
-    package_native_code = col_double(),
-    package_r_code = col_double(),
-    package_code = col_double(),
-    runnable_code = col_double(),
-    runnable_files = col_double(),
-    runnable_code_examples = col_double(),
-    runnable_code_vignettes = col_double(),
-    runnable_code_tests = col_double(),
-    runnable_files_examples = col_double(),
-    runnable_files_vignettes = col_double(),
-    runnable_files_tests = col_double(),
-    coverage_expr = col_double(),
-    coverage_line = col_double(),
-    runnable_success = col_double(),
-    runnable_error = col_double(),
-    runnable_time = col_double(),
-    runnable_rate = col_double(),
-    funs_private = col_double(),
-    funs_public = col_double(),
-    funs = col_double(),
-    revdeps = col_double(),
-    funs_with_assert = col_double(),
-    asserts = col_double(),
-    uses_roxygen = col_logical()
-  )
-)
+pkgs <- read_csv(CRAN_PACKAGES_DETAILS)
 ```
@@
     r(paste(prefix, "with coverage"), w_coverage),
-    r(paste(prefix, "mean expr coverage"), percent(mean(w_coverage))),
-    r(paste(prefix, "median expr coverage"), percent(median(w_coverage))),
+    r(paste(prefix, "mean expr coverage"), percent(mean(w_coverage$coverage_expr))),
+    r(paste(prefix, "median expr coverage"), percent(median(w_coverage$coverage_expr))),
+    r(paste(prefix, "covered functions ratio"), ratio(sum(w_coverage$funs_with_coverage),sum(w_coverage$funs))),
 
@@
     r(paste(prefix, "private functions"), sum(df$funs_private, na.rm=T)),
+    r(paste(prefix, "S three functions"), sum(df$funs_s3, na.rm=T)),
+    r(paste(prefix, "S three classes"), sum(df$s3_classes, na.rm=T)),
 
     r(paste(prefix, "w revdes"), w_revdeps),
-    r(paste(prefix, "mean revdes"), mean(w_revdeps)),
-    r(paste(prefix, "median revdes"), median(w_revdeps)),
+    r(paste(prefix, "revdes"), sum(w_revdeps$revdeps)),
+    r(paste(prefix, "mean revdes"), mean(w_revdeps$revdeps)),
+    r(paste(prefix, "median revdes"), median(w_revdeps$revdeps)),
     r(paste(prefix, "at least min revdep"), filter(w_revdeps, revdeps >= params$min_revdeps)),
+    r(paste(prefix, "revdep runnable code"), sum(w_revdeps$revdeps_runnable_code, na.rm=T)),
+    r(paste(prefix, "revdep mean code coverage"), percent(mean(w_revdeps$revdep_coverage_expr)/100)),
+    r(paste(prefix, "revdep median code coverage"), percent(median(w_revdeps$revdep_coverage_expr)/100)),
 

### 2020-05-15 paper/main.tex
Updated 5

@@
 For this paper we have selected \CorpusLoadable packages consisting of
-\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native
-code (C/C++/Fortran). Figure~\ref{fig:corpus} shows all the packages, the
-size of the dots reflects the size of the project in terms of both R and
-native lines of code\footnote{ Lines of source code reported excludes
-  comments and blank lines, counted by \emph{cloc}, \cf
-  \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the
-expression code coverage in percent and the y-axis gives the number of
-reverse dependencies in log scale. Dotted lines indicate means.  Packages
-with over \PackageSizeOutierRnd lines of code are annotated.
+\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native code
+(C/C++/Fortran). Figure~\ref{fig:corpus} shows these packages, the size of the
+dots reflects the size of the project in terms of lines of code including both R
+and native code\footnote{ Lines of source code reported excludes comments and
+  blank lines, counted by \emph{cloc}, \cf
+  \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the expression
+code coverage in percents and the y-axis gives the number of reverse
+dependencies in log scale. Dotted lines indicate means. Packages with over
+\PackageSizeOutierRnd lines of code are annotated.
 
 \begin{figure}[!h]  \centering
-  \includegraphics[width=.98\linewidth]{plots/corpus.pdf}
+  \includegraphics[width=.8\linewidth]{plots/corpus.pdf}
   \caption{Corpus overview}\label{fig:corpus}
@@
 These packages come from the Comprehensive R Achive Network
-(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of
-R code with over \AllCranRnd packages\footnote{CRAN receives about 6 new
-  package submissions a day~\cite{Ligges2017}} containing over \AllRCodeRnd
-and \AllNativeCodeRnd R and native code respectively. Unlike other open code
-repositories like GitHub, CRAN is a curated repository. Each submitted
-package must abide to a number of well-formedness rules that are
-automatically checked asserting certain quality. Most relevant for this work
-is that all of the runnable code (including code snippets from examples and
-vignettes) is tested and only a successfully running package is admitted in
-the archive.
+(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of R
+code with over \AllCranRnd packages\footnote{CRAN receives about 6 new package
+  submissions a day~\cite{Ligges2017}} containing over \AllRCodeRnd and
+\AllNativeCodeRnd R and native code respectively. Unlike other open code
+repositories such as GitHub, CRAN is a curated repository. Each submitted
+package must abide to a number of well-formedness rules that are automatically
+checked asserting certain quality. Most relevant for this work is that all of
+the runnable code (including code snippets from examples and vignettes) is
+tested and only a successfully running package is admitted in the archive.
 
 We have downloaded and installed all available CRAN packages. Out of the
-\AllCranRnd packages, we managed to install \AllLoadableRnd. The main reason
-for this is that \rdt is based on R 3.5.0 and some of the packages are no
-longer compatible with it. Some packages also require extra native
-dependencies which were not present on our servers.  We have two criteria
-for including a package into the corpus:
+\AllCranRnd packages, we managed to install \AllLoadableRnd. The main reason for
+this is that \rdt is based on R 3.5.0 and some of the packages are no longer
+compatible with this version. Some packages also require extra native
+dependencies which were not present on our servers. We defined two criteria for
+including a package into the corpus:
 \begin{inparaenum}[(1)]
-\item have runnable code that covers significant part of the package source
-  code from which type signatures could be inferred, and
-\item have some reverse dependencies that allows us to evaluate the inferred
-  types on runnable code from these dependent packages.
+  \item have a runnable code that covers a significant part of the package
+  source code from which type signatures could be inferred, and
+  \item have some reverse dependencies that will allows us to evaluate the
+  inferred types using the runnable code from these dependencies.
 \end{inparaenum}
-The concrete thresholds were set to at least \ThresholdCodeCoverage of
-expression coverage and at minimum \ThresholdRevdeps reverse dependencies.
-
-The code coverage was computed for each package using
-\covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R
-tool.  The reverse package dependencies were extracted from the package
-metadata using builtin function.
-
-%
+The concrete thresholds used were: at least \ThresholdCodeCoverage of expression
+coverage and at minimum \ThresholdRevdeps reverse dependencies. The code
+coverage was computed for each package using
+\covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R tool.
+The reverse package dependencies were extracted from the package metadata using
+builtin function.
+
+The \CorpusLoadable selected packages contain \CorpusRunnableCodeRnd lines of
+runnable code in examples (\CorpusExamplesCodeRnd), tests (\CorpusTestsCodeRnd)
+and vignettes (\CorpusVignettesCodeRnd). Running this code results on average in
+\CorpusMeanExprCoverage package code coverage (CRAN average is
+\AllMeanExprCoverage). Together, there is \CorpusRevdesRnd (on average
+\CorpusMeanRevdes, median \CorpusMedianRevdes; CRAN average is \AllMeanRevdes,
+median \AllMedianRevdes) reverse dependencies with \CorpusRevdepRunnableCodeRnd
+runnable lines of code resulting in \CorpusRevdepMeanCodeCoverage coverage (on
+average) of the corpus packages.
+
+Together there are \CorpusFunctionsRnd defined R functions
+(\CorpusPublicFunctionsRnd are from packages' public API).
+\CorpusSThreeFunctionsRnd are S3 functions. Packages in the corpus define
+\CorpusSThreeClasses S3 classes.
+
 %

### 2020-05-15 paper/main.tex
unlock 6 7

--- 2020-05-15 paper/main.tex
@@
 missing arguments: type them as \any or type them as some unit type.  As we
-are performing a dynamic analysis, we conservatively type them as \any.
+are performing a dynamic analysis, we conservatively type them as
+\LIST{\any}.
 
@@
 
-%
-%
-%
-\section{Left overs}
-
-SOME OF THIS STUFF CAN BE USED IN THE EVALUATION
-
-%
-%
-\subsubsection{Possibly-NA Primitives}
-
-A common primitive usage pattern is related to the presence of NAs:
-programmers explicitly check for NAs in primitive vectors, possibly
-sanitizing them if they are present.  NAs complicate practically every
-computation using primitive vectors, a fact that many programmers appear to
-be acutely aware of. For instance, consider the code in
-Figure~\ref{fig:na-example}.
-
-\begin{figure}[htbp]
-\begin{center}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Evaluation} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newcommand{\FIX}[1]{{\bf #1}\xspace}
+
+\subsection{Expressiveness}
+
+The first part of our evaluation attempts to shed light on how good a fit
+our proposed type system is with respect to common programming patterns
+occuring in widely used R libraries.
+
+First we look at the share of monomorphic arguments and function signatures.
+Monomorphic in this context means that the type is not relying on \any or
+inlcuding a union.  The import of monorphism in this context is that it
+means our type language can accurately capture an argument's type or a
+function signature.
+
+<<Insert proportion of monorphic types in argument and in function
+signature. Perhaps a table of monomorphic type signatures that are
+popular...>>
+
+\begin{figure}[!h]  \centering
+  \includegraphics[width=\linewidth]{plots/union_frequency.pdf}
+  \caption{Size of unions}\label{fig:unionfreq}
+\end{figure}
+
+We observe that XXX of the positions have a single type ascribed to them,
+and, only XXX have four or more types ascribed to them. Parameter PARAMNAME
+of function XXXXCOLONCOLONXXX has the maximum, XXX, types ascribed to it.
+Hence, most functions admit or return a uni-typed value.
+
+Figure 1 provides a breakdown of the different type categories observed at
+uni-typed positions.
+
+\input{tables/uni-typed-breakdown.tex}
+
+Figure XXX shows ten most common type categories that occur by
+themselves. We observe that three scalar types, double, logical, and
+character are most frequent. This is followed by XXX vararg parameter
+positions. Next, XX\% of the parameter types are any because arguments at
+those positions were either missing or unused in at least one call to those
+functions. Together, these types amount to more than 50\%. This is followed
+by double vector, the unit type, null, character vector, and classes matrix
+and data.frame respectively.
+
+Figure 1 shows the number of alternatives in the inferred type of function
+parameter and return values.
+
+XXX\% of the parameter positions are monomorphic.  \typetracer observes XXX
+calls to XXX functions with XXX arguments. This translates to XXX parameters
+and XXX return types.  A parameter or return type is from unioning all the
+distinct observed types of
+
+Figure 1 shows the distribution of parameter and return types for these
+functions.
+
+class is the most common type in our corpus, accounting for XXX\% of all the
+inferred types. This is followed by an almost equal proportion,
+28\%, scalar types and, surprisingly, only 13\% vector types. Vectors are the
+workhorse of R, most primitive operations in R are vectorized and literals are
+treated as single element vectors. It is unexpected to observe twice as many
+scalars as vectors. It is even more unexpected to observe class as the most
+common type. class of an object is used for dynamic dispatch by the S3 object
+system in R. This suggests that usage of S3 is prevalent in our corpus.
+
+null types. Together, these four types make up for 80\% of the observed
+types in our corpus.  The other types, occurring only XXX\% of the times
+include XXX environment types, XXXX expression types, XXXX pairlist types
+and lastly, XXX externalptr types.
+
+In order to count the occurrences of each type, we subdivided them into {\it
+  kinds}: For our purposes, the {\it kind} of a type is equivalent to it's
+top-level type constructor, so all classes have kind ``class'', all lists
+have kind ``list'', and so on.  We found that classes are the most common
+kind of type, accounting for roughly 31\% of types.  In order, the most
+common classes are matrices (12\%), data.frames (7.5\%), formulas (2\%),
+factors (2\%), and tibbles (2\%).  Roughly 25\% of classes are part of R's
+base language implementation, the others being defined as part of some
+package's functionality.
+
+Besides classes, scalars and vectors are the next most common kind, making
+up 41\% of types. with scalars making up 28.1\% of types and vectors
+12.65\%.  Nulls and lists follow at 8\% and 7\% respectively, and the vararg
+type \ldots makes up 6.5\% of arguments.  This all totals up to over 90\% of
+types.  We will discuss the any type in more detail in
+Section~\ref{subsec:any}.
+
+The polymorphism in R is best described by {\it ad hoc} polymorphism,
+wherein polymorphism is implemented by function and operator overloading
+(see S3 dispatch) rather than in some more fundamental way in the language
+design.  For the purposes of this evaluation, we will say that an argument
+is polymorphic if its type is a union of two or more types, in other words
+the argument has been observed to take more than one type of value.
+
+We found that 15\% of argument positions were polymorphic, and that X\% of
+functions were polymorphic.
+
+Our type annotation language does not support sophisticated class typing,
+and much of the polymorphism in R is due to these classes, namely through
+the S3 dispatch mechanism.  In future work, we aim to more fully handle S3
+and objects in R.  For now, if we count out
+
+\subsubsection{Discussion}
+
+\paragraph{NAs}
+Our data supports the choice of making the presence of NAs explicit.  In
+practice, programmers explicitly check for NAs in primitive vectors,
+possibly sanitizing them if they are present.
 
@@
 
-\caption{NA checking example from the \code{binom} package.}
-\label{fig:na-example}
-\end{center}
-\end{figure}
-
 The \code{binom} package is a popular R package for computing confidence
-intervals for binomial experiments, and the \code{binom.profile} function
-does so using the profile likelihood, the details of which are not relevant
-here.  This code snippet highlights a data sanitization pattern appearing at
-the beginning of the function: the programmer first binds the vectors into a
-matrix (line 2), with one column for each of \code{x} and \code{n}, then
-finds rows where both columns are not NA (line 3), then extracts non-NA
-values and stores them into \code{x} and \code{n} respectively (lines 4-5).
-Importantly, this process is entirely quiet, and the user of
-\code{binom.profile} will not be notified if their data happens to contain
-NAs, and the NAs will be silently ignored.  In reality, the presence of NAs
-in data could signify some sort of data corruption, and quietly ignoring
-issues in data could lead users (read: statisticians) to draw flawed
-conclusions from erroneous data.
-
-To combat this, we introduce a modifier on primitive types: the type
-\code{^T} signifies that primitive type \code{T} might contain an NA value,
-while \code{T} signifies that \code{T} is NA-free.  The NA-free case has a
-shorter syntax as NA-free vectors are \AT{more common} than the alternative.
-If the \code{binom} developers were to annotate arguments \code{x} and
-\code{n} with such an NA-free type, \contractr will notify users when
-\code{binom.profile} is called with NA-full values, rather than just quietly
-ignoring them.
-
-Note that when combining call traces involving \code{^T} and \code{T} types,
-we subsume \code{T} types into \code{^T} types: if NA-free and NA-full
-vectors are passed to an argument, then we say that the argument can handle
-NA-full vectors.  This is equivalent to saying \code{T <: ^T}.
-
-%
-%
-\subsubsection{Scalar Primitives}
-
-Initial analyses of our data revealed that programmers \AT{often} use
-scalars (as explicitly as they can, anyway), and often do high-level
-dimensionality checks on their data.  To illustrate, consider the code in
-Figure~\ref{fig:scalar-vector-example}.
-
-\begin{figure}[htbp]
-\begin{center}
+intervals, and the \code{binom.profile} function does so using the profile
+likelihood.  This code snippet highlights a data sanitization pattern: the
+programmer first binds the vectors into a matrix, with one column for each
+of \code{x} and \code{n}, then finds rows where both columns are not \NA,
+then extracts non-\NA values and stores them into \code{x} and \code{n}
+respectively.
+
+\paragraph{Scalars}
+The data also suggests that programmers often use scalars, and do
+dimensionality checks on their data.
 
@@
     ...
-    if ( n != trunc( n ) )
-        stop( "argument n ix not an integer" )
-    if ( !is.vector( x ) )
-        stop( "argument x is not a vector" )
+    if ( n != trunc( n ) )  stop( "argument n ix not an integer" )
+    if ( !is.vector( x ) )  stop( "argument x is not a vector" )
     m <- length( x )
-    if ( m < n )
-        stop( "length of argument x is less than n" )
+    if ( m < n )  stop( "length of argument x is less than n" )
     ...
@@
 
-\caption{Scalar/vector example from the \code{matrixcalc} package.}
-\label{fig:scalar-vector-example}
-\end{center}
-\end{figure}
-
-The \code{hankel.matrix} function takes two arguments (as described in the
-comment in the code), and returns a Hankel matrix, which essentially is a
-matrix where the skew-diagonals have a desirable property.  The function
-body has a number of interesting checks: on lines 6-7, \code{n} undergoes an
-integer ``type check''; on lines 8-9, \code{x} undergoes a vector ``type
-check''; on line 11, \code{n} is assumed to be a scalar, as if it were a
-vector, the code \code{m < n} would essentially check \code{m < n[i]} for
-each element in \code{n}, returning a vector of comparison results, and the
-conditional of an if must be a scalar else R generates a warning.
-
-These checks exemplify functionality that we would like to include in our
-type annotation language, namely, we would like our type language to
-differentiate between scalars and vectors.  To that end, we introduce the
-following types into our type language: for some primitive type \code{T},
-the user may modify \code{T} with a modifier \code{[]} to indicate that it
-is a vector of \code{T}.  Perhaps surprisingly, we found that scalars
-\AT{occurred more frequently} than vectors, and for that reason we chose the
-shorter scalar type syntax as the default.  The NA-free modifier works as
-expected: \code{^T[]} indicates a vector of \code{T} that is free of NAs,
-while \code{^T} indicates a scalar which is not NA.
-
-%
-%
-\subsubsection{Matrices}
+The \code{hankel.matrix} function takes two arguments.  The function has a
+number of interesting checks: \code{n} undergoes an integer type check;
+\code{x} undergoes a vector type check; and \code{n} is expected to be a
+scalar, as if it were a vector, the code \code{m < n} would essentially
+check \code{m < n[i]} for each element in \code{n}, returning a vector of
+comparison results, and the conditional of an if must be a scalar else R
+generates a warning.
+
+\paragraph{Matrices}
 
@@
 vectors to matrices automatically.  Relying on automatic coercion is unwise,
-and many functions check to ensure that their arguments are matrices.  For
-instance, consider the code in Figure~\ref{fig:matrix-example}.
-
-\begin{figure}[htbp]
-\begin{center}
+and many functions check to ensure that their arguments are matrices.
 
@@
 
-\caption{Matrix example from the \code{matrixStats} package.}
-\label{fig:matrix-example}
-\end{center}
-\end{figure}
-
 The \code{rowWeightedMeans} function calculates the weighted means of rows
@@
 
-%
-%
-\subsubsection{Forgoing Dimensionality}
-
 Thus far we've seen instances where we introduce types that subsume existing
 argument validation code, but our intention is not to subsume {\it all} such
-code.  As an example of something we {\it do not} want to do, consider the
-code in Figure~\ref{fig:general-validation-example}.
-
-\begin{figure}[htbp]
-\begin{center}
+code.
 
@@
 
-\caption{General dimensionality check example from the \code{matrixStats} package.}
-\label{fig:general-validation-example}
-\end{center}
-\end{figure}
-
 This code is from later in the same \code{rowWeightedMeans} function
@@
 
-%
-%
-%
-%
-%
-\subsubsection{Relationship Between Lists and Vectors}
-
-The ease with which vectors may be transformed into lists is perhaps
-indicative of a relationship between the two.  In R, all vectors can be
-converted to lists with a simple call to \code{as.list}, and lists to
-vectors with \code{unlist}, though care must be taken when unlisting a
-non-homogeneous list, as coercions will be performed where possible, and the
-function will simply do nothing if coercion is impossible.  That said, lists
-and vectors have different access syntax, so vectors cannot stand-in for
-lists without the \code{as.list} call above.
-
-In sum, we opted not to include a subtyping relationship between vectors and
-lists, keeping both types separate.  In practice, \AT{how often do \code{T[]
-    | list<T>} types occur?}.  During our evaluation we also found it rare
-that programmers would attempt to pass list-typed values to vector-typed
-arguments and vice versa, and we will discuss this further in
-Section~\ref{sec:evaluation}.
-
-%
-%
-%
-\subsubsection{Structs}
-\label{subsec:structs}
+\paragraph{Structs.}
 
@@
   functions}.  In R, ``fields'' of named lists can be accessed using the
-\code{$} operator, as illustrated in the code snippet in
-Figure~\ref{fig:struct-ex}.
-
-\begin{figure}[htbp]
-\begin{center}
+\code{$} operator.
 
@@
 \end{lstlisting}
-
-\caption{Struct example from a test in the \code{agricolae} package.}
-\label{fig:struct-ex}
-\end{center}
-\end{figure}
 
@@
   often} structs-typed values had a {\it class}, which helped motivate our
-inclusion of class-like types in our type annotation language.  Classes are
-discussed in more detail in Section~\ref{subsec:classes}.
+inclusion of class-like types in our type annotation language.
 
 %
-\subsubsection{Data Frames}
-
-One of the most popular classes in R is the \code{data.frame} class \AT{we
-  have numbers here, grab em}.  Data frames and the derivative \code{tibble}
-(from the \code{tidyverse} ecosystem) and \code{data.table} (from the
-\code{data.table} package) types underpin a huge amount of data analysis in
-R.
+\paragraph{Data Frames}
+
+One of the most popular classes in R is the \code{data.frame} class.  Data
+frames and the derivative \code{tibble} (from the \code{tidyverse}
+ecosystem) and \code{data.table} (from the \code{data.table} package) types
+underpin a huge amount of data analysis in R.
 
@@
 column names to select rows from the frame (unless only a single column is
-of interest, wherein the \code{$} syntax is appropriate).  We include some
-examples in Figure~\ref{fig:data-frames-bad} to illustrate the complexities.
-
-\begin{figure}[htbp]
-\begin{center}
+of interest, wherein the \code{$} syntax is appropriate).
 
@@
 \end{lstlisting}
-
-\caption{Select examples of idiomatic \code{data.frame} use.}
-\label{fig:data-frames-bad}
-\end{center}
-\end{figure}
 
@@
 
-%
-%
-%
-%
-\subsection{Classes in R}
-\label{subsec:classes}
-
-As discussed in Section~\ref{sec:R}, the R runtime has more than one notion
-of type.  The classes of function arguments are used by R as part of its
-dispatch mechanism, the most prevalent of which (\AT{we have that number,
-  too!}) is S3 dispatch, in which R dispatches on the (S3) class of the
-first argument to a function, selecting the appropriate method to call for
-the class.  To quickly illustrate, consider the code snippet in
-Figure~\ref{fig:dispatch-ex}.
-
-\begin{figure}[htbp]
-\begin{center}
-
-\begin{lstlisting}
-# Quick example of S3 dispatch.
-\end{lstlisting}
-
-\caption{S3 dispatch in R.}
-\label{fig:dispatch-ex}
-\end{center}
-\end{figure}
-
-Classes are very prevalent in R \AT{number}, and so we include a
-\code{class<class-name>} ... \AT{todo} \typetracer was outfit to report the
-class of values, but some care needed to be taken so as to not overreport
-class types, as {\it all} values have a class in R.  The main classes we
-were interested in were matrix, array (higher dimensional matrices), and
-data.frames, in addition to any user-defined classes.  In many cases
-\AT{number?}, values with a user-defined class are in reality lists with
-named elements, or structs.  Including class types, and having \typetracer
-report them in place of structs, significantly reduced the noise in the
-analysis and allowed us to gather more focused data.  To get a sense for
-this difference, consider the code in Figure~\ref{fig:class-vs-struct}.
-
-\begin{figure}[htbp]
-\begin{center}
-
-\begin{lstlisting}
-# Example of gross struct that turns into class.
-\end{lstlisting}
-
-\caption{S3 dispatch in R.}
-\label{fig:class-vs-struct}
-\end{center}
-\end{figure}
-
-When set to ignore classes, \typetracer reports \AT{type} as the type of the function, as compared with \AT{other type} when configured to report classes.
-
-%
-%
-\subsubsection{Multi-Classes}
-
-A value can have multiple classes in R.  This is common practice in the
-\code{dplyr} package, which extends the functionality of data frames: a
-\code{data.frame} that passes through \code{dplyr} functions gains two
-additional classes (\code{tbl} and \code{tbl_df}), bringing the total to at
-least 3, in addition to whichever other classes that data frame happened to
-have.  Note that when looking to perform S3 dispatch, R will try to match
-{\it any} of the classes of the first argument to a dispatched function, so
-each of these classes are important.
-
-We found these so-called multi-classes to be surprisingly common, making up
-\AT{occurrence} of the occurrences of class-typed values.  One possible
-reason for this is that values can gain classes quite easily (e.g. the
-\code{dplyr} example discussed above), somewhat polluting the output of
-\typetracer.  Nevertheless, we found \AT{number} of occurrences of
-monomorphic multi-class arguments, meaning that the function was always
-called with a multi-class value in that argument position.
-
-In the end, we support these multi-class types by allowing \code{class<...>}
-types to contain a list of class names, so for instance a \code{dplyr}
-\code{tibble} would have type \code{class<data.frame, tbl, tbl_df>} in our
-type language, and when performing related contract assertions \contractr
-will check that the value has every class that make up its type.
-
-%
-%
-\subsubsection{Forgoing R's Object Systems}
+\paragraph{Objects}
 
@@
 
-%
-%
-\section{Evaluation}
-\label{sec:evaluation}
-
-To evaluate the effectiveness of our type annotation language and contract
-checking framework, we conducted a large-scale experiment based on the
-corpus of 412 packages we used to inform our language design.  This corpus
-represents some of the most widely used and thoroughly tested R libraries on
-CRAN.
-
-We configured \typetracer to infer the types of each of the functions that
-are visible to package clients.  Recall that \typetracer's primary function
-is to collect call signatures from observed function calls, and we ran
-\typetracer on the test, example, and vignette code of all packages in the
-aforementioned corpus.  These call traces were consolidated into a single
-function type for each visible package function: \typetracer employs
-unification strategies discussed in Section~\ref{sec:typesystemdesign} to
-keep the size of signatures in check and only generates an \code{any} type
-when missing and unused arguments were recored, and when a longjump caused
-function execution to abort prematurely.  The types generated by \typetracer
-will be analyzed in more depth in Section~\ref{subsec:stats}, and a
-large-scale evaluation of the quality of these types will be discussed in
-Section~\ref{subsec:eval}.
-
-%
-%
-%
-%
-\subsection{Language Statistics}
-\label{subsec:stats}
-
-\subsection{Is our simple type language expressive enough to type R code?} Functions in
-R can support arguments of multiple types by reflecting on the type to
-perform appropriate conversions and computations. On the one hand, this form
-of ad-hoc polymorphism makes R functions expressive; on the other hand, it
-requires support for polymorphism in the typing scheme, making the type
-system complex.  Our system lacks explicit support for
-polymorphism. Instead, it supports union, \any and \VARGS types. A union type is
-inferred when values of multiple types are observed at a function parameter
-and return position, \any type is inferred when an argument is missing or
-unused, and \VARGS is inferred for variable length argument list positions. Furthermore, list element
-types can be inferred as \any if they have more than 5 different types. Values with many different types can result in
-a large number of alternatives in the type inferred for that position,
-making it hard to read and unwieldy for end-users. To evaluate this
-''expressivity gap`' between our type inference scheme and R functions, we
-look at the number of different types in the parameter and return types of
-the XXX functions in our corpus.
-
-\begin{figure}[!h]  \centering
-  \includegraphics[width=\linewidth]{plots/union_frequency.pdf}
-  \caption{Types inferred for each position}\label{fig:unionfreq}
-\end{figure}
-
-We observe that XXX of the positions have a single type ascribed to them, and,
-only XXX have four or more types ascribed to them. Parameter PARAMNAME of function
-XXXXCOLONCOLONXXX has the maximum, XXX, types ascribed to it.
-Hence, most functions admit or return a uni-typed value.
-
-Figure 1 provides a breakdown of the different type categories observed at
-uni-typed positions.
-
-\input{tables/uni-typed-breakdown.tex}
-
-Figure XXX shows ten most common type categories that occur by themselves. We
-observe that three scalar types, double, logical, and character are most
-frequent. This is followed by XXX vararg parameter positions. Next, XX% of the
-parameter types are any because arguments at those positions were either missing
-or unused in at least one call to those functions. Together, these types amount
-to more than 50%. This is followed by double vector, the unit type, null,
-character vector, and classes matrix and data.frame respectively.
-
-Figure 1 shows the number of alternatives in the inferred type of function
-parameter and return values.
-
-XXX\% of the parameter positions are monomorphic. 
-\typetracer observes XXX calls to XXX functions with XXX arguments. This
-translates to XXX parameters and XXX return types. 
-A parameter or return type is  from unioning all the distinct observed types of
-
-Figure 1 shows the distribution of parameter and return types for these
-functions.
-
-class is the most common type in our corpus, accounting for XXX% of all the
-inferred types. This is followed by an almost equal proportion,
-28%, scalar types and, surprisingly, only 13% vector types. Vectors are the
-workhorse of R, most primitive operations in R are vectorized and literals are
-treated as single element vectors. It is unexpected to observe twice as many
-scalars as vectors. It is even more unexpected to observe class as the most
-common type. class of an object is used for dynamic dispatch by the S3 object
-system in R. This suggests that usage of S3 is prevalent in our corpus.
-
-null types. Together, these four types make up for 80% of the observed types in
-our corpus.
-The other types, occurring only XXX% of the times include XXX environment types,
-XXXX expression types, XXXX pairlist types and lastly, XXX externalptr types.
-
-We analyzed the type signatures generated by \typetracer in order to build
-up a profile of our type annotation language.  We were interested in broad,
-high-level information, such as the distribution of generated types, the
-level of polymorphism observed in the types, and the distribution of sizes
-of union types.
-
-%
-%
-\subsubsection{Type Distribution}
-
-\AT{Include a figure? Histogram(s) of types?}
-
-In order to count the occurrences of each type, we subdivided them into {\it kinds}: 
-For our purposes, the {\it kind} of a type is equivalent to it's top-level type constructor, so all classes have kind ``class'', all lists have kind ``list'', and so on.
-We found that classes are the most common kind of type, accounting for roughly 31\% of types.
-\AT{Make sure to transform class<function> into any=>any.}
-In order, the most common classes are matrices (12\%), data.frames (7.5\%), formulas (2\%), factors (2\%), and tibbles (2\%).
-Roughly 25\% of classes are part of R's base language implementation, the others being defined as part of some package's functionality.
-
-Besides classes, scalars and vectors are the next most common kind, making up 41\% of types. with scalars making up 28.1\% of types and vectors 12.65\%.
-Nulls and lists follow at 8\% and 7\% respectively, and the vararg type \ldots makes up 6.5\% of arguments.
-This all totals up to over 90\% of types.
-We will discuss the any type in more detail in Section~\ref{subsec:any}.
-
-It would appear that our simple type annotations nicely account for the landscape of values in R.
-\AT{Not sure what to say about classes.}
-% Classes are a common kind of type in R, and for the purposes of this type annotation language allowing users to specify class names as part of class types is welcome: we saw that the breakdown of the class kind does not reveal many outlier classes,
-
-%
-%
-\subsubsection{Polymorphism}
-
-The polymorphism in R is best described by {\it ad hoc} polymorphism, wherein polymorphism is implemented by function and operator overloading (see S3 dispatch) rather than in some more fundamental way in the language design.
-For the purposes of this evaluation, we will say that an argument is polymorphic if its type is a union of two or more types, in other words the argument has been observed to take more than one type of value.
-
-\AT{We need better numbers on this.}
-We found that 15\% of argument positions were polymorphic, and that X\% of functions were polymorphic.
-
-Our type annotation language does not support sophisticated class typing, and much of the polymorphism in R is due to these classes, namely through the S3 dispatch mechanism. 
-In future work, we aim to more fully handle S3 and objects in R.
-For now, if we count out
-
-%
-%
-\subsubsection{Size of Union Types}
-
-%
-%
-\subsubsection{Use of \code{any}}
-\label{subsec:any}
-
-%
-%
-%
-%
-\subsection{RQ2: How good are type signatures generated from traces?}
-\label{subsec:eval}
-
-In the last section, we mainly discussed the function types that \typetracer generated, focusing on analyzing the polymorphism present in the types.
-Now, we're interested in knowing how representative our trace-generated types are of the functions they were generated for.
-To measure this, we conducted a large-scale experiment: for each package in the corpus discussed in Section~\ref{sec:corpus}, we ran \typetracer on its test, example, and vignette code, obtaining types for package functions, and then subsequently ran the test, example, and vignette code of each of the package's reverse dependencies.
-We collected failed contract assertions in order to analyze them and perhaps identify weaknesses in our type annotation language.
-
-%
-%
-\subsubsection{Overall Results}
-
-We ran our evaluation on \NUMPKGSEVAL packages and recorded \NUMASSERTIONS total assertions.
-Overall, we found that only \PERCFAILEDASSERTIONSWUNDEF of contract assertions failed.
-We made a simplification in our analysis, capping the number of function arguments we generate types for to 20, and this resulted in only \PERCASSERTIONSUNDEF of assertions failing: this suggests that the simplification we made did not greatly affect the results.
-Once we controlled for these, we found that \PERCFAILEDASSERTIONS of all assertions failed for reasons related to our generated types.
-Additionally, we found that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of functions types were never violated.
-The number of immaculate function types increases to \PERCSUCCFUNNOSTHREE if we count out S3 functions:
-These S3 functions represent user-defined polymorphism which we do not tackle in this work.
-Overall, these numbers are promising, and suggest that the type signatures generated by running \typetracer on our corpus are indeed representative, and capture intended function behaviour.
-
-%
-%
-\subsubsection{Breakdown of Contract Assertion Failures}
-
-We breakdown the failed assertions by type in Table~\ref{tbl:fail-breakdown-0}.
-By far the most commonly failed assertion is one asserting that a value with \VEC\dbl type has type \CLASS{matrix}.
-This was a deliberate design decision, as coercion from vectors to matrices is ad hoc at best, and surely not a practice codified in the language.
-In a similar vein, another popular failing assertion is checking if a \VEC\dbl has type \VEC\int, another case of commonly performed coercion.
-We did not include these types of coercions in our type annotation framework as programmers cannot rely on them, and it is not always the case that the coercions are safe to perform.
+\subsection{Robustness}
+
+In the last section, we evaluated our type annotation language based on a
+set of function types that were automatically generated by \typetracer.
+Now, we're interested in knowing how representative our trace-generated
+types are of the functions they were generated for.  To measure this, we
+conducted a large-scale experiment: for each package in the corpus discussed
+in Section~\ref{sec:corpus}, we ran \typetracer on its test, example, and
+vignette code, obtaining types for package functions, and then subsequently
+ran the test, example, and vignette code of each of the package's reverse
+dependencies (or clients).  We collected failed contract assertions in order
+to analyze them and perhaps identify weaknesses in our type annotation
+language.
+
+We ran our evaluation on \NUMPKGSEVAL packages and recorded \NUMASSERTIONS
+total assertions.  Overall, we found that only \PERCFAILEDASSERTIONSWUNDEF
+of contract assertions failed.  We made a simplification in our analysis,
+capping the number of function arguments we generate types for to 20, and
+this resulted in only \PERCASSERTIONSUNDEF failed assertions, which suggests
+that the simplification we made did not greatly affect the results.  Once we
+controlled for these, we found that \PERCFAILEDASSERTIONS of all assertions
+failed for reasons related to our generated types.  Additionally, we found
+that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of functions types
+were never violated.  The number of immaculate function types increases to
+\PERCSUCCFUNNOSTHREE if we count out S3 functions: These S3 functions
+represent user-defined polymorphism which we do not tackle in this work.
+Overall, these numbers are promising, and suggest that the type signatures
+generated by running \typetracer on our corpus are indeed representative,
+and capture intended function behaviour.
+
+We breakdown the failed assertions by type in
+Table~\ref{tbl:fail-breakdown-0}.  By far the most commonly failed assertion
+is one asserting that a value with ${\bf dbl}[]$ type has type ${\bf
+  class}\langle matrix \rangle$.  This was a deliberate design decision, as
+coercion from vectors to matrices is ad hoc at best, and surely not a
+practice codified in the language.  In a similar vein, another popular
+failing assertion is checking if a ${\bf dbl}[]$ has type ${\bf int}[]$,
+another case of commonly performed coercion.  We did not include these types
+of coercions in our type annotation framework as programmers cannot rely on
+them, and it is not always the case that the coercions are safe to perform.
 
@@
 
-In addition to the raw number of failing contract checks, we were interested in how many functions had a parameter where a contract check failed, and overall we found this to be the case in \PROPFUNSFAILEDCHECK of functions.
-Digging a little deeper, we noticed that many of these functions were performing S3 dispatch, which is part of R's object-orientation framework: thus, these functions implement user-defined polymorphism which we do not handle in our type framework.
-Removing those functions, we see that \PROPFUNSFAILEDCHECKNOSTHREE of functions exhibited failing contract checks.
-These are functions which were under-tested, as calls to these functions represent only \PERCCALLSBADFUNSINTESTS of recorded calls during \typetracer's run on the core corpus to infer types.
-
-Turning our attention now to arguments, we found that only \PROPARGSFAILINGASSERTS of function parameters exhibited a failed assertion.
-Table~\ref{tbl:fail-breakdown-0} showed the raw results from counting runtime occurrences, but that data alone does not tell the full story, as some failures may be overrepresented if e.g.
-a failing contract assertion was in a loop.
-We were interested in knowing for each of the most common violations in Table~\ref{tbl:fail-breakdown-0}, how many different arguments had that type, and how many of those exhibited the contract failure in question.
-Table~\ref{tbl:fail-breakdown-1} breaks down the results from Table~\ref{tbl:fail-breakdown-0} by unique parameter position, folding away multiple identical failed contract assertions for the same parameter position.
-The first row of this table reads: a value of type \code{double[]} was passed to 15 different function parameters expecting a matrix, of which there are 1394 in total: 1.08\% of matrix-typed parameters were passed double vector values.
-
-We see that even though the double vector and matrix issue was wildly prevalent in the raw, dynamic contract evaluation numbers, the number of actual function argument types that were violated is very small.
-The story is similar with the double and integer coercion we mentioned earlier: it represents many dynamic contract failures, but very few of the integer vector-typed arguments have their contracts violated by double vector-typed values.
-Rows 5 and 6 are interesting: we see that rather often arguments expecting time series data are passed double vector or matrix values.
-This is a quirk of the \code{timeSeries} package, whose functions often accept matrices and vectors, converting them to time series in an ad hoc manner.
-Note that the code coverage of the \code{timeSeries} tests, examples, and vignettes on the \code{timeSeries} package code is only 58.24\%, which is one possible explanation of why these contract failures are occurring: the types that \typetracer generates are only as good as the test code its run on.
+In addition to the raw number of failing contract checks, we were interested
+in how many functions had a parameter where a contract check failed.
+Overall, we found that \PROPFUNSFAILEDCHECK of functions had a failing
+contract check.  Digging a little deeper, we noticed that many of these
+functions were performing S3 dispatch, which is part of R's
+object-orientation framework which we will tackle in future work.  Removing
+those functions, we see that \PROPFUNSFAILEDCHECKNOSTHREE of functions
+exhibited failing contract checks.  These are functions which were
+under-tested, as calls to these functions represent only
+\PERCCALLSBADFUNSINTESTS of recorded calls during \typetracer's run on the
+core corpus to infer types.
+
+Turning our attention now to arguments, we found that only
+\PROPARGSFAILINGASSERTS of function parameters exhibited a failed assertion.
+Table~\ref{tbl:fail-breakdown-0} showed the raw results from counting
+runtime occurrences, but that data alone does not tell the full story, as
+some failures may be overrepresented if e.g. a failing contract assertion
+was in a loop.  We were interested in knowing for each of the most common
+violations in Table~\ref{tbl:fail-breakdown-0}, how many different arguments
+had that type, and how many of those exhibited the contract failure in
+question.  Table~\ref{tbl:fail-breakdown-1} breaks down the results from
+Table~\ref{tbl:fail-breakdown-0} by unique parameter position, folding away
+multiple identical failed contract assertions for the same parameter
+position.  The first row of this table reads: a value of type
+\code{double[]} was passed to 15 different function parameters expecting a
+matrix, of which there are 1394 in total: 1.08\% of matrix-typed parameters
+were passed double vector values.
+
+We see that even though the double vector and matrix issue was wildly
+prevalent in the raw, dynamic contract evaluation numbers, the number of
+actual function argument types that were violated is very small.  The story
+is similar with the double and integer coercion we mentioned earlier: it
+represents many dynamic contract failures, but very few of the integer
+vector-typed arguments have their contracts violated by double vector-typed
+values.  Rows 5 and 6 are interesting: we see that rather often arguments
+expecting time series data are passed double vector or matrix values.  This
+is a quirk of the \code{timeSeries} package, whose functions often accept
+matrices and vectors, converting them to time series in an ad hoc manner.
+Note that the code coverage of the \code{timeSeries} tests, examples, and
+vignettes on the \code{timeSeries} package code is only 58.24\%, which is
+one possible explanation of why these contract failures are occurring: the
+types that \typetracer generates are only as good as the test code its run
+on.
 
@@
 
-Table~\ref{tbl:fail-breakdown-2} presents data on the most frequently violated contracts amongst the most frequently occurring argument types.
-We selected argument types which were in the 90th percentile of argument type occurrences, computed the most frequent type signature violations among them, and reported the most frequently violated contracts together with the type of the value that violated that contract.
-The first row of the table reads: 44 function arguments with integer vector type are passed double vectors instead, and 657 arguments have integer vector type.
-Had we failed to capture some key usage pattern of R with our type annotation framework, we would likely see it here, and we can see this in action if we consider Table~\ref{tbl:fail-breakdown-3}, which was obtained identically to Table~\ref{tbl:fail-breakdown-2} with a percentile of 80th percentile instead.
-The most frequent argument type violation pattern in that of data frames being passed to arguments expecting tibbles: this occurs in only 22\% of such arguments, and represents cases where tests did not adequately cover all valid function inputs.
-Separate from the issue of testing, we can capture this behaviour with user-defined subtyping, as data frames are subtypes of tibbles, but this necessitates a proper object-oriented type system for R which is outside the scope of this work.
+Table~\ref{tbl:fail-breakdown-2} presents data on the most frequently
+violated contracts amongst the most frequently occurring argument types.  We
+selected argument types which were in the 90th percentile of argument type
+occurrences, computed the most frequent type signature violations among
+them, and reported the most frequently violated contracts together with the
+type of the value that violated that contract.  The first row of the table
+reads: 44 function arguments with integer vector type are passed double
+vectors instead, and 657 arguments have integer vector type.  Had we failed
+to capture some key usage pattern of R with our type annotation framework,
+we would likely see it here, and we can see this in action if we consider
+Table~\ref{tbl:fail-breakdown-3}, which was obtained identically to
+Table~\ref{tbl:fail-breakdown-2} with a percentile of 80th percentile
+instead.  The most frequent argument type violation pattern in that of data
+frames being passed to arguments expecting tibbles: this occurs in only 22\%
+of such arguments, and represents cases where tests did not adequately cover
+all valid function inputs.  Separate from the issue of testing, we can
+capture this behaviour with user-defined subtyping, as data frames are
+subtypes of tibbles, but this necessitates a proper object-oriented type
+system for R which is outside the scope of this work.
 
@@
 
-In sum, we believe that this evaluation shows that the type signatures we generate from traces are quite good.
-Only \PERCFAILEDASSERTIONSWUNDEF of contract assertions failed at runtime, representing failures in as few as \PROPARGSFAILINGASSERTS of argument types.
-Even though \PROPFUNSFAILEDCHECKNOSTHREE of functions had at least one argument type involved in a failing contract check, these functions are under-tested, representing only \PERCCALLSBADFUNSINTESTS of calls observed while inferring types.
+In sum, we believe that this evaluation shows that the type signatures we
+generate from traces are quite good.  Only \PERCFAILEDASSERTIONSWUNDEF of
+contract assertions failed at runtime, representing failures in as few as
+\PROPARGSFAILINGASSERTS of argument types.  Even though
+\PROPFUNSFAILEDCHECKNOSTHREE of functions had at least one argument type
+involved in a failing contract check, these functions are under-tested,
+representing only \PERCCALLSBADFUNSINTESTS of calls observed while inferring
+types.
 

### 2020-05-15 paper/main.tex

@@
 These packages come from the Comprehensive R Achive Network
-(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of
-R code with over \AllCranRnd packages\footnote{CRAN receives about 6 new
-  package submissions a day~\cite{Ligges2017}} containing over \AllRCodeRnd
-and \AllNativeCodeRnd R and native code respectively. Unlike other open code
-repositories like GitHub, CRAN is a curated repository. Each submitted
-package must abide to a number of well-formedness rules that are
-automatically checked asserting certain quality. Most relevant for this work
-is that all of the runnable code (including code snippets from examples and
-vignettes) is tested and only a successfully running package is admitted in
-the archive.
+(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of R
+code with over \AllCranRnd packages\footnote{CRAN receives about 6 new package
+  submissions a day~\cite{Ligges2017}} containing over \AllRCodeRnd and
+\AllNativeCodeRnd R and native code respectively. Unlike other open code
+repositories such as GitHub, CRAN is a curated repository. Each submitted
+package must abide to a number of well-formedness rules that are automatically
+checked asserting certain quality. Most relevant for this work is that all of
+the runnable code (including code snippets from examples and vignettes) is
+tested and only a successfully running package is admitted in the archive.
 
 We have downloaded and installed all available CRAN packages. Out of the
-\AllCranRnd packages, we managed to install \AllLoadableRnd. The main reason
-for this is that \rdt is based on R 3.5.0 and some of the packages are no
-longer compatible with it. Some packages also require extra native
-dependencies which were not present on our servers.  We have two criteria
-for including a package into the corpus:
+\AllCranRnd packages, we managed to install \AllLoadableRnd. The main reason for
+this is that \rdt is based on R 3.5.0 and some of the packages are no longer
+compatible with this version. Some packages also require extra native
+dependencies which were not present on our servers. We defined two criteria for
+including a package into the corpus:
 \begin{inparaenum}[(1)]
-\item have runnable code that covers significant part of the package source
-  code from which type signatures could be inferred, and
-\item have some reverse dependencies that allows us to evaluate the inferred
-  types on runnable code from these dependent packages.
+  \item have a runnable code that covers a significant part of the package
+  source code from which type signatures could be inferred, and
+  \item have some reverse dependencies that will allows us to evaluate the
+  inferred types using the runnable code from these dependencies.
 \end{inparaenum}
-The concrete thresholds were set to at least \ThresholdCodeCoverage of
-expression coverage and at minimum \ThresholdRevdeps reverse dependencies.
+The concrete thresholds used were: at least \ThresholdCodeCoverage of expression
+coverage and at minimum \ThresholdRevdeps reverse dependencies. The code
+coverage was computed for each package using
+\covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R tool.
+The reverse package dependencies were extracted from the package metadata using
+builtin function.
+
+The \CorpusLoadable selected packages contain \CorpusRunnableCodeRnd lines of
+runnable code in examples (\CorpusExamplesCodeRnd), tests (\CorpusTestsCodeRnd)
+and vignettes (\CorpusVignettesCodeRnd). Running this code results on average in
+\CorpusMeanExprCoverage package code coverage (CRAN average is
+\AllMeanExprCoverage). Together, there is \CorpusRevdesRnd (on average
+\CorpusMeanRevdes, median \CorpusMedianRevdes; CRAN average is \AllMeanRevdes,
+median \AllMedianRevdes) reverse dependencies with \CorpusRevdepRunnableCodeRnd
+runnable lines of code resulting in \CorpusRevdepMeanCodeCoverage coverage (on
+average) of the corpus packages.
+
+Together there are \CorpusFunctionsRnd defined R functions
+(\CorpusPublicFunctionsRnd are from packages' public API).
+\CorpusSThreeFunctionsRnd are S3 functions. Packages in the corpus define
+\CorpusSThreeClasses S3 classes.
 
@@
 them, and it is not always the case that the coercions are safe to perform.
+=======
 
@@
 In addition to the raw number of failing contract checks, we were interested
-in how many functions had a parameter where a contract check failed.
-Overall, we found that \PROPFUNSFAILEDCHECK of functions had a failing
-contract check.  Digging a little deeper, we noticed that many of these
-functions were performing S3 dispatch, which is part of R's
-object-orientation framework which we will tackle in future work.  Removing
-those functions, we see that \PROPFUNSFAILEDCHECKNOSTHREE of functions
-exhibited failing contract checks.  These are functions which were
-under-tested, as calls to these functions represent only
-\PERCCALLSBADFUNSINTESTS of recorded calls during \typetracer's run on the
-core corpus to infer types.
+in how many functions had a parameter where a contract check failed, and
+overall we found this to be the case in \PROPFUNSFAILEDCHECK of functions.
+Digging a little deeper, we noticed that many of these functions were
+performing S3 dispatch, which is part of R's object-orientation framework:
+thus, these functions implement user-defined polymorphism which we do not
+handle in our type framework.  Removing those functions, we see that
+\PROPFUNSFAILEDCHECKNOSTHREE of functions exhibited failing contract checks.
+These are functions which were under-tested, as calls to these functions
+represent only \PERCCALLSBADFUNSINTESTS of recorded calls during
+\typetracer's run on the core corpus to infer types.
 

### 2020-05-15 paper/main.tex
minor

@@
 
-\begin{lstlisting}
+\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
   binom.profile <- function(x, n, ...) {
@@
    n <- xn[ok, "n"]
-  ...
 \end{lstlisting}
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
   hankel.matrix <- function( n, x ) { 
-### arguments
   ### n = a positive integer value for the order of the Hankel matrix
@@
     if ( m < n )  stop( "length of argument x is less than n" )
-    ...
 \end{lstlisting}
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
 rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL,
                              na.rm = FALSE, ...) {
-  # - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-  # Validate arguments
-  # - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-  # Argument 'x':
-  if (!is.matrix(x)) {
-    .Defunct(msg = sprintf("Argument 'x' is of class %s, but should be a matrix. 
-    The use of a %s is not supported, the correctness of the result is not guaranteed. 
-    Please update your code accordingly.", sQuote(class(x)[1]), sQuote(class(x)[1])))
-  }
-  ...
+  if (!is.matrix(x)) 
+    .Defunct(msg = sprintf("Argument 'x' should be a matrix.)
 \end{lstlisting}
@@
 
-\begin{lstlisting}
-rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL,
-                             na.rm = FALSE, ...) {
-  ...
+\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
+  rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL, na.rm = FALSE) {
     n <- ncol(x)
-  if (length(w) != n) {
-    stop("The length of argument 'w' is does not match the number of column in 'x': ", 
-    length(w), " != ", n)
-  }
-  ...
+    if (length(w) != n) 
+       stop("The length of 'w' does not match columns in 'x': ")
 \end{lstlisting}
@@
 
-\paragraph{Structs.}
+\paragraph{Structs}
 
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
   agricolae::cv.model <- function(x) {
@@
   }
-
   data(sweetpotato)
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
 data(cars)
-
 # tidyverse: Filter rows where speed is > 20. 
 fast_cars_tidy <- filter(cars, speed > 20)
-
 # base R: Get all rows where the "speed" column has a value > 20.
 fast_cars_base <- cars[cars[, "speed"] > 20, ] # base example
-
 \end{lstlisting}
@@
 them, and it is not always the case that the coercions are safe to perform.
-=======
 

### 2020-05-15 paper/main.tex
spell check

@@
 programming with minimal syntactic and conceptual overhead. This design is
-well suited to interactive data analysis, but a bad firt for tools such as
+well suited to interactive data analysis, but a bad fit for tools such as
 compilers or program analyzers which must generate performant code or catch
 programming errors.  In particular, R has no type annotations, its
-operations are dynamically checked at runtime. The starting point for our
+operations are dynamically checked at run-time. The starting point for our
 work are the twin questions, \emph{what expressive power is needed to
@@
 \item We implemented tooling to automatically extract type signatures from R
-  functions and to intrument R functions with checks based on their declared
+  functions and to instrument R functions with checks based on their declared
   types. The tooling is robust and scalable to the entire R language.
-\item We carried out a large-scale analyis of corpus of \PACKAGES widely
+\item We carried out a large-scale analysis of corpus of \PACKAGES widely
   used and actively maintained libraries to extract function type signatures
@@
   programs that use those functions.
-\item We report on the appropriateness and usefulness of a simpe type
+\item We report on the appropriateness and usefulness of a simple type
   language for the R programming language.
@@
 system can be prototyped and evaluated by applying the type rule to
-execution traces of programs. While the appraoch has the limitation of
+execution traces of programs. While the approach has the limitation of
 dynamic analysis techniques, namely that the results are only as good at the
@@
 There is no previous work for the R programming language. We take
-inspiration in the above mentionned works but focus on adapting them to our
+inspiration in the above mentioned works but focus on adapting them to our
 target language.
@@
 Scalar data types include the distinguished \k{NULL} value, which is also of
-ktype \k{NULL}, instance of classes written using the S4 object system,
+type \k{NULL}, instance of classes written using the S4 object system,
 closures and environments.  The implementation of R has a number of other
@@
 stay close to the R language and only depart in small, hopefully,
-non-controverwsial ways.  Functions types have the form \FUN{A_1, \dots,
+non-controversial ways.  Functions types have the form \FUN{A_1, \dots,
   A_n}\T where each $\A_i$ argument is either a type \T or \VARGS, a
@@
 ensue). We support parametrization of the list type, \LIST\T, but for
-heterogenous lists this type will default to \any.
+heterogeneous lists this type will default to \any.
 
@@
 To obtain raw types, we make use of R's C FFI and use low-level machinery to
-collect type tags and attributes from the R runtime.  The types that we
+collect type tags and attributes from the R run-time.  The types that we
 provide to users are constructed during post-processing, and rely on the
@@
 definition across multiple programs and distills them into a single type
-signature.  The shape of the infered function signatures is:
+signature.  The shape of the inferred function signatures is:
 
@@
 While \contractr's primary logic has been implemented in C++ to reduce its
-runtime overhead, we have not observed a single segfault during our use of
+runtier overhead, we have not observed a single segfault during our use of
 the package.  It has been tested with GNU R-3.5.0 and hardened with a
@@
 
-In terms of useability, our checker is enabled automatically whenever a
+In terms of usability, our checker is enabled automatically whenever a
 package is loaded in R. So, a simple invocation of \code{library(contractr)}
@@
 \paragraph{Checking the values} A number of properties can be checked
-by a simple tag check, namely wether a value is a: null, environment, vector
+by a simple tag check, namely whether a value is a: null, environment, vector
 or scalar. Other properties require an inspection of the contents of a
@@
 
-These packages come from the Comprehensive R Achive Network
+These packages come from the Comprehensive R Archive Network
 (CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of R
@@
 our proposed type system is with respect to common programming patterns
-occuring in widely used R libraries.
+occurring in widely used R libraries.
 
@@
 Monomorphic in this context means that the type is not relying on \any or
-inlcuding a union.  The import of monorphism in this context is that it
+including a union.  The import of monorphism in this context is that it
 means our type language can accurately capture an argument's type or a

### 2020-05-15 paper/main.tex
Add some numbers

@@
 
-<<Insert proportion of monorphic types in argument and in function
-signature. Perhaps a table of monomorphic type signatures that are
-popular...>>
-
 \begin{figure}[!h]  \centering
@@
 
-We observe that XXX of the positions have a single type ascribed to them,
-and, only XXX have four or more types ascribed to them. Parameter PARAMNAME
-of function XXXXCOLONCOLONXXX has the maximum, XXX, types ascribed to it.
-Hence, most functions admit or return a uni-typed value.
+There are \PercUnitypedPositions monomorphic parameter and return value
+positions in function signatures. Only \PercManytypedPositions positions have
+four or more types ascribed to them. Parameter PARAMNAME of function
+XXXXCOLONCOLONXXX has the maximum, XXX, types ascribed to it. Most functions
+admit or return a uni-typed value.
 
@@
 \input{tables/uni-typed-breakdown.tex}
+
+We observe that XXX of the positions have a single type ascribed to them, and,
+only XXX have four or more types ascribed to them.
 

### 2020-05-15 paper/main.tex
corpus table

--- 2020-05-15 paper/main.tex
@@
 
-\newcommand{\NUMFUNCTIONS}{25,000\xspace}  %%% TODO auto-gen
+\newcommand{\NUMFUNCTIONS}{20,214\xspace}  %%% TODO auto-gen
 \newcommand{\NUMPACKAGES}{412\xspace}  %%% TODO auto-gen
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+\subsection{Overview of Types for R}
+
+We ran \typetracer on the \CorpusLoadable packages we discussed in this section, which yields types for \NUMFUNCTIONS functions.
+To build up a picture of what types for R functions might look like, consult Table~\ref{tbl:corpus-type-breakdown}, which lists the 10 most popular argument signatures.
+The first row of the table reads: 13255 arguments have type \dbl, represnting 11.06\% of type signatures, and \typetracer observed 41M values of type \dbl flowing into or out of functions while running on the core package corpus, which represents only 1.59\% of dynamically observed values.
+Of interest are rows 2 and 4, where we see that collectively \lgl and \chr make up only 16.11\% of argument types, while making up nearly 60\% of dynamically observed values.
+
+\begin{table}
+% BEGIN Autogenerated
+
+\begin{tabular}{lrrrr}
+\toprule
+Type & Args with Type & \% of Static Types & Dynamic Observations & \% of Dynamic Types\\
+\midrule
+double & 13,255 & 11.06 & 41,560,797 & 1.59\\
+logical & 10,142 & 8.46 & 712,895,959 & 27.21\\
+null & 9,175 & 7.66 & 122,202,183 & 4.66\\
+character & 9,171 & 7.65 & 794,058,588 & 30.31\\
+double[] & 7,811 & 6.52 & 29,272,232 & 1.12\\
+\addlinespace
+... & 7,759 & 6.48 & 76,369,333 & 2.92\\
+any & 6,243 & 5.21 & 255,539,497 & 9.75\\
+character[] & 4,576 & 3.82 & 88,233,523 & 3.37\\
+class<`matrix`> & 4,522 & 3.77 & 32,452,555 & 1.24\\
+class<`data.frame`> & 2,781 & 2.32 & 7,749,094 & 0.30\\
+\bottomrule
+\end{tabular}
+
+% END Autogenerated
+\label{tbl:corpus-type-breakdown}
+\caption{Top types of contract assertion failures.}
+\end{table}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2020-05-15 evaluation/R/evaluate-type-analysis.Rmd

@@
   table_dirpath: ../../paper/tables/
+  macro_dirpath: ../../paper/macros/
 ---
@@
 suppressPackageStartupMessages(library(kableExtra))
+
+macro_filename <- file.path(params$macro_dirpath, "macros.tex")
 
@@
 
+generate_macro("PercUnitypedPositions", 
+               union_frequency_folded %>% filter(alternative_count == "1") %>% pull(percentage),
+               macro_filename)
+
+generate_macro("PercManytypedPositions", 
+               union_frequency_folded %>% filter(alternative_count == "> 3") %>% pull(percentage),
+               macro_filename)
+
 union_types <-
@@
 group_by(alternative_count) %>%
-mutate(percentage = round(100 * count / sum(count), 2),
-       cumsum_percentage = round(100 * cumsum(count) / sum(count), 2)) %>%
+mutate(percentage = round(100 * count / sum(count), 1),
+       cumsum_percentage = round(100 * cumsum(count) / sum(count), 1)) %>%
 ungroup() %>%

### 2020-05-15 paper/main.tex
Include macros file
6.2 edits

@@
 in Section~\ref{sec:corpus}, we ran \typetracer on its test, example, and
-vignette code, obtaining types for package functions, and then subsequently
+vignette code, obtaining types for package functions. Then, we
 ran the test, example, and vignette code of each of the package's reverse
@@
 were never violated.  The number of immaculate function types increases to
-\PERCSUCCFUNNOSTHREE if we count out S3 functions: These S3 functions
+\PERCSUCCFUNNOSTHREE if we discount S3 functions: These S3 functions
 represent user-defined polymorphism which we do not tackle in this work.
@@
 We breakdown the failed assertions by type in
-Table~\ref{tbl:fail-breakdown-0}.  By far the most commonly failed assertion
-is one asserting that a value with ${\bf dbl}[]$ type has type ${\bf
-  class}\langle matrix \rangle$.  This was a deliberate design decision, as
+Table~\ref{tbl:fail-breakdown-0}.  
+Accounting for 27.37\% of assertion failures, the most common vilation arises when a value with \CLASS{matrix} is passed where a \VEC\dbl is expected.
+Considering these types, we might imagine them to be compatible, however not allowing this coercion was a deliberate design decision:
 coercion from vectors to matrices is ad hoc at best, and surely not a
@@
 had that type, and how many of those exhibited the contract failure in
-question.  Table~\ref{tbl:fail-breakdown-1} breaks down the results from
-Table~\ref{tbl:fail-breakdown-0} by unique parameter position, folding away
-multiple identical failed contract assertions for the same parameter
+question.  Table~\ref{tbl:fail-breakdown-1} breaks down the failed assertions by type,
+folding away multiple identical failed contract assertions for the same parameter
 position.  The first row of this table reads: a value of type
@@
 type of the value that violated that contract.  The first row of the table
-reads: 44 function arguments with integer vector type are passed double
-vectors instead, and 657 arguments have integer vector type.  Had we failed
+reads: 44 function arguments with \VEC\int type are passed \VEC\dbl  instead, and 657 arguments have \VEC\int type.  Had we failed
 to capture some key usage pattern of R with our type annotation framework,

### 2020-05-15 evaluation/R/evaluate-type-analysis.Rmd

@@
 ```
+
+##
+
+```{r, eval = TRUE, verbose = FALSE}
+union_frequency <-
+union_data %>%
+distinct(package_name, function_name, parameter_position) %>%
+mutate(polymorphic = ifelse(alternative_count > 1, true, is_polymorphic(alternative_count)) %>%
+group_by(package_name, function_name, polymorphic_position_count = )
+group_by(alternative_count) %>%
+summarize(parameter_count = n()) %>%
+ungroup() %>%
+arrange(desc(parameter_count)) %>%
+mutate(percentage = round(parameter_count / sum(parameter_count), 4),
+       cumulative_percentage = round(cumsum(parameter_count) / sum(parameter_count), 4))
+```

### 2020-05-15 paper/main.tex
Add 7.3

@@
 \input{corpus.tex}
+\input{macros/macros.tex}
+\input{asserts.tex}
 
@@
 
+There is a number of ways in R to check types of function parameters. The
+default and the most common way\footnote{\CranStopifnotRatio of all runtime
+  checks in the whole of CRAN} is to use \code{stopifnot} function from the R
+base package. It takes a number of R expressions which all should evaluate to
+true otherwise a runtime exception is thrown with a message quoting the
+corresponding failed expression. For example, the following code checks whether
+a given parameter \code{x} is a scalar string:
+%
+\begin{lstlisting}
+stopifnot(is.character(x), length(x) == 1L, !is.na(x))
+\end{lstlisting}
+
+Next to \code{stopifnot}, there are 4 packages in CRAN\footnote{Packages are
+  available on CRAN website: \url{https://cran.r-project.org/web/packages/}}.
+that focus on runtime assertions: \emph{assertive}, \emph{ensurer},
+\emph{assertr} and \emph{assertthat} \emph{assertive} and \emph{ensurer} have
+not been updated 2016 and 2015 respectively. \emph{assertr} is used by only
+\AssertrRevdeps other packages and in the current version focuses on checking
+properties of data frames. Only the \emph{assertthat} is maintained and used
+(with \AssertthatRevdeps reverse dependencies). The advantage of
+\emph{assertthat} over the R's default is that it provides a much better error
+message.
+
+One way to asses the usefulness of our type checking system is to find out how
+many of the existing type checking constrains could be replaced by \contractr.
+To measure this, we have extracted all calls to \code{stopifnot} and
+\code{assertthat} assertions and checked which one of them could be either
+completely replaced by \contractr or at least partially simplified by removing a
+portion of an assertion expression. This is useful, because a common pattern is
+that the first part of the parameter assertion checks its type while the rest
+checks its value. In the example above, the whole expression could be replaced
+by \code{character} type.
+
+Out of the \CorpusLoadable packages, \CorpusAssertsInPackages use runtime
+assertions. Together there are \CorpusAsserts asserts in
+\CorpusAssertsInFunctions functions. Among these, \contractr can replace
+\CorpusTypedAsserts (\CorpusTypedAssertsRatio) assertion calls across
+\CorpusTypedAssertsPackages packages and \CorpusTypedAssertsFunctions functions.
+Furthermore, additional \CorpusPartiallyTypedAsserts
+(\CorpusPartiallyTypedAssertsRatio) asserts in packages
+\CorpusPartiallyTypedAssertsPackages packages and
+\CorpusPartiallyTypedAssertsFunctions functions could have been simplified.
+
+Checking the type of function parameters is not something that is seen often in
+the R code. In the whole of CRAN, there are only \CranAssertsRnd asserts in
+\CranAssertsInFunctionsRnd functions define in \CranAssertsInPackagesRnd
+packages. One might speculate that one of the reasons is that the current way is
+not convenient as it is rather verbose. Our system can infer type annotations
+for existing functions automatically. This can remove or partially remove over
+\CorpusPartiallyTypedAssertsRatio of existing assertions.
+
 \section{Conclusion}

### 2020-05-15 paper/main.tex
put correct type lang in 7.2

--- 2020-05-15 paper/main.tex
@@
 \input{corpus.tex}
+\input{macros/macros.tex}
 
@@
 \midrule
-double & 13,255 & 11.06 & 41,560,797 & 1.59\\
-logical & 10,142 & 8.46 & 712,895,959 & 27.21\\
-null & 9,175 & 7.66 & 122,202,183 & 4.66\\
-character & 9,171 & 7.65 & 794,058,588 & 30.31\\
-double[] & 7,811 & 6.52 & 29,272,232 & 1.12\\
+\dbl & 13,255 & 11.06 & 41,560,797 & 1.59\\
+\lgl & 10,142 & 8.46 & 712,895,959 & 27.21\\
+\null & 9,175 & 7.66 & 122,202,183 & 4.66\\
+\chr & 9,171 & 7.65 & 794,058,588 & 30.31\\
+\VEC\dbl & 7,811 & 6.52 & 29,272,232 & 1.12\\
 \addlinespace
-... & 7,759 & 6.48 & 76,369,333 & 2.92\\
-any & 6,243 & 5.21 & 255,539,497 & 9.75\\
-character[] & 4,576 & 3.82 & 88,233,523 & 3.37\\
-class<`matrix`> & 4,522 & 3.77 & 32,452,555 & 1.24\\
-class<`data.frame`> & 2,781 & 2.32 & 7,749,094 & 0.30\\
+\VARGS & 7,759 & 6.48 & 76,369,333 & 2.92\\
+\any & 6,243 & 5.21 & 255,539,497 & 9.75\\
+\VEC\chr & 4,576 & 3.82 & 88,233,523 & 3.37\\
+\CLASS{matrix} & 4,522 & 3.77 & 32,452,555 & 1.24\\
+\CLASS{data.frame} & 2,781 & 2.32 & 7,749,094 & 0.30\\
 \bottomrule
@@
 \label{tbl:corpus-type-breakdown}
-\caption{Top types of contract assertion failures.}
+\caption{Top types of arguments in R.}
 \end{table}
@@
 \midrule
-double[] & class<`matrix`> & 371896 & 27.37 & 27.37\\
-double[] & integer[] & 117212 & 8.62 & 35.99\\
-character & class<`bignum`> | raw[] & 100100 & 7.37 & 43.36\\
-double & integer | null & 60578 & 4.46 & 47.81\\
-double[] & class<`timeSeries`> & 58872 & 4.33 & 52.15\\
+\VEC\dbl & \CLASS{matrix} & 371,896 & 27.37 & 27.37\\
+\VEC\dbl & \VEC\int & 117,212 & 8.62 & 35.99\\
+\chr & \CLASS{bignum} | \VEC\raw & 100,100 & 7.37 & 43.36\\
+\dbl & \int | \null & 60,578 & 4.46 & 47.81\\
+\VEC\dbl & \CLASS{timeSeries} & 58,872 & 4.33 & 52.15\\
 \addlinespace
-class<matrix> & class<`timeSeries`> & 55005 & 4.05 & 56.19\\
-double[] & double & 53338 & 3.92 & 60.12\\
-double & class<`data.frame`> & 32553 & 2.40 & 62.51\\
-double[] & class<`data.frame`> & 31576 & 2.32 & 64.84\\
-double & integer & 20586 & 1.51 & 66.35\\
+\CLASS{matrix} & \CLASS{timeSeries} & 55,005 & 4.05 & 56.19\\
+\VEC\dbl & \dbl & 53,338 & 3.92 & 60.12\\
+\dbl & \CLASS{data.frame} & 32,553 & 2.40 & 62.51\\
+\VEC\dbl & \CLASS{data.frame} & 31,576 & 2.32 & 64.84\\
+\dbl & \int & 20,586 & 1.51 & 66.35\\
 \bottomrule
@@
 position.  The first row of this table reads: a value of type
-\code{double[]} was passed to 15 different function parameters expecting a
-matrix, of which there are 1394 in total: 1.08\% of matrix-typed parameters
-were passed double vector values.
+\VEC\dbl was passed to 15 different function parameters expecting a
+\CLASS{matrix}, of which there are 1394 in total: 1.08\% of \CLASS{matrix}-typed parameters
+were passed \VEC\dbl values.
 
@@
 is similar with the double and integer coercion we mentioned earlier: it
-represents many dynamic contract failures, but very few of the integer
-vector-typed arguments have their contracts violated by double vector-typed
+represents many dynamic contract failures, but very few of the \VEC\int-typed 
+arguments have their contracts violated by \VEC\dbl-typed
 values.  Rows 5 and 6 are interesting: we see that rather often arguments
-expecting time series data are passed double vector or matrix values.  This
+expecting \CLASS{timeSeries} data are passed \VEC\dbl or \CLASS{matrix} values.  This
 is a quirk of the \code{timeSeries} package, whose functions often accept
@@
 \midrule
-double[] & class<`matrix`> & 15 & 1394 & 1.08\\
-double[] & integer[] & 44 & 657 & 6.70\\
-character & class<`bignum`> | raw[] & 1 & 3 & 33.33\\
-double & integer | null & 5 & 20 & 25.00\\
-double[] & class<`timeSeries`> & 19 & 35 & 54.29\\
+\VEC\dbl & \CLASS{matrix} & 15 & 1394 & 1.08\\
+\VEC\dbl & \VEC\int & 44 & 657 & 6.70\\
+\chr & \CLASS{bignum} | \VEC\raw & 1 & 3 & 33.33\\
+\dbl & \int | \null & 5 & 20 & 25.00\\
+\VEC\dbl & \CLASS{timeSeries} &19 & 35 & 54.29\\
 \addlinespace
-class<matrix> & class<`timeSeries`> & 19 & 35 & 54.29\\
-double[] & double & 100 & 5021 & 1.99\\
-double & class<`data.frame`> & 1 & 900 & 0.11\\
-double[] & class<`data.frame`> & 6 & 900 & 0.67\\
-double & integer & 48 & 529 & 9.07\\
+\CLASS{matrix} & \CLASS{timeSeries} &  19 & 35 & 54.29\\
+\VEC\dbl & \dbl & 100 & 5021 & 1.99\\
+\dbl & \CLASS{data.frame} &  1 & 900 & 0.11\\
+\VEC\dbl & \CLASS{data.frame} &  6 & 900 & 0.67\\
+\dbl & \int & 48 & 529 & 9.07\\
 \bottomrule
@@
 Table~\ref{tbl:fail-breakdown-2} with a percentile of 80th percentile
-instead.  The most frequent argument type violation pattern in that of data
-frames being passed to arguments expecting tibbles: this occurs in only 22\%
+instead.  The most frequent argument type violation pattern in that of \CLASS{data.frame} values
+ passed to arguments expecting tibbles, which are an extension of \CLASS{data.frame} built in the \code{tidyverse} package suite: 
+ this occurs in only 22\%
 of such arguments, and represents cases where tests did not adequately cover
@@
 \midrule
-double[] & integer[] & 44 & 657 & 6.70\\
-\textasciicircum{}logical[] & null & 158 & 2756 & 5.73\\
-character[] & character & 183 & 4102 & 4.46\\
-character & null & 117 & 2756 & 4.25\\
-double & null & 107 & 2756 & 3.88\\
+\VEC\dbl & \VEC\int & 44 & 657 & 6.70\\
+\NAVEC\lgl & \null & 158 & 2756 & 5.73\\
+\VEC\chr & \chr & 183 & 4102 & 4.46\\
+\chr & \null & 117 & 2756 & 4.25\\
+\dbl & \null & 107 & 2756 & 3.88\\
 \bottomrule
@@
 \midrule
-class<data.frame> & class<`data.frame`, `tbl`, `tbl\_df`> & 63 & 284 & 22.18\\
-class<matrix> & class<`array`> & 13 & 70 & 18.57\\
-\textasciicircum{}logical[] & logical | null & 9 & 49 & 18.37\\
-character[] & list<character[]> & 20 & 109 & 18.35\\
-complex[] & complex & 7 & 42 & 16.67\\
+\CLASS{data.frame} & \CLASS{data.frame, tbl, tbl\_df} & 63 & 284 & 22.18\\
+\CLASS{matrix} & \CLASS{array} & 13 & 70 & 18.57\\
+\NAVEC\lgl & \lgl | \null & 9 & 49 & 18.37\\
+\VEC\chr & \LIST{\VEC\chr} & 20 & 109 & 18.35\\
+\VEC\clx & \clx & 7 & 42 & 16.67\\
 \bottomrule

### 2020-05-15 evaluation/R/evaluate-type-analysis.Rmd

--- 2020-05-15 evaluation/R/evaluate-type-analysis.Rmd
@@
 ```
+
+##
+
+```{r, eval = TRUE, verbose = FALSE}
+function_polymorphic_freq <-
+union_data %>%
+distinct(package_name, function_name, parameter_position) %>%
+mutate(polymorphic = ifelse(alternative_count > 1, true, is_polymorphic(alternative_count)) %>%
+group_by(package_name, function_name) %>%
+summarize(polymorphic_position_count = sum(polymorphic)) %>%
+ungroup() %>%
+group_by(polymorphic_position_count) %>%
+summarize(function_count = ()) %>%
+ungroup() %>%
+arrange(polymorphic_position_count)
+
+function_polymorphic_freq %>%
+ggplot(aes(polymorphic_position_count, function_count)) +
+
+```

### 2020-05-15 paper/main.tex

@@
 \input{macros/macros.tex}
+\input{asserts.tex}
 

### 2020-05-15 paper/main.tex
6.1.1 WIP

@@
 \newcommand{\TOTNUMSIGS}{21,730\xspace}
+\newcommand{\PROPSTRUCTSDYN}{44.46\%\xspace}
 
@@
 
+Now, we aim to enhance the quantitative analysis of our language design by looking at select code snippets from the corpus we analyzed.
+We will highlight some of the design decisions that we made, give code examples, and accompany them with function signatures generated by \typetracer.
+
 \paragraph{NAs}
@@
 \begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
-  binom.profile <- function(x, n, ...) {
+#' @type <x: dbl[], n: dbl[], conf.level: dbl, maxsteps: dbl,
+#'         del: dbl, bayes: lgl, plot: lgl, ...> => class<data.frame>
+binom.profile <- function(x, n, conf.level = 0.95, maxsteps = 50,
+                            del = zmax / 5, bayes = TRUE, plot = FALSE, ...) {
    xn <- cbind(x = x, n = n)
@@
 then extracts non-\NA values and stores them into \code{x} and \code{n}
-respectively.
+respectively.We see from the type that arguments \code{x} and \code{n} are intended to be free of \NA values.
+\contractr will check them for compliance, and warn programmers of function misuse.
 
@@
 \begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
+#' @type <n: dbl, x: int[]> => class<matrix>
 hankel.matrix <- function( n, x ) { 
@@
 comparison results, and the conditional of an if must be a scalar else R
-generates a warning.
+generates a warning. The type for this function indicates that \code{x} must be
+a vector, though here the type for \code{n} is found to be \dbl---this is likely due
+to numeric values defaulting to \dbl type in R, as e.g. \code{typeof(2)} is \dbl.
 
@@
 \begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
+#' @type <x: class<matrix>, w: ? ^dbl[], rows: null, cols: null, na.rm: lgl, ...> => ^dbl[]
 rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL,
@@
 type checking arguments, rightly producing a message to the user indicating
-that an explicit matrix should be passed.
-
-To capture cases like this, we include a \code{class<matrix>} type in our
-type annotation language.  If \code{rowWeightedMeans} were outfitted with
-such a type, \contractr can perform the check present in this code
-automatically, notifying the users passing ill-typed arguments.
+that an explicit matrix should be passed. We captured this with our class types, 
+and indeed \typetracer inferred a good type for this function, allowing \contractr to 
+perform the check present in this code automatically, notifying the users passing ill-typed arguments.
 
@@
 argument validation code, but our intention is not to subsume {\it all} such
-code.
+code. Consider:
 
@@
 rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL, na.rm = FALSE) {
+    ...
     n <- ncol(x)
@@
 
-We observed the use of consistently-named lists in \AT{a number of
-  functions}.  In R, ``fields'' of named lists can be accessed using the
+\typetracer observed that \PROPSTRUCTSDYN of lists had a \code{names} attribute, and in R, ``fields'' of named lists can be accessed using the
 \code{$} operator.

### 2020-05-15 evaluation/R/evaluate-type-analysis.Rmd
Create graph for function polymorphism

--- 2020-05-15 evaluation/R/evaluate-type-analysis.Rmd
@@
 ```
+
+```{r, eval = TRUE, verbose = FALSE}
+function_polymorphic_freq <-
+union_data %>%
+distinct(package_name, function_name, parameter_position, alternative_count) %>%
+mutate(polymorphic = ifelse(alternative_count > 1, TRUE, is_polymorphic(alternative_count))) %>%
+group_by(package_name, function_name) %>%
+summarize(polymorphic_position_count = as.integer(sum(polymorphic))) %>%
+ungroup() %>%
+group_by(polymorphic_position_count) %>%
+summarize(function_count = n()) %>%
+ungroup() %>%
+arrange(polymorphic_position_count)
+
+function_polymorphic_freq_more_than_three <-
+function_polymorphic_freq %>%
+filter(polymorphic_position_count > 3) %>%
+summarize(polymorphic_position_count = "> 3", function_count = sum(function_count))
+
+function_polymorphic_freq <-
+function_polymorphic_freq %>%
+filter(polymorphic_position_count <= 3) %>%
+rbind(function_polymorphic_freq_more_than_three) %>%
+mutate(function_percentage = function_count / total_function_count)
+
+(function_polymorphic_freq %>%
+ggplot(aes(polymorphic_position_count, function_percentage)) +
+geom_col() +
+scale_y_continuous(sec.axis = sec_axis(~ . * total_function_count, labels = count_labels), labels = relative_labels) +
+scale_x_discrete(limits=c("> 3", 3, 2, 1, 0)) +
+labs(x="Polymorphic Positions", y="Functions") +
+coord_flip() +
+theme(legend.title = element_blank(), legend.position = "top")) %>%
+save_graph("function_polymorphism.pdf")
+
+generate_macro("CountMonomorphicFunctions", 
+               function_polymorphic_freq %>% filter(polymorphic_position_count == "0") %>% pull(function_count),
+               macro_filename)
+
+generate_macro("PercMonomorphicFunctions", 
+               function_polymorphic_freq %>% filter(polymorphic_position_count == "0") %>% pull(function_percentage),
+               macro_filename)
+
+generate_macro("CountPolymorphicFunctions", 
+               total_function_count - function_polymorphic_freq %>% filter(polymorphic_position_count == "0") %>% pull(function_count),
+               macro_filename)
+
+generate_macro("PercPolymorphicFunctions", 
+               1.0 - function_polymorphic_freq %>% filter(polymorphic_position_count == "0") %>% pull(function_percentage),
+               macro_filename)
+```

### 2020-05-15 paper/main.tex
Add graph

@@
 
+\begin{figure}[!h]  \centering
+  \includegraphics[width=\linewidth]{plots/function_polymorphism.pdf}
+  \caption{Function Polymorphism}\label{fig:funpoly}
+\end{figure}
+
+Figure shows the distribution of functions against the number of polymorphic
+positions. We observe that \PercMonomorphicFunctions
+(\CountMonomorphicFunctions) functions are monomorphic. The remaining
+\PercPolymorphicFunctions (\CountPolymorphicFunctions) have at least one
+polymorphic parameter or return type. Polymorphic functions are spread
+across many packages, \CountMonomorphicPackages out of the 412 packages from our
+corpus export only monomorphic functions.
+
 Figure 1 shows the number of alternatives in the inferred type of function

### 2020-05-16 paper/main.tex
edits

--- 2020-05-15 paper/main.tex
@@
 
-\typetracer observed that \PROPSTRUCTSDYN of lists had a \code{names} attribute, and in R, ``fields'' of named lists can be accessed using the
+In R, lists can have a names attribute, allowing lists elements to be accessed by name using the
 \code{$} operator.
@@
 \begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
+#' @type <x: class<aov, lm> | class<lm>> => dbl;
 agricolae::cv.model <- function(x) {
@@
 In the code snippet, the function \code{cv.model} takes an argument
-\code{x}, which we observed to always be a {\it linear model}, which is part
+\code{x}, which we observed to always be a {\it linear model} (either ANOVA with \CLASS{aov, lm} or a simple linear regression model with \CLASS{lm}) which is part
 of R's built-in statistics functionality.  Internally, linear models are
@@
 elements using the \code{$} syntax, as seen on lines 2-4 in the snippet.  To
-describe these kinds of values, we included a struct type in our type
+describe these kinds of values, we intended to include a struct type in our type
 system, allowing users to specify name, type pairs as part of list type
-definitions.
-
-Unfortunately, the inclusion of structs complicated our analysis by
+definitions, though we found this to be problematic.
+
+The inclusion of structs complicated our analysis by
 polluting the data, illustrated on line 8 with the call
@@
 structs we saw a marked increase in misrepresentative types, as the types of
-functions which use these data sets (such as \code{aov} in
-Figure~\ref{fig:struct-ex}) became polluted with the names from the example
+functions which use these data sets (such as \code{aov} in the code example) became polluted with the names from the example
 sets.  We implemented a unification strategy for struct types, removing them
@@
 
-We still believe struct types to be useful to R programmers, and so our
-language supports the type and \contractr will check them, though we
-configured \typetracer to ignore them as they made the data substantially
-less useful.  On the positive side, we found that \AT{get the number, but
-  often} structs-typed values had a {\it class}, which helped motivate our
-inclusion of class-like types in our type annotation language.
+We found that struct-like data was best described with classes, as illustrated in the code snippet, and many structs actually had a class.
+Even though \code{x} is used like a struct, it in fact has a linear model class, which we believe is a more informative type.
 

### 2020-05-16 evaluation/R/evaluate-type-analysis.Rmd

@@
 ```{r, eval = TRUE, verbose = FALSE}
-function_polymorphic_freq <-
+function_polymorphism <-
 union_data %>%
@@
 summarize(polymorphic_position_count = as.integer(sum(polymorphic))) %>%
-ungroup() %>%
+ungroup()
+
+package_polymorphism <-
+function_polymorphism %>%
+group_by(package_name) %>%
+summarize(monomorphic = all(polymorphic_position_count == 0)) %>%
+ungroup()
+
+function_polymorphic_freq <-
+function_polymorphism %>%
 group_by(polymorphic_position_count) %>%
@@
                macro_filename)
+
+generate_macro("CountMonomorphicPackages", 
+               package_polymorphism %>% pull(monomorphic) %>% sum(),
+               macro_filename)
 ```

### 2020-05-16 paper/main.tex
Fix ordering

@@
 
-There are \PercUnitypedPositions monomorphic parameter and return value
-positions in function signatures. Only \PercManytypedPositions positions have
-four or more types ascribed to them. Parameter PARAMNAME of function
-XXXXCOLONCOLONXXX has the maximum, XXX, types ascribed to it. Most functions
-admit or return a uni-typed value.
-
-Figure 1 provides a breakdown of the different type categories observed at
-uni-typed positions.
+\ref{fig:unionfreq} shows the number of types inferred at each position of a
+function signature. Most functions admit or return a uni-typed value. There are
+\PercUnitypedPositions uni-typed parameter and return value positions in
+function signatures. Only \PercManytypedPositions positions have more than three
+types ascribed to them.
 
 \input{tables/uni-typed-breakdown.tex}
-
-We observe that XXX of the positions have a single type ascribed to them, and,
-only XXX have four or more types ascribed to them.
-
-Figure XXX shows ten most common type categories that occur by
-themselves. We observe that three scalar types, double, logical, and
-character are most frequent. This is followed by XXX vararg parameter
-positions. Next, XX\% of the parameter types are any because arguments at
-those positions were either missing or unused in at least one call to those
-functions. Together, these types amount to more than 50\%. This is followed
-by double vector, the unit type, null, character vector, and classes matrix
-and data.frame respectively.
+\ref{fig:singlefreq} provides a breakdown of the different type categories
+observed at uni-typed positions. \scalar, \class and \vector are
+the most common type categories. The shaded rows correspond to polymorphic
+types. Removing those gives us XXX (XXX) monomorphic positions in a corpus of
+XXX parameter and return positions.
 
@@
 
-Figure shows the distribution of functions against the number of polymorphic
-positions. We observe that \PercMonomorphicFunctions
+\ref{fig:funpoly} shows the distribution of functions against the number of
+polymorphic positions. We observe that \PercMonomorphicFunctions
 (\CountMonomorphicFunctions) functions are monomorphic. The remaining
 \PercPolymorphicFunctions (\CountPolymorphicFunctions) have at least one
-polymorphic parameter or return type. Polymorphic functions are spread
-across many packages, \CountMonomorphicPackages out of the 412 packages from our
-corpus export only monomorphic functions.
+polymorphic parameter or return type. Polymorphic functions are spread across
+many packages, \CountMonomorphicPackages out of the 412 packages from our corpus
+export only monomorphic functions.
 

### 2020-05-16 paper/main.tex
6.1.1

--- 2020-05-16 paper/main.tex
@@
 
-One of the most popular classes in R is the \code{data.frame} class.  Data
+One of the most popular classes in R is the \code{data.frame} class, making up nearly 8\% of observed classes.  Data
 frames and the derivative \code{tibble} (from the \code{tidyverse}
@@
 
-In the context of our type annotation language, we will ascribe the type \code{class<data.frame>} to \code{data.frame}-typed values.
-A more in-depth discussion of class types follows.
+In the context of our type annotation language, we ascribe the type \CLASS{data.frame} to \code{data.frame}-typed values.
+Consider the following code snippet, showing the type signature for \code{dplyr}'s \code{all_equal} function which checks two \CLASS{data.frame} for equality.
+\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
+#' @type <target: class<data.frame>, current: class<data.frame>, ignore_col_order: lgl, 
+#'         ignore_row_order: lgl, convert: lgl, ...> => (chr | lgl)
+dplyr::all_equal <- function(target, current, ignore_col_order = TRUE,
+                      ignore_row_order = TRUE, convert = FALSE, ...) { ... }
+\end{lstlisting}
 
@@
 support for S3, S4, and R5.  Thus far, we've mainly discussed S3 and its
-dispatch mechanism, and indeed our class types roughly capture the S3
-classes of values.  We will not discuss R5 as it is still under development.
-
-For S4, we have an \code{s4} type to denote S4 values, though we did not
-elaborate on this type, and in practice S4 values have a class, and thus a
-\code{class<...>} type.  We stuck to simple class-based types for S4 since
+dispatch mechanism, and indeed our class types capture the S3
+classes of values, but not user-defined polymorphism.  We will not discuss R5 as it is still under development.
+
+S4, like S3, uses the class attribute to store types, we type them with a \CLASS{...} type.  
+We stuck to simple class-based types for S4 since
 doing anything more would be immensely complicated, and outside of the scope
@@
 can define their own class hierarchies that we would need to incorporate in
-our type analysis and contract checking frameworks.  Further, \AT{we found
-  limited use of S4 during our analysis}, with S4 types being heavily used
-in a select subset of packages (\AT{such as ...}), some of which make heavy
-use of R's metaprogramming capabilities.  Coming up with a type system that
+our type analysis and contract checking frameworks.  Further,we found
+  limited use of S4 during our analysis. Coming up with a type system that
 accounts for all of these factors and consolidates multiple
@@
 problem in and of itself, and we aim to extend our type annotation framework
-to account for this in future work.  For now, S3 types suffice: S3 is the
-most common object system in R, with no sophisticated hierarchies, and its
-semantics are simple enough to be captured by our proposed design.
+to account for this in future work.  For now, simple class types for S3 and S4 suffice.
+
+To illustrate informative class types, consider the following code snippet.
+\code{distr} is a package which implements statistical distributions in an object-oriented way, and the \code{Binom} function acts as a constructor for \CLASS{Binom} values.
+Another example is from the \code{Epi} package, which provides functionality for demographic and epidemiological analysis in the Lexis diagram.
+The function we highlighted, \code{timeBand}, extracts time band data from the passed Lexis object.
+We see from the type that the first argument is the Lexis object, and that either a \VEC\dbl or a \CLASS{factor} is returned.
+Often, class members are implemented as attributes on R values, and it's likely that this is the case for Lexis objects as evidenced by the programmer accessing the "breaks" attribute in the function code.
+
+\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
+#' @type <size: dbl, prob: dbl> => class<Binom>
+distr::Binom <- function(size = 1,prob = 0.5) new("Binom", size = size, prob = prob)
+
+#' @type <lex: class<`Lexis`, `data.frame`>, time.scale: chr, type: chr> => 
+#'           class<`factor`> | dbl[];
+Epi::timeBand <- function(lex, time.scale, type="integer") {
+  time.scale <- check.time.scale(lex, time.scale)[1]
+  breaks <- attr(lex, "breaks")[[time.scale]]
+  ...
+\end{lstlisting}
 

### 2020-05-16 paper/main.tex
Add macros
fix merge

@@
 \item We carried out a large-scale analysis of corpus of \PACKAGES widely
-  used and actively maintained libraries to extract function type signatures
+  used and actively maintained libraries to extract function type signatures and
   validated the robustness of the inferred type signatures against XXX

### 2020-05-16 evaluation/R/evaluate-type-analysis.Rmd
Add macros

@@
 
+monomorphic_parameter_count <-
+unityped_positions %>%
+mutate(polymorphic = map_lgl(kind, is_polymorphic)) %>%
+filter(!polymorphic) %>%
+pull(count) %>%
+sum()
+
+generate_macro("CountMonomorphicPositions", monomorphic_parameter_count, macro_filename)
+
+generate_macro("PercMonomorphicPositions", monomorphic_parameter_count / total_parameter_count, macro_filename)
+
+generate_macro("CountPositions", total_parameter_count, macro_filename)
+
 single_union_types <-

### 2020-05-16 paper/main.tex
Add macros
corpus

@@
 \renewcommand{\null}{\texttt{\textbf{null}}\xspace}
+\newcommand{\scalar}{\texttt{\textbf{scalar}}\xspace}
+\renewcommand{\vector}{\texttt{\textbf{vector}}\xspace}
+\newcommand{\class}{\texttt{\textbf{class}}\xspace}
 \newcommand{\any}{\texttt{\textbf{any}}\xspace}
-\newcommand{\environment}{\texttt{\textbf{envir}}\xspace}
+\newcommand{\environment}{\texttt{\textbf{env}}\xspace}
 \newcommand{\expression}{\texttt{\textbf{expression}}\xspace}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-\subsection{Overview of Types for R}
-
-We ran \typetracer on the \CorpusLoadable packages we discussed in this section, which yields types for \NUMFUNCTIONS functions.
-To build up a picture of what types for R functions might look like, consult Table~\ref{tbl:corpus-type-breakdown}, which lists the 10 most popular argument signatures.
-The first row of the table reads: 13255 arguments have type \dbl, represnting 11.06\% of type signatures, and \typetracer observed 41M values of type \dbl flowing into or out of functions while running on the core package corpus, which represents only 1.59\% of dynamically observed values.
-Of interest are rows 2 and 4, where we see that collectively \lgl and \chr make up only 16.11\% of argument types, while making up nearly 60\% of dynamically observed values.
-
-\begin{table}
+\paragraph{Type usage}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Table~\ref{tbl:corpus-type-breakdown} reports on the 10 most frequent types
+occuring in the corpus. Scalar doubles account for 11\% function arguments,
+whereas logicals represent 27\% of observed values.  Together, \lgl and \chr
+account for 60\%.
+
+\begin{table}[!h]
 % BEGIN Autogenerated
-
+\small
 \begin{tabular}{lrrrr}
 \toprule
-Type & Args with Type & \% of Static Types & Dynamic Observations & \% of Dynamic Types\\
+Type & Args & \% of Args & Observations & \% of Observations\\
 \midrule
@@
 \ref{fig:singlefreq} provides a breakdown of the different type categories
-observed at uni-typed positions. \scalar, \class and \vector are
+observed at uni-typed positions. Scalar, class and vector are
 the most common type categories. The shaded rows correspond to polymorphic
-types. Removing those gives us XXX (XXX) monomorphic positions in a corpus of
-XXX parameter and return positions.
+types. Removing those gives us \CountMonomorphicPositions
+(\PercMonomorphicPositions) monomorphic positions in a corpus of \CountPositions
+parameter and return positions.
 

### 2020-05-16 paper/main.tex
merge

--- 2020-05-16 paper/main.tex
@@
 \newcommand{\PROPSTRUCTSDYN}{44.46\%\xspace}
+\newcommand{\NUMFILESRUN}{145,373\xspace}
 
@@
 \item We carried out a large-scale analysis of corpus of \PACKAGES widely
-  used and actively maintained libraries to extract function type signatures
-  validated the robustness of the inferred type signatures against XXX
+  used and actively maintained libraries to extract function type signatures and
+  validated the robustness of the inferred type signatures against \NUMFILESRUN
   programs that use those functions.
@@
 minimize their size by removing redundant types, types that are subsumed by
-subtyping, and \null types.
+subtyping, and \null types. Lists with more than 5 types in their union become \LIST{any}.
+We also only generate types for the first 20 arguments of a function.
 
@@
 To build up a picture of what types for R functions might look like, consult Table~\ref{tbl:corpus-type-breakdown}, which lists the 10 most popular argument signatures.
-The first row of the table reads: 13255 arguments have type \dbl, represnting 11.06\% of type signatures, and \typetracer observed 41M values of type \dbl flowing into or out of functions while running on the core package corpus, which represents only 1.59\% of dynamically observed values.
+The first row of the table reads: 13255 arguments have type \dbl, represnting 11.06\% of argument type signatures, and \typetracer observed 41M values of type \dbl flowing into or out of functions while running on the core package corpus, which represents only 1.59\% of dynamically observed values.
 Of interest are rows 2 and 4, where we see that collectively \lgl and \chr make up only 16.11\% of argument types, while making up nearly 60\% of dynamically observed values.
@@
 \toprule
-Type & Args with Type & \% of Static Types & Dynamic Observations & \% of Dynamic Types\\
+Type & Args with Type & \% Args & Dynamic Observations & \% Dynamic Types\\
 \midrule
@@
 
+\section{Future Work}
+
+We see two major avenues for future work.
+
+\subsection{Types for R Objects}
+
+\subsection{Types for Data Frames}
+
 \section{Conclusion}

### 2020-05-16 paper/main.tex

--- 2020-05-16 paper/main.tex
@@
   used and actively maintained libraries to extract function type signatures and
-  validated the robustness of the inferred type signatures against XXX
+  validated the robustness of the inferred type signatures against \NUMFILESRUN
   programs that use those functions.

### 2020-05-16 paper/main.tex
minoe

@@
 
-Table~\ref{tbl:corpus-type-breakdown} reports on the 10 most frequent types
+Table~\ref{ctb} reports on the 10 most frequent types
 occuring in the corpus. Scalar doubles account for 11\% function arguments,
@@
 \toprule
-Type & Args & \% of Args & Observations & \% of Observations\\
+Type & Args & \% of Args & Observations & \% of Obs.\\
 \midrule
-\dbl & 13,255 & 11.06 & 41,560,797 & 1.59\\
-\lgl & 10,142 & 8.46 & 712,895,959 & 27.21\\
-\null & 9,175 & 7.66 & 122,202,183 & 4.66\\
-\chr & 9,171 & 7.65 & 794,058,588 & 30.31\\
-\VEC\dbl & 7,811 & 6.52 & 29,272,232 & 1.12\\
-\addlinespace
-\VARGS & 7,759 & 6.48 & 76,369,333 & 2.92\\
-\any & 6,243 & 5.21 & 255,539,497 & 9.75\\
-\VEC\chr & 4,576 & 3.82 & 88,233,523 & 3.37\\
-\CLASS{matrix} & 4,522 & 3.77 & 32,452,555 & 1.24\\
-\CLASS{data.frame} & 2,781 & 2.32 & 7,749,094 & 0.30\\
+\dbl & 13,255 &     11.0 & 41,560,797 & 1.5\\
+\lgl & 10,142 &      8.4 & 712,895,959 & 27.2\\
+\null & 9,175 &      7.6 & 122,202,183 & 4.6\\
+\chr & 9,171 &       7.5 & 794,058,588 & 30.3\\
+\VEC\dbl & 7,811 &   6.5 & 29,272,232 & 1.1\\
+\VARGS & 7,759 &     6.4 & 76,369,333 & 2.9\\
+\any & 6,243 &       5.2 & 255,539,497 & 9.0\\
+\VEC\chr & 4,576 &   3.8 & 88,233,523 & 3.3\\
+\CLASS{matrix} & 4,522     & 3.7 & 32,452,555 & 1.2\\
+\CLASS{data.frame} & 2,781 & 2.3 & 7,749,094 & 0.3\\
 \bottomrule
@@
 % END Autogenerated
-\label{tbl:corpus-type-breakdown}
-\caption{Top types of arguments in R.}
+\caption{Top types of arguments in R}\label{ctb}
 \end{table}

### 2020-05-16 paper/main.tex
merge

--- 2020-05-16 paper/main.tex
@@
 
-\ref{fig:unionfreq} shows the number of types inferred at each position of a
+Figure~\ref{fig:unionfreq} shows the number of types inferred at each position of a
 function signature. Most functions admit or return a uni-typed value. There are
@@
 \input{tables/uni-typed-breakdown.tex}
-\ref{fig:singlefreq} provides a breakdown of the different type categories
+Figure~\ref{fig:singlefreq} provides a breakdown of the different type categories
 observed at uni-typed positions. Scalar, class and vector are
@@
 
-\ref{fig:funpoly} shows the distribution of functions against the number of
+Figure~\ref{fig:funpoly} shows the distribution of functions against the number of
 polymorphic positions. We observe that \PercMonomorphicFunctions

### 2020-05-16 paper/main.tex
corpus

@@
 For this paper we have selected \CorpusLoadable packages consisting of
-\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native code
-(C/C++/Fortran). Figure~\ref{fig:corpus} shows these packages, the size of the
-dots reflects the size of the project in terms of lines of code including both R
-and native code\footnote{ Lines of source code reported excludes comments and
+\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native
+code (C/Fortran). Figure~\ref{fig:corpus} shows these packages, the size of
+the dots reflects the project's size in lines of code including both R and
+native code\footnote{Lines of source code reported excludes comments and
   blank lines, counted by \emph{cloc}, \cf
-  \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the expression
-code coverage in percents and the y-axis gives the number of reverse
-dependencies in log scale. Dotted lines indicate means. Packages with over
-\PackageSizeOutierRnd lines of code are annotated.
-
-\begin{figure}[!h]  \centering
-  \includegraphics[width=.8\linewidth]{plots/corpus.pdf}
-  \caption{Corpus overview}\label{fig:corpus}
+  \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the
+expression code coverage in percents and the y-axis gives the number of
+reverse dependencies in log scale. Dotted lines indicate means. Packages
+with over \PackageSizeOutierRnd lines of code are annotated.
+
+\begin{figure}[!h]\centering\includegraphics[width=.8\linewidth]
+  {plots/corpus.pdf}\caption{Corpus}\label{fig:corpus}
 \end{figure}
@@
 average) of the corpus packages.
-
 Together there are \CorpusFunctionsRnd defined R functions
@@
 \CorpusSThreeClasses S3 classes.
-
 The code coverage was computed for each package using
@@
 \paragraph{Type usage}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Table~\ref{ctb} reports on the 10 most frequent types
-occuring in the corpus. Scalar doubles account for 11\% function arguments,
-whereas logicals represent 27\% of observed values.  Together, \lgl and \chr
-account for 60\%.
+During execution 3,147 different types were observed.  Classes are the most
+common types, accounting for roughly 31\% of types of arguments.  The most
+common classes are \c{matrices} (12\%), \c{data.frames} (7.5\%),
+\c{formulas} (2\%), \c{factors} (2\%), and \c{tibbles} (2\%).  Roughly 25\%
+of classes are part of R's base libraries, the others are user-defined.
+Scalars and vectors are the next most common kind, making up 41\% of
+remaining types. with scalars making up 28\% of types and vectors 12\%.
+Nulls and lists follow at 8\% and 7\% respectively, and the vararg type
+makes up 6\% of arguments.  This all totals up to over 90\% of types.
+Table~\ref{ctb} reports on the 10 most frequent types occuring in the
+corpus.
 
@@
 % BEGIN Autogenerated
-\small
-\begin{tabular}{lrrrr}
-\toprule
-Type & Args & \% of Args & Observations & \% of Obs.\\
-\midrule
+\small\begin{tabular}{lrrrr}\toprule
+Type & Args & \% of Args & Observations & \% of Obs.\\\midrule
 \dbl               & 13,255 &     11.0 &  41,560,797 & 1.5\\
@@
 \VEC\chr           &  4,576 &      3.8 &  88,233,523 & 3.3\\
-\CLASS{matrix} & 4,522     & 3.7 & 32,452,555 & 1.2\\
-\CLASS{data.frame} & 2,781 & 2.3 & 7,749,094 & 0.3\\
+\CLASS{\texttt{matrix}}     & 4,522 & 3.7 & 32,452,555 & 1.2\\
+\CLASS{\texttt{data.fram}e} & 2,781 & 2.3 &  7,749,094 & 0.3\\
 \bottomrule
 \end{tabular}
-
-% END Autogenerated
 \caption{Top types of arguments in R}\label{ctb}
@@
 and lastly, XXX externalptr types.
-
-In order to count the occurrences of each type, we subdivided them into {\it
-  kinds}: For our purposes, the {\it kind} of a type is equivalent to it's
-top-level type constructor, so all classes have kind ``class'', all lists
-have kind ``list'', and so on.  We found that classes are the most common
-kind of type, accounting for roughly 31\% of types.  In order, the most
-common classes are matrices (12\%), data.frames (7.5\%), formulas (2\%),
-factors (2\%), and tibbles (2\%).  Roughly 25\% of classes are part of R's
-base language implementation, the others being defined as part of some
-package's functionality.
-
-Besides classes, scalars and vectors are the next most common kind, making
-up 41\% of types. with scalars making up 28.1\% of types and vectors
-12.65\%.  Nulls and lists follow at 8\% and 7\% respectively, and the vararg
-type \ldots makes up 6.5\% of arguments.  This all totals up to over 90\% of
-types.  We will discuss the any type in more detail in
-Section~\ref{subsec:any}.
 

### 2020-05-16 paper/main.tex
merge

--- 2020-05-16 paper/main.tex
@@
 Arguments that are part of a function's varargs (denoted \VARGS) are
-ignored, as type \any is output for the entire varag construct.
+ignored, as type \VARGS is output as the type of the entire varargs construct.
 
@@
 \input{tables/uni-typed-breakdown.tex}
-Figure~\ref{fig:singlefreq} provides a breakdown of the different type categories
+Table~\ref{fig:singlefreq} provides a breakdown of the different type categories
 observed at uni-typed positions. Scalar, class and vector are
@@
 export only monomorphic functions.
-
-Figure 1 shows the number of alternatives in the inferred type of function
-parameter and return values.
-
-XXX\% of the parameter positions are monomorphic.  \typetracer observes XXX
-calls to XXX functions with XXX arguments. This translates to XXX parameters
-and XXX return types.  A parameter or return type is from unioning all the
-distinct observed types of
-
-Figure 1 shows the distribution of parameter and return types for these
-functions.
 

### 2020-05-16 paper/main.tex
removed extra

@@
 
-class is the most common type in our corpus, accounting for XXX\% of all the
-inferred types. This is followed by an almost equal proportion,
-28\%, scalar types and, surprisingly, only 13\% vector types. Vectors are the
-workhorse of R, most primitive operations in R are vectorized and literals are
-treated as single element vectors. It is unexpected to observe twice as many
-scalars as vectors. It is even more unexpected to observe class as the most
-common type. class of an object is used for dynamic dispatch by the S3 object
-system in R. This suggests that usage of S3 is prevalent in our corpus.
-
-null types. Together, these four types make up for 80\% of the observed
-types in our corpus.  The other types, occurring only XXX\% of the times
-include XXX environment types, XXXX expression types, XXXX pairlist types
-and lastly, XXX externalptr types.
-
-The polymorphism in R is best described by {\it ad hoc} polymorphism,
-wherein polymorphism is implemented by function and operator overloading
-(see S3 dispatch) rather than in some more fundamental way in the language
-design.  For the purposes of this evaluation, we will say that an argument
-is polymorphic if its type is a union of two or more types, in other words
-the argument has been observed to take more than one type of value.
-
-We found that 15\% of argument positions were polymorphic, and that X\% of
-functions were polymorphic.
-
 Our type annotation language does not support sophisticated class typing,

### 2020-05-16 evaluation/R/evaluate-type-analysis.Rmd
Fix graph theme

@@
 suppressPackageStartupMessages(library(kableExtra))
+
+old_theme <- theme_set(theme_bw() + theme(text = element_text(size = 15), panel.border = element_blank()))
 
@@
 scale_y_continuous(sec.axis = sec_axis(~ . * total_parameter_count, labels = count_labels), labels = relative_labels) +
-labs(x="Types", y="Parameters") +
+labs(x="Types", y="Positions") +
 scale_x_discrete(limits=c("> 3", 3, 2, 1)) +

### 2020-05-16 paper/main.tex
minor

@@
 means our type language can accurately capture an argument's type or a
-function signature.
+function signature. We get to that number in two steps.
+Fig.~\ref{fig:unionfreq} shows the number of inferred argument types and
+their size (in terms of terms in the union). The figure shows the most
+functions do not require a union at all (\PercUnitypedPositions of arguments
+do not have a union), and only \PercManytypedPositions positions have unions
+with more than three members.
 
@@
 
-Figure~\ref{fig:unionfreq} shows the number of types inferred at each position of a
-function signature. Most functions admit or return a uni-typed value. There are
-\PercUnitypedPositions uni-typed parameter and return value positions in
-function signatures. Only \PercManytypedPositions positions have more than three
-types ascribed to them.
-
 \input{tables/uni-typed-breakdown.tex}
-Table~\ref{fig:singlefreq} provides a breakdown of the different type categories
-observed at uni-typed positions. Scalar, class and vector are
-the most common type categories. The shaded rows correspond to polymorphic
+
+Table~\ref{fig:singlefreq} provides a breakdown of the different type
+categories observed at uni-typed positions. Scalar, class and vector are the
+most common type categories. The shaded rows correspond to polymorphic
 types. Removing those gives us \CountMonomorphicPositions
-(\PercMonomorphicPositions) monomorphic positions in a corpus of \CountPositions
-parameter and return positions.
+(\PercMonomorphicPositions) monomorphic positions in a corpus of
+\CountPositions parameter and return positions.
 

### 2020-05-16 paper/main.tex
minor

@@
 
+The next step, is shown in Table~\ref{fig:singlefreq}, the table provides a
+breakdown of types occuring in arguments without a union. Scalar, class and
+vector are the most common type categories. The shaded rows correspond to
+polymorphic types. Removing those gives us \CountMonomorphicPositions
+(\PercMonomorphicPositions) monomorphic positions in a corpus of
+\CountPositions parameters. With close to 70\% of monomorphic argument or
+return values, it is fair to say that even a simple type language provides
+significant benefits.
+
 \input{tables/uni-typed-breakdown.tex}
-
-Table~\ref{fig:singlefreq} provides a breakdown of the different type
-categories observed at uni-typed positions. Scalar, class and vector are the
-most common type categories. The shaded rows correspond to polymorphic
-types. Removing those gives us \CountMonomorphicPositions
-(\PercMonomorphicPositions) monomorphic positions in a corpus of
-\CountPositions parameter and return positions.
 

### 2020-05-16 evaluation/R/evaluate-type-analysis.Rmd
Increase label size
stuff

@@
 
+Make S3-less data frame, too.
+
+```{r}
+S3_funs <- read_csv("../data/package-functions.csv") %>% select(package, fun, is_s3_dispatch, is_s3_method) %>% filter(is_s3_method | is_s3_method) %>% select(package, fun)
+
+S3_funs <- S3_funs %>% rename("package_name" = "package", "function_name" = "fun")
+
+type_analysis_data_no_S3 <- type_analysis_data %>% anti_join(S3_funs)
+```
+
 ```{r high-level, echo=FALSE, eval=TRUE}
@@
 total_function_count <- nrow(distinct(type_analysis_data, package_name, function_name))
+
+total_package_count_no_S3 <- nrow(distinct(type_analysis_data_no_S3, package_name))
+total_function_count_no_S3 <- nrow(distinct(type_analysis_data_no_S3, package_name, function_name))
 ```

### 2020-05-16 paper/main.tex
mono

@@
 
+\begin{figure}[!h]  \centering
+  \includegraphics[width=.7\linewidth]{plots/union_frequency.pdf}
+  \caption{Size of unions}\label{fig:unionfreq}
+\end{figure}
+
 First we look at the share of monomorphic arguments and function signatures.
@@
 
-\begin{figure}[!h]  \centering
-  \includegraphics[width=\linewidth]{plots/union_frequency.pdf}
-  \caption{Size of unions}\label{fig:unionfreq}
-\end{figure}
+\input{tables/uni-typed-breakdown.tex}
 
@@
 
-\input{tables/uni-typed-breakdown.tex}
-
 \begin{figure}[!h]  \centering
-  \includegraphics[width=\linewidth]{plots/function_polymorphism.pdf}
+  \includegraphics[width=.7\linewidth]{plots/function_polymorphism.pdf}
   \caption{Function Polymorphism}\label{fig:funpoly}
@@
 
-Figure~\ref{fig:funpoly} shows the distribution of functions against the number of
-polymorphic positions. We observe that \PercMonomorphicFunctions
-(\CountMonomorphicFunctions) functions are monomorphic. The remaining
-\PercPolymorphicFunctions (\CountPolymorphicFunctions) have at least one
-polymorphic parameter or return type. Polymorphic functions are spread across
-many packages, \CountMonomorphicPackages out of the 412 packages from our corpus
-export only monomorphic functions.
-
-Our type annotation language does not support sophisticated class typing,
-and much of the polymorphism in R is due to these classes, namely through
-the S3 dispatch mechanism.  In future work, we aim to more fully handle S3
-and objects in R.  For now, if we count out
+If we look at the numbers from the point of view of functions and count how
+many of their arguments are polymorphic, we observe that
+\PercMonomorphicFunctions (\CountMonomorphicFunctions) functions are
+monomorphic. The remaining \PercPolymorphicFunctions
+(\CountPolymorphicFunctions) have at least one polymorphic parameter or
+return type.  Figure~\ref{fig:funpoly} shows the distribution of functions
+against the number of polymorphic arguments. Finally, we count
+\CountMonomorphicPackages out of the 412 packages export only monomorphic
+functions.
 
 \subsubsection{Discussion}
-
-Now, we aim to enhance the quantitative analysis of our language design by looking at select code snippets from the corpus we analyzed.
-We will highlight some of the design decisions that we made, give code examples, and accompany them with function signatures generated by \typetracer.
+A number of lessons can be drawn from the data we have gathered.
 

### 2020-05-16 evaluation/R/evaluate-type-analysis.Rmd
Fix vector na

@@
 
-old_theme <- theme_set(theme_bw() + theme(text = element_text(size = 15), panel.border = element_blank()))
+old_theme <- theme_set(theme_bw() + theme(text = element_text(size = 20), panel.border = element_blank()))
 
@@
 type_analysis_data %>%
-filter(kind == "vector-na") %>%
+filter(kind == "^vector") %>%
 group_by(type) %>%

### 2020-05-16 paper/main.tex
scalara

@@
 \paragraph{NAs}
-Our data supports the choice of making the presence of NAs explicit.  In
-practice, programmers explicitly check for NAs in primitive vectors,
-possibly sanitizing them if they are present.
+Our data supports making the presence of {\NA}s explicit. Only 1,058 (of
+0.9\%) of arguments are marked as possibly having {\NA}s, thus the
+overwhelming majority of types appear to be {\NA}-free.  In practice,
+programmers check for them and sanitize them if they are present.  Consider
+the \code{binom} package for computing confidence intervals and its
+\code{binom.profile} function.  This attached code snippet highlights a data
+sanitization pattern: the programmer first binds the vectors into a matrix,
+then finds rows where both columns are not \NA, extracts non-\NA values and
+stores them into \code{x} and \code{n} respectively.
 
 \begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
-#' @type <x: dbl[], n: dbl[], conf.level: dbl, maxsteps: dbl,
-#'         del: dbl, bayes: lgl, plot: lgl, ...> => class<data.frame>
-binom.profile <- function(x, n, conf.level = 0.95, maxsteps = 50,
-                            del = zmax / 5, bayes = TRUE, plot = FALSE, ...) {
+  binom.profile <- function(x, n, conf.level=0.9, maxsteps=50,...) {
     xn <- cbind(x = x, n = n)
@@
 
-The \code{binom} package is a popular R package for computing confidence
-intervals, and the \code{binom.profile} function does so using the profile
-likelihood.  This code snippet highlights a data sanitization pattern: the
-programmer first binds the vectors into a matrix, with one column for each
-of \code{x} and \code{n}, then finds rows where both columns are not \NA,
-then extracts non-\NA values and stores them into \code{x} and \code{n}
-respectively.We see from the type that arguments \code{x} and \code{n} are intended to be free of \NA values.
-\contractr will check them for compliance, and warn programmers of function misuse.
-
 \paragraph{Scalars}
 The data also suggests that programmers often use scalars, and do
-dimensionality checks on their data.
+dimensionality checks on their data.  In our data 31,533 (or 29\%) of the
+arguments are scalar types. While not completely surprising, this is a
+rather large number. Consider the \code{hankel.matrix} function, it takes
+two arguments and checks that \code{n} is \int, that \code{x} is a vector,
+and also, indirectly, that \code{n} is a a scalar (this comes from the fact
+that it is used in the guard of a conditional which fails if \code{n} is
+not a scalar).
 
 \begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
-#' @type <n: dbl, x: int[]> => class<matrix>
  hankel.matrix <- function( n, x ) { 
@@
   ### x = an order 2 * n + 1 vector of numeric values
-    ...
     if ( n != trunc( n ) )  stop( "argument n ix not an integer" )
@@
 \end{lstlisting}
-
-The \code{hankel.matrix} function takes two arguments.  The function has a
-number of interesting checks: \code{n} undergoes an integer type check;
-\code{x} undergoes a vector type check; and \code{n} is expected to be a
-scalar, as if it were a vector, the code \code{m < n} would essentially
-check \code{m < n[i]} for each element in \code{n}, returning a vector of
-comparison results, and the conditional of an if must be a scalar else R
-generates a warning. The type for this function indicates that \code{x} must be
-a vector, though here the type for \code{n} is found to be \dbl---this is likely due
-to numeric values defaulting to \dbl type in R, as e.g. \code{typeof(2)} is \dbl.
 
@@
 and many functions check to ensure that their arguments are matrices.
+The \code{rowWeightedMeans} function calculates the weighted means of rows
+or columns of a matrix \code{x}.  This function exhibits good practice in
+type checking arguments, rightly producing a message to the user indicating
+that an explicit matrix should be passed. We captured this with our class
+types, and indeed \typetracer inferred a good type for this function,
+allowing \contractr to perform the check present in this code automatically,
+notifying the users passing ill-typed arguments.
 
 \begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
-#' @type <x: class<matrix>, w: ? ^dbl[], rows: null, cols: null, na.rm: lgl, ...> => ^dbl[]
-rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL,
-                             na.rm = FALSE, ...) {
+  rowWeightedMeans <- function(x, w=NULL, rows=NULL, cols=NULL, na.rm=FALSE, ...) {
     if (!is.matrix(x)) 
@@
 
-The \code{rowWeightedMeans} function calculates the weighted means of rows
-or columns of a matrix \code{x}.  This function exhibits good practice in
-type checking arguments, rightly producing a message to the user indicating
-that an explicit matrix should be passed. We captured this with our class types, 
-and indeed \typetracer inferred a good type for this function, allowing \contractr to 
-perform the check present in this code automatically, notifying the users passing ill-typed arguments.
-
 Thus far we've seen instances where we introduce types that subsume existing
@@
   rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL, na.rm = FALSE) {
-    ...
     n <- ncol(x)
@@
 This code is from later in the same \code{rowWeightedMeans} function
-discussed previously.  Here, the length of the weights vector \code{w} is
+pdiscussed previously.  Here, the length of the weights vector \code{w} is
 being checked for equality with the number of columns in the matrix
@@
 
-In R, lists can have a names attribute, allowing lists elements to be accessed by name using the
-\code{$} operator.
+In R, lists can have a names attribute, allowing lists elements to be
+accessed by name using the \code{$} operator.
 
 \begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
-#' @type <x: class<aov, lm> | class<lm>> => dbl;
   agricolae::cv.model <- function(x) {
@@
 In the code snippet, the function \code{cv.model} takes an argument
-\code{x}, which we observed to always be a {\it linear model} (either ANOVA with \CLASS{aov, lm} or a simple linear regression model with \CLASS{lm}) which is part
-of R's built-in statistics functionality.  Internally, linear models are
-represented as lists with named elements, and one may access these named
-elements using the \code{$} syntax, as seen on lines 2-4 in the snippet.  To
-describe these kinds of values, we intended to include a struct type in our type
-system, allowing users to specify name, type pairs as part of list type
-definitions, though we found this to be problematic.
-
-The inclusion of structs complicated our analysis by
-polluting the data, illustrated on line 8 with the call
-\code{data(sweetpotato)}.  It is quite common for R example code to use some
-built-in data sets, such as \code{sweetpotato}, to show off some package
-functionality with ``real data''.  When \typetracer is configured to detect
-structs we saw a marked increase in misrepresentative types, as the types of
-functions which use these data sets (such as \code{aov} in the code example) became polluted with the names from the example
-sets.  We implemented a unification strategy for struct types, removing them
-if they co-occurred with list types, and turning them into lists if two
-co-occurring struct types shared no common names, but even so we found
-ourselves with function signatures that were full of noise introduced by the
-struct names.
-
-We found that struct-like data was best described with classes, as illustrated in the code snippet, and many structs actually had a class.
-Even though \code{x} is used like a struct, it in fact has a linear model class, which we believe is a more informative type.
-
-%
+\code{x}, which we observed to always be a {\it linear model} (either ANOVA
+with \CLASS{aov, lm} or a simple linear regression model with \CLASS{lm})
+which is part of R's built-in statistics functionality.  Internally, linear
+models are represented as lists with named elements, and one may access
+these named elements using the \code{$} syntax, as seen on lines 2-4 in the
+snippet.  To describe these kinds of values, we intended to include a struct
+type in our type system, allowing users to specify name, type pairs as part
+of list type definitions, though we found this to be problematic.
+
+The inclusion of structs complicated our analysis by polluting the data,
+illustrated on line 8 with the call \code{data(sweetpotato)}.  It is quite
+common for R example code to use some built-in data sets, such as
+\code{sweetpotato}, to show off some package functionality with ``real
+data''.  When \typetracer is configured to detect structs we saw a marked
+increase in misrepresentative types, as the types of functions which use
+these data sets (such as \code{aov} in the code example) became polluted
+with the names from the example sets.  We implemented a unification strategy
+for struct types, removing them if they co-occurred with list types, and
+turning them into lists if two co-occurring struct types shared no common
+names, but even so we found ourselves with function signatures that were
+full of noise introduced by the struct names.
+
+We found that struct-like data was best described with classes, as
+illustrated in the code snippet, and many structs actually had a class.
+Even though \code{x} is used like a struct, it in fact has a linear model
+class, which we believe is a more informative type.
+
 \paragraph{Data Frames}

### 2020-05-16 paper/main.tex
Add type signatures

@@
 
+type `data.tree::Traverse` <class<`Node`, `R6`>, character[], any, any> => list<any>;
+
+type `emulator::cprod` <class<`matrix`> | double[], class<`matrix`> | double[]> => class<`matrix`>;
+
+type `lava::lava.options` <...> => list<double | character[]>;
+
+type `coin::\%GE\%` <class<`matrix`> | class<`matrix`, `pvalue`> | double[], double[] | class<`matrix`, `pvalue`>> => (class<`matrix`> | logical[]);
+
+type `emplik::llogp` <class<`matrix`> | double[], double> => (class<`matrix`> |
+double[]);
+
+type `emplik::gradf` <class<`matrix`>, double[], double[]> => double[];
+
+type `fields::rdist` <class<`data.frame`> | class<`matrix`> | double[], class<`data.frame`> | class<`matrix`> | null | double[], logical> => (class<`matrix`> | class<`dist`>);
+
 \bibliography{bib/biblio,bib/jv,bib/r,bib/new,bib/gradual}

### 2020-05-16 paper/main.tex
stuff

@@
 
+\paragraph{Nullables}
+The number of argument which may be \c{NULL} is 3,165 (or 3.5\%). This is a
+relatively small number of occurrences, but it is worth expressing the
+potential for the presence of \c{NULL} as these would likely inhibit
+optimizations.
+
+\paragraph{Structs}
+While experimenting with various design, we consider adding a struct type to
+capture lists with named elements that can be accessed with the \code{$}
+operators. We ended discarding those types as they grew large and were often
+only representative of the example data being manipulated.  Consider
+function \code{cv.model}, its argument \code{x} is observed to be of
+\CLASS{\texttt{aov}, \texttt{lm}} or \CLASS{\texttt{lm}}).  Internally,
+linear models are represented as lists with named elements.  The polution is
+illustrated by the lines after the funtion definition.  They load an example
+data set and test the function \code{cv.model}.  The
+\code{data(sweetpotato)} expression load a sample data set. The fields of
+\code{sweetpotato} will be recorded when \code{cv.model} is called.
+
+\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
+  cv.model <- function(x) {
+    suma2 <- sum(x$residual^2)
+    gl <- x$df.residual
+    promedio <- mean(x$fitted.values)
+    return(sqrt(suma2/gl)*100/promedio)
+  }
+
+  data(sweetpotato)
+  model<-aov(yield~virus, data=sweetpotato)
+  cv.model(model)
+\end{lstlisting}
+
+\paragraph{Objects}
+While we record classes, our analysis does not deal with method dispatch.  R
+has multiple, disparate, object systems, called S3, S4, and R5.  The
+\code{class} attribute is used by these systems to dispatch methods. S3 does
+single dispatch, S4 does multiple dispatch and R5 supports impertative
+objects.  The mechanics of S4 dispatch are more complex than for S3, and
+users can define their own class hierarchies that we would need to
+incorporate in our type analysis and contract checking frameworks.  We found
+limited use of S4 during our analysis. Coming up with a type system that
+accounts for all of these factors and consolidates multiple
+object-orientation frameworks in a single language design is an interesting
+problem in and of itself, one we leave for future work.
+
 \paragraph{Matrices}
@@
 
-\paragraph{Structs}
-
-In R, lists can have a names attribute, allowing lists elements to be
-accessed by name using the \code{$} operator.
-
-\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
-  agricolae::cv.model <- function(x) {
-    suma2 <- sum(x$residual^2)
-    gl <- x$df.residual
-    promedio <- mean(x$fitted.values)
-    return(sqrt(suma2/gl)*100/promedio)
-  }
-  data(sweetpotato)
-  model<-aov(yield~virus, data=sweetpotato)
-  cv.model(model)
-\end{lstlisting}
-
-In the code snippet, the function \code{cv.model} takes an argument
-\code{x}, which we observed to always be a {\it linear model} (either ANOVA
-with \CLASS{aov, lm} or a simple linear regression model with \CLASS{lm})
-which is part of R's built-in statistics functionality.  Internally, linear
-models are represented as lists with named elements, and one may access
-these named elements using the \code{$} syntax, as seen on lines 2-4 in the
-snippet.  To describe these kinds of values, we intended to include a struct
-type in our type system, allowing users to specify name, type pairs as part
-of list type definitions, though we found this to be problematic.
-
-The inclusion of structs complicated our analysis by polluting the data,
-illustrated on line 8 with the call \code{data(sweetpotato)}.  It is quite
-common for R example code to use some built-in data sets, such as
-\code{sweetpotato}, to show off some package functionality with ``real
-data''.  When \typetracer is configured to detect structs we saw a marked
-increase in misrepresentative types, as the types of functions which use
-these data sets (such as \code{aov} in the code example) became polluted
-with the names from the example sets.  We implemented a unification strategy
-for struct types, removing them if they co-occurred with list types, and
-turning them into lists if two co-occurring struct types shared no common
-names, but even so we found ourselves with function signatures that were
-full of noise introduced by the struct names.
-
-We found that struct-like data was best described with classes, as
-illustrated in the code snippet, and many structs actually had a class.
-Even though \code{x} is used like a struct, it in fact has a linear model
-class, which we believe is a more informative type.
-
 \paragraph{Data Frames}
@@
                       ignore_row_order = TRUE, convert = FALSE, ...) { ... }
-\end{lstlisting}
-
-\paragraph{Objects}
-
-R has multiple, disparate object systems, with the language providing direct
-support for S3, S4, and R5.  Thus far, we've mainly discussed S3 and its
-dispatch mechanism, and indeed our class types capture the S3
-classes of values, but not user-defined polymorphism.  We will not discuss R5 as it is still under development.
-
-S4, like S3, uses the class attribute to store types, we type them with a \CLASS{...} type.  
-We stuck to simple class-based types for S4 since
-doing anything more would be immensely complicated, and outside of the scope
-of this work.  For instance, the mechanics of S4 dispatch are more complex
-than for S3 (S4 dispatches on the class of {\it all} arguments), and users
-can define their own class hierarchies that we would need to incorporate in
-our type analysis and contract checking frameworks.  Further,we found
-  limited use of S4 during our analysis. Coming up with a type system that
-accounts for all of these factors and consolidates multiple
-object-orientation frameworks in a single language design is an interesting
-problem in and of itself, and we aim to extend our type annotation framework
-to account for this in future work.  For now, simple class types for S3 and S4 suffice.
-
-To illustrate informative class types, consider the following code snippet.
-\code{distr} is a package which implements statistical distributions in an object-oriented way, and the \code{Binom} function acts as a constructor for \CLASS{Binom} values.
-Another example is from the \code{Epi} package, which provides functionality for demographic and epidemiological analysis in the Lexis diagram.
-The function we highlighted, \code{timeBand}, extracts time band data from the passed Lexis object.
-We see from the type that the first argument is the Lexis object, and that either a \VEC\dbl or a \CLASS{factor} is returned.
-Often, class members are implemented as attributes on R values, and it's likely that this is the case for Lexis objects as evidenced by the programmer accessing the "breaks" attribute in the function code.
-
-\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
-#' @type <size: dbl, prob: dbl> => class<Binom>
-distr::Binom <- function(size = 1,prob = 0.5) new("Binom", size = size, prob = prob)
-
-#' @type <lex: class<`Lexis`, `data.frame`>, time.scale: chr, type: chr> => 
-#'           class<`factor`> | dbl[];
-Epi::timeBand <- function(lex, time.scale, type="integer") {
-  time.scale <- check.time.scale(lex, time.scale)[1]
-  breaks <- attr(lex, "breaks")[[time.scale]]
-  ...
 \end{lstlisting}

### 2020-05-16 paper/main.tex
types

--- 2020-05-16 paper/main.tex
@@
 
+type `stabledist::x.exp.m.x` <double[]> => double[];
+
+type `lava::index` <class<`lvm`> | class<`lvmfit`>, ...> => (null | list<any>);
+
+type `pkgload::compact` <list<any>> => list<any>;
+
+type `diptest::dip` <double[], character | logical, logical, double> => (class<`dip`> | double);
+
+type `bbmle::relist2` <double[], double[] | list<any>> => list<any>;
+
+type `stabledist::cospi2` <double[]> => double[];
+type `stabledist::tanpi2` <double[]> => double[];
+
+type `neuralnet::relist` <double[], class<`array`> | integer[], class<`array`> | integer[]> => list<class<`matrix`>>;
+type `neuralnet::calculate.gradients` <any, integer, list<class<`matrix`>>, any, class<`matrix`>, double, logical> => double[];
+
 \bibliography{bib/biblio,bib/jv,bib/r,bib/new,bib/gradual}

### 2020-05-16 paper/main.tex
types

@@
 
-type `emplik::llogp` <class<`matrix`> | double[], double> => (class<`matrix`> |
-double[]);
+type `emplik::llogp` <class<`matrix`> | double[], double> => (class<`matrix`> | double[]);
 
@@
 
+type `openssl::decrypt_envelope` <raw[], raw[], raw[], class<`key`, `rsa`>, any> => raw[];
+
+dbplyr:type `set_win_current_group` <null | character[]> => (character[] | null);
+
+type `sha256` <raw[], null | raw[]> => raw[];
+
+type `initparam` <? double, any, any, any, character, character, logical, double[], double[], any> => double[];
+
 \bibliography{bib/biblio,bib/jv,bib/r,bib/new,bib/gradual}

### 2020-05-16 paper/main.tex
stuff

@@
 \paragraph{Matrices}
-
-Matrices represent another layer of complexity on top of vectorized
-primitives.  Internally, matrices are simply vectors with class ``matrix''
-and a \code{dims} attribute indicating matrix dimensions, and while not
-codified in the language semantics, many internal functions will coerce
-vectors to matrices automatically.  Relying on automatic coercion is unwise,
-and many functions check to ensure that their arguments are matrices.
-The \code{rowWeightedMeans} function calculates the weighted means of rows
-or columns of a matrix \code{x}.  This function exhibits good practice in
-type checking arguments, rightly producing a message to the user indicating
-that an explicit matrix should be passed. We captured this with our class
-types, and indeed \typetracer inferred a good type for this function,
-allowing \contractr to perform the check present in this code automatically,
-notifying the users passing ill-typed arguments.
+Matrices are instances of the eponymous class. They have a \code{dims}
+attribute indicating dimensions, and while not codified in the language
+semantics, many internal functions coerce vectors to matrices automatically.
+The \code{rowWeightedMeans} function calculates the weighted means of rows.
+The programmer added a type check for \code{x}.
 
@@
   rowWeightedMeans <- function(x, w=NULL, rows=NULL, cols=NULL, na.rm=FALSE, ...) {
-    if (!is.matrix(x)) 
-      .Defunct(msg = sprintf("Argument 'x' should be a matrix.)
+    if (!is.matrix(x))   .Defunct(msg = sprintf("'x' should be a matrix.)
 \end{lstlisting}
 
-Thus far we've seen instances where we introduce types that subsume existing
-argument validation code, but our intention is not to subsume {\it all} such
-code. Consider:
-
-\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
-  rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL, na.rm = FALSE) {
-    n <- ncol(x)
-    if (length(w) != n) 
-       stop("The length of 'w' does not match columns in 'x': ")
-\end{lstlisting}
-
-This code is from later in the same \code{rowWeightedMeans} function
-pdiscussed previously.  Here, the length of the weights vector \code{w} is
-being checked for equality with the number of columns in the matrix
-\code{x}.  Specifying this relationship as a type amounts to {\it dependent
-  typing}, which we believe too complex to retrofit onto a dynamic data
-science language such as R.  Ultimately, we decided to leave these kinds of
-checks up to the programmer.
-
 \paragraph{Data Frames}
-
-One of the most popular classes in R is the \code{data.frame} class, making up nearly 8\% of observed classes.  Data
-frames and the derivative \code{tibble} (from the \code{tidyverse}
-ecosystem) and \code{data.table} (from the \code{data.table} package) types
-underpin a huge amount of data analysis in R.
-
+One of the most popular classes in R is the \code{data.frame} class, making
+up nearly 8\% of observed classes.  Data frames and the derivative
+\code{tibble} and \code{data.table} types underpin many popular uses of R.
 One way to deal with data frames is through the struct type, with a named
 field for each column of the data frame, but as mentioned previously structs
-introduced undue noise into \typetracer's analysis results.  Further
-complicating data frames is that many functions built to operate on them
-operate in a name-agnostic way.  For instance, the \code{tidyverse} package
-ecosystem allows programmers to pass column names to functions which operate
-on their data frames.  In base R, idiomatic data frame use is to use string
-column names to select rows from the frame (unless only a single column is
-of interest, wherein the \code{$} syntax is appropriate).
-
-\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
-data(cars)
-# tidyverse: Filter rows where speed is > 20. 
-fast_cars_tidy <- filter(cars, speed > 20)
-# base R: Get all rows where the "speed" column has a value > 20.
-fast_cars_base <- cars[cars[, "speed"] > 20, ] # base example
-\end{lstlisting}
-
-Doing justice to this type will be an endeavour in and of itself, requiring
-reconciliation of base R's \code{data.frame}, \code{tidyverse}'s
-\code{tibble}, and \code{data.table}'s \code{data.table}.  Developing useful
-types for the \code{tidyverse} ecosystem may well involve
-
-It will also require proper treatment of function types, and perhaps even
-parametric types, which dynamic analyses such as \typetracer are unable to
-discern soundly.
-
-In the context of our type annotation language, we ascribe the type \CLASS{data.frame} to \code{data.frame}-typed values.
-Consider the following code snippet, showing the type signature for \code{dplyr}'s \code{all_equal} function which checks two \CLASS{data.frame} for equality.
-\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
-#' @type <target: class<data.frame>, current: class<data.frame>, ignore_col_order: lgl, 
-#'         ignore_row_order: lgl, convert: lgl, ...> => (chr | lgl)
-dplyr::all_equal <- function(target, current, ignore_col_order = TRUE,
-                      ignore_row_order = TRUE, convert = FALSE, ...) { ... }
-\end{lstlisting}
+introduced undue noise.  Further complicating data frames is that many
+functions built to operate on them operate in a name-agnostic way.  For
+instance, the \code{tidyverse} package ecosystem allows programmers to pass
+column names to functions which operate on their data frames.  In base R,
+idiomatic data frame use is to use string column names to select rows from
+the frame (unless only a single column is of interest, wherein the \code{$}
+syntax is appropriate).
 

### 2020-05-16 paper/main.tex
types in types out

--- 2020-05-16 paper/main.tex
@@
 
+type `dplyr::group\_indices` <class<`data.frame`>, ...> => integer[];
+
+type `lava::index` <class<`lvm`> | class<`lvmfit`>, ...> => (null | list<any>);
+
+type `diptest::dip` <double[], character | logical, logical, double> => (class<`dip`> | double);
+
+type `stabledist::cospi2` <double[]> => double[];
+
+type `emplik::llogp` <class<`matrix`> | double[], double> => (class<`matrix`> | double[]);
+
 type `data.tree::Traverse` <class<`Node`, `R6`>, character[], any, any> => list<any>;
 
-type `emulator::cprod` <class<`matrix`> | double[], class<`matrix`> | double[]> => class<`matrix`>;
-
-type `lava::lava.options` <...> => list<double | character[]>;
-
-type `coin::\%GE\%` <class<`matrix`> | class<`matrix`, `pvalue`> | double[], double[] | class<`matrix`, `pvalue`>> => (class<`matrix`> | logical[]);
-
-type `emplik::llogp` <class<`matrix`> | double[], double> => (class<`matrix`> |
-double[]);
-
-type `emplik::gradf` <class<`matrix`>, double[], double[]> => double[];
-
-type `fields::rdist` <class<`data.frame`> | class<`matrix`> | double[], class<`data.frame`> | class<`matrix`> | null | double[], logical> => (class<`matrix`> | class<`dist`>);
-
-type `stabledist::x.exp.m.x` <double[]> => double[];
-
-type `lava::index` <class<`lvm`> | class<`lvmfit`>, ...> => (null | list<any>);
-
-type `pkgload::compact` <list<any>> => list<any>;
-
-type `diptest::dip` <double[], character | logical, logical, double> => (class<`dip`> | double);
-
-type `bbmle::relist2` <double[], double[] | list<any>> => list<any>;
-
-type `stabledist::cospi2` <double[]> => double[];
-type `stabledist::tanpi2` <double[]> => double[];
-
-type `neuralnet::relist` <double[], class<`array`> | integer[], class<`array`> | integer[]> => list<class<`matrix`>>;
-type `neuralnet::calculate.gradients` <any, integer, list<class<`matrix`>>, any, class<`matrix`>, double, logical> => double[];
+type `openssl::decrypt\_envelope` <raw[], raw[], raw[], class<`key`, `rsa`>, any> => raw[];
+
+dbplyr:type `set\_win\_current\_group` <null | character[]> => (character[] | null);
+
+type `sha256` <raw[], null | raw[]> => raw[];
+
+type `initparam` <? double, any, any, any, character, character, logical, double[], double[], any> => double[];
 

### 2020-05-16 paper/main.tex

--- 2020-05-16 paper/main.tex
@@
 
+type `dplyr::group\_indices` <class<`data.frame`>, ...> => integer[];
+
+type `lava::index` <class<`lvm`> | class<`lvmfit`>, ...> => (null | list<any>);
+
+type `diptest::dip` <double[], character | logical, logical, double> => (class<`dip`> | double);
+
+type `stabledist::cospi2` <double[]> => double[];
+
+type `emplik::llogp` <class<`matrix`> | double[], double> => (class<`matrix`> | double[]);
+
 type `data.tree::Traverse` <class<`Node`, `R6`>, character[], any, any> => list<any>;
 
-type `emulator::cprod` <class<`matrix`> | double[], class<`matrix`> | double[]> => class<`matrix`>;
-
-type `lava::lava.options` <...> => list<double | character[]>;
-
-type `coin::\%GE\%` <class<`matrix`> | class<`matrix`, `pvalue`> | double[], double[] | class<`matrix`, `pvalue`>> => (class<`matrix`> | logical[]);
-
-type `emplik::llogp` <class<`matrix`> | double[], double> => (class<`matrix`> | double[]);
-
-type `emplik::gradf` <class<`matrix`>, double[], double[]> => double[];
-
-type `fields::rdist` <class<`data.frame`> | class<`matrix`> | double[], class<`data.frame`> | class<`matrix`> | null | double[], logical> => (class<`matrix`> | class<`dist`>);
-
-type `stabledist::x.exp.m.x` <double[]> => double[];
-
-type `lava::index` <class<`lvm`> | class<`lvmfit`>, ...> => (null | list<any>);
-
-type `pkgload::compact` <list<any>> => list<any>;
-
-type `diptest::dip` <double[], character | logical, logical, double> => (class<`dip`> | double);
-
-type `bbmle::relist2` <double[], double[] | list<any>> => list<any>;
-
-type `stabledist::cospi2` <double[]> => double[];
-type `stabledist::tanpi2` <double[]> => double[];
-
-type `neuralnet::relist` <double[], class<`array`> | integer[], class<`array`> | integer[]> => list<class<`matrix`>>;
-type `neuralnet::calculate.gradients` <any, integer, list<class<`matrix`>>, any, class<`matrix`>, double, logical> => double[];
-
-type `openssl::decrypt_envelope` <raw[], raw[], raw[], class<`key`, `rsa`>, any> => raw[];
-
-dbplyr:type `set_win_current_group` <null | character[]> => (character[] | null);
+type `openssl::decrypt\_envelope` <raw[], raw[], raw[], class<`key`, `rsa`>, any> => raw[];
+
+dbplyr:type `set\_win\_current\_group` <null | character[]> => (character[] | null);
 

### 2020-05-16 paper/main.tex
Fix types

@@
 
-type `data.tree::Traverse` <class<`Node`, `R6`>, character[], any, any> => list<any>;
-
-type `openssl::decrypt\_envelope` <raw[], raw[], raw[], class<`key`, `rsa`>, any> => raw[];
-
-dbplyr:type `set\_win\_current\_group` <null | character[]> => (character[] | null);
-
-type `sha256` <raw[], null | raw[]> => raw[];
-
-type `initparam` <? double, any, any, any, character, character, logical, double[], double[], any> => double[];
+$$type `data.tree::Traverse` \FUN{\CLASS{`Node`, `R6`}, ~\VEC\chr, \any, \any}\LIST\any;$$
+
+$$type `openssl::decrypt\_envelope` \FUN{~\VEC\raw, ~\VEC\raw, ~\VEC\raw, \CLASS{`key`, `rsa`}, \any}~\VEC\raw;$$
+
+$$dbplyr:type `set\_win\_current\_group` \FUN{? ~\VEC\chr}? ~\VEC\chr;$$
+
+$$type `sha256` \FUN{~\VEC\raw, ? ~\VEC\raw}~\VEC\raw;$$
+
+$$type `initparam` \FUN{? \dbl, \any, \any, \any, \chr, \chr, \lgl, ~\VEC\dbl, ~\VEC\dbl, \any}~\VEC\dbl;$$
 

### 2020-05-16 paper/main.tex
types

--- 2020-05-16 paper/main.tex
@@
 
-type `dplyr::group\_indices` <class<`data.frame`>, ...> => integer[];
-
-type `lava::index` <class<`lvm`> | class<`lvmfit`>, ...> => (null | list<any>);
-
-type `diptest::dip` <double[], character | logical, logical, double> => (class<`dip`> | double);
-
-type `stabledist::cospi2` <double[]> => double[];
-
-type `emplik::llogp` <class<`matrix`> | double[], double> => (class<`matrix`> | double[]);
+\code{dplyr::group_indices}: \FUN{\CLASS{data.frame}, \VARGS}{\VEC\int}
+
+\code{lava::index}: \FUN{\CLASS{lvm} \; | \; \CLASS{lvmfit}, \VARGS}{\; ? \; \LIST{any}}
+
+\code{diptest::dip}: \FUN{\VEC\dbl, \chr \; | \; \lgl, \lgl, \dbl}{\CLASS{dip} \; | \; \dbl}
+
+\code{stabledist::cospi2}: \FUN{\VEC\dbl}{\VEC\dbl}
+
+\code{matrixcalc::matrix.power}: \FUN{\CLASS{matrix}, \dbl}{\CLASS{matrix}}
 

### 2020-05-16 paper/main.tex
intro

--- 2020-05-16 paper/main.tex
@@
 \begin{abstract}
-The R programming language is widely used in a variety of domains for tasks
-related to data science. R was designed to favor an interactive style of
-programming with minimal syntactic and conceptual overhead. This design is
-well suited to interactive data analysis, but a bad fit for tools such as
-compilers or program analyzers which must generate performant code or catch
-programming errors.  In particular, R has no type annotations, its
-operations are dynamically checked at run-time. The starting point for our
-work are the twin questions, \emph{what expressive power is needed to
-  accurately type R code?} and \emph{which type system is the R community
-  willing to adopt?} Both questions are difficult to answer without actually
-designing a type system and attempting to convince users and package
-developers to try the proposed system.  The goal of this paper is to provide
-data that can feed into the design process. To this end, we perform a large
-corpus analysis with aim to gain insights in the degree of polymorphism
-exhibited by idiomatic R code and explore potential benefits that the R
-community could accrue from even a simple type system.  As a starting point,
-we infer type signatures for \NUMFUNCTIONS functions from \NUMPACKAGES
-packages among the most widely used open source R libraries. The signatures
-use a minimal type language with only basic types, vectors and classes.  We
-implement a contract checker that can be used to validate the inferred type
-signatures.
+The R programming language is widely used in a variety of domains.  It was
+designed to favor an interactive style of programming with minimal syntactic
+and conceptual overhead. This design is well suited to interactive data
+analysis, but a bad fit for tools such as compilers or program analyzers
+which must generate native code or catch programming errors.  In particular,
+R has no type annotations, all operations are dynamically checked at
+run-time. The starting point for our work are the twin questions, \emph{what
+  expressive power is needed to accurately type R code?} and \emph{which
+  type system is the R community willing to adopt?} Both questions are
+difficult to answer without actually experimenting with a type system. The
+goal of this paper is to provide data that can feed into that design
+process. To this end, we perform a large corpus analysis to gain insights in
+the degree of polymorphism exhibited by idiomatic R code and explore
+potential benefits that the R community could accrue from a simple type
+system.  As a starting point, we infer type signatures for \NUMFUNCTIONS
+functions from \NUMPACKAGES packages among the most widely used open source
+R libraries.
 \end{abstract} \maketitle
@@
 
-In this section, we will introduce related work and the R programming language.
+In this section, we introduce related work on extending dynamic languages
+with static type system and we give a short primer on the R programming
+language.
 
@@
 
-There is no previous work for the R programming language. We take
+There is no previous work on types for the R programming language. We take
 inspiration in the above mentioned works but focus on adapting them to our
@@
 length of the varargs and of \k x and \k y.  The language does not really
-differentiate between scalar and vectors.
-Some more exotic types that can be encountered include vectors of complex
-number and list of arbitrary types. Function \k f can be invoked with those
-as well.
+differentiate between scalar and vectors.  Some more exotic types that can
+be encountered include vectors of complex number and list of arbitrary
+types. Function \k f can be invoked with those as well.
 

### 2020-05-16 paper/main.tex
update type
Fix types
minor

@@
 
-This section sets out to propose a candidate design for a type language that
-can be used to describe the arguments and return values of functions in R.
-The goal is not to be the final design but rather a starting point for an
-iterative process.
-
+This section sets out to propose a candidate design for a type language to
+describe the arguments and return values of functions in R.  The goal is not
+to get a final design but rather a starting point for an iterative process.
 Fig.~\ref{types} presents our type language. An early design choice was to
@@
 
-\paragraph{Scalar vs. Vectors} Our type system distinguishes between
-vectors of length 1, and vectors of any dimensions. The six base types,
-integers, doubles, complex numbers, logicals (booleans), characters, and
-raws (bytes), can be either vectors (e.g., \VEC\int) or scalars (e.g.,
-\int).  It is worthwhile to note that a vector can happen to be of length 1
-and thus a scalar value is a subtype of the corresponding vector type.
-Vectors are homogeneous, in that a vector of doubles contains only doubles.
-
-\paragraph{Missing values} In R, a missing value is written
-\NA, each of basic types has its specific representation of \NA. Thus there
-is an \int missing observation as well as a double missing, and they are
-different.  The type system allows to distinguish between vectors that may
-contain missing observations (written \NAVEC\int) and those who are
-guarantee to be \NA-free (\VEC\int).  A \NA-free vector can be treated as a
-vector that may have missing observations.  We do not allow scalar values to
-be missings.  The type \raw does not allow {\NA}s.
+\paragraph{Scalar vs. Vectors} We distinguish between
+vectors of length 1, and vectors of any dimension. The base types, \int,
+\dbl, \clx, \lgl (booleans), \chr, and \raw (bytes), can be either vectors
+(e.g., \VEC\int) or scalars (e.g., \int).  A vector can happen to be of
+length 1, and thus a scalar value is also a vector.  Vectors are
+homogeneous, in that a vector of doubles contains only doubles.
+
+\paragraph{NA values} In R, a value can be \emph{not available}, written
+\NA. Each of basic types has its specific \NA. Thus there is an \int \NA as
+well as a double \NA, and they are different.  The type system allows to
+distinguish between vectors that may contain {\NA}s(written \NAVEC\int) and
+those who are guarantee to be \NA-free (\VEC\int).  A \NA-free vector can be
+treated as a vector that may {\NA}s.  We do not allow scalar values to be
+\NA and the type \raw does not allow {\NA}s.
 
 \paragraph{Unions} We support untagged unions of types written
-$\T_1 \k{|} \T_n$.
+$\T_1 \k{~|~} \T_n$.
 
 \paragraph{Nullables} 
-The type system has a \null type that is inhabited by a singleton
-\code{NULL} value. This value is often used as a sentinel or as a stand in
-for arguments that were not provided.  To capture this behavior, we
-introduce a nullable type \code{?}\T.  The difference between the \NA and
-\code{NULL} values is that \NA can be stored in primitive vectors.
+The type system has a \null type inhabited by a singleton \code{NULL}
+value. This value is often used as a sentinel or as a stand in for arguments
+that were not provided.  To capture this behavior, we introduce a nullable
+type \code{?}\T.  The difference between the \NA and \code{NULL} values is
+that \NA can be stored in primitive vectors.
 
@@
 
-\code{dplyr::group_indices}: \FUN{\CLASS{data.frame}, \VARGS}{\VEC\int}
-
-\code{lava::index}: \FUN{\CLASS{lvm} \; | \; \CLASS{lvmfit}, \VARGS}{\; ? \; \LIST{any}}
-
-\code{diptest::dip}: \FUN{\VEC\dbl, \chr \; | \; \lgl, \lgl, \dbl}{\CLASS{dip} \; | \; \dbl}
+\code{dplyr::group_indices}: \FUN{\CLASS{\texttt{data.frame}}, \VARGS}{\VEC\int}
+
+\code{lava::index}: \FUN{\CLASS{\texttt{lvm}} \; | \; \CLASS{\texttt{lvmfit}}, \VARGS}{\; ? \; \LIST{\any}}
+
+\code{diptest::dip}: \FUN{\VEC\dbl, \chr \; | \; \lgl, \lgl, \dbl}{\CLASS{\texttt{dip}} \; | \; \dbl}
 
@@
 
-\code{matrixcalc::matrix.power}: \FUN{\CLASS{matrix}, \dbl}{\CLASS{matrix}}
-
-$$type `data.tree::Traverse` \FUN{\CLASS{`Node`, `R6`}, ~\VEC\chr, \any, \any}\LIST\any;$$
-
-$$type `openssl::decrypt\_envelope` \FUN{~\VEC\raw, ~\VEC\raw, ~\VEC\raw, \CLASS{`key`, `rsa`}, \any}~\VEC\raw;$$
-
-$$dbplyr:type `set\_win\_current\_group` \FUN{? ~\VEC\chr}? ~\VEC\chr;$$
-
-$$type `sha256` \FUN{~\VEC\raw, ? ~\VEC\raw}~\VEC\raw;$$
-
-$$type `initparam` \FUN{? \dbl, \any, \any, \any, \chr, \chr, \lgl, ~\VEC\dbl, ~\VEC\dbl, \any}~\VEC\dbl;$$
+\code{matrixcalc::matrix.power}: \FUN{\CLASS{\texttt{matrix}}, \dbl}{\CLASS{\texttt{matrix}}}
+
+\code{data.tree::Traverse}: \FUN{\CLASS{\texttt{Node}, \texttt{R6}}, ~\VEC\chr, \any, \any}{\LIST{\any}}
+
+\code{openssl::decrypt\_envelope}: \FUN{~\VEC\raw, ~\VEC\raw, ~\VEC\raw, \CLASS{\texttt{key}, \texttt{rsa}}, \any}~\VEC\raw
+
+\code{dbplyr::set\_win\_current\_group} \FUN{? ~\VEC\chr}? ~\VEC\chr
+
+\code{openssl::sha256}: \FUN{~\VEC\raw, ? ~\VEC\raw}~\VEC\raw
+
+\code{forecast::initparam}: \FUN{? \dbl, \any, \any, \any, \chr, \chr, \lgl, ~\VEC\dbl, ~\VEC\dbl, \any}~\VEC\dbl
 

### 2020-05-16 paper/main.tex
reorg
subtype

@@
 \paragraph{Nullables} 
-The type system has a \null type inhabited by a singleton \code{NULL}
-value. This value is often used as a sentinel or as a stand in for arguments
-that were not provided.  To capture this behavior, we introduce a nullable
-type \code{?}\T.  The difference between the \NA and \code{NULL} values is
-that \NA can be stored in primitive vectors.
+The \null type is inhabited by a singleton \code{NULL} value often used as a
+sentinel or a stand in for arguments with no value.  To capture this
+behavior, we introduce a nullable type \code{?}\T.  The difference between
+the \NA and \code{NULL} values is that the latter cannot be stored in
+vectors.
 
 \paragraph{Lists} Heterogeneous collections are implemented using lists.
-Lists and vectors are closely related. Any vector can be converted to a list
-with \code{as.list}, and lists to vectors with \code{unlist} (coercions may
-ensue). We support parametrization of the list type, \LIST\T, but for
-heterogeneous lists this type will default to \any.
+Lists and vectors are closely related: a vector converts to a list with
+\code{as.list}, and lists to vectors with \code{unlist} (coercions may
+ensue). The list type is parameterize, \LIST\T, when the list is
+heterogeneous \T defaults to \any.
 
@@
 environment is performed in-place.  They are used to store variables and to
-escape from the copy-on-write semantics of other data types.  Environments
-are comprised of name, value pairs, and each environment has a reference to
-its parent environment.
-
-%
-%
+escape from the copy-on-write semantics of other data types.
+
 \section{Extracting and Checking Signatures}
@@
   $ \NAVEC{\S_1} \k{~|~}\dots \NAVEC{\S_n}$ &  \\
+  \\
+  \\
+  \NAVEC\S  & <:& \VEC\S \\
+  \T        & <:& \k{?}\T \\
+  \LIST{\T} & <:& \LIST{\T'} & \it{iff} \T <: \T' \\
+  \S        & <:& \VEC\S \\
+  \lgl      & <:& \int\\
+  \int      & <:& \dbl\\
+  \dbl      & <:& \clx\\  
 \end{tabular}
@@
 minimize their size by removing redundant types, types that are subsumed by
-subtyping, and \null types. Lists with more than 5 types in their union become \LIST{any}.
-We also only generate types for the first 20 arguments of a function.
+subtyping, and \null types. For performance reasons, we bound lists to no
+more than five types in their union, above that they become \LIST{any}.
+Functions are alos limited to twenty arguments.
 
@@
 \newcommand{\FIX}[1]{{\bf #1}\xspace}
+
+\code{dplyr::group_indices}: \FUN{\CLASS{\texttt{data.frame}}, \VARGS}{\VEC\int}
+
+\code{lava::index}: \FUN{\CLASS{\texttt{lvm}} \; | \; \CLASS{\texttt{lvmfit}}, \VARGS}{\; ? \; \LIST{\any}}
+
+\code{diptest::dip}: \FUN{\VEC\dbl, \chr \; | \; \lgl, \lgl, \dbl}{\CLASS{\texttt{dip}} \; | \; \dbl}
+
+\code{stabledist::cospi2}: \FUN{\VEC\dbl}{\VEC\dbl}
+
+\code{matrixcalc::matrix.power}: \FUN{\CLASS{\texttt{matrix}}, \dbl}{\CLASS{\texttt{matrix}}}
+
+\code{data.tree::Traverse}: \FUN{\CLASS{\texttt{Node}, \texttt{R6}}, ~\VEC\chr, \any, \any}{\LIST{\any}}
+
+\code{openssl::decrypt\_envelope}: \FUN{~\VEC\raw, ~\VEC\raw, ~\VEC\raw, \CLASS{\texttt{key}, \texttt{rsa}}, \any}{\VEC\raw}
+
+\code{dbplyr::set\_win\_current\_group} \FUN{? ~\VEC\chr}{\; ? \; \VEC\chr}
+
+\code{openssl::sha256}: \FUN{~\VEC\raw, ? ~\VEC\raw}{\VEC\raw}
+
+\code{forecast::initparam}: \FUN{? \dbl, \any, \any, \any, \chr, \chr, \lgl, ~\VEC\dbl, ~\VEC\dbl, \any}{\VEC\dbl}
 
@@
 
-\code{dplyr::group_indices}: \FUN{\CLASS{\texttt{data.frame}}, \VARGS}{\VEC\int}
-
-\code{lava::index}: \FUN{\CLASS{\texttt{lvm}} \; | \; \CLASS{\texttt{lvmfit}}, \VARGS}{\; ? \; \LIST{\any}}
-
-\code{diptest::dip}: \FUN{\VEC\dbl, \chr \; | \; \lgl, \lgl, \dbl}{\CLASS{\texttt{dip}} \; | \; \dbl}
-
-\code{stabledist::cospi2}: \FUN{\VEC\dbl}{\VEC\dbl}
-
-\code{matrixcalc::matrix.power}: \FUN{\CLASS{\texttt{matrix}}, \dbl}{\CLASS{\texttt{matrix}}}
-
-\code{data.tree::Traverse}: \FUN{\CLASS{\texttt{Node}, \texttt{R6}}, ~\VEC\chr, \any, \any}{\LIST{\any}}
-
-\code{openssl::decrypt\_envelope}: \FUN{~\VEC\raw, ~\VEC\raw, ~\VEC\raw, \CLASS{\texttt{key}, \texttt{rsa}}, \any}~\VEC\raw
-
-\code{dbplyr::set\_win\_current\_group} \FUN{? ~\VEC\chr}? ~\VEC\chr
-
-\code{openssl::sha256}: \FUN{~\VEC\raw, ? ~\VEC\raw}~\VEC\raw
-
-\code{forecast::initparam}: \FUN{? \dbl, \any, \any, \any, \chr, \chr, \lgl, ~\VEC\dbl, ~\VEC\dbl, \any}~\VEC\dbl
-
 \bibliography{bib/biblio,bib/jv,bib/r,bib/new,bib/gradual}

### 2020-05-16 paper/main.tex
Fix class name formatting

@@
 Table~\ref{tbl:fail-breakdown-0}.  
-Accounting for 27.37\% of assertion failures, the most common vilation arises when a value with \CLASS{matrix} is passed where a \VEC\dbl is expected.
+Accounting for 27.37\% of assertion failures, the most common vilation arises when a value with \CLASS{\texttt{matrix}} is passed where a \VEC\dbl is expected.
 Considering these types, we might imagine them to be compatible, however not allowing this coercion was a deliberate design decision:
@@
 \midrule
-\VEC\dbl & \CLASS{matrix} & 371,896 & 27.37 & 27.37\\
+\VEC\dbl & \CLASS{\texttt{matrix}} & 371,896 & 27.37 & 27.37\\
 \VEC\dbl & \VEC\int & 117,212 & 8.62 & 35.99\\
-\chr & \CLASS{bignum} | \VEC\raw & 100,100 & 7.37 & 43.36\\
+\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 100,100 & 7.37 & 43.36\\
 \dbl & \int | \null & 60,578 & 4.46 & 47.81\\
-\VEC\dbl & \CLASS{timeSeries} & 58,872 & 4.33 & 52.15\\
+\VEC\dbl & \CLASS{\texttt{timeSeries}} & 58,872 & 4.33 & 52.15\\
 \addlinespace
-\CLASS{matrix} & \CLASS{timeSeries} & 55,005 & 4.05 & 56.19\\
+\CLASS{\texttt{matrix}} & \CLASS{\texttt{timeSeries}} & 55,005 & 4.05 & 56.19\\
 \VEC\dbl & \dbl & 53,338 & 3.92 & 60.12\\
-\dbl & \CLASS{data.frame} & 32,553 & 2.40 & 62.51\\
-\VEC\dbl & \CLASS{data.frame} & 31,576 & 2.32 & 64.84\\
+\dbl & \CLASS{\texttt{data.frame}} & 32,553 & 2.40 & 62.51\\
+\VEC\dbl & \CLASS{\texttt{data.frame}} & 31,576 & 2.32 & 64.84\\
 \dbl & \int & 20,586 & 1.51 & 66.35\\
@@
 \VEC\dbl was passed to 15 different function parameters expecting a
-\CLASS{matrix}, of which there are 1394 in total: 1.08\% of \CLASS{matrix}-typed parameters
+\CLASS{\texttt{matrix}}, of which there are 1394 in total: 1.08\% of \CLASS{\texttt{matrix}}-typed parameters
 were passed \VEC\dbl values.
@@
 values.  Rows 5 and 6 are interesting: we see that rather often arguments
-expecting \CLASS{timeSeries} data are passed \VEC\dbl or \CLASS{matrix} values.  This
+expecting \CLASS{\texttt{timeSeries}} data are passed \VEC\dbl or \CLASS{\texttt{matrix}} values.  This
 is a quirk of the \code{timeSeries} package, whose functions often accept
@@
 \midrule
-\VEC\dbl & \CLASS{matrix} & 15 & 1394 & 1.08\\
+\VEC\dbl & \CLASS{\texttt{matrix}} & 15 & 1394 & 1.08\\
 \VEC\dbl & \VEC\int & 44 & 657 & 6.70\\
-\chr & \CLASS{bignum} | \VEC\raw & 1 & 3 & 33.33\\
+\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 1 & 3 & 33.33\\
 \dbl & \int | \null & 5 & 20 & 25.00\\
-\VEC\dbl & \CLASS{timeSeries} &19 & 35 & 54.29\\
+\VEC\dbl & \CLASS{\texttt{timeSeries}} &19 & 35 & 54.29\\
 \addlinespace
-\CLASS{matrix} & \CLASS{timeSeries} &  19 & 35 & 54.29\\
+\CLASS{\texttt{matrix}} & \CLASS{\texttt{timeSeries}} &  19 & 35 & 54.29\\
 \VEC\dbl & \dbl & 100 & 5021 & 1.99\\
-\dbl & \CLASS{data.frame} &  1 & 900 & 0.11\\
-\VEC\dbl & \CLASS{data.frame} &  6 & 900 & 0.67\\
+\dbl & \CLASS{\texttt{data.frame}} &  1 & 900 & 0.11\\
+\VEC\dbl & \CLASS{\texttt{data.frame}} &  6 & 900 & 0.67\\
 \dbl & \int & 48 & 529 & 9.07\\
@@
 Table~\ref{tbl:fail-breakdown-2} with a percentile of 80th percentile
-instead.  The most frequent argument type violation pattern in that of \CLASS{data.frame} values
- passed to arguments expecting tibbles, which are an extension of \CLASS{data.frame} built in the \code{tidyverse} package suite: 
+instead.  The most frequent argument type violation pattern in that of \CLASS{\texttt{data.frame}} values
+passed to arguments expecting tibbles, which are an extension of \CLASS{\texttt{data.frame}} built in the \code{tidyverse} package suite: 
  this occurs in only 22\%
@@
 \midrule
-\CLASS{data.frame} & \CLASS{data.frame, tbl, tbl\_df} & 63 & 284 & 22.18\\
-\CLASS{matrix} & \CLASS{array} & 13 & 70 & 18.57\\
+\CLASS{\texttt{data.frame}} & \CLASS{\texttt{data.frame}, \texttt{tbl}, \texttt{tbl\_df}} & 63 & 284 & 22.18\\
+\CLASS{\texttt{matrix}} & \CLASS{\texttt{array}} & 13 & 70 & 18.57\\
 \NAVEC\lgl & \lgl | \null & 9 & 49 & 18.37\\

### 2020-05-16 paper/main.tex
stuff

--- 2020-05-16 paper/main.tex
@@
 
+We ran \typetracer on the test, example, and vignette code of the aforementioned corpus of \PACKAGES packages and inferred types for \NUMFUNCTIONS functions.
+We have selected and will discuss a few interesting signatures.
+
+\code{dplyr::group_indices}: \FUN{\CLASS{\texttt{data.frame}}, \VARGS}{\VEC\int}
+
+\code{lava::index}: \FUN{\CLASS{\texttt{lvm}} \; | \; \CLASS{\texttt{lvmfit}}, \VARGS}{\; ? \; \LIST{\any}}
+
+\code{diptest::dip}: \FUN{\VEC\dbl, \chr \; | \; \lgl, \lgl, \dbl}{\CLASS{\texttt{dip}} \; | \; \dbl}
+
+\code{stabledist::cospi2}: \FUN{\VEC\dbl}{\VEC\dbl}
+
+\code{matrixcalc::matrix.power}: \FUN{\CLASS{\texttt{matrix}}, \dbl}{\CLASS{\texttt{matrix}}}
+
+\code{data.tree::Traverse}: \FUN{\CLASS{\texttt{Node}, \texttt{R6}}, ~\VEC\chr, \any, \any}{\LIST{\any}}
+
+\code{openssl::decrypt\_envelope}: \FUN{~\VEC\raw, ~\VEC\raw, ~\VEC\raw, \CLASS{\texttt{key}, \texttt{rsa}}, \any}{\VEC\raw}
+
+\code{dbplyr::set\_win\_current\_group} \FUN{? ~\VEC\chr}{\; ? \; \VEC\chr}
+
+\code{openssl::sha256}: \FUN{~\VEC\raw, ? ~\VEC\raw}{\VEC\raw}
+
+\code{forecast::initparam}: \FUN{? \dbl, \any, \any, \any, \chr, \chr, \lgl, ~\VEC\dbl, ~\VEC\dbl, \any}{\VEC\dbl}
+
+Many of the features of our type language are represented here.
+Some of these signatures are telling of the function's behaviour.
+For example, consider the signature for \code{openssl::decrypt_envelope}: the firs three parameters of the function are byte arrays, and the fourth argument is an RSA key, likely one used to decrypt some of the inputs, and the output of the function is another byte array.
+As another example, consider \code{data.tree::Traverse}: according to the function documentation, it takes the root of a tree and traverses it in an order specified by the second argument.
+
 \subsection{Expressiveness}

### 2020-05-16 paper/main.tex

--- 2020-05-16 paper/main.tex
@@
 more than five types in their union, above that they become \LIST{any}.
-Functions are alos limited to twenty arguments.
+Functions are alos limited to twenty arguments. Higher-order functions are
+approximated by \any to \any.
 
@@
 and restores them to their original state when it is unloaded.
-
 The type signatures can be provided externally, in a file, thus avoiding the
@@
 mechanism to register a hook for a custom \code{@type} tag. 
-
 The \contractr API allows users to explicitly insert contracts during
@@
 functions in this version of the tool.
-
 When argument values are wrapped in promises (this is not always the case
@@
 after that function has returned.
-
 Return values require care as well.  Functions return the last expression
@@
 
+We ran \typetracer on the test, example, and vignette code of the aforementioned corpus of \PACKAGES packages and inferred types for \NUMFUNCTIONS functions.
+We have selected and will discuss a few interesting signatures.
+
 \code{dplyr::group_indices}: \FUN{\CLASS{\texttt{data.frame}}, \VARGS}{\VEC\int}
@@
 \code{forecast::initparam}: \FUN{? \dbl, \any, \any, \any, \chr, \chr, \lgl, ~\VEC\dbl, ~\VEC\dbl, \any}{\VEC\dbl}
+
+Many of the features of our type language are represented here.
+Some of these signatures are telling of the function's behaviour.
+For example, consider the signature for \code{openssl::decrypt_envelope}: the firs three parameters of the function are byte arrays, and the fourth argument is an RSA key, likely one used to decrypt some of the inputs, and the output of the function is another byte array.
+As another example, consider \code{data.tree::Traverse}: according to the function documentation, it takes the root of a tree and traverses it in an order specified by the second argument.
 

### 2020-05-16 paper/main.tex
small discussion of types

@@
 Some of these signatures are telling of the function's behaviour.
-For example, consider the signature for \code{openssl::decrypt_envelope}: the firs three parameters of the function are byte arrays, and the fourth argument is an RSA key, likely one used to decrypt some of the inputs, and the output of the function is another byte array.
+For example, consider the signature for \code{openssl::decrypt_envelope}: the firs three parameters of the function are byte arrays, and the fourth argument is an RSA key, used to decrypt some of the inputs, and the output of the function is another byte array.
 As another example, consider \code{data.tree::Traverse}: according to the function documentation, it takes the root of a tree and traverses it in an order specified by the second argument.

### 2020-05-16 paper/main.tex
wird
Tabulate type signatures
numbers in conclusion

@@
 
-\section{Future Work}
-
-We see two major avenues for future work.
-
-\subsection{Types for R Objects}
-
-\subsection{Types for Data Frames}
-
 \section{Conclusion}
 
-\ldots
+Our goal with this work was to ...
+
+Overall, we found that our proposed design fits quite well with the existing language:
+Over 82\% of functions are either monomorphic or have only one single polymorphic argument, indicating that our design fits quite well with the language.
+When we tested the types inferred by \typetracer during our evaluation, we found that only \PERCFAILEDASSERTIONSWUNDEF of contract assertions failed, and found no pathological patterns in argument types with failed assertions.
+Furthermore, we found that our type language and contract checking framework could eliminate or otherwise simplify \CorpusPartiallyTypedAssertsRatio of existing type checks and assertions in user code.
 

### 2020-05-16 paper/main.tex
minor

@@
 \begin{wrapfigure}{r}{6.1cm}
-\footnotesize\begin{tabular}{l|l@{}}\hline
-\multicolumn{2}{l}{\bf Vectorized data types:}  \\\hline
+\footnotesize\begin{tabular}{l@{}l@{}}\hline
+\bf Vectors:\\\hline
 \k{logical}   & vector of boolean values\\
@@
 \k{list}      & vector of values of any type\\\hline
-\multicolumn{2}{l}{\bf Scalar data types:}\\\hline
+{\bf Scalars:}\\\hline
 \k{NULL}      &  singleton null value\\
@@
 \k{environment}& mapping from symbol to value \\\hline
-\multicolumn{2}{l}{\bf Implementation data types:}\\\hline
-\multicolumn{2}{l}{\k{special},
-\k{builtin},
-\k{symbol},
-\k{pairlist},
-\k{promise}}\\
-\multicolumn{2}{l}{
-\k{language},
-\k{char},
-\k{...},
-\k{any},
-\k{expression},
-}\\
-\multicolumn{2}{l}{
-\k{externalprt},
-\k{bytecode},
-\k{weakref}}\\\hline
+{\bf Implementation:}\\\hline
+\k{special},
+\k{builtin}, &\k{symbol}, \k{pairlist}, \k{promise}\\
+\k{language}, \k{char}, \k{...}, & \k{~any}, \k{expression},\\
+\k{externalprt},& \k{bytecode}, \k{weakref}\\\hline
 \end{tabular}\caption{Builtin Types}\label{rtypes}\end{wrapfigure}
@@
 
-\code{dplyr::group_indices}: \FUN{\CLASS{\texttt{data.frame}}, \VARGS}{\VEC\int}
-
-\code{lava::index}: \FUN{\CLASS{\texttt{lvm}} \; | \; \CLASS{\texttt{lvmfit}}, \VARGS}{\; ? \; \LIST{\any}}
-
-\code{diptest::dip}: \FUN{\VEC\dbl, \chr \; | \; \lgl, \lgl, \dbl}{\CLASS{\texttt{dip}} \; | \; \dbl}
-
-\code{stabledist::cospi2}: \FUN{\VEC\dbl}{\VEC\dbl}
-
-\code{matrixcalc::matrix.power}: \FUN{\CLASS{\texttt{matrix}}, \dbl}{\CLASS{\texttt{matrix}}}
-
-\code{data.tree::Traverse}: \FUN{\CLASS{\texttt{Node}, \texttt{R6}}, ~\VEC\chr, \any, \any}{\LIST{\any}}
-
-\code{openssl::decrypt\_envelope}: \FUN{~\VEC\raw, ~\VEC\raw, ~\VEC\raw, \CLASS{\texttt{key}, \texttt{rsa}}, \any}{\VEC\raw}
-
-\code{dbplyr::set\_win\_current\_group} \FUN{? ~\VEC\chr}{\; ? \; \VEC\chr}
-
-\code{openssl::sha256}: \FUN{~\VEC\raw, ? ~\VEC\raw}{\VEC\raw}
-
-\code{forecast::initparam}: \FUN{? \dbl, \any, \any, \any, \chr, \chr, \lgl, ~\VEC\dbl, ~\VEC\dbl, \any}{\VEC\dbl}
+\begin{table}[H]
+  \centering
+  \resizebox{.8\linewidth}{!}{
+  \begin{tabular}{llrrr}
+    \toprule
+    Function & Type Signature \\
+    \midrule
+    \code{dplyr::group_indices} & \FUN{\CLASS{\texttt{data.frame}}, \VARGS}{\VEC\int}\\
+    \code{moments::all.cumulants} & \FUN{\CLASS{\texttt{matrix}} \; | \; \VEC\dbl}{\CLASS{\texttt{matrix}}\; | \; \VEC{\dbl}}\\
+    \code{diptest::dip}& \FUN{\VEC\dbl, \chr \; | \; \lgl, \lgl,
+                         \dbl}{\CLASS{\texttt{dip}} \; | \; \dbl} \\
+    \code{stabledist::cospi2}& \FUN{\VEC\dbl}{\VEC\dbl} \\
+    \code{matrixcalc::matrix.power}& \FUN{\CLASS{\texttt{matrix}},
+                                     \dbl}{\CLASS{\texttt{matrix}}} \\
+    \code{data.tree::Traverse}& \FUN{\CLASS{\texttt{Node}, \texttt{R6}}, ~\VEC\chr,
+                                \any, \any}{\LIST{\any}} \\
+    \code{openssl::decrypt\_envelope}& \FUN{~\VEC\raw, ~\VEC\raw, ~\VEC\raw,
+                                       \CLASS{\texttt{key}, \texttt{rsa}}, \any}{\VEC\raw} \\
+    \code{dbplyr::set\_win\_current\_group}& \FUN{? ~\VEC\chr}{\; ? \; \VEC\chr} \\
+    \code{openssl::sha256}& \FUN{~\VEC\raw, ? ~\VEC\raw}{\VEC\raw} \\
+    \code{forecast::initparam}& \FUN{? \dbl, \any, \any, \any, \chr, \chr, \lgl,
+                                ~\VEC\dbl, ~\VEC\dbl, \any}{\VEC\dbl} \\
+    \bottomrule
+  \end{tabular}}
+\caption{Select Type Signatures}\label{tbl:typesig}
+\end{table}
 
@@
 
-In the last section, we evaluated our type annotation language based on a
-set of function types that were automatically generated by \typetracer.
-Now, we're interested in knowing how representative our trace-generated
-types are of the functions they were generated for.  To measure this, we
-conducted a large-scale experiment: for each package in the corpus discussed
-in Section~\ref{sec:corpus}, we ran \typetracer on its test, example, and
-vignette code, obtaining types for package functions. Then, we
-ran the test, example, and vignette code of each of the package's reverse
-dependencies (or clients).  We collected failed contract assertions in order
-to analyze them and perhaps identify weaknesses in our type annotation
-language.
-
-We ran our evaluation on \NUMPKGSEVAL packages and recorded \NUMASSERTIONS
-total assertions.  Overall, we found that only \PERCFAILEDASSERTIONSWUNDEF
-of contract assertions failed.  We made a simplification in our analysis,
-capping the number of function arguments we generate types for to 20, and
-this resulted in only \PERCASSERTIONSUNDEF failed assertions, which suggests
-that the simplification we made did not greatly affect the results.  Once we
-controlled for these, we found that \PERCFAILEDASSERTIONS of all assertions
-failed for reasons related to our generated types.  Additionally, we found
-that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of functions types
-were never violated.  The number of immaculate function types increases to
-\PERCSUCCFUNNOSTHREE if we discount S3 functions: These S3 functions
-represent user-defined polymorphism which we do not tackle in this work.
+How robust are the inferred types?  To measure this, we conducted another
+large-scale experiment: for each package in the corpus, we ran used the
+inferred type signatures as contracts and ran all of the CRAN reverse
+dependencies for that package. In total we ran \NUMPKGSEVAL packages and
+recorded \NUMASSERTIONS total assertions.  Overall, we found that only
+\PERCFAILEDASSERTIONSWUNDEF of contract failed.  The limit on number of
+arguments (we record only 20) accounted for \PERCASSERTIONSUNDEF failed
+assertion. We found that \PERCSUCCARG of parameter types and \PERCSUCCFUNS
+of functions types never failed.  The number of immaculate function types
+increases to \PERCSUCCFUNNOSTHREE if we discount S3 object method dispatch.
 Overall, these numbers are promising, and suggest that the type signatures
-generated by running \typetracer on our corpus are indeed representative,
-and capture intended function behaviour.
+are robust.
 
 We breakdown the failed assertions by type in
-Table~\ref{tbl:fail-breakdown-0}.  
-Accounting for 27.37\% of assertion failures, the most common vilation arises when a value with \CLASS{\texttt{matrix}} is passed where a \VEC\dbl is expected.
-Considering these types, we might imagine them to be compatible, however not allowing this coercion was a deliberate design decision:
+Table~\ref{tbl:fail-breakdown-0}.  Accounting for 27\% of assertion
+failures, is when \CLASS{\texttt{matrix}} is passed where a \VEC\dbl is
+expected.  Considering these types, we might imagine them to be compatible,
+however not allowing this coercion was a deliberate design decision:
 coercion from vectors to matrices is ad hoc at best, and surely not a
@@
 % BEGIN Autogenerated
-
+\small
 \begin{tabular}{llrrr}
@@
 % END Autogenerated
-\label{tbl:fail-breakdown-0}
-\caption{Top types of contract assertion failures.}
-\end{table}
-
-In addition to the raw number of failing contract checks, we were interested
-in how many functions had a parameter where a contract check failed, and
+\label{tbl:fail-breakdown-0}\caption{Top contract failures}\end{table}
+
+In addition to the number of failed contract checks, we were interested in
+how many functions had a parameter where a contract check failed, and
 overall we found this to be the case in \PROPFUNSFAILEDCHECK of functions.
 Digging a little deeper, we noticed that many of these functions were
-performing S3 dispatch, which is part of R's object-orientation framework:
-thus, these functions implement user-defined polymorphism which we do not
-handle in our type framework.  Removing those functions, we see that
-\PROPFUNSFAILEDCHECKNOSTHREE of functions exhibited failing contract checks.
-These are functions which were under-tested, as calls to these functions
-represent only \PERCCALLSBADFUNSINTESTS of recorded calls during
-\typetracer's run on the core corpus to infer types.
+performing S3 dispatch which we do not handle.  Removing those functions, we
+see that \PROPFUNSFAILEDCHECKNOSTHREE of functions exhibited failing
+contract checks.  These are functions were likely under-tested, as calls to
+these functions represent only \PERCCALLSBADFUNSINTESTS of recorded calls
+during \typetracer's run on the core corpus to infer types.
 
 Turning our attention now to arguments, we found that only
-\PROPARGSFAILINGASSERTS of function parameters exhibited a failed assertion.
-Table~\ref{tbl:fail-breakdown-0} showed the raw results from counting
-runtime occurrences, but that data alone does not tell the full story, as
-some failures may be overrepresented if e.g. a failing contract assertion
-was in a loop.  We were interested in knowing for each of the most common
-violations in Table~\ref{tbl:fail-breakdown-0}, how many different arguments
-had that type, and how many of those exhibited the contract failure in
-question.  Table~\ref{tbl:fail-breakdown-1} breaks down the failed assertions by type,
-folding away multiple identical failed contract assertions for the same parameter
-position.  The first row of this table reads: a value of type
+\PROPARGSFAILINGASSERTS of parameters failed.
+Table~\ref{tbl:fail-breakdown-0} showed the runtime occurrences, but that
+data alone does not tell the full story, as some failures may be
+overrepresented if e.g. a failing contract assertion was in a loop.  We were
+interested in knowing for each of the most common violations in
+Table~\ref{tbl:fail-breakdown-0}, how many different arguments had that
+type, and how many of those exhibited the contract failure in question.
+Table~\ref{tbl:fail-breakdown-1} breaks down the failed assertions by type,
+folding away multiple identical failed contract assertions for the same
+parameter position.  The first row of this table reads: a value of type
 \VEC\dbl was passed to 15 different function parameters expecting a
-\CLASS{\texttt{matrix}}, of which there are 1394 in total: 1.08\% of \CLASS{\texttt{matrix}}-typed parameters
-were passed \VEC\dbl values.
+\CLASS{\texttt{matrix}}, of which there are 1394 in total: 1\% of
+\CLASS{\texttt{matrix}}-typed parameters were passed \VEC\dbl values.
 
@@
 is similar with the double and integer coercion we mentioned earlier: it
-represents many dynamic contract failures, but very few of the \VEC\int-typed 
-arguments have their contracts violated by \VEC\dbl-typed
+represents many dynamic contract failures, but very few of the
+\VEC\int-typed arguments have their contracts violated by \VEC\dbl-typed
 values.  Rows 5 and 6 are interesting: we see that rather often arguments
-expecting \CLASS{\texttt{timeSeries}} data are passed \VEC\dbl or \CLASS{\texttt{matrix}} values.  This
-is a quirk of the \code{timeSeries} package, whose functions often accept
-matrices and vectors, converting them to time series in an ad hoc manner.
-Note that the code coverage of the \code{timeSeries} tests, examples, and
-vignettes on the \code{timeSeries} package code is only 58.24\%, which is
-one possible explanation of why these contract failures are occurring: the
-types that \typetracer generates are only as good as the test code its run
-on.
+expecting \CLASS{\texttt{timeSeries}} data are passed \VEC\dbl or
+\CLASS{\texttt{matrix}} values.  This is a quirk of the \code{timeSeries}
+package, whose functions often accept matrices and vectors, converting them
+to time series in an ad hoc manner.  Note that the code coverage of the
+\code{timeSeries} tests, examples, and vignettes on the \code{timeSeries}
+package code is only 58\%, which is one possible explanation of why these
+contract failures are occurring: the types that \typetracer generates are
+only as good as the test code its run on.
 
@@
  \label{tbl:fail-breakdown-1}
- \caption{Results from Table~\ref{tbl:fail-breakdown-0}, broken down by occurrences of the expected type as a parameter type.}
+ \caption{Results from Table~\ref{tbl:fail-breakdown-0}, broken down by occurrences of the expected type as a parameter type}
  \end{table}
@@
 type of the value that violated that contract.  The first row of the table
-reads: 44 function arguments with \VEC\int type are passed \VEC\dbl  instead, and 657 arguments have \VEC\int type.  Had we failed
-to capture some key usage pattern of R with our type annotation framework,
-we would likely see it here, and we can see this in action if we consider
+reads: 44 function arguments with \VEC\int type are passed \VEC\dbl instead,
+and 657 arguments have \VEC\int type.  Had we failed to capture some key
+usage pattern of R with our type annotation framework, we would likely see
+it here, and we can see this in action if we consider
 Table~\ref{tbl:fail-breakdown-3}, which was obtained identically to
 Table~\ref{tbl:fail-breakdown-2} with a percentile of 80th percentile
-instead.  The most frequent argument type violation pattern in that of \CLASS{\texttt{data.frame}} values
-passed to arguments expecting tibbles, which are an extension of \CLASS{\texttt{data.frame}} built in the \code{tidyverse} package suite: 
- this occurs in only 22\%
-of such arguments, and represents cases where tests did not adequately cover
-all valid function inputs.  Separate from the issue of testing, we can
-capture this behaviour with user-defined subtyping, as data frames are
-subtypes of tibbles, but this necessitates a proper object-oriented type
-system for R which is outside the scope of this work.
+instead.  The most frequent argument type violation pattern in that of
+\CLASS{\texttt{data.frame}} values passed to arguments expecting tibbles,
+which are an extension of \CLASS{\texttt{data.frame}} built in the
+\code{tidyverse} package suite: this occurs in only 22\% of such arguments,
+and represents cases where tests did not adequately cover all valid function
+inputs.  Separate from the issue of testing, we can capture this behaviour
+with user-defined subtyping, as data frames are subtypes of tibbles, but
+this necessitates a proper object-oriented type system for R which is
+outside the scope of this work.
 
  \begin{table}
-% BEGIN Autogenerated
-
+% BEGIN Autogenerated\
+\small
 \begin{tabular}{llrrr}
@@
  \label{tbl:fail-breakdown-2}
- \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 90th percentile.}
+ \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 90th percentile}
  \end{table}
@@
 % BEGIN Autogenerated
-
-\begin{tabular}{llrrr}
+\small
+\begin{tabular}{ll@{}rrr@{}}
 \toprule
-Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
+Passed & Arg Type & \#Args Fail & \#Args w Type & \% Failure\\
 \midrule
@@
  \label{tbl:fail-breakdown-3}
- \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 80th percentile.}
+ \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 80th percentile}
  \end{table}

### 2020-05-16 paper/main.tex
minor

@@
  \label{tbl:fail-breakdown-1}
- \caption{Results from Table~\ref{tbl:fail-breakdown-0}, broken down by occurrences of the expected type as a parameter type}
+ \caption{Results from Table 4, broken down by occurrences of the expected type as a parameter type}
  \end{table}
@@
 Overall, we found that our proposed design fits quite well with the existing language:
-Over 82\% of functions are either monomorphic or have only one single polymorphic argument, indicating that our design fits quite well with the language.
+Over 82\% of functions are either monomorphic or have only one single polymorphic argument.
 When we tested the types inferred by \typetracer during our evaluation, we found that only \PERCFAILEDASSERTIONSWUNDEF of contract assertions failed, and found no pathological patterns in argument types with failed assertions.

### 2020-05-16 paper/main.tex
Add higher-order functions
Add related work

@@
 quantify how much of a code base can be type-checked.  Other approaches that
-infer types for dynamic analysis include the work of \citet{FurrAF2009} for
-Ruby.
+infer types for dynamic analysis include the work of \citet{FurrAF2009} and
+\citet{hicks2011ruby} for Ruby.
 
@@
 
+\paragraph{Higher-Order Functions}
+\typetracer assigns the type \CLASS{\texttt{function}} to function values.
+The number of positions that receive a function is 1,831, which is just
+2\% of all the positions for which we infer types. Given the small number of
+occurrences, it is not worth complicating the inferred types with a complete
+signature for these functions.
+
 \paragraph{Structs}

### 2020-05-16 paper/main.tex
conc wip

--- 2020-05-16 paper/main.tex
@@
 
-Our goal with this work was to ...
-
-Overall, we found that our proposed design fits quite well with the existing language:
+Our ultimate goal is to design a type system for R which accounts for the language's intricacies and eccentricities whilst being usable and desirable for R programmers.
+This is no small feat, as our ability to statically make sense of R code is complicated by a language design which fosters an interactive programming style by minimizing programmer overhead.
+Designing a type system which accounts for language features such as reflective table manipulation, extensive type reflection, and ad hoc polymorphism, as well as consolidating R's many disparate object systems is sure to be a long road.
+
+This work focuses on a simpler problem: instead of an entire type system, we limited the scope of our investigation to ascribing types to function signatures.
+To this end, we designed a simple type language which found a compromise between simplicity and usefulness.
+Our language focuses on the most widely used features of R and omits features such as parametric polymorphism and user-defined subtyping.
+We presented \typetracer, a tool for inferring types for function signatures from runnable code, and \contractr, an easy-to-use package for R which allows users to specify function type signatures and have function arguments checked for compliance at runtime.
+
+We evaluated our design by running \typetracer on a corpus of 412 of the most widely used R packages on CRAN, inferring signatures for exported functions, and testing those inferred signatures on the \NUMPKGSEVAL reverse dependencies of the corpus.
+Overall, we found that our simple design fits quite well with the existing language:
 Over 82\% of functions are either monomorphic or have only one single polymorphic argument.
 When we tested the types inferred by \typetracer during our evaluation, we found that only \PERCFAILEDASSERTIONSWUNDEF of contract assertions failed, and found no pathological patterns in argument types with failed assertions.
-Furthermore, we found that our type language and contract checking framework could eliminate or otherwise simplify \CorpusPartiallyTypedAssertsRatio of existing type checks and assertions in user code.
+Furthermore, we found that our type language and contract checking framework would be useful to programmers, eliminate or otherwise simplify \CorpusPartiallyTypedAssertsRatio of existing type checks and assertions in user code.
 

### 2020-05-16 paper/main.tex
minor

@@
 
-Our ultimate goal is to design a type system for R which accounts for the language's intricacies and eccentricities whilst being usable and desirable for R programmers.
-This is no small feat, as our ability to statically make sense of R code is complicated by a language design which fosters an interactive programming style by minimizing programmer overhead.
-Designing a type system which accounts for language features such as reflective table manipulation, extensive type reflection, and ad hoc polymorphism, as well as consolidating R's many disparate object systems is sure to be a long road.
-
+Designing a type system for R is no small feat, as statically make sense of R code is complicated by a language design which fosters an interactive programming style by minimizing programmer overhead.
 This work focuses on a simpler problem: instead of an entire type system, we limited the scope of our investigation to ascribing types to function signatures.

### 2020-05-16 paper/main.tex
conc

@@
 
-Designing a type system for R is no small feat, as statically make sense of R code is complicated by a language design which fosters an interactive programming style by minimizing programmer overhead.
-This work focuses on a simpler problem: instead of an entire type system, we limited the scope of our investigation to ascribing types to function signatures.
+Designing a type system for R is no small feat, as statically making sense of R code is complicated by a language design which fosters an interactive programming style by minimizing programmer overhead.
+Our intent is to eventually propose changes to the language, but we are aware that for any changes to be accepted by the community, they must have clear benefits without endangering backwards compatibility.
+As a step towards this, we focus on a simpler problem: instead of an entire type system, we limited the scope of our investigation to ascribing types to function signatures.
 To this end, we designed a simple type language which found a compromise between simplicity and usefulness.

### 2020-05-16 paper/main.tex
Add first conclusion para

@@
 \end{document}
+
+Retrofitting a type system for the interactive and exploratory programming style
+of R is hard. It is even challenging to gain acceptance by the user community.
+To this end, we have begun our journey with a simple type language, one that we
+believe captures the essence of R while remaining simple and useful. We focus on
+the basic language features for generating the type signatures, and find that

### 2020-05-16 paper/main.tex
merge

--- 2020-05-16 paper/main.tex
@@
 As a step towards this, we focus on a simpler problem: instead of an entire type system, we limited the scope of our investigation to ascribing types to function signatures.
-To this end, we designed a simple type language which found a compromise between simplicity and usefulness.
-Our language focuses on the most widely used features of R and omits features such as parametric polymorphism and user-defined subtyping.
+To this end, we designed a simple type language which found a compromise between simplicity and usefulness by focusing  on the most widely used features of R.
 We presented \typetracer, a tool for inferring types for function signatures from runnable code, and \contractr, an easy-to-use package for R which allows users to specify function type signatures and have function arguments checked for compliance at runtime.

### 2020-05-16 paper/main.tex
Add conclusion
merge
minorconc

--- 2020-05-16 paper/main.tex
@@
 
-Designing a type system for R is no small feat, as statically making sense of R code is complicated by a language design which fosters an interactive programming style by minimizing programmer overhead.
+Retrofitting a type system for the interactive and exploratory programming style of R is hard: The language is poorly specified and builds upon an eclectic mix of features such as laziness, reflection, dynamic evaluation and ad-hoc object systems.
 Our intent is to eventually propose changes to the language, but we are aware that for any changes to be accepted by the community, they must have clear benefits without endangering backwards compatibility.

### 2020-05-16 paper/main.tex
minorconc

@@
 Retrofitting a type system for the interactive and exploratory programming style of R is hard: The language is poorly specified and builds upon an eclectic mix of features such as laziness, reflection, dynamic evaluation and ad-hoc object systems.
-Our intent is to eventually propose changes to the language, but we are aware that for any changes to be accepted by the community, they must have clear benefits without endangering backwards compatibility.
+Our intent is to eventually propose a type system for inclusion in the language, but we are aware that for any changes to be accepted by the community, they must have clear benefits without endangering backwards compatibility.
 As a step towards this, we focus on a simpler problem: instead of an entire type system, we limited the scope of our investigation to ascribing types to function signatures.
@@
 Furthermore, we found that our type language and contract checking framework would be useful to programmers, eliminate or otherwise simplify \CorpusPartiallyTypedAssertsRatio of existing type checks and assertions in user code.
+In sum, we believe that our simple type language design is a solid foundation for the eventual type system for R.
 

### 2020-05-16 paper/main.tex
related work

@@
 
+Recent work has gone into adding types to Python, the other eminent data science language.
+Typilus is an interesting piece of recent work which explores using neural networks to infer types for Python programs~\cite{allamanis2020typilus}, and Python itself added support for type hints in Python 3.5~\cite{pythonspectypes}.
 There is no previous work on types for the R programming language. We take

### 2020-05-16 paper/main.tex
Typos

@@
 
-Our community builds, improves, and reasons about programming languages.  To
-make design decisions that benefit users, we need to understand our target
-language as well as the real-world needs it answers. Often, we can appeal to
-our intuition, as many languages are intended for general purpose
-programming tasks. Unfortunately, intuition may fail when looking at
-domain-specific languages as these languages designed for a specific group
-of users to solve very specific needs. This is the case of the data science
-language R.
+Our community builds, improves, and reasons about programming languages. To make
+design decisions that benefit users, we need to understand our target language
+as well as the real-world needs it answers. Often, we can appeal to our
+intuition, as many languages are intended for general purpose programming tasks.
+Unfortunately, intuition may fail when looking at domain-specific languages as
+these languages were designed for a specific group of users to solve very
+specific needs. This is the case of the data science language R.
 
@@
 Many of the design decisions that gave us R were intended to foster an
-interactive, exploratory, programming style. These include, to name a few,
+interactive and exploratory programming style. These include, to name a few,
 the lack of type annotations, the ability to use syntactic shortcut, and the
@@
 for a language as baroque as R. For example, one of the most popular data
-type, the \code{data.frame}, is manipulated through reflective operations --
-a data frame is a table whose columns can be added or removed on the fly.
+type, the \code{data.frame}, is manipulated through reflective operations---a data frame is a table whose columns can be added or removed on the fly.
 Second, but just as crucially, designing a type system that will be adopted
@@
 One way to validate the inferred type annotations is to check that different
-programs using one of the functions for which a type signature has been
-inferred respect the signature.  \contractr is an R package that allows to
-decorate function with assertions. We use it to insert type checks of the
-arguments and return values. As usual with R, this is not straightforward.
+programs using one of the functions for which a type signature has been inferred
+respect the signature. For this purpose, we developed \contractr, an R package
+that allows one to decorate function with assertions. We use it to insert type
+checks of the arguments and return values. As usual with R, this is not
+straightforward.
 
 While \contractr's primary logic has been implemented in C++ to reduce its
-runtier overhead, we have not observed a single segfault during our use of
+runtime overhead, we have not observed a single segfault during our use of
 the package.  It has been tested with GNU R-3.5.0 and hardened with a
@@
 \AllCranRnd packages, we managed to install \AllLoadableRnd. The main reason for
-this is that \rdt is based on R 3.5.0 and some of the packages are no longer
+this is that \rdt is based on GNU-R 3.5.0 and some of the packages are no longer
 compatible with this version. Some packages also require extra native
@@
 How robust are the inferred types? To measure this, we conducted another
-large-scale experiment: for each package in the corpus, we ran used the
-inferred type signatures as contracts and ran all of the CRAN reverse
-dependencies for that package. In total we ran \NUMPKGSEVAL packages and
-recorded \NUMASSERTIONS total assertions.  Overall, we found that only
-\PERCFAILEDASSERTIONSWUNDEF of contract failed.  The limit on number of
-arguments (we record only 20) accounted for \PERCASSERTIONSUNDEF failed
-assertion. We found that \PERCSUCCARG of parameter types and \PERCSUCCFUNS
-of functions types never failed.  The number of immaculate function types
-increases to \PERCSUCCFUNNOSTHREE if we discount S3 object method dispatch.
-Overall, these numbers are promising, and suggest that the type signatures
-are robust.
+large-scale experiment: for each package in the corpus, using the inferred type
+signatures as contracts we ran all of the CRAN reverse dependencies for that
+package. In total we ran \NUMPKGSEVAL unique packages and recorded
+\NUMASSERTIONS total assertions. Overall, we found that only
+\PERCFAILEDASSERTIONSWUNDEF of contract failed. The limit on number of arguments
+(we record only 20) accounted for \PERCASSERTIONSUNDEF failed assertion. We
+found that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of functions types
+never failed. The number of immaculate function types increases to
+\PERCSUCCFUNNOSTHREE if we discount S3 object method dispatch. Overall, these
+numbers are promising, and suggest that the type signatures are robust.
 
@@
 % END Autogenerated
-\label{tbl:fail-breakdown-0}\caption{Top contract failures}\end{table}
+\caption{Top contract failures}
+\label{tbl:fail-breakdown-0}
+\end{table}
 
@@
 % END Autogenerated
+ \caption{Results from Table 4, broken down by occurrences of the expected type as a parameter type}
  \label{tbl:fail-breakdown-1}
- \caption{Results from Table 4, broken down by occurrences of the expected type as a parameter type}
  \end{table}
@@
 % END Autogenerated
+ \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 90th percentile}
  \label{tbl:fail-breakdown-2}
- \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 90th percentile}
  \end{table}
@@
 % END Autogenerated
+ \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 80th percentile}
  \label{tbl:fail-breakdown-3}
- \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 80th percentile}
  \end{table}

### 2020-05-16 paper/main.tex
prefinal
minor

@@
 Our community builds, improves, and reasons about programming languages. To make
-design decisions that benefit users, we need to understand our target language
+design decisions that benefit users, we need to understand the target language
 as well as the real-world needs it answers. Often, we can appeal to our
@@
 intersperse typed and untyped code in a fine-grained manner.
+Other work looks at retrofitting types onto Lua~\cite{typedlua2014}.
 
@@
 that allows one to decorate function with assertions. We use it to insert type
-checks of the arguments and return values. As usual with R, this is not
-straightforward.
-
-While \contractr's primary logic has been implemented in C++ to reduce its
-runtime overhead, we have not observed a single segfault during our use of
+checks of the arguments and return values.
+
+\contractr's primary logic has been implemented in C++ to reduce its
+runtime overhead, and we have not observed a single segfault during our use of
 the package.  It has been tested with GNU R-3.5.0 and hardened with a

### 2020-05-16 paper/main.tex
minor wording, specifics

@@
 overall we found this to be the case in \PROPFUNSFAILEDCHECK of functions.
-Digging a little deeper, we noticed that many of these functions were
-performing S3 dispatch which we do not handle.  Removing those functions, we
-see that \PROPFUNSFAILEDCHECKNOSTHREE of functions exhibited failing
-contract checks.  These are functions were likely under-tested, as calls to
+To subdivide this number, we discounted functions that were performing S3
+dispatch, as they exhibit user-defined polymorphism which we do not handle.
+Removing those functions, we see that the proportion of functions with failed contract checks falls to 
+\PROPFUNSFAILEDCHECKNOSTHREE. These remaining functions were under-tested, as calls to
 these functions represent only \PERCCALLSBADFUNSINTESTS of recorded calls

### 2020-05-16 paper/main.tex
minor wording

@@
 Some of these signatures are telling of the function's behaviour.
-For example, consider the signature for \code{openssl::decrypt_envelope}: the firs three parameters of the function are byte arrays, and the fourth argument is an RSA key, used to decrypt some of the inputs, and the output of the function is another byte array.
+For example, consider the signature for \code{openssl::decrypt_envelope}: the first three parameters of the function are byte arrays, and the fourth argument is an RSA key, used to decrypt some of the inputs, and the output of the function is the decrypted byte array.
 As another example, consider \code{data.tree::Traverse}: according to the function documentation, it takes the root of a tree and traverses it in an order specified by the second argument.
+We see that reflected in the type, where the first argument has type \CLASS{\texttt{Node}, \texttt{R6}} and the second argument had type \VEC\chr, representing the traversal order.
 

### 2020-05-16 paper/main.tex
Updated 4.3

@@
 
-In terms of usability, our checker is enabled automatically whenever a
-package is loaded in R. So, a simple invocation of \code{library(contractr)}
-causes contracts to be injected in all packages.  On loading, \contractr
-scans all packages in the user's workspace and inserts contracts in
-functions for which type signatures are available. Then \contractr sets up
-package load hooks which are executed when new packages are loaded.
-Furthermore, \contractr automatically removes contracts from all functions
-and restores them to their original state when it is unloaded.
-The type signatures can be provided externally, in a file, thus avoiding the
-need to change the source code of checked packages.  Type declarations can
-also be written alongside a \code{@type} section inside \roxygen function
-comment blocks. \roxygen is an R package that enables authors to add
-documentation. The documentation sections are processed by \roxygen before
-package build and a registered method for each custom tag section is
-invoked with the section data and documented object. \contractr uses this
-mechanism to register a hook for a custom \code{@type} tag. 
-The \contractr API allows users to explicitly insert contracts during
-interactive development by supplying the type signature as a string.
-Contracts can be removed by calling \code{remove\_contract}.  Contracts can
-be selectively enabled or disabled for code blocks by wrapping them in calls
-to \code{ignore\_contracts} and \code{capture\_contracts}.  These functions
-return a data frame that contains all the information about failed and
-successful contract assertions in the wrapped code blocks.  Multiple
-failures can introduce noise; this is avoided by setting
-\code{contractr.severity} to \code{'silence'}.  \contractr still performs
-type-checking so that results can be obtained by calling
-\code{get\_contracts}. Setting \code{contractr.severity} flag to
-\code{'error'} turns the type-checking warnings to errors and
-halt the program.
+In terms of usability, our checker is enabled automatically whenever a package
+is loaded in R. So, a simple invocation of \code{library(contractr)} causes
+contracts to be injected in all packages. On loading, \contractr scans all
+packages in the user's workspace and inserts contracts in functions for which
+type signatures are available. Then \contractr sets up package load hooks which
+are executed when new packages are loaded. Furthermore, \contractr automatically
+removes contracts from all functions and restores them to their original state
+when it is unloaded. The type signatures can be provided externally, in a file,
+thus avoiding the need to change the source code of checked packages. Type
+declarations can also be written directly in the adjunct function comments using
+\roxygen annotation tags. \contractr introduces a new use \code{@type} tag. When
+a package these type signatures are extracted and stored in the appropriate
+location. The following is an example of a type annotation using \roxygen:
+
+\begin{lstlisting}
+#' @type <chr> => int
+#' @export
+file_size <- function(f) { ... }
+\end{lstlisting}
+
+The \contractr provides an extensive API that allows users to explicitly insert
+contracts during interactive development by supplying the type signature as a
+string, selectively disable contract checking and inspecting type assertions.
+
+% Contracts can be removed by calling \code{remove\_contract}.  Contracts can
+% be selectively enabled or disabled for code blocks by wrapping them in calls
+% to \code{ignore\_contracts} and \code{capture\_contracts}.  These functions
+% return a data frame that contains all the information about failed and
+% successful contract assertions in the wrapped code blocks.  Multiple
+% failures can introduce noise; this is avoided by setting
+% \code{contractr.severity} to \code{'silence'}.  \contractr still performs
+% type-checking so that results can be obtained by calling
+% \code{get\_contracts}. Setting \code{contractr.severity} flag to
+% \code{'error'} turns the type-checking warnings to errors and
+% halt the program.
 

### 2020-05-16 paper/main.tex
stuff

@@
 
-Our community builds, improves, and reasons about programming languages. To make
-design decisions that benefit users, we need to understand the target language
-as well as the real-world needs it answers. Often, we can appeal to our
-intuition, as many languages are intended for general purpose programming tasks.
-Unfortunately, intuition may fail when looking at domain-specific languages as
-these languages were designed for a specific group of users to solve very
-specific needs. This is the case of the data science language R.
+Our community builds, improves, and reasons about programming languages. To
+make design decisions that benefit users, we need to understand our target
+language as well as the real-world needs it answers. Often, we can appeal to
+our intuition, as many languages are intended for general purpose
+programming tasks.  Unfortunately, intuition may fail when looking at
+domain-specific languages as these languages were designed for a specific
+group of users to solve very specific needs. This is the case of the data
+science language R.
 
@@
 
-Retrofitting a type system to the R programming language would increase our
-assurance in the result of data analysis. But, we are faced with two
-challenges. First, it is unclear what would be the \emph{right} type system
-for a language as baroque as R. For example, one of the most popular data
-type, the \code{data.frame}, is manipulated through reflective operations---a data frame is a table whose columns can be added or removed on the fly.
-Second, but just as crucially, designing a type system that will be adopted
-would require overcoming some prejudices and educating large numbers of
-users.
+Retrofitting a type system to the R language would increase our assurance in
+the result of data analysis. But, we are faced with two challenges. First,
+it is unclear what would be the \emph{right} type system for a language as
+baroque as R. For example, one of the most popular data type, the
+\code{data.frame}, is manipulated through reflective operations---a data
+frame is a table whose columns can be added or removed on the fly.  Second,
+but just as crucially, designing a type system that will be adopted would
+require overcoming some prejudices and educating large numbers of users.
 
@@
 backwards compatibility. Our goal is thus to find a compromise between
-simplicity and usefulness; any proposed type system should cover common
-programming idioms while remaining easy to learn and to use.
+simplicity and usefulness; any proposal should cover common programming
+idioms while remaining easy to learn and to use.
 
 This paper focuses on a simpler problem than an entire type system, instead,
-we limit the scope of our investigation to giving types to function
-signatures. In order to do this, we design a simple type language, one that
-matches the R data types but omits features such as parametric polymorphism
-and subtyping between user-defined data types. We then extract type
-signature from execution traces of a corpus of widely used libraries.  This
-allows us to see how far one can get with the simple type language and
-identify limitations of our this particular design. We validate the
-robustness of the extracted type signature by implementing a contact system
-that weaves the types around their respective functions, and use a large
-number of clients of the target packages for validation. The contract system
-can be used by the R community to experiment with our type signatures and to
+we limit our scope to giving types to function signatures. For this, we
+design a simple type language, one that matches the R data types but omits
+features such as parametric polymorphism and subtyping between user-defined
+data types. We then extract type signature from execution traces of a corpus
+of widely used libraries.  This allows us to see how far one can get with
+the simple type language and identify limitations of the design. We validate
+the robustness of the extracted type signature by implementing a contact
+checker that weaves the types around their respective functions, and use a
+large number of clients of the target packages for validation. The contract
+system can be used by the R community to experiment with type signatures and
 replace ad hoc error checking code.
 
-To sum up, our paper makes the following main contributions:
+To sum up, our paper makes the following contributions:
 \begin{itemize}
@@
 object-oriented features of PHP and JavaScript, respectively. They allow to
-intersperse typed and untyped code in a fine-grained manner.
-Other work looks at retrofitting types onto Lua~\cite{typedlua2014}.
+intersperse typed and untyped code in a fine-grained manner.  Other work
+looks at retrofitting types onto Lua~\cite{typedlua2014}.
 
@@
 
-Recent work has gone into adding types to Python, the other eminent data science language.
-Typilus is an interesting piece of recent work which explores using neural networks to infer types for Python programs~\cite{allamanis2020typilus}, and Python itself added support for type hints in Python 3.5~\cite{pythonspectypes}.
-There is no previous work on types for the R programming language. We take
-inspiration in the above mentioned works but focus on adapting them to our
-target language.
+Recent work has gone into adding types to Python, the other eminent data
+science language.  Typilus is an interesting piece of recent work which
+explores using neural networks to infer types for Python
+programs~\cite{allamanis2020typilus}, and Python itself added support for
+type hints in Python 3.5~\cite{pythonspectypes}.  There is no previous work
+on types for R. We take inspiration in the above mentioned works but focus
+on adapting them to our target language.
 
@@
 While this task is seemingly simple, the details and their proverbial devil
-are surprisingly tricky to get right and to scale to long running programs.
-To avoid starting from scratch, our implementation reuses an open source
-dynamic analysis framework for R named \rdt~\cite{oopsla19} which consists
-of an instrumented R Virtual Machine based on GNU-R version 3.5.0. The
-framework exposes hooks at key points in the R interpreter to which user
-defined callbacks can be attached. These hooks include function entry and
-exit, method dispatch for the S3 and S4 object systems, the longjumps used
-by the interpreter to implement non-local exit, creation and forcing of
-promises (lazy evaluation), variable definition, value creation, mutation
-and garbage collection.
+are surprisingly tricky to get right and to scale to large, long running,
+programs.  To avoid starting from scratch, our implementation reuses \rdt,
+an open source dynamic analysis framework for R~\cite{oopsla19} which
+consists of an instrumented R Virtual Machine based on GNU-R version
+3.5.0. The framework exposes hooks in the interpreter to which user defined
+callbacks can be attached. These hooks include function entry and exit,
+method dispatch for the S3 and S4 object systems, the longjumps used by the
+interpreter to implement non-local exit, creation and forcing of promises,
+variable definition, value creation, mutation and garbage collection.
 
@@
 The type information output by the tool includes the \emph{type tag} of each
-value. The R internal types are translated to names in the proposed type
-system. The next bit of information output is the \emph{class} which is an
-optional list of names, it may be absent, and, in some cases, it may be
-implicit (i.e. the interpreter blesses some values with the \k{matrix} and
-\k{array} classes even when they have no attributes). 
-Depending on the value's type tag, the tool collects further information:
+value. Internal types are translated to names in the proposed type
+system. The next bit of informatio is the \emph{class}, an optional list of
+names that may be absent, and, in some cases, is implicit (i.e. the
+interpreter blesses some values with the \k{matrix} and \k{array} classes
+even without attributes).  Depending on a value's type, the tool collects
+further information:
 
 \begin{itemize}
-\item For vectors, the number of dimensions and the present of \NA values.
-\item For lists, a recursive traversal collects element types.
-\item For promises, we attempt to establish the type of its result or output
-  \any.
+\item For vectors, the presence of \NA values.
+\item For lists, element types by a recursive traversal.
+\item For promises, an approximation of the expression type.
 \end{itemize}
@@
 \noindent
-To obtain raw types, we make use of R's C FFI and use low-level machinery to
-collect type tags and attributes from the R run-time.  The types that we
-provide to users are constructed during post-processing, and rely on the
-detailed information made available by these low-level reflection
-mechanisms.
-
-\paragraph{Varargs}
-Arguments that are part of a function's varargs (denoted \VARGS) are
-ignored, as type \VARGS is output as the type of the entire varargs construct.
+To obtain these raw types, we make use of R's C FFI and use low-level
+machinery to collect information from the R language run-time.  Types are
+conmpleted during post-processing, and rely on the detailed information made
+available by these low-level reflection mechanisms.
 
 \paragraph{Promises}
-The fact that arguments are evaluated lazily (in that expression are packed
-into promises and evaluated on first access) complicates the information
-gathering.  For example, some promises remain un-evaluated, and it would be
-erroneous to force them as they may cause side-effects that will affect the
-programs execution.  To deal with unevaluated arguments, we make an initial
-guess for each argument at function entry.  If the promise is later forced,
-we simply update the recorded type for the argument.
+The fact that arguments are lazy (expression are packed into promises and
+only evaluated on first access) complicates information gathering.  For
+example, some promises may remain un-evaluated, and it would be erroneous to
+force them as they may side-effect and change programs execution.  To deal
+with unevaluated arguments, we make an initial guess for each argument at
+function entry.  If the promise is later forced, we simply update the
+recorded type for the argument.
 
@@
 \paragraph{Non-local returns}
-When a function exits with a longjump there is no return value to speak
-of. In order to ensure that call traces are valid when a longjump occurs, we
-intercept the unwinding process initiated by the longjump and mimic the
-functionality of the functions being exited. When no return value is present
-for a call, we record a special \k{jumped} return type.
+When a function exits with a longjump, there is no return value to speak
+of. To ensure call traces are valid when a longjump occurs, we intercept the
+unwinding process and record a special \k{jumped} return type for function
+returns that are skipped.
+
+\paragraph{Varargs}
+Arguments that are part of a function's varargs (denoted \VARGS) are
+ignored. We do not attempt to give varags a type in this type system.
 
@@
 apply some transformation on the types to keep the size of types in check.
+The following overviews the main simplification rules as well as the simple
+notion of subtyping that we have adopted here.
 
@@
   $ \NAVEC{\S_1} \k{~|~}\dots \NAVEC{\S_n}$ &  \\
+  \\
   \\
@@
 minimize their size by removing redundant types, types that are subsumed by
-subtyping, and \null types. For performance reasons, we bound lists to no
-more than five types in their union, above that they become \LIST{any}.
-Functions are alos limited to twenty arguments. Higher-order functions are
-approximated by \any to \any.
+subtyping, immutable lists, and remove \null type to replace them with
+nullables. For performance reasons, we bound lists to no more than five
+types in their union, above that they become \LIST{any}.  Functions are alos
+limited to twenty arguments. Higher-order functions are approximated by \any
+to \any.
 
@@
 
+\newpage
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\newcommand{\FIX}[1]{{\bf #1}\xspace}
-
-We ran \typetracer on the test, example, and vignette code of the aforementioned corpus of \PACKAGES packages and inferred types for \NUMFUNCTIONS functions.
-We have selected and will discuss a few interesting signatures.
+We ran \typetracer on the test, example, and vignette code of the
+aforementioned corpus of \PACKAGES packages and successfully inferred types
+for \NUMFUNCTIONS functions.  Table~\ref{tbl:typesig} illustrates the
+process with ten representative signatures.  Many of the features of our
+type language are represented here.  Some signatures are telling of the
+function's behaviour.  For example, consider \code{decrypt_envelope}: the
+firs three parameters of the function are byte arrays, and the fourth
+argument is an RSA key, used to decrypt some of the inputs, and the output
+of the function is another byte array.  As another example, consider
+\code{Traverse}: according to the function documentation, it takes the root
+of a tree and traverses it in an order specified by the second argument.  We
+see that reflected in the type, where the first argument has type
+\CLASS{\texttt{Node}, \texttt{R6}} and the second argument had type
+\VEC\chr, representing the traversal order.
 
@@
   \centering
-  \resizebox{.8\linewidth}{!}{
+  \resizebox{.98\linewidth}{!}{
   \begin{tabular}{llrrr}
@@
 
-Many of the features of our type language are represented here.
-Some of these signatures are telling of the function's behaviour.
-For example, consider the signature for \code{openssl::decrypt_envelope}: the first three parameters of the function are byte arrays, and the fourth argument is an RSA key, used to decrypt some of the inputs, and the output of the function is the decrypted byte array.
-As another example, consider \code{data.tree::Traverse}: according to the function documentation, it takes the root of a tree and traverses it in an order specified by the second argument.
-We see that reflected in the type, where the first argument has type \CLASS{\texttt{Node}, \texttt{R6}} and the second argument had type \VEC\chr, representing the traversal order.
+This section attempts to evaluate how well the proposed type system is able
+to describe the actual type signatures of functions. For this we focus on
+how often there is a single type for a particular argument; this because
+union types and \any are less accurate (and would likely require a more
+refined notion of subtyping or parametric polymorphism). Then, we evaluate
+how robust the inferred signatures are by checking that they remain valid
+for other inputs.  Lastly, we try to see if the current proposal would be
+useful to programmers by allowing them to remove ad hoc checks and providing
+useful documentation.
 

### 2020-05-16 paper/main.tex
stuff
minor

@@
 
-% Contracts can be removed by calling \code{remove\_contract}.  Contracts can
-% be selectively enabled or disabled for code blocks by wrapping them in calls
-% to \code{ignore\_contracts} and \code{capture\_contracts}.  These functions
-% return a data frame that contains all the information about failed and
-% successful contract assertions in the wrapped code blocks.  Multiple
-% failures can introduce noise; this is avoided by setting
-% \code{contractr.severity} to \code{'silence'}.  \contractr still performs
-% type-checking so that results can be obtained by calling
-% \code{get\_contracts}. Setting \code{contractr.severity} flag to
-% \code{'error'} turns the type-checking warnings to errors and
-% halt the program.
-
 \paragraph{Checking the values} A number of properties can be checked
@@
 value: the absence of \NA or the type of the elements of list. Union types
-require checking for each of the alternatives. We do not check higher-order
-functions in this version of the tool.
-When argument values are wrapped in promises (this is not always the case
+require checking for each of the alternatives. We do not check higher-order functions in this version of the tool. Checking
+them could be done by dynamically wrapping functions with checks. The usage
+of higher-order functions remains limited in R. When argument values are wrapped in promises (this is not always the case
 due to compiler optimizations), in order to retain the non-strict semantics
@@
 after that function has returned.
+
 Return values require care as well.  Functions return the last expression
@@
 
-\begin{figure}[!h]\centering\includegraphics[width=.8\linewidth]
+\begin{figure}[!h]\centering\includegraphics[width=.9\linewidth]
   {plots/corpus.pdf}\caption{Corpus}\label{fig:corpus}

### 2020-05-16 paper/main.tex
Added pipeline figure

@@
 
-For this paper, we have built tooling to (\emph{a}) automate the extraction
-of raw type signatures from execution traces, (\emph{b}) infer type
-signatures from a set of raw types, and (\emph{c}) validate the inferred
-signatures by the means of contracts. This section presents each step
-of this pipeline.
+For this paper, we have built tooling to (\emph{a}) automate the extraction of
+raw type signatures from execution traces, (\emph{b}) infer type signatures from
+a set of raw types, and (\emph{c}) validate the inferred signatures by the means
+of contracts. This section presents each step of this pipeline (\emph{cf.}
+Figure~\ref{fig:pipeline}). The pipeline is run in parallel using GNU
+parallel~\cite{tange2011_parallel} on Intel Xeon 6140, 2.30GHz with 72 cores and
+256GB of RAM).
+
+\begin{figure}[t!]
+  \vspace{2mm}
+  \includegraphics[width=\textwidth]{plots/analysis-pipeline.pdf}
+  \caption{The Analysis Pipeline. (1) the rather long time is because we had to
+    extract metadata from all CRAN packages before we could compose our
+    \PACKAGES package corpus.}
+  \label{fig:pipeline}
+\end{figure}
 

### 2020-05-16 paper/main.tex
Update spaces around stuff

@@
 \subsection{Inferring type signatures from raw types}
-
+\enlargethispage{4mm}
 The output of the \typetracer tool consists of a set of tuples of raw types,
@@
 
-\medskip
+\vspace*{1mm}
 \FUN{ T_{1,1} \k{~|~} T_{1,i}, \dots, T_{n,1} \k{~|~} T_{n,j}}{T_1\k{~|~} T_k}
-\medskip
+\vspace*{1mm}
 
@@
 
-\medskip
+\vspace*{1mm}
 \begin{tabular}{rclll}
@@
   $ \NAVEC{\S_1} \k{~|~}\dots \NAVEC{\S_n}$ &  \\
-  \\
-  \\
   \\
@@
 
-\newpage
+% \newpage
 

### 2020-05-16 paper/main.tex
Move simplification rules into a figure

@@
 \noindent
-In other words, we take the union of the types occurring at individual
-argument positions rather than an union of function types. Furthermore, we
-apply some transformation on the types to keep the size of types in check.
-The following overviews the main simplification rules as well as the simple
-notion of subtyping that we have adopted here.
-
-\vspace*{1mm}
+In other words, we take the union of the types occurring at individual argument
+positions rather than an union of function types. Furthermore, we apply some
+transformation on the types to keep the size of types in check.
+Figure~\ref{fig:siplification-rules} overviews the main simplification rules as
+well as the simple notion of subtyping that we have adopted here.
+
+\begin{figure}[h!tb]
 \begin{tabular}{rclll}
@@
 \end{tabular}
-\medskip
-
-\noindent
+\caption{Simplification rules and subtyping}
+\label{fig:siplification-rules}
+\end{figure}
+
 Assuming that type sequences can be reordered freely, we rewrite types to
@@
 
-\begin{figure}[!h]\centering\includegraphics[width=.9\linewidth]
+\begin{figure}[!h]\centering\includegraphics[width=.8\linewidth]
   {plots/corpus.pdf}\caption{Corpus}\label{fig:corpus}

### 2020-05-16 paper/main.tex
Fixed text

@@
 a set of raw types, and (\emph{c}) validate the inferred signatures by the means
-of contracts. This section presents each step of this pipeline (\emph{cf.}
-Figure~\ref{fig:pipeline}). The pipeline is run in parallel using GNU
+of contracts. Figure~\ref{fig:pipeline} shows an overview of this pipeline. This
+section gives details of the main steps. The pipeline is run using GNU
 parallel~\cite{tange2011_parallel} on Intel Xeon 6140, 2.30GHz with 72 cores and
-256GB of RAM).
+256GB of RAM.
 

### 2020-09-04 paper/main.tex
Added a section about kaggle

@@
 \input{asserts.tex}
+\input{kaggle.tex}
 
@@
 
+\paragraph{User code}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+To represent end-user code in the corpus, we turned to Kaggle, an online
+platform for data-science and machine-learning. The website allows people to
+share data-science competitions, data-analysis problems together with data, for
+which users try to find the best solution. The solutions, called \emph{kernels},
+are then shared back to the platform in the form of either plain scripts or as
+notebooks. One of the most popular competition is about predicting passenger
+survival on Titanic\footnote{\url{https://www.kaggle.com/c/titanic}} with
+\KaggleDownloaded kernels in R (over 1/4 of all available R kernels) which we
+used for our corpus.
+
+Unlike CRAN, Kaggle is not a curated repository and therefore there are no
+guarantees about the quality of the code. After downloading all of the
+\KaggleDownloaded kernels and extracting the R code from the various
+formats,\footnote{We use {\sf rmarkdown} to convert from notebooks to R.} we
+found that \KaggleDuplicated were whole-file duplicates
+(\KaggleDuplicatedRatio). From the resulting \KaggleRunnable kernels,
+\KaggleFail failed to execute. Next to various runtime exceptions, common
+problems were missing libraries (no longer available for R 3.5), parse errors
+and misspelled package names. The final set contains \KaggleSuccess kernels with
+\KaggleSuccessCodeRnd lines of R code. The Kaggle kernels are used for additonal
+validation of the inferred types.
+
 \paragraph{Type usage}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2020-09-10 paper/main.tex
front matter

@@
-\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[acmsmall, screen,review]{acmart}
+\setlength{\pdfpagewidth}{8.5in}
+\setlength{\pdfpageheight}{11in}
+%\settopmatter{printfolios=true,printccs=true,printacmref=true}
+
+\setcopyright{rightsretained}
+\acmPrice{}
+\acmDOI{10.1145/3360579}
+\acmYear{2020}
+\copyrightyear{2020}
 \acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{OOPSLA} % CONF = POPL or ICFP or OOPSLA
-\acmArticle{1}
-\acmYear{2018}
-\acmMonth{1}
-\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\setcopyright{none}
+\acmVolume{3}
+\acmNumber{OOPSLA}
+\acmArticle{153}
+\acmMonth{10}
+
 \bibliographystyle{ACM-Reference-Format}
@@
 \title{Designing Types for R, Empirically}
+\author{Aviral Goel}\affiliation{\institution{Northeastern University}}
+\author{Filip Krikava}\affiliation{\institution{Czech Technical University in Prague}}
+\author{Alexi Turcotte}\affiliation{\institution{Northeastern University}}
+\author{Jan Vitek}\affiliation{\institution{Norteastern / Czech Technical University}}
 
@@
 \newcommand{\NUMFILESRUN}{145,373\xspace}
+
+\keywords{type declarations, dynamic languages, R}
+
+\begin{CCSXML}
+	<ccs2012>
+	<concept>
+	<concept_id>10011007.10011006.10011008.10011024</concept_id>
+	<concept_desc>Software and its engineering~Language features</concept_desc>
+	<concept_significance>500</concept_significance>
+	</concept>
+	<concept>
+	<concept_id>10011007.10011006.10011008</concept_id>
+	<concept_desc>Software and its engineering~General programming languages</concept_desc>
+	<concept_significance>300</concept_significance>
+	</concept>
+	<concept>
+	<concept_id>10011007.10011006.10011008.10011009.10011021</concept_id>
+	<concept_desc>Software and its engineering~Multiparadigm languages</concept_desc>
+	<concept_significance>100</concept_significance>
+	</concept>
+	</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering~Language features}
+\ccsdesc[300]{Software and its engineering~General programming languages}
 

### 2020-09-10 paper/main.tex
minor edits and merge

--- 2020-09-04 paper/main.tex
@@
 
+\AT{TODO: Softer intro to R.
+People always get confused about c() and stuff like that.
+We have some space, we can take it.}
+
 The R Project is a key tool for data analysis.  At the heart of the R
@@
 (\PercMonomorphicPositions) monomorphic positions in a corpus of
-\CountPositions parameters. With close to 70\% of monomorphic argument or
+\CountPositions parameters. With close to 65\% of monomorphic argument or
 return values, it is fair to say that even a simple type language provides
 significant benefits.
+
+We would like to point out the number of occurrences of the singleton {\tt null} type, and discuss why we treat it as an {\it any}-type.
+In R, it is common for programmers to include default values for arguments, and in many cases this value happens to be {\tt NULL}.
+Our type analysis will report a {\tt null} type for these arguments if they are never passed a value during testing.
+Thus, we interpret these as cases where we cannot be sure of the argument.
 
@@
 capture lists with named elements that can be accessed with the \code{$}
-operators. We ended discarding those types as they grew large and were often
+operators. We ended up discarding those types as they grew large and were often
 only representative of the example data being manipulated.  Consider
 function \code{cv.model}, its argument \code{x} is observed to be of
-\CLASS{\texttt{aov}, \texttt{lm}} or \CLASS{\texttt{lm}}).  Internally,
+\CLASS{\texttt{aov}, \texttt{lm}} or \CLASS{\texttt{lm}}.  Internally,
 linear models are represented as lists with named elements.  The polution is
@@
 \code{class} attribute is used by these systems to dispatch methods. S3 does
-single dispatch, S4 does multiple dispatch and R5 supports impertative
+single dispatch, S4 does multiple dispatch and R5 supports imperative
 objects.  The mechanics of S4 dispatch are more complex than for S3, and
@@
 
+\AT{Do we want to expand on this evaluation?
+Or just expand in a more general section later.}
+
+ %
+ %
+ \subsubsection{Kaggle}
+
+ \AT{TODO: Fill in and discuss Kaggle results.}
+
+ To further validate our inferred types, we repeated the experiment discussed in this section on Kaggle code.
+ By-and-large, we saw no meaningful difference in the two data sets, with the contract assertion failure patterns being repeated from the reverse dependencies.
+
+ %
+ %
+ \subsubsection{Working Failed Assertions into Types}
+
+ \AT{TODO: compare with types that we get when looking at reverse dependencies, and discuss implications.}
+
+Throughout this section, we have seen instances where the reverse dependencies of a package exercise its functions in a way that was not accounted for in the package's own test and  example code.
+These do not always represent errors, and may represent cases of oversight by package writers.
+
+If we {\it include} running a package's reverse dependencies in our type inference process, we will obtain different types for certain functions.
+In doing so, we observe \ldots
+
 %

### 2020-09-11 paper/main.tex
layout for follow-up discussion on kaggle, other patterns, and revdeps types. will aim to have draft prose/bullets in by EOD tmr

@@
 data alone does not tell the full story, as some failures may be
-overrepresented if e.g. a failing contract assertion was in a loop.  We were
+overrepresented if, e.g., a failing contract assertion was in a loop.  We were
 interested in knowing for each of the most common violations in
@@
 
-\AT{Do we want to expand on this evaluation?
-Or just expand in a more general section later.}
-
+ %
+ %
+ \subsubsection{Other Observations}
+
+ \AT{Kinds of failures: vectors passed to scalars, wrong things passed to classes, nulls passed to non-nullable arguments.}
+
+ We observed a number of other interesting patterns in the contract assertion failure data.
+
+ First, we were curious about how many \null-typed values flowed into non-nullable arguments.
+ We found that \AT{fill in numbers}.
+ Of these, there were three main kinds of ``type errors''.
+
+ \begin{itemize}
+   \item One kind of type error occurs when an argument with a {\tt NULL} default value is not given a value by the programmer.
+   In testing, the programmer only tested the function by passing values to these null-by-default arguments.
+   \AT{Check correlation with coverage.}
+
+   \item Another kind of type error occurs when {\tt NULL} is passed to an argument which is not expected to be {\tt NULL}, but that does not cause an error.
+   One example is \AT{see as.integer example}.
+
+   \item Finally, in some cases the {\tt NULL}s resulted in errors, either unhandled or handled by a check and a conditional.
+ \end{itemize}
+
+ Next, we were curious about how often vectors were passed to arguments expecting scalars.
+ We found \AT{finish analysis, numbers}.
+ Besides an outright error, this kind of contract assertion failure might indicate that a function was not well-tested, in that it was only ever tested with unit-length vectors being passed to an argument which is intended to have a vector type.
+ Further, these errors may reveal functions that were not designed with a vector-typed argument in mind, but can in fact handle vectors of values (in R, most functions that can accept scalars can also accept vectors).
  %
@@
  By-and-large, we saw no meaningful difference in the two data sets, with the contract assertion failure patterns being repeated from the reverse dependencies.
+
+ \AT{Discuss how things match up between Kaggle and revdeps.}
 

### 2020-09-11 paper/main.tex
added alternate class tyoe unification discussion

@@
 
-In this paper, we will attempt infer type signature for functions. We now
-introduce some relevant concepts.  Functions can be called with named
-parameters, arguments can have default values, and R also support variable
-argument lists.  To illustrate all of this in a single example, consider:
+In this paper, we will attempt infer type signature for functions. 
+We begin by introducing some relevant concepts.
+
+The basic building blocks of R are vectors of primitives.
+Characters (strings), integers, logicals (booleans), doubles, complex numbers, and raw bytes make up R's primitive values.
+When we say that R is {\it vectorized}, we mean that primitive values are {\it always} vectors.
+For example, in R one might see \code{2.5} and think it to be a scalar double, but it is in actuality a unit-length vector containing a single double.
+Vectors can also be explicitly constructed using the vector constructor \code{c(...)}, 
+For example, \code{c(1L, 2L, 3L)} creates a vector of three integers (in R, integer literals are denoted by an \code{L}).
+
+The main object of our work is R functions, which have a number of quirks.
+Arguments to functions in R can be assigned default values (which can be arbitrary expressions!), and R functions can support variable argument lists, and functions can be called by specifying argument names explicitly in the call.
+To illustrate all of this in a single example, consider:
 
 \begin{lstlisting}
-  f <- function(x, ..., y=if(z==0) 1, z=0) x + y + if (missing(...)) 0 else c(...)
+f <- function(x, ..., y=if(z==0) 1, z=0) { 
+  x + y + if (missing(...)) 0 else c(...)
+}
 \end{lstlisting}
@@
 \noindent
-The above hints at the polymorphism of the language, \k f, may return a
-vector of integer or of double, the length of the vector depends on the
-length of the varargs and of \k x and \k y.  The language does not really
+The above hints at the polymorphism of the language: \k f may return a
+vector of integers or of doubles, the length of the vector depends on the
+length of the varargs, and of \k x and \k y.  As we mentioned, the language does not really
 differentiate between scalar and vectors.  Some more exotic types that can
@@
 function. Figure~\ref{rtypes} lists all of the builtin types that are
-provided by the language. They are the possible return values of
-\k{typeof}. There is no intrinsic notion of subtyping in R. But, in many
+provided by the language, which are the possible return values of
+\k{typeof}. There is no intrinsic notion of subtyping in R, but in many
 context a \k{logical} will convert to \k{integer}, and an \k{integer} will
@@
 
- We observed a number of other interesting patterns in the contract assertion failure data.
+ We drew a number of other observations from the contract assertion failure data.
 
@@
 
+%
+%
+%
+%
+\subsection{Discussion}
+
+\AT{Talk about how we thought do to a different unification for class types.}
+
+As a closing point of discussion, we would like to highlight an alternate strategy for unifying class types that we considered.
+Recall that in our proposed type annotation language, classes are represented as \CLASS{{\tt A}} for something of class {\tt A}, and \CLASS{{\tt A}, {\tt B}} for something with two classes {\tt A} and {\tt B}.
+
+During the robustness phase of our evaluation, we noted a number of class-based type errors such as a \CLASS{{\tt data.frame}} being passed to an argument expecting, e.g., a \CLASS{{\tt data.frame}, {\tt tbl}, {\tt tbl\_df}}.
+This represents something of a limitation of our approach, as data frames and {\tt dplyr} tibbles are related and convertible (you can think of a tibble as a subclass of a data frame)---a fact not represented by our types.
+One option to mitigate this would be to include a type such as \CLASS{{\tt data.frame}, ...} that must include {\it at least} the {\tt data.frame} class, but this would not altogether solve this problem.
+For one, to generate this type we would either need to a) include some type hierarchy in our type system, which is outside of the scope of this work, or b) somehow infer the type with \typetracer.
+To create this type, \typetracer would have to observe values of {\it both} type \CLASS{{\tt data.frame}} and \CLASS{{\tt data.frame}, {\tt tbl}, {\tt tbl\_df}}, which would not solve the aforementioned problem as many {\tt dplyr} functions are only called with tibbles.
+
+Further, the data table, another data frame alternative, while being coercible to and from a data frame, only has \CLASS{{\tt data.table}} (not including {\tt data.frame} as part of its class).
+This alternate class type would also not adequately capture this behaviour, and even worse, if a \CLASS{{\tt data.frame}} and \CLASS{{\tt data.table}} were to co-occur in an argument type, we would unify it to \CLASS{...}, which is not very informative.
+
+For these reasons, we believe our unification strategy for class types to be adequate until we extend our type system to handle R's wild and varied object orientation frameworks.
+
+%
+%
+%
+%
+%
+%
 \section{Conclusion}

### 2020-09-11 paper/main.tex
cleanup in new discussion. will work out notebooks and fill in numbers + examples when data available

@@
  We found that \AT{fill in numbers}.
- Of these, there were three main kinds of ``type errors''.
-
- \begin{itemize}
-   \item One kind of type error occurs when an argument with a {\tt NULL} default value is not given a value by the programmer.
-   In testing, the programmer only tested the function by passing values to these null-by-default arguments.
-   \AT{Check correlation with coverage.}
-
-   \item Another kind of type error occurs when {\tt NULL} is passed to an argument which is not expected to be {\tt NULL}, but that does not cause an error.
-   One example is \AT{see as.integer example}.
-
-   \item Finally, in some cases the {\tt NULL}s resulted in errors, either unhandled or handled by a check and a conditional.
- \end{itemize}
+ We manually inspected some of the offending functions and observed three main patterns.
+
+ First, we found that many of these errors occurred in arguments that have a {\tt NULL} default value.
+ This would be the case when the programmer only tested the function by passing values to these null-by-default arguments, and clients of the package make use of the default.
+ In fact, we observed \AT{something about coverage.}
+
+ The other two patterns are for arguments with no default value, where either the call results in an error (perhaps explicitly handled by the programmer), or resulting in valid function behaviour that was untested by the original package designer.
+ Here, the first case can be explained by a lack of testing (or generally incorrect use in a client), and the second case is explained by programmers not fully understanding R's language semantics (\AT{as who would expect as.integer(NULL) to yield integer(0)?}).
 
@@
  Further, these errors may reveal functions that were not designed with a vector-typed argument in mind, but can in fact handle vectors of values (in R, most functions that can accept scalars can also accept vectors).
+
  %

### 2020-09-13 paper/main.tex
lots of prose and comments, can trim. waiting on last data to finish kaggle and revdeps discussion. need to add NA discussion to background

@@
 \title{Designing Types for R, Empirically}
+\author{Alexi Turcotte}\affiliation{\institution{Northeastern University}}
 \author{Aviral Goel}\affiliation{\institution{Northeastern University}}
 \author{Filip Krikava}\affiliation{\institution{Czech Technical University in Prague}}
-\author{Alexi Turcotte}\affiliation{\institution{Northeastern University}}
 \author{Jan Vitek}\affiliation{\institution{Norteastern / Czech Technical University}}
@@
 
- \AT{Kinds of failures: vectors passed to scalars, wrong things passed to classes, nulls passed to non-nullable arguments.}
-
  We drew a number of other observations from the contract assertion failure data.
@@
  This would be the case when the programmer only tested the function by passing values to these null-by-default arguments, and clients of the package make use of the default.
+ As an example, we observed a contract assertion failure when {\tt inner} and {\tt labels} were {\tt NULL} in the following function:
+ \begin{lstlisting}
+nlme::nfGroupedData <- function (formula, data = NULL, order.groups = TRUE, 
+    FUN = function(x) max(x, na.rm = TRUE), outer = NULL, inner = NULL, 
+    labels = NULL, units = NULL) {
+  # ...
+}
+ \end{lstlisting}
  In fact, we observed \AT{something about coverage.}
@@
  The other two patterns are for arguments with no default value, where either the call results in an error (perhaps explicitly handled by the programmer), or resulting in valid function behaviour that was untested by the original package designer.
- Here, the first case can be explained by a lack of testing (or generally incorrect use in a client), and the second case is explained by programmers not fully understanding R's language semantics (\AT{as who would expect as.integer(NULL) to yield integer(0)?}).
+ Here, the first case can be explained by a lack of testing (or generally incorrect use in a client), and the second case is explained by programmers not fully understanding R's language semantics.
+ For example, we observed this kind of error in the following function: 
+ \begin{lstlisting}
+BBmisc::convertIntegers <- function (x) {
+    if (is.integer(x)) 
+        return(x)
+    if (length(x) == 0L || (is.atomic(x) && all(is.na(x)))) 
+        return(as.integer(x))
+    if (is.numeric(x)) {
+        # ...
+    }
+    return(x)
+}
+ \end{lstlisting}
+Here, if {\tt x} is {\tt NULL}, the second branch of the conditional will be triggered (as in R, \code{length(NULL) == 0}), and the function will return \code{as.integer(NULL)}, which curiously returns \code{integer(0)}, a zero-length vector of integers (one might expect it to return the integer \code{NA} value, or error).
 
@@
  Further, these errors may reveal functions that were not designed with a vector-typed argument in mind, but can in fact handle vectors of values (in R, most functions that can accept scalars can also accept vectors).
+ As an example, consider the function \code{BBmisc::strrepeat}, which takes a string and repeats it a specified number of times:
+ \begin{lstlisting}
+BBmisc::strrepeat <- function (x, n, sep = "") {
+    paste0(rep.int(x, n), collapse = sep)
+}
+
+BBmisc::strrepeat("a", 3) # => "aaa"
+BBmisc::strrepeat(c("a", "b"), 3) # => "ababab"
+ \end{lstlisting}
+ This function was only ever tested with unit-length vectors passed to {\tt x}, even though technically it can handle longer vectors, as per the two sample calls above.
+ This could be attributed to poor quality testing, or misunderstanding language semantics (e.g., misunderstanding the semantics of \code{paste0} and \code{rep.int}), but we found other instances of type errors where the functions really ought to have been tested with the offending type, for instance:
+ \begin{lstlisting}
+combinat::permn <- function (x, fun = NULL, ...) {
+    if (is.numeric(x) && length(x) == 1 && x > 0 && trunc(x) == x) 
+        x <- seq(x)
+    # ...
+}
+ \end{lstlisting}
+ As per the documentation, this function is intended to take a vector, and generate all permutations of elements in that vector.
+If given a scalar ``integer'' $n$, it will generate all permutations for the list $[1, 2, ..., n]$.
+Curiously, the function was only tested by passing an integer, and not ever with a vector (even though, presumably, that is the main utility of the function).
+
+Finally, we were curious to see what patterns of errors occurred in arguments expecting classes.
+\AT{Overall, the numbers are ...}.
+Many of these errors appear to be related to ``subclasses'' being passed to arguments expecting a ``superclass''.
+For instance, there are \AT{many} instances of \CLASS{{\tt data.frame}} being passed to arguments expecting tibbles or data.tables (which can be readily converted).
+\AT{Check dispatch, and maybe count it out when doing this.}
+
+Other common patterns include \CLASS{{\tt matrix}}, \CLASS{{\tt data.frame}}, and vectors being passed to arguments expecting \CLASS{{\tt array}}, which is \AT{elaborate on arrays}.
+\AT{Fill rest of this in: vectors and data frames passed to matrices. reiterate how this coercion is ad hoc. maybe talk about coercion semantics.}
 
@@
  \AT{Discuss how things match up between Kaggle and revdeps.}
+
+\begin{itemize}
+  \item compare which functions have failures 
+  \item look at the coverage 
+  \item types of errors?
+\end{itemize}
 

### 2020-09-13 paper/main.tex
numbers and tables updated, smoothing wording now and fleshing out extra discussions

@@
 \newcommand{\NAVEC}[1]{\k{\string^}\!#1\k{[]}\xspace}
+\newcommand{\NASCA}[1]{\k{\string^}\!#1\xspace}
 
@@
 \author{Filip Krikava}\affiliation{\institution{Czech Technical University in Prague}}
-\author{Jan Vitek}\affiliation{\institution{Norteastern / Czech Technical University}}
-
-\newcommand{\NUMFUNCTIONS}{20,214\xspace}  %%% TODO auto-gen
+\author{Jan Vitek}\affiliation{\institution{Nortjeastern / Czech Technical University}}
+
+\newcommand{\NUMFUNCTIONS}{25,215\xspace}  %%% TODO auto-gen
 \newcommand{\NUMPACKAGES}{412\xspace}  %%% TODO auto-gen
@@
 \newcommand{\YEARS}{20\xspace} %%TODO fix
-\newcommand{\PERCFAILEDASSERTIONS}{2.19\%\xspace}
-\newcommand{\PERCFAILEDASSERTIONSWUNDEF}{2.23\%\xspace}
-\newcommand{\PERCASSERTIONSUNDEF}{0.04\%\xspace}
-\newcommand{\NUMPKGSEVAL}{7485\xspace} % autogenerate
-\newcommand{\NUMASSERTIONS}{62,171,573\xspace} %autogenerate
-\newcommand{\PROPFUNSFAILEDCHECK}{21.21\%\xspace}
-\newcommand{\PROPFUNSFAILEDCHECKNOSTHREE}{15.48\%\xspace} 
-\newcommand{\PROPARGSFAILINGASSERTS}{3.53\%\xspace} % autogen....
-\newcommand{\PERCSUCCARG}{96.46\%\xspace}
-\newcommand{\PERCSUCCFUNNOSTHREE}{84.52\%\xspace}
-\newcommand{\PERCSUCCFUNS}{78.79\%\xspace}
-\newcommand{\PERCCALLSBADFUNSINTESTS}{5.38\%\xspace}
+\newcommand{\PERCFAILEDASSERTIONS}{2.05\%\xspace}
+\newcommand{\PERCFAILEDASSERTIONSWUNDEF}{1.98\%\xspace}
+\newcommand{\PERCASSERTIONSUNDEF}{0.07\%\xspace}
+\newcommand{\NUMPKGSEVAL}{8,694\xspace} % autogenerate
+\newcommand{\NUMASSERTIONS}{98,105,161\xspace} %autogenerate
+\newcommand{\PROPFUNSFAILEDCHECK}{12.29\%\xspace}
+\newcommand{\PROPFUNSFAILEDCHECKNOSTHREE}{10.30\%\xspace} 
+\newcommand{\PROPARGSFAILINGASSERTS}{2.40\%\xspace} % autogen....
+\newcommand{\PERCSUCCARG}{97.60\%\xspace}
+\newcommand{\PERCSUCCFUNNOSTHREE}{89.70\%\xspace}
+\newcommand{\PERCSUCCFUNS}{87.70\%\xspace}
+\newcommand{\PERCCALLSBADFUNSINTESTS}{2.73\%\xspace}
 \newcommand{\TOTNUMSIGS}{21,730\xspace}
-\newcommand{\PROPSTRUCTSDYN}{44.46\%\xspace}
-\newcommand{\NUMFILESRUN}{145,373\xspace}
+\newcommand{\PROPSTRUCTSDYN}{65.97\%\xspace}
+\newcommand{\NUMFILESRUN}{160,379\xspace}
+\newcommand{\NUMFILESREVDEPS}{153,806\xspace}
+\newcommand{\NUMFILESKAGGLE}{6,573\xspace}
 
@@
 between scalars and vectors (they are all vectors), so \code{typeof(5) ==}
-\code{typeof(c(5)) == typeof(c(5,5))} \code{ == "double"}. Finally all
-vectorized data types have a distinguished missing value denoted by
-\code{NA}. The default type of \code{NA} is \k{logical}. We can see that
-\code{typeof(NA)=="logical"}, but NA inhabits every type:
-\code{typeof(c(1,NA)[2])=="double"}.
+\code{typeof(c(5)) == typeof(c(5,5))} \code{ == "double"}.
+
+All vectorized data types have a distinguished missing value denoted by \code{NA}. 
+The default type of \code{NA} is \k{logical}. 
+We can see that \code{typeof(NA)=="logical"}, but NA inhabits every type: \code{typeof(c(1,NA)[2])=="double"}.
+In addition to \code{NA}, R has a \code{NULL} value, and to understand why R has both a \code{NULL} value and \code{NA} values it is useful to remember that R is primarily designed for statistics and data science.
+In experimental data science, it is useful to have a notion of a ``missing observation'', which is often NaN (not a number, or similar) in, e.g., MATLAB and Python.
+Since vectors are unityped in R, it {\it must} have a ``NaN'' value for each primitive type, hence the \code{NA}s, while there is only a single \code{NULL} null-typed value in the language to encompass all other cases.
 
@@
 Type & Args & \% of Args & Observations & \% of Obs.\\\midrule
-\dbl               & 13,255 &     11.0 &  41,560,797 & 1.5\\
-\lgl               & 10,142 &      8.4 & 712,895,959 & 27.2\\
-\null              &  9,175 &      7.6 & 122,202,183 & 4.6\\
-\chr               &  9,171 &      7.5 & 794,058,588 & 30.3\\
-\VEC\dbl           &  7,811 &      6.5 &  29,272,232 & 1.1\\
-\VARGS             &  7,759 &      6.4 &  76,369,333 & 2.9\\
-\any               &  6,243 &      5.2 & 255,539,497 & 9.0\\
-\VEC\chr           &  4,576 &      3.8 &  88,233,523 & 3.3\\
-\CLASS{\texttt{matrix}}     & 4,522 & 3.7 & 32,452,555 & 1.2\\
-\CLASS{\texttt{data.fram}e} & 2,781 & 2.3 &  7,749,094 & 0.3\\
+\dbl               & 12,298 &     11.24 &  12,152,787 & 20.3 \\
+\lgl               & 9,366 &      8.7 & 6,650,294 & 11.1 \\
+\null              &  8,799 &      8.0 & 2,187,611 & 3.7 \\
+\chr               &  8,727 &      8.0 & 2,564,207 & 4.3 \\
+\VEC\dbl           &  7,190 &      6.6 &  4,934,773 & 8.2 \\
+\VARGS             &  6,611 &      6.0 &  6,075,874 & 10.1 \\
+\any               &  6,120 &      5.6 & 339,299 & 0.6 \\
+\VEC\chr           &  4,325 &      4.0 &  1,060,466 & 1.8 \\
+\CLASS{\texttt{matrix}}     & 4,152 & 3.8 & 2,805,718 & 4.7 \\
+\CLASS{\texttt{data.frame}} & 2,608 & 2.4 &  352,655 & 0.6 \\
 \bottomrule
@@
 \midrule
-\VEC\dbl & \CLASS{\texttt{matrix}} & 371,896 & 27.37 & 27.37\\
-\VEC\dbl & \VEC\int & 117,212 & 8.62 & 35.99\\
-\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 100,100 & 7.37 & 43.36\\
-\dbl & \int | \null & 60,578 & 4.46 & 47.81\\
-\VEC\dbl & \CLASS{\texttt{timeSeries}} & 58,872 & 4.33 & 52.15\\
-\addlinespace
-\CLASS{\texttt{matrix}} & \CLASS{\texttt{timeSeries}} & 55,005 & 4.05 & 56.19\\
-\VEC\dbl & \dbl & 53,338 & 3.92 & 60.12\\
-\dbl & \CLASS{\texttt{data.frame}} & 32,553 & 2.40 & 62.51\\
-\VEC\dbl & \CLASS{\texttt{data.frame}} & 31,576 & 2.32 & 64.84\\
-\dbl & \int & 20,586 & 1.51 & 66.35\\
+\VEC\dbl & \CLASS{\texttt{matrix}} & 705,036 & 36.36 & 36.36 \\ 
+\VEC\dbl & \VEC\int & 189,800 & 9.79 & 46.15 \\ 
+\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 100,100 & 5.16 & 51.31 \\
+\CLASS{{\tt simpleError, error, condition}} & \CLASS{{\tt data.frame}} | \CLASS{{\tt matrix}} | \CLASS{{\tt randomForest}} | \VEC\dbl & 78,197 & 4.03 & 55.34 \\ 
+  \CLASS{{\tt data.frame}} & \CLASS{{\tt data.frame, tbl, tbl\_df}} & 58,809 & 3.03 & 58.37 \\ 
+  \CLASS{{\tt matrix}} & \CLASS{{\tt timeSeries}} & 53,482 & 2.76 & 61.13 \\ 
+ \VEC\dbl & \VEC\int & 33,350 & 1.72 & 62.85 \\ 
+ \dbl & \CLASS{{\tt data.frame}} & 32261 & 1.66 & 64.52 \\ 
+  \VEC\dbl & \CLASS{{\tt data.frame}} & 31361 & 1.62 & 66.13 \\ 
+  \CLASS{{\tt xml\_node}} & \CLASS{{\tt xml\_missing}} $|$ \CLASS{{\tt xml\_nodeset}} $|$ \CLASS{{\tt xml\_document, xml\_node}} & 30,330 & 1.56 & 67.70 \\ 
 \bottomrule
@@
 Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
-\midrule
-\VEC\dbl & \CLASS{\texttt{matrix}} & 15 & 1394 & 1.08\\
-\VEC\dbl & \VEC\int & 44 & 657 & 6.70\\
-\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 1 & 3 & 33.33\\
-\dbl & \int | \null & 5 & 20 & 25.00\\
-\VEC\dbl & \CLASS{\texttt{timeSeries}} &19 & 35 & 54.29\\
-\addlinespace
-\CLASS{\texttt{matrix}} & \CLASS{\texttt{timeSeries}} &  19 & 35 & 54.29\\
-\VEC\dbl & \dbl & 100 & 5021 & 1.99\\
-\dbl & \CLASS{\texttt{data.frame}} &  1 & 900 & 0.11\\
-\VEC\dbl & \CLASS{\texttt{data.frame}} &  6 & 900 & 0.67\\
-\dbl & \int & 48 & 529 & 9.07\\
+\VEC\dbl & \CLASS{\texttt{matrix}} & 18 & 1522 & 1.18 \\ 
+\VEC\dbl & \VEC\int  & 66 & 5865 & 1.13 \\ 
+\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 1 & 3 & 33.33 \\ 
+\CLASS{{\tt simpleError, error, condition}} & \CLASS{{\tt data.frame}} | \CLASS{{\tt matrix}} | \CLASS{{\tt randomForest}} | \VEC\dbl & 2 & 2 & 100 \\ 
+  \CLASS{{\tt data.frame}} & \CLASS{{\tt data.frame, tbl, tbl\_df}} & 23 & 196 & 11.73 \\ 
+  \CLASS{{\tt matrix}} & \CLASS{{\tt timeSeries}} & 21 & 39 & 53.85 \\ 
+ \VEC\dbl & \VEC\int & 31 & 624 & 4.97 \\ 
+ \dbl & \CLASS{{\tt data.frame}} & 2 & 1025 & 0.20 \\ 
+  \VEC\dbl & \CLASS{{\tt data.frame}}  & 6 & 1025 & 0.59 \\ 
+  \CLASS{{\tt xml\_node}} & \CLASS{{\tt xml\_missing}} $|$ \CLASS{{\tt xml\_nodeset}} $|$ \CLASS{{\tt xml\_document, xml\_node}}  & 1 & 1 & 100 \\ 
 \bottomrule
@@
 \midrule
-\VEC\dbl & \VEC\int & 44 & 657 & 6.70\\
-\NAVEC\lgl & \null & 158 & 2756 & 5.73\\
-\VEC\chr & \chr & 183 & 4102 & 4.46\\
-\chr & \null & 117 & 2756 & 4.25\\
-\dbl & \null & 107 & 2756 & 3.88\\
+\VEC\dbl & \VEC\int & 31 & 624 & 4.97 \\
+\dbl & \int & 21 & 519 & 4.05\\
+\VEC\chr & \chr | \null & 10 & 256 & 3.91 \\
+\NAVEC\lgl & \VEC\lgl & 8  & 219 & 3.65 \\
+\NASCA\lgl & \VEC\lgl & 5 & 219 & 2.28 \\
 \bottomrule
@@
 \midrule
-\CLASS{\texttt{data.frame}} & \CLASS{\texttt{data.frame}, \texttt{tbl}, \texttt{tbl\_df}} & 63 & 284 & 22.18\\
-\CLASS{\texttt{matrix}} & \CLASS{\texttt{array}} & 13 & 70 & 18.57\\
-\NAVEC\lgl & \lgl | \null & 9 & 49 & 18.37\\
-\VEC\chr & \LIST{\VEC\chr} & 20 & 109 & 18.35\\
-\VEC\clx & \clx & 7 & 42 & 16.67\\
+\CLASS{{\tt matrix}} & \CLASS{{\tt timeSeries}} & 21 & 39 & 35.90 \\
+\VEC\dbl & \CLASS{{\tt timeSeries}} & 21 & 39 & 35.90 \\
+\CLASS{{\tt data.frame}} & \CLASS{{\tt timeSeries}} & 14 & 39 & 35.90 \\
+\CLASS{{\tt dtplyr\_step\_first, dtplyr\_step}} & \CLASS{{\tt data.frame}} | Grouped Tibble | Tibble & 10 & 45 & 22.22 \\
+\chr & \raw & 6 & 31 & 19.35 \\
+% \CLASS{\texttt{data.frame}} & \CLASS{\texttt{data.frame}, \texttt{tbl}, \texttt{tbl\_df}} & 63 & 284 & 22.18\\
+% \CLASS{\texttt{matrix}} & \CLASS{\texttt{array}} & 13 & 70 & 18.57\\
+% \NAVEC\lgl & \lgl | \null & 9 & 49 & 18.37\\
+% \VEC\chr & \LIST{\VEC\chr} & 20 & 109 & 18.35\\
+% \VEC\clx & \clx & 7 & 42 & 16.67\\
 \bottomrule
@@
 % END Autogenerated
- \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 80th percentile}
+ \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 80th percentile.
+ *: Grouped Tibble is \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}} , and Tibble is \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}.
+ \AT{TODO Fix this table.}}
  \label{tbl:fail-breakdown-3}

### 2020-09-14 paper/main.tex
new words in. let the edits commence

@@
 \author{Filip Krikava}\affiliation{\institution{Czech Technical University in Prague}}
-\author{Jan Vitek}\affiliation{\institution{Nortjeastern / Czech Technical University}}
+\author{Jan Vitek}\affiliation{\institution{Northeastern / Czech Technical University}}
 
@@
 \newcommand{\NUMFILESKAGGLE}{6,573\xspace}
+\newcommand{\NUMNULLASSERTERRORS}{6.46\%\xspace}
+\newcommand{\NUMFUNSWITHNULLASSERTERROR}{141\xspace}
+\newcommand{\PERCDYNAMICVECERRORS}{12.73\%\xspace}
+\newcommand{\NUMFUNSWITHVECASSERTERROR}{114\xspace}
+\newcommand{\PERCFAILASSCLASS}{81.44\%\xspace}
+\newcommand{\PERCFAILEDONECLASS}{51.13\%\xspace}
+
+% Kaggle
+ \newcommand{\PERCFAILEDASSERTIONSKAGGLE}{2.14\%\xspace}
+\newcommand{\PERCFAILEDASSERTIONSKAGGLENOUNDEF}{0.42\%\xspace}
+ \newcommand{\PROPFUNSWITHFAILEDASSERTKAGGLE}{15.98\%\xspace}
+ \newcommand{\NUMTOTALASSERTIONSKAGGLE}{1,9038,496\xspace}
+ \newcommand{\NUMASSERTEDFUNCTIONSKAGGLE}{970\xspace}
 
@@
 \subsection{The R Programming Language} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\AT{TODO: Softer intro to R.
-People always get confused about c() and stuff like that.
-We have some space, we can take it.}
 
@@
 
+\AT{TODO: Update plots, reread and readjust wording and examples since some tables have changed.}
+
 The first part of our evaluation attempts to shed light on how good a fit
@@
 % BEGIN Autogenerated
-\small
+\scriptsize
 \begin{tabular}{llrrr}
@@
 % BEGIN Autogenerated\
-\small
+\scriptsize
 \begin{tabular}{llrrr}
@@
 % BEGIN Autogenerated
-\small
+\scriptsize
 \begin{tabular}{ll@{}rrr@{}}
@@
  \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 80th percentile.
- *: Grouped Tibble is \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}} , and Tibble is \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}.
- \AT{TODO Fix this table.}}
+ *: Grouped Tibble is \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}} , and Tibble is \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}.}
  \label{tbl:fail-breakdown-3}
@@
 
- First, we were curious about how many \null-typed values flowed into non-nullable arguments.
- We found that \AT{fill in numbers}.
+ First, we were curious about how many \null-typed values flowed into non-nullable arguments, and we found that they accounted for \NUMNULLASSERTERRORS of contract assertion failures in \NUMFUNSWITHNULLASSERTERROR functions.
  We manually inspected some of the offending functions and observed three main patterns.
@@
  \end{lstlisting}
- In fact, we observed \AT{something about coverage.}
 
@@
  Next, we were curious about how often vectors were passed to arguments expecting scalars.
- We found \AT{finish analysis, numbers}.
+ We found that \PERCDYNAMICVECERRORS of dynamic contract assertion failures were of this type, and these errors were present in \NUMFUNSWITHVECASSERTERROR functions.
  Besides an outright error, this kind of contract assertion failure might indicate that a function was not well-tested, in that it was only ever tested with unit-length vectors being passed to an argument which is intended to have a vector type.
@@
 Finally, we were curious to see what patterns of errors occurred in arguments expecting classes.
-\AT{Overall, the numbers are ...}.
-Many of these errors appear to be related to ``subclasses'' being passed to arguments expecting a ``superclass''.
-For instance, there are \AT{many} instances of \CLASS{{\tt data.frame}} being passed to arguments expecting tibbles or data.tables (which can be readily converted).
-\AT{Check dispatch, and maybe count it out when doing this.}
-
-Other common patterns include \CLASS{{\tt matrix}}, \CLASS{{\tt data.frame}}, and vectors being passed to arguments expecting \CLASS{{\tt array}}, which is \AT{elaborate on arrays}.
-\AT{Fill rest of this in: vectors and data frames passed to matrices. reiterate how this coercion is ad hoc. maybe talk about coercion semantics.}
+Overall we found that \PERCFAILASSCLASS of assertion failures were on arguments which were expecting a class in some way (\PERCFAILEDONECLASS of assertion failures were on monomorphic arguments expecting a class, the remainder on polymorphic arguments with at least one option being a class).
+
+There are two broad divisions which account for most of the class-related contract assertion failures that are not outright errors.
+First, we observe a class of errors related to ``subclasses'' being passed to arguments expecting a ``superclass''.
+For instance, we observed \CLASS{{\tt data.frame}} values being passed to arguments expecting tibbles or data.tables (which can be readily converted).
+Second, we observe a class of errors related more to coercion between simple data types and classes.
+As an example, consider the aforementioned assertion failures in the \code{timeSeries} package, and as a further example we found many instances of \CLASS{{\tt matrix}}, \CLASS{{\tt data.frame}}, and vectors being passed to arguments expecting \CLASS{{\tt array}}, a generalization of matrices.
 
@@
 
- \AT{TODO: Fill in and discuss Kaggle results.}
-
- To further validate our inferred types, we repeated the experiment discussed in this section on Kaggle code.
+ To further validate our inferred types, we repeated the experiment discussed in this section on end-user R code found on the Kaggle competition website.
+
  By-and-large, we saw no meaningful difference in the two data sets, with the contract assertion failure patterns being repeated from the reverse dependencies.
-
- \AT{Discuss how things match up between Kaggle and revdeps.}
-
-\begin{itemize}
-  \item compare which functions have failures 
-  \item look at the coverage 
-  \item types of errors?
-\end{itemize}
-
- %
- %
- \subsubsection{Working Failed Assertions into Types}
-
- \AT{TODO: compare with types that we get when looking at reverse dependencies, and discuss implications.}
-
-Throughout this section, we have seen instances where the reverse dependencies of a package exercise its functions in a way that was not accounted for in the package's own test and  example code.
-These do not always represent errors, and may represent cases of oversight by package writers.
-
-If we {\it include} running a package's reverse dependencies in our type inference process, we will obtain different types for certain functions.
-In doing so, we observe \ldots
+ Overall, we observed that \PERCFAILEDASSERTIONSKAGGLE of all contract assertions failed while running Kaggle code.
+ If we remove assertion failures related to our simplifying assumption that function types will not have more than 20 arguments, that number drops to a mere \PERCFAILEDASSERTIONSKAGGLENOUNDEF.
+ In all, \PROPFUNSWITHFAILEDASSERTKAGGLE functions had at least one contract failure.
+ There were \NUMTOTALASSERTIONSKAGGLE assertions in total, on \NUMASSERTEDFUNCTIONSKAGGLE functions.
+
+ \begin{table}[ht]
+ \scriptsize
+\centering
+\begin{tabular}{llrrr}
+\toprule
+Passed & Arg Type & Occurrences & \% Total & Cumulative \%\\
+\midrule
+ \CLASS{{\tt data.table, data.frame}} & \CLASS{{\tt matrix}} & 20002 & 28.15 & 28.15 \\ 
+ \CLASS{{\tt factor}} & \NAVEC\chr & 18344 & 25.81 & 53.96 \\ 
+ \CLASS{{\tt factor}} & \VEC\chr  & 7519 & 10.58 & 64.54 \\ 
+ \VEC\chr & \LIST{\VEC\int} $|$ ... $|$ \LIST{\CLASS{{\tt formula, quosure}}} & 5385 & 7.58 & 72.12 \\ 
+ \CLASS{{\tt ixforeach, iter}} & \CLASS{{\tt dataframeiter, iter}} $|$ ... $|$ \CLASS{{\tt iforeach, iter}} & 4139 & 5.82 & 77.94 \\ 
+\bottomrule
+\end{tabular}
+\caption{Top contract failures in Kaggle kernels}
+\label{tbl:fail-breakdown-0-kaggle}
+\end{table}
+
+ \begin{table}
+% BEGIN Autogenerated\
+\small
+\begin{tabular}{llrrr}
+\toprule
+Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
+\midrule
+\VEC\chr & \chr & 12 & 304 & 0.04 \\ 
+\CLASS{{\tt factor}} & \VEC\chr & 7 & 199 & 0.04 \\ 
+\VEC\dbl & \VEC\chr & 6 & 199 & 0.03 \\ 
+\NAVEC\chr & \VEC\chr & 4 & 199 & 0.02 \\ 
+\int & \chr & 5 & 304 & 0.02 \\ 
+\bottomrule
+\end{tabular}
+\caption{What is the highest failure rate among popular argument types in Kaggle? For argument signatures whose frequency is in the 90th percentile}
+ \label{tbl:fail-breakdown-2-kaggle}
+\end{table}
+
+ To mirror our analysis of contract assertions on the reverse dependencies of our corpus, we show in Table~\ref{tbl:fail-breakdown-0-kaggle} the most frequently failing contract in Kaggle.
+ While we don't see many overlapping entries per se, the table entries have similar explanations.
+ For instance, values of type \CLASS{{\tt data.table, data.frame}} are often passed to arguments expecting a \CLASS{{\tt matrix}}.
+ \CLASS{{\tt data.table, data.frame}} values are known as ``data tables'', and essentially serve the same purpose as data frames and tibbles.
+ As it happens, many data tables can be coerced to matrices, and programmers will often interchange the two (particularly when data tables have numeric data), as is the case here.
+ One function producing many of these errors is the following:
+ \begin{lstlisting}
+class::knn <- function (train, test, cl, k = 1, l = 0, prob = FALSE, use.all = TRUE)  {
+    train <- as.matrix(train)
+    test <- as.matrix(test)
+    ...
+}
+ \end{lstlisting}
+ We see that \code{class::knn} can take arguments as long as they can be passed to \code{as.matrix}, e.g., data tables, and will perform the coercion.
+ As in similar cases discussed earlier, we did not encode this semantics in our type system as the process of coercion in R is ad hoc.
+ On the topic of coercion, rows two and three of Table~ are interesting as they depict {\it factors} being passed to arguments expecting character vectors.
+ \CLASS{{\tt factor}} typed values are known as factors in R, and they are stored as a vector of integer values corresponding to a set of character values, and their purpose is to allow for quickly deal with categorized data.
+ Factors can be readily converted to characters when needed, as evidenced by these assertion failures.
+
+ Another interesting entry in Table~\ref{tbl:fail-breakdown-0-kaggle} is the fifth row, where a \CLASS{{\tt ixforeach, iter}}  is being passed to an argument expecting a long union of classes, each of the form \CLASS{{\tt X, iter}}.
+ This is likely an instance where the user defined their own class, {\tt ixforeach}, and wanted to use the \code{iterators} package (the user called \code{iterator::iter} with a \CLASS{{\tt ixforeach}}, and it gained the class tag {\tt iter} on return).
+ As we mentioned, accounting for object-orientation in the type system is beyond the scope of this work, and such an inclusion would allow us to better type situations like this.
+
+ Table~\ref{tbl:fail-breakdown-2-kaggle} mirrors Table~\ref{tbl:fail-breakdown-2} in showing contract violations on the most frequently occurring argument types.
+ Here, our manual analysis has revealed similar failure patterns.
+ In the case of vectors being passed to scalars, we find functions which can take vectors but were only tested with scalars (e.g., \code{stringr::str_to_upper} which converts a vector of characters to upper case, and \code{dplyr::anti_join}, which can join by a vector of column names but was only ever tested with a scalar).
+ We also see possibly-NA character vectors being passed to NA-free character vectors.
+ These assertion failures arise from a lack of testing: the offending functions are \code{str_trim}, \code{str_sub}, and \code{str_replace_all} from the \code{stringr} package.
+ These functions are actually wrappers for other functions which have the correct argument type (\NAVEC\chr).
+
+ \subsubsection{Discussion}
+
+ Overall, the analysis discussed in this section has revealed two broad categories of contract assertion failures: those related to coercion, and those related to a class hierarchy.
+ Our type system does not account for coercion as coercion in R is ad hoc at best, and it is implemented on a function-by-function basis, even in the core R packages.
+ As for errors related to a class hierarchy, we aim to tackle this in future work, as designing a full fledged object-oriented type system for a language like R is outside of the scope of this work.
 
@@
 
-\AT{Talk about how we thought do to a different unification for class types.}
-
 As a closing point of discussion, we would like to highlight an alternate strategy for unifying class types that we considered.

### 2020-09-14 paper/tables/uni-typed-breakdown.tex
unlock 6 7
Add graph
new words in. let the edits commence

@@
   \centering
-\resizebox{\linewidth}{!}{
+  \resizebox{.5\linewidth}{!}{
     \begin{tabular}{lrrr}
@@
       \midrule
-scalar & 28045 & 35.80 & 35.80\\
-class & 18510 & 23.63 & 59.43\\
-vector & 10296 & 13.14 & 72.58\\
-\rowcolor{lightgray}  \textcolor{black}{...} & \textcolor{black}{6611} & \textcolor{black}{8.44} & \textcolor{black}{81.02}\\
-\rowcolor{lightgray}  \textcolor{black}{any} & \textcolor{black}{5993} & \textcolor{black}{7.65} & \textcolor{black}{88.67}\\
+      scalar & 29673 & 35.06 & 35.06\\
+      class & 20308 & 23.99 & 59.05\\
+      vector & 11529 & 13.62 & 72.67\\
+      \rowcolor{lightgray}  \textcolor{black}{...} & \textcolor{black}{6611} & \textcolor{black}{7.81} & \textcolor{black}{80.48}\\
+      \rowcolor{lightgray} \textcolor{black}{null} & \textcolor{black}{6094} & \textcolor{black}{7.20} & \textcolor{black}{87.68}\\
         \addlinespace
-null & 4795 & 6.12 & 94.79\\
-\rowcolor{lightgray}  \textcolor{black}{list} & \textcolor{black}{3229} & \textcolor{black}{4.12} & \textcolor{black}{98.91}\\
-\textasciicircum{}vector & 461 & 0.59 & 99.50\\
-environment & 390 & 0.50 & 100.00\\
+   \rowcolor{lightgray}  \textcolor{black}{any} & \textcolor{black}{6011} & \textcolor{black}{7.10} & \textcolor{black}{94.78}\\
+      \rowcolor{lightgray}  \textcolor{black}{list} & \textcolor{black}{3400} & \textcolor{black}{4.02} & \textcolor{black}{98.79}\\
+      \textasciicircum{}vector & 612 & 0.72 & 99.52\\
+      environment & 408 & 0.48 & 100.00\\
       \bottomrule
     \end{tabular}}
+  \caption{Singleton Type Categories}\label{fig:singlefreq}
 \end{table}

### 2020-09-14 paper/main.tex
major edits, minor edit pass now

@@
 \usepackage{booktabs} % To thicken table lines
+\usepackage{makecell}
 
@@
 
-\renewcommand{\null}{\texttt{\textbf{null}}\xspace}
+\newcommand{\nul}{\texttt{\textbf{null}}\xspace}
 \newcommand{\scalar}{\texttt{\textbf{scalar}}\xspace}
@@
 \T& ::= & \any          & \it top type\\
-  & |   & \null         & \it null type\\
+  & |   & \nul         & \it null type\\
   & |   & \environment  & \it environment type \\
@@
 \paragraph{Nullables} 
-The \null type is inhabited by a singleton \code{NULL} value often used as a
+The \nul type is inhabited by a singleton \code{NULL} value often used as a
 sentinel or a stand in for arguments with no value.  To capture this
@@
   \LIST{\T} \k{~|~} $\LIST{T'}$ & $\Rightarrow$ &  \LIST\T & \it{iff} $T' <: T$  \\
-   $\null \k{~|~} \VEC{\S_1} \k{~|~}\dots \VEC{\S_n}$  & $\Rightarrow$ &
+   $\nul \k{~|~} \VEC{\S_1} \k{~|~}\dots \VEC{\S_n}$  & $\Rightarrow$ &
   $ \NAVEC{\S_1} \k{~|~}\dots \NAVEC{\S_n}$ &  \\
@@
 minimize their size by removing redundant types, types that are subsumed by
-subtyping, immutable lists, and remove \null type to replace them with
+subtyping, immutable lists, and remove \nul type to replace them with
 nullables. For performance reasons, we bound lists to no more than five
@@
 \lgl               & 9,366 &      8.7 & 6,650,294 & 11.1 \\
-\null              &  8,799 &      8.0 & 2,187,611 & 3.7 \\
+\nul              &  8,799 &      8.0 & 2,187,611 & 3.7 \\
 \chr               &  8,727 &      8.0 & 2,564,207 & 4.3 \\
@@
 
-\AT{TODO: Update plots, reread and readjust wording and examples since some tables have changed.}
-
 The first part of our evaluation attempts to shed light on how good a fit
@@
 
-The next step, is shown in Table~\ref{fig:singlefreq}, the table provides a
-breakdown of types occuring in arguments without a union. Scalar, class and
+Table~\ref{fig:singlefreq} provides a
+breakdown of types occurring in arguments without a union. Scalar, class and
 vector are the most common type categories. The shaded rows correspond to
-polymorphic types. Removing those gives us \CountMonomorphicPositions
+polymorphic types. 
+When an argument's type is \nul, we say that the argument is polymorphic due to a limitation in our analysis:
+In R, it is common for programmers to include default values for arguments, and in many cases this value happens to be {\tt NULL}.
+Our type analysis will report a {\tt null} type for these arguments if they are {\it never passed a value} during testing.
+We interpret these instances of \nul as polymorphic to capture that we cannot be sure of the actual type.
+
+Removing the aforementioned instances of polymorphism gives us \CountMonomorphicPositions
 (\PercMonomorphicPositions) monomorphic positions in a corpus of
-\CountPositions parameters. With close to 65\% of monomorphic argument or
+\CountPositions parameters. With close to 60\% of monomorphic argument or
 return values, it is fair to say that even a simple type language provides
 significant benefits.
-
-We would like to point out the number of occurrences of the singleton {\tt null} type, and discuss why we treat it as an {\it any}-type.
-In R, it is common for programmers to include default values for arguments, and in many cases this value happens to be {\tt NULL}.
-Our type analysis will report a {\tt null} type for these arguments if they are never passed a value during testing.
-Thus, we interpret these as cases where we cannot be sure of the argument.
 
@@
 monomorphic. The remaining \PercPolymorphicFunctions
-(\CountPolymorphicFunctions) have at least one polymorphic parameter or
+(\CountPolymorphicFunctions) have at least one union or polymorphic parameter or
 return type.  Figure~\ref{fig:funpoly} shows the distribution of functions
-against the number of polymorphic arguments. Finally, we count
+against the number of polymorphic arguments. Finally, we count that
 \CountMonomorphicPackages out of the 412 packages export only monomorphic
@@
 \paragraph{NAs}
-Our data supports making the presence of {\NA}s explicit. Only 1,058 (of
-0.9\%) of arguments are marked as possibly having {\NA}s, thus the
+Our data supports making the presence of {\NA}s explicit. Only 2923 (or
+2.78\%) of arguments are marked as possibly having {\NA}s, thus the
 overwhelming majority of types appear to be {\NA}-free.  In practice,
@@
 \begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
-  binom.profile <- function(x, n, conf.level=0.9, maxsteps=50,...) {
+  binom.profile <- function(x, n, conf.level=0.9, maxsteps=50, ...) {
     xn <- cbind(x = x, n = n)
@@
     n <- xn[ok, "n"]
+    # ...
+  }
 \end{lstlisting}
@@
 The data also suggests that programmers often use scalars, and do
-dimensionality checks on their data.  In our data 31,533 (or 29\%) of the
+dimensionality checks on their data.  In our data 25,064 (or 33.33\%) of the
 arguments are scalar types. While not completely surprising, this is a
@@
     if ( m < n )  stop( "length of argument x is less than n" )
+    # ...
+  }
 \end{lstlisting}
@@
 \paragraph{Nullables}
-The number of argument which may be \c{NULL} is 3,165 (or 3.5\%). This is a
+The number of argument which may be \c{NULL} is 5057 (or 4.44\%). This is a
 relatively small number of occurrences, but it is worth expressing the
@@
 \typetracer assigns the type \CLASS{\texttt{function}} to function values.
-The number of positions that receive a function is 1,831, which is just
-2\% of all the positions for which we infer types. Given the small number of
+The number of positions that receive a function (possibly as part of a union) is 1,705, which is just
+1.51\% of all the positions for which we infer types. Given the small number of
 occurrences, it is not worth complicating the inferred types with a complete
@@
 \paragraph{Matrices}
-Matrices are instances of the eponymous class. They have a \code{dims}
+Matrices are instances of the eponymous class, representing 10.71\% of all classes occurring in types. They have a \code{dims}
 attribute indicating dimensions, and while not codified in the language
@@
     if (!is.matrix(x))   .Defunct(msg = sprintf("'x' should be a matrix.)
+    # ...
+  }
 \end{lstlisting}
@@
 One of the most popular classes in R is the \code{data.frame} class, making
-up nearly 8\% of observed classes.  Data frames and the derivative
-\code{tibble} and \code{data.table} types underpin many popular uses of R.
+up  8.15\% of observed classes.  Data frames and the derivative
+\code{tibble} and \code{data.table} types underpin much of the idiomatic usage of R.
 One way to deal with data frames is through the struct type, with a named
@@
 column names to functions which operate on their data frames.  In base R,
-idiomatic data frame use is to use string column names to select rows from
+typical data frame use is to use string column names to select rows from
 the frame (unless only a single column is of interest, wherein the \code{$}
 syntax is appropriate).
+In sum, data frames are a popular class of R values, and have spawned many derivative data types, such as tibbles and data tables.
+We include a \CLASS{{\tt data.frame}} type to cover most use-cases, and we leave a richer type for future work on a full fledged object-oriented type system.
 
@@
 
-How robust are the inferred types? To measure this, we conducted another
+We now ask {\it how robust are the inferred types}? To measure this, we conducted another
 large-scale experiment: for each package in the corpus, using the inferred type
@@
 \PERCSUCCFUNNOSTHREE if we discount S3 object method dispatch. Overall, these
-numbers are promising, and suggest that the type signatures are robust.
+numbers are promising, and suggest that the type signatures are indeed robust.
 
 We breakdown the failed assertions by type in
-Table~\ref{tbl:fail-breakdown-0}.  Accounting for 27\% of assertion
-failures, is when \CLASS{\texttt{matrix}} is passed where a \VEC\dbl is
+Table~\ref{tbl:fail-breakdown-0}.  Accounting for 36.36\% of assertion
+failures, is when a \VEC\dbl is passed where a \CLASS{\texttt{matrix}} is
 expected.  Considering these types, we might imagine them to be compatible,
 however not allowing this coercion was a deliberate design decision:
-coercion from vectors to matrices is ad hoc at best, and surely not a
-practice codified in the language.  In a similar vein, another popular
+coercion from matrices to vectors is ad hoc at best, and surely not a
+practice codified in the language. For instance, if it is determined that we should coerce the matrix it would need to be flattened, and should this be done row-wise or column-wise?
+
+In a similar vein, another popular
 failing assertion is checking if a ${\bf dbl}[]$ has type ${\bf int}[]$,
@@
 them, and it is not always the case that the coercions are safe to perform.
+
+The second row of the table is exemplary of a pattern where vectors are passed when scalars are expected.
+In these cases, the functions exhibiting these assertion failures were under-tested, and can operate just as well on vectors of values.
+As an example, this failure occurred in functions from the \code{lubridate} package which provides date/time functionality.
+Many functions, e.g., \code{date_decimal} and \code{make_datetime} turn doubles into \CLASS{{\tt POSIXct, POSIXt}} (which are dates in R), and they can easily operate on vectors of doubles, producing lists of dates.
+
+Finally, we point out that assertion failures of, e.g., \CLASS{{\tt data.frame}} values being passed to \CLASS{{\tt data.frame, tbl, tbl\_df}} arguments and \CLASS{{\tt xml\_node}} values being passed to an argument expecting other XML-like classes are related to our simplified take on class types.
+Our type system does not encode user-defined subtyping and coercion, which could help address these mismatches.
 
@@
 \toprule
-Passed & Arg Type & Occurrences & \% Total & Cumulative \%\\
+Passed & Arg Type & Occurrences & \% Total & Cumul. \%\\
 \midrule
 \VEC\dbl & \CLASS{\texttt{matrix}} & 705,036 & 36.36 & 36.36 \\ 
-\VEC\dbl & \VEC\int & 189,800 & 9.79 & 46.15 \\ 
+\VEC\dbl & \dbl & 189,800 & 9.79 & 46.15 \\ 
 \chr & \CLASS{\texttt{bignum}} | \VEC\raw & 100,100 & 5.16 & 51.31 \\
-\CLASS{{\tt simpleError, error, condition}} & \CLASS{{\tt data.frame}} | \CLASS{{\tt matrix}} | \CLASS{{\tt randomForest}} | \VEC\dbl & 78,197 & 4.03 & 55.34 \\ 
+\CLASS{{\tt simpleError, error, condition}} & \makecell[tl]{\CLASS{{\tt data.frame}} | \CLASS{{\tt matrix}} | \\ \CLASS{{\tt randomForest}} | \VEC\dbl} & 78,197 & 4.03 & 55.34 \\ 
   \CLASS{{\tt data.frame}} & \CLASS{{\tt data.frame, tbl, tbl\_df}} & 58,809 & 3.03 & 58.37 \\ 
+  \addlinespace
   \CLASS{{\tt matrix}} & \CLASS{{\tt timeSeries}} & 53,482 & 2.76 & 61.13 \\ 
@@
   \VEC\dbl & \CLASS{{\tt data.frame}} & 31361 & 1.62 & 66.13 \\ 
-  \CLASS{{\tt xml\_node}} & \CLASS{{\tt xml\_missing}} $|$ \CLASS{{\tt xml\_nodeset}} $|$ \CLASS{{\tt xml\_document, xml\_node}} & 30,330 & 1.56 & 67.70 \\ 
+  \CLASS{{\tt xml\_node}} & \makecell[tl]{\CLASS{{\tt xml\_missing}} $|$ \CLASS{{\tt xml\_nodeset}} $|$ \\ \CLASS{{\tt xml\_document, xml\_node}}} & 30,330 & 1.56 & 67.70 \\ 
 \bottomrule
@@
 \VEC\dbl was passed to 15 different function parameters expecting a
-\CLASS{\texttt{matrix}}, of which there are 1394 in total: 1\% of
+\CLASS{\texttt{matrix}}, of which there are 1522 in total: 18 (1.18\%) of these
 \CLASS{\texttt{matrix}}-typed parameters were passed \VEC\dbl values.
@@
 \VEC\int-typed arguments have their contracts violated by \VEC\dbl-typed
-values.  Rows 5 and 6 are interesting: we see that rather often arguments
-expecting \CLASS{\texttt{timeSeries}} data are passed \VEC\dbl or
+values.  Row six is interesting: we see that rather often arguments
+expecting \CLASS{\texttt{timeSeries}} data are passed
 \CLASS{\texttt{matrix}} values.  This is a quirk of the \code{timeSeries}
@@
 to time series in an ad hoc manner.  Note that the code coverage of the
-\code{timeSeries} tests, examples, and vignettes on the \code{timeSeries}
-package code is only 58\%, which is one possible explanation of why these
+\code{timeSeries} tests, examples, and vignettes package code is only 58\%, which is one possible explanation of why these
 contract failures are occurring: the types that \typetracer generates are
@@
 Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
+\midrule
 \VEC\dbl & \CLASS{\texttt{matrix}} & 18 & 1522 & 1.18 \\ 
-\VEC\dbl & \VEC\int  & 66 & 5865 & 1.13 \\ 
+\VEC\dbl & \dbl  & 66 & 5865 & 1.13 \\ 
 \chr & \CLASS{\texttt{bignum}} | \VEC\raw & 1 & 3 & 33.33 \\ 
-\CLASS{{\tt simpleError, error, condition}} & \CLASS{{\tt data.frame}} | \CLASS{{\tt matrix}} | \CLASS{{\tt randomForest}} | \VEC\dbl & 2 & 2 & 100 \\ 
+\CLASS{{\tt simpleError, error, condition}} & \makecell[tl]{\CLASS{{\tt data.frame}} | \CLASS{{\tt matrix}} | \\ \CLASS{{\tt randomForest}} | \VEC\dbl}  & 2 & 2 & 100 \\ 
   \CLASS{{\tt data.frame}} & \CLASS{{\tt data.frame, tbl, tbl\_df}} & 23 & 196 & 11.73 \\ 
+  \addlinespace
   \CLASS{{\tt matrix}} & \CLASS{{\tt timeSeries}} & 21 & 39 & 53.85 \\ 
@@
   \VEC\dbl & \CLASS{{\tt data.frame}}  & 6 & 1025 & 0.59 \\ 
-  \CLASS{{\tt xml\_node}} & \CLASS{{\tt xml\_missing}} $|$ \CLASS{{\tt xml\_nodeset}} $|$ \CLASS{{\tt xml\_document, xml\_node}}  & 1 & 1 & 100 \\ 
+  \CLASS{{\tt xml\_node}} & \makecell[tl]{\CLASS{{\tt xml\_missing}} $|$ \CLASS{{\tt xml\_nodeset}} $|$ \\ \CLASS{{\tt xml\_document, xml\_node}}}   & 1 & 1 & 100 \\ 
 \bottomrule
@@
 type of the value that violated that contract.  The first row of the table
-reads: 44 function arguments with \VEC\int type are passed \VEC\dbl instead,
-and 657 arguments have \VEC\int type.  Had we failed to capture some key
+reads: 31 function arguments with \VEC\int type are passed \VEC\dbl instead,
+and 624 arguments have \VEC\int type, representing a failure rate of 4.97\%.
+
+Had we failed to capture some key
 usage pattern of R with our type annotation framework, we would likely see
@@
 instead.  The most frequent argument type violation pattern in that of
-\CLASS{\texttt{data.frame}} values passed to arguments expecting tibbles,
-which are an extension of \CLASS{\texttt{data.frame}} built in the
-\code{tidyverse} package suite: this occurs in only 22\% of such arguments,
+\CLASS{{\tt matrix}}, \VEC\dbl, and \CLASS{\texttt{data.frame}} values passed to arguments expecting \CLASS{{\tt timeSeries}},
+This occurs in only 35.90\% of such arguments,
 and represents cases where tests did not adequately cover all valid function
 inputs.  Separate from the issue of testing, we can capture this behaviour
-with user-defined subtyping, as data frames are subtypes of tibbles, but
-this necessitates a proper object-oriented type system for R which is
-outside the scope of this work.
+with user-defined subtyping or coercion, as the data types which were passed are readily convertible to \CLASS{{\tt timeSeries}}.
 
@@
 \dbl & \int & 21 & 519 & 4.05\\
-\VEC\chr & \chr | \null & 10 & 256 & 3.91 \\
+\VEC\chr & \chr | \nul & 10 & 256 & 3.91 \\
 \NAVEC\lgl & \VEC\lgl & 8  & 219 & 3.65 \\
@@
 % \CLASS{\texttt{matrix}} & \CLASS{\texttt{array}} & 13 & 70 & 18.57\\
-% \NAVEC\lgl & \lgl | \null & 9 & 49 & 18.37\\
+% \NAVEC\lgl & \lgl | \nul & 9 & 49 & 18.37\\
 % \VEC\chr & \LIST{\VEC\chr} & 20 & 109 & 18.35\\
@@
 
- First, we were curious about how many \null-typed values flowed into non-nullable arguments, and we found that they accounted for \NUMNULLASSERTERRORS of contract assertion failures in \NUMFUNSWITHNULLASSERTERROR functions.
+ First, we were curious about how many \nul-typed values flowed into non-nullable arguments, and we found that they accounted for \NUMNULLASSERTERRORS of contract assertion failures in \NUMFUNSWITHNULLASSERTERROR functions.
  We manually inspected some of the offending functions and observed three main patterns.

### 2020-09-14 paper/tables/uni-typed-breakdown.tex
major edits, minor edit pass now

@@
       \midrule
-      scalar & 29673 & 35.06 & 35.06\\
-      class & 20308 & 23.99 & 59.05\\
-      vector & 11529 & 13.62 & 72.67\\
-      \rowcolor{lightgray}  \textcolor{black}{...} & \textcolor{black}{6611} & \textcolor{black}{7.81} & \textcolor{black}{80.48}\\
-      \rowcolor{lightgray} \textcolor{black}{null} & \textcolor{black}{6094} & \textcolor{black}{7.20} & \textcolor{black}{87.68}\\
+      scalar & 35064 & 33.33 & 33.33 \\
+      class & 24256 & 23.06 & 56.39 \\
+      vector & 13025 & 12.38 & 68.77 \\
+      \rowcolor{lightgray}  \textcolor{black}{...} & \textcolor{black}{9142} & \textcolor{black}{8.69} & \textcolor{black}{77.46}\\
+      \rowcolor{lightgray} \textcolor{black}{null} & \textcolor{black}{7694} & \textcolor{black}{7.31} & \textcolor{black}{84.77}\\
         \addlinespace
-   \rowcolor{lightgray}  \textcolor{black}{any} & \textcolor{black}{6011} & \textcolor{black}{7.10} & \textcolor{black}{94.78}\\
-      \rowcolor{lightgray}  \textcolor{black}{list} & \textcolor{black}{3400} & \textcolor{black}{4.02} & \textcolor{black}{98.79}\\
-      \textasciicircum{}vector & 612 & 0.72 & 99.52\\
-      environment & 408 & 0.48 & 100.00\\
+   \rowcolor{lightgray}  \textcolor{black}{any} & \textcolor{black}{7614} & \textcolor{black}{7.24} & \textcolor{black}{92.01}\\
+      \rowcolor{lightgray}  \textcolor{black}{list} & \textcolor{black}{3558} & \textcolor{black}{3.38} & \textcolor{black}{95.39}\\
+      \textasciicircum{}vector & 2923 & 2.78 & 98.17\\
+      function & 1427 & 1.36 & 99.52 \\
+      environment & 500 & 0.48 & 100.00\\
       \bottomrule

### 2020-09-14 paper/main.tex
updated plots, finalizing edits

@@
  \newcommand{\PROPFUNSWITHFAILEDASSERTKAGGLE}{15.98\%\xspace}
- \newcommand{\NUMTOTALASSERTIONSKAGGLE}{1,9038,496\xspace}
+ \newcommand{\NUMTOTALASSERTIONSKAGGLE}{19,038,496\xspace}
  \newcommand{\NUMASSERTEDFUNCTIONSKAGGLE}{970\xspace}
@@
 which must generate native code or catch programming errors.  In particular,
-R has no type annotations, all operations are dynamically checked at
-run-time. The starting point for our work are the twin questions, \emph{what
+R has no type annotations, and all operations are dynamically checked at
+run-time. The starting point for our work are the twin questions: \emph{what
   expressive power is needed to accurately type R code?} and \emph{which
@@
 functions from \NUMPACKAGES packages among the most widely used open source
-R libraries.
+R libraries. We then conduct an evaluation on \NUMPKGSEVAL clients of these packages, as well as on end-user code found on the Kaggle competition website.
 \end{abstract} \maketitle
@@
 programming tasks.  Unfortunately, intuition may fail when looking at
-domain-specific languages as these languages were designed for a specific
+domain-specific languages designed for a specific
 group of users to solve very specific needs. This is the case of the data
@@
 interactive and exploratory programming style. These include, to name a few,
-the lack of type annotations, the ability to use syntactic shortcut, and the
+the lack of type annotations, the ability to use syntactic shortcuts, and the
 automatic conversion between data types. While these choices have led to a
@@
 Retrofitting a type system to the R language would increase our assurance in
-the result of data analysis. But, we are faced with two challenges. First,
+the result of data analysis, but we are faced with two challenges. First,
 it is unclear what would be the \emph{right} type system for a language as
@@
 checker that weaves the types around their respective functions, and use a
-large number of clients of the target packages for validation. The contract
+both a large number of clients of the target packages as well as end-user code for validation. The contract
 system can be used by the R community to experiment with type signatures and
@@
 Our tools are open source and publicly available on GitHub (link anonymize),
-all results in this paper are reproducible and will be submitted for
-artifact evaluation should this paper be accepted.
+all results in this paper are reproducible and were submitted for
+artifact evaluation.
 %
@@
 intersperse typed and untyped code in a fine-grained manner.  Other work
-looks at retrofitting types onto Lua~\cite{typedlua2014}.
+looks at retrofitting types onto Lua~\cite{typedlua2014} which account for the myriad ways Lua programmers use tables.
 
@@
 
-In this paper, we will attempt infer type signature for functions. 
-We begin by introducing some relevant concepts.
-
-The basic building blocks of R are vectors of primitives.
-Characters (strings), integers, logicals (booleans), doubles, complex numbers, and raw bytes make up R's primitive values.
+The basic building blocks of R are vectors of primitives (primitives in R are characters (strings), integers, logicals (booleans), doubles, complex numbers, and raw bytes).
 When we say that R is {\it vectorized}, we mean that primitive values are {\it always} vectors.
-For example, in R one might see \code{2.5} and think it to be a scalar double, but it is in actuality a unit-length vector containing a single double.
+For example, in R one might see the literal \code{2.5} and think it to be a scalar double, but it is in actuality a unit-length vector containing a single double.
 Vectors can also be explicitly constructed using the vector constructor \code{c(...)}, 
 For example, \code{c(1L, 2L, 3L)} creates a vector of three integers (in R, integer literals are denoted by an \code{L}).
-
-The main object of our work is R functions, which have a number of quirks.
-Arguments to functions in R can be assigned default values (which can be arbitrary expressions!), and R functions can support variable argument lists, and functions can be called by specifying argument names explicitly in the call.
+To illustrate that scalars are vectors, in R \code{(c(2.5) == 2.5) == TRUE}.
+
+The main object of our work is R functions, and functions in R have a number of quirks.
+Arguments can be assigned default values (which can be arbitrary expressions!), and functions can support variable argument lists, and functions can be called by specifying argument names explicitly in the call.
 To illustrate all of this in a single example, consider:
@@
 \noindent
-This function four formal parameters, \k x, \k{...}, \k y and \k z. Argument
+This function has four formal parameters, \k x, \k{...}, \k y and \k z. Argument
 \k x can be bound positionally or passed by name. The vararg argument,
 \k{...}, is always passed positionally. The remaining two arguments must be
-passed by name.  Arguments \k y and \k z have default values, in the case of
+passed by name as they are preceded by the \k{...} parameter.  Arguments \k y and \k z have default values, in the case of
 \k z this is a constant, but \k{y}'s default value is an expression that
@@
  > f(1)    
- [1] 2             % a double vector, y is 0, ... is missing
+   [1] 2             # a double vector, y is 0, ... is missing
  > f(1,2) 
- [1] 4             % a double vector, y is 0, ... is 2
+   [1] 4             # a double vector, y is 0, ... is 2
  > f(1,2,3)
- [1] 4 5           % a double vector, y is 0, ... is 2, 3
+   [1] 4 5           # a double vector, y is 0, ... is 2, 3
  > f(2,3,x=1)
- [1] 4 5           % a double vector, y is 0, ... is 2, 3
+   [1] 4 5           # a double vector, y is 0, ... is 2, 3
  > f(x=1, y= 1)
- [1] 2             % a double vector, y is 1, ... is missing
+   [1] 2             # a double vector, y is 1, ... is missing
  > f(x=1, z= 1)
- numeric(0)        % a double vector of length 0, y is NULL
+   numeric(0)        # a double vector of length 0, y is NULL
  > f(1L,2L,y=1L)
- [1] 4             % an integer vector, y is integer 1, ... is integer 2
+   [1] 4             # an integer vector, y is integer 1, ... is integer 2
  > f(1, y=c(1,2))
- [1] 1 2           % a double vetor, y is 1, 2, ... is missing
+   [1] 1 2           # a double vetor, y is 1, 2, ... is missing
 \end{lstlisting}
@@
  > f(c1)
- [1] 1             % a double vector, x is complex, y is double
+   [1] 1             # a double vector, x is complex, y is double
  > f(c1,y=c2)
- [1] 2+1i          % a complex vector, x and y are complex
+   [1] 2+1i          # a complex vector, x and y are complex
  > l1 <- list(1,2)
@@
  > f(l1) 
- [1] 1             % a double vector, x is a list of doubles
+   [1] 1             # a double vector, x is a list of doubles
  > f(l1,y=l2)
- [[1]]             % a list of complex, x is a list of doubles
- [1] 1+2i          %                      y is a list of complex
+   [[1]]             # a list of complex, x is a list of doubles
+   [1] 1+2i          #                      y is a list of complex
 
@@
 
-R has one builtin notion of type that can be queried by the \k{typeof}
+R has a builtin notion of type that can be queried by the \k{typeof}
 function. Figure~\ref{rtypes} lists all of the builtin types that are
-provided by the language, which are the possible return values of
+provided by the language; these are the possible return values of
 \k{typeof}. There is no intrinsic notion of subtyping in R, but in many
-context a \k{logical} will convert to \k{integer}, and an \k{integer} will
+contexts a \k{logical} will convert to \k{integer}, and an \k{integer} will
 convert to \k{double}.  Some odd conversions can occur in corner cases, such
@@
 
-All vectorized data types have a distinguished missing value denoted by \code{NA}. 
+All vectorized data types have a distinguished missing value denoted by \code{NA} (for ``not available''). 
 The default type of \code{NA} is \k{logical}. 
-We can see that \code{typeof(NA)=="logical"}, but NA inhabits every type: \code{typeof(c(1,NA)[2])=="double"}.
+We can see that \code{typeof(NA)=="logical"}, but \k{NA} inhabits every type: \code{typeof(c(1,NA)[2])} \k{==} \k{"double"}.
 In addition to \code{NA}, R has a \code{NULL} value, and to understand why R has both a \code{NULL} value and \code{NA} values it is useful to remember that R is primarily designed for statistics and data science.
 In experimental data science, it is useful to have a notion of a ``missing observation'', which is often NaN (not a number, or similar) in, e.g., MATLAB and Python.
-Since vectors are unityped in R, it {\it must} have a ``NaN'' value for each primitive type, hence the \code{NA}s, while there is only a single \code{NULL} null-typed value in the language to encompass all other cases.
+Since vectors are unityped in R, it {\it must} have a ``NaN'' value for {\it each} primitive type.
+There single \code{NULL} null-typed value in the language to encompass all other cases.
+In a sense, you can think of \code{NA} as representing a missing data point, and \code{NULL} as a missing data {\it set}.
 
@@
 
-With one exception all vectorized data types are monomorphic, the exception
-is the \k{list} type which can hold values of any other type including
+With one exception, all vectorized data types are monomorphic.
+The exception is the \k{list} type which can hold values of any other type including
 \k{list}. For all monomorphic data types, attempting to store a value of a
 different type will cause a conversion. Either the value is converted to the
-type of vector, or the vector is converted to the type of the value.
+type of the vector, or the vector is converted to the type of the value.
 
@@
 and have added {\it attributes}. The best way to think of attributes is as
-an optional map from name to values that can be attached to any object.
+an optional map from names to values that can be attached to any object.
 Attributes are used to encode various type structures. They can be queried
@@
 pair of numbers as value: \code{attr(x,"dim")<-c(2,2)}.  From that point,
-arithmetic functions will treat \k{x} as a 2x2 matrix. Another attribute
+arithmetic functions will treat \k{x} as a 2x2 matrix.
+
+Another attribute
 that can be set is the \k{class}.  This attribute can be bound to a list of
-class names. For instance, \code{class(x)<-"human"}, set the class of \k{x}
-to be \k{human}.  Attributes are thus used for object-oriented
-programming. The S3 object system support single dispatch on the class of
+class names. For instance, \code{class(x)<-"human"} will set the class of \k{x}
+to be \k{human}.  The main usage of attributes is for object-oriented
+programming. There are three object-orientation frameworks in R: S3, S4, and R5. 
+The S3 object system support single dispatch on the class of
 the first argument of a function, whereas the S4 object system allows
-multiple dispatch (on all arguments). Some of the most widely used data
-type, such as data frames, leverage attributes. A data frame, for instance,
-is a list of vectors with a class and a column name attribute.
-
-Scalar data types include the distinguished \k{NULL} value, which is also of
-type \k{NULL}, instance of classes written using the S4 object system,
-closures and environments.  The implementation of R has a number of other
+multiple dispatch (on all arguments). R5 allows for users to define objects
+in a more imperative style. Some of the most widely used data
+types leverage attributes, e.g., data frames and matrices. A data frame, for instance,
+is a list of vectors with a class and a column name attribute, and matrices are vectors with a \k{dims} attribute.
+
+Scalar data types include the distinguished \k{NULL} value, instances of classes written using the S4 object system, closures, and environments.  The implementation of R has a number of other
 types listed in Figure~\ref{rtypes} for reference.
@@
 
-This section sets out to propose a candidate design for a type language to
-describe the arguments and return values of functions in R.  The goal is not
-to get a final design but rather a starting point for an iterative process.
+In this section, we set out to propose a candidate design for a type language to
+describe the arguments and return values of R functions.  The goal is not
+to propose a final design, but rather to propose a starting point for an iterative design process.
 Fig.~\ref{types} presents our type language. An early design choice was to
-stay close to the R language and only depart in small, hopefully,
+stay close to the R language and only depart in small and hopefully
 non-controversial ways.  Functions types have the form \FUN{A_1, \dots,
@@
 \NA. Each of basic types has its specific \NA. Thus there is an \int \NA as
-well as a double \NA, and they are different.  The type system allows to
-distinguish between vectors that may contain {\NA}s(written \NAVEC\int) and
+well as a \dbl \NA, and they are different.  The type system allows one to
+distinguish between vectors that may contain {\NA}s (written \NAVEC\int) and
 those who are guarantee to be \NA-free (\VEC\int).  A \NA-free vector can be
@@
 \paragraph{Unions} We support untagged unions of types written
-$\T_1 \k{~|~} \T_n$.
+$\T_1 \k{~|~} ... \k{~|~} \T_n$.
 
@@
 sentinel or a stand in for arguments with no value.  To capture this
-behavior, we introduce a nullable type \code{?}\T.  The difference between
+behavior, we introduce a nullable type \code{?}\T.  The functional difference between
 the \NA and \code{NULL} values is that the latter cannot be stored in
@@
 \code{as.list}, and lists to vectors with \code{unlist} (coercions may
-ensue). The list type is parameterize, \LIST\T, when the list is
-heterogeneous \T defaults to \any.
+ensue). The list type is parameterize: \LIST\T.
 
 \paragraph{Classes} 
-R has more than one notion of type. Values can be attributed, one important
+R has more than one notion of type. As we mentioned, values can be attributed, and a very important
 attribute is the class of a value.  A class is a list of names that are used
@@
 While this task is seemingly simple, the details and their proverbial devil
-are surprisingly tricky to get right and to scale to large, long running,
+are surprisingly tricky to get right and to scale to large, long running
 programs.  To avoid starting from scratch, our implementation reuses \rdt,
@@
 machinery to collect information from the R language run-time.  Types are
-conmpleted during post-processing, and rely on the detailed information made
+completed during post-processing, and rely on the detailed information made
 available by these low-level reflection mechanisms.
@@
 are performing a dynamic analysis, we conservatively type them as
-\LIST{\any}.
+\any.
 
@@
 unwinding process and record a special \k{jumped} return type for function
-returns that are skipped.
+returns that are skipped. As we cannot be sure of the intended return value, 
+these \k{jumped} become \any types.
 
@@
 by pushing the call onto a stack.  The calls themselves store a trace
-object, it is that object that holds the type information. As R can perform
+object that holds the type information. As R can perform
 single or multiple dispatch on function arguments depending on their class,
@@
 subtyping, immutable lists, and remove \nul type to replace them with
-nullables. For performance reasons, we bound lists to no more than five
-types in their union, above that they become \LIST{any}.  Functions are alos
-limited to twenty arguments. Higher-order functions are approximated by \any
-to \any.
+nullables.  Functions are
+limited to twenty arguments as a simplifying assumption. Higher-order functions are approximated by \any
+to \any by \typetracer.
 
@@
 
-One way to validate the inferred type annotations is to check that different
-programs using one of the functions for which a type signature has been inferred
-respect the signature. For this purpose, we developed \contractr, an R package
+One way to validate an inferred function type is to check that it is respected in a different program which uses the function.
+For this purpose, we developed \contractr, an R package
 that allows one to decorate function with assertions. We use it to insert type
-checks of the arguments and return values.
+checking of arguments and return values.
 
@@
 removes contracts from all functions and restores them to their original state
-when it is unloaded. The type signatures can be provided externally, in a file,
+when it is unloaded. The type signatures can be provided in an external file,
 thus avoiding the need to change the source code of checked packages. Type
@@
 
-The \contractr provides an extensive API that allows users to explicitly insert
+\contractr provides an extensive API that allows users to explicitly insert
 contracts during interactive development by supplying the type signature as a
@@
 \paragraph{Checking the values} A number of properties can be checked
-by a simple tag check, namely whether a value is a: null, environment, vector
+by a simple tag check, namely whether a value is a null, environment, vector
 or scalar. Other properties require an inspection of the contents of a
-value: the absence of \NA or the type of the elements of list. Union types
-require checking for each of the alternatives. We do not check higher-order functions in this version of the tool. Checking
-them could be done by dynamically wrapping functions with checks. The usage
-of higher-order functions remains limited in R. When argument values are wrapped in promises (this is not always the case
+value, such as checking for the absence of \NA or the type of the elements of list. Union types
+require checking for each of the alternatives. As the usage
+of higher-order functions remains limited in R, we do not check higher-order functions in this version of the tool. Checking
+them could be done by dynamically wrapping functions with checks.  When argument values are wrapped in promises (this is not always the case
 due to compiler optimizations), in order to retain the non-strict semantics
@@
 longjumps which causes active function calls on the stack to be
-discarded. When they are discarded, their exit hooks are called.  But, these
-function do not have a return value to type-check. \contractr deals with
+discarded. When they are discarded, their exit hooks are called but they do not have a return value to type-check. \contractr deals with
 this problem by allocating a unique sentinel object which serves as the
@@
 \CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native
-code (C/Fortran). Figure~\ref{fig:corpus} shows these packages, the size of
+code (C/Fortran). Figure~\ref{fig:corpus} shows these packages: the size of
 the dots reflects the project's size in lines of code including both R and
@@
   \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the
-expression code coverage in percents and the y-axis gives the number of
-reverse dependencies in log scale. Dotted lines indicate means. Packages
+expression code coverage as a percentage and the y-axis gives the number of
+reverse dependencies in on log scale. Dotted lines indicate means. Packages
 with over \PackageSizeOutierRnd lines of code are annotated.
@@
   submissions a day~\cite{Ligges2017}} containing over \AllRCodeRnd and
-\AllNativeCodeRnd R and native code respectively. Unlike other open code
-repositories such as GitHub, CRAN is a curated repository. Each submitted
-package must abide to a number of well-formedness rules that are automatically
-checked asserting certain quality. Most relevant for this work is that all of
-the runnable code (including code snippets from examples and vignettes) is
-tested and only a successfully running package is admitted in the archive.
+\AllNativeCodeRnd lines of R and native code respectively. Unlike other open code
+repositories such as GitHub, CRAN is a curated repository where each submitted
+package must abide by a number of well-formedness rules that are automatically
+checked to assess package quality. 
+Notably, CRAN packages must have a set of {\it runnable} example, test, and vignette code which showcase package functionality.
+The code is run by CRAN, and only a successfully running package is admitted to the archive.
 
@@
 \AllCranRnd packages, we managed to install \AllLoadableRnd. The main reason for
-this is that \rdt is based on GNU-R 3.5.0 and some of the packages are no longer
-compatible with this version. Some packages also require extra native
+this is that \rdt (and, by extension, \typetracer) is based on GNU-R 3.5.0 and some of the packages are not compatible with this version. 
+Some packages also require extra native
 dependencies which were not present on our servers. We defined two criteria for
@@
 \begin{inparaenum}[(1)]
-  \item have a runnable code that covers a significant part of the package
+  \item the package must have a runnable code that covers a significant part of the package
   source code from which type signatures could be inferred, and
-  \item have some reverse dependencies that will allows us to evaluate the
+  \item the package must have some reverse dependencies that will allows us to evaluate the
   inferred types using the runnable code from these dependencies.
@@
 The reverse package dependencies were extracted from the package metadata using
-builtin function.
+built-in functions.
 
@@
 and vignettes (\CorpusVignettesCodeRnd). Running this code results on average in
-\CorpusMeanExprCoverage package code coverage (CRAN average is
+\CorpusMeanExprCoverage package code coverage (the average for all of CRAN is
 \AllMeanExprCoverage). Together, there is \CorpusRevdesRnd (on average
@@
 Together there are \CorpusFunctionsRnd defined R functions
-(\CorpusPublicFunctionsRnd are from packages' public API).
-\CorpusSThreeFunctionsRnd are S3 functions. Packages in the corpus define
+(\CorpusPublicFunctionsRnd are from the packages' public APIs).
+\CorpusSThreeFunctionsRnd are S3 functions, either S3 generics or S3 methods. Packages in the corpus define
 \CorpusSThreeClasses S3 classes.
-The code coverage was computed for each package using
-\covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R
-tool.  The reverse package dependencies were extracted from the package
-metadata using builtin function.
 
@@
 platform for data-science and machine-learning. The website allows people to
-share data-science competitions, data-analysis problems together with data, for
-which users try to find the best solution. The solutions, called \emph{kernels},
-are then shared back to the platform in the form of either plain scripts or as
+share data-science competitions and data-analysis problems together with data for
+which users try to find the best solution (something like a repository of hackathon or datathon code). The solutions, called \emph{kernels},
+are then posted on Kaggle either as plain scripts or as
 notebooks. One of the most popular competition is about predicting passenger
@@
 and misspelled package names. The final set contains \KaggleSuccess kernels with
-\KaggleSuccessCodeRnd lines of R code. The Kaggle kernels are used for additonal
+\KaggleSuccessCodeRnd lines of R code. The Kaggle kernels are used for additional
 validation of the inferred types.
@@
 makes up 6\% of arguments.  This all totals up to over 90\% of types.
-Table~\ref{ctb} reports on the 10 most frequent types occuring in the
+Table~\ref{ctb} reports on the 10 most frequent types occurring in the
 corpus.
@@
 process with ten representative signatures.  Many of the features of our
-type language are represented here.  Some signatures are telling of the
+type language are represented here, and some signatures are telling of the
 function's behaviour.  For example, consider \code{decrypt_envelope}: the
-firs three parameters of the function are byte arrays, and the fourth
+first three parameters of the function are byte arrays, and the fourth
 argument is an RSA key, used to decrypt some of the inputs, and the output
@@
 \begin{figure}[!h]  \centering
-  \includegraphics[width=.7\linewidth]{plots/union_frequency.pdf}
+  \includegraphics[width=.7\linewidth]{plots/argument_polymorphism.pdf}
   \caption{Size of unions}\label{fig:unionfreq}
@@
 Table~\ref{tbl:fail-breakdown-0}.  Accounting for 36.36\% of assertion
-failures, is when a \VEC\dbl is passed where a \CLASS{\texttt{matrix}} is
-expected.  Considering these types, we might imagine them to be compatible,
-however not allowing this coercion was a deliberate design decision:
-coercion from matrices to vectors is ad hoc at best, and surely not a
-practice codified in the language. For instance, if it is determined that we should coerce the matrix it would need to be flattened, and should this be done row-wise or column-wise?
+failures are cases where a \VEC\dbl is passed where a \CLASS{\texttt{matrix}} is
+expected.  Considering these types, we might imagine them to be compatible, as a vector is just a one-dimensional matrix.
+However, not allowing this coercion was a deliberate design decision, as coercion of this kind is ad hoc at best, and unfortunately not a practice codified in the language. 
+For example, if the vector has length $n$, should it be a $1 \times n$ or $n \times 1$ dimensional matrix?
 
@@
 Turning our attention now to arguments, we found that only
-\PROPARGSFAILINGASSERTS of parameters failed.
+\PROPARGSFAILINGASSERTS of argument types failed.
 Table~\ref{tbl:fail-breakdown-0} showed the runtime occurrences, but that
@@
 type of the value that violated that contract.  The first row of the table
-reads: 31 function arguments with \VEC\int type are passed \VEC\dbl instead,
+reads: 31 function arguments with \VEC\int type are passed \VEC\dbl values instead,
 and 624 arguments have \VEC\int type, representing a failure rate of 4.97\%.
@@
 Table~\ref{tbl:fail-breakdown-3}, which was obtained identically to
-Table~\ref{tbl:fail-breakdown-2} with a percentile of 80th percentile
+Table~\ref{tbl:fail-breakdown-2} except selecting arguments in the 80th percentile
 instead.  The most frequent argument type violation pattern in that of
-\CLASS{{\tt matrix}}, \VEC\dbl, and \CLASS{\texttt{data.frame}} values passed to arguments expecting \CLASS{{\tt timeSeries}},
+\CLASS{{\tt matrix}}, \VEC\dbl, and \CLASS{\texttt{data.frame}} values passed to arguments expecting \CLASS{{\tt timeSeries}}.
 This occurs in only 35.90\% of such arguments,

### 2020-09-15 paper/main.tex
2nd round fnl

@@
 \newcommand{\CLASS}[1]{\texttt{\textbf{class}}\ensuremath{\langle #1\rangle}\xspace}
-\newcommand{\VEC}[1]{#1\k{[]}\xspace}
-\newcommand{\NAVEC}[1]{\k{\string^}\!#1\k{[]}\xspace}
+\newcommand{\VEC}[1]{#1\k{[]}}
+\newcommand{\NAVEC}[1]{\k{\string^}\!#1\k{[]}}
 \newcommand{\NASCA}[1]{\k{\string^}\!#1\xspace}
@@
 programs do not teach R itself but simply introduce some of its key
-libraries---they have also created a language where errors can go
+libraries---they have also created a language where errors can easily go
 undetected.
@@
 the simple type language and identify limitations of the design. We validate
-the robustness of the extracted type signature by implementing a contact
+the robustness of the extracted type signature by implementing a contract
 checker that weaves the types around their respective functions, and use a
@@
 
-Our tools are open source and publicly available on GitHub (link anonymize),
+Our tools are open source and publicly available on GitHub\footnote{
+\typetracer: \url{https://github.com/PRL-PRG/propagatr}, \contractr: \url{https://github.com/PRL-PRG/contractr}
+},
 all results in this paper are reproducible and were submitted for
@@
 In this section, we introduce related work on extending dynamic languages
-with static type system and we give a short primer on the R programming
+with static type systems and we give a short primer on the R programming
 language.
@@
 the host languages and to foster a particular programming style. For
-example, Typed Racket emphasizes functional programming and support the
+example, Typed Racket emphasizes functional programming and supports the
 migration from untyped to fully typed code~\cite{tf-popl08},
 Hack~\cite{hack13} and TypeScript~\cite{BAT14} focus on typing
-object-oriented features of PHP and JavaScript, respectively. They allow to
+object-oriented features of PHP and JavaScript, respectively. They allow users to
 intersperse typed and untyped code in a fine-grained manner.  Other work
@@
 dynamic analysis techniques, namely that the results are only as good at the
-coverage of the source code, it allows to quickly test new design and
+coverage of the source code, it allows one to quickly test new design and
 quantify how much of a code base can be type-checked.  Other approaches that
@@
 type hints in Python 3.5~\cite{pythonspectypes}.  There is no previous work
-on types for R. We take inspiration in the above mentioned works but focus
+on types for R, besides a short sketch of issues the type system would need to tackle~\cite{turcotte2019towards}. We take inspiration in the aforementioned works but focus
 on adapting them to our target language.
@@
 programming language with a rather unusual combination of
-features~\cite{ecoop12} designed to ease learning by non-programmer and
+features~\cite{ecoop12} designed to ease learning by non-programmers and
 enable rapid development of new statistical methods.  The language was
@@
 \k z this is a constant, but \k{y}'s default value is an expression that
-depends on the value of \k z. The body of the function will add parameters
+depends on the value of \k z (if \k z is not zero, \k y will default to \k{NULL}). The body of the function will add parameters
 \k x and \k y with either the scalar 0 or the result of concatenating the
@@
    [1] 2             # a double vector, y is 0, ... is missing
- > f(1,2) 
+ > f(1, 2) 
    [1] 4             # a double vector, y is 0, ... is 2
- > f(1,2,3)
+ > f(1, 2, 3)
    [1] 4 5           # a double vector, y is 0, ... is 2, 3
- > f(2,3,x=1)
+ > f(2, 3, x=1)
    [1] 4 5           # a double vector, y is 0, ... is 2, 3
- > f(x=1, y= 1)
+ > f(x=1, y=1)
    [1] 2             # a double vector, y is 1, ... is missing
- > f(x=1, z= 1)
+ > f(x=1, z=1)
    numeric(0)        # a double vector of length 0, y is NULL
- > f(1L,2L,y=1L)
+ > f(1L, 2L, y=1L)
    [1] 4             # an integer vector, y is integer 1, ... is integer 2
@@
    [1] 1             # a double vector, x is complex, y is double
- > f(c1,y=c2)
+ > f(c1, y=c2)
    [1] 2+1i          # a complex vector, x and y are complex
- > l1 <- list(1,2)
- > l2 <- list(c1,c2)
+ > l1 <- list(1, 2)
+ > l2 <- list(c1, c2)
  > f(l1) 
    [1] 1             # a double vector, x is a list of doubles
- > f(l1,y=l2)
+ > f(l1, y=l2)
    [[1]]             # a list of complex, x is a list of doubles
@@
 \k{typeof}. There is no intrinsic notion of subtyping in R, but in many
-contexts a \k{logical} will convert to \k{integer}, and an \k{integer} will
-convert to \k{double}.  Some odd conversions can occur in corner cases, such
+contexts a \k{logical} will be coerced to \k{integer}, and an \k{integer} will
+be coerced to \k{double}.  Some odd conversions can occur in corner cases, such
 as \k{1<"2"} holds and \k{c(1,2)[1.6]} returns the first element of the
@@
   \vspace{2mm}
-  \includegraphics[width=\textwidth]{plots/analysis-pipeline.pdf}
+  \includegraphics[width=\textwidth]{plots/analysis-pipeline-new.pdf}
   \caption{The Analysis Pipeline. (1) the rather long time is because we had to
     extract metadata from all CRAN packages before we could compose our
-    \PACKAGES package corpus.}
+    \PACKAGES package corpus. Here CRAN refers to the Comprehensive R Archive Network, R's repository of packages and libraries.}
   \label{fig:pipeline}
@@
 value. Internal types are translated to names in the proposed type
-system. The next bit of informatio is the \emph{class}, an optional list of
+system. The next bit of information is the \emph{class}, an optional list of
 names that may be absent, and, in some cases, is implicit (i.e. the
@@
 example, some promises may remain un-evaluated, and it would be erroneous to
-force them as they may side-effect and change programs execution.  To deal
+force them as they may side-effect and change program behavior.  To deal
 with unevaluated arguments, we make an initial guess for each argument at
@@
 missing. This occurs when a function was called with too few arguments and
-no default value was specified for missing arguments.  We record a
+no default values were specified for missing arguments.  We record a
 \k{missing} type for such argument.  There are two obvious ways to deal with
@@
 returns that are skipped. As we cannot be sure of the intended return value, 
-these \k{jumped} become \any types.
+these \k{jumped} values become \any types.
 
@@
 \k{builtin\_entry}, \k{builtin\_exit}, \k{special\_entry},
-\k{special\_exit}, \k{promise\_force\_entry}, and
-\k{prom\-ise\_force\_exit}.  The function-related callbacks are used mainly
+\k{special\_exit}, \k{promise\_force\_entry}, and finally
+\k{promise\_force\_exit}.  The function-related callbacks are used mainly
 for bookkeeping: the analysis is notified that a construct has been entered
@@
 \vspace*{1mm}
-\FUN{ T_{1,1} \k{~|~} T_{1,i}, \dots, T_{n,1} \k{~|~} T_{n,j}}{T_1\k{~|~} T_k}
+\FUN{ T_{1,1} \k{~|~} T_{1,i}, \dots, T_{n,1} \k{~|~} T_{n,j}}{T_1 \k{~|~} ... \k{~|~} T_k}
 \vspace*{1mm}
@@
 well as the simple notion of subtyping that we have adopted here.
+Our subtyping rules are both born out of necessity (e.g., scalars are subtypes of vectors since a.
+that is the case in R and b.
+we want to allow types like \dbl | \VEC\dbl to simplify), as well as from prevalent data type conversion with no ambiguity (e.g., \int to \dbl).
 
@@
 nullables.  Functions are
-limited to twenty arguments as a simplifying assumption. Higher-order functions are approximated by \any
-to \any by \typetracer.
+limited to twenty arguments as a simplifying assumption. Higher-order functions are conservatively inferred as \any
+$\rightarrow$ \any by \typetracer.
 
@@
 For this purpose, we developed \contractr, an R package
-that allows one to decorate function with assertions. We use it to insert type
+that allows one to decorate a function with assertions. We use it to insert type
 checking of arguments and return values.
@@
 \contractr's primary logic has been implemented in C++ to reduce its
-runtime overhead, and we have not observed a single segfault during our use of
+runtime overhead, and we have not observed a single segmentation fault during our use of
 the package.  It has been tested with GNU R-3.5.0 and hardened with a
-battery of 400 test cases. It works by modifying function definition to
+battery of 400 test cases. It works by modifying function definitions to
 insert a call to a type-checking function.
@@
 declarations can also be written directly in the adjunct function comments using
-\roxygen annotation tags. \contractr introduces a new use \code{@type} tag. When
+\roxygen~\footnote{\url{https://cran.r-project.org/web/packages/roxygen2/}} annotation tags. \contractr introduces a new use \code{@type} tag. When
 a package these type signatures are extracted and stored in the appropriate
@@
 contracts during interactive development by supplying the type signature as a
-string, selectively disable contract checking and inspecting type assertions.
+string, selectively disable contract checking, and inspecting type assertions.
 
@@
 value, such as checking for the absence of \NA or the type of the elements of list. Union types
-require checking for each of the alternatives. As the usage
+require checking for each of the members of the union. As the usage
 of higher-order functions remains limited in R, we do not check higher-order functions in this version of the tool. Checking
@@
 are then posted on Kaggle either as plain scripts or as
-notebooks. One of the most popular competition is about predicting passenger
+notebooks. One of the most popular competitions is about predicting passenger
 survival on Titanic\footnote{\url{https://www.kaggle.com/c/titanic}} with
@@
 Table~\ref{ctb} reports on the 10 most frequent types occurring in the
-corpus.
+corpus. The first row of the table reads: the \dbl type occurs in 12,298 (11.24\%) argument types, and accounts for over 12 million (20.3\%) of the types observed by \typetracer's dynamic analysis.
 
@@
 to describe the actual type signatures of functions. For this we focus on
-how often there is a single type for a particular argument; this because
+how often there is a single type for a particular argument; this is because
 union types and \any are less accurate (and would likely require a more
@@
 Monomorphic in this context means that the type is not relying on \any or
-including a union.  The import of monorphism in this context is that it
+including a union.  The import of monomorphism in this context is that it
 means our type language can accurately capture an argument's type or a
@@
 Fig.~\ref{fig:unionfreq} shows the number of inferred argument types and
-their size (in terms of terms in the union). The figure shows the most
+their size (in terms of members of the union). The figure shows the most
 functions do not require a union at all (\PercUnitypedPositions of arguments
@@
 \CLASS{\texttt{aov}, \texttt{lm}} or \CLASS{\texttt{lm}}.  Internally,
-linear models are represented as lists with named elements.  The polution is
-illustrated by the lines after the funtion definition.  They load an example
+linear models are represented as lists with named elements.  The pollution is
+illustrated by the lines after the function definition.  They load an example
 data set and test the function \code{cv.model}.  The
-\code{data(sweetpotato)} expression load a sample data set. The fields of
+\code{data(sweetpotato)} expression loads a sample data set. The fields of
 \code{sweetpotato} will be recorded when \code{cv.model} is called.
@@
 While we record classes, our analysis does not deal with method dispatch.  R
-has multiple, disparate, object systems, called S3, S4, and R5.  The
+has multiple disparate object systems called S3, S4, and R5.  The
 \code{class} attribute is used by these systems to dispatch methods. S3 does
@@
 semantics, many internal functions coerce vectors to matrices automatically.
-The \code{rowWeightedMeans} function calculates the weighted means of rows.
+For example, the \code{rowWeightedMeans} function calculates the weighted means of rows.
 The programmer added a type check for \code{x}.
@@
 \NUMASSERTIONS total assertions. Overall, we found that only
-\PERCFAILEDASSERTIONSWUNDEF of contract failed. The limit on number of arguments
-(we record only 20) accounted for \PERCASSERTIONSUNDEF failed assertion. We
-found that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of functions types
+\PERCFAILEDASSERTIONSWUNDEF of contract assertions failed. The limit on number of arguments
+(we record only 20) accounted for \PERCASSERTIONSUNDEF failed assertions. We
+found that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of function types
 never failed. The number of immaculate function types increases to
@@
 
-We breakdown the failed assertions by type in
+We break down the failed assertions by type in
 Table~\ref{tbl:fail-breakdown-0}.  Accounting for 36.36\% of assertion
@@
 \toprule
-Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
+              &                &      \multicolumn{2}{c}{\# Args Types}        &  \\
+ Passed & Arg Type & Failed & Total & \% Failure\\
 \midrule
@@
 \CLASS{{\tt matrix}}, \VEC\dbl, and \CLASS{\texttt{data.frame}} values passed to arguments expecting \CLASS{{\tt timeSeries}}.
-This occurs in only 35.90\% of such arguments,
+This occurs in 35.90\% of such arguments,
 and represents cases where tests did not adequately cover all valid function
@@
 % END Autogenerated
- \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 90th percentile}
+ \caption{Highest failure rate among popular argument types, for argument signatures whose frequency is in the 90th percentile.}
  \label{tbl:fail-breakdown-2}
@@
 \CLASS{{\tt data.frame}} & \CLASS{{\tt timeSeries}} & 14 & 39 & 35.90 \\
-\CLASS{{\tt dtplyr\_step\_first, dtplyr\_step}} & \CLASS{{\tt data.frame}} | Grouped Tibble | Tibble & 10 & 45 & 22.22 \\
+\CLASS{{\tt dtplyr\_step\_first, dtplyr\_step}} & \makecell[tl]{\CLASS{{\tt data.frame}} | \\ \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}} | \\ \CLASS{{\tt data.frame, tbl, tbl\_df}}} & 10 & 45 & 22.22 \\
 \chr & \raw & 6 & 31 & 19.35 \\
@@
 % END Autogenerated
- \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 80th percentile.
- *: Grouped Tibble is \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}} , and Tibble is \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}.}
+ \caption{Highest failure rate among popular argument types, for argument signatures whose frequency is in the 80th percentile.}
+ % *: Grouped Tibble is \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}} , and Tibble is \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}.}
  \label{tbl:fail-breakdown-3}
@@
 
- The other two patterns are for arguments with no default value, where either the call results in an error (perhaps explicitly handled by the programmer), or resulting in valid function behaviour that was untested by the original package designer.
- Here, the first case can be explained by a lack of testing (or generally incorrect use in a client), and the second case is explained by programmers not fully understanding R's language semantics.
+ The other two patterns are for arguments with no default value, where either the call results in an error (perhaps explicitly handled by the programmer), or results in valid function behaviour that was untested by the original package designer.
+ Here, the first case can be explained by a lack of testing, and the second case is explained by programmers not fully understanding R's language semantics.
  For example, we observed this kind of error in the following function: 
@@
         return(as.integer(x))
-    if (is.numeric(x)) {
     # ...
-    }
-    return(x)
   }
@@
 
- Next, we were curious about how often vectors were passed to arguments expecting scalars.
+ Next, we analyzed how often vectors were passed to arguments expecting scalars.
  We found that \PERCDYNAMICVECERRORS of dynamic contract assertion failures were of this type, and these errors were present in \NUMFUNSWITHVECASSERTERROR functions.
@@
 If given a scalar ``integer'' $n$, it will generate all permutations for the list $[1, 2, ..., n]$.
-Curiously, the function was only tested by passing an integer, and not ever with a vector (even though, presumably, that is the main utility of the function).
+Interestingly, the function was only tested by passing an integer, and not ever with a vector (even though, presumably, that is the main utility of the function).
 
@@
 There are two broad divisions which account for most of the class-related contract assertion failures that are not outright errors.
-First, we observe a class of errors related to ``subclasses'' being passed to arguments expecting a ``superclass''.
-For instance, we observed \CLASS{{\tt data.frame}} values being passed to arguments expecting tibbles or data.tables (which can be readily converted).
+First, we observe a class of errors related to classes being passed to arguments expecting a different, yet convertible class.
+For instance, we observed \CLASS{{\tt data.frame}} values being passed to arguments expecting tibbles or data.tables (data frames have a straightforward conversion to these classes).
 Second, we observe a class of errors related more to coercion between simple data types and classes.
@@
  \CLASS{{\tt factor}} & \VEC\chr  & 7519 & 10.58 & 64.54 \\ 
- \VEC\chr & \LIST{\VEC\int} $|$ ... $|$ \LIST{\CLASS{{\tt formula, quosure}}} & 5385 & 7.58 & 72.12 \\ 
- \CLASS{{\tt ixforeach, iter}} & \CLASS{{\tt dataframeiter, iter}} $|$ ... $|$ \CLASS{{\tt iforeach, iter}} & 4139 & 5.82 & 77.94 \\ 
+ \VEC\chr & \makecell[tl]{\LIST{\VEC\int} $|$ ... $|$ \\ \LIST{\CLASS{{\tt formula, quosure}}}} & 5385 & 7.58 & 72.12 \\ 
+ \CLASS{{\tt ixforeach, iter}} & \makecell[tl]{\CLASS{{\tt dataframeiter, iter}} $|$ ... $|$ \\ \CLASS{{\tt iforeach, iter}}} & 4139 & 5.82 & 77.94 \\ 
 \bottomrule
@@
 \end{tabular}
-\caption{What is the highest failure rate among popular argument types in Kaggle? For argument signatures whose frequency is in the 90th percentile}
+\caption{Highest failure rate among popular argument types in Kaggle, for argument signatures whose frequency is in the 90th percentile.}
  \label{tbl:fail-breakdown-2-kaggle}
@@
  To mirror our analysis of contract assertions on the reverse dependencies of our corpus, we show in Table~\ref{tbl:fail-breakdown-0-kaggle} the most frequently failing contract in Kaggle.
- While we don't see many overlapping entries per se, the table entries have similar explanations.
- For instance, values of type \CLASS{{\tt data.table, data.frame}} are often passed to arguments expecting a \CLASS{{\tt matrix}}.
- \CLASS{{\tt data.table, data.frame}} values are known as ``data tables'', and essentially serve the same purpose as data frames and tibbles.
- As it happens, many data tables can be coerced to matrices, and programmers will often interchange the two (particularly when data tables have numeric data), as is the case here.
+ While we don't see many overlapping entries per se, the assertions exhibit similar patterns.
+ For instance, {\it data tables} (which have \CLASS{{\tt data.table, data.frame}}) are often passed to arguments expecting a \CLASS{{\tt matrix}}.
+ Data tables are essentially serve the same purpose as data frames and tibbles.
+ As it happens, data tables can be coerced to matrices if their elements are unityped, and programmers will often interchange the two, as is the case here.
  One function producing many of these errors is the following:
@@
  \end{lstlisting}
- We see that \code{class::knn} can take arguments as long as they can be passed to \code{as.matrix}, e.g., data tables, and will perform the coercion.
- As in similar cases discussed earlier, we did not encode this semantics in our type system as the process of coercion in R is ad hoc.
+ We see that \code{class::knn} coerces its first two arguments to matrices.
  On the topic of coercion, rows two and three of Table~ are interesting as they depict {\it factors} being passed to arguments expecting character vectors.
- \CLASS{{\tt factor}} typed values are known as factors in R, and they are stored as a vector of integer values corresponding to a set of character values, and their purpose is to allow for quickly deal with categorized data.
+ \CLASS{{\tt factor}} typed values are known as factors in R, and they are stored as a vector of integer values corresponding to a set of character values, and their purpose is to allow for R to quickly deal with categorized data.
  Factors can be readily converted to characters when needed, as evidenced by these assertion failures.
@@
  Another interesting entry in Table~\ref{tbl:fail-breakdown-0-kaggle} is the fifth row, where a \CLASS{{\tt ixforeach, iter}}  is being passed to an argument expecting a long union of classes, each of the form \CLASS{{\tt X, iter}}.
- This is likely an instance where the user defined their own class, {\tt ixforeach}, and wanted to use the \code{iterators} package (the user called \code{iterator::iter} with a \CLASS{{\tt ixforeach}}, and it gained the class tag {\tt iter} on return).
+ This is likely an instance where the user defined their own class, {\tt ixforeach}, and wanted to use the \code{iterators} package (the user called \code{iterator::iter} with a \CLASS{{\tt ixforeach}}, and it gained the class {\tt iter} on return).
  As we mentioned, accounting for object-orientation in the type system is beyond the scope of this work, and such an inclusion would allow us to better type situations like this.
@@
 
-There is a number of ways in R to check types of function parameters. The
-default and the most common way\footnote{\CranStopifnotRatio of all runtime
-  checks in the whole of CRAN} is to use \code{stopifnot} function from the R
+There are a number of ways to check the types of function parameters in R.
+ The default and most common way\footnote{\CranStopifnotRatio of all runtime
+  checks in the whole of CRAN} is to use the \code{stopifnot} function from the R
 base package. It takes a number of R expressions which all should evaluate to
@@
 
-Next to \code{stopifnot}, there are 4 packages in CRAN\footnote{Packages are
-  available on CRAN website: \url{https://cran.r-project.org/web/packages/}}.
+Besides \code{stopifnot}, there are 4 packages in CRAN\footnote{Packages are
+  available on the CRAN website: \url{https://cran.r-project.org/web/packages/}}
 that focus on runtime assertions: \emph{assertive}, \emph{ensurer},
-\emph{assertr} and \emph{assertthat} \emph{assertive} and \emph{ensurer} have
-not been updated 2016 and 2015 respectively. \emph{assertr} is used by only
-\AssertrRevdeps other packages and in the current version focuses on checking
-properties of data frames. Only the \emph{assertthat} is maintained and used
+\emph{assertr} and \emph{assertthat}. \emph{assertive} and \emph{ensurer} have
+not been updated since 2016 and 2015 respectively, and \emph{assertr} is used by only
+\AssertrRevdeps other packages and currently focuses on checking
+properties of data frames. Only the \emph{assertthat} package is maintained and used
 (with \AssertthatRevdeps reverse dependencies). The advantage of
-\emph{assertthat} over the R's default is that it provides a much better error
-message.
+\emph{assertthat} over the R's default is that it provides much better error
+messages.
 
@@
 To measure this, we have extracted all calls to \code{stopifnot} and
-\code{assertthat} assertions and checked which one of them could be either
+\code{assertthat} assertions and checked which among them could be either
 completely replaced by \contractr or at least partially simplified by removing a
@@
 checks its value. In the example above, the whole expression could be replaced
-by \code{character} type.
+by a \chr type check.
 
@@
 \CranAssertsInFunctionsRnd functions define in \CranAssertsInPackagesRnd
-packages. One might speculate that one of the reasons is that the current way is
-not convenient as it is rather verbose. Our system can infer type annotations
+packages.
+One may speculate that this is the case due to the verbosity and inconvenience of the existing assertion tools.
+Our system can infer type annotations
 for existing functions automatically. This can remove or partially remove over
@@
 %
-\subsection{Discussion}
-
-As a closing point of discussion, we would like to highlight an alternate strategy for unifying class types that we considered.
-Recall that in our proposed type annotation language, classes are represented as \CLASS{{\tt A}} for something of class {\tt A}, and \CLASS{{\tt A}, {\tt B}} for something with two classes {\tt A} and {\tt B}.
-
-During the robustness phase of our evaluation, we noted a number of class-based type errors such as a \CLASS{{\tt data.frame}} being passed to an argument expecting, e.g., a \CLASS{{\tt data.frame}, {\tt tbl}, {\tt tbl\_df}}.
-This represents something of a limitation of our approach, as data frames and {\tt dplyr} tibbles are related and convertible (you can think of a tibble as a subclass of a data frame)---a fact not represented by our types.
-One option to mitigate this would be to include a type such as \CLASS{{\tt data.frame}, ...} that must include {\it at least} the {\tt data.frame} class, but this would not altogether solve this problem.
-For one, to generate this type we would either need to a) include some type hierarchy in our type system, which is outside of the scope of this work, or b) somehow infer the type with \typetracer.
-To create this type, \typetracer would have to observe values of {\it both} type \CLASS{{\tt data.frame}} and \CLASS{{\tt data.frame}, {\tt tbl}, {\tt tbl\_df}}, which would not solve the aforementioned problem as many {\tt dplyr} functions are only called with tibbles.
-
-Further, the data table, another data frame alternative, while being coercible to and from a data frame, only has \CLASS{{\tt data.table}} (not including {\tt data.frame} as part of its class).
-This alternate class type would also not adequately capture this behaviour, and even worse, if a \CLASS{{\tt data.frame}} and \CLASS{{\tt data.table}} were to co-occur in an argument type, we would unify it to \CLASS{...}, which is not very informative.
-
-For these reasons, we believe our unification strategy for class types to be adequate until we extend our type system to handle R's wild and varied object orientation frameworks.
+% \subsection{Discussion}
+
+% As a closing point of discussion, we would like to highlight an alternate strategy for unifying class types that we considered.
+% Recall that in our proposed type annotation language, classes are represented as \CLASS{{\tt A}} for something of class {\tt A}, and \CLASS{{\tt A}, {\tt B}} for something with two classes {\tt A} and {\tt B}.
+
+% During the robustness phase of our evaluation, we noted a number of class-based type errors such as a \CLASS{{\tt data.frame}} being passed to an argument expecting, e.g., a \CLASS{{\tt data.frame}, {\tt tbl}, {\tt tbl\_df}}.
+% This represents something of a limitation of our approach, as data frames and {\tt dplyr} tibbles are related and convertible (you can think of a tibble as a subclass of a data frame)---a fact not represented by our types.
+% One option to mitigate this would be to include a type such as \CLASS{{\tt data.frame}, ...} that must include {\it at least} the {\tt data.frame} class, but this would not altogether solve this problem.
+% For one, to generate this type we would either need to a) include some type hierarchy in our type system, which is outside of the scope of this work, or b) somehow infer the type with \typetracer.
+% To create this type, \typetracer would have to observe values of {\it both} type \CLASS{{\tt data.frame}} and \CLASS{{\tt data.frame}, {\tt tbl}, {\tt tbl\_df}}, which would not solve the aforementioned problem as many {\tt dplyr} functions are only called with tibbles.
+
+% Further, the data table, another data frame alternative, while being coercible to and from a data frame, only has \CLASS{{\tt data.table}} (not including {\tt data.frame} as part of its class).
+% This alternate class type would also not adequately capture this behaviour, and even worse, if a \CLASS{{\tt data.frame}} and \CLASS{{\tt data.table}} were to co-occur in an argument type, we would unify it to \CLASS{...}, which is not very informative.
+
+% For these reasons, we believe our unification strategy for class types to be adequate until we extend our type system to handle R's wild and varied object orientation frameworks.
 
@@
 Overall, we found that our simple design fits quite well with the existing language:
-Over 82\% of functions are either monomorphic or have only one single polymorphic argument.
-When we tested the types inferred by \typetracer during our evaluation, we found that only \PERCFAILEDASSERTIONSWUNDEF of contract assertions failed, and found no pathological patterns in argument types with failed assertions.
+Nearly 80\% of functions are either monomorphic or have only one single polymorphic argument.
+When we tested the types inferred by \typetracer during our evaluation, we found that only \PERCFAILEDASSERTIONSWUNDEF of contract assertions failed.
 Furthermore, we found that our type language and contract checking framework would be useful to programmers, eliminate or otherwise simplify \CorpusPartiallyTypedAssertsRatio of existing type checks and assertions in user code.
@@
 
-\bibliography{bib/biblio,bib/jv,bib/r,bib/new,bib/gradual}
+\section*{Acknowledgements}
+
+This work was partially funded by NSERC.
+
+\bibliography{bib/biblio,bib/jv,bib/r,bib/new,bib/gradual,bib/types4R}
 \end{document}

### 2020-10-05 paper/main.tex
minor
acks

@@
-\documentclass[acmsmall, screen,review]{acmart}
+\documentclass[acmsmall, screen]{acmart}
 \setlength{\pdfpagewidth}{8.5in}
@@
 %
-% \subsection{Discussion}
-
-% As a closing point of discussion, we would like to highlight an alternate strategy for unifying class types that we considered.
-% Recall that in our proposed type annotation language, classes are represented as \CLASS{{\tt A}} for something of class {\tt A}, and \CLASS{{\tt A}, {\tt B}} for something with two classes {\tt A} and {\tt B}.
-
-% During the robustness phase of our evaluation, we noted a number of class-based type errors such as a \CLASS{{\tt data.frame}} being passed to an argument expecting, e.g., a \CLASS{{\tt data.frame}, {\tt tbl}, {\tt tbl\_df}}.
-% This represents something of a limitation of our approach, as data frames and {\tt dplyr} tibbles are related and convertible (you can think of a tibble as a subclass of a data frame)---a fact not represented by our types.
-% One option to mitigate this would be to include a type such as \CLASS{{\tt data.frame}, ...} that must include {\it at least} the {\tt data.frame} class, but this would not altogether solve this problem.
-% For one, to generate this type we would either need to a) include some type hierarchy in our type system, which is outside of the scope of this work, or b) somehow infer the type with \typetracer.
-% To create this type, \typetracer would have to observe values of {\it both} type \CLASS{{\tt data.frame}} and \CLASS{{\tt data.frame}, {\tt tbl}, {\tt tbl\_df}}, which would not solve the aforementioned problem as many {\tt dplyr} functions are only called with tibbles.
-
-% Further, the data table, another data frame alternative, while being coercible to and from a data frame, only has \CLASS{{\tt data.table}} (not including {\tt data.frame} as part of its class).
-% This alternate class type would also not adequately capture this behaviour, and even worse, if a \CLASS{{\tt data.frame}} and \CLASS{{\tt data.table}} were to co-occur in an argument type, we would unify it to \CLASS{...}, which is not very informative.
-
-% For these reasons, we believe our unification strategy for class types to be adequate until we extend our type system to handle R's wild and varied object orientation frameworks.
-
-%
-%
-%
-%
-%
-%
 \section{Conclusion}
 
-Retrofitting a type system for the interactive and exploratory programming style of R is hard: The language is poorly specified and builds upon an eclectic mix of features such as laziness, reflection, dynamic evaluation and ad-hoc object systems.
-Our intent is to eventually propose a type system for inclusion in the language, but we are aware that for any changes to be accepted by the community, they must have clear benefits without endangering backwards compatibility.
-As a step towards this, we focus on a simpler problem: instead of an entire type system, we limited the scope of our investigation to ascribing types to function signatures.
-To this end, we designed a simple type language which found a compromise between simplicity and usefulness by focusing  on the most widely used features of R.
-We presented \typetracer, a tool for inferring types for function signatures from runnable code, and \contractr, an easy-to-use package for R which allows users to specify function type signatures and have function arguments checked for compliance at runtime.
-
-We evaluated our design by running \typetracer on a corpus of 412 of the most widely used R packages on CRAN, inferring signatures for exported functions, and testing those inferred signatures on the \NUMPKGSEVAL reverse dependencies of the corpus.
-Overall, we found that our simple design fits quite well with the existing language:
-Nearly 80\% of functions are either monomorphic or have only one single polymorphic argument.
-When we tested the types inferred by \typetracer during our evaluation, we found that only \PERCFAILEDASSERTIONSWUNDEF of contract assertions failed.
-Furthermore, we found that our type language and contract checking framework would be useful to programmers, eliminate or otherwise simplify \CorpusPartiallyTypedAssertsRatio of existing type checks and assertions in user code.
-In sum, we believe that our simple type language design is a solid foundation for the eventual type system for R.
-
-\section*{Acknowledgements}
-
-This work was partially funded by NSERC.
+Retrofitting a type system for the interactive and exploratory
+programming style of R is hard: The language is poorly specified and
+builds upon an eclectic mix of features such as laziness, reflection,
+dynamic evaluation and ad-hoc object systems.  Our intent is to
+eventually propose a type system for inclusion in the language, but we
+are aware that for any changes to be accepted by the community, they
+must have clear benefits without endangering backwards compatibility.
+As a step towards this, we focus on a simpler problem: instead of an
+entire type system, we limited the scope of our investigation to
+ascribing types to function signatures.  To this end, we designed a
+simple type language which found a compromise between simplicity and
+usefulness by focusing on the most widely used features of R.  We
+presented \typetracer, a tool for inferring types for function
+signatures from runnable code, and \contractr, an easy-to-use package
+for R which allows users to specify function type signatures and have
+function arguments checked for compliance at runtime.
+
+We evaluated our design by running \typetracer on a corpus of 412 of
+the most widely used R packages on CRAN, inferring signatures for
+exported functions, and testing those inferred signatures on the
+\NUMPKGSEVAL reverse dependencies of the corpus.  Overall, we found
+that our simple design fits quite well with the existing language:
+Nearly 80\% of functions are either monomorphic or have only one
+single polymorphic argument.  When we tested the types inferred by
+\typetracer during our evaluation, we found that only
+\PERCFAILEDASSERTIONSWUNDEF of contract assertions failed.
+Furthermore, we found that our type language and contract checking
+framework would be useful to programmers, eliminate or otherwise
+simplify \CorpusPartiallyTypedAssertsRatio of existing type checks and
+assertions in user code.  In sum, we believe that our simple type
+language design is a solid foundation for the eventual type system for
+R.
+
+\begin{acks}
+This work has received funding from the \grantsponsor{ONR}{Office of
+  Naval Research (ONR)}{} award \grantnum{ONR}{503353}, the
+\grantsponsor{NSF}{National Science Foundation}{} awards
+\grantnum{NSF}{1759736}, \grantnum{NSF}{1544542},
+\grantnum{NSF}{1925644}, and \grantnum{NSF}{1910850}, the
+\grantsponsor{BC}{Czech Ministry of Education, Youth and Sports from
+  the Czech Operational Programme Research, Development, and
+  Education}{}, under grant agreement No.
+\grantnum{BC}{CZ.02.1.01/0.0/0.0/15\_003/0000421}, and the
+\grantsponsor{ELE}{European Research Council (ERC) under the European
+  Union's Horizon 2020 research and innovation programme}{}, under
+grant agreement No.  \grantnum{ELE}{695412}. This work was partially
+funded by NSERC.
+\end{acks}
 

### 2020-10-08 paper/main.tex
reformat

@@
 %\settopmatter{printfolios=true,printccs=true,printacmref=true}
-
 \setcopyright{rightsretained}
@@
 \begin{abstract}
-The R programming language is widely used in a variety of domains.  It was
-designed to favor an interactive style of programming with minimal syntactic
-and conceptual overhead. This design is well suited to interactive data
-analysis, but a bad fit for tools such as compilers or program analyzers
-which must generate native code or catch programming errors.  In particular,
-R has no type annotations, and all operations are dynamically checked at
-run-time. The starting point for our work are the twin questions: \emph{what
-  expressive power is needed to accurately type R code?} and \emph{which
-  type system is the R community willing to adopt?} Both questions are
-difficult to answer without actually experimenting with a type system. The
-goal of this paper is to provide data that can feed into that design
-process. To this end, we perform a large corpus analysis to gain insights in
-the degree of polymorphism exhibited by idiomatic R code and explore
-potential benefits that the R community could accrue from a simple type
-system.  As a starting point, we infer type signatures for \NUMFUNCTIONS
-functions from \NUMPACKAGES packages among the most widely used open source
-R libraries. We then conduct an evaluation on \NUMPKGSEVAL clients of these packages, as well as on end-user code found on the Kaggle competition website.
+The R programming language is widely used in a variety of domains.  It
+was designed to favor an interactive style of programming with minimal
+syntactic and conceptual overhead. This design is well suited to
+interactive data analysis, but a bad fit for tools such as compilers
+or program analyzers which must generate native code or catch
+programming errors.  In particular, R has no type annotations, and all
+operations are dynamically checked at run-time. The starting point for
+our work are the twin questions: \emph{what expressive power is needed
+to accurately type R code?} and \emph{which type system is the R
+community willing to adopt?} Both questions are difficult to answer
+without actually experimenting with a type system. The goal of this
+paper is to provide data that can feed into that design process. To
+this end, we perform a large corpus analysis to gain insights in the
+degree of polymorphism exhibited by idiomatic R code and explore
+potential benefits that the R community could accrue from a simple
+type system.  As a starting point, we infer type signatures for
+\NUMFUNCTIONS functions from \NUMPACKAGES packages among the most
+widely used open source R libraries. We then conduct an evaluation on
+\NUMPKGSEVAL clients of these packages, as well as on end-user code
+found on the Kaggle competition website.
 \end{abstract} \maketitle
@@
 Our tools are open source and publicly available on GitHub\footnote{
-\typetracer: \url{https://github.com/PRL-PRG/propagatr}, \contractr: \url{https://github.com/PRL-PRG/contractr}
-},
-all results in this paper are reproducible and were submitted for
-artifact evaluation.
+\typetracer: \url{https://github.com/PRL-PRG/propagatr}, \contractr:
+\url{https://github.com/PRL-PRG/contractr} }.
 %
@@
 
-Dynamic programming languages such as Racket, JavaScript and PHP have been
-extended post-facto with various static type systems.  In each case, the
-type system was carefully engineered to match the salient characteristics of
-the host languages and to foster a particular programming style. For
-example, Typed Racket emphasizes functional programming and supports the
-migration from untyped to fully typed code~\cite{tf-popl08},
-Hack~\cite{hack13} and TypeScript~\cite{BAT14} focus on typing
-object-oriented features of PHP and JavaScript, respectively. They allow users to
-intersperse typed and untyped code in a fine-grained manner.  Other work
-looks at retrofitting types onto Lua~\cite{typedlua2014} which account for the myriad ways Lua programmers use tables.
+Dynamic programming languages such as Racket, JavaScript and PHP have
+been extended post-facto with various static type systems.  In each
+case, the type system was carefully engineered to match the salient
+characteristics of the host languages and to foster a particular
+programming style. For example, Typed Racket emphasizes functional
+programming and supports the migration from untyped to fully typed
+code~\cite{tf-popl08}, Hack~\cite{hack13} and TypeScript~\cite{BAT14}
+focus on typing object-oriented features of PHP and JavaScript,
+respectively. They allow users to intersperse typed and untyped code
+in a fine-grained manner.  Other work looks at retrofitting types onto
+Lua~\cite{typedlua2014} which account for the myriad ways Lua
+programmers use tables.
 
@@
 
-Recent work has gone into adding types to Python, the other eminent data
-science language.  Typilus is an interesting piece of recent work which
-explores using neural networks to infer types for Python
-programs~\cite{allamanis2020typilus}, and Python itself added support for
-type hints in Python 3.5~\cite{pythonspectypes}.  There is no previous work
-on types for R, besides a short sketch of issues the type system would need to tackle~\cite{turcotte2019towards}. We take inspiration in the aforementioned works but focus
-on adapting them to our target language.
+Recent work has gone into adding types to Python, the other eminent
+data science language.  Typilus is an interesting piece of recent work
+which explores using neural networks to infer types for Python
+programs~\cite{allamanis2020typilus}, and Python itself added support
+for type hints in Python 3.5~\cite{pythonspectypes}.  There is no
+previous work on types for R, besides a short sketch of issues the
+type system would need to tackle~\cite{turcotte2019towards}. We take
+inspiration in the aforementioned works but focus on adapting them to
+our target language.
 
@@
 
-The basic building blocks of R are vectors of primitives (primitives in R are characters (strings), integers, logicals (booleans), doubles, complex numbers, and raw bytes).
-When we say that R is {\it vectorized}, we mean that primitive values are {\it always} vectors.
-For example, in R one might see the literal \code{2.5} and think it to be a scalar double, but it is in actuality a unit-length vector containing a single double.
-Vectors can also be explicitly constructed using the vector constructor \code{c(...)}, 
-For example, \code{c(1L, 2L, 3L)} creates a vector of three integers (in R, integer literals are denoted by an \code{L}).
-To illustrate that scalars are vectors, in R \code{(c(2.5) == 2.5) == TRUE}.
-
-The main object of our work is R functions, and functions in R have a number of quirks.
-Arguments can be assigned default values (which can be arbitrary expressions!), and functions can support variable argument lists, and functions can be called by specifying argument names explicitly in the call.
-To illustrate all of this in a single example, consider:
+The basic building blocks of R are vectors of primitives (primitives
+in R are characters (strings), integers, logicals (booleans), doubles,
+complex numbers, and raw bytes).  When we say that R is {\it
+  vectorized}, we mean that primitive values are {\it always} vectors.
+For example, in R one might see the literal \code{2.5} and think it to
+be a scalar double, but it is in actuality a unit-length vector
+containing a single double.  Vectors can also be explicitly
+constructed using the vector constructor \code{c(...)}, For example,
+\code{c(1L, 2L, 3L)} creates a vector of three integers (in R, integer
+literals are denoted by an \code{L}).  To illustrate that scalars are
+vectors, in R \code{(c(2.5) == 2.5) == TRUE}.
+
+The main object of our work is R functions, and functions in R have a
+number of quirks.  Arguments can be assigned default values (which can
+be arbitrary expressions!), and functions can support variable
+argument lists, and functions can be called by specifying argument
+names explicitly in the call.  To illustrate all of this in a single
+example, consider:
 
@@
 \noindent
-This function has four formal parameters, \k x, \k{...}, \k y and \k z. Argument
-\k x can be bound positionally or passed by name. The vararg argument,
-\k{...}, is always passed positionally. The remaining two arguments must be
-passed by name as they are preceded by the \k{...} parameter.  Arguments \k y and \k z have default values, in the case of
-\k z this is a constant, but \k{y}'s default value is an expression that
-depends on the value of \k z (if \k z is not zero, \k y will default to \k{NULL}). The body of the function will add parameters
-\k x and \k y with either the scalar 0 or the result of concatenating the
-varargs into a primitive vector. The function \k{missing} tests if an
-argument was explicitly passed. The following are some valid invocations
-of \k f:
+This function has four formal parameters, \k x, \k{...}, \k y and \k
+z. Argument \k x can be bound positionally or passed by name. The
+vararg argument, \k{...}, is always passed positionally. The remaining
+two arguments must be passed by name as they are preceded by the
+\k{...} parameter.  Arguments \k y and \k z have default values, in
+the case of \k z this is a constant, but \k{y}'s default value is an
+expression that depends on the value of \k z (if \k z is not zero, \k
+y will default to \k{NULL}). The body of the function will add
+parameters \k x and \k y with either the scalar 0 or the result of
+concatenating the varargs into a primitive vector. The function
+\k{missing} tests if an argument was explicitly passed. The following
+are some valid invocations of \k f:
 
@@
 The above hints at the polymorphism of the language: \k f may return a
-vector of integers or of doubles, the length of the vector depends on the
-length of the varargs, and of \k x and \k y.  As we mentioned, the language does not really
-differentiate between scalar and vectors.  Some more exotic types that can
-be encountered include vectors of complex number and list of arbitrary
-types. Function \k f can be invoked with those as well.
+vector of integers or of doubles, the length of the vector depends on
+the length of the varargs, and of \k x and \k y.  As we mentioned, the
+language does not really differentiate between scalar and vectors.
+Some more exotic types that can be encountered include vectors of
+complex number and list of arbitrary types. Function \k f can be
+invoked with those as well.
 
@@
 
-All vectorized data types have a distinguished missing value denoted by \code{NA} (for ``not available''). 
-The default type of \code{NA} is \k{logical}. 
-We can see that \code{typeof(NA)=="logical"}, but \k{NA} inhabits every type: \code{typeof(c(1,NA)[2])} \k{==} \k{"double"}.
-In addition to \code{NA}, R has a \code{NULL} value, and to understand why R has both a \code{NULL} value and \code{NA} values it is useful to remember that R is primarily designed for statistics and data science.
-In experimental data science, it is useful to have a notion of a ``missing observation'', which is often NaN (not a number, or similar) in, e.g., MATLAB and Python.
-Since vectors are unityped in R, it {\it must} have a ``NaN'' value for {\it each} primitive type.
-There single \code{NULL} null-typed value in the language to encompass all other cases.
-In a sense, you can think of \code{NA} as representing a missing data point, and \code{NULL} as a missing data {\it set}.
+All vectorized data types have a distinguished missing value denoted
+by \code{NA} (for ``not available'').  The default type of \code{NA}
+is \k{logical}.  We can see that \code{typeof(NA)=="logical"}, but
+\k{NA} inhabits every type: \code{typeof(c(1,NA)[2])} \k{==}
+\k{"double"}.  In addition to \code{NA}, R has a \code{NULL} value,
+and to understand why R has both a \code{NULL} value and \code{NA}
+values it is useful to remember that R is primarily designed for
+statistics and data science.  In experimental data science, it is
+useful to have a notion of a ``missing observation'', which is often
+NaN (not a number, or similar) in, e.g., MATLAB and Python.  Since
+vectors are unityped in R, it {\it must} have a ``NaN'' value for {\it
+  each} primitive type.  There single \code{NULL} null-typed value in
+the language to encompass all other cases.  In a sense, you can think
+of \code{NA} as representing a missing data point, and \code{NULL} as
+a missing data {\it set}.
 
@@
 
-With one exception, all vectorized data types are monomorphic.
-The exception is the \k{list} type which can hold values of any other type including
-\k{list}. For all monomorphic data types, attempting to store a value of a
-different type will cause a conversion. Either the value is converted to the
-type of the vector, or the vector is converted to the type of the value.
+With one exception, all vectorized data types are monomorphic.  The
+exception is the \k{list} type which can hold values of any other type
+including \k{list}. For all monomorphic data types, attempting to
+store a value of a different type will cause a conversion. Either the
+value is converted to the type of the vector, or the vector is
+converted to the type of the value.
 
@@
 
-Another attribute
-that can be set is the \k{class}.  This attribute can be bound to a list of
-class names. For instance, \code{class(x)<-"human"} will set the class of \k{x}
-to be \k{human}.  The main usage of attributes is for object-oriented
-programming. There are three object-orientation frameworks in R: S3, S4, and R5. 
-The S3 object system support single dispatch on the class of
-the first argument of a function, whereas the S4 object system allows
-multiple dispatch (on all arguments). R5 allows for users to define objects
-in a more imperative style. Some of the most widely used data
-types leverage attributes, e.g., data frames and matrices. A data frame, for instance,
-is a list of vectors with a class and a column name attribute, and matrices are vectors with a \k{dims} attribute.
-
-Scalar data types include the distinguished \k{NULL} value, instances of classes written using the S4 object system, closures, and environments.  The implementation of R has a number of other
-types listed in Figure~\ref{rtypes} for reference.
+Another attribute that can be set is the \k{class}.  This attribute
+can be bound to a list of class names. For instance,
+\code{class(x)<-"human"} will set the class of \k{x} to be \k{human}.
+The main usage of attributes is for object-oriented programming. There
+are three object-orientation frameworks in R: S3, S4, and R5.  The S3
+object system support single dispatch on the class of the first
+argument of a function, whereas the S4 object system allows multiple
+dispatch (on all arguments). R5 allows for users to define objects in
+a more imperative style. Some of the most widely used data types
+leverage attributes, e.g., data frames and matrices. A data frame, for
+instance, is a list of vectors with a class and a column name
+attribute, and matrices are vectors with a \k{dims} attribute.
+
+Scalar data types include the distinguished \k{NULL} value, instances
+of classes written using the S4 object system, closures, and
+environments.  The implementation of R has a number of other types
+listed in Figure~\ref{rtypes} for reference.
 
@@
 
-In this section, we set out to propose a candidate design for a type language to
-describe the arguments and return values of R functions.  The goal is not
-to propose a final design, but rather to propose a starting point for an iterative design process.
-Fig.~\ref{types} presents our type language. An early design choice was to
-stay close to the R language and only depart in small and hopefully
-non-controversial ways.  Functions types have the form \FUN{A_1, \dots,
-  A_n}\T where each $\A_i$ argument is either a type \T or \VARGS, a
-variable length argument list. The rest of this section details and
-motivates our design.
+In this section, we set out to propose a candidate design for a type
+language to describe the arguments and return values of R functions.
+The goal is not to propose a final design, but rather to propose a
+starting point for an iterative design process.  Fig.~\ref{types}
+presents our type language. An early design choice was to stay close
+to the R language and only depart in small and hopefully
+non-controversial ways.  Functions types have the form \FUN{A_1,
+  \dots, A_n}\T where each $\A_i$ argument is either a type \T or
+\VARGS, a variable length argument list. The rest of this section
+details and motivates our design.
 
@@
 \paragraph{Nullables} 
-The \nul type is inhabited by a singleton \code{NULL} value often used as a
-sentinel or a stand in for arguments with no value.  To capture this
-behavior, we introduce a nullable type \code{?}\T.  The functional difference between
-the \NA and \code{NULL} values is that the latter cannot be stored in
-vectors.
+The \nul type is inhabited by a singleton \code{NULL} value often used
+as a sentinel or a stand in for arguments with no value.  To capture
+this behavior, we introduce a nullable type \code{?}\T.  The
+functional difference between the \NA and \code{NULL} values is that
+the latter cannot be stored in vectors.
 
@@
 \paragraph{Classes} 
-R has more than one notion of type. As we mentioned, values can be attributed, and a very important
-attribute is the class of a value.  A class is a list of names that are used
-to mimic object-oriented programming. The type system includes class types
-written \CLASS{\ID_1,\dots,\ID_n}.
+R has more than one notion of type. As we mentioned, values can be
+attributed, and a very important attribute is the class of a value.  A
+class is a list of names that are used to mimic object-oriented
+programming. The type system includes class types written
+\CLASS{\ID_1,\dots,\ID_n}.
 
@@
 
-For this paper, we have built tooling to (\emph{a}) automate the extraction of
-raw type signatures from execution traces, (\emph{b}) infer type signatures from
-a set of raw types, and (\emph{c}) validate the inferred signatures by the means
-of contracts. Figure~\ref{fig:pipeline} shows an overview of this pipeline. This
-section gives details of the main steps. The pipeline is run using GNU
-parallel~\cite{tange2011_parallel} on Intel Xeon 6140, 2.30GHz with 72 cores and
-256GB of RAM.
+For this paper, we have built tooling to (\emph{a}) automate the
+extraction of raw type signatures from execution traces, (\emph{b})
+infer type signatures from a set of raw types, and (\emph{c}) validate
+the inferred signatures by the means of
+contracts. Figure~\ref{fig:pipeline} shows an overview of this
+pipeline. This section gives details of the main steps. The pipeline
+is run using GNU parallel~\cite{tange2011_parallel} on Intel Xeon
+6140, 2.30GHz with 72 cores and 256GB of RAM.
 
@@
   \includegraphics[width=\textwidth]{plots/analysis-pipeline-new.pdf}
-  \caption{The Analysis Pipeline. (1) the rather long time is because we had to
-    extract metadata from all CRAN packages before we could compose our
-    \PACKAGES package corpus. Here CRAN refers to the Comprehensive R Archive Network, R's repository of packages and libraries.}
+  \caption{The Analysis Pipeline. (1) the rather long time is because
+    we had to extract metadata from all CRAN packages before we could
+    compose our \PACKAGES package corpus. Here CRAN refers to the
+    Comprehensive R Archive Network, R's repository of packages and
+    libraries.}
   \label{fig:pipeline}
@@
 \noindent
-In other words, we take the union of the types occurring at individual argument
-positions rather than an union of function types. Furthermore, we apply some
-transformation on the types to keep the size of types in check.
-Figure~\ref{fig:siplification-rules} overviews the main simplification rules as
-well as the simple notion of subtyping that we have adopted here.
-Our subtyping rules are both born out of necessity (e.g., scalars are subtypes of vectors since a.
-that is the case in R and b.
-we want to allow types like \dbl | \VEC\dbl to simplify), as well as from prevalent data type conversion with no ambiguity (e.g., \int to \dbl).
+In other words, we take the union of the types occurring at individual
+argument positions rather than an union of function
+types. Furthermore, we apply some transformation on the types to keep
+the size of types in check.  Figure~\ref{fig:siplification-rules}
+overviews the main simplification rules as well as the simple notion
+of subtyping that we have adopted here.  Our subtyping rules are both
+born out of necessity (e.g., scalars are subtypes of vectors since
+a. that is the case in R and b. we want to allow types like \dbl |
+\VEC\dbl to simplify), as well as from prevalent data type conversion
+with no ambiguity (e.g., \int to \dbl).
 
@@
 
-Assuming that type sequences can be reordered freely, we rewrite types to
-minimize their size by removing redundant types, types that are subsumed by
-subtyping, immutable lists, and remove \nul type to replace them with
-nullables.  Functions are
-limited to twenty arguments as a simplifying assumption. Higher-order functions are conservatively inferred as \any
-$\rightarrow$ \any by \typetracer.
+Assuming that type sequences can be reordered freely, we rewrite types
+to minimize their size by removing redundant types, types that are
+subsumed by subtyping, immutable lists, and remove \nul type to
+replace them with nullables.  Functions are limited to twenty
+arguments as a simplifying assumption. Higher-order functions are
+conservatively inferred as \any $\rightarrow$ \any by \typetracer.
 
@@
 
-One way to validate an inferred function type is to check that it is respected in a different program which uses the function.
-For this purpose, we developed \contractr, an R package
-that allows one to decorate a function with assertions. We use it to insert type
-checking of arguments and return values.
+One way to validate an inferred function type is to check that it is
+respected in a different program which uses the function.  For this
+purpose, we developed \contractr, an R package that allows one to
+decorate a function with assertions. We use it to insert type checking
+of arguments and return values.
 
 \contractr's primary logic has been implemented in C++ to reduce its
-runtime overhead, and we have not observed a single segmentation fault during our use of
-the package.  It has been tested with GNU R-3.5.0 and hardened with a
-battery of 400 test cases. It works by modifying function definitions to
-insert a call to a type-checking function.
-
-In terms of usability, our checker is enabled automatically whenever a package
-is loaded in R. So, a simple invocation of \code{library(contractr)} causes
-contracts to be injected in all packages. On loading, \contractr scans all
-packages in the user's workspace and inserts contracts in functions for which
-type signatures are available. Then \contractr sets up package load hooks which
-are executed when new packages are loaded. Furthermore, \contractr automatically
-removes contracts from all functions and restores them to their original state
-when it is unloaded. The type signatures can be provided in an external file,
-thus avoiding the need to change the source code of checked packages. Type
-declarations can also be written directly in the adjunct function comments using
-\roxygen~\footnote{\url{https://cran.r-project.org/web/packages/roxygen2/}} annotation tags. \contractr introduces a new use \code{@type} tag. When
-a package these type signatures are extracted and stored in the appropriate
-location. The following is an example of a type annotation using \roxygen:
+runtime overhead, and we have not observed a single segmentation fault
+during our use of the package.  It has been tested with GNU R-3.5.0
+and hardened with a battery of 400 test cases. It works by modifying
+function definitions to insert a call to a type-checking function.
+
+In terms of usability, our checker is enabled automatically whenever a
+package is loaded in R. So, a simple invocation of
+\code{library(contractr)} causes contracts to be injected in all
+packages. On loading, \contractr scans all packages in the user's
+workspace and inserts contracts in functions for which type signatures
+are available. Then \contractr sets up package load hooks which are
+executed when new packages are loaded. Furthermore, \contractr
+automatically removes contracts from all functions and restores them
+to their original state when it is unloaded. The type signatures can
+be provided in an external file, thus avoiding the need to change the
+source code of checked packages. Type declarations can also be written
+directly in the adjunct function comments using
+\roxygen~\footnote{\url{https://cran.r-project.org/web/packages/roxygen2/}}
+annotation tags. \contractr introduces a new use \code{@type}
+tag. When a package these type signatures are extracted and stored in
+the appropriate location. The following is an example of a type
+annotation using \roxygen:
 
@@
 \paragraph{Checking the values} A number of properties can be checked
-by a simple tag check, namely whether a value is a null, environment, vector
-or scalar. Other properties require an inspection of the contents of a
-value, such as checking for the absence of \NA or the type of the elements of list. Union types
-require checking for each of the members of the union. As the usage
-of higher-order functions remains limited in R, we do not check higher-order functions in this version of the tool. Checking
-them could be done by dynamically wrapping functions with checks.  When argument values are wrapped in promises (this is not always the case
-due to compiler optimizations), in order to retain the non-strict semantics
-of R, the expression held in the promise is wrapped in a call to the type
-checker, and type checking is delayed until the promise is forced. This
-leads to corner cases such that the type checking of a function may happen
-after that function has returned.
-
-Return values require care as well.  Functions return the last expression
-they evaluate, thus to avoid having to analyze the code of the called
-function the checker will register a callback on the exit hook. The hook is
-executed in the function call's environment. Another wrinkle is due to
-longjumps which causes active function calls on the stack to be
-discarded. When they are discarded, their exit hooks are called but they do not have a return value to type-check. \contractr deals with
-this problem by allocating a unique sentinel object which serves as the
-return value for calls that are discarded. The exit hook does not call the
-type-checker if it see the sentinel.
+by a simple tag check, namely whether a value is a null, environment,
+vector or scalar. Other properties require an inspection of the
+contents of a value, such as checking for the absence of \NA or the
+type of the elements of list. Union types require checking for each of
+the members of the union. As the usage of higher-order functions
+remains limited in R, we do not check higher-order functions in this
+version of the tool. Checking them could be done by dynamically
+wrapping functions with checks.  When argument values are wrapped in
+promises (this is not always the case due to compiler optimizations),
+in order to retain the non-strict semantics of R, the expression held
+in the promise is wrapped in a call to the type checker, and type
+checking is delayed until the promise is forced. This leads to corner
+cases such that the type checking of a function may happen after that
+function has returned.
+
+Return values require care as well.  Functions return the last
+expression they evaluate, thus to avoid having to analyze the code of
+the called function the checker will register a callback on the exit
+hook. The hook is executed in the function call's environment. Another
+wrinkle is due to longjumps which causes active function calls on the
+stack to be discarded. When they are discarded, their exit hooks are
+called but they do not have a return value to type-check. \contractr
+deals with this problem by allocating a unique sentinel object which
+serves as the return value for calls that are discarded. The exit hook
+does not call the type-checker if it see the sentinel.
 
@@
 These packages come from the Comprehensive R Archive Network
-(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of R
-code with over \AllCranRnd packages\footnote{CRAN receives about 6 new package
-  submissions a day~\cite{Ligges2017}} containing over \AllRCodeRnd and
-\AllNativeCodeRnd lines of R and native code respectively. Unlike other open code
-repositories such as GitHub, CRAN is a curated repository where each submitted
-package must abide by a number of well-formedness rules that are automatically
-checked to assess package quality. 
-Notably, CRAN packages must have a set of {\it runnable} example, test, and vignette code which showcase package functionality.
-The code is run by CRAN, and only a successfully running package is admitted to the archive.
-
-We have downloaded and installed all available CRAN packages. Out of the
-\AllCranRnd packages, we managed to install \AllLoadableRnd. The main reason for
-this is that \rdt (and, by extension, \typetracer) is based on GNU-R 3.5.0 and some of the packages are not compatible with this version. 
-Some packages also require extra native
-dependencies which were not present on our servers. We defined two criteria for
+(CRAN\footnote{\url{http://cran.r-project.org}}), the largest
+repository of R code with over \AllCranRnd packages\footnote{CRAN
+receives about 6 new package submissions a day~\cite{Ligges2017}}
+containing over \AllRCodeRnd and \AllNativeCodeRnd lines of R and
+native code respectively. Unlike other open code repositories such as
+GitHub, CRAN is a curated repository where each submitted package must
+abide by a number of well-formedness rules that are automatically
+checked to assess package quality.  Notably, CRAN packages must have a
+set of {\it runnable} example, test, and vignette code which showcase
+package functionality.  The code is run by CRAN, and only a
+successfully running package is admitted to the archive.
+
+We have downloaded and installed all available CRAN packages. Out of
+the \AllCranRnd packages, we managed to install \AllLoadableRnd. The
+main reason for this is that \rdt (and, by extension, \typetracer) is
+based on GNU-R 3.5.0 and some of the packages are not compatible with
+this version.  Some packages also require extra native dependencies
+which were not present on our servers. We defined two criteria for
 including a package into the corpus:
 \begin{inparaenum}[(1)]
-  \item the package must have a runnable code that covers a significant part of the package
-  source code from which type signatures could be inferred, and
-  \item the package must have some reverse dependencies that will allows us to evaluate the
-  inferred types using the runnable code from these dependencies.
+  \item the package must have a runnable code that covers a
+    significant part of the package source code from which type
+    signatures could be inferred, and
+  \item the package must have some reverse dependencies that will
+    allows us to evaluate the inferred types using the runnable code
+    from these dependencies.
 \end{inparaenum}
-The concrete thresholds used were: at least \ThresholdCodeCoverage of expression
-coverage and at minimum \ThresholdRevdeps reverse dependencies. The code
-coverage was computed for each package using
-\covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R tool.
-The reverse package dependencies were extracted from the package metadata using
-built-in functions.
-
-The \CorpusLoadable selected packages contain \CorpusRunnableCodeRnd lines of
-runnable code in examples (\CorpusExamplesCodeRnd), tests (\CorpusTestsCodeRnd)
-and vignettes (\CorpusVignettesCodeRnd). Running this code results on average in
-\CorpusMeanExprCoverage package code coverage (the average for all of CRAN is
+The concrete thresholds used were: at least \ThresholdCodeCoverage of
+expression coverage and at minimum \ThresholdRevdeps reverse
+dependencies. The code coverage was computed for each package using
+\covr\footnote{\url{https://github.com/r-lib/covr}}, the R code
+coverage R tool.  The reverse package dependencies were extracted from
+the package metadata using built-in functions.
+
+The \CorpusLoadable selected packages contain \CorpusRunnableCodeRnd
+lines of runnable code in examples (\CorpusExamplesCodeRnd), tests
+(\CorpusTestsCodeRnd) and vignettes (\CorpusVignettesCodeRnd). Running
+this code results on average in \CorpusMeanExprCoverage package code
+coverage (the average for all of CRAN is
 \AllMeanExprCoverage). Together, there is \CorpusRevdesRnd (on average
-\CorpusMeanRevdes, median \CorpusMedianRevdes; CRAN average is \AllMeanRevdes,
-median \AllMedianRevdes) reverse dependencies with \CorpusRevdepRunnableCodeRnd
-runnable lines of code resulting in \CorpusRevdepMeanCodeCoverage coverage (on
-average) of the corpus packages.
-Together there are \CorpusFunctionsRnd defined R functions
+\CorpusMeanRevdes, median \CorpusMedianRevdes; CRAN average is
+\AllMeanRevdes, median \AllMedianRevdes) reverse dependencies with
+\CorpusRevdepRunnableCodeRnd runnable lines of code resulting in
+\CorpusRevdepMeanCodeCoverage coverage (on average) of the corpus
+packages.  Together there are \CorpusFunctionsRnd defined R functions
 (\CorpusPublicFunctionsRnd are from the packages' public APIs).
-\CorpusSThreeFunctionsRnd are S3 functions, either S3 generics or S3 methods. Packages in the corpus define
-\CorpusSThreeClasses S3 classes.
+\CorpusSThreeFunctionsRnd are S3 functions, either S3 generics or S3
+methods. Packages in the corpus define \CorpusSThreeClasses S3
+classes.
 
 \paragraph{User code}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-To represent end-user code in the corpus, we turned to Kaggle, an online
-platform for data-science and machine-learning. The website allows people to
-share data-science competitions and data-analysis problems together with data for
-which users try to find the best solution (something like a repository of hackathon or datathon code). The solutions, called \emph{kernels},
-are then posted on Kaggle either as plain scripts or as
-notebooks. One of the most popular competitions is about predicting passenger
-survival on Titanic\footnote{\url{https://www.kaggle.com/c/titanic}} with
-\KaggleDownloaded kernels in R (over 1/4 of all available R kernels) which we
-used for our corpus.
-
-Unlike CRAN, Kaggle is not a curated repository and therefore there are no
-guarantees about the quality of the code. After downloading all of the
-\KaggleDownloaded kernels and extracting the R code from the various
-formats,\footnote{We use {\sf rmarkdown} to convert from notebooks to R.} we
-found that \KaggleDuplicated were whole-file duplicates
-(\KaggleDuplicatedRatio). From the resulting \KaggleRunnable kernels,
-\KaggleFail failed to execute. Next to various runtime exceptions, common
-problems were missing libraries (no longer available for R 3.5), parse errors
-and misspelled package names. The final set contains \KaggleSuccess kernels with
-\KaggleSuccessCodeRnd lines of R code. The Kaggle kernels are used for additional
-validation of the inferred types.
+To represent end-user code in the corpus, we turned to Kaggle, an
+online platform for data-science and machine-learning. The website
+allows people to share data-science competitions and data-analysis
+problems together with data for which users try to find the best
+solution (something like a repository of hackathon or datathon
+code). The solutions, called \emph{kernels}, are then posted on Kaggle
+either as plain scripts or as notebooks. One of the most popular
+competitions is about predicting passenger survival on
+Titanic\footnote{\url{https://www.kaggle.com/c/titanic}} with
+\KaggleDownloaded kernels in R (over 1/4 of all available R kernels)
+which we used for our corpus.
+
+Unlike CRAN, Kaggle is not a curated repository and therefore there
+are no guarantees about the quality of the code. After downloading all
+of the \KaggleDownloaded kernels and extracting the R code from the
+various formats,\footnote{We use {\sf rmarkdown} to convert from
+notebooks to R.} we found that \KaggleDuplicated were whole-file
+duplicates (\KaggleDuplicatedRatio). From the resulting
+\KaggleRunnable kernels, \KaggleFail failed to execute. Next to
+various runtime exceptions, common problems were missing libraries (no
+longer available for R 3.5), parse errors and misspelled package
+names. The final set contains \KaggleSuccess kernels with
+\KaggleSuccessCodeRnd lines of R code. The Kaggle kernels are used for
+additional validation of the inferred types.
 
 \paragraph{Type usage}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-During execution 3,147 different types were observed.  Classes are the most
-common types, accounting for roughly 31\% of types of arguments.  The most
-common classes are \c{matrices} (12\%), \c{data.frames} (7.5\%),
-\c{formulas} (2\%), \c{factors} (2\%), and \c{tibbles} (2\%).  Roughly 25\%
-of classes are part of R's base libraries, the others are user-defined.
-Scalars and vectors are the next most common kind, making up 41\% of
-remaining types. with scalars making up 28\% of types and vectors 12\%.
-Nulls and lists follow at 8\% and 7\% respectively, and the vararg type
-makes up 6\% of arguments.  This all totals up to over 90\% of types.
-Table~\ref{ctb} reports on the 10 most frequent types occurring in the
-corpus. The first row of the table reads: the \dbl type occurs in 12,298 (11.24\%) argument types, and accounts for over 12 million (20.3\%) of the types observed by \typetracer's dynamic analysis.
+During execution 3,147 different types were observed.  Classes are the
+most common types, accounting for roughly 31\% of types of arguments.
+The most common classes are \c{matrices} (12\%), \c{data.frames}
+(7.5\%), \c{formulas} (2\%), \c{factors} (2\%), and \c{tibbles} (2\%).
+Roughly 25\% of classes are part of R's base libraries, the others are
+user-defined.  Scalars and vectors are the next most common kind,
+making up 41\% of remaining types. with scalars making up 28\% of
+types and vectors 12\%.  Nulls and lists follow at 8\% and 7\%
+respectively, and the vararg type makes up 6\% of arguments.  This all
+totals up to over 90\% of types.  Table~\ref{ctb} reports on the 10
+most frequent types occurring in the corpus. The first row of the
+table reads: the \dbl type occurs in 12,298 (11.24\%) argument types,
+and accounts for over 12 million (20.3\%) of the types observed by
+\typetracer's dynamic analysis.
 
@@
 We ran \typetracer on the test, example, and vignette code of the
-aforementioned corpus of \PACKAGES packages and successfully inferred types
-for \NUMFUNCTIONS functions.  Table~\ref{tbl:typesig} illustrates the
-process with ten representative signatures.  Many of the features of our
-type language are represented here, and some signatures are telling of the
-function's behaviour.  For example, consider \code{decrypt_envelope}: the
-first three parameters of the function are byte arrays, and the fourth
-argument is an RSA key, used to decrypt some of the inputs, and the output
-of the function is another byte array.  As another example, consider
-\code{Traverse}: according to the function documentation, it takes the root
-of a tree and traverses it in an order specified by the second argument.  We
-see that reflected in the type, where the first argument has type
+aforementioned corpus of \PACKAGES packages and successfully inferred
+types for \NUMFUNCTIONS functions.  Table~\ref{tbl:typesig}
+illustrates the process with ten representative signatures.  Many of
+the features of our type language are represented here, and some
+signatures are telling of the function's behaviour.  For example,
+consider \code{decrypt_envelope}: the first three parameters of the
+function are byte arrays, and the fourth argument is an RSA key, used
+to decrypt some of the inputs, and the output of the function is
+another byte array.  As another example, consider \code{Traverse}:
+according to the function documentation, it takes the root of a tree
+and traverses it in an order specified by the second argument.  We see
+that reflected in the type, where the first argument has type
 \CLASS{\texttt{Node}, \texttt{R6}} and the second argument had type
@@
 
-Table~\ref{fig:singlefreq} provides a
-breakdown of types occurring in arguments without a union. Scalar, class and
-vector are the most common type categories. The shaded rows correspond to
-polymorphic types. 
-When an argument's type is \nul, we say that the argument is polymorphic due to a limitation in our analysis:
-In R, it is common for programmers to include default values for arguments, and in many cases this value happens to be {\tt NULL}.
-Our type analysis will report a {\tt null} type for these arguments if they are {\it never passed a value} during testing.
-We interpret these instances of \nul as polymorphic to capture that we cannot be sure of the actual type.
-
-Removing the aforementioned instances of polymorphism gives us \CountMonomorphicPositions
-(\PercMonomorphicPositions) monomorphic positions in a corpus of
-\CountPositions parameters. With close to 60\% of monomorphic argument or
-return values, it is fair to say that even a simple type language provides
-significant benefits.
+Table~\ref{fig:singlefreq} provides a breakdown of types occurring in
+arguments without a union. Scalar, class and vector are the most
+common type categories. The shaded rows correspond to polymorphic
+types.  When an argument's type is \nul, we say that the argument is
+polymorphic due to a limitation in our analysis: In R, it is common
+for programmers to include default values for arguments, and in many
+cases this value happens to be {\tt NULL}.  Our type analysis will
+report a {\tt null} type for these arguments if they are {\it never
+  passed a value} during testing.  We interpret these instances of
+\nul as polymorphic to capture that we cannot be sure of the actual
+type.
+
+Removing the aforementioned instances of polymorphism gives us
+\CountMonomorphicPositions (\PercMonomorphicPositions) monomorphic
+positions in a corpus of \CountPositions parameters. With close to
+60\% of monomorphic argument or return values, it is fair to say that
+even a simple type language provides significant benefits.
 
@@
 
-If we look at the numbers from the point of view of functions and count how
-many of their arguments are polymorphic, we observe that
+If we look at the numbers from the point of view of functions and
+count how many of their arguments are polymorphic, we observe that
 \PercMonomorphicFunctions (\CountMonomorphicFunctions) functions are
 monomorphic. The remaining \PercPolymorphicFunctions
-(\CountPolymorphicFunctions) have at least one union or polymorphic parameter or
-return type.  Figure~\ref{fig:funpoly} shows the distribution of functions
-against the number of polymorphic arguments. Finally, we count that
-\CountMonomorphicPackages out of the 412 packages export only monomorphic
-functions.
+(\CountPolymorphicFunctions) have at least one union or polymorphic
+parameter or return type.  Figure~\ref{fig:funpoly} shows the
+distribution of functions against the number of polymorphic
+arguments. Finally, we count that \CountMonomorphicPackages out of the
+412 packages export only monomorphic functions.
 
@@
 \paragraph{Higher-Order Functions}
-\typetracer assigns the type \CLASS{\texttt{function}} to function values.
-The number of positions that receive a function (possibly as part of a union) is 1,705, which is just
-1.51\% of all the positions for which we infer types. Given the small number of
-occurrences, it is not worth complicating the inferred types with a complete
-signature for these functions.
+\typetracer assigns the type \CLASS{\texttt{function}} to function
+values.  The number of positions that receive a function (possibly as
+part of a union) is 1,705, which is just 1.51\% of all the positions
+for which we infer types. Given the small number of occurrences, it is
+not worth complicating the inferred types with a complete signature
+for these functions.
 
 \paragraph{Structs}
-While experimenting with various design, we consider adding a struct type to
-capture lists with named elements that can be accessed with the \code{$}
-operators. We ended up discarding those types as they grew large and were often
-only representative of the example data being manipulated.  Consider
-function \code{cv.model}, its argument \code{x} is observed to be of
-\CLASS{\texttt{aov}, \texttt{lm}} or \CLASS{\texttt{lm}}.  Internally,
-linear models are represented as lists with named elements.  The pollution is
-illustrated by the lines after the function definition.  They load an example
-data set and test the function \code{cv.model}.  The
-\code{data(sweetpotato)} expression loads a sample data set. The fields of
-\code{sweetpotato} will be recorded when \code{cv.model} is called.
+While experimenting with various design, we consider adding a struct
+type to capture lists with named elements that can be accessed with
+the \code{$} operators. We ended up discarding those types as they
+grew large and were often only representative of the example data
+being manipulated.  Consider function \code{cv.model}, its argument
+\code{x} is observed to be of \CLASS{\texttt{aov}, \texttt{lm}} or
+\CLASS{\texttt{lm}}.  Internally, linear models are represented as
+lists with named elements.  The pollution is illustrated by the lines
+after the function definition.  They load an example data set and test
+the function \code{cv.model}.  The \code{data(sweetpotato)} expression
+loads a sample data set. The fields of \code{sweetpotato} will be
+recorded when \code{cv.model} is called.
 
@@
 \paragraph{Objects}
-While we record classes, our analysis does not deal with method dispatch.  R
-has multiple disparate object systems called S3, S4, and R5.  The
-\code{class} attribute is used by these systems to dispatch methods. S3 does
-single dispatch, S4 does multiple dispatch and R5 supports imperative
-objects.  The mechanics of S4 dispatch are more complex than for S3, and
-users can define their own class hierarchies that we would need to
-incorporate in our type analysis and contract checking frameworks.  We found
-limited use of S4 during our analysis. Coming up with a type system that
-accounts for all of these factors and consolidates multiple
-object-orientation frameworks in a single language design is an interesting
-problem in and of itself, one we leave for future work.
+While we record classes, our analysis does not deal with method
+dispatch.  R has multiple disparate object systems called S3, S4, and
+R5.  The \code{class} attribute is used by these systems to dispatch
+methods. S3 does single dispatch, S4 does multiple dispatch and R5
+supports imperative objects.  The mechanics of S4 dispatch are more
+complex than for S3, and users can define their own class hierarchies
+that we would need to incorporate in our type analysis and contract
+checking frameworks.  We found limited use of S4 during our
+analysis. Coming up with a type system that accounts for all of these
+factors and consolidates multiple object-orientation frameworks in a
+single language design is an interesting problem in and of itself, one
+we leave for future work.
 
 \paragraph{Matrices}
-Matrices are instances of the eponymous class, representing 10.71\% of all classes occurring in types. They have a \code{dims}
-attribute indicating dimensions, and while not codified in the language
-semantics, many internal functions coerce vectors to matrices automatically.
-For example, the \code{rowWeightedMeans} function calculates the weighted means of rows.
-The programmer added a type check for \code{x}.
+Matrices are instances of the eponymous class, representing 10.71\% of
+all classes occurring in types. They have a \code{dims} attribute
+indicating dimensions, and while not codified in the language
+semantics, many internal functions coerce vectors to matrices
+automatically.  For example, the \code{rowWeightedMeans} function
+calculates the weighted means of rows.  The programmer added a type
+check for \code{x}.
 
@@
 \paragraph{Data Frames}
-One of the most popular classes in R is the \code{data.frame} class, making
-up  8.15\% of observed classes.  Data frames and the derivative
-\code{tibble} and \code{data.table} types underpin much of the idiomatic usage of R.
-One way to deal with data frames is through the struct type, with a named
-field for each column of the data frame, but as mentioned previously structs
-introduced undue noise.  Further complicating data frames is that many
-functions built to operate on them operate in a name-agnostic way.  For
-instance, the \code{tidyverse} package ecosystem allows programmers to pass
-column names to functions which operate on their data frames.  In base R,
-typical data frame use is to use string column names to select rows from
-the frame (unless only a single column is of interest, wherein the \code{$}
-syntax is appropriate).
-In sum, data frames are a popular class of R values, and have spawned many derivative data types, such as tibbles and data tables.
-We include a \CLASS{{\tt data.frame}} type to cover most use-cases, and we leave a richer type for future work on a full fledged object-oriented type system.
+One of the most popular classes in R is the \code{data.frame} class,
+making up 8.15\% of observed classes.  Data frames and the derivative
+\code{tibble} and \code{data.table} types underpin much of the
+idiomatic usage of R.  One way to deal with data frames is through the
+struct type, with a named field for each column of the data frame, but
+as mentioned previously structs introduced undue noise.  Further
+complicating data frames is that many functions built to operate on
+them operate in a name-agnostic way.  For instance, the
+\code{tidyverse} package ecosystem allows programmers to pass column
+names to functions which operate on their data frames.  In base R,
+typical data frame use is to use string column names to select rows
+from the frame (unless only a single column is of interest, wherein
+the \code{$} syntax is appropriate).  In sum, data frames are a
+popular class of R values, and have spawned many derivative data
+types, such as tibbles and data tables.  We include a \CLASS{{\tt
+    data.frame}} type to cover most use-cases, and we leave a richer
+type for future work on a full fledged object-oriented type system.
 
@@
 
-We now ask {\it how robust are the inferred types}? To measure this, we conducted another
-large-scale experiment: for each package in the corpus, using the inferred type
-signatures as contracts we ran all of the CRAN reverse dependencies for that
-package. In total we ran \NUMPKGSEVAL unique packages and recorded
-\NUMASSERTIONS total assertions. Overall, we found that only
-\PERCFAILEDASSERTIONSWUNDEF of contract assertions failed. The limit on number of arguments
-(we record only 20) accounted for \PERCASSERTIONSUNDEF failed assertions. We
-found that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of function types
-never failed. The number of immaculate function types increases to
-\PERCSUCCFUNNOSTHREE if we discount S3 object method dispatch. Overall, these
-numbers are promising, and suggest that the type signatures are indeed robust.
+We now ask {\it how robust are the inferred types}? To measure this,
+we conducted another large-scale experiment: for each package in the
+corpus, using the inferred type signatures as contracts we ran all of
+the CRAN reverse dependencies for that package. In total we ran
+\NUMPKGSEVAL unique packages and recorded \NUMASSERTIONS total
+assertions. Overall, we found that only \PERCFAILEDASSERTIONSWUNDEF of
+contract assertions failed. The limit on number of arguments (we
+record only 20) accounted for \PERCASSERTIONSUNDEF failed
+assertions. We found that \PERCSUCCARG of parameter types and
+\PERCSUCCFUNS of function types never failed. The number of immaculate
+function types increases to \PERCSUCCFUNNOSTHREE if we discount S3
+object method dispatch. Overall, these numbers are promising, and
+suggest that the type signatures are indeed robust.
 
@@
 Table~\ref{tbl:fail-breakdown-0}.  Accounting for 36.36\% of assertion
-failures are cases where a \VEC\dbl is passed where a \CLASS{\texttt{matrix}} is
-expected.  Considering these types, we might imagine them to be compatible, as a vector is just a one-dimensional matrix.
-However, not allowing this coercion was a deliberate design decision, as coercion of this kind is ad hoc at best, and unfortunately not a practice codified in the language. 
-For example, if the vector has length $n$, should it be a $1 \times n$ or $n \times 1$ dimensional matrix?
-
-In a similar vein, another popular
-failing assertion is checking if a ${\bf dbl}[]$ has type ${\bf int}[]$,
-another case of commonly performed coercion.  We did not include these types
-of coercions in our type annotation framework as programmers cannot rely on
-them, and it is not always the case that the coercions are safe to perform.
-
-The second row of the table is exemplary of a pattern where vectors are passed when scalars are expected.
-In these cases, the functions exhibiting these assertion failures were under-tested, and can operate just as well on vectors of values.
-As an example, this failure occurred in functions from the \code{lubridate} package which provides date/time functionality.
-Many functions, e.g., \code{date_decimal} and \code{make_datetime} turn doubles into \CLASS{{\tt POSIXct, POSIXt}} (which are dates in R), and they can easily operate on vectors of doubles, producing lists of dates.
-
-Finally, we point out that assertion failures of, e.g., \CLASS{{\tt data.frame}} values being passed to \CLASS{{\tt data.frame, tbl, tbl\_df}} arguments and \CLASS{{\tt xml\_node}} values being passed to an argument expecting other XML-like classes are related to our simplified take on class types.
-Our type system does not encode user-defined subtyping and coercion, which could help address these mismatches.
+failures are cases where a \VEC\dbl is passed where a
+\CLASS{\texttt{matrix}} is expected.  Considering these types, we
+might imagine them to be compatible, as a vector is just a
+one-dimensional matrix.  However, not allowing this coercion was a
+deliberate design decision, as coercion of this kind is ad hoc at
+best, and unfortunately not a practice codified in the language.  For
+example, if the vector has length $n$, should it be a $1 \times n$ or
+$n \times 1$ dimensional matrix?
+
+In a similar vein, another popular failing assertion is checking if a
+${\bf dbl}[]$ has type ${\bf int}[]$, another case of commonly
+performed coercion.  We did not include these types of coercions in
+our type annotation framework as programmers cannot rely on them, and
+it is not always the case that the coercions are safe to perform.
+
+The second row of the table is exemplary of a pattern where vectors
+are passed when scalars are expected.  In these cases, the functions
+exhibiting these assertion failures were under-tested, and can operate
+just as well on vectors of values.  As an example, this failure
+occurred in functions from the \code{lubridate} package which provides
+date/time functionality.  Many functions, e.g., \code{date_decimal}
+and \code{make_datetime} turn doubles into \CLASS{{\tt POSIXct,
+    POSIXt}} (which are dates in R), and they can easily operate on
+vectors of doubles, producing lists of dates.
+
+Finally, we point out that assertion failures of, e.g., \CLASS{{\tt
+    data.frame}} values being passed to \CLASS{{\tt data.frame, tbl,
+    tbl\_df}} arguments and \CLASS{{\tt xml\_node}} values being
+passed to an argument expecting other XML-like classes are related to
+our simplified take on class types.  Our type system does not encode
+user-defined subtyping and coercion, which could help address these
+mismatches.
 
@@
 
-In addition to the number of failed contract checks, we were interested in
-how many functions had a parameter where a contract check failed, and
-overall we found this to be the case in \PROPFUNSFAILEDCHECK of functions.
-To subdivide this number, we discounted functions that were performing S3
-dispatch, as they exhibit user-defined polymorphism which we do not handle.
-Removing those functions, we see that the proportion of functions with failed contract checks falls to 
-\PROPFUNSFAILEDCHECKNOSTHREE. These remaining functions were under-tested, as calls to
-these functions represent only \PERCCALLSBADFUNSINTESTS of recorded calls
-during \typetracer's run on the core corpus to infer types.
+In addition to the number of failed contract checks, we were
+interested in how many functions had a parameter where a contract
+check failed, and overall we found this to be the case in
+\PROPFUNSFAILEDCHECK of functions.  To subdivide this number, we
+discounted functions that were performing S3 dispatch, as they exhibit
+user-defined polymorphism which we do not handle.  Removing those
+functions, we see that the proportion of functions with failed
+contract checks falls to \PROPFUNSFAILEDCHECKNOSTHREE. These remaining
+functions were under-tested, as calls to these functions represent
+only \PERCCALLSBADFUNSINTESTS of recorded calls during \typetracer's
+run on the core corpus to infer types.
 
@@
 \PROPARGSFAILINGASSERTS of argument types failed.
-Table~\ref{tbl:fail-breakdown-0} showed the runtime occurrences, but that
-data alone does not tell the full story, as some failures may be
-overrepresented if, e.g., a failing contract assertion was in a loop.  We were
-interested in knowing for each of the most common violations in
-Table~\ref{tbl:fail-breakdown-0}, how many different arguments had that
-type, and how many of those exhibited the contract failure in question.
-Table~\ref{tbl:fail-breakdown-1} breaks down the failed assertions by type,
-folding away multiple identical failed contract assertions for the same
-parameter position.  The first row of this table reads: a value of type
-\VEC\dbl was passed to 15 different function parameters expecting a
-\CLASS{\texttt{matrix}}, of which there are 1522 in total: 18 (1.18\%) of these
+Table~\ref{tbl:fail-breakdown-0} showed the runtime occurrences, but
+that data alone does not tell the full story, as some failures may be
+overrepresented if, e.g., a failing contract assertion was in a loop.
+We were interested in knowing for each of the most common violations
+in Table~\ref{tbl:fail-breakdown-0}, how many different arguments had
+that type, and how many of those exhibited the contract failure in
+question.  Table~\ref{tbl:fail-breakdown-1} breaks down the failed
+assertions by type, folding away multiple identical failed contract
+assertions for the same parameter position.  The first row of this
+table reads: a value of type \VEC\dbl was passed to 15 different
+function parameters expecting a \CLASS{\texttt{matrix}}, of which
+there are 1522 in total: 18 (1.18\%) of these
 \CLASS{\texttt{matrix}}-typed parameters were passed \VEC\dbl values.
@@
 We see that even though the double vector and matrix issue was wildly
-prevalent in the raw, dynamic contract evaluation numbers, the number of
-actual function argument types that were violated is very small.  The story
-is similar with the double and integer coercion we mentioned earlier: it
-represents many dynamic contract failures, but very few of the
-\VEC\int-typed arguments have their contracts violated by \VEC\dbl-typed
-values.  Row six is interesting: we see that rather often arguments
-expecting \CLASS{\texttt{timeSeries}} data are passed
-\CLASS{\texttt{matrix}} values.  This is a quirk of the \code{timeSeries}
-package, whose functions often accept matrices and vectors, converting them
-to time series in an ad hoc manner.  Note that the code coverage of the
-\code{timeSeries} tests, examples, and vignettes package code is only 58\%, which is one possible explanation of why these
-contract failures are occurring: the types that \typetracer generates are
-only as good as the test code its run on.
+prevalent in the raw, dynamic contract evaluation numbers, the number
+of actual function argument types that were violated is very small.
+The story is similar with the double and integer coercion we mentioned
+earlier: it represents many dynamic contract failures, but very few of
+the \VEC\int-typed arguments have their contracts violated by
+\VEC\dbl-typed values.  Row six is interesting: we see that rather
+often arguments expecting \CLASS{\texttt{timeSeries}} data are passed
+\CLASS{\texttt{matrix}} values.  This is a quirk of the
+\code{timeSeries} package, whose functions often accept matrices and
+vectors, converting them to time series in an ad hoc manner.  Note
+that the code coverage of the \code{timeSeries} tests, examples, and
+vignettes package code is only 58\%, which is one possible explanation
+of why these contract failures are occurring: the types that
+\typetracer generates are only as good as the test code its run on.
 
@@
 Table~\ref{tbl:fail-breakdown-2} presents data on the most frequently
-violated contracts amongst the most frequently occurring argument types.  We
-selected argument types which were in the 90th percentile of argument type
-occurrences, computed the most frequent type signature violations among
-them, and reported the most frequently violated contracts together with the
-type of the value that violated that contract.  The first row of the table
-reads: 31 function arguments with \VEC\int type are passed \VEC\dbl values instead,
-and 624 arguments have \VEC\int type, representing a failure rate of 4.97\%.
-
-Had we failed to capture some key
-usage pattern of R with our type annotation framework, we would likely see
-it here, and we can see this in action if we consider
-Table~\ref{tbl:fail-breakdown-3}, which was obtained identically to
-Table~\ref{tbl:fail-breakdown-2} except selecting arguments in the 80th percentile
-instead.  The most frequent argument type violation pattern in that of
-\CLASS{{\tt matrix}}, \VEC\dbl, and \CLASS{\texttt{data.frame}} values passed to arguments expecting \CLASS{{\tt timeSeries}}.
-This occurs in 35.90\% of such arguments,
-and represents cases where tests did not adequately cover all valid function
-inputs.  Separate from the issue of testing, we can capture this behaviour
-with user-defined subtyping or coercion, as the data types which were passed are readily convertible to \CLASS{{\tt timeSeries}}.
+violated contracts amongst the most frequently occurring argument
+types.  We selected argument types which were in the 90th percentile
+of argument type occurrences, computed the most frequent type
+signature violations among them, and reported the most frequently
+violated contracts together with the type of the value that violated
+that contract.  The first row of the table reads: 31 function
+arguments with \VEC\int type are passed \VEC\dbl values instead, and
+624 arguments have \VEC\int type, representing a failure rate of
+4.97\%.
+
+Had we failed to capture some key usage pattern of R with our type
+annotation framework, we would likely see it here, and we can see this
+in action if we consider Table~\ref{tbl:fail-breakdown-3}, which was
+obtained identically to Table~\ref{tbl:fail-breakdown-2} except
+selecting arguments in the 80th percentile instead.  The most frequent
+argument type violation pattern in that of \CLASS{{\tt matrix}},
+\VEC\dbl, and \CLASS{\texttt{data.frame}} values passed to arguments
+expecting \CLASS{{\tt timeSeries}}.  This occurs in 35.90\% of such
+arguments, and represents cases where tests did not adequately cover
+all valid function inputs.  Separate from the issue of testing, we can
+capture this behaviour with user-defined subtyping or coercion, as the
+data types which were passed are readily convertible to \CLASS{{\tt
+    timeSeries}}.
 
@@
 % END Autogenerated
- \caption{Highest failure rate among popular argument types, for argument signatures whose frequency is in the 90th percentile.}
+ \caption{Highest failure rate among popular argument types, for
+   argument signatures whose frequency is in the 90th percentile.}
  \label{tbl:fail-breakdown-2}
@@
 
- We drew a number of other observations from the contract assertion failure data.
-
- First, we were curious about how many \nul-typed values flowed into non-nullable arguments, and we found that they accounted for \NUMNULLASSERTERRORS of contract assertion failures in \NUMFUNSWITHNULLASSERTERROR functions.
- We manually inspected some of the offending functions and observed three main patterns.
-
- First, we found that many of these errors occurred in arguments that have a {\tt NULL} default value.
- This would be the case when the programmer only tested the function by passing values to these null-by-default arguments, and clients of the package make use of the default.
- As an example, we observed a contract assertion failure when {\tt inner} and {\tt labels} were {\tt NULL} in the following function:
+ We drew a number of other observations from the contract assertion
+ failure data.
+
+ First, we were curious about how many \nul-typed values flowed into
+ non-nullable arguments, and we found that they accounted for
+ \NUMNULLASSERTERRORS of contract assertion failures in
+ \NUMFUNSWITHNULLASSERTERROR functions.  We manually inspected some of
+ the offending functions and observed three main patterns.
+
+ First, we found that many of these errors occurred in arguments that
+ have a {\tt NULL} default value.  This would be the case when the
+ programmer only tested the function by passing values to these
+ null-by-default arguments, and clients of the package make use of the
+ default.  As an example, we observed a contract assertion failure
+ when {\tt inner} and {\tt labels} were {\tt NULL} in the following
+ function:
  \begin{lstlisting}
@@
 
- The other two patterns are for arguments with no default value, where either the call results in an error (perhaps explicitly handled by the programmer), or results in valid function behaviour that was untested by the original package designer.
- Here, the first case can be explained by a lack of testing, and the second case is explained by programmers not fully understanding R's language semantics.
- For example, we observed this kind of error in the following function: 
+ The other two patterns are for arguments with no default value, where
+ either the call results in an error (perhaps explicitly handled by
+ the programmer), or results in valid function behaviour that was
+ untested by the original package designer.  Here, the first case can
+ be explained by a lack of testing, and the second case is explained
+ by programmers not fully understanding R's language semantics.  For
+ example, we observed this kind of error in the following function:
  \begin{lstlisting}
@@
  \end{lstlisting}
-Here, if {\tt x} is {\tt NULL}, the second branch of the conditional will be triggered (as in R, \code{length(NULL) == 0}), and the function will return \code{as.integer(NULL)}, which curiously returns \code{integer(0)}, a zero-length vector of integers (one might expect it to return the integer \code{NA} value, or error).
-
- Next, we analyzed how often vectors were passed to arguments expecting scalars.
- We found that \PERCDYNAMICVECERRORS of dynamic contract assertion failures were of this type, and these errors were present in \NUMFUNSWITHVECASSERTERROR functions.
- Besides an outright error, this kind of contract assertion failure might indicate that a function was not well-tested, in that it was only ever tested with unit-length vectors being passed to an argument which is intended to have a vector type.
- Further, these errors may reveal functions that were not designed with a vector-typed argument in mind, but can in fact handle vectors of values (in R, most functions that can accept scalars can also accept vectors).
- As an example, consider the function \code{BBmisc::strrepeat}, which takes a string and repeats it a specified number of times:
+Here, if {\tt x} is {\tt NULL}, the second branch of the conditional
+will be triggered (as in R, \code{length(NULL) == 0}), and the
+function will return \code{as.integer(NULL)}, which curiously returns
+\code{integer(0)}, a zero-length vector of integers (one might expect
+it to return the integer \code{NA} value, or error).
+
+ Next, we analyzed how often vectors were passed to arguments
+ expecting scalars.  We found that \PERCDYNAMICVECERRORS of dynamic
+ contract assertion failures were of this type, and these errors were
+ present in \NUMFUNSWITHVECASSERTERROR functions.  Besides an outright
+ error, this kind of contract assertion failure might indicate that a
+ function was not well-tested, in that it was only ever tested with
+ unit-length vectors being passed to an argument which is intended to
+ have a vector type.  Further, these errors may reveal functions that
+ were not designed with a vector-typed argument in mind, but can in
+ fact handle vectors of values (in R, most functions that can accept
+ scalars can also accept vectors).  As an example, consider the
+ function \code{BBmisc::strrepeat}, which takes a string and repeats
+ it a specified number of times:
  \begin{lstlisting}
@@
  \end{lstlisting}
- This function was only ever tested with unit-length vectors passed to {\tt x}, even though technically it can handle longer vectors, as per the two sample calls above.
- This could be attributed to poor quality testing, or misunderstanding language semantics (e.g., misunderstanding the semantics of \code{paste0} and \code{rep.int}), but we found other instances of type errors where the functions really ought to have been tested with the offending type, for instance:
+ This function was only ever tested with unit-length vectors passed to
+ {\tt x}, even though technically it can handle longer vectors, as per
+ the two sample calls above.  This could be attributed to poor quality
+ testing, or misunderstanding language semantics (e.g.,
+ misunderstanding the semantics of \code{paste0} and \code{rep.int}),
+ but we found other instances of type errors where the functions
+ really ought to have been tested with the offending type, for
+ instance:
  \begin{lstlisting}
@@
  \end{lstlisting}
- As per the documentation, this function is intended to take a vector, and generate all permutations of elements in that vector.
-If given a scalar ``integer'' $n$, it will generate all permutations for the list $[1, 2, ..., n]$.
-Interestingly, the function was only tested by passing an integer, and not ever with a vector (even though, presumably, that is the main utility of the function).
-
-Finally, we were curious to see what patterns of errors occurred in arguments expecting classes.
-Overall we found that \PERCFAILASSCLASS of assertion failures were on arguments which were expecting a class in some way (\PERCFAILEDONECLASS of assertion failures were on monomorphic arguments expecting a class, the remainder on polymorphic arguments with at least one option being a class).
-
-There are two broad divisions which account for most of the class-related contract assertion failures that are not outright errors.
-First, we observe a class of errors related to classes being passed to arguments expecting a different, yet convertible class.
-For instance, we observed \CLASS{{\tt data.frame}} values being passed to arguments expecting tibbles or data.tables (data frames have a straightforward conversion to these classes).
-Second, we observe a class of errors related more to coercion between simple data types and classes.
-As an example, consider the aforementioned assertion failures in the \code{timeSeries} package, and as a further example we found many instances of \CLASS{{\tt matrix}}, \CLASS{{\tt data.frame}}, and vectors being passed to arguments expecting \CLASS{{\tt array}}, a generalization of matrices.
+ As per the documentation, this function is intended to take a vector,
+ and generate all permutations of elements in that vector.  If given a
+ scalar ``integer'' $n$, it will generate all permutations for the
+ list $[1, 2, ..., n]$.  Interestingly, the function was only tested
+ by passing an integer, and not ever with a vector (even though,
+ presumably, that is the main utility of the function).
+
+Finally, we were curious to see what patterns of errors occurred in
+arguments expecting classes.  Overall we found that \PERCFAILASSCLASS
+of assertion failures were on arguments which were expecting a class
+in some way (\PERCFAILEDONECLASS of assertion failures were on
+monomorphic arguments expecting a class, the remainder on polymorphic
+arguments with at least one option being a class).
+
+There are two broad divisions which account for most of the
+class-related contract assertion failures that are not outright
+errors.  First, we observe a class of errors related to classes being
+passed to arguments expecting a different, yet convertible class.  For
+instance, we observed \CLASS{{\tt data.frame}} values being passed to
+arguments expecting tibbles or data.tables (data frames have a
+straightforward conversion to these classes).  Second, we observe a
+class of errors related more to coercion between simple data types and
+classes.  As an example, consider the aforementioned assertion
+failures in the \code{timeSeries} package, and as a further example we
+found many instances of \CLASS{{\tt matrix}}, \CLASS{{\tt
+    data.frame}}, and vectors being passed to arguments expecting
+\CLASS{{\tt array}}, a generalization of matrices.
 
@@
 
- To further validate our inferred types, we repeated the experiment discussed in this section on end-user R code found on the Kaggle competition website.
-
- By-and-large, we saw no meaningful difference in the two data sets, with the contract assertion failure patterns being repeated from the reverse dependencies.
- Overall, we observed that \PERCFAILEDASSERTIONSKAGGLE of all contract assertions failed while running Kaggle code.
- If we remove assertion failures related to our simplifying assumption that function types will not have more than 20 arguments, that number drops to a mere \PERCFAILEDASSERTIONSKAGGLENOUNDEF.
- In all, \PROPFUNSWITHFAILEDASSERTKAGGLE functions had at least one contract failure.
- There were \NUMTOTALASSERTIONSKAGGLE assertions in total, on \NUMASSERTEDFUNCTIONSKAGGLE functions.
+ To further validate our inferred types, we repeated the experiment
+ discussed in this section on end-user R code found on the Kaggle
+ competition website.
+
+ By-and-large, we saw no meaningful difference in the two data sets,
+ with the contract assertion failure patterns being repeated from the
+ reverse dependencies.  Overall, we observed that
+ \PERCFAILEDASSERTIONSKAGGLE of all contract assertions failed while
+ running Kaggle code.  If we remove assertion failures related to our
+ simplifying assumption that function types will not have more than 20
+ arguments, that number drops to a mere
+ \PERCFAILEDASSERTIONSKAGGLENOUNDEF.  In all,
+ \PROPFUNSWITHFAILEDASSERTKAGGLE functions had at least one contract
+ failure.  There were \NUMTOTALASSERTIONSKAGGLE assertions in total,
+ on \NUMASSERTEDFUNCTIONSKAGGLE functions.
 
@@
 
- To mirror our analysis of contract assertions on the reverse dependencies of our corpus, we show in Table~\ref{tbl:fail-breakdown-0-kaggle} the most frequently failing contract in Kaggle.
- While we don't see many overlapping entries per se, the assertions exhibit similar patterns.
- For instance, {\it data tables} (which have \CLASS{{\tt data.table, data.frame}}) are often passed to arguments expecting a \CLASS{{\tt matrix}}.
- Data tables are essentially serve the same purpose as data frames and tibbles.
- As it happens, data tables can be coerced to matrices if their elements are unityped, and programmers will often interchange the two, as is the case here.
- One function producing many of these errors is the following:
+ To mirror our analysis of contract assertions on the reverse
+ dependencies of our corpus, we show in
+ Table~\ref{tbl:fail-breakdown-0-kaggle} the most frequently failing
+ contract in Kaggle.  While we don't see many overlapping entries per
+ se, the assertions exhibit similar patterns.  For instance, {\it data
+   tables} (which have \CLASS{{\tt data.table, data.frame}}) are often
+ passed to arguments expecting a \CLASS{{\tt matrix}}.  Data tables
+ are essentially serve the same purpose as data frames and tibbles.
+ As it happens, data tables can be coerced to matrices if their
+ elements are unityped, and programmers will often interchange the
+ two, as is the case here.  One function producing many of these
+ errors is the following:
  \begin{lstlisting}
@@
 
- Another interesting entry in Table~\ref{tbl:fail-breakdown-0-kaggle} is the fifth row, where a \CLASS{{\tt ixforeach, iter}}  is being passed to an argument expecting a long union of classes, each of the form \CLASS{{\tt X, iter}}.
- This is likely an instance where the user defined their own class, {\tt ixforeach}, and wanted to use the \code{iterators} package (the user called \code{iterator::iter} with a \CLASS{{\tt ixforeach}}, and it gained the class {\tt iter} on return).
- As we mentioned, accounting for object-orientation in the type system is beyond the scope of this work, and such an inclusion would allow us to better type situations like this.
-
- Table~\ref{tbl:fail-breakdown-2-kaggle} mirrors Table~\ref{tbl:fail-breakdown-2} in showing contract violations on the most frequently occurring argument types.
- Here, our manual analysis has revealed similar failure patterns.
- In the case of vectors being passed to scalars, we find functions which can take vectors but were only tested with scalars (e.g., \code{stringr::str_to_upper} which converts a vector of characters to upper case, and \code{dplyr::anti_join}, which can join by a vector of column names but was only ever tested with a scalar).
- We also see possibly-NA character vectors being passed to NA-free character vectors.
- These assertion failures arise from a lack of testing: the offending functions are \code{str_trim}, \code{str_sub}, and \code{str_replace_all} from the \code{stringr} package.
- These functions are actually wrappers for other functions which have the correct argument type (\NAVEC\chr).
+ Another interesting entry in Table~\ref{tbl:fail-breakdown-0-kaggle}
+ is the fifth row, where a \CLASS{{\tt ixforeach, iter}} is being
+ passed to an argument expecting a long union of classes, each of the
+ form \CLASS{{\tt X, iter}}.  This is likely an instance where the
+ user defined their own class, {\tt ixforeach}, and wanted to use the
+ \code{iterators} package (the user called \code{iterator::iter} with
+ a \CLASS{{\tt ixforeach}}, and it gained the class {\tt iter} on
+ return).  As we mentioned, accounting for object-orientation in the
+ type system is beyond the scope of this work, and such an inclusion
+ would allow us to better type situations like this.
+
+ Table~\ref{tbl:fail-breakdown-2-kaggle} mirrors
+ Table~\ref{tbl:fail-breakdown-2} in showing contract violations on
+ the most frequently occurring argument types.  Here, our manual
+ analysis has revealed similar failure patterns.  In the case of
+ vectors being passed to scalars, we find functions which can take
+ vectors but were only tested with scalars (e.g.,
+ \code{stringr::str_to_upper} which converts a vector of characters to
+ upper case, and \code{dplyr::anti_join}, which can join by a vector
+ of column names but was only ever tested with a scalar).  We also see
+ possibly-NA character vectors being passed to NA-free character
+ vectors.  These assertion failures arise from a lack of testing: the
+ offending functions are \code{str_trim}, \code{str_sub}, and
+ \code{str_replace_all} from the \code{stringr} package.  These
+ functions are actually wrappers for other functions which have the
+ correct argument type (\NAVEC\chr).
 
@@
 
- Overall, the analysis discussed in this section has revealed two broad categories of contract assertion failures: those related to coercion, and those related to a class hierarchy.
- Our type system does not account for coercion as coercion in R is ad hoc at best, and it is implemented on a function-by-function basis, even in the core R packages.
- As for errors related to a class hierarchy, we aim to tackle this in future work, as designing a full fledged object-oriented type system for a language like R is outside of the scope of this work.
+ Overall, the analysis discussed in this section has revealed two
+ broad categories of contract assertion failures: those related to
+ coercion, and those related to a class hierarchy.  Our type system
+ does not account for coercion as coercion in R is ad hoc at best, and
+ it is implemented on a function-by-function basis, even in the core R
+ packages.  As for errors related to a class hierarchy, we aim to
+ tackle this in future work, as designing a full fledged
+ object-oriented type system for a language like R is outside of the
+ scope of this work.
 
@@
 
-There are a number of ways to check the types of function parameters in R.
- The default and most common way\footnote{\CranStopifnotRatio of all runtime
-  checks in the whole of CRAN} is to use the \code{stopifnot} function from the R
-base package. It takes a number of R expressions which all should evaluate to
-true otherwise a runtime exception is thrown with a message quoting the
-corresponding failed expression. For example, the following code checks whether
-a given parameter \code{x} is a scalar string:
+There are a number of ways to check the types of function parameters
+in R.  The default and most common way\footnote{\CranStopifnotRatio of
+all runtime checks in the whole of CRAN} is to use the
+\code{stopifnot} function from the R base package. It takes a number
+of R expressions which all should evaluate to true otherwise a runtime
+exception is thrown with a message quoting the corresponding failed
+expression. For example, the following code checks whether a given
+parameter \code{x} is a scalar string:
 %
@@
 
-Besides \code{stopifnot}, there are 4 packages in CRAN\footnote{Packages are
-  available on the CRAN website: \url{https://cran.r-project.org/web/packages/}}
-that focus on runtime assertions: \emph{assertive}, \emph{ensurer},
-\emph{assertr} and \emph{assertthat}. \emph{assertive} and \emph{ensurer} have
-not been updated since 2016 and 2015 respectively, and \emph{assertr} is used by only
-\AssertrRevdeps other packages and currently focuses on checking
-properties of data frames. Only the \emph{assertthat} package is maintained and used
-(with \AssertthatRevdeps reverse dependencies). The advantage of
-\emph{assertthat} over the R's default is that it provides much better error
-messages.
-
-One way to asses the usefulness of our type checking system is to find out how
-many of the existing type checking constrains could be replaced by \contractr.
-To measure this, we have extracted all calls to \code{stopifnot} and
-\code{assertthat} assertions and checked which among them could be either
-completely replaced by \contractr or at least partially simplified by removing a
-portion of an assertion expression. This is useful, because a common pattern is
-that the first part of the parameter assertion checks its type while the rest
-checks its value. In the example above, the whole expression could be replaced
-by a \chr type check.
-
-Out of the \CorpusLoadable packages, \CorpusAssertsInPackages use runtime
-assertions. Together there are \CorpusAsserts asserts in
-\CorpusAssertsInFunctions functions. Among these, \contractr can replace
-\CorpusTypedAsserts (\CorpusTypedAssertsRatio) assertion calls across
-\CorpusTypedAssertsPackages packages and \CorpusTypedAssertsFunctions functions.
-Furthermore, additional \CorpusPartiallyTypedAsserts
-(\CorpusPartiallyTypedAssertsRatio) asserts in packages
-\CorpusPartiallyTypedAssertsPackages packages and
-\CorpusPartiallyTypedAssertsFunctions functions could have been simplified.
-
-Checking the type of function parameters is not something that is seen often in
-the R code. In the whole of CRAN, there are only \CranAssertsRnd asserts in
-\CranAssertsInFunctionsRnd functions define in \CranAssertsInPackagesRnd
-packages.
-One may speculate that this is the case due to the verbosity and inconvenience of the existing assertion tools.
-Our system can infer type annotations
-for existing functions automatically. This can remove or partially remove over
+Besides \code{stopifnot}, there are 4 packages in
+CRAN\footnote{Packages are available on the CRAN website:
+\url{https://cran.r-project.org/web/packages/}} that focus on runtime
+assertions: \emph{assertive}, \emph{ensurer}, \emph{assertr} and
+\emph{assertthat}. \emph{assertive} and \emph{ensurer} have not been
+updated since 2016 and 2015 respectively, and \emph{assertr} is used
+by only \AssertrRevdeps other packages and currently focuses on
+checking properties of data frames. Only the \emph{assertthat} package
+is maintained and used (with \AssertthatRevdeps reverse
+dependencies). The advantage of \emph{assertthat} over the R's default
+is that it provides much better error messages.
+
+One way to asses the usefulness of our type checking system is to find
+out how many of the existing type checking constrains could be
+replaced by \contractr.  To measure this, we have extracted all calls
+to \code{stopifnot} and \code{assertthat} assertions and checked which
+among them could be either completely replaced by \contractr or at
+least partially simplified by removing a portion of an assertion
+expression. This is useful, because a common pattern is that the first
+part of the parameter assertion checks its type while the rest checks
+its value. In the example above, the whole expression could be
+replaced by a \chr type check.
+
+Out of the \CorpusLoadable packages, \CorpusAssertsInPackages use
+runtime assertions. Together there are \CorpusAsserts asserts in
+\CorpusAssertsInFunctions functions. Among these, \contractr can
+replace \CorpusTypedAsserts (\CorpusTypedAssertsRatio) assertion calls
+across \CorpusTypedAssertsPackages packages and
+\CorpusTypedAssertsFunctions functions.  Furthermore, additional
+\CorpusPartiallyTypedAsserts (\CorpusPartiallyTypedAssertsRatio)
+asserts in packages \CorpusPartiallyTypedAssertsPackages packages and
+\CorpusPartiallyTypedAssertsFunctions functions could have been
+simplified.
+
+Checking the type of function parameters is not something that is seen
+often in the R code. In the whole of CRAN, there are only
+\CranAssertsRnd asserts in \CranAssertsInFunctionsRnd functions define
+in \CranAssertsInPackagesRnd packages.  One may speculate that this is
+the case due to the verbosity and inconvenience of the existing
+assertion tools.  Our system can infer type annotations for existing
+functions automatically. This can remove or partially remove over
 \CorpusPartiallyTypedAssertsRatio of existing assertions.

### 2020-10-09 paper/main.tex
prelim discussion of different consolidation options, as itemize

@@
 %
+\section{Discussion}
+
+\AT{Let's discuss alternative strategies for consolidating types here.}
+
+\begin{itemize}
+	\item We employed the following strategy to consolidate types: We collected all of the traces observed for a function, and combined them into a single function type, where the type for each argument position was made up of a union of all of the observed types at that position.
+	\item We then simplified these unions using the rules \AT{discussed previously, ref section.}
+	\item We did not develop this strategy in a vacuum, and there are indeed alternatives worth mentioning.
+
+	\item For starters, one may rightly argue that our strategy results in a loss of precision, as relationships between input and output types are obscured.
+	\item For example, consider a function with two unique call traces, \FUN{\chr}{\chr} and \FUN{\dbl}{\dbl}.
+	\item The strategy we employed results in a combined type of \FUN{\chr \;|\; \dbl}{\chr \;|\; \dbl} for the function, which obscures the (potential) relationship between the argument and return type.
+
+	\item An obvious solution to this is simply to convert each call trace into a type, and generate a union of function types as the overall type for the function.
+	\item This would lead to the type  \FUN{\chr}{\chr} $|$ \FUN{\dbl}{\dbl} for the aforementioned function.
+	\item Unfortunately, this leads to a significant blow-up in the size of types, and makes many types unreadable, due to the myriad ways in which we combine the primitive types, as recall that, e.g., scalars and vectors are not the same type, and we see many types of the form \FUN{\chr}{\dbl} $|$ \FUN{\VEC\chr}{\dbl} $|$ \FUN{\NAVEC\chr}{\dbl}.
+	\item To quickly compare this consolidation strategy against the one employed throughout the paper, we find that the types \AT{put numbers, gist is that it's bigger and more incomprehensible.}
+
+	\item Another option is to employ a hybrid approach, wherein we perform a union of function types {\it after} grouping them together by return type, and performing basic primitive type consolidation on the return types.
+	\item This has the advantage of avoiding most of the blowup of the naive approach, however this hybrid  approach suffers particularly when functions return classes, as effectively consolidating class types is unclear given the limited level of detail in our type system.
+	\item Comparing again to the strategy employed throughout the paper, we see \AT{numbers, it's a bit bigger, not too much. Need numbers of unions, etc. Forthcoming.}
+
+	\item \AT{Will pull in example showcasing the different strategies.}
+\end{itemize}
+
+%
+%
+%
+%
 \section{Conclusion}

### 2020-10-12 paper/main.tex
intro

@@
 was designed to favor an interactive style of programming with minimal
-syntactic and conceptual overhead. This design is well suited to
-interactive data analysis, but a bad fit for tools such as compilers
-or program analyzers which must generate native code or catch
-programming errors.  In particular, R has no type annotations, and all
+syntactic and conceptual overhead. This design is well suited to data
+analysis, but a bad fit for tools such as compilers or program
+analyzers. In particular, R has no type annotations, and all
 operations are dynamically checked at run-time. The starting point for
@@
 \NUMPKGSEVAL clients of these packages, as well as on end-user code
-found on the Kaggle competition website.
+from the Kaggle competition.
 \end{abstract} \maketitle
@@
 
-Our community builds, improves, and reasons about programming languages. To
-make design decisions that benefit users, we need to understand our target
-language as well as the real-world needs it answers. Often, we can appeal to
-our intuition, as many languages are intended for general purpose
-programming tasks.  Unfortunately, intuition may fail when looking at
-domain-specific languages designed for a specific
-group of users to solve very specific needs. This is the case of the data
-science language R.
+Our community builds, improves, and reasons about programming
+languages. To make design decisions that benefit users, we need to
+understand our target language as well as the real-world needs it
+answers. Often, we can appeal to intuition, as many languages are
+intended for general purpose programming tasks.  Unfortunately,
+intuition may fail us when looking at domain-specific languages
+designed for a particular group of users to solve specific needs. This
+is the case of the data science language R.
 
 R and its ancestor S were designed, implemented, and maintained by
-statisticians. Originally they aimed to be glue languages for reading data
-and calling statistical routines written in Fortran. Over three decades they
-became widely used across many fields for data analysis and visualization.
+statisticians. Originally they aimed to be glue languages for 
+statistical routines written in Fortran. Over three decades they
+became widely used across many fields for exploratory data analysis.
 Modern R, as an object of study, is fascinating. It is a vectorized,
@@
 Many of the design decisions that gave us R were intended to foster an
-interactive and exploratory programming style. These include, to name a few,
-the lack of type annotations, the ability to use syntactic shortcuts, and the
-automatic conversion between data types. While these choices have led to a
-language with a low barrier to entry---many data science educational
-programs do not teach R itself but simply introduce some of its key
-libraries---they have also created a language where errors can easily go
+interactive and exploratory programming style. These include, to name
+a few salient ones, the lack of type annotations, the ability to use
+syntactic shortcuts, and widespread conversion between data
+types. While these choices have decreased the barrier to entry---many
+data science educational programs do not teach R itself but simply
+introduce some of its key libraries---they also allow for errors to go
 undetected.
@@
 Retrofitting a type system to the R language would increase our assurance in
-the result of data analysis, but we are faced with two challenges. First,
+the result of data analysis, but this requires facing two challenges. First,
 it is unclear what would be the \emph{right} type system for a language as

### 2020-10-12 paper/main.tex
related work

@@
 
-This paper is a data-driven study of what a type system for the R language
-could look like. Our intention is to eventually propose changes to the
-language, but we are aware that for any changes to be accepted by the user
-community they must clearly benefit the language without endangering
-backwards compatibility. Our goal is thus to find a compromise between
-simplicity and usefulness; any proposal should cover common programming
-idioms while remaining easy to learn and to use.
-
-This paper focuses on a simpler problem than an entire type system, instead,
-we limit our scope to giving types to function signatures. For this, we
-design a simple type language, one that matches the R data types but omits
-features such as parametric polymorphism and subtyping between user-defined
-data types. We then extract type signature from execution traces of a corpus
-of widely used libraries.  This allows us to see how far one can get with
-the simple type language and identify limitations of the design. We validate
-the robustness of the extracted type signature by implementing a contract
-checker that weaves the types around their respective functions, and use a
-both a large number of clients of the target packages as well as end-user code for validation. The contract
-system can be used by the R community to experiment with type signatures and
-replace ad hoc error checking code.
-
-To sum up, our paper makes the following contributions:
+The goal of this paper is to gather data that can be used as input to
+the process of designing a type system for R.  To focus our work, we
+chose to limit the scope of investigation on the design of a language
+for function type signatures.  Thus, this paper will use the collected
+data to document the signature of user-defined functions.  For this,
+we design a simple type language, one that matches the R data types.
+We then extract type signature from execution traces of a corpus of
+widely used libraries.  This allows us to see how far one can get with
+the simple type language and identify limitations of the design. We
+validate the robustness of the extracted type signature by
+implementing a contract checker that weaves the types around their
+respective functions, and use a both a large number of clients of the
+target packages as well as end-user code for validation.  To sum up,
+we make the following contributions:
 \begin{itemize}
-\item We implemented tooling to automatically extract type signatures from R
-  functions and to instrument R functions with checks based on their declared
-  types. The tooling is robust and scalable to the entire R language.
-\item We carried out a large-scale analysis of corpus of \PACKAGES widely
-  used and actively maintained libraries to extract function type signatures and
-  validated the robustness of the inferred type signatures against \NUMFILESRUN
-  programs that use those functions.
+\item We implemented scalable and robust tooling to automatically
+  extract type signatures and instrument R functions with checks
+  based on their declared types.
+\item We carried out a corpus analysis of \PACKAGES widely used and
+  maintained packages to extract function type signatures and
+  validated the robustness of the inferred type signatures against
+  \NUMFILESRUN programs that use those functions.
 \item We report on the appropriateness and usefulness of a simple type
-  language for the R programming language.
+  language for R.
+\item Our data and code are open source and publicly
+  available.\footnote{ \url{github.com/PRL-PRG/propagatr},
+  \url{github.com/PRL-PRG/contractr}, {\bf TODO add DATA link}}
 \end{itemize}
 
-Our tools are open source and publicly available on GitHub\footnote{
-\typetracer: \url{https://github.com/PRL-PRG/propagatr}, \contractr:
-\url{https://github.com/PRL-PRG/contractr} }.
 %
@@
 
-In this section, we introduce related work on extending dynamic languages
-with static type systems and we give a short primer on the R programming
-language.
+This section introduces related work and gives a short primer on R.
 
@@
 
-Dynamic programming languages such as Racket, JavaScript and PHP have
-been extended post-facto with various static type systems.  In each
+Dynamic programming languages such as Racket, JavaScript, PHP and Lua
+have been extended post factum with static type systems.  In each
 case, the type system was carefully engineered to match the salient
-characteristics of the host languages and to foster a particular
+characteristics of its host language and to foster a particular
 programming style. For example, Typed Racket emphasizes functional
-programming and supports the migration from untyped to fully typed
+programming and migration from untyped to fully typed
 code~\cite{tf-popl08}, Hack~\cite{hack13} and TypeScript~\cite{BAT14}
-focus on typing object-oriented features of PHP and JavaScript,
+focus on object-oriented features of PHP and JavaScript,
 respectively. They allow users to intersperse typed and untyped code
-in a fine-grained manner.  Other work looks at retrofitting types onto
-Lua~\cite{typedlua2014} which account for the myriad ways Lua
-programmers use tables.
+in a fine-grained manner.  In the case of Lua~\cite{typedlua2014}, the
+type system tried to account for the myriad ways Lua programmers use
+tables.  Other languages adopted a mix of typed and untyped code by
+design. In Julia, type annotations are needed for method dispatch and
+performance~\cite{oopsla18a}. In Thorn, users could freely move
+between typed and untyped code thanks to the addition of like
+types~\cite{popl10}. Lastly C\# is an example of a statically typed
+language which added a dynamic type~\cite{Bierman10}.
 
@@
 for type hints in Python 3.5~\cite{pythonspectypes}.  There is no
-previous work on types for R, besides a short sketch of issues the
-type system would need to tackle~\cite{turcotte2019towards}. We take
-inspiration in the aforementioned works but focus on adapting them to
-our target language.
+previous work on types for R. We take inspiration in the
+aforementioned works but focus on adapting them to our target
+language.
 
@@
 The R Project is a key tool for data analysis.  At the heart of the R
-project is a \emph{vectorized, dynamic, lazy, functional, object-oriented}
-programming language with a rather unusual combination of
-features~\cite{ecoop12} designed to ease learning by non-programmers and
-enable rapid development of new statistical methods.  The language was
+project is a vectorized, dynamic, lazy, functional, object-oriented
+language with a rather unusual combination of features~\cite{ecoop12}
+designed to ease learning by non-programmers and enable rapid
+development of new statistical methods.  The language was
 designed~\citet{R96} as a successor to S~\cite{S88}.

### 2020-10-12 paper/main.tex
background

@@
 
-\subsection{Related Work}
+\subsection{Related Work} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 But what if the design of the type system is unclear?  \citet{tip} propose
-an intriguing approach called trace typing. With trace typing, a new type
+a promising approach called trace typing. With trace typing, a new type
 system can be prototyped and evaluated by applying the type rule to
@@
 
-\subsection{The R Programming Language} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{The R Language} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 project is a vectorized, dynamic, lazy, functional, object-oriented
-language with a rather unusual combination of features~\cite{ecoop12}
-designed to ease learning by non-programmers and enable rapid
-development of new statistical methods.  The language was
-designed~\citet{R96} as a successor to S~\cite{S88}.
+language with an unusual combination of features~\cite{ecoop12}.  R
+was designed~\citet{R96} as a successor to S~\cite{S88}.
 
@@
 \begin{lstlisting}
-f <- function(x, ..., y=if(z==0) 1, z=0) { 
-  x + y + if (missing(...)) 0 else c(...)
-}
+f <- function(x,..., y=if(z==0) 1, z=0) { x + y + if(missing(...)) 0 else c(...) }
 \end{lstlisting}

### 2020-10-12 paper/main.tex
stuff

@@
 
-The basic building blocks of R are vectors of primitives (primitives
-in R are characters (strings), integers, logicals (booleans), doubles,
-complex numbers, and raw bytes).  When we say that R is {\it
-  vectorized}, we mean that primitive values are {\it always} vectors.
-For example, in R one might see the literal \code{2.5} and think it to
-be a scalar double, but it is in actuality a unit-length vector
-containing a single double.  Vectors can also be explicitly
-constructed using the vector constructor \code{c(...)}, For example,
-\code{c(1L, 2L, 3L)} creates a vector of three integers (in R, integer
-literals are denoted by an \code{L}).  To illustrate that scalars are
-vectors, in R \code{(c(2.5) == 2.5) == TRUE}.
-
-The main object of our work is R functions, and functions in R have a
-number of quirks.  Arguments can be assigned default values (which can
-be arbitrary expressions!), and functions can support variable
+It's basic building blocks are vectors of primitives (primitives in R
+are characters (or strings), integers, logicals (or booleans),
+doubles, complex numbers, and raw bytes).  To say that R is vectorized
+implies that primitive values are {\it always} vectors.  For example,
+the literal \code{2.5} is a unit-length vector containing a single
+double.  Vectors can be explicitly constructed using the constructor
+\code{c(...)}, as in \code{c(1L, 2L, 3L)} which creates a vector of
+three integers.
+
+The main object of our work is functions, and R functions have a
+number of quirks.  Arguments can be assigned arbitrary expressions as
+default values, and functions can support variable
 argument lists, and functions can be called by specifying argument

### 2020-10-13 paper/main.tex
paragraph mode

@@
 
-\begin{itemize}
-	\item We employed the following strategy to consolidate types: We collected all of the traces observed for a function, and combined them into a single function type, where the type for each argument position was made up of a union of all of the observed types at that position.
-	\item We then simplified these unions using the rules \AT{discussed previously, ref section.}
-	\item We did not develop this strategy in a vacuum, and there are indeed alternatives worth mentioning.
-
-	\item For starters, one may rightly argue that our strategy results in a loss of precision, as relationships between input and output types are obscured.
-	\item For example, consider a function with two unique call traces, \FUN{\chr}{\chr} and \FUN{\dbl}{\dbl}.
-	\item The strategy we employed results in a combined type of \FUN{\chr \;|\; \dbl}{\chr \;|\; \dbl} for the function, which obscures the (potential) relationship between the argument and return type.
-
-	\item An obvious solution to this is simply to convert each call trace into a type, and generate a union of function types as the overall type for the function.
-	\item This would lead to the type  \FUN{\chr}{\chr} $|$ \FUN{\dbl}{\dbl} for the aforementioned function.
-	\item Unfortunately, this leads to a significant blow-up in the size of types, and makes many types unreadable, due to the myriad ways in which we combine the primitive types, as recall that, e.g., scalars and vectors are not the same type, and we see many types of the form \FUN{\chr}{\dbl} $|$ \FUN{\VEC\chr}{\dbl} $|$ \FUN{\NAVEC\chr}{\dbl}.
-	\item To quickly compare this consolidation strategy against the one employed throughout the paper, we find that the types \AT{put numbers, gist is that it's bigger and more incomprehensible.}
-
-	\item Another option is to employ a hybrid approach, wherein we perform a union of function types {\it after} grouping them together by return type, and performing basic primitive type consolidation on the return types.
-	\item This has the advantage of avoiding most of the blowup of the naive approach, however this hybrid  approach suffers particularly when functions return classes, as effectively consolidating class types is unclear given the limited level of detail in our type system.
-	\item Comparing again to the strategy employed throughout the paper, we see \AT{numbers, it's a bit bigger, not too much. Need numbers of unions, etc. Forthcoming.}
-
-	\item \AT{Will pull in example showcasing the different strategies.}
-\end{itemize}
+Throughout this paper, we employed the following strategy to consolidate types: We collected all of the traces observed for a function, and combined them into a single function type, where the type for each argument position was made up of a union of all of the observed types at that position.
+We then simplified these unions using the rules presented in Section~\ref{fig:siplification-rules}.
+We call this an {\it arrow of unions}.
+Now, this strategy was not developed in a vacuum, and we will now discuss some of the decisions that we made in choosing this strategy.
+
+Arguably, the primary issue with the arrow of unions strategy is that of a loss of precision, as relationships between argument and return types are obscured when all argument and return types are unioned.
+For example, consider a function with two unique call traces, \FUN{\chr}{\chr} and \FUN{\dbl}{\dbl}.
+The strategy we employed results in a combined type of \FUN{\chr \;|\; \dbl}{\chr \;|\; \dbl} for the function, which obscures the (potential) relationship between the argument and return type.
+
+An obvious solution to this is simply to convert each call trace into a type, and generate a union of function types as the overall type for the function.
+This would lead to the type  \FUN{\chr}{\chr} $|$ \FUN{\dbl}{\dbl} for the aforementioned function.
+Unfortunately, this leads to a significant blow-up in the size of types, and makes many types unreadable, due to the myriad ways in which we combine the primitive types, as recall that, e.g., scalars and vectors are not the same type, and we see many types of the form \FUN{\chr}{\dbl} $|$ \FUN{\VEC\chr}{\dbl} $|$ \FUN{\NAVEC\chr}{\dbl}.
+When comparing this consolidation strategy against the one employed throughout the paper, we find that the types are much larger \AT{character count}, and we count \AT{number} top-level unions.
+Recall that we found \AT{number} functions to be polymorphic when collapsing these unions of arrows into an arrow of unions; with unions of arrows, we find \AT{number} functions (\AT{percentage}) to be polymorphic.
+
+Another option is to employ a hybrid approach, wherein we perform a union of function types {\it after} grouping them together by return type, and performing basic primitive type consolidation on the return types.
+This has the advantage of avoiding most of the blowup of the naive approach, however this hybrid approach suffers particularly when functions return classes, as our type system does not allow us to effectively consolidate class types.
+Comparing again to the strategy employed throughout the paper, we find \AT{number} (\AT{\%}) functions to have a union of arrows, and \AT{number} (\AT{percentage}) of arguments are polymorphic.
+We additionally find that \AT{number} (\AT{\%}) of arrow types have at least one polymorphic argument.
+
+To connect this discussion with actual R types, we chose the \AT{which?} function from the \AT{which?} package, and show what the type of the function would be when using each of these strategies.
+We refer the reader to Figure~\ref{tbl:real-function-different-merge-strats}.
+\AT{We could use stringr::str\_extract\_all?}
+
+\begin{table}[H]
+	\centering
+	 \resizebox{.98\linewidth}{!}{
+		\begin{tabular}{llrrr}
+			\toprule
+			Merge Strategy & Function Type \\
+			\midrule
+			Union of Arrows & \\
+			Hybrid &  \\
+			Arrow of Unions & \\
+			\bottomrule
+	\end{tabular}}
+	\caption{Types of \AT{package::function} with different type merge strategies. }\label{tbl:real-function-different-merge-strats}
+\end{table}
 

### 2020-10-14 paper/main.tex
discussion fnl draft, will add graphs maybe and also function example

@@
 Unfortunately, this leads to a significant blow-up in the size of types, and makes many types unreadable, due to the myriad ways in which we combine the primitive types, as recall that, e.g., scalars and vectors are not the same type, and we see many types of the form \FUN{\chr}{\dbl} $|$ \FUN{\VEC\chr}{\dbl} $|$ \FUN{\NAVEC\chr}{\dbl}.
-When comparing this consolidation strategy against the one employed throughout the paper, we find that the types are much larger \AT{character count}, and we count \AT{number} top-level unions.
-Recall that we found \AT{number} functions to be polymorphic when collapsing these unions of arrows into an arrow of unions; with unions of arrows, we find \AT{number} functions (\AT{percentage}) to be polymorphic.
+When comparing this consolidation strategy against the one employed throughout the paper, we find that the types using this strategy are much larger, with 14,057 (55.06\%) functions have at least one top-level union.
+%, and that 16,745 (65.59\%) functions have two or fewer top-level alternatives.
+We observed 93,229 unique, independent signatures using this strategy.
 
@@
 This has the advantage of avoiding most of the blowup of the naive approach, however this hybrid approach suffers particularly when functions return classes, as our type system does not allow us to effectively consolidate class types.
-Comparing again to the strategy employed throughout the paper, we find \AT{number} (\AT{\%}) functions to have a union of arrows, and \AT{number} (\AT{percentage}) of arguments are polymorphic.
-We additionally find that \AT{number} (\AT{\%}) of arrow types have at least one polymorphic argument.
+Comparing again to the strategy employed throughout the paper, we find 5,317 (20.82\%) functions to have a union of arrows, and 38,650 independent signatures.
+Most functions only have one or two top-level alternatives (23,530, or 92.17\%).
+We additionally find that 38,650 (26.14\%) of arrow types have at least one polymorphic argument.
 

### 2020-10-14 paper/main.tex
sec 2

--- 2020-10-13 paper/main.tex
@@
 we design a simple type language, one that matches the R data types.
-We then extract type signature from execution traces of a corpus of
-widely used libraries.  This allows us to see how far one can get with
-the simple type language and identify limitations of the design. We
-validate the robustness of the extracted type signature by
-implementing a contract checker that weaves the types around their
-respective functions, and use a both a large number of clients of the
-target packages as well as end-user code for validation.  To sum up,
-we make the following contributions:
+We then extract call and return types from execution traces of a
+corpus of widely used libraries, and finally synthesize type
+signatures.  This allows us to see how far one can get with a simple
+type language and identify limitations of our design. We validate the
+robustness of the extracted type signatures by implementing a contract
+checker that weaves types around their respective functions, and use
+clients of the target packages for validation.  To sum up, we make the
+following contributions:
 \begin{itemize}
@@
 \item We carried out a corpus analysis of \PACKAGES widely used and
-  maintained packages to extract function type signatures and
-  validated the robustness of the inferred type signatures against
-  \NUMFILESRUN programs that use those functions.
+  maintained packages to synthesize function type signatures and
+  validated the robustness of the signatures against \NUMFILESRUN
+  programs that use those functions.
 \item We report on the appropriateness and usefulness of a simple type
@@
 focus on object-oriented features of PHP and JavaScript,
-respectively. They allow users to intersperse typed and untyped code
-in a fine-grained manner.  In the case of Lua~\cite{typedlua2014}, the
-type system tried to account for the myriad ways Lua programmers use
+respectively. They allow users to mix typed and untyped code in a
+fine-grained manner.  In the case of Lua~\cite{typedlua2014}, the type
+system tried to account for the myriad ways Lua programmers use
 tables.  Other languages adopted a mix of typed and untyped code by
@@
 
-The R Project is a key tool for data analysis.  At the heart of the R
-project is a vectorized, dynamic, lazy, functional, object-oriented
-language with an unusual combination of features~\cite{ecoop12}.  R
-was designed~\citet{R96} as a successor to S~\cite{S88}.
-
-It's basic building blocks are vectors of primitives (primitives in R
-are characters (or strings), integers, logicals (or booleans),
-doubles, complex numbers, and raw bytes).  To say that R is vectorized
-implies that primitive values are {\it always} vectors.  For example,
-the literal \code{2.5} is a unit-length vector containing a single
-double.  Vectors can be explicitly constructed using the constructor
-\code{c(...)}, as in \code{c(1L, 2L, 3L)} which creates a vector of
-three integers.
-
-The main object of our work is functions, and R functions have a
-number of quirks.  Arguments can be assigned arbitrary expressions as
-default values, and functions can support variable
-argument lists, and functions can be called by specifying argument
-names explicitly in the call.  To illustrate all of this in a single
-example, consider:
-
-\begin{lstlisting}
-f <- function(x,..., y=if(z==0) 1, z=0) { x + y + if(missing(...)) 0 else c(...) }
-\end{lstlisting}
-
-\noindent
-This function has four formal parameters, \k x, \k{...}, \k y and \k
-z. Argument \k x can be bound positionally or passed by name. The
-vararg argument, \k{...}, is always passed positionally. The remaining
-two arguments must be passed by name as they are preceded by the
-\k{...} parameter.  Arguments \k y and \k z have default values, in
-the case of \k z this is a constant, but \k{y}'s default value is an
-expression that depends on the value of \k z (if \k z is not zero, \k
-y will default to \k{NULL}). The body of the function will add
-parameters \k x and \k y with either the scalar 0 or the result of
-concatenating the varargs into a primitive vector. The function
-\k{missing} tests if an argument was explicitly passed. The following
-are some valid invocations of \k f:
-
-\begin{lstlisting}
- > f(1)    
-   [1] 2             # a double vector, y is 0, ... is missing
- > f(1, 2) 
-   [1] 4             # a double vector, y is 0, ... is 2
- > f(1, 2, 3)
-   [1] 4 5           # a double vector, y is 0, ... is 2, 3
- > f(2, 3, x=1)
-   [1] 4 5           # a double vector, y is 0, ... is 2, 3
- > f(x=1, y=1)
-   [1] 2             # a double vector, y is 1, ... is missing
- > f(x=1, z=1)
-   numeric(0)        # a double vector of length 0, y is NULL
- > f(1L, 2L, y=1L)
-   [1] 4             # an integer vector, y is integer 1, ... is integer 2
- > f(1, y=c(1,2))
-   [1] 1 2           # a double vetor, y is 1, 2, ... is missing
-\end{lstlisting}
-
-\noindent
-The above hints at the polymorphism of the language: \k f may return a
-vector of integers or of doubles, the length of the vector depends on
-the length of the varargs, and of \k x and \k y.  As we mentioned, the
-language does not really differentiate between scalar and vectors.
-Some more exotic types that can be encountered include vectors of
-complex number and list of arbitrary types. Function \k f can be
-invoked with those as well.
-
-\begin{lstlisting}
- > c1 <- complex(re=1, im=2)
- > c2 <- complex(re=2, im=1)
- > f(c1)
-   [1] 1             # a double vector, x is complex, y is double
- > f(c1, y=c2)
-   [1] 2+1i          # a complex vector, x and y are complex
- > l1 <- list(1, 2)
- > l2 <- list(c1, c2)
- > f(l1) 
-   [1] 1             # a double vector, x is a list of doubles
- > f(l1, y=l2)
-   [[1]]             # a list of complex, x is a list of doubles
-   [1] 1+2i          #                      y is a list of complex
-
-   [[2]]
-   [1] 2+1i
-\end{lstlisting}
-
-R has a builtin notion of type that can be queried by the \k{typeof}
+The R Project is a key tool for data analysis.  At the heart of R is a
+vectorized, dynamic, lazy, functional, object-oriented language with
+an unusual combination of features~\cite{ecoop12}.  R was designed by
+\citet{R96} as a successor to S~\cite{S88}.
+
+R's main data type is the primitive vector.  Vectors care explicitly
+constructed by the constructor \code{c(...)}, as in \code{c(1L, 2L,
+  3L)} which creates a vector of three integers.  R has a builtin
+notion of type that can be queried by the \k{typeof}
 function. Figure~\ref{rtypes} lists all of the builtin types that are
 provided by the language; these are the possible return values of
-\k{typeof}. There is no intrinsic notion of subtyping in R, but in many
-contexts a \k{logical} will be coerced to \k{integer}, and an \k{integer} will
-be coerced to \k{double}.  Some odd conversions can occur in corner cases, such
-as \k{1<"2"} holds and \k{c(1,2)[1.6]} returns the first element of the
-vector, as the double is converted to an integer. R does not distinguish
-between scalars and vectors (they are all vectors), so \code{typeof(5) ==}
-\code{typeof(c(5)) == typeof(c(5,5))} \code{ == "double"}.
-
-All vectorized data types have a distinguished missing value denoted
-by \code{NA} (for ``not available'').  The default type of \code{NA}
-is \k{logical}.  We can see that \code{typeof(NA)=="logical"}, but
-\k{NA} inhabits every type: \code{typeof(c(1,NA)[2])} \k{==}
-\k{"double"}.  In addition to \code{NA}, R has a \code{NULL} value,
-and to understand why R has both a \code{NULL} value and \code{NA}
-values it is useful to remember that R is primarily designed for
-statistics and data science.  In experimental data science, it is
-useful to have a notion of a ``missing observation'', which is often
-NaN (not a number, or similar) in, e.g., MATLAB and Python.  Since
-vectors are unityped in R, it {\it must} have a ``NaN'' value for {\it
-  each} primitive type.  There single \code{NULL} null-typed value in
-the language to encompass all other cases.  In a sense, you can think
-of \code{NA} as representing a missing data point, and \code{NULL} as
-a missing data {\it set}.
+\k{typeof}. There is no intrinsic notion of subtyping in R, but in
+many contexts a \k{logical} will be coerced to \k{integer}, and an
+\k{integer} will be coerced to \k{double}.  Some odd conversions can
+occur in corner cases, such as \k{1<"2"} holds and \k{c(1,2)[1.6]}
+returns the first element of the vector, as the double is converted to
+an integer. R does not distinguish between scalars and vectors (they
+are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
+  typeof(c(5,5))} \code{ == "double"}.  With the exception of lists,
+all vectorized data types are monomorphic.  A \k{list} can hold values
+of any other type including \k{list}. For all monomorphic data types,
+attempting to store a value of a different type will cause a
+conversion. Either the value is converted to the type of the vector,
+or the vector is converted to the type of the value.
 
@@
 
-With one exception, all vectorized data types are monomorphic.  The
-exception is the \k{list} type which can hold values of any other type
-including \k{list}. For all monomorphic data types, attempting to
-store a value of a different type will cause a conversion. Either the
-value is converted to the type of the vector, or the vector is
-converted to the type of the value.
-
-Over the years, programmers have found the need for a richer type structure
-and have added {\it attributes}. The best way to think of attributes is as
-an optional map from names to values that can be attached to any object.
-Attributes are used to encode various type structures. They can be queried
-with functions such as \k{attributes} and \k{class}.  The addition of
-attributes lets programmers extend the set of types by tagging data with
-user-defined attributes. For example, one could define a vector of four
-values, \code{x<-c(1,2,3,4)} and then attach the attribute \k{dim} with a
-pair of numbers as value: \code{attr(x,"dim")<-c(2,2)}.  From that point,
-arithmetic functions will treat \k{x} as a 2x2 matrix.
-
-Another attribute that can be set is the \k{class}.  This attribute
-can be bound to a list of class names. For instance,
-\code{class(x)<-"human"} will set the class of \k{x} to be \k{human}.
-The main usage of attributes is for object-oriented programming. There
-are three object-orientation frameworks in R: S3, S4, and R5.  The S3
-object system support single dispatch on the class of the first
-argument of a function, whereas the S4 object system allows multiple
-dispatch (on all arguments). R5 allows for users to define objects in
-a more imperative style. Some of the most widely used data types
-leverage attributes, e.g., data frames and matrices. A data frame, for
-instance, is a list of vectors with a class and a column name
-attribute, and matrices are vectors with a \k{dims} attribute.
-
-Scalar data types include the distinguished \k{NULL} value, instances
-of classes written using the S4 object system, closures, and
-environments.  The implementation of R has a number of other types
-listed in Figure~\ref{rtypes} for reference.
+\noindent
+All vectorized data types have a distinguished missing value denoted
+by \code{NA} (for ``not available'').  The type of \code{NA} is
+\k{logical} (\code{typeof(NA)=="logical"}), but \k{NA} inhabits every
+type: \code{typeof(c(1,NA)[2])} \k{==} \k{"double"}.  R also has a
+\code{NULL} value.  In data science, it is useful to have a notion of
+a ``missing observation'', since vectors are monomorphic there is a
+need for a missing value to inhabit each primitive type. In a sense,
+\code{NA} represents a missing data point, and \code{NULL} is a
+missing data set.
+
+Over time, programmers have found the need for a richer type
+structure.  R supports this with {\it attributes}. One may think of
+attributes as an optional map from names to values attached to any
+built in type.  Attributes are used to encode various type
+structures. They are queried with \k{attributes} and \k{class}.  Using
+attributes, programmers can extend the set of types by tagging
+data. For example, take the vector of four values,
+\code{x<-c(1,2,3,4)} and attach attribute \k{dim},
+\code{attr(x,"dim")<-c(2,2)}, to treat \k{x} as a 2x2 matrix.
+
+Another attribute that can be set is the \k{class}, this is one way to
+use R for object-oriented programming.  This attribute can be bound to
+a list of class names. For instance, \code{class(x)<-"human"} will set
+the class of \k{x} to be \k{human}.  There are three
+object-orientation frameworks in R: S3, S4, and R5.  The S3 object
+system support single dispatch on the class of the first argument of a
+function, whereas the S4 object system allows multiple dispatch. R5
+allows for users to define objects in a more imperative style. Some of
+the most widely used data types leverage attributes, e.g., data frames
+and matrices. A data frame, for instance, is a list of vectors with a
+class and a column name attribute, and matrices are vectors with a
+ \k{dims} attribute.
+
+R functions have a number of quirks.  Arguments can be assigned
+arbitrary expressions as default values; functions take variable
+numbers of parameters, and can be called positionally and nominally.
+Consider this example:
+
+\begin{lstlisting}
+f <- function(x,..., y=if(z==0) 1, z=0) { x + y + if(missing(...)) 0 else c(...) }
+\end{lstlisting}
+
+\noindent
+This function has four formal parameters, \k x, dots, \k y and \k
+z. Parameter \k x can be bound positionally or passed by name. The
+dots, \k{...}, is always positional. The remaining two parameters must
+be passed by name as they are preceded by dots; \k y and \k z have
+default values, in the case of \k z this is a constant, but \k{y} is
+bound to an expression that depends on \k z's value (if \k z is not
+zero, \k y defaults to \k{NULL}). The body of the function will add \k
+x and \k y tp either 0 or the result of concatenating the dots into a
+primitive vector. The function \k{missing} tests if a parameter was
+explicitly passed. The following are some valid invocations of \k f:
+
+\begin{lstlisting}
+ > f(1)    
+   [1] 2             # a double vector, y is 0, ... is missing
+ > f(2, 3, x=1)
+   [1] 4 5           # a double vector, y is 0, ... is 2, 3
+ > f(x=1, y=1)
+   [1] 2             # a double vector, y is 1, ... is missing
+ > f(x=1, z=1)
+   numeric(0)        # a double vector of length 0, y is NULL
+ > f(1L, 2L, y=1L)
+   [1] 4             # an integer vector, y is integer 1, ... is integer 2
+ > f(1, y=c(1,2))
+   [1] 1 2           # a double vetor, y is 1, 2, ... is missing
+\end{lstlisting}
+
+\noindent
+The above hints at polymorphism: \k f may return a vector of integers
+or of doubles of length equal to the max length of its arguments.
 

### 2020-10-14 paper/main.tex
minor

@@
 language to describe the arguments and return values of R functions.
-The goal is not to propose a final design, but rather to propose a
-starting point for an iterative design process.  Fig.~\ref{types}
-presents our type language. An early design choice was to stay close
-to the R language and only depart in small and hopefully
-non-controversial ways.  Functions types have the form \FUN{A_1,
-  \dots, A_n}\T where each $\A_i$ argument is either a type \T or
-\VARGS, a variable length argument list. The rest of this section
+The goal is not to propose a final design, but rather a
+starting point for an iterative design process.
+
+Fig.~\ref{types} presents our type language. An early design choice
+was to stay close to the R language and only depart in small and
+hopefully non-controversial ways.  Functions types have the form
+\FUN{A_1, \dots, A_n}\T where each $\A_i$ argument is either a type \T
+or \VARGS, a variable length argument list. The rest of this section
 details and motivates our design.

### 2020-10-14 paper/main.tex
plots

--- 2020-10-14 paper/main.tex
@@
 we design a simple type language, one that matches the R data types.
-We then extract call and return types from execution traces of a
-corpus of widely used libraries, and finally synthesize type
-signatures.  This allows us to see how far one can get with a simple
-type language and identify limitations of our design. We validate the
-robustness of the extracted type signatures by implementing a contract
-checker that weaves types around their respective functions, and use
-clients of the target packages for validation.  To sum up, we make the
-following contributions:
+We then extract type signature from execution traces of a corpus of
+widely used libraries.  This allows us to see how far one can get with
+the simple type language and identify limitations of the design. We
+validate the robustness of the extracted type signature by
+implementing a contract checker that weaves the types around their
+respective functions, and use a both a large number of clients of the
+target packages as well as end-user code for validation.  To sum up,
+we make the following contributions:
 \begin{itemize}
@@
 \item We carried out a corpus analysis of \PACKAGES widely used and
-  maintained packages to synthesize function type signatures and
-  validated the robustness of the signatures against \NUMFILESRUN
-  programs that use those functions.
+  maintained packages to extract function type signatures and
+  validated the robustness of the inferred type signatures against
+  \NUMFILESRUN programs that use those functions.
 \item We report on the appropriateness and usefulness of a simple type
@@
 focus on object-oriented features of PHP and JavaScript,
-respectively. They allow users to mix typed and untyped code in a
-fine-grained manner.  In the case of Lua~\cite{typedlua2014}, the type
-system tried to account for the myriad ways Lua programmers use
+respectively. They allow users to intersperse typed and untyped code
+in a fine-grained manner.  In the case of Lua~\cite{typedlua2014}, the
+type system tried to account for the myriad ways Lua programmers use
 tables.  Other languages adopted a mix of typed and untyped code by
@@
 
-The R Project is a key tool for data analysis.  At the heart of R is a
-vectorized, dynamic, lazy, functional, object-oriented language with
-an unusual combination of features~\cite{ecoop12}.  R was designed by
-\citet{R96} as a successor to S~\cite{S88}.
-
-R's main data type is the primitive vector.  Vectors care explicitly
-constructed by the constructor \code{c(...)}, as in \code{c(1L, 2L,
-  3L)} which creates a vector of three integers.  R has a builtin
-notion of type that can be queried by the \k{typeof}
+The R Project is a key tool for data analysis.  At the heart of the R
+project is a vectorized, dynamic, lazy, functional, object-oriented
+language with an unusual combination of features~\cite{ecoop12}.  R
+was designed~\citet{R96} as a successor to S~\cite{S88}.
+
+It's basic building blocks are vectors of primitives (primitives in R
+are characters (or strings), integers, logicals (or booleans),
+doubles, complex numbers, and raw bytes).  To say that R is vectorized
+implies that primitive values are {\it always} vectors.  For example,
+the literal \code{2.5} is a unit-length vector containing a single
+double.  Vectors can be explicitly constructed using the constructor
+\code{c(...)}, as in \code{c(1L, 2L, 3L)} which creates a vector of
+three integers.
+
+The main object of our work is functions, and R functions have a
+number of quirks.  Arguments can be assigned arbitrary expressions as
+default values, and functions can support variable
+argument lists, and functions can be called by specifying argument
+names explicitly in the call.  To illustrate all of this in a single
+example, consider:
+
+\begin{lstlisting}
+f <- function(x,..., y=if(z==0) 1, z=0) { x + y + if(missing(...)) 0 else c(...) }
+\end{lstlisting}
+
+\noindent
+This function has four formal parameters, \k x, \k{...}, \k y and \k
+z. Argument \k x can be bound positionally or passed by name. The
+vararg argument, \k{...}, is always passed positionally. The remaining
+two arguments must be passed by name as they are preceded by the
+\k{...} parameter.  Arguments \k y and \k z have default values, in
+the case of \k z this is a constant, but \k{y}'s default value is an
+expression that depends on the value of \k z (if \k z is not zero, \k
+y will default to \k{NULL}). The body of the function will add
+parameters \k x and \k y with either the scalar 0 or the result of
+concatenating the varargs into a primitive vector. The function
+\k{missing} tests if an argument was explicitly passed. The following
+are some valid invocations of \k f:
+
+\begin{lstlisting}
+ > f(1)    
+   [1] 2             # a double vector, y is 0, ... is missing
+ > f(1, 2) 
+   [1] 4             # a double vector, y is 0, ... is 2
+ > f(1, 2, 3)
+   [1] 4 5           # a double vector, y is 0, ... is 2, 3
+ > f(2, 3, x=1)
+   [1] 4 5           # a double vector, y is 0, ... is 2, 3
+ > f(x=1, y=1)
+   [1] 2             # a double vector, y is 1, ... is missing
+ > f(x=1, z=1)
+   numeric(0)        # a double vector of length 0, y is NULL
+ > f(1L, 2L, y=1L)
+   [1] 4             # an integer vector, y is integer 1, ... is integer 2
+ > f(1, y=c(1,2))
+   [1] 1 2           # a double vetor, y is 1, 2, ... is missing
+\end{lstlisting}
+
+\noindent
+The above hints at the polymorphism of the language: \k f may return a
+vector of integers or of doubles, the length of the vector depends on
+the length of the varargs, and of \k x and \k y.  As we mentioned, the
+language does not really differentiate between scalar and vectors.
+Some more exotic types that can be encountered include vectors of
+complex number and list of arbitrary types. Function \k f can be
+invoked with those as well.
+
+\begin{lstlisting}
+ > c1 <- complex(re=1, im=2)
+ > c2 <- complex(re=2, im=1)
+ > f(c1)
+   [1] 1             # a double vector, x is complex, y is double
+ > f(c1, y=c2)
+   [1] 2+1i          # a complex vector, x and y are complex
+ > l1 <- list(1, 2)
+ > l2 <- list(c1, c2)
+ > f(l1) 
+   [1] 1             # a double vector, x is a list of doubles
+ > f(l1, y=l2)
+   [[1]]             # a list of complex, x is a list of doubles
+   [1] 1+2i          #                      y is a list of complex
+
+   [[2]]
+   [1] 2+1i
+\end{lstlisting}
+
+R has a builtin notion of type that can be queried by the \k{typeof}
 function. Figure~\ref{rtypes} lists all of the builtin types that are
 provided by the language; these are the possible return values of
-\k{typeof}. There is no intrinsic notion of subtyping in R, but in
-many contexts a \k{logical} will be coerced to \k{integer}, and an
-\k{integer} will be coerced to \k{double}.  Some odd conversions can
-occur in corner cases, such as \k{1<"2"} holds and \k{c(1,2)[1.6]}
-returns the first element of the vector, as the double is converted to
-an integer. R does not distinguish between scalars and vectors (they
-are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
-  typeof(c(5,5))} \code{ == "double"}.  With the exception of lists,
-all vectorized data types are monomorphic.  A \k{list} can hold values
-of any other type including \k{list}. For all monomorphic data types,
-attempting to store a value of a different type will cause a
-conversion. Either the value is converted to the type of the vector,
-or the vector is converted to the type of the value.
+\k{typeof}. There is no intrinsic notion of subtyping in R, but in many
+contexts a \k{logical} will be coerced to \k{integer}, and an \k{integer} will
+be coerced to \k{double}.  Some odd conversions can occur in corner cases, such
+as \k{1<"2"} holds and \k{c(1,2)[1.6]} returns the first element of the
+vector, as the double is converted to an integer. R does not distinguish
+between scalars and vectors (they are all vectors), so \code{typeof(5) ==}
+\code{typeof(c(5)) == typeof(c(5,5))} \code{ == "double"}.
+
+All vectorized data types have a distinguished missing value denoted
+by \code{NA} (for ``not available'').  The default type of \code{NA}
+is \k{logical}.  We can see that \code{typeof(NA)=="logical"}, but
+\k{NA} inhabits every type: \code{typeof(c(1,NA)[2])} \k{==}
+\k{"double"}.  In addition to \code{NA}, R has a \code{NULL} value,
+and to understand why R has both a \code{NULL} value and \code{NA}
+values it is useful to remember that R is primarily designed for
+statistics and data science.  In experimental data science, it is
+useful to have a notion of a ``missing observation'', which is often
+NaN (not a number, or similar) in, e.g., MATLAB and Python.  Since
+vectors are unityped in R, it {\it must} have a ``NaN'' value for {\it
+  each} primitive type.  There single \code{NULL} null-typed value in
+the language to encompass all other cases.  In a sense, you can think
+of \code{NA} as representing a missing data point, and \code{NULL} as
+a missing data {\it set}.
 
@@
 
-\noindent
-All vectorized data types have a distinguished missing value denoted
-by \code{NA} (for ``not available'').  The type of \code{NA} is
-\k{logical} (\code{typeof(NA)=="logical"}), but \k{NA} inhabits every
-type: \code{typeof(c(1,NA)[2])} \k{==} \k{"double"}.  R also has a
-\code{NULL} value.  In data science, it is useful to have a notion of
-a ``missing observation'', since vectors are monomorphic there is a
-need for a missing value to inhabit each primitive type. In a sense,
-\code{NA} represents a missing data point, and \code{NULL} is a
-missing data set.
-
-Over time, programmers have found the need for a richer type
-structure.  R supports this with {\it attributes}. One may think of
-attributes as an optional map from names to values attached to any
-built in type.  Attributes are used to encode various type
-structures. They are queried with \k{attributes} and \k{class}.  Using
-attributes, programmers can extend the set of types by tagging
-data. For example, take the vector of four values,
-\code{x<-c(1,2,3,4)} and attach attribute \k{dim},
-\code{attr(x,"dim")<-c(2,2)}, to treat \k{x} as a 2x2 matrix.
-
-Another attribute that can be set is the \k{class}, this is one way to
-use R for object-oriented programming.  This attribute can be bound to
-a list of class names. For instance, \code{class(x)<-"human"} will set
-the class of \k{x} to be \k{human}.  There are three
-object-orientation frameworks in R: S3, S4, and R5.  The S3 object
-system support single dispatch on the class of the first argument of a
-function, whereas the S4 object system allows multiple dispatch. R5
-allows for users to define objects in a more imperative style. Some of
-the most widely used data types leverage attributes, e.g., data frames
-and matrices. A data frame, for instance, is a list of vectors with a
-class and a column name attribute, and matrices are vectors with a
- \k{dims} attribute.
-
-R functions have a number of quirks.  Arguments can be assigned
-arbitrary expressions as default values; functions take variable
-numbers of parameters, and can be called positionally and nominally.
-Consider this example:
-
-\begin{lstlisting}
-f <- function(x,..., y=if(z==0) 1, z=0) { x + y + if(missing(...)) 0 else c(...) }
-\end{lstlisting}
-
-\noindent
-This function has four formal parameters, \k x, dots, \k y and \k
-z. Parameter \k x can be bound positionally or passed by name. The
-dots, \k{...}, is always positional. The remaining two parameters must
-be passed by name as they are preceded by dots; \k y and \k z have
-default values, in the case of \k z this is a constant, but \k{y} is
-bound to an expression that depends on \k z's value (if \k z is not
-zero, \k y defaults to \k{NULL}). The body of the function will add \k
-x and \k y tp either 0 or the result of concatenating the dots into a
-primitive vector. The function \k{missing} tests if a parameter was
-explicitly passed. The following are some valid invocations of \k f:
-
-\begin{lstlisting}
- > f(1)    
-   [1] 2             # a double vector, y is 0, ... is missing
- > f(2, 3, x=1)
-   [1] 4 5           # a double vector, y is 0, ... is 2, 3
- > f(x=1, y=1)
-   [1] 2             # a double vector, y is 1, ... is missing
- > f(x=1, z=1)
-   numeric(0)        # a double vector of length 0, y is NULL
- > f(1L, 2L, y=1L)
-   [1] 4             # an integer vector, y is integer 1, ... is integer 2
- > f(1, y=c(1,2))
-   [1] 1 2           # a double vetor, y is 1, 2, ... is missing
-\end{lstlisting}
-
-\noindent
-The above hints at polymorphism: \k f may return a vector of integers
-or of doubles of length equal to the max length of its arguments.
+With one exception, all vectorized data types are monomorphic.  The
+exception is the \k{list} type which can hold values of any other type
+including \k{list}. For all monomorphic data types, attempting to
+store a value of a different type will cause a conversion. Either the
+value is converted to the type of the vector, or the vector is
+converted to the type of the value.
+
+Over the years, programmers have found the need for a richer type structure
+and have added {\it attributes}. The best way to think of attributes is as
+an optional map from names to values that can be attached to any object.
+Attributes are used to encode various type structures. They can be queried
+with functions such as \k{attributes} and \k{class}.  The addition of
+attributes lets programmers extend the set of types by tagging data with
+user-defined attributes. For example, one could define a vector of four
+values, \code{x<-c(1,2,3,4)} and then attach the attribute \k{dim} with a
+pair of numbers as value: \code{attr(x,"dim")<-c(2,2)}.  From that point,
+arithmetic functions will treat \k{x} as a 2x2 matrix.
+
+Another attribute that can be set is the \k{class}.  This attribute
+can be bound to a list of class names. For instance,
+\code{class(x)<-"human"} will set the class of \k{x} to be \k{human}.
+The main usage of attributes is for object-oriented programming. There
+are three object-orientation frameworks in R: S3, S4, and R5.  The S3
+object system support single dispatch on the class of the first
+argument of a function, whereas the S4 object system allows multiple
+dispatch (on all arguments). R5 allows for users to define objects in
+a more imperative style. Some of the most widely used data types
+leverage attributes, e.g., data frames and matrices. A data frame, for
+instance, is a list of vectors with a class and a column name
+attribute, and matrices are vectors with a \k{dims} attribute.
+
+Scalar data types include the distinguished \k{NULL} value, instances
+of classes written using the S4 object system, closures, and
+environments.  The implementation of R has a number of other types
+listed in Figure~\ref{rtypes} for reference.
 
@@
 \begin{figure}[!h]  \centering
-  \includegraphics[width=.7\linewidth]{plots/argument_polymorphism.pdf}
+  \includegraphics[width=.8\linewidth]{plots/argument_polymorphism.pdf}
   \caption{Size of unions}\label{fig:unionfreq}
@@
 \begin{figure}[!h]  \centering
-  \includegraphics[width=.7\linewidth]{plots/function_polymorphism.pdf}
+  \includegraphics[width=.8\linewidth]{plots/function_polymorphism.pdf}
   \caption{Function Polymorphism}\label{fig:funpoly}
@@
 
+\begin{figure}[!h]  \centering
+	\includegraphics[width=.8\linewidth]{plots/merge_strategy_comp.pdf}
+	\caption{Comparison of top-level function type counts across different merge strategies.}\label{fig:mergecomp}
+\end{figure}
+
 To connect this discussion with actual R types, we chose the \AT{which?} function from the \AT{which?} package, and show what the type of the function would be when using each of these strategies.

### 2020-10-14 paper/main.tex
edits, tables

--- 2020-10-14 paper/main.tex
@@
 
-\AT{Let's discuss alternative strategies for consolidating types here.}
-
 Throughout this paper, we employed the following strategy to consolidate types: We collected all of the traces observed for a function, and combined them into a single function type, where the type for each argument position was made up of a union of all of the observed types at that position.
@@
 
-To connect this discussion with actual R types, we chose the \AT{which?} function from the \AT{which?} package, and show what the type of the function would be when using each of these strategies.
+\begin{figure}[!h]  \centering
+	\includegraphics[width=.8\linewidth]{plots/merge_strategy_comp.pdf}
+	\caption{Comparison of top-level function type counts across different merge strategies.}\label{fig:mergecomp}
+\end{figure}
+
+% To connect this discussion with actual R types, we chose the \code{discretize} function from the \code{arules} package, and show what the type of the function would be when using each of these strategies.
+% The purpose of the function is to convert a continuous variable into a categorical variable, using several unsupervised learning methods.
+To connect this discussion with actual R types, we chose the \code{correlation} function from the \code{agricolae} package, and show what the type of the function would be when using each of these strategies.
+\code{agricolae::correlation} obtains the coefficients of correlation and p-value between all variables of some input data table, and uses a method of the user's choosing.
+It returns the correlation matrix and the probability together in a list.
 We refer the reader to Figure~\ref{tbl:real-function-different-merge-strats}.
-\AT{We could use stringr::str\_extract\_all?}
+We see that the hybrid approach produces a much smaller type, with only two arrow types in the top-level union.
+Moreover, we see that no real precision is lost when further reducing the function type to a pure arrow of unions, as the two options in the type obtained with the hybrid approach are not very informative.
 
@@
 			\midrule
-			Union of Arrows & \\
-			Hybrid &  \\
-			Arrow of Unions & \\
+			Union of Arrows & \makecell[tl]{
+				\FUN{\VEC{\dbl}, \VEC{\dbl}, \chr, \chr}{\nul} | \\
+				\FUN{\VEC{\int}, \VEC{\dbl}, \chr, \chr}{\nul} | \\
+				\FUN{\CLASS{{\tt data.frame}}, \nul, \VEC\chr, \chr}{\LIST{\CLASS{{\tt matrix}} \; | \; \dbl}} | \\
+				\FUN{\CLASS{{\tt data.frame}}, \nul, \chr, \chr}{\LIST{\CLASS{{\tt matrix}} \; | \; \dbl}} | \\
+				\FUN{\VEC\dbl, \CLASS{{\tt data.frame}}, \VEC\chr, \chr}{\LIST{\CLASS{{\tt matrix}} \; | \; \dbl}} | \\
+				\FUN{\VEC\dbl, \CLASS{{\tt data.frame}}, \chr, \chr}{\LIST{\CLASS{{\tt matrix}} \; | \; \dbl}}
+			} \\
+		\addlinespace
+			Hybrid & \makecell[tl]{
+				\FUN{\VEC{\dbl}, \VEC{\dbl}, {\chr}, {\chr}}{\nul} | \\
+				\FUN{\CLASS{{\tt data.frame}} \; | \; \VEC{\dbl}, ? \; \CLASS{{\tt data.frame}}, \VEC{\chr}, {\chr}}{\LIST{\CLASS{{\tt matrix}} \; | \; \dbl}}} \\
+			\addlinespace
+			Arrow of Unions & \FUN{\CLASS{{\tt data.frame}} \; | \; \VEC{\dbl}, ? \; \VEC{\dbl} \; | \; \CLASS{{\tt data.frame}}, \VEC{\chr}, {\chr}}{\, ? \; \LIST{\CLASS{{\tt matrix}} \; | \; \dbl}} \\
 			\bottomrule
 	\end{tabular}}
-	\caption{Types of \AT{package::function} with different type merge strategies. }\label{tbl:real-function-different-merge-strats}
+	\caption{Types of \code{agricolae::correlation} with different type merge strategies. }\label{tbl:real-function-different-merge-strats}
+\end{table}
+
+On the other hand, there are real cases where a meaningful loss of precision occurs.
+Consider instead the \code{rename} function from the \code{dplyr} package, with types shown in Figure~\ref{tbl:real-function-different-merge-strats-2}.
+This function takes a data frame and renames selected columns.
+We see that with both the union-of-arrows and hybrid approach, the types are the same (with only three unique function signatures observed), and can glean from these types that \code{dplyr::rename} produces a data frame of the {\it same class} as the input.
+This information is lost in the arrow-of-unions merge strategy.
+That said, once we extend the type system with proper user-defined type hierarchies, the union-of-arrows type will be much more clear.
+
+\begin{table}[H]
+	\centering
+	\resizebox{.98\linewidth}{!}{
+		\begin{tabular}{llrrr}
+			\toprule
+			Merge Strategy & Function Type \\
+			\midrule
+			Union of Arrows &  \makecell[tl]{
+		      \FUN{\CLASS{{\tt data.frame}}, ...}{\CLASS{{\tt data.frame}}}	| \\
+             \FUN{\CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}, ...}{\CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}} | \\
+             \FUN{\CLASS{{\tt data.frame, tbl, tbl\_df}}, ...}{\CLASS{{\tt data.frame, tbl, tbl\_df}}}
+			} \\
+		\addlinespace
+			Hybrid & \makecell[tl]{
+				\FUN{\CLASS{{\tt data.frame}}, ...}{\CLASS{{\tt data.frame}}}	| \\
+				\FUN{\CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}, ...}{\CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}} | \\
+				\FUN{\CLASS{{\tt data.frame, tbl, tbl\_df}}, ...}{\CLASS{{\tt data.frame, tbl, tbl\_df}}}
+			} \\
+		\addlinespace
+			Arrow of Unions & \makecell[tl]{
+				$\langle \CLASS{{\tt data.frame}} \;|\; \CLASS{{\tt data.frame, tbl, tbl\_df}} \;|\; \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}, ...} \rangle$ $\rightarrow$ \\
+				$\CLASS{{\tt data.frame}} \;|\; \CLASS{{\tt data.frame, tbl, tbl\_df}} \;|\; \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}$
+			} \\
+			\bottomrule
+	\end{tabular}}
+	\caption{Types of \code{X::Y} with different type merge strategies. }\label{tbl:real-function-different-merge-strats-2}
 \end{table}

### 2020-10-15 paper/main.tex
types

--- 2020-10-14 paper/main.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Designing A Type Language for R}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{A Type Language for R}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 In this section, we set out to propose a candidate design for a type
-language to describe the arguments and return values of R functions.
-The goal is not to propose a final design, but rather a
-starting point for an iterative design process.
-
-Fig.~\ref{types} presents our type language. An early design choice
-was to stay close to the R language and only depart in small and
-hopefully non-controversial ways.  Functions types have the form
-\FUN{A_1, \dots, A_n}\T where each $\A_i$ argument is either a type \T
-or \VARGS, a variable length argument list. The rest of this section
-details and motivates our design.
+language to describe the signatures of R functions.  The goal is not
+to propose a final design, but rather a starting point for an
+iterative process.
+
+Given the pecularities of the language there are a number of design
+choices that need to be reviewed and evaluated. It is uncontroversial
+to include types that cover the six kinds of primitive vectors,
+furthermore environments and the distinguished null types are commonly
+used and must be included. Fig.~\ref{types} presents our type language.
 
@@
 
-\paragraph{Scalar vs. Vectors} We distinguish between
+\paragraph{Scalars} We distinguish between
 vectors of length 1, and vectors of any dimension. The base types, \int,
@@
 
-\paragraph{NA values} In R, a value can be \emph{not available}, written
-\NA. Each of basic types has its specific \NA. Thus there is an \int \NA as
-well as a \dbl \NA, and they are different.  The type system allows one to
-distinguish between vectors that may contain {\NA}s (written \NAVEC\int) and
-those who are guarantee to be \NA-free (\VEC\int).  A \NA-free vector can be
-treated as a vector that may {\NA}s.  We do not allow scalar values to be
-\NA and the type \raw does not allow {\NA}s.
+\paragraph{Missing} 
+Each of basic types has its specific \NA. Thus there is an \int \NA as
+well as a \dbl \NA, and they are different.  The type system allows
+one to distinguish between vectors that may contain {\NA}s (written
+\NAVEC\int) and those who are guarantee to be \NA-free (\VEC\int).  A
+\NA-free vector can be treated as a vector that may {\NA}s.  We do not
+allow scalar values to be \NA and the type \raw does not allow {\NA}s.
 
 \paragraph{Unions} We support untagged unions of types written
-$\T_1 \k{~|~} ... \k{~|~} \T_n$.
+$\T_1 \k{~|~} ... \k{~|~} \T_n$. It is noteworthy that the elements
+of a union are not necessarily disjoint. For instance a \NAVEV\int
 
@@
 escape from the copy-on-write semantics of other data types.
+
+\paragraph{Functions.}
+Functions signatures the form \FUN{A_1, \dots, A_n}\T where each
+$\A_i$ argument is either a type \T or \VARGS, a variable length
+argument list. Moreover, a single function's signature can be the
+disjunction of a number of individual signatures.
 

### 2020-10-15 paper/main.tex
typo
stuff about base, will clean up tomorrow

--- 2020-10-14 paper/main.tex
@@
 We call this an {\it arrow of unions}.
-Now, this strategy was not developed in a vacuum, and we will now discuss some of the decisions that we made in choosing this strategy.
+Now, this strategy was not developed in a vacuum, and we will now discuss some of the decisions that we made in choosing it.
 
@@
 
-An obvious solution to this is simply to convert each call trace into a type, and generate a union of function types as the overall type for the function.
+One solution to this is simply to convert each call trace into a type, and generate a union of function types as the overall type for the function.
 This would lead to the type  \FUN{\chr}{\chr} $|$ \FUN{\dbl}{\dbl} for the aforementioned function.
@@
 
+Finally, we want to briefly discuss types for base package functions (in R, functions like \code{+}, \code{-}, vector access, etc. are part of the {\it base package}).
+One can find the implementation of these functions in the R source code\footnote{A read-only mirror of the R source can be found at \url{https://github.com/wch/r-source}.}, where they are implemented in C.
+As we have alluded to, these functions coerce mismatched arguments in an ad hoc manner, with no real defined semantics.
+Further, functions like \code{+} often act on values based on their type (according to \code{typeof}), and ignore the class unless some package extended \code{+} to dispatch on some new class.
+This leads to incredibly large infered signatures for these functions.
+Even after counting out traces related to S3 and S4 dispatch, the type of \code{+} using the hybrid approach has 29 top-level alternatives, 22 of which have a class-typed argument.
+We don't believe this type to be particularly useful to a programmer, but a compiler might find it quite useful, as given a set of types for inputs, it can select the appropriate arrow type from the top-level alternatives.
+
 %

### 2020-10-15 paper/main.tex

--- 2020-10-15 paper/main.tex
@@
 $\T_1 \k{~|~} ... \k{~|~} \T_n$. It is noteworthy that the elements
-of a union are not necessarily disjoint. For instance a \NAVEV\int
+of a union are not necessarily disjoint. For instance a \NAVEC(\int)
 

### 2020-10-15 paper/main.tex
wip

--- 2020-10-15 paper/main.tex
@@
 furthermore environments and the distinguished null types are commonly
-used and must be included. Fig.~\ref{types} presents our type language.
+used and must be included. For simplicity, we omit some of the data
+types that closer to the implementation of the
+language. Fig.~\ref{types} presents our type language.
 
@@
   & |   & \k{?} \T      & \it nullable type \\
-  & |   & \FUN{A_1, ..., A_n}\T    & \it function type \\
+  & |   & \FUN{A_1, \dots~ A_n}\T    & \it function type \\
   & |   & \LIST\T                  & \it list type\\
-  & |   & \CLASS{ID_1, ..., ID_n}  & \it class type\vspace{5pt}\\
+  & |   & \CLASS{ID_1, \dots~ ID_n}  & \it class type\vspace{5pt}\\
 \end{tabular}\end{minipage}   \hfill
@@
 \A & ::= & \T          & \it arguments \\
-   & |   & \VARGS      & \vspace{5pt} \\
-\V & ::= & ~\VEC\S     & \\
+   & |   & \VARGS      & \it dots \vspace{5pt} \\
+\V & ::= & ~\VEC\S     & \it vector types\\
    & |   & \NAVEC\S    & \it na vector types \vspace{5pt}\\
-\S & ::= & \int        & \\
-   & |   & \chr \\
-   & |   & \dbl \\
-   & |   & \lgl \\
-   & |   & \clx \\
-   & |   & \raw \vspace{5pt}\\
+\S & ::= & \int        & \it integer\\
+   & |   & \chr        & \it character\\
+   & |   & \dbl        & \it double \\
+   & |   & \lgl        & \it logical\\
+   & |   & \clx        & \it complex \\
+   & |   & \raw        & {\it raw } \vspace{5pt}\\%
 \end{tabular}\end{minipage}\caption{The R type language}\label{types}
@@
 
-\paragraph{Scalars} We distinguish between
-vectors of length 1, and vectors of any dimension. The base types, \int,
-\dbl, \clx, \lgl (booleans), \chr, and \raw (bytes), can be either vectors
-(e.g., \VEC\int) or scalars (e.g., \int).  A vector can happen to be of
-length 1, and thus a scalar value is also a vector.  Vectors are
-homogeneous, in that a vector of doubles contains only doubles.
+\paragraph{Scalars} While R does not have scalar data types, there are
+cases where functions except scalar values, for example a conditional
+takes a single logical (and will complain if more values are passed).
+We considered tracking dimensions of data structures, but decided
+against it. Instead, the type language differentiates between vectors
+of lenght 1 and vectors of any dimension.  The primitive types can be
+either vectors (e.g., \VEC\int) or scalars (e.g., \int).  A vector can
+happen to be of length 1, and thus a scalar is also a vector.  Vectors
+are monomorphic, a vector of doubles contains only doubles.
 
 \paragraph{Missing} 
-Each of basic types has its specific \NA. Thus there is an \int \NA as
-well as a \dbl \NA, and they are different.  The type system allows
-one to distinguish between vectors that may contain {\NA}s (written
-\NAVEC\int) and those who are guarantee to be \NA-free (\VEC\int).  A
-\NA-free vector can be treated as a vector that may {\NA}s.  We do not
-allow scalar values to be \NA and the type \raw does not allow {\NA}s.
+Each of basic types has its specific \NA. Many built-in functions,
+especially those implemented in C or Fortran, do not support \NA
+values.  It is thus advantageous to distinguish between vector that
+can contain missing values and those that are guaranteed not to.  In
+our experience, functions that expect scalar values tend to not admit
+{\NA}s, thus scalar types are treated as being \NA-free. The type
+language does allows to write \VEC\int to specify that a vector of
+integers may have {\NA}s and \NAVEC\int to say that a vector must not
+have missing values. Of course, a value of type \VEC\int may happen to
+not have any \NA and thus be of type \NAVEC\int.  The type \raw does
+not allow {\NA}s, so \VEC\raw is \NA-free.
 

### 2020-10-15 paper/main.tex
wip

@@
 Given the pecularities of the language there are a number of design
-choices that need to be reviewed and evaluated. It is uncontroversial
-to include types that cover the six kinds of primitive vectors,
-furthermore environments and the distinguished null types are commonly
-used and must be included. For simplicity, we omit some of the data
-types that closer to the implementation of the
+choices that need to be reviewed and evaluated. It is not
+controversial to include types that cover the six kinds of primitive
+vectors, furthermore environments and the distinguished null types are
+commonly used and must be included.  Environments are lists with
+reference semantics: mutating a value in an environment is performed
+in-place.  They are used to store variables and to escape from the
+copy-on-write semantics of other data types.  For simplicity, we omit
+some of the data types that closer to the implementation of the
 language. Fig.~\ref{types} presents our type language.
@@
 
-\paragraph{Scalars} While R does not have scalar data types, there are
+\paragraph{Scalar} While R does not have scalar data types, there are
 cases where functions except scalar values, for example a conditional
@@
 
-\paragraph{Unions} We support untagged unions of types written
-$\T_1 \k{~|~} ... \k{~|~} \T_n$. It is noteworthy that the elements
-of a union are not necessarily disjoint. For instance a \NAVEC(\int)
-
-\paragraph{Nullables} 
+\paragraph{Nullable} 
 The \nul type is inhabited by a singleton \code{NULL} value often used
-as a sentinel or a stand in for arguments with no value.  To capture
-this behavior, we introduce a nullable type \code{?}\T.  The
-functional difference between the \NA and \code{NULL} values is that
-the latter cannot be stored in vectors.
+as a sentinel. Unlike in some other lanugage, \k{NULL} is not the
+default value of uninitialized variables. R has different notion for
+that (which we do not cover here).  To capture common uses, of
+\k{NULL}, the type language has a nullable type, written \code{?}\T.
+Values of this type can be either values of type \T or \k{NULL}.
 
 \paragraph{Lists} Heterogeneous collections are implemented using lists.
-Lists and vectors are closely related: a vector converts to a list with
-\code{as.list}, and lists to vectors with \code{unlist} (coercions may
-ensue). The list type is parameterize: \LIST\T.
-
-\paragraph{Classes} 
-R has more than one notion of type. As we mentioned, values can be
-attributed, and a very important attribute is the class of a value.  A
-class is a list of names that are used to mimic object-oriented
-programming. The type system includes class types written
-\CLASS{\ID_1,\dots,\ID_n}.
-
-\paragraph{Environments}
-Environments are lists with reference semantics: mutating a value in an
-environment is performed in-place.  They are used to store variables and to
-escape from the copy-on-write semantics of other data types.
-
-\paragraph{Functions.}
+Lists and vectors are closely related: a vector converts to a list
+with \code{as.list}, and lists to vectors with \code{unlist}
+(coercions may ensue). The type language allows to specify that a
+value is a list containing element of some type \T, written \LIST\T. R
+does not have built-in type tests for this purpose, to establish the
+type of a value requires traversal and checking individual elements.
+
+\paragraph{Class}
+R has three objects systems, code-named S3, S4 and R5. All of them
+operate by adding a class attribute to values. The most widely used
+system, by far, is S3 which supports single dispatch and multiple
+classes~\cite{ecoop12}. The challenge from a type system point of view
+is that a value such as integer 5, could be attributed with a class.
+Code that performs dispatch would use the class attribute while code
+that does not would view the value as an integer. The type language
+focuses on the attribute and will hide the underlying type of the
+value. While that seems to match common usage it does represent a loss
+of expressiveness. The type language also focuses on S3, we leave the
+other object systems for future work. S3 has no notion of inheritance,
+each value has a list of classes. The type language thus allows to
+write \CLASS{\ID_1,\dots,\ID_n} to denote values tagged with the class
+names $\ID_1$ to $\ID_n$.
+
+\paragraph{Union} We support untagged unions of types written
+$\T_1 \k{~|~} ... \k{~|~} \T_n$. The elements of unions are not
+disjoint, 1L is both an \int, a \VEC\int and a \NAVEC\int. R does not
+provide a built-in type testing mechanism, e.g. for the case of
+\NA-free data types, it is necessary to scan vectors to find out if
+they have missing values.
+
+\paragraph{Function}
 Functions signatures the form \FUN{A_1, \dots, A_n}\T where each
@@
 
-\section{Extracting and Checking Signatures}
+\subsection{Subtyping}
+
+The choice of type language follows the structure of values.  The
+presence of \NA-free data types and scalars are two choices that must
+be validated in practice.  Nullable types are just a special case of
+unions.
+
+While R does not support the notion of subtyping between values.  The
+conversions between primitive types give us a starting point (e.g., an
+\int is always accepted where a \dbl is expected). Furthermore, the
+types introduced above induce some the rules of Figure~\ref{fig:sub}:
+a vector without \NA is a subtype of a vector, a value of type \T is a
+subtype of a nullable \T, a list is subtype of another list if their
+elements are subtypes, and a scalar is a subtype of a vector of the
+same primitive type.
+
+\begin{figure}[h!tb]\begin{tabular}{rclll}
+  \NAVEC\S  & <:& \VEC\S \\
+  \T        & <:& \k{?}\T \\
+  \LIST{\T} & <:& \LIST{\T'} & \it{iff} \T <: \T' \\
+  \S        & <:& \VEC\S \\
+  \lgl      & <:& \int\\
+  \int      & <:& \dbl\\
+  \dbl      & <:& \clx\\  
+\end{tabular}
+\caption{Subtyping}\label{fig:sub}
+\end{figure}
+
+\subsection{Synthesizing signatures}
+
+The output of the \typetracer tool consists of a set of tuples of raw
+types, each representing a function call in some program's
+execution. The inference step consolidates the different tuples
+corresponding to a particular function definition across multiple
+programs and distills them into a single type signature.  The shape of
+the inferred function signatures is:
+
+\vspace*{1mm}
+\FUN{ T_{1,1} \k{~|~} T_{1,i}, \dots, T_{n,1} \k{~|~} T_{n,j}}{T_1 \k{~|~} ... \k{~|~} T_k}
+\vspace*{1mm}
+
+\noindent
+In other words, we take the union of the types occurring at individual
+argument positions rather than an union of function
+types. Furthermore, we apply some transformation on the types to keep
+the size of types in check.  Figure~\ref{fig:siplification-rules}
+overviews the main simplification rules as well as the simple notion
+of subtyping that we have adopted here.
+
+\begin{figure}[h!tb]\begin{tabular}{rclll}
+  \T \k{~|~} \T  & $\Rightarrow$ &  \T &  \\
+  \T \k{~|~} $T'$ & $\Rightarrow$ &  \T & \it{iff} $T' <: T$  \\
+  \LIST{\T} \k{~|~} $\LIST{T'}$ & $\Rightarrow$ &  \LIST\T & \it{iff} $T' <: T$  \\
+   $\nul \k{~|~} \VEC{\S_1} \k{~|~}\dots \VEC{\S_n}$  & $\Rightarrow$ &
+  $ \NAVEC{\S_1} \k{~|~}\dots \NAVEC{\S_n}$ &  \\
+\end{tabular}
+\caption{Simplification rules}\label{fig:siplification-rules}
+\end{figure}
+
+Assuming that type sequences can be reordered freely, we rewrite types
+to minimize their size by removing redundant types, types that are
+subsumed by subtyping, immutable lists, and remove \nul type to
+replace them with nullables.  Functions are limited to twenty
+arguments as a simplifying assumption. Higher-order functions are
+conservatively inferred as \any $\rightarrow$ \any by \typetracer.
+
+\section{Type Extraction and Checking Pipeline}
 
@@
 
-\begin{figure}[t!]
-  \vspace{2mm}
+\begin{figure}[h!]  \vspace{2mm}
   \includegraphics[width=\textwidth]{plots/analysis-pipeline-new.pdf}
-  \caption{The Analysis Pipeline. (1) the rather long time is because
-    we had to extract metadata from all CRAN packages before we could
-    compose our \PACKAGES package corpus. Here CRAN refers to the
-    Comprehensive R Archive Network, R's repository of packages and
-    libraries.}
-  \label{fig:pipeline}
+  \caption{The Analysis Pipeline; $^1$ metadata has to be extracted
+    for \emph{all} CRAN packages.}\label{fig:pipeline}
 \end{figure}
@@
 the relevant information is kept by the {\tt \_entry} variants.
-
-\subsection{Inferring type signatures from raw types}
-\enlargethispage{4mm}
-The output of the \typetracer tool consists of a set of tuples of raw types,
-each representing a function call in some program's execution. The inference
-step consolidates the different tuples corresponding to a particular function
-definition across multiple programs and distills them into a single type
-signature.  The shape of the inferred function signatures is:
-
-\vspace*{1mm}
-\FUN{ T_{1,1} \k{~|~} T_{1,i}, \dots, T_{n,1} \k{~|~} T_{n,j}}{T_1 \k{~|~} ... \k{~|~} T_k}
-\vspace*{1mm}
-
-\noindent
-In other words, we take the union of the types occurring at individual
-argument positions rather than an union of function
-types. Furthermore, we apply some transformation on the types to keep
-the size of types in check.  Figure~\ref{fig:siplification-rules}
-overviews the main simplification rules as well as the simple notion
-of subtyping that we have adopted here.  Our subtyping rules are both
-born out of necessity (e.g., scalars are subtypes of vectors since
-a. that is the case in R and b. we want to allow types like \dbl |
-\VEC\dbl to simplify), as well as from prevalent data type conversion
-with no ambiguity (e.g., \int to \dbl).
-
-\begin{figure}[h!tb]
-\begin{tabular}{rclll}
-  \T \k{~|~} \T  & $\Rightarrow$ &  \T &  \\
-  \T \k{~|~} $T'$ & $\Rightarrow$ &  \T & \it{iff} $T' <: T$  \\
-  \LIST{\T} \k{~|~} $\LIST{T'}$ & $\Rightarrow$ &  \LIST\T & \it{iff} $T' <: T$  \\
-   $\nul \k{~|~} \VEC{\S_1} \k{~|~}\dots \VEC{\S_n}$  & $\Rightarrow$ &
-  $ \NAVEC{\S_1} \k{~|~}\dots \NAVEC{\S_n}$ &  \\
-  \\
-  \NAVEC\S  & <:& \VEC\S \\
-  \T        & <:& \k{?}\T \\
-  \LIST{\T} & <:& \LIST{\T'} & \it{iff} \T <: \T' \\
-  \S        & <:& \VEC\S \\
-  \lgl      & <:& \int\\
-  \int      & <:& \dbl\\
-  \dbl      & <:& \clx\\  
-\end{tabular}
-\caption{Simplification rules and subtyping}
-\label{fig:siplification-rules}
-\end{figure}
-
-Assuming that type sequences can be reordered freely, we rewrite types
-to minimize their size by removing redundant types, types that are
-subsumed by subtyping, immutable lists, and remove \nul type to
-replace them with nullables.  Functions are limited to twenty
-arguments as a simplifying assumption. Higher-order functions are
-conservatively inferred as \any $\rightarrow$ \any by \typetracer.
 
@@
 
-\contractr provides an extensive API that allows users to explicitly insert
-contracts during interactive development by supplying the type signature as a
-string, selectively disable contract checking, and inspecting type assertions.
+\contractr provides an extensive API that allows users to explicitly
+insert contracts during interactive development by supplying the type
+signature as a string, selectively disable contract checking, and
+inspecting type assertions.
 
@@
 does not call the type-checker if it see the sentinel.
-
-% \newpage
 
@@
 
-\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2020-10-15 paper/main.tex
minor

--- 2020-10-15 paper/main.tex
@@
 furthermore environments and the distinguished null types are commonly
-used and must be included. For simplicity, we omit some of the data
-types that closer to the implementation of the
-language. Fig.~\ref{types} presents our type language.
+used and must be included. Fig.~\ref{types} presents our type language.
 
@@
   & |   & \k{?} \T      & \it nullable type \\
-  & |   & \FUN{A_1, \dots~ A_n}\T    & \it function type \\
+  & |   & \FUN{A_1, ..., A_n}\T    & \it function type \\
   & |   & \LIST\T                  & \it list type\\
-  & |   & \CLASS{ID_1, \dots~ ID_n}  & \it class type\vspace{5pt}\\
+  & |   & \CLASS{ID_1, ..., ID_n}  & \it class type\vspace{5pt}\\
 \end{tabular}\end{minipage}   \hfill
@@
 \A & ::= & \T          & \it arguments \\
-   & |   & \VARGS      & \it dots \vspace{5pt} \\
-\V & ::= & ~\VEC\S     & \it vector types\\
+   & |   & \VARGS      & \vspace{5pt} \\
+\V & ::= & ~\VEC\S     & \\
    & |   & \NAVEC\S    & \it na vector types \vspace{5pt}\\
-\S & ::= & \int        & \it integer\\
-   & |   & \chr        & \it character\\
-   & |   & \dbl        & \it double \\
-   & |   & \lgl        & \it logical\\
-   & |   & \clx        & \it complex \\
-   & |   & \raw        & {\it raw } \vspace{5pt}\\%
+\S & ::= & \int        & \\
+   & |   & \chr \\
+   & |   & \dbl \\
+   & |   & \lgl \\
+   & |   & \clx \\
+   & |   & \raw \vspace{5pt}\\
 \end{tabular}\end{minipage}\caption{The R type language}\label{types}
@@
 
-\paragraph{Scalars} While R does not have scalar data types, there are
-cases where functions except scalar values, for example a conditional
-takes a single logical (and will complain if more values are passed).
-We considered tracking dimensions of data structures, but decided
-against it. Instead, the type language differentiates between vectors
-of lenght 1 and vectors of any dimension.  The primitive types can be
-either vectors (e.g., \VEC\int) or scalars (e.g., \int).  A vector can
-happen to be of length 1, and thus a scalar is also a vector.  Vectors
-are monomorphic, a vector of doubles contains only doubles.
+\paragraph{Scalars} We distinguish between
+vectors of length 1, and vectors of any dimension. The base types, \int,
+\dbl, \clx, \lgl (booleans), \chr, and \raw (bytes), can be either vectors
+(e.g., \VEC\int) or scalars (e.g., \int).  A vector can happen to be of
+length 1, and thus a scalar value is also a vector.  Vectors are
+homogeneous, in that a vector of doubles contains only doubles.
 
 \paragraph{Missing} 
-Each of basic types has its specific \NA. Many built-in functions,
-especially those implemented in C or Fortran, do not support \NA
-values.  It is thus advantageous to distinguish between vector that
-can contain missing values and those that are guaranteed not to.  In
-our experience, functions that expect scalar values tend to not admit
-{\NA}s, thus scalar types are treated as being \NA-free. The type
-language does allows to write \VEC\int to specify that a vector of
-integers may have {\NA}s and \NAVEC\int to say that a vector must not
-have missing values. Of course, a value of type \VEC\int may happen to
-not have any \NA and thus be of type \NAVEC\int.  The type \raw does
-not allow {\NA}s, so \VEC\raw is \NA-free.
+Each of basic types has its specific \NA. Thus there is an \int \NA as
+well as a \dbl \NA, and they are different.  The type system allows
+one to distinguish between vectors that may contain {\NA}s (written
+\NAVEC\int) and those who are guarantee to be \NA-free (\VEC\int).  A
+\NA-free vector can be treated as a vector that may {\NA}s.  We do not
+allow scalar values to be \NA and the type \raw does not allow {\NA}s.
 
@@
 	\end{tabular}}
-	\caption{Types of \code{X::Y} with different type merge strategies. }\label{tbl:real-function-different-merge-strats-2}
+	\caption{Types of \code{dplyr::rename} with different type merge strategies. }\label{tbl:real-function-different-merge-strats-2}
 \end{table}

### 2020-10-15 paper/main.tex
discussion edits

--- 2020-10-15 paper/main.tex
@@
 We call this an {\it arrow of unions}.
-Now, this strategy was not developed in a vacuum, and we will now discuss some of the decisions that we made in choosing it.
-
-Arguably, the primary issue with the arrow of unions strategy is that of a loss of precision, as relationships between argument and return types are obscured when all argument and return types are unioned.
+This strategy was not developed in a vacuum, and we will now discuss some of the decisions that we made in choosing it.
+
+Arguably, the primary issue with the arrow of unions strategy is that of a loss of precision, as relationships between argument and return types are obscured when taking a per-argument union of types.
 For example, consider a function with two unique call traces, \FUN{\chr}{\chr} and \FUN{\dbl}{\dbl}.
-The strategy we employed results in a combined type of \FUN{\chr \;|\; \dbl}{\chr \;|\; \dbl} for the function, which obscures the (potential) relationship between the argument and return type.
-
-One solution to this is simply to convert each call trace into a type, and generate a union of function types as the overall type for the function.
+The strategy we employed results in a combined type of \FUN{\chr \;|\; \dbl}{\chr \;|\; \dbl} for the function, which obscures the (potential) relationship between the argument and return type (e.g., it is not as clear that the function maps character inputs to character outputs).
+
+One solution to this is simply to convert each call trace into a type, and generate a union of arrow types as the overall type for the function.
 This would lead to the type  \FUN{\chr}{\chr} $|$ \FUN{\dbl}{\dbl} for the aforementioned function.
-Unfortunately, this leads to a significant blow-up in the size of types, and makes many types unreadable, due to the myriad ways in which we combine the primitive types, as recall that, e.g., scalars and vectors are not the same type, and we see many types of the form \FUN{\chr}{\dbl} $|$ \FUN{\VEC\chr}{\dbl} $|$ \FUN{\NAVEC\chr}{\dbl}.
-When comparing this consolidation strategy against the one employed throughout the paper, we find that the types using this strategy are much larger, with 14,057 (55.06\%) functions have at least one top-level union.
+We call this the {\it basic union of arrows} strategy.
+Unfortunately, this leads to a significant blow-up in the size of types, and makes many types unreadable due to the myriad ways in which we combine the primitive types, as recall that, e.g., scalars and vectors are not the same type, and we see many types of the form \FUN{\chr}{\dbl} $|$ \FUN{\VEC\chr}{\dbl} $|$ \FUN{\NAVEC\chr}{\dbl}.
+When comparing this consolidation strategy against the one employed throughout the paper, we find that the types using are much larger, with 14,057 (55.06\%) functions have at least one top-level union.
 %, and that 16,745 (65.59\%) functions have two or fewer top-level alternatives.
-We observed 93,229 unique, independent signatures using this strategy.
-
-Another option is to employ a hybrid approach, wherein we perform a union of function types {\it after} grouping them together by return type, and performing basic primitive type consolidation on the return types.
+We observed 93,229 independent signatures using this strategy (i.e., the function types using the union of arrows strategy are made up of some ~93K arrow types).
+
+Another option is to employ a hybrid approach, wherein we group function traces by return type, and apply the arrow of unions strategy to these groups.
+We also perform basic primitive type consolidation on return types, so we would, e.g., merge two function traces \FUN{\dbl}{\int} and \FUN{\dbl}{\VEC\int} into a single arrow type, \FUN{\dbl}{\VEC\int}.
+% wherein we perform a union of function types {\it after} grouping them together by return type, and performing basic primitive type consolidation on the return types.
+We refer to this as the {\it hybrid union of arrows} strategy.
 This has the advantage of avoiding most of the blowup of the naive approach, however this hybrid approach suffers particularly when functions return classes, as our type system does not allow us to effectively consolidate class types.
-Comparing again to the strategy employed throughout the paper, we find 5,317 (20.82\%) functions to have a union of arrows, and 38,650 independent signatures.
+Comparing again to the strategy employed throughout the paper, we find 5,317 (20.82\%) functions to have a union of arrows, and we find 38,650 independent signatures.
 Most functions only have one or two top-level alternatives (23,530, or 92.17\%).
 We additionally find that 38,650 (26.14\%) of arrow types have at least one polymorphic argument.
+
+Figure~\ref{fig:mergecomp} shows a comparison of these two alternative merge strategies.
+Functions are grouped by the number of {\it top-level alternatives} in their types, and the number of top-level alternatives in a function type is the number of arrow types in the union of arrows.
+For example, the function type \FUN{\chr}{\dbl} $|$ \FUN{\VEC\chr}{\dbl} has two top-level alternatives.
+We see that just over 40\% of functions have a single arrow type at the top level using the basic union of arrows strategy, which climbs to nearly 80\% when employing the hybrid approach.
+Further, over 5\% of function types have $>9$ top-level alternatives using the basic strategy.
 
@@
 It returns the correlation matrix and the probability together in a list.
-We refer the reader to Figure~\ref{tbl:real-function-different-merge-strats}.
+We refer the reader to Figure~\ref{tbl:real-function-different-merge-strats}:
 We see that the hybrid approach produces a much smaller type, with only two arrow types in the top-level union.
-Moreover, we see that no real precision is lost when further reducing the function type to a pure arrow of unions, as the two options in the type obtained with the hybrid approach are not very informative.
+Moreover, we see that no real precision is lost when further reducing the function type to a pure arrow of unions, as the two top-level alternatives in the type obtained with the hybrid approach are not very informative.
 
@@
 
-On the other hand, there are real cases where a meaningful loss of precision occurs.
+In contrast, there are real cases where a meaningful loss of precision occurs.
 Consider instead the \code{rename} function from the \code{dplyr} package, with types shown in Figure~\ref{tbl:real-function-different-merge-strats-2}.
@@
 We see that with both the union-of-arrows and hybrid approach, the types are the same (with only three unique function signatures observed), and can glean from these types that \code{dplyr::rename} produces a data frame of the {\it same class} as the input.
-This information is lost in the arrow-of-unions merge strategy.
-That said, once we extend the type system with proper user-defined type hierarchies, the union-of-arrows type will be much more clear.
+This information is lost in the arrow of unions merge strategy.
+We believe that the arrow of unions type will be much improved once the type system is extended with user-defined type hierarchies.
 
@@
 Further, functions like \code{+} often act on values based on their type (according to \code{typeof}), and ignore the class unless some package extended \code{+} to dispatch on some new class.
-This leads to incredibly large infered signatures for these functions.
+This leads to incredibly large infered signatures for these functions:
 Even after counting out traces related to S3 and S4 dispatch, the type of \code{+} using the hybrid approach has 29 top-level alternatives, 22 of which have a class-typed argument.
-We don't believe this type to be particularly useful to a programmer, but a compiler might find it quite useful, as given a set of types for inputs, it can select the appropriate arrow type from the top-level alternatives.
+We don't believe these types to be particularly useful to a programmer, but a compiler may find them useful, as given a set of types for inputs, it can select the appropriate arrow type from the top-level alternatives.
 

### 2020-10-15 paper/main.tex
wip

--- 2020-10-15 paper/main.tex
@@
 
-\subsection{Synthesizing signatures}
-
-The output of the \typetracer tool consists of a set of tuples of raw
-types, each representing a function call in some program's
-execution. The inference step consolidates the different tuples
-corresponding to a particular function definition across multiple
-programs and distills them into a single type signature.  The shape of
-the inferred function signatures is:
+\subsection{Synthesizing function type signatures}
+
+In order to keep the type signatures compact, we will compact
+union of arrows into a single top-level arrow with unions
+at each argument position. Thus, the shape of function
+signatures will be as follows:
 
@@
 the size of types in check.  Figure~\ref{fig:siplification-rules}
-overviews the main simplification rules as well as the simple notion
-of subtyping that we have adopted here.
+overviews the main simplification rules that we have adopted here.
 
@@
 subsumed by subtyping, immutable lists, and remove \nul type to
-replace them with nullables.  Functions are limited to twenty
-arguments as a simplifying assumption. Higher-order functions are
-conservatively inferred as \any $\rightarrow$ \any by \typetracer.
+replace them with nullables.  Higher-order functions are
+conservatively treated as \any $\rightarrow$ \any.
 

### 2020-10-15 paper/main.tex
wip

--- 2020-10-15 paper/main.tex
@@
 
-\subsection{Synthesizing function type signatures}
-
-In order to keep the type signatures compact, we will compact
-union of arrows into a single top-level arrow with unions
-at each argument position. Thus, the shape of function
-signatures will be as follows:
+\subsection{Synthesizing signatures}
+
+In order to keep the type signatures compact, we will compact union of
+arrows into a single top-level arrow with unions at each argument
+position. Thus, the shape of function signatures will be as follows:
 
@@
 
-\section{Type Extraction and Checking Pipeline}
+It is noteworthy that by performing this compaction we are loosing
+precision, that is to say, the synthesized signatures will suggest
+that the function accepts combination of types that were not observed.
+
+\section{Analysis and Instrumentation Pipeline}
 
@@
 
-\subsection{Extracting raw type signature from traces}
-
-We implemented \typetracer, an automated tool for extracting raw type
-signatures from execution traces of R programs. The goal of this tool is to
-output a tuple $\langle f, t_1, \dots, t_n, t\rangle$ for each function call
-during the execution of a program, where $f$ is an identifier for a
-function, $t_i$ are type-level summaries of the arguments and $t$ is a
-summary of the return value.
-
-While this task is seemingly simple, the details and their proverbial devil
-are surprisingly tricky to get right and to scale to large, long running
-programs.  To avoid starting from scratch, our implementation reuses \rdt,
-an open source dynamic analysis framework for R~\cite{oopsla19} which
-consists of an instrumented R Virtual Machine based on GNU-R version
-3.5.0. The framework exposes hooks in the interpreter to which user defined
-callbacks can be attached. These hooks include function entry and exit,
-method dispatch for the S3 and S4 object systems, the longjumps used by the
-interpreter to implement non-local exit, creation and forcing of promises,
-variable definition, value creation, mutation and garbage collection.
-
-\paragraph{Raw types}
-The type information output by the tool includes the \emph{type tag} of each
-value. Internal types are translated to names in the proposed type
-system. The next bit of information is the \emph{class}, an optional list of
-names that may be absent, and, in some cases, is implicit (i.e. the
-interpreter blesses some values with the \k{matrix} and \k{array} classes
-even without attributes).  Depending on a value's type, the tool collects
-further information:
-
-\begin{itemize}
-\item For vectors, the presence of \NA values.
-\item For lists, element types by a recursive traversal.
-\item For promises, an approximation of the expression type.
-\end{itemize}
-
-\noindent
-To obtain these raw types, we make use of R's C FFI and use low-level
-machinery to collect information from the R language run-time.  Types are
-completed during post-processing, and rely on the detailed information made
-available by these low-level reflection mechanisms.
+\subsection{Types from traces}
+
+We implemented \typetracer, an automated tool for extracting types
+from execution traces of R programs. The goal of this tool is to
+output a tuple $\langle f, t_1, \dots, t_n, t\rangle$ for each
+function call during the execution of a program, where $f$ is an
+identifier for a function, $t_i$ are type-level summaries of the
+arguments and $t$ is a summary of the return value.
+
+While seemingly simple, the details and their proverbial devil are
+surprisingly tricky to get right at scale.  Our implementation reuses
+\rdt, an open source dynamic analysis framework for R~\cite{oopsla19}
+which consists of an instrumented R Virtual Machine based on GNU-R
+version 3.5.0. The framework exposes hooks in the interpreter to which
+user defined callbacks can be attached. These hooks include function
+entry and exit, method dispatch for the S3 and S4 object systems, the
+longjumps used by the interpreter to implement non-local exit,
+creation and forcing of promises, variable definition, value creation,
+mutation and garbage collection.
+
+\paragraph{Types}
+The type information output by the tool includes the \emph{type tag}
+of each value. Internal types are translated to names in the proposed
+type system. The next bit of information is the \emph{class}, an
+optional list of names that may be absent, and, in some cases, is
+implicit (i.e. the interpreter blesses some values with the \k{matrix}
+and \k{array} classes even without attributes).  Depending on a
+value's type, the tool collects further information: (a) for vectors,
+the presence of \NA values, (b) for lists, element types by a
+recursive traversal, and (c) for promises, an approximation of the
+expression type.  To obtain these types, we make use of R's C FFI and
+use low-level machinery to collect information from the R run-time.
+Types are completed during post-processing, and rely on the detailed
+information made available by these reflection mechanisms.
 
 \paragraph{Promises}
-The fact that arguments are lazy (expression are packed into promises and
-only evaluated on first access) complicates information gathering.  For
-example, some promises may remain un-evaluated, and it would be erroneous to
-force them as they may side-effect and change program behavior.  To deal
-with unevaluated arguments, we make an initial guess for each argument at
-function entry.  If the promise is later forced, we simply update the
-recorded type for the argument.
+The fact that arguments are lazy (expressions are packed into promises
+and only evaluated on first access) complicates information gathering.
+For example, some promises may remain unevaluated, and it would be
+erroneous to force them as they may side-effect and change program
+behavior.  To deal with unevaluated arguments, we make an initial
+guess for each argument at function entry and update the recorded type
+if the promise is forced.
 
 \paragraph{Missing arguments}
-Parameters which receive no values when the function is called are termed
-missing. This occurs when a function was called with too few arguments and
-no default values were specified for missing arguments.  We record a
-\k{missing} type for such argument.  There are two obvious ways to deal with
-missing arguments: type them as \any or type them as some unit type.  As we
-are performing a dynamic analysis, we conservatively type them as
-\any.
+Parameters which receive no values when the function is called are
+termed missing (not to be confused with \NA). This occurs when a
+function is called with too few arguments and no default values are
+specified for those missing arguments.  We record a \k{missing} type
+for such argument.  There are two obvious ways to deal with missing
+arguments: type them as \any or type them as some unit type.  We
+conservatively type them as \any.
 
@@
 
-\paragraph{Varargs}
-Arguments that are part of a function's varargs (denoted \VARGS) are
-ignored. We do not attempt to give varags a type in this type system.
+\paragraph{Dots}
+Arguments that are part of a dots parameter (denoted \VARGS) are
+ignored. We do not attempt to give dots a type.
 
 \paragraph{Implementation details.}
-We primarily rely on eight callbacks: \k{closure\_entry}, \k{closure\_exit},
-\k{builtin\_entry}, \k{builtin\_exit}, \k{special\_entry},
-\k{special\_exit}, \k{promise\_force\_entry}, and finally
-\k{promise\_force\_exit}.  The function-related callbacks are used mainly
-for bookkeeping: the analysis is notified that a construct has been entered
-by pushing the call onto a stack.  The calls themselves store a trace
-object that holds the type information. As R can perform
-single or multiple dispatch on function arguments depending on their class,
-the relevant information is kept by the {\tt \_entry} variants.
-
-\subsection{Checking types signatures with contracts}
-
-One way to validate an inferred function type is to check that it is
-respected in a different program which uses the function.  For this
-purpose, we developed \contractr, an R package that allows one to
-decorate a function with assertions. We use it to insert type checking
-of arguments and return values.
-
-\contractr's primary logic has been implemented in C++ to reduce its
-runtime overhead, and we have not observed a single segmentation fault
-during our use of the package.  It has been tested with GNU R-3.5.0
-and hardened with a battery of 400 test cases. It works by modifying
-function definitions to insert a call to a type-checking function.
-
-In terms of usability, our checker is enabled automatically whenever a
-package is loaded in R. So, a simple invocation of
-\code{library(contractr)} causes contracts to be injected in all
-packages. On loading, \contractr scans all packages in the user's
-workspace and inserts contracts in functions for which type signatures
-are available. Then \contractr sets up package load hooks which are
-executed when new packages are loaded. Furthermore, \contractr
-automatically removes contracts from all functions and restores them
-to their original state when it is unloaded. The type signatures can
-be provided in an external file, thus avoiding the need to change the
-source code of checked packages. Type declarations can also be written
-directly in the adjunct function comments using
-\roxygen~\footnote{\url{https://cran.r-project.org/web/packages/roxygen2/}}
-annotation tags. \contractr introduces a new use \code{@type}
-tag. When a package these type signatures are extracted and stored in
-the appropriate location. The following is an example of a type
-annotation using \roxygen:
+We primarily rely on eight callbacks: \k{closure\_entry},
+\k{closure\_exit}, \k{builtin\_entry}, \k{builtin\_exit},
+\k{special\_entry}, \k{special\_exit}, \k{promise\_force\_entry}, and
+finally \k{promise\_force\_exit}.  The function-related callbacks are
+used mainly for bookkeeping: the analysis is notified that a construct
+has been entered by pushing the call onto a stack.  The calls
+themselves store a trace object that holds the type information. As R
+can perform single or multiple dispatch on function arguments
+depending on their class, the relevant information is kept by the {\tt
+  \_entry} variants.
+
+\subsection{Checking signatures with contracts}
+
+One can validate a function's type signature by checking that it is
+respected in all programs that call the function.  For this, we
+developed \contractr, an R package that decorates functions with
+assertions. We use it to insert type checking code around functions.
+For speed, \contractr's primary logic is implemented in C++.  It has
+been tested with GNU R-3.5.0 and hardened with a battery of 400 test
+cases.  An invocation of \code{library(contractr)} causes contracts to
+be injected. \contractr scans all packages in the workspace and
+inserts contracts in functions for which type signatures are
+available. Package load hooks are executed when new packages are
+loaded. \contractr automatically removes contracts from all functions
+and restores them to their original state when it is unloaded. The
+type signatures can be provided in an external file, thus avoiding the
+need to change the source code of checked packages. Type declarations
+can also be written in comments using \roxygen annotations, using the
+\code{@type} tag:
 
@@
 
-\contractr provides an extensive API that allows users to explicitly
-insert contracts during interactive development by supplying the type
-signature as a string, selectively disable contract checking, and
-inspecting type assertions.
-
-\paragraph{Checking the values} A number of properties can be checked
-by a simple tag check, namely whether a value is a null, environment,
-vector or scalar. Other properties require an inspection of the
-contents of a value, such as checking for the absence of \NA or the
-type of the elements of list. Union types require checking for each of
-the members of the union. As the usage of higher-order functions
-remains limited in R, we do not check higher-order functions in this
-version of the tool. Checking them could be done by dynamically
-wrapping functions with checks.  When argument values are wrapped in
-promises (this is not always the case due to compiler optimizations),
-in order to retain the non-strict semantics of R, the expression held
-in the promise is wrapped in a call to the type checker, and type
-checking is delayed until the promise is forced. This leads to corner
-cases such that the type checking of a function may happen after that
-function has returned.
-
-Return values require care as well.  Functions return the last
-expression they evaluate, thus to avoid having to analyze the code of
-the called function the checker will register a callback on the exit
-hook. The hook is executed in the function call's environment. Another
-wrinkle is due to longjumps which causes active function calls on the
-stack to be discarded. When they are discarded, their exit hooks are
-called but they do not have a return value to type-check. \contractr
-deals with this problem by allocating a unique sentinel object which
-serves as the return value for calls that are discarded. The exit hook
-does not call the type-checker if it see the sentinel.
+The injected contracts check arguments with a simple tag check when
+possible. Some properties require traversing data structures, such as
+the absence of \NA. For union types, multiple checks may be needed, at
+worst one per member of the union.  In order to retain the non-strict
+semantics of R, the expression held in a promise is wrapped in a call
+to the type checker, and type checking is delayed until the promise is
+forced. This leads to corner cases such that the type checking of a
+function may happen after that function has returned.  Return values
+require care as well.  Functions return the last expression they
+evaluate, thus a callback is added to the exit hook.  Another wrinkle
+is due to longjumps which causes active function calls on the stack to
+be discarded. When they are discarded, their exit hooks are called but
+they do not have a return value to type-check. \contractr deals with
+this problem by allocating a unique sentinel object which serves as
+the return value for calls that are discarded. The exit hook does not
+call the type-checker if it see the sentinel.
 

### 2020-10-15 paper/main.tex
merge

--- 2020-10-15 paper/main.tex
@@
 \usepackage{makecell}
+\usepackage{microtype}
 
@@
 			Arrow of Unions & \makecell[tl]{
-				$\langle \CLASS{{\tt data.frame}} \;|\; \CLASS{{\tt data.frame, tbl, tbl\_df}} \;|\; \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}, ...} \rangle$ $\rightarrow$ \\
+				$\langle \CLASS{{\tt data.frame}} \;|\; \CLASS{{\tt data.frame, tbl, tbl\_df}} \;|\; \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}, ... \rangle$ $\rightarrow$ \\
 				$\CLASS{{\tt data.frame}} \;|\; \CLASS{{\tt data.frame, tbl, tbl\_df}} \;|\; \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}$

### 2020-10-16 paper/main.tex
edits that somehow disappeared?

@@
   available.\footnote{ \url{github.com/PRL-PRG/propagatr},
-  \url{github.com/PRL-PRG/contractr}, {\bf TODO add DATA link}}
+  \url{github.com/PRL-PRG/contractr}, data: \url{doi.org/10.5281/zenodo.4091818}, artifact: \url{doi.org/10.5281/zenodo.4037278}}
 \end{itemize}
@@
 
+\pagebreak
+
  %
@@
 
-Throughout this paper, we employed the following strategy to consolidate types: We collected all of the traces observed for a function, and combined them into a single function type, where the type for each argument position was made up of a union of all of the observed types at that position.
+Throughout this paper, we employed the following strategy to consolidate types: We collected all of the traces observed for a function, and combined them into a single function type, where the type for each argument position was made up of a union of all the observed types at that position.
 We then simplified these unions using the rules presented in Section~\ref{fig:siplification-rules}.
 We call this an {\it arrow of unions}.
-Now, this strategy was not developed in a vacuum, and we will now discuss some of the decisions that we made in choosing it.
+This strategy was not developed in a vacuum, and we will now discuss some of the decisions that went into it.
 
@@
 
-One solution to this is simply to convert each call trace into a type, and generate a union of function types as the overall type for the function.
+One solution to this is simply to convert each call trace into an arrow type, and generate a union of arrow types as the overall type for the function, dubbed the basic {\it union of arrows} strategy.
 This would lead to the type  \FUN{\chr}{\chr} $|$ \FUN{\dbl}{\dbl} for the aforementioned function.
@@
 %, and that 16,745 (65.59\%) functions have two or fewer top-level alternatives.
-We observed 93,229 unique, independent signatures using this strategy.
-
-Another option is to employ a hybrid approach, wherein we perform a union of function types {\it after} grouping them together by return type, and performing basic primitive type consolidation on the return types.
-This has the advantage of avoiding most of the blowup of the naive approach, however this hybrid approach suffers particularly when functions return classes, as our type system does not allow us to effectively consolidate class types.
+We observed 93,229 independent signatures using this strategy.
+
+Another option is to employ a hybrid approach, wherein we perform the union of arrow types {\it after} grouping them together by return type (to further simplify we also combine some primitive return types together, such as \dbl and \VEC\dbl).
+While this has the advantage of reducing most of the blow-up of the previous approach, it suffers particularly when functions return classes, as our type system does not allow us to effectively consolidate class types.
 Comparing again to the strategy employed throughout the paper, we find 5,317 (20.82\%) functions to have a union of arrows, and 38,650 independent signatures.
@@
 We additionally find that 38,650 (26.14\%) of arrow types have at least one polymorphic argument.
+
+These findings are summarized in Figure~\ref{fig:mergecomp}, which shows a breakdown of the number of {\it top-level alternatives} in the function types obtained with both of these strategies.
+The term ``top-level alternative'' signifies an arrow type in the union of arrows, e.g., \FUN{\chr}{\chr} and \FUN{\dbl}{\dbl} are top-level alternatives in the type \FUN{\chr}{\chr} $|$ \FUN{\dbl}{\dbl}.
+We see that the hybrid approach greatly increases the number of functions with no union of arrows at the function level, and nearly 5\% of function types obtained with the basic union of arrows strategy have over 9 top-level alternatives.
 
@@
 It returns the correlation matrix and the probability together in a list.
-We refer the reader to Figure~\ref{tbl:real-function-different-merge-strats}.
-We see that the hybrid approach produces a much smaller type, with only two arrow types in the top-level union.
-Moreover, we see that no real precision is lost when further reducing the function type to a pure arrow of unions, as the two options in the type obtained with the hybrid approach are not very informative.
+We refer the reader to Figure~\ref{tbl:real-function-different-merge-strats}, where we see that the hybrid approach produces a much smaller type, with only two arrow types in the top-level union.
+Moreover, we see that no real precision is lost when further reducing the function type to a pure arrow of unions, as the two top-level alternatives in the type obtained with the hybrid approach are not very informative.
 
@@
 This information is lost in the arrow-of-unions merge strategy.
-That said, once we extend the type system with proper user-defined type hierarchies, the union-of-arrows type will be much more clear.
+That said, once the type system is extend with proper user-defined type hierarchies, the arrow of unions type will be much smaller and more informative (as data frames, grouped tibbles, and tibbles are likely all related types).
 

### 2020-10-16 paper/main.tex
edits

@@
 \paragraph{Scalar} While R does not have scalar data types, there are
-cases where functions except scalar values, for example a conditional
+cases where functions expect scalar values, for example a conditional
 takes a single logical (and will complain if more values are passed).
@@
 
-Arguably, the primary issue with the arrow of unions strategy is that of a loss of precision, as relationships between argument and return types are obscured when all argument and return types are unioned.
+Arguably, the primary issue with the arrow of unions strategy is that of a loss of precision, as relationships between argument and return types can be obscured.
 For example, consider a function with two unique call traces, \FUN{\chr}{\chr} and \FUN{\dbl}{\dbl}.
-The strategy we employed results in a combined type of \FUN{\chr \;|\; \dbl}{\chr \;|\; \dbl} for the function, which obscures the (potential) relationship between the argument and return type.
+The arrow of union strategy results in a combined type of \FUN{\chr \;|\; \dbl}{\chr \;|\; \dbl} for the function, which hides the (potential) relationship between the argument and return type.
 
@@
 This would lead to the type  \FUN{\chr}{\chr} $|$ \FUN{\dbl}{\dbl} for the aforementioned function.
-Unfortunately, this leads to a significant blow-up in the size of types, and makes many types unreadable, due to the myriad ways in which we combine the primitive types, as recall that, e.g., scalars and vectors are not the same type, and we see many types of the form \FUN{\chr}{\dbl} $|$ \FUN{\VEC\chr}{\dbl} $|$ \FUN{\NAVEC\chr}{\dbl}.
-When comparing this consolidation strategy against the one employed throughout the paper, we find that the types using this strategy are much larger, with 14,057 (55.06\%) functions have at least one top-level union.
+Unfortunately, this leads to a significant blow-up in the size of types, and makes many types unreadable due to the myriad ways in which we combine the primitive types, as recall that, e.g., scalars and vectors are not the same type, and we see many types of the form \FUN{\chr}{\dbl} $|$ \FUN{\VEC\chr}{\dbl} $|$ \FUN{\NAVEC\chr}{\dbl}.
+When comparing this consolidation strategy against the one employed throughout the paper, we find that the types using this strategy are much larger, with 14,057 (55.06\%) functions having at least one top-level union.
 %, and that 16,745 (65.59\%) functions have two or fewer top-level alternatives.
-We observed 93,229 independent signatures using this strategy.
+We observed 93,229 independent arrow types using this strategy.
 
@@
 While this has the advantage of reducing most of the blow-up of the previous approach, it suffers particularly when functions return classes, as our type system does not allow us to effectively consolidate class types.
-Comparing again to the strategy employed throughout the paper, we find 5,317 (20.82\%) functions to have a union of arrows, and 38,650 independent signatures.
-Most functions only have one or two top-level alternatives (23,530, or 92.17\%).
+This strategy would reduce arrow types \FUN{\chr}{\dbl}, \FUN{\VEC\chr}{\dbl}, and \FUN{\NAVEC\chr}{\dbl} into the function type \FUN{\NAVEC\chr}{\dbl}.
+Comparing again to the strategy employed throughout the paper, we find 5,317 (20.82\%) functions to have a union of arrows, and we found 38,650 independent arrow types.
 We additionally find that 38,650 (26.14\%) of arrow types have at least one polymorphic argument.
@@
 The term ``top-level alternative'' signifies an arrow type in the union of arrows, e.g., \FUN{\chr}{\chr} and \FUN{\dbl}{\dbl} are top-level alternatives in the type \FUN{\chr}{\chr} $|$ \FUN{\dbl}{\dbl}.
-We see that the hybrid approach greatly increases the number of functions with no union of arrows at the function level, and nearly 5\% of function types obtained with the basic union of arrows strategy have over 9 top-level alternatives.
+We see that the hybrid approach greatly increases the number of functions with no union of arrows at the function level, where 92\% of functions only have one or two top-level alternatives.
+Also, nearly 5\% of function types obtained with the basic union of arrows strategy have over 9 top-level alternatives.
 
@@
 % The purpose of the function is to convert a continuous variable into a categorical variable, using several unsupervised learning methods.
-To connect this discussion with actual R types, we chose the \code{correlation} function from the \code{agricolae} package, and show what the type of the function would be when using each of these strategies.
-\code{agricolae::correlation} obtains the coefficients of correlation and p-value between all variables of some input data table, and uses a method of the user's choosing.
-It returns the correlation matrix and the probability together in a list.
+To connect this discussion with actual R types, we show what the types of the \code{correlation} function from the \code{agricolae} would be when using each of these strategies.
+\code{agricolae::correlation} obtains the coefficients of correlation and p-value between all variables of some input data table, using a method of the user's choosing.
+It returns the correlation matrix and the p-value (a double) together in a list.
 We refer the reader to Figure~\ref{tbl:real-function-different-merge-strats}, where we see that the hybrid approach produces a much smaller type, with only two arrow types in the top-level union.
@@
 This function takes a data frame and renames selected columns.
-We see that with both the union-of-arrows and hybrid approach, the types are the same (with only three unique function signatures observed), and can glean from these types that \code{dplyr::rename} produces a data frame of the {\it same class} as the input.
+We see that the types are the same with both the union of arrows and hybrid approach (with only three unique function signatures observed), and can glean from these types that \code{dplyr::rename} produces a data frame of the {\it same class} as the input.
 This information is lost in the arrow-of-unions merge strategy.
-That said, once the type system is extend with proper user-defined type hierarchies, the arrow of unions type will be much smaller and more informative (as data frames, grouped tibbles, and tibbles are likely all related types).
+That said, once the type system is extend with proper user-defined type hierarchies, the arrow of unions type will be much smaller and more informative (as data frames, grouped tibbles, and tibbles all related types).
 
@@
 
-Finally, we want to briefly discuss types for base package functions (in R, functions like \code{+}, \code{-}, vector access, etc. are part of the {\it base package}).
-One can find the implementation of these functions in the R source code\footnote{A read-only mirror of the R source can be found at \url{https://github.com/wch/r-source}.}, where they are implemented in C.
+Finally, we want to briefly discuss types for base package functions (in R, functions like \code{+}, \code{-}, and vector access are part of the {\it base package}).
 As we have alluded to, these functions coerce mismatched arguments in an ad hoc manner, with no real defined semantics.
-Further, functions like \code{+} often act on values based on their type (according to \code{typeof}), and ignore the class unless some package extended \code{+} to dispatch on some new class.
+Further, functions like \code{+} often act on values based on their type (according to \code{typeof}), and ignore the class, unless a package extended \code{+} to dispatch on the class.
 This leads to incredibly large infered signatures for these functions.

### 2020-10-16 paper/main.tex
formatting
minor

@@
 
+\pagebreak
+
 \begin{lstlisting}[basicstyle=\footnotesize\ttfamily] 
@@
  a \CLASS{{\tt ixforeach}}, and it gained the class {\tt iter} on
- return).  As we mentioned, accounting for object-orientation in the
+ return).  As mentioned, accounting for object-orientation in the
  type system is beyond the scope of this work, and such an inclusion

### 2020-10-31 paper/main.tex
edits

@@
 \setcopyright{rightsretained} 
-\acmPrice{}
-\acmDOI{10.1145/3360579}
-\acmYear{2020}
-\copyrightyear{2020}
 \acmJournal{PACMPL}
-\acmVolume{3}
-\acmNumber{OOPSLA}
-\acmArticle{153}
-\acmMonth{10}
+\acmYear{2020} \acmVolume{4} \acmNumber{OOPSLA} \acmArticle{181} \acmMonth{11} \acmPrice{}\acmDOI{10.1145/3428249}
 
@@
 \newcommand{\LIST}[1]{\texttt{\textbf{list}}\ensuremath{\langle #1\rangle}\xspace}
-\newcommand{\CLASS}[1]{\texttt{\textbf{class}}\ensuremath{\langle #1\rangle}\xspace}
+\newcommand{\CLASS}[1]{\texttt{\textbf{class}}\ensuremath{\langle #1\rangle}}
 \newcommand{\VEC}[1]{#1\k{[]}}
@@
 
-In order to keep the type signatures compact, we will compact union of
-arrows into a single top-level arrow with unions at each argument
-position. Thus, the shape of function signatures will be as follows:
+Later in this paper, we will introduce a tool to synthesize function type signatures from running R code, wherein each invocation of a function will generate a trace representing the types of function arguments and returns.
+These traces will be combined into an overall arrow type for the function, and in order to keep these signatures compact we compact the traces into a single top-level arrow with unions at each argument
+position. Thus, the shape of function signatures will be:
 
@@
 the size of types in check.  Figure~\ref{fig:siplification-rules}
-overviews the main simplification rules that we have adopted here.
+overviews the main simplification rules we adopted.
 
@@
 to minimize their size by removing redundant types, types that are
-subsumed by subtyping, immutable lists, and remove \nul type to
-replace them with nullables.  Higher-order functions are
+subsumed by subtyping, immutable lists, and replace \nul types with nullables.  Higher-order functions are
 conservatively treated as \any $\rightarrow$ \any.
 
-It is noteworthy that by performing this compaction we are loosing
-precision, that is to say, the synthesized signatures will suggest
-that the function accepts combination of types that were not observed.
+It is noteworthy that by performing this compaction we are losing
+precision, and the synthesized signatures will suggest combinations of argument types that were not observed.
 
@@
 of each value. Internal types are translated to names in the proposed
-type system. The next bit of information is the \emph{class}, an
+type language. The next bit of information is the \emph{class}, an
 optional list of names that may be absent, and, in some cases, is
@@
 specified for those missing arguments.  We record a \k{missing} type
-for such argument.  There are two obvious ways to deal with missing
+for such arguments.  There are two obvious ways to deal with missing
 arguments: type them as \any or type them as some unit type.  We
@@
 
+Our aim is to gather data with which we can validate our type language design.
+To that end, we propose an experiment wherein we use \typetracer to synthesize types for some core corpus of packages, and validate those types by using \contractr and installing contracts on core corpus functions and running the client code of this corpus.
+
 For this paper we have selected \CorpusLoadable packages consisting of
@@
 this version.  Some packages also require extra native dependencies
-which were not present on our servers. We defined two criteria for
+which were not present on our servers.
+
+We defined two criteria for
 including a package into the corpus:
 \begin{inparaenum}[(1)]
-  \item the package must have a runnable code that covers a
+  \item the package must have runnable code that covers a
     significant part of the package source code from which type
@@
   \item the package must have some reverse dependencies that will
-    allows us to evaluate the inferred types using the runnable code
+    allow us to evaluate the inferred types, using the runnable code
     from these dependencies.
@@
 To represent end-user code in the corpus, we turned to Kaggle, an
-online platform for data-science and machine-learning. The website
-allows people to share data-science competitions and data-analysis
+online platform for data science and machine learning. The website
+allows people to share data science competitions and data analysis
 problems together with data for which users try to find the best
@@
 various runtime exceptions, common problems were missing libraries (no
-longer available for R 3.5), parse errors and misspelled package
+longer available for R 3.5), parse errors, and misspelled package
 names. The final set contains \KaggleSuccess kernels with
@@
 \paragraph{Type usage}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-During execution 3,147 different types were observed.  Classes are the
+During execution, 3,147 different types were observed.  Classes are the
 most common types, accounting for roughly 31\% of types of arguments.
@@
 
-This section attempts to evaluate how well the proposed type system is able
+This section attempts to evaluate how well the proposed type language is able
 to describe the actual type signatures of functions. For this we focus on
@@
 The first part of our evaluation attempts to shed light on how good a fit
-our proposed type system is with respect to common programming patterns
+our proposed type language is with respect to common programming patterns
 occurring in widely used R libraries.
@@
 means our type language can accurately capture an argument's type or a
-function signature. We get to that number in two steps.
+function's signature. We get to that number in two steps.
 Fig.~\ref{fig:unionfreq} shows the number of inferred argument types and
-their size (in terms of members of the union). The figure shows the most
+their size (in terms of members of the union). The figure shows that most
 functions do not require a union at all (\PercUnitypedPositions of arguments
-do not have a union), and only \PercManytypedPositions positions have unions
+do not have a union), and only \PercManytypedPositions of positions have unions
 with more than three members.
@@
 common type categories. The shaded rows correspond to polymorphic
-types.  When an argument's type is \nul, we say that the argument is
+types.  When an argument's type is inferred to be \nul, we say that the argument is
 polymorphic due to a limitation in our analysis: In R, it is common
@@
 cases this value happens to be {\tt NULL}.  Our type analysis will
-report a {\tt null} type for these arguments if they are {\it never
+report a \nul type for these arguments if they are {\it never
   passed a value} during testing.  We interpret these instances of
@@
 lists with named elements.  The pollution is illustrated by the lines
-after the function definition.  They load an example data set and test
-the function \code{cv.model}.  The \code{data(sweetpotato)} expression
-loads a sample data set. The fields of \code{sweetpotato} will be
+after the function definition, where the \code{data(sweetpotato)} expression
+loads a sample data set to test the function.  The fields of \code{sweetpotato} will be
 recorded when \code{cv.model} is called.
@@
 R5.  The \code{class} attribute is used by these systems to dispatch
-methods. S3 does single dispatch, S4 does multiple dispatch and R5
+methods. S3 does single dispatch, S4 does multiple dispatch, and R5
 supports imperative objects.  The mechanics of S4 dispatch are more
@@
 passed to an argument expecting other XML-like classes are related to
-our simplified take on class types.  Our type system does not encode
+our simplified take on class types.  Our type language does not encode
 user-defined subtyping and coercion, which could help address these
@@
 We see that even though the double vector and matrix issue was wildly
-prevalent in the raw, dynamic contract evaluation numbers, the number
+prevalent in the dynamic contract evaluation numbers, the number
 of actual function argument types that were violated is very small.
@@
 often arguments expecting \CLASS{\texttt{timeSeries}} data are passed
-\CLASS{\texttt{matrix}} values.  This is a quirk of the
+\CLASS{\texttt{matrix}} values.  These are all from the
 \code{timeSeries} package, whose functions often accept matrices and
@@
 vignettes package code is only 58\%, which is one possible explanation
-of why these contract failures are occurring: the types that
+for these contract failures: the types that
 \typetracer generates are only as good as the test code its run on.
@@
 Had we failed to capture some key usage pattern of R with our type
-annotation framework, we would likely see it here, and we can see this
-in action if we consider Table~\ref{tbl:fail-breakdown-3}, which was
+annotation framework, we would likely see it here. For example, consider Table~\ref{tbl:fail-breakdown-3}, which was
 obtained identically to Table~\ref{tbl:fail-breakdown-2} except
@@
  and generate all permutations of elements in that vector.  If given a
- scalar ``integer'' $n$, it will generate all permutations for the
+ scalar integer $n$, it will generate all permutations for the
  list $[1, 2, ..., n]$.  Interestingly, the function was only tested
@@
 
- By-and-large, we saw no meaningful difference in the two data sets,
+ By-and-large, this evaluation did not reveal any new insights,
  with the contract assertion failure patterns being repeated from the
@@
  passed to an argument expecting a long union of classes, each of the
- form \CLASS{{\tt X, iter}}.  This is likely an instance where the
+ form \CLASS{{\tt \_, iter}}.  This is likely an instance where the
  user defined their own class, {\tt ixforeach}, and wanted to use the
@@
 \code{stopifnot} function from the R base package. It takes a number
-of R expressions which all should evaluate to true otherwise a runtime
-exception is thrown with a message quoting the corresponding failed
+of R expressions which should all evaluate to \code{TRUE}, otherwise a runtime
+exception is thrown with a message quoting the failed
 expression. For example, the following code checks whether a given
@@
 
-Besides \code{stopifnot}, there are 4 packages in
+Beyond \code{stopifnot}, there are 4 packages in
 CRAN\footnote{Packages are available on the CRAN website:
 \url{https://cran.r-project.org/web/packages/}} that focus on runtime
-assertions: \emph{assertive}, \emph{ensurer}, \emph{assertr} and
+assertions: \emph{assertive}, \emph{ensurer}, \emph{assertr}, and
 \emph{assertthat}. \emph{assertive} and \emph{ensurer} have not been
@@
 by only \AssertrRevdeps other packages and currently focuses on
-checking properties of data frames. Only the \emph{assertthat} package
-is maintained and used (with \AssertthatRevdeps reverse
-dependencies). The advantage of \emph{assertthat} over the R's default
+checking properties of data frames. Only \emph{assertthat} is maintained and used (with \AssertthatRevdeps reverse
+dependencies). The advantage of \emph{assertthat} over the R default
 is that it provides much better error messages.
 
-One way to asses the usefulness of our type checking system is to find
-out how many of the existing type checking constrains could be
-replaced by \contractr.  To measure this, we have extracted all calls
-to \code{stopifnot} and \code{assertthat} assertions and checked which
-among them could be either completely replaced by \contractr or at
+One way to assess the usefulness of our type checking framework is to see how many existing user-defined type checks can be replaced by \contractr.
+To measure this, we extracted all calls
+to \code{stopifnot} and \code{assertthat} assertions, and checked which
+among them could either be completely replaced by \contractr, or at
 least partially simplified by removing a portion of an assertion
-expression. This is useful, because a common pattern is that the first
-part of the parameter assertion checks its type while the rest checks
-its value. In the example above, the whole expression could be
+expression. Partial simplification is useful, as a common pattern in this ad hoc type checking is that the first
+part of the assertion checks the value's type, while the rest checks more detailed properties. In the example above, the whole expression could be
 replaced by a \chr type check.
@@
 Out of the \CorpusLoadable packages, \CorpusAssertsInPackages use
-runtime assertions. Together there are \CorpusAsserts asserts in
+runtime assertions. Altogether, there are \CorpusAsserts assertions in
 \CorpusAssertsInFunctions functions. Among these, \contractr can
@@
 across \CorpusTypedAssertsPackages packages and
-\CorpusTypedAssertsFunctions functions.  Furthermore, additional
+\CorpusTypedAssertsFunctions functions.  Furthermore, an additional
 \CorpusPartiallyTypedAsserts (\CorpusPartiallyTypedAssertsRatio)
-asserts in packages \CorpusPartiallyTypedAssertsPackages packages and
+asserts in \CorpusPartiallyTypedAssertsPackages packages and
 \CorpusPartiallyTypedAssertsFunctions functions could have been
@@
 often in the R code. In the whole of CRAN, there are only
-\CranAssertsRnd asserts in \CranAssertsInFunctionsRnd functions define
+\CranAssertsRnd asserts in \CranAssertsInFunctionsRnd functions defined
 in \CranAssertsInPackagesRnd packages.  One may speculate that this is
@@
 We call this an {\it arrow of unions}.
-This strategy was not developed in a vacuum, and we will now discuss some of the decisions that went into it.
+We will elaborate on some of the design decisions that went into developing this method.
 
@@
 
-Another option is to employ a hybrid approach, wherein we perform the union of arrow types {\it after} grouping them together by return type (to further simplify we also combine some primitive return types together, such as \dbl and \VEC\dbl).
+Another option is to employ a hybrid approach, wherein we perform the union of arrow types {\it after} grouping them together by return type (to further simplify, we also combine some primitive return types together, such as \dbl and \VEC\dbl).
 While this has the advantage of reducing most of the blow-up of the previous approach, it suffers particularly when functions return classes, as our type system does not allow us to effectively consolidate class types.
@@
 
-These findings are summarized in Figure~\ref{fig:mergecomp}, which shows a breakdown of the number of {\it top-level alternatives} in the function types obtained with both of these strategies.
+These findings are summarized in Figure~\ref{fig:mergecomp}, which shows a breakdown of the number of {\it top-level alternatives} in the function types obtained with both of these strategies:
 The term ``top-level alternative'' signifies an arrow type in the union of arrows, e.g., \FUN{\chr}{\chr} and \FUN{\dbl}{\dbl} are top-level alternatives in the type \FUN{\chr}{\chr} $|$ \FUN{\dbl}{\dbl}.
@@
 This information is lost in the arrow-of-unions merge strategy.
-That said, once the type system is extend with proper user-defined type hierarchies, the arrow of unions type will be much smaller and more informative (as data frames, grouped tibbles, and tibbles all related types).
+That said, once the type language is extend with proper user-defined type hierarchies, the arrow of unions type will be much smaller and more informative (as data frames, grouped tibbles, and tibbles are all related types).
 
@@
 Finally, we want to briefly discuss types for base package functions (in R, functions like \code{+}, \code{-}, and vector access are part of the {\it base package}).
-As we have alluded to, these functions coerce mismatched arguments in an ad hoc manner, with no real defined semantics.
+These functions coerce mismatched arguments in an ad hoc manner, with no real defined semantics.
 Further, functions like \code{+} often act on values based on their type (according to \code{typeof}), and ignore the class, unless a package extended \code{+} to dispatch on the class.
-This leads to incredibly large infered signatures for these functions.
+This leads to incredibly large inferred signatures for these functions.
 Even after counting out traces related to S3 and S4 dispatch, the type of \code{+} using the hybrid approach has 29 top-level alternatives, 22 of which have a class-typed argument.
-We don't believe this type to be particularly useful to a programmer, but a compiler might find it quite useful, as given a set of types for inputs, it can select the appropriate arrow type from the top-level alternatives.
+We don't believe this type to be particularly useful to a programmer, but a compiler might find it quite useful, as it can determine the type of the return value of a function given the types of its arguments.
 

### 2020-11-01 paper/main.tex
first part edits

--- 2020-10-16 paper/main.tex
@@
 
-R's main data type is the primitive vector.  Vectors care explicitly
+R's main data type is the primitive vector.  Vectors are explicitly
 constructed by the constructor \code{c(...)}, as in \code{c(1L, 2L,
   3L)} which creates a vector of three integers.  R has a builtin
-notion of type that can be queried by the \k{typeof}
-function. Figure~\ref{rtypes} lists all of the builtin types that are
+notion of type that can be queried with the \k{typeof}
+function. Figure~\ref{rtypes} lists all of the builtin types
 provided by the language; these are the possible return values of
 \k{typeof}. There is no intrinsic notion of subtyping in R, but in
-many contexts a \k{logical} will be coerced to \k{integer}, and an
-\k{integer} will be coerced to \k{double}.  Some odd conversions can
-occur in corner cases, such as \k{1<"2"} holds and \k{c(1,2)[1.6]}
+many contexts a \k{logical} will be coerced to \k{integer}, an
+\k{integer} will be coerced to \k{double}, and a \k{double} will be coerced to \k{complex}.  
+Some odd conversions occur in corner cases, such as \k{1<"2"} holds and \k{c(1,2)[1.6]}
 returns the first element of the vector, as the double is converted to
@@
 are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
-  typeof(c(5,5))} \code{ == "double"}.  With the exception of lists,
-all vectorized data types are monomorphic.  A \k{list} can hold values
-of any other type including \k{list}. For all monomorphic data types,
-attempting to store a value of a different type will cause a
-conversion. Either the value is converted to the type of the vector,
-or the vector is converted to the type of the value.
+  typeof(c(5,5))} \code{ == "double"}. All vectorized data types are monomorphic, except \k{list} which can hold values
+of any type, including \k{list}. For all monomorphic data types,
+attempting to store a differently-typed value will cause a
+conversion: Either the value is converted to the type of the vector, or vice versa.
 
@@
 
-\noindent
 All vectorized data types have a distinguished missing value denoted
@@
 need for a missing value to inhabit each primitive type. In a sense,
-\code{NA} represents a missing data point, and \code{NULL} is a
+\code{NA} represents a missing data point, whereas \code{NULL} represents a
 missing data set.
@@
 built in type.  Attributes are used to encode various type
-structures. They are queried with \k{attributes} and \k{class}.  Using
+structures, and they are queried with \k{attributes}.  Using
 attributes, programmers can extend the set of types by tagging
@@
 
-Another attribute that can be set is the \k{class}, this is one way to
-use R for object-oriented programming.  This attribute can be bound to
-a list of class names. For instance, \code{class(x)<-"human"} will set
-the class of \k{x} to be \k{human}.  There are three
+Another attribute that can be set is the \k{class}, which is one way to
+use R for object-oriented programming. The class attribute can be queried with \k{class}, and it can be bound to
+a list of class names: For instance, \code{class(x)<-c("human", "friend")} will set
+the class of \k{x} to be \k{human} and \k{friend}.  There are three
 object-orientation frameworks in R: S3, S4, and R5.  The S3 object
@@
 zero, \k y defaults to \k{NULL}). The body of the function will add \k
-x and \k y tp either 0 or the result of concatenating the dots into a
+x and \k y to either 0 or the result of concatenating the dots into a
 primitive vector. The function \k{missing} tests if a parameter was
@@
 
-Given the pecularities of the language there are a number of design
+Given the peculiarities of the language there are a number of design
 choices that need to be reviewed and evaluated. It is not
@@
 cases where functions expect scalar values, for example a conditional
-takes a single logical (and will complain if more values are passed).
+takes a single logical and will complain if more values are passed.
 We considered tracking dimensions of data structures, but decided
@@
 either vectors (e.g., \VEC\int) or scalars (e.g., \int).  A vector can
-happen to be of length 1, and thus a scalar is also a vector.  Vectors
+happen to be of length 1, which we handle with subtyping, shown later.  Vectors
 are monomorphic, a vector of doubles contains only doubles.
@@
 \paragraph{Missing} 
-Each of basic types has its specific \NA. Many built-in functions,
+Each primitive type has its specific \NA. Many built-in functions,
 especially those implemented in C or Fortran, do not support \NA
-values.  It is thus advantageous to distinguish between vector that
-can contain missing values and those that are guaranteed not to.  In
+values.  It is thus advantageous to distinguish between vectors that
+can and cannot contain missing values.  In
 our experience, functions that expect scalar values tend to not admit
 {\NA}s, thus scalar types are treated as being \NA-free. The type
-language does allows to write \VEC\int to specify that a vector of
-integers may have {\NA}s and \NAVEC\int to say that a vector must not
-have missing values. Of course, a value of type \VEC\int may happen to
-not have any \NA and thus be of type \NAVEC\int.  The type \raw does
-not allow {\NA}s, so \VEC\raw is \NA-free.
+language allows one to write \NAVEC\int to specify that a vector of
+integers {\it may} have {\NA}s and \VEC\int to say that a vector must not
+have missing values.  The type \raw does
+not allow {\NA}s, so \NAVEC{\raw} is \NA-free.
+R does not provide a built-in type testing mechanism, e.g., for the case of
+\NA-free data types, it is necessary to scan vectors to find out if
+they have missing values.
 
@@
 default value of uninitialized variables. R has different notion for
-that (which we do not cover here).  To capture common uses, of
+that (which we do not cover here).  To capture common uses of
 \k{NULL}, the type language has a nullable type, written \code{?}\T.
@@
 with \code{as.list}, and lists to vectors with \code{unlist}
-(coercions may ensue). The type language allows to specify that a
+(coercions may ensue). The type language allows one to specify that a
 value is a list containing element of some type \T, written \LIST\T. R
@@
 operate by adding a class attribute to values. The most widely used
-system, by far, is S3 which supports single dispatch and multiple
+system by far is S3, which supports single dispatch and multiple
 classes~\cite{ecoop12}. The challenge from a type system point of view
-is that a value such as integer 5, could be attributed with a class.
+is that a value (such as an integer \k{5L}) could be attributed with a class.
 Code that performs dispatch would use the class attribute while code
 that does not would view the value as an integer. The type language
-focuses on the attribute and will hide the underlying type of the
-value. While that seems to match common usage it does represent a loss
-of expressiveness. The type language also focuses on S3, we leave the
+focuses on the attribute, and will hide the underlying type of the
+value. While this seems to match common usage, it does represent a loss
+of expressiveness. The type language also focuses on S3, and we leave the
 other object systems for future work. S3 has no notion of inheritance,
@@
 $\T_1 \k{~|~} ... \k{~|~} \T_n$. The elements of unions are not
-disjoint, 1L is both an \int, a \VEC\int and a \NAVEC\int. R does not
-provide a built-in type testing mechanism, e.g. for the case of
-\NA-free data types, it is necessary to scan vectors to find out if
-they have missing values.
+disjoint, e.g., 1L is both an \int, a \VEC\int and a \NAVEC\int.
 
@@
 Functions signatures the form \FUN{A_1, \dots, A_n}\T where each
-$\A_i$ argument is either a type \T or \VARGS, a variable length
+$\A_i$ argument is either a type \T or dots (\VARGS), a variable length
 argument list. Moreover, a single function's signature can be the
@@
 
+Overall, the choice of type language follows the structure of values.  The
+presence of \NA-free data types and scalars are two choices that must
+be validated in practice.  Note that nullable types are really just a special case of
+unions (i.e., \code{?}\T is shorthand for $\T \k{~|~} \k{NULL}$).
+
 \subsection{Subtyping}
 
-The choice of type language follows the structure of values.  The
-presence of \NA-free data types and scalars are two choices that must
-be validated in practice.  Nullable types are just a special case of
-unions.
-
-While R does not support the notion of subtyping between values.  The
+R does not support the notion of subtyping between values, but we include a few simple rules in
+our framework to capture widespread coercion.  The
 conversions between primitive types give us a starting point (e.g., an
@@
 types introduced above induce some the rules of Figure~\ref{fig:sub}:
-a vector without \NA is a subtype of a vector, a value of type \T is a
+an \NA-free vector is a subtype of its \NA-ful equivalent, a value of type \T is a
 subtype of a nullable \T, a list is subtype of another list if their
@@
 \begin{figure}[h!tb]\begin{tabular}{rclll}
-  \NAVEC\S  & <:& \VEC\S \\
+  \VEC\S    & <:& \NAVEC\S \\
   \T        & <:& \k{?}\T \\

### 2020-11-03 paper/main.tex
publisher edits

@@
 \title{Designing Types for R, Empirically}
-\author{Alexi Turcotte}\affiliation{\institution{Northeastern University}}
-\author{Aviral Goel}\affiliation{\institution{Northeastern University}}
-\author{Filip Krikava}\affiliation{\institution{Czech Technical University in Prague}}
-\author{Jan Vitek}\affiliation{\institution{Northeastern / Czech Technical University}}
+\author{Alexi Turcotte}\affiliation{\institution{Northeastern University, USA}}
+\author{Aviral Goel}\affiliation{\institution{Northeastern University, USA}}
+\author{Filip Křikava}\affiliation{\institution{Czech Technical University, Czechia}}
+\author{Jan Vitek}\affiliation{\institution{Northeastern University, USA / Czech Technical University, Czechia}}
 
@@
 analyzers. In particular, R has no type annotations, and all
-operations are dynamically checked at run-time. The starting point for
-our work are the twin questions: \emph{what expressive power is needed
+operations are dynamically checked at runtime. The starting point for
+our work are the two questions: \emph{what expressive power is needed
 to accurately type R code?} and \emph{which type system is the R
@@
 \NUMPKGSEVAL clients of these packages, as well as on end-user code
-from the Kaggle competition.
+from the Kaggle data science competition website.
 \end{abstract} \maketitle
@@
 became widely used across many fields for exploratory data analysis.
-Modern R, as an object of study, is fascinating. It is a vectorized,
+Modern R is fascinating as an object of study. It is a vectorized,
 dynamically typed, lazy functional language with limited side-effects,
@@
 for function type signatures.  Thus, this paper will use the collected
-data to document the signature of user-defined functions.  For this,
-we design a simple type language, one that matches the R data types.
+data to document the signatures of user-defined functions.  For this,
+we design a simple type language that matches the R data types.
 We then extract call and return types from execution traces of a
@@
 checker that weaves types around their respective functions, and use
-clients of the target packages for validation.  To sum up, we make the
+clients of the target packages for validation, as well as end-user code from data science competitions.
+
+To sum up, we make the
 following contributions:
@@
 
-\subsection{Synthesizing signatures}
+\subsection{Synthesizing Signatures}
 
@@
 
-\subsection{Types from traces}
+\subsection{Types from Traces}
 
@@
 
-\paragraph{Missing arguments}
+\paragraph{Missing Arguments}
 Parameters which receive no values when the function is called are
@@
 
-\paragraph{Non-local returns}
+\paragraph{Non-local Returns}
 When a function exits with a longjump, there is no return value to speak
@@
 
-\paragraph{Implementation details.}
+\paragraph{Implementation Details.}
 We primarily rely on eight callbacks: \k{closure\_entry},
@@
 
-\subsection{Checking signatures with contracts}
+\subsection{Checking Signatures with Contracts}
 
@@
 
-\paragraph{User code}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\paragraph{User Code}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 To represent end-user code in the corpus, we turned to Kaggle, an
@@
 
-\paragraph{Type usage}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\paragraph{Type Usage}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 During execution, 3,147 different types were observed.  Classes are the
@@
 \begin{table}[!h]
+\caption{Top types of arguments in R}\label{ctb}
 % BEGIN Autogenerated
@@
 \end{tabular}
-\caption{Top types of arguments in R}\label{ctb}
 \end{table}
@@
 the features of our type language are represented here, and some
-signatures are telling of the function's behaviour.  For example,
+signatures are telling of the function's behavior.  For example,
 consider \code{decrypt_envelope}: the first three parameters of the
@@
   \centering
+  \caption{Select Type Signatures}\label{tbl:typesig}
   \resizebox{.98\linewidth}{!}{
@@
   \end{tabular}}
-\caption{Select Type Signatures}\label{tbl:typesig}
 \end{table}
@@
 \scriptsize
+\caption{Top contract failures}
 \begin{tabular}{llrrr}
@@
 % END Autogenerated
-\caption{Top contract failures}
+
 \label{tbl:fail-breakdown-0}
@@
 \scriptsize
+ \caption{Results from Table 4, broken down by occurrences of the expected type as a parameter type}
 \begin{tabular}{llrrr}
 \toprule
-              &                &      \multicolumn{2}{c}{\# Args Types}        &  \\
+              &                &      \multicolumn{2}{c}{\# Arg Types}        &  \\
  Passed & Arg Type & Failed & Total & \% Failure\\
@@
 % END Autogenerated
- \caption{Results from Table 4, broken down by occurrences of the expected type as a parameter type}
+
  \label{tbl:fail-breakdown-1}
@@
 all valid function inputs.  Separate from the issue of testing, we can
-capture this behaviour with user-defined subtyping or coercion, as the
+capture this behavior with user-defined subtyping or coercion, as the
 data types which were passed are readily convertible to \CLASS{{\tt
@@
 \scriptsize
+ \caption{Highest failure rate among popular argument types, for
+   argument signatures whose frequency is in the 90th percentile.}
 \begin{tabular}{llrrr}
@@
 % END Autogenerated
- \caption{Highest failure rate among popular argument types, for
-   argument signatures whose frequency is in the 90th percentile.}
+
  \label{tbl:fail-breakdown-2}
@@
 \scriptsize
+ \caption{Highest failure rate among popular argument types, for argument signatures whose frequency is in the 80th percentile.}
 \begin{tabular}{ll@{}rrr@{}}
 \toprule
-Passed & Arg Type & \#Args Fail & \#Args w Type & \% Failure\\
+              &                &      \multicolumn{2}{c}{\# Arg Types}        &  \\
+ Passed & Arg Type & Failed & Total & \% Failure\\
 \midrule
@@
 % END Autogenerated
- \caption{Highest failure rate among popular argument types, for argument signatures whose frequency is in the 80th percentile.}
  % *: Grouped Tibble is \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}} , and Tibble is \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}.}
@@
  either the call results in an error (perhaps explicitly handled by
- the programmer), or results in valid function behaviour that was
+ the programmer), or results in valid function behavior that was
  untested by the original package designer.  Here, the first case can
@@
  \scriptsize
+ \caption{Top contract failures in Kaggle kernels}
 \centering
@@
 \end{tabular}
-\caption{Top contract failures in Kaggle kernels}
 \label{tbl:fail-breakdown-0-kaggle}
@@
 \small
+\caption{Highest failure rate among popular argument types in Kaggle, for argument signatures whose frequency is in the 90th percentile.}
 \begin{tabular}{llrrr}
@@
 \end{tabular}
-\caption{Highest failure rate among popular argument types in Kaggle, for argument signatures whose frequency is in the 90th percentile.}
  \label{tbl:fail-breakdown-2-kaggle}
@@
 	\centering
+	\caption{Types of \code{agricolae::correlation} with different type merge strategies. }\label{tbl:real-function-different-merge-strats}
 	 \resizebox{.98\linewidth}{!}{
@@
 	\end{tabular}}
-	\caption{Types of \code{agricolae::correlation} with different type merge strategies. }\label{tbl:real-function-different-merge-strats}
 \end{table}
@@
 	\centering
+	\caption{Types of \code{dplyr::rename} with different type merge strategies. }\label{tbl:real-function-different-merge-strats-2}
 	\resizebox{.98\linewidth}{!}{
@@
 	\end{tabular}}
-	\caption{Types of \code{X::Y} with different type merge strategies. }\label{tbl:real-function-different-merge-strats-2}
 \end{table}
